---
title: "Subsampling"
output: html_notebook
---

# Load library
```{r load library}
library(tidyr)
library(dplyr)
library(pROC)
library(caret)
library(clue)
library(ordinal)
library(corrplot)
```

# Load codebase and data
```{r load codebase and data}
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source(paste0(codebase_path,"/codebase_subtype.R"))

mcia_results = readRDS("/scratch/data-integration/molecular_subtyping/DimensionReduction/MCIA_phase123_20230726.rds")
organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")
base_obj = readRDS('/scratch/data-integration/cleanedup/mod1_interpretation_base.rds')
extended_data = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/extended_data.rds")
subtype_labels = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
clin_data_used = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/clin_data_aligned.rds")
```

# Data preparation
```{r data preparation}
# clinical data
clin_data_all = organized_data$clinical_data[, setdiff(colnames(organized_data$clinical_data), c("sample_id","sample_type", "phase1_2","samples_collected","phase","enrollment_site"))]
clin_data_all = unique(clin_data_all)
clin_data_aligned = data_frame(event_id = rownames(organized_data$datasets$plasma_proteomics_targeted))
clin_data_aligned = clin_data_aligned %>% left_join(clin_data_all, by = c("event_id"))
ll = which(clin_data_aligned$event_type %in% paste0("Visit ",1:6))
clin_data_aligned = clin_data_aligned[ll,]

# factor scores
mcia_factors = mcia_results$mcia_factors
rownames(mcia_factors) = gsub('-([0-9])$', '-0\\1', rownames(organized_data$datasets_imputed[[1]]))
mcia_factors = mcia_factors[ll,]
n_factor = 40
factor_scores = mcia_factors[,1:n_factor]
factor_scores = apply(factor_scores, 2,function(z) qqnorm(z, plot.it = F)$x)
# update factors that explain more consistent variation across assays
aa = (apply(sqrt(mcia_results$mcia_factor_evaluation$ave_variance_explained[1:n_factor,]),1,sum))^2
factor_scores = t(apply(factor_scores,1,function(z) z*sqrt(aa)))
rownames(factor_scores) <- rownames(mcia_factors)
factor_scores <- as.data.frame(factor_scores)
factor_scores$participant_id <- substr(rownames(factor_scores), start = 1, stop = 8)

# ema of factor scores
mov_ave <- function(a){
  mu <- 1 - 2/(length(a)+2)
  res <- a[1]
  for (k in 1:length(a))  res <- mu*a[k] + (1-mu)*res
  return(res)
}
factor_scores_ema <- aggregate(. ~ participant_id, data = factor_scores, FUN = mov_ave)
rownames(factor_scores_ema) <- factor_scores_ema$participant_id
factor_scores_ema$participant_id <- NULL

# V1 of factor scores
factor_scores_v1 <- factor_scores[!duplicated(factor_scores$participant_id),]
rownames(factor_scores_v1) <- factor_scores_v1$participant_id
factor_scores_v1$participant_id <- NULL
```

# Soft assignment for full sample
```{r soft assignment for full sample}
subtype_labels$merged_labels <- subtype_labels$label
label_map <- c(A = 1, B = 2, C = 3, D = 4, E = 5, F = 6)
subtype_labels$merged_labels <- label_map[subtype_labels$merged_labels]

# subtype center
factor_scores$visit <- clin_data_aligned$event_type
factor_scores$subtype <- subtype_labels[factor_scores$participant_id,]$merged_labels
factor_scores <- na.omit(factor_scores)
subtype_centers <- list()
for (i in 1:6){
  tmp <- factor_scores[factor_scores$subtype == i,]
  visits <- unique(tmp$visit)
  subtype_centers[[i]] <- do.call(rbind, lapply(visits, function(v) {
    col_means <- colMeans(tmp[tmp$visit == v, 1:n_factor])
    return(col_means)
  }))
  rownames(subtype_centers[[i]]) <- visits
}

dist_to_subtype <- data.frame(participant_id = unique(factor_scores$participant_id))
for (j in 1:6) {
  copy <- factor_scores
  copy$dist <- rowSums((copy[, 1:n_factor] - subtype_centers[[j]][copy$visit, ])^2)
  copy <- na.omit(copy)
  avg_dist <- aggregate(dist ~ participant_id, data = copy, FUN = mean)
  colnames(avg_dist) <- c("participant_id", paste0("S",j))
  dist_to_subtype <- left_join(dist_to_subtype, avg_dist, by = "participant_id")
}
rownames(dist_to_subtype) <- dist_to_subtype$participant_id
dist_to_subtype$participant_id <- NULL
softmax <- function(x) {
  exp_x <- exp(x - max(x))
  return(exp_x / sum(exp_x))
}
probs <- as.data.frame(t(apply(-dist_to_subtype, FUN = softmax, MARGIN = 1)))
colnames(probs) <- c("p1","p2","p3","p4","p5","p6")
log_probs <- log(probs)

# check soft assignment and hard assignment
print(sum(apply(probs,1,which.max)==subtype_labels$merged_labels)/dim(probs)[1])
```

# Generate subsample
```{r generate subsample, warning=FALSE}
idx_exclD = which(subtype_labels$label != "D")
X = extended_data$X
n_factor = 40; visit_max = 6
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]

set.seed(123)
nclust_tree_sub <- 7
n_subsample <- 100
percent_subsample <- 0.8
subsample_clusters <- list()
for (i in 1:n_subsample) {
  print(i)
  idx_sample <- sample(1:length(idx_exclD), replace = FALSE,
                       size = floor(length(idx_exclD)*percent_subsample))
  # exclude SubD from subsampling 
  sub_X <- X[idx_exclD, ][idx_sample, ]
  dist_mat_sub <- calculate_distance_matrix(sub_X, n_factor, visit_max)
  hcl_sub <- hclust(d = dist_mat_sub, method = "ward.D2")
  label_tmp <- cutree(hcl_sub, k = nclust_tree_sub)
  label_tmp <- label_tmp[match(clin_data_patient$participant_id, names(label_tmp))]
  idx_nona <- which(!is.na(label_tmp))
  label_tmp <- data.frame(full = label_tmp[idx_nona],
                          TG =  factor(clin_data_patient$trajectory_group[idx_nona]))
  # identify 2 subtypes with the most severe TG
  tg_test <- test_ordinal_demo(
    subtype = label_tmp$full,
    response = label_tmp$TG,
    demo_data = demo_data_adjust[idx_nona,])
  selected_subtype <- names(sort(tg_test$est_ordinal, decreasing = TRUE))
  selected_subtype <- as.integer(sub("Subtype", "", selected_subtype))
  # separate the 2 most severe subtypes
  cut_full = cutree(hcl_sub, k = nclust_tree_sub)
  table_full = table(cut_full)
  c1 = names(which(cut_full == selected_subtype[1]))
  c2 = names(which(cut_full == selected_subtype[2]))
  merged = FALSE
  c3 = NULL
  for (j in 1:(nclust_tree_sub-1)) {
    k = nclust_tree_sub-j-1
    # print(paste0("K=", k))
    cut_sub = cutree(hcl_sub, k = k)
    cut_sub[names(cut_sub) %in% c1] = k+1
    cut_sub[names(cut_sub) %in% c2] = k+2
    if (length(unique(cut_sub)) != nclust_tree_sub) {
      merged = TRUE
    }
    if (merged && is.null(c3)) {
      merged_id = cut_sub[which(cut_sub != cut_full)[1]]
      c3 = names(which(cut_sub == merged_id))
    }
    cut_sub[names(cut_sub) %in% c3] = k+3
    if (length(unique(cut_sub)) == 5) {
      merged_label = cut_sub
      break
    }
  }
  merged_label = factor(merged_label)
  levels(merged_label) = 1:(nclust_tree_sub-2)
  merged_label = merged_label[order(names(merged_label))]
  subsample_clusters[[i]] = merged_label
  names(subsample_clusters[[i]]) = names(merged_label)
}
label_subD = array(6, sum(subtype_labels$label == "D"))
names(label_subD) = rownames(subtype_labels)[subtype_labels$label == "D"]
subsample_clusters_full = subsample_clusters
subsample_clusters_full = lapply(subsample_clusters_full, 
                                 function(x) c(x, label_subD))
subsample_clusters_full = lapply(subsample_clusters_full, 
                                 function(cluster) cluster[order(names(cluster))])
```

# Soft assignment for subsample
```{r soft assignment for subsample}
soft_prob <- function(subsample_subtype) {
  sub_label = subtype_labels
  df_subsample_subtype = data.frame(participant_id = names(subsample_subtype),
                                    merged_labels = subsample_subtype)
  sub_label$participant_id = rownames(sub_label)
  sub_label = sub_label %>% dplyr::select(-merged_labels)
  sub_label = left_join(sub_label, df_subsample_subtype, by = "participant_id")
  rownames(sub_label) = sub_label$participant_id
  
  sub_factor_scores = factor_scores
  sub_factor_scores$subtype = sub_label[sub_factor_scores$participant_id,]$merged_labels
  sub_factor_scores = na.omit(sub_factor_scores)
  subtype_centers = list()
  for (i in 1:6){
    tmp <- sub_factor_scores[sub_factor_scores$subtype == i,]
    visits <- unique(tmp$visit)
    subtype_centers[[i]] <- do.call(rbind, lapply(visits, function(v) {
      col_means <- colMeans(tmp[tmp$visit == v, 1:n_factor])
      return(col_means)
    }))
    rownames(subtype_centers[[i]]) <- visits
  }
  
  dist_to_subtype <- data.frame(participant_id = unique(factor_scores$participant_id))
  for (j in 1:6) {
    if (nrow(subtype_centers[[j]]) != 6) {
      missing_visit = setdiff(paste0("Visit ", 1:6), rownames(subtype_centers[[j]]))
      missing_mat = matrix(NA, nrow = length(missing_visit), 
                           ncol = ncol(subtype_centers[[j]]))
      rownames(missing_mat) <- missing_visit
      subtype_centers[[j]] = rbind(subtype_centers[[j]], missing_mat)
    }
    copy <- factor_scores
    copy$dist <- rowSums((copy[, 1:n_factor] - subtype_centers[[j]][copy$visit, ])^2)
    copy <- na.omit(copy)
    avg_dist <- aggregate(dist ~ participant_id, data = copy, FUN = mean)
    colnames(avg_dist) <- c("participant_id", paste0("S",j))
    dist_to_subtype <- left_join(dist_to_subtype, avg_dist, by = "participant_id")
  }
  rownames(dist_to_subtype) <- dist_to_subtype$participant_id
  dist_to_subtype$participant_id <- NULL
  sub_probs <- as.data.frame(t(apply(-dist_to_subtype, FUN = softmax, MARGIN = 1)))
  colnames(sub_probs) <- paste0("p", 1:6)
  sub_log_probs <- log(sub_probs)
  
  cor_mat = matrix(NA, nrow = 6, ncol = 6)
  for (s in 1:6) {
    for (t in 1:6) {
      cor_mat[s, t] = cor(probs[, s], sub_probs[, t])
    }
  }
  cost_mat = 1 - cor_mat
  assignment = clue::solve_LSAP(cost_mat)
  cor_arr = cor_mat[cbind(seq_along(assignment), assignment)]
  cor_mat = cor_mat[, assignment]
  return(list(cor_arr = cor_arr, cor_mat = cor_mat, 
              sub_log_probs = sub_log_probs, assignment = assignment))
}

cor_list = list()
cor_mat_list = list()
sub_log_probs_list = list()
assignment_list = list()
sum_cor_diag = array(NA, n_subsample)
for (k in 1:n_subsample) {
  print(k)
  soft_prob_list = soft_prob(subsample_clusters_full[[k]])
  cor_list[[k]] = soft_prob_list$cor_arr
  cor_mat_list[[k]] = soft_prob_list$cor_mat
  sum_cor_diag[k] = sum(soft_prob_list$cor_arr)
  sub_log_probs_list[[k]] = soft_prob_list$sub_log_probs
  assignment_list[[k]] = soft_prob_list$assignment
}

sum_cor <- Reduce("+", cor_list)
avg_cor <- sum_cor / length(cor_list)
avg_cor

sum_cor_mat <- Reduce("+", cor_mat_list)
avg_cor_mat <- sum_cor_mat / length(cor_mat_list)
avg_cor_mat

org_cor_mat <- matrix(NA, nrow = 6, ncol = 6)
for (s in 1:6) {
  for (t in 1:6) {
    org_cor_mat[s, t] = cor(probs[, s], probs[, t])
  }
}
org_cor_mat

quantile_values <- quantile(sum_cor_diag, probs = c(0.10, 0.50, 0.90))
quantile_idx <- sapply(quantile_values, function(q) {
    which.min(abs(sum_cor_diag - q))
})
names(quantile_idx) <- c("10%", "50%", "90%")

ci_cor = c()
for (i in 1:6) {
  cor_tmp = unlist(lapply(cor_list, function(x) x[i]))
  ci_tmp = paste0("[", round(quantile(cor_tmp, probs = 0.25), 4), ", ", 
                  round(quantile(cor_tmp, probs = 0.75), 4), "]")
  ci_cor = c(ci_cor, ci_tmp)
}
ci_cor
```

# Correlation matrix plot
```{r correlation matrix plot, fig.height=6, fig.width=6}
rownames(avg_cor_mat) = LETTERS[1:6]
colnames(avg_cor_mat) = LETTERS[1:6]
corrplot(avg_cor_mat, method = "circle", order = "original",
         tl.col = "black", addCoef.col = "black",
         tl.cex = 1.25, cl.cex = 1, number.cex = 1, number.font = 1,
         col = colorRampPalette(c("#284AA0", "white", "#BC4A4A"))(200))

rownames(org_cor_mat) = LETTERS[1:6]
colnames(org_cor_mat) = LETTERS[1:6]
corrplot(org_cor_mat, method = "circle", order = "original",
         tl.col = "black", addCoef.col = "black",
         tl.cex = 1.25, cl.cex = 1, number.cex = 1, number.font = 1,
         col = colorRampPalette(c("#284AA0", "white", "#BC4A4A"))(200))
```

# Hard assignment evaluation
```{r hard assignment evaluation}
jaccard_index_list = list()
for (i in 1:n_subsample) {
  df_sub = data.frame(participant_id = names(subsample_clusters_full[[i]]),
                      label = subsample_clusters_full[[i]])
  mapping = setNames(LETTERS[seq_along(assignment_list[[i]])],
                     as.character(assignment_list[[i]]))
  df_sub$label = mapping[as.character(df_sub$label)]
  subsample_label = data.frame(participant_id = rownames(subtype_labels))
  subsample_label = left_join(subsample_label, df_sub, by = "participant_id")
  subsample_label$full = subtype_labels$label
  subsample_label = na.omit(subsample_label)
  jaccard_index = array(NA, 6)
  for (k in 1:6) {
    current_sub = LETTERS[k]
    set1 = subsample_label$participant_id[subsample_label$label == current_sub]
    set2 = subsample_label$participant_id[subsample_label$full == current_sub]
    intersection = length(intersect(set1, set2))
    union = length(union(set1, set2))
    jaccard_index[k] = intersection / union
  }
  jaccard_index_list[[i]] = jaccard_index
}
Reduce("+", jaccard_index_list) / length(jaccard_index_list)
```

# Factor trajectory
```{r factor trajectory, fig.height=6, fig.width=24, warning=FALSE}
k = quantile_idx[2]
sub_label = subtype_labels
subsample_label = data.frame(participant_id = rownames(subtype_labels))
df_sub = data.frame(participant_id = names(subsample_clusters_full[[k]]),
                    label = subsample_clusters_full[[k]])
mapping = setNames(LETTERS[seq_along(assignment_list[[k]])], 
                   as.character(assignment_list[[k]]))
df_sub$label = mapping[as.character(df_sub$label)]

subsample_label = left_join(subsample_label, df_sub, by = "participant_id")
rownames(subsample_label) = subsample_label$participant_id
idx_nona = which(!is.na(subsample_label$label))


mod_ids = c(1, 2, 3, 10)
negation = c(-1, 1, 1, 1)
# negate factors
factors = t(t(mcia_results$mcia_factors[,mod_ids]) * negation)
factors=apply(factors,2, function(z) qqnorm(z, plot.it = F)$x)
factors = data.frame(factors,row.names = rownames(organized_data$datasets$plasma_proteomics_targeted))
factors = factors[clin_data_aligned$event_id,]
colnames(factors) = paste0("Factor", mod_ids)

assay_mat = factors
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl = subsample_label,
                                 clin_data = clin_data_used)
clin_colnames_select = setdiff(colnames(assay_mat), colnames(factors))

factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "label", use_endpoint_levels = c("A", "C", "E", "F"),
  max_date = 40, text_position = 2)
ggarrange(plotlist = factor_trajectory_analysis_results$test_fig_list, 
          nrow = 1, common.legend = T, legend = "bottom")
```

# Correlation matrix plot (50% case)
```{r correlation matrix plot (50% case), fig.height=6, fig.width=6}
select_corr_mat = cor_mat_list[[k]]
rownames(select_corr_mat) = LETTERS[1:6]
colnames(select_corr_mat) = LETTERS[1:6]
corrplot(select_corr_mat, method = "circle", order = "original",
         tl.col = "black", addCoef.col = "black",
         tl.cex = 1.25, cl.cex = 1, number.cex = 1, number.font = 1,
         col = colorRampPalette(c("#284AA0", "white", "#BC4A4A"))(200))
```

# TG barplot (50% case)
```{r TG barplot (50% case), fig.height=6, fig.width=8}
k = quantile_idx[2]
sub_label = subtype_labels
subsample_label = data.frame(participant_id = rownames(subtype_labels),
                             TG = subtype_labels$TG)
df_sub = data.frame(participant_id = names(subsample_clusters_full[[k]]),
                    label = subsample_clusters_full[[k]])
mapping = setNames(LETTERS[seq_along(assignment_list[[k]])], 
                   as.character(assignment_list[[k]]))
df_sub$label = mapping[as.character(df_sub$label)]
subsample_label = left_join(subsample_label, df_sub, by = "participant_id")
rownames(subsample_label) = subsample_label$participant_id

df_summary = data.frame(Subtype = subsample_label$label,
                        TG = subsample_label$TG)
df_summary = na.omit(df_summary)
df_summary = df_summary[df_summary$Subtype != "B",]
df_summary$Subtype = factor(df_summary$Subtype)
df_summary2 <- df_summary %>%
  dplyr::group_by(Subtype, TG) %>%
  dplyr::summarise(Count = n(), .groups = "drop")
total_counts2 <- df_summary2 %>%
  group_by(Subtype) %>%
  summarise(Total = sum(Count))
df_summary2 <- df_summary2 %>%
  dplyr::group_by(Subtype) %>% 
  mutate(percent = Count/sum(Count))
df_summary2$Subtype <- factor(df_summary2$Subtype, 
                              levels = rev(levels(df_summary2$Subtype)))

ggplot(df_summary2, aes(y = Subtype, x = percent, fill = TG)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
    geom_text(data = total_counts2,
              aes(y = Subtype, x = -0.06,label = sprintf(" %d", Total)),
            position = position_dodge(width = 0.9),
            inherit.aes = FALSE, size = 7)+
  labs(y = "Subtype",
       x = "Percent") +
  theme(axis.title.x = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text.x = element_text(color = "black", size = 15),
        axis.text.y = element_blank(),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.key.size = unit(2, "lines"),
        plot.margin = margin(b = 55, t = 5)) +
  scale_fill_manual(values = c("1" = "#639A21", "2" = "#39828C", 
                               "3" = "#6371AD", "4" = "#BD7D31", 
                               "5" = "#9C3418"))
```

# Subsample TG test
```{r subsample TG test, fig.height=6, fig.width=2.3}
subsample_clust_test = list(); subsample_clust_test[["TG"]] = list()
idx_nona = which(!is.na(subsample_label$label) & (subsample_label$label != "B"))
subsample_clust_test[["TG"]][["ordinal_demo"]] = test_ordinal_demo(
  subtype = subsample_label$label[idx_nona],
  response = subsample_label$TG[idx_nona],
  demo_data = demo_data_adjust[idx_nona, ])

heatmap_ordinal(subsample_clust_test[["TG"]][["ordinal_demo"]])
```

# Barplot colored by PASC
```{r barplot colored by PASC, fig.height=6, fig.width=8}
k = quantile_idx[2]
sub_label = subtype_labels
subsample_label = data.frame(participant_id = rownames(subtype_labels),
                             TG = subtype_labels$TG,
                             PASC = subtype_labels$PASC)
df_sub = data.frame(participant_id = names(subsample_clusters_full[[k]]),
                    label = subsample_clusters_full[[k]])
mapping = setNames(LETTERS[seq_along(assignment_list[[k]])], 
                   as.character(assignment_list[[k]]))
df_sub$label = mapping[as.character(df_sub$label)]
subsample_label = left_join(subsample_label, df_sub, by = "participant_id")
rownames(subsample_label) = subsample_label$participant_id

df_summary <- data.frame(Subtype = factor(subsample_label$label),
                         PASC = subsample_label$PASC,
                         TG =  subsample_label$TG)
df_summary = na.omit(df_summary)
df_summary = df_summary[df_summary$Subtype != "B" &
                          df_summary$Subtype != "D",]
df_summary1 <- df_summary %>%
  dplyr::group_by(Subtype, PASC) %>%
  dplyr::summarise(Count = n(), .groups = "drop")
df_summary1 <- df_summary1[!is.na(df_summary1$PASC),]
total_counts1 <- df_summary1 %>%
  group_by(Subtype) %>%
  summarise(Total = sum(Count))
df_summary1 <- df_summary1 %>%
  dplyr::group_by(Subtype) %>%
  mutate(percent = Count/sum(Count)*100)

df_summary1 <- df_summary1[df_summary1$Subtype != "D", ]
total_counts1 <- total_counts1[total_counts1$Subtype != "D", ]
df_summary1$Subtype <- factor(as.character(df_summary1$Subtype))
df_summary1$PASC <- factor(df_summary1$PASC, labels = c("Minimal", "Physical", 
                                                        "Cognitive", "Multiple"))
df_summary1$Subtype <- factor(df_summary1$Subtype, 
                              levels = rev(levels(df_summary1$Subtype)))

ggplot(df_summary1, aes(y = Subtype, x = percent, fill = PASC)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  geom_text(data = total_counts1,
            aes(y = Subtype, x = -6, label = sprintf(" %d", Total)),
            position = position_dodge(width = 0.9),
            inherit.aes = FALSE, size = 7)+
  labs(y = "Subtype", x = "Percent") +
  theme(axis.title.x = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text.x = element_text(color = "black", size = 15),
        axis.text.y = element_blank(),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.key.size = unit(2, "lines"),
        plot.margin = margin(b = 55, t = 5)) +
  scale_fill_manual(values = c("Minimal" = "#999999", "Physical" = "#58B5EA", 
                               "Cognitive" = "#EAAE2A", "Multiple" = "#FF1820"))
```

# PASC test
```{r PASC test, fig.height=6, fig.width=6}
k = quantile_idx[2]
sub_label = subtype_labels
subsample_label = data.frame(participant_id = rownames(subtype_labels),
                             TG = subtype_labels$TG,
                             PASC = subtype_labels$PASC)
df_sub = data.frame(participant_id = names(subsample_clusters_full[[k]]),
                    label = subsample_clusters_full[[k]])
mapping = setNames(LETTERS[seq_along(assignment_list[[k]])], 
                   as.character(assignment_list[[k]]))
df_sub$label = mapping[as.character(df_sub$label)]
subsample_label = left_join(subsample_label, df_sub, by = "participant_id")
rownames(subsample_label) = subsample_label$participant_id


id = which(!is.na(subsample_label$PASC) & 
             !is.na(subsample_label$label) & 
             subsample_label$label != "D" &
             subsample_label$label != "B")
full_clust_test= list(); full_clust_test[["PASC"]] = list()
ww = ifelse(subsample_label$PASC[id] == "minimal", 0, 1)
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]

full_clust_test[["PASC"]][["fisher"]] = test_clinical_fisher_exact(
  subtype = subsample_label$label[id],
  response = subsample_label$PASC[id],
  response_binary = ww, seed = 123)

heatmap_fisher(full_clust_test[["PASC"]][["fisher"]])
```

# PASC global test
```{r PASC global test, fig.height=6, fig.width=2}
heatmap_fisher_global(full_clust_test[["PASC"]][["fisher"]])
```

# Clinical names correspondence
```{r clinical names correspondence}
clin_names <- c("sex" = "Sex (Female)",
                "ethnicity" = "Ethnicity (Hispanic or Latino)",
                "admit_age" = "Admission age",
                "discretized_admit_age_quantile" = "Admission age quantile",
                "raceAmerican Indian / Alaska Native" = "American Indian/Alaska Native",
                "raceAsian" = "Asian",
                "raceBlack / African American" = "Black/African American",
                "raceBlack...African.American" = "Black/African American",
                "raceMultiple" = "Multiple Races",
                "raceNative Hawaiian / Pacific Islander" = "Native Hawaiian/Pacific Islander",
                "raceWhite" = "White",
                "bmicat" = "BMI category",
                "comorb_htn" = "Hypertension",
                "comorb_isaric_dm" = "Diabetes",
                "comorb_anyresp_noasthma" = "Pulmonary (excludes asthma)",
                "comorb_isaric_asthma" = "Asthma",
                "comorb_isaric_cardiac" = "Chronic cardiac disease",
                "comorb_isaric_ckd" = "Chronic kidney disease",
                "comorb_isaric_neoplasm" = "Malignant neoplasm",
                "comorb_neuro" = "Chronic neurological disorder",
                "comorb_isaric_liver" = "Liver disease",
                "comorb_hxtrans" = "History of solid organ or bone marrow transplant",
                "comorb_hiv2" = "HIV",
                "comorb_eversmkvape" = "Current or former smoking and/or vaping",
                "comorb_anysubstance" = "Drug or alcohol abuse or cannabis use",
                "comp_seizure" = "Seizures",
                "comp_meningitis" = "Meningitis/encephalitis",
                "comp_stroke" = "Stroke/CVA",
                "comp_afib" = "Atrial fibrillation",
                "comp_ventricular_arr" = "Ventricular arrhythmia",
                "comp_endo" = "Endocarditis",
                "comp_myocard" = "Myocarditis/pericarditis",
                "comp_cardarrest" = "Cardiac arrest",
                "comp_stemi" = "ST-elevation MI (STEMI)",
                "comp_nstemi" = "Non-ST elevation MI (NSTEMI) or other cardiac ischemia",
                "comp_chf" = "CHF/cardiomyopathy",
                "comp_pneumonia" = "Pneumonia",
                "comp_bronch" = "Bronchiolitis",
                "comp_pneumo" = "Pneumothorax",
                "comp_pneumomed" = "Pneumomediastinum",
                "comp_aat" = "Acute arterial thromboembolism excluding stroke/CVA or myocardial infarction",
                "comp_avthromb" = "Acute venous thromboembolism",
                "comp_shock" = "Shock (use of vasopressors)",
                "comp_bac" = "Bacteremia",
                "comp_coag" = "Coagulation disorder/disseminated intravascular coagulation",
                "comp_hypogly" = "Symptomatic hypoglycemia",
                "comp_bleed" = "Bleeding",
                "comp_anemia" = "Anemia",
                "comp_pancreas" = "Pancreatitis",
                "comp_renal" = "Acute renal injury/failure",
                "comp_liver" = "Liver dysfunction/failure",
                "comp_othlab" = "Other lab abnormality",
                "comp_electrolyte" = "Electrolyte abnormality",
                "comp_othinf" = "Other infection (non-pulmonary)",
                "comp_othpulm" = "Other pulmonary complication",
                "comp_unddisease" = "Exacerbation of underlying disease",
                "comp_enceph" = "Encephalopathy",
                "comp_othneuro" = "Other neurologic complication",
                "comp_hypergly" = "Hyperglycemia",
                "comp_psych" = "Psychiatric complication",
                "comp_mis" = "Multisystem inflammatory syndrome (MIS)",
                "comp_gastroent" = "Gastroenteritis",
                "comp_pulmvas" = "Pulmonary vascular disease",
                "ever_azithro" = "Azithromycin",
                "ever_icu" = "Ever in ICU",
                "ever_otherabx" = "Other Antibiotics",
                "ever_remdes" = "Remdesivir",
                "ever_steroids" = "Steroids",
                "ever_ever_esc" = "Ever required escalation of care",
                "ever_immsupp" = "Immunosuppressives",
                "baseline_anyimg" = "Any chest imaging at baseline",
                "baseline_img_infil" = "Radiographic findings on chest imaging",
                "baseline_lymph_abn" = "Abnormal Lymphocyte count",
                "baseline_platelets_abn" = "Abnormal Platelets",
                "baseline_alt_abn" = "Abnormal ALT",
                "baseline_cr_abn" = "Abnormal Creatinine",
                "baseline_crp_abn" = "Abnormal CRP",
                "baseline_ddimer_abn" = "Abnormal D-dimer",
                "baseline_trop_abn" = "Abnormal Troponin")
```

```{r}
interest_clin_feature = c(
  "comorb_htn", "comorb_anyresp_noasthma", "comorb_isaric_cardiac", 
  "comorb_isaric_ckd", "comorb_hxtrans", "comp_stroke", "comp_afib",
  "comp_ventricular_arr", "comp_cardarrest", "comp_chf", "comp_pneumonia", 
  "comp_avthromb", "comp_shock", "comp_bac", "comp_bleed", "comp_anemia", 
  "comp_renal", "ever_icu", "ever_ever_esc", "sex", "ethnicity", 
  "raceBlack / African American", "raceWhite", "baseline_lymph_abn", 
  "baseline_platelets_abn", "baseline_alt_abn", "baseline_cr_abn", 
  "baseline_crp_abn", "baseline_ddimer_abn", "baseline_trop_abn", 
  "bmicat", "discretized_admit_age_quantile", "baseline_img_infil")
```

# Demographic, comorbidity, baseline, and complication
```{r clinical heatmap, fig.height=7.7, fig.width=14, warning=FALSE}
clin_data_patient <- clin_data_used[clin_data_used$event_type == "Visit 1",]
demo_data <- clin_data_patient[, c("admit_age", "sex", "ethnicity",
                                   "discretized_admit_age_quantile",
                                   "bmicat", "enrollment_site", "race")]
demo_data$ethnicity[demo_data$ethnicity == "Not Specified"] <- NA
demo_data$bmicat[demo_data$bmicat == 999] <- NA
demo_data$bmicat_test <- demo_data$bmicat
# group obesity together
demo_data$bmicat_test[demo_data$bmicat_test == 5] <- 4
demo_data$race <- as.factor(demo_data$race)
race_one_hot <- model.matrix(~ race-1, data = demo_data)
race_one_hot <- race_one_hot[, !(colnames(race_one_hot) %in%
                                   c("raceOther / Declined", 
                                     "raceUnknown / Unavailable"))]
demo_data_adjust <- demo_data[, c("discretized_admit_age_quantile", "sex")]

clin_data_logit <- clin_data_patient %>%
  dplyr::select(matches("^(comorb|comp)"), ever_icu, ever_ever_esc) %>%
  dplyr::select(-comorb_count, -comorb_countcat, -comp_other2) %>%
  select_if(~sum(.x, na.rm = TRUE) > 0)
clin_data_demo_logit <- cbind(demo_data[, c("sex", "ethnicity")], race_one_hot)
clin_data_demo_logit$sex <- ifelse(clin_data_demo_logit$sex == "Female", 1, 0)
clin_data_demo_logit$ethnicity <- ifelse(clin_data_demo_logit$ethnicity == 
                                           "Hispanic or Latino", 1, 0)

clin_data_baseline <- clin_data_patient %>%
  dplyr::select(matches("^baseline")) %>%
  dplyr::select(-baseline_anyimg) %>%
  mutate(across(everything(), ~replace(., . == 999, NA)))
clin_data_baseline_logit <- clin_data_baseline %>%
  dplyr::select(ends_with("_abn"))


k = quantile_idx[2]
sub_label = subtype_labels
subsample_label = data.frame(participant_id = rownames(subtype_labels),
                             TG = subtype_labels$TG,
                             PASC = subtype_labels$PASC)
df_sub = data.frame(participant_id = names(subsample_clusters_full[[k]]),
                    label = subsample_clusters_full[[k]])
mapping = setNames(LETTERS[seq_along(assignment_list[[k]])], 
                   as.character(assignment_list[[k]]))
df_sub$label = mapping[as.character(df_sub$label)]
subsample_label = left_join(subsample_label, df_sub, by = "participant_id")
rownames(subsample_label) = subsample_label$participant_id


sub_idx <- which(!is.na(subsample_label$label) &
                   subsample_label$label != "D" &
                   subsample_label$label != "B")
clin_test_logit <- test_clinical_logit_demo(
  subsample_label$label[sub_idx],
  clin_data_logit[sub_idx,],
  demo_data_adjust[sub_idx,])
demo_test_logit <- test_clinical_logit(
  subsample_label$label[sub_idx],
  clin_data_demo_logit[sub_idx,])
baseline_test_logit <- test_clinical_logit_demo(
  subsample_label$label[sub_idx],
  clin_data_baseline_logit[sub_idx,],
  demo_data_adjust[sub_idx,])
bmi_test_ordinal <- test_ordinal(
  subsample_label$label[sub_idx],
  demo_data$bmicat_test[sub_idx])
age_test_ordinal <- test_ordinal(
  subsample_label$label[sub_idx],
  demo_data$discretized_admit_age_quantile[sub_idx])
baseline_test_ordinal <- test_ordinal_demo(
  subsample_label$label[sub_idx],
  clin_data_baseline$baseline_img_infil[sub_idx],
  demo_data_adjust[sub_idx,])

clin_test_logit$p_lr_adj <- p.adjust(clin_test_logit$p_lr, method = "BH")
demo_test_logit$p_lr_adj <- p.adjust(demo_test_logit$p_lr, method = "BH")

# clinical test featuring variables of interest
clin_test <- clin_test_logit
# interest_clin_idx <- which(clin_test_logit$p_lr_adj < 0.001)
# interest_demo_idx <- which(demo_test_logit$p_lr_adj < 0.001)
clin_test$est_logit <- cbind(clin_test_logit$est_logit, 
                             demo_test_logit$est_logit,
                             baseline_test_logit$est_logit,
                             bmi_test_ordinal$est_ordinal,
                             age_test_ordinal$est_ordinal,
                             baseline_test_ordinal$est_ordinal)
colnames(clin_test$est_logit)[ncol(clin_test$est_logit)-2] <- "bmicat"
colnames(clin_test$est_logit)[ncol(clin_test$est_logit)-1] <- 
  "discretized_admit_age_quantile"
colnames(clin_test$est_logit)[ncol(clin_test$est_logit)] <- "baseline_img_infil"
clin_test$est_logit = clin_test$est_logit[,interest_clin_feature]

clin_test$p_logit <- cbind(clin_test_logit$p_logit, 
                           demo_test_logit$p_logit,
                           baseline_test_logit$p_logit,
                           bmi_test_ordinal$p_ordinal,
                           age_test_ordinal$p_ordinal,
                           baseline_test_ordinal$p_ordinal)
colnames(clin_test$p_logit)[ncol(clin_test$p_logit)-2] <- "bmicat"
colnames(clin_test$p_logit)[ncol(clin_test$p_logit)-1] <- 
  "discretized_admit_age_quantile"
colnames(clin_test$p_logit)[ncol(clin_test$p_logit)] <- "baseline_img_infil"
clin_test$p_logit = clin_test$p_logit[,interest_clin_feature]

clin_test$p_adj <- apply(clin_test$p_logit, 2,
                         function(p) p.adjust(p, method = "BH"))

heatmap_clinical(clin_test, clin_names, adjust_p = TRUE)
```

# Prediction

## Clinical data
```{r}
# clinical data
clinical_data <- clin_data_aligned[!duplicated(clin_data_aligned$participant_id),]
clinical_data <- as.data.frame(clinical_data)
rownames(clinical_data) <- clinical_data$participant_id
clinical_data <- clinical_data %>% dplyr::select(participant_id, admit_age, sex,  
                                                  bmicat, los, ncomp, 
                                                  baseline_sofa_score_new, 
                                                  matches("comorb"))
clinical_data$comorb_countcat <- NULL
clinical_data[clinical_data$sex=='Female',]$sex <- 1
clinical_data[clinical_data$sex=='Male',]$sex <- 0
clinical_data$sex <- as.numeric(clinical_data$sex)
spike_IgG <- rbind(base_obj$others$serum_rbd_abtiters$train$serum_rbd_abtiters_counts,
                   base_obj$others$serum_rbd_abtiters$test$serum_rbd_abtiters_counts)

spike_IgG$N1_CT <- rbind(base_obj$others$nasal_viralload$train$nasal_viralload_counts,
                         base_obj$others$nasal_viralload$test$nasal_viralload_counts)$N1_CT
spike_IgG <- spike_IgG[,c('AUC Spike IgG','N1_CT')]
spike_IgG <- na.omit(spike_IgG)
spike_IgG$participant_id <- substr(rownames(spike_IgG), start = 1, stop = 8)

# ema of viral loads and antibody
spike_IgG_ema <- aggregate(. ~ participant_id, data = spike_IgG, FUN=mov_ave)
rownames(spike_IgG_ema) <- spike_IgG_ema$participant_id

# V1 of viral loads and antibody
spike_IgG_v1 <- spike_IgG[!duplicated(spike_IgG$participant_id),]
# spike_IgG_v1 <- spike_IgG[substr(rownames(spike_IgG), start = 10, stop = 11)=="01",]
rownames(spike_IgG_v1) <- spike_IgG_v1$participant_id

clinical_data_ema <- merge(clinical_data, spike_IgG_ema, by='participant_id')
rownames(clinical_data_ema) <- clinical_data_ema$participant_id
clinical_data_ema$participant_id <- NULL
clinical_data_ema <- na.omit(clinical_data_ema)
clinical_data_ema <- clinical_data_ema[clinical_data_ema$bmicat!=999,]

clinical_data_v1 <- merge(clinical_data, spike_IgG_v1, by='participant_id')
rownames(clinical_data_v1) <- clinical_data_v1$participant_id
clinical_data_v1$participant_id <- NULL
clinical_data_v1 <- na.omit(clinical_data_v1)
clinical_data_v1 <- clinical_data_v1[clinical_data_v1$bmicat!=999,]
```

## PASC data
```{r}
# PASC
conval_raw <- organized_data$clinical_data$conv_cluster6
names(conval_raw) <- organized_data$clinical_data$event_id
conval_raw <- na.omit(conval_raw)
conval_raw[conval_raw == 3 | conval_raw == 4] = 1
conval_raw[conval_raw == 5] = 3
conval_raw[conval_raw == 6] = 4
conval_raw <- conval_raw[!duplicated(gsub('-[0-9]+$', '', names(conval_raw)))]
names(conval_raw) <- gsub('-[0-9]+$', '', names(conval_raw))
conval_raw <- conval_raw[sort(names(conval_raw))]
```

## Data for prediction
```{r}
# data for prediction
# ema data
id <- Reduce(intersect, list(rownames(clinical_data_ema), 
                             rownames(factor_scores_ema), 
                             rownames(log_probs),
                             names(conval_raw)))
data_pred_ema <- list()
data_pred_ema[['conval']] <- conval_raw[id]
data_pred_ema[['clinical']] <- clinical_data_ema[id,]
data_pred_ema[['subtype']] <- log_probs[id,]
data_pred_ema[['factor']] <- factor_scores_ema[id,]

# V1 data
id_v1 <- Reduce(intersect, list(rownames(clinical_data_v1), 
                             rownames(factor_scores_v1), 
                             rownames(log_probs),
                             names(conval_raw)))
data_pred_v1 <- list()
data_pred_v1[['conval']] <- conval_raw[id_v1]
data_pred_v1[['clinical']] <- clinical_data_v1[id_v1,]
data_pred_v1[['subtype']] <- log_probs[id_v1,]
data_pred_v1[['factor']] <- factor_scores_v1[id_v1,]
```

## Prediction function
```{r}
prediction <- function(data, pred_model) { 
  models <- list()
  test_sets <- list()
  conval <- data$conval
  repeats <- 10
  rocs <- vector("list", repeats)
  prcs <- vector("list", repeats)
  aucs <- rep(0, repeats)
  auprcs <- rep(0, repeats)
  set.seed(3261)
  for (s in 1:repeats) {
    # split data
    ind_train <- sample(1:length(conval), floor(length(conval)*0.7))
    conval_train <- conval[ind_train]
    conval_test <- conval[-ind_train]

    # binary response
    conval01_train <- conval_train
    conval01_train[conval01_train != 1] <- 0
    conval01_test <- conval_test
    conval01_test[conval01_test != 1] <- 0

    d_train <- data.frame(Column1 = 1:length(conval01_train))
    d_test <- data.frame(Column1 = 1:length(conval01_test))
    rownames(d_train) <- names(conval01_train)
    rownames(d_test) <- names(conval01_test)
    # clinical
    d_train <- cbind(d_train, data$clinical[rownames(d_train),])
    d_test <- cbind(d_test, data$clinical[rownames(d_test),])
    # subtype
    if (pred_model != "CLINICAL") {
      d_train <- cbind(d_train, data$subtype[rownames(d_train),])
      d_test <- cbind(d_test, data$subtype[rownames(d_test),])
    }
    # factor scores
    if (pred_model == "CLINICAL-SUBTYPE+") {
      d_train <- cbind(d_train, data$factor[rownames(d_train),])
      d_test <- cbind(d_test, data$factor[rownames(d_test),])
    }
    
    d_train$Column1 <- NULL
    d_test$Column1 <- NULL

    d_train$conval <- conval01_train
    d_train$conval[d_train$conval == 0] <- "level0"
    d_train$conval[d_train$conval == 1] <- "level1"
    d_train$conval <- factor(d_train$conval)
    d_test$conval <- conval01_test
    d_test$conval[d_test$conval == 0] <- "level0"
    d_test$conval[d_test$conval == 1] <- "level1"
    d_test$conval <- factor(d_test$conval)

    control <- trainControl(method = "cv", number = 5, summaryFunction = twoClassSummary, classProbs = TRUE)
    model = train(conval ~ ., data = d_train, trControl = control, method = "svmRadial", metric = "ROC")
    pred <- predict(model, d_test, type = 'prob')[,2]

    rocs[[s]] <- pROC::roc(d_test$conval, pred)
    aucs[s] <- rocs[[s]]$auc
    print(aucs[s])
    
    prcs[[s]] <- PRROC::pr.curve(scores.class0 = 1-pred, weights.class0 = 1-conval01_test, curve = TRUE)
    auprcs[s] <- prcs[[s]]$auc.integral
    print(auprcs[s])
    
    models[[s]] <- model
    test_sets[[s]] <- d_test
  }
  print(paste0("auc mean: ", mean(aucs)))
  print(paste0("auc std: ", sd(aucs)))
  print(paste0("auprc mean: ", mean(auprcs)))
  print(paste0("auprc std: ", sd(auprcs)))
  res <- list(aucs, rocs, auprcs, prcs, models, test_sets)
  names(res) <- c("aucs", "rocs", "auprcs", "prcs", "models", "test_sets")
  return(res)
}
```

## Subsample data for prediction
```{r}
aucs_clinical_list <- list()
auprcs_clinical_list <- list()
aucs_clinical_prob_list <- list()
auprcs_clinical_prob_list <- list()
aucs_clinical_factor_prob_list <- list()
auprcs_clinical_factor_prob_list <- list()

for (i in 1:n_subsample) {
  # ema data
  print(i)
  id <- Reduce(intersect, list(rownames(clinical_data_ema), 
                               rownames(factor_scores_ema),
                               rownames(sub_log_probs_list[[i]]),
                               names(conval_raw)))
  data_pred_ema <- list()
  data_pred_ema[['conval']] <- conval_raw[id]
  data_pred_ema[['clinical']] <- clinical_data_ema[id,]
  data_pred_ema[['subtype']] <- sub_log_probs_list[[i]][id,]
  data_pred_ema[['factor']] <- factor_scores_ema[id,]
  
  res <- prediction(data_pred_ema, "CLINICAL")
  aucs_clinical <- res$aucs
  rocs_clinical <- res$rocs
  auprcs_clinical <- res$auprcs
  prcs_clinical <- res$prcs
  models_clinical <- res$models
  test_sets_clinical <- res$test_sets
  
  res <- prediction(data_pred_ema, "CLINICAL-SUBTYPE")
  aucs_clinical_prob <- res$aucs
  rocs_clinical_prob <- res$rocs
  auprcs_clinical_prob <- res$auprcs
  prcs_clinical_prob <- res$prcs
  models_clinical_prob <- res$models
  test_sets_clinical_prob <- res$test_sets
  
  res <- prediction(data_pred_ema, "CLINICAL-SUBTYPE+")
  aucs_clinical_factor_prob <- res$aucs
  rocs_clinical_factor_prob <- res$rocs
  auprcs_clinical_factor_prob <- res$auprcs
  prcs_clinical_factor_prob <- res$prcs
  models_clinical_factor_prob <- res$models
  test_sets_clinical_factor_prob <- res$test_sets
  
  aucs_clinical_list[[i]] <- aucs_clinical
  auprcs_clinical_list[[i]] <- auprcs_clinical
  aucs_clinical_prob_list[[i]] <- aucs_clinical_prob
  auprcs_clinical_prob_list[[i]] <- auprcs_clinical_prob
  aucs_clinical_factor_prob_list[[i]] <- aucs_clinical_factor_prob
  auprcs_clinical_factor_prob_list[[i]] <- auprcs_clinical_factor_prob
}
```

# Visualization
## Barplot of AUROC, AUPRC
```{r fig.height=6, fig.width=6}
aucs_clinical <- unlist(aucs_clinical_list)
auprcs_clinical <- unlist(auprcs_clinical_list)
aucs_clinical_prob <- unlist(aucs_clinical_prob_list)
auprcs_clinical_prob <- unlist(auprcs_clinical_prob_list)
aucs_clinical_factor_prob <- unlist(aucs_clinical_factor_prob_list)
auprcs_clinical_factor_prob <- unlist(auprcs_clinical_factor_prob_list)


df <- data.frame(
  category = c("Average AUROC", "Average AUPRC"),
  value1 = c(mean(aucs_clinical), mean(auprcs_clinical)),
  value2 = c(mean(aucs_clinical_prob), mean(auprcs_clinical_prob)),
  value3 = c(mean(aucs_clinical_factor_prob), mean(auprcs_clinical_factor_prob)),
  value4 = c(0.5, 0.3715517)
)

df <- tidyr::pivot_longer(df, cols = c(value1, value2, value3, value4), 
                          names_to = "model", values_to = "value")
# df$category <- factor(df$category, levels = c("Average AUROC", "Average AUPRC"))

aucs <- list(aucs_clinical, aucs_clinical_prob, aucs_clinical_factor_prob,
             auprcs_clinical, auprcs_clinical_prob, auprcs_clinical_factor_prob)
texts <- c()
for (i in 1:length(aucs)){
  aucs_svm <- aucs[[i]]
  M <- mean(aucs_svm)
  SD <- sd(aucs_svm)
  texts[i] <- paste0(sprintf("%.3f",round(M,3)), "\n(SD ", sprintf("%.3f",round(SD,3)), ")")
}
texts <- c(texts[1:3], "0.500", texts[4:6], "0.372")
df$text <- texts

# Create the bar plot
ggplot(df, aes(x = category, y = value, fill = model)) +
  geom_bar(data = subset(df, model != "value4"), 
           stat = "identity", position = "dodge") +
  geom_bar(data = subset(df, model == "value4"),
           stat = "identity", position = "dodge", alpha = 0.8) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.8), 
                     breaks = seq(0, 0.8, by = 0.2)) +
  scale_fill_manual(values = c(value1="#F8766D",value2="#7CAE00",
                               value3="#00BFC4",value4="darkgray"),
                    labels = c("CLINICAL","CLINICAL-SUBTYPE",
                               "CLINICAL-SUBTYPE+",
                               "Chance (random classifier)")) +
  guides(fill = guide_legend(ncol = 2)) + 
  labs(x = "", y = "", fill = "") +
  theme_minimal() +
  theme(
        panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank(),  
        panel.border = element_blank(),  
        axis.line.x = element_line(linewidth = 1),  
        axis.line.y = element_line(linewidth = 1),  
        axis.ticks = element_line(linewidth = 1),  
        axis.text.x = element_text(color = "black", size = 15),  
        axis.text.y = element_text(color = "black", size = 15),  
        axis.title = element_text(color = "black", size = 15),  
        axis.ticks.length = unit(0.15, "cm"),  
        legend.position = "top", 
        legend.justification = "left",
        legend.direction = "vertical",
        legend.text = element_text(size = 15),
        plot.margin = margin(r = 20, l = 10),
        aspect.ratio = 0.85) +
  geom_text(data = subset(df, model != "value4"),
            aes(label = text), position = position_dodge(width = 0.9),  
            vjust = -0.5, size = 3) +
  geom_text(data = subset(df, model == "value4"),
            aes(label = text), position = position_dodge(width = 0.9),  
            vjust = -0.5, size = 3)
```

# boxplot of AUROC and AUPRC increment
```{r fig.height=6, fig.width=6}
# Categories
categories <- rep(c("AUROC", "AUPRC"), each = 10 * n_subsample)

# Values for the first box
values1 <- c(unlist(aucs_clinical_prob_list) - unlist(aucs_clinical_list),
             unlist(auprcs_clinical_prob_list) - unlist(auprcs_clinical_list))

# Values for the second box
values2 <- c(unlist(aucs_clinical_factor_prob_list) - unlist(aucs_clinical_list),
             unlist(auprcs_clinical_factor_prob_list) - unlist(auprcs_clinical_list))

# Combine into a data frame
df <- data.frame(category = rep(categories, 2),
                 value = c(values1, values2),
                 group = rep(c("CLINICAL-SUBTYPE", "CLINICAL-SUBTYPE+"), 
                             each = 20*n_subsample))

# add p-values
p.val <- function(x){
  return(wilcox.test(rep(0,10*n_subsample), x, 
                     paired = TRUE, alternative = "less")$p.value)
}
p_values <- aggregate(value ~ category + group, data = df, FUN = p.val)
p_values$value <- c("<0.001 ***", "<0.001 ***", "<0.001 ***", "<0.001 ***")

ggplot(df, aes(x = category, y = value, color = group)) +
  geom_boxplot(position = position_dodge(width = 0.8), size = 1, width = 0.7, outlier.shape = NA) +
  # geom_jitter(aes(color = group), position = position_dodge(width = 0.8), size = 2.5, alpha = 0.5) +
  scale_y_continuous(limits = c(-0.015, 0.1), breaks = seq(0, 0.1, by = 0.02)) + 
  labs(x = "", y = "Increment over CLINICAL", color = NULL) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank(),  
        panel.border = element_blank(),  
        axis.line.x = element_line(linewidth = 1),  
        axis.line.y = element_line(linewidth = 1),  
        axis.ticks = element_line(linewidth = 1),  
        axis.text.x = element_text(color = "black", size = 15),  
        axis.text.y = element_text(color = "black", size = 15),  
        axis.title = element_text(color = "black", size = 15),  
        axis.ticks.length = unit(0.15, "cm"),  
        legend.position = "top",  
        # legend.justification = "left",
        # legend.direction = "vertical",
        legend.text = element_text(size = 15),
        plot.margin = margin(r = 20, l = 10),
        aspect.ratio = 0.85
  ) + 
  geom_text(data = p_values, aes(label = value, y = 0.1, x=c(0.8,1.82,1.22,2.22)), size = 5, vjust = 0, 
            position = position_dodge(width = 0.8), color = "black") +
  coord_cartesian(clip = "off")
```




