---
title: "20250320_bulk_RNAseq_pbmc"
output: html_document
---

Here I analyze host gene expression vs antibiotics exposure.

R version 4.3.3

```{r}
library(tidyverse)
library(ggplot2)
# library(pheatmap)
library(patchwork)

library(limma)
# library(edgeR)

library(fgsea)

# library(gamm4)

setwd("~/20240208_IMPACC_ABX")

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )
which.script <- "20250320_bulk_RNAseq_pbmc.Rmd"
```


# Data import

## Clinical metadata

```{r}
# Patient metadata
clin.individ <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-locked.csv",
  header = TRUE)

# Sample
clin.sample <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-sample-locked.csv",
  header = TRUE)

# Event
clin.event <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-event-locked.csv",
  header = TRUE)
```

Supplemental data for steroid.

```{r}
clin.supp <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-supp-locked.csv",
  header=TRUE
)
```

Convert categorical variables to factor

```{r}
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

Any empty steroid fields?

```{r}
print(any(is.na(clin.supp$ever_steroids)))

clin.supp$ever_steroids <- as.factor(clin.supp$ever_steroids)
```

### Merge clinical metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

Add steroid info

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event.individ,
  clin.supp[,c("participant_id","ever_steroids")],
  by="participant_id",
  all=FALSE,
  sort=FALSE
)
```

## ABX metadata 

Antibiotics data prepared by Cole

```{r}
abx <- read.csv(
  "/scratch/metagenomics/antibiotic_data/pbmc_group_assignment_v2.csv"
)

# Merge antibiotics data with clinical metadata
clin.sample.event.individ <- merge(
  clin.sample.event.individ,
  abx[,c("sample_id","group_id","timepoint")],
  by="sample_id",
  all=FALSE
) %>%
  droplevels()

print(dim(clin.sample.event.individ))
```

How many samples does each Visit have?

```{r}
print(table(clin.sample.event.individ$event_type, clin.sample.event.individ$timepoint))
```

I will remove Control donor visit, without filtering out other event types.

```{r}
clin.sample.event.individ <- clin.sample.event.individ %>%
  subset(event_type != "Control donor visit")
print(dim(clin.sample.event.individ))
print(table(clin.sample.event.individ$event_type, clin.sample.event.individ$timepoint))
```

Do samples from the same patients show up in the same time point?

```{r}
clin.sample.event.individ %>%
  group_by(participant_id, timepoint) %>%
  summarise(n=n()) %>%
  arrange(desc(n))
```

I will need to model participant_id as random effects.

## RNA-seq data

Import gene counts.

```{r}
counts <- read.csv(
  "/data/pbmc-transcriptomics/legacy/2022-06-15/pbmc-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1,
  check.names = FALSE
)
counts <- t(counts)
print(counts[1:6,1:6])

counts.metadata <- read.csv(
  "/data/pbmc-transcriptomics/legacy/2022-06-15/pbmc-transcriptomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
print(head(counts.metadata))

rData <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-RowFeature.csv",
  header = FALSE, row.names = NULL,
  check.names = FALSE
)
stopifnot(rownames(counts)==rData$gene_id)
colnames(rData) <- rData[2,]
rData <- rData[-(1:2),]
print(head(rData))
```

```{r}
# RNA-seq QC
counts.metadata <- subset(
  counts.metadata,
  sample_status == "IMPACC sample assayed and passed QC"
)

# Filter clinical metadata data frame
clin.sample.event.individ <- clin.sample.event.individ %>%
  subset(sample_id %in% counts.metadata$sample_id)
```

```{r}
print(dim(counts))
```

## Load pathways for GSEA

```{r}
# Reactome
reactome <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "C2", subcategory = "CP:REACTOME")
reactome.list <- split(x = reactome$ensembl_gene, f = reactome$gs_name) # for fgsea
reactome.list2 <- reactome %>% select(gs_name, ensembl_gene) # for clusterProfiler

# # Hallmark gene sets
# hallmark <- msigdbr::msigdbr(species = "Homo sapiens",
#                              category = "H")
# hallmark.list <- split(x = hallmark$ensembl_gene, f = hallmark$gs_name)
# hallmark.list2 <- hallmark %>% select(gs_name, ensembl_gene)
```

# Session info

```{r}
sessionInfo()
```

# Azithro vs no Abx at both time points

## Day 1-2

Subset timepoint 1-2, and remove "Other ABX" samples

```{r}
print(dim(clin.sample.event.individ))
clin.sample.event.individ.1a <- clin.sample.event.individ %>%
  subset(group_id!="Other ABX" & timepoint=="Day 1-2")
print(dim(clin.sample.event.individ.1a))
```

I should get 1847 samples before removing Other ABX, and 1006 samples after removing it.

Count patients by group and timepoint

```{r}
clin.sample.event.individ.1a %>%
  group_by(participant_id, group_id, timepoint) %>%
  summarise(n=n()) %>%
  arrange(desc(n))
```

Make "No ABX" the first level

```{r}
clin.sample.event.individ.1a$group_id <- factor(
  clin.sample.event.individ.1a$group_id,
  levels=c("No ABX","Azithromycin")
)
```

Subset gene counts

```{r}
# Subset count data
counts.1a <- counts[,clin.sample.event.individ.1a$sample_id]

# Count number of patients
print(dim(counts.1a))
print(table(clin.sample.event.individ.1a$event_type))
```

Gene filtering.

```{r}
# Keep only protein_coding genes
counts.1a <- counts.1a[rData[rData$gene_biotype=="protein_coding", "gene_id"],]

# Keep genes with at least 10 counts in 20% of samples
keep <- rowSums(counts.1a>=10) >= (0.2*ncol(counts.1a))
counts.1a <- counts.1a[keep,]
```

There should be 1006 samples in total.

```{r}
print(dim(counts.1a))
```

Design matrix

```{r}
stopifnot(colnames(counts.1a)==clin.sample.event.individ.1a$sample_id)

design <- model.matrix(
  ~ group_id + admit_age + sex + trajectory_group + ever_steroids + event_date,
  data=clin.sample.event.individ.1a)
print(colnames(design))
colnames(design)[2] <- "Azithro"
print(colnames(design))
```

Run voom with duplicateCorrelation twice: https://support.bioconductor.org/p/59700/ and Section 18.1.9 in the [User Guide](https://www.bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)

duplicateCorrelation is slow!

```{r}
vwts <- voom(
  counts.1a, 
  design = design,
  normalize.method = "quantile",
  plot = T
)
corfit <- duplicateCorrelation(
  vwts,
  design = design,
  block = clin.sample.event.individ.1a$participant_id
)
print(paste0("Correlation, 1st iteration: ", corfit$consensus))
vwts <- voom(
  counts.1a, 
  design = design,
  normalize.method = "quantile",
  block = clin.sample.event.individ.1a$participant_id,
  correlation = corfit$consensus,
  plot = T
)
corfit <- duplicateCorrelation(
  vwts,
  design = design,
  block = clin.sample.event.individ.1a$participant_id
)
print(paste0("Correlation, 2nd iteration: ", corfit$consensus))

vfit <- lmFit(
  vwts,
  design = design,
  block = clin.sample.event.individ.1a$participant_id,
  correlation = corfit$consensus
)
vfit <- eBayes(vfit)
```

Extract DE result

```{r}
res.1a <- topTable(
  vfit,
  coef = "Azithro",
  sort.by = "none",
  number = Inf,
  adjust.method = "BH",
  p.value = 1
)

head(res.1a)
```

Count number of significant genes

```{r}
print(sum(res.1a$adj.P.Val<0.05))
```

Add gene symbols, then export the DE result

```{r}
temp <- merge(
  res.1a, rData[,c("gene_id","gene_name")],
  by.x="row.names", by.y="gene_id",
  all.x=TRUE, all.y=FALSE,
  sort=FALSE
)
colnames(temp)[1] <- "gene_id"

write.csv(
  temp,
  "20250320_bulk_RNAseq_PBMC_Day1-2_DE.csv",
  row.names=FALSE
)
```

Volcano plot

```{r}
# Identify direction
res.toplot <- res.1a
res.toplot$direction <- ifelse(
  res.toplot$adj.P.Val>=0.05, "Not significant",
  ifelse(
    res.toplot$logFC>0, "Up", "Down"
  )
)

# Limit minimum P value to 1e-10
res.toplot[res.toplot$adj.P.Val<1e-10,"adj.P.Val"] <- 1e-10

p <- ggplot(data=res.toplot) +
  geom_point(aes(x=logFC, y=-log10(adj.P.Val), color=direction), size=0.8) +
  scale_color_manual(
    values=c("Up"="darkred","Down"="dodgerblue4","Not significant"="gray75"),
    name="", guide="none") +
  scale_x_continuous(limits=c(-2,2), expand=c(0,0)) +
  labs(x="log2FC", y="-log10(adjusted P)",
       title="PBMC genes associated\nwith Azithromycin at Day 1-2") +
  my.theme
```

Reactome GSEA

```{r}
# Sort gene list in descending order
gene.ranks <- res.1a$t
names(gene.ranks) <- rownames(res.1a)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
```

## Day 4-6

Subset timepoint 4-6, and remove "Other ABX" samples

```{r}
print(dim(clin.sample.event.individ))
clin.sample.event.individ.1b <- clin.sample.event.individ %>%
  subset((group_id!="Other ABX") & (timepoint=="Day 4-6"))
print(dim(clin.sample.event.individ.1b))
```

I should get 847 samples before removing Other ABX, and 238 samples after removing it.

Make "No ABX" the first level

```{r}
clin.sample.event.individ.1b$group_id <- factor(
  clin.sample.event.individ.1b$group_id,
  levels=c("No ABX","Azithromycin")
)
```

Subset gene counts

```{r}
# Subset count data
counts.1b <- counts[,clin.sample.event.individ.1b$sample_id]

# Count number of patients
print(dim(counts.1b))
print(table(clin.sample.event.individ.1b$event_type))
```

Gene filtering.

```{r}
# Keep only protein_coding genes
counts.1b <- counts.1b[rData[rData$gene_biotype=="protein_coding", "gene_id"],]

# Keep genes with at least 10 counts in 20% of samples
keep <- rowSums(counts.1b>=10) >= (0.2*ncol(counts.1b))
counts.1b <- counts.1b[keep,]
```

There should be 238 samples in total

```{r}
print(dim(counts.1b))
```

Design matrix

```{r}
stopifnot(colnames(counts.1b)==clin.sample.event.individ.1b$sample_id)

design <- model.matrix(
  ~ group_id + admit_age + sex + trajectory_group + ever_steroids + event_date,
  data=clin.sample.event.individ.1b)
print(colnames(design))
colnames(design)[2] <- "Azithro"
print(colnames(design))
```

Count patients by group and timpeoint

```{r}
clin.sample.event.individ.1b %>%
  group_by(participant_id, group_id, timepoint) %>%
  summarise(n=n()) %>%
  arrange(desc(n))
```

Run voom with duplicateCorrelation twice: https://support.bioconductor.org/p/59700/ and Section 18.1.9 in the [User Guide](https://www.bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)

duplicateCorrelation is slow!

```{r}
vwts <- voom(
  counts.1b, 
  design = design,
  normalize.method = "quantile",
  plot = T
)
corfit <- duplicateCorrelation(
  vwts,
  design = design,
  block = clin.sample.event.individ.1b$participant_id
)
print(paste0("Correlation, 1st iteration: ", corfit$consensus))
vwts <- voom(
  counts.1b, 
  design = design,
  normalize.method = "quantile",
  block = clin.sample.event.individ.1b$participant_id,
  correlation = corfit$consensus,
  plot = T
)
corfit <- duplicateCorrelation(
  vwts,
  design = design,
  block = clin.sample.event.individ.1b$participant_id
)
print(paste0("Correlation, 2nd iteration: ", corfit$consensus))

vfit <- lmFit(
  vwts,
  design = design,
  block = clin.sample.event.individ.1b$participant_id,
  correlation = corfit$consensus
)
vfit <- eBayes(vfit)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

```{r}
res.1b <- topTable(
  vfit,
  coef = "Azithro",
  sort.by = "none",
  number = Inf,
  adjust.method = "BH",
  p.value = 1
)

head(res.1b)
```

Count number of significant genes

```{r}
print(sum(res.1b$adj.P.Val<0.05))
```

Add gene symbols, then export the DE result

```{r}
temp <- merge(
  res.1b, rData[,c("gene_id","gene_name")],
  by.x="row.names", by.y="gene_id",
  all.x=TRUE, all.y=FALSE,
  sort=FALSE
)
colnames(temp)[1] <- "gene_id"

write.csv(
  temp,
  "20250320_bulk_RNAseq_PBMC_Day4-6_DE.csv",
  row.names=FALSE
)
```

Volcano plot

```{r}
# Identify direction
res.toplot <- res.1b
res.toplot$direction <- ifelse(
  res.toplot$adj.P.Val>=0.05, "Not significant",
  ifelse(
    res.toplot$logFC>0, "Up", "Down"
  )
)

p <- ggplot(data=res.toplot) +
  geom_point(aes(x=logFC, y=-log10(adj.P.Val), color=direction), size=0.8) +
  scale_color_manual(
    values=c("Up"="darkred","Down"="dodgerblue4","Not significant"="gray75"),
    name="", guide="none") +
  scale_x_continuous(limits=c(-2,2), expand=c(0,0)) +
  labs(x="log2FC", y="-log10(adjusted P)",
       title="PBMC genes associated\nwith Azithromycin at Day 0-2") +
  my.theme
```

Reactome GSEA

```{r}
# Sort gene list in descending order
gene.ranks <- res.1b$t
names(gene.ranks) <- rownames(res.1b)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.1b <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
```

