---
title: "Severity-longitudinal cis-eQTLs: genotype processing and cohort setup"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

This notebook prepares the severity-stratified, longitudinal cis-eQTL cohort and
its genotype covariates. The IMPACC hospitalized COVID-19 cohort is split by
trajectory group into a rapid-recovery cohort (TG1-3, coded "control") and a
prolonged-critical/fatal cohort (TG4-5, coded "case"), and is followed across
nine visits (V1-V4, V6-V10; V5 excluded later in the pipeline). Here we define
that phenotype, write the published GWAS lead-SNP ranges for genotype
extraction, and compute PC-AiR genotype principal components for each
severity-by-visit stratum.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(dplyr)
library(data.table)
library(tidyverse)
library(tidyr)
library(tibble)
library(readr)
library(readxl)
library(magrittr)
library(qs2)
library(SNPRelate)
library(GENESIS)
library(GWASTools)
library(nFactors)
```

```{r paths-and-parameters}
# ---- EDIT THESE PATHS FOR YOUR ENVIRONMENT --------------------------------
# base_dir            : root for all inputs and outputs of this track.
# data_env_path       : the IMPACC acute-phase data environment (RDS), holding
#                       clinical_data, pbmc_transcriptomics_counts / _metadata /
#                       _rowfeature.
# admixture_path      : per-sample ancestry/admixture table linking
#                       participant_id to the genotype core_specific_id (Sample).
# published_snps_path : Excel workbook of published critical COVID-19 GWAS hits.
# results_dir         : where range lists and covariate objects are written.
# All other locations are built from these with file.path().
base_dir            <- path.expand("/path/to/eqtl_data")
data_env_path       <- file.path(base_dir, "expression_data",
                                 "impacc_full-cohort_acute_data_env_030625.RDS")
admixture_path      <- file.path(base_dir, "genotype_data",
                                 "impacc_gwas_cohort_admixture.txt")
published_snps_path <- file.path(base_dir, "tg_impacc_GWAS",
                                 "All_Severe_COVID_GWAS_hits.xlsx")
results_dir         <- file.path(base_dir, "tg_eQTL_with_convalescent")
# ----------------------------------------------------------------------------

genotype_data_dir <- file.path(results_dir, "genotype_data")
covariates_dir    <- file.path(results_dir, "covariates")
published_snps_dir <- file.path(results_dir, "published_gwas_snps")
for (d in c(genotype_data_dir, covariates_dir, published_snps_dir)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

# R 4.3.3
```

## Load the raw PBMC transcriptomics and clinical data

The data environment bundles the clinical table and the PBMC RNA-seq counts,
metadata, and row features used throughout the pipeline.

```{r}
data_env <- readRDS(data_env_path)
```

## Set up the GWAS cohort with trajectory group as phenotype

We take the post-QC GWAS samples, attach the genotype core IDs from the
admixture table, and code the severity phenotype: TG1-3 as "control" and TG4-5
as "case". This mapping is preserved exactly throughout the analysis.

```{r}
clinical_gwas <- data_env$clinical_data %>% # 1255 GWAS samples
    filter(sample_type == "GWAS") %>%
    select("participant_id", "resp_status_v1", "resp_status_v2", "resp_status_v3",
    "respiratory_status_day14", "resp_status_v5", "respiratory_status_day28", "trajectory_group")

# medians: date_v1 = 2 | date_v2 = 5 | date_v3 = 8 | date_v4 = 15 | date_v5 = 22 | date_v6 = 29

gwas_cohort <- clinical_gwas %>% # 1112 distinct GWAS samples
    select(participant_id, trajectory_group) %>%
    distinct()

############### append core_sample_ids and filter to include only post-QC GWAS cohort ###################

# post-QC IMPACC cohort
fam <- fread(admixture_path) %>%
    select(participant_id, Sample) %>%
    rename(core_specific_id = Sample)

final_gwas_cohort <- gwas_cohort %>% # 1031 samples
    inner_join(fam, by = "participant_id") %>%
    relocate(core_specific_id, .before = "trajectory_group") %>%
    mutate(pheno = case_when(
        trajectory_group %in% 1:3 ~ "control",
        trajectory_group %in% 4:5 ~ "case",
        TRUE ~ as.character(trajectory_group)
    ))

# table(final_gwas_cohort$trajectory_group)
#   1   2   3   4   5
#  204 230 303 195  99
# case = 294
# control = 737

qs_save(final_gwas_cohort, file.path(genotype_data_dir, "impacc_tg_pheno.qs2"))

###### LOAD HERE
gwas_pheno <- qs_read(file.path(genotype_data_dir, "impacc_tg_pheno.qs2"))
```

## Non-PC covariates for MatrixEQTL

These are the fixed demographic covariates (age, sex) joined to the post-QC
cohort. Sex is coded 0 = Female, 1 = Male.

```{r}
### For Covariate Use Only (matrixeQTL)

nonPC_covars <- data_env$clinical_data %>% # 1255 GWAS samples
    filter(sample_type == "GWAS") %>%
    select("participant_id", "trajectory_group", "admit_age", "sex") %>%
    arrange(participant_id) %>%
    mutate(
    # Convert sex to a numeric variable where 0 = Female and 1 = Male.
    sex = ifelse(sex == "Male", 1, 0)) %>%
    distinct()

############### append core_sample_ids and filter to include only post-QC GWAS cohort ###################

# post-QC IMPACC cohort
fam <- fread(admixture_path) %>%
    select(participant_id, Sample) %>%
    rename(core_specific_id = Sample)

nonPC_covars_final <- nonPC_covars %>% # 1031 samples
    inner_join(fam, by = "participant_id") %>%
    relocate(core_specific_id, .after = "participant_id")

qs_save(nonPC_covars_final, file.path(covariates_dir, "final_nonPC_covars.qs2"))
```

## Write the published GWAS lead-SNP ranges for genotype extraction

We read the published critical COVID-19 GWAS hits (GenOMICC PMID 37198478 and
HGI PMID 37674002), write a single-position SNP list and a +/- 1 Mb range list,
and record which of the 69 SNPs are not present in the IMPACC genotype set.
The PLINK extractions themselves run in `00-extract_gwas_loci.sh`, which reads
`snps_to_extract.txt` and `ranges_to_extract.txt` written below. This does not
need to be re-run for every analysis.

```{r}
published_snps <- read_excel(published_snps_path)

snps_to_extract <- published_snps %>%
        pivot_longer(cols = everything(),
                    values_to = "chr:pos") %>%
        filter(!is.na(`chr:pos`)) %>%
        distinct(`chr:pos`) %>%
        separate(`chr:pos`, into = c("chr", "pos"), sep = ":") %>%
        mutate(pos = as.numeric(pos),
                start = pos,
                end = pos) %>%
        arrange(chr, start) %>%
        mutate(ranges = paste0("R", 1:n())) %>%
        select(chr, start, end, ranges)

# 69 published, non-overlapping critical COVID-19 GWAS SNPs
write.table(snps_to_extract,
            file = file.path(results_dir, "snps_to_extract.txt"),
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE,
            sep = "\t")

# After running the single-position extraction in 00-extract_gwas_loci.sh
# (extracted_snps.bim), 57 of the 69 SNPs are found in our data. Identify the
# 12 SNPs that are not matched in our IMPACC cohort.
impacc_extracted_snps <- fread(file.path(genotype_data_dir, "extracted_snps.bim")) %>%
    select(V1, V4) %>%
    unite(`grch38`, "V1":"V4", sep = ":")

reported_snps <- published_snps %>%
        pivot_longer(cols = everything(),
                    values_to = "grch38") %>%
        filter(!is.na(`grch38`)) %>%
        distinct(`grch38`)

missing_SNPS <- anti_join(reported_snps, impacc_extracted_snps, by = "grch38")

write.table(missing_SNPS,
            file = file.path(genotype_data_dir, "twelve_SNPs_not_in_IMPACC_cohort.txt"),
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE,
            sep = "\t")

# Next, expand the extraction range to include 2 Mb (+/- 1 Mb) around all SNPs.
ranges_to_extract <- published_snps %>%
        pivot_longer(cols = everything(),
                    values_to = "chr:pos") %>%
        filter(!is.na(`chr:pos`)) %>%
        distinct(`chr:pos`) %>%
        separate(`chr:pos`, into = c("chr", "pos"), sep = ":") %>%
        mutate(pos = as.numeric(pos),
                start = pos - 1000000,
                end = pos + 1000000) %>%
        arrange(chr, start) %>%
        mutate(ranges = paste0("R", 1:n())) %>%
        select(chr, start, end, ranges)

write.table(ranges_to_extract,
            file = file.path(results_dir, "ranges_to_extract.txt"),
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE,
            sep = "\t")
# 299,380 variants extracted in those 69 ranges.
```

## Genotype principal components per severity-by-visit stratum (PC-AiR)

For each visit and severity group we run PC-AiR in GENESIS to derive ancestry
principal components that are robust to cryptic relatedness and population
structure. The per-stratum PLINK binary sets (`*_pca.bed/.bim/.fam`) are
produced by the per-stratum extraction loop in `00-extract_gwas_loci.sh`, which
runs after `02-expression_processing.Rmd` has written each stratum's sample
keep-list. PC-AiR uses LD-pruned, KING-derived kinship, and nScree (optimal
coordinates) selects the number of PCs to retain per stratum.

```{r}
genotype_base_dir <- genotype_data_dir

# We'll store our PC-AiR results in a nested list: visit -> group (case/control)
pcair_results <- list()

# List the visit folders (e.g., "visit_1", "visit_2", ..., "visit_10")
visit_dirs <- list.dirs(genotype_base_dir, recursive = FALSE, full.names = TRUE)

for (visit_dir in visit_dirs) {

  visit_name <- basename(visit_dir)  # e.g., "visit_1"
  # Create an entry for this visit
  pcair_results[[visit_name]] <- list()

  # List the group folders within the visit folder (e.g., "case", "control")
  group_dirs <- list.dirs(visit_dir, recursive = FALSE, full.names = TRUE)

  for (group_dir in group_dirs) {

    group_name <- basename(group_dir)  # e.g., "case" or "control"

    # Construct the file paths for the PLINK files.
    # The PLINK PCA files follow the pattern:
    #   <visit>_<group>_pca.bed, .fam, .bim
    bed.fn <- file.path(group_dir, paste0(visit_name, "_", group_name, "_pca.bed"))
    fam.fn <- file.path(group_dir, paste0(visit_name, "_", group_name, "_pca.fam"))
    bim.fn <- file.path(group_dir, paste0(visit_name, "_", group_name, "_pca.bim"))
    out.gdsfn <- file.path(group_dir, paste0(visit_name, "_", group_name, "_pca.gds"))

    # Convert PLINK files to GDS.
    # For speed, set compress.annotation = "NONE" (lowest compression) and leave compress.geno empty.
    snpgdsBED2GDS(bed.fn, fam.fn, bim.fn, out.gdsfn,
                  family = FALSE, snpfirstdim = NA,
                  compress.annotation = "",
                  compress.geno = "",
                  option = NULL,
                  cvt.chr = c("int", "char"),
                  cvt.snpid = c("auto", "int"),
                  verbose = TRUE)

    # Open the GDS file
    gds_obj <- snpgdsOpen(out.gdsfn, readonly = TRUE)

    # LD-pruning: get a pruned set of SNPs to use in PCA (reduces redundancy)
    set.seed(1234)
    ld_pruned <- snpgdsLDpruning(gds_obj, ld.threshold = 0.1, missing.rate = 0.015,
                                 maf = 0.025, method = "r", num.thread = 46)
    pruned <- unlist(ld_pruned, use.names = FALSE)

    # Read sample IDs from the GDS file
    sample.id <- read.gdsn(index.gdsn(gds_obj, "sample.id"))

    # Compute pairwise relatedness using the KING method (robust to stratification)
    ibd.robust <- snpgdsIBDKING(gds_obj, sample.id = sample.id, family.id = sample.id, num.thread = 46)

    # Convert the KING output to a kinship matrix
    kingmat <- kingToMatrix(ibd.robust, sample.include = NULL, thresh = 2^(-11/2), verbose = TRUE)

    # Close the GDS file to free up resources
    snpgdsClose(gds_obj)

    # Re-open the GDS file as a GenotypeData object
    geno <- GdsGenotypeReader(filename = out.gdsfn) %>% GenotypeData(.)

    # Run PC-AiR using the kinship matrix and pruned SNPs
    mypcair <- pcair(geno, kinobj = kingmat, divobj = kingmat, snp.include = pruned, num.cores = 46)

    # Determine how many PCs were computed:
    nComputed <- ncol(mypcair$vectors)
    nToExtract <- min(32, nComputed)

    geno_mypcs <- data.frame(mypcair$vectors[, 1:nToExtract])
    eig <- mypcair$values[1:nToExtract]

    # Run nScree to get a data-driven recommendation for the number of PCs
    scree_out <- nScree(eig)
    # We'll use the "optimal coordinates" (noc) as our recommendation.
    num_pcs <- scree_out$Components$noc
    message(paste("For", visit_name, group_name, "nScree recommends", num_pcs, "PCs."))

    # Extract the recommended number of PCs from the PC-AiR output.
    # (Here, we extract columns 1 to num_pcs from the geno_mypcs data frame.)
    geno_covariates <- geno_mypcs[, 1:num_pcs, drop = FALSE] %>%
      as.data.frame() %>%
      setNames(paste0("geno_pc", 1:num_pcs)) %>%
      t() %>% as.data.frame()

    # Save the result in our nested list:
    pcair_results[[visit_name]][[group_name]] <- list(
      numOfPCsChosen = num_pcs,
      pcs = geno_covariates
    )

    # Clean up objects to free memory and force garbage collection
    rm(gds_obj, geno, mypcair, geno_mypcs, ld_pruned, pruned, kingmat, ibd.robust)
    gc()

    # Remove the temporary .gds file now that covariates have been extracted
    if (file.exists(out.gdsfn)) {
      file.remove(out.gdsfn)
    }

  } # end group loop

} # end visit loop

# Now, pcair_results is a nested list structured as:
# pcair_results$visit_1$case, pcair_results$visit_1$control, ... up to visit_10.
# Each element contains:
#   numOfPCsChosen: the number of PCs recommended by nScree,
#   pcs: the covariate matrix (with PCs as rows) for MatrixeQTL.

qs_save(pcair_results, file.path(covariates_dir, "final_genotype_pcs.qs2"))
```
