---
title: "Severity-longitudinal cis-eQTLs: MatrixEQTL mapping across strata"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

This notebook assembles the final MatrixEQTL input files for each
severity-by-visit stratum and runs the cis-eQTL mapping. We first merge the
genotype PCs, expression PCs, and demographic covariates into a single
covariate matrix per stratum, then enforce that genes, SNPs, and samples are in
identical order across the expression, genotype, and covariate files. Finally
we run Matrix_eQTL_main with an additive linear model across all strata
(visit_5 control is skipped, as that stratum is too small).

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(MatrixEQTL)
library(data.table)
library(dplyr)
library(tibble)
library(qs2)
```

```{r paths-and-parameters}
# ---- EDIT THESE PATHS FOR YOUR ENVIRONMENT --------------------------------
# base_dir    : root for all inputs and outputs of this track.
# results_dir : holds expression_data/, genotype_data/, covariates/, and is
#               where final_files/ and results/ are written.
base_dir    <- path.expand("/path/to/eqtl_data")
results_dir <- file.path(base_dir, "tg_eQTL_with_convalescent")
# ----------------------------------------------------------------------------

# All MatrixEQTL strata: 18 severity-by-visit combinations (visit_5 control
# excluded). Severity groups: case = TG4-5, control = TG1-3.
visits <- c("visit_1", "visit_2", "visit_3", "visit_4", "visit_5",
            "visit_6", "visit_7", "visit_8", "visit_9", "visit_10")
groups <- c("case", "control")
```

## Combine the genotype, expression, and demographic covariates per stratum

For each stratum we transpose and merge the genotype PCs and expression PCs on
the sample ID, add the demographic covariates (age, sex), and write a single
`_Covariates.txt` file.

```{r}
###### LOAD HERE
geno_covars_list <- qs_read(file.path(results_dir, "covariates", "final_genotype_pcs.qs2"))
rna_covars_list <- qs_read(file.path(results_dir, "covariates", "final_expression_pcs.qs2"))

nonPC_covars_with_dummies <- qs_read(file.path(results_dir, "covariates", "final_nonPC_covars.qs2")) %>%
  select(-participant_id, -trajectory_group)

# Define the output directory for the combined covariates
combined_dir <- file.path(results_dir, "covariates")

# Define visits and groups.
# For rna_covars_list, keys are "Visit 1", etc.
# For geno_covars_list, keys are "visit_1", etc.
visits_rna <- c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6", "Visit 7", "Visit 8", "Visit 9", "Visit 10")
visits_geno <- c("visit_1", "visit_2", "visit_3", "visit_4", "visit_5", "visit_6", "visit_7", "visit_8", "visit_9", "visit_10")

# Initialize a nested list to store combined covariates if needed
combined_covars <- list()

for(i in seq_along(visits_rna)){
        rna_key <- visits_rna[i]
        geno_key <- visits_geno[i]

        # Create an output folder for this visit under the combined directory
        visit_out_dir <- file.path(combined_dir, geno_key)
        if (!dir.exists(visit_out_dir)) {
            dir.create(visit_out_dir, recursive = TRUE)
        }

        # Create separate subdirectories for "case" and "control"
        for(group in groups){
            group_out_dir <- file.path(visit_out_dir, group)
            if (!dir.exists(group_out_dir)) {
            dir.create(group_out_dir, recursive = TRUE)
            }
        }

        combined_covars[[geno_key]] <- list()

        for(group in groups){
            # Try to extract data for this visit/group - use tryCatch to handle potential missing combinations
            tryCatch({
                # Extract the RNA covariates for this visit and group.
                # The covariates are stored with PCs as rows. We transpose to have sample IDs as rows.
                rna_df <- rna_covars_list[[rna_key]][[group]]$pcs %>%
                t() %>%
                as.data.frame() %>%
                rownames_to_column(var = "core_specific_id")

                # Extract the genotype covariates for this visit and group.
                geno_df <- geno_covars_list[[geno_key]][[group]]$pcs %>%
                t() %>%
                as.data.frame() %>%
                rownames_to_column(var = "core_specific_id")

                # Merge the two dataframes by "core_specific_id" to ensure matching samples.
                # This inner join ensures only samples present in both datasets are retained.
                merged_df <- inner_join(geno_df, rna_df, by = "core_specific_id")

                # Add the demographic covariates by merging with nonPC_covars_with_dummies
                # Inner join to ensure we only keep samples with all necessary data
                merged_df_with_demo <- merged_df %>%
                  inner_join(nonPC_covars_with_dummies, by = "core_specific_id")

                # Set the sample IDs as row names and remove the helper column.
                merged_df_final <- merged_df_with_demo %>%
                column_to_rownames(var = "core_specific_id") %>%
                t()

                # Create an output file name: e.g., "visit_1_case_Covariates.txt"
                # Use the group subdirectory for output.
                group_out_dir <- file.path(visit_out_dir, group)
                out_file <- file.path(group_out_dir, paste0(geno_key, "_", group, "_Covariates.txt"))

                # Write the merged dataframe to file
                write.table(merged_df_final, file = out_file, sep = "\t", quote = FALSE,
                            row.names = TRUE, col.names = NA)

                # Store in the nested list (optional)
                combined_covars[[geno_key]][[group]] <- merged_df_final

                # Print success message
                message(paste("Successfully processed:", geno_key, "-", group))

            }, error = function(e) {
                # Handle the case where data might not exist for this visit/group combination
                message(paste("Warning: Unable to process data for", geno_key, "-", group))
                message(paste("Error message:", e$message))
            })
        }
}

# Save the nested list of combined covariates as an qs2 file
qs_save(combined_covars, file.path(combined_dir, "combined_covariates.qs2"))
```

## Column-ordering checks (CRITICAL: sample IDs must be in the same exact order)

MatrixEQTL requires the expression, genotype, and covariate matrices to share an
identical column (sample) order, and the location files to match the matrix
rows. This step reorders all five files per stratum and writes them to
`final_files/`.

```{r}
# Define directories for the original files:
ge_dir    <- file.path(results_dir, "expression_data")    # contains _GE.txt and _geneloc.txt files
snp_dir   <- file.path(results_dir, "genotype_data")        # contains _SNP.txt and _snploc.txt files
covar_dir <- file.path(results_dir, "covariates")           # contains _Covariates.txt files

# Define the output directory for the final, reordered files.
final_dir <- file.path(results_dir, "final_files")
if (!dir.exists(final_dir)) {
  dir.create(final_dir, recursive = TRUE)
}

# Loop over visits and groups to process and write new files.
for(visit in visits) {
  for(gr in groups) {
    # Construct input file paths using the naming convention.
    ge_file      <- file.path(ge_dir, visit, gr, paste0(visit, "_", gr, "_GE.txt"))
    geneloc_file <- file.path(ge_dir, visit, gr, paste0(visit, "_", gr, "_geneloc.txt"))
    snp_file     <- file.path(snp_dir, visit, gr, paste0(visit, "_", gr, "_SNP.txt"))
    snploc_file  <- file.path(snp_dir, visit, gr, paste0(visit, "_", gr, "_snploc.txt"))
    covar_file   <- file.path(covar_dir, visit, gr, paste0(visit, "_", gr, "_Covariates.txt"))

    message("Processing ", visit, " ", gr)

    # Read input files.
    ge      <- fread(ge_file) %>% column_to_rownames(var = "V1")
    geneloc <- fread(geneloc_file)
    snp     <- fread(snp_file) %>% column_to_rownames(var = "V1")
    snploc  <- fread(snploc_file)
    covar   <- fread(covar_file) %>% column_to_rownames(var = "V1")

    # ----- 1. Ensure Gene Order Matches -----
    if(!(all(geneloc$geneid %in% rownames(ge)) && all(rownames(ge) %in% geneloc$geneid))) {
      stop("Gene sets do not match for ", visit, " ", gr)
    }
    ge <- ge[match(geneloc$geneid, rownames(ge)), ]

    # ----- 2. Ensure SNP Order Matches -----
    if(!(all(snploc$snp %in% rownames(snp)) && all(rownames(snp) %in% snploc$snp))) {
      stop("SNP sets do not match for ", visit, " ", gr)
    }
    snp <- snp[match(snploc$snp, rownames(snp)), ]

    # ----- 3. Ensure Sample (Column) Order Matches -----
    common_samples <- Reduce(intersect, list(colnames(ge), colnames(snp), colnames(covar)))
    if(length(common_samples) == 0) {
      stop("No common sample IDs found for ", visit, " ", gr)
    }
    ge    <- ge[, common_samples, drop = FALSE]
    snp   <- snp[, common_samples, drop = FALSE]
    covar <- covar[, common_samples, drop = FALSE]

    if(!(identical(colnames(ge), colnames(snp)) && identical(colnames(ge), colnames(covar)))) {
      stop("Sample order mismatch for ", visit, " ", gr)
    }

    # ----- 4. Write the Corrected Files to the final_files directory -----
    # Create output subdirectory for this visit and group.
    final_out_dir <- file.path(final_dir, visit, gr)
    if (!dir.exists(final_out_dir)) {
      dir.create(final_out_dir, recursive = TRUE)
    }

    # Construct output file paths (preserve naming convention).
    out_ge_file      <- file.path(final_out_dir, paste0(visit, "_", gr, "_GE.txt"))
    out_snp_file     <- file.path(final_out_dir, paste0(visit, "_", gr, "_SNP.txt"))
    out_covar_file   <- file.path(final_out_dir, paste0(visit, "_", gr, "_Covariates.txt"))
    out_geneloc_file <- file.path(final_out_dir, paste0(visit, "_", gr, "_geneloc.txt"))
    out_snploc_file  <- file.path(final_out_dir, paste0(visit, "_", gr, "_snploc.txt"))

    # Write files using fwrite for speed.
    # fwrite by default writes column names and does not output rownames.
    fwrite(ge %>% rownames_to_column(var = "ID"), out_ge_file, sep = "\t", quote = FALSE)
    fwrite(snp %>% rownames_to_column(var = "ID"), out_snp_file, sep = "\t", quote = FALSE)
    fwrite(covar %>% rownames_to_column(var = "ID"), out_covar_file, sep = "\t", quote = FALSE)
    fwrite(geneloc, out_geneloc_file, sep = "\t", quote = FALSE)
    fwrite(snploc, out_snploc_file, sep = "\t", quote = FALSE)

    message("Finished processing ", visit, " ", gr)
  }
}
```

## MatrixEQTL run (loop across all visit-by-severity strata)

We map cis-eQTLs within a 1 Mb window using an additive linear model, saving
all cis associations (p-value threshold 1) per stratum.

```{r}
# Define the base directory for final files (the output from the previous step)
final_dir <- file.path(results_dir, "final_files")
# Define the output directory for Matrix eQTL results
matrix_results_dir <- file.path(results_dir, "results")
if (!dir.exists(matrix_results_dir)) {
  dir.create(matrix_results_dir, recursive = TRUE)
}

# Settings for Matrix eQTL
useModel <- modelLINEAR  # using linear model
pvOutputThreshold_cis <- 1
cisDist <- 1e6
errorCovariance <- numeric()

# Loop over each visit and each group
for (visit in visits) {
  for (group in groups) {

    # Skip visit_5 control only
    if(visit == "visit_5" && group == "control") {
      message("Skipping ", visit, " ", group)
      next
    }

    message("Running Matrix eQTL for ", visit, " ", group)

    # Construct file paths for the five necessary files.
    SNP_file_name <- file.path(final_dir, visit, group, paste0(visit, "_", group, "_SNP.txt"))
    snps_location_file_name <- file.path(final_dir, visit, group, paste0(visit, "_", group, "_snploc.txt"))
    expression_file_name <- file.path(final_dir, visit, group, paste0(visit, "_", group, "_GE.txt"))
    gene_location_file_name <- file.path(final_dir, visit, group, paste0(visit, "_", group, "_geneloc.txt"))
    covariates_file_name <- file.path(final_dir, visit, group, paste0(visit, "_", group, "_Covariates.txt"))

    # Create SlicedData objects for SNP, gene expression, and covariates.
    snps <- SlicedData$new()
    snps$fileDelimiter <- "\t"
    snps$fileOmitCharacters <- "NA"
    snps$fileSkipRows <- 1
    snps$fileSkipColumns <- 1
    snps$fileSliceSize <- 2000
    snps$LoadFile(SNP_file_name)

    gene <- SlicedData$new()
    gene$fileDelimiter <- "\t"
    gene$fileOmitCharacters <- "NA"
    gene$fileSkipRows <- 1
    gene$fileSkipColumns <- 1
    gene$fileSliceSize <- 2000
    gene$LoadFile(expression_file_name)

    cvrt <- SlicedData$new()
    cvrt$fileDelimiter <- "\t"
    cvrt$fileOmitCharacters <- "NA"
    cvrt$fileSkipRows <- 1
    cvrt$fileSkipColumns <- 1
    cvrt$fileSliceSize <- 2000
    if (file.exists(covariates_file_name)) {
      cvrt$LoadFile(covariates_file_name)
    }

    snpspos <- read.table(snps_location_file_name, header = TRUE, stringsAsFactors = FALSE)
    genepos <- read.table(gene_location_file_name, header = TRUE, stringsAsFactors = FALSE)

    # Create temporary file names for cis and trans outputs.
    output_file_name_cis <- tempfile()
    output_file_name_tra <- tempfile()

    # Run Matrix_eQTL_main.
    me <- Matrix_eQTL_main(
      snps = snps,
      gene = gene,
      cvrt = cvrt,
      output_file_name = output_file_name_tra,
      pvOutputThreshold = 0,
      useModel = useModel,
      errorCovariance = errorCovariance,
      verbose = TRUE,
      output_file_name.cis = output_file_name_cis,
      pvOutputThreshold.cis = pvOutputThreshold_cis,
      snpspos = snpspos,
      genepos = genepos,
      cisDist = cisDist,
      pvalue.hist = "qqplot",
      min.pv.by.genesnp = FALSE,
      noFDRsaveMemory = FALSE
    )

    # Remove the trans output file.
    unlink(output_file_name_tra)

    # Directly extract the cis-eQTL results from the me object.
    cis_eqtls <- me$cis$eqtls

    # Create an output folder for this visit and group in the results directory.
    out_dir <- file.path(matrix_results_dir, visit, group)
    if (!dir.exists(out_dir)) {
      dir.create(out_dir, recursive = TRUE)
    }

    # Construct the output file name: e.g., "visit_1_case_cis_eQTL.txt"
    out_file <- file.path(out_dir, paste0(visit, "_", group, "_cis_eQTL.txt"))
    write.table(cis_eqtls, file = out_file, sep = "\t", quote = FALSE,
                row.names = FALSE, col.names = TRUE)

    message("Finished Matrix eQTL for ", visit, " ", group, " in ", me$time.in.sec, " seconds")

    # Clean up temporary cis output file.
    unlink(output_file_name_cis)

    rm(snps, gene, cvrt, snpspos, genepos, me, cis_eqtls)
    gc()
  }
}
```
