---
title: "Severity-longitudinal cis-eQTLs: colocalization with GenOMICC"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

This notebook tests whether the severity-stratified cis-eQTLs colocalize with
the GenOMICC international critical COVID-19 GWAS using coloc.abf. We define
+/- 500 kb windows around the significant lead SNPs, compute per-stratum
allele frequencies (PLINK), harmonize allele orientation between the eQTL and
GWAS summary statistics, and run coloc.abf per SNP-gene pair across every
stratum where the eQTL is significant (eigenMT < 0.05). The set of significant
SNP-gene pairs and the residual-joined summary statistics
(`signif_eqtls_final_tg.qs2`) are produced upstream in
`06-regulatory_pattern_analysis.Rmd`.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(coloc)
library(data.table)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(qs2)
```

```{r paths-and-parameters}
# ---- EDIT THESE PATHS FOR YOUR ENVIRONMENT --------------------------------
# base_dir            : root for all inputs and outputs of this track.
# results_dir         : holds tg_sumstats/, coloc/, genotype_data/, etc.
# gwas_sumstats_path  : GenOMICC international meta-analysis (release 3,
#                       critical) summary statistics, downloadable from
#                       https://datashare.ed.ac.uk/handle/10283/8536 .
base_dir           <- path.expand("/path/to/eqtl_data")
results_dir        <- file.path(base_dir, "tg_eQTL_with_convalescent")
gwas_sumstats_path <- file.path(base_dir, "GenOMICC_sumstats",
                                "GenOMICC_international_meta_release3_critical.tsv")
# ----------------------------------------------------------------------------

sumstats_dir <- file.path(results_dir, "tg_sumstats")
coloc_dir    <- file.path(results_dir, "coloc")
genotype_data_dir <- file.path(results_dir, "genotype_data")
combined_dir <- file.path(results_dir, "eigenMT_FDR", "eigenMT_combined_final_result")
for (d in c(coloc_dir, file.path(coloc_dir, "snps_for_coloc"))) if (!dir.exists(d)) dir.create(d, recursive = TRUE)

# 41 unique SNPs and 73 unique genes feed the colocalization.
all_sig_eqtl_residuals_full <- qs_read(file = file.path(sumstats_dir, "signif_eqtls_final_tg.qs2"))
```

## Define the +/- 500 kb colocalization windows

We take the distinct significant lead SNPs and write a PLINK range file
spanning +/- 500 kb around each, used to extract per-stratum allele frequencies.

```{r}
ranges_for_coloc <- all_sig_eqtl_residuals_full %>%
  select(grch38) %>%
  distinct() %>%
  separate(grch38, into = c("chr", "pos"), sep = ":", remove = TRUE) %>%
  mutate(pos = as.numeric(pos),
  start = pos - 500000,
  end = pos + 500000) %>%
  mutate(ranges = paste0("R", 1:n())) %>%
  select(chr, start, end, ranges)

write.table(ranges_for_coloc,
            file = file.path(genotype_data_dir, "ranges_for_coloc.txt"),
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE,
            sep = "\t")
```

Compute per-stratum allele frequencies now: the PLINK `--freq` extraction over
these coloc windows is performed by the coloc allele-frequency loop in
`00-extract_gwas_loci.sh`, which reads `ranges_for_coloc.txt` written above and
writes one `.frq`/`.bim` per stratum under `coloc/visit_*/{case,control}/`.

## Assemble per-stratum allele frequencies and the eQTL coloc inputs

We read all per-stratum `.frq` and `.bim` files, attach GRCh37 and GRCh38 IDs,
join the combined cis-eQTL summary statistics and eigenMT TESTS, and compute the
standard error and varbeta needed by coloc.

```{r}
# Find all .frq files within the coloc directory recursively
frq_files <- list.files(path = coloc_dir, pattern = "\\.frq$", recursive = TRUE, full.names = TRUE)

# Initialize an empty list to store individual data frames
frq_list <- list()

# Load and process each .frq file individually
for (file in frq_files) {

  # Extract 'visit' and 'group' from the file path
  path_parts <- str_split(file, pattern = "/", simplify = TRUE)
  visit <- path_parts[length(path_parts) - 2]  # e.g., "visit_1"
  group <- path_parts[length(path_parts) - 1]  # e.g., "case" or "control"

  # Read .frq file
  frq_df <- fread(file) %>%
    mutate(visit = visit,
           group = group,
           N = NCHROBS / 2) %>%
    rename(grch37 = SNP,
           effect_allele = A1,
           other_allele = A2) %>%
    select(-NCHROBS)

  # Append processed dataframe to the list
  frq_list[[length(frq_list) + 1]] <- frq_df
}

# Combine all frequency data into a single dataframe
impacc_maf_combined <- bind_rows(frq_list)

# Now, read and combine corresponding .bim files for grch38 data
bim_list <- list()

for (file in frq_files) {
  bim_file <- sub("\\.frq$", ".bim", file)

  # Extract 'visit' and 'group' from the file path
  path_parts <- str_split(bim_file, pattern = "/", simplify = TRUE)
  visit <- path_parts[length(path_parts) - 2]  # e.g., "visit_1"
  group <- path_parts[length(path_parts) - 1]  # e.g., "case" or "control"

  bim_df <- fread(bim_file) %>%
    relocate(V4, .after = V1) %>%
    unite("grch38", V1:V4, sep = ":", remove = TRUE) %>%
    select(-V3) %>%
    rename(grch37 = V2,
           effect_allele = V5,
           other_allele = V6) %>%
    mutate(visit = visit,
           group = group)
  bim_list[[length(bim_list) + 1]] <- bim_df
}

impacc_grch38 <- bind_rows(bim_list)

# Merge grch37 and grch38 information
impacc_geno_dat <- inner_join(impacc_maf_combined, impacc_grch38,
                              by = c("grch37", "effect_allele", "other_allele", "visit", "group")) %>%
                              select(-"CHR", -"effect_allele", -"other_allele") %>%
                              relocate(grch38, .after=grch37)

# 124,583 SNPs fall within +/- 500 kb of the lead SNPs (124,583 * 20 strata in
# impacc_geno_dat).

# Load the combined cis-eQTL summary statistics (faster via qs_read).
combined_eqtls_final <- qs_read(file.path(sumstats_dir, "combined_eqtls_tg_final.qs2"))

eigenMT_final <- fread(file.path(combined_dir, "final_all_merged_eigenMT.txt")) %>%
  select(gene, TESTS, visit, group)

# Final preparation for coloc analysis
coloc_snps <- impacc_geno_dat %>%
              inner_join(combined_eqtls_final, by = c("visit", "group", "grch37", "grch38")) %>%
              inner_join(eigenMT_final, by = c("gene", "visit", "group")) %>%
              mutate(eigenMT = pvalue*TESTS) %>%
              relocate(eigenMT, .before="FDR") %>%
              rename(eigenMT_tests = "TESTS") %>%
              mutate(se = abs(beta / statistic),
                     varbeta = (beta / statistic)^2)
# dim(coloc_snps): 92377420 rows by 18 columns

qs_save(coloc_snps, file.path(coloc_dir, "snps_for_coloc", "coloc_snps.qs2"))

### LOAD HERE
coloc_snps <- qs_read(file.path(coloc_dir, "snps_for_coloc", "coloc_snps.qs2"))
```

## Harmonize the GenOMICC GWAS to the eQTL allele orientation

We overlap the GenOMICC critical COVID-19 GWAS with the eQTL SNPs, drop
ambiguous palindromic SNPs (MAF > 0.40), harmonize allele orientation, and flip
the eQTL beta and MAF where the minor allele matches the GWAS alternate allele.

```{r}
unique_snps <- (unique(coloc_snps %>% select("grch38"))) %>%
                separate(grch38, into = c("CHR", "POSb38"), sep = ":", remove = FALSE) %>%
                mutate(CHR = as.integer(CHR),
                POSb38 = as.integer(POSb38))
# 124,583 unique SNPs flanking +/- 500 kb from the 41 lead SNPs.

central_snps_published <- qs_read(file = file.path(sumstats_dir, "signif_eqtls_final_tg.qs2")) %>%
          select(grch38) %>%
          distinct() # 41 SNPs to test (all published critical COVID-19 SNPs).

# GenOMICC critical COVID-19 GWAS summary statistics:
# wget https://datashare.ed.ac.uk/bitstream/handle/10283/8536/GenOMICC_international_meta_release3_critical.tsv.gz
genomicc_sumstats <- fread(gwas_sumstats_path) %>%
                     select(-ID) %>%
                     inner_join(unique_snps, by = c("CHR", "POSb38")) # 74321 overlapping SNPs.

# remove all palindromic SNPs explicitly
setDT(genomicc_sumstats)

# Step 1: Identify palindromic SNPs explicitly
genomicc_sumstats[, palindromic := ((Allele1 == "A" & Allele2 == "T") |
                                    (Allele1 == "T" & Allele2 == "A") |
                                    (Allele1 == "C" & Allele2 == "G") |
                                    (Allele1 == "G" & Allele2 == "C"))]

# Step 2: Mark ambiguous SNPs using GWAS MAF threshold (defensible: >0.40)
ambig_thresh <- 0.40  # Scientifically defensible threshold
genomicc_sumstats[, ambiguous := palindromic & (Freq1 > ambig_thresh & Freq1 < (1 - ambig_thresh))]

cat("Ambiguous SNPs identified for removal:", sum(genomicc_sumstats$ambiguous), "\n")
# Ambiguous SNPs identified for removal: 1379

###################################
## (A) Harmonize the GenOMICC data ##
###################################
harmonized_genomicc_sumstats <- genomicc_sumstats %>%
  filter(ambiguous == FALSE) %>%
  # rename columns to keep consistent naming
  rename(
    allele1_gwas   = Allele1,
    allele2_gwas   = Allele2,
    beta_gwas      = Effect,
    se_gwas        = StdErr,
    pval_gwas      = P.value,
    freq1_gwas     = Freq1,
    n_gwas         = TotalSampleSize
  ) %>%
  mutate(
    varbeta_gwas = se_gwas^2
  )
# 73,058 SNPs left after removing potentially ambiguous SNPs.

########################################
## (B) Join to the impacc eQTL data   ##
########################################
impacc_eqtl <- qs_read(file.path(coloc_dir, "snps_for_coloc", "coloc_snps.qs2"))

joined_df <- impacc_eqtl %>%
  inner_join(harmonized_genomicc_sumstats, by = c("grch38"))
# 59175457 rows by 32 columns.

#######################################
## (C) Harmonize eQTL to GWAS
#######################################
setDT(joined_df)

# Step 1: Compute an initial flip flag based on allele identity.
joined_df[, flip_initial := fifelse(minor == allele1_gwas, FALSE,
                             fifelse(minor == allele2_gwas, TRUE, NA))]

# Remove any rows where we still have NA (mismatches)
harmonized_df <- joined_df[!is.na(flip_initial)]
# 59114208 rows by 32 columns

#######################################
## (D) Flip eQTL statistics to match
##           GWAS allele orientation
#######################################
# Assign harmonized eQTL columns consistently for all rows.
# If flip_initial is TRUE, flip the beta and MAF; if FALSE, keep as is.
harmonized_df[, `:=`(
  beta_eqtl = ifelse(flip_initial, -beta, beta),
  freq_eqtl = ifelse(flip_initial, 1 - MAF, MAF)
)]

# Also, assign the allele columns so that they reflect the GWAS orientation:
harmonized_df[, `:=`(
  effect_allele = ifelse(flip_initial, allele1_gwas, minor),
  other_allele  = ifelse(flip_initial, allele2_gwas, major)
)]

# Remove the original beta, MAF, minor, and major columns if they're no longer needed.
harmonized_df <- harmonized_df[, !c("MAF", "beta", "minor", "major"), with = FALSE]

# Report flipping count: 35,454 flips and 46,324 no-flips.
num_flips <- harmonized_df %>% select(grch38, flip_initial) %>% distinct()

qs_save(harmonized_df, file.path(sumstats_dir, "harmonized_eqtl_gwas_final_tg.qs2"))
```

## Run coloc.abf across strata

For each significant SNP-gene pair we test every stratum where the eQTL is
significant (eigenMT < 0.05), within a +/- 250 kb window. The eQTL is modeled as
quantitative and the GWAS as case-control (s = 23707 / 1075129 from GenOMICC),
with prior p12 = 5.5e-6 (halfway between 1e-6 and 1e-5). Pairs with PP.H4 > 0.5
are treated as colocalized.

```{r}
## LOAD DATA
harmonized_df <- qs_read(file.path(sumstats_dir, "harmonized_eqtl_gwas_final_tg.qs2")) %>%
  filter(visit != "visit_5")

# Convert to data.table and set keys for efficient subsetting.
harmonized_dt <- as.data.table(harmonized_df)
setkey(harmonized_dt, gene, visit, group, CHR, POSb38)
rm(harmonized_df)

# ---------------------------
# STEP 0: Prepare the SNP-Gene pairs dataframe
# ---------------------------
snp_gene_pairs_to_coloc <- qs_read(file = file.path(sumstats_dir, "signif_eqtls_final_tg.qs2")) %>%
          select(grch38, gene, gene_name) %>%
          distinct() %>% # 92 distinct SNP-gene pairs to test for coloc.
          separate(grch38, into = c("chr", "pos"), sep = ":", convert = TRUE)

# -----------------------------------------
# Step 1: Initialize a dataframe to store coloc results.
# We keep all columns from harmonized_df plus PP.H1-PP.H4.
coloc_results_dat <- harmonized_dt[0, ] %>%
  mutate(PP.H1 = numeric(), PP.H2 = numeric(), PP.H3 = numeric(), PP.H4 = numeric())

# -----------------------------------------
# Step 2: Loop over each SNP-gene pair in snp_gene_pairs_to_coloc
# -----------------------------------------
for(i in 1:nrow(snp_gene_pairs_to_coloc)){

  # Extract the lead SNP and gene information from the current row.
  lead_chr    <- as.character(snp_gene_pairs_to_coloc$chr[i])
  lead_pos    <- snp_gene_pairs_to_coloc$pos[i]
  lead_grch38 <- paste0(lead_chr, ":", lead_pos)
  gene_id     <- snp_gene_pairs_to_coloc$gene[i]
  gene_name   <- snp_gene_pairs_to_coloc$gene_name[i]

  # Define cis window boundaries: +/- 250 kb around the lead SNP position.
  win_lower <- lead_pos - 250000
  win_upper <- lead_pos + 250000

  # -----------------------------------------
  # Step 3: Identify all candidate conditions (visit and group) for this SNP-gene pair, ensuring eigenMT < 0.05.
  candidate_conditions <- harmonized_dt %>%
    filter(gene == gene_id,
           grch38 == lead_grch38,
           eigenMT < 0.05)

  # If no condition meets the significance threshold, skip this SNP-gene pair.
  if(nrow(candidate_conditions) == 0){
    message("No significant eQTL (eigenMT < 0.01) found for ", lead_grch38, " - ", gene_name, "; skipping.")
    next
  }

  # -----------------------------------------
  # Step 4: Loop over each candidate condition
  for(j in 1:nrow(candidate_conditions)){
    condition <- candidate_conditions %>% slice(j)
    cond_visit <- condition$visit
    cond_group <- condition$group

    message("For SNP-gene pair ", lead_grch38, " - ", gene_name,
            " processing condition: ", cond_visit, " ", cond_group)

    # -----------------------------------------
    # Step 5: Extract all SNP data within the +/- 250 kb window for this condition and gene.
    dat_subset <- harmonized_dt[
      gene == gene_id &
      visit == cond_visit &
      group == cond_group &
      CHR == lead_chr &
      POSb38 >= win_lower &
      POSb38 <= win_upper
    ]

    # Remove rows with missing varbeta values to avoid check_dataset error.
    dat_subset <- dat_subset %>% filter(!is.na(varbeta))

    # If no rows remain after filtering, skip this condition.
    if(nrow(dat_subset) == 0){
      message("No eQTL SNPs in window after filtering out missing varbeta for ", lead_grch38, " - ", gene_name,
              " in ", cond_visit, " ", cond_group, "; skipping condition.")
      next
    }

    # -----------------------------------------
    # Step 6: Format the eQTL and GWAS datasets for coloc.
    # Format the eQTL dataset (quantitative).
    eqtl_dataset <- list(
      snp = dat_subset$grch38,
      beta = dat_subset$beta_eqtl,        # use the harmonized beta
      varbeta = dat_subset$varbeta,
      pvalues = dat_subset$pvalue,
      MAF = dat_subset$freq_eqtl,         # use the harmonized MAF
      N = dat_subset$N,
      position = dat_subset$POSb38,
      type = "quant"
    )

    # Format the GWAS dataset (case-control).
    gwas_dataset <- list(
      snp = dat_subset$grch38,
      beta = dat_subset$beta_gwas,
      varbeta = dat_subset$varbeta_gwas,
      pvalues = dat_subset$pval_gwas,
      MAF = dat_subset$freq1_gwas,
      N = dat_subset$n_gwas,
      type = "cc",
      s = (23707 / 1075129),  # defined by the case-control numbers from GenoMICC.
      position = dat_subset$POSb38
    )

    # Run diagnostic checks (optional)
    check_dataset(eqtl_dataset)
    check_dataset(gwas_dataset)

    # -----------------------------------------
    # Step 7: Run coloc.abf using the formatted datasets.
    coloc_res <- coloc.abf(
      dataset1 = eqtl_dataset,
      dataset2 = gwas_dataset,
      p12=5.5e-6) # halfway in between 1e-6 and 1e-5

    # Extract posterior probabilities.
    pp_h1 <- coloc_res$summary["PP.H1.abf"]
    pp_h2 <- coloc_res$summary["PP.H2.abf"]
    pp_h3 <- coloc_res$summary["PP.H3.abf"]
    pp_h4 <- coloc_res$summary["PP.H4.abf"]

    # -----------------------------------------
    # Step 8: Record the results.
    condition_out <- condition %>%
      mutate(PP.H1 = as.numeric(pp_h1),
             PP.H2 = as.numeric(pp_h2),
             PP.H3 = as.numeric(pp_h3),
             PP.H4 = as.numeric(pp_h4))

    # Append the row to the final results dataframe.
    coloc_results_dat <- bind_rows(coloc_results_dat, condition_out)

  } # end inner loop over candidate conditions
} # end outer loop over SNP-gene pairs

# coloc_results_dat now contains one row per candidate condition (visit-group)
# where eigenMT < 0.05 for a SNP-gene pair, with coloc PP.H1-PP.H4.
# With PP.H4 > 0.5: 21 unique SNPs and 24 unique genes.

qs_save(coloc_results_dat, file.path(sumstats_dir, "coloc_results_dat_tg_05.qs2"))
```
