---
title: "Severity-longitudinal cis-eQTLs: regulatory pattern analysis"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

This notebook classifies the colocalized SNP-gene pairs at the published critical
COVID-19 GWAS loci into four regulatory patterns by severity group (case = TG4-5,
control = TG1-3): constitutively active, TG1-3 only, TG4-5 only, and dynamic. It
consumes the significant-eQTL set produced in 04-eigenmt_correction.Rmd and the
colocalization results from 05-colocalization.Rmd, then saves the classified
per-locus table (fig3_clean_classified.qs2) that the Figure 3 notebook uses to
split the heatmap rows and draw panels 3B and 3C.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(data.table)
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(tibble)
library(stringr)
library(qs2)
```

```{r paths-and-parameters}
# ---- EDIT THESE PATHS FOR YOUR ENVIRONMENT --------------------------------
# base_dir            : root for all inputs and outputs of this track.
# results_dir         : holds final_files/, eigenMT_FDR/, tg_sumstats/, etc.
# published_snps_path : Excel workbook of published critical COVID-19 GWAS hits.
base_dir            <- path.expand("/path/to/eqtl_data")
results_dir         <- file.path(base_dir, "tg_eQTL_with_convalescent")
published_snps_path <- file.path(base_dir, "tg_impacc_GWAS",
                                 "All_Severe_COVID_GWAS_hits.xlsx")
# ----------------------------------------------------------------------------

combined_dir <- file.path(results_dir, "eigenMT_FDR", "eigenMT_combined_final_result")
final_dir    <- file.path(results_dir, "final_files")
sumstats_dir <- file.path(results_dir, "tg_sumstats")
residuals_dir <- file.path(results_dir, "dynamic_eQTL_analysis", "residuals")
for (d in c(sumstats_dir, residuals_dir)) if (!dir.exists(d)) dir.create(d, recursive = TRUE)

```

## Classify colocalized SNP-gene pairs into four regulatory patterns

Using the colocalization results (produced in `05-colocalization.Rmd`), we
restrict to the published-locus eQTLs, flag colocalized significant
associations (PP.H4 > 0.5), and classify each SNP-gene pair by its severity
pattern. The four-pattern classifier counts the number of significant
(eigenMT < 0.05) detections in each severity group and assigns one of:
constitutively active (>= 5 detections in both groups), TG1-3 only, TG4-5 only,
or dynamic. This `detection_pattern` is consumed by the Figure 3 notebook to
split the heatmap rows and to draw the pattern panels (3B, 3C).

```{r}
# Inputs: the full combined cis-eQTL summary stats, eigenMT TESTS, the published
# SNP list, and the coloc results (visit, group, SNP, gene, PP.H4).
eigenmt_dir <- file.path(results_dir, "eigenMT_FDR", "eigenMT_combined_final_result")
eqtl_dir    <- sumstats_dir

published_snps_h <- read_excel(published_snps_path) %>%
        pivot_longer(cols = everything(), values_to = "grch38") %>%
        filter(!is.na(grch38)) %>%
        distinct(grch38) %>%
        bind_rows(tibble(grch38 = "12:112945172"))

combined_eqtls_final <- qs_read(file.path(eqtl_dir, "combined_eqtls_tg_final.qs2"))

eigenMT_final2 <- fread(file.path(eigenmt_dir, "final_all_merged_eigenMT.txt")) %>%
  select(gene, TESTS, visit, group)

coloc_results_dat <- qs_read(file.path(eqtl_dir, "coloc_results_dat_tg_05.qs2")) %>%
  select(visit, group, grch37, grch38, gene, gene_name, `PP.H4`)

eqtls_for_published_snps <- combined_eqtls_final %>%
      inner_join(eigenMT_final2, by = c("gene", "visit", "group")) %>%
      mutate(eigenMT = pvalue*TESTS) %>%
      inner_join(published_snps_h, by = "grch38")

fig3_dat <- left_join(eqtls_for_published_snps, coloc_results_dat,
                     by = c("visit", "group", "grch37", "grch38", "gene", "gene_name"))

# STEP 1: Identify SNP-gene pairs with at least one significant eQTL.
significant_pairs <- fig3_dat %>%
  filter(eigenMT < 0.05,
         as.numeric(gsub("visit_", "", visit)) != 5) %>%
  distinct(grch38, gene_name) %>%
  mutate(pair_id = paste0(gene_name, "_", grch38))

# STEP 2: Get ALL data for those pairs (significant and non-significant).
fig3_clean <- fig3_dat %>%
  filter(as.numeric(gsub("visit_", "", visit)) != 5) %>%  # Exclude visit 5
  mutate(pair_id = paste0(gene_name, "_", grch38)) %>%
  # Keep only pairs that have at least one significant eQTL
  filter(pair_id %in% significant_pairs$pair_id) %>%
  mutate(
    group = ifelse(group == "case", "TG4-5", "TG1-3"),
    visit_num = as.numeric(gsub("visit_", "", visit)),
    col_key = paste0("V", visit_num, "_", group),
    neglog10P = pmin(-log10(eigenMT), 10),
    signed_beta = beta,  # ALL betas, not just significant ones
    abs_beta = abs(beta),
    is_significant = eigenMT < 0.05,
    # Colocalization flag only for significant eQTLs
    coloc_flag = is_significant & !is.na(PP.H4) & PP.H4 > 0.5,
    row_id = paste0(gene_name, " | ", grch38)
  ) %>%
  arrange(gene_name, as.numeric(gsub(".*:", "", grch38))) %>%
  mutate(
    row_id = paste0(gene_name, " | ", grch38),
    gene_group = gene_name
  ) %>%
  mutate(signed_beta_clipped = pmax(pmin(signed_beta, 1), -1))

# Function to classify detection patterns (the four regulatory patterns).
classify_detection_pattern <- function(data) {
  data %>%
    group_by(row_id) %>%
    summarise(
      tg13_detections = sum(is_significant & group == "TG1-3", na.rm = TRUE),
      tg45_detections = sum(is_significant & group == "TG4-5", na.rm = TRUE),
      total_detections = sum(is_significant, na.rm = TRUE),
      .groups = 'drop'
    ) %>%
    mutate(
      detection_pattern = case_when(
        tg13_detections >= 5 & tg45_detections >= 5 ~ "Constitutively Active",
        tg13_detections >= 1 & tg45_detections == 0 ~ "TG1-3 Only",
        tg45_detections >= 1 & tg13_detections == 0 ~ "TG4-5 Only",
        TRUE ~ "Dynamic"
      )
    )
}

# Apply classification
detection_patterns <- classify_detection_pattern(fig3_clean)

# Add detection patterns back to main data
fig3_clean <- fig3_clean %>%
  left_join(detection_patterns %>% select(row_id, detection_pattern), by = "row_id") %>%
  # Reorder by detection pattern groups
  arrange(
    factor(detection_pattern, levels = c("Constitutively Active", "TG1-3 Only", "TG4-5 Only", "Dynamic")),
    gene_name,
    as.numeric(gsub(".*:", "", grch38))
  )

# Save the classified per-locus table for the Figure 3 notebook to consume.
qs_save(fig3_clean, file.path(sumstats_dir, "fig3_clean_classified.qs2"))
qs_save(detection_patterns, file.path(sumstats_dir, "detection_patterns.qs2"))
```
