---
title: "Figure 3: longitudinal severity-stratified cis-eQTL landscape"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

This notebook draws the four panels of Figure 3 from the classified per-locus
table produced in `06-regulatory_pattern_analysis.Rmd`:

- 3A: the paired severity effect-size heatmap across visits V1-V10 (V5 excluded),
  with TG1-3 and TG4-5 column blocks, rows split by regulatory detection pattern,
  and colocalized significant eQTLs marked.
- 3B: the severity-group distribution of the colocalized eQTLs (cohort split).
- 3C: representative genes for each detection pattern.
- 3D: the temporal distribution of significant, colocalized cis-eQTLs by visit
  and severity group.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(qs2)
library(dplyr)
library(tidyr)
library(tibble)
library(data.table)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(gridExtra)
library(cowplot)
library(patchwork)
library(ggplot2)
```

```{r paths-and-parameters}
# ---- EDIT THESE PATHS FOR YOUR ENVIRONMENT --------------------------------
# base_dir    : root for all inputs and outputs of this track.
# results_dir : holds tg_sumstats/ with the classified per-locus table written
#               by 06-regulatory_pattern_analysis.Rmd.
base_dir    <- path.expand("/path/to/eqtl_data")
results_dir <- file.path(base_dir, "tg_eQTL_with_convalescent")
# ----------------------------------------------------------------------------

sumstats_dir <- file.path(results_dir, "tg_sumstats")

# fig3_clean is the per-locus table classified into the four regulatory
# detection patterns (see 06-regulatory_pattern_analysis.Rmd).
fig3_clean <- qs_read(file.path(sumstats_dir, "fig3_clean_classified.qs2"))
```

## Panel 3A: paired severity effect-size heatmap across visits

We build three aligned matrices (signed effect size, significance, and
colocalization flag) keyed by SNP-gene pair and visit-by-severity column, then
draw a ComplexHeatmap with TG1-3 / TG4-5 column blocks, rows split by detection
pattern, single diagonals marking significant eQTLs, and full crosses marking
colocalized eQTLs.

```{r}
# UNIFIED MATRIX CREATION FUNCTION
create_matrix <- function(data, value_col, fill_value = 0) {
  data %>%
    select(row_id, col_key, !!sym(value_col)) %>%
    pivot_wider(names_from = col_key, values_from = !!sym(value_col), values_fill = fill_value) %>%
    column_to_rownames("row_id") %>%
    as.matrix()
}

# Create three matrices
mat_beta <- create_matrix(fig3_clean, "signed_beta_clipped", 0)
mat_significance <- create_matrix(fig3_clean, "is_significant", FALSE)
mat_coloc <- create_matrix(fig3_clean, "coloc_flag", FALSE)

# Column ordering
visits_to_include <- c(1, 2, 3, 4, 6, 7, 8, 9, 10)
col_order <- c(paste0("V", visits_to_include, "_TG1-3"),
               paste0("V", visits_to_include, "_TG4-5"))
col_order <- col_order[col_order %in% colnames(mat_beta)]

# Reorder matrices
mat_beta <- mat_beta[, col_order]
mat_significance <- mat_significance[, col_order]
mat_coloc <- mat_coloc[, col_order]

# Column info for annotations
column_info <- data.frame(
  col_name = colnames(mat_beta),
  stringsAsFactors = FALSE
) %>%
  mutate(
    visit = as.numeric(gsub("V(\\d+)_.*", "\\1", col_name)),
    group = gsub(".*_(TG.*)$", "\\1", col_name),
    phase = ifelse(visit %in% c(1, 2, 3, 4, 6), "Acute", "Convalescent")
  )

# UNIFIED CELL FUNCTION
create_cell_function <- function(beta_mat, coloc_mat, significance_mat, show_arrows = TRUE) {
  function(j, i, x, y, width, height, fill) {
    # Base color for all betas
    if(!is.na(beta_mat[i, j])) {
      grid.rect(x, y, width, height, gp = gpar(fill = fill, col = NA))
    }

    # Light grid for all cells
    grid.rect(x, y, width, height, gp = gpar(col = "grey80", fill = NA, lwd = 2.15))

    # Significance and colocalization marks
    if(!is.na(significance_mat[i, j]) && significance_mat[i, j]) {

      if(!is.na(coloc_mat[i, j]) && coloc_mat[i, j]) {
        # FULL X for colocalized eQTLs (both diagonals) - CORNER TO CORNER
        grid.lines(c(x - width*0.41, x + width*0.41), c(y - height*0.41, y + height*0.41),
                  gp = gpar(col = "grey26", lwd = 1.5))
        grid.lines(c(x - width*0.41, x + width*0.41), c(y + height*0.41, y - height*0.41),
                  gp = gpar(col = "grey26", lwd = 1.5))
      } else {
        # SINGLE DIAGONAL for significant eQTLs only - CORNER TO CORNER
        grid.lines(c(x - width*0.41, x + width*0.41), c(y - height*0.41, y + height*0.41),
                  gp = gpar(col = "grey26", lwd = 1.5))
      }
    }
  }
}

# Heatmap builder. filter_coloc keeps only rows with a colocalization event.
create_heatmap <- function(beta_mat, coloc_mat, significance_mat, fig_data, filter_coloc = FALSE, show_arrows = TRUE) {

    if(filter_coloc) {
        row_has_coloc <- apply(coloc_mat, 1, function(x) any(x, na.rm = TRUE))
        beta_mat <- beta_mat[row_has_coloc, ]
        coloc_mat <- coloc_mat[row_has_coloc, ]
        significance_mat <- significance_mat[row_has_coloc, ]
        # Filter fig_data to match
        fig_data <- fig_data %>% filter(row_id %in% rownames(beta_mat))
        title_suffix <- "with Colocalization"
    } else {
        title_suffix <- ""
    }

    # Get detection patterns for row splitting
    row_patterns <- fig_data %>%
      distinct(row_id, detection_pattern) %>%
      filter(row_id %in% rownames(beta_mat))

    detection_factor <- factor(
      row_patterns$detection_pattern[match(rownames(beta_mat), row_patterns$row_id)],
      levels = c("TG4-5 Only", "TG1-3 Only",  "Dynamic", "Constitutively Active"),  # Original levels
      labels = c("TG4-5 Only (Risk)","TG1-3 Only (Protective)",  "Dynamic", "Constitutively\nActive")   # Line break only in labels
    )

    col_fun <- colorRamp2(c(-1, 0, 1), c("#2C7BB6", "white", "#D7191C"))

    Heatmap(
        beta_mat,
        name = "Effect Size",
        col = col_fun,
        na_col = "white",

        cluster_rows = TRUE,
        clustering_distance_rows = "pearson",
        # clustering_method_rows = "ward.D2",
        show_row_dend = FALSE,
        cluster_columns = FALSE,

        # ROW SPLITTING BY DETECTION PATTERN
        row_split = detection_factor,
        row_gap = unit(3, "mm"),
        row_title_rot = 90,  # Rotate group titles 90 degrees
        row_title_gp = gpar(fontsize = 11, fontface = "bold"),
        cluster_row_slices = FALSE,  # prevents group reordering


        show_row_names = TRUE,
        show_column_names = TRUE,
        row_names_gp = gpar(fontsize = 11, fontface = "italic"),
        column_names_gp = gpar(fontsize = 11, fontface = "bold"),
        column_names_rot = 0,
        column_names_centered = TRUE,
        column_labels = gsub("_TG.*", "", colnames(beta_mat)),

        # Cell function
        cell_fun = create_cell_function(beta_mat, coloc_mat, significance_mat, show_arrows),

        # Column grouping
        column_split = factor(column_info$group, levels = c("TG1-3", "TG4-5")),
        column_gap = unit(12, "mm"),
        column_title = c("TG1-3", "TG4-5"),
        column_title_gp = gpar(fontsize = 14, fontface = "bold"),

        width = unit(200, "mm"),
        height = unit(ifelse(filter_coloc, 242, 428), "mm"),

        heatmap_legend_param = list(
            title = "Effect Size",
            title_gp = gpar(fontsize = 12.5, fontface = "bold"),
            title_position = "topcenter",
            labels_gp = gpar(fontsize = 10),
            direction = "horizontal",
            at = c(-1, -0.5, 0, 0.5, 1),
            labels = c("-1", "-0.5", "0", "0.5", "1"),
            legend_height = unit(3.1, "cm")
        )
    )
}

# Build the full and colocalization-only heatmaps.
ht_full <- create_heatmap(mat_beta, mat_coloc, mat_significance, fig3_clean, filter_coloc = FALSE)
ht_coloc <- create_heatmap(mat_beta, mat_coloc, mat_significance, fig3_clean, filter_coloc = TRUE, show_arrows = FALSE)

# Draw the colocalization-only heatmap (Figure 3A).
draw(ht_coloc, column_title = "COVID-19 GWAS Variant eQTL Effects Across Disease Course and Trajectory Groups",
     column_title_gp = gpar(fontsize = 15.5, fontface = "bold"),
     heatmap_legend_side = "bottom", # Move to bottom
     padding = unit(c(20, 2, 2, 2), "mm"))
```

## Panel 3B: severity-group distribution of colocalized eQTLs

Among the colocalized SNP-gene pairs we classify each as TG1-3 only, TG4-5 only,
or significant in both groups, and show the cohort breakdown as a stacked bar.

```{r}
# First, identify which SNP-gene pairs have at least one colocalization event
colocalized_pairs <- fig3_clean %>%
  filter(coloc_flag == TRUE) %>%  # Only pairs with colocalization evidence
  distinct(row_id) %>%
  pull(row_id)

# Create severity group classification for ONLY the colocalized pairs
severity_distribution <- fig3_clean %>%
  filter(row_id %in% colocalized_pairs,  # Only the colocalized pairs
         is_significant) %>%  # Only significant detections within those pairs
  group_by(row_id) %>%
  summarise(
    groups_present = paste(sort(unique(group)), collapse = "_"),
    .groups = "drop"
  ) %>%
  mutate(
    severity_pattern = case_when(
      groups_present == "TG1-3" ~ "TG1-3 Only",
      groups_present == "TG4-5" ~ "TG4-5 Only",
      groups_present == "TG1-3_TG4-5" ~ "Both Groups",
      TRUE ~ "Other"
    )
  )

# Create summary for visualization
severity_summary <- severity_distribution %>%
  count(severity_pattern, name = "n_pairs") %>%
  mutate(
    prop = n_pairs / sum(n_pairs) * 100,
    total_n = sum(n_pairs),
    group = "All Pairs"
  )

# Reorder factor levels
severity_summary$severity_pattern <- factor(
  severity_summary$severity_pattern,
  levels = c("TG1-3 Only", "Both Groups", "TG4-5 Only")
)

# Create stacked bar chart
fig3d <- ggplot(severity_summary, aes(x = group, y = prop, fill = severity_pattern)) +
  geom_bar(stat = "identity", width = 0.6, color = "black", size = 0.3) +

  geom_text(
    aes(label = sprintf("%.1f%%\n(n=%d)", prop, n_pairs)),
    position = position_stack(vjust = 0.5),
    size = 4,
    fontface = "bold",
    color = "white"
  ) +

  geom_text(
    aes(y = 105, label = paste0("Total: ", total_n, " colocalized pairs")),
    size = 4.5,
    fontface = "bold",
    color = "black"
  ) +

  scale_fill_manual(
    values = c(
      "TG1-3 Only" = "#E69F00",      # Orange
      "Both Groups" = "#56B4E9",     # Sky blue
      "TG4-5 Only" = "#009E73"       # Green
    ),
    name = "Detection Pattern"
  ) +

  scale_y_continuous(
    limits = c(0, 115),
    breaks = seq(0, 100, 25),
    expand = c(0, 0),
    labels = function(x) paste0(x, "%")
  ) +

  scale_x_discrete(labels = "COVID-19 GWAS\nColocalized eQTLs") +

  labs(
    x = NULL,
    y = "Percentage of eQTLs",
    title = "Severity Group Distribution\nof Colocalized eQTLs"
  ) +

  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black", size = 1),
    axis.text = element_text(size = 11, color = "black"),
    axis.text.x = element_text(face = "bold", size = 11),
    axis.title = element_text(face = "bold", size = 12),
    legend.position = "bottom",
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 10),
    plot.title = element_text(face = "bold", size = 14),
    plot.margin = margin(t = 10, r = 15, b = 10, l = 10)
  )

# Display the plot
fig3d
```

## Panel 3C: representative genes by detection pattern

For the colocalized pairs we take the top genes in each detection pattern
(ranked by number of timepoints detected and effect size) and show how many
timepoints each was detected at.

```{r}
# Representative genes from each detection pattern
pattern_genes <- fig3_clean %>%
  filter(row_id %in% colocalized_pairs, is_significant) %>%
  group_by(gene_name, detection_pattern) %>%
  summarise(
    n_timepoints = n_distinct(visit_num),
    mean_effect = mean(abs(signed_beta_clipped), na.rm = TRUE),
    max_effect = max(abs(signed_beta_clipped), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # Get top 2-3 genes per pattern
  group_by(detection_pattern) %>%
  arrange(desc(n_timepoints), desc(max_effect)) %>%
  slice_head(n = 3) %>%
  ungroup() %>%
  mutate(
    gene_name = factor(gene_name, levels = rev(gene_name)),
    pattern_colors = case_when(
      detection_pattern == "Constitutively Active" ~ "#2C7BB6",
      detection_pattern == "TG1-3 Only" ~ "#E69F00",
      detection_pattern == "TG4-5 Only" ~ "#009E73",
      detection_pattern == "Dynamic" ~ "#D7191C"
    )
  )

fig3e <- ggplot(pattern_genes, aes(x = n_timepoints, y = gene_name, fill = pattern_colors)) +
  geom_bar(stat = "identity", alpha = 0.8, color = "black", size = 0.75) +
  geom_text(aes(label = n_timepoints), hjust = -0.2, size = 4, fontface = "bold") +
  scale_fill_identity() +  # Use exact colors
  scale_x_continuous(
    breaks = function(x) seq(0, ceiling(max(x)), by = 1),  # Only whole numbers
    expand = expansion(mult = c(0, 0.1))
  ) +
  facet_wrap(~detection_pattern, scales = "free_y", ncol = 1) +
  labs(
    x = "Number of Timepoints Detected",
    y = NULL,
    title = "Representative Genes by Detection Pattern") +
  theme_minimal(base_size = 15) +
  theme(
    panel.grid = element_blank(),
    axis.line.x = element_line(color = "black", size = 1),
    strip.text = element_text(face = "bold", size = 13),
    strip.background = element_rect(fill = "grey90"),
    axis.text.y = element_text(face = "italic", size = 16),  # MUCH BIGGER gene names
    axis.text.x = element_text(size = 12),  # X-axis numbers
    plot.title = element_text(face = "bold", size = 15)
  )

# Display the plot
fig3e
```

## Panel 3D: temporal distribution of significant colocalized eQTLs

We count the unique colocalized SNP-gene pairs detected at each visit within
each severity group, filling absent visit-by-group combinations with zero, and
show them as dodged bars with an acute / convalescent divider.

```{r}
groups_vec <- c("TG1-3", "TG4-5")
visits_vec <- c(1, 2, 3, 4, 6, 7, 8, 9, 10)  # Explicit visits (no V5)

temporal_detection <- fig3_clean %>%
  filter(row_id %in% colocalized_pairs,  # Only colocalized pairs
         is_significant) %>%  # Only significant detections
  group_by(visit_num, group) %>%
  summarise(n_eqtls = n_distinct(row_id),  # Count unique SNP-gene pairs
            .groups = "drop") %>%
  # Fill missing visit x group combinations
  complete(visit_num = visits_vec,
           group = groups_vec,
           fill = list(n_eqtls = 0)) %>%
  mutate(
    phase = ifelse(visit_num %in% c(1,2,3,4,6), "Acute", "Convalescent"),
    group = factor(group, levels = groups_vec),
    visit_label = paste0("V", visit_num)
  )

# Create the plot
fig3f <- ggplot(temporal_detection,
                   aes(x = factor(visit_num),
                       y = n_eqtls,
                       fill = group)) +
  geom_bar(stat = "identity",
           position = position_dodge(width = 0.7),
           width = 0.7,
           colour = "black", size = 0.3, alpha = 0.85) +
  geom_text(aes(label = n_eqtls),
            position = position_dodge(width = 0.7),
            vjust = -0.35, size = 3.5, fontface = "bold") +

  # Phase divider (between acute V6 and convalescent V7)
  geom_vline(xintercept = 5.5, linetype = "dashed",  # Adjusted for missing V5
             colour = "grey40", size = 0.8) +

  scale_fill_manual(values = c("TG1-3" = "#E69F00",
                               "TG4-5" = "#009E73"),
                    name = "Severity Group") +
  scale_x_discrete(labels = paste0("V", visits_vec)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12)),
                     breaks = function(x) seq(0, ceiling(max(x)), by = 2)) +  # Whole numbers only

  annotate("text", x = 3, y = max(temporal_detection$n_eqtls)*0.92,
           label = "Acute", fontface = "bold", size = 4) +
  annotate("text", x = 7.5, y = max(temporal_detection$n_eqtls)*0.92,
           label = "Convalescent", fontface = "bold", size = 4) +

  labs(x = "Visit",
       y = "Colocalized eQTL Pairs Detected",
       title = "Temporal Activity of Causal eQTLs") +

  theme_minimal(base_size = 11) +
  theme(panel.grid = element_blank(),
        axis.line = element_line(colour = "black", linewidth = 0.9),
        axis.text = element_text(face = "bold", size = 12),
        axis.title = element_text(face = "bold"),
        legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 15))

fig3f
```

## Assemble Figure 3 panels B, C, and D

The heatmap (3A) is drawn on its own above; panels 3B, 3C, and 3D are combined
into the multi-panel layout below.

```{r}
# Create 2x1 grid with custom widths
right_panels <- (fig3d + fig3e + plot_layout(widths = c(1, 2))) /
                (fig3f) +
  theme(plot.tag = element_text(size = 16, face = "bold"))

right_panels
```
