---
title: "Figure 4: representative colocalized loci (TMIGD2, OAS1, HLA-G, ABO)"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

This notebook draws Figure 4 for the four representative colocalized loci
(TMIGD2 at 19:4717660, OAS1 at 12:112945172, HLA-G at 6:29947491, and ABO at
9:133271182):

- 4A: genotype-dosage-stratified, covariate-adjusted residual-expression
  boxplots across visits V1-V10 (V5 excluded), faceted by gene and severity
  group (TG1-3 vs TG4-5), with eigenMT significance stars.
- 4B: loess-smoothed dose-response curves of residual expression over time,
  by genotype within each severity group (and a severity-group comparison
  pooled across genotypes).

The covariate-adjusted residuals are precomputed in
`06-regulatory_pattern_analysis.Rmd` and carried in `signif_eqtls_final_tg.qs2`,
which this notebook loads.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(data.table)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(tidyverse)
library(qs2)
library(ggplot2)
```

```{r paths-and-parameters}
# ---- EDIT THESE PATHS FOR YOUR ENVIRONMENT --------------------------------
# base_dir    : root for all inputs and outputs of this track.
# results_dir : holds tg_sumstats/ with the residual-joined significant-eQTL
#               table and the coloc results.
base_dir    <- path.expand("/path/to/eqtl_data")
results_dir <- file.path(base_dir, "tg_eQTL_with_convalescent")
# ----------------------------------------------------------------------------

sumstats_dir <- file.path(results_dir, "tg_sumstats")

# Combined cis-eQTL summary statistics and colocalization results.
combined_eqtl <- qs_read(file.path(sumstats_dir, "combined_eqtls_tg_final.qs2"))
coloc_results_dat <- qs_read(file.path(sumstats_dir, "coloc_results_dat_tg_05.qs2"))
```

## Build the plotting data for the colocalized pairs

We restrict the residual-joined significant-eQTL table to the colocalized pairs
(PP.H4 > 0.5) and strip the "visit_" prefix for plotting.

```{r}
true_coloc_pairs <- coloc_results_dat %>%
  filter(`PP.H4` > 0.5) %>%
  select(grch38, gene, gene_name) %>%
  distinct()
# 33 pairs (23 SNPs, 26 genes)

box_plot_dat <- qs_read(file = file.path(sumstats_dir, "signif_eqtls_final_tg.qs2")) %>%
  inner_join(true_coloc_pairs, by = c("grch38", "gene", "gene_name")) %>%
  mutate(visit = str_replace(visit, "visit_", ""))
```

## Panel 4A: genotype-stratified residual-expression boxplots

For the four representative loci we draw covariate-adjusted residual expression
across visits, filled by genotype, faceted by gene (rows) and severity group
(columns), with a dashed acute / convalescent divider and eigenMT significance
stars positioned above the whiskers.

```{r}
# The four exemplar SNP-gene pairs.
pairs_to_plot <- tribble(
  ~snp,          ~gene,
  "19:4717660",  "TMIGD2",
  "12:112945172", "OAS1",
  "6:29947491",  "HLA-G",
  "9:133271182", "ABO"
)

# Filter data for exact SNP-gene combinations
combined_dat <- box_plot_dat %>%
  inner_join(pairs_to_plot, by = c("grch38" = "snp", "gene_name" = "gene")) %>%
  filter(
    visit %in% c("1","2","3","4","6","7","8","9","10"),
    !is.na(genotype)
  ) %>%
  mutate(
    genotype = factor(genotype, levels = c("0","1","2")),
    visit = factor(visit, levels = c("1","2","3","4","6","7","8","9","10"), ordered = TRUE),
    group = factor(case_when(
      group == "control" ~ "TG1-3",
      group == "case" ~ "TG4-5"
    ), levels = c("TG1-3", "TG4-5")),
    gene_label = case_when(
      gene_name == "TMIGD2" & grch38 == "19:4717660" ~ "TMIGD2 (19:4717660)",
      gene_name == "OAS1" & grch38 == "12:112945172" ~ "OAS1 (12:112945172)",
      gene_name == "HLA-G" & grch38 == "6:29947491" ~ "HLA-G (6:29947491)",
      gene_name == "ABO" & grch38 == "9:133271182" ~ "ABO (9:133271182)"
    ),
    gene_label = factor(gene_label, levels = c(
      "TMIGD2 (19:4717660)",
      "OAS1 (12:112945172)",
      "HLA-G (6:29947491)",
      "ABO (9:133271182)"
    ))
  )

# Get significance data for exact SNP-gene pairs
pval_data_combined <- box_plot_dat %>%
  inner_join(pairs_to_plot, by = c("grch38" = "snp", "gene_name" = "gene")) %>%
  filter(visit %in% c("1","2","3","4","6","7","8","9","10")) %>%
  group_by(gene_name, grch38, visit, group) %>%
  summarize(eigenMT = min(eigenMT, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    group = case_when(
      group == "control" ~ "TG1-3",
      group == "case" ~ "TG4-5"
    ),
    group = factor(group, levels = c("TG1-3", "TG4-5")),
    sig_stars = case_when(
      eigenMT <= 0.001 ~ "***",
      eigenMT <= 0.01 ~ "**",
      eigenMT <= 0.05 ~ "*",
      TRUE ~ ""
    ),
    visit = factor(visit, levels = c("1","2","3","4","6","7","8","9","10"), ordered = TRUE),
    gene_label = case_when(
      gene_name == "TMIGD2" & grch38 == "19:4717660" ~ "TMIGD2 (19:4717660)",
      gene_name == "OAS1" & grch38 == "12:112945172" ~ "OAS1 (12:112945172)",
      gene_name == "HLA-G" & grch38 == "6:29947491" ~ "HLA-G (6:29947491)",
      gene_name == "ABO" & grch38 == "9:133271182" ~ "ABO (9:133271182)"
    ),
    gene_label = factor(gene_label, levels = c(
      "TMIGD2 (19:4717660)",
      "OAS1 (12:112945172)",
      "HLA-G (6:29947491)",
      "ABO (9:133271182)"
    ))
  ) %>%
  filter(sig_stars != "")

# Get y-position data for significance stars (using combined_dat which already has exact pairs)
y_stats_combined <- combined_dat %>%
  group_by(gene_label, visit, group) %>%
  summarize(
    q3 = quantile(residuals, 0.75, na.rm = TRUE),
    iqr = IQR(residuals, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    upper_whisker = q3 + 1.5 * iqr,
    y_position = upper_whisker + 0.5 * iqr
  )

# Join significance with positioning data
pval_data_combined <- pval_data_combined %>%
  left_join(y_stats_combined, by = c("gene_label", "visit", "group"))

# Create the unified plot
figure4 <- ggplot(combined_dat, aes(x = visit, y = residuals, fill = genotype)) +
  geom_boxplot(width = 0.65, outlier.shape = NA) +
  facet_grid(gene_label ~ group, scales = "free_y") +
  scale_fill_manual(values = c("0" = "#E69F00", "1" = "#56B4E9", "2" = "#009E73"), name = "Genotype") +
  geom_vline(xintercept = 5.5, linetype = "dashed", color = "black", alpha = 0.8) +
  labs(
    x = "Visit",
    y = "Residual expression (covariate-adjusted)",
    title = "Dynamic eQTLs Across COVID-19 Disease Progression"
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 12, face = "bold"),
    axis.title.y = element_text(size = 14, face = "bold"),
    strip.background = element_rect(colour = "black", fill = "grey90", linewidth = 0.8),
    strip.text = element_text(size = 12, face = "bold", margin = margin(4, 4, 4, 4)),
    legend.title = element_text(size = 13, face = "bold"),
    legend.text = element_text(size = 11),
    legend.position = "bottom",
    panel.spacing = unit(0.5, "lines"),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5)
  )

# Add significance stars
if (nrow(pval_data_combined) > 0) {
  figure4 <- figure4 +
    geom_text(data = pval_data_combined,
              aes(x = visit, y = y_position, label = sig_stars),
              size = 6.5, fontface = "bold", inherit.aes = FALSE)
}

# Add caption
figure4 <- figure4 +
  labs(caption = "Visits 1-6: Acute phase | Visits 7-10: Convalescent phase\n* FDR < 0.05  ** FDR < 0.01  *** FDR < 0.001 (eigenMT)") +
  theme(plot.caption = element_text(size = 12, hjust = 0.5, face = "italic"))

# saves as width 1100 height 1200
figure4
```

## Panel 4B: loess-smoothed dose-response curves

We fit loess curves of residual expression over visit for the four exemplar
loci: first by genotype within each severity group (and pooled as TG1-5 All),
and separately a severity-group comparison pooled across genotypes.

```{r}
# Define the 4 exemplar gene-SNP pairs.
exemplar_pairs <- tribble(
  ~gene_name, ~snp,
  "TMIGD2",   "19:4717660",
  "OAS1",     "12:112945172",
  "HLA-G",    "6:29947491",
  "ABO",      "9:133271182"
)

# Step 1: base data with the exemplar genes and correct facet order.
spaghetti_data_base <- box_plot_dat %>%
  filter(
    gene_name %in% exemplar_pairs$gene_name,
    grch38 %in% exemplar_pairs$snp,
    visit %in% c("1","2","3","4","6","7","8","9","10"),
    !is.na(genotype), !is.na(residuals)
  ) %>%
  mutate(
    group_clean = case_when(
      group == "control" ~ "TG1-3",
      group == "case" ~ "TG4-5",
      TRUE ~ as.character(group)
    ),
    visit_num = as.numeric(visit),
    gene_snp_label = paste0(gene_name, " (", grch38, ")"),
    genotype_label = factor(genotype,
                           levels = c("0", "1", "2"),
                           labels = c("GT0", "GT1", "GT2")),
    gene_snp_label = factor(gene_snp_label, levels = c(
      "TMIGD2 (19:4717660)",
      "OAS1 (12:112945172)",
      "HLA-G (6:29947491)",
      "ABO (9:133271182)"
    ))
  )

# Step 2: the "All" dataset by duplicating and relabeling.
spaghetti_data_all <- spaghetti_data_base %>%
  mutate(group_clean = "TG1-5 (All)")

# Step 3: combine both datasets.
spaghetti_data_with_all <- bind_rows(
  spaghetti_data_base,   # Original TG1-3 and TG4-5 data
  spaghetti_data_all     # Same data labeled as "All"
) %>%
  mutate(
    group_clean = factor(group_clean,
                        levels = c("TG1-5 (All)", "TG1-3", "TG4-5"))
  )

# PLOT 1: Severity Group Comparison (loess, pooled across genotypes)
severity_comparison_plot <- spaghetti_data_base %>%
  ggplot(aes(x = visit_num, y = residuals)) +

  geom_smooth(aes(color = group_clean, fill = group_clean),
              method = "loess", span = 0.8, se = TRUE,
              alpha = 0.1, size = 1.5, linetype = "solid") +

  facet_wrap(~ gene_snp_label, scales = "free_y", ncol = 2) +

  scale_color_manual(
    name = "Severity Group",
    values = c("TG1-3" = "#4472C4", "TG4-5" = "#7030A0")
  ) +

  scale_fill_manual(
    name = "Severity Group",
    values = c("TG1-3" = "#4472C4", "TG4-5" = "#7030A0"),
    guide = "none"
  ) +

  scale_x_continuous(
    breaks = c(1,2,3,4,6,7,8,9,10),
    labels = paste0("V", c(1,2,3,4,6,7,8,9,10))
  ) +

  labs(
    title = "Severity Group Differences in Expression Patterns",
    subtitle = "Overall expression trajectories by disease severity (pooled across genotypes)",
    x = "Visit",
    y = "Residual Expression",
    caption = "Blue = TG1-3 (mild) | Purple = TG4-5 (severe)"
  ) +

  theme_bw() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12),
    strip.text = element_text(size = 11, face = "bold"),
    axis.title = element_text(size = 11, face = "bold"),
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 0.5),
    legend.position = "bottom",
    panel.grid.major = element_line(color = "grey93", size = 0.5),
    panel.grid.minor = element_blank()
  )

# PLOT 2: Genotype dose-response by severity group (loess)
spaghetti_genotype_plot <- ggplot(spaghetti_data_with_all,
                                  aes(x = visit_num, y = residuals)) +

    geom_smooth(aes(color = factor(genotype), fill = factor(genotype)),
                method = "loess", span = 0.8, se = TRUE,
                alpha = 0.1, size = 1.2, linetype = "solid") +

    facet_grid(gene_snp_label ~ group_clean, scales = "free_y") +

    geom_text(
        data = spaghetti_data_with_all %>%
               filter(group_clean == "TG1-5 (All)") %>%
               group_by(gene_snp_label) %>%
               slice_head(n = 1),  # One row per gene
        aes(x = -Inf, y = Inf, label = "Combined Analysis"),
        hjust = -0.1, vjust = 1.1,
        size = 3, color = "grey30", fontface = "italic",
        inherit.aes = FALSE
    ) +

    scale_color_manual(
        name = "Genotype",
        values = c("0" = "#E69F00", "1" = "#56B4E9", "2" = "#009E73"),
        labels = c("0" = "GT0", "1" = "GT1", "2" = "GT2")
    ) +

    scale_fill_manual(
        name = "Genotype",
        values = c("0" = "#E69F00", "1" = "#56B4E9", "2" = "#009E73"),
        guide = "none"
    ) +

    scale_x_continuous(
        breaks = c(1,2,3,4,6,7,8,9,10),
        labels = paste0("V", c(1,2,3,4,6,7,8,9,10))
    ) +

    labs(
        title = "Temporal Expression Patterns: Genotype Effects by Severity Group",
        subtitle = "Genotype dose-response curves showing population vs subgroup effects",
        x = "Visit", y = "Residual Expression",
        caption = "Left = TG1-5 (all patients) | Middle = TG1-3 | Right = TG4-5"
    ) +

    theme_bw() +
    theme(
        plot.title = element_text(size = 14, face = "bold"),
        plot.subtitle = element_text(size = 12),
        strip.text = element_text(size = 10, face = "bold"),
        axis.title = element_text(size = 11, face = "bold"),
        axis.text = element_text(size = 9),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "bottom",
        panel.grid.major = element_line(color = "grey93", size = 0.5),
        panel.grid.minor = element_blank()
    )

# Display both plots
severity_comparison_plot
spaghetti_genotype_plot

# saves as width 1100 height 1200
```
