---
title: "nasal_rnaseq"
output: html_document
---

Source: 20230125_bulk_RNAseq_nasal

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

library(limma)
library(edgeR)

library(fgsea)

library(gamm4)

setwd("~/20220902_IMPACC_aging")

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```

# Data import

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE
clin.event$steroids <- as.factor(clin.event$steroids)

head(clin.event)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

### Merge clinical metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ, by = "participant_id", all = FALSE, sort = FALSE
)
```

Add patients' individual metadata.

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

## RNA-seq data

```{r}
counts <- read.csv(
  "/data/nasal-transcriptomics/current/nasal-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1,
  check.names = FALSE
)
counts <- t(counts)
print(counts[1:6,1:6])

counts.metadata <- read.csv(
  "/data/nasal-transcriptomics/current/nasal-transcriptomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
print(head(counts.metadata))

rData <- read.csv(
  "/data/nasal-transcriptomics/current/nasal-transcriptomics-RowFeature.csv",
  header = FALSE, row.names = NULL,
  check.names = FALSE
)
colnames(rData) <- rData[2,]
rData <- rData[-(1:2),]
print(head(rData))
```

## Viral load data

SARS-CoV-2 viral load is readily available in nasal-metagenomics-Metadata.csv, sc2_rpm column.

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)

# mngs <- read.csv(
#   "/data/nasal-metagenomics/current/nasal-metagenomics-AllGenera-Counts_RPM.csv",
#   header = TRUE, row.names = 1,
#   check.names = FALSE
# )

# mngs.rData <- read.csv(
#   "/data/nasal-metagenomics/current/nasal-metagenomics-AllGenera-RowFeatures.csv",
#   skip = 1, header = TRUE, row.names = 1,
#   check.names = FALSE
# )
# sc2.id <- mngs.rData %>%
#   subset(name == "Betacoronavirus") %>%
#   rownames()
```

## Only keep samples that pass QC

```{r}
# RNA-seq QC
counts.metadata <- subset(counts.metadata, sample_status == "IMPACC sample assayed and passed QC")

# mNGS QC
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")
# Remove rows without sc2_rpm
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))

# Get a list of samples for RNA-seq count
pass.qc <- Reduce(
  intersect, 
  list(colnames(counts), counts.metadata$sample_id, clin.sample.event.individ$sample_id)
)

counts <- counts[,pass.qc]
clin.sample.event.individ <- clin.sample.event.individ %>%
  subset(sample_id %in% pass.qc) %>%
  droplevels()
```

## Load pathways

```{r}
# Reactome
reactome <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "C2", subcategory = "CP:REACTOME")
reactome.list <- split(x = reactome$ensembl_gene, f = reactome$gs_name)

# Hallmark gene sets
hallmark <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "H")
hallmark.list <- split(x = hallmark$ensembl_gene, f = hallmark$gs_name)
```

# Hospital visit 1 samples

## Prepare data
```{r}
# Keep visit 1 samples for RNA-seq
clin.sample.event.individ.1 <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Keep visit 1 samples for viral load
clin.sample.event.individ.sc2.1 <- clin.sample.event.individ.sc2 %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Add viral load
clin.sample.event.individ.1 <- merge(
  clin.sample.event.individ.1, clin.sample.event.individ.sc2.1[,c("participant_id","sc2_rpm")],
  by = "participant_id", all = FALSE, sort = FALSE
)
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.1$sc2_rpm <- log10(clin.sample.event.individ.1$sc2_rpm + 1)

# Drop unused levels
clin.sample.event.individ.1 <- droplevels(clin.sample.event.individ.1)

# Any row with no sc2_rpm?
print(any(is.na(clin.sample.event.individ.1$sc2_rpm)))

# Any duplicated participant?
print(any(duplicated(clin.sample.event.individ.1$participant_id)))
```

```{r}
# Subset count data
counts.1 <- counts[,clin.sample.event.individ.1$sample_id]
```

QC.

```{r}
# Keep only protein_coding genes
counts.1 <- counts.1[rData[rData$gene_biotype=="protein_coding", "gene_id"],]

# Keep genes with at least 10 counts in 20% of samples
keep <- rowSums(counts.1>=10) >= 0.2*dim(counts.1)[2]
counts.1 <- counts.1[keep,]
```

Export normalized gene expression at visit 1.

```{r}
# Normalize counts at visit 1 without taking into account any covariate
temp <- counts.1
colnames(temp) <- clin.sample.event.individ.1$event_id
temp <- voom(temp, 
             normalize.method = "quantile",
             plot = T)
# write.csv(temp$E, "20230125_bulk_RNAseq_nasal_visit1_counts_voom-quantile.csv")

# Export rows data
temp.rData <- rData
rownames(temp.rData) <- temp.rData$gene_id
# write.csv(temp.rData[rownames(temp),],
#           "20230125_bulk_RNAseq_nasal_visit1_RowFeatures_voom-quantile.csv",
#           row.names = FALSE)

# Save multple objects as RDS
# visit1.norm.expr <- temp$E
# visit1.rData <- temp.rData
# save(visit1.norm.expr, visit1.rData, clin.sample.event.individ.1, 
#      file = "20230125_bulk_RNAseq_nasal_visit1_voom.RData")

# remove(temp, temp.rData)
```

## Explore data

```{r}
# Age distribution
hist(clin.sample.event.individ.1$admit_age, breaks=20,
     xlab = "Admit age", main = "nasal")

# Age by sex
boxplot(admit_age~sex, data=clin.sample.event.individ.1, main = "nasal")
```

# Visit 1: Effects of age

## Without controlling for viral load

```{r}
design <- model.matrix(~ admit_age + sex + respiratory_status,
                       data=clin.sample.event.individ.1)

vwts <- voom(counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

```{r}
res <- topTable(vfit, coef = "admit_age",
                sort.by = "none", number = Inf,
                adjust.method = "BH", p.value = 1)
head(res)

# Save results
# write.csv(res,
#           "20230125_bulk_RNAseq_nasal_visit1_DE_no-viralload.csv")
```

Volcano plot

```{r}
# Figure 6A
# Identify direction
res.toplot <- merge(res, rData[,c("gene_id","gene_name")],
                    all.x = TRUE, all.y = FALSE,
                    by.x = "row.names", by.y = "gene_id",
                    sort = FALSE)
res.toplot$direction <- "Not significant"
res.toplot[res.toplot$adj.P.Val<0.05 & res.toplot$logFC>0,"direction"] <- "Up with age"
res.toplot[res.toplot$adj.P.Val<0.05 & res.toplot$logFC<0,"direction"] <- "Down with age"
res.toplot$direction <- factor(
  res.toplot$direction,
  levels=c("Up with age","Down with age","Not significant"))

p <- ggplot(data=res.toplot) +
  geom_point(aes(x=logFC, y=-log10(adj.P.Val), color=direction), size=0.8) +
  # geom_hline(yintercept=-log10(0.05), color="red", linetype="dashed") +
  scale_color_manual(values=c("darkred","dodgerblue4","gray75"), # c("#d6604d","#4393c3","gray75")
                     name="") +
  scale_x_continuous(limits=c(-0.06,0.06)) +
  scale_y_continuous(limits=c(-0.3,10), expand=c(0,0)) +
  # xlim(-0.06, 0.06) + 
  labs(x="Slope \n(change per 1 year of age)", y="-log10(adjusted P)",
       title="Nasal swab genes associated\nwith age at Visit 1") +
  my.theme # RNAseq_nasal_visit1_age_no-viralload, 4.8 x 3.6 in or clipboard 550 x 440
```

MDS plot: check if data clusters by sex.

```{r}
lcpm <- cpm(vwts, log = TRUE)

group <- clin.sample.event.individ.1$sex
col.group <- factor(group, labels=c("red","blue")) %>% as.character()
# p <- plotMDS(lcpm, labels=group, col=col.group)
```

Reactome GSEA

```{r}
# Sort gene list in descending order
gene.ranks <- res$t
names(gene.ranks) <- rownames(res)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)

# # How many pathways are significant?
# print(sum(reactome.gsea$padj < 0.05))
# # What are they?
# reactome.gsea.sig <- subset(reactome.gsea, padj < 0.05)

# Save the result for comparison later
reactome.gsea.noviral <- reactome.gsea
```

## Control for viral load

```{r}
design <- model.matrix(~ admit_age + sc2_rpm + sex + respiratory_status,
                       data=clin.sample.event.individ.1)

vwts <- voom(counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

```{r}
res <- topTable(vfit, coef = "admit_age",
                sort.by = "none", number = Inf,
                adjust.method = "BH", p.value = 1)
head(res)
```

Volcano plot

```{r}
ggplot(data=res) +
  geom_point(aes(x=logFC, y=-log10(adj.P.Val), color=(adj.P.Val<0.05)), size=1) +
  # geom_hline(yintercept=-log10(0.05), color="red", linetype="dashed") +
  scale_color_manual(values=c("FALSE"="black", "TRUE"="red")) +
  xlim(-0.04, 0.04) + 
  labs(x="logFC (change per 1 year of age)", y="-log10 (adjusted P)") +
  my.theme + theme(legend.position = "bottom")
```

Reactome GSEA.

```{r}
# Reactome GSEA
# Sort gene list in descending order
gene.ranks <- res$t
names(gene.ranks) <- rownames(res)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)

# # How many pathways are significant?
# print(sum(reactome.gsea$padj < 0.05))
```

Plot selected reactome pathways, with and without controlling for viral load

```{r}
# Get list of pathways that are significant in at least one analysis
my.pathways <- union(
  reactome.gsea.noviral[reactome.gsea.noviral$padj<0.05,"pathway"],
  reactome.gsea[reactome.gsea$padj<0.05,"pathway"]
) # save as 20230125_bulk_RNAseq_nasal_visit1_age_gsea-reactome_yes-vs-no-viralload_padj0.05.csv if needed
```

```{r}
# Manually choose pathways
my.pathways <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING",
                 "REACTOME_INTERFERON_GAMMA_SIGNALING",
                 "REACTOME_INTERLEUKIN_10_SIGNALING","REACTOME_INTERLEUKIN_17_SIGNALING",
                 "REACTOME_INTERLEUKIN_1_FAMILY_SIGNALING",
                 "REACTOME_INTERLEUKIN_4_AND_INTERLEUKIN_13_SIGNALING",
                 "REACTOME_TOLL_LIKE_RECEPTOR_TLR1_TLR2_CASCADE",
                 "REACTOME_MYD88_INDEPENDENT_TLR4_CASCADE",
                 "REACTOME_TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE",
                 "REACTOME_COMPLEMENT_CASCADE",
                 "REACTOME_CASPASE_ACTIVATION_VIA_DEATH_RECEPTORS_IN_THE_PRESENCE_OF_LIGAND",
                 "REACTOME_TCR_SIGNALING","REACTOME_COSTIMULATION_BY_THE_CD28_FAMILY",
                 # "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR",
                 "REACTOME_PD_1_SIGNALING")

reactome.gsea.toplot <- rbind(
  subset(reactome.gsea.noviral, pathway %in% my.pathways) %>% mutate(viral="No"),
  subset(reactome.gsea, pathway %in% my.pathways) %>% mutate(viral="Yes")
)
reactome.gsea.toplot$sig <- as.factor(reactome.gsea.toplot$padj)

# Reformat pathway names
reactome.gsea.toplot$pathway <- gsub("^REACTOME_","",reactome.gsea.toplot$pathway)
reactome.gsea.toplot$pathway <- gsub("_"," ",reactome.gsea.toplot$pathway)
reactome.gsea.toplot$pathway <- stringr::str_to_sentence(reactome.gsea.toplot$pathway)

# Fix capitalization
my.dict <- list("tlr"="TLR", "Myd88"="MYD88",
                "Pd 1"="PD1", "Tcr"="TCR", "cd28"="CD28",
                "Caspase activation via death receptors in the presence of ligand" = "Caspase activation via depth receptors")
for (i in names(my.dict)) {
  reactome.gsea.toplot$pathway <- gsub(i, my.dict[[i]], reactome.gsea.toplot$pathway)
}

# Figure 6B
p <- ggplot(data = reactome.gsea.toplot, 
            aes(x=reorder(pathway,NES,FUN=function(x) x[1]), 
                y=NES)) +
  # geom_segment(aes(xend=reorder(pathway,NES)), yend=0) +
  geom_point(aes(color=as.factor(NES>0), 
                 shape=as.factor(padj<0.05)),
             size=2.5, stroke=1) +
  geom_hline(yintercept = 0, size = 0.7, color = "black") +
  scale_color_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  scale_shape_manual(values=c("FALSE"=1,"TRUE"=19), guide="none") +
  # scale_color_manual(values=c("FALSE"=alpha("black",0),"TRUE"=alpha("black",1)), guide="none") +
  scale_y_continuous(limits = c(-4,4), expand = c(0,0)) +
  labs(x="", y="GSEA normalized\nenrichment score",
       title="Pathways upregulated\nwith age at Visit 1") +
  facet_wrap(~viral) +
  coord_flip(ylim = c(-4,4)) + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    panel.grid.minor=element_blank()
    # legend.key = element_rect(fill = "gray90")
  ) # RNAseq_nasal_visit1_age_gsea-reactome_yes-vs-no-viralload, 5.8 x 4.6 in
```

Save DE and GSEA results

```{r}
# write.csv(
#   res,
#   "20220914_bulk_RNAseq_nasal_admit-age_limma-voom.csv"
# )
# 
# data.table::fwrite(
#   reactome.gsea,
#   file="20220914_bulk_RNAseq_nasal_admit-age_reactome-gsea.tsv",
#   sep="\t", sep2=c("", " ", "")
# )
# 
# data.table::fwrite(
#   hallmark.gsea,
#   file="20220914_bulk_RNAseq_nasal_admit-age_hallmark-gsea.tsv",
#   sep="\t", sep2=c("", " ", "")
# )
```

## Age and severity: stratified by age (combined design)

Here I combine both age groups and both severity levels into one analysis

```{r}
clin.sample.event.individ.1.yo <- clin.sample.event.individ.1 %>% 
  subset((admit_age <= 46) | (admit_age > 70))
clin.sample.event.individ.1.yo$age_group <- "Young"
clin.sample.event.individ.1.yo[clin.sample.event.individ.1.yo$admit_age > 70,"age_group"] <- "Old"

# Mild is 3 & 4, severe is 5 & 6
clin.sample.event.individ.1.yo$severity <- "Mild"
clin.sample.event.individ.1.yo[clin.sample.event.individ.1.yo$respiratory_status %in% c("5","6"), "severity"] <- "Severe"

# Combine age and severity into one factor
clin.sample.event.individ.1.yo$age.severity <- paste(
  clin.sample.event.individ.1.yo$age_group,
  clin.sample.event.individ.1.yo$severity,
  sep = "."
)

counts.1.yo <- counts.1[,clin.sample.event.individ.1.yo$sample_id]
```

### Not control for viral load

DE analysis

```{r}
design <- model.matrix(~ 0 + age.severity + sex,
                       data=clin.sample.event.individ.1.yo)
# Rename age.severity column names
colnames(design) <- gsub("age.severity", "", colnames(design))
print(colnames(design))

vwts <- voom(counts.1.yo, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts, design = design) # don't actually need to specify design here after voom
cont.matrix <- makeContrasts(
  SvsMinY = Young.Severe - Young.Mild,
  SvsMinO = Old.Severe - Old.Mild,
  OvsYinS = Old.Severe - Young.Severe,
  OvsYinM = Old.Mild - Young.Mild,
  levels = design)
vfit2 <- contrasts.fit(vfit, cont.matrix)
vfit2 <- eBayes(vfit2)

print(summary(decideTests(vfit2)))
```

Extract DE results

```{r}
res.SvsMinY <- topTable(vfit2, coef = "SvsMinY", 
                        sort.by = "none", number = Inf,
                        adjust.method = "BH", p.value = 1)
res.SvsMinO <- topTable(vfit2, coef = "SvsMinO", 
                        sort.by = "none", number = Inf,
                        adjust.method = "BH", p.value = 1)
```

Reactome GSEA

```{r}
# Young
gene.ranks <- res.SvsMinY$t
names(gene.ranks) <- rownames(res.SvsMinY)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.SvsMinY <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)

# Old
gene.ranks <- res.SvsMinO$t
names(gene.ranks) <- rownames(res.SvsMinO)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.SvsMinO <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)
```

Find Reactome pathways that are significant in either young or old, or both.

```{r}
merged.reactome.gsea <- rbind(
  reactome.gsea.SvsMinY %>% mutate(age_group="[18,46]"), 
  reactome.gsea.SvsMinO %>% mutate(age_group="[71,96]")
)

# Get list of pathways that are significant in at least one analysis
my.pathways <- union(
  reactome.gsea.SvsMinY[reactome.gsea.SvsMinY$padj<0.05,"pathway"],
  reactome.gsea.SvsMinO[reactome.gsea.SvsMinO$padj<0.05,"pathway"]
)

# merged.reactome.gsea <- merged.reactome.gsea %>%
#   subset(pathway %in% my.pathways$pathway)
```

Save merged GSEA result

```{r}
data.table::fwrite(
  merged.reactome.gsea,
  file="20230125_bulk_RNAseq_nasal_visit1_age-group_gsea-reactome_severe-vs-mild_combined-design.tsv",
  sep = "\t", sep2 = c("", " ", "")
)
```
