---
title: "pbmc-nasal_rnaseq_correlation"
output: html_document
---

Source: 20230125_data_integration_pbmc-ns.Rmd

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

library(limma)
library(edgeR)

library(fgsea)

library(gamm4)

setwd("~/20220902_IMPACC_aging")

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```


# Data import

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE
clin.event$steroids <- as.factor(clin.event$steroids)

head(clin.event)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

### Merge clinical metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ, by = "participant_id", 
  all = FALSE, sort = FALSE
)
```

Add patients' individual metadata.

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

## RNA-seq data

PBMC RNA-seq

```{r}
pbmc.counts <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1,
  check.names = FALSE
)
pbmc.counts <- t(pbmc.counts)
print(pbmc.counts[1:6,1:6])

pbmc.metadata <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
print(head(pbmc.metadata))

pbmc.rData <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-RowFeature.csv",
  header = FALSE, row.names = NULL,
  check.names = FALSE
)
colnames(pbmc.rData) <- pbmc.rData[2,]
pbmc.rData <- pbmc.rData[-(1:2),]
print(head(pbmc.rData))
```

NS RNA-seq

```{r}
nasal.counts <- read.csv(
  "/data/nasal-transcriptomics/current/nasal-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1,
  check.names = FALSE
)
nasal.counts <- t(nasal.counts)
print(nasal.counts[1:6,1:6])

nasal.metadata <- read.csv(
  "/data/nasal-transcriptomics/current/nasal-transcriptomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
print(head(nasal.metadata))

nasal.rData <- read.csv(
  "/data/nasal-transcriptomics/current/nasal-transcriptomics-RowFeature.csv",
  header = FALSE, row.names = NULL,
  check.names = FALSE
)
colnames(nasal.rData) <- nasal.rData[2,]
nasal.rData <- nasal.rData[-(1:2),]
print(head(nasal.rData))
```

## Viral load data

SARS-CoV-2 viral load is readily available in nasal-metagenomics-Metadata.csv, sc2_rpm column.

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)

# mngs <- read.csv(
#   "/data/nasal-metagenomics/current/nasal-metagenomics-AllGenera-Counts_RPM.csv",
#   header = TRUE, row.names = 1,
#   check.names = FALSE
# )

# mngs.rData <- read.csv(
#   "/data/nasal-metagenomics/current/nasal-metagenomics-AllGenera-RowFeatures.csv",
#   skip = 1, header = TRUE, row.names = 1,
#   check.names = FALSE
# )
# sc2.id <- mngs.rData %>%
#   subset(name == "Betacoronavirus") %>%
#   rownames()
```

## Only keep samples that pass QC

```{r}
# RNA-seq QC
pbmc.metadata <- subset(pbmc.metadata, sample_status == "IMPACC sample assayed and passed QC")
# Olink QC
nasal.metadata <- subset(nasal.metadata, sample_status=="IMPACC sample assayed and passed QC")

# mNGS QC
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")
# Remove rows without sc2_rpm
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))

# Filter samples that pass QC
clin.sample.event.individ <- clin.sample.event.individ %>%
  subset((sample_id %in% pbmc.metadata$sample_id) |
           (sample_id %in% nasal.metadata$sample_id))
```

## Load pathways for GSEA

```{r}
# Reactome
reactome <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "C2", subcategory = "CP:REACTOME")
reactome.list <- split(x = reactome$ensembl_gene, f = reactome$gs_name)
reactome.list2 <- reactome %>% select(gs_name, ensembl_gene)

# Hallmark gene sets
hallmark <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "H")
hallmark.list <- split(x = hallmark$ensembl_gene, f = hallmark$gs_name)
hallmark.list2 <- hallmark %>% select(gs_name, ensembl_gene)

# Transcription factor
tft <- msigdbr::msigdbr(species = "Homo sapiens",
                        category = "C3")
tft.list2 <- tft %>% select(gs_name, ensembl_gene)
```

# Hospital visit 1 samples

```{r}
# Keep hospital visit 1 samples
clin.sample.event.individ.1 <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Add viral load
clin.sample.event.individ.1 <- merge(
  clin.sample.event.individ.1, clin.sample.event.individ.sc2[,c("event_id","sc2_rpm")],
  by = "event_id", all = FALSE, sort = FALSE
)
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.1$sc2_rpm <- log10(clin.sample.event.individ.1$sc2_rpm + 1)

# Drop unused levels
clin.sample.event.individ.1 <- droplevels(clin.sample.event.individ.1)

# Any row with no sc2_rpm?
print(any(is.na(clin.sample.event.individ.1$sc2_rpm)))
```

```{r}
# Subset PBMC RNA-seq visit 1 samples
pbmc.clin.sample.event.individ.1 <- clin.sample.event.individ.1 %>%
  subset(sample_id %in% colnames(pbmc.counts))

# Subset NS RNA-seq visit 1 samples
nasal.clin.sample.event.individ.1 <- clin.sample.event.individ.1 %>%
  subset(sample_id %in% colnames(nasal.counts))

# Any duplicated participant_id?
print(any(duplicated(pbmc.clin.sample.event.individ.1$participant_id)))
print(any(duplicated(nasal.clin.sample.event.individ.1$participant_id)))
```

Find overlapping event_id.

```{r}
my.event_id <- intersect(
  pbmc.clin.sample.event.individ.1$event_id,
  nasal.clin.sample.event.individ.1$event_id
)

# Subset
pbmc.clin.sample.event.individ.1 <- pbmc.clin.sample.event.individ.1 %>%
  subset(event_id %in% my.event_id)
nasal.clin.sample.event.individ.1 <- nasal.clin.sample.event.individ.1 %>%
  subset(event_id %in% my.event_id)

# Reorder based on event_id
pbmc.clin.sample.event.individ.1 <- pbmc.clin.sample.event.individ.1[order(pbmc.clin.sample.event.individ.1$event_id),]
nasal.clin.sample.event.individ.1 <- nasal.clin.sample.event.individ.1[order(nasal.clin.sample.event.individ.1$event_id),]

# Check if they are the same
print(all.equal(
  pbmc.clin.sample.event.individ.1,
  nasal.clin.sample.event.individ.1
  )
)
```

Extract PBMC and nasal data

```{r}
pbmc.counts.1 <- pbmc.counts[,pbmc.clin.sample.event.individ.1$sample_id]
nasal.counts.1 <- nasal.counts[,nasal.clin.sample.event.individ.1$sample_id]

# Rename sample_id to event_id
colnames(pbmc.counts.1) <- pbmc.clin.sample.event.individ.1$event_id
colnames(nasal.counts.1) <- nasal.clin.sample.event.individ.1$event_id
print(all(colnames(pbmc.counts.1) == colnames(nasal.counts.1)))

# QC on RNA-seq count data
pbmc.counts.1 <- pbmc.counts.1[pbmc.rData[pbmc.rData$gene_biotype=="protein_coding", "gene_id"],]
nasal.counts.1 <- nasal.counts.1[nasal.rData[nasal.rData$gene_biotype=="protein_coding", "gene_id"],]

# Keep genes with at least 10 counts in 20% of samples
keep <- rowSums(pbmc.counts.1>=10) >= 0.2*ncol(pbmc.counts.1)
pbmc.counts.1 <- pbmc.counts.1[keep,]
keep <- rowSums(nasal.counts.1>=10) >= 0.2*ncol(nasal.counts.1)
nasal.counts.1 <- nasal.counts.1[keep,]
```

## Explore data

```{r}
# Age distribution
hist(pbmc.clin.sample.event.individ.1$admit_age, breaks=20,
     xlab = "Admit age", main = "PBMC")

# Age by sex
boxplot(admit_age~sex, data=pbmc.clin.sample.event.individ.1, main = "PBMC")
```

Normalize data.

```{r}
# Normalize RNA-seq
pbmc.vwts <- voom(pbmc.counts.1,
                  normalize.method = "quantile",
                  plot = T)
nasal.vwts <- voom(nasal.counts.1,
                   normalize.method = "quantile",
                   plot = T)
```

## Gene-gene correlation

```{r}
res <- as.data.frame(matrix(NA, nrow=length(my.genes), ncol=4))
colnames(res) <- c("gene_id","all.corr","young.corr","old.corr")
rownames(res) <- my.genes
res$gene_id <- my.genes

for (i in rownames(res)) {
  res[i,"all.corr"] <- cor(pbmc.vwts$E[i,], nasal.vwts$E[i,],
                           method = "pearson")
  res[i,"young.corr"] <- cor(pbmc.vwts$E[i,pbmc.clin.sample.event.individ.1$admit_age<=46],
                             nasal.vwts$E[i,nasal.clin.sample.event.individ.1$admit_age<=46],
                             method = "pearson")
  res[i,"old.corr"] <- cor(pbmc.vwts$E[i,pbmc.clin.sample.event.individ.1$admit_age>=71],
                           nasal.vwts$E[i,nasal.clin.sample.event.individ.1$admit_age>=71],
                           method = "pearson")
}

# Add gene name
res <- merge(res, pbmc.rData[,c("gene_id","gene_name")],
             all.x = TRUE, all.y = FALSE,
             by = "gene_id", sort = FALSE)
```

Plot

```{r}
hist(res$all.corr,
     xlab="Pearson's correlation coefficient",
     main="Correlation of gene expression\nbetween PBMC and NS data")

plot(old.corr~young.corr, data=res,
     xlab="Pearson's correlation <= 46 yo",
     ylab="Pearson's correlation >= 71 yo",
     main="Correlation in gene expression between PBMC and NS")
```

Plot pretty

```{r}
# Figure 6D
p <- ggplot() +
  geom_point(data = res,
             aes(x=young.corr, y=old.corr,
                 color=as.factor(young.corr>0.5 & old.corr>0.5)),
             size=1) +
  geom_point(
    data=subset(res, gene_name %in% c("IFI6","IFI44","IFIT3","HLA-DQA2","HLA-DRB5")),
    aes(x=young.corr, y=old.corr), color="red", size=1
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  scale_color_manual(values=c("TRUE"="black", "FALSE"="gray75"),
                     guide="none") +
  scale_x_continuous(breaks=c(-0.3,0,0.5,1), limits=c(-0.37,1), expand=c(0,0)) +
  scale_y_continuous(breaks=c(-0.3,0,0.5,1), limits=c(-0.37,1), expand=c(0,0)) +
  labs(
    x="Correlation coefficient in [18,46]",
    y="Correlation coefficient in [71,96]",
    title="PBMC-Nasal gene expression\ncorrelation at Visit 1"
  ) + my.theme + theme(aspect.ratio = 1) # pbmc_nasal_visit1_corr_age-group, 360 x 360
```

Enriched pathway and transcription factor.

```{r}
ora <- clusterProfiler::enricher(
  gene = res[res$young.corr>0.5 & res$old.corr>0.5,"gene_id"],
  TERM2GENE = reactome.list2,
  # TERM2GENE = hallmark.list2,
  universe = res$gene_id,
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  pAdjustMethod = "fdr")
ora.res <- ora@result

ora.c3 <- clusterProfiler::enricher(
  gene = res[res$young.corr>0.5 & res$old.corr>0.5,"gene_id"],
  TERM2GENE = tft.list2,
  # TERM2GENE = hallmark.list2,
  universe = res$gene_id,
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  pAdjustMethod = "fdr")
ora.c3.res <- ora.c3@result
```

Plot IFN gamma and alpha/beta pathways.

```{r}
temp <- ora.res %>%
  subset(ID %in% c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING","REACTOME_INTERFERON_GAMMA_SIGNALING"))
temp$GeneRatio.calc <- sapply(temp$GeneRatio, FUN = function(x) as.numeric(strsplit(x,"/")[[1]][1]) / as.numeric(strsplit(x,"/")[[1]][2]))

# Reformat pathway names
temp$ID <- gsub("^REACTOME_","",temp$ID)
temp$ID <- gsub("_"," ",temp$ID)
temp$ID <- stringr::str_to_sentence(temp$ID)

ggplot(data = temp) +
  geom_point(aes(y=reorder(ID, GeneRatio.calc), 
                 x=-log10(p.adjust), 
                 size=GeneRatio.calc)) +
  scale_x_continuous(limit=c(0,16), expand=c(0,0)) +
  scale_size(limits=c(0.1,0.4), range=c(1,6), name = "Gene ratio") +
  labs(x="-log10(adjusted P)", y="", 
       title="Enriched pathways") +
  guides(size = guide_legend(ncol = 2)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    panel.grid.minor=element_blank(),
    legend.direction = "horizontal", legend.position="bottom"
    # legend.key = element_rect(fill = "gray90")
  ) # pbmc_nasal_visit1_corr_age-group_ora, 4.5 x 4 in
```

### Correlation in all 5 age quintiles

```{r}
res <- as.data.frame(matrix(NA, nrow=length(my.genes), ncol=4))
colnames(res) <- c("gene_id","all.corr","young.corr","old.corr")
rownames(res) <- my.genes
res$gene_id <- my.genes

my.age <- cut(pbmc.clin.sample.event.individ.1$admit_age,
              breaks=c(17,46,54,62,70,96))

for (i in rownames(res)) {
  res[i,"corr.all"] <- cor(pbmc.vwts$E[i,], nasal.vwts$E[i,],
                           method = "pearson")
  res[i,"corr.1"] <- cor(pbmc.vwts$E[i,my.age=="(17,46]"],
                         nasal.vwts$E[i,my.age=="(17,46]"],
                         method = "pearson")
  res[i,"corr.2"] <- cor(pbmc.vwts$E[i,my.age=="(46,54]"],
                         nasal.vwts$E[i,my.age=="(46,54]"],
                         method = "pearson")
  res[i,"corr.3"] <- cor(pbmc.vwts$E[i,my.age=="(54,62]"],
                         nasal.vwts$E[i,my.age=="(54,62]"],
                         method = "pearson")
  res[i,"corr.4"] <- cor(pbmc.vwts$E[i,my.age=="(62,70]"],
                         nasal.vwts$E[i,my.age=="(62,70]"],
                         method = "pearson")
  res[i,"corr.5"] <- cor(pbmc.vwts$E[i,my.age=="(70,96]"],
                         nasal.vwts$E[i,my.age=="(70,96]"],
                         method = "pearson")
}

# Add gene name
res <- merge(res, pbmc.rData[,c("gene_id","gene_name")],
             all.x = TRUE, all.y = FALSE,
             by = "gene_id", sort = FALSE)
```

