---
title: "Conv Olink Result"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "GSA"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))

# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```

### Load codebase.R and set path to your local output directory
```{r load-codebase}
source(file = "~/conv/data_analysis_template_codebase_conv.R")
```

### Functions: data processing
```{r data-processing}

preprocess_serum_olink <- function(serum_olink_counts, removeOutliers = TRUE) {
  require(package = "impute")
  
  # list of outlier samples to remove
  flag <- NULL
  if (removeOutliers) {
    flag <- c("0865-002K3G00-002", 
              "0865-002NVJ00-002", 
              "0865-0041HH00-002", 
              "0865-0062BG00-002") # consistent with core paper
  }
  serum_olink_counts_processed <- 
    serum_olink_counts[rowMeans(is.na(serum_olink_counts)) < 1 &
                       !(rownames(serum_olink_counts) %in% flag), ] %>%
    as.matrix() %>%
    impute.knn() %>%
    .$data %>%
    return(value = .)
}
```

## data processing
```{r}
data_env <- load_IMPACC_datasets(DATA_VERSION                = "2023-01-01", 
                                 KEEP_COVID19_POS            = TRUE, 
                                 FILTER_BY_CORE_ASSAY_COHORT = TRUE)
for (n in grep(pattern = "olink|clinical_data", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}


# preprocessing of the clinical data
serum_olink_counts_processed <- preprocess_serum_olink(serum_olink_counts)
```


## pull conv cluster info
```{r}
conv_cluster <- read.csv("/scratch/clinical/current/impacc-convalescent.csv")
conv_cluster$conv_cluster4 <- ifelse(conv_cluster$conv_cluster6 %in% c(1, 3, 4), 1, conv_cluster$conv_cluster6)


clinical_data <-  clinical_data %>%
          merge(conv_cluster, by = "participant_id")

case_ctrl <- read.csv("/scratch/clinical/current/impacc_conv_paired51.csv")%>%
  gather(-matchid, key = "case_ctrl", value = "participant_id")

clinical_data <-  clinical_data %>%
          left_join(case_ctrl)

clinical_data$trajectory_group_0 <- clinical_data$trajectory_group

conv_symtom <- read.csv("/scratch/clinical/current/impacc_conv_symptoms.csv")

clinical_data <-  clinical_data %>%
          left_join(conv_symtom)

#colnames(clinical_data)
```

### set outcome
```{r}
clinical_data$trajectory_group <- clinical_data$conv_cluster4
table(clinical_data$conv_cluster4, clinical_data$convsx_2plus_count)
```

### GAM: longi olink univar
```{r panel-c-analysis, message=FALSE, results="hide"}
data_use <- serum_olink_counts_processed %>%
  as.data.frame()%>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    event_type,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group", "event_type",
                         "discretized_admit_age_quantile", "sex")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))%>%
  filter(event_date <= 28)

## smooth spline takes a long time, demonstrating by cutting down to just ten factors
univar_smooth_spline_model_loop <- model_loop(data_use, 
                                       modelType = "lme",  # smoothSpline lme
                                       endpoint  = "trajectory_group")

univar_smooth_spline_model_loop <- mutate_all(univar_smooth_spline_model_loop, .funs = function(x) signif(x, 2))

plot_var <- univar_smooth_spline_model_loop[univar_smooth_spline_model_loop$adjp.slope <= 0.05|univar_smooth_spline_model_loop$adjp.intercept <= 0.05, ] %>%
  select(adjp.slope, adjp.intercept)

univar_smooth_spline_model_loop["FGF21", grepl("((intercept|slope)\\.adj|adjp.slope$|adjp.intercept$)", colnames(univar_smooth_spline_model_loop))]
```

### Fig.5
```{r}
TrajGrpCol <- c("1" = "#999999", "2" = "#56B4E9", "5" = "#E69F00", "6" = "#FF0000")
library(ggbeeswarm)

res_table_pairwise <- univar_smooth_spline_model_loop[ , grepl("p.intercept_", colnames(univar_smooth_spline_model_loop))]%>%
  mutate_all(.funs = function(x) signif(x, 2))

res_table_pairwise_long_df <- res_table_pairwise %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "feature") %>%
  pivot_longer(cols = -c("feature")) %>%
  rename(Pvalue = value) %>%
  mutate(name = gsub("p.intercept_", "", name))%>%
  tidyr::separate(col = "name", into = c("group1", "group2"), remove = FALSE, sep = "v")

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "FGF21") %>%
  filter(Pvalue < 0.1) %>%
  #mutate(Pvalue = signif(Pvalue,2)) %>%
  select(group1, group2, Pvalue) %>%
  #mutate(`y.position` = c(12, 13, 14)) %>%
  mutate(`y.position` = c(12, 13, 14, 15)) %>%
  as.tibble()

library(ggprism)

df_pooled_mean <- data_use %>%
  filter(name == "FGF21")%>%
  group_by(trajectory_group) %>%
  summarise(n = n(),
            FGF21_mean = mean(value))

boxplot_FGF21 <- ggplot(data_use %>%
  filter(name == "FGF21"), aes(x = trajectory_group, y = value)) + 
  geom_boxplot(aes(x = trajectory_group, y = value, color = trajectory_group), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = trajectory_group, y = value, color = trajectory_group), width = 0.2, alpha = 0.7, size = 1) +
  # geom_point(data = df_pooled_mean, aes(x = conv_cluster4, y = N1_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("FGF21") + xlab("") + #labs(tag = "B")+
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "Pvalue",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  #coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('MIN', 'PHY', 'COG', 'MLT'))

boxplot_FGF21


#ggsave(
#  filename = format(Sys.Date(), "conv/%y%m%d_boxplot_FGF21_PRO_clusters.pdf"),
#  width = 2.5,
#  height = 3
#)
```


### Fig.8S
```{r panel-c-output-gen, fig.height=4, fig.width=15}
lab = "FGF21"
exampleDF <- data_use %>%
  filter(name %in% lab) 

fig5.b <- plot_model(exampleDF, 
                          model_loop     = univar_smooth_spline_model_loop, 
                          modelType      = "lme", 
                          p_adjust       = F, 
                          colors = c("#999999", "#56B4E9",  "#E69F00","#FF0000"),
                          endpoint       = "trajectory_group", 
                          remove_NS_threshold = 0.1,
                          signif_markers = FALSE,
                          remove_NS      = TRUE, 
                          bar_height     = 6.5)

```


```{r, echo=F}
sessionInfo()
```
R version 4.0.2 (2020-06-22)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 18.04.6 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so

locale:
 [1] LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8   
 [6] LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C           LC_TELEPHONE=C        
[11] LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

attached base packages:
[1] parallel  grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] GSA_1.03.1           igraph_1.2.6         ggprism_1.0.4        scales_1.1.1         doParallel_1.0.16    doRNG_1.8.2         
 [7] rngtools_1.5         missForest_1.4       itertools_0.1-3      iterators_1.0.13     foreach_1.5.1        randomForest_4.6-14 
[13] ComplexHeatmap_2.6.2 corrr_0.4.3          gridExtra_2.3        rlang_1.0.6          RColorBrewer_1.1-2   cowplot_1.1.1       
[19] pals_1.6             forcats_0.5.1        stringr_1.4.0        dplyr_1.0.5          purrr_0.3.4          readr_1.4.0         
[25] tidyr_1.1.3          tibble_3.1.2         tidyverse_1.3.0      ggeffects_1.1.1      gamm4_0.2-6          mgcv_1.8-31         
[31] nlme_3.1-148         impute_1.64.0        pvca_0.1.0           magrittr_2.0.1       DT_0.16              survminer_0.4.9     
[37] patchwork_1.1.0      survival_3.2-9       ggpubr_0.4.0         ggbeeswarm_0.6.0     ggplot2_3.3.5        ordinal_2019.12-10  
[43] lme4_1.1-26          Matrix_1.2-18        qvalue_2.22.0        knitr_1.30      

```{r, eval=F, echo=F}
library(knitr)
options(knitr.duplicate.label = "allow")
rmarkdown::render("~/conv/conv_olink_result.Rmd", rmarkdown::html_document(pandoc_args = "--self-contained", toc = T))
```

