---
title: "Blood CyTOF univariate analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

### Load libraries
```{r load-packages}
rm(list = ls())
.libPaths( c( .libPaths(), "~/R.old/x86_64-pc-linux-gnu-library/4.0") )
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "RColorBrewer"))
suppressPackageStartupMessages(library(package = "ComplexHeatmap"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "scales"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "tidyverse"))

# set session options
options(show_col_types         = FALSE,
        dplyr.summarise.inform = FALSE)
VarCorr <- lme4::VarCorr # for PVCA
```

### Load codebase.R, load data matrices and set path to a local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# load codebase.R, olink data and clinical information
source(file = "~/conv/data_analysis_template_codebase_conv.R")
data_env <- load_IMPACC_datasets(DATA_VERSION                = "2023-01-01", 
                                 KEEP_COVID19_POS            = TRUE, 
                                 FILTER_BY_CORE_ASSAY_COHORT = TRUE)
for (n in grep(pattern = "bld_cytof|clinical_data", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}


# Define the outlier samples, if any, that should be removed.
outlier_sample_ids <- 
  rownames(bld_cytof_metadata)[bld_cytof_metadata$sample_status %in% "IMPACC sample assayed with questionable QC"]

```

### Functions: data processing
```{r data-processing}
# normalize parent on non-granulocyte counts
preprocess_bld_cytof_parent_nongran <- function(input_counts, input_rowfeature) {
  output_counts <- input_counts %>%
    rownames_to_column(var = "sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "rowname") %>%
    filter(!sample_id %in% outlier_sample_ids) %>%
    merge(y = rownames_to_column(input_rowfeature), by = "rowname") %>%
    group_by(sample_id, `Broad Parent Population`) %>%
    summarize(value = sum(value)) %>%
    pivot_wider(names_from = `Broad Parent Population`) %>%
    select(-contains(match = "Debris"), 
           -contains(match = "Multiplet"),
           -`Platelets`, 
           -`RBC`, 
           -`Undefined`,
           -`Neutrophil`,
           - Eosinophil,
           -`Basophil`) %>%
    column_to_rownames(var = "sample_id")
  output_counts <- output_counts/rowSums(output_counts)
 
  return(value = output_counts)
} 

# normalize children on non-granulocyte counts
preprocess_bld_cytof_child_nongran <- function(input_counts) {
  output_counts <-  input_counts %>%
    rownames_to_column(var = "sample_id") %>%
    filter(!sample_id %in% outlier_sample_ids) %>%
    select(-contains(match = "Debris"), 
           -contains(match = "Multiplet"),
           -`Platelets`, 
           -`RBC`, 
           -`Tier1_Undefined`,
           -`Tier2_undefined_Undefined`,
           -`Neutrophil (CD16low)`,
           -`Neutrophil (CD16hi)`,
           -`Eosinophil`,
           -`Basophil`) %>%
    column_to_rownames(var = "sample_id")
  output_counts <- output_counts/rowSums(output_counts)
  
  return(value = output_counts)
}

```

## modified code
```{r}
source(file = "~/conv/data_analysis_template_codebase_conv.R")
```

## pull conv cluster info
```{r}
conv_cluster <- read.csv("/scratch/clinical/current/impacc-convalescent.csv")
conv_cluster$conv_cluster4 <- ifelse(conv_cluster$conv_cluster6 %in% c(1, 3, 4), 1, conv_cluster$conv_cluster6)


clinical_data <-  clinical_data %>%
          merge(conv_cluster, by = "participant_id")

case_ctrl <- read.csv("/scratch/clinical/current/impacc_conv_paired51.csv")%>%
  gather(-matchid, key = "case_ctrl", value = "participant_id")

clinical_data <-  clinical_data %>%
          left_join(case_ctrl)

clinical_data$trajectory_group_0 <- clinical_data$trajectory_group

conv_symtom <- read.csv("/scratch/clinical/current/impacc_conv_symptoms.csv")

clinical_data <-  clinical_data %>%
          left_join(conv_symtom)

#colnames(clinical_data)
```
  
### set outcome
```{r}
clinical_data$trajectory_group <- clinical_data$conv_cluster4
table(clinical_data$conv_cluster4, clinical_data$convsx_2plus_count)
```


### Repeated measurment: bld_cytof_counts_processed_parent_nongran
```{r panel-a-analysis, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# non-granulocytes parent counts
bld_cytof_counts_processed_parent_nongran <- 
  preprocess_bld_cytof_parent_nongran(bld_cytof_counts, bld_cytof_rowfeature)

data_use_parent_nongran <- bld_cytof_counts_processed_parent_nongran %>%
  rownames_to_column(var = "sample_id") %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "+",
                     replacement = "pos", 
                     fixed       = TRUE)) %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "-",
                     replacement = "neg", 
                     fixed       = TRUE)) %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "\\(|\\)| ",
                     replacement = "_"
                     )) %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    event_type,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  column_to_rownames(var = "sample_id")%>%
  filter(event_date <= 28)

# a model with only intercept and random effect across sites

data_use <- data_use_parent_nongran
data_use$endpoints <- factor(data_use$trajectory_group)
data_use$trajectory_group <- NULL

res_table_pairwise <- NULL
res_table_ordinal <- NULL

query <- setdiff(names(data_use), names(clinical_data))
query <- query[query != "endpoints"]


for (varName in query) {
  my.formula0 <- paste0(varName,
                        '~(1|enrollment_site)+(1|participant_id)+event_type+',
                        'discretized_admit_age_quantile+sex')
  my.formula1 <- paste0(varName,
                        '~endpoints +(1|enrollment_site)+(1|participant_id)+event_type+',
                        'discretized_admit_age_quantile+sex')

  res_pairwise <- mixed_pairwise(my.formula0, 
                       my.formula1,
                       data_use = data_use)
  
  av = anova(lme4::lmer(my.formula0,data = data_use),
                 lme4::lmer(my.formula1,data = data_use))
  
  res_table_pairwise <- rbind(res_table_pairwise, res_pairwise)
  res_table_ordinal <- rbind(res_table_ordinal, data.frame("var" = varName, "pval" = av$`Pr(>Chisq)`[2]))
}


rownames(res_table_pairwise) <- query

res_table_pairwise_qvalue <- res_table_pairwise %>%
  as.data.frame() %>%
  mutate_all(.funs = function(x) qvalue::qvalue(x, fdr.level = 0.05, pi0 = 1)$qvalues)

res_table_pairwise <- res_table_pairwise %>%
  as.data.frame() 

res_table_ordinal <- res_table_ordinal%>%
  as.data.frame()%>%
  column_to_rownames(var = "var")%>%
  mutate(qval = qvalue::qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues)

res_table_ordinal["B_Cell", ]
res_table_pairwise["B_Cell", ]

```

### Fig.4
```{r}
colnames(data_use) <- gsub("_$", "", colnames(data_use))
res_table_ordinal$qval <- res_table_ordinal$pval

TrajGrpCol <- c("1" = "#999999", "2" = "#56B4E9", "5" = "#E69F00", "6" = "#FF0000")
library(ggbeeswarm)

res_table_pairwise_long_df <- res_table_pairwise %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "feature") %>%
  pivot_longer(cols = -c("feature")) %>%
  rename(Pvalue = value) %>%
  tidyr::separate(col = "name", into = c("group1", "group2"), remove = FALSE)

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "B_Cell") %>%
  filter(Pvalue < 0.1) %>%
  mutate(Pvalue = signif(Pvalue,2)) %>%
  select(group1, group2, Pvalue) %>%
  mutate(`y.position` = c(0.5, 0.55, 0.6, 0.65)) %>%
  as.tibble()

library(ggprism)

df_pooled_mean <- data_use %>%
  group_by(endpoints) %>%
  summarise(n = n(),
            B_Cell_mean = mean(B_Cell))

boxplot_B_Cell <- ggplot(data_use, aes(x = endpoints, y = B_Cell)) + 
  geom_boxplot(aes(x = endpoints, y = B_Cell, color = endpoints), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = endpoints, y = B_Cell, color = endpoints), width = 0.2, alpha = 0.7, size = 1) +
  # geom_point(data = df_pooled_mean, aes(x = conv_cluster4, y = N1_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("B_Cell") + xlab("") + #labs(tag = "A")+
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "Pvalue",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  #coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('MIN', 'PHY', 'COG', 'MLT'))

boxplot_B_Cell


#ggsave(
#  filename = format(Sys.Date(), "~/conv/B_Cell_boxplot_PRO_clusters.png"),
#  width = 2.5,
#  height = 3
#)
```

### Repeated measurment: data_use_child_nongran
```{r panel-b-analysis, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# ratio children to parent
bld_cytof_counts_processed_child_nongran <- 
  preprocess_bld_cytof_child_nongran(bld_cytof_counts)

data_use_child_nongran <- bld_cytof_counts_processed_child_nongran %>%
  rownames_to_column(var = "sample_id") %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "+",
                     replacement = "pos", 
                     fixed       = TRUE)) %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "-",
                     replacement = "neg", 
                     fixed       = TRUE)) %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "\\(|\\)| ",
                     replacement = "_"
                     )) %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    event_type,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  column_to_rownames(var = "sample_id")%>%
  filter(event_date <= 28)


data_use <- data_use_child_nongran
data_use$endpoints <- factor(data_use$trajectory_group)
data_use$trajectory_group <- NULL
query <- setdiff(names(data_use), names(clinical_data))
query <- query[query != "endpoints"]

res_table_pairwise <- NULL
res_table_ordinal <- NULL

query <- query[!grepl("Undefined", query)]
for (varName in query) {
  my.formula0 <- paste0(varName,
                        '~(1|enrollment_site)+(1|participant_id)+event_type+',
                        'discretized_admit_age_quantile+sex')
  my.formula1 <- paste0(varName,
                        '~endpoints +(1|enrollment_site)+(1|participant_id)+event_type+',
                        'discretized_admit_age_quantile+sex')

  res_pairwise <- mixed_pairwise(my.formula0, 
                       my.formula1,
                       data_use = data_use)
  
  av = anova(lme4::lmer(my.formula0,data = data_use),
                 lme4::lmer(my.formula1,data = data_use))
  
  res_table_pairwise <- rbind(res_table_pairwise, res_pairwise)
  res_table_ordinal <- rbind(res_table_ordinal, data.frame("var" = varName, "pval" = av$`Pr(>Chisq)`[2]))
}


rownames(res_table_pairwise) <- gsub("_$", "", query)

res_table_pairwise_qvalue <- res_table_pairwise %>%
  as.data.frame() #%>%
  #mutate_all(.funs = function(x) qvalue::qvalue(x, fdr.level = 0.05, pi0 = 1)$qvalues)

res_table_pairwise <- res_table_pairwise %>%
  as.data.frame() 

res_table_ordinal <- res_table_ordinal%>%
#  mutate(var = gsub("_$", "", var))%>%
  column_to_rownames(var = "var")%>%
  mutate(qval = qvalue::qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues)

res_table_pairwise_qvalue[apply(res_table_pairwise_qvalue, 1, function(x) sum(x < 0.05) > 0), ]

plot_var <- query[apply(res_table_pairwise_qvalue, 1, function(x) sum(x < 0.05) > 0)]

# b cell subset
res_table_ordinal[grepl("B_Cell__naive", rownames(res_table_ordinal)), ]
res_table_pairwise[grepl("B_Cell__naive", rownames(res_table_pairwise)), ]
```

### Fig.6S A
```{r}
colnames(data_use) <- gsub("_$", "", colnames(data_use))
res_table_ordinal$qval <- res_table_ordinal$pval

TrajGrpCol <- c("1" = "#999999", "2" = "#56B4E9", "5" = "#E69F00", "6" = "#FF0000")
library(ggbeeswarm)

res_table_pairwise_long_df <- res_table_pairwise %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "feature") %>%
  pivot_longer(cols = -c("feature")) %>%
  rename(Pvalue = value) %>%
  tidyr::separate(col = "name", into = c("group1", "group2"), remove = FALSE)

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "B_Cell__naive") %>%
  filter(Pvalue < 0.1) %>%
  mutate(Pvalue = signif(Pvalue,2)) %>%
  select(group1, group2, Pvalue) %>%
  mutate(`y.position` = c(0.25, 0.3)) %>%
  as.tibble()

library(ggprism)

df_pooled_mean <- data_use %>%
  group_by(endpoints) %>%
  summarise(n = n(),
            B_Cell__naive_mean = mean(B_Cell__naive))

boxplot_B_Cell__naive <- ggplot(data_use, aes(x = endpoints, y = B_Cell__naive)) + 
  geom_boxplot(aes(x = endpoints, y = B_Cell__naive, color = endpoints), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = endpoints, y = B_Cell__naive, color = endpoints), width = 0.2, alpha = 0.7, size = 1) +
  # geom_point(data = df_pooled_mean, aes(x = conv_cluster4, y = N1_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("B Cell (naive)") + xlab("") + #labs(tag = "B")+
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "Pvalue",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  #coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('MIN', 'PHY', 'COG', 'MLT'))

boxplot_B_Cell__naive


#ggsave(
#  filename = format(Sys.Date(), "conv/%y%m%d_B_Cell_boxplot_PRO_clusters.pdf"),
#  width = 2.5,
#  height = 3
#)
```


### GAM: bld_cytof_counts_processed_parent_nongran
```{r panel-c-analysis, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# children normalized on total number of events
data_use <- bld_cytof_counts_processed_parent_nongran %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "discretized_admit_age_quantile", "sex")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))%>%
  filter(event_date <= 28)

smooth_spline_model_loop <- model_loop(inputDF   = data_use, 
                                             modelType = "lme",
                                             endpoint  = "trajectory_group")

smooth_spline_model_loop["B Cell", ]
```

### Fig.6S.B
```{r}
lab = "B Cell"
exampleDF <- data_use %>%
  filter(name %in% lab) 

fig6s.b <- plot_model(exampleDF, 
                          model_loop     = smooth_spline_model_loop, 
                          modelType      = "lme", 
                          p_adjust       = F, 
                          colors = c("#999999", "#56B4E9",  "#E69F00","#FF0000"),
                          endpoint       = "trajectory_group", 
                          remove_NS_threshold = 0.1,
                          signif_markers = FALSE,
                          remove_NS      = TRUE, 
                          bar_height     = 6.5)

```

### GAM: bld_cytof_counts_processed_child_nongran
```{r panel-c-analysis, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# children normalized on total number of events
data_use <- bld_cytof_counts_processed_child_nongran %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "discretized_admit_age_quantile", "sex")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))%>%
  filter(event_date <= 28)

smooth_spline_model_loop <- model_loop(inputDF   = data_use, 
                                             modelType = "lme",
                                             endpoint  = "trajectory_group")
plot_var <- smooth_spline_model_loop[smooth_spline_model_loop$adjp.slope <= 0.05|smooth_spline_model_loop$adjp.intercept <= 0.05, ] %>%
  select(adjp.slope, adjp.intercept)
```

### supp plots
```{r}
lab = "B Cell (naive)"
exampleDF <- data_use %>%
  filter(name %in% lab) 

fig6s.c <- plot_model(exampleDF, 
                          model_loop     = smooth_spline_model_loop, 
                          modelType      = "lme", 
                          p_adjust       = F, 
                          colors = c("#999999", "#56B4E9",  "#E69F00","#FF0000"),
                          endpoint       = "trajectory_group", 
                          remove_NS_threshold = 0.1,
                          signif_markers = FALSE,
                          remove_NS      = TRUE, 
                          bar_height     = 6.5)



```
### Fig.6S
```{r}
library(patchwork)
layout <- '
AABBBB
AACCCC
'
wrap_plots(boxplot_B_Cell__naive, fig6s.b, fig6s.c)+
  plot_layout(design = layout, guides = "collect")+
  plot_annotation(tag_levels = 'A')
```


### Session info
```{r, print-session-info}
sessionInfo()
```
R version 4.0.2 (2020-06-22)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 18.04.6 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so

locale:
 [1] LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8   
 [6] LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C           LC_TELEPHONE=C        
[11] LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

attached base packages:
[1] parallel  grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] GSA_1.03.1           igraph_1.2.6         ggprism_1.0.4        scales_1.1.1         doParallel_1.0.16   
 [6] doRNG_1.8.2          rngtools_1.5         missForest_1.4       itertools_0.1-3      iterators_1.0.13    
[11] foreach_1.5.1        randomForest_4.6-14  ComplexHeatmap_2.6.2 corrr_0.4.3          gridExtra_2.3       
[16] rlang_1.0.6          RColorBrewer_1.1-2   cowplot_1.1.1        pals_1.6             forcats_0.5.1       
[21] stringr_1.4.0        dplyr_1.0.5          purrr_0.3.4          readr_1.4.0          tidyr_1.1.3         
[26] tibble_3.1.2         tidyverse_1.3.0      ggeffects_1.1.1      gamm4_0.2-6          mgcv_1.8-31         
[31] nlme_3.1-148         impute_1.64.0        pvca_0.1.0           magrittr_2.0.1       DT_0.16             
[36] survminer_0.4.9      patchwork_1.1.0      survival_3.2-9       ggpubr_0.4.0         ggbeeswarm_0.6.0    
[41] ggplot2_3.3.5        ordinal_2019.12-10   lme4_1.1-26          Matrix_1.2-18        qvalue_2.22.0       
[46] knitr_1.30          

loaded via a namespace (and not attached):
  [1] readxl_1.3.1        backports_1.2.0     circlize_0.4.12     plyr_1.8.6          splines_4.0.2      
  [6] digest_0.6.27       htmltools_0.5.1.1   fansi_0.4.1         cluster_2.1.0       openxlsx_4.2.3     
 [11] modelr_0.1.8        matrixStats_0.58.0  colorspace_2.0-0    rvest_1.0.0         haven_2.3.1        
 [16] xfun_0.22           crayon_1.3.4        jsonlite_1.7.1      zoo_1.8-9           glue_1.4.2         
 [21] gtable_0.3.0        GetoptLong_1.0.5    car_3.0-10          shape_1.4.5         BiocGenerics_0.36.0
 [26] maps_3.3.0          abind_1.4-5         DBI_1.1.1           rstatix_0.7.0       Rcpp_1.0.5         
 [31] xtable_1.8-4        clue_0.3-59         foreign_0.8-80      mapproj_1.2.7       km.ci_0.5-2        
 [36] stats4_4.0.2        htmlwidgets_1.5.3   httr_1.4.2          ellipsis_0.3.2      farver_2.1.0       
 [41] pkgconfig_2.0.3     sass_0.4.0          dbplyr_2.1.0        utf8_1.1.4          labeling_0.4.2     
 [46] tidyselect_1.1.0    reshape2_1.4.4      munsell_0.5.0       cellranger_1.1.0    tools_4.0.2        
 [51] cli_3.4.1           generics_0.1.0      sjlabelled_1.1.8    broom_0.7.5         evaluate_0.14      
 [56] yaml_2.2.1          fs_1.5.2            zip_2.1.1           survMisc_0.5.5      egg_0.4.5          
 [61] xml2_1.3.2          compiler_4.0.2      rstudioapi_0.13     beeswarm_0.4.0      curl_4.3           
 [66] png_0.1-7           ggsignif_0.6.1      reprex_1.0.0        statmod_1.4.35      bslib_0.2.5.1      
 [71] stringi_1.5.3       lattice_0.20-44     nloptr_1.2.2.2      KMsurv_0.1-5        vctrs_0.3.8        
 [76] pillar_1.6.1        lifecycle_1.0.0     jquerylib_0.1.4     GlobalOptions_0.1.2 ucminf_1.1-4       
 [81] insight_0.16.0      data.table_1.14.0   R6_2.5.0            rio_0.5.26          vipor_0.4.5        
 [86] IRanges_2.24.1      codetools_0.2-16    dichromat_2.0-0     boot_1.3-25         MASS_7.3-51.6      
 [91] assertthat_0.2.1    rjson_0.2.20        withr_2.3.0         S4Vectors_0.28.1    hms_1.0.0          
 [96] minqa_1.2.4         snakecase_0.11.0    rmarkdown_2.7       carData_3.0-4       Cairo_1.5-12.2     
[101] numDeriv_2016.8-1.1 lubridate_1.7.10   

```{r, eval=F, echo=F}
library(knitr)
options(knitr.duplicate.label = "allow")
rmarkdown::render("~/conv/conv_cytof_result.Rmd", rmarkdown::html_document(pandoc_args = "--self-contained", toc = T))
```
