
---
title: "Serum auto-antibody against IFN's data analysis for convalescent clinical manuscript: case control analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

This is including phases 1-3 data

### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ggnetwork"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "GSA"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))
suppressPackageStartupMessages(library(package = "gtsummary"))

# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```

### Load codebase.R, load data matrices and set path to your local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE}
# load codebase.R, serum-rbd-abtiters data and clinical information
# source("~/bitbucket/impacc/Analysis/data_analysis_template_codebase.R")
source("/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/data_analysis_template_codebase.R")
data_env <- load_IMPACC_datasets(DATA_VERSION                = "current", 
                                 KEEP_COVID19_POS            = TRUE, 
                                 FILTER_BY_CORE_ASSAY_COHORT = FALSE,
                                 USE_LOCKED_CLINICAL = TRUE)
for (n in grep(pattern = "serum_autoantibody|clinical_data|nasal_viralload_counts|nasal_viralload_rowfeature|serum_rbd_abtiters_counts|serum_rbd_abtiters_rowfeature", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}


# Setup the assay specific output directory
# out_dir = "../output"



```


#### Read clinical convalescent TGs data

```{r PrepareViralogyData, echo=FALSE}

# read in the convalescent TGs
clinical_virology_conv_file <- file.path("/scratch/clinical/current", "impacc-convalescent.csv")
clinical_virology_conv_data <- read.csv(clinical_virology_conv_file)
  
# add convalescent trajectory groups
clinical_data <- clinical_data %>%
  # merge convalescent TGs
  inner_join(clinical_virology_conv_data) %>%
  # rename(trajectory_group_original = trajectory_group) %>%
  mutate(conv_cluster4 = recode(conv_cluster6, `1` = 1, `3` = 1, `4` = 1, `2` = 2, `5` = 5, `6` = 6)) %>%
  mutate(conv_cluster4 = factor(conv_cluster4, levels = c("1", "2", "5", "6")))

```



&nbsp;

&nbsp;

#### Combine clinical data and serum auto-antobody data

```{r Combine clinical data and serum auto-antobody data, echo=FALSE}

serum_autoantibody_metadata <- read.csv(file = file.path("/data/serum-autoantibody/legacy/2023-10-10", "SerumAutoantibody-Metadata.csv"))

serum_autoantibody_metadata_select <- serum_autoantibody_metadata %>%
  # rownames_to_column(var = "sample_id") %>%
  filter(sample_status == "IMPACC sample assayed and passed QC") %>%
  select(sample_id, 
         block.IFNa = positive.for.blocking.anti.IFNa.,
         block.IFNb = positive.for.blocking.anti.IFNb.,
         block.IFNw = positive.for.blocking.anti.IFNw.,
         block.IFNabw = positive.for.blocking.autoantibodies.against.at.least.one.type.I.IFN..at.earliest.timepoint..) %>%
  replace_na(list(block.IFNa = "no", block.IFNb = "no", block.IFNw = "no"))


clinical_autoantibody_meta_data <- clinical_data %>%
  inner_join(serum_autoantibody_metadata_select, by = "sample_id") 

# First select 1 earliest sample for each donor based on event date
sample_per_partcipant <- clinical_autoantibody_meta_data %>%
  arrange(participant_id, event_date, event_type) %>%
  distinct(participant_id, .keep_all = TRUE) %>%
  select(sample_id, participant_id, enrollment_site, sex, race, ethnicity, discretized_admit_age_quantile, event_date, event_type, conv_cluster3, conv_cluster6, conv_cluster4, block.IFNa, block.IFNb, block.IFNw, block.IFNabw) 

IFNabw_pos_partcipants_list <- sample_per_partcipant %>%
  filter(block.IFNabw == "yes")
```

&nbsp;

&nbsp;

#### Read case control participants data

```{r Read case control participants data, echo=FALSE}

case_contol_data <- read.csv(file = "autoantibody_matches.csv")

# just make sure that 24 participants are the same from this file and my data
length(intersect(case_contol_data$case_id, IFNabw_pos_partcipants_list$participant_id)) #23

# CDCC said that they can't find a controls for 1 participant due his/her age is 96 years

case_df <- data.frame(participant_id = unique(case_contol_data$case_id), participant_category = "case")
```

&nbsp;

&nbsp;

#### Prepare nasal-viral load (qPCR) data

```{r PrepareNasalViralLoadData, echo=FALSE}

# counts data
nasal_viral_qPCR_data <- nasal_viralload_counts %>%
    tibble::rownames_to_column(var = "sample_id") %>%
    # select only relevant columns
    dplyr::select(sample_id, N1_CT, N2_CT, N1_CTRL, N2_CTRL, 
           RP_CT, RP_CTRL, HPRT_CT, HPRT_CTRL) %>% as.data.frame()

clinical_viralload_data <- nasal_viral_qPCR_data %>% 
  inner_join(clinical_data, by = c("sample_id"))

# replace NA (undetermined) to 40
imputeCT <- 40
clinical_viralload_data <- clinical_viralload_data %>% 
  mutate(N1_CT = replace_na(N1_CT, imputeCT), 
                N2_CT = replace_na(N2_CT, imputeCT), 
                RP_CT = replace_na(RP_CT, imputeCT),
                HPRT_CT = replace_na(HPRT_CT, imputeCT)) %>% 
  # replace values > 40 to 40
  mutate(N1_CT = if_else(N1_CT > imputeCT, imputeCT, N1_CT),
         N2_CT = if_else(N2_CT > imputeCT, imputeCT, N2_CT),
         RP_CT = if_else(RP_CT > imputeCT, imputeCT, RP_CT),
         HPRT_CT = if_else(HPRT_CT > imputeCT, imputeCT, HPRT_CT)) 
# remove samples with undetected RP_CT values
clinical_viralload_data <- clinical_viralload_data %>% 
  filter(RP_CT != 40)

# add days from symptom onset, "symptom_onset_days" as event_date - symptom_date
clinical_viralload_data <- clinical_viralload_data %>%
  mutate(symptom_onset_days = event_date - symptom_date)

clinical_viralload_data_filtered <- clinical_viralload_data %>%
  filter(participant_id %in% c(case_contol_data$control_id, case_contol_data$case_id)) %>%
  mutate(case_control = if_else(participant_id %in% case_contol_data$case_id, "Case", "Control"))

clinical_viralload_data_1_sample_per_participant <- clinical_viralload_data_filtered %>%
  arrange(participant_id, event_date, event_type) %>%
  distinct(participant_id, .keep_all = TRUE) 
  
table(clinical_viralload_data_1_sample_per_participant$conv_cluster4, clinical_viralload_data_1_sample_per_participant$case_control)

# save source data
source_data <- clinical_viralload_data_1_sample_per_participant %>%
  mutate(endpoints = recode(conv_cluster4, `1` = "MIN", `2` = "PHY", `5` = "COG", `6` = "MLT")) %>%
  # tibble::rownames_to_column(var = "sample_id") %>%
  select(sample_id, participant_id, enrollment_site, event_type, discretized_admit_age_quantile, sex, endpoints, case_control, N1_CT, N2_CT)

file_name_1 <- "/scratch/nasal-transcriptomics/impacc/Analysis/serum-autoantibody-IFN/convalescent_clinical/source_data/Source_data_serum_autoantibody_IFN_viralload_case_control_analysis.txt"
write.table(source_data, file = file_name_1, quote = FALSE, sep = "\t", row.names = FALSE)



```


&nbsp;

&nbsp;

&nbsp;

#### Viral load early visit analysis figures: case control classification
```{r PlotViralDataByClinicalTrajectoryGroupV2, fig.width = 8, fig.height = 8, warning=FALSE, echo=FALSE, eval=TRUE}

CaseControlCol <- c("Case" = "#0072B2", "Control" = "#009E73")
# TrajGrpCol <- c("1" = "#639A21", "2" = "#39828C", "5" = "#9C3418", "6" = "pink")

library(ggbeeswarm)

library(ggprism)

df_visit1_mean <- clinical_viralload_data_1_sample_per_participant %>%
  group_by(case_control) %>%
  summarise(n = n(),
            N1_CT_mean = mean(N1_CT),
            N2_CT_mean = mean(N2_CT))

## N1_CT figures
df_p_val <- rstatix::t_test(clinical_viralload_data_1_sample_per_participant, N1_CT ~ case_control) %>% rstatix::add_x_position() %>%
  mutate(p = round(p, 3)) %>%
  mutate(`y.position` = 10)

boxplot_N1 <- ggplot(clinical_viralload_data_1_sample_per_participant, aes(x = case_control, y = N1_CT)) + 
  geom_boxplot(aes(x = case_control, y = N1_CT, color = case_control), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = case_control, y = N1_CT, color = case_control), width = 0.2, alpha = 0.7, size = 1) +
  theme_bw() +
  ylab("N1 Ct") + xlab("") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  scale_color_manual(values = CaseControlCol) +
  add_pvalue(df_p_val,
             label = "p",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('Cases', 'Controls'))

boxplot_N1


ggsave(
  filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/serum-autoantibody-IFN/convalescent_clinical/Figures/%y%m%d-N1_CT_IFN_autoantibody_boxplot_cases_controls.pdf"),
  width = 2.5,
  height = 3
)


## N2_CT figures
df_p_val <- rstatix::t_test(clinical_viralload_data_1_sample_per_participant, N2_CT ~ case_control) %>% rstatix::add_x_position() %>%
  mutate(p = round(p, 3)) %>%
  mutate(`y.position` = 10)

boxplot_N2 <- ggplot(clinical_viralload_data_1_sample_per_participant, aes(x = case_control, y = N2_CT)) + 
  geom_boxplot(aes(x = case_control, y = N2_CT, color = case_control), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = case_control, y = N2_CT, color = case_control), width = 0.2, alpha = 0.7, size = 1) +
  theme_bw() +
  ylab("N2 Ct") + xlab("") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  scale_color_manual(values = CaseControlCol) +
  add_pvalue(df_p_val,
             label = "p",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('Cases', 'Controls'))

boxplot_N2


ggsave(
  filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/serum-autoantibody-IFN/convalescent_clinical/Figures/%y%m%d-N2_CT_IFN_autoantibody_boxplot_cases_controls.pdf"),
  width = 2.5,
  height = 3
)








```


#### Prepare serum rbd titers data

```{r PrepareSerumRBDTitersData, echo=FALSE}

# counts data
serum_rbd_raw_data <- serum_rbd_abtiters_counts %>% 
  tibble::rownames_to_column(var = "sample_id") %>% 
  # select only relevant columns 
  dplyr::select(sample_id, 
                AUC.RBD.IgG = `AUC RBD IgG`,
                AUC.Spike.IgG = `AUC Spike IgG`) %>%
  as.data.frame()

clinical_titers_counts_data <- clinical_data %>%
  inner_join(serum_rbd_raw_data, by = "sample_id") %>%
  mutate(symptom_onset_days = event_date - symptom_date)

clinical_titers_counts_data_filtered <- clinical_titers_counts_data %>%
  filter(participant_id %in% c(case_contol_data$control_id, case_contol_data$case_id)) %>%
  mutate(case_control = if_else(participant_id %in% case_contol_data$case_id, "Case", "Control")) %>%
  mutate(RBD.IgG = log2(AUC.RBD.IgG),
         Spike.IgG = log2(AUC.Spike.IgG))

clinical_titers_counts_data_1_sample_per_participant <- clinical_titers_counts_data_filtered %>%
  arrange(participant_id, event_date, event_type) %>%
  distinct(participant_id, .keep_all = TRUE)
  
table(clinical_titers_counts_data_1_sample_per_participant$conv_cluster4, clinical_titers_counts_data_1_sample_per_participant$case_control)


# save source data
source_data <- clinical_titers_counts_data_1_sample_per_participant %>%
  mutate(endpoints = recode(conv_cluster4, `1` = "MIN", `2` = "PHY", `5` = "COG", `6` = "MLT")) %>%
  # tibble::rownames_to_column(var = "sample_id") %>%
  select(sample_id, participant_id, enrollment_site, event_type, discretized_admit_age_quantile, sex, endpoints, case_control, RBD.IgG, Spike.IgG)

file_name_1 <- "/scratch/nasal-transcriptomics/impacc/Analysis/serum-autoantibody-IFN/convalescent_clinical/source_data/Source_data_serum_autoantibody_IFN_RBD_titers_case_control_analysis.txt"
write.table(source_data, file = file_name_1, quote = FALSE, sep = "\t", row.names = FALSE)




```

&nbsp;

&nbsp;

&nbsp;

#### RBD titers early visit analysis figures: case control classification
```{r PlotViralDataByClinicalTrajectoryGroupV2, fig.width = 8, fig.height = 8, warning=FALSE, echo=FALSE, eval=TRUE}

CaseControlCol <- c("Case" = "#0072B2", "Control" = "#009E73")
# TrajGrpCol <- c("Case" = "blue", "Control" = "#83858a")

library(ggbeeswarm)

## RBD.IgG figures

df_visit1_mean <- clinical_titers_counts_data_1_sample_per_participant %>%
  group_by(case_control) %>%
  summarise(n = n(),
            RBD.IgG_mean = mean(RBD.IgG),
            Spike.IgG_mean = mean(Spike.IgG))


library(ggprism)

df_p_val <- rstatix::t_test(clinical_titers_counts_data_1_sample_per_participant, RBD.IgG ~ case_control) %>% rstatix::add_x_position() %>%
  mutate(p = round(p, 3)) %>%
  mutate(`y.position` = 18)

boxplot_RBD.IgG <- ggplot(clinical_titers_counts_data_1_sample_per_participant, aes(x = case_control, y = RBD.IgG)) + 
  geom_boxplot(aes(x = case_control, y = RBD.IgG, color = case_control), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = case_control, y = RBD.IgG, color = case_control), width = 0.2, alpha = 0.7, size = 1) +
  theme_bw() +
  ylab("anti-RBD IgG") + xlab("") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  scale_color_manual(values = CaseControlCol) +
  add_pvalue(df_p_val,
             label = "p",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  # coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('Cases', 'Controls'))

boxplot_RBD.IgG


ggsave(
  filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/serum-autoantibody-IFN/convalescent_clinical/Figures/%y%m%d-RBD_IgG_IFN_autoantibody_boxplot_cases_controls.pdf"),
  width = 2.5,
  height = 3
)


## Spike.IgG figures
wilcox.test(Spike.IgG ~ case_control, clinical_titers_counts_data_1_sample_per_participant)$p.value

# for some reason this function computes P-values slightly different from wilcox test
df_p_val <- rstatix::t_test(clinical_titers_counts_data_1_sample_per_participant, Spike.IgG ~ case_control) %>% rstatix::add_x_position() %>%
  mutate(p = wilcox.test(Spike.IgG ~ case_control, clinical_titers_counts_data_1_sample_per_participant)$p.value) %>%
  mutate(p = round(p, 3)) %>%
  mutate(`y.position` = 18) 

boxplot_Spike.IgG <- ggplot(clinical_titers_counts_data_1_sample_per_participant, aes(x = case_control, y = Spike.IgG)) + 
  geom_boxplot(aes(x = case_control, y = Spike.IgG, color = case_control), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = case_control, y = Spike.IgG, color = case_control), width = 0.2, alpha = 0.7, size = 1) +
  theme_bw() +
  ylab("anti-Spike IgG") + xlab("") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  scale_color_manual(values = CaseControlCol) +
  add_pvalue(df_p_val,
             label = "p",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  # coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('Cases', 'Controls'))

boxplot_Spike.IgG


ggsave(
  filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/serum-autoantibody-IFN/convalescent_clinical/Figures/%y%m%d-Spike_IgG_IFN_autoantibody_boxplot_cases_controls.pdf"),
  width = 2.5,
  height = 3
)




## Spike.IgG figures

boxplot_Spike <- ggplot(clinical_titers_counts_data_1_sample_per_participant, aes(x = case_control, y = Spike.IgG)) + 
  geom_quasirandom(aes(x = case_control, y = Spike.IgG, color = case_control), alpha = 0.3) +
  geom_boxplot(aes(x = case_control, y = Spike.IgG, fill = case_control), size = 1, alpha = 0.3, outlier.shape = NA) +
  geom_point(data = df_visit1_mean, aes(x = case_control, y = Spike.IgG_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("Anti-Spike IgG") + xlab("Case control groups") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=16)) +
  scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  stat_compare_means(method = "wilcox.test", label.y = 17)
  # add_pvalue(res_table_pairwise_long_df_temp, 
  #            label = "Pvalue")

boxplot_Spike


```



### Session info
```{r, print-session-info}
sessionInfo()
```
<!--
R version 4.0.2 (2020-06-22)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 18.04.4 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so

locale:
 [1] LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8    
 [5] LC_MONETARY=C.UTF-8    LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C             
 [9] LC_ADDRESS=C           LC_TELEPHONE=C         LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

attached base packages:
[1] parallel  grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] gtsummary_1.5.2       doParallel_1.0.17     doRNG_1.8.2           rngtools_1.5.2       
 [5] missForest_1.4        itertools_0.1-3       iterators_1.0.14      foreach_1.5.2        
 [9] randomForest_4.6-10   ComplexHeatmap_2.11.1 corrr_0.4.3           gridExtra_2.3        
[13] rlang_1.0.1           RColorBrewer_1.1-2    cowplot_1.1.1         ggpubr_0.4.0         
[17] pals_1.7              forcats_0.5.1         stringr_1.4.0         dplyr_1.0.8          
[21] purrr_0.3.4           readr_2.1.2           tidyr_1.2.0           tibble_3.1.6         
[25] tidyverse_1.3.1       nlme_3.1-147          lme4_1.1-28           Matrix_1.2-18        
[29] GSA_1.03.2            pvca_0.1.0            ggeffects_1.1.1       ggbeeswarm_0.6.0     
[33] qvalue_2.22.0         ordinal_2019.12-10    ggnetwork_0.5.10      ggplot2_3.3.5        
[37] igraph_1.3.0          impute_1.64.0         knitr_1.37           

loaded via a namespace (and not attached):
 [1] minqa_1.2.4         colorspace_2.0-3    ggsignif_0.6.3      rjson_0.2.21       
 [5] ellipsis_0.3.2      circlize_0.4.14     GlobalOptions_0.1.2 fs_1.5.2           
 [9] dichromat_2.0-0     clue_0.3-60         rstudioapi_0.13     DT_0.15            
[13] fansi_1.0.2         lubridate_1.8.0     xml2_1.3.2          codetools_0.2-16   
[17] splines_4.0.2       jsonlite_1.7.3      nloptr_2.0.0        gt_0.4.0           
[21] broom_0.7.12        cluster_2.1.0       dbplyr_2.1.1        png_0.1-7          
[25] mapproj_1.2.8       compiler_4.0.2      httr_1.4.2          backports_1.4.1    
[29] assertthat_0.2.1    fastmap_1.1.0       cli_3.2.0           htmltools_0.5.2    
[33] tools_4.0.2         gtable_0.3.0        glue_1.6.1          reshape2_1.4.4     
[37] maps_3.4.0          Rcpp_1.0.8          carData_3.0-5       cellranger_1.1.0   
[41] vctrs_0.3.8         broom.helpers_1.6.0 xfun_0.29           rvest_1.0.2        
[45] lifecycle_1.0.1     rstatix_0.7.0       MASS_7.3-51.6       scales_1.1.1       
[49] hms_1.1.1           yaml_2.3.5          sass_0.4.0          stringi_1.7.6      
[53] ucminf_1.1-4        S4Vectors_0.28.1    checkmate_2.0.0     BiocGenerics_0.36.1
[57] boot_1.3-25         shape_1.4.6         commonmark_1.7      pkgconfig_2.0.3    
[61] matrixStats_0.61.0  evaluate_0.14       lattice_0.20-41     htmlwidgets_1.5.4  
[65] tidyselect_1.1.2    plyr_1.8.6          magrittr_2.0.2      R6_2.5.1           
[69] IRanges_2.24.1      generics_0.1.2      DBI_1.1.2           pillar_1.7.0       
[73] haven_2.4.3         withr_2.4.3         abind_1.4-5         modelr_0.1.8       
[77] crayon_1.5.0        car_3.0-12          utf8_1.2.2          tzdb_0.2.0         
[81] rmarkdown_2.11      GetoptLong_1.0.5    readxl_1.3.1        reprex_2.0.1       
[85] digest_0.6.29       numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
[89] beeswarm_0.4.0      vipor_0.4.5
-->



