---
title: "Integrating nasal viral load data with clinical data: convalescent PRO clusters"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ggnetwork"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "GSA"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))

# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```

### Load codebase.R, load data matrices and set path to your local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE}
# load codebase.R, serum-rbd-abtiters data and clinical information
# source("~/bitbucket/impacc/Analysis/data_analysis_template_codebase.R")
source("/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/data_analysis_template_codebase.R")
source("/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/clinical/codebase_conv_immune.R")
source("/users/ndjayavelu/njayavelu/Scripts/plot_model_smooth_conv.R")

data_env <- load_IMPACC_datasets(DATA_VERSION               = "current",
                                KEEP_COVID19_POS            = FALSE,
                                FILTER_BY_CORE_ASSAY_COHORT = FALSE,
                                USE_LOCKED_CLINICAL = TRUE)

for (n in grep(pattern = "nasal_viralload_counts|nasal_viralload_rowfeature|clinical_data", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}



```


&nbsp;

&nbsp;

&nbsp;


#### Read clinical PRO clusters data

```{r PrepareViralogyData, echo=FALSE}

# read in the convalescent PRO clusters
clinical_virology_conv_file <- file.path("/scratch/clinical/current", "impacc-convalescent.csv")
clinical_virology_conv_data <- read.csv(clinical_virology_conv_file)
  
# add convalescent PRO clusters
clinical_virology_data <- clinical_data %>%
  inner_join(clinical_virology_conv_data)

```

&nbsp;

&nbsp;

&nbsp;

#### Prepare nasal-viral load (qPCR) data

```{r PrepareNasalViralLoadData, echo=FALSE}

# counts data
nasal_viral_qPCR_data <- nasal_viralload_counts %>%
    tibble::rownames_to_column(var = "sample_id") %>%
    # select only relevant columns
    dplyr::select(sample_id, N1_CT, N2_CT, N1_CTRL, N2_CTRL, 
           RP_CT, RP_CTRL, HPRT_CT, HPRT_CTRL) %>% as.data.frame()

```

&nbsp;

&nbsp;

&nbsp;

#### Combine clinical data and nasal-viral load data

```{r CombineClinicalNasalViralLoadData, echo=FALSE}

clinical_viralload_data <- nasal_viral_qPCR_data %>% 
  inner_join(clinical_virology_data, by = c("sample_id")) 

# replace NA (undetermined) to 40
imputeCT <- 40
clinical_viralload_data <- clinical_viralload_data %>% 
  mutate(N1_CT = replace_na(N1_CT, imputeCT), 
                N2_CT = replace_na(N2_CT, imputeCT), 
                RP_CT = replace_na(RP_CT, imputeCT),
                HPRT_CT = replace_na(HPRT_CT, imputeCT)) %>% 
  # replace values > 40 to 40
  mutate(N1_CT = if_else(N1_CT > imputeCT, imputeCT, N1_CT),
         N2_CT = if_else(N2_CT > imputeCT, imputeCT, N2_CT),
         RP_CT = if_else(RP_CT > imputeCT, imputeCT, RP_CT),
         HPRT_CT = if_else(HPRT_CT > imputeCT, imputeCT, HPRT_CT)) 
# remove samples with undetected RP_CT values
clinical_viralload_data <- clinical_viralload_data %>% 
  filter(RP_CT != 40)

# add days from symptom onset, "symptom_onset_days" as event_date - symptom_date
clinical_viralload_data <- clinical_viralload_data %>%
  mutate(symptom_onset_days = event_date - symptom_date)


```


&nbsp;

&nbsp;

&nbsp;


### Pooled analysis code
```{r pooled analysis code, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

data_use_pooled_4TG <- clinical_viralload_data %>%
  mutate(conv_cluster4 = recode(conv_cluster6, `1` = 1, `3` = 1, `4` = 1, `2` = 2, `5` = 5, `6` = 6)) %>%
  select(sample_id, N1_CT, N2_CT, 
         enrollment_site,
         conv_cluster4,
         participant_id,
         event_type,
         event_date,
         discretized_admit_age_quantile,
         sex = sex) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(conv_cluster4 = factor(conv_cluster4, levels = c("1", "2", "5", "6"))) %>%
  column_to_rownames(var = "sample_id")

# save source data
source_data <- data_use_pooled_4TG %>%
  mutate(endpoints = recode(conv_cluster4, `1` = "MIN", `2` = "PHY", `5` = "COG", `6` = "MLT")) %>%
  tibble::rownames_to_column(var = "sample_id") %>%
  select(sample_id, participant_id, enrollment_site, event_type, discretized_admit_age_quantile, sex, endpoints, N1_CT, N2_CT)

file_name_1 <- "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/source_data/Source_data_Viralload_pooled_analysis.txt"
write.table(source_data, file = file_name_1, quote = FALSE, sep = "\t", row.names = FALSE)


immune_assay <- data_use_pooled_4TG %>%
  select(N1_CT, N2_CT)
immune_meta <- data_use_pooled_4TG %>% 
  tibble::rownames_to_column(var = "sample_id") %>%
  select(sample_id, 
         participant_id,
         event_type,
         event_date,
         discretized_admit_age_quantile,
         sex,
         enrollment_site,
         conv_cluster4)


if(!(all(rownames(immune_assay) == immune_meta$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}


response_name = "conv_cluster4"
random_effect_list = c("participant_id", "enrollment_site")
fixed_effect_list = c("sex", "discretized_admit_age_quantile")
# fixed_effect_list = c("event_type","sex", "discretized_admit_age_quantile")

# over all association

module_names <- c("N1_CT", "N2_CT")

pval_overall <- NULL

for (moduleName in module_names) {
  my.formula0 <- paste0(moduleName, '~(1|enrollment_site)+(1|participant_id)+discretized_admit_age_quantile+sex')
  my.formula1 <- paste0(moduleName,
                        '~(1|enrollment_site)+(1|participant_id)+discretized_admit_age_quantile+sex+conv_cluster4')
  av = anova(lme4::lmer(my.formula0, data = data_use_pooled_4TG),
                 lme4::lmer(my.formula1, data = data_use_pooled_4TG))
  res <- av$`Pr(>Chisq)`[2]
  pval_overall <- rbind(pval_overall, res)
}

res_table_overall <- data.frame(feature = module_names, 
                                pval = as.vector(pval_overall)) %>% 
  mutate(qval =  qvalue::qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues) %>%
  tibble::column_to_rownames(var = "feature")


# pairwise comparisons
pooled_resDF_pairwise <- NULL
for(feature_name in colnames(immune_assay)){
  res0 = pooled_test(immune_assay=immune_assay, immune_meta=immune_meta, 
                     feature_name = feature_name, response_name=response_name, 
                     random_effect_list=random_effect_list , fix_effect_list=fixed_effect_list)
  pooled_resDF_pairwise <- rbind(pooled_resDF_pairwise,res0)
}
rownames(pooled_resDF_pairwise) = colnames(immune_assay)

res_table_pooled_all <- cbind(res_table_overall, pooled_resDF_pairwise)


```

&nbsp;

&nbsp;

&nbsp;

#### Pooled analysis figures: 4 cluster convalescent TGs classification
```{r PlotViralDataByClinicalTrajectoryGroupV2, fig.width = 8, fig.height = 8, warning=FALSE, echo=FALSE, eval=TRUE}

TrajGrpCol <- c("1" = "#999999", "2" = "#56B4E9", "5" = "#E69F00", "6" = "#FF0000")
# col_sel <- c("MIN" = "#999999", "PHY" = "#56B4E9", "COG" = "#E69F00", "MLT" = "#FF0000")

library(ggbeeswarm)

res_table_pairwise_long_df <- pooled_resDF_pairwise %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "feature") %>%
  pivot_longer(cols = -c("feature")) %>%
  rename(Pvalue = value) %>%
  tidyr::separate(col = "name", into = c("group1", "group2"), remove = FALSE)

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "N1_CT") %>%
  filter(Pvalue < 0.05) %>%
  # mutate(Pvalue = round(Pvalue,4)) %>%
  select(group1, group2, Pvalue) %>%
  mutate(`y.position` = c(9,11)) %>%
  mutate(label = case_when(
  Pvalue < 0.001 ~ "***",
  (Pvalue > 0.001 & Pvalue < 0.01) ~ "**",
  (Pvalue > 0.01 & Pvalue < 0.05) ~ "*",
  .default = "NS")) %>%
  as.tibble()
  
  
library(ggprism)


## N1_CT figures

df_pooled_mean <- data_use_pooled_4TG %>%
  group_by(conv_cluster4) %>%
  summarise(n = n(),
            N1_CT_mean = mean(N1_CT),
            N2_CT_mean = mean(N2_CT))

boxplot_N1 <- ggplot(data_use_pooled_4TG, aes(x = conv_cluster4, y = N1_CT)) + 
  geom_boxplot(aes(x = conv_cluster4, y = N1_CT, color = conv_cluster4), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = conv_cluster4, y = N1_CT, color = conv_cluster4), width = 0.2, alpha = 0.7, size = 1) +
  # geom_point(data = df_pooled_mean, aes(x = conv_cluster4, y = N1_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("N1 Ct") + xlab("") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  # scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "label",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 6) +
  coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('MIN', 'PHY', 'COG', 'MLT'))

boxplot_N1


ggsave(
  filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/Figures/%y%m%d-N1_CT_pooled_boxplot_PRO_clusters.pdf"),
  width = 2.5,
  height = 3
)



## N2_CT figures

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "N2_CT") %>%
  filter(Pvalue < 0.05) %>%
  # mutate(Pvalue = round(Pvalue,4)) %>%
  select(group1, group2, Pvalue) %>%
  mutate(`y.position` = c(9,11)) %>%
  mutate(label = case_when(
  Pvalue < 0.001 ~ "***",
  (Pvalue > 0.001 & Pvalue < 0.01) ~ "**",
  (Pvalue > 0.01 & Pvalue < 0.05) ~ "*",
  .default = "NS")) %>%
  as.tibble()

boxplot_N2 <- ggplot(data_use_pooled_4TG, aes(x = conv_cluster4, y = N2_CT)) + 
  geom_boxplot(aes(x = conv_cluster4, y = N2_CT, color = conv_cluster4), size = 0.8, outlier.shape = NA) + 
  geom_quasirandom(aes(x = conv_cluster4, y = N2_CT, color = conv_cluster4), width = 0.2, alpha = 0.7, size = 1) +
  # geom_point(data = df_pooled_mean, aes(x = conv_cluster4, y = N1_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("N2 Ct") + xlab("") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10, color = "black"),
        axis.text.y = element_text(size = 10, color = "black"),
        text = element_text(size=10)) +
  # scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "label",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 6) +
  coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('MIN', 'PHY', 'COG', 'MLT'))

boxplot_N2


ggsave(
  filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/Figures/%y%m%d-N2_CT_pooled_boxplot_PRO_clusters.pdf"),
  width = 2.5,
  height = 3
)




```



### Longitudinal analysis code: 4 cluster convalescent TGs classification
```{r Longitudinal-analysis, message=FALSE, results="hide"}

data_use_longitudinal_4TG <- clinical_viralload_data %>%
  rename(trajectory_group_original = trajectory_group) %>%
  mutate(conv_cluster4 = recode(conv_cluster6, `1` = "MIN", `3` = "MIN", `4` = "MIN", `2` = "PHY", `5` = "COG", `6` = "MLT")) %>%
  rename(trajectory_group = conv_cluster4) %>%
  tidyr::drop_na(trajectory_group) %>%
  filter(event_date <= 28) %>%
  select(sample_id, N1_CT, N2_CT, participant_id,  
         enrollment_site,
         trajectory_group, event_date, event_type, sex, discretized_admit_age_quantile) %>% 
  mutate(participant_id = factor(participant_id)) %>% 
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group, levels = c("MIN", "PHY", "COG", "MLT"))) 
  
# save source data
source_data <- data_use_longitudinal_4TG %>%
  mutate(endpoints = trajectory_group) %>%
  # tibble::rownames_to_column(var = "sample_id") %>%
  select(sample_id, participant_id, enrollment_site, event_type, event_date, discretized_admit_age_quantile, sex, endpoints, N1_CT, N2_CT)

file_name_1 <- "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/source_data/Source_data_Viralload_longitudinal_analysis.txt"
write.table(source_data, file = file_name_1, quote = FALSE, sep = "\t", row.names = FALSE)



inputDF <- data_use_longitudinal_4TG %>%
  pivot_longer(cols = -c("sample_id", "event_date", "event_type", "participant_id",
                         "enrollment_site", "trajectory_group", "sex", "discretized_admit_age_quantile"
                         ))


# smooth spline
smooth_spline_model_loop_4TG <- model_loop(inputDF, 
                                       modelType = "smoothSpline", 
                                       endpoint = "trajectory_group")

smooth_spline_model_loop_df_4TG <- smooth_spline_model_loop_4TG %>%
  rownames_to_column(var = "feature")

# filename_long = format(Sys.Date(), "/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/nasal-viralload/clinGroup/Figures/conv_symps/%y%m%d-P_values_longitudinal_smoothSpline_with_age_sex_for_conv_symps.csv")
# write.csv(smooth_spline_model_loop_df, 
#            file = filename_long, row.names = FALSE)

```



### Longitudinal analysis figures: template code
```{r Longitudinal-analysis figures, message=FALSE, results="hide"}

plotExample <- inputDF %>%
  filter(name %in% "N1_CT") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)),
                                    levels = c("MIN", "PHY", "COG", "MLT"))) 

# col_sel <- c("MIN" = "#999999", "PHY" = "#CC79A7", "COG" = "#0072B2", "MLT" = "#D55E00")
# col_sel <- c("MIN" = "#009E73", "PHY" = "#CC79A7", "COG" = "#0072B2", "MLT" = "#D55E00")
# col_sel <- c("MIN" = "#009E73", "PHY" = "#CC79A7", "COG" = "#56B4E9", "MLT" = "#E69F00")

# from wesanderson colors
# col_sel <- c("MIN" = "#899DA4", "PHY" = "#CC79A7", "COG" = "#56B4E9", "MLT" = "#C93312")
# col_sel <- c("MIN" = "#74A089", "PHY" = "#9A8822", "COG" = "#56B4E9", "MLT" = "#C93312")
# col_sel <- c("MIN" = "#00A08A", "PHY" = "#CC79A7", "COG" = "#56B4E9", "MLT" = "#FF0000")

# final selection
col_sel <- c("MIN" = "#999999", "PHY" = "#56B4E9", "COG" = "#E69F00", "MLT" = "#FF0000")

temp_plot <- plot_model_smooth_conv(plotExample, model_loop = smooth_spline_model_loop_4TG, modelType = "smoothSpline",
           p_adjust = F, endpoint = "trajectory_group", signif_markers = T, remove_NS = T,
           title = "",
           individual_points_size = 1,
           individual_paths_size = 0.5,
           group_trendline_line_width = 1.5,
           title_size = 12,
           bar_height = 5,
           custom_theme_graph = 'text = element_text(size=10)',
           # custom_theme_bars = 'text = element_text(size=6)',
           knot_lines = F,
           ylabel = "N1 Ct",
           CI = FALSE,
           individual_trendlines = T, individual_points = T, individual_paths = T,
           p_value_text_size = 6,
           return_multi_obj = F,
           colors = col_sel,
           y_axis_reverse = TRUE,
           trajectory_labels = c("MIN", "PHY", "COG", "MLT"))
 

print(temp_plot)
ggsave(filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/Figures/%y%m%d-N1_CT_longitudinal_PRO_clusters.pdf"),
         plot = temp_plot,
         width = 3.5,
         height = 2.5)

# N2 CT
plotExample <- inputDF %>%
  filter(name %in% "N2_CT") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)),
                                    levels = c("MIN", "PHY", "COG", "MLT"))) 
temp_plot <- plot_model_smooth_conv(plotExample, model_loop = smooth_spline_model_loop_4TG, modelType = "smoothSpline",
           p_adjust = F, endpoint = "trajectory_group", signif_markers = T, remove_NS = T,
           title = "",
           individual_points_size = 1,
           individual_paths_size = 0.5,
           group_trendline_line_width = 1.5,
           title_size = 12,
           bar_height = 5,
           custom_theme_graph = 'text = element_text(size=10)',
           # custom_theme_bars = 'text = element_text(size=6)',
           knot_lines = F,
           ylabel = "N2 Ct",
           CI = FALSE,
           individual_trendlines = T, individual_points = T, individual_paths = T,
           p_value_text_size = 6,
           return_multi_obj = F,
           colors = col_sel,
           y_axis_reverse = TRUE,
           trajectory_labels = c("MIN", "PHY", "COG", "MLT"))
 


print(temp_plot)
ggsave(filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/Figures/%y%m%d-N2_CT_longitudinal_PRO_clusters.pdf"),
         plot = temp_plot,
         width = 3.5,
         height = 2.5)

```









