---
title: "Integrating nasal viral load data with clinical data: convalescent PRO clusters"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ggnetwork"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "GSA"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))

# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```

### Load codebase.R, load data matrices and set path to your local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE}
# load codebase.R, serum-rbd-abtiters data and clinical information
# source("~/bitbucket/impacc/Analysis/data_analysis_template_codebase.R")
source("/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/data_analysis_template_codebase.R")
source("/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/clinical/codebase_conv_immune.R")
source("/users/ndjayavelu/njayavelu/Scripts/plot_model_smooth_conv.R")

data_env <- load_IMPACC_datasets(DATA_VERSION               = "current",
                                KEEP_COVID19_POS            = FALSE,
                                FILTER_BY_CORE_ASSAY_COHORT = FALSE)

for (n in grep(pattern = "nasal_viralload_counts|nasal_viralload_rowfeature|clinical_data", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}



```


&nbsp;

&nbsp;

&nbsp;


#### Read clinical PRO clusters data

```{r PrepareViralogyData, echo=FALSE}

# read in the convalescent PRO clusters
clinical_virology_conv_file <- file.path("/scratch/clinical/current", "impacc-convalescent.csv")
clinical_virology_conv_data <- read.csv(clinical_virology_conv_file)
  
# add convalescent PRO clusters
clinical_virology_data <- clinical_data %>%
  inner_join(clinical_virology_conv_data)

```

&nbsp;

&nbsp;

&nbsp;

#### Prepare nasal-viral load (qPCR) data

```{r PrepareNasalViralLoadData, echo=FALSE}

# counts data
nasal_viral_qPCR_data <- nasal_viralload_counts %>%
    tibble::rownames_to_column(var = "sample_id") %>%
    # select only relevant columns
    dplyr::select(sample_id, N1_CT, N2_CT, N1_CTRL, N2_CTRL, 
           RP_CT, RP_CTRL, HPRT_CT, HPRT_CTRL) %>% as.data.frame()

```

&nbsp;

&nbsp;

&nbsp;

#### Combine clinical data and nasal-viral load data

```{r CombineClinicalNasalViralLoadData, echo=FALSE}

clinical_viralload_data <- nasal_viral_qPCR_data %>% 
  inner_join(clinical_virology_data, by = c("sample_id")) 

# replace NA (undetermined) to 40
imputeCT <- 40
clinical_viralload_data <- clinical_viralload_data %>% 
  mutate(N1_CT = replace_na(N1_CT, imputeCT), 
                N2_CT = replace_na(N2_CT, imputeCT), 
                RP_CT = replace_na(RP_CT, imputeCT),
                HPRT_CT = replace_na(HPRT_CT, imputeCT)) %>% 
  # replace values > 40 to 40
  mutate(N1_CT = if_else(N1_CT > imputeCT, imputeCT, N1_CT),
         N2_CT = if_else(N2_CT > imputeCT, imputeCT, N2_CT),
         RP_CT = if_else(RP_CT > imputeCT, imputeCT, RP_CT),
         HPRT_CT = if_else(HPRT_CT > imputeCT, imputeCT, HPRT_CT)) 
# remove samples with undetected RP_CT values
clinical_viralload_data <- clinical_viralload_data %>% 
  filter(RP_CT != 40)

# add days from symptom onset, "symptom_onset_days" as event_date - symptom_date
clinical_viralload_data <- clinical_viralload_data %>%
  mutate(symptom_onset_days = event_date - symptom_date)


```


&nbsp;

&nbsp;

&nbsp;

#### visit 1 analysis code: 4 cluster convalescent TGs classification

```{r visit 1-analysis, warning=FALSE}

data_use_visit1_4TG <- clinical_viralload_data %>%
  rename(trajectory_group_original = trajectory_group) %>%
  mutate(conv_cluster4 = recode(conv_cluster6, `1` = 1, `3` = 1, `4` = 1, `2` = 2, `5` = 5, `6` = 6)) %>%
  rename(trajectory_group = conv_cluster4) %>%
  tidyr::drop_na(trajectory_group) %>% 
  filter(event_type == "Visit 1") %>%
  select(sample_id, N1_CT, N2_CT, 
         enrollment_site,
         trajectory_group,
         discretized_admit_age_quantile,
         sex = sex) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group, levels = c("1", "2", "5", "6"))) %>%
  column_to_rownames(var = "sample_id")


module_names <- c("N1_CT", "N2_CT")

res_table_ordinal <- NULL
data_use <- data_use_visit1_4TG

for (moduleName in module_names) {
  my.formula0 <- paste0('trajectory_group~ 1+(1|enrollment_site)+',
                        'discretized_admit_age_quantile+sex')
  my.formula1 <- paste0('trajectory_group~',
                        moduleName,
                        '+(1|enrollment_site)+',
                        'discretized_admit_age_quantile+sex')
  res <- mixed_ordinal(my.formula0, 
                       my.formula1,
                       data_use)
  coef(clmm(formula(my.formula1), 
            data = data_use))[moduleName] %>%
    unname() %>%
    c(res, coef = .) -> res
  res_table_ordinal <- rbind(res_table_ordinal, res)
}
rownames(res_table_ordinal) <- module_names
res_table_ordinal <- res_table_ordinal %>%
  as.data.frame() %>%
   mutate(qval =  qvalue::qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues)


res_table_ordinal_df <- res_table_ordinal %>%
  rownames_to_column(var = "feature")

# pairwise comparisons
data_use_visit1_4TG_pairwise <- data_use_visit1_4TG %>%
  rename(endpoints = trajectory_group)

for(j in 1:length(module_names)){
  print(j)
  my.formula0 = paste0(module_names[j],"~ 1+(1|enrollment_site)+discretized_admit_age_quantile+sex")
  my.formula1 = paste0(module_names[j],"~ endpoints + (1|enrollment_site)+discretized_admit_age_quantile+sex")
  if(j == 1){
    tmp = mixed_pairwise(my.formula0, my.formula1, data_use_visit1_4TG_pairwise)
    res_table_pairwise = data.frame(matrix(0, ncol = length(tmp), nrow = length(module_names)))
    res_table_pairwise[j,] = tmp
    colnames(res_table_pairwise) = names(tmp)
  }else{
    res_table_pairwise[j,] = mixed_pairwise(my.formula0, my.formula1, data_use_visit1_4TG_pairwise)
  }
}
rownames(res_table_pairwise) = module_names


res_table_ordinal_df <- cbind(res_table_ordinal_df, res_table_pairwise)

res_table_pairwise_long_df <- res_table_pairwise %>%
  tibble::rownames_to_column(var = "feature") %>%
  pivot_longer(cols = -c("feature")) %>%
  rename(Pvalue = value) %>%
  tidyr::separate(col = "name", into = c("group1", "group2"), remove = FALSE)


# filename_long = format(Sys.Date(), "/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/nasal-viralload/clinGroup/Figures/conv_symps/%y%m%d-P_values_clmm_visit1_with_age_sex_for_conv_symps.csv")
# write.csv(res_table_ordinal_df, 
#            file = filename_long, row.names = FALSE)



```

&nbsp;

&nbsp;

&nbsp;

#### visit 1 analysis figures: 4 cluster convalescent TGs classification
```{r PlotViralDataByClinicalTrajectoryGroupV2, fig.width = 8, fig.height = 8, warning=FALSE, echo=FALSE, eval=TRUE}

TrajGrpCol <- c("1" = "#639A21", "2" = "#39828C", "5" = "#9C3418", "6" = "pink")

library(ggbeeswarm)

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "N1_CT") %>%
  filter(Pvalue < 0.05) %>%
  mutate(Pvalue = round(Pvalue,4)) %>%
  select(group1, group2, Pvalue) %>%
  mutate(`y.position` = c(41,44)) %>%
  as.tibble()
  
  
library(ggprism)


## N1_CT figures

df_visit1_mean <- data_use_visit1_4TG %>%
  group_by(trajectory_group) %>%
  summarise(n = n(),
            N1_CT_mean = mean(N1_CT),
            N2_CT_mean = mean(N2_CT))



boxplot_N1 <- ggplot(data_use_visit1_4TG, aes(x = trajectory_group, y = N1_CT)) + 
  geom_quasirandom(aes(x = trajectory_group, y = N1_CT, color = trajectory_group), alpha = 0.3) +
  geom_boxplot(aes(x = trajectory_group, y = N1_CT, fill = trajectory_group), size = 1, alpha = 0.3, outlier.shape = NA) +
  geom_point(data = df_visit1_mean, aes(x = trajectory_group, y = N1_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("N1 CT") + xlab("Convalescent\ntrajectory groups") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=16)) +
  # scale_fill_brewer(palette = "Set1") +
  # scale_color_brewer(palette = "Set1") +
  scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "Pvalue",
             vjust = 2.5) +
  coord_cartesian(ylim = c(46, 8))

boxplot_N1


# ggsave(
#   filename = format(Sys.Date(), "/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/nasal-viralload/clinGroup/Figures/conv_symps/%y%m%d-N1_CT_baseline_boxplot_conv_symps.pdf"),
#   width = 2,
#   height = 3
# )



## N2_CT figures

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "N2_CT") %>%
  filter(Pvalue < 0.05) %>%
  mutate(Pvalue = round(Pvalue,4)) %>%
  select(group1, group2, Pvalue) %>%
  mutate(`y.position` = c(41,44)) %>%
  as.tibble()

boxplot_N2 <- ggplot(data_use_visit1_4TG, aes(x = trajectory_group, y = N2_CT)) + 
  geom_quasirandom(aes(x = trajectory_group, y = N2_CT, color = trajectory_group), alpha = 0.3) +
  geom_boxplot(aes(x = trajectory_group, y = N2_CT, fill = trajectory_group), size = 1, alpha = 0.3, outlier.shape = NA) +
  geom_point(data = df_visit1_mean, aes(x = trajectory_group, y = N2_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("N2 CT") + xlab("Convalescent\ntrajectory groups") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=16)) +
  # scale_fill_brewer(palette = "Set1") +
  # scale_color_brewer(palette = "Set1") +
  scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "Pvalue",
             vjust = 2.5) +
  coord_cartesian(ylim = c(46, 8))

boxplot_N2


# ggsave(
#   filename = format(Sys.Date(), "/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/nasal-viralload/clinGroup/Figures/conv_symps/%y%m%d-N2_CT_baseline_boxplot_conv_symps.pdf"),
#   width = 2,
#   height = 3
# )



```


### Pooled analysis code
```{r pooled analysis code, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

data_use_pooled_4TG <- clinical_viralload_data %>%
  mutate(conv_cluster4 = recode(conv_cluster6, `1` = 1, `3` = 1, `4` = 1, `2` = 2, `5` = 5, `6` = 6)) %>%
  select(sample_id, N1_CT, N2_CT, 
         enrollment_site,
         conv_cluster4,
         participant_id,
         event_type,
         event_date,
         discretized_admit_age_quantile,
         sex = sex) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(conv_cluster4 = factor(conv_cluster4, levels = c("1", "2", "5", "6"))) %>%
  column_to_rownames(var = "sample_id")

immune_assay <- data_use_pooled_4TG %>%
  select(N1_CT, N2_CT)
immune_meta <- data_use_pooled_4TG %>% 
  tibble::rownames_to_column(var = "sample_id") %>%
  select(sample_id, 
         participant_id,
         event_type,
         event_date,
         discretized_admit_age_quantile,
         sex,
         enrollment_site,
         conv_cluster4)


if(!(all(rownames(immune_assay) == immune_meta$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}


response_name = "conv_cluster4"
random_effect_list = c("participant_id", "enrollment_site")
fixed_effect_list = c("sex", "discretized_admit_age_quantile")
# fixed_effect_list = c("event_type","sex", "discretized_admit_age_quantile")

# over all association

module_names <- c("N1_CT", "N2_CT")

pval_overall <- NULL

for (moduleName in module_names) {
  my.formula0 <- paste0(moduleName, '~(1|enrollment_site)+(1|participant_id)+discretized_admit_age_quantile+sex')
  my.formula1 <- paste0(moduleName,
                        '~(1|enrollment_site)+(1|participant_id)+discretized_admit_age_quantile+sex+conv_cluster4')
  av = anova(lme4::lmer(my.formula0, data = data_use_pooled_4TG),
                 lme4::lmer(my.formula1, data = data_use_pooled_4TG))
  res <- av$`Pr(>Chisq)`[2]
  pval_overall <- rbind(pval_overall, res)
}

res_table_overall <- data.frame(feature = module_names, 
                                pval = as.vector(pval_overall)) %>% 
  mutate(qval =  qvalue::qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues) %>%
  tibble::column_to_rownames(var = "feature")


# pairwise comparisons
pooled_resDF_pairwise <- NULL
for(feature_name in colnames(immune_assay)){
  res0 = pooled_test(immune_assay=immune_assay, immune_meta=immune_meta, 
                     feature_name = feature_name, response_name=response_name, 
                     random_effect_list=random_effect_list , fix_effect_list=fixed_effect_list)
  pooled_resDF_pairwise <- rbind(pooled_resDF_pairwise,res0)
}
rownames(pooled_resDF_pairwise) = colnames(immune_assay)

res_table_pooled_all <- cbind(res_table_overall, pooled_resDF_pairwise)


```

&nbsp;

&nbsp;

&nbsp;

#### Pooled analysis figures: 4 cluster convalescent TGs classification
```{r PlotViralDataByClinicalTrajectoryGroupV2, fig.width = 8, fig.height = 8, warning=FALSE, echo=FALSE, eval=TRUE}

TrajGrpCol <- c("1" = "#999999", "2" = "#56B4E9", "5" = "#E69F00", "6" = "#FF0000")
# col_sel <- c("MIN" = "#999999", "PHY" = "#56B4E9", "COG" = "#E69F00", "MLT" = "#FF0000")

library(ggbeeswarm)

res_table_pairwise_long_df <- pooled_resDF_pairwise %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "feature") %>%
  pivot_longer(cols = -c("feature")) %>%
  rename(Pvalue = value) %>%
  tidyr::separate(col = "name", into = c("group1", "group2"), remove = FALSE)

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "N1_CT") %>%
  filter(Pvalue < 0.05) %>%
  mutate(Pvalue = round(Pvalue,4)) %>%
  select(group1, group2, Pvalue) %>%
  mutate(`y.position` = c(9,11)) %>%
  as.tibble()
  
  
library(ggprism)


## N1_CT figures

df_pooled_mean <- data_use_pooled_4TG %>%
  group_by(conv_cluster4) %>%
  summarise(n = n(),
            N1_CT_mean = mean(N1_CT),
            N2_CT_mean = mean(N2_CT))

boxplot_N1 <- ggplot(data_use_pooled_4TG, aes(x = conv_cluster4, y = N1_CT)) + 
  geom_boxplot(aes(x = conv_cluster4, y = N1_CT, color = conv_cluster4), size = 0.8, outlier.shape = NA) +
  geom_quasirandom(aes(x = conv_cluster4, y = N1_CT, color = conv_cluster4), width = 0.2, alpha = 0.7, size = 1) +
  # geom_point(data = df_pooled_mean, aes(x = conv_cluster4, y = N1_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("N1 Ct") + xlab("") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10),
        axis.text.y = element_text(size = 10),
        text = element_text(size=10)) +
  # scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "Pvalue",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('MIN', 'PHY', 'COG', 'MLT'))

boxplot_N1


ggsave(
  filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/Figures/%y%m%d-N1_CT_pooled_boxplot_PRO_clusters.pdf"),
  width = 2.5,
  height = 3
)



## N2_CT figures

res_table_pairwise_long_df_temp <- res_table_pairwise_long_df %>%
  filter(feature == "N2_CT") %>%
  filter(Pvalue < 0.05) %>%
  mutate(Pvalue = round(Pvalue,4)) %>%
  select(group1, group2, Pvalue) %>%
  mutate(`y.position` = c(9,11)) %>%
  as.tibble()

boxplot_N2 <- ggplot(data_use_pooled_4TG, aes(x = conv_cluster4, y = N2_CT)) + 
  geom_boxplot(aes(x = conv_cluster4, y = N2_CT, color = conv_cluster4), size = 0.8, outlier.shape = NA) + 
  geom_quasirandom(aes(x = conv_cluster4, y = N2_CT, color = conv_cluster4), width = 0.2, alpha = 0.7, size = 1) +
  # geom_point(data = df_pooled_mean, aes(x = conv_cluster4, y = N1_CT_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  ylab("N2 Ct") + xlab("") +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 10, color = "black"),
        axis.text.y = element_text(size = 10, color = "black"),
        text = element_text(size=10)) +
  # scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  add_pvalue(res_table_pairwise_long_df_temp, 
             label = "Pvalue",
             vjust = 0,
             coord.flip = FALSE,
             tip.length = 0,
             bracket.nudge.y = 0,
             label.size = 3) +
  coord_cartesian(ylim = c(40, 8)) +
  scale_x_discrete(labels=c('MIN', 'PHY', 'COG', 'MLT'))

boxplot_N2


ggsave(
  filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/Figures/%y%m%d-N2_CT_pooled_boxplot_PRO_clusters.pdf"),
  width = 2.5,
  height = 3
)




```



### Longitudinal analysis code: 4 cluster convalescent TGs classification
```{r Longitudinal-analysis, message=FALSE, results="hide"}

data_use_longitudinal_4TG <- clinical_viralload_data %>%
  rename(trajectory_group_original = trajectory_group) %>%
  mutate(conv_cluster4 = recode(conv_cluster6, `1` = "MIN", `3` = "MIN", `4` = "MIN", `2` = "PHY", `5` = "COG", `6` = "MLT")) %>%
  rename(trajectory_group = conv_cluster4) %>%
  tidyr::drop_na(trajectory_group) %>%
  filter(event_date <= 28) %>%
  select(sample_id, N1_CT, N2_CT, participant_id,  
         enrollment_site,
         trajectory_group, event_date, event_type, sex, discretized_admit_age_quantile) %>% 
  mutate(participant_id = factor(participant_id)) %>% 
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group, levels = c("MIN", "PHY", "COG", "MLT"))) 
  
  
inputDF <- data_use_longitudinal_4TG %>%
  pivot_longer(cols = -c("sample_id", "event_date", "event_type", "participant_id",
                         "enrollment_site", "trajectory_group", "sex", "discretized_admit_age_quantile"
                         ))


# smooth spline
smooth_spline_model_loop_4TG <- model_loop(inputDF, 
                                       modelType = "smoothSpline", 
                                       endpoint = "trajectory_group")

smooth_spline_model_loop_df_4TG <- smooth_spline_model_loop_4TG %>%
  rownames_to_column(var = "feature")

# filename_long = format(Sys.Date(), "/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/nasal-viralload/clinGroup/Figures/conv_symps/%y%m%d-P_values_longitudinal_smoothSpline_with_age_sex_for_conv_symps.csv")
# write.csv(smooth_spline_model_loop_df, 
#            file = filename_long, row.names = FALSE)

```



### Longitudinal analysis figures: template code
```{r Longitudinal-analysis figures, message=FALSE, results="hide"}

plotExample <- inputDF %>%
  filter(name %in% "N1_CT") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)),
                                    levels = c("MIN", "PHY", "COG", "MLT"))) 

# col_sel <- c("MIN" = "#999999", "PHY" = "#CC79A7", "COG" = "#0072B2", "MLT" = "#D55E00")
# col_sel <- c("MIN" = "#009E73", "PHY" = "#CC79A7", "COG" = "#0072B2", "MLT" = "#D55E00")
# col_sel <- c("MIN" = "#009E73", "PHY" = "#CC79A7", "COG" = "#56B4E9", "MLT" = "#E69F00")

# from wesanderson colors
# col_sel <- c("MIN" = "#899DA4", "PHY" = "#CC79A7", "COG" = "#56B4E9", "MLT" = "#C93312")
# col_sel <- c("MIN" = "#74A089", "PHY" = "#9A8822", "COG" = "#56B4E9", "MLT" = "#C93312")
# col_sel <- c("MIN" = "#00A08A", "PHY" = "#CC79A7", "COG" = "#56B4E9", "MLT" = "#FF0000")

# final selection
col_sel <- c("MIN" = "#999999", "PHY" = "#56B4E9", "COG" = "#E69F00", "MLT" = "#FF0000")

temp_plot <- plot_model_smooth_conv(plotExample, model_loop = smooth_spline_model_loop_4TG, modelType = "smoothSpline",
           p_adjust = F, endpoint = "trajectory_group", signif_markers = F, remove_NS = T,
           title = "",
           individual_points_size = 1,
           individual_paths_size = 0.5,
           group_trendline_line_width = 1.5,
           title_size = 12,
           bar_height = 5,
           custom_theme_graph = 'text = element_text(size=10)',
           # custom_theme_bars = 'text = element_text(size=6)',
           knot_lines = F,
           ylabel = "N1 Ct",
           CI = FALSE,
           individual_trendlines = T, individual_points = T, individual_paths = T,
           p_value_text_size = 3,
           return_multi_obj = F,
           colors = col_sel,
           y_axis_reverse = TRUE,
           trajectory_labels = c("MIN", "PHY", "COG", "MLT"))
 

print(temp_plot)
ggsave(filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/Figures/%y%m%d-N1_CT_longitudinal_PRO_clusters.pdf"),
         plot = temp_plot,
         width = 3.5,
         height = 2.5)

# N2 CT
plotExample <- inputDF %>%
  filter(name %in% "N2_CT") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)),
                                    levels = c("MIN", "PHY", "COG", "MLT"))) 
temp_plot <- plot_model_smooth_conv(plotExample, model_loop = smooth_spline_model_loop_4TG, modelType = "smoothSpline",
           p_adjust = F, endpoint = "trajectory_group", signif_markers = F, remove_NS = T,
           title = "",
           individual_points_size = 1,
           individual_paths_size = 0.5,
           group_trendline_line_width = 1.5,
           title_size = 12,
           bar_height = 5,
           custom_theme_graph = 'text = element_text(size=10)',
           # custom_theme_bars = 'text = element_text(size=6)',
           knot_lines = F,
           ylabel = "N2 Ct",
           CI = FALSE,
           individual_trendlines = T, individual_points = T, individual_paths = T,
           p_value_text_size = 3,
           return_multi_obj = F,
           colors = col_sel,
           y_axis_reverse = TRUE,
           trajectory_labels = c("MIN", "PHY", "COG", "MLT"))
 


print(temp_plot)
ggsave(filename = format(Sys.Date(), "/scratch/nasal-transcriptomics/impacc/Analysis/nasal-viralload/clinical_conv/Figures/%y%m%d-N2_CT_longitudinal_PRO_clusters.pdf"),
         plot = temp_plot,
         width = 3.5,
         height = 2.5)

```



### Longitudinal analysis figures
```{r Longitudinal-analysis figures, message=FALSE, results="hide"}


pred_smooth_spline_gamm4 <- function(plotDF = plotDF, age_sex = F){
  if(age_sex){
    print("Model is adjusting for age and sex")
    formula_use = formula(paste0("value~ sex + discretized_admit_age_quantile + s(event_date, bs = 'cr')+
              s(event_date, bs = 'cr', by =", endpoint, ")+", endpoint))
    fit_1 <- mgcv::gam(formula_use, data = plotDF)
  
    gamm4_pred <- mgcv::predict.gam(fit_1, plotDF, se.fit = TRUE)
    yhat_max <- as.numeric(gamm4_pred$fit + 1.96*gamm4_pred$se.fit)
    yhat_min <- as.numeric(gamm4_pred$fit - 1.96*gamm4_pred$se.fit)
  
    plotDF_v2 <- plotDF %>%
    mutate(yhat = as.numeric(gamm4_pred$fit),
           yhat_max = yhat_max,
           yhat_min = yhat_min)
    
  } else {print("Model is NOT adjusting for age and sex")
    formula_use = formula(paste0("value~ s(event_date, bs = 'cr')+
              s(event_date, bs = 'cr', by =", endpoint, ")+", endpoint))
    fit_1 <- mgcv::gam(formula_use, data = plotDF)
  
    gamm4_pred <- mgcv::predict.gam(fit_1, plotDF, type = "response", se.fit = TRUE)
    yhat_max <- as.numeric(gamm4_pred$fit + 1.96*gamm4_pred$se.fit)
    yhat_min <- as.numeric(gamm4_pred$fit - 1.96*gamm4_pred$se.fit)
  
    plotDF_v2 <- plotDF %>%
    mutate(yhat = as.numeric(gamm4_pred$fit),
           yhat_max = yhat_max,
           yhat_min = yhat_min)
    }
  
  return(plotDF_v2)


}

inputDF_plot <- data_use_longitudinal_4TG %>%
  # mutate(conv_symp = if_else(symp == 1, "Yes", "No")) %>%
  # mutate(conv_symp = factor(conv_symp)) %>%
  mutate(participant_id = factor(participant_id)) %>%
  mutate(discretized_admit_age_quantile = factor(discretized_admit_age_quantile)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "event_type", "participant_id",
                         "enrollment_site", "trajectory_group", "sex", "discretized_admit_age_quantile"
                         )) 

endpoint <- "trajectory_group"
inputDF_plot_N1 <- inputDF_plot %>% 
  filter(event_date <= 28) %>%
  filter(name == "N1_CT")

plotDF_N1 <- pred_smooth_spline(plotDF = inputDF_plot_N1, age_sex = F)

longi_plot_N1 <- plotDF_N1 %>%
  ggplot() +
  xlab("Days from admission") +
  ylab("N1 CT") +
  # scale_fill_manual(values = color_remdes) +
  # scale_color_manual(values = color_remdes) +
  geom_jitter(data = plotDF_N1, aes(x = event_date, y = value, color = trajectory_group), alpha = 0.3, width = 0.25, height = 0.25) +
  # geom_ribbon(data = plotDF_N1, aes(x = event_date, y = yhat, fill = trajectory_group, ymin = yhat_min, ymax = yhat_max), alpha = 0.3) +
  geom_line(data = plotDF_N1, aes(x = event_date, y = yhat, color = trajectory_group), alpha = 0.7, size = 2) +
theme_bw() + theme( 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=16)) +
  scale_x_continuous(breaks = c(0,7,14,21,28)) +
  scale_y_continuous(breaks = c(20, 30, 40)) +
  coord_cartesian(ylim = c(40,18)) +
  labs(color='Convalescent\ntrajectory groups') +
  # scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  guides(fill = "none")

longi_plot_N1
# legend.position = "none",

# ggsave(
#   filename = format(Sys.Date(), "/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/nasal-viralload/clinGroup/Figures/conv_symps//%y%m%d-N1_CT_longitudinal_conv_symp.pdf"),
#   width = 6,
#   height = 3
# )


inputDF_plot_N2 <- inputDF_plot %>% 
  filter(event_date <= 28) %>%
  filter(name == "N2_CT")

plotDF_N2 <- pred_smooth_spline(plotDF = inputDF_plot_N2, age_sex = F)

longi_plot_N2 <- plotDF_N2 %>%
  ggplot() +
  xlab("Days from admission") +
  ylab("N2 CT") +
  # scale_fill_manual(values = color_remdes) +
  # scale_color_manual(values = color_remdes) +
  geom_jitter(data = plotDF_N2, aes(x = event_date, y = value, color = trajectory_group), alpha = 0.3, width = 0.25, height = 0.25) +
  # geom_ribbon(data = plotDF_N2, aes(x = event_date, y = yhat, fill = trajectory_group, ymin = yhat_min, ymax = yhat_max), alpha = 0.3) +
  geom_line(data = plotDF_N2, aes(x = event_date, y = yhat, color = trajectory_group), alpha = 0.7, size = 2) +
theme_bw() + theme( 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=16)) +
  scale_x_continuous(breaks = c(0,7,14,21,28)) +
  scale_y_continuous(breaks = c(20, 30, 40)) +
  coord_cartesian(ylim = c(40,18)) +
  # scale_fill_manual(values = TrajGrpCol) +
  scale_color_manual(values = TrajGrpCol) +
  labs(color='Convalescent\ntrajectory groups') +
  guides(fill = "none")

longi_plot_N2

# ggsave(
#   filename = format(Sys.Date(), "/users/ndjayavelu/njayavelu/Scripts/impacc/Analysis/nasal-viralload/clinGroup/Figures/conv_symps//%y%m%d-N2_CT_longitudinal_conv_symp.pdf"),
#   width = 6,
#   height = 3
# )



```

&nbsp;

&nbsp;

&nbsp;



&nbsp;

&nbsp;

&nbsp;

### Convalescent symptoms analysis
```{r Convalescent symptoms analysis, message=FALSE, results="hide"}
# read symptoms data
file_conv_symp = file.path("/scratch/clinical/current", "impacc_conv_symptoms.csv")
conv_symp_data <- read.csv(file_conv_symp) %>%
  mutate(convsx_cardio_1plus = factor(convsx_cardio_1plus, levels = c("0", "1")),
         convsx_systemic_1plus = factor(convsx_systemic_1plus, levels = c("0", "1")),
         convsx_other_1plus = factor(convsx_other_1plus, levels = c("0", "1")),
         convsx_1plus_count = factor(convsx_1plus_count, levels = c("0", "1", "2", "3")),
         convsx_cardio_2plus = factor(convsx_cardio_2plus, levels = c("0", "1")),
         convsx_systemic_2plus = factor(convsx_systemic_2plus, levels = c("0", "1")),
         convsx_other_2plus = factor(convsx_other_2plus, levels = c("0", "1")),
         convsx_2plus_count = factor(convsx_2plus_count, levels = c("0", "1", "2", "3"))) 
 
clinical_symp_viralload_data <- data_use_pooled_4TG %>%
  inner_join(conv_symp_data)

# convsx_cardio_1plus
my.formula1 <- paste0("convsx_cardio_1plus~Ratio_RBD_N1+discretized_admit_age_quantile+sex+(1|enrollment_site)+(1|participant_id)")
convsx_cardio_1plus_fit <- glmer(my.formula1, data = clinical_symp_viralload_titers_counts_data, family = binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 1)
summary(convsx_cardio_1plus_fit)

# convsx_systemic_1plus
my.formula1 <- paste0("convsx_systemic_1plus~Ratio_RBD_N1+discretized_admit_age_quantile+sex+(1|enrollment_site)+(1|participant_id)")
convsx_systemic_1plus_fit <- glmer(my.formula1, data = clinical_symp_viralload_titers_counts_data, family = binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 1)
summary(convsx_systemic_1plus_fit)


# convsx_1plus_count
my.formula0 <- paste0("convsx_1plus_count~discretized_admit_age_quantile+sex+(1|enrollment_site)+(1|participant_id)")
my.formula1 <- paste0("convsx_1plus_count~Ratio_RBD_N1+discretized_admit_age_quantile+sex+(1|enrollment_site)+(1|participant_id)")
convsx_1plus_count_fit0 <- glmer(my.formula0, data = clinical_symp_viralload_titers_counts_data, family = binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 1)
convsx_1plus_count_fit1 <- glmer(my.formula1, data = clinical_symp_viralload_titers_counts_data, family = binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 1)
anova(convsx_1plus_count_fit0, convsx_1plus_count_fit1)
summary(convsx_1plus_count_fit)

# over all association
# N1 CT
module_names <- c("convsx_cardio_1plus", "convsx_systemic_1plus", "convsx_other_1plus",
                     "convsx_cardio_2plus", "convsx_systemic_2plus", "convsx_other_2plus",
                  "convsx_1plus_count", "convsx_2plus_count")
pval_overall <- NULL

moduleName <- module_names[1]

for (moduleName in module_names) {
  my.formula0 <- paste0(moduleName, '~(1|enrollment_site)+(1|participant_id)+discretized_admit_age_quantile+sex')
  my.formula1 <- paste0(moduleName,
                        '~(1|enrollment_site)+(1|participant_id)+discretized_admit_age_quantile+sex+N1_CT')
  fit0 <- glmer(my.formula0, data = clinical_symp_viralload_data, family = binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 1)
  fit1 <- glmer(my.formula1, data = clinical_symp_viralload_data, family = binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 1)
  av = anova(fit0, fit1)
  res <- av$`Pr(>Chisq)`[2]
  pval_overall <- rbind(pval_overall, res)
}

res_table_overall <- data.frame(feature = module_names, 
                                pval = as.vector(pval_overall)) # %>% 
  mutate(qval =  qvalue::qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues) %>%
  tibble::column_to_rownames(var = "feature")

# N2_CT
module_names <- c("convsx_cardio_1plus", "convsx_systemic_1plus", "convsx_other_1plus",
                     "convsx_cardio_2plus", "convsx_systemic_2plus", "convsx_other_2plus",
                  "convsx_1plus_count", "convsx_2plus_count")
pval_overall <- NULL

for (moduleName in module_names) {
  my.formula0 <- paste0(moduleName, '~(1|enrollment_site)+(1|participant_id)+discretized_admit_age_quantile+sex')
  my.formula1 <- paste0(moduleName,
                        '~(1|enrollment_site)+(1|participant_id)+discretized_admit_age_quantile+sex+N2_CT')
  fit0 <- glmer(my.formula0, data = clinical_symp_viralload_data, family = binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 1)
  fit1 <- glmer(my.formula1, data = clinical_symp_viralload_data, family = binomial, control = glmerControl(optimizer = "bobyqa"), nAGQ = 1)
  av = anova(fit0, fit1)
  res <- av$`Pr(>Chisq)`[2]
  pval_overall <- rbind(pval_overall, res)
}

res_table_overall <- data.frame(feature = module_names, 
                                pval = as.vector(pval_overall)) # %>% 
  mutate(qval =  qvalue::qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues) %>%
  tibble::column_to_rownames(var = "feature")


```

&nbsp;

&nbsp;

&nbsp;

### Convalescent symptoms figures
```{r Convalescent symptoms figures, message=FALSE, results="hide"}

data_plot_long_df <- clinical_symp_viralload_data %>%
  select(participant_id, event_type, N1_CT, N2_CT, convsx_cardio_1plus, convsx_systemic_1plus, convsx_other_1plus, convsx_cardio_2plus, convsx_systemic_2plus, convsx_other_2plus, convsx_1plus_count, convsx_2plus_count) %>%
  pivot_longer(cols = -c("event_type", "participant_id", "N1_CT", "N2_CT"))


# N1_CT, 1plus and 2 plus
boxplot_N1_conv_symp_1plus_2plus <- data_plot_long_df %>%
  filter(name %in% c("convsx_cardio_1plus", "convsx_systemic_1plus", "convsx_other_1plus",
                     "convsx_cardio_2plus", "convsx_systemic_2plus", "convsx_other_2plus")) %>%
  mutate(name = factor(name, levels = c("convsx_cardio_1plus", "convsx_systemic_1plus", "convsx_other_1plus", "convsx_cardio_2plus", "convsx_systemic_2plus", "convsx_other_2plus"))) %>%
  drop_na(value) %>%
  ggplot(aes(x = value, y = N1_CT, fill = value)) + 
  geom_boxplot(aes(x = value, y = N1_CT, fill = value), size = 1, alpha = 0.6) +
  facet_wrap(~name, ncol = 3) +
  theme_bw() +
  ylab("N1 CT") + xlab("Convalescent symptom groups") +
  # coord_cartesian(ylim = c(0, 4))
  # scale_fill_manual(values = color_remdes) +
  scale_y_reverse()
  # stat_compare_means(comparisons = my_comparisons)

# N1_CT, 1plus and 2 plus count
boxplot_N1_conv_symp_1plus_2plus_count <- data_plot_long_df %>%
  filter(name %in% c("convsx_1plus_count",
                     "convsx_2plus_count")) %>%
  mutate(name = factor(name, levels = c("convsx_1plus_count", "convsx_2plus_count"))) %>%
  drop_na(value) %>%
  ggplot(aes(x = value, y = N1_CT, fill = value)) + 
  geom_boxplot(aes(x = value, y = N1_CT, fill = value), size = 1, alpha = 0.6) +
  facet_wrap(~name, ncol = 1) +
  theme_bw() +
  ylab("N1 CT") + xlab("Convalescent symptom groups") +
  # coord_cartesian(ylim = c(0, 4)) +
  scale_y_reverse()





# N2_CT, 1plus and 2 plus
boxplot_N2_conv_symp_1plus_2plus <- data_plot_long_df %>%
  filter(name %in% c("convsx_cardio_1plus", "convsx_systemic_1plus", "convsx_other_1plus",
                     "convsx_cardio_2plus", "convsx_systemic_2plus", "convsx_other_2plus")) %>%
  mutate(name = factor(name, levels = c("convsx_cardio_1plus", "convsx_systemic_1plus", "convsx_other_1plus", "convsx_cardio_2plus", "convsx_systemic_2plus", "convsx_other_2plus"))) %>%
  drop_na(value) %>%
  ggplot(aes(x = value, y = N2_CT, fill = value)) + 
  geom_boxplot(aes(x = value, y = N2_CT, fill = value), size = 1, alpha = 0.6) +
  facet_wrap(~name, ncol = 3) +
  theme_bw() +
  ylab("N2 CT") + xlab("Convalescent symptom groups") +
  # coord_cartesian(ylim = c(0, 4))
  # scale_fill_manual(values = color_remdes) +
  scale_y_reverse()
  # stat_compare_means(comparisons = my_comparisons)

# N2_CT, 1plus and 2 plus count
boxplot_N2_conv_symp_1plus_2plus_count <- data_plot_long_df %>%
  filter(name %in% c("convsx_1plus_count",
                     "convsx_2plus_count")) %>%
  mutate(name = factor(name, levels = c("convsx_1plus_count", "convsx_2plus_count"))) %>%
  drop_na(value) %>%
  ggplot(aes(x = value, y = N2_CT, fill = value)) + 
  geom_boxplot(aes(x = value, y = N2_CT, fill = value), size = 1, alpha = 0.6) +
  facet_wrap(~name, ncol = 1) +
  theme_bw() +
  ylab("N2 CT") + xlab("Convalescent symptom groups") +
  # coord_cartesian(ylim = c(0, 4)) +
  scale_y_reverse()

```





&nbsp;

&nbsp;

&nbsp;


### Effect of steroid use on viral load 
```{r Longitudinal-analysis figures, message=FALSE, results="hide"}

# read steroids data
file_steroids = file.path("/data/clinical/current", "impacc-clin-individ-supp.csv")
ind_supp_data <- read.csv(file_steroids) %>%
  select(participant_id, ever_steroids, steroids_start, ever_remdes, remdes_start) %>%
  mutate(ever_steroids = factor(ever_steroids, levels = c("0", "1"))) %>%
  mutate(ever_remdes = factor(ever_remdes, levels = c("0", "1")))
 
clinical_viralload_data_steroid <- clinical_viralload_data %>%
  inner_join(ind_supp_data)

df_visit1 <- clinical_viralload_data_steroid %>%
  filter(event_type == "Visit 1")

my_comparisons <- list(c("1", "0"))

# overall
boxplot_steroid_viral_N1 <- df_visit1 %>% 
  tidyr::drop_na(ever_steroids) %>%
  ggplot(aes(x = ever_steroids, y = N1_CT, fill = ever_steroids)) + 
  geom_boxplot(aes(x = ever_steroids, y = N1_CT, fill = ever_steroids), size = 1, alpha = 0.6) +
  theme_bw() +
  # ggtitle("N1 CT values distribution \nby steroid use") +
  ylab("N1_CT") + xlab("") +
  # scale_fill_manual(values = color_remdes) +
  scale_y_reverse() +
  stat_compare_means(comparisons = my_comparisons)

# conv 6TG
boxplot_steroid_viral_N1_by_conv6TG <- df_visit1 %>% 
  tidyr::drop_na(ever_steroids) %>%
  ggplot(aes(x = ever_steroids, y = N1_CT, fill = ever_steroids)) + 
  geom_boxplot(aes(x = ever_steroids, y = N1_CT, fill = ever_steroids), size = 1, alpha = 0.6) +
  facet_wrap(~conv_cluster6) +
  theme_bw() +
  # ggtitle("N1 CT values distribution \nby steroid use") +
  ylab("N1_CT") + xlab("") +
  # scale_fill_manual(values = color_remdes) +
  scale_y_reverse() +
  stat_compare_means(comparisons = my_comparisons)

# conv 3TG
boxplot_steroid_viral_N1_by_conv3TG <- df_visit1 %>% 
  tidyr::drop_na(ever_steroids) %>%
  ggplot(aes(x = ever_steroids, y = N1_CT, fill = ever_steroids)) + 
  geom_boxplot(aes(x = ever_steroids, y = N1_CT, fill = ever_steroids), size = 1, alpha = 0.6) +
  facet_wrap(~conv_cluster3) +
  theme_bw() +
  # ggtitle("N1 CT values distribution \nby steroid use") +
  ylab("N1_CT") + xlab("") +
  # scale_fill_manual(values = color_remdes) +
  scale_y_reverse() +
  stat_compare_means(comparisons = my_comparisons)


```





