---
title: "Transplant nasal transcriptomics"
output: html_document
date: "2024-01-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(biomaRt)
library(dplyr)
library(ggplot2)
library(patchwork)
library(gamm4)
library(limma)
library(clusterProfiler)
library(edgeR)
library(ReactomePA)
library(RColorBrewer)
library(pheatmap)

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```

# Data import

## Clinical metadata

```{r}
# Patient metadata
clin.individ <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-locked.csv",
  header = TRUE)

# Sample
clin.sample <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-sample-locked.csv",
  header = TRUE)

# Event
clin.event <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-event-locked.csv",
  header = TRUE)
```

Merge metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

Manually correct some typos.

```{r}
# Correct the symptom_date of patient 008-0103: change from -376 to -13 (confirmed with data management team)
clin.sample.event.individ[clin.sample.event.individ$participant_id=="008-0103","symptom_date"] <- -13

# Correct event_date of 004-0012-8: change from 2 to 32 (it's 32 in 2021-11-11-frozen data)
clin.sample.event.individ[clin.sample.event.individ$event_id=="004-0012-8","event_date"] <- 32
```

## Transplant metadata 

264 patients included in transplant study (use new metadata table that Harry will share)

```{r}
transplant <- read.delim(
  "/scratch/immunosuppression/2023.03.09_txcasecontrol.txt",
  header = TRUE)
newTG <- read.delim(
  "/scratch/immunosuppression/2023.10.03_new_tgroups.txt",
  header = TRUE)

# Merge new TG assignment
transplant <- merge(
  transplant, newTG,
  by="studyid", all=FALSE
)

# Format factor
transplant$transplant_final <- as.factor(transplant$transplant_final)

```

Remove BMT patients and their controls.

```{r}
transplant <- transplant %>%
  subset(!(studyid %in% c("002-0022","007-0008","001-0013","010-0009","004-0041","005-0047")))

```

Only keep transplant patients in the clinical metadata

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event.individ[,c("participant_id","event_id","sample_id","event_date","event_type","event_location","respiratory_status")], 
  transplant,
  by.x="participant_id", by.y="studyid",
  all=FALSE
)

```

## Import
```{r}
#nasal transcriptomics data
nasal <- read.csv("/data/nasal-transcriptomics/current/nasal-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1
)
#metadata
nasal.metadata <- read.csv("/data/nasal-transcriptomics/current/nasal-transcriptomics-Metadata.csv",
  header = TRUE, row.names = 1
)

#gene info
genes=read.delim("/data/nasal-transcriptomics/current/nasal-transcriptomics-RowFeature.csv",header=T,sep=",",skip = 1)

#add entrez-id
ensembl=useMart("ensembl")
ensembl=useDataset("hsapiens_gene_ensembl", mart = ensembl)
new=getBM(attributes = c("ensembl_gene_id", "entrezgene_id","external_gene_name"),filters = "ensembl_gene_id",values=genes$gene_id,mart = ensembl)
genes$entrez=new$entrezgene_id[match(genes$gene_id,new$ensembl_gene_id)]


# Only keep samples that pass QC
nasal.metadata <- subset(nasal.metadata, sample_status=="IMPACC sample assayed and passed QC")
clin.sample.event.individ <- clin.sample.event.individ %>%
  subset(sample_id %in% rownames(nasal.metadata))
```

```{r}
# Filter nasal data to samples of interest
nasal <- nasal[clin.sample.event.individ$sample_id,]
```

## Import viral load data

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  # "/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv",
  "/data/nasal-metagenomics/legacy/2022-09-03/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)

mngs.metadata <- mngs.metadata %>%
  filter(!is.na(sc2_rpm))

# mNGS QC
mngs.metadata <- mngs.metadata %>%
  filter(sample_status == "IMPACC sample assayed and passed QC")
```

```{r}
# Reformat visit_num: decapitalize "VISIT"
mngs.metadata$visit_num <- gsub(
  "^VISIT", "Visit", mngs.metadata$visit_num
)

# Merge with sample metadata
clin.sample.event.individ <- merge(
  clin.sample.event.individ, mngs.metadata[,c("participant_id","visit_num","sc2_rpm")],
  by.x=c("participant_id","event_type"), by.y=c("participant_id","visit_num"),
  all.x=TRUE, all.y=FALSE # keep samples that don't pass SARS-CoV-2 metatranscriptomics QC
  # all=FALSE # discard samples that don't pass SARS-CoV-2 metatranscriptomics QC <<< do NOT do this
)

# log-transform viral load
clin.sample.event.individ$sc2_rpm_log10 <- log10(clin.sample.event.individ$sc2_rpm+1)
```

# Analysis: Visit 1 only
```{r}
# Sample metadata
clin.sample.event.individ.visit1 <- clin.sample.event.individ %>%
  subset(event_type == "Visit 1")

# Subset nasal data
nasal.visit1 <- nasal[clin.sample.event.individ.visit1$sample_id,]

```

```{r}
#run regression
dge=DGEList(t(nasal.visit1),group=clin.sample.event.individ.visit1$transplant_final,samples=clin.sample.event.individ.visit1)
dge=calcNormFactors(dge)
design=model.matrix(~group,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
vfit=lmFit(v,design)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
res.nasal=topTreat(tfit,coef=2,n=Inf)
res.nasal$nam=rownames(res.nasal)

#get normalised and log-transformed counts for downstream analysis
ge.nasal.visit1=cpm(dge,log=TRUE)

#repeat with vrld adjustment
dge=DGEList(t(nasal.visit1[!is.na(clin.sample.event.individ.visit1$sc2_rpm),]),group=clin.sample.event.individ.visit1$transplant_final[!is.na(clin.sample.event.individ.visit1$sc2_rpm)],samples=clin.sample.event.individ.visit1[!is.na(clin.sample.event.individ.visit1$sc2_rpm),])
dge=calcNormFactors(dge)
design=model.matrix(~group+sc2_rpm,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
vfit=lmFit(v,design)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
res.nasal.vrld=topTreat(tfit,coef=2,n=Inf)
res.nasal.vrld$nam=rownames(res.nasal.vrld)

```

```{r}
##pathway enrichment
res.nasal$entrez=genes$entrez[match(rownames(res.nasal),genes$gene_id)]
res.nasal$name=genes$gene_name[match(rownames(res.nasal),genes$gene_id)]
res.nasal.vrld$entrez=genes$entrez[match(rownames(res.nasal.vrld),genes$gene_id)]
res.nasal.vrld$name=genes$gene_name[match(rownames(res.nasal.vrld),genes$gene_id)]

#setup and run
res.nasal.genes2=res.nasal$logFC
names(res.nasal.genes2)=res.nasal$entrez
res.nasal.genes2=sort(res.nasal.genes2,decreasing = T)
res.nasal.gsea=gsePathway(res.nasal.genes2,maxGSSize=1000,minGSSize=3,organism="human",pvalueCutoff=0.1,pAdjustMethod="fdr")

res.nasal.vrld.genes2=res.nasal.vrld$logFC
names(res.nasal.vrld.genes2)=res.nasal.vrld$entrez
res.nasal.vrld.genes2=sort(res.nasal.vrld.genes2,decreasing = T)
res.nasal.vrld.gsea=gsePathway(res.nasal.vrld.genes2,maxGSSize=1000,minGSSize=3,organism="human",pvalueCutoff=0.1,pAdjustMethod="fdr")

#subset to top pathways, removing redundancy
#merge with results from vrld adjustment
tmp=res.nasal.gsea@result[c(1,4,5,7,20,24,27,71,74,100,110,123),] #requires manual curation
tmpv=res.nasal.vrld.gsea[match(tmp$Description,res.nasal.vrld.gsea@result$Description),]
tmpv$p.adjust[is.na(tmpv$Description)]=1 #if not enriched after vrld adjustment, make p-value 1
tmpv$NES[is.na(tmpv$Description)]=0 #as above, but make the NES zero
tmpv$Description[is.na(tmpv$Description)]=tmp$Description[is.na(tmpv$Description)]
tmp$control="No"
tmpv$control="Yes"
tmpp=rbind(tmp,tmpv)
tmpp$control=as.factor(tmpp$control)
tmpp$nam=tmpp$Description #these next lines might need editing depending on the pathways being plotted
#the purpose is to shorten the pathway names to make visualisation easier
tmpp$nam[12]="Antigen Presentation: Class I MHC"
tmpp$nam[24]="Antigen Presentation: Class I MHC"
tmpp$nam[5]="Interactions between Lymphoid:non-Lymphoid cell"
tmpp$nam[17]="Interactions between Lymphoid:non-Lymphoid cell"

#plot as barplot/Figure 6A and Supp. Figure 5C
ggplot(data=tmpp[tmpp$control=="No",], #switch to 'Yes' for vrld adjustment 
             aes(x=NES, y=reorder(nam, NES, FUN = function(x) x[1]))) +
  geom_col(aes(color=as.factor(NES>0)), fill=NA, width=0.7) +
  geom_col(
    data=. %>% subset(p.adjust<0.05),
    aes(fill=as.factor(NES>0)), width=0.7) +
  geom_vline(xintercept = 0, size = 0.4, color = "black") +
  #geom_text(
   # data=. %>% subset(NES>0),
    #aes(x=0,label=sprintf("P = %.1e", p.adjust)), 
    #hjust=1.06, size=2.8) +
  #geom_text(
   # data=. %>% subset(NES<0),
    #aes(x=0, label=sprintf("P = %.1e", p.adjust)),
    #hjust=-0.06, size=2.8) +
  labs(x="GSEA Normalized Enrichment Score", y="",
       title="Differential nasal pathways at Visit 1") +
  scale_x_continuous(
    limits=c(-3,3), expand=c(0,0)) +
  scale_color_manual(values=c("TRUE"="orange","FALSE"="cornflowerblue"),
                     guide="none") +
  scale_fill_manual(values=c("TRUE"="orange","FALSE"="cornflowerblue"),
                    guide="none") +
  #facet_wrap(~ control) + # separate into 2 panels: Yes/No w.r.t. controlling for viral load
  theme_bw() + theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank(),
    panel.spacing = unit(0.9, "cm"),
  )

```


```{r}
##compare nasal populations by trajectory group

#make split TG variable
clin.sample.event.individ.visit1$TG45=clin.sample.event.individ.visit1$trajectory_group>3

#run regression for control and tx separately
dge=DGEList(t(nasal.visit1[clin.sample.event.individ.visit1$transplant_final==0,]),group=clin.sample.event.individ.visit1$TG45[clin.sample.event.individ.visit1$transplant_final==0],samples=clin.sample.event.individ.visit1[clin.sample.event.individ.visit1$transplant_final==0,])
dge=calcNormFactors(dge)
design=model.matrix(~group,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
vfit=lmFit(v,design)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
res.nasal.ctrl=topTreat(tfit,coef=2,n=Inf)
res.nasal.ctrl$nam=rownames(res.nasal.ctrl)

dge=DGEList(t(nasal.visit1[clin.sample.event.individ.visit1$transplant_final==1,]),group=clin.sample.event.individ.visit1$TG45[clin.sample.event.individ.visit1$transplant_final==1],samples=clin.sample.event.individ.visit1[clin.sample.event.individ.visit1$transplant_final==1,])
dge=calcNormFactors(dge)
design=model.matrix(~group,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
vfit=lmFit(v,design)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
res.nasal.tx=topTreat(tfit,coef=2,n=Inf)
res.nasal.tx$nam=rownames(res.nasal.tx)

##pathway enrichment
res.nasal.ctrl$entrez=genes$entrez[match(rownames(res.nasal.ctrl),genes$gene_id)]
res.nasal.ctrl$name=genes$gene_name[match(rownames(res.nasal.ctrl),genes$gene_id)]
res.nasal.tx$entrez=genes$entrez[match(rownames(res.nasal.tx),genes$gene_id)]
res.nasal.tx$name=genes$gene_name[match(rownames(res.nasal.tx),genes$gene_id)]

#setup and run
res.nasal.genes2c=res.nasal.ctrl$logFC
names(res.nasal.genes2c)=res.nasal.ctrl$entrez
res.nasal.genes2c=sort(res.nasal.genes2c,decreasing = T)
res.nasal.gsea.ctrl=gsePathway(res.nasal.genes2c,maxGSSize=1000,minGSSize=3,organism="human",pvalueCutoff=0.1,pAdjustMethod="fdr")

res.nasal.genes2t=res.nasal.tx$logFC
names(res.nasal.genes2t)=res.nasal.tx$entrez
res.nasal.genes2t=sort(res.nasal.genes2t,decreasing = T)
res.nasal.gsea.tx=gsePathway(res.nasal.genes2t,maxGSSize=1000,minGSSize=3,organism="human",pvalueCutoff=0.1,pAdjustMethod="fdr")

#pick a subset of pathways and combine
#next two lines pick top pathways from each enrichment analysis, removing redundancy
tmp=res.nasal.gsea.ctrl@result[c(1,4,9,10,13,14,19,34,35,41,45,47,75,89,127,243),]
tmpv=res.nasal.gsea.tx@result[c(1,14,30,36,51,54,58,104),]
tmpp=unique(c(tmp$Description,tmpv$Description))

tmp=res.nasal.gsea.ctrl@result[match(tmpp,res.nasal.gsea.ctrl@result$Description),]
tmpv=res.nasal.gsea.tx@result[match(tmpp,res.nasal.gsea.tx@result$Description),]
tmpv$p.adjust[is.na(tmpv$Description)]=1 #if not enriched in transplant, make p-value 1
tmpv$NES[is.na(tmpv$Description)]=0 #as above, but make the NES zero
tmpv$Description[is.na(tmpv$Description)]=tmp$Description[is.na(tmpv$Description)]
tmp$p.adjust[is.na(tmp$Description)]=1 #as above, but for controls
tmp$NES[is.na(tmp$Description)]=0
tmp$Description[is.na(tmp$Description)]=tmpv$Description[is.na(tmp$Description)]
tmp$control="Control"
tmpv$control="Transplant"
tmpp=rbind(tmp,tmpv)
tmpp$control=as.factor(tmpp$control)
tmpp$nam=tmpp$Description #shorten names where necessary
tmpp$nam[7]="Interactions between Lymphoid:non-Lymphoid cell"
tmpp$nam[30]="Interactions between Lymphoid:non-Lymphoid cell"

#plot as barplot/Figure 7D
ggplot(data=tmpp,
aes(x=NES, y=reorder(nam, NES, FUN = function(x) x[1]))) +
geom_col(aes(color=as.factor(NES>0)), fill=NA, width=0.7) +
geom_col(
data=. %>% subset(p.adjust<0.05),
aes(fill=as.factor(NES>0)), width=0.7) +
geom_vline(xintercept = 0, size = 0.4, color = "black") +
geom_text(
data=. %>% subset(NES>0),
aes(x=0,label=sprintf("P = %.1e", p.adjust)),
hjust=1.06, size=2.8) +
geom_text(
data=. %>% subset(NES<0),
aes(x=0, label=sprintf("P = %.1e", p.adjust)),
hjust=-0.06, size=2.8) +
labs(x="GSEA Normalized Enrichment Score", y="",
title="Nasal pathways associated\nwith severity at Visit 1") +
scale_x_continuous(
limits=c(-3,3.3), expand=c(0,0)) +
scale_color_manual(values=c("TRUE"="coral2","FALSE"="chartreuse3"),
guide="none") +
scale_fill_manual(values=c("TRUE"="coral2","FALSE"="chartreuse3"),
guide="none") +
facet_wrap(~ control) + # separate into 2 panels: Yes/No w.r.t. controlling for viral load
theme_bw() + theme(
plot.title = element_text(hjust = 0.5, size=12, face="plain"),
axis.text.x = element_text(size=11, color="black"),
axis.text.y = element_text(size=11, color="black"),
text = element_text(size=12, family="Arial"),
plot.margin = unit(c(0,1,1,0), "cm"),
panel.grid.minor=element_blank(),
panel.spacing = unit(0.9, "cm"),
)
```


# Analysis: vrld vs IFN signaling by transplant status
```{r}
#read in genes in IFN signaling pathway
ifns=read.delim("/scratch/immunosuppression/2023.08.17_ifn_signaling.txt",header=T,sep="\t")
nasal.visit1.ifns=nasal.visit1[,colnames(nasal.visit1)%in%ifns$ensembl]

#for controls and Tx, run regression to determine the association between IFN genes and vrld
gectrl=nasal.visit1.ifns[!is.na(clin.sample.event.individ.visit1$sc2_rpm_log10) & clin.sample.event.individ.visit1$transplant_final==0,]
infoctrl=clin.sample.event.individ.visit1[!is.na(clin.sample.event.individ.visit1$sc2_rpm_log10) & clin.sample.event.individ.visit1$transplant_final==0,]

dge=DGEList(t(gectrl),samples=infoctrl)
dge=calcNormFactors(dge)
design=model.matrix(~sc2_rpm_log10,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
vfit=lmFit(v,design)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
nasalres.ctrl.limma=topTreat(tfit,coef=2,n=Inf)
nasalres.ctrl.limma$nam=rownames(nasalres.ctrl.limma)
gectrl.cpm=cpm(dge,log=TRUE)

getx=nasal.visit1.ifns[!is.na(clin.sample.event.individ.visit1$sc2_rpm_log10) & clin.sample.event.individ.visit1$transplant_final==1,]
infotx=clin.sample.event.individ.visit1[!is.na(clin.sample.event.individ.visit1$sc2_rpm_log10) & clin.sample.event.individ.visit1$transplant_final==1,]

dge=DGEList(t(getx),samples=infotx)
dge=calcNormFactors(dge)
design=model.matrix(~sc2_rpm_log10,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
vfit=lmFit(v,design)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
nasalres.tx.limma=topTreat(tfit,coef=2,n=Inf)
nasalres.tx.limma$nam=rownames(nasalres.tx.limma)
getx.cpm=cpm(dge,log=TRUE)

```

```{r}
#setup plot
nasalres.tx.limma.matched=nasalres.tx.limma[match(rownames(nasalres.ctrl.limma),rownames(nasalres.tx.limma)),]
nasalres.ctrl.limma$col="black"
nasalres.ctrl.limma$bg="azure3"
nasalres.ctrl.limma$bg[nasalres.ctrl.limma$adj.P.Val<0.05]="cornflowerblue"
nasalres.ctrl.limma$name=genes$gene_name[match(nasalres.ctrl.limma$nam,genes$gene_id)]

#plot/Supp. Figure 7B
plot(nasalres.ctrl.limma$logFC,nasalres.tx.limma.matched$logFC,pch=21,xlab="Control - Log2(Fold-change)",ylab="Transplant - Log2(Fold-change)",xlim=c(-0.65,0.65),ylim=c(-0.65,0.65),type='n')
abline(h=0,v=0,lty=3,col=adjustcolor("azure3",0.8))
par(new=T)
plot(nasalres.ctrl.limma$logFC[nasalres.ctrl.limma$bg!="cornflowerblue"],nasalres.tx.limma.matched$logFC[nasalres.ctrl.limma$bg!="cornflowerblue"],pch=21,bg=nasalres.ctrl.limma$bg[nasalres.ctrl.limma$bg!="cornflowerblue"],col=nasalres.ctrl.limma$col[nasalres.ctrl.limma$bg!="cornflowerblue"],xlab="",ylab="",xlim=c(-0.65,0.65),ylim=c(-0.65,0.65))
par(new=T)
plot(nasalres.ctrl.limma$logFC[nasalres.ctrl.limma$bg=="cornflowerblue"],nasalres.tx.limma.matched$logFC[nasalres.ctrl.limma$bg=="cornflowerblue"],pch=21,bg=nasalres.ctrl.limma$bg[nasalres.ctrl.limma$bg=="cornflowerblue"],col=nasalres.ctrl.limma$col[nasalres.ctrl.limma$bg=="cornflowerblue"],xlab="",ylab="",xlim=c(-0.65,0.65),ylim=c(-0.65,0.65))
abline(a=0,b=1)

#plot examples/Supp. Figure 7D
gege=rbind(gectrl,getx)
gege2=data.frame(t(cbind(gectrl.cpm,getx.cpm)))
infoinfo=rbind(infoctrl,infotx)

infoinfo$txtx="Control"
infoinfo$txtx[infoinfo$transplant_final==1]="Transplant"

z=nasalres.ctrl.limma$nam[nasalres.ctrl.limma$adj.P.Val<0.05]
zzzz=gege2[,z]

zz=nasalres.ctrl.limma$name[match(z,nasalres.ctrl.limma$nam)]
zsub=z[c(4,8,22)] #manually select few genes to plot

p <- list()
for (i in zsub) {

ii=zz[match(i,z)]

temp <- data.frame(
value = zzzz[,i],
event_date = infoinfo$event_date,
txstatus = infoinfo$txtx,
sc2 = infoinfo$sc2_rpm_log10,
#sex = clin.sample.event.individ.long$sex,
#tg = clin.sample.event.individ.long$trajectory_group,
#site = clin.sample.event.individ.long$enrollment_site,
pid = infoinfo$participant_id
) %>%
subset(!is.na(value) & !is.na(sc2) & !is.infinite(value)) %>%
droplevels()
# Predict
fit <- lm(
value ~ sc2 + sc2:txstatus + txstatus,
data = temp)
my.pred <- ggeffects::ggpredict(fit, terms=c("sc2","txstatus"),
type="fixed")
colnames(my.pred)[colnames(my.pred)=="group"] <- "txstatus"
temp$yhat <- predict(fit)
# Calculate IQR and quartile values for limiting y-axes
q1 <- quantile(temp$value, 0.25)
q3 <- quantile(temp$value, 0.75)
iqr <- IQR(temp$value)
p[[i]]=ggplot(data = temp, aes(x=sc2, y=value, color=txstatus)) +
geom_line(aes(group=pid), color="gray", alpha=0.4) +
#geom_line(inherit.aes=FALSE, aes(x=event_date, y=yhat, group=pid),
#         color="#5F5F5F", alpha=0.9) +
geom_point(aes(color="black"), shape=20, size=1) +
stat_smooth(method = "lm", formula = y ~ x,aes(color=txstatus,fill=txstatus),linewidth=1.5) +
#geom_line(inherit.aes=FALSE, data=my.pred, aes(x,predicted),
#         color="black", size=1.5) +
coord_cartesian(ylim = c(q1-1.5*iqr, q3+1.5*iqr)) +
labs(x="SARS-CoV-2 log10(rpM)", y="log2(normalized counts)",
title=paste0(ii,": ",sprintf("%s-P=%.1e","Ctrl",nasalres.ctrl.limma$adj.P.Val[nasalres.ctrl.limma$nam==i])," / ",sprintf("%s-P=%.1e","Tx",nasalres.tx.limma.matched$adj.P.Val[nasalres.ctrl.limma$nam==i]))) +
scale_color_viridis_d(begin=0.4, guide="none") +
# scale_color_brewer(palette="Paired", guide="none") +
scale_color_manual(values=c("black","cornflowerblue","orange")) +
scale_fill_manual(values=c("cornflowerblue","orange")) +
# scale_color_discrete(guide="none") +
facet_wrap(~ txstatus, nrow=1) + theme_bw() +
theme(
plot.title = element_text(hjust = 0.5, size=11, face="plain"),
axis.text.x = element_text(size=10, color="black"),
axis.text.y = element_text(size=10, color="black"),
text = element_text(size=12, family="Arial"),
plot.margin = unit(c(0.2,1,0.7,0.2), "cm"),
panel.spacing = unit(0.35, "cm"),
legend.position = "none",
panel.grid.minor=element_blank()
)
}

```

# Analysis: vrld vs all GE by transplant status
```{r}
#for controls and Tx, run regression to determine the association between all genes and vrld
gectrl=nasal.visit1[!is.na(clin.sample.event.individ.visit1$sc2_rpm_log10) & clin.sample.event.individ.visit1$transplant_final==0,]
infoctrl=clin.sample.event.individ.visit1[!is.na(clin.sample.event.individ.visit1$sc2_rpm_log10) & clin.sample.event.individ.visit1$transplant_final==0,]

dge=DGEList(t(gectrl),samples=infoctrl)
dge=calcNormFactors(dge)
design=model.matrix(~sc2_rpm_log10,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
vfit=lmFit(v,design)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
nasalres.ctrl.limma=topTreat(tfit,coef=2,n=Inf)
nasalres.ctrl.limma$nam=rownames(nasalres.ctrl.limma)
gectrl.cpm=cpm(dge,log=TRUE)

getx=nasal.visit1[!is.na(clin.sample.event.individ.visit1$sc2_rpm_log10) & clin.sample.event.individ.visit1$transplant_final==1,]
infotx=clin.sample.event.individ.visit1[!is.na(clin.sample.event.individ.visit1$sc2_rpm_log10) & clin.sample.event.individ.visit1$transplant_final==1,]

dge=DGEList(t(getx),samples=infotx)
dge=calcNormFactors(dge)
design=model.matrix(~sc2_rpm_log10,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
vfit=lmFit(v,design)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
nasalres.tx.limma=topTreat(tfit,coef=2,n=Inf)
nasalres.tx.limma$nam=rownames(nasalres.tx.limma)
getx.cpm=cpm(dge,log=TRUE)

#get pathways and plot
nres.genes=nasalres.ctrl.limma$logFC
names(nres.genes)=genes$entrez[match(nasalres.ctrl.limma$nam,genes$gene_id)]
nres.genes=sort(nres.genes,decreasing = T)
nres.ctrl.gsea=gsePathway(nres.genes,maxGSSize=1000,minGSSize=3,organism="human",pvalueCutoff=1,pAdjustMethod="fdr")
nres.genes=nasalres.tx.limma$logFC
names(nres.genes)=genes$entrez[match(nasalres.tx.limma$nam,genes$gene_id)]
nres.genes=sort(nres.genes,decreasing = T)
nres.tx.gsea=gsePathway(nres.genes,maxGSSize=1000,minGSSize=3,organism="human",pvalueCutoff=1,pAdjustMethod="fdr")

nc=c(1,2,3,5,8,9,11,12,13,16,18,20,24,33,38,40,47,48,51,67)
nt=c(6,7,16,17,18,19,21,24,26,27,29,30,31,40,45,52,54,56,61,62)
ncp=nres.ctrl.gsea@result$Description[nc]
ntp=nres.tx.gsea@result$Description[nt]
pways=unique(c(ncp,ntp))

ncpall=nres.ctrl.gsea@result[match(pways,nres.ctrl.gsea@result$Description),]
ncpall$NES[is.na(ncpall$NES)]=0
ncpall$p.adjust[is.na(ncpall$p.adjust)]=1
ncpall$status="Non-SOT"
ncpall$Description=pways

ntpall=nres.tx.gsea@result[match(pways,nres.tx.gsea@result$Description),]
ntpall$NES[is.na(ntpall$NES)]=0
ntpall$p.adjust[is.na(ntpall$p.adjust)]=1
ntpall$status="SOT"
ntpall$Description=pways

ncombo=rbind(ncpall,ntpall)

ggplot(data=ncombo,
       aes(x=NES, y=reorder(Description, NES, FUN = function(x) x[1]))) +
  geom_col(aes(color=as.factor(NES>0)), fill=NA, width=0.7) +
  geom_col(
    data=. %>% subset(p.adjust<0.05),
    aes(fill=as.factor(NES>0)), width=0.7) +
  geom_vline(xintercept = 0, size = 0.4, color = "black") +
  #geom_text(
   # data=. %>% subset(NES>0),
    #aes(x=0,label=sprintf("P = %.1e", p.adjust)),
    #hjust=1.06, size=2.8) +
  #geom_text(
   # data=. %>% subset(NES<0),
    #aes(x=0, label=sprintf("P = %.1e", p.adjust)),
    #hjust=-0.06, size=2.8) +
  labs(x="GSEA Normalized Enrichment Score", y="",
       title="Nasal pathways associated \nwith SARS-CoV-2 rpM at Visit 1") +
  scale_x_continuous(
    limits=c(-3.1,3.3), expand=c(0,0)) +
  scale_color_manual(values=c("TRUE"="azure4","FALSE"="azure3"),
                     guide="none") +
  scale_fill_manual(values=c("TRUE"="azure4","FALSE"="azure3"),
                    guide="none") +
  facet_wrap(~ status) + # separate into 2 panels: Yes/No w.r.t. controlling for viral load
  theme_bw() + theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank(),
    panel.spacing = unit(0.9, "cm"),
  )

```

# Analysis: longitudinal

```{r}
#Sample metadata
clin.sample.event.individ.long <- clin.sample.event.individ %>%
  subset(event_type %in% c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6"))

# Try restricting to 30 days post admission
clin.sample.event.individ.long <- clin.sample.event.individ.long %>%
  subset(event_date <= 30)

nasal.long <- nasal[clin.sample.event.individ.long$sample_id,]

```

```{r}
#run regression
dge=DGEList(t(nasal.long),group=clin.sample.event.individ.long$transplant_final,samples=clin.sample.event.individ.long)
dge=calcNormFactors(dge)
design=model.matrix(~group,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
dupcor=duplicateCorrelation(v,design,block=dge2$samples$participant_id)
v2=voom(dge2,design,block=dge2$samples$studyid,correlation=dupcor$consensus)
dupcor=duplicateCorrelation(v2,design,block=dge2$samples$participant_id)
vfit=lmFit(v2,design,block=dge2$samples$participant_id, correlation=dupcor$consensus)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
res.nasal.long=topTreat(tfit,coef=2,n=Inf)
res.nasal.long$nam=rownames(res.nasal.long)

#get normalised and log-transformed counts for downstream analysis
ge.nasal.long=cpm(dge,log=TRUE)

#repeat with vrld adjustment
dge=DGEList(t(nasal.long[!is.na(clin.sample.event.individ.long$sc2_rpm),]),group=clin.sample.event.individ.long$transplant_final[!is.na(clin.sample.event.individ.long$sc2_rpm)],samples=clin.sample.event.individ.long[!is.na(clin.sample.event.individ.long$sc2_rpm),])
dge=calcNormFactors(dge)
design=model.matrix(~group+sc2_rpm,data=dge$samples)
dge2=dge[filterByExpr(dge),]
v=voom(dge2,design)
dupcor=duplicateCorrelation(v,design,block=dge2$samples$participant_id)
v2=voom(dge2,design,block=dge2$samples$studyid,correlation=dupcor$consensus)
dupcor=duplicateCorrelation(v2,design,block=dge2$samples$participant_id)
vfit=lmFit(v2,design,block=dge2$samples$participant_id, correlation=dupcor$consensus)
efit=eBayes(vfit)
tfit=treat(vfit,lfc=0)
res.nasal.long.vrld=topTreat(tfit,coef=2,n=Inf)
res.nasal.long.vrld$nam=rownames(res.nasal.long.vrld)

```

```{r}
##pathway enrichment
res.nasal.long$entrez=genes$entrez[match(rownames(res.nasal.long),genes$gene_id)]
res.nasal.long$name=genes$gene_name[match(rownames(res.nasal.long),genes$gene_id)]
res.nasal.long.vrld$entrez=genes$entrez[match(rownames(res.nasal.long.vrld),genes$gene_id)]
res.nasal.long.vrld$name=genes$gene_name[match(rownames(res.nasal.long.vrld),genes$gene_id)]

#setup and run
res.nasal.long.genes2=res.nasal.long$logFC
names(res.nasal.long.genes2)=res.nasal.long$entrez
res.nasal.long.genes2=sort(res.nasal.long.genes2,decreasing = T)
res.nasal.long.gsea=gsePathway(res.nasal.long.genes2,maxGSSize=1000,minGSSize=3,organism="human",pvalueCutoff=0.1,pAdjustMethod="fdr")

res.nasal.long.vrld.genes2=res.nasal.long.vrld$logFC
names(res.nasal.long.vrld.genes2)=res.nasal.long.vrld$entrez
res.nasal.long.vrld.genes2=sort(res.nasal.long.vrld.genes2,decreasing = T)
res.nasal.long.vrld.gsea=gsePathway(res.nasal.long.vrld.genes2,maxGSSize=1000,minGSSize=3,organism="human",pvalueCutoff=0.1,pAdjustMethod="fdr")

#subset to top pathways, removing redundancy
#merge with results from vrld adjustment
tmp=res.nasal.long.gsea@result[c(1,2,3,5,11,13,18,22,51,56,64,95),]
tmpv=res.nasal.long.vrld.gsea[match(tmp$Description,res.nasal.long.vrld.gsea@result$Description),]
tmpv$p.adjust[is.na(tmpv$Description)]=1
tmpv$NES[is.na(tmpv$Description)]=0
tmpv$Description[is.na(tmpv$Description)]=tmp$Description[is.na(tmpv$Description)]
tmp$gene=NULL
tmpv$keep=tmp$keep
tmp$control="No"
tmpv$control="Yes"
tmpp=rbind(tmp,tmpv)
tmpp$control=as.factor(tmpp$control)
tmpp$nam=tmpp$Description
tmpp$nam[12]="Antigen activates BCR"
tmpp$nam[24]="Antigen activates BCR"
tmpp$nam[2]="Interactions between Lymphoid:non-Lymphoid cell"
tmpp$nam[14]="Interactions between Lymphoid:non-Lymphoid cell"

#plot/Figure 6B and Supp. Figure 5D
kpord=tmpp$nam[1:12][order(tmpp$NES[1:12])]
kpord2=match(tmpp$nam[1:12],kpord)

ggplot(data=tmpp[tmpp$control=="No",], #switch to 'Yes' for vrld adjustment
aes(x=NES, y=reorder(nam, NES, FUN = function(x) x[1]))) +
geom_point(shape=21, size=1,alpha=0)+
geom_segment(aes(xend=NES, x=0,yend=kpord2),size=0.25) +
geom_point(shape=21,
aes(color=as.factor(NES>0)), size=4) +
geom_point(data=. %>% subset(p.adjust<0.05),
aes(color=as.factor(NES>0)), fill=NA, size=4) +
geom_vline(xintercept = 0, size = 0.4, color = "black") +
geom_text(
data=. %>% subset(NES>0),
aes(x=0,label=sprintf("P = %.1e", p.adjust)),
hjust=1.06, size=2.8) +
geom_text(
data=. %>% subset(NES<0),
aes(x=0, label=sprintf("P = %.1e", p.adjust)),
hjust=-0.06, size=2.8) +
labs(x="GSEA Normalized Enrichment Score", y="",
title="Longitudinal dynamics of nasal pathways") +
scale_x_continuous(
limits=c(-3.1,3.1), expand=c(0,0)) +
scale_color_manual(values=c("TRUE"="orange","FALSE"="cornflowerblue"),
guide="none") +
scale_fill_manual(values=c("TRUE"="orange","FALSE"="cornflowerblue"),
guide="none") +
#facet_wrap(~ control) + # separate into 2 panels: Yes/No w.r.t. controlling for viral load
theme_bw() + theme(
plot.title = element_text(hjust = 0.5, size=12, face="plain"),
axis.text.x = element_text(size=11, color="black"),
axis.text.y = element_text(size=11, color="black"),
text = element_text(size=12, family="Arial"),
plot.margin = unit(c(0,1,1,0), "cm"),
panel.grid.minor=element_blank(),
panel.spacing = unit(0.9, "cm"),
)

```

```{r}
#example longitudinal plots
z=unique(tmpp$Description)#get the pathways from the earlier plot
zzzz=data.frame(matrix(nrow=ncol(ge.nasal.long),ncol=length(z)))
colnames(zzzz)=z
for(i in 1:length(z)){zz=ge.nasal.long[genes$gene_id[match(unlist(strsplit(res.nasal.long.gsea@result$core_enrichment[res.nasal.long.gsea@result$Description==z[i]],"/")),genes$entrez)],]
zzz=prcomp(t(zz),scale=T)
if(mean(cor(t(zz),zzz$x[,1]))<0)
{
  zzzz[,i]=zzz$x[,1]*-1
}else{
  zzzz[,i]=zzz$x[,1]}
}

clin.sample.event.individ.long$tx="Control"
clin.sample.event.individ.long$tx[clin.sample.event.individ.long$transplant_final==1]="Transplant"

zz=z[c(4,8,10)]

#plot pathways/Figure 6C
p <- list()
for (i in zz) {
  temp <- data.frame(
    value = scale(zzzz[,i]),
    event_date = clin.sample.event.individ.long$event_date,
    txstatus = clin.sample.event.individ.long$tx,
    #sc2 = tx.all$sc2_rpm,
    #sex = clin.sample.event.individ.long$sex,
    #tg = clin.sample.event.individ.long$trajectory_group,
    #site = clin.sample.event.individ.long$enrollment_site,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value) & !is.na(event_date)) %>%
    droplevels()
  # Predict
  fit <- lmer(
    value ~ event_date + event_date:txstatus + txstatus +(1|pid),
    data = temp, REML = FALSE)
  my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","txstatus"),
                                  type="fixed")
  colnames(my.pred)[colnames(my.pred)=="group"] <- "txstatus"
  temp$yhat <- predict(fit)
  # Calculate IQR and quartile values for limiting y-axes
  q1 <- quantile(temp$value, 0.25)
  q3 <- quantile(temp$value, 0.75)
  iqr <- IQR(temp$value)
  p[[i]] <- ggplot(data = temp, aes(x=event_date, y=value, color=txstatus)) +
    geom_line(aes(group=pid), color="gray", alpha=0.4) +
    #geom_line(inherit.aes=FALSE, aes(x=event_date, y=yhat, group=pid),
     #         color="#5F5F5F", alpha=0.9) +
    geom_point(aes(color="black"), shape=20, size=1) +
    geom_line(inherit.aes=FALSE, data=my.pred, aes(x,predicted,color=txstatus),
              size=1.5) +
    geom_ribbon(inherit.aes=FALSE,data=my.pred,aes(x, predicted, ymax=conf.high, ymin=conf.low,fill=txstatus),alpha=0.2) +
    coord_cartesian(ylim = c(q1-1.5*iqr, q3+1.5*iqr)) +
    labs(x="Days from admission", y="Normalized expression",
         title=sprintf("%s (P=%.1e)",i,res.nasal.long.gsea@result$p.adjust[res.nasal.long.gsea@result$Description==i])) +
    scale_color_viridis_d(begin=0.4, guide="none") +
    # scale_color_brewer(palette="Paired", guide="none") +
scale_color_manual(values=c("black","cornflowerblue","orange")) +
scale_fill_manual(values=c("cornflowerblue","orange")) +
    # scale_color_discrete(guide="none") +
    facet_wrap(~ txstatus, nrow=1) + theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5, size=11, face="plain"),
      axis.text.x = element_text(size=10, color="black"),
      axis.text.y = element_text(size=10, color="black"),
      text = element_text(size=12, family="Arial"),
      plot.margin = unit(c(0.2,1,0.7,0.2), "cm"),
      panel.spacing = unit(0.35, "cm"),
      legend.position = "none",
      panel.grid.minor=element_blank()
    )
}

```