---
title: "blood_cytof"
output: html_document
editor_options: 
  chunk_output_type: inline
---

Source: 20230124_cytof.Rmd

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

library(gamm4)

# ggplot theme
my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, family="Arial", color="black"),
    text = element_text(size=12, family="Arial", color="black"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```

# Import data

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

### Clinical metadata processing

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE
clin.event$steroids <- as.factor(clin.event$steroids)

head(clin.event)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert sex to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

### Merge clinical metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ, by = "participant_id", all = FALSE, sort = FALSE
)
```

## Blood sample

```{r}
# Count data
bld <- read.csv(
  "/data/bld-cytof/current/bld-cytof-Counts.csv",
  skip = 1, header = FALSE, row.names = 1
)
colnames(bld) <- 1:ncol(bld)
bld <- t(bld)

# Cell type data
bld.rData <- read.csv(
  "/data/bld-cytof/current/bld-cytof-RowFeature.csv",
  skip = 1, header = TRUE, row.names = NULL
)

# QC data
bld.metadata <- read.csv(
  "/data/bld-cytof/current/bld-cytof-Metadata.csv",
  header = TRUE, row.names = NULL
)
```

```{r}
# Sum up the cell count
bld.2 <- rowsum(bld, bld.rData$Broad.Parent.Population)

# Only keep samples that pass QC
bld.metadata <- subset(bld.metadata, sample_status=="IMPACC sample assayed and passed QC")
bld.pass.qc <- Reduce(
  intersect, 
  list(colnames(bld), bld.metadata$sample_id, clin.sample.event.individ$sample_id)
)
bld.2 <- bld.2[, bld.pass.qc]

# Replace NA with 0
bld.2[is.na(bld.2)] <- 0

# Remove neutrophils
bld.2 <- bld.2[rownames(bld.2) != "Neutrophil",]
# Remove multiplets
bld.2 <- bld.2[!grepl("^multiplet", rownames(bld.2), ignore.case = TRUE),]
# Remove debris
bld.2 <- bld.2[!grepl("^debris", rownames(bld.2), ignore.case = TRUE),]
# Remove Undefined cells
bld.2 <- bld.2[rownames(bld.2) != "Undefined",]
```

Fine cell type annotation.

```{r}
# Remove neutrophils
temp <- subset(bld.rData, Final.Annotated.Population != "Neutrophil (CD16hi)")
# Remove multiplets
temp <- temp[!grepl("^multiplet", temp$Final.Annotated.Population, ignore.case = TRUE),]
# Remove debris
temp <- temp[!grepl("^debris", temp$Final.Annotated.Population, ignore.case = TRUE),]
# Remove Undefined cells, including undefined T cells
temp <- temp[!grepl("Undefined", temp$Final.Annotated.Population, ignore.case = TRUE),]

bld.2f <- bld[rownames(temp),]
```

Do I need to filter more cell types? For each cell type, check the number of patients in which that cell type is detected.

```{r}
# Histogram of number of detected patients per cell type
hist(rowSums(bld.2>0))
```

Blood sample clinical metadata.

```{r}
bld.clin.sample.event.individ <- clin.sample.event.individ %>%
  subset(sample_id %in% bld.pass.qc)
```

## Viral load data

SARS-CoV-2 viral load is readily available in nasal-metagenomics-Metadata.csv, sc2_rpm column.

```{r}

# mNGS QC
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")
# Remove rows without sc2_rpm
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))
```

# Visit 1 analysis

```{r}
# Keep visit 1 samples for viral load
clin.sample.event.individ.sc2.1 <- clin.sample.event.individ.sc2 %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.sc2.1$sc2_rpm <- log10(clin.sample.event.individ.sc2.1$sc2_rpm + 1)

# Blood
bld.clin.sample.event.individ.1 <- bld.clin.sample.event.individ %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")
# Add viral load
bld.clin.sample.event.individ.1 <- merge(
  bld.clin.sample.event.individ.1, clin.sample.event.individ.sc2.1[,c("participant_id","sc2_rpm")],
  by = "participant_id", all = FALSE, sort = FALSE
)
bld.clin.sample.event.individ.1 <- droplevels(bld.clin.sample.event.individ.1)

# EA
# ea.clin.sample.event.individ.1 <- ea.clin.sample.event.individ %>%
#   subset(event_type=="Visit 1" & event_location=="Hospital")
# # Add viral load
# ea.clin.sample.event.individ.1 <- merge(
#   ea.clin.sample.event.individ.1, clin.sample.event.individ.sc2.1[,c("participant_id","sc2_rpm")],
#   by = "participant_id", all = FALSE, sort = FALSE
# )
# ea.clin.sample.event.individ.1 <- droplevels(ea.clin.sample.event.individ.1)

# Subset samples
bld.3 <- bld.2[,bld.clin.sample.event.individ.1$sample_id]
# ea.3 <- ea.2[,ea.clin.sample.event.individ.1$sample_id]
```

## Linear regression on age effect: blood

### Fine cell type annotation

```{r}
# Convert cell count to cell type proportion
bld.3f <- apply(bld.2f, MARGIN = 2,
                FUN = function(x) x/sum(x)) # t(t(bld.2f)/colSums(bld.2f)) 

# Subset visit 1 samples
bld.3f <- bld.3f[, bld.clin.sample.event.individ.1$sample_id]

# CLR transform
bld.4f <- log(bld.3f*100 + 1) # Add 1 to avoid log of zero
bld.4f <- apply(bld.4f, MARGIN = 2,
                FUN = function(x) x - mean(x)) # t(t(bld.4f) - colMeans(bld.4f))
```

### Without controlling for viral load

```{r}
# Linear regression on clr-transformed cell type proportion
res.clr <- data.frame(matrix(NA, nrow=nrow(bld.4f), ncol=4))
colnames(res.clr) <- c("cell.type","slope","PVal","adj.PVal")
rownames(res.clr) <- rownames(bld.4f)
res.clr$cell.type <- bld.rData[rownames(res.clr),"Final.Annotated.Population"]

for (i in rownames(res.clr)) {
  temp <- data.frame(clr=bld.4f[i,], 
                     age=bld.clin.sample.event.individ.1$admit_age,
                     sex=bld.clin.sample.event.individ.1$sex,
                     sc2=bld.clin.sample.event.individ.1$sc2_rpm,
                     resp=bld.clin.sample.event.individ.1$respiratory_status)%>%
    subset(!is.na(clr))
  my.lm <- lm(clr ~ age + sex + resp, data=temp)
  res.clr[i,"slope"] <- summary(my.lm)$coefficients["age",1]
  res.clr[i,"PVal"] <- summary(my.lm)$coefficients["age",4]
  
  remove(temp, my.lm)
}

# Correct P-values
res.clr$adj.PVal <- p.adjust(res.clr$PVal, method = "BH")

# Add cell type
res.clr$cell.type <- bld.rData[rownames(res.clr),"Final.Annotated.Population"]
```

Plot slopes of significant cell types.

```{r}
# Figure 3b
p <- ggplot() +
  geom_col(data = subset(res.clr, adj.PVal<0.05),
           aes(x = slope, y = reorder(cell.type,slope),
               fill = as.factor(slope>0)),
           width = 0.7) +
  geom_vline(xintercept=0, size=0.5, color="black") +
  scale_fill_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  # coord_flip() +
  labs(x = "Slope\n(change per 1 year of age)",
       y = "", title = "Cell types associated\nwith age at Visit 1") +
  xlim(-0.022, 0.022) + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank()
  ) # cytof_visit1_age_no-viralload, 6 x 6.2 in (w x h)
```

Plot examples.

```{r}
# Figure 3c
p <- list()
counter <- 1
my.colors <- c("darkred","darkred","dodgerblue4","dodgerblue4")
for (i in c("Monocytes (CD14+CD16+)","NK Cell (CD56low CD16hi CD57hi)","CD8+ T Cell (naive)","B Cell (naive)")) {
  # Row number that corresponds to cell type i
  row.i <- rownames(bld.rData)[bld.rData$Final.Annotated.Population==i]
  temp <- data.frame(clr=bld.4f[row.i,], 
                     age=bld.clin.sample.event.individ.1$admit_age,
                     sex=bld.clin.sample.event.individ.1$sex,
                     # steroids=bld.clin.sample.event.individ.1$steroids,
                     sc2=bld.clin.sample.event.individ.1$sc2_rpm,
                     resp=bld.clin.sample.event.individ.1$respiratory_status) %>%
    subset(!is.na(clr))
  my.lm <- lm(clr ~ age + sex + resp, data=temp)
  my.pred <- ggeffects::ggpredict(my.lm, "age")
  
  p[[i]] <- ggplot() +
    geom_point(data = temp, aes(x=age,y=clr), size = 0.5, color="black") +
    geom_line(data = my.pred, aes(x,predicted), size = 1, color=my.colors[counter]) +
    geom_ribbon(data = my.pred, 
                aes(x = x, y = predicted, ymin = conf.low, ymax = conf.high),
                fill = my.colors[counter], alpha = .4) +
    # annotate(geom = "text", label = sprintf("P = %.1e", summary(my.lm)$coefficients["age",4]),
    #          x = 0.7*min(temp$age), y = 1.2*min(temp$clr), hjust=0) +
    scale_x_continuous(limits=c(0,100), expand=c(0,0)) +
    scale_y_continuous(limits=c(-1,max(temp$clr))) +
    labs(x = "Age",
         y = "Cell type abundance\n(CLR-transformed)",
         title = sprintf("%s",i)) +
    theme_classic() + 
    theme(
      plot.title = element_text(hjust = 0.5, size=12, face="plain"),
      axis.text = element_text(size=12, family="Arial", color="black"),
      text = element_text(size=12, family="Arial", color="black"),
      plot.margin = unit(c(0,1.2,0.5,0), "cm")
    )
  counter <- counter + 1
}

p[[1]]+p[[2]]+p[[3]]+p[[4]] + plot_layout(ncol=2) # cytof_visit1_age_no-viralload_examples, 6.6 x 6.2 in
```

### Controlled for viral load

```{r}
# Linear regression on clr-transformed cell type proportion
res.clr.sc2 <- data.frame(matrix(NA, nrow=nrow(bld.4f), ncol=4))
colnames(res.clr.sc2) <- c("cell.type","slope","PVal","adj.PVal")
rownames(res.clr.sc2) <- rownames(bld.4f)
res.clr.sc2$cell.type <- bld.rData[rownames(res.clr.sc2),"Final.Annotated.Population"]

for (i in rownames(res.clr.sc2)) {
  temp <- data.frame(clr=bld.4f[i,], 
                     age=bld.clin.sample.event.individ.1$admit_age,
                     sex=bld.clin.sample.event.individ.1$sex,
                     # steroids=bld.clin.sample.event.individ.1$steroids,
                     sc2=bld.clin.sample.event.individ.1$sc2_rpm,
                     resp=bld.clin.sample.event.individ.1$respiratory_status)
  my.lm <- lm(clr ~ age + sc2 + sex + resp, data=temp)
  res.clr.sc2[i,"slope"] <- summary(my.lm)$coefficients["age",1]
  res.clr.sc2[i,"PVal"] <- summary(my.lm)$coefficients["age",4]
  
  remove(temp, my.lm)
}

# Correct P-values
res.clr.sc2$adj.PVal <- p.adjust(res.clr.sc2$PVal, method = "BH")
```

Plot slopes of significant cell types.

```{r}
# Supplementary figure 4
p <- ggplot() +
  geom_col(data = subset(res.clr.sc2, adj.PVal<0.05),
           aes(x = slope, y = reorder(cell.type,slope),
               fill = as.factor(slope>0)),
           width = 0.7) +
  geom_vline(xintercept=0, size=0.5, color="black") +
  scale_fill_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  # coord_flip() +
  labs(x = "Slope\n(change per 1 year of age)",
       y = "", title = "Cell types associated\nwith age at Visit 1\n(controlled for viral load)") +
  xlim(-0.022, 0.022) + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank()
  ) # cytof_visit1_age_viralload, 6 x 7 in (w x h)
```
