---
title: "patient_characteristics"
output: html_document
---

Here I will analyze age vs some patient characteristics.

Source: 20231024_table1.Rmd

```{r}
library(tidyverse)
library(ggplot2)
library(patchwork)

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```

# Import data

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

### Clinical metadata processing

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start","ever_steroids","diedever","timetodeath","los")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE
clin.event$steroids <- as.factor(clin.event$steroids)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert some columns to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

## Viral load data

Get SARS-CoV-2 viral load from nasal-metagenomics-AllGenera-Counts_RPM.csv, sc2_rpm column.

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/legacy/2022-09-03/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
```

# Merge metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

```{r}
# mNGS QC
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))

# Keep samples within 30 days of hospitalization
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2, event_date <= 30
)
```

# Get metadata of unique patients

```{r}
# Keep unique patients
clin.sample.event.individ.sc2.unique <- clin.sample.event.individ.sc2[!duplicated(clin.sample.event.individ.sc2$participant_id),]

# Add age quintile
clin.sample.event.individ.sc2.unique$age_group <- cut(
  clin.sample.event.individ.sc2.unique$admit_age,
  breaks=c(17,46,54,62,70,96)
  )
# Rename levels: convert round bracket to square bracket
levels(clin.sample.event.individ.sc2.unique$age_group) <- c("[18,46]","[47,54]","[55,62]","[63,70]","[71,96]")
```

Extract visit 1 samples for reference

```{r}
clin.sample.event.individ.visit1.unique <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1") %>%
  subset(!duplicated(participant_id))
```

## Age

```{r}
# Age quintiles
print("Age quintiles")
print(
  table(clin.sample.event.individ.sc2.unique$age_group)
)

# Age
print("Age: median")
print(median(clin.sample.event.individ.sc2.unique$admit_age))
print("Age: 1st quartile")
print(quantile(clin.sample.event.individ.sc2.unique$admit_age, 0.25))
print("Age: 3rd quartile")
print(quantile(clin.sample.event.individ.sc2.unique$admit_age, 0.75))
```

```{r}
print(
  clin.sample.event.individ.sc2.unique %>%
    group_by(age_group) %>%
    summarise(med = median(admit_age),
              q1 = quantile(admit_age, 0.25),
              q3 = quantile(admit_age, 0.75))
)
```

Plot age histogram.

```{r}
# Figure 1B
p1 <- ggplot(data=clin.sample.event.individ.sc2.unique,
            aes(x=admit_age, fill=age_group)) +
  geom_histogram(breaks=seq(18,96,by=2), color="black") +
  labs(x="Age", y="Number of individuals") +
  scale_fill_viridis_d(guide="none") +
  scale_x_continuous(breaks=seq(20,90,10)) +
  scale_y_continuous(limits=c(0,65), expand=c(0,0)) +
  my.theme # age_histogram.pdf, 3.5 x 3 in
```

Plot age vs TG.

```{r}
# Figure 1C
red.palette <- colorRampPalette(c("#fcbba1", "#a50f15"), space="Lab")

p2 <- ggplot(clin.sample.event.individ.sc2.unique,
            aes(x=trajectory_group,
                y=admit_age,
                color=trajectory_group)) + 
  stat_boxplot(geom ='errorbar', width=0.3) + 
  geom_boxplot(fill="white", outlier.shape=NA) +
  ggbeeswarm::geom_quasirandom(width=0.35, size = 0.5) +
  # geom_jitter(width = 0.2, size = 1) +
  labs(x="Trajectory group", y="Age") +
  # scale_fill_brewer(palette="Oranges") +
  scale_color_manual(values=red.palette(5)) +
  scale_y_continuous(breaks=seq(20,100,20), limits = c(15, 100)) +
  my.theme + theme(legend.position = "none") # age-vs-TG.pdf, 3.3 x 3 in
      

print(kruskal.test(admit_age~trajectory_group, data=clin.sample.event.individ.sc2.unique))
```

Plot age vs 28-day mortality.

```{r}
# Overall mortality
p3 <- ggplot(clin.sample.event.individ.sc2.unique,
             aes(x=as.factor(diedever),
                 y=admit_age,
                 fill=as.factor(diedever))) + 
  stat_boxplot(geom ='errorbar', width=0.3) + 
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(width = 0.2, size = 0.5) +
  # geom_jitter(width = 0.2, size = 1) +
  labs(x="Overall mortality", y="Age") +
  scale_fill_brewer(palette="Oranges") +
  scale_y_continuous(breaks=seq(20,100,20), limits = c(15, 100)) +
  my.theme + theme(legend.position = "none")
print(kruskal.test(admit_age~as.factor(diedever), data=clin.sample.event.individ.sc2.unique))

# Figure 1D
# 28-day mortality
clin.sample.event.individ.sc2.unique$died28d <- ifelse(
  is.na(clin.sample.event.individ.sc2.unique$timetodeath) | clin.sample.event.individ.sc2.unique$timetodeath>28,
  "No", "Yes")
p3 <- ggplot(clin.sample.event.individ.sc2.unique,
             aes(x=died28d,
                 y=admit_age,
                 color=died28d)) + 
  stat_boxplot(geom ='errorbar', width=0.25) + 
  geom_boxplot(fill="white", outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(width=0.25, size=0.3) +
  # geom_jitter(width = 0.2, size = 1) +
  labs(x="28-day mortality", y="Age") +
  # scale_fill_brewer(palette="Oranges") +
  scale_color_manual(values=red.palette(2)) +
  scale_y_continuous(breaks=seq(20,100,20), limits = c(15, 100)) +
  my.theme + theme(legend.position = "none") # age-vs-died28d.pdf, 3 x 3 in
print(kruskal.test(admit_age~as.factor(diedever), data=clin.sample.event.individ.sc2.unique))
```
