---
title: "convalescent"
output: pdf_document
---

```{r load code and data, warning=FALSE, message=FALSE}
# codebase directory
require(openxlsx)
library(factoextra)
library(cluster)
require(nlme) # directly provides p-values, unlike lme4
require(splines)
require(dplyr)
library(fgsea)
library(knitr)
library(kableExtra)
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source(paste0(codebase_path,"/codebase_subtype.R"))

conv_data_env <- readRDS("/scratch/metagenomics/viral_reactivation/2024-02-02_obj.RDS")
acute_plus_convalescent_preprocessed_data <- readRDS("/scratch/data-integration/molecular_subtyping/Revisions/acute_plus_convalescent_preprocessed_data.RDS")
convalescent_factors <- readRDS("/scratch/data-integration/molecular_subtyping/Revisions/convalescent_factors_20250401.rds")
subtype_hcl <- readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
```

# Clinical data preparation
```{r clinical data preparation}
tmp = setdiff(colnames(conv_data_env$clinical_data),
              c("sample_id", "sample_type", "phase1_2",
                "samples_collected", "phase"))
clinical_data = conv_data_env$clinical_data[, tmp]
clinical_data = unique(clinical_data)
clin_data_aligned = data.frame(event_id = rownames(convalescent_factors))
clin_data_aligned = left_join(clin_data_aligned, clinical_data,
                                  by = "event_id")
assay_subtype = subtype_hcl
assay_subtype$participant_id <- rownames(assay_subtype)
assay_subtype = assay_subtype[,c("participant_id", "PASC")]
clin_data_aligned = left_join(clin_data_aligned, assay_subtype, 
                                  by = "participant_id")
clin_data_aligned$PASC_group = clin_data_aligned$PASC

mod_ids = c(1, 2, 3, 10)
negation = c(-1, 1, 1, 1)
# negate factors
mcia_factors = t(t(convalescent_factors[, mod_ids]) * negation)
mcia_factors = apply(mcia_factors, 2, function(z) qqnorm(z, plot.it = F)$x)
mcia_factors = data.frame(mcia_factors, row.names = rownames(convalescent_factors))
mcia_factors = mcia_factors[clin_data_aligned$event_id, ]
colnames(mcia_factors) = paste0("Factor", mod_ids)
```

# Acute kinetics of shared samples across subtypes
```{r acute kinetics of shared samples across subtypes, fig.height=6, fig.width=24}
idx = which(clin_data_aligned$event_type %in% paste0("Visit ", 1:6))

assay_mat = mcia_factors[idx,]
assay_mat$event_id = rownames(assay_mat)
assay_mat = assay_mat[order(assay_mat$event_id), ]
assay_mat = left_join(
  assay_mat, clin_data_aligned[,c("event_id", "event_type", "enrollment_site",
                                  "sex", "discretized_admit_age_quantile",
                                  "participant_id", "event_date",
                                  "symptom_date", "PASC_group")], by = "event_id")
assay_mat = assay_mat[!is.na(assay_mat$event_type), ]
assay_subtype = subtype_hcl
assay_subtype$participant_id <- rownames(assay_subtype)
assay_mat = left_join(assay_mat, assay_subtype, by = "participant_id")
assay_mat = assay_mat[!is.na(assay_mat$label), ]

clin_colnames_select = setdiff(colnames(assay_mat), colnames(mcia_factors))

factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "label", use_endpoint_levels = c("A", "B", "C", "E", "F"),
  max_date = 40, text_position = 2)

ggarrange(plotlist = factor_trajectory_analysis_results$test_fig_list,
          nrow = 1, common.legend = T, legend = "bottom")
```

# Acute kinetics of participants without LC measure across subtypes
```{r acute kinetics of participants without LC measure across subtypes, fig.height=6, fig.width=24}
idx = which(clin_data_aligned$event_type %in% paste0("Visit ", 1:6) &
              is.na(clin_data_aligned$PASC_group))

assay_mat = mcia_factors[idx,]
assay_mat$event_id = rownames(assay_mat)
assay_mat = assay_mat[order(assay_mat$event_id), ]
assay_mat = left_join(
  assay_mat, clin_data_aligned[,c("event_id", "event_type", "enrollment_site",
                                  "sex", "discretized_admit_age_quantile",
                                  "participant_id", "event_date",
                                  "symptom_date", "PASC_group")], by = "event_id")
assay_mat = assay_mat[!is.na(assay_mat$event_type), ]
assay_subtype = subtype_hcl
assay_subtype$participant_id <- rownames(assay_subtype)
assay_mat = left_join(assay_mat, assay_subtype, by = "participant_id")
assay_mat = assay_mat[!is.na(assay_mat$label), ]

clin_colnames_select = setdiff(colnames(assay_mat), colnames(mcia_factors))

factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "label", use_endpoint_levels = c("A", "B", "C", "E", "F"),
  max_date = 40, text_position = 2)

ggarrange(plotlist = factor_trajectory_analysis_results$test_fig_list, 
          nrow = 1, common.legend = T, legend = "bottom")
```

# Acute kinetics of participants with LC measure across subtypes (Fig 6G)
```{r acute kinetics of participants with LC measure across subtypes, fig.height=5.5, fig.width=24}
idx = which(clin_data_aligned$event_type %in% paste0("Visit ", 1:6) &
              !is.na(clin_data_aligned$PASC_group))

assay_mat = mcia_factors[idx,]
assay_mat$event_id = rownames(assay_mat)
assay_mat = assay_mat[order(assay_mat$event_id), ]
assay_mat = left_join(
  assay_mat, clin_data_aligned[,c("event_id", "event_type", "enrollment_site",
                                  "sex", "discretized_admit_age_quantile",
                                  "participant_id", "event_date",
                                  "symptom_date", "PASC_group")], by = "event_id")
assay_mat = assay_mat[!is.na(assay_mat$event_type), ]
assay_subtype = subtype_hcl
assay_subtype$participant_id <- rownames(assay_subtype)
assay_mat = left_join(assay_mat, assay_subtype, by = "participant_id")
assay_mat = assay_mat[!is.na(assay_mat$label), ]

clin_colnames_select = setdiff(colnames(assay_mat), colnames(mcia_factors))

factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "label", use_endpoint_levels = c("A", "B", "C", "E", "F"),
  max_date = 40, text_position = 2)

ggarrange(plotlist = factor_trajectory_analysis_results$test_fig_list, 
          nrow = 1, common.legend = T, legend = "bottom")
```

# Convalescent kinetics of participants with LC measure across subtypes (Fig 6H)
```{r convalescent kinetics of participants with LC measure across subtypes, fig.height=6, fig.width=24}
idx = which(clin_data_aligned$event_type %in% paste0("Visit ", 7:10) &
              !is.na(clin_data_aligned$PASC_group))

assay_mat = mcia_factors[idx,]
assay_mat$event_id = rownames(assay_mat)
assay_mat = assay_mat[order(assay_mat$event_id), ]
assay_mat = left_join(
  assay_mat, clin_data_aligned[,c("event_id", "event_type", "enrollment_site",
                                  "sex", "discretized_admit_age_quantile",
                                  "participant_id", "event_date",
                                  "symptom_date", "PASC_group")], by = "event_id")
assay_mat = assay_mat[!is.na(assay_mat$event_type), ]
assay_subtype = subtype_hcl
assay_subtype$participant_id <- rownames(assay_subtype)
assay_mat = left_join(assay_mat, assay_subtype, by = "participant_id")
assay_mat = assay_mat[!is.na(assay_mat$label), ]

clin_colnames_select = setdiff(colnames(assay_mat), colnames(mcia_factors))

factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "label", use_endpoint_levels = c("A", "B", "C", "E", "F"),
  max_date = 420, min_date = 80, text_position = 2)

ggarrange(plotlist = factor_trajectory_analysis_results$test_fig_list, 
          nrow = 1, common.legend = T, legend = "bottom")
```

# Convalescent barplot of participants with LC measure across subtypes (Fig 6I)
```{r, fig.height=6, fig.width=24}
fnames = paste0("Factor", c(1, 2, 3, 10))
znames = setdiff(colnames(assay_mat), fnames)
conv_barplot = list()

for (i in 1:length(fnames)) {
  df_plot = assay_mat[, c(fnames[i], znames)]
  df_plot = df_plot[df_plot$label != "D", ]
  df_plot$event_type = factor(
    df_plot$event_type,
    levels = c("Visit 7", "Visit 8", "Visit 9","Visit 10"))
  ftitle = fnames[i]
  ftitle = gsub("Factor", "Factor ", ftitle)
  
  conv_barplot[[i]] = ggplot(df_plot, aes(x = label, y = !!sym(fnames[i]))) +
  geom_boxplot(aes(fill = label), width = 0.9, outlier.size = 0.3, 
               size = 0.6, alpha = 0.3) +
  geom_jitter(aes(fill = label), colour = "grey40",
              width = 0.3, size = 0.8, alpha = 0.8, shape = 21) +
  facet_wrap(~ event_type, scales = "free_x", nrow = 1) +
  theme_minimal() +
  labs(y = "Factor Score", title = ftitle) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_text(size = 15, color = "black"),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 15, color = "black"),
        legend.position = "bottom",
        panel.spacing = unit(0.5, "lines"),
        strip.text.x = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.key.size = unit(2, "lines"),
        axis.line.x = element_line(linewidth = 1),
        axis.line.y = element_line(linewidth = 1),
        plot.title = element_text(size = 15, hjust = 0.5)) +
  scale_fill_manual(values = c("#2ca02c", "#1f77b4", "#ff7f0e",
                               "#9C3418", "#9467bd"), name = "Subtype") +
  scale_colour_manual(values = c("#2ca02c", "#1f77b4", "#ff7f0e",
                                 "#9C3418", "#9467bd"), guide = "none")
}

ggarrange(plotlist = conv_barplot,
          nrow = 1, common.legend = T, legend = "bottom")
```

# Convalescent kinetics of participants with LC measure across LC groups
```{r, fig.height=6, fig.width=6}
# idx = which(clin_data_aligned$event_type %in% paste0("Visit ", 1:6) &
#               !is.na(clin_data_aligned$PASC_group))
idx = which(clin_data_aligned$event_type %in% paste0("Visit ", 7:10) &
              !is.na(clin_data_aligned$PASC_group))

assay_mat = mcia_factors[idx,]
assay_mat$event_id = rownames(assay_mat)
assay_mat = assay_mat[order(assay_mat$event_id), ]
assay_mat = left_join(
  assay_mat, clin_data_aligned[,c("event_id", "event_type", "enrollment_site",
                                  "sex", "discretized_admit_age_quantile",
                                  "participant_id", "event_date",
                                  "symptom_date", "PASC_group")], by = "event_id")
assay_mat = assay_mat[!is.na(assay_mat$event_type), ]
assay_subtype = subtype_hcl
assay_subtype$participant_id <- rownames(assay_subtype)
assay_mat = left_join(assay_mat, assay_subtype, by = "participant_id")
assay_mat = assay_mat[!is.na(assay_mat$label), ]
assay_mat = assay_mat[!is.na(assay_mat$PASC),]
assay_mat$PASC = ifelse(assay_mat$PASC == "minimal", 
                         "Minimal Deficit", "Other Deficits")

clin_colnames_select = setdiff(colnames(assay_mat), colnames(mcia_factors))

factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "PASC", use_endpoint_levels = c("Minimal Deficit", "Other Deficits"),
  subgroup_cr = c("Other Deficits" = "#9C3418", "Minimal Deficit" = "#39828C"),
  max_date = 420, min_date = 80, legend_name = "PASC", multiple_correction = FALSE, 
  text_position = 2)

factor_trajectory_analysis_results$test_fig_list$Factor30
```



