---
title: "PRO Scale data cleaning and preparation"
author: "Jessica Maciuch"
date: "`r format(Sys.time(), '%b %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

workDir <- "~/impacc-full-cohort/Analysis/convalescent-project/Spear/clinical_associations/"
setwd(workDir) # For troubleshooting in the command line
knitr::opts_knit$set(root.dir = workDir)

library(tidyverse)
library(magrittr)
```

# Prepare PRO Scales

For explanation of PRO variables, see /scratch/clinical/current/PRO_scores_dates_updated_29Aug23_README.txt

Needed to convert all PRO columns to numeric because of conflicting data type issues in the original file.

**Expect an "NAs induced by coercion" warning because some NA values were input as the character string "#N/A".**

There are 76 observations in the data with duplicate event IDs where the PROMIS Physical Scores and PROMIS Psycho Scores appear to have been input separately from the rest of the PRO scores. Since these duplicate event IDs are non-overlapping, all values for the same event ID were merged, and the "survey date"/"diff date" columns were eliminated in the wide-form document to avoid merging conflicts (since they were different between duplicate event IDs in some cases). 

```{r Prepare PRO Scores}

# Load PRO_scores
  PRO_scores <- read.csv("/data/clinical/current/2023-01-01-locked/PRO_scores_dates_updated.csv")

# Make sure scores are numeric to avoid datatype issues down the line
# NOTE: this will generate an "NAs induced by coercion" warning
  PRO_scores %<>% mutate(across(PROMIS_Dyspnea_score:PROMIS_Psycho_score, as.numeric))
  
# Change name of participant ID variable to align with clinical data
  names(PRO_scores)[names(PRO_scores) == "PID"] <- "participant_id"
  
# Merge duplicate event ID scores
  # Separate each score
  PRO_scores_separated <- PRO_scores %>% select(PROMIS_Dyspnea_score:PROMIS_Psycho_score) %>% as.list()
  
  # Append event ID to each separated list 
  create_df <- function(scores, score_name) {
    df <- data.frame(PRO_scores$participant_id,
                     PRO_scores$event_id,
                     scores)
    
    colnames(df) <- c("participant_id", "event_id", score_name)
    
    df %<>% filter(!is.na(.data[[score_name]]))
    return(df)
  }
  
  PRO_scores_separated <- list(PRO_scores_separated, names(PRO_scores_separated)) |>  purrr::pmap(\(scores, score_name) create_df(scores, score_name))
  
  PRO_scores <- PRO_scores_separated |> purrr::reduce(full_join, by = c("participant_id", "event_id"), na_matches = "na")
  
# Save output as csv
  write_csv(PRO_scores, file = "~/impacc-full-cohort/Analysis/convalescent-project/Spear/clinical_associations/PRO_score_preprocessing/20240229_PRO_Scores_Cleaned.csv")

```

# Long form version with updated names for manuscript

```{r}

# Load PRO_scores
  PRO_scores <- read.csv("/data/clinical/current/2023-01-01-locked/PRO_scores_dates_updated.csv")

# Make sure scores are numeric to avoid datatype issues down the line
# NOTE: this will generate an "NAs induced by coercion" warning
  PRO_scores %<>% mutate(across(PROMIS_Dyspnea_score:PROMIS_Psycho_score, as.numeric))
  
# Convert to long form
  PRO_scores %<>% pivot_longer(cols = PROMIS_Dyspnea_score:PROMIS_Psycho_score,
                               names_to = "PRO_scale")
    
# Remove missing scores
  PRO_scores %<>% filter(!is.na(value))
  
# Drop participant ID to prevent merging conflicts
  PRO_scores %<>% select(-c(PID))
```

# Append readable names

```{r}

# Generate dictionary for PRO scale names
  PRO_dict <- data.frame("PRO_scale" = unique(PRO_scores$PRO_scale))

  PRO_dict %<>% mutate(PRO_scale_manuscript = case_match(PRO_scale,
                                                         "PROMIS_Physical_score" ~ "PROMIS Physical Score",
                                                         "PROMIS_Dyspnea_score" ~ "PROMIS Dyspnea Score",
                                                         "PROMIS_Mental_score" ~ "PROMIS Mental Score",
                                                         "PROMIS_Impact_score" ~ "PROMIS Psychosocial Impact Score",
                                                         "PROMIS_Psycho_score" ~ "PROMIS Cognitive Score",
                                                         "health_score" ~ "Health Score",
                                                         "eq5d5l_score" ~ "EQ-5D-5L Score"))
  
# Order PRO scales
  PRO_dict$PRO_scale_manuscript %<>% factor(., levels = c("PROMIS Physical Score",
                                                          "EQ-5D-5L Score",
                                                          "PROMIS Mental Score",
                                                          "PROMIS Psychosocial Impact Score",
                                                          "PROMIS Cognitive Score",
                                                          "PROMIS Dyspnea Score",
                                                          "Health Score"))
  
# Combine with all scores
  PRO_scores %<>% left_join(x = .,
                            y = PRO_dict,
                            by = "PRO_scale")
  
# Order scores 
  PRO_scores %<>% arrange(PRO_scale_manuscript)
  
# Save long form
  write_csv(PRO_scores, file = "./PRO_score_preprocessing/20240913_PRO_Scores_long.csv")
```