---
title: "Acute data preparation and imputation"
author: "Gisela Gabernet (gisela.gabernet@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

```{r}
library(MOFA2)
#library(MOFAdata)
library(data.table)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(kableExtra)
library(RColorBrewer)


knitr::opts_chunk$set(warning = F, message=F)

print_kable <- function(table, kable_height = "500px", kable_width = "100%") {
  kable(table, digits=3) %>%
    kable_styling("striped") #%>%
    #scroll_box(height = kable_height, width = kable_width)
}

# set random seed
set.seed(42)
```


# Get MOFA model and impute values (with CyTOF)

## Load MOFA model
```{r fig.height=5, fig.width=10}

# Set MOFA model name
mofa_model_name_train <- "MOFA_acute_train_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240331.rds"
mofa_model_name_test <- "MOFA_acute_test_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240331.rds"
results_dir <- "/scratch/data-fullcohort/spear/acute_preprocessed_data_20240323/"

dir.create(results_dir)

# Import data env
data_env_train <- readRDS('/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_train_Mar_23_2024.RDS')
data_env_test <- readRDS('/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_test_Mar_23_2024.RDS')

# Import preprocessed assays
preprocessed_assays_train <- readRDS("/scratch/data-fullcohort/batch-effects/preprocessed_assays_acute_train_Mar_23_2024.RDS")
preprocessed_assays_test <- readRDS("/scratch/data-fullcohort/batch-effects/preprocessed_assays_acute_test_Mar_23_2024.RDS")

clinical_data_train <- data_env_train$clinical_data
clinical_data_test <- data_env_test$clinical_data

event_id_train <- unique(clinical_data_train %>% pull(event_id))
event_id_test <- unique(clinical_data_test %>% pull(event_id))

# Get selected features from convalescent assays train
preprocessed_assays_conv_train <- readRDS("/scratch/data-fullcohort/batch-effects/preprocessed_assays_train_Feb_11_2024.RDS")

# Get features for test data feature selection
PPT_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$plasma_proteomics_targeted)
PPG_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$plasma_proteomics_global_dda)

#cytof MSI features
BCT_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$bld_msi_cytof)

# PBMC transcriptomics features
PGX_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$pbmc_transcriptomics)

# Plasma metabolomics features
PMG_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$plasma_metabolomics_global)
```

## Aggregate datasets according to event_id

## Prepare training data

```{r}
omic_name <- c()
dim_samples <- c()
dim_features <- c()

sample_event <- clinical_data_train %>% dplyr::select(sample_id, event_id) %>% distinct()

# Get list of event_ids across modalities
pbmc_transcriptomics <- preprocessed_assays_train$preprocessed_data$pbmc_transcriptomics %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)
omic_name <- append(omic_name, "pbmc_transcriptomics")
dim_samples <- append(dim_samples, nrow(pbmc_transcriptomics))
dim_features <- append(dim_features, ncol(pbmc_transcriptomics)-1)

plasma_proteomics_targeted <- preprocessed_assays_train$preprocessed_data$plasma_proteomics_targeted %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)

omic_name <- append(omic_name, "plasma_proteomics_targeted")
dim_samples <- append(dim_samples, nrow(plasma_proteomics_targeted))
dim_features <- append(dim_features, ncol(plasma_proteomics_targeted)-1)

# Filter out features from serum olink and plasma proteomics global that are already in plasma proteomics targeted
rm_dups_ppt <- colnames(preprocessed_assays_train$preprocessed_data$plasma_proteomics_global_dda) %in% colnames(plasma_proteomics_targeted)

rm_dups_so <- colnames(preprocessed_assays_train$preprocessed_data$plasma_proteomics_global_dda) %in% colnames(preprocessed_assays_train$preprocessed_data$serum_olink)

dups_names_ppt <- colnames(preprocessed_assays_train$preprocessed_data$plasma_proteomics_global_dda)[rm_dups_ppt]
dups_names_so <- colnames(preprocessed_assays_train$preprocessed_data$plasma_proteomics_global_dda)[rm_dups_so]

plasma_proteomics_global <- preprocessed_assays_train$preprocessed_data$plasma_proteomics_global_dda %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id) %>%
                          dplyr::select(-all_of(c(dups_names_ppt, dups_names_so)))
omic_name <- append(omic_name, "plasma_proteomics_global")
dim_samples <- append(dim_samples, nrow(plasma_proteomics_global))
dim_features <- append(dim_features, ncol(plasma_proteomics_global)-1)

plasma_metabolomics_global <- preprocessed_assays_train$preprocessed_data$plasma_metabolomics_global %>%
                          dplyr::select(all_of(PMG_selected_features)) %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)
omic_name <- append(omic_name, "plasma_metabolomics_global")
dim_samples <- append(dim_samples, nrow(plasma_metabolomics_global))
dim_features <- append(dim_features, ncol(plasma_metabolomics_global)-1)

serum_olink <- preprocessed_assays_train$preprocessed_data$serum_olink %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)
omic_name <- append(omic_name, "serum_olink")
dim_samples <- append(dim_samples, nrow(serum_olink))
dim_features <- append(dim_features, ncol(serum_olink)-1)

bld_msi_cytof <- preprocessed_assays_train$preprocessed_data$bld_msi_cytof %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)
omic_name <- append(omic_name, "bld_msi_cytof")
dim_samples <- append(dim_samples, nrow(bld_msi_cytof))
dim_features <- append(dim_features, ncol(bld_msi_cytof)-1)

all_event_ids_train <- unique(c(pbmc_transcriptomics$event_id,
                           plasma_proteomics_targeted$event_id,
                           plasma_proteomics_global$event_id,
                           plasma_metabolomics_global$event_id,
                           serum_olink$event_id,
                            bld_msi_cytof$event_id
                           ))

all_sample_ids_train <- unique(c(row.names(preprocessed_assays_train$preprocessed_data$pbmc_transcriptomics),
                           row.names(preprocessed_assays_train$preprocessed_data$plasma_proteomics_targeted),
                           row.names(preprocessed_assays_train$preprocessed_data$plasma_proteomics_global),
                           row.names(preprocessed_assays_train$preprocessed_data$plasma_metabolomics_global),
                           row.names(preprocessed_assays_train$preprocessed_data$serum_olink),
                           row.names(preprocessed_assays_train$bld_msi_cytof)
                           ))

pbmc_transcriptomics <- pbmc_transcriptomics %>%
  complete(event_id = all_event_ids_train) %>% 
  column_to_rownames("event_id")

plasma_proteomics_targeted <- plasma_proteomics_targeted %>%
  complete(event_id = all_event_ids_train) %>% 
  column_to_rownames("event_id")

plasma_proteomics_global <- plasma_proteomics_global %>%
  complete(event_id = all_event_ids_train) %>% 
  column_to_rownames("event_id")

plasma_metabolomics_global <- plasma_metabolomics_global %>%
  complete(event_id = all_event_ids_train) %>% 
  column_to_rownames("event_id")

serum_olink <- serum_olink %>%
  complete(event_id = all_event_ids_train) %>% 
  column_to_rownames("event_id")

bld_msi_cytof <- bld_msi_cytof %>%
  complete(event_id = all_event_ids_train) %>%
  column_to_rownames("event_id")


df_dims <- data.frame(omic_name, dim_samples, dim_features)
print_kable(df_dims)
```


## Check train matrix dimensions

Train sample dimensions

```{r}
train_data = list(
  SO = t(as.matrix(serum_olink %>% filter(rownames(serum_olink) %in% event_id_train))),
  PPT = t(as.matrix(plasma_proteomics_targeted %>% filter(rownames(plasma_proteomics_targeted) %in% event_id_train))),
  PPG = t(as.matrix(plasma_proteomics_global %>% filter(rownames(plasma_proteomics_global) %in% event_id_train))),
  PMG = t(as.matrix(plasma_metabolomics_global %>% filter(rownames(plasma_metabolomics_global) %in% event_id_train))),
  PGX = t(as.matrix(pbmc_transcriptomics %>% filter(rownames(pbmc_transcriptomics) %in% event_id_train))),
  BCT = t(as.matrix(bld_msi_cytof %>% filter(rownames(bld_msi_cytof) %in% event_id_train)))
)


lapply(train_data,dim)


```
## Prepare test data

```{r}
omic_name <- c()
dim_samples <- c()
dim_features <- c()

sample_event <- clinical_data_test %>% dplyr::select(sample_id, event_id) %>% distinct()

# Get list of event_ids across modalities
# check that they are the same features as for training data

# PGX
stopifnot(ncol(preprocessed_assays_test$preprocessed_data$pbmc_transcriptomics) == ncol(preprocessed_assays_train$preprocessed_data$pbmc_transcriptomics))

stopifnot(sum(colnames(preprocessed_assays_test$preprocessed_data$pbmc_transcriptomics ) == colnames(preprocessed_assays_train$preprocessed_data$pbmc_transcriptomics)) == ncol(preprocessed_assays_test$processed_data$pbmc_transcriptomic))

pbmc_transcriptomics <- preprocessed_assays_test$preprocessed_data$pbmc_transcriptomics %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)
omic_name <- append(omic_name, "pbmc_transcriptomics")
dim_samples <- append(dim_samples, nrow(pbmc_transcriptomics))
dim_features <- append(dim_features, ncol(pbmc_transcriptomics)-1)


# PPT

stopifnot(ncol(preprocessed_assays_test$preprocessed_data$plasma_proteomics_targeted ) == ncol(preprocessed_assays_train$preprocessed_data$plasma_proteomics_targeted))

stopifnot(sum(colnames(preprocessed_assays_test$preprocessed_data$plasma_proteomics_targeted) == colnames(preprocessed_assays_train$preprocessed_data$plasma_proteomics_targeted)) == ncol(preprocessed_assays_test$preprocessed_data$plasma_proteomics_targeted))

plasma_proteomics_targeted <- preprocessed_assays_test$preprocessed_data$plasma_proteomics_targeted %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)

omic_name <- append(omic_name, "plasma_proteomics_targeted")
dim_samples <- append(dim_samples, nrow(plasma_proteomics_targeted))
dim_features <- append(dim_features, ncol(plasma_proteomics_targeted)-1)


## PPG
# Overlapping features in PPG + PPT training data were removed so they have to be removed there as well

plasma_proteomics_global_test <- preprocessed_assays_test$preprocessed_data$plasma_proteomics_global_dda %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id) %>%
                          dplyr::select(-all_of(c(dups_names_ppt, dups_names_so)))

omic_name <- append(omic_name, "plasma_proteomics_global")
dim_samples <- append(dim_samples, nrow(plasma_proteomics_global_test))
dim_features <- append(dim_features, ncol(plasma_proteomics_global_test)-1)

## PMG

plasma_metabolomics_global <- preprocessed_assays_test$preprocessed_data$plasma_metabolomics_global %>%
                          dplyr::select(all_of(PMG_selected_features)) %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)
omic_name <- append(omic_name, "plasma_metabolomics_global")
dim_samples <- append(dim_samples, nrow(plasma_metabolomics_global))
dim_features <- append(dim_features, ncol(plasma_metabolomics_global)-1)

#SO
stopifnot(ncol(preprocessed_assays_test$preprocessed_data$serum_olink ) == ncol(preprocessed_assays_train$preprocessed_data$serum_olink))

stopifnot(sum(colnames(preprocessed_assays_test$preprocessed_data$serum_olink) == colnames(preprocessed_assays_train$preprocessed_data$serum_olink)) == ncol(preprocessed_assays_test$preprocessed_data$serum_olink))

serum_olink <- preprocessed_assays_test$preprocessed_data$serum_olink %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)
omic_name <- append(omic_name, "serum_olink")
dim_samples <- append(dim_samples, nrow(serum_olink))
dim_features <- append(dim_features, ncol(serum_olink)-1)

#BCT
stopifnot(ncol(preprocessed_assays_test$preprocessed_data$bld_msi_cytof ) == ncol(preprocessed_assays_train$preprocessed_data$bld_msi_cytof))

stopifnot(sum(colnames(preprocessed_assays_test$preprocessed_data$bld_msi_cytof) == colnames(preprocessed_assays_train$preprocessed_data$bld_msi_cytof)) == ncol(preprocessed_assays_test$preprocessed_data$bld_msi_cytof))

bld_msi_cytof <- preprocessed_assays_test$preprocessed_data$bld_msi_cytof %>%
                          rownames_to_column("sample_id") %>%
                          left_join(sample_event, by="sample_id") %>%
                          dplyr::select(-sample_id)
omic_name <- append(omic_name, "bld_msi_cytof")
dim_samples <- append(dim_samples, nrow(bld_msi_cytof))
dim_features <- append(dim_features, ncol(bld_msi_cytof)-1)


all_event_ids_test <- unique(c(pbmc_transcriptomics$event_id,
                           plasma_proteomics_targeted$event_id,
                           plasma_proteomics_global_test$event_id,
                           plasma_metabolomics_global$event_id,
                           serum_olink$event_id,
                           bld_msi_cytof$event_id
                           ))

all_sample_ids_test <- unique(c(row.names(preprocessed_assays_test$preprocessed_data$pbmc_transcriptomics),
                           row.names(preprocessed_assays_test$preprocessed_data$plasma_proteomics_targeted),
                           row.names(preprocessed_assays_test$preprocessed_data$plasma_proteomics_global),
                           row.names(preprocessed_assays_test$preprocessed_data$plasma_metabolomics_global),
                           row.names(preprocessed_assays_test$preprocessed_data$serum_olink),
                           row.names(preprocessed_assays_test$preprocessed_data$bld_msi_cytof)
                           ))

pbmc_transcriptomics <- pbmc_transcriptomics %>%
  complete(event_id = all_event_ids_test) %>% 
  column_to_rownames("event_id")

plasma_proteomics_targeted <- plasma_proteomics_targeted %>%
  complete(event_id = all_event_ids_test) %>% 
  column_to_rownames("event_id")

plasma_proteomics_global_test <- plasma_proteomics_global_test %>%
  complete(event_id = all_event_ids_test) %>% 
  column_to_rownames("event_id")

plasma_metabolomics_global <- plasma_metabolomics_global %>%
  complete(event_id = all_event_ids_test) %>% 
  column_to_rownames("event_id")

serum_olink <- serum_olink %>%
  complete(event_id = all_event_ids_test) %>% 
  column_to_rownames("event_id")

bld_msi_cytof <- bld_msi_cytof %>%
  complete(event_id = all_event_ids_test) %>%
  column_to_rownames("event_id")


df_dims <- data.frame(omic_name, dim_samples, dim_features)
print_kable(df_dims)
```

## Check test matrix dimensions

Test sample dimensions

```{r}
test_data = list(
  SO = t(as.matrix(serum_olink %>% filter(rownames(serum_olink) %in% event_id_test))),
  PPT = t(as.matrix(plasma_proteomics_targeted %>% filter(rownames(plasma_proteomics_targeted) %in% event_id_test))),
  PPG = t(as.matrix(plasma_proteomics_global_test %>% filter(rownames(plasma_proteomics_global_test) %in% event_id_test))),
  PMG = t(as.matrix(plasma_metabolomics_global %>% filter(rownames(plasma_metabolomics_global) %in% event_id_test))),
  PGX = t(as.matrix(pbmc_transcriptomics %>% filter(rownames(pbmc_transcriptomics) %in% event_id_test))),
  BCT = t(as.matrix(bld_msi_cytof %>% filter(rownames(bld_msi_cytof) %in% event_id_test)))
)


lapply(test_data,dim)
```


## Save acute train and test data

```{r}
acute_preprocessed_data <- list(train = train_data,
                                test = test_data)

saveRDS(acute_preprocessed_data, file.path(results_dir,"acute_preprocessed_data.rds"))
```


# Train MOFA training data

```{r fig.height=10, eval=F}

# Defining the multi-group framework to distinguish the visits
MOFAobject_base <- MOFA2::create_mofa(train_data)

plot_data_overview(MOFAobject_base)

#data options
data_opts <- get_default_data_options(MOFAobject_base)

# Model options
model_opts_F300 <- get_default_model_options(MOFAobject_base)
model_opts_F300$num_factors <- 250

## train options
train_opts <- get_default_training_options(MOFAobject_base)
train_opts$convergence_mode <- "fast"
train_opts$seed <- 42

# Automatically drop factors that explain less than 1% of variance in all omics
train_opts$drop_factor_threshold <- 0.01

```

## Run MOFA train


```{r use_basilisk = TRUE, eval=F}
MOFAobject_f300 <- prepare_mofa(MOFAobject_base,
  data_options = data_opts,
  model_options = model_opts_F300,
  training_options = train_opts
)

MOFAobject_f300 <- run_mofa(MOFAobject_f300, use_basilisk = TRUE)

```


```{r eval=F}
correlation_metadata <- clinical_data_train %>%
  dplyr::select(event_id, resp_status_v1, sex, admit_age, event_after_first_vaccine_dose,
                race, enrollment_site, event_date, comorb_anyresp_noasthma,
                event_date_3weeks_after_dose, pro_group_labels, event_type,
                discretized_admit_age_quantile, participant_id, partition)  %>%
  dplyr::distinct() %>%
  dplyr::mutate(sample = event_id) %>%
  filter(sample %in% samples_metadata(MOFAobject_f300)$sample)

freq = correlation_metadata %>%
  group_by(sample) %>%
  summarise(n = n())

correlation_metadata_join <- samples_metadata(MOFAobject_f300) %>%
  left_join(correlation_metadata, by="sample")

samples_metadata(MOFAobject_f300) <- correlation_metadata_join

# Save MOFA object
saveRDS(MOFAobject_f300, file = paste0(results_dir, mofa_model_name_train ))

```

# Train MOFA test data

```{r fig.height=10, eval=F}

# Defining the multi-group framework to distinguish the visits
MOFAobject_base_test <- MOFA2::create_mofa(test_data)

plot_data_overview(MOFAobject_base_test)

```

## Run MOFA test

```{r use_basilisk = TRUE, eval=F}
MOFAobject_f300_test <- prepare_mofa(MOFAobject_base_test,
  data_options = data_opts,
  model_options = model_opts_F300,
  training_options = train_opts
)

MOFAobject_f300_test <- run_mofa(MOFAobject_f300_test, use_basilisk = TRUE)

## Add metadata to MOFA object
correlation_metadata <- clinical_data_test %>%
  dplyr::select(event_id, resp_status_v1, sex, admit_age, event_after_first_vaccine_dose,
                race, enrollment_site, event_date, comorb_anyresp_noasthma,
                event_date_3weeks_after_dose, pro_group_labels, event_type,
                discretized_admit_age_quantile, participant_id, partition)  %>%
  dplyr::distinct() %>%
  dplyr::mutate(sample = event_id) %>%
  filter(sample %in% samples_metadata(MOFAobject_f300_test)$sample)

freq = correlation_metadata %>%
  group_by(sample) %>%
  summarise(n = n())

correlation_metadata_join <- samples_metadata(MOFAobject_f300_test) %>%
  left_join(correlation_metadata, by="sample")

samples_metadata(MOFAobject_f300_test) <- correlation_metadata_join


# Save MOFA object
saveRDS(MOFAobject_f300_test, file = paste0(results_dir, mofa_model_name_test))

```



## Perform imputation 

```{r}
# Impute train data using MOFA
imputed_data_train = impute(MOFAobject_f300, views = "all")
data_train = imputed_data_train@imputed_data
imputed_train <- list(
  SO = as.data.frame(t(as.matrix(data_train$SO$group1))), # serum olink
  PPT = as.data.frame(t(as.matrix(data_train$PPT$group1))), # targeted plasma proteomics
  PPG = as.data.frame(t(as.matrix(data_train$PPG$group1))), # global plasma proteomics
  PMG = as.data.frame(t(as.matrix(data_train$PMG$group1))), # plasma metabolomics global
  PGX = as.data.frame(t(as.matrix(data_train$PGX$group1))), # pbmc trnascriptomics
  #NGX = as.data.frame(t(as.matrix(data_train$NGX$group1))), # nasal transcriptomics
  BCT = as.data.frame(t(as.matrix(data_train$BCT$group1)))
)


# Impute test data using MOFA
imputed_data_test = impute(MOFAobject_f300_test, views = "all")
data_test = imputed_data_test@imputed_data
imputed_test <- list(
  SO = as.data.frame(t(as.matrix(data_test$SO$group1))), # serum olink
  PPT = as.data.frame(t(as.matrix(data_test$PPT$group1))), # targeted plasma proteomics
  PPG = as.data.frame(t(as.matrix(data_test$PPG$group1))), # global plasma proteomics
  PMG = as.data.frame(t(as.matrix(data_test$PMG$group1))), # plasma metabolomics global
  PGX = as.data.frame(t(as.matrix(data_test$PGX$group1))), # pbmc trnascriptomics
  #NGX = as.data.frame(t(as.matrix(data_test$NGX$group1))), # nasal transcriptomics
  BCT = as.data.frame(t(as.matrix(data_test$BCT$group1)))
)

acute_imputed_data <- list(train = imputed_train,
                                test = imputed_test)

saveRDS(acute_imputed_data, file.path(results_dir,"acute_imputed_data.rds"))
```


# Tools versions

```{r}
sessionInfo()
```

