---
title: "Acute full cohort"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 5
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
author: Gisela Gabernet (gisela.gabernet@yale.edu)
date: "`r format(Sys.time(), '%b %d %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(caret)
```

<!-- ################################################################################################### -->
<!-- Mandatory code starts here -->

```{r paths and dependencies}
# Change this to the location where the Bitbucket has been cloned in your home directory (~):
path_to_bitbucket <- "~/repos/impacc-full-cohort"
path_to_codebase <- paste0(path_to_bitbucket,
                           "/Analysis/IMPACC_data_analysis_codebase.R")
source(path_to_codebase)
```


## Read convalescent annotations file
    
```{r read and process convalescent annotations}

convalescent_pro_groups <- read.csv("/scratch/clinical/current/impacc-convalescent.csv")
participant_status <- read.csv("/scratch/clinical/current/impacc_participant_status.csv")

# Convert PRO group cluster numbers to PRO groups as described in paper
# This mapping was communicated by Leying Guan via Teams on May 15th as the 
# original table is unclear about the exact PRO groups mappings
    
convalescent_pro_groups <- convalescent_pro_groups %>% 
  dplyr::mutate( pro_group_labels = case_match( 
    conv_cluster6,
    c(1,3,4) ~ "MIN",
    c(2) ~ "PHY",
    c(5) ~ "COG",
    c(6) ~ "MLT"
  ))

# Check number of patients in each of the groups
conv_cohort_size <- length(unique(convalescent_pro_groups$participant_id))
cat("There are ", conv_cohort_size,
        " participants in the convalescent cohort.")

clindat_integration <- read.csv("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-locked.csv")
integration_participants <- clindat_integration[clindat_integration$integration_cohort==TRUE,]$participant_id
  
cohort_integration_convalescent <- union(integration_participants, convalescent_pro_groups$participant_id)
cat("There are ", length(cohort_integration_convalescent),
        " participants in the union of the integration and convalescent cohorts.")
  
# We select participant IDs in the integration cohort, and also in the convalescent cohort (union), as two participants in the convalescent cohort were not part of the integration cohort.
```


```{r define dataset version, echo = FALSE}
##### Authors: Ravi K. Patel, Jeremy Gygi;

# Indicate the version of the data to use ("current" or a date in "YYYY-MM-DD" format)
# If "current", the data files in "current" directories are used, otherwise, the data files in "legacy" directories that were current on the specified date are used.
DATA_VERSION <- "current"

# Indicate the phases that you want to include.
# NOTE: PHASES = 4 refers to 4a
#.      PHASES = 5 refers to 4b
#.      PHASES = 9 refers to phase 9 (leftover samples)
PHASES <- c(1,2,3)

# Allowed QC? Defaults to pass (including questionable)
ALLOWED_SMPL_STATUS <- c("IMPACC sample assayed and passed QC", 
                         "IMPACC sample assayed with questionable QC")

# Keep only COVID-19 Positive individuals. Defaults to TRUE
# (TRUE will remove all COVID-19 Negative individuals)
KEEP_COVID19_POS <- TRUE
# Keep Escalation visits (Escalation 1, Escalation 2)? Defaults to FALSE
KEEP_ESCALATIONS = FALSE
# Keep Emory Controls? Defaults to FALSE
KEEP_EMORY_CONTROLS = FALSE

# Upper limit for event_date.
# Using higher values (e.g. 999) will include all visits (see KEEP_ESCALATIONS
#   for inclusion of Escalation visits).
EVENT_DATE_UPPER = 999

# Upper limit for event_type. For example, 6 would use up to 'Visit 6'.
# Using higher values (e.g. 999) will include all visits (see KEEP_ESCALATIONS
#   for inclusion of Escalation visits).
VISIT_LOWER = 1
VISIT_UPPER = 6
SELECT_VISITS = NULL

# Use 01/01/2023 locked data in "/data/clinical/current"?
USE_LOCKED_CLINICAL <- TRUE

# Filtering by manuscript cohorts:
# Currently supports core assay / integration manuscript cohort filtering.
FILTER_BY_CORE_ASSAY_COHORT = FALSE
FILTER_BY_INTEGRATION_COHORT = FALSE

# Filter by convalescent cohort participant ID
SELECT_PARTICIPANT_ID = cohort_integration_convalescent

# Exclude samples
EXCLUDE_SAMPLES = NULL
```

<!-- ################################################################################################### -->
<!--           No need to edit any of the code below this point until end of Mandatory Code chunk        -->
<!-- ################################################################################################### -->

```{r Prepare environment, echo=FALSE}
##### Authors: Ravi K. Patel, Jeremy Gygi.
# Set options
options(encoding = "UTF-8") # instead of ANSI
options(scipen = 999) # non-scientific output
```


### ------------------------------------
#### **Study:** IMPACC
#### **Analysis title:** IMPACC Convalescent Data Loading
#### **Analysis description:** Loading combined data from all phases for analyses.
#### **License:** 
#### **Author(s):** Ravi K. Patel (ravi.patel2@ucsf.edu), Jeremy Gygi (jeremy.gygi@yale.edu), Steven Kleinstein (steven.kleinstein@yale.edu), Leying Guan (leying.guan@yale.edu)
#### **Date Created:** `r paste(format(Sys.time(), "%Y-%d-%m %X"), "UTC")`

#### **Source file:** `r knitr::current_input()`
#### **Dataset version:** `r ifelse(DATA_VERSION == "current", paste0("current (date: ", format(Sys.time(),"%Y-%m-%d"), ")"), paste0("legacy (date: ", DATA_VERSION, ")") )` `r paste0("(phases: ", paste0(PHASES, collapse=","), ")") `
#### **R version:** `r paste(R.Version()$version.string)`
#### **Platform:** `r paste(R.Version()$platform)`
#### **Running under**: `r paste(sessionInfo()$running)`

#### **Notes:** The observations for which the clinical data is unavailable or have failed QC are removed.
### ------------------------------------

```{r load header template codebase, echo=FALSE}
# Source the codebase file:
source(path_to_codebase)
```

```{r load datasets, message=FALSE, echo=FALSE}
# Load all available IMPACC datasets associated with DATA_VERSION.
data_env <- load_IMPACC_datasets(# Data version:
                                 DATA_VERSION = DATA_VERSION,
                                 # QC Sample Filtering:
                                 ALLOWED_SMPL_STATUS = ALLOWED_SMPL_STATUS,
                                 # COVID-19 Positive
                                 KEEP_COVID19_POS = KEEP_COVID19_POS,
                                 # Emory controls:
                                 KEEP_EMORY_CONTROLS = KEEP_EMORY_CONTROLS,
                                 # Escalations:
                                 KEEP_ESCALATIONS = KEEP_ESCALATIONS,
                                 # Use the locked clinical data (01/01/2023)
                                 USE_LOCKED_CLINICAL = USE_LOCKED_CLINICAL,
                                 
                                 # Don't restrict filtering by manuscript:
                                 FILTER_BY_CORE_ASSAY_COHORT = FILTER_BY_CORE_ASSAY_COHORT,
                                 FILTER_BY_INTEGRATION_COHORT = FILTER_BY_INTEGRATION_COHORT,
                                 
                                 # Select participant IDs
                                 SELECT_PARTICIPANT_ID = SELECT_PARTICIPANT_ID,
                                 
                                 # Exclude samples
                                 EXCLUDE_SAMPLES = EXCLUDE_SAMPLES,
                                 
                                 # event_type / visit upper limits
                                 EVENT_DATE_UPPER = EVENT_DATE_UPPER,
                                 VISIT_LOWER = VISIT_LOWER,
                                 VISIT_UPPER = VISIT_UPPER,
                                 SELECT_VISITS = SELECT_VISITS,
                                 
                                 # Phases
                                 PHASES = PHASES)

# Load and assign the data objects into the environment
# User can call upon any assay via the name e.g. serum_olink_counts
for( n in names(data_env) ) {
  assign( n, data_env[[n]] )
}

original_clinical_data <- data_env$clinical_data
```

Data summary table

```{r Print file and object names, echo=FALSE}

df <- get_data_summary(data_env)

DT::datatable(df) %>% DT::formatStyle(columns = c(0:4), fontSize = '80%')
```

<!-- Mandatory code ends here -->
<!-- ################################################################################################### -->
#### Add convalescent PRO group info to clinical_data

```{r join clinical data, echo=FALSE, fig.width = 10}
clinical_data <- left_join(clinical_data, convalescent_pro_groups, by="participant_id") %>%
                dplyr::mutate(pro_2_groups =
                    case_match( pro_group_labels,
                    "MIN" ~ "MIN",
                    c("PHY", "COG", "MLT")  ~ "OTHER"
                    )
                )

data_env$clinical_data <- clinical_data
```

#### Add vaccine group info to clinical_data

```{r join vaccination data, echo=FALSE, fig.width = 10}
vaccine_data <- read.csv("~/repos/impacc-full-cohort/Analysis/convalescent-project/vaccination/Convalescent_vaccination_data_with_event_dates.csv")

clinical_data <- left_join(clinical_data, vaccine_data, by=join_by("participant_id"==id))

clinical_data <- clinical_data %>% 
  rowwise() %>%
  mutate(event_after_first_vaccine_dose_NA = 
        dplyr::if_else(event_date >= date_anydose,
                       "yes",
                       "no"),
        event_after_first_vaccine_dose = dplyr::if_else(vac_reported_corrected=="No", 
                                                        "no", 
                                                        event_after_first_vaccine_dose_NA),
        event_date_relative_to_first_dose = event_date - date_anyfirstdose,
        event_date_relative_to_second_dose = event_date - Second,
        event_date_relative_to_booster_dose = event_date - Booster,
        event_date_3weeks_after_first_dose = ifelse((event_date_relative_to_first_dose >= 0 && 
                                                       event_date_relative_to_first_dose <= 21),
                                        1, 0),
        event_date_3weeks_after_second_dose = ifelse((event_date_relative_to_second_dose >= 0 &&
                                                       event_date_relative_to_second_dose <= 21),
                                        1, 0),
        event_date_3weeks_after_booster = ifelse((event_date_relative_to_booster_dose >= 0 &&
                                                       event_date_relative_to_booster_dose <= 21),
                                        1, 0),
        event_date_3weeks_after_dose_NA = ifelse(( event_date_3weeks_after_first_dose |  event_date_3weeks_after_second_dose | event_date_3weeks_after_booster ), "yes", "no"),
        event_date_3weeks_after_dose = ifelse(is.na(event_date_3weeks_after_dose_NA),
                                                    ifelse(event_after_first_vaccine_dose == "no", "no", event_date_3weeks_after_dose_NA),
                                                    event_date_3weeks_after_dose_NA),
        event_date_3weeks_after_dose = ifelse(is.na(event_date_3weeks_after_dose),
                                                    ifelse(vaccinated == "No", "no", event_date_3weeks_after_dose),
                                                    event_date_3weeks_after_dose)
        )

```

#### Predefined color schemes

```{r Define color schemes, echo=FALSE, fig.width = 10}
##### Authors: Ravi K. Patel, Jeremy Gygi
DT::datatable(color_schemes)

### Display color schemes
  legends <- list( 
                  display_color_scheme(color_sex, "Sex"), 
                  display_color_scheme(color_respiratory_status, "Respiratory status"), 
                  display_color_scheme(color_sample_type, "Sample type"), 
                  display_color_scheme(color_race, "Race"), 
                  display_color_scheme(color_event_type, "Event type") 
                )
  
  #ggarrange(plotlist = legends, ncol = 5, nrow = 1)
```



&nbsp;

## Factorizing visits

```{r Subset convalescent, echo=FALSE, fig.height=6, fig.width=10, warning=FALSE}

### Color scheme for pro_group
pro_group_levels <- unique( sort( clinical_data$pro_group_labels ) )
color_pro_group <- get_categorical_color_vector( pro_group_levels )

### Color scheme for visits (adapted so that Visit 10 comes after Visit 9)
all_visits <- c("Visit 1",
             "Visit 2",
             "Visit 3",
             "Visit 4",
             "Visit 5",
             "Visit 6")
clinical_data$event_type <- factor(clinical_data$event_type, levels=all_visits)
visit_levels <- levels(clinical_data$event_type)
color_event_type <- get_ordinal_color_vector( visit_levels )

### Color scheme for phases
phase_levels <- unique(sort(clinical_data$phase))
color_phase <- get_ordinal_color_vector(phase_levels)

# Color scheme for vaccines
vaccine_levels <- unique(sort(clinical_data$vaccinated))
color_vaccinated <- get_categorical_color_vector(vaccine_levels)

# Color scheme for event after first vaccine
event_after_vaccine_levels <- unique(sort(clinical_data$event_after_first_vaccine_dose))
color_event_after_vaccine_levels <- get_categorical_color_vector(event_after_vaccine_levels)

# Color scheme for event 3 weeks after dose
event_3weeks_after_dose_levels <- unique(sort(clinical_data$event_date_3weeks_after_dose))
color_event_3weeks_after_dose <- get_categorical_color_vector(event_3weeks_after_dose_levels)

# Color scheme for resp status v1
traj_group_levels <- unique(sort(as.factor(clinical_data$trajectory_group)))
color_traj_group <- get_ordinal_color_vector(traj_group_levels)
```

# General cohort characteristics

```{r Demographics data, fig.width = 10, fig.height = 9, warning=FALSE, echo=FALSE}
#generate_sample_plots <- function() {

  # Distribution of samples by sex
  bar_sex <- clinical_data %>% count(sex) %>% 
    ggplot( aes(x="Sex", y=n, fill=sex) ) + 
    geom_bar(position="stack", stat="identity") + 
    scale_fill_manual( values = color_sex ) + theme_classic() + 
    labs(x="", y="Event Count", title = "", fill="")
  
  
  # Distribution of samples by respiratory_status
  # bar_respiratory_status <- clinical_data %>% count(resp_status_v1) %>% 
  #   ggplot( aes(x="Respiratory status v1", y=n, fill = as.factor( resp_status_v1 ) ) ) + 
  #   geom_bar(position="stack", stat="identity") + 
  #   scale_fill_manual( values = color_resp_status_v1_levels, na.value = color_na ) + theme_classic() + 
  #   labs(x="", y="Event Count", title = "Respiratory status v1", fill="")

  bar_trajectory_group <- clinical_data %>% count(trajectory_group) %>% 
  ggplot( aes(x="Trajectory group", y=n, fill = as.factor( trajectory_group ) ) ) + 
  geom_bar(position="stack", stat="identity") + 
  scale_fill_manual( values = color_traj_group, na.value = color_na ) + theme_classic() + 
  labs(x="", y="Event Count", title = "Trajectory group", fill="")
  
  # Distribution of samples by PRO group
  bar_PRO_group <- clinical_data %>% count(pro_group_labels) %>% 
    ggplot( aes(x="PRO group", y=n, fill = as.factor( pro_group_labels ) ) ) + 
    geom_bar(position="stack", stat="identity") + 
    scale_fill_manual( values = color_pro_group, na.value = color_na ) + theme_classic() + 
    labs(x="", y="Event Count", title = "PRO group", fill="")
  
  # Adapt factor of clinical data event type so Visit 10 comes after Visit 9
  
  # Distribution of samples by visit
  bar_visit <- clinical_data %>% count(event_type) %>%
    ggplot( aes(x="Event type", y=n, fill = as.factor( event_type ) ) ) + 
    geom_bar(position="stack", stat="identity") + 
    scale_fill_manual( values = color_event_type, na.value = color_na ) + theme_classic() + 
    labs(x="", y="Event Count", title = "Visit", fill="")
  
  # Distribution of samples by phase
  bar_phase <- clinical_data %>% count(phase) %>%
    ggplot( aes(x="Phase", y=n, fill = as.factor( phase ))) + 
    geom_bar(position="stack", stat="identity") + 
    scale_fill_manual( values = color_phase, na.value = color_na ) + theme_classic() + 
    labs(x="", y="Event Count", title = "Phase", fill="")
  
  # Distribution of samples by vaccine state
  bar_vaccinated <- clinical_data %>% count(vaccinated) %>%
    ggplot( aes(x="Vaccinated", y=n, fill = as.factor( vaccinated ) ) ) + 
    geom_bar(position="stack", stat="identity") + 
    scale_fill_manual( values = color_vaccinated, na.value = color_na ) + theme_classic() + 
    labs(x="", y="Event Count", title = "Vaccinated", fill="")
  
  bar_event_after_first_vaccine <- clinical_data %>% count(event_after_first_vaccine_dose) %>%
    ggplot( aes(x="Event_after_first_vaccine", y=n, fill = as.factor(event_after_first_vaccine_dose)) ) +
    geom_bar(position="stack", stat = "identity") +
    scale_fill_manual( values = color_event_after_vaccine_levels, na.value = color_na) + theme_classic() +
    labs(x="", y="Event Count", title = "After first vac dose", fill="")
  
  bar_event_3weeks_after_dose <- clinical_data %>% count(event_date_3weeks_after_dose) %>%
    ggplot( aes(x="Event_within_3weeks_after_dose", y=n, fill = as.factor(event_date_3weeks_after_dose)) ) +
    geom_bar(position="stack", stat = "identity") +
    scale_fill_manual( values = color_event_3weeks_after_dose, na.value = color_na) + theme_classic() +
    labs(x="", y="Event Count", title = "3w after vac dose", fill="")
  
  
  print(ggarrange(bar_sex, bar_trajectory_group, bar_PRO_group, 
                  bar_visit, bar_phase, bar_vaccinated,
                  bar_event_after_first_vaccine, bar_event_3weeks_after_dose,
            labels = c("A", "B", "C", "D", "E", "F", "G", "H"), ncol = 4, nrow = 3))
```

```{r Respiratory status heatmap}
  # Generate heatmap:
  melted.df <-
    tidyr::pivot_longer(clinical_data, cols = respiratory_status, )
  melted.df <-
    mutate(melted.df, respiratory_status = as.character(value))
  ggplot(melted.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = respiratory_status)) +
    # add color scale:
    scale_fill_manual(values = color_respiratory_status, na.value = color_na) +
    ggtitle("Respiratory status heatmap") +
    theme_bw() +
    theme(axis.text.x = element_blank())
  
#}
```

```{r Phase heatmap}
  # Generate heatmap:
  melted.df <-
    tidyr::pivot_longer(clinical_data, cols = phase, )
  melted.df <-
    mutate(melted.df, phase = as.character(value))
  ggplot(melted.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = phase)) +
    # add color scale:
    scale_fill_manual(values = color_phase, na.value = color_na) +
    ggtitle("Phase heatmap") +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
#}
```


```{r Respiratory status by phase}
  melted.df <-
    mutate(clinical_data, respiratory_status = as.character(respiratory_status))
  ggplot(melted.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = respiratory_status)) +
    # add color scale:
    scale_fill_manual(values = color_respiratory_status, na.value = color_na) +
    ggtitle("Respiratory status heatmap split by phase") +
    facet_wrap(~phase, ncol=6) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
#}
```

# Samples per omics layer

### Targeted plasma proteomics checks

```{r Targeted plasma proteomics}
  modality <- plasma_proteomics_targeted_counts
  modality_name <- "Targeted plasma proteomics"

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality))
  ggplot(plot.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = sample_type)) +
    # add color scale:
    scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
```

```{r Targeted plasma proteomics histogram}

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality)) %>%
                group_by(event_id) %>%
                summarize(sample_per_event_count = n())
  ggplot(plot.df, aes(x = sample_per_event_count)) +
    geom_histogram(binwidth = 1) +
    stat_count(geom = 'text',
             color = 'black',
             aes(label = after_stat(count)),
           position = position_stack(vjust = 1)) +
    # add color scale:
    #scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw()
```

```{r Targeted plasma proteomics check 2 samples per event id}
  count_gt1 <- plot.df %>% filter(sample_per_event_count >1)
  tab <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            filter(event_id %in% count_gt1$event_id) %>%
            dplyr::select(sample_id, sample_type, event_id, event_type, participant_id, phase)
  DT::datatable(tab) %>% DT::formatStyle(columns = c(1:ncol(tab)), fontSize = '80%')
  
```

```{r Targeted plasma proteomics event histogram}

plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality)) %>%
              dplyr::select(participant_id, event_id, phase, event_type) %>%
              distinct() %>%
              group_by(participant_id, event_type, event_id) %>%
              summarize(event_per_visit_count = n())
ggplot(plot.df, aes(x = event_per_visit_count)) +
  geom_histogram(binwidth = 1) +
  stat_count(geom = 'text',
           color = 'black',
           aes(label = after_stat(count)),
         position = position_stack(vjust = 1)) +
  # add color scale:
  #scale_fill_manual(values = color_sample_type, na.value = color_na) +
  ggtitle(modality_name) +
  theme_bw()
```

```{r Targeted plasma proteomics check event_id and event_type unique}
  count_gt1 <- plot.df %>% filter(event_per_visit_count >1)
  tab <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            filter(event_id %in% count_gt1$event_id) %>%
            dplyr::select(sample_id, sample_type, event_id, event_type, participant_id, phase)
  DT::datatable(tab) %>% DT::formatStyle(columns = c(1:ncol(tab)), fontSize = '80%')
  
```

```{r Targeted plasma proteomics number of samples per visit}
  tab_targeted_plasma_proteomics <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            group_by(event_type) %>%
            summarize(modality = n()) %>%
            pivot_wider(names_from = event_type, values_from = modality) %>%
            mutate( omics = modality_name )
  DT::datatable(tab_targeted_plasma_proteomics) %>% DT::formatStyle(columns = c(1:ncol(tab)))
  
```

### Global proteomics checks

```{r Global plasma proteomics}
  modality <- plasma_proteomics_global_dda_counts
  modality_name <- "Global plasma proteomics"

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality))
  ggplot(plot.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = sample_type)) +
    # add color scale:
    scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
```

```{r Global plasma proteomics histogram}

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality)) %>%
                group_by(event_id) %>%
                summarize(sample_per_event_count = n())
  ggplot(plot.df, aes(x = sample_per_event_count)) +
    geom_histogram(binwidth = 1) +
    stat_count(geom = 'text',
             color = 'black',
             aes(label = after_stat(count)),
           position = position_stack(vjust = 1)) +
    # add color scale:
    #scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw()
```

```{r Global plasma proteomics check 2 samples per event id}
  count_gt1 <- plot.df %>% filter(sample_per_event_count >1)
  tab <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            filter(event_id %in% count_gt1$event_id) %>%
            dplyr::select(sample_id, sample_type, event_id, event_type, participant_id, phase)
  DT::datatable(tab) %>% DT::formatStyle(columns = c(1:ncol(tab)), fontSize = '80%')
  
```

```{r Global plasma proteomics number of samples per visit}
  tab_global_plasma_proteomics <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            group_by(event_type) %>%
            summarize(modality = n()) %>%
          pivot_wider(names_from = event_type, values_from = modality) %>%
          mutate( omics = modality_name )
  DT::datatable(tab_global_plasma_proteomics) %>% DT::formatStyle(columns = c(1:ncol(tab)))
  
```

### PBMC transcriptomics

```{r PBMC transcriptomics}
  modality <- pbmc_transcriptomics_counts
  modality_name <- "PBMC transcriptomics"
  
  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality))
  ggplot(plot.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = sample_type)) +
    # add color scale:
    scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
```

```{r PBMC transcriptomics histogram}

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality)) %>%
                group_by(event_id) %>%
                summarize(sample_per_event_count = n())
  ggplot(plot.df, aes(x = sample_per_event_count)) +
    geom_histogram(binwidth = 1) +
    stat_count(geom = 'text',
             color = 'black',
             aes(label = after_stat(count)),
           position = position_stack(vjust = 1)) +
    # add color scale:
    #scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw()
```

```{r PBMC transcriptomics check 2 samples per event id}
  count_gt1 <- plot.df %>% filter(sample_per_event_count >1)
  tab <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            filter(event_id %in% count_gt1$event_id) %>%
            dplyr::select(sample_id, sample_type, event_id, event_type, participant_id, phase)
  DT::datatable(tab) %>% DT::formatStyle(columns = c(1:ncol(tab)), fontSize = '80%')
  
```

```{r PBMC transcriptomics number of samples per visit}
  tab_pbmc_transcriptomics <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            group_by(event_type) %>%
            summarize(modality = n()) %>%
            pivot_wider(names_from = event_type, values_from = modality) %>%
            mutate( omics = modality_name )
  DT::datatable(tab_pbmc_transcriptomics) %>% DT::formatStyle(columns = c(1:ncol(tab)))
  
```

### Olink targeted serum

```{r serum Olink targeted}
  modality <- serum_olink_counts
  modality_name <- "Serum Olink"

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality))
  ggplot(plot.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = sample_type)) +
    # add color scale:
    scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
```

```{r Serum Olink histogram}

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality)) %>%
                group_by(event_id) %>%
                summarize(sample_per_event_count = n())
  ggplot(plot.df, aes(x = sample_per_event_count)) +
    geom_histogram(binwidth = 1) +
    stat_count(geom = 'text',
             color = 'black',
             aes(label = after_stat(count)),
           position = position_stack(vjust = 1)) +
    # add color scale:
    #scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw()
```

```{r Serum Olink check 2 samples per event id}
  count_gt1 <- plot.df %>% filter(sample_per_event_count >1)
  tab <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            filter(event_id %in% count_gt1$event_id) %>%
            dplyr::select(sample_id, sample_type, event_id, event_type, participant_id, phase)
  DT::datatable(tab) %>% DT::formatStyle(columns = c(1:ncol(tab)), fontSize = '80%')
  
```

```{r Serum Olink number of samples per visit}
  tab_olink_targeted_serum <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            group_by(event_type) %>%
            summarize( modality = n()) %>%
            pivot_wider(names_from = event_type, values_from = modality) %>%
            mutate( omics = modality_name )
  DT::datatable(tab_olink_targeted_serum) %>% DT::formatStyle(columns = c(1:ncol(tab)))
  
```

### Plasma metabolomics

```{r Plasma metabolomics}
  modality <- plasma_metabolomics_global_counts
  modality_name <- "Plasma metabolomics"

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality))
  ggplot(plot.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = sample_type)) +
    # add color scale:
    scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
```

```{r Plasma metabolomics histogram}

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality)) %>%
                group_by(event_id) %>%
                summarize(sample_per_event_count = n())
  ggplot(plot.df, aes(x = sample_per_event_count)) +
    geom_histogram(binwidth = 1) +
    stat_count(geom = 'text',
             color = 'black',
             aes(label = after_stat(count)),
           position = position_stack(vjust = 1)) +
    # add color scale:
    #scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw()
```

```{r Plasma metabolomics check 2 samples per event id}
  count_gt1 <- plot.df %>% filter(sample_per_event_count >1)
  tab <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            filter(event_id %in% count_gt1$event_id) %>%
            dplyr::select(sample_id, sample_type, event_id, event_type, participant_id, phase)
  DT::datatable(tab) %>% DT::formatStyle(columns = c(1:ncol(tab)), fontSize = '80%')
  
```

```{r plasma metabolomics number of samples per visit}
  tab_plasma_metabolomics <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            group_by(event_type) %>%
            summarize( modality = n()) %>%
            pivot_wider(names_from = event_type, values_from = modality) %>%
            mutate( omics = modality_name )
  DT::datatable(tab_plasma_metabolomics) %>% DT::formatStyle(columns = c(1:ncol(tab)))
  
```

### Nasal transcriptomics

```{r Nasal transcriptomics}
  modality <- nasal_transcriptomics_counts
  modality_name <- "Nasal transcriptomics"

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality))
  ggplot(plot.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = sample_type)) +
    # add color scale:
    scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
```

```{r Nasal transcriptomics histogram}

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality)) %>%
                group_by(event_id) %>%
                summarize(sample_per_event_count = n())
  ggplot(plot.df, aes(x = sample_per_event_count)) +
    geom_histogram(binwidth = 1) +
    stat_count(geom = 'text',
             color = 'black',
             aes(label = after_stat(count)),
           position = position_stack(vjust = 1)) +
    # add color scale:
    #scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw()
```

```{r Nasal transcriptomics check 2 samples per event id}
  count_gt1 <- plot.df %>% filter(sample_per_event_count >1)
  tab <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            filter(event_id %in% count_gt1$event_id) %>%
            dplyr::select(sample_id, sample_type, event_id, event_type, participant_id, phase)
  DT::datatable(tab) %>% DT::formatStyle(columns = c(1:ncol(tab)), fontSize = '80%')
  
```

```{r Nasal transcriptomics number of samples per visit}
  tab_nasal_transcriptomics <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            group_by(event_type) %>%
            summarize( modality = n()) %>%
            pivot_wider(names_from = event_type, values_from = modality) %>%
            mutate( omics = modality_name )
  DT::datatable(tab_nasal_transcriptomics) %>% DT::formatStyle(columns = c(1:ncol(tab)))
  
```

### Blood cytof

```{r Blood cytof}
  modality <- bld_cytof_msi_counts
  modality_name <- "Blood cytof"

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality))
  ggplot(plot.df) +
    geom_tile(aes(x = participant_id, y = event_type, fill = sample_type)) +
    # add color scale:
    scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  
```

```{r Blood cytof histogram}

  plot.df <- clinical_data %>% filter(sample_id %in% row.names(modality)) %>%
                group_by(event_id) %>%
                summarize(sample_per_event_count = n())
  ggplot(plot.df, aes(x = sample_per_event_count)) +
    geom_histogram(binwidth = 1) +
    stat_count(geom = 'text',
             color = 'black',
             aes(label = after_stat(count)),
           position = position_stack(vjust = 1)) +
    # add color scale:
    #scale_fill_manual(values = color_sample_type, na.value = color_na) +
    ggtitle(modality_name) +
    theme_bw()
```

```{r Blood cytof check 2 samples per event id}
  count_gt1 <- plot.df %>% filter(sample_per_event_count >1)
  tab <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            filter(event_id %in% count_gt1$event_id) %>%
            dplyr::select(sample_id, sample_type, event_id, event_type, participant_id, phase)
  DT::datatable(tab) %>% DT::formatStyle(columns = c(1:ncol(tab)), fontSize = '80%')
  
```

```{r Blood cytof number of samples per visit}
  tab_blood_cytof <- clinical_data %>% 
            filter(sample_id %in% row.names(modality)) %>% 
            group_by(event_type) %>%
            summarize( modality = n()) %>%
            pivot_wider(names_from = event_type, values_from = modality) %>%
            mutate( omics = modality_name )
  DT::datatable(tab_blood_cytof) %>% DT::formatStyle(columns = c(1:ncol(tab)))
  
```

# Summary of samples per omics modality

```{r Summary omics modality}
df_omics <- bind_rows(
  tab_pbmc_transcriptomics,
  tab_global_plasma_proteomics,
  tab_targeted_plasma_proteomics,
  tab_nasal_transcriptomics,
  tab_olink_targeted_serum,
  tab_plasma_metabolomics,
  tab_blood_cytof
) %>% 
  relocate(omics) %>%
  rowwise() %>%
  mutate(total = sum(`Visit 1`, `Visit 2`, `Visit 3`, `Visit 4`, `Visit 6`)) %>%
  ungroup()

DT::datatable(df_omics) %>% DT::formatStyle(columns = c(1:ncol(tab)))
```

## Total number of samples per visit

```{r Summary visits}
df_visits <- df_omics %>% 
                dplyr::select(!c("omics")) %>%
                summarise_all(sum) %>%
                mutate(total = rowSums(.))
                

DT::datatable(df_visits) %>% DT::formatStyle(columns = c(1:ncol(tab)))
```
```{r}
so_exclude <- c("0865-002K3G00-002", "0865-0041HH00-002", "0865-00BJWD00-002", 
                                                   "0865-002NVJ00-002", "0865-0062BG00-002", "0865-00DH6J00-002",
                                                   "0865-00G24H00-002", "0865-00G66B00-002", "0865-00HMPH00-002")
ppg_exclude <- c("0865-006X1H00-005", "0865-006X9D00-004", "0865-0071BB00-004", "0865-0071FH00-005", "0865-0071JJ00-005")
ppt_exclude <- c("0865-006V0J00-005", "0865-006X1H00-005", "0865-006X9D00-004", "0865-0071JJ00-005")
pmg_exclude <- c("0865-006V0J00-005", "0865-006X1H00-005", "0865-006X9D00-004", "0865-0071FH00-005", "0865-0071JJ00-005")
ngx_exclude <- c("0865-00CD0B00-001", "	0865-004D4K00-001")
bct_exclude <- c()

serum_olink_counts <- serum_olink_counts[!row.names(serum_olink_counts) %in% so_exclude,]
plasma_proteomics_global_dda_counts <- plasma_proteomics_global_dda_counts[!row.names(plasma_proteomics_global_dda_counts) %in% ppg_exclude,]
plasma_proteomics_targeted_counts <- plasma_proteomics_targeted_counts[!row.names(plasma_proteomics_targeted_counts) %in% ppt_exclude,]
plasma_metabolomics_global_counts <- plasma_metabolomics_global_counts[!row.names(plasma_metabolomics_global_counts) %in% pmg_exclude,]
nasal_transcriptomics_counts <- nasal_transcriptomics_counts[!row.names(nasal_transcriptomics_counts) %in% ngx_exclude,]
bld_cytof_msi_counts <- bld_cytof_msi_counts[!row.names(bld_cytof_msi_counts) %in% bct_exclude,]

```

```{r}
all_sample_ids <- unique(c(row.names(pbmc_transcriptomics_counts),
                           row.names(plasma_proteomics_targeted_counts),
                           row.names(plasma_proteomics_global_dda_counts),
                           row.names(plasma_metabolomics_global_counts),
                           row.names(serum_olink_counts),
                          row.names(bld_cytof_msi_counts)
                           ))
```


# Load data by train test

```{r train-test split}
data_env_convalescent <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_convalescent_data_env_all_Feb_11_2024.RDS")

original_partition <- data_env_convalescent$clinical_data %>% dplyr::select(participant_id, partition)

train <- original_partition %>% filter(partition == "train")

# treat the rest of participants / events as test
clinical_data <- clinical_data %>% mutate(partition = 
                                            case_when (
                                              participant_id %in% train$participant_id ~ "train",
                                              .default = "test"
                                            ))
data_env$clinical_data <- clinical_data
event_id_train <- clinical_data %>% filter(partition == "train") %>% pull(event_id)
event_id_test <- clinical_data %>% filter(partition == "test") %>% pull(event_id)
participant_id_train <- clinical_data %>% filter(partition == "train") %>% pull(participant_id)
participant_id_test <- clinical_data %>% filter(partition == "test") %>% pull(participant_id)
```

```{r}
#The extra added metadata is needed

data_env_train <- load_IMPACC_datasets(# Data version:
                                 DATA_VERSION = DATA_VERSION,
                                 ALLOWED_SMPL_STATUS = ALLOWED_SMPL_STATUS,
                                 KEEP_COVID19_POS = KEEP_COVID19_POS,
                                 KEEP_EMORY_CONTROLS = KEEP_EMORY_CONTROLS,
                                 KEEP_ESCALATIONS = KEEP_ESCALATIONS,
                                 USE_LOCKED_CLINICAL = USE_LOCKED_CLINICAL,
                                 FILTER_BY_CORE_ASSAY_COHORT = FILTER_BY_CORE_ASSAY_COHORT,
                                 FILTER_BY_INTEGRATION_COHORT = FILTER_BY_INTEGRATION_COHORT,
                                 # Select train participant IDs
                                 SELECT_PARTICIPANT_ID = participant_id_train,
                                 EXCLUDE_SAMPLES = EXCLUDE_SAMPLES,
                                 EVENT_DATE_UPPER = EVENT_DATE_UPPER,
                                 VISIT_LOWER = VISIT_LOWER,
                                 VISIT_UPPER = VISIT_UPPER,
                                 SELECT_VISITS = SELECT_VISITS,
                                 PHASES = PHASES)

# Extra metadata is needed
data_env_train$clinical_data <- clinical_data %>% filter(event_id %in% event_id_train)

# Save matrices with excluded samples
data_env_train$serum_olink_counts <- data_env_train$serum_olink_counts[!row.names(data_env_train$serum_olink_counts) %in% so_exclude,]
data_env_train$plasma_proteomics_global_dda_counts <- data_env_train$plasma_proteomics_global_dda_counts[!row.names(data_env_train$plasma_proteomics_global_dda_counts) %in% ppg_exclude,]
data_env_train$plasma_proteomics_targeted_counts <- data_env_train$plasma_proteomics_targeted_counts[!row.names(data_env_train$plasma_proteomics_targeted_counts) %in% ppt_exclude,]
data_env_train$plasma_metabolomics_global_counts <- data_env_train$plasma_metabolomics_global_counts[!row.names(data_env_train$plasma_metabolomics_global_counts) %in% pmg_exclude,]
data_env_train$nasal_transcriptomics_counts <- data_env_train$nasal_transcriptomics_counts[!row.names(data_env_train$nasal_transcriptomics_counts) %in% ngx_exclude,]
data_env_train$bld_cytof_msi_counts <- data_env_train$bld_cytof_msi_counts[!row.names(data_env_train$bld_cytof_msi_counts) %in% bct_exclude,]

data_env_test <- load_IMPACC_datasets(# Data version:
                                 DATA_VERSION = DATA_VERSION,
                                 ALLOWED_SMPL_STATUS = ALLOWED_SMPL_STATUS,
                                 KEEP_COVID19_POS = KEEP_COVID19_POS,
                                 KEEP_EMORY_CONTROLS = KEEP_EMORY_CONTROLS,
                                 KEEP_ESCALATIONS = KEEP_ESCALATIONS,
                                 USE_LOCKED_CLINICAL = USE_LOCKED_CLINICAL,
                                 FILTER_BY_CORE_ASSAY_COHORT = FILTER_BY_CORE_ASSAY_COHORT,
                                 FILTER_BY_INTEGRATION_COHORT = FILTER_BY_INTEGRATION_COHORT,
                                 # select test participant IDs
                                 SELECT_PARTICIPANT_ID = participant_id_test,
                                 EXCLUDE_SAMPLES = EXCLUDE_SAMPLES,
                                 EVENT_DATE_UPPER = EVENT_DATE_UPPER,
                                 VISIT_LOWER = VISIT_LOWER,
                                 VISIT_UPPER = VISIT_UPPER,
                                 SELECT_VISITS = SELECT_VISITS,
                                 PHASES = PHASES)

# Extra metadata is needed
data_env_test$clinical_data <- clinical_data %>% filter(event_id %in% event_id_test)

# Save matrices with excluded samples
data_env_test$serum_olink_counts <- data_env_test$serum_olink_counts[!row.names(data_env_test$serum_olink_counts) %in% so_exclude,]
data_env_test$plasma_proteomics_global_dda_counts <- data_env_test$plasma_proteomics_global_dda_counts[!row.names(data_env_test$plasma_proteomics_global_dda_counts) %in% ppg_exclude,]
data_env_test$plasma_proteomics_targeted_counts <- data_env_test$plasma_proteomics_targeted_counts[!row.names(data_env_test$plasma_proteomics_targeted_counts) %in% ppt_exclude,]
data_env_test$plasma_metabolomics_global_counts <- data_env_test$plasma_metabolomics_global_counts[!row.names(data_env_test$plasma_metabolomics_global_counts) %in% pmg_exclude,]
data_env_test$nasal_transcriptomics_counts <- data_env_test$nasal_transcriptomics_counts[!row.names(data_env_test$nasal_transcriptomics_counts) %in% ngx_exclude,]
data_env_test$bld_cytof_msi_counts <- data_env_test$bld_cytof_msi_counts[!row.names(data_env_test$bld_cytof_msi_counts) %in% bct_exclude,]
```

# Preprocess data by train test


```{r preprocess assays, eval=T, warning=F}
# Get selected features from convalescent assays train
preprocessed_assays_conv_train <- readRDS("/scratch/data-fullcohort/batch-effects/preprocessed_assays_train_Feb_11_2024.RDS")

# Get features for test data feature selection
PPT_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$plasma_proteomics_targeted)
PPG_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$plasma_proteomics_global_dda)

#cytof MSI features
BCT_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$bld_msi_cytof)

# PBMC transcriptomics features
PGX_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$pbmc_transcriptomics)

# Plasma metabolomics features
PMG_selected_features <- colnames(preprocessed_assays_conv_train$preprocessed_data$plasma_metabolomics_global)
```

```{r preprocess acute train data}
preprocessed_assays_train <- preprocess_IMPACC_datasets(data_env = data_env_train,
                                              assays_to_process = c("plasma_proteomics_targeted",
                                                                    "pbmc_transcriptomics",
                                                                    "serum_olink",
                                                                    "bld_msi_cytof",
                                                                    "plasma_metabolomics_global"
                                                                    ),
                                              transcriptomics_batch_correction = TRUE,
                                              #transcriptomics_expression_filter_grouping = "pro_group_labels",
                                              #Cannot select pro_group_labels as there are NAs
                                              transcriptomics_limma_matrix_design = NULL,
                                              transcriptomics_limma_matrix_design_variables = NULL,
                                              transcriptomics_selected_features = PGX_selected_features,
                                              proteomics_selected_features = PPT_selected_features,
                                              proteomics_batch_correction = T,
                                              proteomics_combat_matrix_design = NULL,
                                              proteomics_combat_matrix_design_variables = NULL,
                                              metabolomics_selected_features = PMG_selected_features,
                                              additional_pvca_variables = c("event_type"),
                                              cytof_msi_batch_correction = TRUE,
                                              cytof_msi_limma_matrix_design = NULL,
                                              cytof_msi_limma_matrix_design_variables = NULL,
                                              cytof_msi_selected_features = BCT_selected_features,
                                              run_pvca=T)

preprocessed_PPG_train <- preprocess_IMPACC_datasets(data_env = data_env_train,
                                              assays_to_process = c(
                                                                    "plasma_proteomics_global_dda"
                                                                    ),
                                              transcriptomics_batch_correction = TRUE,
                                              transcriptomics_limma_matrix_design = NULL,
                                              transcriptomics_limma_matrix_design_variables = NULL,
                                              proteomics_selected_features = PPG_selected_features,
                                              proteomics_batch_correction = T,
                                              proteomics_combat_matrix_design = NULL,
                                              proteomics_combat_matrix_design_variables = NULL,
                                              additional_pvca_variables = c("event_type"),
                                              run_pvca=T)
                                                                      
preprocessed_assays_train$preprocessed_data$plasma_proteomics_global_dda <- preprocessed_PPG_train$preprocessed_data$plasma_proteomics_global_dda
preprocessed_assays_train$pvca_data <- bind_rows(preprocessed_assays_train$pvca_data, preprocessed_PPG_train$pvca_data)
preprocessed_assays_train$pvca_plots$plasma_proteomics_global_dda <- preprocessed_PPG_train$pvca_plots$plasma_proteomics_global_dda
```



```{r preprocess acute test data}
preprocessed_assays_test <- preprocess_IMPACC_datasets(data_env = data_env_test,
                                              assays_to_process = c("plasma_proteomics_targeted",
                                                                    "pbmc_transcriptomics",
                                                                    "serum_olink",
                                                                    "bld_msi_cytof",
                                                                    "plasma_metabolomics_global"
                                                                    ),
                                              transcriptomics_batch_correction = TRUE,
                                              transcriptomics_limma_matrix_design = NULL,
                                              transcriptomics_limma_matrix_design_variables = NULL,
                                              transcriptomics_selected_features = PGX_selected_features,
                                              proteomics_selected_features = PPT_selected_features,
                                              proteomics_batch_correction = T,
                                              proteomics_combat_matrix_design = NULL,
                                              proteomics_combat_matrix_design_variables = NULL,
                                              metabolomics_selected_features = PMG_selected_features,
                                              additional_pvca_variables = c("event_type"),
                                              cytof_msi_batch_correction = TRUE,
                                              cytof_msi_limma_matrix_design = NULL,
                                              cytof_msi_limma_matrix_design_variables = NULL,
                                              cytof_msi_selected_features = BCT_selected_features,
                                              run_pvca=T)

preprocessed_PPG_test <- preprocess_IMPACC_datasets(data_env = data_env_test,
                                              assays_to_process = c(
                                                                    "plasma_proteomics_global_dda"
                                                                    ),
                                              transcriptomics_batch_correction = TRUE,
                                              proteomics_selected_features = PPG_selected_features,
                                              proteomics_batch_correction = T,
                                              proteomics_combat_matrix_design = NULL,
                                              proteomics_combat_matrix_design_variables = NULL,
                                              additional_pvca_variables = c("event_type"),
                                              run_pvca=T)
                                                                      
preprocessed_assays_test$preprocessed_data$plasma_proteomics_global_dda <- preprocessed_PPG_test$preprocessed_data$plasma_proteomics_global_dda
preprocessed_assays_test$pvca_data <- bind_rows(preprocessed_assays_test$pvca_data, preprocessed_PPG_test$pvca_data)
preprocessed_assays_test$pvca_plots$plasma_proteomics_global_dda <- preprocessed_PPG_test$pvca_plots$plasma_proteomics_global_dda
```

```{r plot_assays, eval = T}
dir.create("plot_assays")

assays_processed = c("plasma_proteomics_targeted",
                      "plasma_proteomics_global_dda",
                      "plasma_metabolomics_global",
                      "pbmc_transcriptomics",
                      "serum_olink",
                      "bld_msi_cytof"
                      )


```


# Save objects
```{r eval=T}
saveRDS(data_env, file=paste0("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_allcohort_all_",
                             format(Sys.time(), '%b_%d_%Y'),".RDS"))
saveRDS(data_env_train, file=paste0("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_allcohort_train_",
                             format(Sys.time(), '%b_%d_%Y'),".RDS"))
saveRDS(data_env_test, file=paste0("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_allcohort_test_",
                             format(Sys.time(), '%b_%d_%Y'),".RDS"))
saveRDS(preprocessed_assays_train, file=paste0("/scratch/data-fullcohort/batch-effects/preprocessed_assays_acute_allcohort_train_",
                            format(Sys.time(), '%b_%d_%Y'),".RDS"))
saveRDS(preprocessed_assays_test, file=paste0("/scratch/data-fullcohort/batch-effects/preprocessed_assays_acute_allcohort_test_",
                            format(Sys.time(), '%b_%d_%Y'),".RDS"))
```


```{r eval=F}
ds = c("plasma_proteomics_targeted",
            "plasma_proteomics_global_dda",
            "plasma_metabolomics_global",
            "pbmc_transcriptomics",
            #"nasal_transcriptomics",
            "serum_olink",
           "bld_msi_cytof"
            )

# save train
preprocessed_data_train = preprocessed_assays_train$preprocessed_data
pvca_data = preprocessed_assays_train$pvca_data %>% dplyr::filter(str_remove(preprocessed_assays_train$pvca_data$assay, "_[^_]+$") %in% ds)
pvca_data = pivot_wider(pvca_data, names_from = effect, values_from = explained) %>% column_to_rownames("assay")

pvca_dir <- "pvca_plots_preprocess_acute_allcohort_train"
dir.create(pvca_dir)

prefix = "pvca_"
lapply(names(preprocessed_assays_train$pvca_plots), 
  function(x)ggsave(filename=paste(pvca_dir, "/",prefix, x,"_p0_",format(Sys.time(), '%b_%d_%Y'),".png",sep=""),
                    plot=preprocessed_assays_train$pvca_plots[[x]]$p0, width = 10, height = 7))
lapply(names(preprocessed_assays_train$pvca_plots), 
  function(x)ggsave(filename=paste(pvca_dir,"/",prefix,x,"_p1_",format(Sys.time(), '%b_%d_%Y'),".png",sep=""),
                    plot=preprocessed_assays_train$pvca_plots[[x]]$p1, width = 10, height = 7))

pvca_dir <- "pvca_plots_preprocess_acute_allcohort_test"
dir.create(pvca_dir)
# save test 
preprocessed_data_test = preprocessed_assays_test$preprocessed_data
pvca_data = preprocessed_assays_test$pvca_data %>% dplyr::filter(str_remove(preprocessed_assays_test$pvca_data$assay, "_[^_]+$") %in% ds)
pvca_data = pivot_wider(pvca_data, names_from = effect, values_from = explained) %>% column_to_rownames("assay")


prefix = "pvca_"
lapply(names(preprocessed_assays_test$pvca_plots), 
  function(x)ggsave(filename=paste(pvca_dir, "/",prefix, x,"_p0_",format(Sys.time(), '%b_%d_%Y'),".png",sep=""),
                    plot=preprocessed_assays_test$pvca_plots[[x]]$p0, width = 10, height = 7))
lapply(names(preprocessed_assays_test$pvca_plots), 
  function(x)ggsave(filename=paste(pvca_dir,"/",prefix,x,"_p1_",format(Sys.time(), '%b_%d_%Y'),".png",sep=""),
                    plot=preprocessed_assays_test$pvca_plots[[x]]$p1, width = 10, height = 7))
```


```{r print session info}
sessionInfo()
```

