---
title: "Figure3_4"
author: "Abigail Glascock"
date: "2025-09-24"
output: html_document
---
## Figure 3 - Nasal AMR Results

```{r data import and setup}
#| echo: false
#| warning: false

### load packages ###
library(nlme)
library(lme4)
library(lmerTest)
library(ggeffects)
library(tidyverse)
library(rstatix)
library(ggpubr) ## For stat_cor
library(scales) ## For log scaling of y axis (makes it look pretty)
library(ggridges) ## geom_density_ridges is really nice for distribution graphs
library(ggbeeswarm) ## For geom_quasirandom(), the best looking jitter plots
library(readxl)
library(vegan)
library(RColorBrewer)

### Load clinical data ###  
clinical_data_ext <- read_csv("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-supp-locked.csv")
clinical_data1 <- read_csv("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-sample-locked.csv")
clinical_data2 <- read_csv("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-locked.csv")
clinical_data3 <- read_csv("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-event-locked.csv")
clinical_data <- merge(clinical_data1, clinical_data2, by = "participant_id")
clinical_data <- merge(clinical_data, clinical_data3, by= c("event_id", "participant_id"))
clinical_data<-merge(clinical_data,clinical_data_ext, by="participant_id")
clinical_data$discretized_admit_age_quantile <- cut(clinical_data$admit_age, breaks = c(17,46,54,62,70, 96), include.lowest = FALSE)
rm(clinical_data1, clinical_data2, clinical_data3)

days_hosp<-clinical_data %>% dplyr::filter(event_date<=10) %>% dplyr::select(sample_id)

### Load Metagenomics Metadata ###
simplified.taxa.out.name <- read.csv("/scratch/metagenomics/all_ncbi_tax_id/taxonomy_ready_common_names.csv", row.names = 1)
nasal_metagenomics_metadata <- read_csv("/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv")

### load ABX info, excluded samples, colors, etc. ###
TrajGrpCol <- c("1" = "#639A21", "2" = "#39828C", "3" = "#6371AD", "4" = "#BD7D31", "5" = "#9C3418")

antibiotic_data <- read_csv("/scratch/metagenomics/antibiotic_data/antibiotics_data.csv")
colnames(antibiotic_data)[1] <- "participant_id" ## Make this matching to make merging easier with other IMPACC dataframes

ABXcoverage <- read_excel("/scratch/metagenomics/antibiotic_data/Antibiotic coverage.xls", sheet = 1)%>%
  dplyr::select(-X_ANTIMICROBIALMED, -scale) %>%
  dplyr::filter(!is.na(`Score Total`)) %>%
  rowwise() %>%
  dplyr::mutate(`Score Total` = sum(`GP Coverage`, `GN coverage`, `Anaerobic Coverage`)) %>%
  distinct()
Gerbercoverage <- read_excel("/scratch/metagenomics/antibiotic_data/Antibiotic coverage.xls", sheet = 2)

antibiotic_class_mapping <- read_excel("/scratch/metagenomics/antibiotic_data/Antibiotic_data_table_CL.xlsx")  %>%
  dplyr::filter(!grepl("Ignore", `Med group`)) %>%
  left_join(ABXcoverage) %>%
  rename(med_group = `Med group`)

n <- 100
qual_col_pals = brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector = unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))

exclude_participant <- antibiotic_data %>% dplyr::filter(conmed_dose_start > conmed_dose_stop)
exclude_ABX_metadata <- plyr::rbind.fill(nasal_metagenomics_metadata) %>%
  dplyr::select(sample_id) %>%
  left_join(clinical_data %>% dplyr::select(sample_id, participant_id)) %>%
  dplyr::filter(participant_id %in% exclude_participant$participant_id)
exclude_samples = c("0865-00F2MG00-001", "0865-003Y5D00-001_REMOVE", "0865-004KMH00-001", "0865-00DGJG00-001")
exclude_samples <- unique(c(exclude_samples, exclude_ABX_metadata$sample_id ))

CI_interval = 0.95

### Metagenomics Data 
nasal_long <- read_csv("/scratch/metagenomics/batch_corrected/NS_genus_data_090723.csv") %>%
  dplyr::rename("core_specific_ID_SUFFIX" = "sample_name") %>%
  inner_join(nasal_metagenomics_metadata) %>%
  dplyr::filter(!sample_id %in% exclude_samples) %>%
  dplyr::filter(internal_control != "yes") %>%
  dplyr::filter(!sample_id %in% exclude_samples) %>%
  left_join(clinical_data %>% dplyr::select(sample_id, event_date, participant_id)) %>%
  dplyr::filter(!is.na(event_date)) %>% dplyr::select(-event_date)

nasal_long_species <- read_csv("/scratch/metagenomics/batch_corrected/NS_species_data_090723.csv") %>%
  dplyr::rename("core_specific_ID_SUFFIX" = "sample_name") %>%
  inner_join(nasal_metagenomics_metadata) %>%
  dplyr::filter(internal_control != "yes") %>%
  dplyr::filter(!sample_id %in% exclude_samples) %>%
  left_join(clinical_data %>% dplyr::select(sample_id, event_date, participant_id)) %>%
  dplyr::filter(!is.na(event_date)) %>% dplyr::select(-event_date)

### Load clinical ABX data ###
ABX_days_NS_single_new<-read.csv("/scratch/metagenomics/antibiotic_data/ABX_days_unweighted_v4.csv", header=TRUE)

### Limit to ten days abx and 10 days hospitalization
ABX_days_NS_10abx_new <-ABX_days_NS_single_new %>% 
  dplyr::mutate(keep= case_when(Cefepime>10 ~"no", Vancomycin>10 ~ "no", Ceftazidime>10 ~ "no", Ceftriaxone>10 ~ "no",
    Doxycycline>10 ~ "no", Azithromycin>10 ~ "no", Piperacillin_Tazobactam>10 ~ "no", Metronidazole>10 ~ "no",
    Cefuroxime>10 ~ "no", Meropenem>10 ~ "no", Unclassed>10 ~ "no", Ampiciillin_Sulbactam>10 ~ "no", 
    Ertapenem>10 ~ "no", Levofloxacin>10 ~ "no", Aztreonam>10 ~ "no", Sulfamethoxazole_Trimethoprim>10 ~ "no",
    Linezolid>10 ~ "no", Ampicillin>10 ~ "no", Ciprofloxacin>10 ~ "no", Daptomycin>10 ~ "no", Cefazolin>10 ~ "no", 
    Tobramycin>10 ~ "no", Cefpodoxime>10 ~ "no", Rifaximin>10 ~ "no", Erythromycin>10 ~ "no", Gentamicin>10 ~ "no", 
    Clindamycin>10 ~ "no", .default = "yes")) %>% 
  dplyr::filter(keep=="yes")

ABX_days_NS_GAMM_new<-ABX_days_NS_10abx_new %>% dplyr::filter(sample_id %in% days_hosp$sample_id)

### Load in AMR data ###
### select samples and genes to include ###
AMR_ns_raw<-read_csv("/scratch/metagenomics/AMR_data/NS_AMR_090723.csv")
AMR_ns_trimmed<- AMR_ns_raw %>%
  left_join(nasal_metagenomics_metadata %>% distinct(sample_id, core_specific_ID_SUFFIX), by = c("sample_name" = "core_specific_ID_SUFFIX")) %>%
  dplyr::filter(!is.na(sample_id)) %>%
  dplyr::filter(!sample_id %in% exclude_samples) %>%
  dplyr::filter(!grepl("water|pc-|QC", sample_id)) 

total_samples<-nrow(AMR_ns_trimmed %>% dplyr::select(sample_id) %>% distinct())
  
### prevalence cutoff ###
AMR_ns_prevalent<-AMR_ns_trimmed %>%
  group_by(gene_name) %>%
  dplyr::mutate(gene_count=n()) %>%
  dplyr::mutate(gene_prevalence=(gene_count/total_samples)) %>%
  ungroup() %>%
  dplyr::filter(gene_prevalence>0.05 | dpm>=1 & num_reads>=10)


# trim to only include samples with at least 10 reads #
AMR_NS_10reads<-AMR_ns_prevalent %>%
  group_by(sample_id) %>%
  dplyr::mutate(total_amr_reads=sum(num_reads)) %>%
  dplyr::filter(total_amr_reads>=10) %>%
  ungroup()

### add in drug_class info ###
AMR_NS_withclass <- AMR_NS_10reads %>%
  dplyr::filter(read_coverage_breadth>=5) %>%
  # mechanism categories
  dplyr::mutate(mech_efflux = case_when(grepl("efflux", resistance_mechanism) ~ 1, TRUE ~ 0),
         mech_influx = case_when(grepl("permeability", resistance_mechanism) ~ 1, TRUE ~ 0),
         mech_abxinact = case_when(grepl("inactivation", resistance_mechanism) ~ 1, TRUE ~ 0), 
         mech_targalt = case_when(grepl("target", resistance_mechanism) ~ 1, TRUE ~ 0)) %>%
  dplyr::mutate(mech_num = dplyr::select(., starts_with("mech_")) %>% rowSums()) %>%
  dplyr::mutate(amr_yn = 1,
         #         amr_Acm = case_when(grepl("aminocoumarin", drug_class) ~ 1, TRUE ~ 0),
         amr_AGly = case_when(grepl("aminoglycoside", drug_class) ~ 1, TRUE ~ 0),
         amr_Phe = case_when(grepl("phenicol", drug_class) ~ 1, TRUE ~ 0),
         amr_Bla = case_when(grepl("carbapenem|penem|penam|cephalosporin|cephamycin|monobactam", drug_class) ~ 1, TRUE ~ 0),
         amr_Flq = case_when(grepl("fluoroquinolone", drug_class) ~ 1, TRUE ~ 0),
         amr_Fcyn = case_when(grepl("fosfomycin", drug_class) ~ 1, TRUE ~ 0),
         amr_Glyco = case_when(grepl("glycopeptide", drug_class) ~ 1, TRUE ~ 0), #vancomycin
         #amr_Lca = case_when(grepl("lincosamide", drug_class) ~ 1, TRUE ~ 0), #clindamycin
         amr_Mls = case_when(grepl("macrolide|lincosamide|streptogramin", drug_class) ~ 1, TRUE ~ 0),
         amr_Nfn = case_when(grepl("nitrofuran", drug_class) ~ 1, TRUE ~ 0),
         amr_Oxa = case_when(grepl("oxazolidinone", drug_class) ~ 1, TRUE ~ 0), #linezolid
         amr_Rif = case_when(grepl("rifamycin", drug_class) ~ 1, TRUE ~ 0),
         amr_Sulf = case_when(grepl("sulfonamide", drug_class) ~ 1, TRUE ~ 0),
         #amr_Strep = case_when(grepl("streptogramin", drug_class) ~ 1, TRUE ~ 0),
         #streptogramin
         amr_Mcr = case_when(grepl("mcr", gene_name) ~ 1, TRUE ~ 0),
         amr_Tet = case_when(grepl("glycylcycline|tetracycline", drug_class) ~ 1, TRUE ~ 0))%>%
  dplyr::mutate(amr_class_num = dplyr::select(., starts_with("amr_")) %>% rowSums() -1) %>% #-1 subtracts the 1 being added in from abx_yn
  dplyr::mutate(abx_class=case_when(amr_class_num==1 & amr_AGly==1 ~ "aminoglycoside",
                           amr_class_num==1 & amr_Phe==1 ~ "chloramphenicol",
                           amr_class_num==1 & amr_Bla==1 ~ "beta-lactam",
                           amr_class_num==1 & amr_Flq==1 ~ "fluoroquinolone",
                           amr_class_num==1 & amr_Fcyn==1 ~ "fosfomycin",
                           amr_class_num==1 & amr_Glyco==1 ~ "glycopeptide",
                           #amr_class_num==1 & amr_Lca==1 ~ "MLS", 
                           amr_class_num==1 & amr_Mls==1 ~ "MLS",
                           #amr_class_num==1 & amr_Strep==1 ~ "MLS",
                           amr_class_num==1 & amr_Nfn==1 ~ "nitrofurantoin",
                           amr_class_num==1 & amr_Oxa==1 ~ "oxazolidinone",
                           amr_class_num==1 & amr_Rif==1 ~ "rifampin",
                           amr_class_num==1 & amr_Sulf==1 ~ "sulfonamide",
                           amr_class_num==1 & amr_Tet==1 ~ "tetracycline",
                           amr_class_num==1 & amr_Mcr==1 ~ "colistin",
                           amr_class_num==0 ~ "Other",
                           amr_class_num>1 & mech_efflux==1 ~ "multi-drug efflux pump",
                           drug_class %in% c("macrolide antibiotic; lincosamide antibiotic; streptogramin antibiotic","macrolide antibiotic; lincosamide antibiotic; streptogramin antibiotic; streptogramin A antibiotic; streptogramin B antibiotic", "macrolide antibiotic; lincosamide antibiotic", "lincosamide antibiotic; streptogramin antibiotic; oxazolidinone antibiotic; phenicol antibiotic","lincosamide antibiotic; streptogramin antibiotic; streptogramin A antibiotic; streptogramin B antibiotic; oxazolidinone antibiotic; phenicol antibiotic; pleuromutilin antibiotic") ~ "MLS",
                           drug_class=="fluoroquinolone antibiotic; aminoglycoside antibiotic" ~ "fluoroquinolone/aminoglycoside",
                           gene_name=="POXTA" ~ "phenicol-oxazolidinone-tetracycline"))

 

AMR_NS_withclass2 <- AMR_NS_withclass %>%
  dplyr::mutate(amr_Tet = case_when(amr_Tet==1 & amr_class_num>1 ~ 2, TRUE ~ amr_Tet)) %>% #abx_tet=0 when no tet gene, =1 when only tet resistance, =2 when carried with other abx resistance
  dplyr::mutate(Tet_class = case_when(amr_Tet==0 ~ "non-Tet", amr_Tet==1 ~ "Tet only", amr_Tet==2 ~ "Tet+")) %>%
  rename(depth = read_coverage_depth,amr_total_reads = total_reads) %>%
  group_by(sample_id,abx_class) %>%
  dplyr::mutate(total_class_reads=sum(num_reads)) %>%
  ungroup() %>%
  dplyr::mutate(class_perc_of_resistome=(total_class_reads/total_amr_reads)*100) %>%
  dplyr::mutate(gene_perc_of_resistome=(num_reads/total_amr_reads)*100) 

# clean names, calc sum dpm/rpm, gene richness -> finalize single timepoint
AMR_NS_single <-AMR_NS_withclass2 %>% 
  dplyr::mutate(gene_name2 = case_when(gene_name=="23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)" ~ "erma", gene_name=="bifidobacterium adolescentis rpob mutants conferring resistance to rifampicin" ~ "rpob", gene_name =="staphylococcus aureus mupa conferring resistance to mupirocin" ~ "mupa", TRUE ~ gene_name)) %>%
  group_by(sample_name)%>%
  dplyr::mutate(sum_rpm = sum(rpm), sum_dpm=sum(dpm), gene_richness=n())%>%
  ungroup() 

### Trim to 5 days or less of Azithromycin 
ABX_NS_azithro5<-ABX_days_NS_GAMM_new %>% filter(Azithromycin<=5) %>% dplyr::select(sample_id)

### For GAMM data, limit to ten days abx and 10 days hosp ###
AMR_NS_GAMM <-AMR_NS_single %>% 
    dplyr::filter(sample_id %in% days_hosp$sample_id) %>%
    dplyr::filter(sample_id %in% ABX_days_NS_10abx_new$sample_id) %>%
    dplyr::filter(sample_id %in% ABX_NS_azithro5$sample_id)
```

NOTE: Group and time point data is stored on the R studio server (rstudio.impaccstudy.org) at /scratch/metagenomics/antibiotic_data/.

```{r make group and timepoint data}
#| echo: false
#| warning: false

ABX_days_NS_ST_new <-ABX_days_NS_single_new  #dplyr::mutate(ABX_days=rowSums(across(where(is.numeric))))

ABX_colors<-c("#4268C6", "#E07E3D", "#399960")

NS_group_timepoint_data1_new <-AMR_NS_single %>%
  dplyr::select(sample_id) %>% 
  distinct() %>%
  left_join(clinical_data %>% dplyr::select(sample_id, participant_id, event_date)) %>%
  dplyr::filter(!is.na(participant_id)) %>% # remove samples wtith no participant info
  left_join(ABX_days_NS_ST_new %>% dplyr::select(sample_id, Azithromycin, ABX_days)) %>%
  dplyr::filter(!is.na(ABX_days)) %>% # remove samples with no ABX info
  dplyr::mutate(group_id = ifelse(Azithromycin > 0, "Azithromycin",
                           ifelse(ABX_days == 0, "No ABX", "Other ABX"))) %>%
  dplyr::mutate(timepoint = ifelse((group_id == "Azithromycin" & Azithromycin<=2)|
                              (group_id == "Other ABX" & ABX_days<=2) |
                              (group_id=="No ABX" & event_date<=40 & row_number()%%2==1), "Day 0-2",
                            ifelse((group_id == "Azithromycin" & Azithromycin>=4  & Azithromycin<=6)|
                                     (group_id == "Other ABX" & ABX_days<=6 & ABX_days>=4)|
                                     (group_id=="No ABX" & event_date>=3 & event_date<=40 & row_number()%%2==0), "Day 4-6", "Other"))) %>%
  dplyr::filter(timepoint!="Other")

### Rarefy No ABX samples 50% ###
trim_new<-NS_group_timepoint_data1_new %>% 
  dplyr::filter(group_id=="No ABX") %>% 
  arrange(timepoint) %>%
  dplyr::mutate(keep=case_when(row_number()%%2==1 ~ "yes", row_number()%%2==0 ~ "no"))

NS_group_timepoint_data_new<-NS_group_timepoint_data1_new %>%
  left_join(trim_new)%>%
  dplyr::filter(keep=="yes" | is.na(keep)) %>%
  dplyr::select(sample_id, group_id, timepoint) %>%
  distinct()
  

### Identify participants in azithro, other, no abx groups
groupdata_cole<-read.csv("/scratch/metagenomics/antibiotic_data/timepoint_group_assignment_2025_03_05.csv", header=TRUE)
group_assign<-groupdata_cole %>% select(participant_id,group_id)
agroup<-group_assign %>% filter(group_id=="Azithromycin") %>% select(participant_id) %>% distinct()
ogroup<-group_assign %>% filter(group_id=="Other ABX") %>% select(participant_id) %>% distinct()
ngroup<-group_assign %>% filter(group_id=="No ABX") %>% select(participant_id) %>% distinct()

### assign AMR samples to group

AMR_newgroups <-AMR_NS_single %>%
  dplyr::select(sample_id) %>% 
  distinct() %>%
  left_join(clinical_data %>% dplyr::select(sample_id, participant_id, event_date)) %>%
  dplyr::filter(!is.na(participant_id)) %>% # remove samples with no participant info
  left_join(ABX_days_NS_ST_new) %>%
  dplyr::filter(!is.na(ABX_days)) %>% # remove samples with no ABX info
  dplyr::mutate(group_id = ifelse(participant_id %in% agroup$participant_id, "Azithromycin",
                           ifelse(participant_id %in% ogroup$participant_id, "Other ABX", "No ABX"))) %>%
  dplyr::mutate(timepoint = ifelse((group_id == "Azithromycin" & Azithromycin<=2)|
                              (group_id == "Other ABX" & ABX_days<=2) |
                              (group_id=="No ABX" & event_date<=40 & row_number()%%2==1), "Day 0-2",
                            ifelse((group_id == "Azithromycin" & Azithromycin>=4  & Azithromycin<=6)|
                                     (group_id == "Other ABX" & ABX_days<=6 & ABX_days>=4)|
                                     (group_id=="No ABX" & event_date>=3 & event_date<=40 & row_number()%%2==0), "Day 4-6", "Other"))) %>%
  dplyr::filter(timepoint!="Other")

### Rarefy No ABX samples 50% ###
trimmed_AMR_newgroups<-AMR_newgroups %>% 
  dplyr::filter(group_id=="No ABX") %>% 
  arrange(timepoint) %>%
  dplyr::mutate(keep=case_when(row_number()%%2==1 ~ "yes", row_number()%%2==0 ~ "no"))

NS_group_timepoint_data_0306<-AMR_newgroups %>%
  left_join(trimmed_AMR_newgroups)%>%
  dplyr::filter(keep=="yes" | is.na(keep)) %>%
  dplyr::select(sample_id, group_id, timepoint) %>%
  distinct()

startstop<-read.csv("/scratch/metagenomics/antibiotic_data/ABX_cleaned_start_stop.csv", header=TRUE)

write.csv(NS_group_timepoint_data_0306,"final_group_assign_resubmit_033025.csv")

```

NOTE: Group and time point data is stored on the R studio server (rstudio.impaccstudy.org) at /scratch/metagenomics/antibiotic_data/.

### Supplement: AMR Negative controls

```{r}
AMR_ns_neg<- AMR_ns_raw %>%
  left_join(nasal_metagenomics_metadata %>% distinct(sample_id, core_specific_ID_SUFFIX), by = c("sample_name" = "core_specific_ID_SUFFIX")) %>%
  dplyr::filter(!is.na(sample_id)) %>%
  dplyr::filter(!sample_id %in% exclude_samples) %>%
  dplyr::filter(grepl("pc-", sample_id)) %>%
  pivot_wider(names_from = sample_id,values_from = num_reads,values_fill = 0,id_cols="gene_name")

amr_neg_trimmed<-AMR_ns_neg[,-1]
rownames(amr_neg_trimmed)<-AMR_ns_neg$gene_name
colnames(amr_neg_trimmed) <- paste0("Negative_control", 1:22)
write.csv(amr_neg_trimmed, "Supplement_IMPACC_NS_neg_control_AMR.csv", row.names = T)
```

### 3A - Alpha Diversity

```{r, alpha_div}
new <- as_labeller(c("Day 0-2" = "1 \u00b1 1 Days", "Day 4-6" = "5 \u00b1 1 Days"))

library(vegan)
AMR_prep_new<-AMR_NS_single %>%
  pivot_wider(names_from = gene_name, values_from = dpm, values_fill = 0, id_cols = sample_id) %>%
  as.data.frame()
rownames(AMR_prep_new)<-AMR_prep_new$sample_id
input_new<-AMR_prep_new[,-1]

shannon_new<-diversity(input_new,index="shannon") %>% as.data.frame()
shan_new<-shannon_new %>% dplyr::mutate(sample_id= rownames(shannon_new)) %>% set_names("shannon", "sample_id") %>% distinct()

adiv_data_new <- AMR_NS_single %>%
  left_join(shan_new, by="sample_id") %>%
  dplyr::select(sample_id, shannon) %>%
  distinct() %>%
  left_join(clinical_data %>% dplyr::select(sample_id, participant_id, event_date, ever_steroids, trajectory_group, discretized_admit_age_quantile, sex, enrollment_site)) %>%
  dplyr::filter(!is.na(participant_id)) %>%
  left_join(ABX_days_NS_ST_new %>% dplyr::select(sample_id, Azithromycin, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, ABX_days)) %>%
  dplyr::filter(!is.na(ABX_days)) %>%
  dplyr::filter(sample_id %in% NS_group_timepoint_data_0306$sample_id) %>%
  left_join(NS_group_timepoint_data_0306 %>% dplyr::select(sample_id, group_id, timepoint)) %>%
  dplyr::filter(!is.na(group_id), !is.na(timepoint)) %>%
  dplyr::mutate(Azith_other_one_zero = ifelse(group_id == "Azithromycin", 1,
                                       ifelse(group_id == "Other ABX", 0, NA))) %>%
  dplyr::select(sample_id,group_id,timepoint,shannon, event_date, participant_id, ever_steroids, trajectory_group, discretized_admit_age_quantile, sex, Azithromycin, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, ABX_days,Azith_other_one_zero, enrollment_site) %>%
  distinct()

stats_df_adiv_new <- adiv_data_new %>%
  group_by(timepoint) %>%
  wilcox_test(shannon ~ group_id) %>%
  dplyr::mutate(y.position=c(2.7,3.2,3.7,2.7,3.2,3.7)) %>%
  dplyr::mutate(p.adj = p.adjust(p = p, method = "fdr")) %>%
  add_significance(p.col = "p.adj") %>%
  dplyr::mutate(p_char = case_when(p.adj<0.01 ~ format(p.adj, scientific=TRUE, digits=2), p.adj>0.045 & p.adj<0.05 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), p.adj>0.0095 & p.adj<0.01 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p.adj,digits=2), nsmall = 2))))
  
sample_sizes_new <-adiv_data_new %>%
  group_by(group_id,timepoint) %>%
  summarise(n = n(), .groups = "drop")

### linear v1 ###
out_df <- data.frame()
for(comparison in list(c("Azithromycin", "No ABX"), c("Azithromycin", "Other ABX"), c("No ABX", "Other ABX"))){
  for(act_timepoint in unique(adiv_data_new$timepoint)){
    data<-adiv_data_new %>% 
      ungroup() %>%
      filter(timepoint == act_timepoint) %>%
      filter(group_id %in% comparison)
    fit <- lme4::lmer(data=data, formula = shannon ~ group_id + event_date + ever_steroids + trajectory_group + discretized_admit_age_quantile + sex + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|participant_id))
    fit2 <- lme4::lmer(data=data, formula = shannon ~ event_date + ever_steroids + trajectory_group + discretized_admit_age_quantile + sex + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|participant_id))
    an <- anova(fit, fit2)
    out_df <- rbind.data.frame(out_df, data.frame(name = "shannon", timepoint = act_timepoint, group1 = comparison[1],group2 = comparison[2],p = an$`Pr(>Chisq)`[2]))
    }}
out_df <- out_df %>% 
  group_by(name, timepoint) %>% 
  mutate(p.adj = p.adjust(p, "fdr"))
stats_df_adiv_new2 <- stats_df_adiv_new %>% 
  select(-p.adj, -p) %>% 
  left_join(out_df) %>%
  dplyr::mutate(p_char = case_when(p.adj<0.01 ~ format(p.adj, scientific=TRUE, digits=2), p.adj>0.045 & p.adj<0.05 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), p.adj>0.0095 & p.adj<0.01 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p.adj,digits=2), nsmall = 2))))


### linear v2 ###

out_df <- data.frame()
for(comparison in list(c("No ABX", "Azithromycin"),
                       c("Azithromycin", "Other ABX"),
                       c("No ABX", "Other ABX"))){
  for(act_timepoint in unique(adiv_data_new$timepoint)){
      if(all(comparison == c("No ABX", "Azithromycin"))){
        formula_fit1 <- formula("shannon ~ group_id + sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|enrollment_site/participant_id)")
        formula_fit2 <- formula("shannon ~ sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|enrollment_site/participant_id)")
      } else if(all(comparison == c("No ABX", "Other ABX"))){
        formula_fit1 <- formula("shannon ~ group_id + sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + (1|enrollment_site/participant_id)")
        formula_fit2 <- formula("shannon ~ sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + (1|enrollment_site/participant_id)")
      } else if(all(comparison ==  c("Azithromycin", "Other ABX"))){
        ## This is a model that only regresses the other six most common antibiotics for the Azithro group and not the Other ABX group (because these are the other antibiotics)
        formula_fit1 <- formula("shannon ~ group_id + sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + Azith_other_one_zero:Ceftriaxone + Azith_other_one_zero:Vancomycin + Azith_other_one_zero:Cefepime + Azith_other_one_zero:Piperacillin_Tazobactam + Azith_other_one_zero:Doxycycline + Azith_other_one_zero:Meropenem + (1|enrollment_site/participant_id)")
        formula_fit2 <- formula("shannon ~ sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + Azith_other_one_zero:Ceftriaxone + Azith_other_one_zero:Vancomycin + Azith_other_one_zero:Cefepime + Azith_other_one_zero:Piperacillin_Tazobactam + Azith_other_one_zero:Doxycycline + Azith_other_one_zero:Meropenem + (1|enrollment_site/participant_id)")
      }
      
      fit <- lme4::lmer(formula = formula_fit1,
       data = adiv_data_new %>% 
         ungroup() %>%
         #filter(name == shannon) %>%
         filter(timepoint==act_timepoint) %>%
         filter(group_id %in% comparison))
      
      fit2 <- lme4::lmer(formula = formula_fit2,
           data = adiv_data_new %>% 
             ungroup() %>%
             #filter(name == shannon) %>%
             filter(timepoint == act_timepoint) %>%
             filter(group_id %in% comparison))
      
      an <- anova(fit, fit2)
      out_df <- rbind.data.frame(out_df,
                                 data.frame(timepoint = act_timepoint,
                                            group1 = comparison[1],
                                            group2 = comparison[2],
                                            p = an$`Pr(>Chisq)`[2]))
  }
  }

out_df <- out_df %>%
  group_by(timepoint) %>%
  mutate(p.adj = p.adjust(p, "fdr"))%>%
  ungroup()
write.csv(out_df, paste0("Single_timepoint_model_results_adiv.csv"))

stats_df_adiv_new2 <- out_df %>%
  dplyr::mutate(p_char = case_when(p.adj<0.01 ~ format(p.adj, scientific=TRUE, digits=2), p.adj>0.045 & p.adj<0.05 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), p.adj>0.0095 & p.adj<0.01 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p.adj,digits=2), nsmall = 2)))) %>%
  dplyr::mutate(y.position=c(2.7,3.2,3.7,2.7,3.2,3.7)) %>% 
  add_significance(p.col = "p.adj") 
  

gg3a <- adiv_data_new%>%
  dplyr::mutate(group_id = factor(group_id, c("No ABX", "Azithromycin", "Other ABX"))) %>%
  ggplot(aes(x=group_id, y=shannon, color = group_id,group = group_id)) +
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(size = 1, alpha=0.5) +
  facet_grid(~timepoint,scales = "free", labeller=new) +
  theme_classic() +
  theme(axis.text.x = element_text(  vjust = 1, hjust=1, size=12),
        axis.text.y=element_text(size=12),
        legend.text=element_text(size=12)) +
  scale_color_manual(values = ABX_colors) +
  stat_pvalue_manual(stats_df_adiv_new2, label="p_char", size=4)+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  ylab("Shannon Diversity Index") + xlab("") + 
  geom_text(data = sample_sizes_new, aes(x = group_id, y =-0.5, label = paste("n =", n)), vjust = 1.5, color = "black")+
  coord_cartesian(ylim=c(-1,3.5)) 

pdf(file = "3A.pdf", width=5, height=2.5)
adiv_data_new%>%
  dplyr::mutate(group_id = factor(group_id, c("No ABX", "Azithromycin", "Other ABX"))) %>%
  ggplot(aes(x=group_id, y=shannon, color = group_id,group = group_id)) +
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(size = 1, alpha=0.5) +
  facet_grid(~timepoint,scales = "free", labeller=new) +
  theme_classic() +
  theme(axis.text.x = element_text(  vjust = 1, hjust=1, size=12),
        axis.text.y=element_text(size=12),
        legend.text=element_text(size=12)) +
  scale_color_manual(values = ABX_colors) +
  stat_pvalue_manual(stats_df_adiv_new2, label="p_char", size=4)+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  ylab("Shannon Diversity Index") + xlab("") + 
  geom_text(data = sample_sizes_new, aes(x = group_id, y =-0.5, label = paste("n =", n)), vjust = 1.5, color = "black")+
  coord_cartesian(ylim=c(-1,3.5))
invisible(capture.output(dev.off()))

mydata <-adiv_data_new %>% 
  select(group_id, timepoint, shannon)

write.csv(mydata, "3A.csv")

write.csv(mydata %>%
  group_by(group_id, timepoint) %>%
  dplyr::summarize(mean = mean(shannon),
                   median = median(shannon),
                   std_dev = sd(shannon),
                   count = n(),
                   se = std_dev / sqrt(n()),
                   CI_lower = mean - (1.96 * se),
                   CI_upper = mean + (1.96 * se)), "3A_effect_sizes.csv")
```

### 3C - MLS ARG Richness

```{r}
gr_data_new <-AMR_NS_single %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(mls_gene_count=n()) %>%
  ungroup() %>%
  dplyr::filter(abx_class=="MLS") %>%
  dplyr::select(sample_id,mls_gene_count) %>%
  distinct()

gr_data2_new <- AMR_NS_single %>%
  left_join(gr_data_new) %>%
  dplyr::mutate(mls_gene_count = ifelse(is.na(mls_gene_count),0,mls_gene_count))%>%
  distinct(sample_id, mls_gene_count) %>%
  left_join(clinical_data %>% dplyr::select(sample_id, participant_id, event_date, ever_steroids,trajectory_group, discretized_admit_age_quantile, sex, enrollment_site)) %>%
  dplyr::filter(!is.na(participant_id)) %>%
  left_join(ABX_days_NS_ST_new %>% dplyr::select(sample_id, Azithromycin, ABX_days, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem)) %>%
  dplyr::filter(!is.na(ABX_days)) %>%
  dplyr::filter(sample_id %in% NS_group_timepoint_data_0306$sample_id) %>%
  left_join(NS_group_timepoint_data_0306 %>% dplyr::select(sample_id,group_id,timepoint)) %>%
  dplyr::filter(!is.na(timepoint), !is.na(group_id)) %>%
    dplyr::mutate(Azith_other_one_zero = ifelse(group_id == "Azithromycin", 1,
                                       ifelse(group_id == "Other ABX", 0, NA))) %>%
  dplyr::select(sample_id, group_id, timepoint, mls_gene_count, event_date, participant_id, ever_steroids,trajectory_group, discretized_admit_age_quantile, sex, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, ABX_days, enrollment_site, Azith_other_one_zero) %>%
  distinct()

sample_sizes_new <- gr_data2_new %>%
  dplyr::select(sample_id, group_id, timepoint) %>%
  distinct() %>%
  group_by(group_id,timepoint) %>%
  summarise(n = n(), .groups = "drop")

stats_df_mls_gene_new <- gr_data2_new %>%
  group_by(timepoint) %>%
  wilcox_test(mls_gene_count ~ group_id) %>%
  dplyr::mutate(p.adj = p.adjust(p = p, method = "fdr"), y.position=c(8,9,10,8,9,10)) %>%
  add_significance(p.col = "p.adj") %>%
  dplyr::mutate(p_char = case_when(p.adj<0.01 ~ format(p.adj, scientific=TRUE, digits=2), p.adj>0.045 & p.adj<0.05 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), p.adj>0.0095 & p.adj<0.01 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p.adj,digits=2), nsmall = 2))))

### linear ###
out_df <- data.frame()
for(comparison in list(c("Azithromycin", "No ABX"), c("Azithromycin", "Other ABX"), c("No ABX", "Other ABX"))){
  for(act_timepoint in unique(gr_data2_new$timepoint)){
    data<-gr_data2_new %>% ungroup() %>%
      filter(timepoint == act_timepoint) %>%
      filter(group_id %in% comparison)
    fit <- lme4::lmer(data=data, formula = mls_gene_count ~ group_id + event_date + ever_steroids + trajectory_group + discretized_admit_age_quantile + sex + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|participant_id))
    fit2 <- lme4::lmer(data=data, formula = mls_gene_count ~ event_date + ever_steroids + trajectory_group + discretized_admit_age_quantile + sex + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|participant_id))
    an <- anova(fit, fit2)
    out_df <- rbind.data.frame(out_df, data.frame(name = "mls_gene_count", timepoint = act_timepoint,
                                          group1 = comparison[1],group2 = comparison[2],
                                          p = an$`Pr(>Chisq)`[2]))}}
out_df <- out_df %>% group_by(name, timepoint) %>% mutate(p.adj = p.adjust(p, "fdr"))
stats_df_mls_gene_new2 <- stats_df_mls_gene_new %>% select(-p.adj, -p) %>% left_join(out_df) %>%
  dplyr::mutate(p_char = case_when(p.adj<0.01 ~ format(p.adj, scientific=TRUE, digits=2), p.adj>0.045 & p.adj<0.05 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), p.adj>0.0095 & p.adj<0.01 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p.adj,digits=2), nsmall = 2))))


### linear model v2 ###
out_df <- data.frame()
for(comparison in list(c("No ABX", "Azithromycin"),
                       c("Azithromycin", "Other ABX"),
                       c("No ABX", "Other ABX"))){
  for(act_timepoint in unique(gr_data2_new$timepoint)){
      if(all(comparison == c("No ABX", "Azithromycin"))){
        formula_fit1 <- formula("mls_gene_count ~ group_id + sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|enrollment_site/participant_id)")
        formula_fit2 <- formula("mls_gene_count ~ sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|enrollment_site/participant_id)")
      } else if(all(comparison == c("No ABX", "Other ABX"))){
        formula_fit1 <- formula("mls_gene_count ~ group_id + sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + (1|enrollment_site/participant_id)")
        formula_fit2 <- formula("mls_gene_count ~ sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + (1|enrollment_site/participant_id)")
      } else if(all(comparison ==  c("Azithromycin", "Other ABX"))){
        ## This is a model that only regresses the other six most common antibiotics for the Azithro group and not the Other ABX group (because these are the other antibiotics)
        formula_fit1 <- formula("mls_gene_count ~ group_id + sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + Azith_other_one_zero:Ceftriaxone + Azith_other_one_zero:Vancomycin + Azith_other_one_zero:Cefepime + Azith_other_one_zero:Piperacillin_Tazobactam + Azith_other_one_zero:Doxycycline + Azith_other_one_zero:Meropenem + (1|enrollment_site/participant_id)")
        formula_fit2 <- formula("mls_gene_count ~ sex + discretized_admit_age_quantile + event_date + ever_steroids + trajectory_group + Azith_other_one_zero:Ceftriaxone + Azith_other_one_zero:Vancomycin + Azith_other_one_zero:Cefepime + Azith_other_one_zero:Piperacillin_Tazobactam + Azith_other_one_zero:Doxycycline + Azith_other_one_zero:Meropenem + (1|enrollment_site/participant_id)")
      }
      
      fit <- lme4::lmer(formula = formula_fit1,
       data = gr_data2_new %>% 
         ungroup() %>%
         filter(timepoint==act_timepoint) %>%
         filter(group_id %in% comparison))
      
      fit2 <- lme4::lmer(formula = formula_fit2,
           data = gr_data2_new %>% 
             ungroup() %>%
             filter(timepoint == act_timepoint) %>%
             filter(group_id %in% comparison))
      
      an <- anova(fit, fit2)
      out_df <- rbind.data.frame(out_df,
                                 data.frame(timepoint = act_timepoint,
                                            group1 = comparison[1],
                                            group2 = comparison[2],
                                            p = an$`Pr(>Chisq)`[2]))
  }
  }

out_df <- out_df %>%
  group_by(timepoint) %>%
  mutate(p.adj = p.adjust(p, "fdr"))%>%
  ungroup()
write.csv(out_df, paste0("Single_timepoint_model_results_richness.csv"))

my_stats <- out_df %>%
  dplyr::mutate(p_char = case_when(p.adj<0.01 ~ format(p.adj, scientific=TRUE, digits=2), p.adj>0.045 & p.adj<0.05 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), p.adj>0.0095 & p.adj<0.01 ~ as.character(format(round(p.adj,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p.adj,digits=2), nsmall = 2)))) %>%
  dplyr::mutate(y.position=c(8,9,10,8,9,10)) %>% 
  add_significance(p.col = "p.adj") 

gr_data2_new %>%
  dplyr::mutate(group_id = factor(group_id, c("No ABX", "Azithromycin", "Other ABX"))) %>%
  ggplot(aes(x=group_id, y=mls_gene_count, color = group_id, group = group_id)) +
  geom_violin(trim=FALSE,alpha=0.5) +
  geom_boxplot(outlier.shape = NA, width=0.1) +
  geom_jitter(size=1, alpha=0.2, width=0.4) +
  facet_grid(~timepoint,scales = "free", labeller=new) +
  theme_classic() +
  theme(axis.text.x = element_text(  vjust = 1, hjust=1, size=12), 
        axis.text.y = element_text(size=12), 
        legend.text = element_text(size=12)) +
  stat_pvalue_manual(my_stats, label="p_char", size=4) +
  scale_color_manual(values = ABX_colors) +
  ylab("Number of Detectably Expressed MLS ARGs") + xlab("") + 
  geom_text(data = sample_sizes_new, aes(x = group_id, y = -0.5, label = paste("n =", n)), vjust = 1.5, color = "black") + coord_cartesian(ylim=c(-1,10.5))

pdf(file = "3C.pdf", width=5, height=2.5)
gr_data2_new %>%
  dplyr::mutate(group_id = factor(group_id, c("No ABX", "Azithromycin", "Other ABX"))) %>%
  ggplot(aes(x=group_id, y=mls_gene_count, color = group_id, group = group_id)) +
  geom_violin(trim=FALSE,alpha=0.5) +
  geom_boxplot(outlier.shape = NA, width=0.1) +
  geom_jitter(size=1, alpha=0.2, width=0.4) +
  facet_grid(~timepoint,scales = "free", labeller=new) +
  theme_classic() +
  theme(axis.text.x = element_text(  vjust = 1, hjust=1, size=12), 
        axis.text.y = element_text(size=12), 
        legend.text = element_text(size=12)) +
  stat_pvalue_manual(my_stats, label="p_char", size=4) +
  scale_color_manual(values = ABX_colors) +
  ylab("Number of Detectably Expressed MLS ARGs") + xlab("") + 
  geom_text(data = sample_sizes_new, aes(x = group_id, y = -0.5, label = paste("n =", n)), vjust = 1.5, color = "black") + coord_cartesian(ylim=c(-1,10.5))
invisible(capture.output(dev.off()))

mydata <-gr_data2_new %>% 
  select(group_id, timepoint, mls_gene_count)

write.csv(mydata, "3C.csv")

write.csv(mydata %>%
  group_by(group_id, timepoint) %>%
  dplyr::summarize(mean = mean(mls_gene_count),
                   median = median(mls_gene_count),
                   std_dev = sd(mls_gene_count),
                   count = n(),
                   se = std_dev / sqrt(n()),
                   CI_lower = mean - (1.96 * se),
                   CI_upper = mean + (1.96 * se)), "3C_effect_sizes.csv")
```

### 4A - Bar Chart of individual MLS genes

```{r bar chart}
### ID genes to collapse
AMR_NS_single%>% 
  mutate(total=sum(dpm)) %>% ### all reads in entire cohort
  group_by(gene_name2) %>%
  mutate(sum=sum(dpm)) %>% ### sum reads for each gene
  mutate(perc=sum/total*100) %>%  ### generate percent
  filter(abx_class=="MLS") %>%  ### only look at MLS genes
  dplyr::select(gene_name2, perc) %>%  ###show gene name and prevalence
  distinct()

### trim dataset
AMR_NS_MLS_single<-AMR_NS_single %>%
  group_by(sample_id) %>%
  mutate(all_amr=sum(num_reads)) %>%
  ungroup() %>%
  dplyr::filter(abx_class=="MLS")

### Collapse <1% prevalence
AMR_NS_MLS_single2 <-AMR_NS_MLS_single%>%
  dplyr::mutate(gene_name3=case_when(gene_name2 == "lsac" |
                                       gene_name2 == "vgaalc" |
                                       gene_name2 == "vgaa" |
                                       gene_name2 == "ermf" |
                                       gene_name2=="lsaa" |
                                       gene_name2=="lnuc" | 
                                       gene_name2=="mel" |
                                       gene_name2=="lnua" ~ "Other", TRUE ~ gene_name2)) %>%
  group_by(sample_id, gene_name3) %>%
  dplyr::mutate(samplegenesum=sum(num_reads)) %>%
  dplyr::mutate(gene_perc=(samplegenesum/all_amr)*100) %>%
  ungroup() %>%
  group_by(sample_id,abx_class) %>%
  mutate(class_total_reads=sum(num_reads)) %>%
  mutate(gene_perc_class=(samplegenesum/class_total_reads)*100) %>%
  ungroup() %>%
  dplyr::select(sample_id, gene_name3, abx_class, class_total_reads, gene_perc_class, all_amr, samplegenesum) %>%
  distinct()

### Add timepoints 
data_new <-AMR_NS_MLS_single2 %>% 
  left_join(NS_group_timepoint_data_0306) 


### Barplot of full resistome 
custom_colors <- c("msra" = "#C49A00", "erma" = "#F8766D", "ermc" = "#53B400", "mphc" = "#00C094", "ermx" = "#00B6EB","ermb" = "#A58AFF", "Other" = "#FB61D7")
data_new %>%
  filter(timepoint=="Day 4-6", group_id=="Azithromycin") %>%
  mutate(all_mls=sum(samplegenesum)) %>%
  group_by(gene_name3) %>%
  mutate(gene_total=sum(samplegenesum)) %>%
  mutate(gene_perc_all=(gene_total/all_mls)*100) %>%
  dplyr::select(gene_name3,gene_perc_all) %>%
  distinct() %>%
  mutate(gene_name4=as.factor(gene_name3)) %>%
  ggplot(aes(x=fct_reorder(gene_name4, gene_perc_all,.fun = sum, .desc=TRUE), y=gene_perc_all, fill=gene_name4)) + geom_bar(stat="identity") +
  labs(color="ARG Name") +
  scale_fill_manual(values=custom_colors) +
  theme_classic() +  
  labs( x = "ARG", y = "Percent of MLS resistome") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),
        axis.text.y = element_text(size=12), legend.position="none")

pdf(file = "4A.pdf", width=5, height=2.5)
data_new %>%
  filter(timepoint=="Day 4-6", group_id=="Azithromycin") %>%
  mutate(all_mls=sum(samplegenesum)) %>%
  group_by(gene_name3) %>%
  mutate(gene_total=sum(samplegenesum)) %>%
  mutate(gene_perc_all=(gene_total/all_mls)*100) %>%
  dplyr::select(gene_name3,gene_perc_all) %>%
  distinct() %>%
  mutate(gene_name4=as.factor(gene_name3)) %>%
  ggplot(aes(x=fct_reorder(gene_name4, gene_perc_all,.fun = sum, .desc=TRUE), y=gene_perc_all, fill=gene_name4)) + geom_bar(stat="identity") +
  labs(color="ARG Name") +
  scale_fill_manual(values=custom_colors) +
  theme_classic() +  
  labs( x = "ARG", y = "Percent of MLS resistome") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size=12),
        axis.text.y = element_text(size=12), legend.position="none")
invisible(capture.output(dev.off()))

my_data<-data_new %>%
  filter(timepoint=="Day 4-6", group_id=="Azithromycin") %>%
  mutate(all_mls=sum(samplegenesum)) %>%
  group_by(gene_name3) %>%
  mutate(gene_total=sum(samplegenesum)) %>%
  mutate(gene_perc_all=(gene_total/all_mls)*100) %>%
  dplyr::select(gene_name3,gene_perc_all) %>%
  distinct() %>%
  mutate(gene_name4=as.factor(gene_name3)) %>% 
  ungroup() %>%
  select(gene_name4,gene_perc_all) %>% 
  distinct()

write.csv(my_data, "4A.csv")

```

### Figure 3F, S5 - Post-Azithro Analysis

```{r postazithro}
stop_dates<-startstop %>% filter(med_group=="Azithromycin")

post_data_new<-AMR_NS_single %>%
  left_join(NS_group_timepoint_data1_new) %>%
  left_join(clinical_data) %>%
  left_join(stop_dates)

### 37 participants
mid_participants_new<-post_data_new %>%
  filter(group_id=="Azithromycin", timepoint=="Day 4-6") %>%
  dplyr::select(participant_id) %>%
  distinct()

### 62 samples
mid_samples_new<-post_data_new %>%
  filter(group_id=="Azithromycin", timepoint=="Day 4-6") %>%
  dplyr::select(sample_id) %>%
  distinct()

### full 4-6 data
mid_data_new<-post_data_new %>%
  filter(group_id=="Azithromycin", timepoint=="Day 4-6")

### All samples post-azithro - 42 samples
all_late_new<-post_data_new %>%
  filter(participant_id %in% mid_participants_new$participant_id) %>% 
  left_join(ABX_days_NS_ST_new %>% dplyr::select(sample_id, Azithromycin, ABX_days, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem)) %>%
  dplyr::select(participant_id, sample_id,event_date, Azithromycin, admit_month, event_type, conmed_dose_stop, ever_steroids, trajectory_group, discretized_admit_age_quantile, sex, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem)%>%
  distinct() %>%
  mutate(days_since_stop=event_date-conmed_dose_stop) %>%
  filter(days_since_stop>0)

### 12 samples
late_710_new<-all_late_new %>%
  filter(days_since_stop>=7 & days_since_stop<=10)

compare_new<-post_data_new %>% 
  mutate(late_compare_new=case_when(sample_id %in% late_710_new$sample_id ~ "7-10 Days\nAfter Cessation", (group_id=="Azithromycin" & timepoint=="Day 4-6") ~ "5 \u00b1 1 Days\nof Exposure", TRUE ~ "None")) %>%
  filter(late_compare_new!="None") %>%
  left_join(ABX_days_NS_ST_new %>% dplyr::select(sample_id, Azithromycin, ABX_days, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem)) %>%
  dplyr::select(sample_id, late_compare_new, event_date, participant_id, ever_steroids,trajectory_group, discretized_admit_age_quantile, sex, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem) %>%
  distinct()

compare_all_new<-post_data_new %>% 
  mutate(late_compare_new=case_when(sample_id %in% all_late_new$sample_id ~ "After Azithromycin\nCessation", (group_id=="Azithromycin" & timepoint=="Day 4-6") ~ "5 \u00b1 1 Days\nof Exposure", TRUE ~ "None")) %>%
  filter(late_compare_new!="None") %>%
  left_join(ABX_days_NS_ST_new %>% dplyr::select(sample_id, Azithromycin, ABX_days, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem)) %>%
  dplyr::select(sample_id, late_compare_new, event_date, participant_id, ever_steroids,trajectory_group, discretized_admit_age_quantile, sex, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem) %>%
  distinct()

### Figure 3F
gr_data3_new<-gr_data2_new %>%
  filter(sample_id %in% compare_new$sample_id) %>%
  left_join(compare_new) %>%
  dplyr::select(sample_id, late_compare_new, mls_gene_count, event_date, participant_id, ever_steroids,trajectory_group, discretized_admit_age_quantile, sex, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem) %>%
  distinct()

sample_sizes_new <- gr_data3_new %>%
  dplyr::select(sample_id, late_compare_new) %>%
  distinct() %>%
  group_by(late_compare_new) %>%
  summarise(n = n(), .groups = "drop")

stats_df_mls_gene_new <- gr_data3_new %>%
  wilcox_test(mls_gene_count ~ late_compare_new) %>%
  dplyr::mutate(y.position=9) %>%
  add_significance(p.col = "p") %>%
  dplyr::mutate(p_char = case_when(p<0.01 ~ format(p, scientific=TRUE, digits=2), p>0.045 & p<0.05 ~ as.character(format(round(p,digits=3), nsmall = 2)), p>0.0095 & p<0.01 ~ as.character(format(round(p,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p,digits=2), nsmall = 2))))

### linear stats ###
out_df <- data.frame()
fit <- lme4::lmer(data=gr_data3_new, formula = mls_gene_count ~ late_compare_new + event_date + ever_steroids + trajectory_group + discretized_admit_age_quantile + sex + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|participant_id))
fit2 <- lme4::lmer(data=gr_data3_new, formula = mls_gene_count ~ event_date + ever_steroids + trajectory_group + discretized_admit_age_quantile + sex + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|participant_id))
an <- anova(fit, fit2)
out_df <- rbind.data.frame(out_df, data.frame(name = "mls_gene_count", group1 ="5 \u00b1 1 Days\nof Exposure", group2 = "7-10 Days\nAfter Cessation", p = an$`Pr(>Chisq)`[2]))
out_df2<-out_df %>% dplyr::mutate(p_char = case_when(p<0.01 ~ format(p, scientific=TRUE, digits=2), p>0.045 & p<0.05 ~ as.character(format(round(p,digits=3), nsmall = 2)), p>0.0095 & p<0.01 ~ as.character(format(round(p,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p,digits=2), nsmall = 2)))) %>% dplyr::mutate(y.position=9) 
write.csv(out_df2, "3F_stats.csv")
gr_data3_new %>%
  dplyr::mutate(late_compare_new = factor(late_compare_new, c("5 \u00b1 1 Days\nof Exposure", "7-10 Days\nAfter Cessation"))) %>%
  ggplot(aes(x=late_compare_new, y=mls_gene_count, color = late_compare_new, group = late_compare_new)) +
  geom_violin(trim=FALSE,alpha=0.5) +
  geom_boxplot(outlier.shape = NA, width=0.1) +
  geom_jitter(size=1,alpha=0.2, width=0.4) +
  #ggbeeswarm::geom_quasirandom(size = 1, alpha=0.5) +
  theme_classic() +
  labs(color="Timepoint") +
  #scale_y_log10()+
  theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10), 
        legend.text = element_text(size=10)) +
  stat_pvalue_manual(out_df2, label="p_char", size=4) +
  ylab("Number of Detectably Expressed MLS ARGs") + xlab("") + 
  geom_text(data = sample_sizes_new, aes(x = late_compare_new, y = -0.5, label = paste("n =", n)), vjust = 1.5, color = "black") + coord_cartesian(ylim=c(-1,10))

pdf( file = "3F.pdf", width=5, height=2.5)
gr_data3_new %>%
  dplyr::mutate(late_compare_new = factor(late_compare_new, c("5 \u00b1 1 Days\nof Exposure", "7-10 Days\nAfter Cessation"))) %>%
  ggplot(aes(x=late_compare_new, y=mls_gene_count, color = late_compare_new, group = late_compare_new)) +
  geom_violin(trim=FALSE,alpha=0.5) +
  geom_boxplot(outlier.shape = NA, width=0.1) +
  geom_jitter(size=1,alpha=0.2, width=0.4) +
  #ggbeeswarm::geom_quasirandom(size = 1, alpha=0.5) +
  theme_classic() +
  labs(color="Timepoint") +
  #scale_y_log10()+
  theme(axis.text.x = element_text(size=10), 
        axis.text.y = element_text(size=10), 
        legend.text = element_text(size=10)) +
  stat_pvalue_manual(out_df2, label="p_char", size=4) +
  ylab("Number of Detectably Expressed MLS ARGs") + xlab("") + 
  geom_text(data = sample_sizes_new, aes(x = late_compare_new, y = -0.5, label = paste("n =", n)), vjust = 1.5, color = "black") + coord_cartesian(ylim=c(-1,10))
invisible(capture.output(dev.off()))

my_data<-gr_data3_new %>%
  mutate(late_compare_new = factor(late_compare_new, c("5 \u00b1 1 Days\nof Exposure", "7-10 Days\nAfter Cessation"))) %>%
  select(late_compare_new, mls_gene_count)

write.csv(my_data, "3F.csv")

write.csv(my_data %>%
  group_by(late_compare_new) %>%
  dplyr::summarize(mean = mean(mls_gene_count),
                   median = median(mls_gene_count),
                   std_dev = sd(mls_gene_count),
                   count = n(),
                   se = std_dev / sqrt(n()),
                   CI_lower = mean - (1.96 * se),
                   CI_upper = mean + (1.96 * se)), "3F_effect_sizes.csv")

### Figure S5
perc_data_new <- AMR_NS_single %>%
  dplyr::filter(sample_id %in% compare_new$sample_id)%>%
  group_by(sample_id) %>%
  dplyr::mutate(sample_total_reads=sum(num_reads)) %>% ### store total reads in each sample
  ungroup() 

mls_perc_new<-perc_data_new %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(class_reads = sum(num_reads)) %>% ### store total reads in each sample/class 
  dplyr::mutate(class_perc=(class_reads/sample_total_reads)*100) %>% ### convert to percent of resistome
  dplyr::mutate(class_perc = ifelse(is.na(class_perc), 0, class_perc)) %>%
  ungroup() %>%
  filter(abx_class=="MLS") %>%
  dplyr::select(sample_id, class_perc) %>%
  distinct()

perc_data_2_new<-perc_data_new %>%
  left_join(mls_perc_new, by="sample_id") %>%
  dplyr::mutate(class_perc=ifelse(is.na(class_perc), 0, class_perc)) %>% ###for samples with no mls 
  dplyr::select(sample_id,class_perc) %>%
  distinct() %>%
  left_join(clinical_data %>% dplyr::select(sample_id, participant_id, event_date, ever_steroids,trajectory_group, discretized_admit_age_quantile, sex)) %>%
  dplyr::filter(!is.na(participant_id)) %>%
  left_join(ABX_days_NS_ST_new %>% dplyr::select(sample_id, Azithromycin, ABX_days, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem)) %>%
  dplyr::filter(!is.na(ABX_days)) %>%
  left_join(compare_new %>% dplyr::select(sample_id,late_compare_new)) %>%
  dplyr::filter(!is.na(late_compare_new)) %>%
  dplyr::select(sample_id,late_compare_new, class_perc, event_date, participant_id, ever_steroids,trajectory_group, discretized_admit_age_quantile, sex, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem) %>%
  distinct()

sample_sizes_new <- perc_data_2_new %>% 
  group_by(late_compare_new) %>%
  summarise(n = n(), .groups = "drop")

stats_df_perc_new <- perc_data_2_new %>%
  wilcox_test(class_perc ~ late_compare_new) %>%
  dplyr::mutate(y.position=110) %>%
  add_significance(p.col = "p") %>%
  dplyr::mutate(p_char = case_when(p<0.01 ~ format(p, scientific=TRUE, digits=2), p>0.045 & p<0.05 ~ as.character(format(round(p,digits=3), nsmall = 2)), p>0.0095 & p<0.01 ~ as.character(format(round(p,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p,digits=2), nsmall = 2))))

### linear 
out_df <- data.frame()
fit <- lme4::lmer(data=perc_data_2_new, formula = class_perc ~ late_compare_new + event_date + ever_steroids + trajectory_group + discretized_admit_age_quantile + sex + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|participant_id))
fit2 <- lme4::lmer(data=perc_data_2_new, formula = class_perc ~ event_date + ever_steroids + trajectory_group + discretized_admit_age_quantile + sex + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam + Doxycycline + Meropenem + (1|participant_id))
an <- anova(fit, fit2)
out_df <- rbind.data.frame(out_df, data.frame(name = "class_perc", group1 ="5 \u00b1 1 Days\nof Exposure", group2 = "7-10 Days\nAfter Cessation", p = an$`Pr(>Chisq)`[2]))
out_df2<-out_df %>% dplyr::mutate(p_char = case_when(p<0.01 ~ format(p, scientific=TRUE, digits=2), p>0.045 & p<0.05 ~ as.character(format(round(p,digits=3), nsmall = 2)), p>0.0095 & p<0.01 ~ as.character(format(round(p,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(p,digits=2), nsmall = 2)))) %>% dplyr::mutate(y.position=110) 
write.csv(out_df2, "S5_stats.csv")
perc_data_2_new %>%
  dplyr::mutate(late_compare_new = factor(late_compare_new, c("5 \u00b1 1 Days\nof Exposure", "7-10 Days\nAfter Cessation"))) %>%
  ggplot(aes(x=late_compare_new, y=class_perc, color = late_compare_new, group = late_compare_new)) +
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(size = 1, alpha=0.5) +
  theme_classic() +
  labs(color="Timepoint") +
  theme(axis.text.x = element_text(size=12),
        axis.text.y = element_text(size=12),
        legend.text = element_text(size=12)) +
  stat_pvalue_manual(out_df2, label="p_char", size=4) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  ylab("Percent of Resistome - MLS") + xlab("") + 
  geom_text(data = sample_sizes_new, aes(x = late_compare_new, y =-5, label = paste("n =", n)), vjust = 1.5, color = "black")+
  coord_cartesian(ylim=c(-10,120))

pdf( file = "S5.pdf", width=5, height=2.5)
perc_data_2_new %>%
  dplyr::mutate(late_compare_new = factor(late_compare_new, c("5 \u00b1 1 Days\nof Exposure", "7-10 Days\nAfter Cessation"))) %>%
  ggplot(aes(x=late_compare_new, y=class_perc, color = late_compare_new, group = late_compare_new)) +
  geom_boxplot(outlier.shape = NA) +
  ggbeeswarm::geom_quasirandom(size = 1, alpha=0.5) +
  theme_classic() +
  labs(color="Timepoint") +
  theme(axis.text.x = element_text(size=12),
        axis.text.y = element_text(size=12),
        legend.text = element_text(size=12)) +
  stat_pvalue_manual(out_df2, label="p_char", size=4) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  ylab("Percent of Resistome - MLS") + xlab("") + 
  geom_text(data = sample_sizes_new, aes(x = late_compare_new, y =-5, label = paste("n =", n)), vjust = 1.5, color = "black")+
  coord_cartesian(ylim=c(-10,120))
invisible(capture.output(dev.off()))

my_data<-perc_data_2_new %>%
  dplyr::mutate(late_compare_new = factor(late_compare_new, c("5 \u00b1 1 Days\nof Exposure", "7-10 Days\nAfter Cessation"))) %>%
  select(late_compare_new, class_perc)

write.csv(my_data, "S5.csv")

write.csv(my_data %>%
  group_by(late_compare_new) %>%
  dplyr::summarize(mean = mean(class_perc),
                   median = median(class_perc),
                   std_dev = sd(class_perc),
                   count = n(),
                   se = std_dev / sqrt(n()),
                   CI_lower = mean - (1.96 * se),
                   CI_upper = mean + (1.96 * se)), "S5_effect_sizes.csv")
```

### Figure S5B&C - MLS Richness Overtime

```{r, fig.width = 4, fig.height = 2.5}

## Richness of Genes
all_late_new %>%
  left_join(gr_data_new) %>%
  mutate(mls_gene_count = ifelse(is.na(mls_gene_count), 0, mls_gene_count)) %>%
  ggplot(aes(x=days_since_stop, y= mls_gene_count)) +
  geom_line(aes(group = participant_id, color = participant_id), alpha =0.4) +
  theme_classic() +
  ylab("Number of Detectablyn\nExpressed MLS ARGs") +
  xlab("Days since Cessiation of Azithromycin") +
  theme(legend.position = "none")
ggsave("FigureS5C.pdf", width = 4, height = 2.5)

write.csv(file = "source_data_FigureS5C.csv",
          x = all_late_new %>%
  left_join(gr_data_new) %>%
    mutate(mls_gene_count = ifelse(is.na(mls_gene_count), 0, mls_gene_count)) %>%
  mutate(participant_id = as.numeric(factor(participant_id))) %>%
    distinct(participant_id, mls_gene_count, days_since_stop))

## Percent of Resistome
all_late_new %>%
  left_join(mls_perc_new) %>%
  mutate(class_perc = ifelse(is.na(class_perc), 0, class_perc)) %>%
  ggplot(aes(x=days_since_stop, y= class_perc)) +
  geom_line(aes(group = participant_id, color = participant_id), alpha =0.4) +
  theme_classic() +
  ylab("Percent of Resistome MLS (%)") +
  xlab("Days since Cessiation of Azithromycin") +
  theme(legend.position = "none")
ggsave("FigureS5B.pdf", width = 4, height = 2.5)

write.csv(file = "source_data_FigureS5B.csv",
          x = all_late_new %>%
  left_join(mls_perc_new) %>%
    mutate(class_perc = ifelse(is.na(class_perc), 0, class_perc)) %>%
  mutate(participant_id = as.numeric(factor(participant_id))) %>%
    distinct(participant_id, class_perc, days_since_stop))
```


### Figures 3D-3E

```{r GAMM summary stats new}
#| echo: false
#| warning: false

ABX_days_NS_GAMM_plus_new <-ABX_days_NS_GAMM_new %>% 
  dplyr::select(-keep) #%>% 
  #rowwise() %>% 
  #dplyr::mutate(ABX_days=rowSums(across(where(is.numeric))))

mlsdpmgamm_new <- AMR_NS_GAMM %>%
  left_join(clinical_data) %>%
  left_join(ABX_days_NS_GAMM_plus_new) %>% #add additional info
  left_join(NS_group_timepoint_data_0306) %>%
  group_by(sample_id) %>%
  dplyr::mutate(sum_dpm=sum(dpm)) %>%
  dplyr::mutate(sum_reads=sum(num_reads)) %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(class_dpm=sum(dpm)) %>% #calc sum DPM MLS
  dplyr::mutate(class_reads=sum(num_reads)) %>%
  dplyr::mutate(pseudo=class_dpm+1) %>%
  dplyr::mutate(logmlsdpm=log10(pseudo)) %>%
  ungroup() %>%
  #dplyr::mutate(logdpm=log10(class_dpm+1)) %>% #transform class DPM
  dplyr::mutate(perc=(class_dpm/sum_dpm)*100) %>% #calc perc MLS of resistome
  dplyr::mutate(perc2=case_when(is.na(perc) ~ 0, TRUE ~ perc)) %>% #convert NA to 0
  dplyr::mutate(perc3=(class_reads/sum_reads)*100) %>% #calc perc MLS of resistome
  dplyr::mutate(perc4=case_when(is.na(perc3) ~ 0, TRUE ~ perc3)) %>% #convert NA to 0
  dplyr::mutate(logmlsdpm=case_when(is.na(logmlsdpm) ~ 0, TRUE ~ logmlsdpm)) %>% #convert NA to 0
  dplyr::filter(abx_class=="MLS") %>% #filter to only MLS
  ungroup() %>%
  dplyr::select(sample_id, ABX_days,logmlsdpm, class_dpm, sum_dpm, perc, perc2, perc3, perc4,Azithromycin, event_date, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, trajectory_group, sex, discretized_admit_age_quantile, participant_id, enrollment_site, group_id, ever_steroids) %>% 
  distinct() 


### Figure 3E
noabx_perc_new<-mlsdpmgamm_new %>% filter(group_id=="No ABX")
  #filter(ABX_days==0)
noavg_perc_new<-mean(noabx_perc_new$perc4)

formula_fit <- formula(perc4 ~ s(Azithromycin, bs = 'cr', k= 4) +s(event_date, bs = 'cr') + s(Ceftriaxone, bs = 'cr', k= 4) + s(Vancomycin, bs = 'cr', k= 4) + s(Cefepime, bs = 'cr', k= 4) + s(Piperacillin_Tazobactam, bs = 'cr', k= 4) + s(Doxycycline, bs = 'cr', k= 4) + s(Meropenem, bs = 'cr', k= 4) + trajectory_group + sex + discretized_admit_age_quantile + ever_steroids)

fit <- gamm4::gamm4(data = as.data.frame(mlsdpmgamm_new), formula_fit, random = ~(1|enrollment_site/participant_id))

tmp <- data.frame("feature" = "Azithromycin", "p.value" = anova(fit$gam)[["s.pv"]][1])
#tmp
pred <- ggpredict(fit, c("Azithromycin"), type = "fixed", ci_level = CI_interval)
write.csv(pred, "3E_stats.csv")
p3 <-pred %>%
  ggplot(mapping = aes(x = x, y = predicted, fill=group)) +
  geom_line(mapping = aes(y = predicted, x=x), size = 0.8, color="#E07E3D") +
  geom_line(y=noavg_perc_new, color="#4268C6", linetype=2)+
  geom_ribbon(aes(x=x, ymin = conf.low, ymax = conf.high),alpha = 0.3, colour=NA) +
  theme_classic(base_size=14) +
  scale_color_manual(values=c("#E07E3D")) + scale_fill_manual(values=c("#E07E3D")) +
  ylab("MLS Percent of Resistome") + xlab("Days of Azithromycin Exposure") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 5), ylim=c(0,70)) +
  geom_text(label=paste("p=",case_when(tmp$p.value<0.01 ~ format(tmp$p.value, scientific=TRUE, digits=2),tmp$p.value>0.045 & tmp$p.value<0.05 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)),  tmp$p.value>0.0095 & tmp$p.value<0.01 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(tmp$p.value,digits=2), nsmall = 2)))), x=0.5, y=68, size=5)
plot(p3)

my_data<-mlsdpmgamm_new %>%
  select(Azithromycin,perc4)

write.csv(my_data, "3E.csv")

### Figure 3D

richnessgamm_new <- AMR_NS_GAMM %>%
  left_join(clinical_data) %>%
  left_join(ABX_days_NS_GAMM_plus_new) %>% #add additional data
  left_join(NS_group_timepoint_data_0306) %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(richness=n()) %>% #calc richness by class
  ungroup() %>%
  dplyr::filter(abx_class=="MLS") %>% #grab only MLS richness
  dplyr::select(sample_id, ABX_days,richness,Azithromycin, event_date, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, trajectory_group, sex, discretized_admit_age_quantile, participant_id, enrollment_site, group_id, ever_steroids) %>% 
  distinct()
  
noabx_rich_new<-richnessgamm_new %>% filter(group_id=="No ABX")
  #filter(ABX_days==0)
noavg_rich_new<-mean(noabx_rich_new$richness)

 formula_fit <- formula(richness ~ s(Azithromycin, bs = 'cr', k= 4) +s(event_date, bs = 'cr') + s(Ceftriaxone, bs = 'cr', k= 4) + s(Vancomycin, bs = 'cr', k= 4) + s(Cefepime, bs = 'cr', k= 4) + s(Piperacillin_Tazobactam, bs = 'cr', k= 4) + s(Doxycycline, bs = 'cr', k= 4) + s(Meropenem, bs = 'cr', k= 4) + trajectory_group + sex + discretized_admit_age_quantile + ever_steroids)

fit <- gamm4::gamm4(data = as.data.frame(richnessgamm_new), formula_fit, random = ~(1|enrollment_site/participant_id))

tmp <- data.frame("feature" = "Azithromycin", "p.value" = anova(fit$gam)[["s.pv"]][1])
#tmp
pred <- ggpredict(fit, c("Azithromycin"), type = "fixed", ci_level = CI_interval)
write.csv(pred, "3D_stats.csv")
p4 <-pred %>%
      ggplot(mapping = aes(x = x, y = predicted, fill=group)) +
      geom_line(mapping = aes(y = predicted, x=x), size = 0.8, color="#E07E3D") +
      geom_line(y=noavg_rich_new, color="#4268C6", linetype=2)+
      geom_ribbon(aes(x=x, ymin = conf.low, ymax = conf.high),alpha = 0.3, colour=NA) +
      theme_classic(base_size  = 14) +
      scale_color_manual(values=c("#E07E3D")) +
      scale_fill_manual(values=c("#E07E3D")) +
      ylab("Number of Detectably Expressed MLS ARGs") +
      xlab("Days of Azithromycin Exposure") +
      theme(legend.position = "none") +
      coord_cartesian(xlim = c(0, 5)) +
      geom_text(label=paste("p=",case_when(tmp$p.value<0.01 ~ format(tmp$p.value, scientific=TRUE, digits=2),tmp$p.value>0.045 & tmp$p.value<0.05 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)),  tmp$p.value>0.0095 & tmp$p.value<0.01 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(tmp$p.value,digits=2), nsmall = 2)))), x=0.5, y=3.3, size=5)
plot(p4)

my_data<-richnessgamm_new %>%
  select(Azithromycin, richness)

write.csv(my_data, "3D.csv")

pdf(file="3E.pdf", width=4, height=2.5)
plot(p3)
invisible(capture.output(dev.off()))

pdf(file="3D.pdf", width=4, height=2.5)
plot(p4)
invisible(capture.output(dev.off()))

```

### Figure 3B - Beta Diversity

```{r beta div}
library(vegan)

amr_wide_new <- AMR_NS_single %>%
  left_join(NS_group_timepoint_data_0306) %>%
  dplyr::filter(group_id!="Other ABX", !is.na(group_id)) %>%
  pivot_wider(id_cols = "sample_id", names_from = gene_name, values_from = num_reads) %>%
  column_to_rownames("sample_id") %>%
  mutate_if(is.numeric, ~replace_na(., 0)) 

samples<-row.names(amr_wide_new)

metadata_new<-data.frame(samples) %>% 
  set_names("sample_id") %>%
  left_join(NS_group_timepoint_data_0306) %>%
  dplyr::mutate(group2=case_when(group_id=="Azithromycin" & timepoint=="Day 4-6" ~ "Late Azithromycin", group_id=="Azithromycin" & timepoint=="Day 0-2" ~ "Early Azithromycin", TRUE ~ "No ABX")) %>%
  left_join(clinical_data %>% select(sample_id, participant_id, event_date, discretized_admit_age_quantile, sex, trajectory_group, ever_steroids)) %>%
  left_join(ABX_days_NS_ST_new %>% select(sample_id, Azithromycin, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem)) %>%
  dplyr::select(sample_id, group2, participant_id, event_date, discretized_admit_age_quantile, sex, trajectory_group, ever_steroids,Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem)

dist_bc <- vegan::vegdist(amr_wide_new, method = "bray")
AMR.dispersion<-betadisper(dist_bc, group=metadata_new$group2)
AMR.permutest<-permutest(AMR.dispersion)
#plot(AMR.dispersion, hull=FALSE, ellipse=TRUE)
set.seed(100)
AMR.div<-adonis2(dist_bc ~ group2 + event_date + participant_id + discretized_admit_age_quantile + sex + trajectory_group + ever_steroids + Ceftriaxone + Vancomycin + Cefepime + Piperacillin_Tazobactam +  Doxycycline + Meropenem, data=metadata_new, permutations = 1000, method="bray")
#print(AMR.div)

pcoa_result <- cmdscale(dist_bc, k = 2, eig = TRUE)
#plot(pcoa_result$points, xlab = "PCoA1", ylab = "PCoA2", main = "PCoA Plot")
PC1 <- pcoa_result$points[,1]
PC2 <- pcoa_result$points[,2]
plot<-cbind(metadata_new, PC1, PC2)
cent <- aggregate(cbind(PC1, PC2) ~ group2, data = metadata_new, FUN = mean)
segs<-merge(plot,setNames(cent, c("group2", "oPC1", "oPC2")), by = "group2", sort = FALSE)

PCA_plot<-ggplot(plot %>% dplyr::mutate(group2=factor(group2, levels=c("Early Azithromycin","Late Azithromycin", "No ABX"))), aes(PC1, PC2, color=group2)) +
  geom_point(alpha=0.5) +
  stat_ellipse(geom = "polygon", type="t",size =1, alpha=0.1, aes(fill=group2)) +  ##draws 95% confidence interval ellipses
  #xlim(-1,1) +
  geom_segment(data = segs, mapping = aes(xend = oPC1, yend = oPC2, alpha=0.1)) + # spiders
  geom_point(data = cent, size = 4) +                         # centroids
#  coord_fixed() +                                              # same axis scaling
  scale_color_manual(labels=c("Early Azithromycin", "Late Azithromycin", "No ABX"), values=c("#E07E3D","brown4","#4268C6")) +
  scale_fill_manual(labels=c("Early Azithromycin","Late Azithromycin","No ABX"), values=c("#E07E3D", "brown4","#4268C6")) +
  annotate("text", x=0.75, y=0.5, label="Early Azithromycin", hjust = 1, color=c("#E07E3D"), size=5) +
  annotate("text", x=-0.1, y=0.5, label="Late Azithromycin", hjust = 1, color=c("brown4"), size=5) +
  annotate("text", x=0.7, y=-0.4, label="No ABX", hjust = 1, color=c("#4268C6"), size=5) +
  annotate("text", x=0.1, y=-0.5, label=paste("PERMANOVA p=0.001"), hjust = 0, size=5) +
  theme(legend.position="none") + theme_bw()


PCA_plot

pdf(file="3B.pdf", width=5.5, height=2.5)
plot(PCA_plot)
invisible(capture.output(dev.off()))

rownames(amr_wide_new)<-NULL
write.csv(amr_wide_new, "3B.csv")

```

## Figure 4B
```{r corrplot prep 4b}
library(ggcorrplot)
library(corrplot)
### select samples
samples<-AMR_NS_GAMM %>%
  left_join(ABX_days_NS_GAMM_new, by="sample_id") %>%
  dplyr::filter(Azithromycin>0) %>%
  dplyr::select(sample_id) %>% 
  distinct()

### AMR data
AMR_corr_mlsdpm_gene<-AMR_NS_GAMM %>%
  dplyr::filter(sample_id %in% samples$sample_id, abx_class=="MLS") %>%
  dplyr::mutate(pseudo=dpm+1) %>%
  dplyr::mutate(logdpm=log10(pseudo)) %>%
  dplyr::select(sample_id, gene_name2, logdpm) %>%
  distinct() %>%
  pivot_wider(names_from = gene_name2, values_from = logdpm, values_fill = 0) 

AMR_corr_mlsperc_gene<-AMR_NS_GAMM %>%
  dplyr::filter(sample_id %in% samples$sample_id) %>%
  group_by(sample_id)%>%
  dplyr::mutate(sum_dpm = sum(dpm)) %>%
  ungroup%>%
  dplyr::filter(abx_class=="MLS") %>%
  dplyr::mutate(perc=(dpm/sum_dpm)*100) %>%
  dplyr::select(sample_id, gene_name2, perc) %>%
  distinct() %>%
  pivot_wider(names_from = gene_name2, values_from = perc, values_fill = 0) 

AMR_corr_mlsdpm_class<-AMR_NS_GAMM %>%
  dplyr::filter(sample_id %in% samples$sample_id) %>%
  group_by(sample_id, abx_class)%>%
  dplyr::mutate(sum_classdpm = sum(dpm)) %>%
  ungroup() %>%
  dplyr::filter(abx_class=="MLS") %>%
  dplyr::mutate(pseudo=sum_classdpm+1) %>%
  dplyr::mutate(logsumdpm=log10(pseudo)) %>%
  dplyr::select(sample_id, abx_class, logsumdpm) %>% 
  distinct() %>%
  pivot_wider(names_from = abx_class, values_from = logsumdpm, values_fill = 0) 

AMR_corr_mlsperc_class<-AMR_NS_GAMM %>%
  dplyr::filter(sample_id %in% samples$sample_id) %>%
  group_by(sample_id)%>%
  dplyr::mutate(sum_dpm = sum(dpm)) %>%
  ungroup%>%
  group_by(sample_id, abx_class)%>%
  dplyr::mutate(sum_dpm_class= sum(dpm)) %>%
  ungroup() %>%
  dplyr::mutate(classperc=(sum_dpm_class/sum_dpm)*100) %>%
  dplyr::filter(abx_class=="MLS") %>%
  dplyr::select(sample_id, abx_class, classperc) %>%
  distinct() %>%
  pivot_wider(names_from = abx_class, values_from = classperc, values_fill = 0) 


### Bacterial Data
Bact_corr<-nasal_long %>%
  dplyr::filter(sample_id %in% samples$sample_id, 
         category=="bacteria", 
         nt_alignment_length>=70, 
         nt_count>=10, 
         nr_count>=1, 
         p_adj<0.1) %>%
  dplyr::select(sample_id, name, nt_rpm) %>%
  pivot_wider(names_from = name, values_from = nt_rpm, values_fill=0) 

microbes <-Bact_corr[c(2, 3:length(Bact_corr))]
tdf<-t(microbes)
data<-as.data.frame(tdf)
with_total<-data %>% dplyr::mutate(sum = rowSums(.)) %>% arrange(desc(sum)) %>% do(head(., n=30))
trimmed_list_microbes<-rownames(with_total)
top30_data<-Bact_corr %>% dplyr::select(sample_id, trimmed_list_microbes)
top30_data[sapply(top30_data, is.numeric)] <- top30_data[sapply(top30_data, is.numeric)] + 1
top30_data[sapply(top30_data, is.numeric)] <-sapply(top30_data[sapply(top30_data, is.numeric)], log10)

combined_corr_dpmclass<-left_join(AMR_corr_mlsdpm_class, top30_data, by="sample_id") %>%
  replace(is.na(.), 0)

combined_corr_percclass<-left_join(AMR_corr_mlsperc_class, top30_data, by="sample_id") %>%
  replace(is.na(.), 0)

combined_corr_dpmgene<-left_join(AMR_corr_mlsdpm_gene, top30_data, by="sample_id") %>%
  replace(is.na(.), 0)

combined_corr_percgene<-left_join(AMR_corr_mlsperc_gene, top30_data, by="sample_id") %>%
  replace(is.na(.), 0)


library(reshape2)

# Subset the columns
first_set <- combined_corr_dpmclass[, 2]
second_set <- combined_corr_dpmclass[, 3:32]
data_subset <- combined_corr_dpmclass[, c(2, 3:32)]

# Perform pairwise correlation between columns of the two subsets
cor_matrix <- cor(first_set, second_set, method="spearman")
cmatrix<-as.matrix(cor_matrix)
p_matrix<-cor_pmat(data_subset)[1,2:31]
adjusted<-p.adjust(p_matrix, method="BH")
trans<-t(adjusted)
rownames(trans)<-c("MLS")
pmatrix<-as.matrix(trans)
```

```{r corrplot viz 4b}
### calc sum MLS_dpm
all_data<-combined_corr_dpmgene %>%
  left_join(AMR_corr_mlsdpm_class, by="sample_id")

reorg<-cbind(all_data[1:8], all_data[39], all_data[9:38])

# Subset the columns
first_set <- reorg[, 2:9]
second_set <- reorg[, 10:39]
data_subset <- reorg[, c(2:9, 10:39)]

# Perform pairwise correlation between columns of the two subsets
cor_matrix <- cor(first_set, second_set, method="spearman")
cmatrix<-as.matrix(cor_matrix)
p_matrix<-cor_pmat(data_subset)[1:8,9:38]
pmatrix<-as.matrix(p_matrix) 
p_values_vector <- as.vector(pmatrix)
adjusted_p_values <- p.adjust(p_values_vector, method = "BH") 
adjusted_p_matrix <- matrix(adjusted_p_values, nrow = nrow(pmatrix), ncol = ncol(pmatrix))
rownames(adjusted_p_matrix)<-row.names(cmatrix)
colnames(adjusted_p_matrix)<-colnames(cmatrix)
cmatrix[is.na(cmatrix)]<-0
adjusted_p_matrix[is.na(adjusted_p_matrix)]<-1

write.csv(cmatrix,"source_data_correlation_values_Figure4B.csv")
write.csv(pmatrix, "source_data_pvalues_Figure4B.csv")


corrplot(cmatrix, method="circle", type='full', tl.col='black', p.mat=adjusted_p_matrix, sig.level = c(0.001, 0.01, 0.05), pch.cex = 0.6, insig = 'label_sig', pch.col = 'grey20', tl.cex = 0.6,  col=colorRampPalette(c("#3B9AB2","#EEEEEE","#F21A00"))(200))

pdf(file="4B.pdf", width=6, height=2.5)
corrplot(cmatrix, method="circle", type='full', tl.col='black', p.mat=adjusted_p_matrix, sig.level = c(0.001, 0.01, 0.05), pch.cex = 0.6, insig='label_sig', pch.col = 'grey20', tl.cex = 0.6, col=colorRampPalette(c("#3B9AB2","#EEEEEE","#F21A00"))(200))
invisible(capture.output(dev.off()))
```

### Figure S6 - Sensitivity Analysis

```{r GAMM by TG, fig.height=5, fig.width=8}
library(gamm4)
### TG 1-3
richnessgamm_new <- AMR_NS_GAMM %>%
  left_join(clinical_data) %>%
  left_join(ABX_days_NS_GAMM_plus_new) %>% #add additional data
  left_join(NS_group_timepoint_data_0306) %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(richness=n()) %>% #calc richness by class
  ungroup() %>%
  dplyr::filter(abx_class=="MLS") %>% #grab only MLS richness
  dplyr::select(sample_id, ABX_days,richness,Azithromycin, event_date, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, trajectory_group, sex, discretized_admit_age_quantile, participant_id, enrollment_site, group_id, ever_steroids) %>% 
  distinct() %>% filter(trajectory_group<=3, !is.na(trajectory_group))
  
noabx_rich_new<-richnessgamm_new %>% filter(group_id=="No ABX")
  #filter(ABX_days==0)
noavg_rich_new<-mean(noabx_rich_new$richness)

 formula_fit <- formula(richness ~ s(Azithromycin, bs = 'cr', k= 4) +s(event_date, bs = 'cr') + s(Ceftriaxone, bs = 'cr', k= 4) + s(Vancomycin, bs = 'cr', k= 4) + s(Cefepime, bs = 'cr', k= 4) + s(Piperacillin_Tazobactam, bs = 'cr', k= 4) + s(Doxycycline, bs = 'cr', k= 4) + s(Meropenem, bs = 'cr', k= 4) + trajectory_group + sex + discretized_admit_age_quantile + ever_steroids)

fit <- gamm4::gamm4(data = as.data.frame(richnessgamm_new), formula_fit, random = ~(1|enrollment_site/participant_id))

tmp <- data.frame("feature" = "Azithromycin", "p.value" = anova(fit$gam)[["s.pv"]][1])
#tmp
pred1 <- ggpredict(fit, c("Azithromycin"), type = "fixed", ci_level = CI_interval)

p20 <-pred1 %>%
      ggplot(mapping = aes(x = x, y = predicted, fill=group)) +
      geom_line(mapping = aes(y = predicted, x=x), size = 0.8, color="#E07E3D") +
      geom_line(y=noavg_rich_new, color="#4268C6", linetype=2)+
      geom_ribbon(aes(x=x, ymin = conf.low, ymax = conf.high),alpha = 0.3, colour=NA) +
      theme_classic(base_size  = 14) +
      scale_color_manual(values=c("#E07E3D")) +
      scale_fill_manual(values=c("#E07E3D")) +
      ylab("Number of Detectably Expressed MLS ARGs") +
      xlab("Days of Azithromycin Exposure") +
      theme(legend.position = "none") +
      coord_cartesian(xlim = c(0, 5)) +
      geom_text(label=paste("p=",case_when(tmp$p.value<0.01 ~ format(tmp$p.value, scientific=TRUE, digits=2),tmp$p.value>0.045 & tmp$p.value<0.05 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)),  tmp$p.value>0.0095 & tmp$p.value<0.01 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(tmp$p.value,digits=2), nsmall = 2)))), x=0.5, y=3.3, size=5)
plot(p20)


### TG 4-5

richnessgamm_new <- AMR_NS_GAMM %>%
  left_join(clinical_data) %>%
  left_join(ABX_days_NS_GAMM_plus_new) %>% #add additional data
  left_join(NS_group_timepoint_data_0306) %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(richness=n()) %>% #calc richness by class
  ungroup() %>%
  dplyr::filter(abx_class=="MLS") %>% #grab only MLS richness
  dplyr::select(sample_id, ABX_days,richness,Azithromycin, event_date, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, trajectory_group, sex, discretized_admit_age_quantile, participant_id, enrollment_site, group_id, ever_steroids) %>% 
  distinct() %>% filter(trajectory_group>3, !is.na(trajectory_group))
  
noabx_rich_new<-richnessgamm_new %>% filter(group_id=="No ABX")
  #filter(ABX_days==0)
noavg_rich_new<-mean(noabx_rich_new$richness)

 formula_fit <- formula(richness ~ s(Azithromycin, bs = 'cr', k= 4) +s(event_date, bs = 'cr') + s(Ceftriaxone, bs = 'cr', k= 4) + s(Vancomycin, bs = 'cr', k= 4) + s(Cefepime, bs = 'cr', k= 4) + s(Piperacillin_Tazobactam, bs = 'cr', k= 4) + s(Doxycycline, bs = 'cr', k= 4) + s(Meropenem, bs = 'cr', k= 4) + trajectory_group + sex + discretized_admit_age_quantile + ever_steroids)

fit <- gamm4::gamm4(data = as.data.frame(richnessgamm_new), formula_fit, random = ~(1|enrollment_site/participant_id))

tmp <- data.frame("feature" = "Azithromycin", "p.value" = anova(fit$gam)[["s.pv"]][1])
#tmp
pred2 <- ggpredict(fit, c("Azithromycin"), type = "fixed", ci_level = CI_interval)

p21 <-pred2 %>%
      ggplot(mapping = aes(x = x, y = predicted, fill=group)) +
      geom_line(mapping = aes(y = predicted, x=x), size = 0.8, color="#E07E3D") +
      geom_line(y=noavg_rich_new, color="#4268C6", linetype=2)+
      geom_ribbon(aes(x=x, ymin = conf.low, ymax = conf.high),alpha = 0.3, colour=NA) +
      theme_classic(base_size  = 14) +
      scale_color_manual(values=c("#E07E3D")) +
      scale_fill_manual(values=c("#E07E3D")) +
      ylab("Number of Detectably Expressed MLS ARGs") +
      xlab("Days of Azithromycin Exposure") +
      theme(legend.position = "none") +
      coord_cartesian(xlim = c(0, 5)) +
      geom_text(label=paste("p=",case_when(tmp$p.value<0.01 ~ format(tmp$p.value, scientific=TRUE, digits=2),tmp$p.value>0.045 & tmp$p.value<0.05 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)),  tmp$p.value>0.0095 & tmp$p.value<0.01 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(tmp$p.value,digits=2), nsmall = 2)))), x=0.5, y=3.3, size=5)
plot(p21)

# Combine plots into one
preds_13 <- pred1 %>% mutate(TG = "1-3")
preds_45 <- pred2 %>% mutate(TG = "4-5")
combined_preds <- bind_rows(preds_13, preds_45)

p <-ggplot(combined_preds, aes(x = x, y = predicted, color = TG)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = TG),
              alpha = 0.2, color = NA) +
  theme_classic(base_size = 14) +
  xlab("Days of Azithromycin Exposure") +
  ylab("Predicted MLS ARG Richness") +
  coord_cartesian(xlim = c(0, 5)) +
  guides(fill = "none") +
  scale_color_brewer(palette = "Set1") +
  annotate("text", x = 0.35, y = 3.5, label = paste0("p = 0.86"),hjust = 0, size = 5)

p

## 1) Build one data frame with both strata
dat <- AMR_NS_GAMM %>%
  left_join(clinical_data) %>%
  left_join(ABX_days_NS_GAMM_plus_new) %>%
  left_join(NS_group_timepoint_data_0306) %>%
  group_by(sample_id, abx_class) %>%
  mutate(richness = n()) %>%
  ungroup() %>%
  filter(abx_class == "MLS") %>%
  select(sample_id, ABX_days, richness, Azithromycin, event_date,
         Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam,
         Doxycycline, Meropenem, trajectory_group, sex,
         discretized_admit_age_quantile, participant_id, enrollment_site,
         group_id, ever_steroids) %>%
  distinct() %>%
  mutate(TG = factor(ifelse(trajectory_group <= 3, "1-3", "4-5"))) %>%
  filter(!is.na(TG))

## Reduced model: ONE common smooth of Azithromycin
form_red  <- as.formula(paste("richness ~ s(Azithromycin, bs='cr', k=4) + s(event_date, bs= 'cr') + s(Ceftriaxone, bs = 'cr', k = 4) + s(Vancomycin, bs = 'cr', k = 4) +
  s(Cefepime, bs = 'cr', k = 4) + s(Piperacillin_Tazobactam, bs = 'cr', k = 4) +
  s(Doxycycline, bs = 'cr', k = 4) + s(Meropenem, bs = 'cr', k = 4) +
  TG + sex + discretized_admit_age_quantile + ever_steroids"))
fit_red   <- gamm4(form_red, random = ~(1|enrollment_site/participant_id), data = dat)

form_full <- as.formula(paste("richness ~ s(Azithromycin, bs='cr', k=4) + s(Azithromycin, by=TG, bs='cr', k=4) + s(event_date, bs= 'cr') + s(Ceftriaxone, bs = 'cr', k = 4) + s(Vancomycin, bs = 'cr', k = 4) + s(Cefepime, bs = 'cr', k = 4) + s(Piperacillin_Tazobactam, bs = 'cr', k = 4) +
s(Doxycycline, bs = 'cr', k = 4) + s(Meropenem, bs = 'cr', k = 4) +
TG + sex + discretized_admit_age_quantile + ever_steroids"))
fit_full  <- gamm4(form_full, random = ~(1|enrollment_site/participant_id), data = dat)


cmp <- anova(fit_red$mer, fit_full$mer)   # from lme4
cmp
p_global <- cmp$`Pr(>Chisq)`[2]
p_global


p
ggsave("FigureS6A.pdf", width = 4, height = 2.5)
  write.csv(combined_preds, "source_data_FigureS6A.csv")
```


```{r MLS percent compare TGs}
### TG 1-3
mlsdpmgamm_new <- AMR_NS_GAMM %>%
  left_join(clinical_data) %>%
  left_join(ABX_days_NS_GAMM_plus_new) %>% #add additional info
  left_join(NS_group_timepoint_data_0306) %>%
  group_by(sample_id) %>%
  dplyr::mutate(sum_dpm=sum(dpm)) %>%
  dplyr::mutate(sum_reads=sum(num_reads)) %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(class_dpm=sum(dpm)) %>% #calc sum DPM MLS
  dplyr::mutate(class_reads=sum(num_reads)) %>%
  dplyr::mutate(pseudo=class_dpm+1) %>%
  dplyr::mutate(logmlsdpm=log10(pseudo)) %>%
  ungroup() %>%
  #dplyr::mutate(logdpm=log10(class_dpm+1)) %>% #transform class DPM
  dplyr::mutate(perc=(class_dpm/sum_dpm)*100) %>% #calc perc MLS of resistome
  dplyr::mutate(perc2=case_when(is.na(perc) ~ 0, TRUE ~ perc)) %>% #convert NA to 0
  dplyr::mutate(perc3=(class_reads/sum_reads)*100) %>% #calc perc MLS of resistome
  dplyr::mutate(perc4=case_when(is.na(perc3) ~ 0, TRUE ~ perc3)) %>% #convert NA to 0
  dplyr::mutate(logmlsdpm=case_when(is.na(logmlsdpm) ~ 0, TRUE ~ logmlsdpm)) %>% #convert NA to 0
  dplyr::filter(abx_class=="MLS") %>% #filter to only MLS
  ungroup() %>%
  dplyr::select(sample_id, ABX_days,logmlsdpm, class_dpm, sum_dpm, perc, perc2, perc3, perc4,Azithromycin, event_date, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, trajectory_group, sex, discretized_admit_age_quantile, participant_id, enrollment_site, group_id, ever_steroids) %>% 
  distinct() %>%
  filter(trajectory_group<=3, !is.na(trajectory_group))

### Figure 3E
noabx_perc_new<-mlsdpmgamm_new %>% filter(group_id=="No ABX")
  #filter(ABX_days==0)
noavg_perc_new<-mean(noabx_perc_new$perc4)

formula_fit <- formula(perc4 ~ s(Azithromycin, bs = 'cr', k= 4) +s(event_date, bs = 'cr') + s(Ceftriaxone, bs = 'cr', k= 4) + s(Vancomycin, bs = 'cr', k= 4) + s(Cefepime, bs = 'cr', k= 4) + s(Piperacillin_Tazobactam, bs = 'cr', k= 4) + s(Doxycycline, bs = 'cr', k= 4) + s(Meropenem, bs = 'cr', k= 4) + trajectory_group + sex + discretized_admit_age_quantile + ever_steroids)

fit <- gamm4::gamm4(data = as.data.frame(mlsdpmgamm_new), formula_fit, random = ~(1|enrollment_site/participant_id))

tmp <- data.frame("feature" = "Azithromycin", "p.value" = anova(fit$gam)[["s.pv"]][1])
#tmp
pred13 <- ggpredict(fit, c("Azithromycin"), type = "fixed", ci_level = CI_interval)

p13 <-pred13 %>%
  ggplot(mapping = aes(x = x, y = predicted, fill=group)) +
  geom_line(mapping = aes(y = predicted, x=x), size = 0.8, color="#E07E3D") +
  geom_line(y=noavg_perc_new, color="#4268C6", linetype=2)+
  geom_ribbon(aes(x=x, ymin = conf.low, ymax = conf.high),alpha = 0.3, colour=NA) +
  theme_classic(base_size=14) +
  scale_color_manual(values=c("#E07E3D")) + scale_fill_manual(values=c("#E07E3D")) +
  ylab("MLS Percent of Resistome") + xlab("Days of Azithromycin Exposure") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 5), ylim=c(0,70)) +
  geom_text(label=paste("p=",case_when(tmp$p.value<0.01 ~ format(tmp$p.value, scientific=TRUE, digits=2),tmp$p.value>0.045 & tmp$p.value<0.05 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)),  tmp$p.value>0.0095 & tmp$p.value<0.01 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(tmp$p.value,digits=2), nsmall = 2)))), x=0.5, y=68, size=5)
plot(p13)

### TG 4-5
mlsdpmgamm_new <- AMR_NS_GAMM %>%
  left_join(clinical_data) %>%
  left_join(ABX_days_NS_GAMM_plus_new) %>% #add additional info
  left_join(NS_group_timepoint_data_0306) %>%
  group_by(sample_id) %>%
  dplyr::mutate(sum_dpm=sum(dpm)) %>%
  dplyr::mutate(sum_reads=sum(num_reads)) %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(class_dpm=sum(dpm)) %>% #calc sum DPM MLS
  dplyr::mutate(class_reads=sum(num_reads)) %>%
  dplyr::mutate(pseudo=class_dpm+1) %>%
  dplyr::mutate(logmlsdpm=log10(pseudo)) %>%
  ungroup() %>%
  #dplyr::mutate(logdpm=log10(class_dpm+1)) %>% #transform class DPM
  dplyr::mutate(perc=(class_dpm/sum_dpm)*100) %>% #calc perc MLS of resistome
  dplyr::mutate(perc2=case_when(is.na(perc) ~ 0, TRUE ~ perc)) %>% #convert NA to 0
  dplyr::mutate(perc3=(class_reads/sum_reads)*100) %>% #calc perc MLS of resistome
  dplyr::mutate(perc4=case_when(is.na(perc3) ~ 0, TRUE ~ perc3)) %>% #convert NA to 0
  dplyr::mutate(logmlsdpm=case_when(is.na(logmlsdpm) ~ 0, TRUE ~ logmlsdpm)) %>% #convert NA to 0
  dplyr::filter(abx_class=="MLS") %>% #filter to only MLS
  ungroup() %>%
  dplyr::select(sample_id, ABX_days,logmlsdpm, class_dpm, sum_dpm, perc, perc2, perc3, perc4,Azithromycin, event_date, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, trajectory_group, sex, discretized_admit_age_quantile, participant_id, enrollment_site, group_id, ever_steroids) %>% 
  distinct() %>%
  filter(trajectory_group>3, !is.na(trajectory_group))

### Figure 3E
noabx_perc_new<-mlsdpmgamm_new %>% filter(group_id=="No ABX")
  #filter(ABX_days==0)
noavg_perc_new<-mean(noabx_perc_new$perc4)

formula_fit <- formula(perc4 ~ s(Azithromycin, bs = 'cr', k= 4) +s(event_date, bs = 'cr') + s(Ceftriaxone, bs = 'cr', k= 4) + s(Vancomycin, bs = 'cr', k= 4) + s(Cefepime, bs = 'cr', k= 4) + s(Piperacillin_Tazobactam, bs = 'cr', k= 4) + s(Doxycycline, bs = 'cr', k= 4) + s(Meropenem, bs = 'cr', k= 4) + trajectory_group + sex + discretized_admit_age_quantile + ever_steroids)

fit <- gamm4::gamm4(data = as.data.frame(mlsdpmgamm_new), formula_fit, random = ~(1|enrollment_site/participant_id))

tmp <- data.frame("feature" = "Azithromycin", "p.value" = anova(fit$gam)[["s.pv"]][1])
#tmp
pred45 <- ggpredict(fit, c("Azithromycin"), type = "fixed", ci_level = CI_interval)

p45 <-pred45 %>%
  ggplot(mapping = aes(x = x, y = predicted, fill=group)) +
  geom_line(mapping = aes(y = predicted, x=x), size = 0.8, color="#E07E3D") +
  geom_line(y=noavg_perc_new, color="#4268C6", linetype=2)+
  geom_ribbon(aes(x=x, ymin = conf.low, ymax = conf.high),alpha = 0.3, colour=NA) +
  theme_classic(base_size=14) +
  scale_color_manual(values=c("#E07E3D")) + scale_fill_manual(values=c("#E07E3D")) +
  ylab("MLS Percent of Resistome") + xlab("Days of Azithromycin Exposure") +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0, 5), ylim=c(0,70)) +
  geom_text(label=paste("p=",case_when(tmp$p.value<0.01 ~ format(tmp$p.value, scientific=TRUE, digits=2),tmp$p.value>0.045 & tmp$p.value<0.05 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)),  tmp$p.value>0.0095 & tmp$p.value<0.01 ~ as.character(format(round(tmp$p.value,digits=3), nsmall = 2)), TRUE ~ as.character(format(round(tmp$p.value,digits=2), nsmall = 2)))), x=0.5, y=68, size=5)
plot(p45)


# Combine plots into one
preds_13 <- pred13 %>% mutate(TG = "1-3")
preds_45 <- pred45 %>% mutate(TG = "4-5")
combined_preds <- bind_rows(preds_13, preds_45)

p <-ggplot(combined_preds, aes(x = x, y = predicted, color = TG)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = TG),
              alpha = 0.2, color = NA) +
  theme_classic(base_size = 14) +
  xlab("Days of Azithromycin Exposure") +
  ylab("Predicted MLS Percent of Resistome") +
  coord_cartesian(xlim = c(0, 5)) +
  guides(fill = "none") +
  scale_color_brewer(palette = "Set1") +  
  annotate("text", x = 0.35, y = 65, label = paste0("p = 0.54"),hjust = 0, size = 5)

p

## 1) Build one data frame with both strata
dat <- AMR_NS_GAMM %>%
  left_join(clinical_data) %>%
  left_join(ABX_days_NS_GAMM_plus_new) %>% #add additional info
  left_join(NS_group_timepoint_data_0306) %>%
  group_by(sample_id) %>%
  dplyr::mutate(sum_dpm=sum(dpm)) %>%
  dplyr::mutate(sum_reads=sum(num_reads)) %>%
  group_by(sample_id, abx_class) %>%
  dplyr::mutate(class_dpm=sum(dpm)) %>% #calc sum DPM MLS
  dplyr::mutate(class_reads=sum(num_reads)) %>%
  dplyr::mutate(pseudo=class_dpm+1) %>%
  dplyr::mutate(logmlsdpm=log10(pseudo)) %>%
  ungroup() %>%
  #dplyr::mutate(logdpm=log10(class_dpm+1)) %>% #transform class DPM
  dplyr::mutate(perc=(class_dpm/sum_dpm)*100) %>% #calc perc MLS of resistome
  dplyr::mutate(perc2=case_when(is.na(perc) ~ 0, TRUE ~ perc)) %>% #convert NA to 0
  dplyr::mutate(perc3=(class_reads/sum_reads)*100) %>% #calc perc MLS of resistome
  dplyr::mutate(perc4=case_when(is.na(perc3) ~ 0, TRUE ~ perc3)) %>% #convert NA to 0
  dplyr::mutate(logmlsdpm=case_when(is.na(logmlsdpm) ~ 0, TRUE ~ logmlsdpm)) %>% #convert NA to 0
  dplyr::filter(abx_class=="MLS") %>% #filter to only MLS
  ungroup() %>%
  dplyr::select(sample_id, ABX_days,logmlsdpm, class_dpm, sum_dpm, perc, perc2, perc3, perc4,Azithromycin, event_date, Ceftriaxone, Vancomycin, Cefepime, Piperacillin_Tazobactam, Doxycycline, Meropenem, trajectory_group, sex, discretized_admit_age_quantile, participant_id, enrollment_site, group_id, ever_steroids) %>% 
  distinct() %>%
  mutate(TG = factor(ifelse(trajectory_group <= 3, "1-3", "4-5"))) %>%
  filter(!is.na(TG))

## Reduced model: ONE common smooth of Azithromycin
form_red  <- as.formula(paste("perc4 ~ s(Azithromycin, bs='cr', k=4) + s(event_date, bs= 'cr') + s(Ceftriaxone, bs = 'cr', k = 4) + s(Vancomycin, bs = 'cr', k = 4) +
  s(Cefepime, bs = 'cr', k = 4) + s(Piperacillin_Tazobactam, bs = 'cr', k = 4) +
  s(Doxycycline, bs = 'cr', k = 4) + s(Meropenem, bs = 'cr', k = 4) +
  TG + sex + discretized_admit_age_quantile + ever_steroids"))

fit_red   <- gamm4(form_red, random = ~(1|enrollment_site/participant_id), data = dat)

form_full <- as.formula(paste("perc4 ~ s(Azithromycin, bs='cr', k=4) + s(Azithromycin, by=TG, bs='cr', k=4) + s(event_date, bs= 'cr') + s(Ceftriaxone, bs = 'cr', k = 4) + s(Vancomycin, bs = 'cr', k = 4) + s(Cefepime, bs = 'cr', k = 4) + s(Piperacillin_Tazobactam, bs = 'cr', k = 4) +
s(Doxycycline, bs = 'cr', k = 4) + s(Meropenem, bs = 'cr', k = 4) +
TG + sex + discretized_admit_age_quantile + ever_steroids"))
fit_full  <- gamm4(form_full, random = ~(1|enrollment_site/participant_id), data = dat)


cmp <- anova(fit_red$mer, fit_full$mer)   # from lme4
cmp
p_global <- cmp$`Pr(>Chisq)`[2]
p_global

p
ggsave("FigureS6B.pdf", width = 4, height = 2.5)
write.csv(combined_preds, "source_data_FigureS6B.csv")
```

## Figure S7 - Species-level assignment of ARGS

```{r species top 30 }
Bact_corr_sp<-nasal_long_species %>%
  dplyr::filter(sample_id %in% samples$sample_id, 
         category=="bacteria", 
         nt_alignment_length>=70, 
         nt_count>=10, 
         nr_count>=1, 
         p_adj<0.1) %>%
  dplyr::select(sample_id, name, nt_rpm) %>%
  pivot_wider(names_from = name, values_from = nt_rpm, values_fill=0) 

### select columns with sp 
microbes_sp <-Bact_corr_sp[c(2, 3:length(Bact_corr_sp))]

### transpose
tdf_sp<-t(microbes_sp)

### Top 30 most abundant species
data_sp<-as.data.frame(tdf_sp)

with_total_sp30<-data_sp %>%
  dplyr::mutate(sum = rowSums(.)) %>% 
  arrange(desc(sum)) %>% 
  do(head(., n=30))

### list microbes
trimmed_list_microbes_sp30<-rownames(with_total_sp30)

### subset data
top30_data_sp<-Bact_corr_sp %>% dplyr::select(sample_id, trimmed_list_microbes_sp30)
top30_data_sp[sapply(top30_data_sp, is.numeric)] <- top30_data_sp[sapply(top30_data_sp, is.numeric)] + 1
top30_data_sp[sapply(top30_data_sp, is.numeric)] <-sapply(top30_data_sp[sapply(top30_data_sp, is.numeric)], log10)

combined_corr_dpmclass_sp30<-left_join(AMR_corr_mlsdpm_class, top30_data_sp, by="sample_id") %>%
  replace(is.na(.), 0)

combined_corr_percclass_sp30<-left_join(AMR_corr_mlsperc_class, top30_data_sp, by="sample_id") %>%
  replace(is.na(.), 0)

combined_corr_dpmgene_sp30<-left_join(AMR_corr_mlsdpm_gene, top30_data_sp, by="sample_id") %>%
  replace(is.na(.), 0)

combined_corr_percgene_sp30<-left_join(AMR_corr_mlsperc_gene, top30_data_sp, by="sample_id") %>%
  replace(is.na(.), 0)


library(reshape2)

# Subset the columns
first_set_sp30 <- combined_corr_dpmclass_sp30[, 2]
second_set_sp30 <- combined_corr_dpmclass_sp30[, 3:32]
data_subset_sp30 <- combined_corr_dpmclass_sp30[, c(2, 3:32)]

# Perform pairwise correlation between columns of the two subsets
cor_matrix_sp30 <- cor(first_set_sp30, second_set_sp30, method="spearman")
cmatrix_sp30<-as.matrix(cor_matrix_sp30)
p_matrix_sp30<-cor_pmat(data_subset_sp30)[1,2:31]
adjusted_sp30<-p.adjust(p_matrix_sp30, method="BH")
trans_sp30<-t(adjusted_sp30)
rownames(trans_sp30)<-c("MLS")
pmatrix_sp30<-as.matrix(trans_sp30)
```

```{r corrplot viz}
### calc sum MLS_dpm
all_data_sp30<-combined_corr_dpmgene_sp30 %>%
  left_join(AMR_corr_mlsdpm_class, by="sample_id")

reorg_sp30<-cbind(all_data_sp30[1:8], all_data_sp30[39], all_data_sp30[9:38])

# Subset the columns
first_set_sp30 <- reorg_sp30[, 2:9]
second_set_sp30 <- reorg_sp30[, 10:39]
data_subset_sp30 <- reorg_sp30[, c(2:9, 10:39)]

# Perform pairwise correlation between columns of the two subsets
cor_matrix_sp30 <- cor(first_set_sp30, second_set_sp30, method="spearman")
cmatrix_sp30<-as.matrix(cor_matrix_sp30)
p_matrix_sp30<-cor_pmat(data_subset_sp30)[1:8,9:38]
pmatrix_sp30<-as.matrix(p_matrix_sp30) 
p_values_vector_sp30 <- as.vector(pmatrix_sp30)
adjusted_p_values_sp30 <- p.adjust(p_values_vector_sp30, method = "BH") 
adjusted_p_matrix_sp30 <- matrix(adjusted_p_values_sp30, nrow = nrow(pmatrix_sp30), ncol = ncol(pmatrix_sp30))
rownames(adjusted_p_matrix_sp30)<-row.names(cmatrix_sp30)
colnames(adjusted_p_matrix_sp30)<-colnames(cmatrix_sp30)
cmatrix_sp30[is.na(cmatrix_sp30)]<-0
adjusted_p_matrix_sp30[is.na(adjusted_p_matrix_sp30)]<-1

write.csv(cmatrix_sp30,"source_data_correlation_values_species30_FigureS7A.csv")
write.csv(adjusted_p_matrix_sp30, "source_data_pvalues_species30_FigureS7A.csv")

mat <- cmatrix_sp30[, order(colnames(cmatrix_sp30)), drop = FALSE]
matp <- adjusted_p_matrix_sp30[, order(colnames(adjusted_p_matrix_sp30)), drop = FALSE]

row_order <- c("erma", "ermb", "ermc", "ermx", "mel", "mphc", "msra", "MLS")

# keep only names that exist, in that order
row_idx <- na.omit(match(row_order, rownames(mat)))
cor_rows_only <- mat[row_idx, , drop = FALSE]
p_rows_only <- matp[row_idx, , drop = FALSE]

corrplot(cor_rows_only, is.corr=FALSE,method="circle", type='full', tl.col='black', p.mat=p_rows_only, sig.level = c(0.001, 0.01, 0.05), pch.cex = 0.6, insig='label_sig', pch.col = 'grey20', tl.cex = 0.6,col=colorRampPalette(c("#3B9AB2","#EEEEEE","#F21A00"))(200), cl.cex=0.6, col.lim=c(-1,1))

pdf(file="S7A_sp.pdf", width=6, height=2.5)
corrplot(cmatrix_sp30, method="circle", type='full', tl.col='black', p.mat=adjusted_p_matrix_sp30, sig.level = c(0.001, 0.01, 0.05), pch.cex = 0.6, insig='label_sig', pch.col = 'grey20', tl.cex = 0.6, col=colorRampPalette(c("#3B9AB2","#EEEEEE","#F21A00"))(200))
invisible(capture.output(dev.off()))
```

```{r CARD Path calls}
library(pheatmap)
AMR_NS_species<-AMR_NS_GAMM %>%
  select(sample_name,gene_name,read_species,rpm,dpm,sample_id,abx_class) %>%
  filter(!is.na(read_species), abx_class=="MLS") %>% 
  left_join(NS_group_timepoint_data_0306, by="sample_id") %>%
  filter(group_id=="Azithromycin")

AMR_NS_sp_02<- AMR_NS_species %>%
  filter(timepoint=="Day 0-2") %>%
  select(sample_id, gene_name, rpm, dpm, read_species) %>%
  group_by(gene_name,read_species) %>%
  mutate(species_sumdpm=sum(dpm)) %>%
  ungroup() %>%
  select(gene_name, read_species, species_sumdpm) %>%
  distinct()

AMR_NS_sp_46<- AMR_NS_species %>%
  filter(timepoint=="Day 4-6") %>%
  select(sample_id, gene_name, rpm, dpm, read_species) %>%
  group_by(gene_name,read_species) %>%
  mutate(species_sumdpm=sum(dpm)) %>%
  ungroup() %>%
  select(gene_name, read_species, species_sumdpm) %>%
  distinct()

matrix02<-AMR_NS_sp_02 %>%
  pivot_wider(names_from = gene_name, values_from = species_sumdpm, id_cols=read_species, values_fill = 0) %>%
    mutate(erma=`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`) %>%
  select(-`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`)
df1<-matrix02 %>% select(-read_species)
rownames(df1)<-matrix02$read_species

matrix46<-AMR_NS_sp_46 %>%
  pivot_wider(names_from = gene_name, values_from = species_sumdpm, id_cols=read_species, values_fill = 0) %>%
  mutate(erma=`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`) %>%
  select(-`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`)
df2<-matrix46 %>% select(-read_species)
rownames(df2)<-matrix46$read_species

```

```{r cleaned calls}
df_long <- AMR_NS_species %>%
  # Separate multiple species into individual rows
  separate_rows(read_species, sep = ";") %>%
  # Remove leading/trailing whitespace
  mutate(read_species2 = str_trim(read_species)) %>%
  # Extract only the genus + species part (before first parenthesis)
  mutate(species_clean = str_extract(read_species2,  "^.*\\)")) %>%
  # remove empties
  filter(!is.na(species_clean)) %>%
  # Trim again in case there was a trailing space
  mutate(species_clean2 = str_trim(species_clean)) %>%
  select(gene_name, sample_id, species_clean2, dpm) %>%
  distinct() %>%
  left_join(NS_group_timepoint_data_0306)
```
```{r split new data into groups}
AMR_NS_sp_02<- df_long %>%
  filter(timepoint=="Day 0-2") %>%
  select(sample_id, gene_name, dpm, species_clean2) %>%
  group_by(gene_name,species_clean2) %>%
  mutate(species_sumdpm=sum(dpm)) %>%
  ungroup() %>%
  select(gene_name, species_clean2, species_sumdpm) %>%
  distinct()

AMR_NS_sp_46<- df_long %>%
  filter(timepoint=="Day 4-6") %>%
  select(sample_id, gene_name, dpm, species_clean2) %>%
  group_by(gene_name,species_clean2) %>%
  mutate(species_sumdpm=sum(dpm)) %>%
  ungroup() %>%
  select(gene_name, species_clean2, species_sumdpm) %>%
  distinct()

matrix02<-AMR_NS_sp_02 %>%
  pivot_wider(names_from = gene_name, values_from = species_sumdpm, id_cols=species_clean2, values_fill = 0) %>%
    mutate(erma=`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`) %>%
  select(-`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`)
df1<-matrix02 %>% select(-species_clean2)
rownames(df1)<-matrix02$species_clean2

matrix46<-AMR_NS_sp_46 %>%
  pivot_wider(names_from = gene_name, values_from = species_sumdpm, id_cols=species_clean2, values_fill = 0) %>%
  mutate(erma=`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`) %>%
  select(-`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`)
df2<-matrix46 %>% select(-species_clean2)
rownames(df2)<-matrix46$species_clean2
```

```{r}

### select samples
samples<-AMR_NS_GAMM %>%
left_join(ABX_days_NS_GAMM_new, by="sample_id") %>%
dplyr::filter(Azithromycin>0) %>%
dplyr::select(sample_id) %>%
distinct()

df_long_azith<-AMR_NS_species %>%
  filter(sample_id %in% samples$sample_id) %>%
  separate_rows(read_species, sep = ";") %>%
  mutate(read_species2 = str_trim(read_species)) %>%
  mutate(species_clean = str_extract(read_species2, "^.*\\)")) %>%
  filter(!is.na(species_clean)) %>%
  mutate(species_clean2 = str_trim(species_clean)) %>%
  select(gene_name, sample_id, species_clean2, dpm) %>%
  distinct()

new<- df_long_azith %>%
  select(sample_id, gene_name, dpm, species_clean2) %>%
  group_by(gene_name,species_clean2) %>%
  mutate(species_sumdpm=sum(dpm)) %>%
  ungroup() %>%
  select(gene_name, species_clean2, species_sumdpm) %>%
  distinct()

matrix_match<-new %>%
  pivot_wider(names_from = gene_name, values_from = species_sumdpm, id_cols=species_clean2, values_fill = 0) %>%
    mutate(erma=`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`) %>%
  select(-`23s rrna (adenine(2058)-n(6))-methyltransferase erm(a)`)
df3<-matrix_match %>% select(-species_clean2)
rownames(df3)<-matrix_match$species_clean2

pheatmap(df3,legend=TRUE,
         color = colorRampPalette(c("white", "red"))(100),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "none",
         fontsize_row = 8,
         main="MLS ARG Sum DPM"
         )

pdf(file="FigureS7B_sp.pdf", width=5, height=5)
pheatmap(df3,legend=TRUE,
         color = colorRampPalette(c("white", "red"))(100),
         clustering_distance_rows = "euclidean",
         clustering_distance_cols = "euclidean",
         clustering_method = "complete",
         scale = "none",
         fontsize_row = 8,
         main="MLS ARG Sum DPM"
         )
invisible(capture.output(dev.off()))
write.csv(df3, "source_data_FigureS7B.csv")
```

## Session Info

```{r}
sessionInfo()
```
