---
title: "Extended Data Fig7"
output: pdf_document
---

```{r load library}
set.seed(123)
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source(paste0(codebase_path,"/codebase_subtype.R"))

# load subtype assignment and clinical data
nclust_tree_all = 6
subtype_hcl = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
clin_data_aligned = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/clin_data_aligned.rds")
organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")
```

# Enrichment preparation
```{r enrichment preparation}
library(clusterProfiler)
data_path <- "/scratch/data-integration/pipelines/processedData/"
data_env0 <- readRDS(paste0(data_path, "data_env_ravi_20220513.RDS"))

intermediate_results_outdir_DR = "/scratch/data-integration/molecular_subtyping/DimensionReduction/"
mcia_results = readRDS(paste0(intermediate_results_outdir_DR,"MCIA_phase123_20230726.rds"))

kegg_database = readRDS(file = "/scratch/data-integration/pipelines/Interpretation/remapped_databases/kegg0522.rds")
hallmark_database = readRDS( file = "/scratch/data-integration/pipelines/Interpretation/remapped_databases/hallmark0522.rds")
metabolite_row_DF = data_env0$plasma_metabolomics_global_rowfeature
rownames(metabolite_row_DF) = metabolite_row_DF[["CHEMICAL_NAME"]]
metabolite_row_DF=metabolite_row_DF[,c("CHEMICAL_NAME","SUB_PATHWAY","SUPER_PATHWAY")]
subpath_databse = list()
subpath_databse[["pmg"]] = data.frame(description =metabolite_row_DF[["SUB_PATHWAY"]],
                                      CHEMICAL_NAME = metabolite_row_DF[["CHEMICAL_NAME"]])

mcia_output = mcia_results
omics_used_DR = c(
  "ppt" = "plasma_proteomics_targeted",
  "ppgd" = "plasma_proteomics_global_dda",
  "so" = "serum_olink",
  "pmg" = "plasma_metabolomics_global",
  "nt" = "nasal_transcriptomics",
  "pbmc" = "pbmc_transcriptomics"
)
omic_short_names = names(omics_used_DR) %>% setNames(omics_used_DR)
omic_names =c("plasma_proteomics_targeted","plasma_proteomics_global_dda",
              "serum_olink","plasma_metabolomics_global","nasal_transcriptomics",
              "pbmc_transcriptomics")  
regression_coef = mcia_output$mcia_model$pb
for(d in 1:length(regression_coef)){
  regression_coef[[d]] = data.frame(regression_coef[[d]])
  rownames( regression_coef[[d]]) = rownames(mcia_output$mcia_factor_evaluation$projection_coef[[d]])
}
projection_coef = convert_mciaObj_to_annotation_format(mcia_output$mcia_factor_evaluation$projection_coef, omic_names = names(omics_used_DR))
projection_pval = convert_mciaObj_to_annotation_format(mcia_output$mcia_factor_evaluation$projection_pval, omic_names = names(omics_used_DR))
regression_coef = convert_mciaObj_to_annotation_format(regression_coef, omic_names = names(omics_used_DR))
projection_coef=projection_coef[1:40]
projection_pval=projection_pval[1:40]
regression_coef= regression_coef[1:40]
for(mod_id in c(1:40)){
  mod_name = paste0("mod", mod_id)
  print(mod_name)
  projection_coef[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = projection_coef[[mod_id]], omic_names = omic_names) 
  projection_pval[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = projection_pval[[mod_id]], omic_names = omic_names)
  regression_coef[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = regression_coef[[mod_id]], omic_names = omic_names)
}
```

# Factor trajectory (ExtendedData 7A left)
```{r factor trajectory, fig.height=6, fig.width=6}
mod_ids = c(1,2,3,10,30)
negation = c(-1,1,1,1,1)
# negate factors
mcia_factors = t(t(mcia_results$mcia_factors[,mod_ids]) * negation)
mcia_factors = apply(mcia_factors, 2, function(z) qqnorm(z, plot.it = F)$x)
mcia_factors = data.frame(
  mcia_factors,
  row.names = rownames(organized_data$datasets$plasma_proteomics_targeted))
mcia_factors = mcia_factors[clin_data_aligned$event_id,]
mcia_factors = data.frame(
  mcia_factors,
  row.names = clin_data_aligned$event_id)
colnames(mcia_factors) = paste0("Factor", mod_ids)

assay_mat = mcia_factors
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl = subtype_hcl,
                                 clin_data = clin_data_aligned)
assay_mat = assay_mat[!is.na(assay_mat$PASC),]
assay_mat$PASC = ifelse(assay_mat$PASC == "minimal", 
                         "Minimal Deficit", "Other Deficits")
clin_colnames_select = setdiff(colnames(assay_mat), colnames(mcia_factors))
factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "PASC", use_endpoint_levels = c("Minimal Deficit", "Other Deficits"),
  subgroup_cr = c("Other Deficits" = "#9C3418", "Minimal Deficit" = "#39828C"),
  max_date = 40, legend_name = "PASC", multiple_correction = FALSE, 
  text_position = 2)
factor_trajectory_analysis_results$test_fig_list$Factor30
```

# Early/late Factor 30 two-group comparison (Extended Data Fig7A right)
```{r early/late Factor 30 two-group comparison, fig.height=6, fig.width=3}
assay_mat = mcia_factors
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl = subtype_hcl,
                                 clin_data = clin_data_aligned)
assay_mat = assay_mat[!is.na(assay_mat$PASC),]
assay_mat$PASC = ifelse(assay_mat$PASC == "minimal", 
                         "Minimal Deficit", "Other Deficits")
discretized_test = list()
feature_name = c("Factor30")
discretized_test[["early"]] = two_pasc_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V1", "V2", "V3"))
discretized_test[["late"]] = two_pasc_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V4", "V5", "V6"))

# Factor 30
est_mat = cbind(discretized_test$early$est["Factor30"],
                discretized_test$late$est["Factor30"])
rownames(est_mat) = "Any vs Minimal Deficit"
colnames(est_mat) = c("V1-V3", "V4-V6")
pval_mat = cbind(discretized_test$early$pval["Factor30"],
                 discretized_test$late$pval["Factor30"])
colnames(pval_mat) = c("V1-V3", "V4-V6")
adj_p = matrix(p.adjust(pval_mat, method = "BH"), nrow = 1)

factor30_multi_comparison = ComplexHeatmap::Heatmap(
  as.matrix(t(est_mat)), row_names_gp = gpar(fontsize = 15), 
  column_names_gp = gpar(fontsize = 15, lineheight = 1.5), 
  column_names_rot = 0,
  column_names_centered = T,
  heatmap_legend_param = list(title = "Coefficient", direction = "horizontal", 
                              title_position = "topcenter",
                              title_gp = gpar(fontsize = 15), 
                              labels_gp = gpar(fontsize = 15)),
  col = circlize::colorRamp2(c(-max(abs(est_mat)), 0, max(abs(est_mat))), 
                             c("#284AA0", "white", "#BC4A4A")), 
  na_col = "grey90", cluster_columns = F, cluster_rows = F,
  cell_fun = function(j, i, x, y, w, h, fill) {
    if(t(pval_mat)[i, j] < 0.001) {
      grid.text("***", x, y-h*0.08, gp = gpar(cex=4))
      } else if(t(pval_mat)[i, j] < 0.01) {
        grid.text("**", x, y-h*0.08, gp = gpar(cex=4))
        } else if(t(pval_mat)[i, j] < 0.05) {
          grid.text("*", x, y-h*0.08, gp = gpar(cex=4))
          }
    })
ComplexHeatmap::draw(factor30_multi_comparison, heatmap_legend_side = "top")
```

# GSEA
```{r GSEA}
enrichment_gsea_list = list()
enrichment_gsea_list[["kegg"]] = 
  runGSEA(projection_coef = regression_coef[[30]],
          Term2GeneList = kegg_database)
enrichment_gsea_list[["hallmark"]]= 
  runGSEA(projection_coef = regression_coef[[30]],
          Term2GeneList = hallmark_database)
enrichment_gsea_list[["subpath"]]= 
  runGSEA(projection_coef = regression_coef[[30]],
          Term2GeneList = subpath_databse)
```

# Enrichment plot (full)
```{r enrichment plot (full), fig.height=20, fig.width=10}
thr = 0.05
shorten = FALSE
data_base_names = c("kegg", "hallmark", "subpath")
KEGG_max = 10; Hallmark_max = 10; Subpath_max = 5
KEGG_metabo_max = 2; pathway_max = NULL
size_max = c(KEGG_max, Hallmark_max, Subpath_max)
pathway_counts = data.frame(Factor30 = array(NA, 6))

aggregated_mat = enrichment_aggregation_subtype_gsea(
  enrichment_gsea_list, adj.p.thr = thr, thr = .2, type = "gsea")

temp = aggregated_mat[, c(1:6)]
temp[is.na(temp)] = 0
pathway_counts = apply(abs(temp) > (-log(thr, base = 10)), 2, sum)
if(shorten) {
  aggregated_mat_short = pathway_shortlist_selection(
    aggregated_mat, data_base_names = data_base_names,
    size_max = size_max, KEGG_metabo_max = KEGG_metabo_max)
} else {
  aggregated_mat_short =  aggregated_mat
}
aggregated_mat_short = aggregated_mat_short[order(-aggregated_mat_short$joint),]

plot_title = paste0("Factor30")
plotDF_df = gather(
  aggregated_mat_short[1:min(nrow(aggregated_mat_short), pathway_max),],
  key = "dataset", value = "log10.sig.val", -joint, -pathway, -database)
plotDF_df = plotDF_df[!is.na(plotDF_df$log10.sig.val),]
plotDF_df$direction = NA
plotDF_df$direction[plotDF_df$log10.sig.val>0] = "pos"
plotDF_df$direction[plotDF_df$log10.sig.val<0] = "neg"
plotDF_df$direction[plotDF_df$dataset == "joint_adj"] = "unsigned"
plotDF_df$log10.sig.val = abs(plotDF_df$log10.sig.val)

plotDF_df$pathway = factor(
  plotDF_df$pathway,
  levels = aggregated_mat_short$pathway[length(aggregated_mat$pathway):1])
  
plotDF_df$dataset = factor(plotDF_df$dataset,
                           levels = c(omic_short_names, "joint_adj"))
plotDF_df$database = factor(plotDF_df$database,
                            levels = data_base_names)
test_enrichment = plotDF_df
plot_list = ggplot(data = plotDF_df, 
                   aes(x = dataset, y = pathway,
                       size = log10.sig.val, col = direction)) +
  geom_point(color = "white") + 
  geom_point(shape = 1) + 
  geom_point(data =. %>% dplyr::filter(log10.sig.val > -log10(thr))) + 
  ggtitle(plot_title) + 
  theme_classic() +
  theme(panel.grid.major = element_line(color="grey", size=0.1), 
        axis.text.x = element_text(angle = 90, hjust=1, vjust=0.5, 
                                    size = 15, color = "black"), 
        axis.text.y = element_text(size = 15, color = "black"),
        strip.text = element_text(color = "black", size = 12.5),
        plot.title = element_text(color = "black", size = 15),
        legend.title = element_text(color = "black", size = 15),
        legend.text = element_text(color = "black", size = 15)) + 
  ylab("") + 
  xlab("") +
  scale_color_manual(values = c("neg" ="#284AA0", "pos" ="#BC4A4A",
                                "unsigned"= "#808080"),
                      labels = c("negative", "positive", "unsigned"),
                      name = "Direction") +
  facet_grid(database~., scales = "free_y", space = "free_y") +
  guides(size = guide_legend(title = expression("Signed "*log[10]*"(p.val)"),
                             order = 1))
plot_list
```

# Core enrichment table
```{r core enrichment table}
library(openxlsx)
subpath_fname = rownames(aggregated_mat_short[aggregated_mat_short$database == "subpath",])
table_subpath = lapply(names(enrichment_gsea_list[["subpath"]]), function(name) {
  tmp <- enrichment_gsea_list[["subpath"]][[name]]
  tmp <- tmp[, c("NES", "pvalue", "p.adjust", "core_enrichment")]
  tmp <- cbind(ID = rownames(tmp), omic = name, tmp)
})
table_subpath = do.call(rbind, table_subpath)
table_subpath = do.call(rbind, lapply(subpath_fname, function(id) {
  subset(table_subpath, ID == id)
}))

kegg_fname = rownames(aggregated_mat_short[aggregated_mat_short$database == "kegg",])
table_kegg = lapply(names(enrichment_gsea_list[["kegg"]]), function(name) {
  tmp <- enrichment_gsea_list[["kegg"]][[name]]
  tmp <- tmp[, c("NES", "pvalue", "p.adjust", "core_enrichment")]
  tmp <- cbind(ID = rownames(tmp), omic = name, tmp)
})
table_kegg = do.call(rbind, table_kegg)
table_kegg = do.call(rbind, lapply(kegg_fname, function(id) {
  subset(table_kegg, ID == id)
}))

hallmark_fname = rownames(aggregated_mat_short[aggregated_mat_short$database == "hallmark",])
table_hallmark = lapply(names(enrichment_gsea_list[["hallmark"]]), function(name) {
  tmp <- enrichment_gsea_list[["hallmark"]][[name]]
  tmp <- tmp[, c("NES", "pvalue", "p.adjust", "core_enrichment")]
  tmp <- cbind(ID = rownames(tmp), omic = name, tmp)
})
table_hallmark = do.call(rbind, table_hallmark)
table_hallmark = do.call(rbind, lapply(hallmark_fname, function(id) {
  subset(table_hallmark, ID == id)
}))
```

# Filtering
```{r filtering}
pathway_mat = aggregated_mat_short[aggregated_mat_short$database == "kegg",]
# all(diff(pathway_mat$joint_adj) <= 0)
# all(diff(pathway_mat$joint) <= 0)
pathway_mat = pathway_mat[,1:6]
sig_feature = rownames(pathway_mat)
disease_name = sig_feature[grep("disease", sig_feature, ignore.case = FALSE)]
other_feature = sig_feature[!(sig_feature %in% disease_name)]
# move disease related pathways to later ranking
sig_feature = c(other_feature, disease_name)
pathway_mat = pathway_mat[sig_feature,]
omics_id = apply(pathway_mat, 1, function(x) {
  valid_idx = which(abs(x) > -log10(0.05))
  if (length(valid_idx) > 0) {
    return(colnames(pathway_mat)[valid_idx])
  } else {
    return(NA)
  }
})
omics_id = lapply(names(omics_id), function(pathway_name) {
  data.frame(pathway = pathway_name, omics = omics_id[[pathway_name]], 
             stringsAsFactors = FALSE)
})
omics_id = do.call(rbind, omics_id)
unique_omic_names = unique(omics_id$omics)

core_enrichment = data.frame()
for (database_name in names(enrichment_gsea_list)) {
  for (omic_name in names(enrichment_gsea_list[[database_name]])) {
    tmp = as.data.frame(enrichment_gsea_list[[database_name]][[omic_name]]@result[, c("ID", "core_enrichment")])
    tmp$database = database_name
    tmp$omics = omic_name
    core_enrichment = rbind(core_enrichment, tmp)
  }
}
core_enrichment = core_enrichment[core_enrichment$ID %in% omics_id$pathway, ]

similarity_mat = data.frame(omics_id)
similarity_mat$similarity = NA
for (k in 1:length(unique_omic_names)) {
  omic_name = unique_omic_names[k]
  fnames = omics_id$pathway[omics_id$omics == omic_name]
  current_set = c()
  for (i in 1:length(fnames)) {
    enrich = core_enrichment[core_enrichment$ID == fnames[i] &
                               core_enrichment$omics == omic_name,]
    core_list = unlist(strsplit(enrich$core_enrichment, "/"))
    if (i != 1) {
      similarity_mat$similarity[similarity_mat$pathway == fnames[i] &
                                  similarity_mat$omics == omic_name] =
        length(intersect(core_list, current_set))/length(core_list)
      current_set = unique(c(current_set, core_list))
    } else {
      similarity_mat$similarity[similarity_mat$pathway == fnames[i] &
                                  similarity_mat$omics == omic_name] = 0
      current_set = unique(core_list)
    }
  }
}
similarity_mat = similarity_mat %>%
  group_by(pathway) %>%
  slice(which.min(similarity))
similarity_mat = similarity_mat[match(sig_feature, similarity_mat$pathway), ]
```

# Enrichment plot (Extended Data Fig7D)
```{r enrichment plot, fig.height=16, fig.width=10}
similarity_mat$keep_pathway = ifelse(similarity_mat$similarity<0.5, TRUE, FALSE)
remove_pathway = similarity_mat$pathway[similarity_mat$keep_pathway == FALSE]
plotDF_df_sub = plotDF_df[!(plotDF_df$pathway %in% remove_pathway), ]

plot_list_sub = ggplot(data = plotDF_df_sub, 
                   aes(x = dataset, y = pathway,
                       size = log10.sig.val, col = direction)) +
  geom_point(color = "white") + 
  geom_point(shape = 1) + 
  geom_point(data =. %>% dplyr::filter(log10.sig.val > -log10(thr))) + 
  ggtitle(plot_title) + 
  theme_classic() +
  theme(panel.grid.major = element_line(color="grey", size=0.1), 
        axis.text.x = element_text(angle = 90, hjust=1, vjust=0.5, 
                                    size = 15, color = "black"), 
        axis.text.y = element_text(size = 15, color = "black"),
        strip.text = element_text(color = "black", size = 12.5),
        plot.title = element_text(color = "black", size = 15),
        legend.title = element_text(color = "black", size = 15),
        legend.text = element_text(color = "black", size = 15)) + 
  ylab("") + 
  xlab("") +
  scale_color_manual(values = c("neg" ="#284AA0", "pos" ="#BC4A4A",
                                "unsigned"= "#808080"),
                      labels = c("negative", "positive", "unsigned"),
                      name = "Direction") +
  facet_grid(database~., scales = "free_y", space = "free_y") +
  guides(size = guide_legend(title = expression("Signed "*log[10]*"(p.val)"),
                             order = 1))
plot_list_sub
```

# Complement and coagulation cascades - ppgd
```{r CC - ppgd}
pathway_mat = aggregated_mat_short[aggregated_mat_short$database == "kegg",]
pathway_mat = pathway_mat[,1:6]
mod_id = 30
pathways0 = c("Complement and coagulation cascades", "Ribosome")

ppgd_list = unique(kegg_database$ppgd[kegg_database$ppgd$description%in%pathways0, ])
flist = rbind(ppgd_list)

weight_comparisons = list()
for(dataset_id in c(1:6)){
  temp = intersect(rownames(projection_coef$mod1[[dataset_id]]), flist$gene_name)
  weight_comparisons[[dataset_id]] = data.frame(feature = temp)
  for(mod_id in c(30)){
    weight_comparisons[[dataset_id]][[paste0("Factor", mod_id)]] =
      projection_coef[[mod_id]][[dataset_id]][temp,]
  }
  clist = c()
  for(gene in weight_comparisons[[dataset_id]][["feature"]]) {
    temp2 = flist[flist$gene_name == gene,]
    colnames(temp2) = c("pathway", "feature")
    clist = rbind(clist, temp2)
  }
  if (!is.null(clist)) {
     weight_comparisons[[dataset_id]] =
       left_join(clist, weight_comparisons[[dataset_id]], by = "feature")
  }
}
names(weight_comparisons) = names(projection_coef$mod1)

# ppgd
dataDF = organized_data$datasets$plasma_proteomics_global_dda
colnames(dataDF) = rownames(projection_coef$mod1$ppgd)
thrs = matrix(0, ncol = 1, nrow = length(pathways0))
thrs[,1] = sign(pathway_mat[pathways0, "ppgd"])
pathway_score_ppgd = NULL
pathways1 = pathways0
i=0
for(mod_id in c(30)){
  i=i+1
  for(j in 1:length(pathways1)){
    temp1 = weight_comparisons$ppgd[weight_comparisons$ppgd$pathway==pathways1[j],]
    temp1 = temp1[sign(temp1[,paste0("Factor",mod_id)]) == thrs[j,i],]
    temp2 = dataDF[,(temp1$feature)]
    temp3 = as.matrix(temp2) %*% abs(temp1[[paste0("Factor",mod_id)]])
    temp3 = data.frame(value = temp3)
    colnames(temp3)= pathways1[j]
    if(is.null(pathway_score_ppgd)){
     pathway_score_ppgd = temp3
    }else{
      pathway_score_ppgd = cbind(pathway_score_ppgd, temp3)
    }
  }
}
ll = apply(pathway_score_ppgd,2,var, na.rm = T)
pathway_score_ppgd = pathway_score_ppgd[,ll>0]
ii = !is.na(pathway_score_ppgd[,1])
pathway_score_ppgd[ii,] = apply(pathway_score_ppgd[ii,], 2,
                                function(z) qqnorm(z, plot.it = F)$x)
pathway_score_ppgd$event_id = rownames(dataDF)
```

# Long Chain Polyunsaturated Fatty Acid - pmg
```{r LCPUFA - pmg}
pathway_mat = aggregated_mat_short[aggregated_mat_short$database == "subpath",]
pathway_mat = pathway_mat[,1:6]
mod_id = 30
pathways0 = c("Long Chain Polyunsaturated Fatty Acid (n3 and n6)", "Lysophospholipid")

pmg_list = unique(subpath_databse$pmg[subpath_databse$pmg$description%in%pathways0, ])
colnames(pmg_list) = c("description", "gene_name")
flist = rbind(pmg_list)

weight_comparisons = list()
for(dataset_id in c(1:6)){
  temp = intersect(rownames(projection_coef$mod1[[dataset_id]]), flist$gene_name)
  weight_comparisons[[dataset_id]] = data.frame(feature = temp)
  for(mod_id in c(30)){
    weight_comparisons[[dataset_id]][[paste0("Factor", mod_id)]] =
      projection_coef[[mod_id]][[dataset_id]][temp,]
  }
  clist = c()
  for(gene in weight_comparisons[[dataset_id]][["feature"]]) {
    temp2 = flist[flist$gene_name == gene,]
    colnames(temp2) = c("pathway", "feature")
    clist = rbind(clist, temp2)
  }
  if (!is.null(clist)) {
     weight_comparisons[[dataset_id]] =
       left_join(clist, weight_comparisons[[dataset_id]], by = "feature")
  }
}
names(weight_comparisons) = names(projection_coef$mod1)

# pmg
dataDF = organized_data$datasets$plasma_metabolomics_global
colnames(dataDF) = rownames(projection_coef$mod1$pmg)
thrs = matrix(0, ncol = 1, nrow = length(pathways0))
thrs[,1] = sign(pathway_mat[pathways0, "pmg"])
pathway_score_pmg = NULL
pathways1 = pathways0
i=0
for(mod_id in c(30)){
  i=i+1
  for(j in 1:length(pathways1)){
    temp1 = weight_comparisons$pmg[weight_comparisons$pmg$pathway==pathways1[j],]
    temp1 = temp1[sign(temp1[,paste0("Factor",mod_id)]) == thrs[j,i],]
    temp2 = dataDF[,(temp1$feature)]
    temp3 = as.matrix(temp2) %*% abs(temp1[[paste0("Factor",mod_id)]])
    temp3 = data.frame(value = temp3)
    colnames(temp3)= pathways1[j]
    if(is.null(pathway_score_pmg)){
     pathway_score_pmg = temp3
    }else{
      pathway_score_pmg = cbind(pathway_score_pmg, temp3)
    }
  }
}
ll = apply(pathway_score_pmg,2,var, na.rm = T)
pathway_score_pmg = pathway_score_pmg[,ll>0]
ii = !is.na(pathway_score_pmg[,1])
pathway_score_pmg[ii,] = apply(pathway_score_pmg[ii,], 2,
                                function(z) qqnorm(z, plot.it = F)$x)
pathway_score_pmg$event_id = rownames(dataDF)
```

# CC and LCPUFA PASC comparison (Extended Data Fig7E)
```{r, fig.height=2.5, fig.width=8.5}
# ppgd
assay_mat = pathway_score_ppgd
hallmark_feature = colnames(assay_mat)
hallmark_feature = setdiff(hallmark_feature, "event_id")
rownames(assay_mat) = assay_mat$event_id
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl = subtype_hcl,
                                 clin_data = clin_data_aligned)
assay_mat = assay_mat[!is.na(assay_mat$PASC),]
assay_mat$PASC = ifelse(assay_mat$PASC == "minimal", 
                         "Minimal Deficit", "Other Deficits")

ppgd_test = list()
feature_name = hallmark_feature
ppgd_test[["early"]] = two_pasc_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V1", "V2", "V3"))
ppgd_test[["late"]] = two_pasc_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V4", "V5", "V6"))

# pmg
assay_mat = pathway_score_pmg
hallmark_feature = colnames(assay_mat)
hallmark_feature = setdiff(hallmark_feature, "event_id")
rownames(assay_mat) = assay_mat$event_id
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl = subtype_hcl,
                                 clin_data = clin_data_aligned)
assay_mat = assay_mat[!is.na(assay_mat$PASC),]
assay_mat$PASC = ifelse(assay_mat$PASC == "minimal", 
                         "Minimal Deficit", "Other Deficits")

pmg_test = list()
feature_name = hallmark_feature
pmg_test[["early"]] = two_pasc_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V1", "V2", "V3"))
pmg_test[["late"]] = two_pasc_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V4", "V5", "V6"))

est_mat = rbind(
  c(ppgd_test$early$est["Complement and coagulation cascades"], NA,
    ppgd_test$late$est["Complement and coagulation cascades"], NA),
  c(NA, pmg_test$early$est["Long Chain Polyunsaturated Fatty Acid (n3 and n6)"],
    NA, pmg_test$late$est["Long Chain Polyunsaturated Fatty Acid (n3 and n6)"]))
colnames(est_mat) = c("ppgd", "pmg", "ppgd", "pmg")
rownames(est_mat) = c("Complement and coagulation cascades",
                      "Long Chain Polyunsaturated Fatty Acid\n(n3 and n6)")
pval_mat = rbind(
  c(ppgd_test$early$pval["Complement and coagulation cascades"], NA,
    ppgd_test$late$pval["Complement and coagulation cascades"], NA),
  c(NA, pmg_test$early$pval["Long Chain Polyunsaturated Fatty Acid (n3 and n6)"],
    NA, pmg_test$late$pval["Long Chain Polyunsaturated Fatty Acid (n3 and n6)"]))
colnames(pval_mat) = c("ppgd", "pmg", "ppgd", "pmg")
rownames(est_mat) = c("Complement and coagulation cascades",
                      "Long Chain Polyunsaturated Fatty Acid\n(n3 and n6)")
test_type = factor(colnames(est_mat), 
                   levels = c("ppgd", "pmg"))
colnames(est_mat) = rep(c("V1-V3","V4-V6"), each = 2)
colnames(pval_mat) = rep(c("V1-V3","V4-V6"), each = 2)

select_pathway_comparison = ComplexHeatmap::Heatmap(
  as.matrix(est_mat), 
  col = circlize::colorRamp2(c(-max(abs(est_mat), na.rm = TRUE), 
                               0, max(abs(est_mat), na.rm = TRUE)),
                             c("#284AA0", "white", "#BC4A4A")),
  show_heatmap_legend = TRUE, na_col = "grey90", border=T, 
  column_title_gp = gpar(fontsize = 15), 
  border_gp = gpar(col = "darkgrey", lwd = 0.5), 
  rect_gp = gpar(col = "darkgrey", lwd = 0.5),
  row_names_gp = gpar(fontsize = 15), 
  column_names_gp = gpar(fontsize = 15),
  heatmap_legend_param = list(
    title = "Coefficient",
    direction = "vertical", title_position = "leftcenter-rot",
    title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
  column_split = test_type, cluster_columns = F,
  width = unit(60, "mm"),
  cluster_rows = F, cell_fun = function(j, i, x, y, w, h, fill) {
  if(!is.na(pval_mat[i, j]) & pval_mat[i, j] < 0.001) {
    grid.text("***", x, y-h*0.1, gp = gpar(cex=2))
  }else if(!is.na(pval_mat[i, j]) & pval_mat[i, j] < 0.01) {
    grid.text("**", x, y-h*0.1, gp = gpar(cex=2))
  }else if(!is.na(pval_mat[i, j]) & pval_mat[i, j] < 0.05) {
    grid.text("*", x, y-h*0.1, gp = gpar(cex=2))
  }
})
ComplexHeatmap::draw(select_pathway_comparison, heatmap_legend_side = "left")
```

# PBMC-Nasal opposite inflammation components (Extended Data Fig7G)
```{r PBMC-Nasal comparison, fig.height=6, fig.width=10}
select_genes = unique(hallmark_database$pbmc$gene_name[hallmark_database$pbmc$description %in% c("HALLMARK_INTERFERON_ALPHA_RESPONSE", "HALLMARK_INTERFERON_GAMMA_RESPONSE", "HALLMARK_TNFA_SIGNALING_VIA_NFKB")])
select_genes = intersect(intersect(select_genes,
                                   rownames(projection_coef$mod30$nt)),
                         rownames(projection_coef$mod30$pbmc))
cor.test(projection_coef$mod30$nt[select_genes,], projection_coef$mod30$pbmc[select_genes,])
temp_coef = data.frame(gene = select_genes,
                       "PBMC" = projection_coef$mod30$pbmc[select_genes,],
                       "nt" = projection_coef$mod30$nt[select_genes,])
temp_coef$diff = temp_coef$PBMC-temp_coef$nt

thr = 0.075
bottom_right = temp_coef[temp_coef$PBMC>0 & temp_coef$diff>0,]
bottom_right = bottom_right[bottom_right$diff > thr,]
bottom_right = bottom_right[order(-bottom_right$diff),]
top_left = temp_coef[temp_coef$nt>0 & temp_coef$diff<0,]
top_left = top_left[top_left$diff < -thr,]
labeled_genes = rbind(bottom_right, top_left)
temp_coef$col<- ifelse(temp_coef$gene %in% labeled_genes$gene, "labeled", "not_labeled")
max_lim <- max(abs(c(temp_coef$PBMC, temp_coef$nt)))
max_lim1 <- max(abs(temp_coef$PBMC))
max_lim2 <- max(abs(temp_coef$nt))

ggplot(temp_coef, aes(x = PBMC, y = nt)) +
  geom_point(aes(color = col, shape = col, size = col)) +
  scale_size_manual(values = c("labeled" = 2, "not_labeled" = 0.4)) +
  scale_shape_manual(values = c("labeled" = 16, "not_labeled" = 1)) +
  scale_color_manual(values = c("labeled" = "#284AA0", "not_labeled" = "grey40")) +
    geom_text(data = labeled_genes, aes(label = gene),
            nudge_x = 0, nudge_y = 0.003, size = 3.5, color = "#284AA0") +
  labs(x = "PBMC Coefficient", y = "NT Coefficient", 
       title = "Coefficient Comparison Between PBMC and Nasal Transcriptomics")+
  xlim(-max_lim1, max_lim1) +
  ylim(-max_lim2, max_lim2) +
  expand_limits(x = 0, y = 0) + 
  geom_hline(yintercept = 0, color = "black", size = 0.5) +
  geom_vline(xintercept = 0, color = "black", size = 0.5) +
  theme_minimal()+
  theme(panel.grid = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(size = 15, color = "black"),
        axis.title.y = element_text(size = 15, color = "black"),
        axis.text.x = element_text(size = 15, color = "black"),
        axis.text.y = element_text(size = 15, color = "black"),
        axis.line.x = element_line(linewidth = 1),
        axis.line.y = element_line(linewidth = 1),
        title = element_text(size = 15, color = "black"))
```

# Assay correlation with factor 30 (Extended Data Fig7B left)
```{r assay correlation with factor 30, fig.height=6, fig.width=3.5}
mod_id = 30
factor_level = mcia_results$mcia_factors[, mod_id]
assay_cor = matrix(NA, nrow = 6, ncol = 2)
rownames(assay_cor) = names(projection_coef[[mod_id]])
colnames(assay_cor) = c("cor", "pval")
assay_weighted_list = NULL
for (dataset_id in c(1:6)) {
  temp = projection_coef[[mod_id]][[dataset_id]]
  assay_weighted = as.matrix(organized_data$datasets[[dataset_id]]) %*% temp[, 1]
  cor_test = cor.test(assay_weighted, factor_level)
  assay_cor[dataset_id, "cor"] = cor_test$estimate
  assay_cor[dataset_id, "pval"] = cor_test$p.value
  assay_weighted_list = cbind(assay_weighted_list, assay_weighted)
}
colnames(assay_weighted_list) = paste0("weighted_", names(projection_coef[[mod_id]]))

df_assay_cor = data.frame(assay = names(projection_coef[[mod_id]]),
                          corr = assay_cor[, "cor"])
df_assay_cor$assay = factor(df_assay_cor$assay, 
                            levels = names(projection_coef[[mod_id]]))

ggplot(df_assay_cor, aes(x = assay, y = corr)) +
  geom_bar(stat = "identity", fill = "#BC4A4A") +
  coord_flip() +
  labs(x = "Assay", y = "Correlation with Factor 30") +
  theme_classic() +
  scale_y_continuous(breaks = c(0, 0.5, 1), limits = c(0, 1)) +
  theme(axis.title.x = element_text(size = 15, color = "black"),
        axis.title.y = element_text(size = 15, color = "black"),
        axis.text.x = element_text(size = 15, color = "black"),
        axis.text.y = element_text(size = 15, color = "black"),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        axis.line.x = element_line(linewidth = 1),
        axis.line.y = element_line(linewidth = 1))
```

# Hallmark pathways - pbmc and nt
```{r hallmark pathways - pbmc and nt}
pathway_mat = aggregated_mat_short[aggregated_mat_short$database == "hallmark",]
pathway_mat = pathway_mat[,1:6]
mod_id = 30
pathways0 = rownames(pathway_mat)
hallmark_feature = rownames(pathway_mat)

pbmc_list = unique(hallmark_database$pbmc[hallmark_database$pbmc$description%in%pathways0, ])
flist = rbind(pbmc_list)

weight_comparisons = list()
for(dataset_id in c(1:6)){
  temp = intersect(rownames(projection_coef$mod1[[dataset_id]]), flist$gene_name)
  weight_comparisons[[dataset_id]] = data.frame(feature = temp)
  for(mod_id in c(30)){
    weight_comparisons[[dataset_id]][[paste0("Factor", mod_id)]] =
      projection_coef[[mod_id]][[dataset_id]][temp,]
  }
  clist = c()
  for(gene in weight_comparisons[[dataset_id]][["feature"]]) {
    temp2 = flist[flist$gene_name == gene,]
    colnames(temp2) = c("pathway", "feature")
    clist = rbind(clist, temp2)
  }
  if (!is.null(clist)) {
     weight_comparisons[[dataset_id]] =
       left_join(clist, weight_comparisons[[dataset_id]], by = "feature")
  }
}
names(weight_comparisons) = names(projection_coef$mod1)

# pbmc
dataDF = organized_data$datasets$pbmc_transcriptomics
colnames(dataDF) = rownames(projection_coef$mod1$pbmc)
thrs = matrix(0, ncol = 1, nrow = length(pathways0))
thrs[,1] = sign(pathway_mat[pathways0, "pbmc"])
pathway_score_pbmc = NULL
pathways1 = pathways0
i=0
for(mod_id in c(30)){
  i=i+1
  for(j in 1:length(pathways1)){
    temp1 = weight_comparisons$pbmc[weight_comparisons$pbmc$pathway==pathways1[j],]
    temp1 = temp1[sign(temp1[,paste0("Factor",mod_id)]) == thrs[j,i],]
    temp2 = dataDF[,(temp1$feature)]
    temp3 = as.matrix(temp2) %*% abs(temp1[[paste0("Factor",mod_id)]])
    temp3 = data.frame(value = temp3)
    colnames(temp3)= pathways1[j]
    if(is.null(pathway_score_pbmc)){
     pathway_score_pbmc = temp3
    }else{
      pathway_score_pbmc = cbind(pathway_score_pbmc, temp3)
    }
  }
}
ll = apply(pathway_score_pbmc,2,var, na.rm = T)
pathway_score_pbmc=pathway_score_pbmc[,ll>0]
ii = !is.na(pathway_score_pbmc[,1])
pathway_score_pbmc[ii,] = apply(pathway_score_pbmc[ii,], 2,
                                function(z) qqnorm(z, plot.it = F)$x)

# nt
dataDF = organized_data$datasets$nasal_transcriptomics
colnames(dataDF) = rownames(projection_coef$mod1$nt)
thrs = matrix(0, ncol = 1, nrow = length(pathways0))
thrs[,1] = sign(pathway_mat[pathways0, "nt"])
pathway_score_nt = NULL
pathways1 = pathways0
i=0
for(mod_id in 30){
  i=i+1
  for(j in 1:length(pathways1)){
    temp1 = weight_comparisons$nt[weight_comparisons$nt$pathway==pathways1[j],]
    temp1 = temp1[sign(temp1[,paste0("Factor",mod_id)]) == thrs[j,i],]
    temp2 = dataDF[,(temp1$feature)]
    temp3 = as.matrix(temp2)%*%abs(temp1[[paste0("Factor",mod_id)]])
    temp3 = data.frame(value = temp3)
    colnames(temp3)= pathways1[j]
    if(is.null(pathway_score_nt)){
     pathway_score_nt = temp3 
    }else{
      pathway_score_nt = cbind(pathway_score_nt, temp3)
    }
  }
}
ll = apply(pathway_score_nt,2,var, na.rm = T)
pathway_score_nt=pathway_score_nt[,ll>0]
ii = !is.na(pathway_score_nt[,1])
pathway_score_nt[ii,] = apply(pathway_score_nt[ii,], 2,
                                function(z) qqnorm(z, plot.it = F)$x)
```

# Hallmark correlation with Factor 30 (Extended Data Fig7F)
```{r hallmark correlation with Factor 30, fig.height=6, fig.width=6.5}
mod_id = 30
factor_level = mcia_results$mcia_factors[, mod_id]
hallmark_cor = matrix(NA, nrow = length(hallmark_feature), ncol = 2)
rownames(hallmark_cor) = hallmark_feature
colnames(hallmark_cor) = c("pbmc", "nt")
hallmark_pval = matrix(NA, nrow = length(hallmark_feature), ncol = 2)
rownames(hallmark_pval) = hallmark_feature
colnames(hallmark_pval) = c("pbmc", "nt")

for (hm_id in 1:length(hallmark_feature)) {
  h1 = pathway_score_pbmc[, hm_id]
  h2 = pathway_score_nt[, hm_id]
  cor_test1 = cor.test(h1, factor_level)
  cor_test2 = cor.test(h2, factor_level)
  hallmark_cor[hm_id, "pbmc"] = cor_test1$estimate
  hallmark_cor[hm_id, "nt"] = cor_test2$estimate
  hallmark_pval[hm_id, "pbmc"] = cor_test1$p.value
  hallmark_pval[hm_id, "nt"] = cor_test2$p.value
}
hallmark_adj_pval = apply(hallmark_pval, 2, 
                          function(p) p.adjust(p, method = "BH"))
heatmap_hallmark_cor(hallmark_cor, hallmark_adj_pval, hallmark_feature)
```

# Weighted serum olink trajectory (Extended Data Fig7C left)
```{r weighted serum olink trajectory, fig.height=6, fig.width=6}
assay_mat = as.data.frame(assay_weighted_list)
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl = subtype_hcl,
                                 clin_data = clin_data_aligned)
assay_mat = assay_mat[!is.na(assay_mat$PASC),]
assay_mat$PASC = ifelse(assay_mat$PASC == "minimal", 
                         "Minimal Deficit", "Other Deficits")
clin_colnames_select = setdiff(colnames(assay_mat), colnames(assay_weighted_list))
assay_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "PASC", use_endpoint_levels = c("Minimal Deficit", "Other Deficits"),
  subgroup_cr = c("Other Deficits" = "#9C3418", "Minimal Deficit" = "#39828C"),
  max_date = 40, legend_name = "PASC", multiple_correction = FALSE)
assay_trajectory_analysis_results$test_fig_list$weighted_so
```

# Early/late weighted so two-group comparison (Extended Data Fig7C right)
```{r early/late weighted so two-group comparison, fig.height=6, fig.width=3}
assay_mat = as.data.frame(assay_weighted_list)
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl = subtype_hcl,
                                 clin_data = clin_data_aligned)
assay_mat = assay_mat[!is.na(assay_mat$PASC),]
assay_mat$PASC = ifelse(assay_mat$PASC == "minimal", 
                         "Minimal Deficit", "Other Deficits")
so_test = list()
feature_name = c("weighted_so")
so_test[["early"]] = two_pasc_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V1", "V2", "V3"))
so_test[["late"]] = two_pasc_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V4", "V5", "V6"))

# Factor 30
est_mat = cbind(so_test$early$est["weighted_so"],
                so_test$late$est["weighted_so"])
rownames(est_mat) = "Any vs Minimal Deficit"
colnames(est_mat) = c("V1-V3", "V4-V6")
pval_mat = cbind(so_test$early$pval["weighted_so"],
                 so_test$late$pval["weighted_so"])
colnames(pval_mat) = c("V1-V3", "V4-V6")
adj_p = matrix(p.adjust(pval_mat, method = "BH"), nrow = 1)

weighted_so_comparison = ComplexHeatmap::Heatmap(
  as.matrix(t(est_mat)), row_names_gp = gpar(fontsize = 15), 
  column_names_gp = gpar(fontsize = 15, lineheight = 1.5), 
  column_names_rot = 0,
  column_names_max_height = unit(6, "cm"),
  column_names_centered = T, 
  heatmap_legend_param = list(title = "Coefficient", direction = "horizontal", 
                              title_position = "topcenter",
                              title_gp = gpar(fontsize = 15), 
                              labels_gp = gpar(fontsize = 15)),
  col = circlize::colorRamp2(c(-max(abs(est_mat)), 0, max(abs(est_mat))), 
                             c("#284AA0", "white", "#BC4A4A")), 
  na_col = "grey90", cluster_columns = F, cluster_rows = F,
  cell_fun = function(j, i, x, y, w, h, fill) {
    if(t(pval_mat)[i, j] < 0.001) {
      grid.text("***", x, y-h*0.08, gp = gpar(cex=4))
      } else if(t(pval_mat)[i, j] < 0.01) {
        grid.text("**", x, y-h*0.08, gp = gpar(cex=4))
        } else if(t(pval_mat)[i, j] < 0.05) {
          grid.text("*", x, y-h*0.08, gp = gpar(cex=4))
          }
    })
ComplexHeatmap::draw(weighted_so_comparison, heatmap_legend_side = "top")
```

# Top cytokine coefficient (Extended Data Fig7B right)
```{r top cytokine coefficient, fig.height=6, fig.width=4.2}
mod_id = 30; coef_thr = 0.1
coef_mat = as.matrix(projection_coef[[mod_id]][["so"]])
coef_mat = coef_mat[abs(coef_mat) > coef_thr,]
plotDF = data.frame(feature = names(coef_mat),
                    Factor30 = coef_mat)
plotDF = plotDF[order(plotDF$Factor30, decreasing = TRUE),]
feature_levels = plotDF$feature
plotDF$feature = factor(plotDF$feature, 
                        levels = feature_levels[length(feature_levels):1])

plotDF = gather(plotDF, key = "Factor", value = "coefficient", -feature)
plotDF$abs_coef = abs(plotDF$coefficient)
plotDF$sign = ifelse(plotDF$coefficient > 0, "Positive", "Negative")

ggplot(plotDF, aes(x = coefficient, y = feature, group = feature, color = sign)) +
  geom_segment(aes(xend = 0, yend =  feature), linewidth = 1.5) +
  geom_point(aes(x = coefficient), size = 3) +
  scale_color_manual(values = c("Positive" = "#BC4A4A", 
                                "Negative" = "#284AA0")) +
  labs(x = "Absolute Value of Coefficient", y = "Feature", 
       title = "") +
  theme_minimal()  +
  geom_vline(xintercept = 0, linetype = "solid") +
  geom_hline(yintercept = -Inf, linetype = "solid") +
  theme(axis.ticks.x = element_line(linewidth = 0.5),
        axis.line.x = element_line(linewidth = 1),
        axis.line.y = element_line(linewidth = 1),
        axis.ticks = element_line(linewidth = 1),
        axis.text.x = element_text(color = "black", size = 15),  
        axis.text.y = element_text(color = "black", size = 15),  
        axis.title = element_text(color = "black", size = 15),
        axis.ticks.length = unit(0.15, "cm"),
        plot.subtitle = element_text(color = "black", hjust = 0.5, 
                                     size = 15),
        legend.text = element_text(color = "black", size = 15),
        legend.title = element_text(color = "black", size = 15)) +
  scale_x_continuous(labels = function(x) abs(x), limits = c(-0.3, 0.3))+
  theme(legend.position = "bottom")
```



