---
title: "Blood CyTOF univariate analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "RColorBrewer"))
suppressPackageStartupMessages(library(package = "ComplexHeatmap"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "scales"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "tidyverse"))

# set session options
options(show_col_types         = FALSE,
        dplyr.summarise.inform = FALSE)
VarCorr <- lme4::VarCorr # for PVCA
```

### Load codebase.R, load data matrices and set path to a local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# load codebase.R, olink data and clinical information
source(file = "../../Codebase/codebase_v2.R")
data_env <- load_IMPACC_datasets(DATA_VERSION                = "2022-01-01", 
                                 KEEP_COVID19_POS            = TRUE, 
                                 FILTER_BY_CORE_ASSAY_COHORT = TRUE,
                                 ASSAY_NAME = "bld_cytof")
for (n in grep(pattern = "bld_cytof|clinical_data", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}

# Will be used later in the code.
data_env_withHC <- load_IMPACC_datasets(DATA_VERSION                = "2022-01-01", 
                                        KEEP_COVID19_POS            = FALSE, 
                                        FILTER_BY_CORE_ASSAY_COHORT = FALSE,
                                        ASSAY_NAME = "bld_cytof")


# Setup the assay specific output directory
out_dir <- "../output"
if(!dir.exists(out_dir))
  dir.create(out_dir)


### Misc
# MFI of blood CyTOF annotation training samples
mfiFile <- "impacc_mfi_heatmap.csv"

# Define the outlier samples, if any, that should be removed.
outlier_sample_ids <- 
  rownames(bld_cytof_metadata)[bld_cytof_metadata$sample_status %in%
                               "IMPACC sample assayed with questionable QC"]

```

### Functions: data processing
```{r data-processing}
# normalize parent on total cell counts
preprocess_bld_cytof_parent <- function(input_counts, input_rowfeature) {
  output_counts <-  input_counts %>%
    rownames_to_column(var = "sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "rowname") %>%
    filter(!sample_id %in% outlier_sample_ids) %>%
    merge(y = rownames_to_column(input_rowfeature), by = "rowname") %>%
    group_by(sample_id, `Broad Parent Population`) %>%
    summarize(value = sum(value)) %>%
    pivot_wider(names_from = `Broad Parent Population`) %>%
    select(-contains(match = "Debris"), 
           -contains(match = "Multiplet"),
           -`Platelets`, 
           -`RBC`, 
           -`Undefined`) %>%
    column_to_rownames(var = "sample_id")
  output_counts <- output_counts/rowSums(output_counts)

  return(value = output_counts)
}

# normalize children on total cell counts
#' @param input_counts data.frame of event counts for each samples (rows) per children subset (cols)
preprocess_bld_cytof_child <- function(input_counts) {
  output_counts <-  input_counts %>%
    rownames_to_column(var = "sample_id") %>%
    filter(!sample_id %in% outlier_sample_ids) %>%
    select(-contains(match = "Debris"), 
           -contains(match = "Multiplet"),
           -`Platelets`, 
           -`RBC`, 
           -`Tier1_Undefined`,
           -`Tier2_undefined_Undefined`) %>%
    column_to_rownames(var = "sample_id")
  output_counts <- output_counts/rowSums(output_counts)

  return(value = output_counts)
}

# normalize parent on non-granulocyte counts
preprocess_bld_cytof_parent_nongran <- function(input_counts, input_rowfeature) {
  output_counts <- input_counts %>%
    rownames_to_column(var = "sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "rowname") %>%
    filter(!sample_id %in% outlier_sample_ids) %>%
    merge(y = rownames_to_column(input_rowfeature), by = "rowname") %>%
    group_by(sample_id, `Broad Parent Population`) %>%
    summarize(value = sum(value)) %>%
    pivot_wider(names_from = `Broad Parent Population`) %>%
    select(-contains(match = "Debris"), 
           -contains(match = "Multiplet"),
           -`Platelets`, 
           -`RBC`, 
           -`Undefined`,
           -`Neutrophil`,
           - Eosinophil,
           -`Basophil`) %>%
    column_to_rownames(var = "sample_id")
  output_counts <- output_counts/rowSums(output_counts)
 
  return(value = output_counts)
} 

# normalize children on non-granulocyte counts
preprocess_bld_cytof_child_nongran <- function(input_counts) {
  output_counts <-  input_counts %>%
    rownames_to_column(var = "sample_id") %>%
    filter(!sample_id %in% outlier_sample_ids) %>%
    select(-contains(match = "Debris"), 
           -contains(match = "Multiplet"),
           -`Platelets`, 
           -`RBC`, 
           -`Tier1_Undefined`,
           -`Tier2_undefined_Undefined`,
           -`Neutrophil (CD16low)`,
           -`Neutrophil (CD16hi)`,
           -`Eosinophil`,
           -`Basophil`) %>%
    column_to_rownames(var = "sample_id")
  output_counts <- output_counts/rowSums(output_counts)
  
  return(value = output_counts)
}

# normalize child to parent population
#' @param input_counts data.frame of event counts for each samples (rows) per children subset (cols)
preprocess_bld_cytof_child_to_parent_ratio <- function(input_counts, input_rowfeature) {
  # step sum counts of parent pop
  input_parent_counts <- input_counts %>%
    rownames_to_column(var = "sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "rowname") %>%
    filter(!sample_id %in% outlier_sample_ids) %>%
    merge(y = rownames_to_column(input_rowfeature), by = "rowname") %>%
    group_by(sample_id, `Broad Parent Population`) %>%
    summarize(value = sum(value)) %>%
    pivot_wider(names_from = `Broad Parent Population`) %>%
    column_to_rownames(var = "sample_id")

  output_child_to_parent_ratio <- input_counts %>%
    rownames_to_column(var = "sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "rowname") %>%
    merge(y = rownames_to_column(input_rowfeature), by = "rowname")

  input_parent_counts %>%
    rownames_to_column(var = "sample_id") %>%
    pivot_longer(cols = -sample_id, names_to = "Broad Parent Population", values_to = "parent.value") %>%
    merge(x = output_child_to_parent_ratio, by = c("sample_id", "Broad Parent Population")) %>%
    mutate(value = (value + 0.0001)/(parent.value + 0.0001)) %>%
    select(sample_id, rowname, value) %>%
    pivot_wider(values_from = value, names_from = rowname) %>%
    select(-contains(match = "Debris"), 
           -contains(match = "Multiplet"),
           -`Platelets`, 
           -`RBC`, 
           -`Tier1_Undefined`,
           -`Tier2_undefined_Undefined`) %>%
    column_to_rownames(var = "sample_id") -> output_child_to_parent_ratio
  
  # dropping populations with only 1 child to parent
  output_child_to_parent_ratio <- output_child_to_parent_ratio[, apply(output_child_to_parent_ratio,
                                                                       MARGIN = 2,
                                                                       FUN = var) > 0]

  return(value = output_child_to_parent_ratio)
}
```

### Functions: data analysis
```{r data-analysis}
# subfunction use to fit ordinal regression and extract reg. coef
mixed_pairwise_coef <- function(my.formula0, my.formula1, data_use) {
  endpoints0 <- sort(unique(data_use$endpoints))
  pair_names <- c()
  res_table <- c()
  for(i in 1:(length(endpoints0)-1)){
    for(j in (i+1):(length(endpoints0))){
      pair_names <- c(pair_names, paste0(endpoints0[i],"|", endpoints0[j]))
      data_tmp <- data_use[data_use$endpoints %in% endpoints0[c(i, j)], ]
      fit <- lme4::lmer(my.formula1, data = data_tmp)
      res_table <- c(res_table, 
                     unique(coef(fit)$sites[,grep(pattern = "endpoint", 
                                               names(coef(fit)$sites))]))
    }
  }
  names(res_table) <- pair_names
  return(value = res_table)
}
```

### Panel A: analysis code
```{r panel-a-analysis, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# non-granulocytes parent counts
bld_cytof_counts_processed_parent_nongran <- 
  preprocess_bld_cytof_parent_nongran(bld_cytof_counts, bld_cytof_rowfeature)

data_use_parent_nongran_visit1 <- bld_cytof_counts_processed_parent_nongran %>%
  rownames_to_column(var = "sample_id") %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "+",
                     replacement = "pos", 
                     fixed       = TRUE)) %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "-",
                     replacement = "neg", 
                     fixed       = TRUE)) %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "\\(|\\)| ",
                     replacement = "_"
                     )) %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_type,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex),
        by = "sample_id") %>%
  filter(event_type %in% "Visit 1" &
         !is.na(trajectory_group)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  select(-event_type) %>%
  column_to_rownames(var = "sample_id")

# a model with only intercept and random effect across sites
res_table_ordinal <- NULL
data_use <- data_use_parent_nongran_visit1
for (varName in setdiff(names(data_use), names(clinical_data))) {
  my.formula0 <- paste0('trajectory_group~ 1+(1|enrollment_site)+',
                        'discretized_admit_age_quantile+sex')
  my.formula1 <- paste0('trajectory_group~',
                        varName,
                        '+(1|enrollment_site)+',
                        'discretized_admit_age_quantile+sex')
  res <- mixed_ordinal(my.formula0, 
                       my.formula1,
                       data_use = data_use)
  coef(clmm(formula(my.formula1), 
            data = data_use))[varName] %>%
    unname() %>%
    c(res, coef = .) -> res
  res_table_ordinal <- rbind(res_table_ordinal, res)
}
rownames(res_table_ordinal) <- setdiff(names(data_use), names(clinical_data))
res_table_ordinal <- res_table_ordinal %>%
  as.data.frame() %>%
  mutate(qval =  qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues)

res_table_ordinal_parent_nongran_visit1 <- res_table_ordinal
filter(res_table_ordinal_parent_nongran_visit1, qval <= 0.05) %>%
  mutate_if(is.numeric, signif, digits = 3) %>%
  kable()
```

### Panel A: output generation code
```{r panel-a-output-gen, fig.width=4, fig.height=4}
plotDF <- data_use_parent_nongran_visit1 %>%
  select(Hematopoetic_progrenitor_Cell, trajectory_group) %>%
  pivot_longer(cols = -trajectory_group)

fig1a <- ggplot(data    = plotDF,
                mapping = aes(x = trajectory_group, 
                              y = value)) +
  geom_beeswarm(cex = 2, size = 2, aes(color = trajectory_group)) +
  geom_boxplot(fill = "transparent", outlier.colour = "transparent") +
  labs(x = "Clinical trajectory groups",
       y = "Frequency of non-granulocytes (%)", tag = "A") +
  facet_wrap(facets = ~name, scale = "free", nrow = 1) +
  scale_color_manual(values= c("#639A21", "#39828C", "#6371AD", "#BD7D31", "#9C3418")) +
  theme_bw() +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(0,5), ylim = c(0, 0.01)) +
  scale_y_continuous(labels = percent)
print(fig1a)

pdf(file.path(out_dir, "main_A.pdf"), width=8, height = 8)
print(fig1a)
dev.off()
```

### Panel B: analysis code
```{r panel-b-analysis, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# ratio children to parent
bld_cytof_counts_processed_child_to_parent_ratio <-  
  preprocess_bld_cytof_child_to_parent_ratio(bld_cytof_counts,
                                             bld_cytof_rowfeature)

data_use_child_to_parent_ratio_visit1 <- bld_cytof_counts_processed_child_to_parent_ratio %>%
  rownames_to_column(var = "sample_id") %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "+",
                     replacement = "pos", 
                     fixed       = TRUE)) %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "-",
                     replacement = "neg", 
                     fixed       = TRUE)) %>%
  setNames(nm = gsub(x           = names(.), 
                     pattern     = "\\(|\\)| ",
                     replacement = "_"
                     )) %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_type,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex),
        by = "sample_id") %>%
  filter(event_type %in% "Visit 1" &
         !is.na(trajectory_group)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  select(-event_type) %>%
  column_to_rownames(var = "sample_id")

res_table_ordinal <- NULL
data_use <- data_use_child_to_parent_ratio_visit1
for (varName in setdiff(names(data_use), names(clinical_data))) {
  my.formula0 <- paste0('trajectory_group~ 1+(1|enrollment_site)+',
                        'discretized_admit_age_quantile+sex')
  my.formula1 <- paste0('trajectory_group~',
                        varName,
                        '+(1|enrollment_site)+',
                        'discretized_admit_age_quantile+sex')
  res <- mixed_ordinal(my.formula0, 
                       my.formula1,
                       data_use = data_use)
  coef(clmm(formula(my.formula1), 
            data = data_use))[varName] %>%
    unname() %>%
    c(res, coef = .) -> res
  res_table_ordinal <- rbind(res_table_ordinal, res)
}
rownames(res_table_ordinal) <- setdiff(names(data_use), names(clinical_data))
res_table_ordinal <- res_table_ordinal %>%
  as.data.frame() %>%
   mutate(qval =  qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues)

res_table_ordinal_child_to_parent_ratio_visit1 <- res_table_ordinal
filter(res_table_ordinal_child_to_parent_ratio_visit1, qval <= 0.05) %>%
  mutate_if(is.numeric, signif, digits = 3) %>%
  kable()
```

### Panel B: output generation code
```{r panel-b-output-gen, fig.width=8, fig.height=3}
plotDF <- data_use_child_to_parent_ratio_visit1 %>%
  select(matches(match = "^Monocytes"), trajectory_group) %>%
  pivot_longer(cols = -trajectory_group) %>%
  mutate(name = gsub(pattern = "__", replacement = "_", name),
         name = gsub(pattern = "_$", replacement = "", name),
         name = recode(name, Monocytes_CD14negCD16pos = "Monocytes_CD14dimCD16pos"),
         name = factor(name, levels =c("Monocytes_CD14posCD16neg",
                                       "Monocytes_CD14posCD16pos",
                                       "Monocytes_CD14dimCD16pos")))
fig1b1 <- ggplot(data = filter(plotDF, name == "Monocytes_CD14posCD16neg"),
       mapping = aes(x = trajectory_group, y = value)) +
  geom_beeswarm(cex = 2, size = 2, mapping = aes(color = trajectory_group)) +
  geom_boxplot(fill = "transparent", outlier.colour = "transparent") +
  labs(y = "Frequency of Monocytes (%)", x = "Clinical trajectory group") +
  facet_wrap(facets = ~name, scale = "free") +
  scale_color_manual(values= c("#639A21", "#39828C", "#6371AD", "#BD7D31", "#9C3418")) +
  scale_y_continuous(labels = percent) +
  theme_bw() +
  theme(legend.position = "none")

fig1b2 <- ggplot(data = filter(plotDF, name == "Monocytes_CD14posCD16pos"),
       mapping = aes(x = trajectory_group, y = value)) +
  geom_beeswarm(cex = 2, size = 2, mapping = aes(color = trajectory_group)) +
  geom_boxplot(fill = "transparent", outlier.colour = "transparent") +
  labs(y = NULL, x = "Clinical trajectory group") +
  facet_wrap(facets = ~name, scale = "free") +
  scale_color_manual(values= c("#639A21", "#39828C", "#6371AD", "#BD7D31", "#9C3418")) +
  scale_y_continuous(labels = percent) +
  coord_cartesian(ylim = c(0, 0.2)) +
  theme_bw() +
  theme(legend.position = "none")

fig1b3 <- ggplot(data = filter(plotDF, name == "Monocytes_CD14dimCD16pos"),
       mapping = aes(x = trajectory_group, y = value)) +
  geom_beeswarm(cex = 2, size = 2, mapping = aes(color = trajectory_group)) +
  geom_boxplot(fill = "transparent", outlier.colour = "transparent") +
  labs(y = NULL, x = "Clinical trajectory group") +
  facet_wrap(facets = ~name, scale = "free") +
  scale_color_manual(values= c("#639A21", "#39828C", "#6371AD", "#BD7D31", "#9C3418")) +
  scale_y_continuous(labels = percent) +
  coord_cartesian(ylim = c(0, 0.1)) +
  theme_bw() +
  theme(legend.position = "none")

ggarrange(fig1b1, fig1b2, fig1b3, nrow = 1, labels = "B")

pdf(file.path(out_dir, "main_B.pdf"), width=8, height = 8)
ggarrange(fig1b1, fig1b2, fig1b3, nrow = 1, labels = "B")
dev.off()
```

### Panel C: analysis code
```{r panel-c-analysis, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# children normalized on total number of events
bld_cytof_counts_processed_child <- 
  preprocess_bld_cytof_child(bld_cytof_counts)

data_use_child <- bld_cytof_counts_processed_child %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "discretized_admit_age_quantile", "sex")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))

smooth_spline_model_loop_child <- model_loop(inputDF   = data_use_child, 
                                             modelType = "smoothSpline",
                                             endpoint  = "trajectory_group") %>%
  ungroup() %>%
  mutate(adjp.shape = p.adjust(p.slope,method = "BH"),
         adjp.average = p.adjust(p.intercept,method = "BH"))

```

### Panel C: output generation code
```{r panel-c-output-gen, message=FALSE, fig.width=7, fig.height=4.5, warning=FALSE}
exampleDF <- data_use_child %>%
  filter(name %in% "Neutrophil (CD16hi)") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5")))

plotExample <- plot_model(exampleDF, 
                          model_loop     = smooth_spline_model_loop_child, 
                          modelType      = "smoothSpline", 
                          p_adjust       = NA, 
                          endpoint       = "trajectory_group", 
                          signif_markers = FALSE,
                          remove_NS      = TRUE,
                          bar_height     = 6.5)

fig1c1 <- plotExample$data %>%
  mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = event_date, y = value)) +
  geom_line(mapping = aes(group = participant_id, y = value), 
            color   = "gray",
            alpha   = 0.4) +
  geom_line(mapping = aes(group = participant_id, y = yhat), 
            size    = 0.6, 
            alpha   = 0.5, 
            color   = "black") +
  geom_point(mapping = aes(color = trajectory_group)) +
  geom_line(data    = mutate(get("pred", plotExample$plot_env),
                             trajectory_group = paste0("TG", trajectory_group)),
            mapping = aes(group = trajectory_group, y = predicted),
            color   = "black", 
            size    = 1.5) +
  facet_wrap(facets = ~trajectory_group, nrow = 1) +
  labs(y = "Neutrophil CD16hi (%)",  x = "Days from admission",  tag = "C") +
  scale_color_manual(values= c("TG1" = "#639A21",
                               "TG2" = "#39828C",
                               "TG3" = "#6371AD", 
                               "TG4" = "#BD7D31",
                               "TG5" = "#9C3418")) +
  coord_cartesian(xlim = c(0, 42)) +
  theme_bw() +
  theme(legend.pos = "none",
        panel.grid.minor = element_blank())

exampleDF <- data_use_child %>%
  filter(name %in% "Neutrophil (CD16low)") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5")))
plotExample <- plot_model(exampleDF, 
                          model_loop     = smooth_spline_model_loop_child, 
                          modelType      = "smoothSpline", 
                          p_adjust       = NA, 
                          endpoint       = "trajectory_group", 
                          signif_markers = FALSE,
                          remove_NS      = TRUE,
                          bar_height     = 6.5)

fig1c2 <- plotExample$data %>%
 mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = event_date, y = value)) +
  geom_line(mapping = aes(group = participant_id, y = value), 
            color   = "gray",
            alpha   = 0.4) +
  geom_line(mapping = aes(group = participant_id, y = yhat), 
            size    = 0.6, 
            alpha   = 0.5, 
            color   = "black") +
  geom_point(mapping = aes(color = trajectory_group)) +
  geom_line(data    = mutate(get("pred", plotExample$plot_env),
                             trajectory_group = paste0("TG", trajectory_group)),
            mapping = aes(group = trajectory_group, y = predicted),
            color   = "black", 
            size    = 1.5) +
  facet_wrap(facets = ~trajectory_group, nrow = 1) +
  labs(y = "Neutrophil CD16low (%))",  x = "Days from admission", tag = " ") +
  scale_color_manual(values= c("TG1" = "#639A21",
                               "TG2" = "#39828C",
                               "TG3" = "#6371AD", 
                               "TG4" = "#BD7D31",
                               "TG5" = "#9C3418")) +
  coord_cartesian(xlim = c(0, 42)) +
  theme_bw() +
  theme(legend.pos = "none",
        panel.grid.minor = element_blank())

grid.arrange(fig1c1, fig1c2, ncol = 1)

pdf(file.path(out_dir, "main_C.pdf"), width=8, height = 8)
grid.arrange(fig1c1, fig1c2, ncol = 1)
dev.off()
```

### Panel D: analysis code
```{r panel-d-analysis-gen-v2, results='hide', fig.show="hide", message=FALSE, warning=FALSE}
# combine all T cells
bld_cytof_counts_processed_child$`T Cell` <-
  rowSums(bld_cytof_counts_processed_child[, grepl(pattern = "T Cell", colnames(bld_cytof_counts_processed_child))])

data_use_child <- bld_cytof_counts_processed_child %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "discretized_admit_age_quantile", "sex")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))

smooth_spline_model_loop_child <- model_loop(inputDF   = data_use_child, 
                                             modelType = "smoothSpline",
                                             endpoint  = "trajectory_group")
```

### Panel D: output generation code
```{r panel-d-output-gen-v2, results='hide', message=FALSE, fig.width=8.5, fig.height=2.5}
exampleDF <- data_use_child %>%
  filter(name %in% "T Cell") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5")))

plotExample <- plot_model(exampleDF, 
                          model_loop     = smooth_spline_model_loop_child, 
                          modelType      = "smoothSpline", 
                          p_adjust       = NA, 
                          endpoint       = "trajectory_group", 
                          signif_markers = FALSE,
                          remove_NS      = TRUE,
                          bar_height     = 6.5)

fig1dAlt <- plotExample$data %>%
  mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = event_date, y = value)) +
  geom_line(mapping = aes(group = participant_id, y = value), 
            color   = "gray",
            alpha   = 0.4) +
  geom_line(mapping = aes(group = participant_id, y = yhat), 
            size    = 0.6, 
            alpha   = 0.5, 
            color   = "black") +
  geom_point(mapping = aes(color = trajectory_group)) +
  geom_line(data    = mutate(get("pred", plotExample$plot_env),
                             trajectory_group = paste0("TG", trajectory_group)),
            mapping = aes(group = trajectory_group, y = predicted),
            color   = "black", 
            size    = 1.5) +
  facet_wrap(facets = ~trajectory_group, nrow = 1) +
  labs(y = "T cells (%)",  x = "Days from admission",  tag = "C") +
  scale_color_manual(values= c("TG1" = "#639A21",
                               "TG2" = "#39828C",
                               "TG3" = "#6371AD", 
                               "TG4" = "#BD7D31",
                               "TG5" = "#9C3418")) +
  coord_cartesian(xlim = c(0, 42)) +
  theme_bw() +
  theme(legend.pos = "none",
        panel.grid.minor = element_blank())

print(fig1dAlt)

pdf(file.path(out_dir, "main_D.pdf"), width=8, height = 8)
print(fig1dAlt)
dev.off()
```

```{r panel-d-analysis, results="hide", fig.show="hide", message=FALSE, echo=TRUE}
# parent count excluding granulocytes
bld_cytof_counts_processed_parent_nongran <- 
  preprocess_bld_cytof_parent_nongran(bld_cytof_counts, bld_cytof_rowfeature)
# combine all T cells
bld_cytof_counts_processed_parent_nongran$`T Cell` <-
  rowSums(bld_cytof_counts_processed_parent_nongran[, grepl(pattern = "T Cell", colnames(bld_cytof_counts_processed_parent_nongran))])

data_use_parent_nongran <- bld_cytof_counts_processed_parent_nongran %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "discretized_admit_age_quantile", "sex")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))

smooth_spline_model_loop_parent_nongran <- model_loop(inputDF   = data_use_parent_nongran,
                                                       modelType = "smoothSpline",
                                                       endpoint  = "trajectory_group")
```

```{r panel-d-output-gen, message=FALSE, results="hide", fig.width=8.5, fig.height=2.5, echo=TRUE, eval=TRUE}
exampleDF <- data_use_parent_nongran %>%
  filter(name %in% "T Cell") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5")))

plotExample <- plot_model(exampleDF, 
                          model_loop     = smooth_spline_model_loop_parent_nongran, 
                          modelType      = "smoothSpline", 
                          p_adjust       = NA, 
                          endpoint       = "trajectory_group", 
                          signif_markers = FALSE,
                          remove_NS      = TRUE,
                          bar_height     = 6.5)

fig1d <- plotExample$data %>%
  mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = event_date, y = value)) +
  geom_line(mapping = aes(group = participant_id, y = value), 
            color   = "gray",
            alpha   = 0.4) +
  geom_line(mapping = aes(group = participant_id, y = yhat), 
            size    = 0.6, 
            alpha   = 0.5, 
            color   = "black") +
  geom_point(mapping = aes(color = trajectory_group)) +
  geom_line(data    = mutate(get("pred", plotExample$plot_env),
                             trajectory_group = paste0("TG", trajectory_group)),
            mapping = aes(group = trajectory_group, y = predicted),
            color   = "black", 
            size    = 1.5) +
  facet_wrap(facets = ~trajectory_group, nrow = 1) +
  labs(y = "T Cell\n(% of non granulocytes)",  x = "Days from admission",  tag = "D") +
  scale_color_manual(values= c("TG1" = "#639A21",
                               "TG2" = "#39828C",
                               "TG3" = "#6371AD", 
                               "TG4" = "#BD7D31",
                               "TG5" = "#9C3418")) +
  coord_cartesian(xlim = c(0, 42)) +
  theme_bw() +
  theme(legend.pos = "none",
        panel.grid.minor = element_blank())

print(fig1d)

pdf(file.path(out_dir, "main_D.pdf"), width=8, height = 8)
print(fig1d)
dev.off()
```

### Panel E: analysis code
```{r panel-e-analysis, results="hide", fig.show="hide", message=FALSE}
bld_cytof_counts_processed_child_nongran <-
  preprocess_bld_cytof_child_nongran(bld_cytof_counts)

data_use_child_nongran <- bld_cytof_counts_processed_child_nongran %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    participant_id),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "discretized_admit_age_quantile", "sex")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))

smooth_spline_model_loop_child_nongran <- model_loop(inputDF   = data_use_child_nongran,
                                                     modelType = "smoothSpline",
                                                     endpoint  = "trajectory_group") %>%
  ungroup() %>%
  mutate(adjp.shape = p.adjust(p.slope,method = "BH"),
         adjp.average = p.adjust(p.intercept,method = "BH"))

```

### Panel E: output generation code
```{r panel-e-output-gen, message=FALSE, results="hide", fig.width=8.25, fig.height=2.5}
exampleDF <- data_use_child_nongran %>%
  filter(name %in% "NK Cell (CD56low CD16hi CD57low)") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5")))

plotExample <- plot_model(exampleDF, 
                          model_loop     = smooth_spline_model_loop_child_nongran, 
                          modelType      = "smoothSpline", 
                          p_adjust       = NA, 
                          endpoint       = "trajectory_group", 
                          signif_markers = FALSE,
                          remove_NS      = TRUE,
                          bar_height     = 6.5)

fig1e <- plotExample$data %>%
  mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = event_date, y = value)) +
  geom_line(mapping = aes(group = participant_id, y = value), 
            color   = "gray",
            alpha   = 0.4) +
  geom_line(mapping = aes(group = participant_id, y = yhat), 
            size    = 0.6, 
            alpha   = 0.5, 
            color   = "black") +
  geom_point(mapping = aes(color = trajectory_group)) +
  geom_line(data    = mutate(get("pred", plotExample$plot_env),
                             trajectory_group = paste0("TG", trajectory_group)),
            mapping = aes(group = trajectory_group, y = predicted),
            color   = "black", 
            size    = 1.5) +
  facet_wrap(facets = ~trajectory_group, nrow = 1) +
  labs(y   = "NK Cell CD56low CD16hi CD57low\n(% of non granulocytes)", 
       x   = "Days from admission",  
       tag = "E") +
  scale_color_manual(values= c("TG1" = "#639A21",
                               "TG2" = "#39828C",
                               "TG3" = "#6371AD", 
                               "TG4" = "#BD7D31",
                               "TG5" = "#9C3418")) +
  coord_cartesian(xlim = c(0, 42)) +
  theme_bw() +
  theme(legend.pos = "none",
        panel.grid.minor = element_blank(),
        axis.title.y = element_text(size = 8))

print(fig1e)

pdf(file.path(out_dir, "main_E.pdf"), width=8, height = 8)
print(fig1e)
dev.off()
```

### Supplementary Panel A: output generation code
```{r sup-panel-a-output-gen, fig.height=9, fig.width=10}
mfiDF <- read_csv(file = mfiFile, show_col_types = FALSE)

colAnnotDF <- mfiDF %>%
  rowid_to_column() %>%
  select(rowid, Predicted_celltype) %>%
  # correct typo
  mutate(Predicted_celltype = gsub(pattern = "Activated",
                                   replacement = "activated",
                                   Predicted_celltype)) %>%
  merge(y = rownames_to_column(bld_cytof_rowfeature,
                               var = "Predicted_celltype"),
        by = "Predicted_celltype",
        all.x = TRUE) %>%
  mutate(`Broad Parent Population` = ifelse(test = is.na(`Broad Parent Population`),
                                            yes  = gsub(pattern = " \\(.+",
                                                        replacement = "",
                                                        Predicted_celltype),
                                            no   = `Broad Parent Population`)) %>%
  rename(Parent_Celltype = `Broad Parent Population`,
         Child_Celltype  = Predicted_celltype) %>%
  select(rowid, Parent_Celltype, Child_Celltype) %>%
  data.frame() %>%
  arrange(rowid) %>%
  column_to_rownames(var = "rowid")

mfiMat <- mfiDF %>%
  rowid_to_column() %>%
  select(-Tier1, -Predicted_celltype) %>%
  mutate_at(.vars = vars(-("rowid")), scale) %>%
  as.data.frame() %>%
  column_to_rownames(var = "rowid") %>%
  t() %>%
  as.matrix()

# exclude Multiplets
mfiMat <- mfiMat[, !grepl(pattern = "Multiplet", colAnnotDF$Parent_Celltype)]
colAnnotDF <- colAnnotDF[colnames(mfiMat), ]

# generate vector of discrete colors 
qual_col_pals <- brewer.pal.info[brewer.pal.info$category %in% 'qual',]
col_vector <- unlist(mapply(FUN = brewer.pal, n = qual_col_pals$maxcolors, name = rownames(qual_col_pals)))

# assign parent a color
set.seed(seed = 1)
parent_col <- sample(col_vector, size = length(unique(colAnnotDF$Parent_Celltype))) %>%
  setNames(nm = unique(colAnnotDF$Parent_Celltype))

# assign parent a color
set.seed(seed = 2)
child_col <- sample(col_vector, size = length(unique(colAnnotDF$Child_Celltype))) %>%
  setNames(nm = unique(colAnnotDF$Child_Celltype))

top_anno <- HeatmapAnnotation(df                      = colAnnotDF,
                              show_legend             = TRUE,
                              col                     = list(Parent_Celltype = parent_col,
                                                             Child_Celltype = child_col),
                              annotation_legend_param = list(legend_direction = "horizontal",
                                                             ncol = 4))

figs1a <- Heatmap(matrix               = mfiMat,
                  show_column_names    = FALSE,
                  cluster_columns      = FALSE,
                  name                 = "scaled\nMFI",
                  top_annotation =  top_anno)

draw(figs1a, 
     heatmap_legend_side    = "right",
     annotation_legend_side = "bottom",
     legend_grouping        = "adjusted")

pdf(file.path(out_dir, "suppl_A.pdf"), width=8, height = 8)
draw(figs1a, 
     heatmap_legend_side    = "right",
     annotation_legend_side = "bottom",
     legend_grouping        = "adjusted")
dev.off()
```

### Supplementary Panel B: analysis code
```{r sup-panel-b-analysis, results="hide", message=FALSE}
for (n in grep(pattern = "bld_cytof|clinical_data", 
               names(data_env_withHC), 
               value   = TRUE)) {
  assign(paste0(n, "_withHC"), value = data_env_withHC[[n]])
}

# remove outliers
bld_cytof_counts_processed_parent_withHC <- 
  preprocess_bld_cytof_parent(bld_cytof_counts_withHC,
                              bld_cytof_rowfeature_withHC)

# PCA
pc <- prcomp(bld_cytof_counts_processed_parent_withHC)
```

### Supplementary Panel B: output generation code
```{r sup-panel-b-output-gen, fig.width=3.8, fig.height=3.75}
plotDF <- pc$x[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = clinical_data_withHC, by = "sample_id") %>%
  merge(y  = rownames_to_column(bld_cytof_metadata_withHC, var = "sample_id") %>%
          select(sample_id, Core),
        by = "sample_id") %>%
  mutate(core_lab        = gsub(pattern = ".+_", replacement = "", Core),
         enrollment_site = ifelse(test = participant_type %in% "Healthy control (Emory)",
                                  yes  = "Emory-Ctrl", 
                                  no   = enrollment_site),
         enrollment_site = factor(enrollment_site),
         enrollment_site = relevel(enrollment_site, ref = "Emory-Ctrl"))
enrollmentSite2color <- rainbow(n = length(unique(plotDF$enrollment_site))) %>%
  setNames(nm = unique(plotDF$enrollment_site))
enrollmentSite2color["Emory-Ctrl"] <- "black"
figs1b <- ggplot(data    = plotDF,
                 mapping = aes(x = PC1, y = PC2, color = enrollment_site))+
  geom_point(size = 3, alpha = 0.7, shape = 16) +
  scale_color_manual(values = enrollmentSite2color) +
  stat_ellipse()+
  labs(x     = paste0("1st dimension (",
                      round((pc$sdev^2/sum(pc$sdev^2))[1] * 100),
                      "%)"),
       y     = paste0("2nd dimension (",
                      round((pc$sdev^2/sum(pc$sdev^2))[2] * 100),
                      "%)"))+
  theme_classic() +
  theme(legend.text     = element_text(size = 6),
        legend.key.size = unit(0.01, units = "npc"),
        legend.position = "none") 

print(figs1b)

pdf(file.path(out_dir, "suppl_B.pdf"), width=8, height = 8)
print(figs1b)
dev.off()
```

### Supplementary Panel C: analysis code
```{r sup-panel-c-analysis}
# step 2: specify which assay is going to be used as input
bld_cytof_counts_processed_parent <- 
  preprocess_bld_cytof_parent(bld_cytof_counts, bld_cytof_rowfeature)
pvca_input_matrix <- t(bld_cytof_counts_processed_parent)
pvca_input_raw <- t(bld_cytof_counts_processed_parent)
pvca_input_metadata <- bld_cytof_metadata

# step3: select variable to be included in PVCA
#   event_data_week: day from admission to to the hospital coded as week (categorical variable)
#   sympt_date_week: day from onset of symptoms coded as week (categorical variable)
#   discretized_admit_age_quantile: age at admission coded as quintiles 
pvca_input_phenodata <- clinical_data %>%
  mutate(event_date_week = findInterval(event_date, 
                                        vec        = c(0, 7, 14, 21, 28), 
                                        all.inside = TRUE),
         day_from_sympt  = event_date - symptom_date,
         sympt_date_week = findInterval(day_from_sympt, 
                                        vec        = c(0, 7, 14, 21, 28, Inf),
                                        all.inside = TRUE),
         death_day28 = respiratory_status_day28 %in% 7) %>%
    select(sample_id, event_date_week, event_type, respiratory_status,
           sex, death_day28, sympt_date_week, event_location,
           discretized_admit_age_quantile, ethnicity, participant_id, race,
           trajectory_group, enrollment_site)

# step 4 append meta data to pvca phenodata
#   core site: where are the sample processed (this chunk of code need to be modified for each core assay)
#   plate = interaction term of phase and plate
pvca_input_phenodata <- pvca_input_metadata %>%
  rownames_to_column(var = "sample_id") %>%
  mutate(core_site = gsub(pattern = ".+_", replacement = "", Core),
         plate     = interaction(phase, plate_num, drop = TRUE)) %>%
  select(sample_id, plate, phase, core_site) %>%
  merge(x     = pvca_input_phenodata,
        by    = "sample_id",
        all.x = TRUE)

# append rownames
pvca_input_phenodata <- pvca_input_phenodata %>%
  column_to_rownames(var = "sample_id")
pvca_input_phenodata <- pvca_input_phenodata[colnames(pvca_input_matrix), , drop = FALSE]

# removing core_site (only 1 core) and event location (only Hospitalized)
pvca_input_phenodata$event_location <- NULL
pvca_input_phenodata$core_site <- NULL

# step 5: run PVCA
fit <- PVCA(counts    = pvca_input_matrix,
            meta      = pvca_input_phenodata,
            inter     = FALSE,
            threshold = 0.6)
```

### Supplementary Panel C: output generation code
```{r sup-panel-c-output-gen, fig.width=5.25, fig.height=3.75}
# step 6: plot barplot with PVCA results
pvca_barplot_data <- data.frame(explained = as.vector(fit),
                                effect    = names(fit)) %>%
  arrange(explained) %>%
  mutate(effect = factor(effect, levels = effect))

figs1c <- ggplot(data    = pvca_barplot_data,
                 mapping = aes(x = effect, y = explained)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = signif(explained, digits = 2)),
            nudge_y   = 0.03,
            size      = 2) +
  labs(x = NULL, y = "Proportion of the variance explained", tag = "B") + 
  theme_classic() +
  coord_flip() +
    geom_hline(yintercept = 0.1, linetype = 2)

print(figs1c)

pdf(file.path(out_dir, "suppl_C.pdf"), width=8, height = 8)
print(figs1c)
dev.off()
```

### Supplementary Panel D: analysis code
```{r sup-panel-d-analysis, message=FALSE}
# add lymphocyte and myeloid and ratio 
bld_cytof_counts_processed_parent$Myeloid <-
  rowSums(bld_cytof_counts_processed_parent[, c("Basophil", "Conventional DC", "Eosinophil",
                                                        "Monocytes", "Neutrophil")])
bld_cytof_counts_processed_parent$Lymphocyte <-
  rowSums(bld_cytof_counts_processed_parent[, c("B Cell", "CD4+ T Cell", "CD8+ T Cell", "gd T Cell",
                                                "ILC", "NK Cell", "T Cell")])

bld_cytof_counts_processed_parent$MyeloidToLymphocyte <-
  (bld_cytof_counts_processed_parent$Myeloid + 0.0001) / 
  (bld_cytof_counts_processed_parent$Lymphocyte + 0.0001)

data_use_parent <- bld_cytof_counts_processed_parent %>%
    rownames_to_column(var = "sample_id") %>%
    merge(y  = select(clinical_data, 
                      sample_id, 
                      event_date,
                      trajectory_group,
                      enrollment_site,
                      discretized_admit_age_quantile,
                      sex,
                      participant_id),
          by = "sample_id") %>%
    filter(!is.na(trajectory_group)) %>%
    filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
    mutate_if(is.character, as.factor) %>%
    mutate(trajectory_group = factor(trajectory_group)) %>%
    pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                           "enrollment_site", "trajectory_group",
                           "discretized_admit_age_quantile", "sex")) %>%
    mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
    filter(!is.na(value))

  smooth_spline_model_loop_parent <- model_loop(inputDF   = data_use_parent, 
                                         modelType = "smoothSpline",
                                         endpoint  = "trajectory_group")
  # exclude Lymphocyte, Myeloid, MyeloidToLymphocyte when correction for multiple testing
adjpDF <- smooth_spline_model_loop_parent %>%
  rownames_to_column() %>%
  filter(!(rowname %in% c("Lymphocyte", "Myeloid", "MyeloidToLymphocyte"))) %>%
  ungroup() %>%
  mutate(adjp.shape = p.adjust(p.slope,method = "BH"),
         adjp.average = p.adjust(p.intercept,method = "BH")) %>%
  select(rowname, adjp.shape, adjp.average)

smooth_spline_model_loop_parent %>%
  rownames_to_column() %>%
  merge(y = adjpDF, by = "rowname", all.x = TRUE) %>%
  column_to_rownames(var = "rowname") -> smooth_spline_model_loop_parent

```

### Supplementary Panel D: output generation code
```{r sup-panel-d-output-gen, fig.width=8.5, fig.height=2.5}
exampleDF <- data_use_parent %>%
  filter(name %in% "MyeloidToLymphocyte") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5")))

plotExample <- plot_model(exampleDF, 
                          model_loop     = smooth_spline_model_loop_parent, 
                          modelType      = "smoothSpline", 
                          p_adjust       = NA, 
                          endpoint       = "trajectory_group", 
                          signif_markers = FALSE,
                          remove_NS      = TRUE,
                          bar_height     = 6.5)

figs1d <- plotExample$data %>%
  mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = event_date, y = value)) +
  geom_line(mapping = aes(group = participant_id, y = value), 
            color   = "gray",
            alpha   = 0.4) +
  geom_line(mapping = aes(group = participant_id, y = yhat), 
            size    = 0.6, 
            alpha   = 0.5, 
            color   = "black") +
  geom_point(mapping = aes(color = trajectory_group)) +
  geom_line(data    = mutate(get("pred", plotExample$plot_env),
                             trajectory_group = paste0("TG", trajectory_group)),
            mapping = aes(group = trajectory_group, y = predicted),
            color   = "black", 
            size    = 1.5) +
  facet_wrap(facets = ~trajectory_group, nrow = 1) +
  labs(y = "Myeloid/Lymphocyte",  x = "Days from admission",  tag = "D") +
  scale_color_manual(values= c("TG1" = "#639A21",
                               "TG2" = "#39828C",
                               "TG3" = "#6371AD", 
                               "TG4" = "#BD7D31",
                               "TG5" = "#9C3418")) +
  coord_cartesian(xlim = c(0, 42)) +
  theme_bw() +
  theme(legend.pos = "none",
        panel.grid.minor = element_blank())
print(figs1d)

pdf(file.path(out_dir, "suppl_D.pdf"), width=8, height = 8)
print(figs1d)
dev.off()

```

### Supplementary Table A: analysis code
```{r sup-table-a-analysis, message=FALSE, warning=FALSE, results="hide", eval=TRUE}
for (SUFFIX in c("parent", "child", "parent_nongran", "child_nongran", "child_to_parent_ratio")) {
  eval(expr = parse(text = paste0("bld_cytof_counts_processed <- bld_cytof_counts_processed_",
                                  SUFFIX)))

  # visit 1 analysis
  data_use_visit1 <- bld_cytof_counts_processed %>%
    rownames_to_column(var = "sample_id") %>%
    setNames(nm = gsub(x           = names(.), 
                       pattern     = "+",
                       replacement = "pos", 
                       fixed       = TRUE)) %>%
    setNames(nm = gsub(x           = names(.), 
                       pattern     = "-",
                       replacement = "neg", 
                       fixed       = TRUE)) %>%
    setNames(nm = gsub(x           = names(.), 
                       pattern     = "\\(|\\)| ",
                       replacement = "_"
                       )) %>%
    merge(y  = select(clinical_data, 
                      sample_id, 
                      event_type,
                      trajectory_group,
                      enrollment_site,
                      discretized_admit_age_quantile,
                      sex),
          by = "sample_id") %>%
    filter(event_type %in% "Visit 1" &
           !is.na(trajectory_group)) %>%
    mutate_if(is.character, as.factor) %>%
    mutate(trajectory_group = factor(trajectory_group)) %>%
    select(-event_type) %>%
    column_to_rownames(var = "sample_id")

  res_table_ordinal_visit1 <- NULL
  data_use <- data_use_visit1
  for (varName in setdiff(names(data_use_visit1), names(clinical_data))) {
    my.formula0 <- paste0('trajectory_group~ 1+(1|enrollment_site)+',
                          'discretized_admit_age_quantile+sex')
    my.formula1 <- paste0('trajectory_group~',
                          varName,
                          '+(1|enrollment_site)+',
                          'discretized_admit_age_quantile+sex')
    res <- mixed_ordinal(my.formula0, 
                         my.formula1,
                         data_use = data_use_visit1)
    coef(clmm(formula(my.formula1), 
              data = data_use_visit1))[varName] %>%
      unname() %>%
      c(res, coef = .) -> res
    res_table_ordinal_visit1 <- rbind(res_table_ordinal_visit1, res)
  }
  rm(data_use)
  rownames(res_table_ordinal_visit1) <- setdiff(names(data_use_visit1), names(clinical_data))
  res_table_ordinal_visit1 <- res_table_ordinal_visit1 %>%
    as.data.frame() %>%
    mutate(qval =  qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues,
           adjp = p.adjust(pval, method = "BH"))
  eval(expr = parse(text = paste0("res_table_ordinal_visit1_",
                                  SUFFIX,
                                  " <- res_table_ordinal_visit1")))
  
  # do pairwise comparison at visit 1 and extract coefficient of regression
  data_use_visit1_tmp <- data_use_visit1 %>%
    rename(endpoints = trajectory_group,
           sites     = enrollment_site,
           age       = discretized_admit_age_quantile)
  res_table_pairwise_visit1 <- NULL
  res_table_pairwise_coef_visit1 <- NULL
  varNames <- setdiff(names(data_use_visit1_tmp), 
                      c(names(clinical_data),
                        c("endpoints", "sites", "age")))
  for(varName in varNames){
    my.formula0 <- paste0(varName,
                          "~(1|sites)",
                          "+age",
                          "+sex")
    my.formula1 <- paste0(varName,
                          "~endpoints",
                          "+(1|sites)",
                          "+age",
                          "+sex")
    res_table_pairwise_visit1 <- rbind(res_table_pairwise_visit1,
                                              mixed_pairwise(my.formula0, 
                                                             my.formula1, 
                                                             data_use_visit1_tmp))
    res_table_pairwise_coef_visit1 <- rbind(res_table_pairwise_coef_visit1,
                                                   mixed_pairwise_coef(my.formula0, 
                                                                       my.formula1, 
                                                                       data_use_visit1_tmp))
  }
  rownames(res_table_pairwise_visit1) <- varNames
  rownames(res_table_pairwise_coef_visit1) <- varNames
  eval(expr = parse(text = paste0("res_table_pairwise_visit1_",
                                  SUFFIX,
                                  " <- res_table_pairwise_visit1")))
  eval(expr = parse(text = paste0("res_table_pairwise_coef_visit1_",
				  SUFFIX,
				  " <- res_table_pairwise_coef_visit1")))

  # parent smooth spline
  data_use <- bld_cytof_counts_processed %>%
    rownames_to_column(var = "sample_id") %>%
    merge(y  = select(clinical_data, 
                      sample_id, 
                      event_date,
                      trajectory_group,
                      enrollment_site,
                      discretized_admit_age_quantile,
                      sex,
                      participant_id),
          by = "sample_id") %>%
    filter(!is.na(trajectory_group)) %>%
    filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
    mutate_if(is.character, as.factor) %>%
    mutate(trajectory_group = factor(trajectory_group)) %>%
    pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                           "enrollment_site", "trajectory_group",
                           "discretized_admit_age_quantile", "sex")) %>%
    mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
    filter(!is.na(value))

  smooth_spline_model_loop <- model_loop(inputDF   = data_use, 
                                         modelType = "smoothSpline",
                                         endpoint  = "trajectory_group")
  
  # for longitudinal analysis fetch directionality from lme (TG1 vs TG5)
  smooth_spline_model_loop$trajectory_group5 <- NA
  smooth_spline_model_loop$"event_date:trajectory_group5" <- NA
  for (modName in rownames(smooth_spline_model_loop)) {
    fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + 
                 discretized_admit_age_quantile,
               random = ~1|enrollment_site/participant_id,
               data   = filter(data_use, name %in% modName))
    smooth_spline_model_loop[modName, "trajectory_group5"] <- 
      unique(coef(fit)["trajectory_group5"])
    smooth_spline_model_loop[modName, "event_date:trajectory_group5"] <- 
      unique(coef(fit)["event_date:trajectory_group5"])
  }
  # correct adjp
  smooth_spline_model_loop$adjp.slope <- p.adjust(smooth_spline_model_loop$p.slope,
                                                 method = "BH")
  smooth_spline_model_loop$adjp.intercept <- p.adjust(smooth_spline_model_loop$p.intercept,
                                                 method = "BH")
  eval(expr = parse(text = paste0("smooth_spline_model_loop_",
                                  SUFFIX,
                                  " <- smooth_spline_model_loop")))

  # for longitudinal analysis, do pairwise regression and extract coefficient from
  # linear model
  combMat <- combn(unique(data_use$trajectory_group), m = 2)
  coefCombMat <- apply(combMat, MARGIN = 2, FUN = function(comparison) {
    # print(comparison)
    data_use_tmp <- filter(data_use, trajectory_group %in% comparison)
    
    coefDF <- lapply(rownames(smooth_spline_model_loop), FUN = function(modName) {
      #fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + 
      #             discretized_admit_age_quantile,
      #           random = ~1|enrollment_site/participant_id,
      #           data   = filter(data_use_tmp, name %in% modName))
      
      fit <- try(lme(fixed  = value ~ event_date * trajectory_group + sex + 
                   discretized_admit_age_quantile,
                 random = ~1|enrollment_site/participant_id,
                 data   = filter(data_use_tmp, name %in% modName)), silent = TRUE)
      if (!("try-error" %in% class(fit))) {
        average <- unique(coef(fit)[grep(pattern = "^trajectory_group",
                                         colnames(coef(fit)))]) %>%
          unlist() %>%
          unname()
        shape <- unique(coef(fit)[grep(pattern = "^event_date:trajectory_group",
                                       colnames(coef(fit)))]) %>%
          unlist() %>%
          unname()
      } else {
        average <- NA
        shape <- NA
      }
      
      return(value = data.frame(`Module (or Feature)` = modName,
                                average, 
                                shape,
                                comparison = paste(sort(comparison), 
                                                   collapse = "|"),
                                check.names = FALSE))
    }) %>%
      do.call(what = rbind)
    return(value = coefDF)
  }) %>%
    do.call(what = rbind)
  eval(expr = parse(text = paste0("coefCombMat_",
                                  SUFFIX,
                                  " <- coefCombMat")))

  # generate input data.frame for longitudinal DFSO analysis 
  # (event_date=event_date-symptom_date)
  data_use_dfso <- bld_cytof_counts_processed %>%
    rownames_to_column(var = "sample_id") %>%
    merge(y  = select(clinical_data, 
                      sample_id, 
                      event_date,
                      trajectory_group,
                      enrollment_site,
                      discretized_admit_age_quantile,
                      sex,
                      participant_id,
                      symptom_date),
          by = "sample_id") %>%
    filter(!is.na(trajectory_group)) %>%
    filter(duplicated(participant_id) | 
             duplicated(participant_id, fromLast = TRUE)) %>%
    mutate_if(is.character, as.factor) %>%
    mutate(trajectory_group = factor(trajectory_group)) %>%
    pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                           "enrollment_site", "trajectory_group",
                           "discretized_admit_age_quantile", "sex", 
                           "symptom_date")) %>%
    mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile),
           event_date = event_date - symptom_date) %>%
    filter(!is.na(value) & !is.na(event_date))

  # generate smooth spline regression with symptom onset as reference and impute 
  # directionality from linear regression
  smooth_spline_model_loop_dfso <- model_loop(data_use_dfso, 
                                              modelType = "smoothSpline", 
                                              endpoint  = "trajectory_group")

  # fetch directionality from lme
  smooth_spline_model_loop_dfso$trajectory_group5 <- NA
  smooth_spline_model_loop_dfso$"event_date:trajectory_group5" <- NA
  for (modName in rownames(smooth_spline_model_loop_dfso)) {
    fit <- lme(fixed= value ~ event_date * trajectory_group + sex + 
                 discretized_admit_age_quantile,
               random = ~1|enrollment_site/participant_id,
               data = filter(data_use_dfso, name %in% modName))
    smooth_spline_model_loop_dfso[modName, "trajectory_group5"] <- 
      unique(coef(fit)["trajectory_group5"])
    smooth_spline_model_loop_dfso[modName, "event_date:trajectory_group5"] <- 
      unique(coef(fit)["event_date:trajectory_group5"])
  }
  smooth_spline_model_loop_dfso$adjp.slope <- p.adjust(smooth_spline_model_loop_dfso$p.slope,
                                                  method = "BH")
  smooth_spline_model_loop_dfso$adjp.intercept <- p.adjust(smooth_spline_model_loop_dfso$p.intercept,
                                                  method = "BH")
  eval(expr = parse(text = paste0("smooth_spline_model_loop_dfso_",
				  SUFFIX,
				  " <- smooth_spline_model_loop_dfso")))

  # do pairwise regression with DFSO as time and extract coefficient from linear 
  # model
  # use linear regression to infer direction
  combMat <- combn(unique(data_use_dfso$trajectory_group), m = 2)
  coefCombMat_dfso <- apply(combMat, MARGIN = 2, FUN = function(comparison) {
    # print(comparison)
    data_use_tmp <- filter(data_use_dfso, trajectory_group %in% comparison)
    
    coefDF <- lapply(rownames(smooth_spline_model_loop_dfso), 
                     FUN = function(modName) {
                       fit <- try(lme(fixed  = value ~ event_date * trajectory_group + sex + 
                                    discretized_admit_age_quantile,
                                  random = ~1|enrollment_site/participant_id,
                                  data   = filter(data_use_tmp, name %in% modName)),
                                  silent = TRUE)
                       if (!("try-error" %in% class(fit))) {
                         average <- unique(coef(fit)[grep(pattern = "^trajectory_group",
                                                        colnames(coef(fit)))]) %>%
                         unlist() %>%
                         unname()
                       shape <- unique(coef(fit)[grep(pattern = "^event_date:trajectory_group",
                                                      colnames(coef(fit)))]) %>%
                         unlist() %>%
                         unname()
                       } else {
                         average <- shape <- NA
                       }
                       return(value = data.frame(`Module (or Feature)` = modName,
                                                 average, 
                                                 shape,
                                                 comparison = paste(sort(comparison), collapse = "|"),
                                                 check.names = FALSE))
                     }) %>%
      do.call(what = rbind)
    return(value = coefDF)
  }) %>%
    do.call(what = rbind)
  eval(expr = parse(text = paste0("coefCombMat_dfso_",
				  SUFFIX,
				  "<- coefCombMat_dfso")))
}
```

### Supplementary Table A: output generation code
```{r sup-table-a-output-gen, results='asis', eval=TRUE}
suffixLS <- c("parent", "child", "parent_nongran", "child_nongran", "child_to_parent_ratio")
longNameLS <- c("parent"                = "Parent population",
                "child"                 = "Child population",
                "parent_nongran"        = "Parent population of non granulocytes",
                "child_nongran"         = "Child population of non granulocytes",
                "child_to_parent_ratio" = "Child to parent ratio")

for (SUFFIX in suffixLS) {
  eval(expr = parse(text = paste0("res_table_ordinal_visit1 <- res_table_ordinal_visit1_",
                                   SUFFIX)))
  supTab_resultsVisit1 <- res_table_ordinal_visit1 %>%
    rownames_to_column(var = "Module (or Feature)") %>%
    filter(pval <= 0.05) %>%
    mutate(Analysis              = paste("Visit 1 analysis",
                                        longNameLS[SUFFIX],
                                        "- overall"),
           Direction             = ifelse(test = sign(coef) %in% 1,
                                          yes  = "Severe",
                                          no   = "Mild")) %>%
    rename(`P value` = pval,
           `Adj. p value` = adjp) %>%
    select(`Module (or Feature)`,
           Analysis,
           `P value`,
           `Adj. p value`,
           Direction) %>%
    `rownames<-`(NULL)
  
  supTab_resultsVisit1 %>%
    head() %>%
    kable() %>%
    print()
  
  # append pairwise comparison to visit1 results
  eval(expr = parse(text = paste0("res_table_pairwise_visit1 <- res_table_pairwise_visit1_",
                                  SUFFIX)))
  eval(expr = parse(text = paste0("res_table_pairwise_coef_visit1 <- res_table_pairwise_coef_visit1_",
                                  SUFFIX)))
  supTab_resultsVisit1_tmp <- res_table_pairwise_visit1 %>%
    as.data.frame() %>%
    rownames_to_column(var = "Module (or Feature)") %>%
    pivot_longer(cols      = -`Module (or Feature)`, 
                 names_to  = "Analysis", 
                 values_to = "P value") %>%
    mutate(`Adj. p value` = p.adjust(`P value`, method = "BH"))
  res_table_pairwise_coef_visit1 %>%
    as.data.frame() %>%
    rownames_to_column(var = "Module (or Feature)") %>%
    pivot_longer(cols      = -`Module (or Feature)`,
                 names_to  = "Analysis",
                 values_to = "Direction") %>%
    merge(x  = supTab_resultsVisit1_tmp, 
          by = c("Module (or Feature)", "Analysis")) -> supTab_resultsVisit1_tmp
  
  supTab_resultsVisit1_tmp <- supTab_resultsVisit1_tmp %>%
    mutate(Analysis              = paste("Visit 1 analysis",
                                         longNameLS[SUFFIX],
                                         "-", 
                                         Analysis),
           Direction             = ifelse(test = sign(Direction) %in% 1,
                                          yes  = "Severe",
                                          no   = "Mild")) %>%
    filter(`P value` <= 0.05 & 
             `Module (or Feature)` %in% supTab_resultsVisit1$"Module (or Feature)")
  
  supTab_resultsVisit1 <- rbind(supTab_resultsVisit1,
                                       supTab_resultsVisit1_tmp)
  
  supTab_resultsVisit1 %>%
    tail() %>%
    kable() %>%
    print()
  eval(expr = parse(text = paste0("supTab_resultsVisit1_",
                                  SUFFIX,
                                  " <- supTab_resultsVisit1")))
  
  # generate supplementary table with longitudinal analaysis overall
  eval(expr = parse(text = paste0("smooth_spline_model_loop <- smooth_spline_model_loop_",
                                  SUFFIX)))
  supTab_resultsLongitudinal <- smooth_spline_model_loop %>%
    rownames_to_column(var = "Module (or Feature)") %>%
    select(`Module (or Feature)`, p.slope, p.intercept, adjp.slope, adjp.intercept, trajectory_group5, `event_date:trajectory_group5`) %>%
    rename(coef.intercept = trajectory_group5, 
           coef.slope = `event_date:trajectory_group5`) %>%
    pivot_longer(cols = -`Module (or Feature)`, names_to = c("cname", "Analysis"), names_pattern = "(.*)\\.(.*)") %>%
    pivot_wider(names_from = cname, values_from = value) %>%
    filter(p <= 0.05) %>%
    mutate(Analysis = recode(Analysis,
                             `slope` = paste("Longitudinal analysis",
                                             longNameLS[SUFFIX],
                                             "- overall shape"),
                             `intercept` = paste("Longitudinal analysis",
                                                 longNameLS[SUFFIX],
                                                 "- overall average")),
           Direction = ifelse(test = sign(coef) %in% 1,
                              yes  = "Severe",
                              no   = "Mild")) %>%
    rename(`P value` = p,
           `Adj. p value` = adjp) %>%
    select(`Module (or Feature)`,
           Analysis,
           `P value`,
           `Adj. p value`,
           Direction) %>%
    `rownames<-`(NULL)
  
  supTab_resultsLongitudinal  %>% 
    head() %>%
    kable() %>%
    print()
  
  # append pairwise comparison to longitudinal results
  eval(expr = parse(text = paste0("coefCombMat <- coefCombMat_",
                                  SUFFIX)))
  supTab_resultsLongitudinalPairwise <- smooth_spline_model_loop %>%
    rownames_to_column(var = "Module (or Feature)") %>%
    select(`Module (or Feature)`, 
           matches(match = "^p.slope_.v.$"),
           matches(match = "^p.intercept_.v.$")) %>%
    pivot_longer(cols      = -`Module (or Feature)`,
                 names_to  = "Analysis",
                 values_to = "P value") %>%
    group_by(Analysis) %>%
    mutate(`Adj. p value` = p.adjust(`P value`, method = "BH")) %>%
    ungroup() %>%
    mutate(Analysis = gsub(pattern     = "slope",
                           replacement = "shape",
                           Analysis),
           Analysis = gsub(pattern     = "intercept",
                           replacement = "average",
                           Analysis),
           Analysis = gsub(pattern = "^p\\.(.+)_(.)v(.)$",
                           replacement = paste("Longitudinal analysis",
                                               longNameLS[SUFFIX],
                                               "- \\2|\\3 \\1"),
                           Analysis))
  coefCombMat %>%
    pivot_longer(cols      = -c(`Module (or Feature)`, comparison),
                 names_to  = "Analysis",
                 values_to = "Direction") %>%
    mutate(Analysis = paste("Longitudinal analysis",
                            longNameLS[SUFFIX],
                            "-",
                            comparison,
                            Analysis),
           Direction = ifelse(test = sign(Direction) %in% 1,
                              yes  = "Severe",
                              no   = "Mild")) %>%
    select(-comparison) %>%
    merge(x  = supTab_resultsLongitudinalPairwise,
          by = c("Module (or Feature)", "Analysis")) -> 
    supTab_resultsLongitudinalPairwise
  
  # append to overall results
  supTab_resultsLongitudinalPairwise %>%
    filter(`P value` <= 0.05 &
             `Module (or Feature)` %in% 
             supTab_resultsLongitudinal$"Module (or Feature)") %>%
    rbind(supTab_resultsLongitudinal, .) -> supTab_resultsLongitudinal
  
  supTab_resultsLongitudinal %>%
    tail() %>%
    kable() %>%
    print()
  eval(expr = parse(text = paste0("supTab_resultsLongitudinal_",
                                  SUFFIX,
                                  " <- supTab_resultsLongitudinal")))
  
  # generate supplementary table with longitudinal analaysis DFSO overall
  eval(expr = parse(text = paste0("smooth_spline_model_loop_dfso <- smooth_spline_model_loop_dfso_",
                                  SUFFIX)))
  supTab_resultsLongitudinalDFSO <- smooth_spline_model_loop_dfso %>%
    rownames_to_column(var = "Module (or Feature)") %>%
    select(`Module (or Feature)`, p.slope, p.intercept, adjp.slope, adjp.intercept, trajectory_group5, 
           `event_date:trajectory_group5`) %>%
    rename(coef.intercept = trajectory_group5, 
           coef.slope = `event_date:trajectory_group5`) %>%
    pivot_longer(cols = -`Module (or Feature)`, names_to = c("cname", "Analysis"), 
                 names_pattern = "(.*)\\.(.*)") %>%
    pivot_wider(names_from = cname, values_from = value) %>%
    filter(p <= 0.05) %>%
    mutate(Analysis = recode(Analysis,
                             `slope`     = 
                               paste("Longitudinal analysis DFSO",
                                     longNameLS[SUFFIX],
                                     "- overall shape"),
                             `intercept` = 
                               paste("Longitudinal analysis DFSO",
                                     longNameLS[SUFFIX],
                                     "- overall average")),
           Direction = ifelse(test = sign(coef) %in% 1,
                              yes  = "Severe",
                              no   = "Mild")) %>%
    rename(`P value` = p,
           `Adj. p value` = adjp) %>%
    select(`Module (or Feature)`,
           Analysis,
           `P value`,
           `Adj. p value`,
           Direction) %>%
    `rownames<-`(NULL)
  
  supTab_resultsLongitudinalDFSO %>%
    head() %>%
    kable() %>%
    print()
  
  # append pairwise comparison to longitudinal results DFSO
  eval(expr = parse(text = paste0("coefCombMat_dfso <- coefCombMat_dfso_",
                                  SUFFIX)))
  supTab_resultsLongitudinalDFSOPairwise <- smooth_spline_model_loop_dfso %>%
    rownames_to_column(var = "Module (or Feature)") %>%
    select(`Module (or Feature)`, 
           matches(match = "^p.slope_.v.$"),
           matches(match = "^p.intercept_.v.$")) %>%
    pivot_longer(cols      = -`Module (or Feature)`,
                 names_to  = "Analysis",
                 values_to = "P value") %>%
    group_by(Analysis) %>%
    mutate(`Adj. p value` = p.adjust(`P value`, method = "BH")) %>%
    ungroup() %>%
    mutate(Analysis = gsub(pattern     = "slope",
                           replacement = "shape",
                           Analysis),
           Analysis = gsub(pattern     = "intercept",
                           replacement = "average",
                           Analysis),
           Analysis = gsub(pattern     = "^p\\.(.+)_(.)v(.)$",
                           replacement = 
                             paste("Longitudinal analysis DFSO",
                             longNameLS[SUFFIX],
                             "- \\2|\\3 \\1"),
                           Analysis))
  coefCombMat_dfso %>%
    pivot_longer(cols      = -c(`Module (or Feature)`, comparison),
                 names_to  = "Analysis",
                 values_to = "Direction") %>%
    mutate(Analysis = paste("Longitudinal analysis DFSO",
                            longNameLS[SUFFIX],
                            "-",
                            comparison,
                            Analysis),
           Direction = ifelse(test = sign(Direction) %in% 1,
                              yes  = "Severe",
                              no   = "Mild")) %>%
    select(-comparison) %>%
    merge(x  = supTab_resultsLongitudinalDFSOPairwise,
          by = c("Module (or Feature)", "Analysis")) -> 
    supTab_resultsLongitudinalDFSOPairwise
  
  # append to overall results
  supTab_resultsLongitudinalDFSOPairwise %>%
    filter(`P value` <= 0.05 &
             `Module (or Feature)` %in% 
             supTab_resultsLongitudinalDFSO$"Module (or Feature)") %>%
    rbind(supTab_resultsLongitudinalDFSO, .) -> supTab_resultsLongitudinalDFSO
  
  supTab_resultsLongitudinalDFSO %>%
    tail() %>%
    kable() %>%
    print()
  eval(expr = parse(text = paste0("supTab_resultsLongitudinalDFSO_",
                                  SUFFIX,
                                  " <- supTab_resultsLongitudinalDFSO")))
}

# merge all results table into one csv file
supTab_results <- rbind(supTab_resultsVisit1_parent,
                        supTab_resultsVisit1_child,
                        supTab_resultsVisit1_parent_nongran,
                        supTab_resultsVisit1_child_nongran,
                        supTab_resultsVisit1_child_to_parent_ratio,
                        supTab_resultsLongitudinal_parent,
                        supTab_resultsLongitudinal_child,
                        supTab_resultsLongitudinal_parent_nongran,
                        supTab_resultsLongitudinal_child_nongran,
                        supTab_resultsLongitudinal_child_to_parent_ratio,
                        supTab_resultsLongitudinalDFSO_parent,
                        supTab_resultsLongitudinalDFSO_child,
                        supTab_resultsLongitudinalDFSO_parent_nongran,
                        supTab_resultsLongitudinalDFSO_child_nongran,
                        supTab_resultsLongitudinalDFSO_child_to_parent_ratio)

write_csv(supTab_results, file = file.path(out_dir, "BloodCyTOF_Results.csv"))
```

### Session info
```{r, print-session-info}
sessionInfo()
```
<!--
R version 4.0.2 (2020-06-22)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 18.04.4 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so

locale:
 [1] LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8   
 [6] LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C           LC_TELEPHONE=C        
[11] LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

attached base packages:
 [1] grid      stats4    parallel  stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] GSA_1.03.2            pvca_0.1.0            WGCNA_1.69-81         fastcluster_1.2.3     dynamicTreeCut_1.63-1
 [6] doParallel_1.0.17     doRNG_1.8.2           rngtools_1.5.2        missForest_1.4        itertools_0.1-3      
[11] iterators_1.0.14      foreach_1.5.2         randomForest_4.6-14   ComplexHeatmap_2.6.2  corrr_0.4.3          
[16] gridExtra_2.3         rlang_1.0.2           RColorBrewer_1.1-2    cowplot_1.1.1         ggpubr_0.4.0         
[21] pals_1.7              ggeffects_1.1.1       ggbeeswarm_0.6.0      qvalue_2.22.0         ordinal_2019.12-10   
[26] ggnetwork_0.5.10      igraph_1.2.11         impute_1.64.0         knitr_1.38            forcats_0.5.1        
[31] stringr_1.4.0         dplyr_1.0.8           purrr_0.3.4           readr_2.1.2           tidyr_1.2.0          
[36] tibble_3.1.6          ggplot2_3.3.5         tidyverse_1.3.1       nlme_3.1-147          lme4_1.1-28          
[41] Matrix_1.2-18         GSEABase_1.52.1       graph_1.68.0          annotate_1.68.0       XML_3.99-0.9         
[46] AnnotationDbi_1.52.0  IRanges_2.24.1        S4Vectors_0.28.1      Biobase_2.50.0        BiocGenerics_0.36.1  
[51] edgeR_3.32.1          limma_3.46.0         

loaded via a namespace (and not attached):
  [1] readxl_1.3.1          backports_1.4.1       circlize_0.4.14       Hmisc_4.6-0           plyr_1.8.7           
  [6] splines_4.0.2         digest_0.6.29         htmltools_0.5.2       GO.db_3.12.1          fansi_1.0.3          
 [11] checkmate_2.0.0       magrittr_2.0.2        memoise_1.1.0         cluster_2.1.0         tzdb_0.2.0           
 [16] modelr_0.1.8          matrixStats_0.61.0    jpeg_0.1-9            colorspace_2.0-3      blob_1.2.2           
 [21] rvest_1.0.2           haven_2.4.3           xfun_0.30             crayon_1.5.1          jsonlite_1.8.0       
 [26] survival_3.1-12       glue_1.6.2            gtable_0.3.0          GetoptLong_1.0.5      car_3.0-12           
 [31] shape_1.4.6           maps_3.4.0            abind_1.4-5           scales_1.1.1          DBI_1.1.2            
 [36] rstatix_0.7.0         Rcpp_1.0.8.3          htmlTable_2.4.0       xtable_1.8-4          clue_0.3-60          
 [41] foreign_0.8-79        bit_4.0.4             mapproj_1.2.8         preprocessCore_1.52.1 Formula_1.2-4        
 [46] DT_0.15               htmlwidgets_1.5.1     httr_1.4.2            ellipsis_0.3.2        pkgconfig_2.0.3      
 [51] nnet_7.3-14           dbplyr_2.1.1          locfit_1.5-9.4        utf8_1.2.2            tidyselect_1.1.2     
 [56] reshape2_1.4.4        munsell_0.5.0         cellranger_1.1.0      tools_4.0.2           cli_3.2.0            
 [61] generics_0.1.2        RSQLite_2.2.11        broom_0.7.12          evaluate_0.15         fastmap_1.1.0        
 [66] yaml_2.3.5            bit64_4.0.5           fs_1.5.2              xml2_1.3.3            compiler_4.0.2       
 [71] rstudioapi_0.13       beeswarm_0.4.0        gamm4_0.2-6           png_0.1-7             ggsignif_0.6.3       
 [76] reprex_2.0.1          stringi_1.7.6         highr_0.9             lattice_0.20-41       nloptr_2.0.0         
 [81] vctrs_0.3.8           pillar_1.7.0          lifecycle_1.0.1       GlobalOptions_0.1.2   ucminf_1.1-4         
 [86] data.table_1.14.2     latticeExtra_0.6-29   R6_2.5.1              vipor_0.4.5           codetools_0.2-16     
 [91] dichromat_2.0-0       boot_1.3-25           MASS_7.3-51.6         assertthat_0.2.1      rjson_0.2.21         
 [96] withr_2.5.0           mgcv_1.8-31           hms_1.1.1             rpart_4.1-15          minqa_1.2.4          
[101] rmarkdown_2.13        carData_3.0-5         Cairo_1.5-15          base64enc_0.1-3       numDeriv_2016.8-1.1  
[106] lubridate_1.8.0
-->

