---
title: "Fig5"
output: pdf_document
---

```{r load library}
require(openxlsx)
library(factoextra)
library(cluster)
require(nlme) # directly provides p-values, unlike lme4
require(splines)
require(dplyr)
library(fgsea)
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source(paste0(codebase_path,"/codebase_subtype.R"))

organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")

intermediate_results_outdir_DR = "/scratch/data-integration/molecular_subtyping/DimensionReduction/"
mcia_results = readRDS(paste0(intermediate_results_outdir_DR,"MCIA_phase123_20230726.rds"))

data_path <- "/scratch/data-integration/pipelines/processedData/"
data_env0 <- readRDS(paste0(data_path, "data_env_ravi_20220513.RDS"))

base_obj = readRDS(file = "/scratch/data-integration/cleanedup/mod1_interpretation_base.rds")
```

# Load preliminary results
```{r load subtype assignment and clinical data}
# load subtype assignment and clinical data
nclust_tree_all = 6
subtype_hcl = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
clin_data_aligned = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/clin_data_aligned.rds")

# load omics used
omics_used_DR = c(
  "ppt" = "plasma_proteomics_targeted",
  "ppgd" = "plasma_proteomics_global_dda",
  "so" = "serum_olink",
  "pmg" = "plasma_metabolomics_global",
  "nt" = "nasal_transcriptomics",
  "pbmc" = "pbmc_transcriptomics"
)
omic_short_names = names(omics_used_DR) %>% setNames(omics_used_DR)
omic_names =c("plasma_proteomics_targeted","plasma_proteomics_global_dda",
              "serum_olink","plasma_metabolomics_global","nasal_transcriptomics",
              "pbmc_transcriptomics")

# load all databases used for enrichment
kegg_database = readRDS(file = "/scratch/data-integration/pipelines/Interpretation/remapped_databases/kegg0522.rds")
hallmark_database = readRDS( file = "/scratch/data-integration/pipelines/Interpretation/remapped_databases/hallmark0522.rds")
metabolite_row_DF = data_env0$plasma_metabolomics_global_rowfeature
rownames(metabolite_row_DF) = metabolite_row_DF[["CHEMICAL_NAME"]]
metabolite_row_DF=metabolite_row_DF[,c("CHEMICAL_NAME","SUB_PATHWAY","SUPER_PATHWAY")]
subpath_databse = list()
subpath_databse[["pmg"]] = data.frame(description =metabolite_row_DF[["SUB_PATHWAY"]],
                                      CHEMICAL_NAME = metabolite_row_DF[["CHEMICAL_NAME"]])
for(k in 1:length(kegg_database)){
  kegg_database[[k]]=kegg_database[[k]][which(kegg_database[[k]]!="Ig"),]
}

# load enrichment results
mag_thr=0.3
enrichment_mhg_list0 = readRDS(file=paste0(intermediate_results_outdir_DR, "pathway_enrichment_mhgF1_2_3_10_LG", mag_thr,".rds"))

projection_coef0 = convert_mciaObj_to_annotation_format(mcia_results$mcia_factor_evaluation$projection_coef, omic_names = names(omics_used_DR))
projection_pval0 = convert_mciaObj_to_annotation_format(mcia_results$mcia_factor_evaluation$projection_pval, omic_names = names(omics_used_DR))
K = 10
projection_coef=projection_coef0[1:K]
projection_pval=projection_pval0[1:K]
for(mod_id in c(1:K)){
  mod_name = paste0("mod", mod_id)
  print(mod_name)
  projection_coef[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = projection_coef[[mod_id]], omic_names = omic_names) 
  projection_pval[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = projection_pval[[mod_id]], omic_names = omic_names)
}

###### FACTOR NEGATION IF NEEDED
mod_ids = c(1,2, 3,10)
negation = c(-1, 1, 1, 1)
# negate factor coefficients
for(l in 1:length(mod_ids)){
  mod_id = mod_ids[l]
  mod_name = paste0("mod", mod_id)
  print(mod_name)
  for(i in 1:length(projection_coef[[mod_id]])){
    projection_coef[[mod_id]][[i]][,1] = negation[l]*
      projection_coef[[mod_id]][[i]][,1]
  }
}
# negate enrichment results
enrichment_mhg_list = enrichment_mhg_list0
for(l in 1:length(enrichment_mhg_list)){
  for(i in 1:length(mod_ids)){
    mod_id =mod_ids[i]
    if(negation[i] == -1){
      enrichment_mhg_list[[l]][[mod_id]]$pos = enrichment_mhg_list0[[l]][[mod_id]]$neg
      enrichment_mhg_list[[l]][[mod_id]]$neg = enrichment_mhg_list0[[l]][[mod_id]]$pos
    }
  }
}
mod_ids = c(2, 3)
negation = c(1, 1)
# negate factors
mcia_factors = t(t(mcia_results$mcia_factors[,mod_ids]) * negation)
mcia_factors=apply(mcia_factors,2, function(z) qqnorm(z, plot.it = F)$x)
mcia_factors = data.frame(mcia_factors,row.names = rownames(organized_data$datasets$plasma_proteomics_targeted))
mcia_factors = mcia_factors[clin_data_aligned$event_id,]
colnames(mcia_factors) = paste0("Factor", mod_ids)
```

# Factor trajectory (Figure 5A left & Figure 5E left)
```{r factor trajectory, fig.height=6, fig.width=6, warning=FALSE}
mod_ids = c(1, 2, 3, 10)
negation = c(-1, 1, 1, 1)
# negate factors
mcia_factors = t(t(mcia_results$mcia_factors[,mod_ids]) * negation)
mcia_factors=apply(mcia_factors,2, function(z) qqnorm(z, plot.it = F)$x)
mcia_factors = data.frame(mcia_factors,row.names = rownames(organized_data$datasets$plasma_proteomics_targeted))
mcia_factors = mcia_factors[clin_data_aligned$event_id,]
colnames(mcia_factors) = paste0("Factor", mod_ids)

assay_mat = mcia_factors
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl =subtype_hcl,
                                 clin_data = clin_data_aligned)
clin_colnames_select = setdiff(colnames(assay_mat), colnames(mcia_factors))

factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "label", use_endpoint_levels = c("A", "B", "C", "E", "F"),
  max_date = 40, text_position = 2)
test_Fig5_factor_traj = factor_trajectory_analysis_results$test_pval_list
factor_trajectory_analysis_results$test_fig_list$Factor2
factor_trajectory_analysis_results$test_fig_list$Factor3
```

# Early/late Factor 2 multi-group comparison (Figure 5A right)
```{r early/late Factor 2 multi-group comparison, fig.height=6, fig.width=4}
mod_ids = c(2, 3)
negation = c(1, 1)
# negate factors
mcia_factors = t(t(mcia_results$mcia_factors[,mod_ids]) * negation)
mcia_factors=apply(mcia_factors,2, function(z) qqnorm(z, plot.it = F)$x)
mcia_factors = data.frame(mcia_factors,row.names = rownames(organized_data$datasets$plasma_proteomics_targeted))
mcia_factors = mcia_factors[clin_data_aligned$event_id,]
colnames(mcia_factors) = paste0("Factor", mod_ids)

assay_mat = mcia_factors
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl =subtype_hcl,
                                 clin_data = clin_data_aligned)

discretized_test = list()
feature_name = c("Factor2","Factor3")
discretized_test[["early"]] = multi_group_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V1", "V2", "V3"))
discretized_test[["late"]] = multi_group_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V4", "V5", "V6"))

# Factor 2
est_mat = cbind(discretized_test$early$est[,1],
                discretized_test$late$est[,1])
colnames(est_mat) = c("V1-V3", "V4-V6")
pval_mat = cbind(discretized_test$early$pval[,1],
                 discretized_test$late$pval[,1])
colnames(pval_mat) = c("V1-V3", "V4-V6")
adj_p = apply(pval_mat, 2, function(p) p.adjust(p, method = "BH"))
test_Fig5A = list()
test_Fig5A$estimate = est_mat
test_Fig5A$pval = pval_mat
test_Fig5A$adj_p = adj_p

factor2_multi_comparison = ComplexHeatmap::Heatmap(
  as.matrix(est_mat), row_names_gp = gpar(fontsize = 15), 
  column_names_gp = gpar(fontsize = 15), column_names_rot = 0, 
  column_names_centered = T, 
  heatmap_legend_param = list(title = "Coefficient", direction = "horizontal", 
                              title_position = "topcenter",
                              title_gp = gpar(fontsize = 15), 
                              labels_gp = gpar(fontsize = 15)),
  col = circlize::colorRamp2(c(-max(abs(est_mat)), 0, max(abs(est_mat))), 
                             c("#284AA0", "white", "#BC4A4A")), 
  na_col = "grey90", cluster_columns = F, cluster_rows = F, 
  cell_fun = function(j, i, x, y, w, h, fill) {
    if(adj_p[i, j] < 0.001) {
      grid.text("***", x, y-h*0.08, gp = gpar(cex=2))
      } else if(adj_p[i, j] < 0.01) {
        grid.text("**", x, y-h*0.08, gp = gpar(cex=2))
        } else if(adj_p[i, j] < 0.05) {
          grid.text("*", x, y-h*0.08, gp = gpar(cex=2))
          }
    })
ComplexHeatmap::draw(factor2_multi_comparison, heatmap_legend_side = "top")
```

# Early/late Factor 3 multi-group comparison (Figure 5E right)
```{r early/late Factor 3 multi-group comparison, fig.height=6, fig.width=4}
# Factor 3
est_mat = cbind(discretized_test$early$est[,2],
                discretized_test$late$est[,2])
colnames(est_mat) = c("V1-V3", "V4-V6")
pval_mat = cbind(discretized_test$early$pval[,2],
                 discretized_test$late$pval[,2])
colnames(pval_mat) = c("V1-V3", "V4-V6")
adj_p = apply(pval_mat, 2, function(p) p.adjust(p, method = "BH"))
test_Fig5E = list()
test_Fig5E$estimate = est_mat
test_Fig5E$pval = pval_mat
test_Fig5E$adj_p = adj_p

factor3_multi_comparison = ComplexHeatmap::Heatmap(
  as.matrix(est_mat), row_names_gp = gpar(fontsize = 15), 
  column_names_gp = gpar(fontsize = 15), column_names_rot = 0, 
  column_names_centered = T, 
  heatmap_legend_param = list(title = "Coefficient", direction = "horizontal", 
                              title_position = "topcenter",
                              title_gp = gpar(fontsize = 15), 
                              labels_gp = gpar(fontsize = 15)),
  col = circlize::colorRamp2(c(-max(abs(est_mat)), 0, max(abs(est_mat))), 
                             c("#284AA0", "white", "#BC4A4A")), 
  na_col = "grey90", cluster_columns = F, cluster_rows = F, 
  cell_fun = function(j, i, x, y, w, h, fill) {
    if(adj_p[i, j] < 0.001) {
      grid.text("***", x, y-h*0.08, gp = gpar(cex=2))
      } else if(adj_p[i, j] < 0.01) {
        grid.text("**", x, y-h*0.08, gp = gpar(cex=2))
        } else if(adj_p[i, j] < 0.05) {
          grid.text("*", x, y-h*0.08, gp = gpar(cex=2))
          }
    })
ComplexHeatmap::draw(factor3_multi_comparison, heatmap_legend_side = "top")
```

# Enrichment analysis - Factor 2 (Figure 5B)
```{r enrichment analysis - factor 2, fig.height=6, fig.width=8.5}
KEGG_max = 10; Hallmark_max = 10; Subpath_max = 5;
KEGG_metabo_max = 2; pathway_max = 30; date = 20231201
kegg_pathnames = unique(kegg_database$pbmc$description)
kegg_metabopathnames = kegg_pathnames[grepl("metabolism", kegg_pathnames)]
data_base_names = c("kegg", "hallmark", "subpath")
size_max = c(KEGG_max, Hallmark_max, Subpath_max)
shorten = FALSE
plot_list = list()
test_enrichment = list()
thr = 0.05
pathway_counts = data.frame(matrix(NA, ncol = 6, nrow = length(mod_ids)))
colnames(pathway_counts) = names(enrichment_mhg_list$kegg[[1]]$pos)
rownames(pathway_counts) = paste0("Factor", mod_ids)
for(i in 1:length(mod_ids)){
  mod_id = mod_ids[i]
  factor_enrichment = lapply(enrichment_mhg_list, function(z) z[[mod_id]])
  aggregated_mat = enrichment_aggregation_subtype(factor_enrichment[1:3], 
                                                  adj.p.thr = thr, 
                                                  thr = .2, type = "mhg")
  temp = aggregated_mat[,c(1:6)]
  temp[is.na(temp)] = 0
  pathway_counts[i,]=apply(abs(temp) > (-log(thr, base = 10)), 2, sum)
  if(shorten) {
    aggregated_mat_short = pathway_shortlist_selection(
      aggregated_mat, data_base_names = data_base_names, 
      size_max = size_max, KEGG_metabo_max = KEGG_metabo_max)
  } else {
    aggregated_mat_short =  aggregated_mat
  }
  aggregated_mat_short = aggregated_mat_short
  aggregated_mat_short = aggregated_mat_short[order(-aggregated_mat_short$joint),]
  
  plot_title = paste0("Factor",mod_id)
  plotDF_df = gather(
    aggregated_mat_short[1:min(nrow(aggregated_mat_short), pathway_max),], 
    key = "dataset", value = "log10.sig.val", -joint, -pathway, -database)
  plotDF_df = plotDF_df[!is.na(plotDF_df$log10.sig.val),]
  plotDF_df$direction = NA
  plotDF_df$direction[plotDF_df$log10.sig.val>0] = "pos"
  plotDF_df$direction[plotDF_df$log10.sig.val<0] = "neg"
  plotDF_df$direction[plotDF_df$dataset == "joint_adj"] = "unsigned"
  plotDF_df$log10.sig.val = abs(plotDF_df$log10.sig.val)
  
  plotDF_df$pathway = factor(
    plotDF_df$pathway, 
    levels = aggregated_mat_short$pathway[length(aggregated_mat$pathway):1])
  
  levels(plotDF_df$pathway) = gsub("HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION", 
                                  "HALLMARK_EPITHELIAL_\nMESENCHYMAL_TRANSITION", 
                                  levels(plotDF_df$pathway))
  
  plotDF_df$dataset = factor(plotDF_df$dataset, 
                             levels = c(omic_short_names, "joint_adj"))
  plotDF_df$database = factor(plotDF_df$database, 
                              levels = data_base_names)
  test_enrichment[[i]] = plotDF_df
  plot_list[[i]] = 
    ggplot(data = plotDF_df, aes(x = dataset, y = pathway, 
                                 size = log10.sig.val, col = direction)) +
    geom_point(color = "white") + 
    geom_point(shape = 1) + 
    geom_point(data =. %>% dplyr::filter(log10.sig.val > -log10(thr))) + 
    ggtitle(plot_title) + 
    theme_classic() +
    theme(panel.grid.major = element_line(color="grey", size=0.1), 
          axis.text.x = element_text(angle = 90, hjust=1, vjust=0.5, 
                                     size = 15, color = "black"), 
          axis.text.y = element_text(size = 15, color = "black"),
          strip.text = element_text(color = "black", size = 12.5),
          plot.title = element_text(color = "black", size = 15),
          legend.title = element_text(color = "black", size = 15),
          legend.text = element_text(color = "black", size = 15)) + 
    ylab("") + 
    xlab("") +
    scale_color_manual(values = c("neg" ="#284AA0", "pos" ="#BC4A4A",
                                  "unsigned"= "#808080"),
                       labels = c("negative", "positive", "unsigned"),
                       name = "Direction") +
    facet_grid(database~., scales = "free_y", space = "free_y") +
    guides(size = guide_legend(title = expression("Signed "*log[10]*"(p.val)"), 
                               order = 1))
  
  plotDF_df$pathway = factor(
    plotDF_df$pathway,
    levels = aggregated_mat_short$pathway[1:length(aggregated_mat$pathway)])
   plotDF_df$dataset = factor(
     plotDF_df$dataset,
     levels = c(omic_short_names[length(omic_short_names):1], "joint_adj"))
  p_horizontal = ggplot(data = plotDF_df, aes(x = pathway, y = dataset,
                                              size = log10.sig.val,
                                              col = direction)) +
    geom_point(color = "white") + 
    geom_point(shape = 1) + 
    geom_point(data =. %>% dplyr::filter(log10.sig.val > -log10(thr))) + 
    ggtitle(plot_title) + 
    theme_classic() +
    theme(panel.grid.major = element_line(color = "grey", size = 0.1), 
          axis.text.x = element_text(angle = 90, hjust = 1, size = rel(1.5)), 
          axis.text.y = element_text(size = rel(1.5))) +
    ylab(NULL) + 
    scale_color_manual(values = c("neg" ="#284AA0", "pos" = "#BC4A4A", 
                                  "unsigned"= "#808080")) +
    facet_grid(.~database, scales = "free_x", space = "free_x")
  # extract core enrichment
  aggregated_mat_core_enrichment0 = aggregated_mat_short[,c(1:6, 9, 10)]
  aggregated_mat_core_enrichment = aggregated_mat_short[,c(1:6, 9, 10)]
  for(j in 1:nrow(aggregated_mat_core_enrichment)){
    database0 = aggregated_mat_core_enrichment$database[j]
    pathway0 = aggregated_mat_core_enrichment$pathway[j]
    temp_pos = lapply(enrichment_mhg_list[[database0]][[mod_id]]$pos, 
                      function(z) z[z$ID==pathway0,])
    temp_neg = lapply(enrichment_mhg_list[[database0]][[mod_id]]$neg, 
                      function(z) z[z$ID==pathway0,])
    for(dataset_id in 1:6){
      dataset_name = omic_short_names[dataset_id]
      if(!is.na(aggregated_mat_core_enrichment0[j,dataset_id])){
        if((aggregated_mat_core_enrichment0[j,dataset_id]) > 0){
           aggregated_mat_core_enrichment[j,dataset_id] = 
             temp_pos[[dataset_name]]$core_enrichment
        }else if((aggregated_mat_core_enrichment0[j,dataset_id]) < 0){
          aggregated_mat_core_enrichment[j,dataset_id] = 
            temp_neg[[dataset_name]]$core_enrichment
        }
      }
    }
  }
  colnames(aggregated_mat_core_enrichment)[1:6] = 
    paste0("core_enrich_",colnames(aggregated_mat_core_enrichment)[1:6])
  aggregated_mat_short = left_join(aggregated_mat_short, 
                                   aggregated_mat_core_enrichment,
                                   by = c("pathway","database"))
}
#         ppt ppgd so pmg nt pbmc
#Factor2   1    4  0   7  0    0
#Factor3   1    4  1   3  0    0
print(pathway_counts)
test_Fig5B = test_enrichment[[1]]
test_Fig5F = test_enrichment[[2]]

plot_list[[1]]
```

# Enrichment analysis - Factor 3 (Figure 5F)
```{r enrichment analysis - factor 3, fig.height=6, fig.width=7}
plot_list[[2]]
```

# Pathway activity trajectory - Factor 2 (Figure 5D)
```{r pathway activity trajectory - factor 2, fig.height=6, fig.width=12, warning=FALSE}
pathway_prot = c("Complement and coagulation cascades",
                 "Cardiac muscle contraction" )
pathway_met = c("Plasmalogen",
                "Long Chain Polyunsaturated Fatty Acid (n3 and n6)", 
                "Tryptophan Metabolism",
                "Acetylated Peptides")
pathways0 = c(pathway_prot,pathway_met)
protein_list = kegg_database$ppgd[kegg_database$ppgd$description%in%pathways0,]
metabolite_list = subpath_databse$pmg[subpath_databse$pmg$description%in%pathways0,]
protein_list = unique(protein_list)
metabolite_list = unique(metabolite_list)
colnames(metabolite_list) = colnames(protein_list)
flist = rbind(protein_list,metabolite_list)
# for factor 1 and factor 4 and the weights
weight_comparisons = list()
for(dataset_id in c(1:6)){
  temp =intersect(rownames(projection_coef$mod1[[dataset_id]]),flist$gene_name)
  weight_comparisons[[dataset_id]] = data.frame(feature =   temp)
  for(mod_id in (1:3)){
    weight_comparisons[[dataset_id]][[paste0("Factor", mod_id)]] =
      projection_coef[[mod_id]][[dataset_id]][ temp,]
  }
  temp2 = flist
  rownames( temp2) = temp2$gene_name
  weight_comparisons[[dataset_id]]$pathway = temp2[temp,]$description
}
names(weight_comparisons) = names(projection_coef$mod1)

dataDF = organized_data$datasets$plasma_proteomics_global_dda
colnames(dataDF) = rownames(projection_coef$mod1$ppgd)
thrs = matrix(0, ncol =2, nrow = 2)
thrs[1,] = c(1,-1) #cog
thrs[2,] = c(-1,1) #muscle
thrs = thrs*mag_thr
pathway_score_prot = NULL
pathways1 = pathway_prot
i=0
for(mod_id in c(2,3)){
  i=i+1
  for(j in 1:length(pathways1)){
    temp1 = weight_comparisons[[2]][weight_comparisons[[2]]$pathway==pathways1[j],]
    temp1 = temp1[temp1[,paste0("Factor",mod_id)]/thrs[j,i] >=1,]
    temp2 = dataDF[,(temp1$feature)]
    temp3 = as.matrix(temp2)%*%abs(temp1[[paste0("Factor",mod_id)]])
    temp3 = data.frame(value = temp3)
    colnames(temp3)= paste0(pathways1[j],"_Factor",mod_id)
    if(is.null(pathway_score_prot)){
     pathway_score_prot = temp3 
    }else{
      pathway_score_prot = cbind(pathway_score_prot, temp3 )
    }
  }
}
ll = apply(pathway_score_prot,2,var, na.rm = T)
pathway_score_prot=pathway_score_prot[,ll>0]
ii = !is.na(pathway_score_prot[,1])
pathway_score_prot[ii,] = apply(pathway_score_prot[ii,], 2,
                                function(z) qqnorm(z, plot.it = F)$x)
pathway_score_prot$event_id = rownames(dataDF)


dataDF = organized_data$datasets$plasma_metabolomics_global
colnames(dataDF) = rownames(projection_coef$mod1$pmg)
thrs = matrix(0, ncol =3, nrow = 4)
thrs[1,] = c(-1,1,1) # Plasmalogen
thrs[2,] = c(1,-1,-1) # fatty acid
thrs[3,] = c(1,-1,1) # Peptide
thrs[4,] = c(1,-1,1) # Trptophan
thrs = thrs*mag_thr
pathway_score_met = NULL
pathways1 = pathway_met
for(mod_id in c(2,3)){
  i = mod_id
  for(j in 1:length(pathways1)){
    temp1 = weight_comparisons[[4]][weight_comparisons[[4]]$pathway==pathways1[j],]
    temp1 = temp1[temp1[,paste0("Factor",mod_id)]/thrs[j,i] >=1,]
    temp2 = dataDF[,(temp1$feature)]
    temp3 = as.matrix(temp2)%*%abs(temp1[[paste0("Factor",mod_id)]])
    temp3 = data.frame(value = temp3)
    colnames(temp3)= paste0(pathways1[j],"_Factor",mod_id)
    if(is.null(pathway_score_met)){
     pathway_score_met = temp3 
    } else {
      pathway_score_met = cbind(pathway_score_met, temp3 )
    }
  }
}
ll = apply(pathway_score_met,2,var, na.rm = T)
pathway_score_met = pathway_score_met[,ll>0]
ii = !is.na(pathway_score_met[,1])
pathway_score_met[ii,] = apply(pathway_score_met[ii,], 2,
                               function(z) qqnorm(z, plot.it = F)$x)

pathway_score_met$event_id = rownames(dataDF)

pathway_score_select = cbind(pathway_score_prot[,c(1,3,4)], 
                             pathway_score_met[,c(1,2,3)])
colnames(pathway_score_select ) = c("ComplementAndCoagulation_Factor2",
                                    "CardiacMuscleContraction_Factor3",
                                    "event_id",
                                    "LC_PolyunsaturatedFattyAcid_Factor2",
                                    "TryptophanMetabolism_Factor3", 
                                    "AcetylatedPeptides_Factor3")
  
dataDF = pathway_score_select
dataDF = dataDF[clin_data_aligned$event_id,]
dataDF = cytof_mat_preprocess(cytof_mat = dataDF,
                              subtype_hcl = subtype_hcl,
                              clin_data = clin_data_aligned)
assay_mat = dataDF
clin_colnames_select = c(setdiff(colnames(assay_mat),
                                 colnames(pathway_score_select)),"event_id")

pathway_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "label", use_endpoint_levels = c("A", "B", "C", "E", "F"),
  max_date = 40, feature_with_ylab = c("ComplementAndCoagulation_Factor2",
                                       "CardiacMuscleContraction_Factor3"),
  text_position = 2)
test_Fig5_pathway_traj = pathway_trajectory_analysis_results$test_pval_list
ggarrange(plotlist = pathway_trajectory_analysis_results$test_fig_list[
  c("ComplementAndCoagulation_Factor2", "LC_PolyunsaturatedFattyAcid_Factor2")
], nrow = 1, common.legend = T, legend = "bottom")
```

# Pathway activity trajectory - Factor 3 (Figure 5G)
```{r pathway activity trajectory - factor 3, fig.height=6, fig.width=18, warning=FALSE}
ggarrange(plotlist = pathway_trajectory_analysis_results$test_fig_list[
  c("CardiacMuscleContraction_Factor3", "TryptophanMetabolism_Factor3",
    "AcetylatedPeptides_Factor3")], 
  nrow = 1, common.legend = T, legend = "bottom")
```

# CC protein comparison for Factor 1 and Factor 2 (Figure 5C)
```{r CC protein comparison, fig.height=6, fig.width=4.5}
coef_mat = weight_comparisons[[2]][weight_comparisons[[2]]$pathway %in% 
                                     c("Complement and coagulation cascades"),
                                   c(1,2,3)]
coef_mat = coef_mat[apply(abs(coef_mat[,-1]),1,max)>mag_thr,]
coef_mat = coef_mat[(rank(-coef_mat[,3])<=5)|(rank(-coef_mat[,2])<=5),]
temp1 = coef_mat[(rank(-coef_mat[,3])<=5),]
temp1 = temp1[order(-temp1$Factor2),]
feature_levels = temp1$feature
temp1 = coef_mat[(rank(-coef_mat[,2])<=5),]
temp1 = temp1[order(temp1$Factor1),]
feature_levels = c(feature_levels, setdiff(temp1$feature,feature_levels))
plotDF = coef_mat
plotDF$feature = factor(plotDF$feature, 
                        levels = feature_levels[length(feature_levels):1])

plotDF = gather(plotDF, key = "Factor", value = "coefficient", -feature)
plotDF$abs_coef = abs(plotDF$coefficient)
plotDF$sign = ifelse(plotDF$coefficient > 0, "Positive", "Negative")
plotDF$plot_coef = ifelse(plotDF$Factor == "Factor2", 
                          -plotDF$abs_coef, plotDF$abs_coef)
test_Fig5C = plotDF
ggplot(plotDF, aes(x =plot_coef , y = feature, group = feature, color = sign)) +
  geom_segment(aes(xend = 0, yend =  feature), linewidth = 1.5) +
  geom_point(aes(x = plot_coef), size = 3) +
  scale_color_manual(values = c("Positive" = "#BC4A4A", 
                                "Negative" = "#284AA0")) +
  labs(x = "Absolute Value of Coefficient", y = "Feature", 
       title = "") +
  theme_minimal()  +
  geom_vline(xintercept = 0, linetype = "solid") +
  geom_hline(yintercept = -Inf, linetype = "solid") +
  theme(axis.ticks.x = element_line(linewidth = 0.5),
        axis.line.x = element_line(linewidth = 1),
        axis.line.y = element_line(linewidth = 1),
        axis.ticks = element_line(linewidth = 1),
        axis.text.x = element_text(color = "black", size = 15),  
        axis.text.y = element_text(color = "black", size = 15),  
        axis.title = element_text(color = "black", size = 15),
        axis.ticks.length = unit(0.15, "cm"),
        plot.subtitle = element_text(color = "black", hjust = 0.5, 
                                     size = 15),
        legend.text = element_text(color = "black", size = 15),
        legend.title = element_text(color = "black", size = 15)) +
  annotate("text", x = -0.2, y = Inf, label = "Factor 2", vjust = 1, 
           hjust = 1, size = 5.5) +
  annotate("text", x = 0.2, y = Inf, label = "Factor 1", vjust = 1, 
           hjust = 0, size = 5.5) + 
  scale_x_continuous(labels = function(x) abs(x))+
  annotate("segment", x = -.5, xend = -.7, y = -Inf, yend = -Inf) +
  annotate("segment", x = .5, xend = .7, y = -Inf, yend = -Inf) + 
  theme(legend.position = "bottom")
```




