---
title: "Hospital-admission cis-eQTLs: cohort definition and genotype extraction"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(dplyr)
library(data.table)
library(tidyverse)
library(tidyr)
library(tibble)
library(readxl)
library(qs2)
library(SNPRelate)
library(GENESIS)
library(magrittr)
library(GWASTools)
library(nFactors)
```

## Paths and parameters

All hardcoded absolute paths from the original notebook are collected here.
Edit these values to match your environment; everything downstream is built
from them with `file.path()`.

```{r paths-and-parameters}
# ---- EDIT THESE FOR YOUR ENVIRONMENT ----
# Root of the raw IMPACC genotype/expression data env.
base_dir <- "/path/to/impacc_data/"

# Serialized IMPACC data environment (clinical + omics).
data_env_path <- file.path(base_dir, "impacc_data_env.RDS")

# Post-QC admixture phenotype file (maps participant_id to core_specific_id).
admixture_pheno_path <- "/path/to/impacc_admixture_pheno.txt"

# Published critical COVID-19 GWAS lead-SNP table.
published_snps_path <- "/path/to/All_Severe_COVID_GWAS_hits.xlsx"

# Analysis output / results root (this is the original panEQTL_dir).
results_dir <- "/path/to/hospital_admission_results"
panEQTL_dir <- results_dir
# -----------------------------------------

data_env <- readRDS(data_env_path)
```

## Set up the GWAS cohort (no trajectory-group stratification)

Define the unstratified GWAS cohort: all GWAS samples that passed genotyping
QC, with age and sex recorded. Sex is recoded numerically (0 = Female,
1 = Male), and the cohort is intersected with the post-QC admixture sample list.

```{r gwas-cohort}
gwas_cohort <- data_env$clinical_data %>% # 1255 GWAS samples
    filter(sample_type == "GWAS") %>%
    select(participant_id, trajectory_group, admit_age, sex) %>%
    arrange(participant_id) %>%
    mutate(
    # Convert sex to a numeric variable where 0 = Female and 1 = Male.
    sex = ifelse(sex == "Male", 1, 0)) %>%
    distinct() #1112 distinct GWAS samples

# post-QC IMPACC cohort
fam <- fread(admixture_pheno_path) %>%
    select(participant_id, FID) %>%
    rename(core_specific_id = FID)

final_gwas_cohort <- gwas_cohort %>% # 1031 samples 
    inner_join(fam, by = "participant_id") %>%
    relocate(core_specific_id, .before = "trajectory_group")

#   1   2   3   4   5 
# 204 230 303 195  99 

qs_save(final_gwas_cohort, file.path(panEQTL_dir, "impacc_panEQTL_cohort.qs2"))
```

## Convert the genotype GDS to PLINK format

Convert the IMPACC GDS genotype file to PLINK BED/BIM/FAM using SNPRelate.

```{r gds-to-plink}
## ---- gds-to-plink ---------------------------------------------------------

# path to your GDS file
gds_fn <- file.path(base_dir, "impacc_geno.gds")

# create an output directory for PLINK files
plink_dir <- file.path(base_dir, "plink")
if (!dir.exists(plink_dir)) dir.create(plink_dir, recursive = TRUE)

# prefix for PLINK output (creates *.bed, *.bim, *.fam)
plink_prefix <- file.path(plink_dir, "impacc_geno")

# open the GDS file
gds <- snpgdsOpen(gds_fn, readonly = TRUE)

# optional: subset samples or SNPs here
# samp_keep <- readLines("sample_ids.txt")      # example
# snp_keep  <- readLines("snp_ids.txt")         # example

## THIS FIRST STEP CAUSES an ALT/REF Allele error.

# convert to PLINK
snpgdsGDS2BED(
  gds,
  bed.fn = paste0(plink_prefix),   # .bim and .fam are created automatically
  # sample.id = samp_keep,                 # uncomment if sub-setting
  # snp.id    = snp_keep,
  verbose = TRUE
)

snpgdsClose(gds)
```

## Define the extraction ranges around the published GWAS loci

Build the +/- 1 Mb range list around each published critical COVID-19 GWAS
lead SNP (GenOMICC PMID 37198478, HGI PMID 37674002). The published list is 69
non-overlapping SNPs. Both a point-position list (snps_to_extract.txt) and a
2 Mb-window list (ranges_to_extract.txt) are written for PLINK. This step does
not need to be re-run every time.

```{r gwas-ranges}
published_snps <- read_excel(published_snps_path)

snps_to_extract <- published_snps %>%
        pivot_longer(cols = everything(), 
                    values_to = "chr:pos") %>%
        filter(!is.na(`chr:pos`)) %>%
        distinct(`chr:pos`) %>%
        separate(`chr:pos`, into = c("chr", "pos"), sep = ":") %>%
        mutate(pos = as.numeric(pos),
                start = pos,
                end = pos) %>%
        arrange(chr, start) %>%
        mutate(ranges = paste0("R", 1:n())) %>%
        select(chr, start, end, ranges)

# 69 published, non-overlapping critical COVID-19 GWAS SNPs  
write.table(snps_to_extract, 
            file = file.path(panEQTL_dir, "snps_to_extract.txt"),
            quote = FALSE, 
            row.names = FALSE, 
            col.names = FALSE, 
            sep = "\t")

# Expand the extraction range to include 2 Mb around all SNPs
ranges_to_extract <- published_snps %>%
        pivot_longer(cols = everything(), 
                    values_to = "chr:pos") %>%
        filter(!is.na(`chr:pos`)) %>%
        distinct(`chr:pos`) %>%
        separate(`chr:pos`, into = c("chr", "pos"), sep = ":") %>%
        mutate(pos = as.numeric(pos),
                start = pos - 1000000,
                end = pos + 1000000) %>%
        arrange(chr, start) %>%
        mutate(ranges = paste0("R", 1:n())) %>%
        select(chr, start, end, ranges)

write.table(ranges_to_extract, 
            file = file.path(panEQTL_dir, "ranges_to_extract.txt"),
            quote = FALSE, 
            row.names = FALSE, 
            col.names = FALSE, 
            sep = "\t")        
```

The actual PLINK `--extract range` run that consumes `ranges_to_extract.txt`
(299,380 variants extracted across the 69 ranges) lives in
`00-extract_gwas_loci.sh`.

