---
title: "Hospital-admission cis-eQTLs: genotype covariates and cis-eQTL mapping"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(MatrixEQTL)
library(dplyr)
library(data.table)
library(tidyr)
library(tibble)
library(readxl)
library(qs2)
library(SNPRelate)
library(GENESIS)
library(magrittr)
library(GWASTools)
library(nFactors)
```

## Paths and parameters

All hardcoded absolute paths from the original notebook are collected here.
Edit these values to match your environment; everything downstream is built
from them with `file.path()`.

```{r paths-and-parameters}
# ---- EDIT THESE FOR YOUR ENVIRONMENT ----
# Analysis output / results root (this is the original panEQTL_dir).
results_dir <- "/path/to/hospital_admission_results"
panEQTL_dir <- results_dir

# Directory holding the order-checked MatrixEQTL inputs (written by the order-check step
# in this notebook).
final_dir <- file.path(panEQTL_dir, "final_files")

# Filename prefix shared by all MatrixEQTL input files for this visit.
prefix <- "impacc_panEQTL_visit1"
# -----------------------------------------
```



## Filter and format the genotype data for MatrixEQTL

Write the list of core-specific sample IDs to keep, then (in PLINK, below)
produce the A-transpose genotype matrix, the binary set for PCA, and the MAP
file of SNP locations. The R block then reformats the `.traw` genotype matrix
and `.map` file into the SNP.txt and snploc.txt inputs MatrixEQTL expects.

```{r matrixeqtl-sample-list}
# Sample IDs are taken from the inverse-normalized expression matrix written by
# 02-expression_processing.Rmd, so run that notebook first.
core_specific_ids <- colnames(fread(file.path(panEQTL_dir, paste0(prefix, "_GE.txt"))))[-1]  # sample IDs from 02-expression_processing.Rmd

sample_ids_df <- data.frame(
    Column1 = rep(0, length(core_specific_ids)),   # all zeros
    Column2 = core_specific_ids                    # core-specific IDs
)
            
# Write the sample extraction file (tab-delimited, no quotes, no row/column names)
write.table(sample_ids_df, file = file.path(panEQTL_dir, "visit1_panEQTL_coreIDs.txt"), quote = FALSE, col.names = FALSE, row.names = FALSE)
```

The following PLINK calls build the MatrixEQTL genotype inputs. They were
inline bash in the original notebook; edit the leading variables to match your
environment before running.

```{bash matrixeqtl-genotype-plink, eval=FALSE}
# Common paths (edit for your environment)
bfile=/path/to/impacc_data/plink/impacc_geno
keep=/path/to/hospital_admission_results/visit1_panEQTL_coreIDs.txt
ranges=/path/to/hospital_admission_results/ranges_to_extract.txt
outdir=/path/to/hospital_admission_results

# 1) Genotype matrix for MatrixEQTL (A-transpose) 

## FLIP FLIP here: https://www.cog-genomics.org/plink/2.0/formats#traw vs https://www.cog-genomics.org/plink/1.9/formats#traw

plink2 --bfile "$bfile" \
       --keep  "$keep" \
       --extract range "$ranges" \
       --recode A-transpose \
       --out   "$outdir/visit1_panEQTL"

# 2) Binary files for PCA
plink2 --bfile "$bfile" \
       --keep  "$keep" \
       --make-bed \
       --out   "$outdir/visit1_panEQTL_PCA"

# 3) MAP file for SNP locations (PED discarded afterwards)
plink2 --bfile "$bfile" \
       --keep  "$keep" \
       --extract range "$ranges" \
       --export ped \
       --out   "$outdir/visit1_panEQTL_ped"  # produces .ped + .map
```

```{r format-snp-files}
######### Create the SNP.txt genotype file and the _snploc.txt file. 

traw_file <- file.path(panEQTL_dir, "visit1_panEQTL.traw")
map_file  <- file.path(panEQTL_dir, "visit1_panEQTL_ped.map")

prefix <- "impacc_panEQTL_visit1"

# Build output file paths:
snp_out_file    <- file.path(panEQTL_dir, paste0(prefix, "_SNP.txt"))
snploc_out_file <- file.path(panEQTL_dir, paste0(prefix, "_snploc.txt"))
            
# ----- Process the .traw file -----
# Read the .traw file and remove unwanted columns.
geno <- fread(traw_file) %>%
        select(-CHR, -`(C)M`, -POS, -COUNTED, -ALT) %>%
        column_to_rownames(var = "SNP")

# Adjust column names: drop leading "0_" and collapse duplicated fragments like "ID_ID"
new_colnames <- vapply(colnames(geno), function(name) {
  # drop leading "0_" if present
  if (startsWith(name, "0_"))
    name <- sub("^0_", "", name)

  # collapse duplicated fragments like "ID_ID"
  parts <- strsplit(name, "_", fixed = TRUE)[[1]]
  if (length(parts) > 1 && parts[1] == parts[2])
    return(parts[1])

  name
}, character(1))
colnames(geno) <- new_colnames

# Write the reformatted genotype (SNP) file
write.table(geno, file = snp_out_file, sep = "\t", col.names = NA, quote = FALSE)

# ----- Process the .map file for SNP locations -----
snploc <- fread(map_file) %>%
select(-V3) %>%
rename(chr = V1, snp = V2, pos = V4) %>%
select(snp, chr, pos) %>%
mutate(chr = paste0("chr", chr))

# Reorder snploc so that SNPs are in the same order as the row names in 'geno'
snploc <- snploc[match(rownames(geno), snploc$snp), ]

# Write the snploc file
write.table(snploc, file = snploc_out_file, sep = "\t", col.names = TRUE, 
            quote = FALSE, row.names = FALSE)
```

## Genotype principal components via PC-AiR

Compute genotype PCs with PC-AiR (GENESIS) to capture population structure
while accounting for cryptic relatedness. The PLINK PCA binaries are converted
to GDS, LD-pruned, and a KING kinship matrix is built; PC-AiR is then run, and
nScree picks the number of PCs to retain via optimal coordinates.

```{r genotype-pcs-pcair}
# Do this in GENESIS; R 4.4.3

# Construct the file paths for the PLINK files.
# The PLINK PCA files follow the pattern:
#   visit1_panEQTL_PCA.bed, .fam, .bim

prefix <- "visit1_panEQTL_PCA"

bed.fn <- file.path(panEQTL_dir, paste0(prefix, ".bed"))
fam.fn <- file.path(panEQTL_dir, paste0(prefix, ".fam"))
bim.fn <- file.path(panEQTL_dir, paste0(prefix, ".bim"))
out.gdsfn <- file.path(panEQTL_dir, paste0(prefix, ".gds"))
    
# Convert PLINK files to GDS.
# For speed, set compress.annotation = "NONE" (lowest compression) and leave compress.geno empty.
snpgdsBED2GDS(bed.fn, fam.fn, bim.fn, out.gdsfn,
                family = FALSE, snpfirstdim = NA, 
                compress.annotation = "",
                compress.geno = "",
                option = NULL,
                cvt.chr = c("int", "char"),
                cvt.snpid = c("auto", "int"),
                verbose = TRUE)

# Open the GDS file
gds_obj <- snpgdsOpen(out.gdsfn, readonly = TRUE)

# LD-pruning: get a pruned set of SNPs to use in PCA (reduces redundancy)
set.seed(1234)
ld_pruned <- snpgdsLDpruning(gds_obj, ld.threshold = 0.1, missing.rate = 0.015,
                                maf = 0.025, method = "r", num.thread = 4)
pruned <- unlist(ld_pruned, use.names = FALSE)

# Read sample IDs from the GDS file
sample.id <- read.gdsn(index.gdsn(gds_obj, "sample.id"))

# Compute pairwise relatedness using the KING method (robust to stratification)
ibd.robust <- snpgdsIBDKING(gds_obj, sample.id = sample.id, family.id = sample.id, num.thread = 4)

# Convert the KING output to a kinship matrix
kingmat <- kingToMatrix(ibd.robust, sample.include = NULL, thresh = 2^(-11/2), verbose = TRUE)

# Close the GDS file to free up resources
snpgdsClose(gds_obj)

# Re-open the GDS file as a GenotypeData object
geno <- GdsGenotypeReader(filename = out.gdsfn) %>% GenotypeData(.)

# Run PC-AiR using the kinship matrix and pruned SNPs
mypcair <- pcair(geno, kinobj = kingmat, divobj = kingmat, snp.include = pruned, num.cores = 4)

# Determine how many PCs were computed:
nComputed <- ncol(mypcair$vectors)
nToExtract <- min(32, nComputed)

geno_mypcs <- data.frame(mypcair$vectors[, 1:nToExtract])
eig <- mypcair$values[1:nToExtract]

# Run nScree to get a data-driven recommendation for the number of PCs
scree_out <- nScree(eig)
# We'll use the "optimal coordinates" (noc) as our recommendation.
num_pcs <- scree_out$Components$noc
    
# Extract the recommended number of PCs from the PC-AiR output.
# (Here, we extract columns 1 to num_pcs from the geno_mypcs data frame.)
geno_covariates <- geno_mypcs[, 1:num_pcs, drop = FALSE] %>%
    as.data.frame() %>%
    setNames(paste0("geno_pc", 1:num_pcs)) %>%
    t() %>% as.data.frame()

# Clean up objects to free memory and force garbage collection
rm(gds_obj, geno, mypcair, geno_mypcs, ld_pruned, pruned, kingmat, ibd.robust)
gc()
        
qs_save(geno_covariates, file.path(panEQTL_dir, "final_genotype_pcs.qs2"))
```

## Assemble the final covariate file

Combine the demographic covariates (age, sex), the genotype PCs, and the
expression PCs (written by 02-expression_processing.Rmd) into one covariate
matrix, keeping only samples present in all three.

```{r combine-covariates}
nonPC_covars <- qs_read(file.path(panEQTL_dir, "impacc_panEQTL_cohort.qs2")) %>%
                select(-participant_id, -trajectory_group) 

rna_covars <- qs_read(file.path(panEQTL_dir, "final_expression_pcs.qs2")) %>%
              t() %>%  
              as.data.frame() %>% 
              rownames_to_column(var = "core_specific_id")

geno_covars <- qs_read(file.path(panEQTL_dir, "final_genotype_pcs.qs2")) %>%
               t() %>%  
               as.data.frame() %>% 
               rownames_to_column(var = "core_specific_id")

# Merge the two dataframes by "core_specific_id" to ensure matching samples.
# This inner join ensures only samples present in both datasets are retained.
merged_df <- inner_join(geno_covars, rna_covars, by = "core_specific_id") %>%
# Add the demographic covariates by merging with nonPC_covars
# Inner join to ensure we only keep samples with all necessary data
             inner_join(nonPC_covars, by = "core_specific_id") %>%
             column_to_rownames(var = "core_specific_id") %>% 
             t()

write.table(merged_df, file = file.path(panEQTL_dir, "impacc_panEQTL_visit1_Covariates.txt"), 
                       sep = "\t", quote = FALSE, row.names = TRUE, col.names = NA)
```

## Column-ordering checks and final MatrixEQTL inputs

Sample columns must appear in exactly the same order across the expression,
genotype, and covariate matrices, and genes/SNPs must match their location
tables. This chunk reads the per-file inputs, enforces matching gene, SNP, and
sample order (stopping on any mismatch), and writes the reordered files into
`final_files/`, which the MatrixEQTL run below loads.

```{r column-ordering-checks}
prefix <- "impacc_panEQTL_visit1"

# Define the output directory for the final, reordered files.
final_dir <- file.path(panEQTL_dir, "final_files")
if (!dir.exists(final_dir)) {
  dir.create(final_dir, recursive = TRUE)
}

# Construct input file paths using the naming convention.
ge_file      <- file.path(panEQTL_dir, paste0(prefix, "_GE.txt"))
geneloc_file <- file.path(panEQTL_dir, paste0(prefix, "_geneloc.txt"))
snp_file     <- file.path(panEQTL_dir, paste0(prefix, "_SNP.txt"))
snploc_file  <- file.path(panEQTL_dir, paste0(prefix, "_snploc.txt"))
covar_file   <- file.path(panEQTL_dir, paste0(prefix, "_Covariates.txt"))
    
# Read input files.
ge      <- fread(ge_file) %>% column_to_rownames(var = "V1")
geneloc <- fread(geneloc_file)
snp     <- fread(snp_file) %>% column_to_rownames(var = "V1")
snploc  <- fread(snploc_file)
covar   <- fread(covar_file) %>% column_to_rownames(var = "V1")
    
# ----- 1. Ensure Gene Order Matches -----
if(!(all(geneloc$geneid %in% rownames(ge)) && all(rownames(ge) %in% geneloc$geneid))) {
    stop("Gene sets do not match")
}
ge <- ge[match(geneloc$geneid, rownames(ge)), ]

# ----- 2. Ensure SNP Order Matches -----
if(!(all(snploc$snp %in% rownames(snp)) && all(rownames(snp) %in% snploc$snp))) {
    stop("SNP sets do not match")
}
snp <- snp[match(snploc$snp, rownames(snp)), ]

# ----- 3. Ensure Sample (Column) Order Matches -----
common_samples <- Reduce(intersect, list(colnames(ge), colnames(snp), colnames(covar)))
if(length(common_samples) == 0) {
    stop("No common sample IDs found")
}
ge    <- ge[, common_samples, drop = FALSE]
snp   <- snp[, common_samples, drop = FALSE]
covar <- covar[, common_samples, drop = FALSE]

if(!(identical(colnames(ge), colnames(snp)) && identical(colnames(ge), colnames(covar)))) {
    stop("Sample order mismatch")
}

# Construct output file paths (preserve naming convention).
out_ge_file      <- file.path(final_dir, paste0(prefix, "_GE.txt"))
out_snp_file     <- file.path(final_dir, paste0(prefix, "_SNP.txt"))
out_covar_file   <- file.path(final_dir, paste0(prefix, "_Covariates.txt"))
out_geneloc_file <- file.path(final_dir, paste0(prefix, "_geneloc.txt"))
out_snploc_file  <- file.path(final_dir, paste0(prefix, "_snploc.txt"))

# Write files using fwrite for speed. 
# fwrite by default writes column names and does not output rownames.
fwrite(ge %>% rownames_to_column(var = "ID"), out_ge_file, sep = "\t", quote = FALSE)
fwrite(snp %>% rownames_to_column(var = "ID"), out_snp_file, sep = "\t", quote = FALSE)
fwrite(covar %>% rownames_to_column(var = "ID"), out_covar_file, sep = "\t", quote = FALSE)
fwrite(geneloc, out_geneloc_file, sep = "\t", quote = FALSE)
fwrite(snploc, out_snploc_file, sep = "\t", quote = FALSE)
```

## Run MatrixEQTL at Visit 1

Run cis-eQTL mapping with a linear model. The cis distance is 1 Mb, all cis
p-values are kept (`pvOutputThreshold_cis <- 1`), and no trans tests are run.
The five order-checked input files (SNP, snploc, expression, geneloc,
covariates) are loaded as SlicedData objects and passed to `Matrix_eQTL_main`;
the resulting cis-eQTL table is written to the results directory.

```{r matrixeqtl-run}
# 05.05.2025 
# R Version 4.4.3

# Define the output directory for Matrix eQTL results
matrixeqtl_results_dir <- file.path(panEQTL_dir, "panEQTL_results")
if (!dir.exists(matrixeqtl_results_dir)) {
  dir.create(matrixeqtl_results_dir, recursive = TRUE)
}

# Settings for Matrix eQTL
useModel <- modelLINEAR  # using linear model
pvOutputThreshold_cis <- 1
cisDist <- 1e6
errorCovariance <- numeric()

# Construct file paths for the five necessary files.
SNP_file_name           <- file.path(final_dir, paste0(prefix, "_SNP.txt"))
snps_location_file_name <- file.path(final_dir, paste0(prefix, "_snploc.txt"))
expression_file_name    <- file.path(final_dir, paste0(prefix, "_GE.txt"))
gene_location_file_name <- file.path(final_dir, paste0(prefix, "_geneloc.txt"))
covariates_file_name    <- file.path(final_dir, paste0(prefix, "_Covariates.txt"))

# Create SlicedData objects for SNP, gene expression, and covariates.
snps <- SlicedData$new()
snps$fileDelimiter <- "\t"
snps$fileOmitCharacters <- "NA"
snps$fileSkipRows <- 1
snps$fileSkipColumns <- 1
snps$fileSliceSize <- 2000
snps$LoadFile(SNP_file_name)

gene <- SlicedData$new()
gene$fileDelimiter <- "\t"
gene$fileOmitCharacters <- "NA"
gene$fileSkipRows <- 1
gene$fileSkipColumns <- 1
gene$fileSliceSize <- 2000
gene$LoadFile(expression_file_name)

cvrt <- SlicedData$new()
cvrt$fileDelimiter <- "\t"
cvrt$fileOmitCharacters <- "NA"
cvrt$fileSkipRows <- 1
cvrt$fileSkipColumns <- 1
cvrt$fileSliceSize <- 2000
if (file.exists(covariates_file_name)) {
    cvrt$LoadFile(covariates_file_name)
}

snpspos <- read.table(snps_location_file_name, header = TRUE, stringsAsFactors = FALSE)
genepos <- read.table(gene_location_file_name, header = TRUE, stringsAsFactors = FALSE)
    
# Create temporary file names for cis and trans outputs.
output_file_name_cis <- tempfile()
output_file_name_tra <- tempfile()

# Run Matrix_eQTL_main.
me <- Matrix_eQTL_main(
    snps = snps,
    gene = gene,
    cvrt = cvrt,
    output_file_name = output_file_name_tra,
    pvOutputThreshold = 0,
    useModel = useModel,
    errorCovariance = errorCovariance,
    verbose = TRUE,
    output_file_name.cis = output_file_name_cis,
    pvOutputThreshold.cis = pvOutputThreshold_cis,
    snpspos = snpspos,
    genepos = genepos,
    cisDist = cisDist,
    pvalue.hist = "qqplot",
    min.pv.by.genesnp = FALSE,
    noFDRsaveMemory = FALSE
)

# Remove the trans output file.
unlink(output_file_name_tra)

# Directly extract the cis-eQTL results from the me object.
cis_eqtls <- me$cis$eqtls

write.table(cis_eqtls, file = file.path(matrixeqtl_results_dir, "panEQTL_results.txt"), sep = "\t", quote = FALSE,
            row.names = FALSE, col.names = TRUE)

# Clean up temporary cis output file.
unlink(output_file_name_cis)

rm(snps, gene, cvrt, snpspos, genepos, me, cis_eqtls)
gc()
```
