---
title: "Hospital-admission cis-eQTLs: eigenMT correction and significance"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(data.table)
library(dplyr)
library(tidyverse)
library(tibble)
library(readxl)
library(qs2)
```

## Paths and parameters

All hardcoded absolute paths from the original notebook are collected here.
Edit these values to match your environment; everything downstream is built
from them with `file.path()`.

```{r paths-and-parameters}
# ---- EDIT THESE FOR YOUR ENVIRONMENT ----
# Root of the raw IMPACC genotype/expression data env.
base_dir <- "/path/to/impacc_data/"

# Serialized IMPACC data environment (clinical + omics).
data_env_path <- file.path(base_dir, "impacc_data_env.RDS")

# PLINK genotype prefix (its .bim provides grch37 -> grch38 mapping).
plink_geno_prefix <- file.path(base_dir, "plink", "impacc_geno")

# Published critical COVID-19 GWAS lead-SNP table.
published_snps_path <- "/path/to/All_Severe_COVID_GWAS_hits.xlsx"

# Analysis output / results root (this is the original panEQTL_dir).
results_dir <- "/path/to/hospital_admission_results"
panEQTL_dir <- results_dir

# Subdirectories built from the results root.
final_dir          <- file.path(panEQTL_dir, "final_files")
matrixeqtl_results <- file.path(panEQTL_dir, "panEQTL_results")  # holds the cis-eQTL file
formatted_dir      <- file.path(panEQTL_dir, "eigenMT_formatted_files")
prefix <- "impacc_panEQTL_visit1"
# -----------------------------------------
```

## Format the MatrixEQTL output for eigenMT

eigenMT is a computationally efficient proxy for permutation-based FDR control.
This chunk rewrites the MatrixEQTL inputs and outputs into the layout eigenMT
expects: chromosome labels lose the "chr" prefix, beta is moved before the test
statistic, and the cis-eQTL columns are renamed (SNP, t-stat, p-value).

```{r format-for-eigenmt}
# 1 10 11 12 13 16 17 19  2 20 21  3  4  5  6  7  8  9  X (19 unique chromosomes)

## Process snploc file
snploc_in <- file.path(final_dir, paste0(prefix, "_snploc.txt"))
snploc_out <- file.path(formatted_dir, paste0(prefix, "_snploc_eigenMT.txt"))

dt_snploc <- fread(snploc_in) %>%
    # Remove the "chr" prefix from the chr column
    mutate(chr = gsub("chr", "", chr))

fwrite(dt_snploc, snploc_out, sep = "\t", quote = FALSE)

## Process SNP file
snp_in <- file.path(final_dir, paste0(prefix, "_SNP.txt"))
snp_out <- file.path(formatted_dir, paste0(prefix, "_SNP_eigenMT.txt"))

dt_snp <- fread(snp_in)

fwrite(dt_snp, snp_out, sep = "\t", quote = FALSE)

## Process cis_eQTL file (from the 'results' directory)
qtl_in <- file.path(matrixeqtl_results, "panEQTL_results.txt")
qtl_out <- file.path(formatted_dir, paste0(prefix, "_cis_eQTL_eigenMT.txt"))

dt_qtl <- fread(qtl_in) %>%
    # Move the beta column to come before statistic
    relocate(beta, .before = statistic) %>%
    # Rename columns as needed for eigenMT
    rename(SNP = snps,
            `t-stat` = statistic,
            `p-value` = pvalue)

fwrite(dt_qtl, qtl_out, sep = "\t", quote = FALSE)

## Process geneloc file
geneloc_in <- file.path(final_dir, paste0(prefix, "_geneloc.txt"))
geneloc_out <- file.path(formatted_dir, paste0(prefix, "_geneloc_eigenMT.txt"))

dt_geneloc <- fread(geneloc_in) %>%
    # Remove the "chr" prefix from the chr column
    mutate(chr = gsub("chr", "", chr))

fwrite(dt_geneloc, geneloc_out, sep = "\t", quote = FALSE)
```

The per-chromosome eigenMT run that consumes these formatted files lives in
`04-run_eigenmt.sh`. Run that script before continuing, then combine its
per-chromosome outputs below.

## Combine the per-chromosome eigenMT results

Read every per-chromosome eigenMT output, tag each with its chromosome, and
bind them into one table.

```{r combine-eigenmt}
eigenMT_results_dir <- file.path(panEQTL_dir, "eigenMT_results/visit1")

# Find all chromosome-specific eigenMT files
eigenMT_files <- list.files(path = eigenMT_results_dir, 
                           pattern = "chr.*_eigenMT\\.txt$", 
                           full.names = TRUE)

# Check if files were found
if (length(eigenMT_files) == 0) {
  stop("No eigenMT files found in ", eigenMT_results_dir)
}

message("Found ", length(eigenMT_files), " eigenMT files to combine")

# Read and combine all files
all_data <- lapply(eigenMT_files, function(file) {
  message("Reading file: ", basename(file))
  dt <- fread(file)
  
  # Extract chromosome number from filename
  chr <- gsub(".*chr([0-9XY]+)_eigenMT.txt", "\\1", basename(file))
  
  # Add chromosome column if needed (if not already in the data)
  if (!"chr" %in% names(dt)) {
    dt[, chr := chr]
  }
  
  return(dt)
}) %>% rbindlist(fill = TRUE)  # fill=TRUE allows for different column structures if needed

# Save the combined data
output_file <- file.path(panEQTL_dir, "panEQTL_results", "panEQTL_eigenMT.txt")
fwrite(all_data, file = output_file, sep = "\t", quote = FALSE)
```

## Annotate the cis-eQTL results

Add gene names and the grch38 coordinate to the MatrixEQTL output, using the
PBMC row features and the PLINK `.bim` (which maps grch37 IDs to grch38
coordinates and records the minor/major alleles). The annotated table is saved
for downstream colocalization, and the published GWAS SNPs are joined in for
inspection.

```{r annotate-eqtls}
data_env <- readRDS(data_env_path)

rowfeature <- data_env$pbmc_transcriptomics_rowfeature %>%
        rownames_to_column(var = "gene") %>%
        select(gene, gene_name)

grch38_to_37 <- fread(paste0(plink_geno_prefix, ".bim")) %>%
    select(-"V3") %>%
    rename(grch37 = "V2", 
           minor = "V5",
           major = "V6") %>%
           unite(`grch38`, c("V1","V4"), sep = ":")

combined_eqtls_final <- fread(file.path(matrixeqtl_results, "panEQTL_results.txt")) %>%
  rename(grch37 = snps) %>%
  inner_join(rowfeature, by = "gene") %>%
  relocate(gene_name, .before = "statistic") %>%
  inner_join(grch38_to_37, by = "grch37") %>%
  relocate(grch38, .before = "gene")

# Save the combined cis-eQTL results as an RDS file and as a tab-delimited text file.
qs_save(combined_eqtls_final, file.path(matrixeqtl_results, "panEQTL_final.qs2"))

published_snps <- read_excel(published_snps_path) %>%
  pivot_longer(cols = everything(), 
              values_to = "grch38") %>%
  select(-"name") %>%
  filter(!is.na(`grch38`)) %>%
  distinct(grch38)  

eQTL_snps <- published_snps %>%
    inner_join(combined_eqtls_final, by = "grch38")
```

## eigenMT correction and significant eQTL counts

Join the per-gene eigenMT correction factor (`TESTS`) onto the cis-eQTL table
and compute the eigenMT-corrected p-value as `pvalue * TESTS`. Restrict to the
published SNPs (plus the chr12 proxy variant 12:112945172), save the
significant set, and report the eQTL, unique-SNP, and unique-gene counts at
eigenMT < 0.01.

```{r eigenmt-significance}
combined_eqtls_final <- qs_read(file.path(matrixeqtl_results, "panEQTL_final.qs2"))

# Load published SNP data and reformat
published_snps <- read_excel(published_snps_path) %>%
  pivot_longer(cols = everything(), values_to = "grch38") %>%
  filter(!is.na(grch38)) %>%
  group_by(grch38) %>%
  mutate(source = if(n() > 1) "both" else as.character(name)) %>%
  ungroup() %>%
  select(-name) %>%
  distinct(grch38, source)

# Create list of SNPs of interest (70 SNPs) and add extra proxy variant for chr12
published_snps_plus_chr12 <- published_snps %>%
  distinct(grch38) %>%
  bind_rows(tibble(grch38 = "12:112945172"))

eigenMT_final <- fread(file.path(matrixeqtl_results, "panEQTL_eigenMT.txt")) %>%
  select(gene, TESTS)

# Convert to data.table
combined_eqtls_dt <- as.data.table(combined_eqtls_final)
eigenMT_dt <- as.data.table(eigenMT_final)
published_snps_dt <- as.data.table(published_snps_plus_chr12)

# Set keys for faster joins
setkey(combined_eqtls_dt, gene)
setkey(eigenMT_dt, gene)
setkey(published_snps_dt, grch38)

# Join eQTLs with eigenMT correction factors and calculate corrected p-values
eqtls_with_correction <- combined_eqtls_dt[eigenMT_dt, nomatch=0
  ][, eigenMT := pvalue*TESTS]

# Filter for SNPs of interest only
setkey(eqtls_with_correction, grch38)
significant_eqtls <- eqtls_with_correction[published_snps_dt, nomatch=0]

# Save results
qs_save(significant_eqtls, file.path(matrixeqtl_results, "panEQTL_significant_published_eqtls.qs2"))

significant_eqtls_0.01 <- significant_eqtls %>% filter(eigenMT < 0.01)

# Print summary
cat(sprintf("Found %d eQTLs for published SNPs\n", nrow(`significant_eqtls_0.01`)))
cat(sprintf("This includes %d unique SNPs and %d unique genes\n", 
            length(unique(`significant_eqtls_0.01`$grch38)), 
            length(unique(`significant_eqtls_0.01`$gene))))

# Found 61 eQTLs (snp-gene pairs) for published SNPs
# This includes 32 unique SNPs and 52 unique genes
```
