---
title: "Hospital-admission cis-eQTLs: colocalization with GenOMICC"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(coloc)
library(data.table)
library(dplyr)
library(tidyr)
library(stringr)
library(qs2)
library(tibble)
library(ggplot2)
```

## Paths and parameters

All hardcoded absolute paths from the original notebook are collected here.
Edit these values to match your environment; everything downstream is built
from them with `file.path()`. The GenOMICC critical-illness summary statistics
can be downloaded from the URL noted in the harmonization chunk below.

```{r paths-and-parameters}
# ---- EDIT THESE FOR YOUR ENVIRONMENT ----
# Root of the raw IMPACC genotype/expression data env.
base_dir <- "/path/to/impacc_data/"

# Serialized IMPACC data environment (clinical + omics).
data_env_path <- file.path(base_dir, "impacc_data_env.RDS")

# GenOMICC international critical-illness GWAS summary statistics (TSV).
gwas_sumstats_path <- file.path(base_dir, "GenOMICC_international_meta_release3_critical.tsv")

# Analysis output / results root (this is the original panEQTL_dir).
results_dir <- "/path/to/hospital_admission_results"
panEQTL_dir <- results_dir

# Subdirectories built from the results root.
matrixeqtl_results <- file.path(panEQTL_dir, "panEQTL_results")
final_dir          <- file.path(panEQTL_dir, "final_files")
output_dir         <- file.path(panEQTL_dir, "residuals_analysis")
prefix <- "impacc_panEQTL_visit1"
# -----------------------------------------
```

## Compute genotype-conditioned expression residuals

For each significant SNP-gene pair (eigenMT < 0.01), regress the gene's
inverse-normal expression on all covariates (excluding genotype) and keep the
residuals alongside the genotype, for downstream plotting. 61 SNP-gene pairs
(32 unique SNPs, 52 unique genes) are processed.

```{r compute-residuals}
# Load the significant eQTLs we filtered in previous step
significant_eqtls <- qs_read(file.path(matrixeqtl_results, "panEQTL_significant_published_eqtls.qs2"))

# Further filter to eigenMT < 0.01 (significant after multiple testing correction)
significant_eqtls_filtered <- significant_eqtls[eigenMT < 0.01]

# Extract the target SNP-gene pairs for residual calculation
target_pairs <- significant_eqtls_filtered %>%
  distinct(grch37, gene)

# Print summary of what we're computing
cat(sprintf("Computing residuals for %d SNP-gene pairs (eigenMT < 0.01)\n", nrow(target_pairs)))
# Computing residuals for 61 SNP-gene pairs (eigenMT < 0.01)

# Define file paths for residual calculations (preserve naming convention).
ge_file      <- file.path(final_dir, paste0(prefix, "_GE.txt"))
snp_file     <- file.path(final_dir, paste0(prefix, "_SNP.txt"))
covar_file   <- file.path(final_dir, paste0(prefix, "_Covariates.txt"))

# Check that all required files exist
if(!all(file.exists(ge_file, covar_file, snp_file))) {
  stop("One or more required files missing. Check paths.")
}

# Read the data files
ge <- fread(ge_file) %>% column_to_rownames(var = "ID")
covar <- fread(covar_file) %>% column_to_rownames(var = "ID")
snp <- fread(snp_file) %>% column_to_rownames(var = "ID")

# Initialize a list to store residual results for each SNP-gene pair
residuals_list <- list()

# Loop over each SNP-gene pair
for (i in seq_len(nrow(target_pairs))) {
  target_snp <- target_pairs$grch37[i]    # SNP identifier
  target_gene <- target_pairs$gene[i]     # Gene name
  pair_id <- paste0(target_snp, "_", target_gene)
  
  # Extract expression for target gene
  if(!(target_gene %in% rownames(ge))) {
    message("Target gene ", target_gene, " not found in GE file. Skipping.")
    next
  }
  
  expr_vals <- t(ge[target_gene,]) %>% 
              as.data.frame() %>%
              setNames("expression") %>%
              rownames_to_column(var = "sample_id")
      
  # Extract genotype for target SNP
  if(!(target_snp %in% rownames(snp))) {
    message("Target SNP ", target_snp, " not found in SNP file. Skipping.")
    next
  }
  
  geno_vals <- t(snp[target_snp,]) %>% 
              as.data.frame() %>%
              setNames("genotype") %>%
              rownames_to_column(var = "sample_id")
      
  covar_final <- t(covar) %>% 
              as.data.frame() %>%
              rownames_to_column(var = "sample_id")
      
  # Combine into a data frame for regression
  dat <- expr_vals %>%
        inner_join(geno_vals, by = "sample_id") %>%
        inner_join(covar_final, by = "sample_id") %>%
        column_to_rownames(var = "sample_id")

  # Build the linear model formula: expression ~ covariates (excluding genotype)
  covar_names <- colnames(covar_final)[-1]
  formula_str <- paste("expression ~", paste(covar_names, collapse = " + "))

  # Fit the linear model
  model <- lm(as.formula(formula_str), data = dat)
  
  # Extract residuals
  resids <- residuals(model) %>%
        as.data.frame() %>%
        setNames("residuals") %>%
        rownames_to_column(var = "sample_id")

  # Add genotype information for plotting
  geno_for_plotting <- dat %>%
                       select(genotype) %>%
                       rownames_to_column(var = "sample_id")

  # Create final residuals data frame for this SNP-gene pair
  res_df <- resids %>%
        inner_join(geno_for_plotting, by = "sample_id") %>%
        mutate(snp = target_snp,
               gene = target_gene)
  
  # Store in list
  residuals_list[[pair_id]] <- res_df
  
  # Print progress
  if (i %% 10 == 0 || i == nrow(target_pairs)) {
    cat(sprintf("Processed %d/%d SNP-gene pairs\n", i, nrow(target_pairs)))
  }
}

# Combine all residuals into one data frame
combined_residuals <- bind_rows(residuals_list, .id = "pair_id") %>%
                      separate(pair_id, into = c("grch37", "gene_id"), sep = "_", remove = TRUE) %>%
                      select(-snp, -gene_id)

# Print summary
cat("Combined residuals dataset dimensions:", dim(combined_residuals), "\n") # 55998 7 (918 people X 61 unique eQTL PAIRS)
cat("Number of unique SNPs:", length(unique(combined_residuals$grch37)), "\n") # 32
cat("Number of unique genes:", length(unique(combined_residuals$gene_id)), "\n") # 52

# Save results
qs_save(combined_residuals, file.path(output_dir, "panEQTL_residuals.qs2"))
```

## Define colocalization ranges and compute allele frequencies

Define a +/- 500 kb window around each of the 32 unique significant SNPs and
write the range list for PLINK. PLINK then computes allele frequencies over
those windows (107,803 SNPs fall within +/- 500 kb of the 32 lead SNPs).

```{r coloc-ranges}
data_env <- readRDS(data_env_path)

sig_eQTLs_no_residuals <- qs_read(file = file.path(matrixeqtl_results, "panEQTL_significant_published_eqtls.qs2"))

all_sig_eqtl_residuals_full <- qs_read(file = file.path(matrixeqtl_results, "panEQTL_residuals.qs2")) %>%
    select(-gene_id) %>%
    inner_join(sig_eQTLs_no_residuals, by = c("grch37", "gene"))

ranges_for_coloc <- all_sig_eqtl_residuals_full %>%
  select(grch38) %>%
  distinct() %>%
  separate(grch38, into = c("chr", "pos"), sep = ":", remove = TRUE) %>%
  mutate(pos = as.numeric(pos),
  start = pos - 500000,
  end = pos + 500000) %>%
  mutate(ranges = paste0("R", 1:n())) %>%
  select(chr, start, end, ranges)

write.table(ranges_for_coloc, 
            file = file.path(panEQTL_dir, "ranges_for_coloc.txt"),
            quote = FALSE, 
            row.names = FALSE, 
            col.names = FALSE, 
            sep = "\t")        
```

The allele-frequency computation was inline PLINK in the original notebook.
Edit the leading variables and run it before the harmonization chunk.

```{bash coloc-allele-freq, eval=FALSE}
# Common paths (edit for your environment)
base_dir=/path/to/hospital_admission_results
keep="$base_dir/visit1_panEQTL_coreIDs.txt"
ranges="$base_dir/ranges_for_coloc.txt"
bfile=/path/to/impacc_data/plink/impacc_geno

# PLINK run: Calculate allele frequencies (--freq)
plink   --bfile "$bfile" \
        --keep "$keep" \
        --extract range "$ranges" \
        --freq \
        --make-bed \
        --out "$base_dir/visit1_panEQTL_coloc"
```

## Harmonize eQTL and GenOMICC data

Merge PLINK allele frequencies with the variant table, attach the cis-eQTL
results and eigenMT correction, and compute the eQTL standard error and varbeta.
The GenOMICC summary statistics are then read, palindromic/ambiguous SNPs are
removed (ambiguous defined as palindromic with GWAS MAF between 0.40 and 0.60),
the GWAS is harmonized, and eQTL effect alleles are flipped to match the GWAS
allele orientation.

```{r harmonize-coloc}
# Read .afreq file
frq_df <- fread(file.path(panEQTL_dir, "visit1_panEQTL_coloc.afreq")) %>%
            mutate(N = OBS_CT / 2) %>%
            rename(grch37 = ID, 
                   CHR = "#CHROM", 
                   MAF = ALT_FREQS) %>%
            select(-OBS_CT, -"PROVISIONAL_REF?")

bim_df <- fread(file.path(panEQTL_dir, "visit1_panEQTL_coloc.bim")) %>%
            relocate(V4, .after = V1) %>%
            unite("grch38", V1:V4, sep = ":", remove = TRUE) %>%
            select(-V3) %>%
            rename(grch37 = V2,
                    ALT = V5,
                    REF = V6) 

# Merge grch37 and grch38 information
impacc_geno_dat <- inner_join(frq_df, bim_df,
                              by = c("grch37", "REF", "ALT")) %>%
                              select(-"CHR", -"ALT", -"REF") %>%
                              relocate(grch38, .after=grch37)

# 107,803 SNPs fall within the range of +/-500kb of the 32 unique SNPs.

# Now, we need to get the relevant cis-eQTL results data from all of these snps

# Load with qs_read -- much faster
combined_eqtls_final <- qs_read(file.path(matrixeqtl_results, "panEQTL_final.qs2"))

eigenMT_final <- fread(file.path(matrixeqtl_results, "panEQTL_eigenMT.txt")) %>%
  select(gene, TESTS)

# Final preparation for coloc analysis
coloc_snps <- impacc_geno_dat %>%
              inner_join(combined_eqtls_final, by = c("grch37", "grch38")) %>%
              inner_join(eigenMT_final, by = c("gene")) %>%
              mutate(eigenMT = pvalue*TESTS) %>%
              relocate(eigenMT, .before="FDR") %>%
              rename(eigenMT_tests = "TESTS") %>%
              mutate(se = abs(beta / statistic),
                     varbeta = (beta / statistic)^2)

# these are all of the variants (with all corresponding data columns) flanking (+/- 500kb) the 32 unique SNPs with eQTLs
qs_save(coloc_snps, file.path(panEQTL_dir, "panEQTL_coloc_snps.qs2"))

unique_snps <- (unique(coloc_snps %>% select("grch38"))) %>%
                separate(grch38, into = c("CHR", "POSb38"), sep = ":", remove = FALSE) %>%
                mutate(CHR = as.integer(CHR),
                POSb38 = as.integer(POSb38))
# 107,803 unique SNPs flanking +/- 500kb from the 32 SNPs. 

# Download the GenOMICC summary statistics once, e.g.:
# wget https://datashare.ed.ac.uk/bitstream/handle/10283/8536/GenOMICC_international_meta_release3_critical.tsv.gz
# curl -O https://datashare.ed.ac.uk/bitstream/handle/10283/8536/GenOMICC_international_meta_release3_critical.tsv.gz

genomicc_sumstats <- fread(gwas_sumstats_path) %>%
                     select(-ID) %>%
                     inner_join(unique_snps, by = c("CHR", "POSb38")) # 72,157 overlapping SNPs between eQTL & GenoMICC. 

# remove all palindromic SNPs explicitly
setDT(genomicc_sumstats)

# Step 1: Identify palindromic SNPs explicitly
genomicc_sumstats[, palindromic := ((Allele1 == "A" & Allele2 == "T") |
                                    (Allele1 == "T" & Allele2 == "A") |
                                    (Allele1 == "C" & Allele2 == "G") |
                                    (Allele1 == "G" & Allele2 == "C"))]

# Step 2: Mark ambiguous SNPs using GWAS MAF threshold (defensible: >0.40)
ambig_thresh <- 0.40  # Scientifically defensible threshold
genomicc_sumstats[, ambiguous := palindromic & (Freq1 > ambig_thresh & Freq1 < (1 - ambig_thresh))]

cat("Ambiguous SNPs identified for removal:", sum(genomicc_sumstats$ambiguous), "\n")

# Ambiguous SNPs identified for removal: 1204 


###################################
## (A) Harmonize the GenOMICC data ##
###################################
harmonized_genomicc_sumstats <- genomicc_sumstats %>%
  filter(ambiguous == FALSE) %>%
  # rename columns to keep consistent naming
  rename(
    allele1_gwas   = Allele1,
    allele2_gwas   = Allele2,
    beta_gwas      = Effect,
    se_gwas        = StdErr,
    pval_gwas      = P.value,
    freq1_gwas     = Freq1,
    n_gwas         = TotalSampleSize
  ) %>%
  mutate(
    varbeta_gwas = se_gwas^2
  )
#70,953 SNPs left in the genomicc_sumstats after removing potentially Ambiguous SNPs. 

########################################
## (B) Join to your impacc_eQTL data  ##
########################################
# eQTL data is: impacc_eqtl
impacc_eqtl <- qs_read(file.path(panEQTL_dir, "panEQTL_coloc_snps.qs2"))

joined_df <- impacc_eqtl %>%
  inner_join(harmonized_genomicc_sumstats, by = c("grch38"))
# 2836614 rows by 29 columns. 

#######################################
## (C) Harmonize eQTL to GWAS
#######################################

setDT(joined_df)

# Step 1: Compute an initial flip flag based on allele identity.
joined_df[, flip_initial := fifelse(minor == allele1_gwas, FALSE,
                             fifelse(minor == allele2_gwas, TRUE, NA))]

# Remove any rows where we still have NA (mismatches)
harmonized_df <- joined_df[!is.na(flip_initial)]
# 2833834 rows      30 columns

#######################################
## (D) Flip eQTL statistics to match 
##           GWAS allele orientation
#######################################

# Assign beta_eqtl and freq_eqtl for all rows:
# Assign harmonized eQTL columns consistently for all rows.
# If flip_initial is TRUE, flip the beta and MAF; if FALSE, keep as is.

harmonized_df[, `:=`(
  beta_eqtl = ifelse(flip_initial, -beta, beta),
  freq_eqtl = ifelse(flip_initial, 1 - MAF, MAF)
)]

# Also, assign the allele columns so that they reflect the GWAS orientation:
harmonized_df[, `:=`(
  effect_allele = ifelse(flip_initial, allele1_gwas, minor),
  other_allele  = ifelse(flip_initial, allele2_gwas, major)
)]

# Remove the original beta, MAF, minor, and major columns if they're no longer needed.
harmonized_df <- harmonized_df[, !c("MAF", "beta", "minor", "major"), with = FALSE]

# Report flipping count
num_flips <- harmonized_df %>% select(grch38, flip_initial) %>% distinct()
# 30,790 FLIPS
# 40,086 NO FLIPS

qs_save(harmonized_df, file.path(panEQTL_dir, "harmonized_panEQTL_gwas_final_tg.qs2"))
```

## Run coloc.abf over each significant SNP-gene pair

For each of the 61 significant SNP-gene pairs (eigenMT < 0.01), extract all
harmonized SNPs in a +/- 250 kb window, format the eQTL (quantitative) and GWAS
(case-control, with case fraction s = 23707 / 1075129) datasets, and run
`coloc.abf` with prior p12 = 5.5e-6. Posterior probabilities PP.H1-PP.H4 are
recorded per pair. With PP.H4 > 0.5 the source reports 11 pairings, 9 unique
SNPs, and 10 unique genes.

```{r coloc-abf}
## LOAD DATA
harmonized_df <- qs_read(file.path(panEQTL_dir, "harmonized_panEQTL_gwas_final_tg.qs2"))

# Convert to data.table and set keys for efficient subsetting.
harmonized_dt <- as.data.table(harmonized_df)
setkey(harmonized_dt, gene, CHR, POSb38)
rm(harmonized_df)
# ---------------------------
# STEP 0: Prepare the SNP-Gene pairs dataframe
# ---------------------------
snp_gene_pairs_to_coloc <- qs_read(file = file.path(matrixeqtl_results, "panEQTL_significant_published_eqtls.qs2")) %>%
          filter(eigenMT < 0.01) %>%
          select(grch38, gene, gene_name) %>%
          distinct() %>% # 61 distinct SNP-gene pairs to test for coloc. 
          separate(grch38, into = c("chr", "pos"), sep = ":", convert = TRUE)

# -----------------------------------------
# Step 1: Initialize a dataframe to store coloc results.
# We want to keep all columns from harmonized_df plus PP.H3 and PP.H4.
# Harmonized_df has the full set of columns as listed.
coloc_results_dat <- harmonized_dt[0, ] %>% 
  mutate(PP.H1 = numeric(), PP.H2 = numeric(), PP.H3 = numeric(), PP.H4 = numeric())

# -----------------------------------------
# Step 2: Loop over each SNP-gene pair in snp_gene_pairs_to_coloc
# -----------------------------------------
for(i in 1:nrow(snp_gene_pairs_to_coloc)){
  
  # Extract the lead SNP and gene information from the current row.
  lead_chr    <- as.character(snp_gene_pairs_to_coloc$chr[i])
  lead_pos    <- snp_gene_pairs_to_coloc$pos[i]
  lead_grch38 <- paste0(lead_chr, ":", lead_pos)
  gene_id     <- snp_gene_pairs_to_coloc$gene[i]
  gene_name   <- snp_gene_pairs_to_coloc$gene_name[i]
  
  # Define cis window boundaries: +/- 250 kb around the lead SNP position.
  win_lower <- lead_pos - 250000
  win_upper <- lead_pos + 250000
  
  # -----------------------------------------
  # Step 3: Identify all candidate conditions (visit and group) for this SNP-gene pair.
  candidate_conditions <- harmonized_dt %>%
    filter(gene == gene_id,
           grch38 == lead_grch38,
           eigenMT < 0.01)
  
  # If no condition meets the significance threshold, skip this SNP-gene pair.
  if(nrow(candidate_conditions) == 0){
    message("No significant eQTL (eigenMT < 0.01) found for ", lead_grch38, " - ", gene_name, "; skipping.")
    next
  }
  
  # -----------------------------------------
  # Step 4: Loop over each candidate condition
  for(j in 1:nrow(candidate_conditions)){
    condition <- candidate_conditions %>% slice(j)
    
    # -----------------------------------------
    # Step 5: Extract all SNP data within the +/- 250 kb window for this condition and gene.

    dat_subset <- harmonized_dt[
      gene == gene_id & 
      CHR == lead_chr & 
      POSb38 >= win_lower & 
      POSb38 <= win_upper
    ]
        
    # Remove rows with missing varbeta values to avoid check_dataset error.
    dat_subset <- dat_subset %>% filter(!is.na(varbeta))

    # If no rows remain after filtering, skip this condition.
    if(nrow(dat_subset) == 0){
      message("No eQTL SNPs in window after filtering out missing varbeta for ", lead_grch38, " - ", gene_name, 
              " in ", cond_visit, " ", cond_group, "; skipping condition.")
      next
    }
    
    # -----------------------------------------
    # Step 6: Format the eQTL and GWAS datasets for coloc.
    # Format the eQTL dataset (quantitative).
    eqtl_dataset <- list(
      snp = dat_subset$grch38,
      beta = dat_subset$beta_eqtl,        # use the harmonized beta
      varbeta = dat_subset$varbeta,
      pvalues = dat_subset$pvalue,
      MAF = dat_subset$freq_eqtl,         # use the harmonized MAF
      N = dat_subset$N,
      position = dat_subset$POSb38,
      type = "quant"
    )
    
    # Format the GWAS dataset (case-control).
    gwas_dataset <- list(
      snp = dat_subset$grch38,
      beta = dat_subset$beta_gwas,
      varbeta = dat_subset$varbeta_gwas,
      pvalues = dat_subset$pval_gwas,
      MAF = dat_subset$freq1_gwas,
      N = dat_subset$n_gwas,
      type = "cc",
      s = (23707 / 1075129),  # defined by the case-control numbers from GenoMICC. 
      position = dat_subset$POSb38
    )
    
    # Run diagnostic checks (optional)
    check_dataset(eqtl_dataset)
    check_dataset(gwas_dataset)
    
    # -----------------------------------------
    # Step 7: Run coloc.abf using the formatted datasets.
    coloc_res <- coloc.abf(
      dataset1 = eqtl_dataset,
      dataset2 = gwas_dataset,
      p12=5.5e-6) # halfway in between 1e-6 and 1e-5
    
    # Extract posterior probabilities.
    pp_h1 <- coloc_res$summary["PP.H1.abf"]
    pp_h2 <- coloc_res$summary["PP.H2.abf"]
    pp_h3 <- coloc_res$summary["PP.H3.abf"]
    pp_h4 <- coloc_res$summary["PP.H4.abf"]
    
    # -----------------------------------------
    # Step 8: Record the results.
    # Add the coloc results (PP.H1, PP.H2, PP.H3, PP.H4) to the candidate condition row.
    condition_out <- condition %>% 
      mutate(PP.H1 = as.numeric(pp_h1),
             PP.H2 = as.numeric(pp_h2),
             PP.H3 = as.numeric(pp_h3),
             PP.H4 = as.numeric(pp_h4))
    
    # Append the row to the final results dataframe.
    coloc_results_dat <- bind_rows(coloc_results_dat, condition_out)
    
  } # end inner loop over candidate conditions
} # end outer loop over SNP-gene pairs

# -----------------------------------------
# Final Output:
# coloc_results_dat now contains one row for each candidate condition (visit-group) 
# where eigenMT < 0.01 for a SNP-gene pair, along with coloc PP.H1, PP.H2, PP.H3, PP.H4.
# ----------------------------------------- 

qs_save(coloc_results_dat, file.path(matrixeqtl_results, "panEQTL_coloc_results.qs2"))

# with PPH4 >0.5
# 11 pairings 
# 9 unique SNPs
# 10 unique genes
```
