---
title: "Hospital-admission cis-eQTLs: Figure 2"
author: "Ernie Chen (ernie.chen@yale.edu)"
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(dplyr)
library(tidyr)
library(tibble)
library(data.table)
library(ComplexHeatmap)   # if not installed: BiocManager::install("ComplexHeatmap")
library(circlize)         # for colour ramps
library(ggplot2)
library(scales)
library(ggbreak)
library(qs2)
```

## Paths and parameters

All hardcoded absolute paths from the original notebook are collected here.
Edit these values to match your environment; everything downstream is built
from them with `file.path()`.

```{r paths-and-parameters}
# ---- EDIT THESE FOR YOUR ENVIRONMENT ----
# Analysis output / results root (this is the original panEQTL_dir).
results_root <- "/path/to/hospital_admission_results"
panEQTL_dir <- results_root

# Directory holding the eigenMT-significant eQTL table and coloc results.
results_dir <- file.path(panEQTL_dir, "panEQTL_results")
# -----------------------------------------

cis_eqtls_of_published_snps <- qs_read(file.path(results_dir, "panEQTL_significant_published_eqtls.qs2")) 

coloc_dat <- qs_read(file.path(results_dir, "panEQTL_coloc_results.qs2")) 
```

## Figure 2B: Visit-1 SNP x gene effect-size heatmap

Merge the colocalization posterior probabilities onto the significant cis-eQTL
table, keep only SNP-gene pairs with eigenMT < 0.01 (61 pairs, 52 genes,
32 SNPs), and build the SNP-by-gene effect-size matrix. SNPs are ordered by
genomic position and genes by hierarchical clustering of beta. The heatmap fill
is the signed effect size; each significant cell carries a diagonal mark, and
colocalized pairs (PP.H4 > 0.5) get a full X plus a border box, with their SNP
and gene labels highlighted.

```{r figure2b-heatmap}
############################################################
## Figure 2: Visit-1 cis-eQTL landscape
##   • Panel A: dot-heat-map (58 SNPs × 52 genes)
##   • Panel B: discovery/colocalisation bar
############################################################
## ──────────────────────────────────────────────
## 1.  Merge coloc results back onto cis-eQTL table
##     (coloc columns are NA for the >800 non-significant pairs)
## ──────────────────────────────────────────────
eqtl_v1 <- cis_eqtls_of_published_snps %>% 
  left_join(    # adds PP.H1-H4 where available
    coloc_dat   %>%   
      select(grch38, gene, starts_with("PP.H")), 
    by = c("grch38", "gene")
  )

## ──────────────────────────────────────────────
## 2.  Keep only SNP-gene pairs with eigenMT < 0.01
##     → 61 pairs, 52 unique genes, 32 unique SNPs
## ──────────────────────────────────────────────
eqtl_sig <- eqtl_v1 %>% 
  filter(eigenMT < 0.01)

## ──────────────────────────────────────────────
## 3.  Derive plotting helpers
## ──────────────────────────────────────────────
eqtl_sig <- eqtl_sig %>% 
  mutate(
    neglog10P  = pmin(-log10(eigenMT), 10),
    signed     = ifelse(beta >= 0,  neglog10P, -neglog10P),
    coloc_flag = ifelse(!is.na(PP.H4) & PP.H4 > 0.5, TRUE, FALSE)  # NA → FALSE
  )

## Ordering rows (SNPs) by genomic position …
snp_order <- eqtl_sig %>% 
  distinct(grch37, grch38) %>% 
  separate(grch38, into = c("chr","pos"), sep = ":", convert = TRUE, remove = FALSE) %>% 
  arrange(chr, pos) %>% 
  pull(grch38)

## …and columns (genes) by hierarchical clustering of beta
gene_order <- eqtl_sig %>% 
  select(grch38, gene_name, beta) %>% 
  pivot_wider(names_from = gene_name, values_from = beta, values_fill = 0) %>% 
  column_to_rownames("grch38") %>% 
  as.matrix() %>% 
  t() %>%                                  # genes in rows
  { 
    hc <- hclust(dist(., method = "euclidean"), method = "ward.D2")
    hc$labels[hc$order]                    # ordered gene names
  }

## ──────────────────────────────────────────────
## 4.  Build matrices for ComplexHeatmap
## ──────────────────────────────────────────────
mat_beta <- eqtl_sig %>%
  select(grch38, gene_name, beta) %>%
  pivot_wider(names_from = gene_name, values_from = beta, values_fill = 0) %>%
  column_to_rownames("grch38") %>%
  as.matrix() %>%
  .[snp_order, gene_order]

mat_coloc <- eqtl_sig %>%
  select(grch38, gene_name, coloc_flag) %>%
  pivot_wider(names_from = gene_name, values_from = coloc_flag, values_fill = FALSE) %>%
  column_to_rownames("grch38") %>%
  as.matrix() %>%
  .[snp_order, gene_order]

# First, identify which SNPs and genes have any colocalization
snps_with_coloc <- rownames(mat_coloc)[rowSums(mat_coloc, na.rm = TRUE) > 0]
genes_with_coloc <- colnames(mat_coloc)[colSums(mat_coloc, na.rm = TRUE) > 0]

# Create vectors to store the text colors and font styles
row_colors <- ifelse(rownames(mat_beta) %in% snps_with_coloc, "red", "black")
row_faces <- ifelse(rownames(mat_beta) %in% snps_with_coloc, "bold", "plain") 
row_sizes <- ifelse(rownames(mat_beta) %in% snps_with_coloc, 9.5, 8)  # Larger font for colocalized SNPs

column_colors <- ifelse(colnames(mat_beta) %in% genes_with_coloc, "red", "black")
column_faces <- ifelse(colnames(mat_beta) %in% genes_with_coloc, "bold", "plain")
column_sizes <- ifelse(colnames(mat_beta) %in% genes_with_coloc, 9.5, 8)  # Larger font for colocalized genes

## ---- per-SNP significant-gene counts (for right-hand bar) ----
row_counts <- rowSums(mat_beta != 0)

ha_right <- rowAnnotation(
  # Change title from "Sig\nGenes" to "Sig Genes" (single line)
  `Sig. Genes` = anno_barplot(
    row_counts,
    border = FALSE,
    gp      = gpar(fill = "#8DA0CB"),
    # Increase width from 15mm to 25mm
    width   = unit(25, "mm"),
    axis_param = list(
      at = seq(0, max(row_counts) + (max(row_counts) %% 2), by = 2),
      labels = as.character(seq(0, max(row_counts) + (max(row_counts) %% 2), by = 2)),
      labels_rot = 0,
      # Increase axis label font size
      gp = gpar(fontsize = 9)
    )
  ),
  annotation_name_side = "top",
  annotation_name_rot = 0,
  # Increase title font size from 8 to 10
  annotation_name_gp = gpar(fontsize = 12, fontface = "bold")
)

## ──────────────────────────────────────────────
## Figure 2B: Visit-1 cis-eQTL effect-size heatmap
## ──────────────────────────────────────────────
# Determine reasonable beta range for color scale
beta_range <- max(abs(range(mat_beta, na.rm = TRUE)))
col_fun <- colorRamp2(c(-beta_range, 0, beta_range), c("#2C7BB6", "white", "#D7191C"))  

ht <- Heatmap(
  mat_beta,
  name             = "Effect Size",
  col              = col_fun,
  na_col           = "grey90",
  cluster_rows     = FALSE,
  cluster_columns  = FALSE,
  
  # Use custom row name parameters with varied sizes
  show_row_names   = TRUE,
  row_names_gp     = gpar(fontsize = row_sizes, col = row_colors, fontface = row_faces),
  
  # Use custom column name parameters with varied sizes
  show_column_names= TRUE, 
  column_names_gp  = gpar(fontsize = column_sizes, col = column_colors, fontface = column_faces),
  
  heatmap_width    = unit(264, "mm"), # 120
  heatmap_height   = unit(166,  "mm"), # 80
  right_annotation = ha_right,

    cell_fun = function(j, i, x, y, width, height, fill) {
      # Draw the background fill for the cell
      if(!is.na(fill)) {
        grid.rect(x, y, width, height, 
                  gp = gpar(fill = fill, col = NA))
      }
      
      # Check if this cell has a significant eQTL (non-zero beta)
      if(!is.na(mat_beta[i, j]) && mat_beta[i, j] != 0) {
        
        # Check if it's also colocalized
        if(!is.na(mat_coloc[i, j]) && mat_coloc[i, j]) {
          # Full "X" for colocalized eQTLs
          grid.lines(
            x = c(x - width*0.4, x + width*0.4),
            y = c(y - height*0.4, y + height*0.4),
            gp = gpar(col = "black", lwd = 1.5)
          )
          grid.lines(
            x = c(x - width*0.4, x + width*0.4),
            y = c(y + height*0.4, y - height*0.4),
            gp = gpar(col = "black", lwd = 1.5)
          )
          
          # ADD THIS: Border box around colocalized pairs
          grid.rect(x, y, width*0.85, height*0.85,
                    gp = gpar(col = "black", fill = NA, lwd = 2))
          
        } else {
          # Single "\" for significant cis-eQTLs only
          grid.lines(
            x = c(x - width*0.4, x + width*0.4),
            y = c(y - height*0.4, y + height*0.4),
            gp = gpar(col = "black", lwd = 1.5)
          )
        }
      }
      
      # Light grey border around all cells
      grid.rect(x, y, width, height,
                gp = gpar(col = "grey90", fill = NA, lwd = 0.4))
    },

  # Customize the titles
  column_title = "Significant cis-eQTLs at Visit 1",
  column_title_gp = gpar(fontsize = 16, fontface = "bold"),  # Larger bold font for column title
  
  row_title = "Critical COVID-19 GWAS lead SNPs",
  row_title_gp = gpar(fontsize = 14),  # Larger bold font for row title
  
  # You can also customize the legend title
  heatmap_legend_param = list(
    title_gp = gpar(fontsize = 10, fontface = "bold"),  # Bold legend title
    labels_gp = gpar(fontsize = 9),  # Normal font for legend labels
    title_position = "topcenter"
  )
)
draw(ht)
```

## Figure 2A: SNP and gene attrition bars

Summarize the attrition from all tested SNPs and genes through eigenMT
significance to colocalization, as percentages of the total tested. The
broken-axis bar plot accommodates the 887 total tested genes alongside the much
smaller significant and colocalized counts.

```{r figure2a-attrition}
## Figure 2A: SNP and gene attrition and colocalization bars
## -----------------------------------------------------------------------
## Counts derived outside the plot for full reproducibility

# Clean data structure - counts only
panelB_df <- tibble::tribble(
  ~Category, ~Status,        ~Count,
  "SNPs",    "Total Tested",       58,
  "SNPs",    "Significant",        32,
  "SNPs",    "Colocalized",         9,
  "Genes",   "Total Tested",      887,
  "Genes",   "Significant",        52,
  "Genes",   "Colocalized",        10
) %>% 
  mutate(
    Category = factor(Category, levels = c("SNPs", "Genes")),
    # Reorder for filtering pipeline: Total (top) → Significant (middle) → Colocalized (bottom)
    Status = factor(Status, levels = c("Colocalized", "Significant", "Total Tested"))
  ) %>%
  group_by(Category) %>%
  mutate(
    Percentage = round(Count / first(Count[Status == "Total Tested"]) * 100, 1)
  ) %>%
  ungroup()

# More intuitive color scheme: light → dark = less → more specific
journal_colors <- c("Total Tested" = "#969696",   # Light gray for all tested
                    "Significant" = "#377EB8",     # Blue for significant eQTLs  
                    "Colocalized" = "#053061")     # Dark blue for colocalized

# Create improved publication-quality plot
bar_plots <- ggplot(panelB_df, aes(x = Status, y = Count, fill = Status)) +
  facet_wrap(~Category, scales = "free_y", ncol = 1, strip.position = "left") +
  
  # Main bars
  geom_bar(stat = "identity", width = 0.9, color = "white", size = 0.75) +
  
  # Percentage labels positioned next to bars
  geom_text(data = filter(panelB_df, Status != "Total Tested"),
            aes(label = paste0(round(Percentage, 1), "%"),
                y = Count + 0.5),  # Small offset to the right of bars
            size = 4.5, hjust = 0, vjust = 0.5, fontface = "bold") +
  
#  Count labels with conditional formatting
  geom_text(data = filter(panelB_df, Status == "Total Tested"),
            aes(label = comma(Count, accuracy = 1)),
            position = position_stack(vjust = 0.5),
            size = 5, fontface = "bold", color = "black") +
            
  geom_text(data = filter(panelB_df, Status == "Significant"),
            aes(label = comma(Count, accuracy = 1)),
            position = position_stack(vjust = 0.5),
            size = 5, fontface = "bold", color = "white") +
            
  geom_text(data = filter(panelB_df, Status == "Colocalized"),
            aes(label = comma(Count, accuracy = 1)),
            position = position_stack(vjust = 0.5),
            size = 4, fontface = "bold", color = "white") +
  
  # Clean, professional theme
  theme_minimal(base_size = 14) +
  theme(
    # Remove y-axis elements for cleaner look
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), 
    axis.title.y = element_blank(),
    axis.text.x = element_text(face = "bold", size = 12, angle = 0),
    axis.line.x = element_line(color = "black", size = 1),
    axis.ticks.x = element_line(color = "black", size = 0.5),
    axis.ticks.length.x = unit(0.25, "cm"),
    
    # Clean panels
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    
    # Improved strip styling
    strip.background = element_rect(fill = "grey90", color = "black", size = 0.5),
    strip.text = element_text(size = 14, face = "bold", margin = margin(5, 5, 5, 5)),
    
    # Legend improvements - make sure it's visible
    legend.position = "right",  # Simple right position
    legend.background = element_rect(fill = "white", color = "black", size = 0.5),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 11),
    legend.key.size = unit(0.8, "cm"),
    legend.margin = margin(8, 8, 8, 8),
    
    # Better spacing
    panel.spacing.y = unit(1.5, "cm"),
    aspect.ratio = 0.25,
    
    # Title styling
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "grey30", margin = margin(b = 15)),
    plot.caption = element_text(size = 10, color = "grey50", hjust = 0),
    plot.margin = margin(15, 25, 15, 25)
  ) +
  
  # Color scheme with proper order for filtering pipeline
  scale_fill_manual(
    name = "Analysis Stage",
    values = journal_colors,
    breaks = c("Total Tested", "Significant", "Colocalized"),  # Legend order
    guide = guide_legend(override.aes = list(color = "white", size = 0.5))
  ) +
  
  # Put back the essential scale break for the 887 value
  scale_y_break(c(60, 400), scales = "free", ticklabels = NULL, expand = c(0.01, 0.01)) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05)),
    breaks = function(x) {
      if(max(x) > 100) seq(0, ceiling(max(x)/100)*100, by = 200)
      else seq(0, ceiling(max(x)/10)*10, by = 20)
    }
  ) +
  
  # Improved labeling with methodological details
  labs(
    x = NULL,
    y = NULL,
    title = "eQTL Colocalization of Critical COVID-19 GWAS Variants",
    subtitle = "PBMC tissue • eigenMT FDR < 0.01 • COLOC posterior probability > 0.5",
    caption = "Colocalization indicates shared causal variants between GWAS and eQTL signals."
  ) +
  
  scale_x_discrete(expand = expansion(mult = c(0.3, 0.3))) +
  coord_flip()

# Display the plot
print(bar_plots)
```
