---
title: "Proteomics-global univariate and WGCNA analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))
suppressPackageStartupMessages(library(package = "msigdbr"))
suppressPackageStartupMessages(library(package = "clusterProfiler"))
suppressPackageStartupMessages(library(package = "ggraph"))
suppressPackageStartupMessages(library(package = "enrichplot"))
suppressPackageStartupMessages(library(package = "ggpubr"))

# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```

### Load codebase.R, load data matrices and set path to local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# Get codebase. Note this is our (i.e. proteomics) own due to some small changes
#source("data_analysis_template_codebase.R")
source("../../Codebase/codebase_v2.R")

#Data for all analysis except PCA plot
data_env <- load_IMPACC_datasets(DATA_VERSION                = "2022-01-01", 
                                 KEEP_COVID19_POS            = TRUE, 
                                 FILTER_BY_CORE_ASSAY_COHORT = TRUE,
                                 ASSAY_NAMES = "plasma_proteomics_global_dda")

# Data for PCA 
data_env_withHC <- load_IMPACC_datasets(DATA_VERSION                = "2022-01-01",
                                        KEEP_COVID19_POS            = FALSE, 
                                        FILTER_BY_CORE_ASSAY_COHORT = TRUE,
                                 ASSAY_NAMES = "plasma_proteomics_global_dda")

# Load the data objects into the environment.
for( n in names(data_env) ) {
  assign( n, data_env[[n]] )
}

# Setup the assay specific output directory
out_dir <- "../output"
```

### Functions: data loading
```{r data-loading}
#Get Enrichment DB
all_gene_sets = msigdbr(species = "Homo sapiens", category = "H")
msigdbr_t2g = all_gene_sets %>%
  dplyr::distinct(gs_name, gene_symbol) %>%
  mutate(gs_name = str_remove_all(gs_name, "HALLMARK_")) %>%
  as.data.frame()
```

### Functions: data processing
```{r data-processing}
#Define half minimum imputation method
impute_half_minimum <- function(dataset){
  #We use a simple imputation. We assume that missing values are below detection threshold.
  #We therefore fill the missing values by taking half of the minimum value for a given protein.
  #the forloop here under fills that.
  for(i in 1:ncol(dataset)) {       # for-loop over columns
    halfmin <- min(dataset[,i], na.rm = TRUE)/2
    dataset[,i][is.na(dataset[,i])] <- halfmin
  }
  
  return(dataset)
}

#copy the data, use 50% cut off filter.
plasma_proteomics_global_dda_counts_filtered <- plasma_proteomics_global_dda_counts %>% purrr::discard(~sum(is.na(.x))/length(.x)* 100 >=50)

#Impute using function above
plasma_proteomics_global_dda_counts_filtered_imputed <- impute_half_minimum(plasma_proteomics_global_dda_counts_filtered)

#Z-score the data.
plasma_proteomics_global_dda_counts_filtered_imputed_normalized <- data.frame(scale(plasma_proteomics_global_dda_counts_filtered_imputed))
```

### Supplementary Panel A: analysis code
```{r pvca analysis, warning=FALSE}
#Input Matrices
pvca_input_matrix <- t(plasma_proteomics_global_dda_counts_filtered_imputed)
pvca_input_raw <- t(plasma_proteomics_global_dda_counts)
pvca_input_metadata <- plasma_proteomics_global_dda_metadata

#Select Variables
pvca_input_phenodata <- clinical_data %>%
  mutate(event_date_week = findInterval(event_date, 
                                        vec        = c(0, 7, 14, 21, 28), 
                                        all.inside = TRUE),
         day_from_sympt  = event_date - symptom_date,
         sympt_date_week = findInterval(day_from_sympt, 
                                        vec        = c(0, 7, 14, 21, 28, Inf),
                                        all.inside = TRUE)) %>%
      select(sample_id, event_date_week, event_type, respiratory_status,
           sex, sympt_date_week, event_location, #death
           discretized_admit_age_quantile, ethnicity, participant_id, race,
           trajectory_group, enrollment_site)


# step 4 append meta data to pvca phenodata
pvca_input_phenodata <- pvca_input_metadata %>%
  rownames_to_column(var = "sample_id") %>%
  mutate(#core_site = gsub(pattern = "_.+", replacement = "", comment),
    
         plate     = interaction(phase, plate_num, drop = TRUE)) %>%
  select(sample_id, plate, phase) %>%
  merge(x     = pvca_input_phenodata,
        by    = "sample_id",
        all.x = TRUE)

# append rownames
pvca_input_phenodata <- pvca_input_phenodata %>%
  column_to_rownames(var = "sample_id")
pvca_input_phenodata <- pvca_input_phenodata[colnames(pvca_input_matrix), , drop = FALSE]

# add percent missing values (if processeed matrix required imputation otherwise comment line below)
pvca_input_phenodata$percent_missing <- factor(colMeans(is.na(pvca_input_raw[, rownames(pvca_input_phenodata)])))

# step 5: run PVCA
fit <- PVCA(counts    = pvca_input_matrix,
            meta      = pvca_input_phenodata,
            inter     = FALSE,
            threshold = 0.6)
```

### Supplementary Panel A: output generation code
```{r pvca, warning=FALSE}
#step 6: plot barplot with PVCA results
pvca_barplot_data <- data.frame(explained = as.vector(fit),
                                effect    = names(fit)) %>%
  arrange(explained) %>%
  mutate(effect = factor(effect, levels = effect))

pvca_barplot <- ggplot(data    = pvca_barplot_data,
       mapping = aes(y = effect, x = explained)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = signif(explained, digits = 3)),
            nudge_x   = 0.03,
            size      = 3) +
  labs(x = NULL, y = "Proportion of the variance explained") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1))

#Write Plot
pdf(file.path(out_dir, "/suppl_D.pdf"))
plot(pvca_barplot)
dev.off()
plot(pvca_barplot)
```

# Supplementary Panel B: analysis code
```{r, pca-no-outlierz b analysis , fig.width=6, fig.height=4.5}
#Load the data specificly for this plot.
for( n in names(data_env_withHC) ) {
  assign( n, data_env_withHC[[n]] )
}

#copy the data, use 50% cut off filter.
plasma_proteomics_global_dda_counts_filtered <- plasma_proteomics_global_dda_counts %>% purrr::discard(~sum(is.na(.x))/length(.x)* 100 >=50)

#Impute using function above
plasma_proteomics_global_dda_counts_filtered_imputed <- impute_half_minimum(plasma_proteomics_global_dda_counts_filtered)

#Z-score the data.
plasma_proteomics_global_dda_counts_filtered_imputed_normalized <- data.frame(scale(plasma_proteomics_global_dda_counts_filtered_imputed))

pc <- prcomp(plasma_proteomics_global_dda_counts_filtered_imputed_normalized)

plotDF <- pc$x[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = clinical_data, by = "sample_id") %>%
  merge(y  = rownames_to_column(plasma_proteomics_global_dda_metadata, var = "sample_id"),
        by = "sample_id") %>%
  mutate( enrollment_site = ifelse(test = participant_type %in% "Healthy control (Emory)",
                                  yes  = "Emory-Ctrl", 
                                  no   = enrollment_site))
enrollmentSite2color <- rainbow(n = length(unique(plotDF$enrollment_site))) %>%
  setNames(nm = unique(plotDF$enrollment_site))
enrollmentSite2color["Emory-Ctrl"] <- "black"
```

# Supplementary Panel B: output generation code
```{r, pca-no-outlierz b output, fig.width=6, fig.height=4.5}

plotPCA <- ggplot(data = plotDF,
       mapping = aes(x = PC1, y = PC2, color = enrollment_site))+
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = enrollmentSite2color) +
  scale_shape_manual(values = c(21, 23))+
  stat_ellipse()+
  labs(x     = paste0("1st dimension (",
                      round((pc$sdev^2/sum(pc$sdev^2))[1] * 100),
                      "%)"),
       y     = paste0("2nd dimension (",
                      round((pc$sdev^2/sum(pc$sdev^2))[2] * 100),
                      "%)"),
       tag   = "")+
  scale_fill_discrete(name = "Enrollment Sites") +
  theme_classic() +
  theme(legend.text     = element_text(size = 6),
        legend.key.size = unit(0.01, units = "npc"))

#Write Plot
pdf(file.path(out_dir, "/suppl_E.pdf"))
plot(plotPCA)
dev.off()
plot(plotPCA)
```

#Supplementary Panel C: analysis code
```{r, pca-no-outlierz c analysis, fig.width=6, fig.height=4.5}
#Load the data to normal back.
for( n in names(data_env) ) {
  assign( n, data_env[[n]] )
}

#copy the data, use 50% cut off filter.
plasma_proteomics_global_dda_counts_filtered <- plasma_proteomics_global_dda_counts %>% purrr::discard(~sum(is.na(.x))/length(.x)* 100 >=50)

#Impute using function above
plasma_proteomics_global_dda_counts_filtered_imputed <- impute_half_minimum(plasma_proteomics_global_dda_counts_filtered)

#Z-score the data.
plasma_proteomics_global_dda_counts_filtered_imputed_normalized <- data.frame(scale(plasma_proteomics_global_dda_counts_filtered_imputed))

#Get SFT values
sft_tuned = tune_soft_threshold_WGCNA(data_df = plasma_proteomics_global_dda_counts_filtered,
           networkType = "signed", corFnc = "cor",powers =c(100:200)/10)

if(!is.na(sft_tuned$powerEstimate)){
  ret =   generate_WGCNA_modules(
                   data_df = plasma_proteomics_global_dda_counts_filtered, 
                   # Preprocessed data in data.frame class
                   networkType = "signed",     
                   # Indicate the type of network to construct
                   power = sft_tuned$powerEstimate,              
                   # Provide a power value chosen from the first step, this parameter is mandatory.
                   minModuleSize = 20,      
                   # Provide a value for the minimum number of features a module can have; use a larger value (> 30) for datasets with thousands of features, and a smaller value otherwise.
                   corType = "pearson",         
                   # Correlation function
                   maxPOutliers = 0.1, 
                   # It is safer to use a smaller value for this parameter when corType = bicor is used.
                   assay_alias = "globalprot",         
                   # Append this assay_alias to the module names as prefix.
                   reassignThreshold = 1e-6,   
                   # This and rest of the parameters use default values, but users may want to tweak them based on the assay type and data.
                   mergeCutHeight = 0.05,
                   minKMEtoStay = 0.25,
                   minCoreKME = 0.1,
                   deepSplit = 4)}

```

#Supplementary Panel C: output generation code
```{r, pca-no-outlierz c output, fig.width=6, fig.height=4.5}
# Opening the graphical device
pdf(file.path(out_dir, "/suppl_F.pdf"))
# Creating a plot
draw(ret$WGCNAPlot1)
dev.off()
draw(ret$WGCNAPlot1)
```

### Panel M: analysis code
```{r}
#Take module 1 proteins, test them against Hallmark DB
globalmod4_extract <- ret$module_membership %>% filter(module == "globalprot_mod4")
globalmod4_enrichment <- enricher(gene = globalmod4_extract$feature, TERM2GENE = msigdbr_t2g)
```

### Panel M: output generation code
```{r}
#Plot using clusterprofiler
globalmod4_enrichmentplot <- cnetplot(globalmod4_enrichment, cex_category = 2, cex_label_category = 0.6, cex__label_gene = 0.6) +
  scale_color_manual(values = c("#3B9AB2", "#F21A00")) +
  theme(legend.position = "none") +
  labs(tag = "M")

#Write 
pdf(file.path(out_dir, "/main_M.pdf"))
plot(globalmod4_enrichmentplot)
dev.off()
plot(globalmod4_enrichmentplot)
```

#Panel N: analysis code


```{r univariate association study with WGCNA modules, echo=TRUE, warning=FALSE}
######################################
#Prepare data for univariate analysis.
######################################
clinical_subsets = clinical_data
rownames(clinical_subsets) = clinical_subsets$sample_id
row_ids = rownames(plasma_proteomics_global_dda_counts_filtered)
clinical_subsets = clinical_subsets[row_ids,]
visit1_row_ids = rownames(clinical_subsets[clinical_subsets$event_type == 'Visit 1',])

# Use the visit1_row_ids to only get samples from 'Visit 1':
modules_filtered = ret$MEs[visit1_row_ids ,]
clinical_subsets = clinical_subsets[visit1_row_ids ,]
data_use = modules_filtered 

# Add the respiratory_status_day14 as $endpoints
data_use$endpoints = as.factor(clinical_subsets$trajectory_group)

# Add the enrollment_site as $sites
data_use$sites = clinical_subsets$enrollment_site
data_use$age =  clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex
# Add the $control, needs to be different for each sample
# required for clmm (always requires some random effect)
data_use$control =  factor(1:nrow(data_use))

# Set module names:
module_names = colnames(modules_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$endpoints),]

######################
# Run ordinal analysis
######################
res_table_ordinal = data.frame(matrix(0, ncol = 3, nrow = length(module_names)+1))
colnames(res_table_ordinal) = c("AIC", "pval", "qval")
rownames(res_table_ordinal) = c(module_names, "combined")
for(j in 1:length(module_names)){
  print(j)
  my.formula0 = paste0('endpoints~ 1+(1|sites)+age+sex')
  my.formula1 = paste0('endpoints~',module_names[j],'+(1|sites)+age+sex')
  res_table_ordinal[j,1:2]=mixed_ordinal(my.formula0, my.formula1,data_use)
}
my.formula0 = paste0('endpoints~ 1+(1|sites)+age+sex')
my.formula1 = paste0('endpoints~',paste(module_names, collapse = " + "),'+ (1|sites)+age+sex')
res_table_ordinal[length(module_names)+1,1:2] = mixed_ordinal(my.formula0, my.formula1,data_use)
res_table_ordinal$qval =  qvalue::qvalue(res_table_ordinal$pval, fdr.level = 0.05, pi0 = 1)$qvalues

#######################
# RUN pairwise analysis
#######################

for(j in 1:length(module_names)){
  print(j)
  my.formula0 = paste0(module_names[j],"~ (1|sites)+age+sex")
  my.formula1 = paste0(module_names[j],"~ endpoints + (1|sites)+age+sex")
  if(j == 1){
    tmp = mixed_pairwise(my.formula0, my.formula1, data_use)
    res_table_pairwise = data.frame(matrix(0, ncol = length(tmp), nrow = length(module_names)))
    res_table_pairwise[j,] = tmp
    colnames(res_table_pairwise) = names(tmp)
  }else{
    res_table_pairwise[j,] = mixed_pairwise(my.formula0, my.formula1, data_use)
  }
}
rownames(res_table_pairwise) = module_names
res_table_pairwise_adj = qvalue::qvalue(as.vector(res_table_pairwise), fdr.level = 0.05, pi0 = 1)$qvalues

########################
#Make the visualization
########################

visit1Mod1 <- data_use %>%
  rownames_to_column(var = "sample_id") %>%
 dplyr::select(sample_id, globalprot_mod4, endpoints) %>%
  pivot_longer(cols = -c(sample_id, endpoints)) %>%
  filter(name %in% "globalprot_mod4") %>%
  mutate(name = "Global.prot.mod6:\nTissue Leakage",###############################
         trajectory_group = paste0("TG", endpoints)) %>%
  ggplot(mapping = aes(x = trajectory_group, y = value)) +
  geom_beeswarm(mapping = aes(color =trajectory_group), cex = 1.5, size = 0.8) +
  geom_boxplot(fill = "transparent", outlier.color= "transparent") +
  facet_wrap(facets= ~name, ncol = 1) +
  theme_bw() +
  theme(legend.pos = "none", strip.text = element_text(size = 7)) + 
  labs(y = "Eigenvalue", x = "Clinical trajectory group", tag = "N") +
  scale_color_manual(values = c("TG1" = "#639A21",
                                                          "TG2" = "#39828C",
                                                          "TG3"     = "#6371AD",
                                                          "TG4"     = "#BD7D31",
                                                          "TG5"      = "#9C3418"))
```

#Panel N: output generation code
```{r}
pdf(file.path(out_dir, "/main_N.pdf"))
plot(visit1Mod1)
dev.off()
plot(visit1Mod1)
```

# Panel O: analysis code
```{r}
########################################################
#Prepare data for longitudinal smoothing spline analysis
########################################################
clinical_subsets = clinical_data
rownames(clinical_subsets) = clinical_subsets$sample_id
row_ids = rownames(plasma_proteomics_global_dda_counts_filtered)
clinical_subsets = clinical_subsets[row_ids,]
modules_scores = ret$MEs
modules_filtered =modules_scores
clinical_subsets = clinical_subsets
data_use = cbind(modules_filtered, clinical_subsets[c("trajectory_group", "respiratory_status_day14", "respiratory_status_day28",  "event_date", "enrollment_site", "participant_id","sex", "discretized_admit_age_quantile")])

inputDF <- ret$MEs %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = dplyr::select(clinical_data, sample_id, event_date, participant_id, respiratory_status_day14, respiratory_status_day28,enrollment_site, trajectory_group, sex, admit_age, enrollment_site, discretized_admit_age_quantile),
        by = "sample_id") %>%
  # mutate_if(is.numeric, funs(ifelse(is.na(symptom_date), 0, .))) %>%  #### DFSO
  # mutate(event_date=event_date-symptom_date) %>%
  # select(-symptom_date) %>%################ DFSO!!
  mutate(outcomeD14  = cut(respiratory_status_day14,
                           breaks = c(1, 2, 4, 6, 7),
                           include.lowest = TRUE)) %>%
  mutate(outcomeD28  = cut(respiratory_status_day28,
                           breaks = c(1, 2, 4, 6, 7),
                           include.lowest = TRUE)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "respiratory_status_day14", "outcomeD14",
                         "respiratory_status_day28", "outcomeD28",
                         "sex", "admit_age", "discretized_admit_age_quantile")) %>%
  filter(!is.na(value) & !is.na(trajectory_group))

## Ensure proper variables are factors
inputDF[["trajectory_group"]] = ordered(as.factor(as.character(inputDF[["trajectory_group"]])), levels =
                                        c("1", "2", "3", "4", "5"))
inputDF$name <- as.factor(inputDF$name)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)

smooth_model_loop <- model_loop(inputDF, age_sex=T, modelType = "smoothSpline", endpoint = "trajectory_group")
print(smooth_model_loop)


#######################
# Create Visualization
#######################

## Extract a single factor, in this case mod1
exampleDF <- inputDF[inputDF$name == "globalprot_mod4",]
plotExample <- plot_model(exampleDF, model_loop = smooth_model_loop, modelType = "smoothSpline", p_adjust = NA, endpoint = "trajectory_group", signif_markers = F,
 remove_NS = T, bar_height = 6.5)

plotMod1 <- plotExample$data %>%
  mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
ggplot(mapping = aes(x = event_date, y = value)) +
  geom_line(mapping = aes(group = participant_id, y = value), color ='gray', alpha = 0.4) +
  geom_line(mapping = aes(group = participant_id, y = yhat), size = 0.6, alpha = 0.5, color = "black") +
  geom_point(mapping = aes(color = trajectory_group)) +
  geom_line(data = mutate(get("pred", plotExample$plot_env),
                              trajectory_group = paste0("TG", trajectory_group)),
            mapping = aes(group = trajectory_group, y = predicted),
                           color = "black", size = 1.5) +
  facet_wrap(facets = ~trajectory_group, nrow = 1) +
  labs(y = "Eigenvalue",  x = "Days from admission", tag = "O") +
  scale_color_manual(values= c("TG1"                          = "#639A21",
                               "TG2"                   = "#39828C",
                               "TG3"  = "#6371AD", 
                               "TG4"                     = "#BD7D31",
                               "TG5"                                         = "#9C3418")) +
  coord_cartesian(xlim = c(0, 42)) +
  theme_bw() +
  theme(legend.pos = "none",
        panel.grid.minor = element_blank())
```


# Panel O: output generation code
```{r final-panels, fig.height=3, fig.width=12, echo=FALSE, eval=TRUE}
pdf(file.path(out_dir, "/main_O.pdf"))
plot(plotMod1)
dev.off()
plot(plotMod1)

ggarrange(globalmod4_enrichmentplot, visit1Mod1, plotMod1, nrow = 1, widths = c(0.4,0.4, 1.2))

```

### Supplementary Table A: output generation code
```{r sup-table-a-output-gen}
supTab_modules <- ret$module_membership %>% 
  arrange(module) %>%
  rename(Module  = module,
         Feature = feature) %>%
  select(Module, Feature) %>%
  `rownames<-`(NULL)

write_csv(supTab_modules, file = file.path(out_dir, "proteomics_global_Modules.csv") )
```

### Supplementary Table B: output generation code
```{r sup-table-b-output-gen}

#Make the Annotations DF based on Hallmark Enrichments
Annotations <- data.frame(matrix(ncol = 10, nrow = 0))
colnames(Annotations) <- c("Module","Genes_In_Module","ID_HALLMARK","GeneRatio","BgRatio","pvalue","p.adjust","qvalue", "geneID", "Count")
num <- 0
for (i in unique(ret$module_membership$module)){
  num <- num + 1
  testmod_loop <- ret$module_membership %>% filter(module == i)
  testmod_loop_results <- enricher(gene = testmod_loop$feature, TERM2GENE = msigdbr_t2g)
  Annotations[num,1] <- i
  Annotations[num,2] <- toString(testmod_loop$feature)
  Annotations[num,3] <- toString(testmod_loop_results$ID)
  Annotations[num,4] <- toString(testmod_loop_results$GeneRatio)
  Annotations[num,5] <- toString(testmod_loop_results$BgRatio)
  Annotations[num,6] <- toString(testmod_loop_results$pvalue)
  Annotations[num,7] <- toString(testmod_loop_results$p.adjust)
  Annotations[num,8] <- toString(testmod_loop_results$qvalue)
  Annotations[num,9] <- toString(testmod_loop_results$geneID)
  Annotations[num,10] <- toString(testmod_loop_results$Count)
}

write_csv(Annotations, file = file.path(out_dir, "proteomics_global_Annotation.csv"))
```

### Supplementary Table C: analysis code
```{r sup-table-c-analysis, message=FALSE, warning=FALSE, results="hide"}
######################################
#Prepare data for univariate analysis.
######################################
clinical_subsets = clinical_data
rownames(clinical_subsets) = clinical_subsets$sample_id
row_ids = rownames(plasma_proteomics_global_dda_counts_filtered)
clinical_subsets = clinical_subsets[row_ids,]
visit1_row_ids = rownames(clinical_subsets[clinical_subsets$event_type == 'Visit 1',])

# Use the visit1_row_ids to only get samples from 'Visit 1':
modules_filtered = ret$MEs[visit1_row_ids ,]
clinical_subsets = clinical_subsets[visit1_row_ids ,]
data_use = modules_filtered 

# Add the respiratory_status_day14 as $endpoints
data_use$endpoints = as.factor(clinical_subsets$trajectory_group)

# Add the enrollment_site as $sites
data_use$sites = clinical_subsets$enrollment_site
data_use$age =  clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex
data_use$control =  factor(1:nrow(data_use))

# Set module names:
module_names = colnames(modules_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$endpoints),]

#############################
# BUILD ALL VISIT 1 DATASETS
#############################

# a model with only intercept and random effect across sites
res_table_ordinal <- data.frame(matrix(NA, ncol = 4, nrow = length(module_names)))
colnames(res_table_ordinal) <- c("AIC", "pval", "qval", "coef")
rownames(res_table_ordinal) <- module_names
for(j in 1:length(module_names)) {
  my.formula0 <- paste0('endpoints~ 1+(1|sites)+age+sex')
  my.formula1 <- paste0('endpoints~',module_names[j],'+(1|sites)+age+sex')
  res_table_ordinal[j,1:2] <- mixed_ordinal(my.formula0, my.formula1,data_use)
  res_table_ordinal[j,4] <- ordinal::clmm(formula(my.formula1), 
                                          data = data_use) %>%
      coef() %>%
    .[module_names[j]]
}

res_table_ordinal$qval <- qvalue::qvalue(res_table_ordinal$pval, 
                                         fdr.level = 0.05, 
                                         pi0       = 1)$qvalues


supTab_resultsVisit1 <- res_table_ordinal %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  filter(pval <= 0.05) %>%
  mutate(`Module (or Feature)` = paste0("Proteomics_global", ".", `Module (or Feature)`),
         Analysis              = "Visit 1 analysis - overall",
         Direction             = ifelse(test = sign(coef) %in% 1,
                                        yes  = "Severe",
                                        no   = "Mild")) %>%
  rename(`P value` = pval) %>%
  select(`Module (or Feature)`,
          Analysis,
         `P value`,
         Direction) %>%
  `rownames<-`(NULL)

# subfunction use to fit ordinal regression and extract reg. coef
mixed_pairwise_coef <- function(my.formula0, my.formula1, data_use) {
  endpoints0 <- sort(unique(data_use$endpoints))
  pair_names <- c()
  res_table <- c()
  for(i in 1:(length(endpoints0)-1)){
    for(j in (i+1):(length(endpoints0))){
      pair_names <- c(pair_names, paste0(endpoints0[i],"|", endpoints0[j]))
      data_tmp <- data_use[data_use$endpoints %in% endpoints0[c(i, j)], ]
      fit <- lme4::lmer(my.formula1, data = data_tmp)
      res_table <- c(res_table, 
                     unique(coef(fit)$sites[,grep(pattern = "endpoint", 
                                               names(coef(fit)$sites))]))
    }
  }
  names(res_table) <- pair_names
  return(value = res_table)
}
  
res_table_pairwise <- NULL
res_table_pairwise_coef <- NULL
for(j in 1:length(module_names)){
  my.formula0 = paste0(module_names[j],"~ (1|sites)+age+sex")
  my.formula1 = paste0(module_names[j],"~ endpoints + (1|sites)+age+sex")
  res_table_pairwise <- rbind(res_table_pairwise,
                              mixed_pairwise(my.formula0, my.formula1, data_use))
  res_table_pairwise_coef <- rbind(res_table_pairwise_coef,
                                   mixed_pairwise_coef(my.formula0, 
                                                       my.formula1, 
                                                       data_use))
}

rownames(res_table_pairwise) <- module_names
rownames(res_table_pairwise_coef) <- module_names

supTab_resultsVisit1tmp <- res_table_pairwise %>%
  as.data.frame() %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  pivot_longer(cols      = -`Module (or Feature)`, 
               names_to  = "Analysis", 
               values_to = "P value")
res_table_pairwise_coef %>%
  as.data.frame() %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  pivot_longer(cols      = -`Module (or Feature)`,
               names_to  = "Analysis",
               values_to = "Direction") %>%
  merge(x  = supTab_resultsVisit1tmp, 
        by = c("Module (or Feature)", "Analysis")) -> supTab_resultsVisit1tmp

supTab_resultsVisit1tmp <- supTab_resultsVisit1tmp %>%
  mutate(`Module (or Feature)` = paste0("Proteomics_global", ".", `Module (or Feature)`),
         Analysis              = paste0("Visit 1 analysis - ", Analysis),
         Direction             = ifelse(test = sign(Direction) %in% 1,
                                        yes  = "Severe",
                                        no   = "Mild")) %>%
  filter(`P value` <= 0.05)

supTab_resultsVisit1 <- rbind(supTab_resultsVisit1,
                              supTab_resultsVisit1tmp)


########################################################
#Prepare data for longitudinal smoothing spline analysis
########################################################
clinical_subsets = clinical_data
rownames(clinical_subsets) = clinical_subsets$sample_id
row_ids = rownames(plasma_proteomics_global_dda_counts_filtered)
clinical_subsets = clinical_subsets[row_ids,]
modules_scores = ret$MEs
modules_filtered =modules_scores
clinical_subsets = clinical_subsets
data_use = cbind(modules_filtered, clinical_subsets[c("trajectory_group", "respiratory_status_day14", "respiratory_status_day28",  "event_date", "enrollment_site", "participant_id","sex", "discretized_admit_age_quantile")])

inputDF <- ret$MEs %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = dplyr::select(clinical_data, sample_id, event_date, participant_id, respiratory_status_day14, respiratory_status_day28,enrollment_site, trajectory_group, sex, admit_age, enrollment_site, discretized_admit_age_quantile),
        by = "sample_id") %>%
  # mutate_if(is.numeric, funs(ifelse(is.na(symptom_date), 0, .))) %>%  #### DFSO
  # mutate(event_date=event_date-symptom_date) %>%
  # select(-symptom_date) %>%################ DFSO!!
  mutate(outcomeD14  = cut(respiratory_status_day14,
                           breaks = c(1, 2, 4, 6, 7),
                           include.lowest = TRUE)) %>%
  mutate(outcomeD28  = cut(respiratory_status_day28,
                           breaks = c(1, 2, 4, 6, 7),
                           include.lowest = TRUE)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "respiratory_status_day14", "outcomeD14",
                         "respiratory_status_day28", "outcomeD28",
                         "sex", "admit_age", "discretized_admit_age_quantile")) %>%
  filter(!is.na(value) & !is.na(trajectory_group))

#Set Factors Correct
inputDF$trajectory_group <- factor(inputDF$trajectory_group)
inputDF$name <- as.factor(inputDF$name)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)

##################
# SMOOTHING SPLINE
##################

## Smooth spline takes a long time, demonstrating by cutting down to just ten factors
smooth_spline_model_loop <- model_loop(inputDF, 
                                       modelType = "smoothSpline", 
                                       endpoint  = "trajectory_group")

# fetch directionality from lme (TG1 vs TG5)
smooth_spline_model_loop$trajectory_group5 <- NA
smooth_spline_model_loop$"event_date:trajectory_group5" <- NA
for (modName in rownames(smooth_spline_model_loop)) {
  fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
             random = ~1|enrollment_site/participant_id,
             data   = dplyr::filter(inputDF, name %in% modName))
  smooth_spline_model_loop[modName, "trajectory_group5"] <-   unique(coef(fit)["trajectory_group5"])
  smooth_spline_model_loop[modName, "event_date:trajectory_group5"] <- unique(coef(fit)["event_date:trajectory_group5"])
}


# generate supplementary table with longitudinal analaysis overall
supTab_resultsLongitudinal <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  select(`Module (or Feature)`, p.slope, p.intercept, trajectory_group5, `event_date:trajectory_group5`) %>%
  rename(coef.intercept = trajectory_group5, 
         coef.slope = `event_date:trajectory_group5`) %>%
  pivot_longer(cols = -`Module (or Feature)`, names_to = c("cname", "Analysis"), names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = cname, values_from = value) %>%
  filter(p <= 0.05) %>%
  mutate(Analysis = recode(Analysis,
                           `slope` = "Longitudinal analysis - overall shape",
                           `intercept` = "Longitudinal analysis - overall average"),
         Direction = ifelse(test = sign(coef) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
    rename(`P value` = p) %>%
  select(`Module (or Feature)`,
         Analysis,
         `P value`,
         Direction) %>%
  `rownames<-`(NULL)


# append pairwise comparison to longitudinal results
supTab_resultsLongitudinalPairwise <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  select(`Module (or Feature)`, 
         matches(match = "^p.slope_.v.$"),
         matches(match = "^p.intercept_.v.$")) %>%
  pivot_longer(cols      = -`Module (or Feature)`,
               names_to  = "Analysis",
               values_to = "P value") %>%
  mutate(Analysis = gsub(pattern     = "slope",
                         replacement = "shape",
                         Analysis),
         Analysis = gsub(pattern     = "intercept",
                         replacement = "average",
                         Analysis),
         Analysis = gsub(pattern = "^p\\.(.+)_(.)v(.)$",
                         replacement = "Longitudinal analysis - \\2|\\3 \\1",
                         Analysis))

combMat <- combn(unique(inputDF$trajectory_group), m = 2)
coefCombMat <- apply(combMat, MARGIN = 2, FUN = function(comparison) {
  inputTemp <- filter(inputDF, trajectory_group %in% comparison)
  
  coefDF <- lapply(rownames(smooth_spline_model_loop), FUN = function(modName) {
    fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
               random = ~1|enrollment_site/participant_id,
               data   = filter(inputTemp, name %in% modName))
    average <- unique(coef(fit)[grep(pattern = "^trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    shape <- unique(coef(fit)[grep(pattern = "^event_date:trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    return(value = data.frame(`Module (or Feature)` = modName,
                              average, 
                              shape,
                              comparison = paste(sort(comparison), collapse = "|"),
                              check.names = FALSE))
  }) %>%
    do.call(what = rbind)
  return(value = coefDF)
}) %>%
  do.call(what = rbind)


coefCombMat %>%
  pivot_longer(cols      = -c(`Module (or Feature)`, comparison),
               names_to  = "Analysis",
               values_to = "Direction") %>%
  mutate(Analysis = paste("Longitudinal analysis -",
                          comparison,
                          Analysis),
         Direction = ifelse(test = sign(Direction) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
  select(-comparison) %>%
  merge(x  = supTab_resultsLongitudinalPairwise,
        by = c("Module (or Feature)", "Analysis")) -> 
  supTab_resultsLongitudinalPairwise

# append to overall results
supTab_resultsLongitudinalPairwise %>%
  filter(`P value` <= 0.05 &
           `Module (or Feature)` %in% 
           supTab_resultsLongitudinal$"Module (or Feature)") %>%
  rbind(supTab_resultsLongitudinal, .) -> supTab_resultsLongitudinal

###############################################################
#Prepare data for longitudinal smoothing spline analysis - DFSO
###############################################################

clinical_subsets = clinical_data
rownames(clinical_subsets) = clinical_subsets$sample_id
row_ids = rownames(plasma_proteomics_global_dda_counts_filtered)
clinical_subsets = clinical_subsets[row_ids,]
modules_scores = ret$MEs
modules_filtered =modules_scores
clinical_subsets = clinical_subsets
data_use = cbind(modules_filtered, clinical_subsets[c("trajectory_group", "respiratory_status_day14", "respiratory_status_day28",  "event_date", "enrollment_site", "participant_id","sex", "discretized_admit_age_quantile")])

inputDF <- ret$MEs %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = dplyr::select(clinical_data, sample_id, event_date, participant_id, respiratory_status_day14, respiratory_status_day28,enrollment_site, trajectory_group, sex, admit_age, symptom_date, enrollment_site, discretized_admit_age_quantile),
        by = "sample_id") %>%
  mutate(event_date = event_date - symptom_date) %>%
  select(-symptom_date) %>%################ DFSO!!
  mutate(outcomeD14  = cut(respiratory_status_day14,
                           breaks = c(1, 2, 4, 6, 7),
                           include.lowest = TRUE)) %>%
  mutate(outcomeD28  = cut(respiratory_status_day28,
                           breaks = c(1, 2, 4, 6, 7),
                           include.lowest = TRUE)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "respiratory_status_day14", "outcomeD14",
                         "respiratory_status_day28", "outcomeD28",
                         "sex", "admit_age", "discretized_admit_age_quantile")) %>%
  filter(!is.na(value) & !is.na(trajectory_group) & !is.na(event_date))

#Set Factors Correct
inputDF$trajectory_group <- factor(inputDF$trajectory_group)
inputDF$name <- as.factor(inputDF$name)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)


##################
# SMOOTHING SPLINE
##################

## Smooth spline takes a long time, demonstrating by cutting down to just ten factors
smooth_spline_model_loop <- model_loop(inputDF, 
                                       modelType = "smoothSpline", 
                                       endpoint  = "trajectory_group")

# fetch directionality from lme
smooth_spline_model_loop$trajectory_group5 <- NA
smooth_spline_model_loop$"event_date:trajectory_group5" <- NA
for (modName in rownames(smooth_spline_model_loop)) {
  fit <- lme(fixed= value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
      random = ~1|enrollment_site/participant_id,
      data = filter(inputDF, name %in% modName))
smooth_spline_model_loop[modName, "trajectory_group5"] <- unique(coef(fit)["trajectory_group5"])
smooth_spline_model_loop[modName, "event_date:trajectory_group5"] <- unique(coef(fit)["event_date:trajectory_group5"])
}

supTab_resultsLongitudinalDFSO <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  select(`Module (or Feature)`, p.slope, p.intercept, trajectory_group5, `event_date:trajectory_group5`) %>%
  rename(coef.intercept = trajectory_group5, 
         coef.slope = `event_date:trajectory_group5`) %>%
  pivot_longer(cols = -`Module (or Feature)`, names_to = c("cname", "Analysis"), names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = cname, values_from = value) %>%
  filter(p <= 0.05) %>%
  mutate(Analysis = recode(Analysis,
                           `slope` = "Longitudinal analysis DFSO - overall shape",
                           `intercept` = "Longitudinal analysis DFSO- overall average"),
         Direction = ifelse(test = sign(coef) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
    rename(`P value` = p) %>%
  select(`Module (or Feature)`,
         Analysis,
         `P value`,
         Direction) %>%
  `rownames<-`(NULL)
               

combMat <- combn(unique(inputDF$trajectory_group), m = 2)
coefCombMat <- apply(combMat, MARGIN = 2, FUN = function(comparison) {
  inputTemp <- filter(inputDF, trajectory_group %in% comparison)
  
  coefDF <- lapply(rownames(smooth_spline_model_loop), FUN = function(modName) {
    fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
               random = ~1|enrollment_site/participant_id,
               data   = filter(inputTemp, name %in% modName))
    average <- unique(coef(fit)[grep(pattern = "^trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    shape <- unique(coef(fit)[grep(pattern = "^event_date:trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    return(value = data.frame(`Module (or Feature)` = modName,
                              average, 
                              shape,
                              comparison = paste(sort(comparison), collapse = "|"),
                              check.names = FALSE))
  }) %>%
    do.call(what = rbind)
  return(value = coefDF)
}) %>%
  do.call(what = rbind)

supTab_resultsLongitudinalDFSOPairwise <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  select(`Module (or Feature)`, 
         matches(match = "^p.slope_.v.$"),
         matches(match = "^p.intercept_.v.$")) %>%
  pivot_longer(cols      = -`Module (or Feature)`,
               names_to  = "Analysis",
               values_to = "P value") %>%
  mutate(Analysis = gsub(pattern     = "slope",
                         replacement = "shape",
                         Analysis),
         Analysis = gsub(pattern     = "intercept",
                         replacement = "average",
                         Analysis),
         Analysis = gsub(pattern = "^p\\.(.+)_(.)v(.)$",
                         replacement = "Longitudinal analysis DFSO - \\2|\\3 \\1",
                         Analysis))
coefCombMat %>%
  pivot_longer(cols      = -c(`Module (or Feature)`, comparison),
               names_to  = "Analysis",
               values_to = "Direction") %>%
  mutate(Analysis = paste("Longitudinal analysis DFSO -",
                          comparison,
                          Analysis),
         Direction = ifelse(test = sign(Direction) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
  select(-comparison) %>%
  merge(x  = supTab_resultsLongitudinalDFSOPairwise,
        by = c("Module (or Feature)", "Analysis")) -> supTab_resultsLongitudinalDFSOPairwise


# append to overall results
supTab_resultsLongitudinalDFSOPairwise %>%
  filter(`P value` <= 0.05) %>%
  rbind(supTab_resultsLongitudinalDFSO, .) -> supTab_resultsLongitudinalDFSO

```

### Supplementary Table C: output generation code
```{r sup-table-c-output-gen}

# combine all sup table results
supTab_results <- rbind(supTab_resultsVisit1,
                        supTab_resultsLongitudinal,
                        supTab_resultsLongitudinalDFSO)

write_csv(supTab_results, file = file.path(out_dir, "proteomics_global_Results.csv"))

```


### Session info
```{r, print-session-info}
sessionInfo()
```
