---
title: "pbmc_rnaseq"
output: html_document
---

Source: 20230123_bulk_RNAseq_pbmc.Rmd

```{r}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(patchwork)

library(limma)
library(edgeR)

library(fgsea)

library(gamm4)

setwd("~/20220902_IMPACC_aging")

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```


# Data import

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE # replace NA with FALSE
clin.event$steroids <- as.factor(clin.event$steroids)

head(clin.event)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

### Merge clinical metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ, by = "participant_id", all = FALSE, sort = FALSE
)
```

Add patients' individual metadata.

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

## RNA-seq data

```{r}
counts <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1,
  check.names = FALSE
)
counts <- t(counts)
print(counts[1:6,1:6])

counts.metadata <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
print(head(counts.metadata))

rData <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-RowFeature.csv",
  header = FALSE, row.names = NULL,
  check.names = FALSE
)
colnames(rData) <- rData[2,]
rData <- rData[-(1:2),]
print(head(rData))
```

## Viral load data

SARS-CoV-2 viral load is readily available in nasal-metagenomics-Metadata.csv, sc2_rpm column.

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)

# mngs <- read.csv(
#   "/data/nasal-metagenomics/current/nasal-metagenomics-AllGenera-Counts_RPM.csv",
#   header = TRUE, row.names = 1,
#   check.names = FALSE
# )

# mngs.rData <- read.csv(
#   "/data/nasal-metagenomics/current/nasal-metagenomics-AllGenera-RowFeatures.csv",
#   skip = 1, header = TRUE, row.names = 1,
#   check.names = FALSE
# )
# sc2.id <- mngs.rData %>%
#   subset(name == "Betacoronavirus") %>%
#   rownames()
```

## Only keep samples that pass QC

```{r}
# RNA-seq QC
counts.metadata <- subset(counts.metadata, sample_status == "IMPACC sample assayed and passed QC")

# mNGS QC
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")
# Remove rows without sc2_rpm
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))

# Get a list of samples for RNA-seq count
pass.qc <- Reduce(
  intersect, 
  list(colnames(counts), counts.metadata$sample_id, clin.sample.event.individ$sample_id)
)

counts <- counts[,pass.qc]
clin.sample.event.individ <- subset(clin.sample.event.individ, sample_id %in% pass.qc)
```

## Load pathways for GSEA

```{r}
# Reactome
reactome <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "C2", subcategory = "CP:REACTOME")
reactome.list <- split(x = reactome$ensembl_gene, f = reactome$gs_name) # for fgsea
reactome.list2 <- reactome %>% select(gs_name, ensembl_gene) # for clusterProfiler

# Hallmark gene sets
hallmark <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "H")
hallmark.list <- split(x = hallmark$ensembl_gene, f = hallmark$gs_name)
hallmark.list2 <- hallmark %>% select(gs_name, ensembl_gene)
```

# Hospital visit 1 samples

## Prepare data

```{r}
# Keep visit 1 samples for RNA-seq
clin.sample.event.individ.1 <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Keep visit 1 samples for viral load
clin.sample.event.individ.sc2.1 <- clin.sample.event.individ.sc2 %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Add viral load
clin.sample.event.individ.1 <- merge(
  clin.sample.event.individ.1, clin.sample.event.individ.sc2.1[,c("participant_id","sc2_rpm")],
  by = "participant_id", all = FALSE, sort = FALSE
)
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.1$sc2_rpm <- log10(clin.sample.event.individ.1$sc2_rpm + 1)

# Drop unused levels
clin.sample.event.individ.1 <- droplevels(clin.sample.event.individ.1)

# Divide into age groups
clin.sample.event.individ.1$age_group <- cut(
  clin.sample.event.individ.1$admit_age,
  breaks=c(17,46,54,62,70,96)
  )

# Rename levels: convert round brack to square bracket
levels(clin.sample.event.individ.1$age_group) <- c("[18,46]","[47,54]","[55,62]","[63,70]","[71,96]")

# Any row with no sc2_rpm?
print(any(is.na(clin.sample.event.individ.1$sc2_rpm)))

# Any duplicated participant?
print(any(duplicated(clin.sample.event.individ.1$participant_id)))
```

```{r}
# Subset count data
counts.1 <- counts[,clin.sample.event.individ.1$sample_id]
```

QC.

```{r}
# Keep only protein_coding genes
counts.1 <- counts.1[rData[rData$gene_biotype=="protein_coding", "gene_id"],]

# Keep genes with at least 10 counts in 20% of samples
keep <- rowSums(counts.1>=10) >= 0.2*ncol(counts.1)
counts.1 <- counts.1[keep,]
```

Export normalized gene expression at visit 1.

```{r}
# Normalize counts at visit 1 without taking into account any covariate
temp <- counts.1
colnames(temp) <- clin.sample.event.individ.1$event_id
temp <- voom(temp, 
             normalize.method = "quantile",
             plot = T)
# write.csv(temp$E, "20230123_bulk_RNAseq_pbmc_visit1_counts_voom-quantile.csv")

# Export rows data
temp.rData <- rData
rownames(temp.rData) <- temp.rData$gene_id
# write.csv(temp.rData[rownames(temp),],
#           "20230123_bulk_RNAseq_pbmc_visit1_RowFeatures_voom-quantile.csv",
#           row.names = FALSE)

remove(temp, temp.rData)
```

## Explore data

```{r}
# Age distribution
hist(clin.sample.event.individ.1$admit_age, breaks=20,
     xlab = "Admit age", main = "PBMC")

# Age by sex
boxplot(admit_age~sex, data=clin.sample.event.individ.1, main = "PBMC")
```

# Visit 1: effects of age

## Is BMI associated with gene expression?

```{r}
clin.sample.event.individ.1.bmi <- clin.sample.event.individ.1 %>%
  subset(!is.na(bmi))
counts.1.bmi <- counts.1[,clin.sample.event.individ.1.bmi$sample_id]

design <- model.matrix(~admit_age + bmi + sex + respiratory_status,
                       data=clin.sample.event.individ.1.bmi)
vwts <- voom(counts.1.bmi, 
             design = design,
             normalize.method = "quantile",
             plot = T)
vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

Answer: no, BMI does not have a strong effect on PBMC gene expression.

## Without controlling for viral load

```{r}
design <- model.matrix(~ admit_age + sex + respiratory_status,
                       data=clin.sample.event.individ.1)

vwts <- voom(counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

```{r}
res <- topTable(vfit, coef = "admit_age",
                sort.by = "none", number = Inf,
                adjust.method = "BH", p.value = 1)
head(res)
```

Save DE results

```{r}
write.csv(
  res,
  "20230123_bulk_RNAseq_pbmc_visit1_DE_no-viralload.csv"
)
```

Volcano plot

```{r}
# Identify direction
res.toplot <- merge(res, rData[,c("gene_id","gene_name")],
                    all.x = TRUE, all.y = FALSE,
                    by.x = "row.names", by.y = "gene_id",
                    sort = FALSE)
res.toplot$direction <- "Not significant"
res.toplot[res.toplot$adj.P.Val<0.05 & res.toplot$logFC>0,"direction"] <- "Increase with age"
res.toplot[res.toplot$adj.P.Val<0.05 & res.toplot$logFC<0,"direction"] <- "Decrease with age"
res.toplot$direction <- factor(
  res.toplot$direction,
  levels=c("Increase with age","Decrease with age","Not significant"))

# Limit minimum P value to 10e-20
res.toplot[res.toplot$adj.P.Val<1e-10,"adj.P.Val"] <- 1e-10

# Figure 4A
p <- ggplot(data=res.toplot) +
  geom_point(aes(x=logFC, y=-log10(adj.P.Val), color=direction), size=0.8) +
  # geom_hline(yintercept=-log10(0.05), color="red", linetype="dashed") +
  scale_color_manual(values=c("darkred","dodgerblue4","gray75"), # c("#d6604d","#4393c3","gray75")
                     name="", guide="none") +
  xlim(-0.06, 0.06) + 
  scale_y_continuous(limits=c(-0.4,10.1), expand=c(0,0),
                     breaks=c(0,5,10), labels=c(0,5,"\u2265 10")) +
  labs(x="Slope \n(change per 1 year of age)", y="-log10(adjusted P)",
       title="PBMC genes associated\nwith age at Visit 1") +
  my.theme # RNAseq_pbmc_visit1_age_no-viralload, 4.8 x 3.6 in or clipboard 470 x 420
```

Reactome GSEA

```{r}
# Sort gene list in descending order
gene.ranks <- res$t
names(gene.ranks) <- rownames(res)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)

# # How many pathways are significant?
# print(sum(reactome.gsea$padj < 0.05))
# # What are they?
# reactome.gsea.sig <- subset(reactome.gsea, padj < 0.05)

# Save the result for comparison later
reactome.gsea.noviral <- reactome.gsea
```

```{r}
# Plot significant reactome pathways
ggplot(data = arrange(reactome.gsea, desc(-padj)) %>% slice(1:50) ,
       aes(x=reorder(pathway,NES), y=NES)) +
  geom_segment(aes(xend=reorder(pathway,NES)), yend=0) +
  geom_point(aes(fill=as.factor(NES>0), color=as.factor(padj<0.05)),
             pch=21, size=2, stroke=1) +
  # ylim(-4, 4) +
  scale_color_manual(values=c("white","black")) +
  labs(x="Reactome pathways", y="NES", title="Top 50 pathways upregulated with age") +
  coord_flip() + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,0,0,0), "cm"),
    legend.key = element_rect(fill = "gray90")
  )
```

Save GSEA results

```{r}
# data.table::fwrite(
#   reactome.gsea.noviral,
#   file="20230123_bulk_RNAseq_pbmc_visit1_admit-age_sex-ordinalscore_reactome-gsea_fdr0.05.tsv",
#   sep="\t", sep2=c("", " ", "")
# )
```

### Gene-concept network plot

```{r}
# Sort gene list in descending order
gene.ranks <- res$t
names(gene.ranks) <- rownames(res)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

# GSEA with clusterProfiler
reactome.cp.res <- clusterProfiler::GSEA(
  geneList = gene.ranks,
  TERM2GENE = reactome.list2,
  minGSSize = 15,
  maxGSSize = 500,
  pvalueCutoff = 1, pAdjustMethod = "BH"
  )

# Try replacing ensemblid with gene symbol
reactome.cp.res@gene2Symbol <- rData$gene_name
names(reactome.cp.res@gene2Symbol) <- rData$gene_id

# gsePathway requires entrez id
# reactome.cp.res2 <- ReactomePA::gsePathway(
#   gene.ranks, 
#   pvalueCutoff = 1,
#   pAdjustMethod = "BH",
#   verbose = FALSE,
#   readable = TRUE)
```

```{r}
# Plot certain pathways only
my.pathways <- c(
  "REACTOME_INTERFERON_ALPHA_BETA_SIGNALING","REACTOME_INTERFERON_GAMMA_SIGNALING",
  "REACTOME_TOLL_LIKE_RECEPTOR_TLR1_TLR2_CASCADE","REACTOME_TCR_SIGNALING",
  "REACTOME_PD_1_SIGNALING")

# Fold change
gene.FC <- res$logFC
names(gene.FC) <- rownames(res)

# Gene-concept network
p1 <- enrichplot::cnetplot(
  reactome.cp.res,
  showCategory = my.pathways,
  node_label="all",
  foldChange = gene.FC,
  cex_label_gene = 0.6, cex_label_category = 0.9) +
  # ggraph::geom_node_text(aes_(label=~name), repel=TRUE, size = 1)
  scale_colour_gradient2(low="#67a9cf", high="#ef8a62", midpoint = 0)

p2 <- enrichplot::cnetplot(
  reactome.cp.res,
  showCategory = my.pathways,
  node_label="none",
  foldChange = gene.FC,
  cex_label_gene = 0.6, cex_label_category = 0.6) +
  scale_colour_gradient2(low="#67a9cf", high="#ef8a62", midpoint = 0)

# Manually convert the gene IDs to gene names
p3 <- p2
p3$data$name[-c(1:5)] <- unname(reactome.cp.res@gene2Symbol[p3$data$name[-c(1:5)]])
p3 <- p3 + 
  ggraph::geom_node_text(
    aes_(label=~name), data = p3$data[-c(1:5),],
    repel=TRUE, size = 3) # RNAseq_pbmc_visit1_no-viralload_gene-concept_gene-annotation.pdf, 10 x 7 in
```

### Try controlling for race

First, check for effects of race on gene expression.

```{r}
design <- model.matrix(~ admit_age + respiratory_status + race,
                       data=clin.sample.event.individ.1)

vwts <- voom(counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

print(summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05)))

res.race <- topTable(vfit, coef = 6:12,
                     sort.by = "none", number = Inf,
                     adjust.method = "BH", p.value = 1)
print(sum(res.race$adj.P.Val < 0.05))
```

Second, try controlling for race in the analysis of age's effect.

```{r}
design <- model.matrix(~ admit_age + sex + respiratory_status + race,
                       data=clin.sample.event.individ.1)

vwts <- voom(counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

print(summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05)))

res.race <- topTable(vfit, coef = "admit_age",
                     sort.by = "none", number = Inf,
                     adjust.method = "BH", p.value = 1)
print(sum(res.race$adj.P.Val < 0.05))
```

GSEA

```{r}
# Sort gene list in descending order
gene.ranks <- res.race$t
names(gene.ranks) <- rownames(res.race)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.race <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
```

Compare with and without controlling for race.

```{r}
# Manually choose pathways
my.pathways <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING","REACTOME_INTERFERON_GAMMA_SIGNALING",
                 "REACTOME_INTERLEUKIN_12_SIGNALING",
                 "REACTOME_INTERLEUKIN_2_FAMILY_SIGNALING","REACTOME_TRAF6_MEDIATED_IRF7_ACTIVATION",
                 "REACTOME_TNFR2_NON_CANONICAL_NF_KB_PATHWAY","REACTOME_TOLL_LIKE_RECEPTOR_CASCADES",
                 "REACTOME_TOLL_LIKE_RECEPTOR_TLR1_TLR2_CASCADE",
                 "REACTOME_CASPASE_ACTIVATION_VIA_DEATH_RECEPTORS_IN_THE_PRESENCE_OF_LIGAND",
                 "REACTOME_CASPASE_ACTIVATION_VIA_EXTRINSIC_APOPTOTIC_SIGNALLING_PATHWAY",
                 "REACTOME_DECTIN_1_MEDIATED_NONCANONICAL_NF_KB_SIGNALING",
                 "REACTOME_REGULATION_OF_RUNX3_EXPRESSION_AND_ACTIVITY",
                 "REACTOME_NEUTROPHIL_DEGRANULATION","REACTOME_TCR_SIGNALING",
                 "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR","REACTOME_PD_1_SIGNALING",
                 "REACTOME_TRANSLATION")

reactome.gsea.toplot.race <- rbind(
  subset(reactome.gsea.race, pathway %in% my.pathways) %>% mutate(race="Yes"),
  subset(reactome.gsea.noviral, pathway %in% my.pathways) %>% mutate(race="No")
)
reactome.gsea.toplot.race$sig <- as.factor(reactome.gsea.toplot.race$padj)

# Reformat pathway names
reactome.gsea.toplot.race$pathway <- gsub("^REACTOME_","",reactome.gsea.toplot.race$pathway)
reactome.gsea.toplot.race$pathway <- gsub("_"," ",reactome.gsea.toplot.race$pathway)
reactome.gsea.toplot.race$pathway <- stringr::str_to_sentence(reactome.gsea.toplot.race$pathway)

# Fix capitalization of pathway names
my.dict <- list("Tnfr2"="TNFR2", "Traf6"="TRAF6", "irf7"="IRF7",
                "runx3"="RUNX3", "nf kb"="NF-kB", "Dectin 1 mediated"="Dectin 1",
                "Pd 1"="PD1", "Tcr"="TCR",
                "bcr"="BCR", "b cell"="B cell",
                "Toll like receptor cascades"="TLR cascades",
                "Toll like receptor tlr1 tlr2 cascade"="TLR1 TLR2 cascade",
                "Caspase activation via death receptors in the presence of ligand" = "Caspase activation via depth receptors",
                "Caspase activation via extrinsic apoptotic signalling pathway" = "Caspase activation via apoptotic signalling")
for (i in names(my.dict)) {
  reactome.gsea.toplot.race$pathway <- gsub(i, my.dict[[i]], reactome.gsea.toplot.race$pathway)
}

p <- ggplot(data = reactome.gsea.toplot.race, aes(x=reorder(pathway,NES), y=NES)) +
  # geom_segment(aes(xend=reorder(pathway,NES)), yend=0) +
  geom_point(aes(color=as.factor(NES>0), 
                 shape=as.factor(padj<0.05)),
             size=2.5, stroke=1) +
  geom_hline(yintercept = 0, size = 0.7, color = "black") +
  scale_color_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  scale_shape_manual(values=c("FALSE"=1,"TRUE"=19), guide="none") +
  # scale_color_manual(values=c("FALSE"=alpha("black",0),"TRUE"=alpha("black",1)), guide="none") +
  # ylim(-3.9, 3.9) +
  scale_y_continuous(limits = c(-4,4), expand = c(0,0)) +
  labs(x="", y="GSEA normalized\nenrichment score",
       title="Pathways upregulated\nwith age at Visit 1") +
  facet_wrap(~race) +
  coord_flip(ylim = c(-4,4)) + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    panel.grid.minor=element_blank()
    # legend.key = element_rect(fill = "gray90")
  )
```

## Controlling for viral load

```{r}
design <- model.matrix(~ admit_age + sc2_rpm + sex + respiratory_status,
                       data=clin.sample.event.individ.1)

vwts <- voom(counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

```{r}
res <- topTable(vfit, coef = "admit_age",
                sort.by = "none", number = Inf,
                adjust.method = "BH", p.value = 1)
head(res)
```

Save DE results

```{r}
# write.csv(
#   res,
#   "20230123_bulk_RNAseq_pbmc_visit1_DE_viralload.csv"
# )
```

Reactome GSEA 

```{r}
# Reactome GSEA
# Sort gene list in descending order
gene.ranks <- res$t
names(gene.ranks) <- rownames(res)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)

# # How many pathways are significant?
# print(sum(reactome.gsea$padj < 0.05))
# # What are they?
# reactome.gsea.sig <- subset(reactome.gsea, padj < 0.05)
```

Save GSEA results

```{r}
data.table::fwrite(
  reactome.gsea.noviral,
  file="20230123_bulk_RNAseq_pbmc_visit1_admit-age_sex-ordinalscore-viralload_reactome-gsea_fdr0.05.tsv",
  sep="\t", sep2=c("", " ", "")
)
```

```{r}
# Plot significant hallmark pathways
ggplot(data = hallmark.gsea, aes(x=reorder(pathway,NES), y=NES)) +
  geom_segment(aes(xend=reorder(pathway,NES)), yend=0) +
  geom_point(aes(fill=as.factor(NES>0), color=as.factor(padj<0.05)),
             pch=21, size=2, stroke=1) +
  scale_color_manual(values=c("white","black")) +
  labs(x="Hallmark pathways", y="NES", title="Pathways upregulated with age") +
  coord_flip() + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    legend.key = element_rect(fill = "gray90")
  )
```

Plot selected reactome pathways, with and without controlling for viral load. Also include translation shutoff (https://doi.org/10.1038/s41586-021-03610-3)

```{r}
# Get list of pathways that are significant in at least one analysis
sig.pathways <- union(
  reactome.gsea.noviral[reactome.gsea.noviral$padj<0.05,"pathway"],
  reactome.gsea[reactome.gsea$padj<0.05,"pathway"]
)
```

```{r}
# Manually choose pathways
my.pathways <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING","REACTOME_INTERFERON_GAMMA_SIGNALING",
                 "REACTOME_INTERLEUKIN_12_SIGNALING",
                 "REACTOME_INTERLEUKIN_2_FAMILY_SIGNALING","REACTOME_TRAF6_MEDIATED_IRF7_ACTIVATION",
                 "REACTOME_TNFR2_NON_CANONICAL_NF_KB_PATHWAY","REACTOME_TOLL_LIKE_RECEPTOR_CASCADES",
                 "REACTOME_TOLL_LIKE_RECEPTOR_TLR1_TLR2_CASCADE",
                 "REACTOME_CASPASE_ACTIVATION_VIA_DEATH_RECEPTORS_IN_THE_PRESENCE_OF_LIGAND",
                 "REACTOME_CASPASE_ACTIVATION_VIA_EXTRINSIC_APOPTOTIC_SIGNALLING_PATHWAY",
                 "REACTOME_DECTIN_1_MEDIATED_NONCANONICAL_NF_KB_SIGNALING",
                 "REACTOME_REGULATION_OF_RUNX3_EXPRESSION_AND_ACTIVITY",
                 "REACTOME_NEUTROPHIL_DEGRANULATION","REACTOME_TCR_SIGNALING",
                 "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR","REACTOME_PD_1_SIGNALING",
                 "REACTOME_TRANSLATION")

reactome.gsea.toplot <- rbind(
  subset(reactome.gsea, pathway %in% my.pathways) %>% mutate(viral="Yes"),
  subset(reactome.gsea.noviral, pathway %in% my.pathways) %>% mutate(viral="No")
)
reactome.gsea.toplot$sig <- as.factor(reactome.gsea.toplot$padj)

# Reformat pathway names
reactome.gsea.toplot$pathway <- gsub("^REACTOME_","",reactome.gsea.toplot$pathway)
reactome.gsea.toplot$pathway <- gsub("_"," ",reactome.gsea.toplot$pathway)
reactome.gsea.toplot$pathway <- stringr::str_to_sentence(reactome.gsea.toplot$pathway)

# Fix capitalization of pathway names
my.dict <- list("Tnfr2"="TNFR2", "Traf6"="TRAF6", "irf7"="IRF7",
                "runx3"="RUNX3", "nf kb"="NF-kB", "Dectin 1 mediated"="Dectin 1",
                "Pd 1"="PD1", "Tcr"="TCR",
                "bcr"="BCR", "b cell"="B cell",
                "Toll like receptor cascades"="TLR cascades",
                "Toll like receptor tlr1 tlr2 cascade"="TLR1 TLR2 cascade",
                "Caspase activation via death receptors in the presence of ligand" = "Caspase activation via depth receptors",
                "Caspase activation via extrinsic apoptotic signalling pathway" = "Caspase activation via apoptotic signalling")
for (i in names(my.dict)) {
  reactome.gsea.toplot$pathway <- gsub(i, my.dict[[i]], reactome.gsea.toplot$pathway)
}

# Figure 4B
p <- ggplot(data = reactome.gsea.toplot, aes(x=reorder(pathway,NES), y=NES)) +
  # geom_segment(aes(xend=reorder(pathway,NES)), yend=0) +
  geom_point(aes(color=as.factor(NES>0), 
                 shape=as.factor(padj<0.05)),
             size=2.5, stroke=1) +
  geom_hline(yintercept = 0, size = 0.7, color = "black") +
  scale_color_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  scale_shape_manual(values=c("FALSE"=1,"TRUE"=19), guide="none") +
  # scale_color_manual(values=c("FALSE"=alpha("black",0),"TRUE"=alpha("black",1)), guide="none") +
  # ylim(-3.9, 3.9) +
  scale_y_continuous(limits = c(-4,4), expand = c(0,0)) +
  labs(x="", y="GSEA normalized\nenrichment score",
       title="Pathways upregulated\nwith age at Visit 1") +
  facet_wrap(~viral) +
  coord_flip(ylim = c(-4,4)) + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    panel.grid.minor=element_blank()
    # legend.key = element_rect(fill = "gray90")
  ) # RNAseq_pbmc_visit1_age_gsea-reactome_yes-vs-no-viralload, 6.1 x 6 in
```

Save DE and GSEA results

```{r}
# write.csv(
#   res,
#   "20220914_bulk_RNAseq_pbmc_admit-age_limma-voom.csv"
# )
# 
# data.table::fwrite(
#   reactome.gsea.sig,
#   file="20220914_bulk_RNAseq_pbmc_visit1_admit-age_viralload-ordinalscore_reactome-gsea.tsv",
#   sep="\t", sep2=c("", " ", "")
# )
# 
# data.table::fwrite(
#   hallmark.gsea,
#   file="20220914_bulk_RNAseq_pbmc_admit-age_hallmark-gsea.tsv",
#   sep="\t", sep2=c("", " ", "")
# )
```

## Age and severity: stratified by age (combined design)

Here I combine both age groups and both severity levels into one analysis

```{r}
clin.sample.event.individ.1.yo <- clin.sample.event.individ.1 %>% 
  subset((admit_age <= 46) | (admit_age > 70))
clin.sample.event.individ.1.yo$age_group <- cut(
  clin.sample.event.individ.1.yo$admit_age,
  breaks = c(17,46,70,96)
) %>% droplevels()
# Rename levels: convert round bracket to square bracket
levels(clin.sample.event.individ.1.yo$age_group) <- c("Young","Old")

# Mild is 3 & 4, severe is 5 & 6
clin.sample.event.individ.1.yo$severity <- "Mild"
clin.sample.event.individ.1.yo[clin.sample.event.individ.1.yo$respiratory_status %in% c("5","6"), "severity"] <- "Severe"

# Combine age and severity into one factor
clin.sample.event.individ.1.yo$age.severity <- paste(
  clin.sample.event.individ.1.yo$age_group,
  clin.sample.event.individ.1.yo$severity,
  sep = "."
)

counts.1.yo <- counts.1[,clin.sample.event.individ.1.yo$sample_id]
```

### Not controlled for viral load

DE analysis

```{r}
design <- model.matrix(~ 0 + age.severity + sex,
                       data=clin.sample.event.individ.1.yo)
# Rename age.severity column names
colnames(design) <- gsub("age.severity", "", colnames(design))
print(colnames(design))

vwts <- voom(counts.1.yo, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts, design = design) # don't actually need to specify design here after voom

cont.matrix <- makeContrasts(
  SvsMinY = Young.Severe - Young.Mild,
  SvsMinO = Old.Severe - Old.Mild,
  OvsYinS = Old.Severe - Young.Severe,
  OvsYinM = Old.Mild - Young.Mild,
  levels = design)
vfit2 <- contrasts.fit(vfit, cont.matrix)
vfit2 <- eBayes(vfit2)

print(summary(decideTests(vfit2)))
```

Extract DE results

```{r}
res.SvsMinY <- topTable(vfit2, coef = "SvsMinY", 
                        sort.by = "none", number = Inf,
                        adjust.method = "BH", p.value = 1)
res.SvsMinO <- topTable(vfit2, coef = "SvsMinO", 
                        sort.by = "none", number = Inf,
                        adjust.method = "BH", p.value = 1)
```

Reactome GSEA

```{r}
# Young
gene.ranks <- res.SvsMinY$t
names(gene.ranks) <- rownames(res.SvsMinY)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.SvsMinY <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)

# Old
gene.ranks <- res.SvsMinO$t
names(gene.ranks) <- rownames(res.SvsMinO)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.SvsMinO <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)
```

Find Reactome pathways that are significant in either young or old, or both.

```{r}
merged.reactome.gsea <- rbind(
  reactome.gsea.SvsMinY %>% mutate(age_group="[18,46]"),
  reactome.gsea.SvsMinO %>% mutate(age_group="[71,96]")
)

# # Get list of pathways that are significant in at least one analysis
# my.pathways <- union(
#   reactome.gsea.SvsMinY[reactome.gsea.SvsMinY$padj<0.05,"pathway"],
#   reactome.gsea.SvsMinO[reactome.gsea.SvsMinO$padj<0.05,"pathway"]
# )

# merged.reactome.gsea <- merged.reactome.gsea %>%
#   subset(pathway %in% my.pathways$pathway)
```

Import merged GSEA results from nasal analysis

```{r}
merged.reactome.gsea.nasal <- data.table::fread(
  "20230125_bulk_RNAseq_nasal_visit1_age-group_gsea-reactome_severe-vs-mild_combined-design.tsv",
  sep="\t", sep2=c("", " ", "")
)
```

Annotate PBMC/NS.

```{r}
merged.merged <- rbind(
  merged.reactome.gsea %>% mutate(type="PBMC"),
  merged.reactome.gsea.nasal %>% mutate(type="Nasal")
)

merged.merged$type <- factor(merged.merged$type, levels=c("PBMC","Nasal"))
```

Pathways that are significant in at least one comparison

```{r}
sig.pathways <- unique(merged.merged[merged.merged$padj < 0.05,"pathway"])
```

Select some pathways to plot.

```{r}
my.pathways <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING","REACTOME_INTERFERON_GAMMA_SIGNALING", 
                 "REACTOME_INTERLEUKIN_1_SIGNALING","REACTOME_INTERLEUKIN_2_FAMILY_SIGNALING",
                 "REACTOME_INTERLEUKIN_7_SIGNALING","REACTOME_INTERLEUKIN_10_SIGNALING",
                 "REACTOME_INTERLEUKIN_3_INTERLEUKIN_5_AND_GM_CSF_SIGNALING",
                 "REACTOME_INTERLEUKIN_4_AND_INTERLEUKIN_13_SIGNALING",
                 "REACTOME_NEUTROPHIL_DEGRANULATION","REACTOME_TOLL_LIKE_RECEPTOR_TLR1_TLR2_CASCADE",
                 "REACTOME_MYD88_INDEPENDENT_TLR4_CASCADE","REACTOME_TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE",
                 "REACTOME_COSTIMULATION_BY_THE_CD28_FAMILY",
                 "REACTOME_THE_NLRP3_INFLAMMASOME",
                 "REACTOME_PD_1_SIGNALING","REACTOME_TCR_SIGNALING"
                 )

merged.merged.toplot <- merged.merged %>%
  subset(pathway %in% my.pathways)

# Reformat pathway names
merged.merged.toplot$pathway <- gsub("^REACTOME_","", merged.merged.toplot$pathway)
merged.merged.toplot$pathway <- gsub("_"," ", merged.merged.toplot$pathway)
merged.merged.toplot$pathway <- stringr::str_to_sentence(merged.merged.toplot$pathway)

# Fix capitalization of pathway names
my.dict <- list("tlr"="TLR", "Myd88"="MyD88",
                "Toll like receptor 9 "="", "Toll like receptor "="",
                "cd28"="CD28",
                "Tcr"="TCR", "Pd 1"="PD1", "gm csf"="GM-CSF",
                "[I|i]nterleukin "="IL", "nlrp3"="NLRP3",
                "IL3 IL5"="IL3, IL5")
for (i in names(my.dict)) {
  merged.merged.toplot$pathway <- gsub(i, my.dict[[i]], 
                                       merged.merged.toplot$pathway)
}

# Dotplot
p <- ggplot(data = merged.merged.toplot,
            aes(x = age_group, 
                y = reorder(pathway, NES),
                color = NES)) +
  geom_point(color = "white", size = 3) +
  geom_point(shape = 1, size = 3, stroke = 1.5) +
  geom_point(data = . %>% subset(padj < 0.05), size = 4) +
  scale_color_gradientn(
    colors = c("limegreen","#f7f7f7","coral2"),
    limits = c(-4,4),
    breaks = c(-4,0,4),
    name = "") +
  # scale_color_distiller(
  #   palette = "RdYlGn",
  #   limits = c(-max(abs(res.toplot$slope)),max(abs(res.toplot$slope))),
  #   breaks = c(-1,0,1), name = "") +
  labs(
    x = "Age group", y = "",
    title = "Pathways associated\nwith severity at Visit 1"
  ) +
  facet_wrap(~ type, nrow=1) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black", angle=45, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    legend.key = element_rect(fill = "gray90"),
    panel.grid.minor=element_blank()
  ) # RNAseq_pbmc-nasal_visit1_severity_gsea-reactome_young-vs-old_dotplot, 6.1 x 5.7 in
```

Plot PBMC and NS separately

```{r}
# Dotplot
p1 <- ggplot(data = subset(merged.merged.toplot, type=="PBMC"),
            aes(x = age_group, 
                y = pathway,
                color = NES)) +
  geom_point(color = "white", size = 3) +
  geom_point(shape = 1, size = 3, stroke = 1.5) +
  geom_point(data = . %>% subset(padj < 0.05), size = 4) +
  scale_color_gradientn(
    colors = c("limegreen","#f7f7f7","coral2"),
    limits = c(-4,4),
    breaks = c(-4,0,4),
    name = "") +
  labs(
    x = "Age group", y = "",
    title = "Pathways upregulated\nin severe patients (PBMS)"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black", angle=45, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    legend.key = element_rect(fill = "gray90"),
    panel.grid.minor=element_blank()
  )

p2 <- ggplot(data = subset(merged.merged.toplot, type=="NS"),
            aes(x = age_group, 
                y = pathway,
                color = NES)) +
  geom_point(color = "white", size = 3) +
  geom_point(shape = 1, size = 3, stroke = 1.5) +
  geom_point(data = . %>% subset(padj < 0.05), size = 4) +
  scale_color_gradientn(
    colors = c("limegreen","#f7f7f7","coral2"),
    limits = c(-4,4),
    breaks = c(-4,0,4),
    name = "") +
  labs(
    x = "Age group", y = "",
    title = "Pathways upregulated\nin severe patients (NS)"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black", angle=45, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    legend.key = element_rect(fill = "gray90"),
    panel.grid.minor=element_blank()
  )

p1+p2 + plot_layout(nrow=1)
```

### Controlled for viral load

DE analysis

```{r}
design <- model.matrix(~ 0 + age.severity + sex + sc2_rpm,
                       data=clin.sample.event.individ.1.yo)
# Rename age.severity column names
colnames(design) <- gsub("age.severity", "", colnames(design))
print(colnames(design))

vwts <- voom(counts.1.yo, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts, design = design) # don't actually need to specify design here after voom

cont.matrix <- makeContrasts(
  SvsMinY = Young.Severe - Young.Mild,
  SvsMinO = Old.Severe - Old.Mild,
  OvsYinS = Old.Severe - Young.Severe,
  OvsYinM = Old.Mild - Young.Mild,
  levels = design)
vfit2 <- contrasts.fit(vfit, cont.matrix)
vfit2 <- eBayes(vfit2)

print(summary(decideTests(vfit2)))
```

Extract DE results

```{r}
res.SvsMinY <- topTable(vfit2, coef = "SvsMinY", 
                        sort.by = "none", number = Inf,
                        adjust.method = "BH", p.value = 1)
res.SvsMinO <- topTable(vfit2, coef = "SvsMinO", 
                        sort.by = "none", number = Inf,
                        adjust.method = "BH", p.value = 1)
```

Reactome GSEA

```{r}
# Young
gene.ranks <- res.SvsMinY$t
names(gene.ranks) <- rownames(res.SvsMinY)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.SvsMinY <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)

# Old
gene.ranks <- res.SvsMinO$t
names(gene.ranks) <- rownames(res.SvsMinO)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.SvsMinO <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)
```

Find Reactome pathways that are significant in either young or old, or both.

```{r}
merged.reactome.gsea <- rbind(
  reactome.gsea.SvsMinY %>% mutate(age_group="[18,46]"),
  reactome.gsea.SvsMinO %>% mutate(age_group="[71,96]")
)

# # Get list of pathways that are significant in at least one analysis
# my.pathways <- union(
#   reactome.gsea.SvsMinY[reactome.gsea.SvsMinY$padj<0.05,"pathway"],
#   reactome.gsea.SvsMinO[reactome.gsea.SvsMinO$padj<0.05,"pathway"]
# )

# merged.reactome.gsea <- merged.reactome.gsea %>%
#   subset(pathway %in% my.pathways$pathway)
```

Import merged GSEA results from nasal analysis

```{r}
merged.reactome.gsea.nasal <- data.table::fread(
  "20230125_bulk_RNAseq_nasal_visit1_age-group_gsea-reactome_severe-vs-mild_viralload_combined-design.tsv",
  sep="\t", sep2=c("", " ", "")
)
```

Annotate PBMC/NS.

```{r}
merged.merged <- rbind(
  merged.reactome.gsea %>% mutate(type="PBMC"),
  merged.reactome.gsea.nasal %>% mutate(type="Nasal")
)

merged.merged$type <- factor(merged.merged$type, levels=c("PBMC","Nasal"))
```

Pathways that are significant in at least one comparison

```{r}
sig.pathways <- unique(merged.merged[merged.merged$padj < 0.05,"pathway"])
```

Select some pathways to plot.

```{r}
my.pathways <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING","REACTOME_INTERFERON_GAMMA_SIGNALING", 
                 "REACTOME_INTERLEUKIN_1_SIGNALING","REACTOME_INTERLEUKIN_2_FAMILY_SIGNALING",
                 "REACTOME_INTERLEUKIN_7_SIGNALING","REACTOME_INTERLEUKIN_10_SIGNALING",
                 "REACTOME_INTERLEUKIN_3_INTERLEUKIN_5_AND_GM_CSF_SIGNALING",
                 "REACTOME_INTERLEUKIN_4_AND_INTERLEUKIN_13_SIGNALING",
                 "REACTOME_NEUTROPHIL_DEGRANULATION","REACTOME_TOLL_LIKE_RECEPTOR_TLR1_TLR2_CASCADE",
                 "REACTOME_MYD88_INDEPENDENT_TLR4_CASCADE","REACTOME_TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE",
                 "REACTOME_COSTIMULATION_BY_THE_CD28_FAMILY",
                 "REACTOME_THE_NLRP3_INFLAMMASOME",
                 "REACTOME_PD_1_SIGNALING","REACTOME_TCR_SIGNALING"
                 )

merged.merged.toplot <- merged.merged %>%
  subset(pathway %in% my.pathways)

# Reformat pathway names
merged.merged.toplot$pathway <- gsub("^REACTOME_","", merged.merged.toplot$pathway)
merged.merged.toplot$pathway <- gsub("_"," ", merged.merged.toplot$pathway)
merged.merged.toplot$pathway <- stringr::str_to_sentence(merged.merged.toplot$pathway)

# Fix capitalization of pathway names
my.dict <- list("tlr"="TLR", "Myd88"="MyD88",
                "Toll like receptor 9 "="", "Toll like receptor "="",
                "cd28"="CD28",
                "Tcr"="TCR", "Pd 1"="PD1", "gm csf"="GM-CSF",
                "[I|i]nterleukin "="IL", "nlrp3"="NLRP3",
                "IL3 IL5"="IL3, IL5")
for (i in names(my.dict)) {
  merged.merged.toplot$pathway <- gsub(i, my.dict[[i]], 
                                       merged.merged.toplot$pathway)
}

# Dotplot
p <- ggplot(data = merged.merged.toplot,
            aes(x = age_group, 
                y = reorder(pathway, NES),
                color = NES)) +
  geom_point(color = "white", size = 3) +
  geom_point(shape = 1, size = 3, stroke = 1.5) +
  geom_point(data = . %>% subset(padj < 0.05), size = 4) +
  scale_color_gradientn(
    colors = c("limegreen","#f7f7f7","coral2"),
    limits = c(-4,4),
    breaks = c(-4,0,4),
    name = "") +
  # scale_color_distiller(
  #   palette = "RdYlGn",
  #   limits = c(-max(abs(res.toplot$slope)),max(abs(res.toplot$slope))),
  #   breaks = c(-1,0,1), name = "") +
  labs(
    x = "Age group", y = "",
    title = "Pathways associateded\nwith severity at Visit 1\n(controlled for viral load)"
  ) +
  facet_wrap(~ type, nrow=1) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black", angle=45, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    legend.key = element_rect(fill = "gray90"),
    panel.grid.minor=element_blank()
  ) # RNAseq_pbmc-nasal_visit1_severity_viralload_gsea-reactome_young-vs-old_dotplot, 6.1 x 5.9 in
```

## Age and severity: stratified by severity (combined design)

Mild is 3 & 4, severe is 5 & 6.

```{r}
clin.sample.event.individ.1.sm <- clin.sample.event.individ.1
clin.sample.event.individ.1.sm$severity <- "Mild"
clin.sample.event.individ.1.sm[clin.sample.event.individ.1.sm$respiratory_status %in% c("5","6"), "severity"] <- "Severe"

counts.1.sm <- counts.1[,clin.sample.event.individ.1.sm$sample_id]
```

DE analysis on continuous age.

```{r}
# Design matrix
design <- model.matrix(~ severity + severity:admit_age + sex,
                       data=clin.sample.event.individ.1.sm)
# Rename
colnames(design) <- gsub("severity", "", colnames(design))
print(head(design))

vwts <- voom(counts.1.sm, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts, design = design)
vfit <- eBayes(vfit)

print(summary(decideTests(vfit)))

res.m <- topTable(vfit, coef = "Mild:admit_age",
                  sort.by = "none", number = Inf,
                  adjust.method = "BH", p.value = 1)
res.s <- topTable(vfit, coef = "Severe:admit_age",
                  sort.by = "none", number = Inf,
                  adjust.method = "BH", p.value = 1)
```

### Reactome GSEA

```{r}
# Severe
gene.ranks <- res.s$t
names(gene.ranks) <- rownames(res.s)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.s <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)

# Mild
gene.ranks <- res.m$t
names(gene.ranks) <- rownames(res.m)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.m <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)
```

Find Reactome pathways that are significant in either severe or mild, or both.

```{r}
merged.reactome.gsea <- rbind(
  reactome.gsea.s %>% mutate(severity="Severe"),
  reactome.gsea.m %>% mutate(severity="Mild")
)

# # Get list of pathways that are significant in at least one analysis
# my.pathways <- union(
#   reactome.gsea.s[reactome.gsea.s$padj<0.05,"pathway"],
#   reactome.gsea.m[reactome.gsea.m$padj<0.05,"pathway"]
# )

# merged.reactome.gsea <- merged.reactome.gsea %>%
#   subset(pathway %in% my.pathways$pathway)
```

Import merged GSEA results from nasal analysis

```{r}
merged.reactome.gsea.nasal <- data.table::fread(
  "20230123_bulk_RNAseq_nasal_visit1_severity_gsea-reactome_admit-age_combined-design.tsv",
  sep="\t", sep2=c("", " ", "")
)
```

Annotate PBMC/NS.

```{r}
merged.merged <- rbind(
  merged.reactome.gsea %>% mutate(type="PBMC"),
  merged.reactome.gsea.nasal %>% mutate(type="NS")
)

merged.merged$type <- factor(merged.merged$type, levels=c("PBMC","NS"))
```

Pathways that are significant in at least one comparison

```{r}
sig.pathways <- unique(merged.merged[merged.merged$padj < 0.05,"pathway"])
```

Select some pathways to plot.

```{r}
my.pathways <- c(
  "REACTOME_INTERFERON_ALPHA_BETA_SIGNALING","REACTOME_INTERFERON_GAMMA_SIGNALING",
  "REACTOME_INTERLEUKIN_1_SIGNALING","REACTOME_INTERLEUKIN_12_SIGNALING",
  "REACTOME_INTERLEUKIN_4_AND_INTERLEUKIN_13_SIGNALING",
  "REACTOME_INTERLEUKIN_2_FAMILY_SIGNALING","REACTOME_INTERLEUKIN_7_SIGNALING",
  "REACTOME_INTERLEUKIN_10_SIGNALING","REACTOME_INTERLEUKIN_17_SIGNALING",
  "REACTOME_NEUTROPHIL_DEGRANULATION","REACTOME_TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE",
  "REACTOME_TOLL_LIKE_RECEPTOR_TLR1_TLR2_CASCADE","REACTOME_PD_1_SIGNALING",
  "REACTOME_TCR_SIGNALING","REACTOME_DECTIN_1_MEDIATED_NONCANONICAL_NF_KB_SIGNALING"
  )

merged.merged.toplot <- merged.merged %>%
  subset(pathway %in% my.pathways)

# Reformat pathway names
merged.merged.toplot$pathway <- gsub("^REACTOME_","", merged.merged.toplot$pathway)
merged.merged.toplot$pathway <- gsub("_"," ", merged.merged.toplot$pathway)
merged.merged.toplot$pathway <- stringr::str_to_sentence(merged.merged.toplot$pathway)

# Fix capitalization of pathway names
my.dict <- list("tlr"="TLR",
                "Toll like receptor 9 "="", "Toll like receptor "="",
                "nf kb"="NF-kB", "Dectin 1 mediated"="Dectin 1",
                "Tcr"="TCR", "Pd 1"="PD1",
                "[I|i]nterleukin "="IL")
for (i in names(my.dict)) {
  merged.merged.toplot$pathway <- gsub(i, my.dict[[i]],
                                       merged.merged.toplot$pathway)
}

# Dotplot
p <- ggplot(data = merged.merged.toplot,
            aes(x = severity, 
                y = reorder(pathway, NES),
                color = NES)) +
  geom_point(color = "white", size = 3) +
  geom_point(shape = 1, size = 3, stroke = 1.5) +
  geom_point(data = . %>% subset(padj < 0.05), size = 4) +
  scale_color_gradientn(
    colors = c("dodgerblue4","#f7f7f7","darkred"),
    limits = c(-4,4),
    breaks = c(-4,0,4),
    name = "") +
  labs(
    x = "Disease severity", y = "",
    title = "Pathways upregulated\nwith age"
  ) +
  facet_wrap(~ type, nrow=1) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black", angle=45, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    legend.key = element_rect(fill = "gray90"),
    panel.grid.minor=element_blank()
  ) # RNAseq_pbmc-nasal_visit1_admit-age_gsea-reactome_severe-vs-mild_dotplot, 6.1 x 5.5 in
```

## Age and ordinal score: stratified by age

```{r}
clin.sample.event.individ.1.y <- clin.sample.event.individ.1 %>% 
  subset((admit_age <= 46) & (respiratory_status=="3" | respiratory_status=="6")) %>%
  droplevels()
clin.sample.event.individ.1.o <- clin.sample.event.individ.1 %>% 
  subset((admit_age > 70) & (respiratory_status=="3" | respiratory_status=="6")) %>%
  droplevels()

counts.1.y <- counts.1[,clin.sample.event.individ.1.y$sample_id]
counts.1.o <- counts.1[,clin.sample.event.individ.1.o$sample_id]
```

Mild is 3, severe is 6.

```{r}
print(table(clin.sample.event.individ.1.y$respiratory_status))
print(table(clin.sample.event.individ.1.o$respiratory_status))
```

DE analysis on respiratory_status.

```{r}
# Young
design <- model.matrix(~ sc2_rpm + sex + respiratory_status,
                       data=clin.sample.event.individ.1.y)
vwts <- voom(counts.1.y, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

res.y <- topTable(vfit, coef = "respiratory_status6",
                  sort.by = "none", number = Inf,
                  adjust.method = "BH", p.value = 1)

# Old
design <- model.matrix(~ sc2_rpm + sex + respiratory_status,
                       data=clin.sample.event.individ.1.o)
vwts <- voom(counts.1.o, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

res.o <- topTable(vfit, coef = "respiratory_status6",
                  sort.by = "none", number = Inf,
                  adjust.method = "BH", p.value = 1)
```

### Hallmark GSEA

```{r}
# Young
gene.ranks <- res.y$t
names(gene.ranks) <- rownames(res.y)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

hallmark.gsea.y <- fgseaMultilevel(
  pathways = hallmark.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)

# Old
gene.ranks <- res.o$t
names(gene.ranks) <- rownames(res.o)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

hallmark.gsea.o <- fgseaMultilevel(
  pathways = hallmark.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)
```

Plot both y and o GSEA results.

```{r}
merged.hallmark.gsea <- rbind(
  hallmark.gsea.y %>% mutate(status="<= 46 yo"), 
  hallmark.gsea.o %>% mutate(status="> 70 yo")
)

# Get list of pathways that are significant in at least one analysis
my.pathways <- union(
  hallmark.gsea.y[hallmark.gsea.y$padj<0.05,"pathway"],
  hallmark.gsea.o[hallmark.gsea.o$padj<0.05,"pathway"]
)

merged.hallmark.gsea <- merged.hallmark.gsea %>%
  subset(pathway %in% my.pathways$pathway)

ggplot(data=merged.hallmark.gsea) +
  geom_point(mapping=aes(x=reorder(pathway,NES), y=NES, fill=as.factor(NES>0),
                         colour=as.factor(padj < 0.05)),
             pch=21, size=2, stroke=1) +
  geom_hline(yintercept=0, color="black") +
  scale_color_manual(values=c("white","black")) +
  scale_y_continuous(breaks=seq(-4,4,2), limits=c(-4,4)) +
  labs(x="Hallmark pathways", y="NES", title="Pathways upregulated in severe covid") +
  coord_flip() + facet_wrap(~ status) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(1, "lines"),
    legend.key = element_rect(fill = "gray90")
  )
```

### Reactome GSEA

```{r}
# Young
gene.ranks <- res.y$t
names(gene.ranks) <- rownames(res.y)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

reactome.gsea.y <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)

# Old
gene.ranks <- res.o$t
names(gene.ranks) <- rownames(res.o)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

reactome.gsea.o <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)
```

Plot Reactome pathways of both mild and severe.

```{r}
merged.reactome.gsea <- rbind(
  reactome.gsea.y %>% mutate(status="<= 46 yo"), 
  reactome.gsea.o %>% mutate(status="> 70 yo")
)

# Get list of pathways that are significant in at least one analysis
my.pathways <- union(
  reactome.gsea.y[reactome.gsea.y$padj<0.05,"pathway"],
  reactome.gsea.o[reactome.gsea.o$padj<0.05,"pathway"]
)

merged.reactome.gsea <- merged.reactome.gsea %>%
  subset(pathway %in% my.pathways$pathway)

my.pathways <- c("REACTOME_ANTIVIRAL_MECHANISM_BY_IFN_STIMULATED_GENES", "REACTOME_INTERFERON_ALPHA_BETA_SIGNALING",
                 "REACTOME_INTERFERON_GAMMA_SIGNALING", "REACTOME_INTERLEUKIN_2_SIGNALING",
                 "REACTOME_NEUTROPHIL_DEGRANULATION", "REACTOME_TCR_SIGNALING",
                 "REACTOME_TNFR2_NON_CANONICAL_NF_KB_PATHWAY")

merged.reactome.gsea.toplot <- merged.reactome.gsea %>%
  subset(pathway %in% my.pathways)

ggplot(data=merged.reactome.gsea.toplot) +
  geom_point(mapping=aes(x=reorder(pathway,NES), y=NES, fill=as.factor(NES>0),
                         colour=as.factor(padj < 0.05)),
             pch=21, size=2, stroke=1) +
  geom_hline(yintercept=0, color="black") +
  scale_color_manual(values=c("white","black")) +
  scale_y_continuous(breaks=seq(-4,4,2), limits=c(-4,4)) +
  labs(x="",
       y="GSEA normalized\nenrichment score",
       title="Pathways upregulated in severe COVID") +
  coord_flip() + facet_wrap(~ status) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(1, "lines"),
    legend.key = element_rect(fill = "gray90")
  )
```

## Age and ordinal score: stratified by severity

```{r}
print(table(clin.sample.event.individ.1$respiratory_status))
```

Mild is 3, severe is 6.

```{r}
clin.sample.event.individ.1.mild <- clin.sample.event.individ.1 %>% 
  subset(respiratory_status=="3")
clin.sample.event.individ.1.severe <- clin.sample.event.individ.1 %>% 
  subset(respiratory_status=="6")

counts.1.mild <- counts.1[,clin.sample.event.individ.1.mild$sample_id]
counts.1.severe <- counts.1[,clin.sample.event.individ.1.severe$sample_id]
```

DE analysis on age effects.

```{r}
# Mild
design <- model.matrix(~ admit_age + sc2_rpm + sex,
                       data=clin.sample.event.individ.1.mild)
vwts <- voom(counts.1.mild, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

res.mild <- topTable(vfit, coef = "admit_age",
                     sort.by = "none", number = Inf,
                     adjust.method = "BH", p.value = 1)

# Severe
design <- model.matrix(~ admit_age + sc2_rpm + sex,
                       data=clin.sample.event.individ.1.severe)
vwts <- voom(counts.1.severe, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts)
vfit <- eBayes(vfit)

res.severe <- topTable(vfit, coef = "admit_age",
                       sort.by = "none", number = Inf,
                       adjust.method = "BH", p.value = 1)
```

### Hallmark GSEA

```{r}
# Mild
gene.ranks <- res.mild$t
names(gene.ranks) <- rownames(res.mild)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
hallmark.gsea.mild <- fgseaMultilevel(
  pathways = hallmark.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)

# Severe
gene.ranks <- res.severe$t
names(gene.ranks) <- rownames(res.severe)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
hallmark.gsea.severe <- fgseaMultilevel(
  pathways = hallmark.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)
```

Plot both mild and severe GSEA results.

```{r}
merged.hallmark.gsea <- rbind(
  hallmark.gsea.mild %>% mutate(status="Mild"), 
  hallmark.gsea.severe %>% mutate(status="Severe")
)

# Get list of pathways that are significant in at least one analysis
my.pathways <- union(
  hallmark.gsea.mild[hallmark.gsea.mild$padj<0.05,"pathway"],
  hallmark.gsea.severe[hallmark.gsea.severe$padj<0.05,"pathway"]
)

merged.hallmark.gsea <- merged.hallmark.gsea %>%
  subset(pathway %in% my.pathways$pathway)

ggplot(data=merged.hallmark.gsea) +
  geom_point(mapping=aes(x=reorder(pathway,NES), y=NES, fill=as.factor(NES>0),
                         colour=as.factor(padj < 0.05)),
             pch=21, size=2, stroke=1) +
  geom_hline(yintercept=0, color="black") +
  scale_color_manual(values=c("white","black")) +
  scale_y_continuous(breaks=seq(-4,4,2), limits=c(-4,4)) +
  labs(x="Hallmark pathways", y="NES", title="Pathways upregulated with age") +
  coord_flip() + facet_wrap(~ status) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(1, "lines"),
    legend.key = element_rect(fill = "gray90")
  )
```

### Reactome GSEA

```{r}
# Mild
gene.ranks <- res.mild$t
names(gene.ranks) <- rownames(res.mild)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.mild <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)

# Severe
gene.ranks <- res.severe$t
names(gene.ranks) <- rownames(res.severe)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.severe <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
remove(gene.ranks)
```

Plot Reactome pathways of both mild and severe.

```{r}
merged.reactome.gsea <- rbind(
  reactome.gsea.mild %>% mutate(status="Mild"), 
  reactome.gsea.severe %>% mutate(status="Severe")
)

# Get list of pathways that are significant in at least one analysis
my.pathways <- union(
  reactome.gsea.mild[reactome.gsea.mild$padj<0.05,"pathway"],
  reactome.gsea.severe[reactome.gsea.severe$padj<0.05,"pathway"]
)

merged.reactome.gsea <- merged.reactome.gsea %>%
  subset(pathway %in% my.pathways$pathway)

my.pathways <- c("REACTOME_ANTIVIRAL_MECHANISM_BY_IFN_STIMULATED_GENES", "REACTOME_DEATH_RECEPTOR_SIGNALLING",
                 "REACTOME_GENE_AND_PROTEIN_EXPRESSION_BY_JAK_STAT_SIGNALING_AFTER_INTERLEUKIN_12_STIMULATION",
                 "REACTOME_INTERFERON_GAMMA_SIGNALING", "REACTOME_INTERLEUKIN_12_SIGNALING",
                 "REACTOME_INTERLEUKIN_2_FAMILY_SIGNALING", "REACTOME_INTERLEUKIN_3_INTERLEUKIN_5_AND_GM_CSF_SIGNALING",
                 "REACTOME_NEUTROPHIL_DEGRANULATION", "REACTOME_TNF_SIGNALING", "REACTOME_TOLL_LIKE_RECEPTOR_CASCADES",
                 "REACTOME_ANTIMICROBIAL_PEPTIDES", "REACTOME_CASPASE_ACTIVATION_VIA_DEATH_RECEPTORS_IN_THE_PRESENCE_OF_LIGAND")

merged.reactome.gsea.toplot <- merged.reactome.gsea %>%
  subset(pathway %in% my.pathways)

ggplot(data=merged.reactome.gsea.toplot) +
  geom_point(mapping=aes(x=reorder(pathway,NES), y=NES, fill=as.factor(NES>0),
                         colour=as.factor(padj < 0.05)),
             pch=21, size=2, stroke=1) +
  geom_hline(yintercept=0, color="black") +
  scale_color_manual(values=c("white","black")) +
  scale_y_continuous(breaks=seq(-4,4,2), limits=c(-4,4)) +
  labs(x="Reactome pathways", y="NES", title="Pathways upregulated with age") +
  coord_flip() + facet_wrap(~ status) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(1, "lines"),
    legend.key = element_rect(fill = "gray90")
  )
```

Save GSEA results.

```{r}
# data.table::fwrite(
#   merged.reactome.gsea,
#   file="20230123_bulk_RNAseq_pbmc_visit1_admit-age_sex-viralload_mildvssevere_reactome-gsea_fdr0.05.tsv",
#   sep="\t", sep2=c("", " ", "")
# )
```

# Longitudinal analysis

```{r}
# Add viral load to clin.sample.event.individ
clin.sample.event.individ.long <- merge(
  clin.sample.event.individ, clin.sample.event.individ.sc2[,c("event_id","sc2_rpm")],
  by = "event_id",
  all = FALSE, sort = FALSE
)
# Log-transform viral load
clin.sample.event.individ.long$sc2_rpm <- log10(clin.sample.event.individ.long$sc2_rpm+1)

# Only keep hospital samples, and remove escalation samples
clin.sample.event.individ.long <- clin.sample.event.individ.long %>%
  subset(!grepl("Escalation", event_type) & event_date<=30)

# Divide into age groups
clin.sample.event.individ.long$age_group <- cut(
  clin.sample.event.individ.long$admit_age,
  breaks=c(17,46,54,62,70,96)
  )
# Rename levels: convert round bracket to square bracket
levels(clin.sample.event.individ.long$age_group) <- c("[18,46]","[47,54]","[55,62]","[63,70]","[71,96]")

# Subset samples
counts.long <- counts[,clin.sample.event.individ.long$sample_id]

# Filter genes
# Keep only protein_coding genes
counts.long <- counts.long[rData[rData$gene_biotype=="protein_coding", "gene_id"],]
# Remove low expression genes
counts.long <- counts.long[rowSums(counts.long>=10) >= 0.2*ncol(counts.long),]
```

Normalize gene counts with limma-voom.

```{r}
# design <- model.matrix(~age_group, clin.sample.event.individ.long)

vwts <- voom(counts.long, 
             # design = design,
             normalize.method = "quantile",
             plot = T)
```

## Without controlling for viral load

### All age groups: lmer

Linear mixed effects model

```{r}
res <- data.frame(matrix(NA, nrow=nrow(vwts), ncol=12))
colnames(res) <- c("gene_id",
                   "slope.2vs1","slope.3vs1","slope.4vs1","slope.5vs1",
                   "slope.1","slope.2","slope.3","slope.4","slope.5",
                   "PVal","adj.PVal")
rownames(res) <- rownames(vwts)
res$gene_id <- rownames(res)

print(sprintf("%s : Started longitudinal analysis", Sys.time()))
counter <- 1
for (i in rownames(res)) {
  temp <- data.frame(
    value = vwts$E[i,],
    event_date = clin.sample.event.individ.long$event_date,
    age_group = clin.sample.event.individ.long$age_group,
    sc2 = clin.sample.event.individ.long$sc2_rpm,
    sex = clin.sample.event.individ.long$sex,
    tg = clin.sample.event.individ.long$trajectory_group,
    site = clin.sample.event.individ.long$enrollment_site,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value)) %>%
    droplevels()
  
  fit1 <- lmer(
    value ~ event_date + event_date:age_group + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  fit2 <- lmer(
    value ~ event_date + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  
  # Get P-values for the interaction term across all age-groups
  aa <- anova(fit2, fit1, test = "LRT")
  res[i,"PVal"] <- aa[2,"Pr(>Chisq)"]
  
  # Get interaction term
  aa <- summary(fit1)
  res[i,"slope.2vs1"] <- coefficients(aa)["event_date:age_group[47,54]","Estimate"]
  res[i,"slope.3vs1"] <- coefficients(aa)["event_date:age_group[55,62]","Estimate"]
  res[i,"slope.4vs1"] <- coefficients(aa)["event_date:age_group[63,70]","Estimate"]
  res[i,"slope.5vs1"] <- coefficients(aa)["event_date:age_group[71,96]","Estimate"]
  
  # Calculate slope
  res[i,"slope.1"] <- coefficients(aa)["event_date","Estimate"]
  res[i,"slope.2"] <- res[i,"slope.1"] + res[i,"slope.2vs1"]
  res[i,"slope.3"] <- res[i,"slope.1"] + res[i,"slope.3vs1"]
  res[i,"slope.4"] <- res[i,"slope.1"] + res[i,"slope.4vs1"]
  res[i,"slope.5"] <- res[i,"slope.1"] + res[i,"slope.5vs1"]
  
  # Timer
  if ((counter %% 1000) == 0) {
    print(sprintf("%s : Processed %d features", Sys.time(), counter))
  }
  counter <- counter + 1
}
print(sprintf("%s : Finished longitudinal analysis", Sys.time()))

# Correct P-values
res$adj.PVal <- p.adjust(res$PVal, method = "BH")

# Add gene names
res <- merge(res, rData[,c("gene_id","gene_name")],
             by = "gene_id", all.x = TRUE, all.y = FALSE,
             sort = FALSE)
```

Save/load results

```{r}
# write.csv(res,
#           "20230123_bulk_RNAseq_pbmc_longitudinal-linear_agegroup-all5.csv")

res <- read.csv("20230123_bulk_RNAseq_pbmc_longitudinal-linear_agegroup-all5.csv",
                row.names = 1)
```

#### Plot all genes

```{r}
# Number of significant genes
print(sum(res$adj.PVal < 0.05))
```

Plot heatmap of significant genes' slopes.

```{r}
res.toplot <- res[, c("slope.1","slope.2","slope.3","slope.4","slope.5")]
colnames(res.toplot) <- levels(clin.sample.event.individ.long$age_group)
rownames(res.toplot) <- res$gene_id
res.toplot <- res.toplot[res$adj.PVal<0.05,]

# Figure 4C
# Heatmap
out <- pheatmap(
  res.toplot, # res.toplot[,c("[18,46]","[71,96]")],
  breaks = seq(-0.1,0.1, length.out = 255),
  legend_breaks = c(-0.1,0,0.1),
  legend_labels = c("\u2264 -0.1","0","\u2265 0.1"),
  # color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(255),
  color = rgb(colorRamp(c ("deepskyblue3", "#F9F4EC", "deeppink4"), space="rgb", interpolate="linear")(0:255/255), maxColorValue=255),
  cluster_cols = FALSE,
  clustering_distance_rows = "euclidean", clustering_method = "ward.D2",
  angle_col = 45,
  border_color = NA,
  show_rownames = FALSE,
  # cutree_rows = 3
) # RNAseq_pbmc_longitudinal_heatmap, 3.8 x 9.7 in
```

#### Plot example genes

Plot heatmap of 6 example HLA genes.

```{r}
res.mhc <- subset(res, adj.PVal<0.05 & grepl("^HLA-", gene_name))
rownames(res.mhc) <- res.mhc$gene_name
res.mhc <- res.mhc[,c("slope.1","slope.2","slope.3","slope.4","slope.5")]
colnames(res.mhc) <- levels(clin.sample.event.individ.long$age_group)

# Figure 4D
# Heatmap
out <- pheatmap(
  res.mhc[c("HLA-A","HLA-B","HLA-C","HLA-DRA","HLA-DRB1","HLA-DPA1"),],
  breaks = seq(-0.1,0.1, length.out = 101),
  legend_breaks = c(-0.1,0,0.1),
  legend_labels = c("\u2264 -0.1","0","\u2265 0.1"),
  color = colorRampPalette(c("deepskyblue3", "#F9F4EC", "deeppink4"),interpolate="linear")(101),
  # color = rgb(colorRamp(c("deepskyblue3", "#F9F4EC", "deeppink4"), space="rgb", interpolate="linear")(0:255/255), maxColorValue=255),
  cluster_cols = FALSE, cluster_rows = FALSE,
  # clustering_distance_rows = "euclidean", clustering_method = "ward.D2",
  angle_col = 45,
  border_color = "gray75",
  # main = "MHC genes"
  # show_rownames = FALSE,
  # cutree_rows = 2
) # RNAseq_pbmc_longitudinal_mhc_heatmap, 3.7 x 3 in
```

Try plotting all 5 trend lines on one plot

```{r}
p <- list()
for (i in c("HLA-A","HLA-DRA")) {
  # gene_id of this gene
  my.id <- rData[rData$gene_name==i,"gene_id"]
  
  temp <- data.frame(
    value = vwts$E[my.id,],
    event_date = clin.sample.event.individ.long$event_date,
    age_group = clin.sample.event.individ.long$age_group,
    sc2 = clin.sample.event.individ.long$sc2_rpm,
    sex = clin.sample.event.individ.long$sex,
    tg = clin.sample.event.individ.long$trajectory_group,
    site = clin.sample.event.individ.long$enrollment_site,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value)) %>%
    droplevels()
  
  # Predict
  fit <- lmer(
    value ~ event_date + event_date:age_group + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","age_group"),
                                  type="fixed")
  colnames(my.pred)[colnames(my.pred)=="group"] <- "age_group"
  
  temp$yhat <- predict(fit)

  
  p[[i]] <- ggplot() +
    # geom_line(aes(group=pid), color="gray", alpha=0.4) +
    # geom_line(inherit.aes=FALSE, aes(x=event_date, y=yhat, group=pid),
              # color="#5F5F5F", alpha=0.9) +
    # geom_point(aes(color=age_group), size=1, alpha=0.8) +
    geom_line(data=my.pred, aes(x,predicted,color=age_group),
              size=1.5) +
    labs(x="Days from admission", y="Normalized expression",
         title=sprintf("%s (P = %.3f)",i,res[res$gene_name==i,"adj.PVal"])) +
    scale_color_manual(values=c("#6699CC","mediumpurple1","#44AA99","#DDCC77","#CC6677")) +
    # scale_color_viridis_d(name = "Age group", guide = "none") +
    theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,0.5,0), "cm")
  )
}

p[[1]]+p[[2]] + plot_layout(ncol = 1) # RNAseq_pbmc_longitudinal_mhc_examples_combined, 4.3 x 6.2 in
```

Plot inflammatory genes.

```{r}
# Get inflammatory genes from hallmark gene set
infl.genes <- subset(hallmark,
                     gs_name=="HALLMARK_INFLAMMATORY_RESPONSE" &
                       ensembl_gene %in% rownames(vwts))

res.infl <- res %>%
  subset(adj.PVal<0.05 & gene_id %in% infl.genes$ensembl_gene)
rownames(res.infl) <- res.infl$gene_name
res.infl <- res.infl[,c("slope.1","slope.2","slope.3","slope.4","slope.5")]
colnames(res.infl) <- levels(clin.sample.event.individ.long$age_group)

# Figure 4D
# Heatmap
out <- pheatmap(
  res.infl[c("IL6","IL1R1","CXCL6","SLAMF1","IL10RA","CCR7","IL2RB","IL7R"),],
  breaks = seq(-0.1,0.1, length.out = 101),
  legend_breaks = c(-0.1,0,0.1),
  legend_labels = c("\u2264 -0.1","0","\u2265 0.1"),
  color = colorRampPalette(c("deepskyblue3", "#F9F4EC", "deeppink4"),interpolate="linear")(101),
  cluster_cols = FALSE, cluster_rows = FALSE,
  # clustering_distance_rows = "euclidean", clustering_method = "ward.D2",
  angle_col = 45,
  border_color = "gray75",
  # show_rownames = FALSE,
  # cutree_rows = 2
) # RNAseq_pbmc_longitudinal_infl_heatmap, 3.5 x 3 in
```

Plot genes in TCR signaling pathway.

```{r}
# Get TCR signaling genes from hallmark gene set
tcr.genes <- subset(reactome,
                    gs_name=="REACTOME_TCR_SIGNALING" &
                      ensembl_gene %in% rownames(vwts))

res.tcr <- res %>%
  subset(adj.PVal<0.05 & gene_id %in% tcr.genes$ensembl_gene)
rownames(res.tcr) <- res.tcr$gene_name
res.tcr <- res.tcr[,c("slope.1","slope.2","slope.3","slope.4","slope.5")]
colnames(res.tcr) <- levels(clin.sample.event.individ.long$age_group)

# Figure 4D
# Heatmap
out <- pheatmap(
  res.tcr[c("PTPRC","MALT1","ITK","CD4","CD3E","ZAP70","LCK","LAT"),], # LCP2 slopes change abruptly for [63,70], so replace it with MALT1
  breaks = seq(-0.1,0.1, length.out = 101),
  legend_breaks = c(-0.1,0,0.1),
  legend_labels = c("\u2264 -0.1","0","\u2265 0.1"),
  color = colorRampPalette(c("deepskyblue3", "#F9F4EC", "deeppink4"),interpolate="linear")(101),
  cluster_cols = FALSE, 
  cluster_rows = FALSE,
  # clustering_distance_rows = "euclidean", clustering_method = "ward.D2",
  angle_col = 45,
  border_color = "gray75",
  # show_rownames = FALSE,
  # cutree_rows = 3
) # RNAseq_pbmc_longitudinal_tcr_heatmap, 3.45 x 3 in
```

Plot example genes (combine all 5 trend lines into 1 plot)

```{r}
# Figure 4E
p <- list()
for (i in c("HLA-A","HLA-DRA","CXCL6","IL10RA","CD3E","LAT")) {
  # gene_id of this gene
  my.id <- rData[rData$gene_name==i,"gene_id"]
  
  temp <- data.frame(
    value = vwts$E[my.id,],
    event_date = clin.sample.event.individ.long$event_date,
    age_group = clin.sample.event.individ.long$age_group,
    sc2 = clin.sample.event.individ.long$sc2_rpm,
    sex = clin.sample.event.individ.long$sex,
    tg = clin.sample.event.individ.long$trajectory_group,
    site = clin.sample.event.individ.long$enrollment_site,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value)) %>%
    droplevels()
  
  # Predict
  fit <- lmer(
    value ~ event_date + event_date:age_group + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","age_group"),
                                  type="fixed")
  colnames(my.pred)[colnames(my.pred)=="group"] <- "age_group"
  
  temp$yhat <- predict(fit)

  
  p[[i]] <- ggplot() +
    # geom_line(aes(group=pid), color="gray", alpha=0.4) +
    # geom_line(inherit.aes=FALSE, aes(x=event_date, y=yhat, group=pid),
              # color="#5F5F5F", alpha=0.9) +
    # geom_point(aes(color=age_group), size=1, alpha=0.8) +
    geom_line(data=my.pred, aes(x,predicted,color=age_group),
              size=1.5) +
    labs(x="Days from admission", y="Normalized expression",
         title=sprintf("%s (P = %.3f)",i,res[res$gene_name==i,"adj.PVal"])) +
    # scale_color_manual(values=c("#6699CC","mediumpurple1","#44AA99","#DDCC77","#CC6677"),
    #                    name="Age group") +
    scale_color_viridis_d(name = "Age group") +
    theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,0.75,0), "cm")
  )
}

p[[1]]+
  p[[2]]+ylim(c(10,11.5)) +
  p[[3]]+p[[4]] + 
  p[[5]]+p[[6]] + plot_layout(nrow = 3) # RNAseq_pbmc_longitudinal_examples_combined, 9.2 x 10 in
```

Plot example genes (separate plot per age group)

```{r}
# Supplementary Figure 6
p <- list()
for (i in c("HLA-A","HLA-DRA","CXCL6","IL10RA","CD3E","LAT")) {
  # gene_id of this gene
  my.id <- rData[rData$gene_name==i,"gene_id"]
  
  temp <- data.frame(
    value = vwts$E[my.id,],
    event_date = clin.sample.event.individ.long$event_date,
    age_group = clin.sample.event.individ.long$age_group,
    sc2 = clin.sample.event.individ.long$sc2_rpm,
    sex = clin.sample.event.individ.long$sex,
    tg = clin.sample.event.individ.long$trajectory_group,
    site = clin.sample.event.individ.long$enrollment_site,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value)) %>%
    droplevels()
  
  # Predict
  fit <- lmer(
    value ~ event_date + event_date:age_group + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","age_group"),
                                  type="fixed")
  colnames(my.pred)[colnames(my.pred)=="group"] <- "age_group"
  
  temp$yhat <- predict(fit)

  # Calculate IQR and quartile values for limiting y-axes
  q1 <- quantile(temp$value, 0.25)
  q3 <- quantile(temp$value, 0.75)
  iqr <- IQR(temp$value)
  p[[i]] <- ggplot(data = temp, aes(x=event_date, y=value, color=age_group)) +
    geom_line(aes(group=pid), color="gray", alpha=0.4) +
    # geom_line(inherit.aes=FALSE, aes(x=event_date, y=yhat, group=pid),
              # color="#5F5F5F", alpha=0.9) +
    geom_point(aes(color=age_group), shape=20, size=0.5) +
    geom_line(inherit.aes=FALSE, data=my.pred, aes(x,predicted),
              color = "black", size=1) +
    geom_ribbon(inherit.aes=FALSE, data=my.pred, 
                aes(x = x, y = predicted, 
                    ymin = conf.low, ymax = conf.high), 
                fill = "black", alpha = 0.5) +
    labs(x="Days from admission", y="Normalized expression",
         title=sprintf("%s (P = %.3f)",i,res[res$gene_name==i,"adj.PVal"])) +
    # scale_color_manual(values=c("#6699CC","mediumpurple1","#44AA99","#DDCC77","#CC6677"),
    #                    name="Age group") +
    scale_color_viridis_d(name = "Age group", guide = "none") +
    coord_cartesian(ylim = c(q1-1.5*iqr, q3+1.5*iqr)) +
    facet_wrap(~ age_group, nrow=1) + 
    theme_bw() + 
    theme(
      plot.title = element_text(hjust = 0.5, size=12, face="plain"),
      axis.text = element_text(size=12, color="black"),
      text = element_text(size=12, family="Arial"),
      plot.margin = unit(c(0,0.85,0.7,0), "cm")
      )
}

p[[1]]+p[[2]]+
  p[[3]]+p[[4]] + 
  p[[5]]+p[[6]] + plot_layout(nrow = 3) # RNAseq_pbmc_longitudinal_examples, 13 x 8.5 in
```
