---
title: "serum_olink"
output: html_document
---

Source: 20231004_olink.Rmd

```{r}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(patchwork)

library(lme4)

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```

# Import data

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

### Clinical metadata processing

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE
clin.event$steroids <- as.factor(clin.event$steroids)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert some columns to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

## Olink concentration

```{r}
# Concentration
olink <- read.csv(
  "/data/serum-olink/legacy/2023-01-06/Olink-Counts.csv",
  header = TRUE, row.names = 1
)
olink <- t(olink) # transpose

# Row data
rData <- read.csv(
  "/data/serum-olink/legacy/2023-01-06/Olink-RowFeature.csv",
  header = TRUE, row.names = 3
)

# Olink metadata
olink.metadata <- read.csv(
  "/data/serum-olink/legacy/2023-01-06/Olink-Metadata.csv",
  header = TRUE, row.names = NULL
)

# Verify the sample_id and protein names match
stopifnot(rownames(olink) == rownames(rData))
stopifnot(colnames(olink) == olink.metadata$sample_id)
```

## Viral load data

Get SARS-CoV-2 viral load from nasal-metagenomics-AllGenera-Counts_RPM.csv, sc2_rpm column.

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/legacy/2022-09-03/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
```

# Merge metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

```{r}
# Only keep samples that pass QC
olink.metadata <- subset(olink.metadata, sample_status=="IMPACC sample assayed and passed QC")
pass.qc <- Reduce(
  intersect, 
  list(colnames(olink), olink.metadata$sample_id, clin.sample.event.individ$sample_id)
)

# mNGS QC
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))

olink <- olink[,pass.qc]
clin.sample.event.individ <- subset(clin.sample.event.individ, sample_id %in% pass.qc)
```

```{r}
# Set concentrations below LOD to 0
lod <- replicate(ncol(olink), rData$LOD)
olink[olink < lod] <- 0

# Only keep cytokines detected in at least 20% of the samples
olink <- olink[rowSums(olink>0, na.rm = TRUE) >= 0.2*ncol(olink),]
```

# Analysis: Visit 1

## Prepare Visit 1 data

```{r}
# Keep visit 1 samples for RNA-seq
clin.sample.event.individ.1 <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Keep visit 1 samples for viral load
clin.sample.event.individ.sc2.1 <- clin.sample.event.individ.sc2 %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Add viral load
clin.sample.event.individ.1 <- merge(
  clin.sample.event.individ.1, clin.sample.event.individ.sc2.1[,c("event_id","sc2_rpm")],
  by = "event_id", all = FALSE, sort = FALSE
)
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.1$sc2_rpm <- log10(clin.sample.event.individ.1$sc2_rpm + 1)

# Drop unused levels
clin.sample.event.individ.1 <- droplevels(clin.sample.event.individ.1)

# Any row with no sc2_rpm?
print(any(is.na(clin.sample.event.individ.1$sc2_rpm)))

# Any duplicated participant?
print(any(duplicated(clin.sample.event.individ.1$participant_id)))
```

```{r}
# Subset visit 1 samples
olink.1 <- olink[,clin.sample.event.individ.1$sample_id]

# Scale protein concentration
olink.1.2 <- t(scale(t(olink.1)))
```

## Continuous age

### Not control for viral load

```{r}
# Linear model
res.noviralload <- data.frame(matrix(NA, nrow=nrow(olink.1.2), ncol=4))
colnames(res.noviralload) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res.noviralload) <- rownames(olink.1.2)
res.noviralload$cytokine <- rownames(res.noviralload)

for (i in rownames(res.noviralload)) {
  temp <- data.frame(conc=olink.1.2[i,], 
                     age=clin.sample.event.individ.1$admit_age,
                     sex=clin.sample.event.individ.1$sex,
                     steroids=clin.sample.event.individ.1$steroids,
                     sc2=clin.sample.event.individ.1$sc2_rpm,
                     resp=clin.sample.event.individ.1$respiratory_status) %>%
    subset(!is.na(conc))
  my.lm <- lm(conc ~ age + sex + resp, data=temp)
  res.noviralload[i,"slope"] <- summary(my.lm)$coefficients["age",1]
  res.noviralload[i,"PVal"] <- summary(my.lm)$coefficients["age",4]
  
  remove(temp, my.lm)
}

# Correct P-values
res.noviralload$adj.PVal <- p.adjust(res.noviralload$PVal, method = "BH")
```

### Control for viral load

```{r}
# Linear model
res <- data.frame(matrix(NA, nrow=nrow(olink.1.2), ncol=4))
colnames(res) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res) <- rownames(olink.1.2)
res$cytokine <- rownames(res)

for (i in rownames(res)) {
  temp <- data.frame(conc=olink.1.2[i,], 
                     age=clin.sample.event.individ.1$admit_age,
                     sex=clin.sample.event.individ.1$sex,
                     steroids=clin.sample.event.individ.1$steroids,
                     sc2=clin.sample.event.individ.1$sc2_rpm,
                     resp=clin.sample.event.individ.1$respiratory_status) %>%
    subset(!is.na(conc))
  my.lm <- lm(conc ~ age + sc2 + sex + resp, data=temp)
  res[i,"slope"] <- summary(my.lm)$coefficients["age",1]
  res[i,"PVal"] <- summary(my.lm)$coefficients["age",4]
  
  remove(temp, my.lm)
}

# Correct P-values
res$adj.PVal <- p.adjust(res$PVal, method = "BH")
```

### Plotting

Plot slope of all significant cytokines, without controlling for viral load.

```{r}
# Figure 4A
p <- ggplot(data = subset(res.noviralload, adj.PVal < 0.05)) +
  geom_col(aes(y = reorder(cytokine,slope), x = slope,
               fill = as.factor(slope>0)),
           width=0.7) +
  geom_vline(xintercept=0, color="black", size=0.5) +
  scale_fill_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  # coord_flip() +
  labs(x = "Slope\n(change per 1 year of age)", y = "",
       title = "Proteins associated\nwith age at Visit 1") +
  scale_x_continuous(limits=c(-0.02, 0.02), expand=c(0,0)) +
  theme_bw() + theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=10, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank()
  ) # olink_visit1_age_no-viralload, 3.7 x 7.5 in (w x h)
```

Plot slope of all significant cytokines, controlling for viral load.

```{r}
# Figure S7A
p <- ggplot(data = subset(res, adj.PVal < 0.05)) +
  geom_col(aes(y = reorder(cytokine,slope), x = slope,
               fill = as.factor(slope>0)),
           width=0.7) +
  geom_vline(xintercept=0, color="black", size=0.5) +
  scale_fill_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  # coord_flip() +
  labs(x = "Slope\n(change per 1 year of age)", y = "",
       title = "Proteins associated\nwith age at Visit 1\n(controlled for viral abundance)") +
  scale_x_continuous(limits=c(-0.02, 0.02), expand=c(0,0)) +
  theme_bw() + theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank()
  ) # olink_visit1_age_viralload, 3.7 x 9.5 in (w x h)
```

Plot some example cytokines (not controlling for viral load).

```{r}
# Figure 4B
p <- list()
counter <- 1
my.colors <- c("darkred","dodgerblue4")
for (i in c("CXCL9","SIRT2")) {
  temp <- data.frame(conc=olink.1.2[i,],
                     age=clin.sample.event.individ.1$admit_age,
                     sex=clin.sample.event.individ.1$sex,
                     steroids=clin.sample.event.individ.1$steroids,
                     sc2=clin.sample.event.individ.1$sc2_rpm,
                     resp=clin.sample.event.individ.1$respiratory_status)
  my.lm <- lm(conc ~ age + sex + resp, data=temp)
  my.pred <- ggeffects::ggpredict(my.lm, "age")
  
  p[[i]] <- ggplot() +
    geom_point(data = temp, aes(age,conc), size = 0.5) +
    geom_line(data = my.pred, aes(x,predicted), size = 1, color = my.colors[counter]) +
    geom_ribbon(data = my.pred, 
                aes(x = x, y = predicted, ymin = conf.low, ymax = conf.high),
                fill = my.colors[counter], alpha = .4) +
    # annotate(geom = "text", label = sprintf("P = %.1e", res.noviralload[i,"adj.PVal"]),
    #          x = 0.6*min(temp$age), y = 1.3*min(temp$conc), hjust=0) +
    labs(x = "Age", y = "NPX (z-score)", 
         title = i) +
    theme_classic() + 
    theme(
      plot.title = element_text(hjust = 0.5, size=12, face="plain"),
      axis.text = element_text(size=12, family="Arial", color="black"),
      text = element_text(size=12, family="Arial", color="black"),
      plot.margin = unit(c(0,1,0.5,0), "cm")
    )
  
  counter <- counter + 1
}

p[[1]]+p[[2]] + plot_layout(ncol = 1) # olink_visit1_age_no-viralload_examples, 3.3 x 7 in
```

## Age and viral load interaction

### Stratified analysis

```{r}
res.y <- data.frame(matrix(NA, nrow=nrow(olink.1), ncol=4))
colnames(res.y) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res.y) <- rownames(olink.1)
res.y$cytokine <- rownames(res.y)
res.o <- data.frame(matrix(NA, nrow=nrow(olink.1), ncol=4))
colnames(res.o) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res.o) <- rownames(olink.1)
res.o$cytokine <- rownames(res.o)

for (i in rownames(olink.1.2)) {
  # Prepare data
  temp <- data.frame(value=scale(olink.1[i,]),
                     age=clin.sample.event.individ.1$admit_age,
                     sex=clin.sample.event.individ.1$sex,
                     steroids=clin.sample.event.individ.1$steroids,
                     sc2=clin.sample.event.individ.1$sc2_rpm,
                     resp=clin.sample.event.individ.1$respiratory_status) %>%
    subset(!is.na(value))
  
  # Young
  temp.y <- temp %>%
    subset(age<=46) %>%
    droplevels()
  # temp.y$value <- scale(temp.y$value)
  my.lm <- lm(value ~ sc2 + sex + resp, data=temp.y)
  aa <- summary(my.lm)
  res.y[i,"slope"] <- aa$coefficients["sc2","Estimate"]
  res.y[i,"PVal"] <- aa$coefficients["sc2","Pr(>|t|)"]
  
  remove(my.lm)
  
  # Old
  temp.o <- temp %>%
    subset(age > 70) %>%
    droplevels()
  # temp.o$value <- scale(temp.o$value)
  my.lm <- lm(value ~ sc2 + sex + resp, data=temp.o)
  aa <- summary(my.lm)
  res.o[i,"slope"] <- aa$coefficients["sc2","Estimate"]
  res.o[i,"PVal"] <- aa$coefficients["sc2","Pr(>|t|)"]
  
  remove(temp, temp.y, temp.o, my.lm)
}

# Correct P-values
res.y$adj.PVal <- p.adjust(res.y$PVal, method = "BH")
res.o$adj.PVal <- p.adjust(res.o$PVal, method = "BH")
```

Plot slope of old vs young.

```{r}
x <- res.y$slope
names(x) <- rownames(res.y)
y <- res.o$slope
names(y) <- rownames(res.o)

print(all(names(x) == names(y)))

temp <- data.frame(
  cytokine=names(x),
  y.slope=x, o.slope=y,
  y.sig=res.y$adj.PVal<0.05,
  o.sig=res.o$adj.PVal<0.05,
  sig = (res.y$adj.PVal<0.05) | (res.o$adj.PVal<0.05)
)
ggplot(data=temp, aes(x=y.slope, y=o.slope)) +
  geom_abline(slope=1, intercept=0,
              linetype="dashed", color="red", size=1) +
  geom_point(aes(color=sig), size=1.2) +
  geom_text(
    data=subset(temp, cytokine %in% c("CXCL10","CX3CL1","CCL23","SLAMF1","CCL8","TNFSF10")),
    aes(label=cytokine), hjust=0.5, vjust=0.1, size=3) +
  labs(x="Slope in age group [18,46]",
       y="Slope in age group [71,96]",
       title="Association with viral abundance\nat visit 1") + 
  my.theme + theme(aspect.ratio = 1)
# olink_visit1_viralload_age1vs5, 3.5 x 3.8 in

# Dot plot of sig cytokines
temp2 <- rbind(
  res.y[(res.y$adj.PVal<0.05) | (res.o$adj.PVal<0.05),c("cytokine","slope","adj.PVal")] %>% mutate(age_group="[18,46]"),
  res.o[(res.y$adj.PVal<0.05) | (res.o$adj.PVal<0.05),c("cytokine","slope","adj.PVal")] %>% mutate(age_group="[71,96]")
)

# Figure 4C
p <- ggplot(data=temp2,
       aes(x=age_group,
           y=reorder(cytokine, slope, FUN=function(x) x[1]),
           color=slope)) +
  geom_point(color="white", size=3) +
  geom_point(shape=1, size=3, stroke=1.5) +
  geom_point(data=. %>% subset(adj.PVal<0.05), size=5) +
  scale_color_distiller(palette = "BrBG",
                        limits=c(-0.2,0.2),
                        breaks=c(-0.2,0,0.2),
                        name="") +
  labs(
    x = "Age group", y = "Protein",
    title = "Proteins associated with\nviral abundance at Visit 1"
  ) +
  theme_bw() + theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black", angle=45, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank(),
    legend.direction = "horizontal",
    legend.position = "bottom"
  ) # olink_visit1_viralload_age1vs5_dotplot, 3 x 4.5 in
```

## Age and severity: stratified by age

```{r}
clin.sample.event.individ.1.y <- clin.sample.event.individ.1 %>% 
  subset((admit_age <= 46)) %>%
  droplevels()
clin.sample.event.individ.1.o <- clin.sample.event.individ.1 %>% 
  subset((admit_age > 70)) %>%
  droplevels()
```

Mild is 3 & 4, severe is 5 & 6.

```{r}
clin.sample.event.individ.1.y$severity <- "Severe"
clin.sample.event.individ.1.y[clin.sample.event.individ.1.y$respiratory_status %in% c("3","4"),"severity"] <- "Mild"

clin.sample.event.individ.1.o$severity <- "Severe"
clin.sample.event.individ.1.o[clin.sample.event.individ.1.o$respiratory_status %in% c("3","4"),"severity"] <- "Mild"

print(table(clin.sample.event.individ.1.y$severity))
print(table(clin.sample.event.individ.1.o$severity))
```

### Not control for viral load

Linear model

```{r}
# Prepare data frame for young patients
res.y <- data.frame(matrix(NA, nrow=nrow(olink.1.2), ncol=4))
colnames(res.y) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res.y) <- rownames(olink.1.2)
res.y$cytokine <- rownames(olink.1.2)

# Prepare data frame for old patients
res.o <- data.frame(matrix(NA, nrow=nrow(olink.1.2), ncol=4))
colnames(res.o) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res.o) <- rownames(olink.1.2)
res.o$cytokine <- rownames(olink.1.2)

for (i in rownames(res.y)) {
  # Young
  temp <- data.frame(value=olink.1.2[i,clin.sample.event.individ.1.y$sample_id],
                     severity=as.factor(clin.sample.event.individ.1.y$severity),
                     sex=clin.sample.event.individ.1.y$sex,
                     sc2=clin.sample.event.individ.1.y$sc2_rpm
                     ) %>%
    subset(!is.na(value))
  my.lm <- lm(value ~ severity + sex, data = temp)
  aa <- summary(my.lm)
  res.y[i,"slope"] <- coefficients(aa)["severitySevere",1]
  res.y[i,"PVal"] <- coefficients(aa)["severitySevere",4]
  
  # Old
  temp <- data.frame(value=olink.1.2[i,clin.sample.event.individ.1.o$sample_id],
                     severity=as.factor(clin.sample.event.individ.1.o$severity),
                     sex=clin.sample.event.individ.1.o$sex,
                     sc2=clin.sample.event.individ.1.o$sc2_rpm
                     ) %>%
    subset(!is.na(value))
  my.lm <- lm(value ~ severity + sex, data = temp)
  aa <- summary(my.lm)
  res.o[i,"slope"] <- coefficients(aa)["severitySevere",1]
  res.o[i,"PVal"] <- coefficients(aa)["severitySevere",4]
}

# BH adjust P value
res.y$adj.PVal <- p.adjust(res.y$PVal, method = "BH")
res.o$adj.PVal <- p.adjust(res.o$PVal, method = "BH")

# Show number of significant cytokines
print(sum(res.y$adj.PVal < 0.05))
print(sum(res.o$adj.PVal < 0.05))
```

Plot significant cytokines only

```{r}
# Which cytokines are significant in at least one age group
my.cytokines <- union(
  res.y[res.y$adj.PVal < 0.05,"cytokine"],
  res.o[res.o$adj.PVal < 0.05,"cytokine"]
)

res.toplot <- rbind(
  res.y[my.cytokines,] %>% mutate(age_group="[18,46]"),
  res.o[my.cytokines,] %>% mutate(age_group="[71,96]")
)


# Dotplot: Figure 6B
p2 <- ggplot(data=res.toplot,
             aes(y=age_group, 
                 x=reorder(cytokine, slope, FUN=function(x) x[1]),
                 color=slope)) +
  geom_point(color="white", size=3) +
  geom_point(shape=1, size=3, stroke=1.5) +
  geom_point(data=. %>% subset(adj.PVal < 0.05), size=4) +
  scale_y_discrete(limits=c("[71,96]","[18,46]")) +
  scale_color_gradientn(
    colors=c("limegreen","#f7f7f7","coral2"),
    limits=c(-1,1),
    breaks=c(-1,0,1),
    name="") +
  labs(
    y="Age group", x="",
    title="Proteins associated with severity at Visit 1"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=10, color="black", angle=45, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,2,1,0), "cm"),
    panel.grid.minor=element_blank(),
    legend.direction = "horizontal",
    legend.position = "bottom"
  ) # olink_visit1_severity_young-vs-old_dotplot, 11 x 3.4 in
```

### Control for viral load

```{r}
# Prepare data frame for young patients
res.y <- data.frame(matrix(NA, nrow=nrow(olink.1.2), ncol=4))
colnames(res.y) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res.y) <- rownames(olink.1.2)
res.y$cytokine <- rownames(olink.1.2)

# Prepare data frame for old patients
res.o <- data.frame(matrix(NA, nrow=nrow(olink.1.2), ncol=4))
colnames(res.o) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res.o) <- rownames(olink.1.2)
res.o$cytokine <- rownames(olink.1.2)

for (i in rownames(res.y)) {
  # Young
  temp <- data.frame(value=olink.1.2[i,clin.sample.event.individ.1.y$sample_id],
                     severity=as.factor(clin.sample.event.individ.1.y$severity),
                     sex=clin.sample.event.individ.1.y$sex,
                     sc2=clin.sample.event.individ.1.y$sc2_rpm
                     ) %>%
    subset(!is.na(value))
  my.lm <- lm(value ~ severity + sex + sc2, data = temp)
  aa <- summary(my.lm)
  res.y[i,"slope"] <- coefficients(aa)["severitySevere",1]
  res.y[i,"PVal"] <- coefficients(aa)["severitySevere",4]
  
  # Old
  temp <- data.frame(value=olink.1.2[i,clin.sample.event.individ.1.o$sample_id],
                     severity=as.factor(clin.sample.event.individ.1.o$severity),
                     sex=clin.sample.event.individ.1.o$sex,
                     sc2=clin.sample.event.individ.1.o$sc2_rpm
                     ) %>%
    subset(!is.na(value))
  my.lm <- lm(value ~ severity + sex + sc2, data = temp)
  aa <- summary(my.lm)
  res.o[i,"slope"] <- coefficients(aa)["severitySevere",1]
  res.o[i,"PVal"] <- coefficients(aa)["severitySevere",4]
}

# BH adjust P value
res.y$adj.PVal <- p.adjust(res.y$PVal, method = "BH")
res.o$adj.PVal <- p.adjust(res.o$PVal, method = "BH")

# Show number of significant cytokines
print(sum(res.y$adj.PVal < 0.05))
print(sum(res.o$adj.PVal < 0.05))
```

Plot significant cytokines only

```{r}
# Which cytokines are significant in at least one age group
my.cytokines <- union(
  res.y[res.y$adj.PVal < 0.05,"cytokine"],
  res.o[res.o$adj.PVal < 0.05,"cytokine"]
)

res.toplot <- rbind(
  res.y[my.cytokines,] %>% mutate(age_group="[18,46]"),
  res.o[my.cytokines,] %>% mutate(age_group="[71,96]")
)

# Figure S10B
# Dotplot
p2 <- ggplot(data=res.toplot,
             aes(x=age_group, 
                 y=reorder(cytokine, slope, FUN=function(x) x[1]),
                 color=slope)) +
  geom_point(color="white", size=3) +
  geom_point(shape=1, size=3, stroke=1.5) +
  geom_point(data=. %>% subset(adj.PVal < 0.05), size=4) +
  scale_color_gradientn(
    colors=c("limegreen","#f7f7f7","coral2"),
    limits=c(-1,1),
    breaks=c(-1,0,1),
    name="") +
  labs(
    x="Age group", y="",
    title="Proteins associated\nwith severity at Visit 1\n(controlled for viral load)"
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black", angle=45, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,2,1,0), "cm"),
    panel.grid.minor=element_blank(),
    legend.direction = "horizontal",
    legend.position = "bottom"
  ) # olink_visit1_severity_viralload_young-vs-old_dotplot, 3 x 11.3 in
```

# Longitudinal analysis

```{r}
# Add viral load to clin.sample.event.individ
clin.sample.event.individ.long <- merge(
  clin.sample.event.individ, clin.sample.event.individ.sc2[,c("event_id","sc2_rpm")],
  by = "event_id",
  all = FALSE, sort = FALSE
)
# Log-transform viral load
clin.sample.event.individ.long$sc2_rpm <- log10(clin.sample.event.individ.long$sc2_rpm+1)

# Remove escalation samples
clin.sample.event.individ.long <- clin.sample.event.individ.long %>%
  subset(!grepl("Escalation", event_type) & event_date<=30)

# Divide into age groups
clin.sample.event.individ.long$age_group <- cut(
  clin.sample.event.individ.long$admit_age,
  breaks=c(17,46,54,62,70,96)
  )

# Rename levels: convert round bracket to square bracket
levels(clin.sample.event.individ.long$age_group) <- c("[18,46]","[47,54]","[55,62]","[63,70]","[71,96]")

# Subset olink samples
olink.long <- olink[,clin.sample.event.individ.long$sample_id]
```

Look at event_date per visit.

```{r}
p1 <- ggplot(data=clin.sample.event.individ.long,
             aes(x=event_type, y=event_date, color=age_group)) +
  geom_boxplot()
```

## Without controlling for viral load

Linear mixed effects model

```{r}
res <- data.frame(matrix(NA, nrow=nrow(olink.long), ncol=11))
colnames(res) <- c("slope.2vs1","slope.3vs1","slope.4vs1","slope.5vs1",
                   "slope.1","slope.2","slope.3","slope.4","slope.5",
                   "PVal","adj.PVal")
rownames(res) <- rownames(olink.long)

for (i in rownames(res)) {
  temp <- data.frame(
    value = scale(olink.long[i,]),
    event_date = clin.sample.event.individ.long$event_date,
    age_group = clin.sample.event.individ.long$age_group,
    sc2 = clin.sample.event.individ.long$sc2_rpm,
    sex = clin.sample.event.individ.long$sex,
    tg = clin.sample.event.individ.long$trajectory_group,
    site = clin.sample.event.individ.long$enrollment_site,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value)) %>%
    droplevels()
  
  fit1 <- lmer(
    value ~ event_date + event_date:age_group + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  fit2 <- lmer(
    value ~ event_date + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  
  # Get P-values for the interaction term across all age-groups
  aa <- anova(fit2, fit1, test = "LRT")
  res[i,"PVal"] <- aa[2,"Pr(>Chisq)"]
  
  # Get interaction term
  aa <- summary(fit1)
  res[i,"slope.2vs1"] <- coefficients(aa)["event_date:age_group[47,54]","Estimate"]
  res[i,"slope.3vs1"] <- coefficients(aa)["event_date:age_group[55,62]","Estimate"]
  res[i,"slope.4vs1"] <- coefficients(aa)["event_date:age_group[63,70]","Estimate"]
  res[i,"slope.5vs1"] <- coefficients(aa)["event_date:age_group[71,96]","Estimate"]
  
  # Calculate slope
  res[i,"slope.1"] <- coefficients(aa)["event_date","Estimate"]
  res[i,"slope.2"] <- res[i,"slope.1"] + res[i,"slope.2vs1"]
  res[i,"slope.3"] <- res[i,"slope.1"] + res[i,"slope.3vs1"]
  res[i,"slope.4"] <- res[i,"slope.1"] + res[i,"slope.4vs1"]
  res[i,"slope.5"] <- res[i,"slope.1"] + res[i,"slope.5vs1"]
}

# Correct P-values
res$adj.PVal <- p.adjust(res$PVal, method = "BH")
```

```{r}
# Display significant cytokines
ggplot() +
  geom_col(data=arrange(res, desc(-adj.PVal)) %>% slice(1:25),
           aes(x=reorder(cytokine, -adj.PVal), 
               y=-log10(adj.PVal),
               fill=as.factor(adj.PVal<0.05))) +
  scale_fill_manual(values = c("FALSE"="gray","TRUE"="#F8766D")) +
  geom_hline(yintercept=-log10(0.05), linetype="dashed") +
  coord_flip() +
  labs(x = "Cytokine", y = "-log10(adjusted P)",
       title = "Top 25 cytokines") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    text = element_text(size=12, family="Arial"),
    axis.text = element_text(size=12, color="black"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    legend.position = "bottom"
  )
```

Plot heatmap of significant cytokines' slopes.

```{r}
# Figure 4D
res.toplot <- res[res$adj.PVal<0.05, c("slope.1","slope.2","slope.3","slope.4","slope.5")]
colnames(res.toplot) <- levels(clin.sample.event.individ.long$age_group)
out <- pheatmap(
  res.toplot,
  breaks = seq(-0.06,0.06, length.out = 255),
  legend_breaks = c(-0.06,-0.03,0,0.03,0.06),
  # color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(99),
  color = rgb(colorRamp(c ("deepskyblue3", "#F9F4EC", "deeppink4"), space="rgb", interpolate="linear")(0:255/255), maxColorValue=255),
  cluster_cols = FALSE,
  clustering_distance_rows = "euclidean", clustering_method = "complete",
  angle_col = 45,
  cutree_rows = 3, border_color = "gray75"
) # olink_longitudinal_heatmap, 4 x 6.5 in
```

Plot example significant cytokines.

```{r}
# Figure S8B
p <- list()
for (i in c("TNFSF11","CXCL8","CXCL9","IL6")) {
  temp <- data.frame(
    value = scale(olink.long[i,]),
    event_date = clin.sample.event.individ.long$event_date,
    age_group = clin.sample.event.individ.long$age_group,
    sc2 = clin.sample.event.individ.long$sc2_rpm,
    sex = clin.sample.event.individ.long$sex,
    tg = clin.sample.event.individ.long$trajectory_group,
    site = clin.sample.event.individ.long$enrollment_site,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value)) %>%
    droplevels()
  
  # Predict
  fit <- lmer(
    value ~ event_date + event_date:age_group + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","age_group"),
                                  type="fixed")
  colnames(my.pred)[colnames(my.pred)=="group"] <- "age_group"
  
  temp$yhat <- predict(fit)

  # Calculate IQR and quartile values for limiting y-axes
  q1 <- quantile(temp$value, 0.25)
  q3 <- quantile(temp$value, 0.75)
  iqr <- IQR(temp$value)
  p[[i]] <- ggplot(data = temp, aes(x=event_date, y=value, color=age_group)) +
    geom_line(aes(group=pid), color="gray", alpha=0.4) +
    geom_point(aes(color=age_group), shape=20, size=0.5) +
    geom_line(inherit.aes=FALSE, data=my.pred, aes(x,predicted),
              color = "black", size=1) +
    geom_ribbon(inherit.aes=FALSE, data=my.pred,
                aes(x = x, y = predicted,
                    ymin = conf.low, ymax = conf.high),
                fill = "black", alpha = 0.5) +
    coord_cartesian(ylim = c(q1-1.5*iqr, q3+1.5*iqr)) +
    labs(x="Days from admission", y="NPX (z-score)",
         title=sprintf("%s (P = %.1e)",i,res[i,"adj.PVal"])) +
    scale_color_viridis_d(begin=0.4, guide="none") +
    facet_wrap(~ age_group, nrow=1) + theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5, size=12, face="plain"),
      axis.text.x = element_text(size=11, color="black"),
      axis.text.y = element_text(size=12, color="black"),
      text = element_text(size=12, family="Arial"),
      plot.margin = unit(c(0,1,0.5,0), "cm"),
      panel.spacing = unit(0.35, "cm"),
      panel.grid.minor=element_blank()
    )
}

p[[1]] +
  p[[2]] +
  p[[3]] + 
  p[[4]] + plot_layout(ncol = 1) # olink_longitudinal, 7 x 10 in
```

Try plotting all 5 trend lines on one plot

```{r}
# Figure 4E
p <- list()
for (i in c("TNFSF11","CXCL8","CXCL9","IL6")) {
  temp <- data.frame(
    value = scale(olink.long[i,]),
    event_date = clin.sample.event.individ.long$event_date,
    age_group = clin.sample.event.individ.long$age_group,
    sc2 = clin.sample.event.individ.long$sc2_rpm,
    sex = clin.sample.event.individ.long$sex,
    tg = clin.sample.event.individ.long$trajectory_group,
    site = clin.sample.event.individ.long$enrollment_site,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value)) %>%
    droplevels()
  
  # Predict
  fit <- lmer(
    value ~ event_date + event_date:age_group + age_group + sex + tg + (1|pid), 
    data = temp, REML = FALSE)
  my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","age_group"),
                                  type="fixed")
  colnames(my.pred)[colnames(my.pred)=="group"] <- "age_group"
  
  temp$yhat <- predict(fit)

  
  p[[i]] <- ggplot() +
    geom_line(data=my.pred, aes(x,predicted,color=age_group),
              size=1.5) +
    labs(x="Days from admission", y="NPX (z-score)",
         title=sprintf("%s (P = %.1e)",i,res[i,"adj.PVal"])) +
    scale_color_viridis_d(name = "Age group", guide = "none") +
    theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )
}

p[[1]]+p[[2]]+p[[3]] + 
  p[[4]]+ylim(-1.5,0) + plot_layout(nrow = 1) # olink_longitudinal_combined, 13.5 x 3.7 in
```
