---
title: "Fig4"
output: html_document
---

```{r load library}
require(openxlsx)
library(factoextra)
library(cluster)
require(nlme) # directly provides p-values, unlike lme4
require(splines)
require(dplyr)
library(fgsea)
library(knitr)
library(kableExtra)
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source("~/impacc/Analysis/molecular_subtyping/preliminary/helpers_LG202301.R")
source("~/impacc/Analysis/molecular_subtyping/interpretation/leying_helpers.R")
source("~/impacc/Analysis/molecular_subtyping/code/Fig4/src/Fig4_helper.R")

date = "20230726"; inner_folds = 10
path_impacc_analysis = "/scratch/data-integration/cleanedup/"
intermediate_results_outdir_DR = "/scratch/data-integration/molecular_subtyping/DimensionReduction/"
intermediate_results_outdir_RESULTS = "/scratch/data-integration/molecular_subtyping/Results/"
organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")
file_mcia = paste0(intermediate_results_outdir_DR,"MCIA_phase123_",date,".rds")
mcia_results = readRDS(file_mcia)

base_obj = readRDS(file ="/scratch/data-integration/cleanedup/mod1_interpretation_base.rds")
file_impacc_analysis = paste0(path_impacc_analysis,"impacc_analysis_record_innerfold", inner_folds,".rds")
impacc_analysis_record=readRDS(file = file_impacc_analysis)
data_path <- "/scratch/data-integration/pipelines/processedData/"
data_env0 <- readRDS(paste0(data_path, "data_env_ravi_20220513.RDS"))
```

# Load preliminary results
```{r load subtype assignment and clinical data}
# load subtype assignment and clinical data
nclust_tree_all = 6
subtype_hcl = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
clin_data_aligned = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/clin_data_aligned.rds")

# load omics used
omics_used_DR = c(
  "ppt" = "plasma_proteomics_targeted",
  "ppgd" = "plasma_proteomics_global_dda",
  "so" = "serum_olink",
  "pmg" = "plasma_metabolomics_global",
  "nt" = "nasal_transcriptomics",
  "pbmc" = "pbmc_transcriptomics"
)
omic_short_names = names(omics_used_DR) %>% setNames(omics_used_DR)
omic_names =c("plasma_proteomics_targeted","plasma_proteomics_global_dda",
              "serum_olink","plasma_metabolomics_global","nasal_transcriptomics",
              "pbmc_transcriptomics")

# load all databases used for enrichment
kegg_database = readRDS(file = "/scratch/data-integration/pipelines/Interpretation/remapped_databases/kegg0522.rds")
hallmark_database = readRDS( file = "/scratch/data-integration/pipelines/Interpretation/remapped_databases/hallmark0522.rds")
metabolite_row_DF = data_env0$plasma_metabolomics_global_rowfeature
rownames(metabolite_row_DF) = metabolite_row_DF[["CHEMICAL_NAME"]]
metabolite_row_DF=metabolite_row_DF[,c("CHEMICAL_NAME","SUB_PATHWAY","SUPER_PATHWAY")]
subpath_databse = list()
subpath_databse[["pmg"]] = data.frame(description =metabolite_row_DF[["SUB_PATHWAY"]],
                                      CHEMICAL_NAME = metabolite_row_DF[["CHEMICAL_NAME"]])
for(k in 1:length(kegg_database)){
  kegg_database[[k]]=kegg_database[[k]][which(kegg_database[[k]]!="Ig"),]
}

# load enrichment results
mag_thr=0.3
enrichment_mhg_list0 = readRDS(file=paste0(intermediate_results_outdir_DR, "pathway_enrichment_mhgF1_2_3_10_LG", mag_thr,".rds"))

projection_coef0 = convert_mciaObj_to_annotation_format(mcia_results$mcia_factor_evaluation$projection_coef, omic_names = names(omics_used_DR))
projection_pval0 = convert_mciaObj_to_annotation_format(mcia_results$mcia_factor_evaluation$projection_pval, omic_names = names(omics_used_DR))
K = 10
projection_coef=projection_coef0[1:K]
projection_pval=projection_pval0[1:K]
for(mod_id in c(1:K)){
  mod_name = paste0("mod", mod_id)
  print(mod_name)
  projection_coef[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = projection_coef[[mod_id]], omic_names = omic_names) 
  projection_pval[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = projection_pval[[mod_id]], omic_names = omic_names)
}

###### FACTOR NEGATION IF NEEDED
mod_ids = c(1, 2, 3,10)
negation = c(-1, 1, 1, 1)
# negate factor coefficients
for(l in 1:length(mod_ids)){
  mod_id = mod_ids[l]
  mod_name = paste0("mod", mod_id)
  print(mod_name)
  for(i in 1:length(projection_coef[[mod_id]])){
    projection_coef[[mod_id]][[i]][,1] = negation[l]*
      projection_coef[[mod_id]][[i]][,1]
  }
}
# negate enrichment results
enrichment_mhg_list = enrichment_mhg_list0
for(l in 1:length(enrichment_mhg_list)){
  for(i in 1:length(mod_ids)){
    mod_id =mod_ids[i]
    if(negation[i] == -1){
      enrichment_mhg_list[[l]][[mod_id]]$pos = enrichment_mhg_list0[[l]][[mod_id]]$neg
      enrichment_mhg_list[[l]][[mod_id]]$neg = enrichment_mhg_list0[[l]][[mod_id]]$pos
    }
  }
}
mod_ids = c(1, 10)
negation = c(-1, 1)
# negate factors
mcia_factors = t(t(mcia_results$mcia_factors[,mod_ids]) * negation)
mcia_factors=apply(mcia_factors,2, function(z) qqnorm(z, plot.it = F)$x)
mcia_factors = data.frame(mcia_factors,row.names = rownames(organized_data$datasets$plasma_proteomics_targeted))
mcia_factors = mcia_factors[clin_data_aligned$event_id,]
colnames(mcia_factors) = paste0("Factor", mod_ids)
```

# Factor trajectory (Figure 4A left & Figure 4C left)
```{r factor trajectory, fig.height=6, fig.width=6, warning=FALSE}
assay_mat = mcia_factors
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl =subtype_hcl,
                                 clin_data = clin_data_aligned)
clin_colnames_select = setdiff(colnames(assay_mat), colnames(mcia_factors))

factor_trajectory_analysis_results = helper_trajectory_data_smooth_spline(
  assay_mat = assay_mat, clin_colnames_select = clin_colnames_select,
  endpoint = "label", use_endpoint_levels = c("A", "B", "C", "E", "F"),
  max_date = 40)
test_Fig4_factor_traj = factor_trajectory_analysis_results$test_pval_list
factor_trajectory_analysis_results$test_fig_list$Factor10
```

# Early/late Factor 1 multi-group comparison (Figure 4A right)
```{r early/late Factor 1 multi-group comparison, fig.height=6, fig.width=3.7}
assay_mat = mcia_factors
assay_mat$event_id = rownames(assay_mat)
assay_mat = cytof_mat_preprocess(cytof_mat = assay_mat, 
                                 subtype_hcl =subtype_hcl,
                                 clin_data = clin_data_aligned)

discretized_test = list()
feature_name = c("Factor1","Factor10")
discretized_test[["early"]] = multi_group_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V1", "V2", "V3"))
discretized_test[["late"]] = multi_group_mixed_test(
  feature_name = feature_name, data_mat = assay_mat, 
  visit_numbers = c("V4", "V5", "V6"))

# Factor 1
est_mat = cbind(discretized_test$early$est[,1],
                discretized_test$late$est[,1])
colnames(est_mat) = c("V1-V3", "V4-V6")
pval_mat = cbind(discretized_test$early$pval[,1],
                 discretized_test$late$pval[,1])
colnames(pval_mat) = c("V1-V3", "V4-V6")
adj_p = apply(pval_mat, 2, function(p) p.adjust(p, method = "BH"))
test_Fig4A = list()
test_Fig4A$estimate = est_mat
test_Fig4A$pval = pval_mat
test_Fig4A$adj_p = adj_p

factor1_multi_comparison = ComplexHeatmap::Heatmap(
  as.matrix(est_mat), row_names_gp = gpar(fontsize = 18), 
  column_names_gp = gpar(fontsize = 18), column_names_rot = 0, 
  column_names_centered = T, 
  heatmap_legend_param = list(title = "Coefficient", direction = "horizontal", 
                              title_position = "topcenter",
                              title_gp = gpar(fontsize = 18), 
                              labels_gp = gpar(fontsize = 18)),
  col = circlize::colorRamp2(c(-max(abs(est_mat)), 0, max(abs(est_mat))), 
                             c("#284AA0", "white", "#BC4A4A")), 
  na_col = "grey90", cluster_columns = F, cluster_rows = F, 
  cell_fun = function(j, i, x, y, w, h, fill) {
    if(adj_p[i, j] < 0.001) {
      grid.text("***", x, y-h*0.08, gp = gpar(cex=2))
      } else if(adj_p[i, j] < 0.01) {
        grid.text("**", x, y-h*0.08, gp = gpar(cex=2))
        } else if(adj_p[i, j] < 0.05) {
          grid.text("*", x, y-h*0.08, gp = gpar(cex=2))
          }
    })
ComplexHeatmap::draw(factor1_multi_comparison, heatmap_legend_side = "top")
```

# Early/late Factor 10 multi-group comparison (Figure 4C right)
```{r early/late Factor 10 multi-group comparison, fig.height=6, fig.width=3.7}
# Factor 10
est_mat = cbind(discretized_test$early$est[,2],
                discretized_test$late$est[,2])
colnames(est_mat) = c("V1-V3", "V4-V6")
pval_mat = cbind(discretized_test$early$pval[,2],
                 discretized_test$late$pval[,2])
colnames(pval_mat) = c("V1-V3", "V4-V6")
adj_p = apply(pval_mat, 2, function(p) p.adjust(p, method = "BH"))
test_Fig4C = list()
test_Fig4C$estimate = est_mat
test_Fig4C$pval = pval_mat
test_Fig4C$adj_p = adj_p

factor10_multi_comparison = ComplexHeatmap::Heatmap(
  as.matrix(est_mat), row_names_gp = gpar(fontsize = 18), 
  column_names_gp = gpar(fontsize = 18), column_names_rot = 0, 
  column_names_centered = T, 
  heatmap_legend_param = list(title = "Coefficient", direction = "horizontal", 
                              title_position = "topcenter",
                              title_gp = gpar(fontsize = 18), 
                              labels_gp = gpar(fontsize = 16)),
  col = circlize::colorRamp2(c(-max(abs(est_mat)), 0, max(abs(est_mat))), 
                             c("#284AA0", "white", "#BC4A4A")), 
  na_col = "grey90", cluster_columns = F, cluster_rows = F, 
  cell_fun = function(j, i, x, y, w, h, fill) {
    if(adj_p[i, j] < 0.001) {
      grid.text("***", x, y-h*0.08, gp = gpar(cex=2))
      } else if(adj_p[i, j] < 0.01) {
        grid.text("**", x, y-h*0.08, gp = gpar(cex=2))
        } else if(adj_p[i, j] < 0.05) {
          grid.text("*", x, y-h*0.08, gp = gpar(cex=2))
          }
    })
ComplexHeatmap::draw(factor10_multi_comparison, heatmap_legend_side = "top")
```

# Enrichment analysis - Factor 1 (Figure 4B)
```{r enrichment analysis - Factor 1, fig.height=10, fig.width=10}
KEGG_max = 10; Hallmark_max = 10; Subpath_max = 5
KEGG_metabo_max = 2; pathway_max = 30; date = 20231201
kegg_pathnames = unique(kegg_database$pbmc$description)
kegg_metabopathnames = kegg_pathnames[grepl("metabolism",kegg_pathnames)]
data_base_names = c( "kegg","hallmark","subpath")
size_max = c(KEGG_max,Hallmark_max,Subpath_max)
shorten = FALSE
plot_list = list()
test_enrichment = list()
thr = 0.05
pathway_counts = data.frame(matrix(NA, ncol = 6, nrow = length(mod_ids)))
colnames(pathway_counts) = names(enrichment_mhg_list$kegg[[1]]$pos)
rownames(pathway_counts) = paste0("Factor", mod_ids)

for(i in 1:length(mod_ids)){
  mod_id = mod_ids[i]
  factor_enrichment = lapply(enrichment_mhg_list, function(z) z[[mod_id]])
  aggregated_mat = enrichment_aggregation_subtype(
    factor_enrichment[1:3], adj.p.thr = thr, thr = .2, type = "mhg")
  temp = aggregated_mat[,c(1:6)]
  temp[is.na(temp)] = 0
  pathway_counts[i,]=apply(abs(temp) > (-log(thr, base = 10)), 2, sum)
  if(shorten){
    aggregated_mat_short = pathway_shortlist_selection(
      aggregated_mat, data_base_names = data_base_names, 
      size_max = size_max, KEGG_metabo_max = KEGG_metabo_max)
  } else {
    aggregated_mat_short = aggregated_mat
  }
  aggregated_mat_short = aggregated_mat_short
  aggregated_mat_short = aggregated_mat_short[order(-aggregated_mat_short$joint),]
  
  plot_title = paste0("Factor",mod_id)
  plotDF_df = gather(
    aggregated_mat_short[1:min(nrow(aggregated_mat_short), pathway_max),],
    key = "dataset", value = "log10.sig.val", -joint, -pathway, -database)
  plotDF_df = plotDF_df[!is.na(plotDF_df$log10.sig.val),]
  plotDF_df$direction = NA
  plotDF_df$direction[plotDF_df$log10.sig.val>0] = "pos"
  plotDF_df$direction[plotDF_df$log10.sig.val<0] = "neg"
  plotDF_df$direction[plotDF_df$dataset == "joint_adj"] = "unsigned"
  plotDF_df$log10.sig.val = abs(plotDF_df$log10.sig.val)
  
  plotDF_df$pathway = factor(
    plotDF_df$pathway,
    levels = aggregated_mat_short$pathway[length(aggregated_mat$pathway):1])
  plotDF_df$dataset = factor(plotDF_df$dataset, 
                             levels = c(omic_short_names, "joint_adj"))
  plotDF_df$database = factor(plotDF_df$database, 
                              levels = data_base_names)
  test_enrichment[[i]] = plotDF_df
  plot_list[[i]] = ggplot(data = plotDF_df, 
                          aes(x = dataset, y = pathway,
                              size = log10.sig.val, col = direction)) +
    geom_point(color="white") +
    geom_point(shape=1) +
    geom_point(data=. %>% dplyr::filter(log10.sig.val > -log10(thr) ) ) +
    ggtitle(plot_title) + 
    theme_classic() +
    theme(panel.grid.major = element_line(color = "grey", size = 0.1), 
          axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, 
                                     size = 15, color = "black"), 
          axis.text.y = element_text(size = 15, color = "black"),
          strip.text = element_text(color = "black", size = 15),
          plot.title = element_text(color = "black", size = 15),
          legend.title = element_text(color = "black", size = 15),
          legend.text = element_text(color = "black", size = 15)) +
    ylab("") +
    xlab("") +
    scale_color_manual(values = c("pos" ="#BC4A4A", "neg" ="#284AA0", 
                                  "unsigned"= "#808080"), 
                       breaks = c("neg", "pos", "unsigned"), 
                       labels = c("negative","positive","unsigned"), 
                       name = "Direction") +
    facet_grid(database~., scales = "free_y", space = "free_y") +
    guides(size = guide_legend(title = expression("Signed "*log[10]*"(p.val)"), 
                               order = 1))
}
#          ppt ppgd so pmg  nt pbmc
# Factor1    8   34 11  12  10   77
# Factor10   0    0  0   0 143    0
print(pathway_counts)
test_Fig4B = test_enrichment[[1]]

plot_list[[1]]
```

# Core analytes
```{r core analytes}
# Truncate the analytes in each assay to keep the most important ones.
# Look at the unsigned list to keep both the positive and the negative ones.

K = 10
mcia_output=mcia_results
omics_used_DR = c(
  "ppt" = "plasma_proteomics_targeted",
  "ppgd" = "plasma_proteomics_global_dda",
  "so" = "serum_olink",
  "pmg" = "plasma_metabolomics_global",
  "nt" = "nasal_transcriptomics",
  "pbmc" = "pbmc_transcriptomics"
)
omic_short_names = names(omics_used_DR) %>% setNames(omics_used_DR)
omic_names =c("plasma_proteomics_targeted","plasma_proteomics_global_dda",
              "serum_olink","plasma_metabolomics_global","nasal_transcriptomics",
              "pbmc_transcriptomics")
projection_coef = convert_mciaObj_to_annotation_format(
  mcia_output$mcia_factor_evaluation$projection_coef, 
  omic_names = names(omics_used_DR))
projection_pval = convert_mciaObj_to_annotation_format(
  mcia_output$mcia_factor_evaluation$projection_pval, 
  omic_names = names(omics_used_DR))
projection_coef=projection_coef[1:K]
projection_pval=projection_pval[1:K]
for(mod_id in c(1:K)){
  mod_name = paste0("mod", mod_id)
  print(mod_name)
  projection_coef[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = projection_coef[[mod_id]], omic_names = omic_names)
  projection_pval[[mod_id]]= coefficient_name_translator(data_env = data_env0, projection_coef_factor = projection_pval[[mod_id]], omic_names = omic_names)
}
short_list_features = list()
mag_thr = 0.3; adjp = 0.01
for(mod_id in c(1:K)){
  ###extract coefficients for relevant module
  tmp_coef = projection_coef[[mod_id]]
  tmp_pval = projection_pval[[mod_id]]
  short_list_features[[mod_id]] = list()
  short_list_features[[mod_id]][["unsigned"]]=cutlist(
    projection_coef = tmp_coef ,
    projection_pval = tmp_pval,
    mag_thr = mag_thr, FWER_thr =adjp, direction = NULL)
  short_list_features[[mod_id]][["pos"]]=cutlist(
    projection_coef = tmp_coef ,
    projection_pval = tmp_pval,
    mag_thr = mag_thr, FWER_thr =adjp, direction =1)
  short_list_features[[mod_id]][["neg"]]=cutlist(
    projection_coef = tmp_coef ,
    projection_pval = tmp_pval,
    mag_thr = mag_thr, FWER_thr =adjp, direction =-1)
}
counts_analyte = lapply(short_list_features,function(z) sapply(z[[1]],dim))
# counts_analyte[[10]]
# short_list_features[[1]]$unsigned$pbmc[order(-abs(short_list_features[[1]]$unsigned$pbmc[,1])),,drop = F]
```

# Enrichment analysis - Factor 10 (Figure 4D bottom)
```{r enrichment analysis - Factor 10, fig.height=8.2, fig.width=10}
KEGG_max = 10; Hallmark_max = 10; Subpath_max = 5
KEGG_metabo_max = 2; pathway_max = 30; date = 20231201
kegg_pathnames = unique(kegg_database$pbmc$description)
kegg_metabopathnames = kegg_pathnames[grepl("metabolism",kegg_pathnames)]
data_base_names = c( "kegg","hallmark","subpath")
size_max = c(KEGG_max,Hallmark_max,Subpath_max)
shorten =FALSE
plot_list = list()
thr = 0.05

i=2
mod_id = mod_ids[i]
factor_enrichment = lapply(enrichment_mhg_list, function(z) z[[mod_id]])
aggregated_mat = enrichment_aggregation_subtype(
  factor_enrichment[1:3], adj.p.thr =thr, thr = .2, type = "mhg")
temp = aggregated_mat[,c(1:6)]
temp[is.na(temp)] = 0
if(shorten){
  aggregated_mat_short = pathway_shortlist_selection(
    aggregated_mat, data_base_names = data_base_names, 
    size_max = size_max, KEGG_metabo_max = KEGG_metabo_max)
} else {
  aggregated_mat_short = aggregated_mat
}
aggregated_mat_short = aggregated_mat_short
aggregated_mat_short = aggregated_mat_short[order(-aggregated_mat_short$joint),]

plot_title = paste0("Factor",mod_id)
plotDF_df = gather(
  aggregated_mat_short[1:min(nrow(aggregated_mat_short), pathway_max),],
  key = "dataset", value = "log10.sig.val", -joint, -pathway, -database)
plotDF_df = plotDF_df[!is.na(plotDF_df$log10.sig.val),]
plotDF_df$direction = NA
plotDF_df$direction[plotDF_df$log10.sig.val>0] = "pos"
plotDF_df$direction[plotDF_df$log10.sig.val<0] = "neg"
plotDF_df$direction[plotDF_df$dataset == "joint_adj"] = "unsigned"
plotDF_df$log10.sig.val = abs(plotDF_df$log10.sig.val)

plotDF_df$pathway = factor(
  plotDF_df$pathway, 
  levels = aggregated_mat_short$pathway[length(aggregated_mat$pathway):1])
plotDF_df$dataset = factor(plotDF_df$dataset, 
                           levels = c(omic_short_names, "joint_adj"))
plotDF_df$database = factor(plotDF_df$database, levels = data_base_names)
plotDF_df = plotDF_df[plotDF_df$dataset=="nt",]

# Count analytes in each pathways
analytes = short_list_features[[10]]$unsigned$nt
nt_kegg = kegg_database$nt
nt_hallmark = hallmark_database$nt
nt_kegg = nt_kegg[nt_kegg$description %in% plotDF_df$pathway,]
nt_hallmark = nt_hallmark[nt_hallmark$description %in% plotDF_df$pathway,]
nt = rbind(nt_kegg, nt_hallmark)
nt = nt[nt$gene_name %in% rownames(analytes),]
counts = as.data.frame(table(nt$description))
colnames(counts) = c("pathway", "count")
plotDF_df = merge(plotDF_df, counts, by = "pathway")
test_Fig4D = plotDF_df
ggplot(data = plotDF_df, aes(x = count, y = pathway, fill = log10.sig.val)) +
  geom_bar(stat = "identity") + 
  ggtitle(plot_title) + 
  theme_classic() +
  theme(panel.grid.major = element_line(color = "grey", size = 0.1), 
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, 
                                   size = 15, color = "black"), 
        axis.text.y = element_text(size = 15, color = "black"),
        axis.title.x = element_text(size = 15, color = "black"),
        strip.text = element_text(color = "black", size = 15),
        plot.title = element_text(color = "black", size = 15),
        legend.title = element_text(color = "black", size = 15),
        legend.text = element_text(color = "black", size = 15),
        ) +
  ylab("") +
  xlab("Analyte count") +
  scale_fill_gradient2(low = "#284AA0", mid="white",  high = "#BC4A4A", 
                       midpoint=0, name = expression("Signed "*log[10]*"(p.val)"),
                       na.value="grey90", 
                       limits = c(-max(abs(plotDF_df$log10.sig.val)), 
                                  max(abs(plotDF_df$log10.sig.val)))) +
  facet_grid(database~., scales = "free_y", space = "free_y")
#          ppt ppgd so pmg  nt pbmc
# Factor1    8   34 11  12  10   77
# Factor10   0    0  0   0 143    0
```

# Analyte and pathway count (Figure 4D top)
```{r analyte and pathway count}
df_count = data.frame(Counts = c("Analyte", "Pathway"),
                      rbind(counts_analyte[[10]][1,],
                            pathway_counts[2,]), row.names = NULL)

kable(df_count) %>%
  kable_styling(full_width = FALSE)
```

# Load viral load/antibody/cytof and cytokine cluster PC
```{r load viral load/antibody/cytof and cytokine cluster PC}
cytof_parent_mat_train = base_obj$others$bld_cytof$train$bld_cytof_counts
cytof_parent_mat_train = 
  base_obj$others$bld_cytof$preprocess_bld_cytof_parent_granulo_separate(
  counts = cytof_parent_mat_train, 
  rowfeature = base_obj$others$bld_cytof$train$bld_cytof_rowfeatures)
cytof_parent_mat_test = base_obj$others$bld_cytof$test$bld_cytof_counts
cytof_parent_mat_test = 
  base_obj$others$bld_cytof$preprocess_bld_cytof_parent_granulo_separate(
    counts = cytof_parent_mat_test, 
    rowfeature = base_obj$others$bld_cytof$test$bld_cytof_rowfeatures)

cytof_parent_mat = rbind(cytof_parent_mat_train, cytof_parent_mat_test)
cytof_feature_names = colnames(cytof_parent_mat)
cytof_parent_mat$event_id = rownames(cytof_parent_mat)
cytof_parent_mat = cytof_mat_preprocess(cytof_parent_mat, 
                                        subtype_hcl = subtype_hcl,  
                                        clin_data = clin_data_aligned)

ct_train = data.frame(
  "event_id" = rownames(base_obj$others$nasal_viralload$train$nasal_viralload_counts),
  "Ct" = base_obj$others$nasal_viralload$train$nasal_viralload_counts$N1_CT,
  "RP_CT" = base_obj$others$nasal_viralload$train$nasal_viralload_counts$RP_CT)
ct_test = data.frame(
  "event_id" = rownames(base_obj$others$nasal_viralload$test$nasal_viralload_counts),
  "Ct" = base_obj$others$nasal_viralload$test$nasal_viralload_counts$N1_CT,
  "RP_CT" = base_obj$others$nasal_viralload$test$nasal_viralload_counts$RP_CT)
ct_df = rbind(ct_train, ct_test)
ct_df = ct_df[order(ct_df$event_id),]
rownames(ct_df) = ct_df$event_id
# remove samples with undetected RP_CT values
ct_df = ct_df[!is.na(ct_df$RP_CT), ]
# replace NA (undetermined) in N1_CT to 40
ct_df$Ct[is.na(ct_df$Ct)] = 40
# replace values > 40 to 40
ct_df$Ct[ct_df$Ct > 40] = 40
# flip the sign of N1_CT
ct_df$Ct = -ct_df$Ct
ct_df = subset(ct_df, select = -c(event_id, RP_CT))
ct_df = quantileNormalizeNoTies(ct_df)
ct_df = as.data.frame(ct_df)
ct_df$event_id = rownames(ct_df)
ct_df = cytof_mat_preprocess(ct_df, subtype_hcl, clin_data_aligned)

auc_train = data.frame(
  "event_id" = rownames(base_obj$others$serum_rbd_abtiters$train$serum_rbd_abtiters_counts),
  "log2_AUC" = base_obj$others$serum_rbd_abtiters$train$serum_rbd_abtiters_counts$`AUC RBD IgG`)
auc_test = data.frame(
  "event_id" = rownames(base_obj$others$serum_rbd_abtiters$test$serum_rbd_abtiters_counts),
  "log2_AUC" = base_obj$others$serum_rbd_abtiters$test$serum_rbd_abtiters_counts$`AUC RBD IgG`)
auc_df = rbind(auc_train, auc_test)
auc_df = auc_df[order(auc_df$event_id),]
rownames(auc_df) = auc_df$event_id
auc_df = subset(auc_df, select = -event_id)
auc_df = quantileNormalizeNoTies(auc_df)
auc_df = as.data.frame(auc_df)
auc_df$event_id = rownames(auc_df)
auc_df = cytof_mat_preprocess(auc_df, subtype_hcl, clin_data_aligned)

cytokine_mat = organized_data$datasets$serum_olink
cytokine_feature_names = colnames(cytokine_mat)
cytokine_mat$event_id = rownames(cytokine_mat)
cytokine_mat = cytof_mat_preprocess(cytokine_mat, 
                                    subtype_hcl = subtype_hcl,  
                                    clin_data = clin_data_aligned)
cytokine_cls = readRDS("~/impacc/Analysis/data-integration/Subtyping/test/olink_cluster.rds")

cytokine_PCs = matrix(NA, ncol = 10, nrow = nrow(cytokine_mat))
for(SPmod in 1:10){
  ii = cytokine_cls$feature[cytokine_cls$category==SPmod]
  temp = cytokine_mat[,ii]
  ii1 = which(!is.na(temp[,1]))
  temp1 = svd(scale(temp[ii1,],scale = F, center = T))
  loading = temp1$v[,1]
  if(sum(loading)<0){
    loading= -loading
  }
  cytokine_PCs[,SPmod] = as.matrix(temp)%*% loading
}
cytokine_PCs = data.frame(cytokine_PCs)
rownames(cytokine_PCs) = rownames(cytokine_mat)
colnames(cytokine_PCs) = paste0("SPmod",1:10)
cytokine_PCs$event_id = cytokine_mat$event_id
```

# Interaction with viral load and cytokine PC (Figure 4E)
```{r interaction with viral load and cytokine PC, fig.height=6, fig.width=9.7}
datDF = left_join(left_join(left_join(left_join(
  assay_mat, ct_df[, c("Ct", "event_id")], by = "event_id"),
  auc_df[, c("log2_AUC", "event_id")], by = "event_id"),
  cytof_parent_mat[, 1:15], by = "event_id"),
                  cytokine_PCs, by = "event_id")
labels_select <- c("Factor1", "Factor10", "Ct", "log2_AUC", paste0("SPmod",1:10))
datDF = datDF[, c(labels_select,
                  "participant_id", "event_type", "event_date", 
                  "enrollment_site", "discretized_admit_age_quantile", 
                  "sex", "label")]
datDF[,1:length(labels_select)] = scale(datDF[,1:length(labels_select)])
datDF0 = datDF[datDF$label %in% c("A", "B", "C"),]
datDF0 = datDF0[apply(is.na(datDF0),1,sum) == 0,]

datDF1 = datDF[datDF$label %in% c("E","F"),]
datDF1 = datDF1[apply(is.na(datDF1),1,sum)==0,]

datDF2 = rbind(datDF0, datDF1)
datDF2 = cbind(datDF2[,1:14], datDF2)
colnames(datDF2)[1:14] = paste0(colnames(datDF2)[15:28],"_delta")
datDF2[datDF2$label %in% c("A","B","C"),1:14] = 0

# both pan-inflammation
formula0 = formula(paste0("Ct~sex+discretized_admit_age_quantile+Factor1+Factor10+log2_AUC+",
                          paste(paste0("SPmod",1:10),collapse = "+")))
formula1 = formula(paste0("Ct~sex+discretized_admit_age_quantile+log2_AUC+",
                          paste(paste0("SPmod",1:10),collapse = "+")))
formula2 = formula(paste0("Ct~sex+discretized_admit_age_quantile+Factor1+Factor10+log2_AUC+",
                          paste(paste0("SPmod",1:10),collapse = "+"), "+", 
                          paste(colnames(datDF2)[1:14][-3],collapse = "+")))
formula3 = formula(paste0("Ct~sex+discretized_admit_age_quantile+log2_AUC+",
                          paste(paste0("SPmod",1:10),collapse = "+"), "+",
                          paste(colnames(datDF2)[4:14],collapse = "+")))

fit1 = nlme::lme(formula0, random = ~1|enrollment_site/participant_id,data=datDF0)
fit2 = nlme::lme(formula0, random = ~1|enrollment_site/participant_id,data=datDF1)
coef_table = data.frame(t(cbind(summary(fit1)$tTable[-(1:6),1], summary(fit2)$tTable[-(1:6),1])))[,1:2]
pval_table = data.frame(t(cbind(summary(fit1)$tTable[-(1:6),5], summary(fit2)$tTable[-(1:6),5])))[,1:2]
fit5 = nlme::lme(formula2, random = ~1|enrollment_site/participant_id,data=datDF2)
coef_table = rbind(coef_table, summary(fit5)$tTable[-(1:6),1][c(14,15)])
pval_table = rbind(pval_table, summary(fit5)$tTable[-(1:6),5][c(14,15)])

fit3 = nlme::lme(formula1, random = ~1|enrollment_site/participant_id,data=datDF0)
fit4 = nlme::lme(formula1, random = ~1|enrollment_site/participant_id,data=datDF1)
fit6 = nlme::lme(formula3, random = ~1|enrollment_site/participant_id,data=datDF2)
coef_table = cbind(coef_table, rbind(data.frame(t(cbind(summary(fit3)$tTable[-(1:6),1], summary(fit4)$tTable[-(1:6),1]))),summary(fit6)$tTable[-(1:6),1][12:22])) 
pval_table = cbind(pval_table, rbind(data.frame(t(cbind(summary(fit3)$tTable[-(1:6),5], summary(fit4)$tTable[-(1:6),5]))),summary(fit6)$tTable[-(1:6),5][12:22]))
rownames(coef_table) = rownames(pval_table) = c("Severe", "Critical", "Critical-Severe")

pval_table_adj = pval_table
pval_table_adj = t(apply(pval_table, 1, function(x) p.adjust(x, method = "BH")))
ii = which(apply(pval_table_adj[1:2,],2,min)<0.1)
pval_table_adj=pval_table_adj[,ii]
colnames(coef_table)[3] <- "AUC RBD IgG"
test_types = c(rep("formula0",2),rep("formula1",length(ii)-2))
test_types = factor(test_types, levels = c("formula0","formula1"))
factor_interaction = ComplexHeatmap::Heatmap(
  as.matrix(coef_table[,ii]), column_title_gp = gpar(fontsize = 25),
  column_title=gt_render("Interaction with viral load"),
  row_names_gp = gpar(fontsize = 18), column_names_gp = gpar(fontsize = 18),
  heatmap_legend_param = list(title = "Coefficient    ",
                              direction = "horizontal", 
                              title_position = "leftcenter",
                              title_gp = gpar(fontsize = 18), 
                              labels_gp = gpar(fontsize = 18)),
  col = circlize::colorRamp2(c(-max(abs(coef_table[,ii])), 0, 
                               max(abs(coef_table[,ii]))), 
                             c("#284AA0", "white", "#BC4A4A")), 
  na_col = "grey90", column_split = test_types, column_gap = unit(2, "mm"),
  cluster_columns = F, cluster_rows = F, 
  cell_fun = function(j, i, x, y, w, h, fill) {
    if(pval_table_adj[i, j] < 0.001) {
      grid.text("***", x, y, gp = gpar(cex=2), vjust=0.7)
      }else if(pval_table_adj[i, j] < 0.01) {
        grid.text("**", x, y, gp = gpar(cex=2), vjust=0.7)
        }else if(pval_table_adj[i, j] < 0.05) {
          grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
          }
    })
test_Fig4E = list()
test_Fig4E$Coefficient = coef_table
test_Fig4E$pval = pval_table
colnames(test_Fig4E$pval)[colnames(test_Fig4E$pval) == "log2_AUC"] = "AUC RBD IgG"
test_Fig4E$adj_pval = t(apply(pval_table,1,p.adjust))
colnames(test_Fig4E$adj_pval)[colnames(test_Fig4E$adj_pval) == "log2_AUC"] = "AUC RBD IgG"
ComplexHeatmap::draw(factor_interaction, heatmap_legend_side = "bottom")
```


