---
title: "Serum antibody titers data analysis for core-assay manuscript"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ggnetwork"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "GSA"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))
suppressPackageStartupMessages(library(package = "ggsignif"))

# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```

### Load codebase.R, load data matrices and set path to your local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE}
# load codebase.R, serum-rbd-abtiters data and clinical information
# source("~/bitbucket/impacc/Analysis/data_analysis_template_codebase.R")
source("../../Codebase/codebase_v2.R")
data_env <- load_IMPACC_datasets(DATA_VERSION                = "2022-01-01", 
                                 KEEP_COVID19_POS            = TRUE, 
                                 FILTER_BY_CORE_ASSAY_COHORT = TRUE,
                                 ASSAY_NAMES = "serum_rbd_abtiters")
#data_env <- readRDS("~/data_env.rds")
for (n in grep(pattern = "serum_rbd_abtiters_counts|clinical_data", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}


# Setup the assay specific output directory
out_dir = "../output"



```


&nbsp;

&nbsp;

#### Combine clinical data and serum rbd titers data

```{r CombineClinicalSerumRBDTitersData, echo=FALSE}
##### Authors: Naresh Doni Jayavelu

# fix column names, replace spaces with dots.
colnames(serum_rbd_abtiters_counts) <- gsub(" ", ".", colnames(serum_rbd_abtiters_counts))

serum_rbd_abtiters_counts_df <- serum_rbd_abtiters_counts %>%
  rownames_to_column(var = "sample_id") %>%
  select(sample_id, 
         AUC.RBD.IgG,
         AUC.Spike.IgG,
         AUC.Spike.IgM,
         AUC.Spike.IgA)


clinical_titers_counts_data <- clinical_data %>%
  inner_join(serum_rbd_abtiters_counts_df, by = "sample_id") 

# convert discretized_admit_age_quantile into a factor
clinical_titers_counts_data <- clinical_titers_counts_data %>%
  mutate(discretized_admit_age_quantile = factor(discretized_admit_age_quantile, levels = c("[19,35]", "[36,50]", "[51,65]", "[66,80]", "[81,95]")))


```


&nbsp;

&nbsp;

&nbsp;

&nbsp;

&nbsp;

#### Panel E analysis code

```{r panel-E-analysis, warning=FALSE}

data_use_visit1 <- clinical_titers_counts_data %>%
  tidyr::drop_na(trajectory_group) %>%
  filter(event_type == "Visit 1") %>%
  mutate(RBD.IgG = log2(AUC.RBD.IgG),
         Spike.IgG = log2(AUC.Spike.IgG)) %>%
  select(sample_id, 
         RBD.IgG,
         Spike.IgG,
         enrollment_site,
         trajectory_group,
         discretized_admit_age_quantile,
         sex) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group))

module_names <- c("RBD.IgG", "Spike.IgG")

res_table_ordinal <- NULL
data_use <- data_use_visit1
for (moduleName in module_names) {
  my.formula0 <- paste0('trajectory_group~ 1+(1|enrollment_site)+',
                        'discretized_admit_age_quantile+sex')
  my.formula1 <- paste0('trajectory_group~',
                        moduleName,
                        '+(1|enrollment_site)+',
                        'discretized_admit_age_quantile+sex')
  res <- mixed_ordinal(my.formula0, 
                       my.formula1,
                       data_use = data_use_visit1)
  coef(clmm(formula(my.formula1), 
            data = data_use_visit1))[moduleName] %>%
    unname() %>%
    c(res, coef = .) -> res
  res_table_ordinal <- rbind(res_table_ordinal, res)
}
rownames(res_table_ordinal) <- module_names
res_table_ordinal <- res_table_ordinal %>%
  as.data.frame() %>%
   mutate(qval =  qvalue::qvalue(pval, fdr.level = 0.05, pi0 = 1)$qvalues)


res_table_ordinal_df <- res_table_ordinal %>%
  rownames_to_column(var = "feature")

# pairwise comparisons
data_use_pairwise <- data_use %>%
  rename(endpoints = trajectory_group)

for(j in 1:length(module_names)){
  print(j)
  my.formula0 = paste0(module_names[j],"~ 1+(1|enrollment_site)+discretized_admit_age_quantile+sex")
  my.formula1 = paste0(module_names[j],"~ endpoints + (1|enrollment_site)+discretized_admit_age_quantile+sex")
  if(j == 1){
    tmp = mixed_pairwise(my.formula0, my.formula1, data_use_pairwise)
    res_table_pairwise = data.frame(matrix(0, ncol = length(tmp), nrow = length(module_names)))
    res_table_pairwise[j,] = tmp
    colnames(res_table_pairwise) = names(tmp)
  }else{
    res_table_pairwise[j,] = mixed_pairwise(my.formula0, my.formula1, data_use_pairwise)
  }
}
rownames(res_table_pairwise) = module_names


res_table_ordinal_df <- cbind(res_table_ordinal_df, res_table_pairwise)


```



### Main Panel E: output generation code
```{r panel-E-output-gen, fig.height=4, fig.width=5}
# RBD IgG
visit1_RBD <- data_use_visit1 %>%
  # rownames_to_column(var = "sample_id") %>%
  select(sample_id, RBD.IgG, trajectory_group) %>%
  pivot_longer(cols = -c(sample_id, trajectory_group)) %>%
  mutate(name             = "anti-RBD IgG",
         trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = trajectory_group, y = value)) +
  geom_quasirandom(mapping = aes(color = trajectory_group), cex = 1.5) +
  geom_boxplot(fill = "transparent", outlier.color= "transparent") +
  facet_wrap(facets= ~name, ncol = 1) + 
  labs(y = "log2(AUC)", x = "Clinical trajectory group") +
  scale_color_manual(values = c("TG1" = "#639A21",
                                "TG2" = "#39828C",
                                "TG3" = "#6371AD",
                                "TG4" = "#BD7D31",
                                "TG5" = "#9C3418")) +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=16)) 

print(visit1_RBD)
ggsave(
 filename = file.path(out_dir, "main_E.pdf"),
 width = 3,
 height = 3
)





```



### Supple Panel C: output generation code
```{r Supple-panel-C-output-gen, fig.height=4, fig.width=5}

# Spike IgG
visit1_Spike <- data_use_visit1 %>%
  # rownames_to_column(var = "sample_id") %>%
  select(sample_id, Spike.IgG, trajectory_group) %>%
  pivot_longer(cols = -c(sample_id, trajectory_group)) %>%
  mutate(name             = "anti-Spike IgG",
         trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = trajectory_group, y = value)) +
  geom_quasirandom(mapping = aes(color = trajectory_group), cex = 1.5) +
  geom_boxplot(fill = "transparent", outlier.color= "transparent") +
  facet_wrap(facets= ~name, ncol = 1) + 
  labs(y = "log2(AUC)", x = "Clinical trajectory group") +
  scale_color_manual(values = c("TG1" = "#639A21",
                                "TG2" = "#39828C",
                                "TG3" = "#6371AD",
                                "TG4" = "#BD7D31",
                                "TG5" = "#9C3418")) +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=16)) 

print(visit1_Spike)
ggsave(
 filename = file.path(out_dir, "suppl_C.pdf"),
 width = 3,
 height = 3
)

```



&nbsp;

&nbsp;



### Panel D: analysis code
```{r panel-D-analysis, message=FALSE, results="hide"}

data_use_longitudinal <- clinical_titers_counts_data %>% 
  tidyr::drop_na(trajectory_group) %>% 
  mutate(sample_id, 
         RBD.IgG = AUC.RBD.IgG, 
         Spike.IgG = AUC.Spike.IgG) %>% 
  mutate(participant_id = factor(participant_id)) %>%
  select(sample_id, 
         RBD.IgG , 
         Spike.IgG, 
         participant_id,  
         enrollment_site,
         trajectory_group, event_date, event_type,
         sex, discretized_admit_age_quantile) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile))


inputDF <- data_use_longitudinal %>%
  pivot_longer(cols = -c("sample_id", "event_date", "event_type", "participant_id",
                         "enrollment_site", "trajectory_group", "sex", "discretized_admit_age_quantile",
                         ))


# smooth spline
smooth_spline_model_loop <- model_loop(inputDF, 
                                       modelType = "smoothSpline", 
                                       endpoint = "trajectory_group")

smooth_spline_model_loop_df <- smooth_spline_model_loop %>%
  rownames_to_column(var = "feature")


```


&nbsp;

&nbsp;


### Main Panel F: output generation code
```{r Main panel-F-output-gen, fig.height=4, fig.width=5}
# RBD IgG
exampleDF <- inputDF %>%
  filter(name %in% "RBD.IgG") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5"))) %>%
  mutate(value = log2(value))

temp_plot <- plot_model(exampleDF, model_loop = smooth_spline_model_loop, modelType = "smoothSpline",
           p_adjust = F, endpoint = "trajectory_group", signif_markers = F, remove_NS = T,
           title = "",
           individual_points_size = 0.5,
           individual_paths_size = 0.5,
           group_trendline_line_width = 1.5,
           title_size = 12,
           bar_height = 5,
           custom_theme_graph = 'text = element_text(size=10)',
           # custom_theme_bars = 'text = element_text(size=3)',
           knot_lines = F,
           ylabel = "log2(AUC)",
           CI = FALSE,
           individual_trendlines = T, individual_points = T, individual_paths = T,
           y_axis_reverse = FALSE, p_value_text_size = 1.5,
           return_multi_obj = T)

print(temp_plot[[1]])
ggsave(
 filename = file.path(out_dir, "main_F.pdf"),
 plot = temp_plot[[1]],
 width = 5,
 height = 2
)




```


### Supple Panel D: output generation code
```{r Supple-panel-D-output-gen, fig.height=4, fig.width=5}

# Spike IgG
exampleDF <- inputDF %>%
  filter(name %in% "Spike.IgG") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5"))) %>%
  mutate(value = log2(value))

temp_plot <- plot_model(exampleDF, model_loop = smooth_spline_model_loop, modelType = "smoothSpline",
           p_adjust = F, endpoint = "trajectory_group", signif_markers = F, remove_NS = T,
           title = "",
           individual_points_size = 0.5,
           individual_paths_size = 0.5,
           group_trendline_line_width = 1.5,
           title_size = 12,
           bar_height = 5,
           custom_theme_graph = 'text = element_text(size=10)',
           # custom_theme_bars = 'text = element_text(size=3)',
           knot_lines = F,
           ylabel = "log2(AUC)",
           CI = FALSE,
           individual_trendlines = T, individual_points = T, individual_paths = T,
           y_axis_reverse = FALSE, p_value_text_size = 1.5,
           return_multi_obj = T)

print(temp_plot[[1]])
ggsave(
 filename = file.path(out_dir, "suppl_D.pdf"),
 plot = temp_plot[[1]],
 width = 5,
 height = 2
)


```


### Session info
```{r, print-session-info}
sessionInfo()
```
<!--
R version 4.0.2 (2020-06-22)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 18.04.4 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so

locale:
 [1] LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8    
 [5] LC_MONETARY=C.UTF-8    LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C             
 [9] LC_ADDRESS=C           LC_TELEPHONE=C         LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

attached base packages:
[1] parallel  grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] doParallel_1.0.17     doRNG_1.8.2           rngtools_1.5.2        missForest_1.4       
 [5] itertools_0.1-3       iterators_1.0.14      foreach_1.5.2         randomForest_4.6-10  
 [9] ComplexHeatmap_2.11.1 corrr_0.4.3           gridExtra_2.3         rlang_1.0.1          
[13] RColorBrewer_1.1-2    cowplot_1.1.1         ggpubr_0.4.0          pals_1.7             
[17] forcats_0.5.1         stringr_1.4.0         dplyr_1.0.8           purrr_0.3.4          
[21] readr_2.1.2           tidyr_1.2.0           tibble_3.1.6          tidyverse_1.3.1      
[25] nlme_3.1-147          lme4_1.1-28           Matrix_1.2-18         GSA_1.03.2           
[29] pvca_0.1.0            ggeffects_1.1.1       ggbeeswarm_0.6.0      qvalue_2.22.0        
[33] ordinal_2019.12-10    ggnetwork_0.5.10      ggplot2_3.3.5         igraph_1.3.0         
[37] impute_1.64.0         knitr_1.37           

loaded via a namespace (and not attached):
 [1] minqa_1.2.4         colorspace_2.0-3    ggsignif_0.6.3      rjson_0.2.21       
 [5] ellipsis_0.3.2      sjlabelled_1.1.8    snakecase_0.11.0    circlize_0.4.14    
 [9] GlobalOptions_0.1.2 fs_1.5.2            dichromat_2.0-0     clue_0.3-60        
[13] rstudioapi_0.13     farver_2.1.0        fansi_1.0.2         lubridate_1.8.0    
[17] xml2_1.3.2          codetools_0.2-16    splines_4.0.2       jsonlite_1.7.3     
[21] nloptr_2.0.0        broom_0.7.12        cluster_2.1.0       dbplyr_2.1.1       
[25] png_0.1-7           mapproj_1.2.8       compiler_4.0.2      httr_1.4.2         
[29] backports_1.4.1     fastmap_1.1.0       assertthat_0.2.1    cli_3.2.0          
[33] htmltools_0.5.2     tools_4.0.2         gtable_0.3.0        glue_1.6.1         
[37] reshape2_1.4.4      maps_3.4.0          Rcpp_1.0.8          carData_3.0-5      
[41] cellranger_1.1.0    vctrs_0.3.8         insight_0.16.0      xfun_0.29          
[45] rvest_1.0.2         lifecycle_1.0.1     rstatix_0.7.0       MASS_7.3-51.6      
[49] scales_1.1.1        hms_1.1.1           gamm4_0.2-6         yaml_2.3.5         
[53] stringi_1.7.6       ucminf_1.1-4        S4Vectors_0.28.1    BiocGenerics_0.36.1
[57] boot_1.3-25         shape_1.4.6         pkgconfig_2.0.3     matrixStats_0.61.0 
[61] evaluate_0.14       lattice_0.20-41     labeling_0.4.2      tidyselect_1.1.2   
[65] plyr_1.8.6          magrittr_2.0.2      R6_2.5.1            IRanges_2.24.1     
[69] generics_0.1.2      DBI_1.1.2           mgcv_1.8-31         pillar_1.7.0       
[73] haven_2.4.3         withr_2.4.3         abind_1.4-5         modelr_0.1.8       
[77] crayon_1.5.0        car_3.0-12          utf8_1.2.2          rmarkdown_2.11     
[81] tzdb_0.2.0          GetoptLong_1.0.5    readxl_1.3.1        reprex_2.0.1       
[85] digest_0.6.29       numDeriv_2016.8-1.1 stats4_4.0.2        munsell_0.5.0      
[89] viridisLite_0.4.0   beeswarm_0.4.0      vipor_0.4.5
-->



