---
title: "Figure2"
format: html
editor: visual
---

## Figure 2

### Boot-up

```{r}
library(tidyverse)
library(data.table)
library(viridis)
library(rstatix)
library(ggpubr)
TrajGrpCol <- c("1" = "#639A21", "2" = "#39828C", "3" = "#6371AD", "4" = "#BD7D31", "5" = "#9C3418")

comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1) %>%
  select(-starts_with("Human betaherpesvirus 6"),
         -starts_with("Human gammaherpesvirus 8"),
         -starts_with("Human betaherpesvirus 7"),
         -starts_with("Respiratory syncytial virus"),
         -starts_with("Human immunodeficiency virus 1"),
         -starts_with("SARS-CoV-2"),
         -starts_with("Enterovirus.pbmc"),
         -starts_with("Enterovirus.ea"),
         -starts_with("Anelloviridae.ea"),
         -starts_with("Anelloviridae.nasal"),
         -starts_with("Human alphaherpesvirus 2 (HSV2).ea"),
         -starts_with("Human alphaherpesvirus 1 (HSV1).pbmc"),
         -starts_with("Human alphaherpesvirus 2 (HSV2).pbmc"))
data_env <- readRDS("/scratch/metagenomics/viral_reactivation/2024-02-02_obj.RDS")
clinical_data <- data_env$clinical_data
rm(data_env)

clinical_data_ext <- read_csv("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-supp-locked.csv")
out_dir <- "../output/"
```

### Figure 2A: Severity (and Figure S6A)

```{r}
library(aplot)
library(rstatix)
library(ordinal)

ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type,
                                     event_date)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0))


graph_data <- ever_reactivate %>%
  left_join(clinical_data %>% distinct(participant_id, trajectory_group)) %>%
  group_by(trajectory_group, name) %>%
  dplyr::summarize(freq = mean(binary) * 100,
                   count = n()) %>%
  filter(count > 10) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
   mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  ungroup() %>%
  complete(trajectory_group, name) %>%
  mutate(freq = ifelse(is.na(freq), 0, freq),
         count = ifelse(is.na(count), 0, count)) %>%
  mutate(positive = freq * count / 100) %>%
  group_by(name) %>%
  mutate(pos_freq = positive / sum(positive) *100)


clm_data <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type, event_date,
                                     sex)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data %>% dplyr::select(participant_id, discretized_admit_age_quantile, enrollment_site, trajectory_group)) %>%
  ungroup() %>%
  filter(!is.na(discretized_admit_age_quantile)) %>%
  filter(!is.na(enrollment_site)) %>%
  filter(!is.na(trajectory_group)) %>%
  distinct() %>%
   mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) 

## TG Association Testing
stat.df <- data.frame()
for(active_virus in unique(graph_data$name)){
  tmp.chi <- graph_data %>%
    filter(name == active_virus) %>%
    ungroup() %>%
    mutate(negative = count - positive) %>%
    select(trajectory_group, positive, negative) %>%
    column_to_rownames("trajectory_group")
  
    tmp.chi <- tmp.chi[4:5,]

  clm_data_tmp <- clm_data %>%
    filter(name == active_virus) %>%
    mutate(trajectory_group = ordered(factor(trajectory_group))) %>%
    mutate(enrollment_site = factor(enrollment_site))
  clmm_res <- clmm2(trajectory_group~binary, data = clm_data_tmp, random = enrollment_site,
                    Hess = TRUE)
  
  stat.df <- rbind.data.frame(stat.df, cbind(chisq_test(tmp.chi) %>%
                                               select(-method, -p,
                                                      -statistic, -df,
                                                      -p.signif, -n) %>%
                                mutate(name = active_virus,
                 clm.p = summary(clmm_res)[["coefficients"]][5,4]),
                 fisher_test(tmp.chi, simulate.p.value = TRUE)))
}

stat.df <- stat.df %>%
  mutate(p.value = ifelse(grepl("\\.ea", name), p, clm.p)) %>%
  mutate(p.adj = p.adjust(p.value, "fdr")) %>%
  mutate(TG4_v_TG5_p.adj = p.adjust(p, "fdr")) %>%
  mutate(p.adj = ifelse(grepl("\\.ea", name), TG4_v_TG5_p.adj, p.adj)) %>%
  add_significance("p.adj") %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  arrange(name) %>%
  mutate(xpos = 11:1) %>%
  mutate(group1 = "filler_text", group2 = "variable") %>%
  mutate(y.position = 40)
write.csv(stat.df, paste0(out_dir, "Supplemental_Data_1_part1.csv"))

ggS6A <- graph_data %>%
  ggplot(aes(y=name, x=pos_freq)) +
  geom_bar(stat = "identity", position = "stack", aes(fill = trajectory_group)) +
  theme_classic() + scale_y_discrete(limits = rev(unique(graph_data$name))) +
  scale_fill_manual(values = TrajGrpCol, name = "Trajectory Group") +
  ylab("Virus and Compartment") +
  xlab("Percent of Participants (%)") +
  theme(axis.text = element_text(color="black"),
        legend.position = "top",
        legend.direction = "vertical",
        axis.ticks = element_line(color = "black"))
ggS6A


gg1 <- graph_data %>%
  mutate(trajectory_group = factor(trajectory_group, levels = 5:1)) %>%
  ggplot(aes(x=name, y=freq)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"), aes(fill = trajectory_group)) +
  theme_classic() + scale_x_discrete(limits = rev(unique(graph_data$name))) +
  scale_fill_manual(values = TrajGrpCol, name = "Trajectory Group") +
  xlab("Virus and Compartment") +
  ylab(" ") +
  stat_pvalue_manual(stat.df %>%
                       filter(!p.adj.signif %in% c("ns", "", " ")) %>%
                       mutate(y.position = c(38, 11, 15, 19, 45, 28)),
                     x= "xpos", label = "p.adj.signif", vjust = 0.7) +
    coord_flip() +
  theme(#axis.text.y = element_blank(),
        #axis.title.y = element_blank(),
        axis.text = element_text(color="black"),
        axis.ticks = element_line(color = "black"))
gg1

```

### Figure 2A: Post-acute Mortality in TG4

```{r, fig.width = 7.5, fig.height = 5.5}
library(rstatix)
library(ggpattern)
library(lme4)
## The ggpattern began to gave issues due to a server update that prevent installation of the sf package
## Thus, this code generates the exact same plots, just without the stripes on the bar for the right
## panel of Figure 2A

ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, trajectory_group, event_type, event_date)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, trajectory_group, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data_ext %>% dplyr::select(participant_id, diedever)) %>%
  ungroup() %>%
  filter(!is.na(trajectory_group)) %>%
  group_by(trajectory_group, name, diedever) %>%
  dplyr::summarize(mean = mean(binary) * 100,
                   count = n(),
                   positive = mean * count /100)

model_df <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, trajectory_group, event_type, event_date)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, trajectory_group, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data_ext %>% dplyr::select(participant_id, diedever, ever_steroids, ever_remdes)) %>%
  left_join(clinical_data %>% dplyr::distinct(participant_id, enrollment_site)) %>%
  ungroup() %>%
  filter(!is.na(trajectory_group)) 

## TG4 Chi.square survival testing
stat.df <- data.frame()
for(active_virus in unique(ever_reactivate$name)){
  tmp.chi <- ever_reactivate %>%
    filter(trajectory_group == 4) %>%
    filter(name == active_virus) %>%
    ungroup() %>%
    mutate(negative = count - positive) %>%
    select(diedever, positive, negative) %>%
    column_to_rownames("diedever")
  
  ## Correction for steroids and remdesivir
  tmp_model_df <- model_df %>%
    filter(trajectory_group == 4) %>%
    filter(name == active_virus) %>%
    mutate(diedever = factor(diedever),
           ever_steroids = factor(ever_steroids),
           ever_remdes = factor(ever_remdes),
           binary = factor(binary))
  
  fit <- glmer(formula(diedever~ever_steroids+ever_remdes+binary+(1|enrollment_site)), data = tmp_model_df, family = "binomial")
  
  aa <- car::Anova(fit, type = "III")
  
  stat.df <- rbind.data.frame(stat.df, chisq_test(tmp.chi) %>%
                                mutate(name = active_virus,
                           med.adj.p = aa[4,3]))
}

stat.df <- stat.df %>%
  mutate(p.adj = p.adjust(p, "fdr")) %>%
  mutate(p.adj.med = p.adjust(med.adj.p, "fdr")) %>%
  add_significance("p.adj.med") %>%
  mutate(group1 = "Died",
         group2 = "Survived",
         `.y.` = "diedever") %>%
  mutate(y.position = 40) %>%
  relocate(name, `.y.`, group1, group2) %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  arrange(name) %>%
  mutate(x=c(11:1), xmin = x-0.2, xmax =x+0.2)
write.csv(stat.df, paste0(out_dir, "Supplemental_Data_1_part2.csv"))

gg2 <- ever_reactivate %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  filter(trajectory_group == "4") %>%
  mutate(diedever = ifelse(diedever == 1, "Died", "Survived")) %>%
  mutate(diedever = factor(diedever, levels = c("Died", "Survived"))) %>%
  ggplot(aes(x=name, y=mean)) +
  geom_bar(aes(fill = diedever, color = diedever), stat = "identity",  width = 0.7,
           position = position_dodge(width = 0.85)) +
  ## The crosshatching package is inconsistent between different R versions
  ## Manually adding back in in illustrator, original code the crosshatch worked with below
  # geom_bar_pattern(aes(pattern = diedever, fill = diedever, color = diedever), stat = "identity",
  #                  pattern_spacing = 0.06, width = 0.7,
  #                  pattern_density = 0.35, position = position_dodge(width = 0.85),
  #                   pattern_fill = "white", pattern_color = "white") +
  theme_classic() +
  scale_x_discrete(limits = rev) +
  scale_color_manual(values = c( "#9E5D10", "#9E5D10"), name = "TG4 >Day 28 \nOutcome") +
  scale_fill_manual(values = c("#9E5D10", "#ffffff00"), name = "TG4 >Day 28 \nOutcome") +
  #scale_pattern_discrete(choices = c("stripe", "none"), name = "TG4 >Day 28 \nOutcome") +
  xlab("Virus & Compartment") +
  ylab("Percent of Patients (%)") + stat_pvalue_manual(
  stat.df %>% filter(p.adj.med.signif != "ns") %>%
    mutate(y.position = c(36, 28, 33, 38, 45)),
  label = "p.adj.med.signif", hjust = -1.1, vjust = 1.5, tip.length = 0.01
  ) +
    coord_flip() +
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.text = element_text(color="black"),
        axis.ticks = element_line(color = "black"))
## 7.5 x 5.5
gg1 %>% insert_right(gg2) 

write_csv(gg1$data, paste0(out_dir, "source_data_Figure2A_part1.csv"))
write_csv(gg2$data, paste0(out_dir, "source_data_Figure2A_part2.csv"))
```

### Figure 2B: Age

```{r}
library(viridis)
library(rstatix)
library(ggpubr)

ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type, event_date,
                                     sex)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data %>% dplyr::select(participant_id, discretized_admit_age_quantile)) %>%
  ungroup() %>%
  filter(!is.na(discretized_admit_age_quantile)) %>%
  group_by(name, discretized_admit_age_quantile) %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  distinct() %>%
  dplyr::summarize(mean = mean(binary) * 100,
                   count = n(),
                   positive = mean * count /100)

clm_data <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type,
                                     event_date, sex)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data %>% dplyr::select(participant_id, discretized_admit_age_quantile, enrollment_site, trajectory_group)) %>%
  ungroup() %>%
  filter(!is.na(discretized_admit_age_quantile)) %>%
  filter(!is.na(enrollment_site)) %>%
  filter(!is.na(trajectory_group)) %>%
  distinct()

## Age Association Testing
stat.df <- data.frame()
library(ordinal)
for(active_virus in unique(clm_data$name)){
  tmp.chi <- ever_reactivate %>%
    filter(name == active_virus) %>%
    ungroup() %>%
    mutate(negative = count - positive) %>%
    select(discretized_admit_age_quantile, positive, negative) %>%
    column_to_rownames("discretized_admit_age_quantile")
  
  ## CLMM testing
  clm_data_tmp <- clm_data %>%
    filter(name == active_virus) %>%
    mutate(discretized_admit_age_quantile = ordered(factor(discretized_admit_age_quantile))) %>%
    mutate(trajectory_group = ordered(factor(trajectory_group))) %>%
    mutate(enrollment_site = factor(enrollment_site))
  clmm_base <- clmm2(discretized_admit_age_quantile~trajectory_group, data = clm_data_tmp, random = enrollment_site,
                    Hess = TRUE)
  clmm_res <- clmm2(discretized_admit_age_quantile~binary+trajectory_group, data = clm_data_tmp, random = enrollment_site,
                    Hess = TRUE)
  
  stat.df <- rbind.data.frame(stat.df, chisq_test(tmp.chi) %>% mutate(name = active_virus) %>%
                                mutate(ordinal.p = anova(clmm_res, clmm_base)$`Pr(Chi)`[2]))
}

stat.df <- stat.df %>%
  mutate(p.adj = p.adjust(ordinal.p, "fdr")) %>%
  add_significance("p.adj") %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  arrange(name) %>%
  mutate(xpos = 11:1) %>%
  mutate(group1 = "filler_text", group2 = "variable") %>%
  mutate(y.position = 40)
write.csv(stat.df, paste0(out_dir, "Supplemental_Data_1_part3.csv"))

gg2B <- ever_reactivate %>%
  mutate(discretized_admit_age_quantile = fct_rev(discretized_admit_age_quantile)) %>%
  ggplot(aes(x=name, y=mean)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"),
           aes(fill = discretized_admit_age_quantile)) +
  theme_classic() + scale_x_discrete(limits = rev(unique(ever_reactivate$name))) +
  ylab("% of Patients") +
  xlab("") +
    coord_flip() +
  viridis::scale_fill_viridis(discrete = TRUE, name = "Age Quantile") +
  stat_pvalue_manual(stat.df %>%
                       filter(!p.adj.signif %in% c("ns", "", " ")) %>%
                       mutate(y.position = c(35)),
                     x= "xpos", label = "p.adj.signif", vjust = 0.7) +
  theme(axis.text = element_text(color="black"),
        axis.ticks = element_line(color = "black"))

gg2B
write_csv(gg2B$data, paste0(out_dir, "source_data_Figure2B.csv"))
```

```{r, fig.width = 8, fig.height = 6}
gg1 %>% insert_right(gg2) %>% insert_right(gg2B)
ggsave(paste0(out_dir, "Figure2A_B.pdf"), width = 8, height = 6)
##10x5
```

### Figure 2C: Demographics, Complications, and Comorbidites

```{r}

ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type,
                                     event_date)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0))

```

```{r}
#| eval: false
### This chunk is very slow, only run it once and the results are already present
require(lme4)

ordered_v <- c("comorb_count", "comorb_countcat", "ncomp")
numeric <- c("")

model_data <- ever_reactivate %>%
  inner_join(clinical_data_ext) %>%
  left_join(clinical_data %>% distinct(participant_id, sex, 
                                       discretized_admit_age_quantile,
                                       trajectory_group,
                                       enrollment_site))

out_model <- data.frame()
out_rates <- data.frame()
out_rates_TG <- data.frame()
for(active_virus in unique(model_data$name)){
  print(paste0("Starting ", active_virus))
  for(active_var in colnames(clinical_data_ext)[grepl("comorb|comp|symptom|ever", colnames(clinical_data_ext))]){
    model_data.tmp <- model_data %>%
      filter(name == active_virus) %>%
      ungroup() %>%
      mutate(binary = factor(binary)) %>%
      mutate(sex = factor(sex),
             discretized_admit_age_quantile = ordered(factor(discretized_admit_age_quantile)),
             trajectory_group = ordered(factor(trajectory_group)))
    model_data.tmp[[active_var]] = factor(model_data.tmp[[active_var]])
    if(active_var %in% ordered_v){
      model_data.tmp[[active_var]] = ordered(model_data.tmp[[active_var]])
      }
    
    formula_use_cont <- formula(paste0(active_var, "~ binary + sex + discretized_admit_age_quantile + trajectory_group + (1 | enrollment_site)"))
    
    fit <- try(glmer(formula_use_cont, data = model_data.tmp,
       family = binomial,
      control = glmerControl(optimizer = "bobyqa"),
      nAGQ = 10), silent = TRUE)
    if(class(fit) == "try-error"){next}
    drop1_ress <- try(drop1(fit, test="Chisq"), silent = TRUE)
    if(class(drop1_ress) == "try-error"){next}
    
    ## Calculate the rate of the event in both groups
    tmp_rates <- model_data %>%
      filter(name == active_virus) %>%
      group_by(binary, .data[[active_var]]) %>%
      dplyr::summarize(count = n()) %>%
      filter(!is.na(.data[[active_var]])) %>%
      ungroup() %>%
      complete(binary, .data[[active_var]]) %>%
      group_by(binary) %>%
      mutate(count = ifelse(is.na(count), 0, count)) %>%
      mutate(frequency = count / sum(count) * 100,
             total = sum(count)) %>%
      mutate(Rate = paste0(round(frequency, 1), "% (", count, "/", total, ")")) %>%
      filter(.data[[active_var]] == 1) %>%
      mutate(binary = ifelse(binary == 1, "of Clinical Event in Virus Positive Participants",
                             "of Clinical Event in Virus Negative Participants")) %>%
      ungroup() %>%
      select(-count, -total) %>%
      select(-.data[[active_var]]) %>%
      mutate(`Clinical Variable` = active_var) %>%
      relocate(`Clinical Variable`) %>%
      pivot_wider(names_from = binary, values_from = c(Rate, frequency),
                  names_sep = " ") %>%
      rowwise() %>%
      mutate(`Percent Difference` = round(`frequency of Clinical Event in Virus Positive Participants` - `frequency of Clinical Event in Virus Negative Participants`, 1)) %>%
      select(c(1:3,6)) %>%
      mutate(virus = active_virus)
    
    ## Calculate the rate of the event in both groups by TG
    tmp_rates_TG <- model_data %>%
      filter(name == active_virus) %>%
      group_by(trajectory_group, binary, .data[[active_var]]) %>%
      dplyr::summarize(count = n()) %>%
      filter(!is.na(.data[[active_var]])) %>%
      ungroup() %>%
      complete(trajectory_group, binary, .data[[active_var]]) %>%
      group_by(trajectory_group, binary) %>%
      mutate(count = ifelse(is.na(count), 0, count)) %>%
      mutate(frequency = count / sum(count) * 100,
             total = sum(count)) %>%
      mutate(Rate = paste0(round(frequency, 1), "% (", count, "/", total, ")")) %>%
      filter(.data[[active_var]] == 1)  %>%
      mutate(binary = ifelse(binary == 1, "of Clinical Event in Virus Positive Participants",
                             "of Clinical Event in Virus Negative Participants")) %>%
      ungroup() %>%
      select(-count, -total) %>%
      select(-.data[[active_var]]) %>%
      mutate(`Clinical Variable` = active_var) %>%
      relocate(`Clinical Variable`) %>%
      pivot_wider(names_from = binary, values_from = c(Rate, frequency),
                  names_sep = " ") %>%
      rowwise() %>%
      mutate(`Percent Difference` = round(`frequency of Clinical Event in Virus Positive Participants` - `frequency of Clinical Event in Virus Negative Participants`, 1)) %>%
      select(c(1:4,7)) %>%
      mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
      pivot_wider(names_from = trajectory_group, values_from = c(`Rate of Clinical Event in Virus Positive Participants`,
                                                   `Rate of Clinical Event in Virus Negative Participants`,
                                                   `Percent Difference`))
    headers <- gsub(".*_", "", colnames(tmp_rates_TG))
    headers[1] <- ""
    colnames(tmp_rates_TG) <- paste0(headers, " ", gsub("_.*", "", colnames(tmp_rates_TG)))
    tmp_rates_TG <- tmp_rates_TG[, sort(colnames(tmp_rates_TG))]
    tmp_rates_TG$virus <- active_virus
    
    out_model <- rbind.data.frame(out_model,
                                drop1_ress %>%
                                  rownames_to_column("term") %>%
                                  mutate(virus = active_virus,
                                         feature = active_var,
                                         coef_virus = unique(coef(fit)$enrollment_site$binary1)))
    
    out_rates <- rbind.data.frame(out_rates, tmp_rates)
    out_rates_TG <- rbind.data.frame(out_rates_TG, tmp_rates_TG)
  }
}

write.csv(out_model %>% distinct(), paste0(out_dir, "Supplemental_Data_1_part4.csv"))
write.csv(out_rates, paste0(out_dir, "Supplemental_Data_1_part4_2.csv"))
write.csv(out_rates_TG, paste0(out_dir, "Supplemental_Data_1_part4_3.csv"))
## This is also reported in the source data (exact same)
```

```{r, width = 8.5, height = 4.3}
library(scales)

order_of_features <- c("CHF" = "comp_chf",
                       "STEMI" = "comp_stemi",
                       "Stroke" = "comp_stroke",
                       "Shock" = "comp_shock",
                       "Acute Venous Thromboembolism" = "comp_avthromb",
                       "Bacteremia" = "comp_bac",
                       "Other Infection" = "comp_othinf",
                       "Pneumothorax" = "comp_pneumo",
                       "Pulm Vascular Disease" = "comp_pulmvas",
                       "Liver Dysfunction/Failure" = "comp_liver",
                       "Liver Disease" = "comorb_isaric_liver",
                       "Renal" = "comp_renal",
                       "Organ/BM Transplant" = "comorb_hxtrans",
                       "Substance Use" = "comorb_anysubstance",
                       "Azithromycin" = "ever_azithro",
                       "Immunosupressed" = "ever_immsupp",
                       "Conv Plasma" =  "ever_conplas",
                       "Steroids" = "ever_steroids",
                       "Remdesivir" = "ever_remdes",
                        "ICU" = "ever_icu",
                       "Number of Complications" = "ncomp")

viral_order <- gsub(".*\\(", "", gsub("\\)", "", colnames(comb_data)[-1][order(colnames(comb_data)[-1])]))
out_model <- read_csv(paste0(out_dir, "Supplemental_Data_1_part4.csv")) %>% select(-1)
library(scales)
library(ggh4x)
adj_sd14 <- as.data.frame(out_model) %>%
  mutate(`Pr(Chi)` = as.numeric(`Pr(Chi)`)) %>%
  group_by(virus) %>%
  mutate(`Pr(Chi)` = p.adjust(`Pr(Chi)`, 'fdr')) %>%
  filter(term == "binary")
write.csv(adj_sd14, paste0(out_dir, "Supplemental_Data_1_part4_adj.csv"))

## 8.5 x 4
tmp <- out_model %>%
  mutate(`Pr(Chi)` = as.numeric(`Pr(Chi)`)) %>%
  filter(feature %in% order_of_features) %>%
   group_by(virus) %>%
  mutate(`Pr(Chi)` = p.adjust(`Pr(Chi)`, 'fdr')) %>%
   filter(term == "binary") %>%
  mutate(direction = (coef_virus/abs(coef_virus)),
         log10p = -log10(`Pr(Chi)`)) %>%
  mutate(signedlog10p = log10p * direction) %>%
  mutate(virus = gsub(".*\\(", "", gsub("\\)", "", virus))) %>%
  filter(!feature %in% c("anycomp", "diedever")) %>%
  left_join(as.data.frame(order_of_features) %>%
              rownames_to_column("Label") %>%
              dplyr::rename("feature" = 2)) %>%
  mutate(Label = factor(Label, levels = names(order_of_features))) %>%
  mutate(panel = gsub("_.*", "", feature)) %>%
  mutate(panel = ifelse(panel == "ncomp",
                        "Complications & Comorbidities", panel)) %>%
  mutate(panel = ifelse(panel == "comp", "Complications & Comorbidities", panel))%>%
  mutate(panel = ifelse(panel == "symptom", "Complications & Comorbidities", panel))%>%
  mutate(panel = ifelse(panel == "comorb", "Complications & Comorbidities", panel)) %>%
  mutate(panel = ifelse(feature %in% c("ever_ever_esc", "ever_icu", "ncomp"),
                        "Clinical Event", panel)) %>%
  filter(feature != "ever_ever_esc") %>%
  mutate(panel = ifelse(panel == "ever", "Medication", panel)) %>%
  ungroup() %>%
  filter(!grepl("Beta", virus)) 

gg2C <- tmp %>% ungroup() %>% filter(virus != "HSV2.nasal") %>%
  ggplot(aes(x=interaction(Label), y=virus)) +
  geom_point(aes(size = log10p, color = signedlog10p), shape = 21, fill = NA) +
  geom_point(data = . %>% filter(`Pr(Chi)` <= 0.05),
             aes(size = log10p, fill = signedlog10p), shape = 21) +
  theme_classic() +
  ylab("Virus and Compartment") +
  facet_grid(~panel,scales = "free_x", space = "free_x") +
  scale_y_discrete(limits = rev(unique(tmp$virus)[order(unique(tmp$virus))])[-1]) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_color_distiller(palette = "RdBu", limits = c(-4,4), oob=squish,
                        breaks = c(-3, -2, -1, 0, 1, 2, 3),
                        labels = c(-0.001, -0.01, -0.1, 1, 0.1, 0.01, 0.001),
                        name = "Signed\nadjusted p-value",
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black"))+
  scale_fill_distiller(palette = "RdBu", limits = c(-4,4), oob=squish,
                        breaks = c(-3, -2, -1, 0, 1, 2, 3),
                        labels = c(-0.001, -0.01, -0.1, 1, 0.1, 0.01, 0.001),
                        name = "Signed\nadjusted p-value",
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")) +
  theme(axis.title.x = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.text = element_text(color = "black")) +
  guides(size = guide_legend(title = "-Log10 adjusted\np-value")) 
gg2C
ggsave(paste0(out_dir, "Figure2C.pdf"), width = 8.5, height = 4.3)

write_csv(gg2C$data, paste0(out_dir, "source_data_Figure2C.csv"))
```

## Supplemental Figure 6

### Figure S6B: Ethnicity

```{r}
################ Ethinicity #########################
ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type,
                                     event_date, sex)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data %>% dplyr::select(participant_id, ethnicity)) %>%
  ungroup() %>%
  filter(!is.na(ethnicity)) %>%
  group_by(name, ethnicity) %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  distinct() %>%
  dplyr::summarize(mean = mean(binary) * 100,
                   count = n(),
                   positive = mean * count /100)

## Ethnicity Association Testing
stat.df <- data.frame()
for(active_virus in unique(ever_reactivate$name)){
  tmp.chi <- ever_reactivate %>%
    filter(ethnicity != "Not Specified") %>%
    filter(name == active_virus) %>%
    ungroup() %>%
    mutate(negative = count - positive) %>%
    select(ethnicity, positive, negative) %>%
    column_to_rownames("ethnicity")
  
  stat.df <- rbind.data.frame(stat.df, chisq_test(tmp.chi) %>% mutate(name = active_virus))
}

stat.df <- stat.df %>%
  mutate(p.adj = p.adjust(p, "fdr")) %>%
  add_significance("p.adj") %>%
  mutate(xpos = 11:1) %>%
  mutate(group1 = "filler_text", group2 = "variable") %>%
  mutate(y.position = 40)

write.csv(stat.df, paste0(out_dir, "Supplemental_Data_1_part5.csv"))

library(viridis)
ggS6B <- ever_reactivate %>%
  ggplot(aes(x=name, y=mean)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"), aes(fill = ethnicity)) +
  theme_classic() + scale_x_discrete(limits = rev(unique(ever_reactivate$name))) +
  xlab("Virus & Compartment") +
  ylab("Percent of Participants (%)") +
    coord_flip() +
  theme(axis.text = element_text(color = "black"),
        #axis.text.y = element_blank(),
        #axis.title.y = element_blank(),
       legend.position = "top",
        legend.direction = "vertical") +
  scale_fill_manual(values = c("#22a6a8", "#a82277", "lightgray"), name = "Ethnicity") +
  ## viridis::scale_fill_viridis(discrete = TRUE, name = "Ethnicity",direction = -1,) +
  stat_pvalue_manual(stat.df %>%
                       filter(!p.adj.signif %in% c("ns", "", " ")) %>%
                       mutate(y.position = c(13,13, 15, 40)),
                     x= "xpos", label = "p.adj.signif", vjust = 0.7)

```

### Figure S6C: Sex

```{r, fig.width = 8, fig.height = 5}

## S6A is from earlier in the markdown (in the severity chunk)

ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type,
                                     event_date, sex)) %>%
  filter(event_type %in% paste0("Visit ", 1:6))  %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data %>% dplyr::select(participant_id, sex)) %>%
  ungroup() %>%
  filter(!is.na(sex)) %>%
  group_by(name, sex) %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  distinct() %>%
  dplyr::summarize(mean = mean(binary) * 100,
                   count = n(),
                   positive = mean * count /100)

## Sex Association Testing
stat.df <- data.frame()
for(active_virus in unique(ever_reactivate$name)){
  tmp.chi <- ever_reactivate %>%
    filter(name == active_virus) %>%
    ungroup() %>%
    mutate(negative = count - positive) %>%
    select(sex, positive, negative) %>%
    column_to_rownames("sex")
  
  stat.df <- rbind.data.frame(stat.df, chisq_test(tmp.chi) %>% mutate(name = active_virus))
}


stat.df <- stat.df %>%
  mutate(p.adj = p.adjust(p, "fdr")) %>%
  add_significance("p.adj") %>%
  mutate(xpos = 11:1) %>%
  mutate(group1 = "filler_text", group2 = "variable") %>%
  mutate(y.position = 40)
## Nothing significant even before corrections
write.csv(stat.df, paste0(out_dir, "Supplemental_Data_1_part6.csv"))

library(viridis)
ggS6C <- ever_reactivate %>%
  ggplot(aes(x=name, y=mean)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"), aes(fill = sex)) +
  theme_classic() + scale_x_discrete(limits = rev(unique(ever_reactivate$name))) +
  ylab("Percent of Participants (%)") +
  xlab("Virus & Compartment") +
    coord_flip() +
  theme(axis.text = element_text(color = "black"),
        legend.direction = "vertical",
        legend.position = "top") +
  scale_fill_manual(values = c("darkred", "darkblue"), name = "Sex")

```

### Figure S6D-E: Treatments

```{r, fig.width = 5, fig.height = 5}
library(rstatix)

ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% distinct(event_id, participant_id, trajectory_group,
                                     event_date, event_type, enrollment_site)) %>%
  filter(event_type %in% paste0("Visit ", 1:6))  %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, trajectory_group, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data_ext %>% dplyr::select(participant_id, ever_steroids)) %>%
  ungroup() %>%
  filter(!is.na(trajectory_group)) %>%
  group_by(name, ever_steroids) %>%
  dplyr::summarize(mean = mean(binary) * 100,
                   count = n(),
                   positive = mean * count /100)

## Steroid Chi.square survival testing (not correcting for severity)
stat.df <- data.frame()
for(active_virus in unique(ever_reactivate$name)){
  tmp.chi <- ever_reactivate %>%
    filter(name == active_virus) %>%
    ungroup() %>%
    mutate(negative = count - positive) %>%
    select(ever_steroids, positive, negative) %>%
    column_to_rownames("ever_steroids")
  
  tmp_model_df <- model_df %>%
    filter(name == active_virus) %>%
    mutate(ever_steroids = factor(ever_steroids),
           binary = factor(binary))
  
  fit <- glmer(formula(diedever~ever_steroids+binary+(1|enrollment_site)), data = tmp_model_df, family = "binomial")
  
  aa <- car::Anova(fit, type = "III")
  
  stat.df <- rbind.data.frame(stat.df, chisq_test(tmp.chi) %>%
                                mutate(name = active_virus,
                           mod.p = aa[3,3]))
}

stat.df <- stat.df %>%
  mutate(p.adj = p.adjust(p, "fdr")) %>%
   mutate(mod.p.adj = p.adjust(mod.p, "fdr")) %>%
  add_significance("p.adj") %>%
  mutate(group1 = "Steroids",
         group2 = "No Steroids",
         `.y.` = "ever_steroids")
write.csv(stat.df, paste0(out_dir, "Supplemental_Data_1_part7.csv"))

gg3_steroids <- ever_reactivate %>%
  mutate(ever_steroids = factor(ever_steroids)) %>%
 # filter(grepl("implex|EBV|CMV", name)) %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  mutate(ever_steroids = ifelse(ever_steroids == 1, "Steroids", "No Steroids")) %>%
  mutate(ever_steroids = factor(ever_steroids, levels = c("No Steroids", "Steroids"))) %>%
  ggplot(aes(x=name, y=mean)) +
  geom_bar(aes(fill = ever_steroids), stat = "identity", position = "dodge") +
  theme_classic() +
  scale_x_discrete(limits = rev) +
  scale_fill_manual(values = c("#2F2D2E", "#E69A8DFF"), name = "Steroid Usage") +
  xlab("Virus & Compartment") +
  ylab("Percent of Participants (%)") +
    coord_flip() +
  theme(legend.position = "top",
        legend.direction = "vertical",
        axis.text = element_text(color="black"),
        axis.ticks = element_line(color = "black"))

gg3_steroids
```

```{r, fig.width = 5, fig.height = 5}
library(rstatix)

ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, trajectory_group,
                                     event_date, event_type)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, trajectory_group, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data_ext %>% dplyr::select(participant_id, ever_remdes)) %>%
  ungroup() %>%
  filter(!is.na(trajectory_group)) %>%
  group_by(name, ever_remdes) %>%
  dplyr::summarize(mean = mean(binary) * 100,
                   count = n(),
                   positive = mean * count /100)

## Steroid Chi.square survival testing (not correcting for severity)
stat.df <- data.frame()
for(active_virus in unique(ever_reactivate$name)){
  tmp.chi <- ever_reactivate %>%
    filter(name == active_virus) %>%
    ungroup() %>%
    mutate(negative = count - positive) %>%
    select(ever_remdes, positive, negative) %>%
    column_to_rownames("ever_remdes")
  
   tmp_model_df <- model_df %>%
    filter(name == active_virus) %>%
    mutate(ever_remdes = factor(ever_remdes),
           binary = factor(binary))
  
  fit <- glmer(formula(diedever~ever_remdes+binary+(1|enrollment_site)), data = tmp_model_df, family = "binomial")
  
  aa <- car::Anova(fit, type = "III")
  
  stat.df <- rbind.data.frame(stat.df, chisq_test(tmp.chi) %>%
                                mutate(name = active_virus,
                           mod.p = aa[3,3]))
}

stat.df <- stat.df %>%
  mutate(p.adj = p.adjust(p, "fdr")) %>%
   mutate(mod.p.adj = p.adjust(mod.p, "fdr")) %>%
  add_significance("p.adj") %>%
  mutate(group1 = "Remdesivir",
         group2 = "No Remdesivir",
         `.y.` = "ever_remdes")
write.csv(stat.df, paste0(out_dir, "Supplemental_Data_1_part8.csv"))

gg3_rem <- ever_reactivate %>%
  mutate(ever_remdes = factor(ever_remdes)) %>%
 # filter(grepl("implex|EBV|CMV", name)) %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  mutate(ever_remdes = ifelse(ever_remdes == 1, "Remdesivir", "No Remdesivir")) %>%
  mutate(ever_remdes = factor(ever_remdes, levels = c("No Remdesivir", "Remdesivir"))) %>%
  ggplot(aes(x=name, y=mean)) +
  geom_bar(aes(fill = ever_remdes), stat = "identity", position = "dodge") +
  theme_classic() +
  scale_x_discrete(limits = rev) +
  scale_fill_manual(values = c("#2F2D2E", "#99C24D"), name = "Remdesivir Usage") +
  xlab("Virus & Compartment") +
  ylab("Percent of Participants (%)") +
    coord_flip() +
  theme(legend.position = "top",
        legend.direction = "vertical",
        axis.text = element_text(color="black"),
        axis.ticks = element_line(color = "black"))

gg3_rem
```

```{r, fig.width = 10, fig.height = 12}

## 10 x 12
ggarrange(ggS6A + theme(legend.position = "top",
                        axis.title.y = element_blank()),
          ggS6B + theme(legend.position = "top",
                        axis.title.y = element_blank()),
          ggS6C + theme(legend.position = "top",
                        axis.title.y = element_blank()),
          gg3_steroids + theme(legend.position = "top",
                        axis.title.y = element_blank()) ,
          gg3_rem + theme(legend.position = "top",
                        axis.title.y = element_blank()),
          align = "hv",nrow = 2, ncol = 3, labels = c("A", "B", "C", "D", "E"), font.label = list(size=18))
ggsave(paste0(out_dir, "FigureS6.pdf"), width = 10, height = 12)
write_csv(ggS6A$data, paste0(out_dir, "source_data_FigureS6A.csv"))
write_csv(ggS6B$data, paste0(out_dir, "source_data_FigureS6B.csv"))
write_csv(ggS6C$data, paste0(out_dir, "source_data_FigureS6C.csv"))
write_csv(gg3_steroids$data, paste0(out_dir, "source_data_FigureS6D.csv"))
write_csv(gg3_rem$data, paste0(out_dir, "source_data_FigureS6E.csv"))
```

### Figure S6F: Percent with Chronic Immunosupression in Anelloviridae+ vs -

```{r, fig.height = 3, fig.width = 3}
core <- comb_data %>%
  left_join(clinical_data %>% distinct(event_id, participant_id, event_type,
                                       event_date)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id) %>%
  slice_max(Anelloviridae.pbmc) %>%
  filter(!is.na(Anelloviridae.pbmc)) %>%
  left_join(clinical_data_ext %>% distinct(participant_id, ever_immsupp)) %>%
  ungroup() %>% rowwise() %>%
  mutate(binary = Anelloviridae.pbmc > 0) %>% ungroup() %>%
  distinct(participant_id, binary, ever_immsupp) %>%
  group_by(ever_immsupp, binary) %>%
  dplyr::summarize(count = n()) %>% ungroup()

ggS6F <- core %>% group_by(binary) %>%
  mutate(freq = count / sum(count) * 100) %>%
  mutate(ever_immsupp = as.factor(ever_immsupp)) %>%
  ggplot(aes(x=binary, y= freq, fill = ever_immsupp)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(freq, 1), "%")),
            position = position_stack(vjust = 0.5),
            color = "black", size = 3) +
  theme_classic() +
  xlab("Anelloviridae in PBMC") +
  ylab("Percent of Participants (%)") +
  scale_fill_manual(values = c("darkgray", "cornflowerblue"))
## Anelloviridae are certainly associated with Immunosupression, but only comprise 17.19% of the patient positive during the acute period
ggS6F
ggsave(paste0(out_dir, "FigureS6F.pdf"), width = 3, height = 3)
write_csv(ggS6F$data, paste0(out_dir, "source_data_FigureS6F.csv"))
```

#### Supplemental Anelloviridae Modeling

```{r}
library(lme4)
model_df <- comb_data %>%
  left_join(clinical_data %>% distinct(event_id, participant_id,
                                       trajectory_group,
                                       event_type,
                                       enrollment_site,
                                       event_date,
                                       discretized_admit_age_quantile,
                                       sex)) %>%
  filter(event_type %in% paste0("Visit ", 1:6)) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id) %>%
  slice_max(Anelloviridae.pbmc) %>%
  filter(!is.na(Anelloviridae.pbmc)) %>%
  left_join(clinical_data_ext %>% distinct(participant_id, ever_immsupp)) %>%
  ungroup() %>% rowwise() %>%
  mutate(binary = Anelloviridae.pbmc > 0) %>% ungroup() %>%
  distinct(participant_id, binary, ever_immsupp, discretized_admit_age_quantile,
           trajectory_group, sex, enrollment_site) %>%
  mutate(trajectory_group = factor(trajectory_group),
         discretized_admit_age_quantile = factor(discretized_admit_age_quantile),
         sex = factor(sex),
         ever_immsupp = factor(ever_immsupp))

## Generalized linear model of all features
fit <- glmer(binary ~ ever_immsupp + trajectory_group + sex + discretized_admit_age_quantile + (1|enrollment_site),
           data = model_df, family = "binomial")

stat.df <- as.data.frame(car::Anova(fit, type = "III")) %>%
  rownames_to_column("feature")
write.csv(stat.df, paste0(out_dir, "Supplemental_Data_1_part9.csv"))

```

#### Subsampling of Visits to only Visits 1, 2, 4, and 6 for Figure 2A

```{r}

ever_reactivate <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type,
                                     event_date)) %>%
  filter(event_type %in% paste0("Visit ", c(1,2,4,6))) %>%
  filter(event_date <= 40) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0))

graph_data <- ever_reactivate %>%
  left_join(clinical_data %>% distinct(participant_id, trajectory_group)) %>%
  group_by(trajectory_group, name) %>%
  dplyr::summarize(freq = mean(binary) * 100,
                   count = n()) %>%
  filter(count > 10) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
   mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  ungroup() %>%
  complete(trajectory_group, name) %>%
  mutate(freq = ifelse(is.na(freq), 0, freq),
         count = ifelse(is.na(count), 0, count)) %>%
  mutate(positive = freq * count / 100) %>%
  group_by(name) %>%
  mutate(pos_freq = positive / sum(positive) *100)


clm_data <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  left_join(clinical_data %>% select(event_id, participant_id, event_type, sex)) %>%
  filter(event_type %in% paste0("Visit ", c(1,2,4,6))) %>%
  group_by(participant_id, name) %>%
  dplyr::summarize(max = max(value, na.rm = TRUE)) %>%
  filter(max != -Inf) %>%
  mutate(binary = ifelse(max > 0, 1, 0)) %>%
  left_join(clinical_data %>% dplyr::select(participant_id, discretized_admit_age_quantile, enrollment_site, trajectory_group)) %>%
  ungroup() %>%
  filter(!is.na(discretized_admit_age_quantile)) %>%
  filter(!is.na(enrollment_site)) %>%
  filter(!is.na(trajectory_group)) %>%
  distinct() %>%
   mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) 

## TG Association Testing
stat.df <- data.frame()
for(active_virus in unique(graph_data$name)){
  tmp.chi <- graph_data %>%
    filter(name == active_virus) %>%
    ungroup() %>%
    mutate(negative = count - positive) %>%
    select(trajectory_group, positive, negative) %>%
    column_to_rownames("trajectory_group")
  
  clm_data_tmp <- clm_data %>%
    filter(name == active_virus) %>%
    mutate(trajectory_group = ordered(factor(trajectory_group))) %>%
    mutate(enrollment_site = factor(enrollment_site))
  clmm_res <- clmm2(trajectory_group~binary, data = clm_data_tmp, random = enrollment_site,
                    Hess = TRUE)
  
  stat.df <- rbind.data.frame(stat.df, chisq_test(tmp.chi) %>%
                                mutate(name = active_virus,
                 clm.p = summary(clmm_res)[["coefficients"]][5,4]))
}

stat.df <- stat.df %>%
  mutate(p.value = ifelse(grepl("\\.ea", name), p, clm.p)) %>%
  mutate(p.adj = p.adjust(p.value, "fdr")) %>%
  add_significance("p.adj") %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
  arrange(name) %>%
  mutate(xpos = 11:1) %>%
  mutate(group1 = "filler_text", group2 = "variable") %>%
  mutate(y.position = 40)

gg1 <- graph_data %>%
  mutate(trajectory_group = factor(trajectory_group, levels = 5:1)) %>%
  ggplot(aes(x=name, y=freq)) +
  geom_bar(stat = "identity", position = position_dodge(preserve = "single"), aes(fill = trajectory_group)) +
  theme_classic() + scale_x_discrete(limits = rev(unique(graph_data$name))) +
  scale_fill_manual(values = TrajGrpCol, name = "Trajectory Group") +
  xlab("Virus and Compartment") +
  ylab(" ") +
  stat_pvalue_manual(stat.df %>%
                       filter(!p.adj.signif %in% c("ns", "", " ")) %>%
                       mutate(y.position = c(38, 11, 15, 19, 45, 28)),
                     x= "xpos", label = "p.adj.signif", vjust = 0.7) +
    coord_flip() +
  theme(#axis.text.y = element_blank(),
        #axis.title.y = element_blank(),
        axis.text = element_text(color="black"),
        axis.ticks = element_line(color = "black"))
gg1

```

# Session Info

```{r}
sessionInfo()
```
