--
title: "Developing machine learning models to predict long COVID outcomes using baseline clinical variables and viral load, titers"
output: 
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
---


### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
# suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ggnetwork"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "GSA"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))

# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```


# Re-running the analysis with feature selection within the training data and preserving 3 long COVID phenotypes in training and testing 

## load the data

```{r load the Data, echo=FALSE}
filename_1 <- "/scratch/nasal-transcriptomics/Naresh/data/MLmodels/temp_240115.rda"

# save(X_all, Y_all, nasal_viralload_copies_per_ul, clinical_data, clin_supp_viral_titers_data_1, clin_supp_viral_titers_data_2, clinical_supp_data_PRO_clust, file = filename_1)

load(filename_1)


```

## Split the data into training and testing

```{r Split the data into training and testing, echo=FALSE}

set.seed(100)
sampling_indices <- caret::createDataPartition(clin_supp_viral_titers_data_2$conv_cluster4, p = 0.7, list = FALSE, times = 15) %>%
  as.data.frame()


```



#### Perform feature selection using LASSO

```{r LASSO feature selection, echo=FALSE}

# X_test <- X_all[-indices$Resample1, ]
# Y_test <- Y_all[-indices$Resample1 ]

temp_data <- clin_supp_viral_titers_data_2 %>%
    select(participant_id, conv_cluster4)
temp_data$outcome <- factor(ifelse(temp_data$conv_cluster4 == 1, "Non_LongCOVID", "LongCOVID"))

sampling_lasso <- vector("list", length = 15)
sampling_lasso_features <- vector("list", length = 15)

for(m in 1:length(sampling_lasso)){
  idx_m <- sampling_indices[, m]
  
  # train data
  X_train <- X_all[idx_m, ]
  temp_data_train <- temp_data[idx_m, ]
  temp_data_train$outcome <- factor(ifelse(temp_data_train$conv_cluster4 == 1, "Non_LongCOVID", "LongCOVID"))
  Y_train <- temp_data_train$outcome
  
  # test data
  # temp_data_test <- temp_data[-idx_m, ]
  # temp_data_test$outcome <- factor(ifelse(temp_data_test$conv_cluster4 == 1, "Non_LongCOVID", "LongCOVID"))
  
  # Run cross-validation step 100 times
  set.seed(100)
  # AUC_vals <- NULL
  nonzero_coefs_lambda_min_all <- NULL
  for (k in 1:100) {
    cvglm_k <-
      glmnet::cv.glmnet(
        x = as.matrix(X_train),
        y = Y_train,
        family = "binomial",
        alpha = 1,
        type.measure = "auc",
        nfolds = 5,
        standardize = TRUE
      )
    # AUC_vals <- cbind(AUC_vals, cvglm_k$cvm)
    nonzero_coefs_lambda_min_k <-
      coef(cvglm_k, s = "lambda.min")[, 1] %>% .[. != 0]
    nonzero_coefs_lambda_min_k <-
      data.frame(feature = names(nonzero_coefs_lambda_min_k),
                 coef = nonzero_coefs_lambda_min_k) %>%
      mutate(iter = k)
    nonzero_coefs_lambda_min_all <-
      rbind(nonzero_coefs_lambda_min_all, nonzero_coefs_lambda_min_k) 
    }
    lasso_features_m <- nonzero_coefs_lambda_min_all %>% 
      group_by(feature) %>%
      summarise(count = n()) %>% 
      filter(feature != "(Intercept)") %>% 
      arrange(desc(count)) %>% 
      top_n(n = 15, wt = count)
    lasso_feature_sel_m <- lasso_features_m$feature
    sampling_lasso[[m]] <- lasso_features_m
    sampling_lasso_features[[m]] <- lasso_feature_sel_m  
  }

filename_1 <- "/scratch/nasal-transcriptomics/Naresh/data/MLmodels/temp_240702_lasso_feature_sel.rda"

# save(sampling_lasso, sampling_lasso_features, file = filename_1)



### do not re-run multiple times
### just 1 time

# sampling_lasso <- vector("list", length = 15)
# sampling_lasso_features <- vector("list", length = 15)
# 
# for(m in 1:length(sampling_lasso)){
#   idx_m <- sampling_indices[, m]
#   X_train <- X_all[idx_m, ]
#   Y_train <- Y_all[idx_m]
#   # Run cross-validation step 100 times
#   set.seed(100)
#   nonzero_coefs_lambda_min_all <- NULL
#   cvglm_k <-
#     glmnet::cv.glmnet(
#       x = as.matrix(X_train),
#       y = Y_train,
#       family = "binomial",
#       alpha = 1,
#       type.measure = "auc",
#       nfolds = 5,
#       standardize = TRUE
#     )
#     
#     nonzero_coefs_lambda_min_k <-
#       coef(cvglm_k, s = "lambda.min")[, 1] %>% .[. != 0]
#     lasso_features_m <- nonzero_coefs_lambda_min_k
#     lasso_feature_sel_m <- setdiff(names(lasso_features_m), "(Intercept)")
#     sampling_lasso[[m]] <- lasso_features_m
#     sampling_lasso_features[[m]] <- lasso_feature_sel_m  
#   }
# 





```


## train the models bGLM and NNET

```{r train the models, echo=FALSE}

tune_control <- caret::trainControl(
  method = "cv", # cross-validation
  number = 3, # with n folds 
  #index = createFolds(tr_treated$Id_clean), # fix the folds
  verboseIter = FALSE, # no training log
  allowParallel = FALSE, # FALSE for reproducible results
  returnResamp = "all", 
  summaryFunction = twoClassSummary, 
  classProbs = TRUE, 
  savePredictions = TRUE
)

tune_control_none <- caret::trainControl(
  method = "none", # cross-validation
  number = 1, # with n folds 
  #index = createFolds(tr_treated$Id_clean), # fix the folds
  verboseIter = FALSE, # no training log
  allowParallel = FALSE, # FALSE for reproducible results
  returnResamp = "all", 
  summaryFunction = twoClassSummary, 
  classProbs = TRUE, 
  savePredictions = TRUE
)



itr <- 15
nitr <- 15
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# Initialize a list to store models
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- sampling_indices[, itr]
  X_train <- X_all[idx_itr, ]
  Y_train <- temp_data$outcome[idx_itr]
  lasso_feature_sel <- sampling_lasso_features[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[-idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  # bGLM model
  
  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    # tuneGrid = final_grid_knn,
    method = "bayesglm",
    metric = "ROC",
    # preProcess = c("center", "scale")
    preProcess = NULL
  )
  
  Y_test <- temp_data$outcome[-idx_itr]
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
}

# save data for this model
bGLM_model_results <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

bGLM_m1 <- bGLM_model_results$model_list[["iteration_11"]]
var_imp_features_bGLM_m1 <- varImp(bGLM_m1)$importance %>%
  rownames_to_column(var = "feature")

X_train_temp <- bGLM_m1$trainingData %>%
  select(-.outcome)
Y_train_temp <- bGLM_m1$trainingData %>%
  pull(.outcome)
pred_train <- predict(bGLM_m1, newdata = X_train_temp, type = "prob") %>% dplyr::mutate(truth = Y_train_temp)
  
auroc_itr <- roc_auc(pred_train, truth, LongCOVID, event_level = "first")$.estimate
auprc_itr <- pr_auc(pred_train, truth, LongCOVID, event_level = "first")$.estimate
  
bGLM_model_results$auroc_itr_all[[11]]
bGLM_model_results$auprc_itr_all[[11]]

ROC_plot_all_itrs <- roc_data_itr_all %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = 1-specificity, y = sensitivity)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 12),
        text = element_text(size = 12),
        plot.title = element_text(size = 12, face = "bold"),
        legend.position = "none") +
  xlab("1-specificity") + 
  ylab("sensitivity") # +
  facet_wrap(~model)
  
PRC_plot_all_itrs <- pr_data_itr_all %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = recall, y = precision)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 12),
        text = element_text(size = 12),
        plot.title = element_text(size = 12, face = "bold"),
        legend.position = "none") +
  xlab("Recall") + 
  ylab("Precision") # +
  facet_wrap(~model)

PRC_plot_all_models



# NNet model

tune_control_none <- caret::trainControl(
  method = "none", # cross-validation
  number = 1, # with n folds 
  #index = createFolds(tr_treated$Id_clean), # fix the folds
  verboseIter = FALSE, # no training log
  allowParallel = FALSE, # FALSE for reproducible results
  returnResamp = "all", 
  summaryFunction = twoClassSummary, 
  classProbs = TRUE, 
  savePredictions = TRUE
)



itr <- 1
nitr <- 15
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# model_list <- vector("list", length = nitr)
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- sampling_indices[, itr]
  X_train <- X_all[idx_itr, ]
  Y_train <- temp_data$outcome[idx_itr]
  lasso_feature_sel <- sampling_lasso_features[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[-idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  # NNET model
  tune_grid1 <- expand.grid(
    size = c(4, 8, 16, 32), 
    decay = c(seq(0,1,0.1)),
    bag = TRUE)

  
  set.seed(100)
  model_tune1 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid1,
    method = "avNNet",
    metric = "ROC",
    nthread = 1,
    # preProcess = c("center", "scale")
    preProcess = NULL
  )
  
  # mean(model_tune1$resample$ROC) # 0.7138039
  
  final_grid <- expand.grid(size = model_tune1$bestTune$size,
                                 decay = model_tune1$bestTune$decay,
                                 bag = model_tune1$bestTune$bag)
  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    tuneGrid = final_grid,
    method = "avNNet",
    metric = "ROC",
    nthread = 1,
    # preProcess = c("center", "scale")
    preProcess = NULL
    )

  Y_test <- temp_data$outcome[-idx_itr]
  
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
  
}

mean(auroc_itr_all)
median(auroc_itr_all)
mean(auprc_itr_all)
median(auprc_itr_all)

# save data for this model
NNET_model_results <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

# save the data
filename_2 <- "/scratch/nasal-transcriptomics/Naresh/data/MLmodels/temp_240703_bGLM_NNET_results.rda"

# save(NNET_model_results, bGLM_model_results, file = filename_2)

NNET_m1 <- NNET_model_results$model_list[["iteration_11"]]
var_imp_features_NNET_m1 <- varImp(NNET_m1)$importance %>%
  rownames_to_column(var = "feature")


```
## train the models svmRadial

```{r train the models, echo=FALSE}

itr <- 1
nitr <- 15
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# model_list <- vector("list", length = nitr)
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- sampling_indices[, itr]
  X_train <- X_all[idx_itr, ]
  Y_train <- temp_data$outcome[idx_itr]
  lasso_feature_sel <- sampling_lasso_features[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[-idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  # svm Radial model
  # first compute sigma parameter
  sigma_svmRadial <- kernlab::sigest(as.matrix(X_train_select), na.action = na.omit, scaled = TRUE)
  # caret by default takes the mean of 90% and 10% values, mean(as.vector(sigma_svmRadial[-2]))
  sigma_svmRadial_2 <- c(as.vector(sigma_svmRadial), mean(as.vector(sigma_svmRadial[-2])))
  tune_grid1 <- expand.grid(C = seq(0.1, 3, by = 0.1),
                                   sigma = sigma_svmRadial_2)

  
  set.seed(100)
  model_tune1 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid1,
    method = "svmRadial",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
  )
  
  final_grid <- expand.grid(C =  model_tune1$bestTune$C,
                            sigma = model_tune1$bestTune$sigma)
  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    tuneGrid = final_grid,
    method = "svmRadial",
    verbose = TRUE,
    metric = "ROC",
    nthread = 1,
    preProcess = NULL
    )

  Y_test <- temp_data$outcome[-idx_itr]
  
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
  
}

mean(auroc_itr_all)
median(auroc_itr_all)
mean(auprc_itr_all)
median(auprc_itr_all)

# save data for this model
svmRadial_model_results <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

# save the data
filename_3 <- "/scratch/nasal-transcriptomics/Naresh/data/MLmodels/temp_240705_bGLM_NNET_svmRadial_results.rda"

# save(svmRadial_model_results, NNET_model_results, bGLM_model_results, file = filename_3)

```


## train the models XBG

```{r train the models, echo=FALSE}

itr <- 1
nitr <- 15
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# model_list <- vector("list", length = nitr)
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- sampling_indices[, itr]
  X_train <- X_all[idx_itr, ]
  Y_train <- temp_data$outcome[idx_itr]
  lasso_feature_sel <- sampling_lasso_features[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[-idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  # note to start nrounds from 200, as smaller learning rates result in errors so
  # big with lower starting points that they'll mess the scales
  tune_grid1 <- expand.grid(
    nrounds = seq(from = 100, to = 500, by = 50),
    eta = c(0.025, 0.05, 0.1, 0.2, 0.3),
    max_depth = c(2, 3, 4, 5),
    gamma = 0,
    colsample_bytree = 1,
    min_child_weight = 1,
    subsample = 1
  )
  

  
  set.seed(100)
  xgb_tune1 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid1,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    # preProcess = c("center", "scale")
  )
  
  tune_grid2 <- expand.grid(
    nrounds = xgb_tune1$bestTune$nrounds,
    eta = xgb_tune1$bestTune$eta,
    max_depth = xgb_tune1$bestTune$max_depth,
    gamma = 0,
    colsample_bytree = 1,
    min_child_weight = c(1, 2, 3, 4, 5),
    subsample = 1
  )
  
  set.seed(100)
  xgb_tune2 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid2,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    # preProcess = c("center", "scale")
  )
  

  tune_grid3 <- expand.grid(
    nrounds = xgb_tune1$bestTune$nrounds,
    eta = xgb_tune1$bestTune$eta,
    max_depth = xgb_tune1$bestTune$max_depth,
    gamma = 0,
    colsample_bytree = seq(from = 0.2, to = 1, by = 0.1),
    min_child_weight = xgb_tune2$bestTune$min_child_weight,
    subsample = c(0.25, 0.5, 0.75, 1.0)
  )
  
  set.seed(100)
  xgb_tune3 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid3,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    # preProcess = c("center", "scale")
  )
  
  tune_grid4 <- expand.grid(
    nrounds = xgb_tune1$bestTune$nrounds,
    eta = xgb_tune1$bestTune$eta,
    max_depth = xgb_tune1$bestTune$max_depth,
    gamma = c(0, 0.05, 0.1, 0.5, 0.7, 0.9, 1.0),
    colsample_bytree = xgb_tune3$bestTune$colsample_bytree,
    min_child_weight = xgb_tune2$bestTune$min_child_weight,
    subsample = xgb_tune3$bestTune$subsample
  )
  
  set.seed(100)
  xgb_tune4 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid4,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    # preProcess = c("center", "scale")
  )
  
  final_grid_xgb <- expand.grid(
    nrounds = xgb_tune1$bestTune$nrounds,
    eta = xgb_tune1$bestTune$eta,
    max_depth = xgb_tune1$bestTune$max_depth,
    gamma = xgb_tune4$bestTune$gamma,
    colsample_bytree = xgb_tune3$bestTune$colsample_bytree,
    min_child_weight = xgb_tune2$bestTune$min_child_weight,
    subsample = xgb_tune3$bestTune$subsample
  )
  

  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    tuneGrid = final_grid_xgb,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    )

  Y_test <- temp_data$outcome[-idx_itr]
  
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
  
}

mean(auroc_itr_all)
median(auroc_itr_all)
mean(auprc_itr_all)
median(auprc_itr_all)

# save data for this model
XGB_model_results <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

# save the data
filename_4 <- "/scratch/nasal-transcriptomics/Naresh/data/MLmodels/temp_240705_bGLM_NNET_svmRadial_XGB_results.rda"

# save(XGB_model_results, svmRadial_model_results, NNET_model_results, bGLM_model_results, file = filename_4)

```


## train the models RF

```{r train the models, echo=FALSE}

itr <- 1
nitr <- 15
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# model_list <- vector("list", length = nitr)
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- sampling_indices[, itr]
  X_train <- X_all[idx_itr, ]
  Y_train <- temp_data$outcome[idx_itr]
  lasso_feature_sel <- sampling_lasso_features[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[-idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  tune_grid1 <- expand.grid(
    mtry = 1:10
  )
  
  
  set.seed(100)
  rf_tune1 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid1,
    method = "rf",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    # preProcess = c("center", "scale"),
    preProcess = NULL
  )
  
  # Iterate over a number of trees
  ntree_values <- c(100, 200, 300, 400, 500, 1000, 2000, 3000, 4000, 5000, 10000)

  # tune_grid <- expand.grid(.mtry = c(sqrt(ncol(dataset))))
  tune_grid <- expand.grid(.mtry = rf_tune1$bestTune)
  modellist_ntree <- list()
  
  #train with different ntree parameters
  for (ntree in ntree_values){
    set.seed(100)
    fit <- caret::train(
      x = X_train_select, 
      y = Y_train, 
      trControl = tune_control, 
      tuneGrid = tune_grid, 
      method = "rf", 
      verbose = TRUE, 
      nthread = 1, 
      ntree = ntree,
      # preProcess = c("center", "scale")
      preProcess = NULL
    )
    key <- toString(ntree)
    modellist_ntree[[key]] <- fit
  }
  
  # Aggregate resampling results
  rf_tune_results <- resamples(modellist_ntree)
  
  # Summarize results
  summary_rf_results <- summary(rf_tune_results)
  
  # Extract median ROC for each ntree value
  median_roc_values <- summary_rf_results$statistics$ROC[,"Median"]
  
  # Find the ntree value with the highest median ROC
  best_ntree <- names(median_roc_values)[which.max(median_roc_values)]
  best_ntree <- as.numeric(best_ntree)
  
  # Print the best ntree value and corresponding median ROC
  cat("Best ntree:", best_ntree, "\n")
  cat("Highest median ROC:", max(median_roc_values), "\n")

  final_grid_rf <- expand.grid(.mtry = rf_tune1$bestTune)
  
  
  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    tuneGrid = final_grid_rf,
    ntree = best_ntree,
    method = "rf",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    )

  Y_test <- temp_data$outcome[-idx_itr]
  
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
  
}

mean(auroc_itr_all)
median(auroc_itr_all)
mean(auprc_itr_all)
median(auprc_itr_all)

# save data for this model
RF_model_results <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

# save the data
filename_5 <- "/scratch/nasal-transcriptomics/Naresh/data/MLmodels/temp_240705_bGLM_NNET_svmRadial_XGB_RF_results.rda"

# save(RF_model_results, XGB_model_results, svmRadial_model_results, NNET_model_results, bGLM_model_results, file = filename_5)

all_model_results_list <- list(bGLM_model_results = bGLM_model_results,
                               NNET_model_results = NNET_model_results,
                               svmRadial_model_results = svmRadial_model_results,
                               XGB_model_results = XGB_model_results,
                               RF_model_results = RF_model_results)

# make data frames for plotting
# Initialize empty lists to store merged data
auroc_auprc_merged <- data.frame()
roc_data_merged <- data.frame()
pr_data_merged <- data.frame()

# Define model names
model_names <- c("bGLM", "NNET", "svmRadial", "XGB", "RF")

# Iterate through each model and combine the data
for (model in model_names) {
  # Extract the model results
  model_results <- all_model_results_list[[paste0(model, "_model_results")]]
  
  # Combine auroc_itr_all and auprc_itr_all with model information
  auroc_auprc_df <- data.frame(
    AUROC = model_results$auroc_itr_all,
    AUPRC = model_results$auprc_itr_all,
    iteration = 1:length(model_results$auroc_itr_all),
    Model = model
  )
  
  # Append to the merged data frame
  auroc_auprc_merged <- rbind(auroc_auprc_merged, auroc_auprc_df)
  
  # Combine roc_data_itr_all with model information
  roc_df <- model_results$roc_data_itr_all
  roc_df$Model <- model
  
  # Append to the merged data frame
  roc_data_merged <- rbind(roc_data_merged, roc_df)
  
  # Combine pr_data_itr_all with model information
  pr_df <- model_results$pr_data_itr_all
  pr_df$Model <- model
  
  # Append to the merged data frame
  pr_data_merged <- rbind(pr_data_merged, pr_df)
}

# Print the merged data frames
View(auroc_auprc_merged)
View(roc_data_merged)
View(pr_data_merged)

df_mean <- auroc_auprc_merged %>% 
  group_by(Model) %>%
  summarise(n = n(),
            AUROC_mean = mean(AUROC),
            AUROC_median = median(AUROC),
            AUPRC_mean = mean(AUPRC),
            AUPRC_median = median(AUPRC)) %>%
  mutate(AUROC_mean_text = round(AUROC_mean, 3),
         AUPRC_mean_text = round(AUPRC_mean, 3))



paletteer::paletteer_d("calecopal::superbloom2")
col_sel <- c("bGLM" = "#DE7424FF", "NNET" = "#496849FF", "RF" = "#F5CA37FF",
             "svmRadial" = "#654783FF", "XGB" = "#AD8D26FF")

# plotting AUROC and AUPRC as box plots
AUROC_plot1 <- auroc_auprc_merged %>%
  ggplot(aes(x = Model, y = AUROC)) +
  geom_boxplot(aes(color = Model), alpha = 1, lwd = 1.2) +
  geom_quasirandom(aes(color = Model), width = 0.2, alpha = 0.7, size = 1.5) +
  geom_point(data = df_mean, aes(x = Model, y = AUROC_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  # xlab("UMAP 1") +
  # ylab("UMAP 2") +
  scale_color_manual(values = col_sel) 


AUPRC_plot1 <- auroc_auprc_merged %>%
  ggplot(aes(x = Model, y = AUPRC)) +
  geom_boxplot(aes(color = Model), alpha = 1, lwd = 1.2) +
  geom_quasirandom(aes(color = Model), width = 0.2, alpha = 0.7, size = 1.5) +
  geom_point(data = df_mean, aes(x = Model, y = AUPRC_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  # xlab("UMAP 1") +
  # ylab("UMAP 2") +
  scale_color_manual(values = col_sel) 

AUROC_AUPRC_plot1 <- cowplot::plot_grid(AUROC_plot1, AUPRC_plot1, ncol = 1, labels = NULL, rel_widths = c(1, 1))

### Plotting ROC and PRC curves
ROC_plot1 <- roc_data_merged %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = 1-specificity, y = sensitivity)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  geom_text(data = df_mean, aes(x = 0.8, y = 0.2, label = AUROC_mean_text), size = 6) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  xlab("1-Specificity") +
  ylab("Sensitivity") +
  # scale_color_manual(values = col_sel) +
  geom_abline(slope = 1, intercept = 0, color = "black", linewidth = 1.5, linetype = "dashed") +
  scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  scale_x_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  facet_wrap(~Model, ncol = 5) +
  theme(legend.position = "none")

PRC_plot1 <- pr_data_merged %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = recall, y = precision)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  geom_text(data = df_mean, aes(x = 0.8, y = 0.2, label = AUPRC_mean_text), size = 6) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  xlab("Recall") +
  ylab("Precision") +
  # scale_color_manual(values = col_sel) +
  # geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1.5, linetype = "dashed") +
  scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  scale_x_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  facet_wrap(~Model, ncol = 5) +
  theme(legend.position = "none")

ROC_PRC_plot1 <- cowplot::plot_grid(ROC_plot1, PRC_plot1, ncol = 1, labels = NULL, rel_widths = c(1, 1))


### Feature importance data
# Initialize empty data frame to store feature importance data
feature_importance_merged <- data.frame()

# Define model names
model_names <- c("bGLM", "NNET", "svmRadial", "XGB", "RF")

# Iterate through each model and extract feature importance for each iteration
for (model in model_names) {
  # Extract the model list
  model_results <- all_model_results_list[[paste0(model, "_model_results")]]
  model_list <- model_results$model_list
  
  # Iterate over each model in the list
  for (iteration in 1:length(model_list)) {
    # Extract the fitted model for the current iteration
    fitted_model <- model_list[[iteration]]
    
    # Extract feature importance
    importance <- varImp(fitted_model, scale = TRUE)
    importance_df <- as.data.frame(importance$importance)
    
    # for XGB and RF models varImp output is different so adjust to be consistent with other models
    if(model == "XGB" | model == "RF"){
      importance_df <- importance_df %>%
        dplyr::rename(LongCOVID = Overall) %>%
        mutate(Non_LongCOVID = LongCOVID)
    }
    
    # Add model and iteration information
    importance_df$Feature <- rownames(importance_df)
    importance_df$Model <- model
    importance_df$Iteration <- iteration
    
    # Append to the merged data frame
    feature_importance_merged <- rbind(feature_importance_merged, importance_df)
  }
  
  # Print the model
  cat("Model finished", model, "\n")
}

# Print the merged feature importance data frame
View(feature_importance_merged)

feature_importance_mean <- feature_importance_merged %>%
  group_by(Model, Feature) %>%
  summarise(count = n(),
            mean_imp = mean(LongCOVID),
            median_imp = median(LongCOVID))

feature_importance_mean_NNET <- feature_importance_mean %>%
  filter(Model == "NNET") %>%
  dplyr::filter(count > 6 & median_imp > 0) %>%
  arrange(desc(median_imp))

# Convert Feature column to factor with levels in the order of descending median_imp
feature_importance_mean <- feature_importance_mean %>% mutate(Feature = factor(Feature, levels = rev(feature_importance_mean_NNET$Feature)))


# variable importance plot
feature_imp_dotplot_all_models <- feature_importance_mean %>%
  dplyr::filter(Model %in% c("XGB", "RF", "svmRadial", "bGLM", "NNET")) %>%
  dplyr::filter(count > 6) %>%
  dplyr::filter(Feature %in% feature_importance_mean_NNET$Feature) %>%
  ggplot(aes(x = Model, y = Feature)) +
  geom_point(aes(color = median_imp, size = median_imp)) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14),
        axis.text.y = element_text(size = 14),
        text = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold")) +
  scale_color_continuous(type = "viridis", name = "Score") +
  guides(size = "none") +
  xlab("") + 
  ylab("")



```


## Model predictions on training data

```{r Model predictions on training data, echo=FALSE}

model_pred_train_itr <- function(model_data = model_data){
  train_data <- model_data$trainingData
  X_train <- train_data %>%
    select(-.outcome)
  Y_train <- train_data %>%
    pull(.outcome)
  # compute probabilities
  pred_train <- predict(model_data, newdata = X_train, type = "prob") %>% dplyr::mutate(truth = Y_train)
  
  auroc_itr <- roc_auc(pred_train, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_train, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_train, truth, LongCOVID, event_level = "first") 
  roc_data_itr <- roc_curve(pred_train, truth, LongCOVID, event_level = "first")
  
  model_output <- list(auroc_itr = auroc_itr,
                       auprc_itr = auprc_itr,
                       pr_data_itr = pr_data_itr,
                       roc_data_itr = roc_data_itr)
  
  return(model_output)
}

tt = model_pred_train_itr(model_data = all_model_results_list$bGLM_model_results$model_list[[1]])$roc_data_itr
tt1 = model_pred_train_itr(model_data = all_model_results_list$bGLM_model_results$model_list[[11]])$roc_data_itr

model_pred_train_itr_all <- function(data_list = all_model_results_list$bGLM_model_results$model_list){
  nitr <- length(data_list)
  auroc_itr_all <- NULL
  auprc_itr_all <- NULL
  pr_data_itr_all <- NULL
  roc_data_itr_all <- NULL
  for(itr in 1:nitr){
    model_output_itr <- model_pred_train_itr(model_data = data_list[[itr]])
    
    auroc_itr <- model_output_itr$auroc_itr
    auprc_itr <- model_output_itr$auprc_itr
    
    pr_data_itr <- model_output_itr$pr_data_itr %>%
      mutate(iteration = itr)
    roc_data_itr <- model_output_itr$roc_data_itr %>%
      mutate(iteration = itr)
    
    auroc_itr_all <- c(auroc_itr_all, auroc_itr)
    auprc_itr_all <- c(auprc_itr_all, auprc_itr)
    
    roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
    pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
    
  }
  
  model_results <- list(auroc_itr_all = auroc_itr_all,
                        auprc_itr_all = auprc_itr_all,
                        roc_data_itr_all = roc_data_itr_all,
                        pr_data_itr_all = pr_data_itr_all)
  
  return(model_results)
}

bGLM_model_results_train <- model_pred_train_itr_all(data_list = all_model_results_list$bGLM_model_results$model_list)
NNET_model_results_train <- model_pred_train_itr_all(data_list = all_model_results_list$NNET_model_results$model_list)
svmRadial_model_results_train <- model_pred_train_itr_all(data_list = all_model_results_list$svmRadial_model_results$model_list)
XGB_model_results_train <- model_pred_train_itr_all(data_list = all_model_results_list$XGB_model_results$model_list)
RF_model_results_train <- model_pred_train_itr_all(data_list = all_model_results_list$RF_model_results$model_list)

all_model_results_list_train <- list(bGLM_model_results = bGLM_model_results_train,
                               NNET_model_results = NNET_model_results_train,
                               svmRadial_model_results = svmRadial_model_results_train,
                               XGB_model_results = XGB_model_results_train,
                               RF_model_results = RF_model_results_train)

# make data frames for plotting
# Initialize empty lists to store merged data
auroc_auprc_merged <- data.frame()
roc_data_merged <- data.frame()
pr_data_merged <- data.frame()

# Define model names
model_names <- c("bGLM", "NNET", "svmRadial", "XGB", "RF")

# Iterate through each model and combine the data
for (model in model_names) {
  # Extract the model results
  model_results <- all_model_results_list_train[[paste0(model, "_model_results")]]
  
  # Combine auroc_itr_all and auprc_itr_all with model information
  auroc_auprc_df <- data.frame(
    AUROC = model_results$auroc_itr_all,
    AUPRC = model_results$auprc_itr_all,
    iteration = 1:length(model_results$auroc_itr_all),
    Model = model
  )
  
  # Append to the merged data frame
  auroc_auprc_merged <- rbind(auroc_auprc_merged, auroc_auprc_df)
  
  # Combine roc_data_itr_all with model information
  roc_df <- model_results$roc_data_itr_all
  roc_df$Model <- model
  
  # Append to the merged data frame
  roc_data_merged <- rbind(roc_data_merged, roc_df)
  
  # Combine pr_data_itr_all with model information
  pr_df <- model_results$pr_data_itr_all
  pr_df$Model <- model
  
  # Append to the merged data frame
  pr_data_merged <- rbind(pr_data_merged, pr_df)
}

# Print the merged data frames
View(auroc_auprc_merged)
View(roc_data_merged)
View(pr_data_merged)

df_mean <- auroc_auprc_merged %>% 
  group_by(Model) %>%
  summarise(n = n(),
            AUROC_mean = mean(AUROC),
            AUROC_median = median(AUROC),
            AUPRC_mean = mean(AUPRC),
            AUPRC_median = median(AUPRC)) %>%
  mutate(AUROC_mean_text = round(AUROC_mean, 3),
         AUPRC_mean_text = round(AUPRC_mean, 3))



paletteer::paletteer_d("calecopal::superbloom2")
col_sel <- c("bGLM" = "#DE7424FF", "NNET" = "#496849FF", "RF" = "#F5CA37FF",
             "svmRadial" = "#654783FF", "XGB" = "#AD8D26FF")

# plotting AUROC and AUPRC as box plots
AUROC_plot1 <- auroc_auprc_merged %>%
  ggplot(aes(x = Model, y = AUROC)) +
  geom_boxplot(aes(color = Model), alpha = 1, lwd = 1.2) +
  geom_quasirandom(aes(color = Model), width = 0.2, alpha = 0.7, size = 1.5) +
  geom_point(data = df_mean, aes(x = Model, y = AUROC_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  # xlab("UMAP 1") +
  # ylab("UMAP 2") +
  scale_color_manual(values = col_sel) +
  scale_y_continuous(limits = c(0.6,1))


AUPRC_plot1 <- auroc_auprc_merged %>%
  ggplot(aes(x = Model, y = AUPRC)) +
  geom_boxplot(aes(color = Model), alpha = 1, lwd = 1.2) +
  geom_quasirandom(aes(color = Model), width = 0.2, alpha = 0.7, size = 1.5) +
  geom_point(data = df_mean, aes(x = Model, y = AUPRC_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  # xlab("UMAP 1") +
  # ylab("UMAP 2") +
  scale_color_manual(values = col_sel) +
  scale_y_continuous(limits = c(0.6,1))

AUROC_AUPRC_plot1 <- cowplot::plot_grid(AUROC_plot1, AUPRC_plot1, ncol = 1, labels = NULL, rel_widths = c(1, 1))

### Plotting ROC and PRC curves
ROC_plot1 <- roc_data_merged %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = 1-specificity, y = sensitivity)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  geom_text(data = df_mean, aes(x = 0.8, y = 0.2, label = AUROC_mean_text), size = 6) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  xlab("1-Specificity") +
  ylab("Sensitivity") +
  # scale_color_manual(values = col_sel) +
  geom_abline(slope = 1, intercept = 0, color = "black", linewidth = 1.5, linetype = "dashed") +
  scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  scale_x_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  facet_wrap(~Model, ncol = 5) +
  theme(legend.position = "none")

PRC_plot1 <- pr_data_merged %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = recall, y = precision)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  geom_text(data = df_mean, aes(x = 0.8, y = 0.2, label = AUPRC_mean_text), size = 6) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  xlab("Recall") +
  ylab("Precision") +
  # scale_color_manual(values = col_sel) +
  # geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1.5, linetype = "dashed") +
  scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  scale_x_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  facet_wrap(~Model, ncol = 5) +
  theme(legend.position = "none")

ROC_PRC_plot1 <- cowplot::plot_grid(ROC_plot1, PRC_plot1, ncol = 1, labels = NULL, rel_widths = c(1, 1))


#




```

## Find key features from the feature importance by univariate assessments by logistic regression
* LASSO selected features
* variable importance by logistic regression


```{r variable importance logistic regression, echo=FALSE}


# make sure rows are ordered correctly
all(rownames(X_all) == nasal_viralload_copies_per_ul$participant_id)

data_temp <- X_all %>%
  select(-N1_CT_au) %>%
  cbind(N1_CT_au = nasal_viralload_copies_per_ul$N1_CT_au) %>%
  mutate_if(is.integer, as.factor) %>% 
  cbind(Y_all)
  
LR_out <- NULL
# feature_names <- setdiff(colnames(data_temp), "Y_all")
feature_names <- feature_importance_mean_NNET$Feature
for (k in 1:length(feature_names)){
  feature_k <- feature_names[k]
  formula_k <- paste0("Y_all~", feature_k)
  LR_out_k <- broom::tidy(glm(formula_k, family = binomial(logit), data = data_temp), exponentiate = TRUE, conf.int = TRUE)[2,] %>%
    mutate(feature_name = feature_k)
  LR_out <- rbind(LR_out, LR_out_k)
}

LR_out2 <- LR_out %>%
  mutate(feature_name = factor(feature_name, levels = rev(feature_importance_mean_NNET$Feature)))
# forest plot

temp_plot <- LR_out2 %>%
  ggplot() + 
  geom_point(data = LR_out, aes(x = feature_name, y = estimate), size = 4, alpha = 1) +
  geom_errorbar(aes(x = feature_name, ymin = conf.low, ymax = conf.high), width = 0.4, colour = "black", alpha = 1, linewidth = 1) + 
  scale_y_continuous(trans='log', breaks=c(0.05, 0.5, 1, 10, 30), labels = c("0.05", "0.5", "1", "10", "30")) + 
  geom_hline(yintercept = 1, linetype="dashed", 
                color = "darkgrey", size = 0.8) +
  theme_bw() +
  ylab("Odds Ratio (95% CI)") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 16),
        axis.text.y = element_text(size = 16),
        text = element_text(size=18)) +
  coord_flip() # +
  # scale_y_continuous(trans='log', breaks=c(0.0625, 1, 16, 256), labels = c("0.0625", "1", "16", "256"))
  
temp_plot


```


### Trying cross-site analysis to see if models performances are comparable across sites.



```{r cross-site analysis, echo=FALSE}
df_site <- data.frame(participant_id = rownames(X_all), outcome = Y_all) %>%
  inner_join(clinical_data %>% select(participant_id, enrollment_site)) %>%
  distinct(participant_id, enrollment_site, .keep_all = TRUE)

table(df_site$enrollment_site)
table(df_site$enrollment_site, df_site$outcome)

# merged_sites <- c("Arizona", "Boston/BWH", "Case Western", "ISMMS (Mt Sinai)", 
#                   "OHSU (Oregon)", "OUHSC (Oklahoma)", "Stanford")

merged_sites <- c("Case Western", "ISMMS (Mt Sinai)", 
                  "OHSU (Oregon)", "OUHSC (Oklahoma)")

df_site_5_groups <- df_site %>%
  mutate(new_site = case_when(
    enrollment_site %in% merged_sites ~ "merged_sites",
    .default = enrollment_site
  ))

table(df_site_5_groups$new_site, df_site_5_groups$outcome)


# custom function to create folds based on site
groupKFold_custom <- function(group, k = length(unique(group)), seed_number = 100){
  g_unique <- unique(group)
  m <- length(g_unique)
  if (k > m) {
    stop("`k` should be less than ", m)
  }
  set.seed(seed_number)
  g_folds <- sample(k, size = m, replace = FALSE)
  out <- split(seq_along(group), g_folds[match(group, g_unique)])
  names(out) <- paste0("Fold", gsub(" ", "0", format(seq_along(out))))
  # lapply(out, function(z) seq_along(group)[-z])
  return(out)
}

# cc <- createFolds(Y_all, k = 2, list = TRUE, returnTrain = TRUE)
site_fold_idx <- groupKFold_custom(df_site$enrollment_site, k = length(unique(df_site$enrollment_site)), seed_number = 100)
# lapply(site_fold_idx, length)

site_fold_idx <- groupKFold_custom(df_site_5_groups$new_site, k = length(unique(df_site_5_groups$new_site)), seed_number = 100)

### Feature selection using lasso  

sampling_lasso_site <- vector("list", length = 12)
sampling_lasso_features_site <- vector("list", length = 12)

for(m in 1:length(sampling_lasso_site)){
  idx_m <- site_fold_idx[[m]]
  
  # train data
  X_train <- X_all[-idx_m, ]
  # temp_data_train <- temp_data[idx_m, ]
  # temp_data_train$outcome <- factor(ifelse(temp_data_train$conv_cluster4 == 1, "Non_LongCOVID", "LongCOVID"))
  # Y_train <- temp_data_train$outcome
  
  Y_train = Y_all[-idx_m]
  
  # test data
  # temp_data_test <- temp_data[-idx_m, ]
  # temp_data_test$outcome <- factor(ifelse(temp_data_test$conv_cluster4 == 1, "Non_LongCOVID", "LongCOVID"))
  
  # Run cross-validation step 100 times
  set.seed(100)
  # AUC_vals <- NULL
  nonzero_coefs_lambda_min_all <- NULL
  for (k in 1:100) {
    cvglm_k <-
      glmnet::cv.glmnet(
        x = as.matrix(X_train),
        y = Y_train,
        family = "binomial",
        alpha = 1,
        type.measure = "auc",
        nfolds = 5,
        standardize = TRUE
      )
    # AUC_vals <- cbind(AUC_vals, cvglm_k$cvm)
    nonzero_coefs_lambda_min_k <-
      coef(cvglm_k, s = "lambda.min")[, 1] %>% .[. != 0]
    nonzero_coefs_lambda_min_k <-
      data.frame(feature = names(nonzero_coefs_lambda_min_k),
                 coef = nonzero_coefs_lambda_min_k) %>%
      mutate(iter = k)
    nonzero_coefs_lambda_min_all <-
      rbind(nonzero_coefs_lambda_min_all, nonzero_coefs_lambda_min_k) 
    }
    lasso_features_m <- nonzero_coefs_lambda_min_all %>% 
      group_by(feature) %>%
      summarise(count = n()) %>% 
      filter(feature != "(Intercept)") %>% 
      arrange(desc(count)) %>% 
      top_n(n = 15, wt = count)
    lasso_feature_sel_m <- lasso_features_m$feature
    sampling_lasso_site[[m]] <- lasso_features_m
    sampling_lasso_features_site[[m]] <- lasso_feature_sel_m  
  }

filename_1_site <- "/scratch/nasal-transcriptomics/Naresh/data/MLmodels/temp_240717_lasso_feature_sel_site.rda"

# save(sampling_lasso_site, sampling_lasso_features_site, file = filename_1_site)




tune_control_site_cv <- caret::trainControl(
  # method = "repeatedcv", # cross-validation
  # number = 5, # with n folds
  # repeats = 3,
  #index = createFolds(tr_treated$Id_clean), # fix the folds
  index = site_fold_idx,
  # sampling = "rose",
  verboseIter = FALSE, # no training log
  allowParallel = FALSE, # FALSE for reproducible results
  returnResamp = "all", 
  summaryFunction = twoClassSummary, 
  classProbs = TRUE, 
  savePredictions = TRUE
)

###################
### bGLM model ####
###################

itr <- 1
nitr <- 12
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# Initialize a list to store models
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- site_fold_idx[[itr]]
  X_train <- X_all[-idx_itr, ]
  # Y_train <- temp_data$outcome[idx_itr]
  Y_train <- factor(Y_all[-idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  lasso_feature_sel <- sampling_lasso_features_site[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  # bGLM model
  
  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    # tuneGrid = final_grid_knn,
    method = "bayesglm",
    metric = "ROC",
    # preProcess = c("center", "scale")
    preProcess = NULL
  )
  
  # Y_test <- temp_data$outcome[-idx_itr]
  Y_test <- factor(Y_all[idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
}

# save data for this model
bGLM_model_results_site <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)
###################
### NNet model ####
###################

itr <- 1
nitr <- 12
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# model_list <- vector("list", length = nitr)
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- site_fold_idx[[itr]]
  X_train <- X_all[-idx_itr, ]
  # Y_train <- temp_data$outcome[idx_itr]
  Y_train <- factor(Y_all[-idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  lasso_feature_sel <- sampling_lasso_features_site[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  # NNET model
  tune_grid1 <- expand.grid(
    size = c(4, 8, 16, 32), 
    decay = c(seq(0,1,0.1)),
    bag = TRUE)

  
  set.seed(100)
  model_tune1 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid1,
    method = "avNNet",
    metric = "ROC",
    nthread = 1,
    # preProcess = c("center", "scale")
    preProcess = NULL
  )
  
  final_grid <- expand.grid(size = model_tune1$bestTune$size,
                                 decay = model_tune1$bestTune$decay,
                                 bag = model_tune1$bestTune$bag)
  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    tuneGrid = final_grid,
    method = "avNNet",
    metric = "ROC",
    nthread = 1,
    # preProcess = c("center", "scale")
    preProcess = NULL
    )

  Y_test <- factor(Y_all[idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
  
}

mean(auroc_itr_all)
median(auroc_itr_all)
mean(auprc_itr_all)
median(auprc_itr_all)

# save data for this model
NNET_model_results_site <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

#######################
### svmRadial #########
#######################

itr <- 1
nitr <- 12
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# model_list <- vector("list", length = nitr)
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- site_fold_idx[[itr]]
  X_train <- X_all[-idx_itr, ]
  # Y_train <- temp_data$outcome[idx_itr]
  Y_train <- factor(Y_all[-idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  lasso_feature_sel <- sampling_lasso_features_site[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  # svm Radial model
  # first compute sigma parameter
  sigma_svmRadial <- kernlab::sigest(as.matrix(X_train_select), na.action = na.omit, scaled = TRUE)
  # caret by default takes the mean of 90% and 10% values, mean(as.vector(sigma_svmRadial[-2]))
  sigma_svmRadial_2 <- c(as.vector(sigma_svmRadial), mean(as.vector(sigma_svmRadial[-2])))
  tune_grid1 <- expand.grid(C = seq(0.1, 3, by = 0.1),
                                   sigma = sigma_svmRadial_2)

  
  set.seed(100)
  model_tune1 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid1,
    method = "svmRadial",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
  )
  
  final_grid <- expand.grid(C =  model_tune1$bestTune$C,
                            sigma = model_tune1$bestTune$sigma)
  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    tuneGrid = final_grid,
    method = "svmRadial",
    verbose = TRUE,
    metric = "ROC",
    nthread = 1,
    preProcess = NULL
    )

  Y_test <- factor(Y_all[idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
  
}

mean(auroc_itr_all)
median(auroc_itr_all)
mean(auprc_itr_all)
median(auprc_itr_all)

# save data for this model
svmRadial_model_results_site <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

################
### XGB ########
################

itr <- 1
nitr <- 12
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# model_list <- vector("list", length = nitr)
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- site_fold_idx[[itr]]
  X_train <- X_all[-idx_itr, ]
  # Y_train <- temp_data$outcome[idx_itr]
  Y_train <- factor(Y_all[-idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  lasso_feature_sel <- sampling_lasso_features_site[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  # note to start nrounds from 200, as smaller learning rates result in errors so
  # big with lower starting points that they'll mess the scales
  tune_grid1 <- expand.grid(
    nrounds = seq(from = 100, to = 500, by = 50),
    eta = c(0.025, 0.05, 0.1, 0.2, 0.3),
    max_depth = c(2, 3, 4, 5),
    gamma = 0,
    colsample_bytree = 1,
    min_child_weight = 1,
    subsample = 1
  )
  

  
  set.seed(100)
  xgb_tune1 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid1,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    # preProcess = c("center", "scale")
  )
  
  tune_grid2 <- expand.grid(
    nrounds = xgb_tune1$bestTune$nrounds,
    eta = xgb_tune1$bestTune$eta,
    max_depth = xgb_tune1$bestTune$max_depth,
    gamma = 0,
    colsample_bytree = 1,
    min_child_weight = c(1, 2, 3, 4, 5),
    subsample = 1
  )
  
  set.seed(100)
  xgb_tune2 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid2,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    # preProcess = c("center", "scale")
  )
  

  tune_grid3 <- expand.grid(
    nrounds = xgb_tune1$bestTune$nrounds,
    eta = xgb_tune1$bestTune$eta,
    max_depth = xgb_tune1$bestTune$max_depth,
    gamma = 0,
    colsample_bytree = seq(from = 0.2, to = 1, by = 0.1),
    min_child_weight = xgb_tune2$bestTune$min_child_weight,
    subsample = c(0.25, 0.5, 0.75, 1.0)
  )
  
  set.seed(100)
  xgb_tune3 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid3,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    # preProcess = c("center", "scale")
  )
  
  tune_grid4 <- expand.grid(
    nrounds = xgb_tune1$bestTune$nrounds,
    eta = xgb_tune1$bestTune$eta,
    max_depth = xgb_tune1$bestTune$max_depth,
    gamma = c(0, 0.05, 0.1, 0.5, 0.7, 0.9, 1.0),
    colsample_bytree = xgb_tune3$bestTune$colsample_bytree,
    min_child_weight = xgb_tune2$bestTune$min_child_weight,
    subsample = xgb_tune3$bestTune$subsample
  )
  
  set.seed(100)
  xgb_tune4 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid4,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    # preProcess = c("center", "scale")
  )
  
  final_grid_xgb <- expand.grid(
    nrounds = xgb_tune1$bestTune$nrounds,
    eta = xgb_tune1$bestTune$eta,
    max_depth = xgb_tune1$bestTune$max_depth,
    gamma = xgb_tune4$bestTune$gamma,
    colsample_bytree = xgb_tune3$bestTune$colsample_bytree,
    min_child_weight = xgb_tune2$bestTune$min_child_weight,
    subsample = xgb_tune3$bestTune$subsample
  )
  

  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    tuneGrid = final_grid_xgb,
    method = "xgbTree",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    )

  Y_test <- factor(Y_all[idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
  
}

mean(auroc_itr_all)
median(auroc_itr_all)
mean(auprc_itr_all)
median(auprc_itr_all)

# save data for this model
XGB_model_results_site <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

################
### RF #########
################

itr <- 1
nitr <- 12
auroc_itr_all <- NULL
auprc_itr_all <- NULL
pr_data_itr_all <- NULL
roc_data_itr_all <- NULL
# model_list <- vector("list", length = nitr)
model_list <- list()

for(itr in 1:nitr){
  idx_itr <- site_fold_idx[[itr]]
  X_train <- X_all[-idx_itr, ]
  # Y_train <- temp_data$outcome[idx_itr]
  Y_train <- factor(Y_all[-idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  lasso_feature_sel <- sampling_lasso_features_site[[itr]]
  X_train_select <- X_train %>% 
    select(all_of(lasso_feature_sel)) 
  
  X_test <- X_all[idx_itr, ]
  X_test <- X_test %>%
    select(all_of(lasso_feature_sel))
  
  tune_grid1 <- expand.grid(
    mtry = 1:10
  )
  
  
  set.seed(100)
  rf_tune1 <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control,
    tuneGrid = tune_grid1,
    method = "rf",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    # preProcess = c("center", "scale"),
    preProcess = NULL
  )
  
  # Iterate over a number of trees
  ntree_values <- c(100, 200, 300, 400, 500, 1000, 2000, 3000, 4000, 5000, 10000)

  # tune_grid <- expand.grid(.mtry = c(sqrt(ncol(dataset))))
  tune_grid <- expand.grid(.mtry = rf_tune1$bestTune)
  modellist_ntree <- list()
  
  #train with different ntree parameters
  for (ntree in ntree_values){
    set.seed(100)
    fit <- caret::train(
      x = X_train_select, 
      y = Y_train, 
      trControl = tune_control, 
      tuneGrid = tune_grid, 
      method = "rf", 
      verbose = TRUE, 
      nthread = 1, 
      ntree = ntree,
      # preProcess = c("center", "scale")
      preProcess = NULL
    )
    key <- toString(ntree)
    modellist_ntree[[key]] <- fit
  }
  
  # Aggregate resampling results
  rf_tune_results <- resamples(modellist_ntree)
  
  # Summarize results
  summary_rf_results <- summary(rf_tune_results)
  
  # Extract median ROC for each ntree value
  median_roc_values <- summary_rf_results$statistics$ROC[,"Median"]
  
  # Find the ntree value with the highest median ROC
  best_ntree <- names(median_roc_values)[which.max(median_roc_values)]
  best_ntree <- as.numeric(best_ntree)
  
  # Print the best ntree value and corresponding median ROC
  cat("Best ntree:", best_ntree, "\n")
  cat("Highest median ROC:", max(median_roc_values), "\n")

  final_grid_rf <- expand.grid(.mtry = rf_tune1$bestTune)
  
  
  set.seed(100)
  model_final <- caret::train(
    x = X_train_select,
    y = Y_train,
    trControl = tune_control_none,
    tuneGrid = final_grid_rf,
    ntree = best_ntree,
    method = "rf",
    verbose = TRUE,
    metric = "ROC", 
    nthread = 1,
    preProcess = NULL
    )

  Y_test <- factor(Y_all[idx_itr], levels = c("LongCOVID", "Non_LongCOVID"))
  
  # compute probabilities
  pred_test <- predict(model_final, newdata = X_test, type = "prob") %>% dplyr::mutate(truth = Y_test)
  
  auroc_itr <- roc_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  auprc_itr <- pr_auc(pred_test, truth, LongCOVID, event_level = "first")$.estimate
  
  pr_data_itr <- pr_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  roc_data_itr <- roc_curve(pred_test, truth, LongCOVID, event_level = "first") %>%
    mutate(iteration = itr)
  
  auroc_itr_all <- c(auroc_itr_all, auroc_itr)
  auprc_itr_all <- c(auprc_itr_all, auprc_itr)
  
  roc_data_itr_all <- rbind(roc_data_itr_all, roc_data_itr)
  pr_data_itr_all <- rbind(pr_data_itr_all, pr_data_itr)
  
  # Add the model to the list
  model_list[[paste0("iteration_", itr)]] <- model_final
  
}

mean(auroc_itr_all)
median(auroc_itr_all)
mean(auprc_itr_all)
median(auprc_itr_all)

# save data for this model
RF_model_results_site <- list(auroc_itr_all = auroc_itr_all,
                           auprc_itr_all = auprc_itr_all,
                           roc_data_itr_all = roc_data_itr_all,
                           pr_data_itr_all = pr_data_itr_all,
                           model_list = model_list)

# save the data
filename_5_site <- "/scratch/nasal-transcriptomics/Naresh/data/MLmodels/temp_240717_bGLM_NNET_svmRadial_XGB_RF_results_cross_site.rda"

# save(RF_model_results_site, XGB_model_results_site, svmRadial_model_results_site, NNET_model_results_site, bGLM_model_results_site, file = filename_5_site)

all_model_results_list_site <- list(bGLM_model_results = bGLM_model_results_site,
                               NNET_model_results = NNET_model_results_site,
                               svmRadial_model_results = svmRadial_model_results_site,
                               XGB_model_results = XGB_model_results_site,
                               RF_model_results = RF_model_results_site)

# make data frames for plotting
# Initialize empty lists to store merged data
auroc_auprc_merged <- data.frame()
roc_data_merged <- data.frame()
pr_data_merged <- data.frame()

# Define model names
model_names <- c("bGLM", "NNET", "svmRadial", "XGB", "RF")

# Iterate through each model and combine the data
for (model in model_names) {
  # Extract the model results
  model_results <- all_model_results_list_site[[paste0(model, "_model_results")]]
  
  # Combine auroc_itr_all and auprc_itr_all with model information
  auroc_auprc_df <- data.frame(
    AUROC = model_results$auroc_itr_all,
    AUPRC = model_results$auprc_itr_all,
    iteration = 1:length(model_results$auroc_itr_all),
    Model = model
  )
  
  # Append to the merged data frame
  auroc_auprc_merged <- rbind(auroc_auprc_merged, auroc_auprc_df)
  
  # Combine roc_data_itr_all with model information
  roc_df <- model_results$roc_data_itr_all
  roc_df$Model <- model
  
  # Append to the merged data frame
  roc_data_merged <- rbind(roc_data_merged, roc_df)
  
  # Combine pr_data_itr_all with model information
  pr_df <- model_results$pr_data_itr_all
  pr_df$Model <- model
  
  # Append to the merged data frame
  pr_data_merged <- rbind(pr_data_merged, pr_df)
}

# Print the merged data frames
View(auroc_auprc_merged)
View(roc_data_merged)
View(pr_data_merged)

df_mean <- auroc_auprc_merged %>% 
  group_by(Model) %>%
  summarise(n = n(),
            AUROC_mean = mean(AUROC),
            AUROC_median = median(AUROC),
            AUPRC_mean = mean(AUPRC),
            AUPRC_median = median(AUPRC)) %>%
  mutate(AUROC_mean_text = round(AUROC_mean, 3),
         AUPRC_mean_text = round(AUPRC_mean, 3))



paletteer::paletteer_d("calecopal::superbloom2")
col_sel <- c("bGLM" = "#DE7424FF", "NNET" = "#496849FF", "RF" = "#F5CA37FF",
             "svmRadial" = "#654783FF", "XGB" = "#AD8D26FF")

# plotting AUROC and AUPRC as box plots
AUROC_plot1 <- auroc_auprc_merged %>%
  ggplot(aes(x = Model, y = AUROC)) +
  geom_boxplot(aes(color = Model), alpha = 1, lwd = 1.2) +
  geom_quasirandom(aes(color = Model), width = 0.2, alpha = 0.7, size = 1.5) +
  geom_point(data = df_mean, aes(x = Model, y = AUROC_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  # xlab("UMAP 1") +
  # ylab("UMAP 2") +
  scale_color_manual(values = col_sel) 


AUPRC_plot1 <- auroc_auprc_merged %>%
  ggplot(aes(x = Model, y = AUPRC)) +
  geom_boxplot(aes(color = Model), alpha = 1, lwd = 1.2) +
  geom_quasirandom(aes(color = Model), width = 0.2, alpha = 0.7, size = 1.5) +
  geom_point(data = df_mean, aes(x = Model, y = AUPRC_mean), color = "black", size = 5, alpha = 0.7) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  # xlab("UMAP 1") +
  # ylab("UMAP 2") +
  scale_color_manual(values = col_sel) 

AUROC_AUPRC_plot1 <- cowplot::plot_grid(AUROC_plot1, AUPRC_plot1, ncol = 1, labels = NULL, rel_widths = c(1, 1))

### Plotting ROC and PRC curves
ROC_plot1 <- roc_data_merged %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = 1-specificity, y = sensitivity)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  geom_text(data = df_mean, aes(x = 0.8, y = 0.2, label = AUROC_mean_text), size = 6) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  xlab("1-Specificity") +
  ylab("Sensitivity") +
  # scale_color_manual(values = col_sel) +
  geom_abline(slope = 1, intercept = 0, color = "black", linewidth = 1.5, linetype = "dashed") +
  scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  scale_x_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  facet_wrap(~Model, ncol = 5) +
  theme(legend.position = "none")

PRC_plot1 <- pr_data_merged %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = recall, y = precision)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  geom_text(data = df_mean, aes(x = 0.8, y = 0.2, label = AUPRC_mean_text), size = 6) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  xlab("Recall") +
  ylab("Precision") +
  # scale_color_manual(values = col_sel) +
  # geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1.5, linetype = "dashed") +
  scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  scale_x_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  facet_wrap(~Model, ncol = 5) +
  theme(legend.position = "none")

ROC_PRC_plot1 <- cowplot::plot_grid(ROC_plot1, PRC_plot1, ncol = 1, labels = NULL, rel_widths = c(1, 1))


### Plotting ROC and PRC curves
ROC_plot1 <- roc_data_merged %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = 1-specificity, y = sensitivity)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  geom_text(data = df_mean, aes(x = 0.8, y = 0.2, label = AUROC_mean_text), size = 6) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  xlab("1-Specificity") +
  ylab("Sensitivity") +
  # scale_color_manual(values = col_sel) +
  geom_abline(slope = 1, intercept = 0, color = "black", linewidth = 1.5, linetype = "dashed") +
  scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  scale_x_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  facet_wrap(~Model, ncol = 5) +
  theme(legend.position = "none")

PRC_plot1 <- pr_data_merged %>%
  mutate(iteration = factor(iteration)) %>%
  ggplot(aes(x = recall, y = precision)) +
  geom_path(aes(color = iteration), linewidth = 0.8) + 
  geom_text(data = df_mean, aes(x = 0.8, y = 0.2, label = AUPRC_mean_text), size = 6) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, size = 14),
        # axis.text.x = element_blank(),
        axis.text.y = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold"),
        axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        strip.text = element_text(size=14),
        legend.title=element_blank(),
        legend.text = element_text(size = 12)) +
  xlab("Recall") +
  ylab("Precision") +
  # scale_color_manual(values = col_sel) +
  # geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1.5, linetype = "dashed") +
  scale_y_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  scale_x_continuous(breaks=c(0, 0.2, 0.4, 0.6, 0.8, 1), labels = c("0", "0.2", "0.4", "0.6", "0.8", "1")) +
  facet_wrap(~Model, ncol = 5) +
  theme(legend.position = "none")

ROC_PRC_plot1 <- cowplot::plot_grid(ROC_plot1, PRC_plot1, ncol = 1, labels = NULL, rel_widths = c(1, 1))


### Feature importance data
# Initialize empty data frame to store feature importance data
feature_importance_merged <- data.frame()

# Define model names
model_names <- c("bGLM", "NNET", "svmRadial", "XGB", "RF")

# Iterate through each model and extract feature importance for each iteration
for (model in model_names) {
  # Extract the model list
  model_results <- all_model_results_list_site[[paste0(model, "_model_results")]]
  model_list <- model_results$model_list
  
  # Iterate over each model in the list
  for (iteration in 1:length(model_list)) {
    # Extract the fitted model for the current iteration
    fitted_model <- model_list[[iteration]]
    
    # Extract feature importance
    importance <- varImp(fitted_model, scale = TRUE)
    importance_df <- as.data.frame(importance$importance)
    
    # for XGB and RF models varImp output is different so adjust to be consistent with other models
    if(model == "XGB" | model == "RF"){
      importance_df <- importance_df %>%
        dplyr::rename(LongCOVID = Overall) %>%
        mutate(Non_LongCOVID = LongCOVID)
    }
    
    # Add model and iteration information
    importance_df$Feature <- rownames(importance_df)
    importance_df$Model <- model
    importance_df$Iteration <- iteration
    
    # Append to the merged data frame
    feature_importance_merged <- rbind(feature_importance_merged, importance_df)
  }
  
  # Print the model
  cat("Model finished", model, "\n")
}

# Print the merged feature importance data frame
View(feature_importance_merged)

feature_importance_mean <- feature_importance_merged %>%
  group_by(Model, Feature) %>%
  summarise(count = n(),
            mean_imp = mean(LongCOVID),
            median_imp = median(LongCOVID))

feature_importance_mean_NNET <- feature_importance_mean %>%
  filter(Model == "NNET") %>%
  dplyr::filter(count > 4 & median_imp > 0) %>%
  arrange(desc(median_imp))

# Convert Feature column to factor with levels in the order of descending median_imp
feature_importance_mean <- feature_importance_mean %>% mutate(Feature = factor(Feature, levels = rev(feature_importance_mean_NNET$Feature)))


# variable importance plot
feature_imp_dotplot_all_models <- feature_importance_mean %>%
  dplyr::filter(Model %in% c("XGB", "RF", "svmRadial", "bGLM", "NNET")) %>%
  dplyr::filter(count > 4) %>%
  dplyr::filter(Feature %in% feature_importance_mean_NNET$Feature) %>%
  ggplot(aes(x = Model, y = Feature)) +
  geom_point(aes(color = median_imp, size = median_imp)) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, size = 14),
        axis.text.y = element_text(size = 14),
        text = element_text(size = 14),
        plot.title = element_text(size = 14, face = "bold")) +
  scale_color_continuous(type = "viridis", name = "Score") +
  guides(size = "none") +
  xlab("") + 
  ylab("")

feature_imp_dotplot_all_models


```




