---
title: "Figure1"
format: html
editor: visual
---

## Figure 1

### PBMC Load

```{r}
library(tidyverse)
library(data.table)
library(viridis)
library(rstatix)
library(ggpubr)
library(corrplot)
library(ComplexHeatmap) 
TrajGrpCol <- c("1" = "#639A21", "2" = "#39828C", "3" = "#6371AD", "4" = "#BD7D31", "5" = "#9C3418")
out_dir <- "../output/"

viruses_of_interest <- c("Human gammaherpesvirus 4 (EBV)",
                         "Human betaherpesvirus 5 (CMV)",
                         "Anelloviridae",
                         "Enterovirus",
                         "Human alphaherpesvirus 2 (HSV2)",
                         "Human alphaherpesvirus 1 (HSV1)",
                         "Human gammaherpesvirus 8", 
                           "Respiratory syncytial virus",
                         "Human immunodeficiency virus 1", "Human betaherpesvirus 7",
                         "Human betaherpesvirus 6",
                               "Betacoronavirus")

Anelloviridae_genera <- c("Alphatorquevirus", "Betatorquevirus",
                         "Gammatorquevirus", "non-genus-specific reads in family Anelloviridae")

clinical_data_sample <- read_csv("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-sample-locked.csv")
fastq_metadata <- read.csv("/scratch/metagenomics/viral_reactivation/pbmc_fastq_metadata.csv")

pbmc_reports <- fread("/scratch/metagenomics/viral_reactivation/data/phase1_4_pbmc_transcriptomics_v2.csv.gz") %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  pivot_wider(id_cols = full_czid_id, names_from = name, values_from = nt_rpm) %>%
  full_join(fastq_metadata %>% distinct(core_specific_ID_filename, sample_id), by = c("full_czid_id" = "core_specific_ID_filename")) %>%
  relocate(sample_id) %>% select(-full_czid_id) %>%
  distinct()

pbmc_reports_subset <- pbmc_reports %>%
  inner_join(clinical_data_sample %>%
               select(sample_id, participant_id, event_id) %>%
               distinct(), by = "sample_id") %>%
  filter(!grepl("C", participant_id)) %>%
  relocate(sample_id, event_id, participant_id) %>%
  rowwise() %>% 
  mutate(Anelloviridae = sum(!!!syms(Anelloviridae_genera),na.rm = T)) %>%
  ungroup()

colnames(pbmc_reports_subset) <- gsub("Cytomegalovirus",
                                      "Human betaherpesvirus 5 (CMV)",
                         gsub("Lymphocryptovirus",
                              "Human gammaherpesvirus 4 (EBV)",
                                   colnames(pbmc_reports_subset)))

pbmc_reports_subset <- pbmc_reports_subset %>%
  select(c(sample_id, event_id, participant_id,  any_of(viruses_of_interest))) %>%
  group_by(event_id) %>%
  select(-sample_id, -participant_id) %>%
  replace(is.na(.), 0) %>%
  summarise_all("mean")

```

### Nasal Load

```{r}
nasal_metagenomics_metadata <- read_csv("/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv")
nasal_transcriptomics_metadata <- read_csv("/data/nasal-transcriptomics/current/nasal-transcriptomics-Metadata.csv")
## See the section Nasal Batch Check Below for the reason this batch is being excluded for HSV1
plate_3B_4 <- nasal_transcriptomics_metadata$sample_id[nasal_transcriptomics_metadata$phase == "3B" & nasal_transcriptomics_metadata$plate == "4"]
plate_3B_4 <- plate_3B_4[!is.na(plate_3B_4)]
nasal_metagenomics_rowfeature <- read_csv("/data/nasal-metagenomics/current/nasal-metagenomics-AllGenera-RowFeatures.csv", skip = 1)
nasal_viral_samples <- fread("/scratch/metagenomics/viral_reactivation/data/nasal_virus_samples_analyzed.csv.gz")

nasal_reports <- fread("/scratch/metagenomics/viral_reactivation/data/nasal_virus_transcriptomic_v1.csv.gz") %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  pivot_wider(id_cols = full_czid_id, names_from = name, values_from = nt_rpm) %>%
  inner_join(nasal_metagenomics_metadata %>%
              filter(core_specific_ID %in% nasal_viral_samples$samples) %>%
              filter(sample_status == "IMPACC sample assayed and passed QC") %>%
              filter(internal_control != "yes") %>%
              distinct(core_specific_ID, sample_id, phase, plate),
            by = c("full_czid_id" = "core_specific_ID")) %>%
  relocate(sample_id, phase, plate) %>% rowwise() %>%
  mutate(`Human alphaherpesvirus 1` = ifelse(sample_id %in% plate_3B_4,
                                             ifelse(`Human alphaherpesvirus 1` > 0, -Inf,
                                                    0), `Human alphaherpesvirus 1`),
        `Human alphaherpesvirus 2` = ifelse(sample_id %in% plate_3B_4,
                                            ifelse(`Human alphaherpesvirus 2` > 0, -Inf,
                                                    0), `Human alphaherpesvirus 2`)) %>%
  ungroup() %>%
  select(-full_czid_id, -phase, -plate) %>%
  distinct()

nasal_reports_subset <- nasal_reports %>%
  inner_join(clinical_data_sample %>%
               select(sample_id, participant_id, event_id) %>%
               distinct(), by = "sample_id") %>%
  filter(!grepl("C", participant_id)) %>%
  relocate(sample_id, event_id, participant_id) %>%
  rowwise() %>% 
  mutate(Anelloviridae = sum(!!!syms(Anelloviridae_genera),na.rm = T)) %>%
  ungroup()

colnames(nasal_reports_subset) <- gsub("Human betaherpesvirus 5",
                                      "Human betaherpesvirus 5 (CMV)",
                         gsub("Human gammaherpesvirus 4",
                              "Human gammaherpesvirus 4 (EBV)",
                        gsub("Human alphaherpesvirus 1",
                             "Human alphaherpesvirus 1 (HSV1)",
                             gsub("Human alphaherpesvirus 2",
                                  "Human alphaherpesvirus 2 (HSV2)",
                                  colnames(nasal_reports_subset)))))

nasal_reports_subset <- nasal_reports_subset %>%
  select(c(sample_id, event_id, participant_id,  any_of(viruses_of_interest))) %>%
  group_by(event_id) %>%
  select(-sample_id, -participant_id) %>%
  replace(is.na(.), 0) %>%
  replace(.==-Inf, NA) %>%
  summarise_all("mean")


###################### ADD IN THE CONVALESCENT
### Run once to generate file (rawfiles no longer on server)
# files <- c("/impacc-study/BRI/nasal-transcriptomics/rawfiles/phase4a/nasal-transcriptomics-phase4a-rawfiles-Metadata_2021-08-21.tsv",
#            "/impacc-study/BRI/nasal-transcriptomics/rawfiles/phase4b/nasal-transcriptomics-phase4b-rawfiles-Metadata_2021-08-21.tsv")
# fastq_metadata_nasal <- data.frame()
# for(i in files){
#     fastq_metadata_nasal <- plyr::rbind.fill(fastq_metadata_nasal, read_tsv(i) %>% mutate(site = "BRI"))
# }
#write.csv(fastq_metadata_nasal, "/scratch/metagenomics/viral_reactivation/nasal_transcriptomic_metadata_phase4_v2.csv")


fastq_metadata_nasal <- read_csv("/scratch/metagenomics/viral_reactivation/nasal_transcriptomic_metadata_phase4_v2.csv")

fastq_metadata_nasal <- fastq_metadata_nasal %>%
  dplyr::select(filename, sample_id) %>%
  mutate(filename = gsub("_R.*", "", filename)) %>%
  distinct()

nasal_long_convalescent <- fread("/scratch/metagenomics/viral_reactivation/data/phase4_nasal_transcriptomics.csv.gz", nThread = 6) %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  #filter(!grepl("CTR", czid_id)) %>%
  pivot_wider(id_cols = full_czid_id, names_from = name, values_from = nt_rpm) %>%
  full_join(fastq_metadata_nasal, by = c("full_czid_id" = "filename")) %>%
  relocate(sample_id) %>% dplyr::select(-full_czid_id) %>%
  filter(sample_id != "0865-00H4YJ00-001") %>%
  rowwise() %>% 
  mutate(Anelloviridae = sum(!!!syms(Anelloviridae_genera[-2:-3]), na.rm = T)) %>%
  ungroup()

nasal_conv_reports <- nasal_long_convalescent %>%
  left_join(clinical_data_sample %>% dplyr::select(sample_id, event_id, participant_id) %>% distinct()) %>%
  relocate(sample_id, event_id, participant_id) %>%
  select(-sample_id, participant_id) %>%
  filter(!is.na(event_id)) %>%
  replace(is.na(.), 0)

colnames(nasal_conv_reports) <- gsub("Human betaherpesvirus 5",
                                      "Human betaherpesvirus 5 (CMV)",
                         gsub("Human gammaherpesvirus 4",
                              "Human gammaherpesvirus 4 (EBV)",
                        gsub("Human alphaherpesvirus 1",
                             "Human alphaherpesvirus 1 (HSV1)",
                             gsub("Human alphaherpesvirus 2",
                                  "Human alphaherpesvirus 2 (HSV2)",
                                  colnames(nasal_conv_reports)))))

nasal_reports_subset2 <- nasal_conv_reports %>%
  select(c(event_id,  any_of(viruses_of_interest)))
nasal_reports_subset <- plyr::rbind.fill(nasal_reports_subset, nasal_reports_subset2)

```

### EA Load

```{r}
ea_metagenomics_metadata <- read_csv("/data/ea-metagenomics/current/ea-metagenomics-Metadata.csv")
ea_metagenomics_rowfeature <- read_csv("/data/ea-metagenomics/current/ea-metagenomics-AllGenera-RowFeatures.csv", skip = 1)
ea_viral_samples <- fread("/scratch/metagenomics/viral_reactivation/data/ea_virus_samples_analyzed.csv.gz")

ea_reports <- fread("/scratch/metagenomics/viral_reactivation/data/ea_virus_transcriptomic_v1.csv.gz") %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  pivot_wider(id_cols = full_czid_id, names_from = name, values_from = nt_rpm) %>%
  inner_join(ea_metagenomics_metadata %>%
              filter(core_specific_ID %in% ea_viral_samples$samples) %>%
              filter(sample_status == "IMPACC sample assayed and passed QC") %>%
              filter(internal_control != "yes") %>%
              distinct(core_specific_ID, sample_id),
            by = c("full_czid_id" = "core_specific_ID")) %>%
  relocate(sample_id) %>% select(-full_czid_id) %>%
  distinct()

ea_reports_subset <- ea_reports %>%
  inner_join(clinical_data_sample %>%
               select(sample_id, participant_id, event_id) %>%
               distinct(), by = "sample_id") %>%
  filter(!grepl("C", participant_id)) %>%
  relocate(sample_id, event_id, participant_id) %>%
  rowwise() %>% 
  mutate(Anelloviridae = sum(!!!syms(Anelloviridae_genera[-2:-3]),na.rm = T)) %>%
  ungroup()

colnames(ea_reports_subset) <- gsub("Human betaherpesvirus 5",
                                      "Human betaherpesvirus 5 (CMV)",
                         gsub("Human gammaherpesvirus 4",
                              "Human gammaherpesvirus 4 (EBV)",
                        gsub("Human alphaherpesvirus 1",
                             "Human alphaherpesvirus 1 (HSV1)",
                             gsub("Human alphaherpesvirus 2",
                                  "Human alphaherpesvirus 2 (HSV2)",
                                  colnames(ea_reports_subset)))))

ea_reports_subset <- ea_reports_subset %>%
  select(c(sample_id, event_id, participant_id,  any_of(viruses_of_interest))) %>%
  group_by(event_id) %>%
  select(-sample_id, -participant_id) %>%
  replace(is.na(.), 0) %>%
  summarise_all("mean")


```

Stitch together a simplified file of the viral loads across their respective omics.

```{r}
## Using event_ids as the identifier
full_comb_data <- pbmc_reports_subset %>%
  rename_with(~paste0(., ".pbmc"), -event_id) %>%
  full_join(nasal_reports_subset %>%
    rename_with(~paste0(., ".nasal"), -event_id),
  by = c("event_id")) %>%
  left_join(ea_reports_subset %>%
    rename_with(~paste0(., ".ea"), -event_id),
  by = c("event_id")) 

HSV1_2_Ratio <- full_comb_data %>%
  filter(!is.na(`Human alphaherpesvirus 1 (HSV1).nasal`)) %>%
  mutate(HSV1_2Ratio = `Human alphaherpesvirus 1 (HSV1).nasal` / (`Human alphaherpesvirus 1 (HSV1).nasal` + `Human alphaherpesvirus 2 (HSV2).nasal`)) %>%
  ggplot(aes(x=HSV1_2Ratio)) +
  geom_histogram(bins = 100) +
  theme_classic()
HSV1_2_Ratio
ggsave(paste0(out_dir, "HSV1_2Ratio_nasal.pdf"))

HSV1_2_Ratio <- full_comb_data %>%
  filter(!is.na(`Human alphaherpesvirus 1 (HSV1).ea`)) %>%
  mutate(HSV1_2Ratio = `Human alphaherpesvirus 1 (HSV1).ea` / (`Human alphaherpesvirus 1 (HSV1).ea` + `Human alphaherpesvirus 2 (HSV2).ea`)) %>%
  ggplot(aes(x=HSV1_2Ratio)) +
  geom_histogram(bins = 100) +
  theme_classic()
HSV1_2_Ratio
ggsave(paste0(out_dir, "HSV1_2Ratio_ea.pdf"))

## Attribute all HSV2 reads to HSV1 if the ratio is high enough (i.e. HSV1 > HSV2,
## as although primative, the histogram above demonstrates a clear separation)
## Doing for both the nasal and ea
for(i in 1:nrow(full_comb_data)){
  if(is.na(full_comb_data$`Human alphaherpesvirus 1 (HSV1).nasal`[i]) | is.na(full_comb_data$`Human alphaherpesvirus 2 (HSV2).nasal`)[i]){
    next
  }
  if(full_comb_data$`Human alphaherpesvirus 1 (HSV1).nasal`[i] > full_comb_data$`Human alphaherpesvirus 2 (HSV2).nasal`[i]){
    full_comb_data$`Human alphaherpesvirus 1 (HSV1).nasal`[i] <- full_comb_data$`Human alphaherpesvirus 1 (HSV1).nasal`[i] +
      full_comb_data$`Human alphaherpesvirus 2 (HSV2).nasal`[i]
     full_comb_data$`Human alphaherpesvirus 2 (HSV2).nasal`[i] <-0
  } else if (full_comb_data$`Human alphaherpesvirus 1 (HSV1).nasal`[i] < full_comb_data$`Human alphaherpesvirus 2 (HSV2).nasal`[i]){
     full_comb_data$`Human alphaherpesvirus 2 (HSV2).nasal`[i] <- full_comb_data$`Human alphaherpesvirus 2 (HSV2).nasal`[i] +
      full_comb_data$`Human alphaherpesvirus 1 (HSV1).nasal`[i]
     full_comb_data$`Human alphaherpesvirus 1 (HSV1).nasal`[i] <-0
  }
}

for(i in 1:nrow(full_comb_data)){
  if(is.na(full_comb_data$`Human alphaherpesvirus 1 (HSV1).ea`[i]) | is.na(full_comb_data$`Human alphaherpesvirus 2 (HSV2).ea`)[i]){
    next
  }
  if(full_comb_data$`Human alphaherpesvirus 1 (HSV1).ea`[i] > full_comb_data$`Human alphaherpesvirus 2 (HSV2).ea`[i]){
    full_comb_data$`Human alphaherpesvirus 1 (HSV1).ea`[i] <- full_comb_data$`Human alphaherpesvirus 1 (HSV1).ea`[i] +
      full_comb_data$`Human alphaherpesvirus 2 (HSV2).ea`[i]
     full_comb_data$`Human alphaherpesvirus 2 (HSV2).ea`[i] <-0
  } else if (full_comb_data$`Human alphaherpesvirus 1 (HSV1).ea`[i] < full_comb_data$`Human alphaherpesvirus 2 (HSV2).ea`[i]){
     full_comb_data$`Human alphaherpesvirus 2 (HSV2).ea`[i] <- full_comb_data$`Human alphaherpesvirus 2 (HSV2).ea`[i] +
      full_comb_data$`Human alphaherpesvirus 1 (HSV1).ea`[i]
     full_comb_data$`Human alphaherpesvirus 1 (HSV1).ea`[i] <-0
  }
}

colnames(full_comb_data) <- gsub("Betacoronavirus", "SARS-CoV-2", colnames(full_comb_data))
write.csv(full_comb_data, "/scratch/metagenomics/viral_reactivation/full_comb_data.csv")
## This is the key file used for all subsequent figures reporting the viral load on a event_id basis

```

Do one final load of the key dataframes (this will be the quicker boot up used for other Figure's code)

```{r}
library(tidyverse)
library(data.table)
library(viridis)
library(rstatix)
library(ggpubr)
TrajGrpCol <- c("1" = "#639A21", "2" = "#39828C", "3" = "#6371AD", "4" = "#BD7D31", "5" = "#9C3418")

comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1)

data_env <- readRDS("/scratch/metagenomics/viral_reactivation/2024-02-02_obj.RDS")
clinical_data <- data_env$clinical_data
nasal_viralload_counts <- data_env$nasal_viralload_counts %>%
  rownames_to_column("sample_id") %>%
  inner_join(clinical_data %>% distinct(sample_id, event_id)) %>%
  dplyr::select(-sample_id) %>%
  column_to_rownames("event_id")
rm(data_env)
clinical_data_ext <- read_csv("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-supp-locked.csv")

out_dir <- "../output/"

## This is the list of participants used in the study, and the ones used to generate 
## The supplemental clinical table.
## Hashed out, as only need to run one time
# write_csv(comb_data %>% left_join(clinical_data %>%
#                                     distinct(event_id, participant_id)) %>%
#             filter(!is.na(participant_id)) %>% distinct(participant_id),
#           "/scratch/metagenomics/viral_reactivation/final_participant_list.csv")
```

## Figure 1B: Biospecimens across visits and omics

```{r, fig.width = 5, fig.height = 2.5}

## From the combined_data file (comb_data), NAs indicate they weren't collected.
## So we can recapitulate the number of specimens
comb_data_tmp <- comb_data %>%
  select(-starts_with("Human betaherpesvirus 6"),
         -starts_with("Human gammaherpesvirus 8"),
         -starts_with("Human betaherpesvirus 7"),
         -starts_with("Respiratory syncytial virus"),
         -starts_with("Human immunodeficiency virus 1"),
         -starts_with("SARS-CoV-2.pbmc"),
         -starts_with("Enterovirus.pbmc"),
         -starts_with("Enterovirus.ea"),
         -starts_with("Anelloviridae.ea"),
         -starts_with("Anelloviridae.nasal"))

count_data <- comb_data_tmp %>%
  mutate(nasal_total = rowSums(select(., ends_with('.nasal'))))%>%
  mutate(pbmc_total = rowSums(select(., ends_with('.pbmc'))))%>%
  mutate(ea_total = rowSums(select(., ends_with('.ea')))) %>%
  select(event_id, nasal_total, pbmc_total, ea_total) %>%
  left_join(clinical_data %>% distinct(event_id, participant_id, event_type)) %>%
  mutate(nasal_total = ifelse(!is.na(nasal_total), 1, 0))%>%
  mutate(pbmc_total = ifelse(!is.na(pbmc_total), 1, 0))%>%
  mutate(ea_total = ifelse(!is.na(ea_total), 1, 0)) %>%
  filter(!is.na(participant_id)) %>%
  group_by(event_type) %>%
  dplyr::summarize(Nasal = sum(nasal_total),
                   PBMC = sum(pbmc_total),
                   EA = sum(ea_total)) %>%
  pivot_longer(cols = -event_type) %>%
  filter(!event_type %in% c("Escalation 1", "Escalation 2")) %>%
  mutate(event_type = factor(event_type, levels = paste0('Visit ', 1:10))) %>%
  mutate(name = factor(name, c("EA", "Nasal", "PBMC"))) %>%
  mutate(percent = value / 1154 * 100)

gg1 <- count_data %>%
  mutate(xmin = match(event_type, levels(count_data$event_type)) - 0.5,
         xmax = match(event_type, levels(count_data$event_type)) + ((percent/100)-0.5) ,
         ymin = match(name, levels(count_data$name)) - 0.5,
         ymax = match(name, levels(count_data$name)) + 0.5) %>%
  ggplot(aes(x=event_type, y=name)) +
  geom_tile(fill = NA, color = "black") +
   geom_rect(aes(xmin=xmin, xmax = xmax,
                ymin = ymin, ymax = ymax), fill = "lightgray",
             color = NA) +
  geom_tile(fill = NA, color = "black") +
  theme_classic() +
  theme(legend.position = "none") +
    geom_text(aes(label=value)) +
  scale_fill_viridis(option = "C") +
 # xlab("Event Type") +
  ylab("Transcriptomic Assay") +
  theme(#axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank())
gg1

visit_annotation <- data.frame(event_type = levels(count_data$event_type),
                               location = c(rep("Inpatient",3),"Inpatient & Outpatient",
                                            "Inpatient", "Inpatient & Outpatient", rep("Outpatient", 4)))
gg2 <- visit_annotation %>%
  mutate(event_type = gsub("Escalation", "Esc.", event_type)) %>%
  mutate(event_type = factor(event_type, c(paste0("Visit ", 1:10)))) %>%
  ggplot(aes(x=event_type, y= 1)) +
  geom_tile(aes(fill = location), color = "black") +
  theme_void() +
  scale_fill_manual(values = c("#0E7A1B", "#35D34F", "#C7EFCD")) +
theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
      axis.title.x = element_text()) +
  xlab("Event Type")

## 5x2.5
ggpubr::ggarrange(gg1, gg2, nrow = 2, ncol = 1,common.legend = T,
                  align = "v", heights = c(6,3), legend = "bottom")
ggsave(paste0(out_dir, "Figure1B.pdf"), width = 5, height = 2.5)
```

## Figure 1C: Heatmap

```{r, fig.width =18, fig.height =7}

cutoff = 0

pbmc_percents <- pbmc_reports_subset %>%
  pivot_longer(cols = -event_id) %>%
  inner_join(clinical_data %>% distinct(participant_id, event_id, event_type)) %>%
  mutate(value = ifelse(is.na(value), 0, value)) %>%
  mutate(value = ifelse(value > cutoff, 1, 0)) %>%
  group_by(event_type, name) %>%
  dplyr::summarize(percent_positive = mean(value) * 100,
                   total = n(),
                   label = paste0(total*percent_positive/100, "/",total)) %>%
  mutate(assay = "PBMC")

nasal_percents <- nasal_reports_subset %>%
  pivot_longer(cols = -event_id) %>%
  inner_join(clinical_data %>% distinct(participant_id, event_id, event_type)) %>%
  mutate(value = ifelse(is.na(value), 0, value)) %>%
  mutate(value = ifelse(value > cutoff, 1, 0)) %>%
  group_by(event_type, name) %>%
  dplyr::summarize(percent_positive = mean(value) * 100,
                   total = n(),
                   label = paste0(total*percent_positive/100, "/",total)) %>%
  mutate(assay = "Nasal")

ea_percents <- ea_reports_subset %>%
  pivot_longer(cols = -event_id) %>%
  inner_join(clinical_data %>% distinct(participant_id, event_id, event_type)) %>%
  mutate(value = ifelse(is.na(value), 0, value)) %>%
  mutate(value = ifelse(value > cutoff, 1, 0)) %>%
  group_by(event_type, name) %>%
  dplyr::summarize(percent_positive = mean(value) * 100,
                   total = n(),
                   label = paste0(total*percent_positive/100, "/",total)) %>%
  mutate(assay = "EA")


omic_summary <- rbind.data.frame(pbmc_percents, nasal_percents,
                                 #nasal_conv_percents,
                                 ea_percents) %>%
  as.data.frame() %>%
  mutate(event_type = factor(event_type, levels = c(paste0('Visit ', 1:10)))) %>%
  filter(!is.na(event_type)) %>%
  arrange(assay, event_type)

heatdf <- omic_summary %>%
    pivot_wider(names_from = c(event_type, assay),
                values_from = percent_positive,
                id_cols = name)
heatdf <- heatdf[heatdf$name %in% viruses_of_interest, ]
heatdf <- heatdf %>% column_to_rownames("name")
heatdf <- heatdf[order(rownames(heatdf)), ]
rownames(heatdf) <- gsub("Betacoronavirus", "SARS-CoV-2", rownames(heatdf))
## Manual reording of rows in heatmap
heatdf <- heatdf[c(2, 4,5,9,6, 7,8, 10,1, 11, 12,3),]


col_ano = columnAnnotation(Assay = gsub(".*_", "", colnames(heatdf)),
                            col = list(Assay = c("EA" = "#FE91AD",
                                      "Nasal" = "#E291FE", "PBMC" = "#91E3FE")))

colors_heat <- c("#045375", "#089099", "#7CCBA2", "#FCDE9C", "#F0746E", "#DC3977")
heatdf_annot <- heatdf
heatdf_annot <- apply(heatdf_annot, 2, function(x){
  ifelse(is.na(x), " ", ifelse(x == 0, " ", paste0(round(x, 1), ""))) ## Took % sign out of paste
})
heatdf[is.na(heatdf)] <- 0
heatdf_f <- heatdf
colnames(heatdf_f) <- gsub("_.*", "", colnames(heatdf_f))

new_labels <- c("Baseline", "Day 4", "Day 7", "Day 14", "Day 21", "Day 28",
                "Baseline", "Day 4", "Day 7", "Day 14", "Day 21", "Day 28",
                "3 Months", "6 Months", "9 Months", "12 Months",
                "Baseline", "Day 4", "Day 7", "Day 14", "Day 21", "Day 28",
                "3 Months", "6 Months", "9 Months", "12 Months")
colnames(heatdf_f) <- new_labels
## Scaling of color palette
heatdf_s <- as.data.frame(apply(heatdf_f, 2, function(x){qqnorm(x, plot.it = F)$x}))

pdf(paste0(out_dir, "Figure1C.pdf"), width = 14.5, height = 5)
Heatmap(as.matrix(heatdf_f), name = "Percent of\nSamples", cluster_rows = F,
        cluster_columns = F, col = colors_heat, top_annotation = col_ano,
        row_names_side = "left",
        show_heatmap_legend = T, column_names_rot = 45,
        border = TRUE, column_split = factor(gsub(".*_", "", colnames(heatdf)),
                                             levels = c("PBMC", "Nasal", "EA")),
    cell_fun = function(j, i, x, y, width, height, fill) {
        grid.text(sprintf("%s", heatdf_annot[i, j]), x, y, gp = gpar(fontsize = 11))
})  
## 14.5 x 5in
dev.off()
```

## Figure 1D: Viral Reactivation Curves

```{r}
kernel_size = 5 ## days

kernel_smoothed_values <- data.frame()
for(days in 1:40){
  event_range = (days-((kernel_size-1) / 2)):(days+((kernel_size-1) / 2))
  tmp <- comb_data %>%
    left_join(clinical_data %>% distinct(event_id, event_date)) %>%
    #select(-ends_with(".ea")) %>%
    pivot_longer(cols = c(-event_id, -event_date)) %>%
    mutate(value = ifelse(value > 0, 1, 0)) %>%
    filter(!is.na(value)) %>%
    filter(event_date %in% event_range) %>%
    group_by(name) %>%
    dplyr::summarize(percent = mean(value) * 100) %>%
    mutate(day = days)
  
  kernel_smoothed_values <- rbind.data.frame(kernel_smoothed_values, tmp)
}

smooth_traj <- kernel_smoothed_values %>%
  ungroup() %>%
  mutate(compartment = gsub(".*\\.", "", name),
         virus = gsub("\\..*", "", name)) %>% 
  filter(compartment != "ea") %>%
  filter(!grepl("SARS-CoV-2", virus)) %>%
  filter(!name %in% c("Simplexvirus.pbmc", "Enterovirus.pbmc")) %>%
  filter(!grepl("7|6|8|immuno", name)) %>%
  ggplot(aes(x=day, y=percent, color = virus, linetype = compartment)) +
  geom_smooth(se = F) +
  theme_classic() +
  xlab("Days from Hospitalization") +
  ylab("Percent of Samples") +
  scale_color_manual(values = c("#86005F", "#01a8a0", "#1b820d", "#cba90a",
                                "#005F86", "#BF5700"),
                     name = "Virus") +
  scale_linetype_manual(values = c("22", "solid"),
                        name = "Compartment") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

smooth_traj +
  coord_cartesian(ylim =c(0,25)) # 6x3.5 in
ggsave(paste0(out_dir, "Figure1D.pdf"), width = 6, height = 3.5)

write.csv(smooth_traj$data, paste0(out_dir, "source_data_Figure1D.csv"))
```

## Figure 1E: Viral Load Correlation

```{r}
## Manually installed an older version of Hmisc for R v4.0.3 just for the rcorr function
# packageurl <- "https://cran.r-project.org/src/contrib/Archive/Hmisc/Hmisc_4.8-0.tar.gz"
# install.packages(packageurl, repos=NULL, type="source")

library(Hmisc)
library(corrplot)
flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}
for_cor_data <- as.matrix(comb_data[,c(2:4, 10:15, 22, 23, 24, 26 )])
for_cor_data <- for_cor_data[,!grepl("Beta", colnames(for_cor_data))]
colnames(for_cor_data) <- gsub("\\)", "", gsub(".* \\(", "", colnames(for_cor_data)))
res <- rcorr(for_cor_data,type = "spearman")
resdf <- flattenCorrMatrix(res$r, res$P)
write.csv(resdf, paste0(out_dir, "source_data_Figure1E.csv"))
tmpP <- res$P
tmpP[is.na(tmpP)] <- 1
res$r[is.nan(res$r)] <- 0
pdf(paste0(out_dir, "Figure1E.pdf"), width = 8, height = 8)
corrplot(res$r, type = "upper", order = "hclust", 
         tl.col = "black",outline = TRUE, tl.srt = 45,
         col=colorRampPalette(c("darkblue","white","darkred"))(100),
          p.mat = tmpP, sig.level = 0.05, insig = "blank")
dev.off()
## 8 x8
```

## Supplemental Figure 1

### Figure S1A-B: RT-qPCR and Metagenomics Correlation for SARS-CoV-2

```{r}
library(ggpubr)
corr_data <- comb_data %>%
  dplyr::select(event_id, `SARS-CoV-2.nasal`) %>%
  left_join(nasal_viralload_counts %>%
  rownames_to_column("event_id") %>%
    select(event_id, N1_CT, N2_CT))
corr_data <- corr_data[complete.cases(corr_data), ]

gga <- corr_data %>%
  mutate(`SARS-CoV-2.nasal` = log10(`SARS-CoV-2.nasal` + 1)) %>%
  ggplot(aes(x=`SARS-CoV-2.nasal`, y=N1_CT)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, color = "red",se = F) +
  stat_cor() +
  theme_classic() +
  scale_y_reverse() +
  scale_x_continuous(labels = c(0, 100,1E4,1E6),breaks = c(0,2,4,6)) +
  ylab("Nucleocapsid 1 Cycle Threshold") +
  xlab("Betacoronavirus RPM")

ggb <- corr_data %>%
  mutate(`SARS-CoV-2.nasal` = log10(`SARS-CoV-2.nasal` + 1)) %>%
  ggplot(aes(x=`SARS-CoV-2.nasal`, y=N2_CT)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, color = "red",se = F) +
  stat_cor() +
  theme_classic() +
  scale_y_reverse() +
  scale_x_continuous(labels = c(0, 100,1E4,1E6),breaks = c(0,2,4,6)) +
  ylab("Nucleocapsid 2 Cycle Threshold") +
  xlab("Betacoronavirus RPM")

write_csv(gga$data %>% select(-event_id), paste0(out_dir, "source_data_FigureS1A.csv"))
write_csv(ggb$data %>% select(-event_id), paste0(out_dir, "source_data_FigureS1B.csv"))

top <- ggarrange(gga, ggb, nrow =1, ncol =2, labels = c("A", "B"))
top
ggsave(paste0(out_dir, "FigureS1A-B.pdf"), width = 8, height = 2.5)
```

### Figure S1C: Taxnonomy of Anelloviridae

```{r}
anneloviridae_reports <- plyr::rbind.fill(fread("/scratch/metagenomics/viral_reactivation/data/phase1_4_pbmc_transcriptomics_v2.csv.gz"),
                                          fread("/scratch/metagenomics/viral_reactivation/data/phase4_nasal_transcriptomics.csv.gz"),
                                          fread("/scratch/metagenomics/viral_reactivation/data/nasal_virus_transcriptomic_v1.csv.gz"),
                                          fread("/scratch/metagenomics/viral_reactivation/data/ea_virus_transcriptomic_v1.csv.gz")) %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  filter(genus_tax_id %in% c("200687329", "687333", "687331", "687332")) %>%
  filter(nt_rpm > 0)
  
cat(paste(unique(anneloviridae_reports$tax_id), collapse = "\n"))
## Paste these into https://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi


```

```{r, fig.width=2, fig.height =4}
library(ggtree)
tree <- read.tree("Anelloviridae_phyliptree_ncbi_taxonomy_2024-08-23.phy")

ggf <- ggtree(tree) +  geom_tiplab(align=TRUE, linesize=.5) + 
  geom_hilight(node=37, fill="gold") + 
  geom_hilight(node=36, fill="purple")+ 
  geom_hilight(node=38, fill="blue") + ggplot2::xlim(0, 60)
ggf
ggsave(paste0(out_dir, "FigureS1C.pdf"), width = 5.9, height = 5.9)
## 5.9x5.9
```

### Figure S1D: Taxnonomy of Enteroviruses

```{r}
enterovirus_reports <- plyr::rbind.fill(fread("/scratch/metagenomics/viral_reactivation/data/phase1_4_pbmc_transcriptomics_v2.csv.gz"),
                                          fread("/scratch/metagenomics/viral_reactivation/data/phase4_nasal_transcriptomics.csv.gz"),
                                          fread("/scratch/metagenomics/viral_reactivation/data/nasal_virus_transcriptomic_v1.csv.gz"),
                                          fread("/scratch/metagenomics/viral_reactivation/data/ea_virus_transcriptomic_v1.csv.gz")) %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  filter(genus_tax_id %in% c("12059")) %>%
  filter(nt_rpm > 0)
  
cat(paste(unique(enterovirus_reports$tax_id), collapse = "\n"))
## Paste these into https://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi
```

```{r, fig.width=5, fig.height =4}
library(ggtree)
tree <- read.tree("enterovirus_2024-08-25_ncbi_taxonomy.phy")

ggg <- ggtree(tree) +  geom_tiplab(align=TRUE, linesize=.5) +
  ggplot2::xlim(0, 60)
ggg
ggsave(paste0(out_dir, "FiguresS1D.pdf"), width = 5, height = 4)
## 5x4
```

### Figure S1E-F: Enrollment site, sequencing core

```{r, fig.height = 3, fig.width = 6}
c15 <- c(
  "dodgerblue2", "#E31A1C", 
  "#6A3D9A",
  "#FF7F00",
  "black", "gold1",
  "skyblue2", "#FB9A99",
  "palegreen2",
  "#CAB2D6",
  "#FDBF6F", 
  "gray70", "khaki2",
  "maroon", "orchid1"
)
event_date_cutoff = 40

comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1) %>%
  select(-starts_with("Human betaherpesvirus 6"),
         -starts_with("Human gammaherpesvirus 8"),
         -starts_with("Human betaherpesvirus 7"),
         -starts_with("Respiratory syncytial virus"),
         -starts_with("Human immunodeficiency virus 1"),
         -starts_with("SARS-CoV-2.pbmc"),
         -starts_with("Enterovirus.ea"),
         -starts_with("Enterovirus.pbmc"),
         -starts_with("Anelloviridae.ea"),
         -starts_with("Anelloviridae.nasal"),
         -ends_with(".ea")) %>%
  dplyr::rename("Betacoronavirus.nasal" = "SARS-CoV-2.nasal")

max_part <- comb_data %>%
    pivot_longer(cols = -event_id) %>%
    left_join(clinical_data %>%
                distinct(event_id, participant_id, event_date)) %>%
    filter(event_date >= 0) %>%
    filter(event_date <= event_date_cutoff) %>%
    group_by(participant_id, name) %>%
    filter(name != "Betacoronavirus.nasal") %>%
    dplyr::summarize(value = max(value, na.rm = TRUE)) %>%
    filter(value != -Inf) %>%
    distinct() %>%
    pivot_wider(names_from = name, values_from = value) %>%
    ungroup() %>%
    dplyr::select(-starts_with("Betacoronavirus"))

ggS1E <- max_part %>%
  pivot_longer(cols = -participant_id) %>%
  left_join(clinical_data %>% distinct(participant_id, enrollment_site)) %>%
    mutate(value = ifelse(value > 0, 1, value)) %>%
    filter(!is.na(value)) %>%
  filter(value == 1) %>%
    group_by(name, value, enrollment_site) %>%
    dplyr::summarize(count = n()) %>%
    mutate(freq = count /sum(count) * 100,
           total = sum(count)) %>% 
  mutate(name = gsub(".*\\(", "", gsub(")", "", name))) %>%
  ggplot(aes(x=freq, y=name, fill = enrollment_site)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  scale_y_discrete(limits = rev) +
  guides(fill = guide_legend(override.aes = list(size = 0.25))) +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"),
        legend.key.size = unit(0.4, 'cm')) +
  scale_fill_manual(values = c15, name = "Enrollment Site") +
  xlab("Percent of Positive Participants (%)") +
  ylab("Virus and Tissue")
ggS1E
ggsave(paste0(out_dir, "FigureS1E.pdf"), width = 6, height = 3)
write_csv(ggS1E$data %>% ungroup() %>% dplyr::select(-value, -.group), paste0(out_dir, "source_data_FigureS1E.csv"))
```

For the PBMC sequencing, it was conducted between two cores, with participants split between the two. Here we compare the rates of PBMC associated viruses between the two and find them comparable.

```{r, fig.width = 4.5, fig.height = 2.5}
comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1) %>%
  select(-starts_with("Human betaherpesvirus 6"),
         -starts_with("Human gammaherpesvirus 8"),
         -starts_with("Human betaherpesvirus 7"),
         -starts_with("Respiratory syncytial virus"),
         -starts_with("Human immunodeficiency virus 1"),
         -starts_with("SARS-CoV-2.pbmc"),
         -starts_with("Enterovirus.ea"),
         -starts_with("Enterovirus.pbmc"),
         -starts_with("Anelloviridae.ea"),
         -starts_with("Anelloviridae.nasal"),
         -ends_with(".ea")) %>%
  dplyr::rename("Betacoronavirus.nasal" = "SARS-CoV-2.nasal")

###########################################################################

pbmc_metadata <- read_csv("/data/pbmc-transcriptomics/current/pbmc-transcriptomics-Metadata.csv") %>%
  distinct(sample_id, Core) %>%
  left_join(clinical_data %>% distinct(sample_id, event_id)) %>%
  distinct(event_id, Core)

ggS1F <- comb_data %>%
    pivot_longer(cols = -event_id) %>%
    left_join(pbmc_metadata) %>%
    filter(grepl(".pbmc", name)) %>%
    mutate(value = ifelse(value > 0, 1, value)) %>%
    filter(!is.na(value)) %>%
    group_by(Core, name, value) %>%
    dplyr::summarize(count = n()) %>%
    mutate(freq = count /sum(count) * 100,
           total = sum(count)) %>%
    filter(value == 1) %>%
    filter(!is.na(Core)) %>%
  mutate(Core = gsub(".*_", "", Core)) %>%
  mutate(name = gsub(".*\\(", "", gsub("\\)", "", name))) %>%
    ggplot(aes(x=Core, y= freq,
               fill = Core)) +
    geom_bar(stat = "identity") +
    facet_wrap( ~ name) +
    theme_classic() +
    ylab("Percent of Samples Positive (%)") +
    xlab("Sequencing Core") +
  geom_text(aes(label = paste0(round(freq, 1), "%")), vjust = -0.5, color = "black") +
  coord_cartesian(ylim = c(0, 14)) +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"),
        legend.position = "none") 
ggS1F
ggsave(paste0(out_dir, "FigureS1F.pdf"), width = 4.5, height = 2.5)
write_csv(ggS1F$data %>% ungroup() %>% dplyr::select(-value, -.group), paste0(out_dir, "source_data_FigureS1F.csv"))
```

### Supplemental Analysis of Evaluation for Plate/Batch Effect

```{r}
## data load of a version without the exclusion of the nasal batch 3B_4
comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/old_full_comb_data.csv") %>% select(-1) %>%
  select(-starts_with("Human betaherpesvirus 6"),
         -starts_with("Human gammaherpesvirus 8"),
         -starts_with("Human betaherpesvirus 7"),
         -starts_with("Respiratory syncytial virus"),
         -starts_with("Human immunodeficiency virus 1"),
         -starts_with("SARS-CoV-2.pbmc"),
         -starts_with("Enterovirus.ea"),
         -starts_with("Enterovirus.pbmc"),
         -starts_with("Anelloviridae.ea"),
         -starts_with("Anelloviridae.nasal"),
         -ends_with(".ea")) %>%
  dplyr::rename("Betacoronavirus.nasal" = "SARS-CoV-2.nasal")

## Has phase and plate information
nasal_metadata <- read_csv("/data/nasal-transcriptomics/current/nasal-transcriptomics-Metadata.csv")
pbmc_metadata <- read_csv("/data/pbmc-transcriptomics/current/pbmc-transcriptomics-Metadata.csv")
ea_metadata <- read_csv("/data/ea-transcriptomics/current/ea-transcriptomics-Metadata.csv")
```

#### Nasal Batch Check

```{r, fig.width = 8, fig.height = 4}

model_df <- comb_data %>%
  select(-`Enterovirus.nasal`) %>%
  pivot_longer(cols = -event_id) %>%
  filter(grepl(".nasal", name)) %>%
  filter(!is.na(value)) %>% ## Removes events without a nasal sample
  left_join(nasal_metadata %>%
              left_join(clinical_data %>% distinct(sample_id, event_id,
                                                   trajectory_group,
                                                   event_date))) %>%
  mutate(plate = paste0(phase, "_", plate)) %>%
  filter(event_date <= 40)

model_df %>% mutate(value2 = ifelse(value > 0, 1, 0 )) %>%
  group_by(name, plate, trajectory_group, value2) %>%
  dplyr::summarize(count = n()) %>%
  group_by(name, plate) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup() %>%
  filter(value2 == 1) %>%
  complete(name, plate) %>%
  mutate(freq = ifelse(is.na(freq), 0, freq)) %>%
  filter(name != "Betacoronavirus.nasal") %>%
  mutate(name = gsub(".*\\(", "", gsub(")", "", name))) %>%
  mutate(trajectory_group = factor(trajectory_group, levels = 5:1)) %>%
  ggplot(aes(x=plate, y=freq * 100, fill = trajectory_group)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  facet_wrap(~name,scales = "free") +
  ylab("Percent of Batch") +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(values = TrajGrpCol, name = "Trajectory Group") +
  xlab("Batch")
ggsave(paste0(out_dir, "Nasal_batch_rates.pdf"), width = 8, height = 4)

```


Exploration of high HSV1 batch
```{r, fig.width = 5, fig.height = 3}

model_df %>%
  group_by(name, plate, trajectory_group, participant_id) %>%
  filter(name == "Human alphaherpesvirus 1 (HSV1).nasal") %>%
  filter(plate == "3B_4") %>%
  mutate(trajectory_group = factor(trajectory_group, levels = 5:1)) %>%
  ggplot(aes(x=event_date, y=value, color = trajectory_group)) +
  geom_line(aes(group = participant_id)) +
  geom_point() +
  theme_classic() +
  ylab("HSV1 RPM") +
  #theme(axis.text.x = element_blank()) +
  scale_color_manual(values = TrajGrpCol) +
  xlab("Event Date")+
  scale_y_continuous(trans='log10')

ggsave(paste0(out_dir, "HSV1_RPM_outlier_batch_excluded.pdf"), width = 5, height = 3)
```

##### Statistical Testing
```{r}
## Excluding the identified plate 3B_4 finds no significant association with batch
model_df <- comb_data %>%
  dplyr::select(-Betacoronavirus.nasal) %>%
  pivot_longer(cols = -event_id) %>%
  filter(grepl(".nasal", name)) %>%
  filter(!is.na(value)) %>% ## Removes events without a nasal sample
  left_join(nasal_metadata %>%
              left_join(clinical_data %>% distinct(sample_id, event_id,
                                                   trajectory_group,
                                                   event_date))) %>%
  mutate(plate = paste0(phase, "_", plate)) %>%
  filter(plate != "3B_4") %>%
  mutate(binary = ifelse(value > 0, 1, 0))

nasal_out <- data.frame()
for(active_virus in unique(model_df$name)){
  fit <- lm(value ~ trajectory_group * event_date + plate, data = model_df %>% filter(name == active_virus))
  nasal_out <- rbind.data.frame(nasal_out,
                                car::Anova(fit, type = "III") %>%
                                  as.data.frame() %>%
                                  rownames_to_column("feature") %>%
                                  mutate(virus = active_virus))
}
write.csv(nasal_out, paste0(out_dir, "nasal_batch_testing.csv"), row.names = F)
```


#### PBMC Batch Check

```{r, fig.width = 6, fig.height = 4.5}

model_df <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  filter(grepl(".pbmc", name)) %>%
  filter(!is.na(value)) %>% ## Removes events without a nasal sample
  left_join(pbmc_metadata %>%
              left_join(clinical_data %>% distinct(sample_id, event_id,
                                                   trajectory_group,
                                                   event_date))) %>%
  mutate(plate = paste0(phase, "_", plate_num)) %>%
  filter(event_date <= 40)

model_df %>% mutate(value2 = ifelse(value > 0, 1, 0 )) %>%
  group_by(name, Core, plate, trajectory_group, value2) %>%
  dplyr::summarize(count = n()) %>%
  group_by(name, plate) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup() %>%
  filter(value2 == 1) %>%
  complete(name, nesting(Core, plate)) %>%
  mutate(freq = ifelse(is.na(freq), 0, freq)) %>%
  filter(name != "Betacoronavirus.nasal") %>%
  mutate(name = gsub(".*\\(", "", gsub(")", "", name))) %>%
  mutate(trajectory_group = factor(trajectory_group, levels = 5:1)) %>%
  ggplot(aes(x=plate, y=freq * 100, fill = trajectory_group)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  facet_grid(name~Core,scales = "free") +
  ylab("Percent of Batch") +
  theme(axis.text.x = element_blank()) +
  #theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  scale_fill_manual(values = TrajGrpCol, name = "Trajectory Group") +
  xlab("Batch")
ggsave(paste0(out_dir, "PBMC_batch_rates.pdf"), width = 6, height = 4.5)
```

#### EA Batch Check

```{r, fig.width = 6, fig.height = 3.5}

comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1) %>%
  select(-starts_with("Human betaherpesvirus 6"),
         -starts_with("Human gammaherpesvirus 8"),
         -starts_with("Human betaherpesvirus 7"),
         -starts_with("Respiratory syncytial virus"),
         -starts_with("Human immunodeficiency virus 1"),
         -starts_with("SARS-CoV-2.pbmc"),
         -starts_with("Enterovirus.ea"),
         -starts_with("Enterovirus.pbmc"),
         -starts_with("Anelloviridae.ea"),
         -starts_with("Anelloviridae.nasal"))

model_df <- comb_data %>%
  pivot_longer(cols = -event_id) %>%
  filter(grepl(".ea", name)) %>%
  filter(!is.na(value)) %>% ## Removes events without a nasal sample
  left_join(ea_metadata %>%
              left_join(clinical_data %>% distinct(sample_id, event_id,
                                                   trajectory_group,
                                                   event_date))) %>%
  mutate(plate = paste0(phase, "_", plate)) %>%
  filter(event_date <= 40)

model_df %>% mutate(value2 = ifelse(value > 0, 1, 0 )) %>%
  group_by(name, plate, trajectory_group, value2) %>%
  dplyr::summarize(count = n()) %>%
  group_by(name, plate) %>%
  mutate(freq = count / sum(count)) %>%
  ungroup() %>%
  filter(value2 == 1) %>%
  complete(name, plate) %>%
  mutate(freq = ifelse(is.na(freq), 0, freq)) %>%
  filter(name != "SARS-CoV-2.ea") %>%
  mutate(name = gsub(".*\\(", "", gsub(")", "", name))) %>%
  mutate(trajectory_group = factor(trajectory_group, levels = 5:1)) %>%
  ggplot(aes(x=plate, y=freq * 100, fill = trajectory_group)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  facet_wrap(~name,scales = "free") +
  ylab("Percent of Batch") +
  theme(axis.text.x = element_blank()) +
  scale_fill_manual(values = TrajGrpCol) +
  xlab("Batch")
ggsave(paste0(out_dir, "EA_batch_rates.pdf"), width = 8, height = 4)

```

### Figure S1G: Checking Read Duplication
#### EBV in PBMCs
```{r, fig.width = 2.5, fig.height = 3}
#| eval: false
## Takes a long time to run (hence eval: false), presaved results in repo
pbmc_metadata <- read_csv("/data/pbmc-transcriptomics/current/pbmc-transcriptomics-Metadata.csv")
ebv_pbmc_reads <- read_csv("/scratch/metagenomics/viral_reactivation/EBV_pbmc_reads.csv") %>%
  mutate(sample_id = gsub(".*_", "", gsub(".*\\/", "", gsub("_S.*", "",i))))
  
key_data <- ebv_pbmc_reads %>%
  left_join(pbmc_metadata) %>%
  mutate(phase = paste0(Core, "_", phase)) %>%
  mutate(plate = paste0(Core, "_", plate_num)) %>%
  mutate(phase_plate = paste0(phase, "__", plate_num))

for(read in 1:length(key_data$tmp_reads)){ #
  print(read)
  test <- key_data %>%
    filter(sample_id != key_data$sample_id[read])
  
  key_data$global_dup[read] <- ifelse(sum(grepl(key_data$tmp_reads[read], test$tmp_reads)) > 0, TRUE, FALSE)
  
  ## Exclude reads from the other core in the case of PBMCs
  test_core <- test %>%
    filter(Core == test$Core[read])
  key_data$core_dup[read] <- ifelse(sum(grepl(key_data$tmp_reads[read], test_core$tmp_reads)) > 0, TRUE, FALSE)
  
  ## Exclude reads from the wrong phase of the project
  test_phase <- test_core %>%
    filter(phase == test$phase[read])
  key_data$phase_dup[read] <- ifelse(sum(grepl(key_data$tmp_reads[read], test_phase$tmp_reads)) > 0, TRUE, FALSE)
  
  ## Exclude reads from other plates/batches from that phase
  test_phase_plate <- test_phase %>%
    filter(phase_plate == test$phase_plate[read])
  key_data$plate_dup[read] <- ifelse(sum(grepl(key_data$tmp_reads[read], test_phase_plate$tmp_reads)) > 0, TRUE, FALSE)
}

tmp1 <- key_data %>%
 dplyr::summarize(count = n(),
                  freq_of_dup = sum(global_dup) / count * 100)

tmp2 <- key_data %>%
  group_by(Core) %>%
 dplyr::summarize(count = n(),
                  freq_of_dup = sum(core_dup) / count * 100)

tmp3 <- key_data %>%
  group_by(phase) %>%
 dplyr::summarize(count = n(),
                  freq_of_dup = sum(phase_dup) / count * 100) %>%
   ungroup() %>%
  dplyr::summarize(freq_of_dup = mean(freq_of_dup))

tmp4 <- key_data %>%
  group_by(phase_plate) %>%
 dplyr::summarize(count = n(),
                  freq_of_dup = sum(plate_dup) / count * 100 )%>%
  ungroup() %>%
  dplyr::summarize(freq_of_dup = mean(freq_of_dup))


ebv_read_dup <- plyr::rbind.fill(tmp1 %>% mutate(Core = "All Reads"),
                 tmp2,
               #  tmp3 %>% mutate(Core = "Within Phase"),
                 tmp4 %>% mutate(Core = "Within Batch")) %>%
  filter(!is.na(Core)) %>%
  mutate(Core = factor(Core, levels = c("All Reads",
                                        "Eckalbar_UCSF", "Bosinger_Emory",
                                        "Within Phase", "Within Batch"))) %>%
  mutate(label = paste0(round(freq_of_dup, 1), "%")) %>%
  ggplot(aes(x=Core, y=freq_of_dup)) +
    geom_bar(stat = "identity", color = "black", fill = "#BF5700") +
  geom_text(aes(label = label), 
    vjust = -0.5
  ) +
    theme_classic() +
  ylab("% of Duplicated Reads") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black")) +
  ggtitle("EBV in PBMCs") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  coord_cartesian(ylim = c(0, 50))
ebv_read_dup
ggsave(paste0(out_dir, "FigureS1G_part1.pdf"), width = 2.5, height = 3)
write.csv(ebv_read_dup$data, paste0(out_dir, "source_data_FigureS1G_part1.csv"))
```

#### CMV in PBMCs
```{r, fig.width = 2.5, fig.height = 3}
#| eval: false
## Takes a long time to run (hence eval: false), presaved results in repo
pbmc_metadata <- read_csv("/data/pbmc-transcriptomics/current/pbmc-transcriptomics-Metadata.csv")
cmv_pbmc_reads <- read_csv("/scratch/metagenomics/viral_reactivation/CMV_pbmc_reads.csv") %>%
  mutate(sample_id = gsub(".*_", "", gsub(".*\\/", "", gsub("_S.*", "",i))))
  
key_data <- cmv_pbmc_reads %>%
  left_join(pbmc_metadata) %>%
  mutate(phase = paste0(Core, "_", phase)) %>%
  mutate(plate = paste0(Core, "_", plate_num)) %>%
  mutate(phase_plate = paste0(phase, "__", plate_num))

for(read in 1:length(key_data$tmp_reads)){ #
  print(read)
  test <- key_data %>%
    filter(sample_id != key_data$sample_id[read])
  
  key_data$global_dup[read] <- ifelse(sum(grepl(key_data$tmp_reads[read], test$tmp_reads)) > 0, TRUE, FALSE)
  
  ## Exclude reads from the other core in the case of PBMCs
  test_core <- test %>%
    filter(Core == test$Core[read])
  key_data$core_dup[read] <- ifelse(sum(grepl(key_data$tmp_reads[read], test_core$tmp_reads)) > 0, TRUE, FALSE)
  
  ## Exclude reads from the wrong phase of the project
  test_phase <- test_core %>%
    filter(phase == test$phase[read])
  key_data$phase_dup[read] <- ifelse(sum(grepl(key_data$tmp_reads[read], test_phase$tmp_reads)) > 0, TRUE, FALSE)
  
  ## Exclude reads from other plates/batches from that phase
  test_phase_plate <- test_phase %>%
    filter(phase_plate == test$phase_plate[read])
  key_data$plate_dup[read] <- ifelse(sum(grepl(key_data$tmp_reads[read], test_phase_plate$tmp_reads)) > 0, TRUE, FALSE)
}

tmp1 <- key_data %>%
 dplyr::summarize(count = n(),
                  freq_of_dup = sum(global_dup) / count * 100)

tmp2 <- key_data %>%
  group_by(Core) %>%
 dplyr::summarize(count = n(),
                  freq_of_dup = sum(core_dup) / count * 100)

tmp3 <- key_data %>%
  group_by(phase) %>%
 dplyr::summarize(count = n(),
                  freq_of_dup = sum(phase_dup) / count * 100) %>%
   ungroup() %>%
  dplyr::summarize(freq_of_dup = mean(freq_of_dup))

tmp4 <- key_data %>%
  group_by(phase_plate) %>%
 dplyr::summarize(count = n(),
                  freq_of_dup = sum(plate_dup) / count * 100 )%>%
  ungroup() %>%
  dplyr::summarize(freq_of_dup = mean(freq_of_dup))


cmv_read_dup <- plyr::rbind.fill(tmp1 %>% mutate(Core = "All Reads"),
                 tmp2,
               #  tmp3 %>% mutate(Core = "Within Phase"),
                 tmp4 %>% mutate(Core = "Within Batch")) %>%
  filter(!is.na(Core)) %>%
  mutate(Core = factor(Core, levels = c("All Reads",
                                        "Eckalbar_UCSF", "Bosinger_Emory",
                                        "Within Phase", "Within Batch"))) %>%
  mutate(label = paste0(round(freq_of_dup, 1), "%")) %>%
  ggplot(aes(x=Core, y=freq_of_dup)) +
    geom_bar(stat = "identity", color = "black", fill = "#005F86") +
  geom_text(aes(label = label), 
    vjust = -0.5
  ) +
    theme_classic() +
  ylab("% of Duplicated Reads") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black")) +
  ggtitle("CMV in PBMCs") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  coord_cartesian(ylim = c(0, 25))

cmv_read_dup
ggsave(paste0(out_dir, "FigureS1G_part2.pdf"), width = 2.5, height = 3)
write.csv(cmv_read_dup$data, paste0(out_dir, "source_data_FigureS1G_part2.csv"))
```


### Figure S1H: Average E Score of Taxonomic Alignment

```{r}
library(scales)

## Report from CZID includes E Score
pbmc_reports <- fread("/scratch/metagenomics/viral_reactivation/data/phase1_4_pbmc_transcriptomics_v2.csv.gz") %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  filter(nt_rpm > 0)

Anelloviridae_genera <- c("Alphatorquevirus", "Betatorquevirus",
                         "Gammatorquevirus", "non-genus-specific reads in family Anelloviridae")
pbmc_ready <- pbmc_reports %>%
  mutate(name = gsub("Cytomegalovirus",
                                      "Human betaherpesvirus 5 (CMV)",
                         gsub("Lymphocryptovirus",
                              "Human gammaherpesvirus 4 (EBV)",
                                   name))) %>%
   mutate(name = stringi::stri_replace_all_regex(name,
                       pattern = Anelloviridae_genera,
                       replacement = rep("Anelloviridae", 4),
                       vectorize_all = TRUE)) %>%
  filter(name %in% viruses_of_interest) %>%
  filter(name != "Betacoronavirus") %>%
  rowwise() %>%
  mutate(nt_e_value2 = (eval(parse(text = nt_e_value)))) %>%
  group_by(czid_id, full_czid_id, name) %>% 
  dplyr::summarize(nt_e_value3 = mean(rep(nt_e_value2, nt_count)))
```

```{r}

### Nasal ####
nasal_metagenomics_metadata <- read_csv("/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv")
nasal_viral_samples <- fread("/scratch/metagenomics/viral_reactivation/data/nasal_virus_samples_analyzed.csv.gz")
nasal_samples <- nasal_metagenomics_metadata %>%
              filter(core_specific_ID %in% nasal_viral_samples$samples) %>%
              filter(sample_status == "IMPACC sample assayed and passed QC") %>%
              filter(internal_control != "yes") %>%
              distinct(core_specific_ID, sample_id)
nasal_reports <- fread("/scratch/metagenomics/viral_reactivation/data/nasal_virus_transcriptomic_v1.csv.gz") %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  filter(nt_rpm > 0) %>%
  filter(full_czid_id %in% nasal_metagenomics_metadata$core_specific_ID)

nasal_ready <- nasal_reports %>%
  mutate(name = gsub("Cytomegalovirus",
                                      "Human betaherpesvirus 5 (CMV)",
                         gsub("Lymphocryptovirus",
                              "Human gammaherpesvirus 4 (EBV)",
                                   name))) %>%
   mutate(name = stringi::stri_replace_all_regex(name,
                       pattern = Anelloviridae_genera,
                       replacement = rep("Anelloviridae", 4),
                       vectorize_all = TRUE)) %>%
  filter(name %in% viruses_of_interest) %>%
  filter(name != "Betacoronavirus") %>%
  rowwise() %>%
  mutate(nt_e_value2 = (eval(parse(text = nt_e_value)))) %>%
  group_by(full_czid_id, name) %>% 
  dplyr::summarize(nt_e_value3 = mean(rep(nt_e_value2, nt_count))) ## weighted average
## Have to combine the Anelloviridae RPMs (with a weighted average)


### EA ####
ea_metagenomics_metadata <- read_csv("/data/ea-metagenomics/current/ea-metagenomics-Metadata.csv")
ea_metagenomics_rowfeature <- read_csv("/data/ea-metagenomics/current/ea-metagenomics-AllGenera-RowFeatures.csv", skip = 1)
ea_viral_samples <- fread("/scratch/metagenomics/viral_reactivation/data/ea_virus_samples_analyzed.csv.gz")

ea_reports <- fread("/scratch/metagenomics/viral_reactivation/data/ea_virus_transcriptomic_v1.csv.gz") %>%
  filter(category == "viruses" & is_phage == FALSE) %>%
  filter(nr_rpm > 0) %>%
  filter(nt_rpm > 0) %>%
  filter(full_czid_id %in% ea_metagenomics_metadata$core_specific_ID)
ea_ready <- ea_reports %>%
  mutate(name = gsub("Cytomegalovirus",
                                      "Human betaherpesvirus 5 (CMV)",
                         gsub("Lymphocryptovirus",
                              "Human gammaherpesvirus 4 (EBV)",
                                   name))) %>%
   mutate(name = stringi::stri_replace_all_regex(name,
                       pattern = Anelloviridae_genera,
                       replacement = rep("Anelloviridae", 4),
                       vectorize_all = TRUE)) %>%
  filter(name %in% viruses_of_interest) %>%
  filter(name != "Betacoronavirus") %>%
  rowwise() %>%
  mutate(nt_e_value2 = (eval(parse(text = nt_e_value)))) %>%
  group_by(full_czid_id, name) %>% 
  dplyr::summarize(nt_e_value3 = mean(rep(nt_e_value2, nt_count)))

##### Combine them all #####
all_ready <- plyr::rbind.fill(pbmc_ready %>% mutate(omic = "PBMC"),
                              nasal_ready %>% mutate(omic = "Nasal"),
                              ea_ready %>% mutate(omic = "EA"))

ggS1H <- all_ready %>%
  mutate(omic = factor(omic, levels = c("PBMC", "Nasal", "EA"))) %>%
  ggplot(aes(y=-log10(nt_e_value3), x= name)) +
  geom_boxplot(outlier.shape = NA) +
  #ggbeeswarm::geom_quasirandom() +
  geom_jitter(size = 0.7) +
  theme_classic() +
  #scale_y_continuous(trans='log10') +
  #coord_trans(y = "log10") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black")) +
  ylab("-log10(Average E Value)") +
  facet_grid(~omic, scales = "free", space = "free") +
  geom_hline(yintercept = -log10(1e-5), color = "red", linetype = "dashed") +
  xlab("Virus")
ggS1H
ggsave(paste0(out_dir, "FigureS1H.pdf"), width =6,height = 4)
write_csv(ggS1H$data %>% select(-czid_id, -full_czid_id),
          paste0(out_dir, "source_data_FigureS1H.csv"))

```

## Supplemental Figure 2

### Figure S2A: Viral Reads per Million Graph Over Acute Period

```{r, fig.width = 10, fig.height = 6}
## This is the same comb_data that was generated at the top of this script
comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1)

ggS2 <- comb_data %>%
  pivot_longer(-event_id) %>%
  mutate(compartment = gsub(".*\\.", "", name),
         virus = gsub("\\..*", "", name)) %>% 
  filter(compartment != "ea") %>%
  filter(!grepl("SARS-CoV-2", virus)) %>%
  filter(!grepl("syn", virus)) %>%
  filter(!name %in% c("Simplexvirus.pbmc", "Enterovirus.pbmc")) %>%
  filter(!grepl("7|6|8|immuno", name)) %>%
  mutate(name = gsub(".*\\(", "", gsub(")", "", name))) %>%
  left_join(clinical_data %>% distinct(event_id, participant_id, event_date)) %>%
  filter(event_date <= 40) %>%
  ggplot(aes(x=event_date, y= value)) +
  geom_line(aes(group = participant_id), alpha = 0.4, color = "darkgray") +
  geom_point(aes(color = virus), alpha = 0.3, size = 1.6, stroke=NA) +
  facet_wrap(~name) +
  theme_classic() +
  scale_y_continuous(trans='log10') +
  xlab("Days from Hospitalization") +
  ylab("Reads per Million") +
  scale_color_manual(values = c("#86005F", "#01a8a0", "#1b820d", "#cba90a",
                                "#005F86", "#BF5700"),
                     name = "Virus") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))
ggS2
ggsave(paste0(out_dir, "FigureS2A.pdf"), width = 10, height = 6)
write_csv(ggS2$data %>%
            mutate(participant_id = as.numeric(as.factor(participant_id))) %>%
            select(-event_id), paste0(out_dir, "source_data_FigureS2A.csv"))
```

Alternative graph as days from symptom onset instead of days from hospitalization

```{r, fig.width = 10, fig.height = 6}
## This is the same comb_data that was generated at the top of this script
comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1)

ggS2_alt <- comb_data %>%
  pivot_longer(-event_id) %>%
  mutate(compartment = gsub(".*\\.", "", name),
         virus = gsub("\\..*", "", name)) %>% 
  filter(compartment != "ea") %>%
  filter(!grepl("SARS-CoV-2", virus)) %>%
  filter(!grepl("syn", virus)) %>%
  filter(!name %in% c("Simplexvirus.pbmc", "Enterovirus.pbmc")) %>%
  filter(!grepl("7|6|8|immuno", name)) %>%
  mutate(name = gsub(".*\\(", "", gsub(")", "", name))) %>%
  left_join(clinical_data %>% distinct(event_id, participant_id, event_date,
                                       symptom_date)) %>%
  rowwise() %>%
  mutate(event_date = event_date-symptom_date) %>% ungroup() %>%
  filter(event_date <= 40) %>%
  ggplot(aes(x=event_date, y= value)) +
  geom_line(aes(group = participant_id), alpha = 0.4, color = "darkgray") +
  geom_point(aes(color = virus), alpha = 0.3, size = 1.6, stroke=NA) +
  facet_wrap(~name) +
  theme_classic() +
  scale_y_continuous(trans='log10') +
  xlab("Days from Symptom Onset") +
  ylab("Reads per Million") +
  scale_color_manual(values = c("#86005F", "#01a8a0", "#1b820d", "#cba90a",
                                "#005F86", "#BF5700"),
                     name = "Virus") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))
ggS2_alt
ggsave(paste0(out_dir, "FigureS2A_DFSO.pdf"), width = 10, height = 6)
write_csv(ggS2_alt$data %>%
            mutate(participant_id = as.numeric(as.factor(participant_id))) %>%
            select(-event_id),
          ## Removing participant_id (has new id system in ImmPort and we cannot)
          ## make participant_id widely available with this data release
          paste0(out_dir, "source_data_FigureS2A_DFSO.csv"))
```

### Figure S2B-E: Expanded Smoothed Reactivation Curves

```{r}
kernel_size = 5 ## days

kernel_smoothed_values <- data.frame()
tg_comp_all <- data.frame()
for(days in 1:40){
  event_range = (days-((kernel_size-1) / 2)):(days+((kernel_size-1) / 2))
  tmp1 <- comb_data %>%
    left_join(clinical_data %>% distinct(event_id, event_date)) %>%
    #select(-ends_with(".ea")) %>%
    pivot_longer(cols = c(-event_id, -event_date)) %>%
    mutate(value = ifelse(value > 0, 1, 0)) %>%
    filter(!is.na(value)) %>%
    filter(event_date %in% event_range) 
  
  tg_comp <- tmp1 %>%
    left_join(clinical_data %>% distinct(event_id, participant_id, trajectory_group)) %>%
    mutate(tissue = gsub(".*\\.", "", name)) %>%
    group_by(tissue, trajectory_group) %>%
    dplyr::summarize(count = n()) %>%
    mutate(freq = count / sum(count) * 100) %>%
    mutate(day = days)
  
  tmp <- tmp1 %>%
    group_by(name) %>%
    dplyr::summarize(percent = mean(value) * 100) %>%
    mutate(day = days)
  
  kernel_smoothed_values <- rbind.data.frame(kernel_smoothed_values, tmp)
  tg_comp_all <- rbind.data.frame(tg_comp_all, tg_comp)
}


smooth_traj <- kernel_smoothed_values %>%
  ungroup() %>%
  mutate(compartment = gsub(".*\\.", "", name),
         virus = gsub("\\..*", "", name)) %>% 
  filter(!name %in% c("Simplexvirus.pbmc", "Enterovirus.pbmc")) %>%
  filter(!grepl("7|6|8|immuno", name)) %>%
  ggplot(aes(x=day, y=percent, color = virus, linetype = compartment)) +
  geom_smooth(se = F) +
  theme_classic() +
  xlab("Days from Hospitalization") +
  ylab("Percent of Samples") +
  scale_color_manual(values = c("Anelloviridae" ="#86005F",
                                "Enterovirus" ="#01a8a0",
                                "Human alphaherpesvirus 1 (HSV1)" ="#1b820d",
                                "Human alphaherpesvirus 2 (HSV2)" ="#cba90a",
                                "Human betaherpesvirus 5 (CMV)" = "#005F86",
                                "Human gammaherpesvirus 4 (EBV)" ="#BF5700",
                                "SARS-CoV-2" ="black"),
                     name = "Virus") +
  scale_linetype_manual(values = c("ea"= "dotted","nasal" = "22","pbmc" =  "solid"),
                        name = "Compartment") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))
gge <- smooth_traj


ggzoom <- smooth_traj
ggh <- ggg <- ggf <- ggzoom
ggf$data <- ggzoom$data %>% filter(grepl(".ea$", name))
ggg$data <- ggzoom$data %>% filter(grepl(".nasal$", name))
ggg <- ggg +
  coord_cartesian(ylim = c(0, 20))
ggh$data <- ggzoom$data %>% filter(grepl(".pbmc$", name))

write_csv(gge$data, paste0(out_dir, "source_data_FigureS2B.csv"))
write_csv(ggf$data, paste0(out_dir, "source_data_FigureS2C.csv"))
write_csv(ggg$data, paste0(out_dir, "source_data_FigureS2D.csv"))
write_csv(ggh$data, paste0(out_dir, "source_data_FigureS2E.csv"))

```

```{r}
tg_comp_all <- as.data.frame(tg_comp_all)
gg_pbmc_comp <- tg_comp_all %>%
  filter(tissue == "pbmc") %>%
  mutate(trajectory_group = as.factor(trajectory_group)) %>%
  ggplot(aes(x=day, y= freq, fill = trajectory_group)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  scale_fill_manual(values = TrajGrpCol)+
  ylab("% of PBMC")+
  xlab("Days from Hospitalization") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

gg_nasal_comp <- tg_comp_all %>%
  filter(tissue == "nasal") %>%
  mutate(trajectory_group = as.factor(trajectory_group)) %>%
  ggplot(aes(x=day, y= freq, fill = trajectory_group)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_classic() +
  scale_fill_manual(values = TrajGrpCol) +
  ylab("% of Nasal") +
  xlab("Days from Hospitalization") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

gg_ea_comp <- tg_comp_all %>%
  filter(tissue == "ea") %>%
  mutate(trajectory_group = as.factor(trajectory_group)) %>%
  ggplot(aes(x=day, y= freq, fill = trajectory_group)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_classic() +
  scale_fill_manual(values = TrajGrpCol) +
  ylab("% of EA") +
  xlab("Days from Hospitalization") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggpubr::ggarrange(gg_pbmc_comp,
                  gg_nasal_comp,
                  gg_ea_comp,
                  align = "hv", nrow = 3, common.legend = T, heights = c(1,1,1))
write_csv(gg_pbmc_comp$data, paste0(out_dir, "source_data_FigureS2C_part2.csv"))
write_csv(gg_pbmc_comp$data, paste0(out_dir, "source_data_FigureS2D_part2.csv"))
write_csv(gg_pbmc_comp$data, paste0(out_dir, "source_data_FigureS2E_part2.csv"))
```

```{r, fig.width = 8, fig.height = 7.5}
bottom <- ggarrange(gge, ggf + ylab("Percent of EA Samples (%)"),
                    NA, gg_ea_comp,
                    align = "hv", heights = c(3,1.5),
                    nrow =2, ncol =2, labels = c("B", "C", "", ""),
                    common.legend = TRUE,legend = "right")
bottom2 <- ggarrange(ggg + ylab("Percent of Nasal Samples (%)"),
                     ggh + ylab("Percent of PBMC Samples (%)"),
                     gg_nasal_comp,
                     gg_pbmc_comp,
                     nrow =2, ncol =2, align = "hv",
                     labels = c("D", "E"),
                      heights = c(3,1.5),
                     common.legend = TRUE,legend = "right")
ggarrange(bottom, bottom2, nrow = 2, ncol =1,
          heights = c(1, 1))
ggsave(paste0(out_dir, "FigureS2_B-E.pdf"), width = 8, height = 7.5)
```

### Figure S2F-I: Days from Symptom Onset Smoothed Reactivation Curves

```{r}
kernel_size = 5 ## days

kernel_smoothed_values <- data.frame()
tg_comp_all <- data.frame()
for(days in 1:40){
  event_range = (days-((kernel_size-1) / 2)):(days+((kernel_size-1) / 2))
  tmp1 <- comb_data %>%
    left_join(clinical_data %>% distinct(event_id, event_date, symptom_date) %>%
                rowwise() %>%
                mutate(event_date = event_date-symptom_date) %>% ungroup() %>%
                distinct(event_id, event_date)) %>%
    #select(-ends_with(".ea")) %>%
    pivot_longer(cols = c(-event_id, -event_date)) %>%
    mutate(value = ifelse(value > 0, 1, 0)) %>%
    filter(!is.na(value)) %>%
    filter(event_date %in% event_range) 
  
  tg_comp <- tmp1 %>%
    left_join(clinical_data %>% distinct(event_id, participant_id, trajectory_group)) %>%
    mutate(tissue = gsub(".*\\.", "", name)) %>%
    group_by(tissue, trajectory_group) %>%
    dplyr::summarize(count = n()) %>%
    mutate(freq = count / sum(count) * 100) %>%
    mutate(day = days)
  
  tmp <- tmp1 %>%
    group_by(name) %>%
    dplyr::summarize(percent = mean(value) * 100) %>%
    mutate(day = days)
  
  kernel_smoothed_values <- rbind.data.frame(kernel_smoothed_values, tmp)
  tg_comp_all <- rbind.data.frame(tg_comp_all, tg_comp)
}


smooth_traj <- kernel_smoothed_values %>%
  ungroup() %>%
  mutate(compartment = gsub(".*\\.", "", name),
         virus = gsub("\\..*", "", name)) %>% 
  filter(!name %in% c("Simplexvirus.pbmc", "Enterovirus.pbmc")) %>%
  filter(!grepl("7|6|8|immuno", name)) %>%
  ggplot(aes(x=day, y=percent, color = virus, linetype = compartment)) +
  geom_smooth(se = F) +
  theme_classic() +
  xlab("Days from Symptom Onset") +
  ylab("Percent of Samples") +
  scale_color_manual(values = c("Anelloviridae" ="#86005F",
                                "Enterovirus" ="#01a8a0",
                                "Human alphaherpesvirus 1 (HSV1)" ="#1b820d",
                                "Human alphaherpesvirus 2 (HSV2)" ="#cba90a",
                                "Human betaherpesvirus 5 (CMV)" = "#005F86",
                                "Human gammaherpesvirus 4 (EBV)" ="#BF5700",
                                "SARS-CoV-2" ="black"),
                     name = "Virus") +
  scale_linetype_manual(values = c("ea"= "dotted","nasal" = "22","pbmc" =  "solid"),
                        name = "Compartment") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))
gge <- smooth_traj


ggzoom <- smooth_traj
ggh <- ggg <- ggf <- ggzoom
ggf$data <- ggzoom$data %>% filter(grepl(".ea$", name))
ggg$data <- ggzoom$data %>% filter(grepl(".nasal$", name))
ggg <- ggg +
  coord_cartesian(ylim = c(0, 20))
ggh$data <- ggzoom$data %>% filter(grepl(".pbmc$", name))

write_csv(gge$data, paste0(out_dir, "source_data_FigureS2F.csv"))
write_csv(ggf$data, paste0(out_dir, "source_data_FigureS2G.csv"))
write_csv(ggg$data, paste0(out_dir, "source_data_FigureS2H.csv"))
write_csv(ggh$data, paste0(out_dir, "source_data_FigureS2I.csv"))

```

```{r}
tg_comp_all <- as.data.frame(tg_comp_all)
gg_pbmc_comp <- tg_comp_all %>%
  filter(tissue == "pbmc") %>%
  mutate(trajectory_group = as.factor(trajectory_group)) %>%
  ggplot(aes(x=day, y= freq, fill = trajectory_group)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  scale_fill_manual(values = TrajGrpCol)+
  ylab("% of PBMC")+
  xlab("Days from Hospitalization") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

gg_nasal_comp <- tg_comp_all %>%
  filter(tissue == "nasal") %>%
  mutate(trajectory_group = as.factor(trajectory_group)) %>%
  ggplot(aes(x=day, y= freq, fill = trajectory_group)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_classic() +
  scale_fill_manual(values = TrajGrpCol) +
  ylab("% of Nasal") +
  xlab("Days from Hospitalization") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

gg_ea_comp <- tg_comp_all %>%
  filter(tissue == "ea") %>%
  mutate(trajectory_group = as.factor(trajectory_group)) %>%
  ggplot(aes(x=day, y= freq, fill = trajectory_group)) +
  geom_bar(position = "stack", stat = "identity") +
  theme_classic() +
  scale_fill_manual(values = TrajGrpCol) +
  ylab("% of EA") +
  xlab("Days from Hospitalization") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"))

ggpubr::ggarrange(gg_pbmc_comp,
                  gg_nasal_comp,
                  gg_ea_comp,
                  align = "hv", nrow = 3, common.legend = T, heights = c(1,1,1))
write_csv(gg_pbmc_comp$data, paste0(out_dir, "source_data_FigureS2G_part2.csv"))
write_csv(gg_pbmc_comp$data, paste0(out_dir, "source_data_FigureS2H_part2.csv"))
write_csv(gg_pbmc_comp$data, paste0(out_dir, "source_data_FigureS2I_part2.csv"))
```

```{r, fig.width = 8, fig.height = 7.5}
bottom <- ggarrange(gge, ggf + ylab("Percent of EA Samples (%)"),
                    NA, gg_ea_comp,
                    align = "hv", heights = c(3,1.5),
                    nrow =2, ncol =2, labels = c("F", "G", "", ""),
                    common.legend = TRUE,legend = "right")
bottom2 <- ggarrange(ggg + ylab("Percent of Nasal Samples (%)"),
                     ggh + ylab("Percent of PBMC Samples (%)"),
                     gg_nasal_comp,
                     gg_pbmc_comp,
                     nrow =2, ncol =2, align = "hv",
                     labels = c("H", "I"),
                      heights = c(3,1.5),
                     common.legend = TRUE,legend = "right")
ggarrange(bottom, bottom2, nrow = 2, ncol =1,
          heights = c(1, 1))
ggsave(paste0(out_dir, "FigureS2_F-I.pdf"), width = 8, height = 7.5)
```

## Supplemental Figure 3

### Figure S3A: Viral Co-incidence on an event basis

```{r}
comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1) 
conincidence_func <- function(matrix){
  out <- data.frame()
  for(i in colnames(matrix)){
    test_tmp <- matrix %>%
      rownames_to_column("sample_id") %>%
      pivot_longer(cols=-sample_id) %>%
      filter(!is.na(value)) 
    
    live <- test_tmp %>%
      filter(name == i) %>%
      filter(value == 1)
    
    test <- test_tmp %>%
      filter(sample_id %in% live$sample_id) %>%
      filter(value == 1) %>%
      group_by(name) %>%
      dplyr::summarize(count = n()) %>%
      mutate(freq = count / nrow(live) * 100) %>%
      dplyr::select(name, freq)
    colnames(test)[2] <- i 
     if(nrow(out) == 0){
       out <- test
     } else {
       out <- out %>%
        full_join(test)
     }
  }
  return(out)
}

conincendence_data <- comb_data[,-1]
conincendence_data <- conincendence_data[,!grepl("8|6|7|immuno|Respir", colnames(conincendence_data))]
colnames(conincendence_data) <- gsub("\\)", "", gsub(".* \\(", "", colnames(conincendence_data)))
conincendence_data[conincendence_data>0] <- 1
conincendence_data <- as.data.frame(conincendence_data)
rownames(conincendence_data) <- comb_data$event_id
pilot <- conincidence_func(conincendence_data)
pilot[is.na(pilot)] <- 0

pilot_2 <-as.data.frame(pilot[,-1])
rownames(pilot_2) <- pilot$name
pilot_2 <- pilot_2[,order(colnames(pilot_2))]
pilot_2 <- pilot_2[order(rownames(pilot_2)),]

write_csv(as.data.frame(t(as.matrix(pilot_2))), paste0(out_dir, "source_data_FigureS3A.csv"))

ggS3A <- as.data.frame(t(as.matrix(pilot_2))) %>%
  rownames_to_column("Rows") %>%
  pivot_longer(cols = -Rows) %>%
  ggplot(aes(x=name, y= Rows)) +
  geom_tile(aes(fill = value)) +
  geom_text(aes(label = round(value, 2))) +
  theme_classic() +
    scale_y_discrete(limits=rev) +
  scale_fill_gradient(low = "white", high = "red", name = "Percent") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  xlab("This virus was present X%") +
  ylab("If the event is positive for this virus")
ggS3A
ggsave(paste0(out_dir, "FigureS3A.pdf"), width = 10, height = 6)

```

### Figure S3B: Viral Co-incidence on a participant basis

##### Key Viruses

```{r}
comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1) %>%
  select(-starts_with("Human betaherpesvirus 6"),
         -starts_with("Human gammaherpesvirus 8"),
         -starts_with("Human betaherpesvirus 7"),
         -starts_with("Respiratory syncytial virus"),
         -starts_with("Human immunodeficiency virus 1"),
         -starts_with("SARS-CoV-2.pbmc"),
         -starts_with("Enterovirus.ea"),
         -starts_with("Enterovirus.pbmc"),
         -starts_with("Anelloviridae.ea"),
         -starts_with("Anelloviridae.nasal"),
         -ends_with(".ea")) %>%
  dplyr::rename("Betacoronavirus.nasal" = "SARS-CoV-2.nasal")
event_date_cutoff = 40
max_part <- comb_data %>%
    pivot_longer(cols = -event_id) %>%
    left_join(clinical_data %>%
                distinct(event_id, participant_id, event_date)) %>%
    filter(event_date >= 0) %>%
    filter(event_date <= event_date_cutoff) %>%
    group_by(participant_id, name) %>%
    filter(name != "Betacoronavirus.nasal") %>%
    dplyr::summarize(value = max(value, na.rm = TRUE)) %>%
    filter(value != -Inf) %>%
    distinct() %>%
    pivot_wider(names_from = name, values_from = value) %>%
    ungroup() %>%
    dplyr::select(-starts_with("Betacoronavirus"))

conincendence_data <- max_part[,-1]
conincendence_data <- conincendence_data[,!grepl("8|6|7|immuno|Respir", colnames(conincendence_data))]
colnames(conincendence_data) <- gsub("\\)", "", gsub(".* \\(", "", colnames(conincendence_data)))
conincendence_data[conincendence_data>0] <- 1
conincendence_data <- as.data.frame(conincendence_data)
rownames(conincendence_data) <- max_part$participant_id
pilot <- conincidence_func(conincendence_data)
pilot[is.na(pilot)] <- 0

pilot_2 <-as.data.frame(pilot[,-1])
rownames(pilot_2) <- pilot$name
pilot_2 <- pilot_2[,order(colnames(pilot_2))]
pilot_2 <- pilot_2[order(rownames(pilot_2)),]

write_csv(as.data.frame(t(as.matrix(pilot_2))), paste0(out_dir, "source_data_FigureS3B.csv"))

ggs3B <- as.data.frame(t(as.matrix(pilot_2))) %>%
  rownames_to_column("Rows") %>%
  pivot_longer(cols = -Rows) %>%
  ggplot(aes(x=name, y= Rows)) +
  geom_tile(aes(fill = value)) +
  geom_text(aes(label = round(value, 1))) +
  theme_classic() +
    scale_y_discrete(limits=rev) +
  scale_fill_gradient(low = "white", high = "red", name = "Percent") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  xlab("Also at one point had") +
  ylab("If they had this virus")
ggs3B
ggsave(paste0(out_dir, "FigureS3B_key_viruses.pdf"), width = 6, height = 4)
```

##### All Viruses

```{r}
comb_data <- read_csv("/scratch/metagenomics/viral_reactivation/full_comb_data.csv") %>% select(-1)
max_part <- comb_data %>%
    pivot_longer(cols = -event_id) %>%
    left_join(clinical_data %>%
                distinct(event_id, participant_id, event_date)) %>%
    filter(event_date >= 0) %>%
    filter(event_date <= event_date_cutoff) %>%
    group_by(participant_id, name) %>%
    filter(name != "Betacoronavirus.nasal") %>%
    dplyr::summarize(value = max(value, na.rm = TRUE)) %>%
    filter(value != -Inf) %>%
    distinct() %>%
    pivot_wider(names_from = name, values_from = value) %>%
    ungroup() %>%
    dplyr::select(-starts_with("Betacoronavirus"))

conincendence_data <- max_part[,-1]
conincendence_data <- conincendence_data[,!grepl("8|6|7|immuno|Respir", colnames(conincendence_data))]
colnames(conincendence_data) <- gsub("\\)", "", gsub(".* \\(", "", colnames(conincendence_data)))
conincendence_data[conincendence_data>0] <- 1
conincendence_data <- as.data.frame(conincendence_data)
rownames(conincendence_data) <- max_part$participant_id
pilot <- conincidence_func(conincendence_data)
pilot[is.na(pilot)] <- 0

pilot_2 <-as.data.frame(pilot[,-1])
rownames(pilot_2) <- pilot$name
pilot_2 <- pilot_2[,order(colnames(pilot_2))]
pilot_2 <- pilot_2[order(rownames(pilot_2)),]

ggs3b_alt <- as.data.frame(t(as.matrix(pilot_2))) %>%
  rownames_to_column("Rows") %>%
  pivot_longer(cols = -Rows) %>%
  ggplot(aes(x=name, y= Rows)) +
  geom_tile(aes(fill = value)) +
  geom_text(aes(label = round(value, 1))) +
  theme_classic() +
    scale_y_discrete(limits=rev) +
  scale_fill_gradient(low = "white", high = "red", name = "Percent") +
  theme(axis.text = element_text(color = "black"),
        axis.ticks = element_line(color = "black"),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  xlab("Also at one point had") +
  ylab("If they had this virus")
ggsave(paste0(out_dir, "FigureS3B_all_viruses.pdf"), width = 10, height = 6)
ggs3b_alt
write_csv(ggs3b_alt$data, paste0(out_dir, "source_data_FigureS3B_all_viruses.csv"))
```

```{r, fig.width = 10, fig.height = 10}

ggarrange(ggS3A,
          ggarrange(ggs3B, NA, nrow = 1, ncol = 2, widths = c(6, 4)),
          ncol = 1, nrow = 2, heights = c(6, 4))
ggsave(paste0(out_dir, "FigureS3.pdf"), width = 10, height = 10)
```

# Session Info

```{r}
sessionInfo()
```
