---
title: "ExtendedData1"
output: html_notebook
---

```{r load library}
library(ade4) #used in mcia
library(mogsa)
library(factoextra)
library(cluster)
library(nnet)
library(survival)
library(ordinal)
library(ggeffects)
library(openxlsx)
library(network)
library(GGally)
library(ggalt)
library(metafor)
library(cowplot)
library(survminer)
require(nlme) # directly provides p-values, unlike lme4
require(splines)
require(dplyr)
library(fgsea)
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source("~/impacc/Analysis/molecular_subtyping/preliminary/helpers_LG202301.R")
source("~/impacc/Analysis/molecular_subtyping/code/ExtendedData1/src/ExtendedData1_helper.R")
date = "20230726"; inner_folds = 10
PACS_group = list(physical = c(2), cognitive = c(5), multiple = c(6))
path_impacc_analysis = "/scratch/data-integration/cleanedup/"
intermediate_results_outdir_DR = "/scratch/data-integration/molecular_subtyping/DimensionReduction/"
organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")
file_mcia = paste0(intermediate_results_outdir_DR,"MCIA_phase123_",date,".rds")
mcia_results = readRDS(file_mcia)

base_obj = readRDS(file ="/scratch/data-integration/cleanedup/mod1_interpretation_base.rds")
file_impacc_analysis = paste0(path_impacc_analysis,"impacc_analysis_record_innerfold", inner_folds,".rds")
impacc_analysis_record=readRDS(file = file_impacc_analysis)
data_path <- "/scratch/data-integration/pipelines/processedData/"
data_env0 <- readRDS(paste0(data_path, "data_env_ravi_20220513.RDS"))

clin_data_all = organized_data$clinical_data[,setdiff(colnames(organized_data$clinical_data), c("sample_id","sample_type", "phase1_2","samples_collected","phase"))]
clin_data_all = unique(clin_data_all)
clin_data_aligned = data_frame(event_id=rownames(organized_data$datasets$plasma_proteomics_targeted))
clin_data_aligned <- clin_data_aligned %>% left_join(clin_data_all, by = c("event_id"))
ll = which(clin_data_aligned$event_type %in% paste0("Visit ",1:6))
clin_data_aligned =clin_data_aligned[ll,]
mcia_factors= mcia_results$mcia_factors[ll,]
clin_data_aligned$conv_cluster6[is.na(clin_data_aligned$conv_cluster6)] = 999
clin_data_aligned$conv_physical_deficit = ifelse(clin_data_aligned$conv_cluster6%in%PACS_group$physical,1,0)
clin_data_aligned$conv_cognitive_deficit = ifelse(clin_data_aligned$conv_cluster6%in%PACS_group$cognitive,1,0)
clin_data_aligned$conv_global_deficit = ifelse(clin_data_aligned$conv_cluster6%in%PACS_group$multiple,1,0)
```

# Clustering data preparation
```{r data preparation}
n_factor = 40
factor_scores = mcia_factors[,1:n_factor]
# normalize factor scores
factor_scores = apply(factor_scores, 2, function(z) qqnorm(z, plot.it = F)$x)
# update factors that explain more consistent variation across assays
aa = (apply(sqrt(mcia_results$mcia_factor_evaluation$ave_variance_explained[1:n_factor,]),
            1, sum))^2
factor_scores = t(apply(factor_scores, 1, function(z) z*sqrt(aa)))
# flip the sign of factor1
factor_scores[,1] = -factor_scores[,1]

visit_max = 6
extended_data = make_data_long(factor_scores, sample_meta = clin_data_aligned, 
                               visit_max = visit_max, mean_removal = F)
X = extended_data$X
# upweight samples with convalesent outcomes
ll0 = which(clin_data_aligned$conv_cluster6 != 999)
extended_data_conv = make_data_long(factor_scores[ll0,], 
                                    sample_meta = clin_data_aligned[ll0,], 
                                    visit_max = visit_max, mean_removal = F)

participant_ids = rownames(extended_data$X)
other_features = colnames(clin_data_aligned)[grepl("comorb",colnames(clin_data_aligned)) |grepl("ever",colnames(clin_data_aligned))|grepl("comp",colnames(clin_data_aligned))]
clin_vars <- c("participant_id", "conv_cluster3", "conv_cluster6", 
               "conv_cognitive_deficit", "conv_physical_deficit", 
               "conv_global_deficit", "ethnicity", other_features)
clin_mat_use = data.frame(unique(clin_data_aligned[, clin_vars]))
rownames(clin_mat_use) = clin_mat_use$participant_id
clin_mat_use = clin_mat_use[participant_ids,]
tmp1 = extended_data$sample_meta_participant_level
tmp1$participant_id = NULL
clin_mat_use = cbind(clin_mat_use, tmp1)

id_PASC = (clin_mat_use$conv_cluster6 != 999)
PASCs = rep(NA, nrow(clin_mat_use))
PASCs[id_PASC] = ifelse(
  clin_mat_use$conv_cluster6[id_PASC]%in%PACS_group$physical, "physical",
  ifelse(clin_mat_use$conv_cluster6[id_PASC]%in%PACS_group$cognitive, "cognitive",
         ifelse(clin_mat_use$conv_cluster6[id_PASC]%in%PACS_group$multiple, "multiple",
                "minimal")))
PASCs = factor(PASCs, c("minimal", "physical", "cognitive", "multiple"))
clin_data_pasc = data.frame("participant_id" = clin_mat_use$participant_id,
                            "PASC_group" = PASCs)
clin_data_aligned = left_join(clin_data_aligned, clin_data_pasc, 
                              by = "participant_id")
```

# Participant size across visits (ExtendedData 1A)
```{r participant size across visits, fig.height=6, fig.width=8}
library(UpSetR)
df_avail <- extended_data$X[, seq(1, by = 40, length.out = visit_max)]
colnames(df_avail) <- paste0("Visit ", 1:visit_max)
df_avail <- as.data.frame(+!is.na(df_avail))

df_avail %>%
  UpSetR::upset(order.by = "freq", nsets = 6, nintersects = NA,
                sets = paste0("Visit ", seq(visit_max, 1, by = -1)), 
                keep.order = TRUE, sets.bar.color = "lightblue",
                main.bar.color = "salmon3", point.size = 2.8, line.size = 0.6,
                text.scale = c(1.8, 1.6, 1.8, 1.6, 1.6, 1.6),
                set_size.show = TRUE, set_size.numbers_size = 6, 
                set_size.scale_max = 1400)
```

# Hierarchical clustering
```{r hierarchical clustering}
dist_mat_all <- calculate_distance_matrix(X, n_factor, visit_max = visit_max)
dist_mat_conv <- calculate_distance_matrix(extended_data_conv$X, n_factor, 
                                           visit_max = visit_max)
participant_id_conv = rownames(extended_data_conv$X)
# Clustering using all participants
hcl_all <- hclust(d = dist_mat_all, method = "ward.D2")
# Clustering using convalesent participants
hcl_conv <- hclust(d = dist_mat_conv, method = "ward.D2")

Kmin = 5; Kmax= 10
k.values <- Kmin:Kmax
sil_values <- c()
avg_sil <- function(dist_obj, cluster) {
  ss <- silhouette(cluster, dist_obj)
  mean(ss[, 3])
}
for (k in k.values) {
  cluster_cut <- cutree(hcl_all, k)
  sil_values[k - Kmin+1] <- avg_sil(dist_obj  = dist(x = X[,1:n_factor]), 
                                    cluster_cut)
}

nclust_tree_conv = nclust_tree_all = k.values[which.max(sil_values)]

subtype_hcl_all = cutree(hcl_all, k = nclust_tree_all)
subtype_hcl_conv =  cutree(hcl_conv, k = nclust_tree_conv)
subtype_hcl = data.frame(full = subtype_hcl_all)
subtype_hcl$conv = NA
rownames(subtype_hcl) = names(subtype_hcl_all)
subtype_hcl[names(subtype_hcl_conv),]$conv = subtype_hcl_conv
subtype_hcl$PASC = PASCs
subtype_hcl$TG = factor(clin_mat_use$trajectory_group)

# S8 S3 S4 S5 S7 S1 S6 S2
# rename: SD, SC1, SC2, SE, SF, SA, SB1, SB2
# merge: SD, SC, SC, SE, SF, SA, SB, SB
mapping_label <- c("1" = "A", "2" = "B", "3" = "C", "4" = "C", 
                   "5" = "E", "6" = "B", "7" = "F", "8" = "D")
subtype_hcl$label <- mapping_label[as.character(subtype_hcl$full)]
# table(subtype_hcl$label, subtype_hcl$full)
nclust_tree_all = 6

mapping_label2 <- c("1" = "A", "2" = "B2", "3" = "C1", "4" = "C2", 
                   "5" = "E", "6" = "B1", "7" = "F", "8" = "D")
subtype_hcl$label_merge <- mapping_label2[as.character(subtype_hcl$full)]
# table(subtype_hcl$label_merge, subtype_hcl$label)
```

# Silhouette values for optimal cluster selection (ExtendedData 1B)
```{r silhouette values, fig.height=6, fig.width=8}
df_sil <- data.frame(k = k.values, sil_values = sil_values)
ggplot(df_sil, aes(x = k, y = sil_values)) +
  geom_vline(xintercept = 8, linetype = "dashed", color = "red") +
  geom_hline(yintercept = max(sil_values), linetype = "dashed", color = "red") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  xlab("Number of clusters (k)") +
  ylab("Average Silhouette Width") +
  theme_minimal() +
  theme(axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15),
        axis.text.x = element_text(color = "black", size = 15),
        axis.text.y = element_text(color = "black", size = 15))
```

# Cluster dendrogram - 8 clusters (ExtendedData 1C)
```{r cluster dendrogram, fig.height=6, fig.width=8.5, warning=FALSE}
cluster_dendrogram <- fviz_dend(hcl_all, k = 8, show_labels = FALSE,
                                k_colors = c("#FFDC91B2", "#E18727B2", 
                                             "#E18727B2", "#EE4C97B2",
                                             "#BC3C29B2", "#20854EB2", 
                                             "#0072B5B2", "#0072B5B2"),
                                rect = TRUE, lwd = 0.6) + 
  ggtitle(paste0("Dendrogram at Selected Cluster Tree Depth")) +
  theme(plot.title = element_text(size = 15),
        axis.title.y = element_text(size = 15, margin = margin(r = 10)),
        axis.title.x = element_text(size = 15),
        axis.text.y = element_text(size = 15, color = "black"),
        axis.text.x = element_blank())
cluster_annot <- data.frame(x = c(15, 70, 230, 420, 535, 710, 890, 1040),
                            y = rep(-2.6, 8),
                            label = c("D", "C1", "C2", "E", 
                                      "F", "A", "B1", "B2"),
                            color = rep("grey50", 8), size = rep(4.5, 8), 
                            fontface = rep("bold", 8))
cluster_dendrogram +
  lapply(seq_len(nrow(cluster_annot)), function(i) {
    annotate("text", x = cluster_annot$x[i], y = cluster_annot$y[i], 
             label = cluster_annot$label[i], color = cluster_annot$color[i],
             size = cluster_annot$size[i], 
             fontface = cluster_annot$fontface[i])})
```

# Network plot - 8 clusters (ExtendedData 1D)
```{r network plot, fig.height=6, fig.width=9}
net_hcl <- as.matrix(dist_mat_all)
net_hcl <- network(net_hcl, directed=FALSE)
net_subtype <- as.factor(subtype_hcl$label_merge)
levels(net_subtype) <- paste("Subtype", levels(net_subtype))
net_hcl %v% "Subtype" <- as.character(net_subtype)

network_plot <- ggnet2(net_hcl, node.size = 1, node.color = "Subtype", 
                       node.alpha = 0.5,
  palette = c("Subtype A" = "#20854EB2", "Subtype B1" = "#005D99", 
              "Subtype B2" = "#3391CC", "Subtype C1" = "#C76D1E",
              "Subtype C2" = "#F29B41", "Subtype D" = "#FFDC91B2",
              "Subtype E" = "#EE4C97B2", "Subtype F" = "#BC3C29B2"),
  edge.size = 0.1, edge.color = c("color", "white"),
  edge.alpha = 0.6, legend.size = 12, mode = "mds",
  layout.par = list(var="user", dist="none", vm=dist_mat_all, exp=0.1))

network_plot_data <- ggplot_build(network_plot)$data[[2]]
major_points <- network_plot_data %>%
    group_by(group) %>%
    mutate(centroid_x = mean(x), 
           centroid_y = mean(y),
           distance = sqrt((x - centroid_x)^2 + (y - centroid_y)^2)) %>%
    arrange(group, distance) %>%
    group_by(group) %>%
    slice_head(prop = 0.6)
network_plot +
  geom_encircle(data = major_points[major_points$group == 2, ], 
                aes(x = x, y = y, group = group), 
                color = "#005D99", fill = "#005D99",
                size = 1.5, linetype = "dashed", alpha = 0.5) +
  geom_encircle(data = major_points[major_points$group == 3, ], 
                aes(x = x, y = y, group = group), 
                color = "#3391CC", fill = "#3391CC",
                size = 1.5, linetype = "dashed", alpha = 0.5) +
  geom_encircle(data = major_points[major_points$group == 4, ], 
                aes(x = x, y = y, group = group), 
                color = "#C76D1E", fill = "#C76D1E",
                size = 1.5, linetype = "dashed", alpha = 0.5) +
  geom_encircle(data = major_points[major_points$group == 5, ], 
                aes(x = x, y = y, group = group), 
                color = "#F29B41", fill = "#F29B41",
                size = 1.5, linetype = "dashed", alpha = 0.5) +
  guides(color = guide_legend(keywidth = 1.5, keyheight = 1.5)) +
  theme(legend.text = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.key.size = unit(1.5, "lines"))
```

# Barplot colored by TG - 8 clusters (ExtendedData 1E right)
```{r barplot colored by TG, fig.height=6, fig.width=8, message=FALSE}
idx_label <- which(subtype_hcl$label_merge == "B1" |
                    subtype_hcl$label_merge == "B2" |
                    subtype_hcl$label_merge == "C1" |
                    subtype_hcl$label_merge == "C2" |
                    subtype_hcl$label_merge == "E" |
                    subtype_hcl$label_merge == "F")
df_summary <- data.frame(Subtype = factor(subtype_hcl$label_merge[idx_label]),
                         PASC =subtype_hcl$PASC[idx_label],
                         TG = subtype_hcl$TG[idx_label])

df_summary2 <- df_summary %>%
  dplyr::group_by(Subtype, TG) %>%
  dplyr::summarise(Count = n(), .groups = "drop")
total_counts2 <- df_summary2 %>%
  group_by(Subtype) %>%
  summarise(Total = sum(Count))
df_summary2 <- df_summary2 %>%
  dplyr::group_by(Subtype) %>%
  mutate(percent = Count/sum(Count))

df_summary2$Subtype <- factor(df_summary2$Subtype, 
                              levels = rev(levels(df_summary2$Subtype)))
ggplot(df_summary2 , aes(y = Subtype, x = percent, fill = TG)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  geom_text(data = total_counts2,
            aes(y = Subtype, x = -0.06, label = sprintf(" %d", Total)),
            position = position_dodge(width = 0.9),
            inherit.aes = FALSE, size = 7) +
  labs(y = "Subtype", x = "Percent") +
  theme(axis.title.x = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text.x = element_text(color = "black", size = 15),
        axis.text.y = element_text(color = "black", size = 15),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        legend.key.size = unit(2, "lines"),
        plot.margin = margin(b = 55, t = 5)) +
  scale_fill_manual(values = c("1" = "#639A21", "2" = "#39828C",
                               "3" = "#6371AD", "4" = "#BD7D31", 
                               "5" = "#9C3418"))
```

# Ordinal regression before merging (ExtendedData 1E left)
```{r ordinal regression before merging, fig.height=6, fig.width=2.5}
idx_label = which(subtype_hcl$label_merge == "B1" |
                    subtype_hcl$label_merge == "B2" |
                    subtype_hcl$label_merge == "C1" |
                    subtype_hcl$label_merge == "C2" |
                    subtype_hcl$label_merge == "E" |
                    subtype_hcl$label_merge == "F")
merge_clust_test= list(); merge_clust_test[["TG"]] = list()
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]

merge_clust_test[["TG"]][["ordinal_demo"]] = rbind(
  test_ordinal_demo_two_group(
    subtype = subtype_hcl$label_merge[idx_label],
    response = subtype_hcl$TG[idx_label],
    demo_data = demo_data_adjust[idx_label,],
    base_group = "B1", test_group = "B2"),
  test_ordinal_demo_two_group(
    subtype = subtype_hcl$label_merge[idx_label],
    response = subtype_hcl$TG[idx_label],
    demo_data = demo_data_adjust[idx_label,],
    base_group = "C1", test_group = "C2"),
  test_ordinal_demo_two_group(
    subtype = subtype_hcl$label_merge[idx_label],
    response = subtype_hcl$TG[idx_label],
    demo_data = demo_data_adjust[idx_label,],
    base_group = "E", test_group = "F"))

test_ExtendedData1E = merge_clust_test[["TG"]][["ordinal_demo"]][,c(1,4)]
colnames(test_ExtendedData1E) = c("Estimate", "p-value")

heatmap_ordinal_merge(merge_clust_test[["TG"]][["ordinal_demo"]])
```

# Forest plot for acute mortality - two group comparison (ExtendedData 1F)
```{r forest plot for acute mortality, fig.height=6, fig.width=8, warning=FALSE}
# adjust for demographics
clin_data_patient <- clin_data_aligned[clin_data_aligned$event_type == "Visit 1",]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", "sex")]

sample_meta_participant = extended_data$sample_meta_participant_level
sample_meta_participant$death_date[sample_meta_participant$death_date<0] = 999
censoring_date_death = 28
death_surv = cbind(sample_meta_participant$death_date,
                   sample_meta_participant$death_date)
death_surv[death_surv[,2]>censoring_date_death,2] = censoring_date_death+1
death_surv[,1] = ifelse(death_surv[,2]<=censoring_date_death,1,0)
death_surv = data.frame(death_surv)
colnames(death_surv) = c("state", "time")

# hazard ratio for B1,B2
hr_idx1 <- which(subtype_hcl$label_merge %in% c("B1", "B2"))
surv_object1 <- Surv(time = death_surv$time[hr_idx1],
                     event = death_surv$state[hr_idx1])
data_surv1 <- cbind(data.frame(
  subtype = factor(subtype_hcl$label_merge[hr_idx1])),
  demo_data_adjust[hr_idx1,])
cox1 <- coxph(surv_object1 ~., data = data_surv1)
# summary(cox1)$coefficients[1,]

# hazard ratio for C1,C2
hr_idx2 <- which(subtype_hcl$label_merge %in% c("C1", "C2"))
surv_object2 <- Surv(time = death_surv$time[hr_idx2 ],
                     event = death_surv$state[hr_idx2 ])
data_surv2 <- cbind(data.frame(
  subtype = factor(subtype_hcl$label_merge[hr_idx2 ])),
  demo_data_adjust[hr_idx2 ,])
cox2 <- coxph(surv_object2 ~., data = data_surv2)
# summary(cox2)$coefficients[1,]

# hazard ratio for E,F
hr_idx3 <- which(subtype_hcl$label_merge %in% c("E", "F"))
surv_object3 <- Surv(time = death_surv$time[hr_idx3],
                     event = death_surv$state[hr_idx3])
data_surv3 <- cbind(data.frame(
  subtype = factor(subtype_hcl$label[hr_idx3])),
  demo_data_adjust[hr_idx3,])
cox3 <- coxph(surv_object3 ~., data = data_surv3)
# summary(cox3)$coefficients[1,]

coef_cox_two_group_comparison <- rbind(
  summary(cox1)$coefficients[1,],
  summary(cox2)$coefficients[1,],
  summary(cox3)$coefficients[1,])

test_ExtendedData1F = cbind(coef_cox_two_group_comparison[,1],
                            coef_cox_two_group_comparison[,3],
                            coef_cox_two_group_comparison[,5])
rownames(test_ExtendedData1F) = c("B2 vs B1", "C2 vs C1", "F vs E")
colnames(test_ExtendedData1F) = c("Coef", "se(Coef)", "p-value")

forest(coef_cox_two_group_comparison[,1], 
       sei = coef_cox_two_group_comparison[,3], refline = 0, 
       xlim = c(-5.5,7), pch = 20, slab = c("B2 vs B1", "C2 vs C1", "F vs E"), 
       shade = "zebra", at = c(-2,-1,0,1,2), cex = 1.4,
       header = c("Subtype", "log(Hazard Ratio) [95% CI]"), xlab = "")
text(0, 6, "log(Hazard Ratio) for Acute Mortality", font = 2, cex = 1.4)
```


