---
title: "Transplant antibodies"
output: html_document
date: "2024-01-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
library(gamm4)


my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```

# Data import

## Clinical metadata

```{r}
# Patient metadata
clin.individ <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-locked.csv",
  header = TRUE)

# Sample
clin.sample <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-sample-locked.csv",
  header = TRUE)

# Event
clin.event <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-event-locked.csv",
  header = TRUE)
```

Merge metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

Manually correct some typos.

```{r}
# Correct the symptom_date of patient 008-0103: change from -376 to -13 (confirmed with data management team)
clin.sample.event.individ[clin.sample.event.individ$participant_id=="008-0103","symptom_date"] <- -13

# Correct event_date of 004-0012-8: change from 2 to 32 (it's 32 in 2021-11-11-frozen data)
clin.sample.event.individ[clin.sample.event.individ$event_id=="004-0012-8","event_date"] <- 32
```

## Transplant metadata 

264 patients included in transplant study (use new metadata table that Harry will share)

```{r}
transplant <- read.delim(
  "/scratch/immunosuppression/2023.03.09_txcasecontrol.txt",
  header = TRUE)
newTG <- read.delim(
  "/scratch/immunosuppression/2023.10.03_new_tgroups.txt",
  header = TRUE)

# Merge new TG assignment
transplant <- merge(
  transplant, newTG,
  by="studyid", all=FALSE
)

# Format factor
transplant$transplant_final <- as.factor(transplant$transplant_final)

```

Remove BMT patients and their controls.

```{r}
transplant <- transplant %>%
  subset(!(studyid %in% c("002-0022","007-0008","001-0013","010-0009","004-0041","005-0047")))

```

Only keep transplant patients in the clinical metadata

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event.individ[,c("participant_id","event_id","sample_id","event_date","event_type","event_location","respiratory_status")], 
  transplant,
  by.x="participant_id", by.y="studyid",
  all=FALSE
)

```

## Import
```{r}
#antibody data
abd <- read.csv("/data/serum-rbd-abtiters/current/serum-rbd-abtiters-Counts.csv",
  header = TRUE, row.names = 1
)
#metadata
abd.metadata <- read.csv("/data/serum-rbd-abtiters/current/serum-rbd-abtiters-Metadata.csv",
  header = TRUE, row.names = 1
)

# Only keep samples that pass QC
abd.metadata <- subset(abd.metadata, sample_status=="IMPACC sample assayed and passed QC")
clin.sample.event.individ <- clin.sample.event.individ %>%
  subset(sample_id %in% rownames(abd.metadata))
```

```{r}
# Filter abd data to samples of interest
abd <- abd[clin.sample.event.individ$sample_id,]
```

## Import viral load data

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  # "/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv",
  "/data/nasal-metagenomics/legacy/2022-09-03/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)

mngs.metadata <- mngs.metadata %>%
  filter(!is.na(sc2_rpm))

# mNGS QC
mngs.metadata <- mngs.metadata %>%
  filter(sample_status == "IMPACC sample assayed and passed QC")
```

```{r}
# Reformat visit_num: decapitalize "VISIT"
mngs.metadata$visit_num <- gsub(
  "^VISIT", "Visit", mngs.metadata$visit_num
)

# Merge with sample metadata
clin.sample.event.individ <- merge(
  clin.sample.event.individ, mngs.metadata[,c("participant_id","visit_num","sc2_rpm")],
  by.x=c("participant_id","event_type"), by.y=c("participant_id","visit_num"),
  all.x=TRUE, all.y=FALSE # keep samples that don't pass SARS-CoV-2 metatranscriptomics QC
  # all=FALSE # discard samples that don't pass SARS-CoV-2 metatranscriptomics QC <<< do NOT do this
)

# log-transform viral load
clin.sample.event.individ$sc2_rpm_log10 <- log10(clin.sample.event.individ$sc2_rpm+1)
```

# Analysis: Visit 1 only
```{r}
# Sample metadata
clin.sample.event.individ.visit1 <- clin.sample.event.individ %>%
  subset(event_type == "Visit 1")

# Subset abd data
abd.visit1 <- abd[clin.sample.event.individ.visit1$sample_id,]

#Spike antibodies/Figure 3C
i=4
p=summary(lm(log2(abd.visit1[,i])~clin.sample.event.individ.visit1$transplant_final))$coefficients[2,4]
boxplot(abd.visit1[,i]~clin.sample.event.individ.visit1$transplant_final,xaxt='n',yaxt='n',col="white",border=c("cornflowerblue","orange"),xlab="",ylab="",las=2,cex.lab=1.5,cex.main=1.5,cex.axis=1.2,font.main=1,log='y',pch="")
axis(1,at=1:2,labels=c("Control","Transplant"),cex.axis=1.2)
axis(2,at=c(1,10,100,1000,10000,100000),labels=c(1,10,100,1000,10000,100000),cex.axis=1.2,las=2)
mtext(side=2,text="Spike IgG (AUC)",cex=1.5,line=5)
stripchart(abd.visit1[clin.sample.event.individ.visit1$transplant_final==0,i]~factor(clin.sample.event.individ.visit1$transplant_final[clin.sample.event.individ.visit1$transplant_final==0]),add=T,vertical=T,method="jitter",pch=20,at=1,col="cornflowerblue")
stripchart(abd.visit1[clin.sample.event.individ.visit1$transplant_final==1,i]~factor(clin.sample.event.individ.visit1$transplant_final[clin.sample.event.individ.visit1$transplant_final==1]),add=T,vertical=T,method="jitter",pch=20,at=2,col="orange")
text(x=1.5,y=0.9*max(abd.visit1[,i],na.rm=T),label=paste0("P=",round(p,4)),cex=1.5)
```

# Analysis: longitudinal

```{r}
#Sample metadata
clin.sample.event.individ.long <- clin.sample.event.individ %>%
  subset(event_type %in% c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6"))

# Try restricting to 30 days post admission
clin.sample.event.individ.long <- clin.sample.event.individ.long %>%
  subset(event_date <= 30)

abd.long <- abd[clin.sample.event.individ.long$sample_id,]

#run regression
res.abs <- data.frame(matrix(NA, nrow=4, ncol=5))
colnames(res.abs) <- c("slope.2vs1","slope.1","slope.2",
                       "PVal","adj.PVal")
res.abs$variable <- colnames(abd.long)[1:4]
rownames(res.abs) <- colnames(abd.long)[1:4]
for(i in 1:4)
{
  clin.sample.event.individ.long$tmp=abd.long[,i]
  fit1=lmer(tmp~event_date+event_date:transplant_final+transplant_final+(1|participant_id),data=clin.sample.event.individ.long,REML=F)
  fit2=lmer(tmp~event_date+transplant_final+(1|participant_id),data=clin.sample.event.individ.long,REML=F)
  aa <- anova(fit2, fit1, test = "LRT")
  res.abs[i,"PVal"] <- aa[2,"Pr(>Chisq)"]
  aa <- summary(fit1)
  res.abs[i,"slope.2vs1"] <- coefficients(aa)["event_date:transplant_final1","Estimate"]
  res.abs[i,"slope.1"] <- coefficients(aa)["event_date","Estimate"]
  res.abs[i,"slope.2"] <- res.abs[i,"slope.1"] + res.abs[i,"slope.2vs1"]
  print(i)
}
res.abs$adj.PVal=p.adjust(res.abs$PVal,"BH")
res.abs$nam=c("RBD IgG","RBd IgG","Spike IgG","Spike IgG")

#plot
i=4 #for Spike antibodies/Figure 3D
clin.sample.event.individ.long$transplant_final_full="Control"
clin.sample.event.individ.long$transplant_final_full[clin.sample.event.individ.long$transplant_final==1]="Transplant"

temp <- data.frame(
value = log2(abd.long[,i]),
event_date = clin.sample.event.individ.long$event_date,
txstatus = clin.sample.event.individ.long$transplant_final_full,
pid = clin.sample.event.individ.long$participant_id
) %>%
subset(!is.na(value) & !is.na(event_date)) %>%
droplevels()
# Predict
fit <- lmer(
value ~ event_date + event_date:txstatus + txstatus +(1|pid),
data = temp, REML = FALSE)
my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","txstatus"),
type="fixed")
colnames(my.pred)[colnames(my.pred)=="group"] <- "txstatus"
temp$yhat <- predict(fit)
# Calculate IQR and quartile values for limiting y-axes
q1 <- quantile(temp$value, 0.25)
q3 <- quantile(temp$value, 0.75)
iqr <- IQR(temp$value)

ggplot(data = temp, aes(x=event_date, y=value, color=txstatus)) +
geom_line(aes(group=pid), color="gray", alpha=0.4) +
#geom_line(inherit.aes=FALSE, aes(x=event_date, y=yhat, group=pid),
#         color="#5F5F5F", alpha=0.9) +
geom_point(aes(color="black"), shape=20, size=1) +
stat_smooth(method = "lm", formula = y ~ poly(x,3),linewidth=1.5,aes(color=txstatus,fill=txstatus)) +
#geom_ribbon(inherit.aes=FALSE,data=my.pred,aes(x, predicted, ymax=conf.high, ymin=conf.low,fill=txstatus),alpha=0.2) +
#geom_line(inherit.aes=FALSE, data=my.pred, aes(x,predicted),
#         color="black", size=1.5) +
coord_cartesian(ylim = c(q1-1.5*iqr, q3+1.5*iqr)) +
labs(x="Days from admission", y="AUC (log transformed)",
title=sprintf("%s (P=%.1e)",res.abs[i,"nam"],res.abs[i,"adj.PVal"])) +
scale_color_viridis_d(begin=0.4, guide="none") +
# scale_color_brewer(palette="Paired", guide="none") +
scale_color_manual(values=c("black","cornflowerblue","orange")) +
scale_fill_manual(values=c("cornflowerblue","orange")) +
# scale_color_discrete(guide="none") +
facet_wrap(~ txstatus, nrow=1) + theme_bw() +
theme(
plot.title = element_text(hjust = 0.5, size=11, face="plain"),
axis.text.x = element_text(size=10, color="black"),
axis.text.y = element_text(size=10, color="black"),
text = element_text(size=12, family="Arial"),
plot.margin = unit(c(0.2,1,0.7,0.2), "cm"),
panel.spacing = unit(0.35, "cm"),
legend.position = "none",
panel.grid.minor=element_blank()
)

```

#Compare days from symptom onset
```{r}
#one correction
transplant[transplant$studyid=="008-0103","symptom_date"] <- -13
#remove BMT + controls
final_transplant <- transplant %>%
  subset(!(studyid %in% c("002-0022","007-0008","001-0013","010-0009","004-0041","005-0047")))

#plot/Supp Figure 2
par(mar=c(5.1,5.1,4.1,2.1))
boxplot(-final_transplant$symptom_date~final_transplant$transplant_final,xaxt='n',yaxt='n',col="white",border=c("cornflowerblue","orange"),xlab="",ylab="",las=2,cex.lab=1.5,cex.main=1.5,cex.axis=1.2,font.main=1,pch="")
p=summary(lm(final_transplant$symptom_date~final_transplant$transplant_final))$coefficients[2,4]
axis(1,at=1:2,labels=c("Control","Transplant"),cex.axis=1.2)
axis(2,at=c(0,10,20,30,40),labels=c(0,10,20,30,""),cex.axis=1.2,las=2)
mtext(side=2,text="Days since symptom onset",cex=1.5,line=3.5)
stripchart(-final_transplant[which(final_transplant$transplant_final==0),"symptom_date"]~factor(final_transplant$transplant_final[which(final_transplant$transplant_final==0)]),add=T,vertical=T,method="jitter",pch=20,at=1,col="cornflowerblue")
stripchart(-final_transplant[which(final_transplant$transplant_final==1),"symptom_date"]~factor(final_transplant$transplant_final[which(final_transplant$transplant_final==1)]),add=T,vertical=T,method="jitter",pch=20,at=2,col="orange")
text(x=1.5,y=0.9*max(-final_transplant$symptom_date[which(final_transplant$symptom_date>(-300))],na.rm=T),label=paste0("P=",round(p,3)),cex=1.5)

```
