---
title: "blood_cytof"
output: html_document
editor_options: 
  chunk_output_type: inline
---

Source: 20231012_cytof.Rmd

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

library(limma)

# ggplot theme
my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, family="Arial", color="black"),
    text = element_text(size=12, family="Arial", color="black"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```

# Import data

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

### Clinical metadata processing

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE
clin.event$steroids <- as.factor(clin.event$steroids)

head(clin.event)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert sex to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

### Merge clinical metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ, by = "participant_id", all = FALSE, sort = FALSE
)
```

## Blood sample

```{r}
# Count data
bld <- read.csv(
  "/data/bld-cytof/legacy/2023-02-06/bld-cytof-Counts.csv",
  skip = 1, header = FALSE, row.names = 1
)
colnames(bld) <- 1:ncol(bld)
bld <- t(bld)

# Cell type data
bld.rData <- read.csv(
  "/data/bld-cytof/legacy/2023-02-06/bld-cytof-RowFeature.csv",
  skip = 1, header = TRUE, row.names = NULL
)

# QC data
bld.metadata <- read.csv(
  "/data/bld-cytof/legacy/2023-02-06/bld-cytof-Metadata.csv",
  header = TRUE, row.names = NULL
)

# Check if sample_id in counts and metadata match
stopifnot(colnames(bld) %in% bld.metadata$sample_id)
```

Fine cell type annotation.

```{r}
# Remove neutrophils
temp <- subset(bld.rData, Final.Annotated.Population != "Neutrophil (CD16hi)")
# Remove multiplets
temp <- temp[!grepl("^multiplet", temp$Final.Annotated.Population, ignore.case = TRUE),]
# Remove debris
temp <- temp[!grepl("^debris", temp$Final.Annotated.Population, ignore.case = TRUE),]
# Remove Undefined cells, including undefined T cells
temp <- temp[!grepl("Undefined", temp$Final.Annotated.Population, ignore.case = TRUE),]

bld.2f <- bld[rownames(temp),]
```

Do I need to filter more cell types? For each cell type, check the number of patients in which that cell type is detected.

```{r}
# Histogram of number of detected patients per cell type
hist(rowSums(bld.2>0))
```

Blood sample clinical metadata.

```{r}
bld.clin.sample.event.individ <- clin.sample.event.individ %>%
  subset(sample_id %in% bld.pass.qc)
```

## Viral load data

SARS-CoV-2 viral load is readily available in nasal-metagenomics-Metadata.csv, sc2_rpm column.

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/legacy/2022-09-03/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
```

## QC

```{r}
# Keep CyTOF data passing QC
bld.metadata <- subset(bld.metadata,
                       sample_status=="IMPACC sample assayed and passed QC")

# Keep viral load data passing QC
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")
# Remove rows without sc2_rpm
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))

# Get a list of samples fpassing QC
pass.qc <- Reduce(
  intersect, 
  list(colnames(bld), bld.metadata$sample_id, clin.sample.event.individ$sample_id)
)

bld <- bld[,pass.qc]
clin.sample.event.individ <- subset(clin.sample.event.individ, sample_id %in% pass.qc)
```

# Visit 1 analysis

## Data preparation

```{r}
# Keep visit 1 samples for viral load
clin.sample.event.individ.sc2.1 <- clin.sample.event.individ.sc2 %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.sc2.1$sc2_rpm <- log10(clin.sample.event.individ.sc2.1$sc2_rpm + 1)

# Blood
clin.sample.event.individ.1 <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")
# Add viral load
clin.sample.event.individ.1 <- merge(
  clin.sample.event.individ.1, clin.sample.event.individ.sc2.1[,c("event_id","sc2_rpm")],
  by = "event_id", all = FALSE, sort = FALSE
)
clin.sample.event.individ.1 <- droplevels(clin.sample.event.individ.1)
```

Fine cell type annotation

```{r}
# Convert cell count to cell proportion
bld.3f <- apply(bld.2f, MARGIN = 2,
                FUN = function(x) x/sum(x)) # t(t(bld.2f)/colSums(bld.2f)) 

# Subset visit 1 samples
bld.3f <- bld.3f[, clin.sample.event.individ.1$sample_id]

# CLR transform
bld.4f <- log(bld.3f*100 + 1) # Add 1 to avoid log of zero
bld.4f <- apply(bld.4f, MARGIN = 2,
                FUN = function(x) x - mean(x)) # t(t(bld.4f) - colMeans(bld.4f))
```

## Linear regression on age effect: blood

### Without controlling for viral load

```{r}
# Linear regression on clr-transformed cell type proportion
res.clr <- data.frame(matrix(NA, nrow=nrow(bld.4f), ncol=4))
colnames(res.clr) <- c("cell.type","slope","PVal","adj.PVal")
rownames(res.clr) <- rownames(bld.4f)
res.clr$cell.type <- bld.rData[rownames(res.clr),"Final.Annotated.Population"]

for (i in rownames(res.clr)) {
  temp <- data.frame(clr=bld.4f[i,], 
                     age=clin.sample.event.individ.1$admit_age,
                     sex=clin.sample.event.individ.1$sex,
                     sc2=clin.sample.event.individ.1$sc2_rpm,
                     resp=clin.sample.event.individ.1$respiratory_status)%>%
    subset(!is.na(clr))
  my.lm <- lm(clr ~ age + sex + resp, data=temp)
  res.clr[i,"slope"] <- summary(my.lm)$coefficients["age",1]
  res.clr[i,"PVal"] <- summary(my.lm)$coefficients["age",4]
  
  remove(temp, my.lm)
}

# Correct P-values
res.clr$adj.PVal <- p.adjust(res.clr$PVal, method = "BH")

# Add cell type
res.clr$cell.type <- bld.rData[rownames(res.clr),"Final.Annotated.Population"]
```

Plot slopes of significant cell types.

```{r}
# Figure 2B
p <- ggplot() +
  geom_col(data = subset(res.clr, adj.PVal<0.05),
           aes(x = slope, y = reorder(cell.type,slope),
               fill = as.factor(slope>0)),
           width = 0.7) +
  geom_vline(xintercept=0, linewidth=0.5, color="black") +
  scale_fill_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  labs(x = "Slope\n(change per 1 year of age)",
       y = "", title = "Cell types associated\nwith age at Visit 1") +
  xlim(-0.022, 0.022) + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank()
  ) # cytof_visit1_age_no-viralload, 6 x 6.2 in (w x h)
```

Plot examples.

```{r}
# Figure 2C
p <- list()
counter <- 1
my.colors <- c("darkred","darkred","dodgerblue4","dodgerblue4")
for (i in c("Monocytes (CD14+CD16+)","NK Cell (CD56low CD16hi CD57hi)","CD8+ T Cell (naive)","B Cell (naive)")) {
  # Row number that corresponds to cell type i
  row.i <- rownames(bld.rData)[bld.rData$Final.Annotated.Population==i]
  temp <- data.frame(clr=bld.4f[row.i,], 
                     age=clin.sample.event.individ.1$admit_age,
                     sex=clin.sample.event.individ.1$sex,
                     sc2=clin.sample.event.individ.1$sc2_rpm,
                     resp=clin.sample.event.individ.1$respiratory_status) %>%
    subset(!is.na(clr))
  my.lm <- lm(clr ~ age + sex + resp, data=temp)
  my.pred <- ggeffects::ggpredict(my.lm, "age")
  
  p[[i]] <- ggplot() +
    geom_point(data = temp, aes(x=age,y=clr), size = 0.5, color="black") +
    geom_line(data = my.pred, aes(x,predicted), size = 1, color=my.colors[counter]) +
    geom_ribbon(data = my.pred, 
                aes(x = x, y = predicted, ymin = conf.low, ymax = conf.high),
                fill = my.colors[counter], alpha = .4) +
    scale_x_continuous(limits=c(0,100), expand=c(0,0)) +
    scale_y_continuous(limits=c(-1,max(temp$clr))) +
    labs(x = "Age",
         y = "Cell type abundance\n(CLR-transformed)",
         title = sprintf("%s",i)) +
    theme_classic() + 
    theme(
      plot.title = element_text(hjust = 0.5, size=12, face="plain"),
      axis.text = element_text(size=12, family="Arial", color="black"),
      text = element_text(size=12, family="Arial", color="black"),
      plot.margin = unit(c(0,1.2,0.5,0), "cm")
    )
  counter <- counter + 1
}

p[[1]]+p[[2]]+p[[3]]+p[[4]] + plot_layout(ncol=2) # cytof_visit1_age_no-viralload_examples, 6.6 x 6.2 in
```

### Controlled for viral load

```{r}
# Linear regression on clr-transformed cell type proportion
res.clr.sc2 <- data.frame(matrix(NA, nrow=nrow(bld.4f), ncol=4))
colnames(res.clr.sc2) <- c("cell.type","slope","PVal","adj.PVal")
rownames(res.clr.sc2) <- rownames(bld.4f)
res.clr.sc2$cell.type <- bld.rData[rownames(res.clr.sc2),"Final.Annotated.Population"]

for (i in rownames(res.clr.sc2)) {
  temp <- data.frame(clr=bld.4f[i,], 
                     age=clin.sample.event.individ.1$admit_age,
                     sex=clin.sample.event.individ.1$sex,
                     sc2=clin.sample.event.individ.1$sc2_rpm,
                     resp=clin.sample.event.individ.1$respiratory_status)
  my.lm <- lm(clr ~ age + sc2 + sex + resp, data=temp)
  res.clr.sc2[i,"slope"] <- summary(my.lm)$coefficients["age",1]
  res.clr.sc2[i,"PVal"] <- summary(my.lm)$coefficients["age",4]
  
  remove(temp, my.lm)
}

# Correct P-values
res.clr.sc2$adj.PVal <- p.adjust(res.clr.sc2$PVal, method = "BH")
```

Plot slopes of significant cell types.

```{r}
# Figure S4
p <- ggplot() +
  geom_col(data = subset(res.clr.sc2, adj.PVal<0.05),
           aes(x = slope, y = reorder(cell.type,slope),
               fill = as.factor(slope>0)),
           width = 0.7) +
  geom_vline(xintercept=0, size=0.5, color="black") +
  scale_fill_manual(values=c("TRUE"="darkred", "FALSE"="dodgerblue4"), guide="none") +
  # coord_flip() +
  labs(x = "Slope\n(change per 1 year of age)",
       y = "", title = "Cell types associated\nwith age at Visit 1\n(controlled for viral load)") +
  xlim(-0.022, 0.022) + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank()
  ) # cytof_visit1_age_viralload, 6 x 7 in (w x h)
```

# Visit 1 analysis: gene expression vs monocytes

Here I will look at what genes are associated with each monocyte subsets.

## Load PBMC RNA-seq data

```{r}
pbmc.counts <- read.csv(
  "/data/pbmc-transcriptomics/legacy/2022-06-15/pbmc-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1,
  check.names = FALSE
)
pbmc.counts <- t(pbmc.counts)
print(pbmc.counts[1:6,1:6])

pbmc.metadata <- read.csv(
  "/data/pbmc-transcriptomics/legacy/2022-06-15/pbmc-transcriptomics-Metadata.csv",
  check.names = FALSE
)
print(head(pbmc.metadata))
stopifnot(colnames(pbmc.counts) %in% pbmc.metadata$sample_id) # do sample_id match

pbmc.rData <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-RowFeature.csv",
  header = FALSE, row.names = NULL,
  check.names = FALSE
)
stopifnot(rownames(pbmc.counts)==pbmc.rData$gene_id) # do gene ID match

colnames(pbmc.rData) <- pbmc.rData[2,]
pbmc.rData <- pbmc.rData[-(1:2),]
print(head(pbmc.rData))
```

QC

```{r}
pbmc.metadata <- pbmc.metadata %>%
  subset(sample_status == "IMPACC sample assayed and passed QC")
```

## Load Reactome pathways

```{r}
# Reactome
reactome <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "C2", subcategory = "CP:REACTOME")
reactome.list <- split(x = reactome$ensembl_gene, f = reactome$gs_name) # for fgsea
```

## Prepare overlapping PBMC and CyTOF data

```{r}
# Prepare temporary clin.sample.event.individ for Visit 1
temp <- merge(
  clin.sample.event, clin.individ, by = "participant_id",
  all = FALSE, sort = FALSE) %>%
  subset((event_type=="Visit 1") & (event_location=="Hospital"))

# Add viral load, also ensuring that I only use samples with valid SARS-CoV-2 viral rpM
temp <- merge(
  temp, clin.sample.event.individ.sc2[,c("event_id","sc2_rpm")],
  by = "event_id", all = FALSE, sort = FALSE)

# Filter for samples passing RNA-seq IMPACC QC
cytof.clin.sample.event.individ.1 <- temp %>%
  subset(sample_id %in% bld.metadata$sample_id)
pbmc.clin.sample.event.individ.1 <- temp %>%
  subset(sample_id %in% pbmc.metadata$sample_id)

# Overlapping patients between cytof and pbmc RNA-seq data
overlap <- intersect(
  cytof.clin.sample.event.individ.1$event_id,
  pbmc.clin.sample.event.individ.1$event_id)
cytof.clin.sample.event.individ.1 <- cytof.clin.sample.event.individ.1 %>%
  subset(event_id %in% overlap)
pbmc.clin.sample.event.individ.1 <- pbmc.clin.sample.event.individ.1 %>%
  subset(event_id %in% overlap)

# Reorder
cytof.clin.sample.event.individ.1 <- cytof.clin.sample.event.individ.1[order(cytof.clin.sample.event.individ.1$event_id),]
pbmc.clin.sample.event.individ.1 <- pbmc.clin.sample.event.individ.1[order(pbmc.clin.sample.event.individ.1$event_id),]

# Verify that the participant_id match
stopifnot(cytof.clin.sample.event.individ.1$participant_id==cytof.clin.sample.event.individ.1$participant_id)
stopifnot(cytof.clin.sample.event.individ.1$event_id==pbmc.clin.sample.event.individ.1$event_id)
stopifnot(!any(duplicated(cytof.clin.sample.event.individ.1$participant_id)))

remove(temp)
```

## Merge monocyte abundance data

Here I will add monocyte abundance data to the pbmc metadata for DE analysis.

```{r}
# Verify that my pre-calculated CLR data contains all the samples I will be using
stopifnot(cytof.clin.sample.event.individ.1$sample_id %in% colnames(bld.4f))

# Extract cytof samples to be merged
temp <- bld.4f[,cytof.clin.sample.event.individ.1$sample_id]
colnames(temp) <- cytof.clin.sample.event.individ.1$event_id
rownames(temp) <- bld.rData[rownames(temp),"Final.Annotated.Population"]

# Reformat cytof data, so that rows are event_id, columns are monocytes
temp <- temp[grep("^Monocyte", rownames(temp), ignore.case=TRUE),] %>%
  t() %>% # transpose
  as.data.frame() %>%
  tibble::rownames_to_column(var="event_id")

# Rename monocytes
colnames(temp)[colnames(temp)=="Monocytes (CD14+CD16+)"] <- "Monocyte_int" # intermediate monocyte
colnames(temp)[colnames(temp)=="Monocytes (CD14+CD16-)"] <- "Monocyte_c" # classical monocyte
colnames(temp)[colnames(temp)=="Monocytes (CD14-CD16+)"] <- "Monocyte_nc" # non-classical monocyte

# Calculate the log ratio of c to nc monocytes
temp <- temp %>%
  mutate(Monocyte_c_to_nc = Monocyte_c - Monocyte_nc)

# Merge cytof data
pbmc.clin.sample.event.individ.1 <- merge(
  pbmc.clin.sample.event.individ.1, temp,
  by="event_id", all=FALSE)
```

## DE vs monocyte abundance in youngest and oldest patients (combined design)

```{r}
pbmc.clin.sample.event.individ.1.yo <- pbmc.clin.sample.event.individ.1 %>%
  subset((admit_age<=46) | (admit_age>=71))

# Age group
pbmc.clin.sample.event.individ.1.yo$age_old <- as.factor(
  pbmc.clin.sample.event.individ.1.yo$admit_age >= 71)

# Drop unused levels
pbmc.clin.sample.event.individ.1.yo$respiratory_status <- droplevels(
  pbmc.clin.sample.event.individ.1.yo$respiratory_status)

```

```{r}
# Filter Visit 1 PBMC gene counts
pbmc.counts.1 <- pbmc.counts[,pbmc.clin.sample.event.individ.1.yo$sample_id]
colnames(pbmc.counts.1) <- pbmc.clin.sample.event.individ.1.yo$event_id

# Filter genes
pbmc.counts.1 <- pbmc.counts.1[pbmc.rData[pbmc.rData$gene_biotype=="protein_coding", "gene_id"],]
keep <- rowSums(pbmc.counts.1>=10) >= 0.2*ncol(pbmc.counts.1)
pbmc.counts.1 <- pbmc.counts.1[keep,]
```

### Classical monocytes

```{r}
design <- model.matrix(~ age_old + age_old:Monocyte_c + sex + respiratory_status,
                       data=pbmc.clin.sample.event.individ.1.yo)

vwts <- voom(pbmc.counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts, design=design)
vfit <- eBayes(vfit)

# DE results in old group
pbmc.res.o <- topTable(
  vfit, coef = "age_oldTRUE:Monocyte_c",
  sort.by = "none", number = Inf,
  adjust.method = "BH", p.value = 1)

# DE results in young group
pbmc.res.y <- topTable(
  vfit, coef = "age_oldFALSE:Monocyte_c",
  sort.by = "none", number = Inf,
  adjust.method = "BH", p.value = 1)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

Volcano plot

```{r}
temp <- rbind(
  pbmc.res.o %>%
    tibble::rownames_to_column(var="gene_id") %>%
    mutate(age="[71,96]"),
  pbmc.res.y %>%
    tibble::rownames_to_column(var="gene_id") %>%
    mutate(age="[18,46]")
)

# Reorder age group order
temp$age <- factor(
  temp$age,
  levels=c("[71,96]","[18,46]")
)

# Limit minimum P value to 1e-10
temp[temp$adj.P.Val<1e-10,"adj.P.Val"] <- 1e-10

# Figure S12A
p <- ggplot(data=temp, aes(x=logFC, y=-log10(adj.P.Val))) +
  geom_point(
    data=. %>% subset(adj.P.Val>=0.05),
    color="gray75", size=0.4) +
  geom_point(
    data=. %>% subset((adj.P.Val<0.05) & (logFC>0)),
    color="#7b3294", size=0.4) +
  geom_point(
    data=. %>% subset((adj.P.Val<0.05) & (logFC<0)),
    color="#008837", size=0.4) +
  facet_wrap(~age, ncol=1, scales="free_x") +
  scale_x_continuous(limits=c(-4,4), expand=c(0,0)) +
  scale_y_continuous(limits=c(-0.2,10.2), expand=c(0,0),
                     breaks=c(0,5,10), labels=c(0,5,"\u2265 10")) +
  labs(x="logFC", y="-log10(adjusted P)",
       title="Classical") +
  my.theme # cytof_rnaseq_c-monocytes_volcano, 3.2w x 5.7h in
```

GSEA

```{r}
# Old patients -----
# Sort gene list in descending order
gene.ranks <- pbmc.res.o$t
names(gene.ranks) <- rownames(pbmc.res.o)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.o.monocyte_c <- fgsea::fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)

# Young patients -----
# Sort gene list in descending order
gene.ranks <- pbmc.res.y$t
names(gene.ranks) <- rownames(pbmc.res.y)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.y.monocyte_c <- fgsea::fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
```

### Non-classical monocytes

```{r}
design <- model.matrix(~ age_old + age_old:Monocyte_nc + sex + respiratory_status,
                       data=pbmc.clin.sample.event.individ.1.yo)

vwts <- voom(pbmc.counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts, design=design)
vfit <- eBayes(vfit)

# DE results in old group
pbmc.res.o <- topTable(
  vfit, coef = "age_oldTRUE:Monocyte_nc",
  sort.by = "none", number = Inf,
  adjust.method = "BH", p.value = 1)

# DE results in young group
pbmc.res.y <- topTable(
  vfit, coef = "age_oldFALSE:Monocyte_nc",
  sort.by = "none", number = Inf,
  adjust.method = "BH", p.value = 1)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

Volcano plot

```{r}
temp <- rbind(
  pbmc.res.o %>%
    tibble::rownames_to_column(var="gene_id") %>%
    mutate(age="[71,96]"),
  pbmc.res.y %>%
    tibble::rownames_to_column(var="gene_id") %>%
    mutate(age="[18,46]")
)

# Reorder age group order
temp$age <- factor(
  temp$age,
  levels=c("[71,96]","[18,46]")
)
# Limit minimum P value to 1e-10
temp[temp$adj.P.Val<1e-10,"adj.P.Val"] <- 1e-10

# Figure S12B
p <- ggplot(data=temp, aes(x=logFC, y=-log10(adj.P.Val))) +
  geom_point(
    data=. %>% subset(adj.P.Val>=0.05),
    color="gray75", size=0.4) +
  geom_point(
    data=. %>% subset((adj.P.Val<0.05) & (logFC>0)),
    color="#7b3294", size=0.4) +
  geom_point(
    data=. %>% subset((adj.P.Val<0.05) & (logFC<0)),
    color="#008837", size=0.4) +
  facet_wrap(~age, ncol=1, scales="free_x") +
  scale_x_continuous(limits=c(-4,4), expand=c(0,0)) +
  scale_y_continuous(limits=c(-0.2,10.2), expand=c(0,0),
                     breaks=c(0,5,10), labels=c(0,5,"\u2265 10")) +
  labs(x="logFC", y="-log10(adjusted P)",
       title="Non-classical") +
  my.theme # cytof_rnaseq_nc-monocytes_volcano, 3.2w x 5.7h in
```

GSEA

```{r}
# Old patients -----
# Sort gene list in descending order
gene.ranks <- pbmc.res.o$t
names(gene.ranks) <- rownames(pbmc.res.o)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.o.monocyte_nc <- fgsea::fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)

# Young patients -----
# Sort gene list in descending order
gene.ranks <- pbmc.res.y$t
names(gene.ranks) <- rownames(pbmc.res.y)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.y.monocyte_nc <- fgsea::fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
```

### Intermediate monocytes

```{r}
design <- model.matrix(~ age_old + age_old:Monocyte_int + sex + respiratory_status,
                       data=pbmc.clin.sample.event.individ.1.yo)

vwts <- voom(pbmc.counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts, design=design)
vfit <- eBayes(vfit)

# DE results in old group
pbmc.res.o <- topTable(
  vfit, coef = "age_oldTRUE:Monocyte_int",
  sort.by = "none", number = Inf,
  adjust.method = "BH", p.value = 1)

# DE results in young group
pbmc.res.y <- topTable(
  vfit, coef = "age_oldFALSE:Monocyte_int",
  sort.by = "none", number = Inf,
  adjust.method = "BH", p.value = 1)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

Volcano plot

```{r}
temp <- rbind(
  pbmc.res.o %>%
    tibble::rownames_to_column(var="gene_id") %>%
    mutate(age="[71,96]"),
  pbmc.res.y %>%
    tibble::rownames_to_column(var="gene_id") %>%
    mutate(age="[18,46]")
)

# Reorder age group order
temp$age <- factor(
  temp$age,
  levels=c("[71,96]","[18,46]")
)

# Limit minimum P value to 1e-10
temp[temp$adj.P.Val<1e-10,"adj.P.Val"] <- 1e-10

# Figure S12C
p <- ggplot(data=temp, aes(x=logFC, y=-log10(adj.P.Val))) +
  geom_point(
    data=. %>% subset(adj.P.Val>=0.05),
    color="gray75", size=0.4) +
  geom_point(
    data=. %>% subset((adj.P.Val<0.05) & (logFC>0)),
    color="#7b3294", size=0.4) +
  geom_point(
    data=. %>% subset((adj.P.Val<0.05) & (logFC<0)),
    color="#008837", size=0.4) +
  facet_wrap(~age, ncol=1, scales="free_x") +
  scale_x_continuous(limits=c(-4,4), expand=c(0,0)) +
  scale_y_continuous(limits=c(-0.2,10.2), expand=c(0,0),
                     breaks=c(0,5,10), labels=c(0,5,"\u2265 10")) +
  labs(x="logFC", y="-log10(adjusted P)",
       title="Intermediate") +
  my.theme # cytof_rnaseq_int-monocytes_volcano, 3.2w x 5.7h in
```

GSEA

```{r}
# Old patients -----
# Sort gene list in descending order
gene.ranks <- pbmc.res.o$t
names(gene.ranks) <- rownames(pbmc.res.o)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.o.monocyte_int <- fgsea::fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)

# Young patients -----
# Sort gene list in descending order
gene.ranks <- pbmc.res.y$t
names(gene.ranks) <- rownames(pbmc.res.y)
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
reactome.gsea.y.monocyte_int <- fgsea::fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
```

### Compare GSEA results of these 3 monocyte subtypes

```{r}
# Manually choose pathways
my.pathways <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING",
                 "REACTOME_INTERFERON_GAMMA_SIGNALING",
                 "REACTOME_INTERLEUKIN_1_FAMILY_SIGNALING",
                 "REACTOME_INTERLEUKIN_10_SIGNALING", "REACTOME_INTERLEUKIN_12_SIGNALING",
                 "REACTOME_TCR_SIGNALING", "REACTOME_PD_1_SIGNALING",
                 "REACTOME_COSTIMULATION_BY_THE_CD28_FAMILY",
                 "REACTOME_ANTIGEN_ACTIVATES_B_CELL_RECEPTOR_BCR_LEADING_TO_GENERATION_OF_SECOND_MESSENGERS",
                 "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR",
                 "REACTOME_DOWNSTREAM_SIGNALING_EVENTS_OF_B_CELL_RECEPTOR_BCR",
                 "REACTOME_NEUTROPHIL_DEGRANULATION",
                 "REACTOME_CLASS_I_MHC_MEDIATED_ANTIGEN_PROCESSING_PRESENTATION"
                 )
```

```{r}
# Combine results
combined.gsea <- rbind(
  reactome.gsea.o.monocyte_c %>%
    subset(pathway %in% my.pathways) %>%
    mutate(subtype="Classical", age="[71,96]"),
  reactome.gsea.o.monocyte_nc %>%
    subset(pathway %in% my.pathways) %>%
    mutate(subtype="Non-classical", age="[71,96]"),
  reactome.gsea.o.monocyte_int %>%
    subset(pathway %in% my.pathways) %>%
    mutate(subtype="Intermediate", age="[71,96]"),
  reactome.gsea.y.monocyte_c %>%
    subset(pathway %in% my.pathways) %>%
    mutate(subtype="Classical", age="[18,46]"),
  reactome.gsea.y.monocyte_nc %>%
    subset(pathway %in% my.pathways) %>%
    mutate(subtype="Non-classical", age="[18,46]"),
  reactome.gsea.y.monocyte_int %>%
    subset(pathway %in% my.pathways) %>%
    mutate(subtype="Intermediate", age="[18,46]")
)

# Reorder subset order
combined.gsea$subtype <- factor(
  combined.gsea$subtype,
  levels=c("Classical","Non-classical","Intermediate")
)
# Reorder age group order
combined.gsea$age <- factor(
  combined.gsea$age,
  levels=c("[71,96]","[18,46]")
)

# Reformat pathway names
combined.gsea$pathway <- gsub("^REACTOME_","",combined.gsea$pathway)
combined.gsea$pathway <- gsub("_"," ",combined.gsea$pathway)
combined.gsea$pathway <- stringr::str_to_sentence(combined.gsea$pathway)

# Fix capitalization
my.dict <- list("Interleukin "="IL", "interleukin "="IL",
                "Pd 1"="PD1", "Tcr"="TCR",
                "cd28"="CD28",
                "b cell receptor "="", "bcr"="BCR",
                "Class i mhc"="MHC class I")
for (i in names(my.dict)) {
  combined.gsea$pathway <- gsub(i, my.dict[[i]], combined.gsea$pathway)
}
```

Plot

```{r}
# Figure S12D
p1 <- ggplot(data=combined.gsea, 
            aes(y=reorder(pathway,NES,FUN=function(x) x[1]),
                x=NES)) +
  geom_point(aes(color=as.factor(NES>0)), fill="white",
             shape=21, size=2.5, stroke=1) +
  geom_point(data=. %>% subset(padj<0.05),
             aes(color=as.factor(NES>0), fill=as.factor(NES>0)),
             shape=21, size=2.5, stroke=1) +
  geom_vline(xintercept = 0, linewidth = 0.7, color = "black") +
  scale_color_manual(values=c("TRUE"="#7b3294", "FALSE"="#008837"), guide="none") +
  scale_fill_manual(values=c("TRUE"="#7b3294", "FALSE"="#008837"), guide="none") +
  scale_x_continuous(limits = c(-3,3), expand = c(0,0)) +
  labs(y="", x="GSEA normalized\nenrichment score",
       title="PBMC pathways associated with\nmonocyte subset abundance at Visit 1") +
  facet_wrap(~age + subtype, nrow=2, scales="free_x") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=11, color="black", family="Arial"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    panel.grid.minor=element_blank()
  ) # cytof_rnaseq_3monocytes_gsea-reactome, 10w x 8.7h in
```
