---
title: "nasal_viralload"
output: html_document
---

Here I will analyze viral load from nasal mNGS data, and compare viral load by mNGS vs qPCR.

Source: 20230426_viralload.Rmd

```{r}
library(dplyr)
library(ggplot2)
library(pheatmap)
library(RColorBrewer)
library(patchwork)

library(lme4)
library(gamm4)

# library(fgsea)

setwd("~/20220902_IMPACC_aging")

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )
```

# Import data

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/current/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

### Clinical metadata processing

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE
clin.event$steroids <- as.factor(clin.event$steroids)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert some columns to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

## Viral load data

Get SARS-CoV-2 viral load from nasal-metagenomics-AllGenera-Counts_RPM.csv, sc2_rpm column.

```{r}
# mNGS metadata, containing viral load by mNGS
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)

# qPCR data
pcr <- read.csv(
  "/data/nasal-viralload/current/nasal-viralload-Counts.csv",
  header=TRUE, row.names=NULL,
  check.names=FALSE
)
pcr.metadata <- read.csv(
 "/data/nasal-viralload/current/nasal-viralload-Metadata.csv",
 header=TRUE, row.names=NULL,
 check.names=FALSE
)
```

# Merge metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

```{r}
# mNGS QC
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))
```

```{r}
# qPCR QC
pcr.metadata <- subset(pcr.metadata,
                       sample_status=="IMPACC sample assayed and passed QC")
pcr <- subset(pcr, (sample_id %in% pcr.metadata$sample_id) &
                !is.na(N1_CT) & !is.na(N2_CT))

# Merge with clinical metadata
clin.sample.event.individ.pcr <- merge(
  clin.sample.event.individ, pcr[,c("sample_id","N1_CT","N2_CT")],
  by="sample_id", all=FALSE, sort=FALSE
)
```

# Visit 1 Analysis

```{r}
# Only keep Visit 1 samples
clin.sample.event.individ.visit1 <- clin.sample.event.individ.sc2 %>%
  subset(event_type=="Visit 1")

# log-transform rpm
clin.sample.event.individ.visit1$sc2_rpm_log10 <- log10(clin.sample.event.individ.visit1$sc2_rpm+1)

# Divide into age groups
clin.sample.event.individ.visit1$age_group <- cut(
  clin.sample.event.individ.visit1$admit_age,
  breaks=c(17,46,54,62,70,96)
  )
# Rename levels: convert round bracket to square bracket
levels(clin.sample.event.individ.visit1$age_group) <- c("[18,46]","[47,54]","[55,62]","[63,70]","[71,96]")
```

Plot

```{r}
# Figure 2D

p1 <- ggplot(clin.sample.event.individ.visit1,
            aes(x=age_group,
                y=sc2_rpm_log10,
                color=age_group)) + 
  stat_boxplot(geom ="errorbar", width=0.3) + 
  geom_boxplot(fill="white", outlier.shape=NA) +
  ggbeeswarm::geom_quasirandom(width=0.35, size=0.5) +
  # geom_jitter(width=0.2, size=0.5, alpha=0.5) +
  labs(x="Age group", y="Viral load") +
  scale_color_viridis_d(guide="none") +
  # scale_y_continuous(breaks = c(20, 40, 60, 80, 100), limits = c(15, 100)) +
  theme_classic() + theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank()
  ) # viralload_visit1, 3.5 x 3 in
```

P value

```{r}
print(kruskal.test(
  sc2_rpm_log10 ~ age_group, data=clin.sample.event.individ.visit1
))
```

## Compare mNGS vs qPCR

```{r}
# Add qPCR data
clin.sample.event.individ.visit1 <- merge(
  clin.sample.event.individ.visit1, clin.sample.event.individ.pcr[,c("event_id","N1_CT","N2_CT")],
  by="event_id", all=FALSE, sort=FALSE
)

print(unique(clin.sample.event.individ.visit1$event_location))
```

```{r}
p1 <- ggplot(data=clin.sample.event.individ.visit1,
             aes(x=N1_CT, y=sc2_rpm_log10)) +
  geom_point(size=1) +
  labs(x="qPCR (N1 CT)", y="Metatranscriptomics\n(log-transformed rpM)") +
  my.theme

p2 <- ggplot(data=clin.sample.event.individ.visit1,
             aes(x=N2_CT, y=sc2_rpm_log10)) +
  geom_point(size=1) +
  labs(x="qPCR (N2 CT)", y="Metatranscriptomics\n(log-transformed rpM)") +
  my.theme

# Supplementary figure 1
p1 + p2 + plot_layout(nrow=1) # viralload_mngs-vs-qpcr.pdf, 6 x 3 in
```

```{r}
print(cor.test(
  clin.sample.event.individ.visit1$N1_CT, clin.sample.event.individ.visit1$sc2_rpm_log10
))

print(cor.test(
  clin.sample.event.individ.visit1$N2_CT, clin.sample.event.individ.visit1$sc2_rpm_log10
))
```

# Longitudinal analysis

```{r}
# Remove escalation samples
clin.sample.event.individ.long <- clin.sample.event.individ.sc2 %>%
  subset(!grepl("Escalation", event_type) & event_date<=30)

# Divide into age groups
clin.sample.event.individ.long$age_group <- cut(
  clin.sample.event.individ.long$admit_age,
  breaks=c(17,46,54,62,70,96)
  )

# Rename levels: convert round bracket to square bracket
levels(clin.sample.event.individ.long$age_group) <- c("[18,46]","[47,54]","[55,62]","[63,70]","[71,96]")

# log-transform viral load
clin.sample.event.individ.long$sc2_rpm_log10 <- log10(clin.sample.event.individ.long$sc2_rpm+1)
```

Generalized additive mixed modeling

```{r}
# Prepare ordered age group
clin.sample.event.individ.long$age_group_ordered <- ordered(clin.sample.event.individ.long$age_group)
fit <- gamm4(
    sc2_rpm_log10 ~ s(event_date, bs="cr") + s(event_date, bs="cr", by=age_group_ordered) + age_group_ordered + trajectory_group + sex, 
    data = clin.sample.event.individ.long, random = ~(1|enrollment_site/participant_id))
  
aa <- anova(fit$gam)
  
# Get P-values for the interaction term across all age-groups
aa1 <- sum(aa$s.table[-c(1), "F"]*aa$s.table[-c(1), "Ref.df"])
aa2 <- sum(aa$s.table[-c(1), "Ref.df"])
pval <- pchisq(aa1, aa2, lower.tail = FALSE)
```

Plot

```{r}
# Predict
my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","age_group_ordered"),
                                type="fixed")
colnames(my.pred)[colnames(my.pred)=="group"] <- "age_group"

clin.sample.event.individ.long$yhat <- predict(fit$mer)

# Figure 2E
# Longitudinal plot
p4 <- ggplot(data=clin.sample.event.individ.long, 
             aes(x=event_date, y=sc2_rpm_log10, color=age_group)) +
  geom_line(aes(group=participant_id), color="gray", alpha=0.4) +
  # geom_line(inherit.aes=FALSE, aes(x=event_date, y=yhat, group=participant_id),
  #           color="#5F5F5F", alpha=0.9) +
  geom_point(aes(color=age_group),
             shape=16, size=1, alpha=0.7) +
  geom_line(inherit.aes=FALSE, data=my.pred, aes(x,predicted),
            color="black", linewidth=1.5) +
  labs(x="Days from admission", y="Viral load",
       title="") +
  facet_wrap(~ age_group, nrow=1) +
  scale_color_viridis_d(guide="none") +
  theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5, size=12, face="plain"),
      axis.text.x = element_text(size=11, color="black"),
      axis.text.y = element_text(size=12, color="black"),
      text = element_text(size=12, family="Arial"),
      plot.margin = unit(c(0,1,0.5,0), "cm"),
      panel.spacing = unit(0.25, "cm"),
      panel.grid.minor=element_blank()
    ) # viralload_longitudinal.pdf, 7 x 3.1 in
```

