---
title: "serum_RBD_IgG"
output: html_document
---

Here I will analyze serum IgG titer.

Source: 20231023_IgG.Rmd

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(patchwork)

library(lme4)
library(gamm4)

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )
```

# Import data

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

### Clinical metadata processing

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert some columns to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

## IgG titer

```{r}
# Count matrix (4434 samples)
# Keep it in sample-by-feature format
igg.counts <- read.csv(
  "/data/serum-rbd-abtiters/legacy/2022-10-11/serum-rbd-abtiters-Counts.csv",
  header = TRUE, row.names = 1
)
# igg.counts <- t(igg.counts) 

# IgG metadata
igg.metadata <- read.csv(
  "/data/serum-rbd-abtiters/legacy/2022-10-11/serum-rbd-abtiters-Metadata.csv",
  header = TRUE, row.names = NULL
)

# Verify the sample_id of the two files match
stopifnot(rownames(igg.counts) == igg.metadata$sample_id)
```

## Viral load data

Get SARS-CoV-2 viral load from nasal-metagenomics-AllGenera-Counts_RPM.csv, sc2_rpm column.

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/legacy/2022-09-03/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
```

## Merge data

```{r}
# Only keep samples that pass QC
igg.metadata <- subset(igg.metadata,
                       sample_status=="IMPACC sample assayed and passed QC")
pass.qc <- intersect(
  igg.metadata$sample_id, clin.sample.event.individ$sample_id
) # 4113 samples

# mNGS QC
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))

igg.counts <- igg.counts[pass.qc,]
clin.sample.event.individ <- subset(clin.sample.event.individ, sample_id %in% pass.qc)
```

# Analysis: Visit 1

## Prepare Visit 1 data

```{r}
# Keep visit 1 samples for RNA-seq
clin.sample.event.individ.1 <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Keep visit 1 samples for viral load
clin.sample.event.individ.sc2.1 <- clin.sample.event.individ.sc2 %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Add viral load
clin.sample.event.individ.1 <- merge(
  clin.sample.event.individ.1, clin.sample.event.individ.sc2.1[,c("event_id","sc2_rpm")],
  by = "event_id", all = FALSE, sort = FALSE
)
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.1$sc2_rpm <- log10(clin.sample.event.individ.1$sc2_rpm + 1)

# Drop unused levels
clin.sample.event.individ.1 <- droplevels(clin.sample.event.individ.1)

# Divide into age groups
clin.sample.event.individ.1$age_group <- cut(
  clin.sample.event.individ.1$admit_age,
  breaks=c(17,46,54,62,70,96)
  )

# Rename levels: convert round bracket to square bracket
levels(clin.sample.event.individ.1$age_group) <- c("[18,46]","[47,54]","[55,62]","[63,70]","[71,96]")

# Any row with no sc2_rpm?
stopifnot(!is.na(clin.sample.event.individ.1$sc2_rpm))

# Any duplicated participant?
stopifnot(!duplicated(clin.sample.event.individ.1$participant_id))
```

```{r}
# Subset Visit 1 samples
igg.counts.1 <- igg.counts[clin.sample.event.individ.1$sample_id,]

# Log-transform (Alexandra used log2 instead)
clin.sample.event.individ.1$log_AUC_RBD <- log10(igg.counts.1$AUC.RBD.IgG)
clin.sample.event.individ.1$log_AUC_Spike <- log10(igg.counts.1$AUC.Spike.IgG)
```

## Continuous age

### Not control for viral load

P-value using Kruskal-Wallis test

```{r}
print(kruskal.test(
  log_AUC_RBD ~ age_group, data=clin.sample.event.individ.1
))
```

Plot

```{r}
# Figure S3A
p <- ggplot(data=clin.sample.event.individ.1, 
            aes(x=age_group, y=log_AUC_RBD,
                color = age_group)) +
  stat_boxplot(geom ='errorbar', width=0.3) +
  geom_boxplot(fill="white", outlier.shape = NA) +
  geom_jitter(size=0.7, width = 0.2) + 
  scale_y_continuous(limits=c(-0.2,6), expand=c(0,0)) +
  labs(x="Age group", y="IgG titer") +
  scale_color_viridis_d(guide="none") +
  my.theme # IgG_visit1.pdf, 4w x 3h in
```

# Longitudinal analysis

```{r}
# Add viral load to clin.sample.event.individ (2364 samples)
clin.sample.event.individ.long <- merge(
  clin.sample.event.individ, clin.sample.event.individ.sc2[,c("event_id","sc2_rpm")],
  by = "event_id",
  all = FALSE, sort = FALSE
)
# Log-transform viral load
clin.sample.event.individ.long$sc2_rpm <- log10(clin.sample.event.individ.long$sc2_rpm+1)

# Remove escalation samples
clin.sample.event.individ.long <- clin.sample.event.individ.long %>%
  subset(!grepl("Escalation", event_type) & event_date<=30)
print(table(clin.sample.event.individ.long$event_type))

# Divide into age groups
clin.sample.event.individ.long$age_group <- cut(
  clin.sample.event.individ.long$admit_age,
  breaks=c(17,46,54,62,70,96)
  )

# Rename levels: convert round bracket to square bracket
levels(clin.sample.event.individ.long$age_group) <- c("[18,46]","[47,54]","[55,62]","[63,70]","[71,96]")
```

```{r}
# Subset longitudinal samples
igg.counts.long <- igg.counts[clin.sample.event.individ.long$sample_id,]

# Log-transform (Alexandra used log2 instead)
clin.sample.event.individ.long$log_AUC_RBD <- log10(igg.counts.long$AUC.RBD.IgG)
clin.sample.event.individ.long$log_AUC_Spike <- log10(igg.counts.long$AUC.Spike.IgG)
```

Look at event_date per visit.

```{r}
p1 <- ggplot(data=clin.sample.event.individ.long,
             aes(x=event_type, y=event_date, color=age_group)) +
  geom_boxplot()
```

## Without controlling for viral load

General additive mixed effects model

```{r}
# Prepare ordered age group
clin.sample.event.individ.long$age_group_ordered <- ordered(clin.sample.event.individ.long$age_group)
fit <- gamm4(
    log_AUC_RBD ~ s(event_date, bs="cr") + s(event_date, bs="cr", by=age_group_ordered) +
      age_group_ordered + trajectory_group + sex, 
    data = clin.sample.event.individ.long, random = ~(1|enrollment_site/participant_id))
  
aa <- anova(fit$gam)
  
# Get P-values for the interaction term across all age-groups
aa1 <- sum(aa$s.table[-c(1), "F"]*aa$s.table[-c(1), "Ref.df"])
aa2 <- sum(aa$s.table[-c(1), "Ref.df"])
pval <- pchisq(aa1, aa2, lower.tail = FALSE)
```

Longitudinal plot.

```{r}
# Figure S3B
my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","age_group_ordered"),
                                type="fixed")
colnames(my.pred)[colnames(my.pred)=="group"] <- "age_group"

clin.sample.event.individ.long$yhat <- predict(fit$mer)

p <- ggplot(data=clin.sample.event.individ.long,
            aes(x=event_date, y=log_AUC_RBD, color=age_group)) +
  geom_line(aes(group=participant_id), color="gray", alpha=0.4) +
  geom_point(aes(color=age_group), 
             shape=16, size=1, alpha=0.7) +
  geom_line(inherit.aes=FALSE, data=my.pred, aes(x,predicted),
            color="black", linewidth=1.5) +
  labs(x="Days from admission", y="IgG titer",
       title="") +
  scale_color_viridis_d(guide="none") + 
  facet_wrap(~ age_group, nrow=1) + 
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=12, color="black"),
    axis.text.y = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,0.5,0), "cm"),
    panel.spacing = unit(0.25, "cm"),
    panel.grid.minor=element_blank()
    ) # IgG_longitudinal, 8.7w x 3h in
```