---
title: "Transplant Olink"
output: html_document
---

Source: 20230413_olink.Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
library(gamm4)


my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```

# Data import

## Clinical metadata

```{r}
# Patient metadata
clin.individ <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-locked.csv",
  header = TRUE)

# Sample
clin.sample <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-sample-locked.csv",
  header = TRUE)

# Event
clin.event <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-event-locked.csv",
  header = TRUE)
```

Merge metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

Manually correct some typos.

```{r}
# Correct the symptom_date of patient 008-0103: change from -376 to -13 (confirmed with data management team)
clin.sample.event.individ[clin.sample.event.individ$participant_id=="008-0103","symptom_date"] <- -13

# Correct event_date of 004-0012-8: change from 2 to 32 (it's 32 in 2021-11-11-frozen data)
clin.sample.event.individ[clin.sample.event.individ$event_id=="004-0012-8","event_date"] <- 32
```

## Transplant metadata 

264 patients included in transplant study (use new metadata table that Harry will share)

```{r}
transplant <- read.delim(
  "/scratch/immunosuppression/2023.03.09_txcasecontrol.txt",
  header = TRUE)
newTG <- read.delim(
  "/scratch/immunosuppression/2023.10.03_new_tgroups.txt",
  header = TRUE)

# Verify that the studyid in the two files are the same
# They are not exactly the same. newTG probably excludes BMT patients
stopifnot(newTG$studyid %in% transplant$studyid)

# Merge new TG assignment
transplant <- merge(
  transplant, newTG,
  by="studyid", all=FALSE
)

# What are the changes in TG?
print(table(transplant$trajectory_group, transplant$newtg))

# Format factor
transplant$transplant_final <- as.factor(transplant$transplant_final)

# How many patients are there?
print(length(unique(transplant$studyid)))
```

Remove BMT patients and their controls.

```{r}
print(unique(transplant$organ_final))

transplant <- transplant %>%
  subset(!(studyid %in% c("002-0022","007-0008","001-0013","010-0009","004-0041","005-0047")))

# How many patients remain?
print(length(unique(transplant$studyid)))

print(unique(transplant$organ_final))
```

Only keep transplant patients in the clinical metadata

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event.individ[,c("participant_id","event_id","sample_id","event_date","event_type","event_location","respiratory_status")], 
  transplant,
  by.x="participant_id", by.y="studyid",
  all=FALSE
)

# How many patients are left?
print(length(unique(clin.sample.event.individ$participant_id)))
```

## Import Olink data

```{r}
# Concentration
# 3655 total samples
olink <- read.csv(
  # "/data/serum-olink/current/Olink-Counts.csv",
  "/data/serum-olink/legacy/2023-01-06/Olink-Counts.csv",
  header = TRUE, row.names = 1
)
olink <- t(olink)

# Row data
rData <- read.csv(
  # "/data/serum-olink/current/Olink-RowFeature.csv",
  "/data/serum-olink/legacy/2023-01-06/Olink-RowFeature.csv",
  header = TRUE, row.names = 3
)
stopifnot(rownames(olink) == rownames(rData))

# Olink metadata
olink.metadata <- read.csv(
  # "/data/serum-olink/current/Olink-Metadata.csv",
  "/data/serum-olink/legacy/2023-01-06/Olink-Metadata.csv",
  header = TRUE, row.names = NULL
)
```

The code below checks for mismatch in the sample_id, which was fixed on 10/17/2023

```{r}
# Check if there are mismatches between olink and olink.metadata
# If there are no mismatch, throw an error (to check if mismatches have been fixed)
# stopifnot(!all(colnames(olink) == olink.metadata$sample_id))

# Export mismatch between olink and olink.metadata
# write.csv(as.data.frame(
#   colnames(olink)[!(colnames(olink) %in% olink.metadata$sample_id)]),
#   "serum-olink_2023-01-06_sample_id_in-counts_not-in-metadata.csv"
# )
# write.csv(as.data.frame(
#   olink.metadata[!(olink.metadata$sample_id %in% colnames(olink)),"sample_id"]),
#   "serum-olink_2023-01-06_sample_id_in-metadata_not-in-counts.csv"
# )
```


```{r}
# Only keep samples that pass QC (3474 samples)
olink.metadata <- subset(olink.metadata, sample_status=="IMPACC sample assayed and passed QC")
clin.sample.event.individ <- clin.sample.event.individ %>%
  subset(sample_id %in% olink.metadata$sample_id)
```

```{r}
# Filter Olink data to samples of interest
olink <- olink[,clin.sample.event.individ$sample_id] # 787 samples
```

```{r}
# Set concentrations below LOD to 0
lod <- replicate(ncol(olink), rData$LOD)
olink[olink < lod] <- 0

# Only keep cytokines detected in at least 20% of the samples (84/92 cytokines)
olink <- olink[rowSums(olink>0, na.rm = TRUE) >= 0.2*ncol(olink),]

# # Assume NA samples as 0
# olink[is.na(olink)] <- 0
```

## Import viral load data

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  # "/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv",
  "/data/nasal-metagenomics/legacy/2022-09-03/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)

mngs.metadata <- mngs.metadata %>%
  filter(!is.na(sc2_rpm))

# mNGS QC
mngs.metadata <- mngs.metadata %>%
  filter(sample_status == "IMPACC sample assayed and passed QC")
```

```{r}
# Reformat visit_num: decapitalize "VISIT"
mngs.metadata$visit_num <- gsub(
  "^VISIT", "Visit", mngs.metadata$visit_num
)

# Merge with sample metadata
clin.sample.event.individ <- merge(
  clin.sample.event.individ, mngs.metadata[,c("participant_id","visit_num","sc2_rpm")],
  by.x=c("participant_id","event_type"), by.y=c("participant_id","visit_num"),
  all.x=TRUE, all.y=FALSE # keep samples that don't pass SARS-CoV-2 metatranscriptomics QC
  # all=FALSE # discard samples that don't pass SARS-CoV-2 metatranscriptomics QC <<< do NOT do this
)

# log-transform viral load
clin.sample.event.individ$sc2_rpm_log10 <- log10(clin.sample.event.individ$sc2_rpm+1)
```

# Analysis: Visit 1 only

```{r}
# Sample metadata (241 samples out of 258 patients)
clin.sample.event.individ.visit1 <- clin.sample.event.individ %>%
  subset(event_type == "Visit 1")

# Remove sample "0865-002K3G00-002", because the vast majority of its cytokines are 0
clin.sample.event.individ.visit1 <- clin.sample.event.individ.visit1 %>%
  subset(sample_id != "0865-002K3G00-002")

# Subset olink data
olink.visit1 <- olink[,clin.sample.event.individ.visit1$sample_id]

# Scale protein concentration
# olink.visit1 <- t(scale(t(olink.visit1)))
olink.visit1 <- t(apply(
  olink.visit1, MARGIN=1, FUN = function(x)
    (x-mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
))
```

```{r}
# #For plotting use this instead
# olink_visit1_plot <- t(olink_visit1)
# 
# #Merge with sample metadata for plotting
# library(tibble)
# 
# olink_visit1_plot <- rownames_to_column(as.data.frame(olink_visit1_plot), "sample_id")
# olink_visit1_merged <- merge(olink_visit1_plot, visit1, by = "sample_id")
```

## Linear regression: no covariates

```{r}
# Set up table for storing results
res <- data.frame(matrix(NA, nrow=nrow(olink.visit1), ncol=4))
colnames(res) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res) <- rownames(olink.visit1)
res$cytokine <- rownames(res)

# Linear regression 
for (i in rownames(res)) {
  temp <- data.frame(conc=olink.visit1[i,], 
                     transplant=clin.sample.event.individ.visit1$transplant_final,
                     steroids=clin.sample.event.individ.visit1$steroids,
                     sc2=clin.sample.event.individ.visit1$sc2_rpm_log10) %>%
    subset(!is.na(conc)) # don't need to subset for viral rpM here
  
  my.lm <- lm(conc ~ transplant, data=temp)
  res[i,"slope"] <- summary(my.lm)$coefficients["transplant1",1]
  res[i,"PVal"] <- summary(my.lm)$coefficients["transplant1",4]
  remove(temp, my.lm)
}

# Correct P-values
res$adj.PVal <- p.adjust(res$PVal, method = "BH")
```

### Plot significant cytokines

```{r}
p1 <- ggplot(data=subset(res, adj.PVal<0.05),
             aes(x=slope, y=reorder(cytokine,slope))) +
  geom_col(aes(fill=as.factor(slope>0)), width=0.6) +
  geom_vline(xintercept=0, linewidth=0.5, color="black") +
  labs(x="Change in\nprotein expression", y="",
       title="Differentially expressed\nproteins at Visit 1") +
  scale_x_continuous(limits=c(-1,1), expand=c(0,0)) +
  scale_fill_manual(values=c("TRUE"="orange","FALSE"="cornflowerblue"),
                    guide="none") +
  theme_bw() + theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    # axis.text.x = element_text(size=11, color="black"),
    # axis.text.y = element_text(size=10, color="black"),
    axis.text = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank()
  ) # olink_visit1_significant_no-viralload.pdf, 3.6 x 5.5 in
```

## Linear regression: control for viral load

```{r}
# Set up table for storing results
res.sc2 <- data.frame(matrix(NA, nrow=nrow(olink.visit1), ncol=4))
colnames(res.sc2) <- c("cytokine","slope","PVal","adj.PVal")
rownames(res.sc2) <- rownames(olink.visit1)
res.sc2$cytokine <- rownames(res.sc2)

# Linear regression 
for (i in rownames(res.sc2)) {
  temp <- data.frame(conc=olink.visit1[i,], 
                     transplant=clin.sample.event.individ.visit1$transplant_final,
                     steroids=clin.sample.event.individ.visit1$steroids,
                     sc2=clin.sample.event.individ.visit1$sc2_rpm_log10) %>%
    subset(!is.na(conc) & !is.na(sc2))
  
  my.lm <- lm(conc ~ transplant + sc2, data=temp)
  res.sc2[i,"slope"] <- summary(my.lm)$coefficients["transplant1",1]
  res.sc2[i,"PVal"] <- summary(my.lm)$coefficients["transplant1",4]
  remove(temp, my.lm)
}

# Correct P-values
res.sc2$adj.PVal <- p.adjust(res.sc2$PVal, method = "BH")
```

## Plot significant cytokines from both analyses

```{r}
my.cytokines <- union(
  subset(res, adj.PVal<0.05)$cytokine,
  subset(res.sc2, adj.PVal<0.05)$cytokine
)

res.both <- rbind(
  res[my.cytokines,] %>% mutate(control="No"),
  res.sc2[my.cytokines,] %>% mutate(control="Yes")
)
```

Plot both

```{r}
p2 <- ggplot(data=res.both,
       aes(x=slope, y=reorder(cytokine, slope, FUN = function(x) x[1]))) +
  geom_col(aes(color=as.factor(slope>0)), fill="white", width=0.6) +
  geom_col(
    data=. %>% subset(adj.PVal<0.05),
    aes(fill=as.factor(slope>0)), width=0.6) +
  geom_vline(xintercept=0, linewidth=0.4, color="black") +
  geom_text(
    data=. %>% subset(slope>0),
    aes(x=0,label=sprintf("P = %.1e", adj.PVal)), 
    hjust=1.06, size=3) +
  geom_text(
    data=. %>% subset(slope<0),
    aes(x=0, label=sprintf("P = %.1e", adj.PVal)),
    hjust=-0.06, size=3) +
  labs(x="Increase in protein expression", y="",
       title="Differential protein expression at Visit 1") +
  scale_x_continuous(
    limits=c(-1,1), expand=c(0,0)) +
  scale_color_manual(values=c("TRUE"="orange","FALSE"="cornflowerblue"),
                    guide="none") +
  scale_fill_manual(values=c("TRUE"="orange","FALSE"="cornflowerblue"),
                    guide="none") +
  facet_wrap(~ control) +
  theme_bw() + theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=11, color="black"),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid.minor=element_blank(),
    # panel.grid.major.y = element_blank(),
    panel.spacing = unit(0.9, "cm"),
  ) # olink_visit1_significant_yes-vs-no-viralload.pdf, 6.1 x 5.8 in
```

### Plot example cytokines

```{r}
p <- list()
for (i in c("CX3CL1","IFNG")) {
  set.seed(0) # for geom_jitter
  
  temp <- data.frame(conc=olink.visit1[i,], 
                     transplant=clin.sample.event.individ.visit1$transplant_final,
                     steroids=clin.sample.event.individ.visit1$steroids,
                     sc2=clin.sample.event.individ.visit1$sc2_rpm_log10) %>%
    subset(!is.na(conc))
  
  p[[i]] <- ggplot(data=temp, aes(x=transplant, y=conc)) +
    stat_boxplot(aes(color=transplant),
                 geom="errorbar", width=0.3) +
    geom_boxplot(aes(color=transplant), fill="white",
                 outlier.shape=NA, width=0.55) +
    # geom_jitter(width=0.15, size=0.6) +
    ggbeeswarm::geom_quasirandom(
      aes(color=transplant),
      size=0.6, width=0.3) +
    scale_color_manual(values=c("1"="orange","0"="cornflowerblue"),
                       guide="none") +
    scale_x_discrete(labels=c("Control","Transplant")) +
    labs(x="", y="Protein expression", title=i) + 
    my.theme
}

p[[1]]+p[[2]] + plot_layout(nrow=1) # olink_visit1_significant_examples.pdf, 5.6w x 3.2h in
```

## Test for association of cytokines with viral load

```{r}
# Set up table for storing results
res.vs.sc2 <- data.frame(matrix(NA, nrow=nrow(olink.visit1), ncol=7))
colnames(res.vs.sc2) <- c("cytokine",
                       "slope.c","PVal.c","adj.PVal.c",
                       "slope.t","PVal.t","adj.PVal.t")
rownames(res.vs.sc2) <- rownames(olink.visit1)
res.vs.sc2$cytokine <- rownames(res.vs.sc2)

for (i in rownames(res.vs.sc2)) {
  temp <- data.frame(
    conc=olink.visit1[i,], 
    transplant=clin.sample.event.individ.visit1$transplant_final,
    sc2=clin.sample.event.individ.visit1$sc2_rpm_log10) %>%
  subset(!is.na(conc) & !is.na(sc2))

  # Control
  temp.c <- subset(temp, transplant=="0")
  lm.c <- lm(conc~sc2, data=temp.c)
  # Transplant
  temp.t <- subset(temp, transplant=="1")
  lm.t <- lm(conc~sc2, data=temp.t)
  
  # Store results
  res.vs.sc2[i,"slope.c"] <- coef(summary(lm.c))["sc2","Estimate"]
  res.vs.sc2[i,"PVal.c"] <- coef(summary(lm.c))["sc2","Pr(>|t|)"]
  res.vs.sc2[i,"slope.t"] <- coef(summary(lm.t))["sc2","Estimate"]
  res.vs.sc2[i,"PVal.t"] <- coef(summary(lm.t))["sc2","Pr(>|t|)"]
}

# Adjust P-values
res.vs.sc2$adj.PVal.c <- p.adjust(res.vs.sc2$PVal.c, method="BH")
res.vs.sc2$adj.PVal.t <- p.adjust(res.vs.sc2$PVal.t, method="BH")
```

Compare slopes of each group

```{r}
p1 <- ggplot(data=res.vs.sc2,
             aes(x=slope.c, y=slope.t),
             color="black") +
  geom_hline(yintercept=0, linetype=3, color="azure3") +
  geom_vline(xintercept=0, linetype=3, color="azure3") +
  geom_point(shape=21, fill="azure3") +
  geom_point(data=. %>% subset(adj.PVal.t<0.05),
             shape=21, fill="orange") +
  geom_abline(slope=1, intercept=0, linewidth=0.3) +
  scale_x_continuous(limits=c(-0.25,0.25), expand=c(0,0)) +
  scale_y_continuous(limits=c(-0.25,0.25), expand=c(0,0)) +
  labs(x="Control:\nprotein vs viral rpM slope",
       y="Transplant:\nprotein vs viral rpM slope") +
  theme_bw() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    plot.margin = unit(c(0,1,1,0), "cm")
  ) # olink_visit1_all_vs_viralload.pdf, 3.4w x 3.3h in
```

Plot CXCL8 vs viral rpM

```{r}
temp <- data.frame(conc=olink.visit1["CXCL8",], 
                   transplant=clin.sample.event.individ.visit1$transplant_final,
                   sc2=clin.sample.event.individ.visit1$sc2_rpm_log10) %>%
  subset(!is.na(conc) & !is.na(sc2))

# Control
temp.c <- subset(temp, transplant=="0")
lm.c <- lm(conc~sc2, data=temp.c)
pred.c <- ggeffects::ggpredict(lm.c, terms=c("sc2"))
print(summary(lm.c))

# Transplant
temp.t <- subset(temp, transplant=="1")
lm.t <- lm(conc~sc2, data=temp.t)
pred.t <- ggeffects::ggpredict(lm.t, terms=c("sc2"))
print(summary(lm.t))

my.pred <- rbind(
  pred.c %>% mutate(transplant="0"),
  pred.t %>% mutate(transplant="1")
)

# Plot
p2 <- ggplot(data=temp) +
  geom_point(aes(x=sc2, y=conc), size=0.5) +
  geom_line(data=my.pred, 
            aes(x,predicted,color=transplant),
            linewidth=1.5) +
  geom_ribbon(inherit.aes=FALSE, data=my.pred,
              aes(x=x, y=predicted,
                  ymin=conf.low, ymax=conf.high,
                  fill=transplant),
              alpha=0.4) +
  scale_color_manual(values=c("cornflowerblue", "orange"),
                     guide="none") +
  scale_fill_manual(values=c("cornflowerblue", "orange"),
                     guide="none") +
  labs(x="SARS-CoV-2 log10(rpM)", 
       y="Protein expression",
       title="CXCL8") +
  facet_wrap(~transplant, nrow=1) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=12, color="black"),
    axis.text.y = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,0.7,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    panel.grid.minor=element_blank()
  ) # olink_visit1_CXCL8_vs_viralload.pdf, 4.9w x 3.1h in
```

## Compare trajectory groups

```{r}
# Trajectory groups 1-3 (FALSE) vs 4-5 (TRUE)
# clin.sample.event.individ.visit1$tg45 <- as.factor(clin.sample.event.individ.visit1$trajectory_group>=4)
clin.sample.event.individ.visit1$tg45 <- as.factor(clin.sample.event.individ.visit1$newtg>=4) # updated TG
```

### Stratified by control/transplant

```{r}
# Compare TG45 vs TG13 in each of control and transplant groups
res.tg <- list(
  ctrl=data.frame(matrix(NA, nrow=nrow(olink.visit1), ncol=4)),
  tx=data.frame(matrix(NA, nrow=nrow(olink.visit1), ncol=4))
)
for (i in names(res.tg)) {
  colnames(res.tg[[i]]) <- c("cytokine","slope","PVal","adj.PVal")
  rownames(res.tg[[i]]) <- rownames(olink.visit1)
  res.tg[[i]]$cytokine <- rownames(res.tg[[i]])
}

# Linear regression 
for (i in rownames(olink.visit1)) { # rownames(res.tg)
  temp <- data.frame(conc=olink.visit1[i,], 
                     transplant=clin.sample.event.individ.visit1$transplant_final,
                     tg=clin.sample.event.individ.visit1$tg45) %>%
    subset(!is.na(conc))
  
  my.lm <- lm(conc ~ transplant + transplant:tg, data=temp)
  res.tg[["ctrl"]][i,"slope"] <- summary(my.lm)$coefficients["transplant0:tgTRUE",1]
  res.tg[["ctrl"]][i,"PVal"] <- summary(my.lm)$coefficients["transplant0:tgTRUE",4]
  res.tg[["tx"]][i,"slope"] <- summary(my.lm)$coefficients["transplant1:tgTRUE",1]
  res.tg[["tx"]][i,"PVal"] <- summary(my.lm)$coefficients["transplant1:tgTRUE",4]
  # remove(temp, my.lm)
}

# Correct P-values
res.tg[["ctrl"]]$adj.PVal <- p.adjust(res.tg[["ctrl"]]$PVal, method="BH")
res.tg[["tx"]]$adj.PVal <- p.adjust(res.tg[["tx"]]$PVal, method="BH")
```

Plot cytokines significant in at least one group

```{r}
# Prepare data frame to plot
my.cytokines <- union(
  res.tg[["ctrl"]][res.tg[["ctrl"]]$adj.PVal<0.05,"cytokine"],
  res.tg[["tx"]][res.tg[["tx"]]$adj.PVal<0.05,"cytokine"]
)
res.toplot <- rbind(
  res.tg[["ctrl"]][my.cytokines,] %>% mutate(group="Control"),
  res.tg[["tx"]][my.cytokines,] %>% mutate(group="Transplant")
)

# Plot
p1 <- ggplot(data=res.toplot, 
             aes(x=group, y=reorder(cytokine,slope,FUN=function(x) x[1]))) +
  geom_point(aes(color=slope), shape=1, size=2.4, stroke=1.2) +
  geom_point(data=. %>% subset(adj.PVal < 0.05), 
             aes(color=slope),
             shape=19, size=2.4) +
  geom_hline(yintercept=seq.int(2,length(my.cytokines))-0.5,
             linewidth=0.4, linetype="dotted", color="grey") +
  scale_color_gradientn(
    colors=c("limegreen","#f7f7f7","coral2"),
    limits=c(-1.1,1.1),
    breaks=c(-1,0,1), 
    name="") +
  labs(x="", y="", title="Proteins associated with\nseverity at Visit 1") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=12, color="black", angle=35, hjust=1),
    axis.text.y = element_text(size=11, color="black"),
    text = element_text(size=12, family="Arial"),
    axis.ticks.y=element_blank(),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.grid=element_blank()
  ) # olink_visit1_TG.pdf, 3.6w x 7.8h in
```

# Longitudinal analysis 

```{r}
#Sample metadata (659 samples)
clin.sample.event.individ.long <- clin.sample.event.individ %>%
  subset(event_type %in% c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6"))

# Remove sample "0865-002K3G00-002", because the vast majority of its cytokines are 0
clin.sample.event.individ.long <- clin.sample.event.individ.long %>%
  subset(sample_id != "0865-002K3G00-002")

# Try restricting to 30 days post admission
clin.sample.event.individ.long <- clin.sample.event.individ.long %>%
  subset(event_date <= 30)

olink.long <- olink[,clin.sample.event.individ.long$sample_id]

# Scale protein concentration
olink.long <- t(apply(
  olink.long, MARGIN=1, FUN = function(x)
    (x-mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)
))

```

## Linear mixed effects model

```{r}
res.long <- data.frame(matrix(NA, nrow=nrow(olink.long), ncol=4))
colnames(res.long) <- c("cytokine","interaction.slope","PVal","adj.PVal")
res.long$cytokine <- rownames(olink.long)
rownames(res.long) <- rownames(olink.long)

for (i in rownames(res.long)) {
  temp <- data.frame(
    value = olink.long[i,],
    event_date = clin.sample.event.individ.long$event_date,
    transplant = clin.sample.event.individ.long$transplant_final,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value))
  
  temp <- droplevels(temp)

  fit1 <- lmer(
      value ~ event_date + transplant + (1|pid), 
      data = temp, REML = FALSE)
   
  fit2 <- lmer(
    value ~ event_date + transplant + event_date:transplant + (1|pid),
    data = temp, REML = FALSE)

  aa <- anova(fit1, fit2, test = "LRT")
  res.long[i,"PVal"] <- aa[2,"Pr(>Chisq)"]
  res.long[i,"interaction.slope"] <- coef(summary(fit2))["event_date:transplant1","Estimate"]
}

# Correct P-values
res.long$adj.PVal <- p.adjust(res.long$PVal, method = "BH")
```

```{r}
subset(res.long, adj.PVal<0.05)
```

## Plotting longitudinal data

```{r}
i <- "CXCL11"
temp <- data.frame(
    value = olink.long[i,],
    event_date = clin.sample.event.individ.long$event_date,
    transplant = clin.sample.event.individ.long$transplant_final,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value))

# Rename transplant levels
levels(temp$transplant) <- c("Control","Transplant")

# Predict
fit <- lmer(
      value ~ event_date + transplant + event_date:transplant + (1|pid), 
      data = temp, REML = FALSE)
  
my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","transplant"),
                                type="fixed")

colnames(my.pred)[colnames(my.pred)=="group"] <- "transplant"

temp$yhat <- predict(fit)

p1 <- ggplot(data = temp, aes(x=event_date, y=value)) +
  geom_line(aes(group=pid), color="gray", alpha=0.4) +
  geom_point(color="black", shape=19, size=0.4, alpha=0.8) +
  geom_line(inherit.aes=FALSE, data=my.pred, 
            aes(x,predicted, color=transplant),
            linewidth=1.5) +
  geom_ribbon(inherit.aes=FALSE, data=my.pred,
              aes(x=x, y=predicted,
                  ymin=conf.low, ymax=conf.high,
                  fill=transplant),
              alpha=0.4) +
  labs(x="Days from admission", 
       y="Protein expression",
       title=i) +
  scale_color_manual(values=c("cornflowerblue", "orange"),
                     guide="none") +
  scale_fill_manual(values=c("cornflowerblue", "orange"),
                     guide="none") +
  facet_wrap(~ transplant, nrow=1) + theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text.x = element_text(size=12, color="black"),
    axis.text.y = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,0.7,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    panel.grid.minor=element_blank()
  ) # olink_longitudinal_CXCL11.pdf, 5.4w x 3.2h in
```

# Longitudinal analysis: control for viral load

## Linear mixed effects model

```{r}
res.long.2 <- data.frame(matrix(NA, nrow=nrow(olink.long), ncol=4))
colnames(res.long.2) <- c("cytokine","interaction.slope","PVal","adj.PVal")
res.long.2$cytokine <- rownames(olink.long)
rownames(res.long.2) <- rownames(olink.long)

for (i in rownames(res.long.2)) {
  temp <- data.frame(
    value = olink.long[i,],
    event_date = clin.sample.event.individ.long$event_date,
    transplant = clin.sample.event.individ.long$transplant_final,
    sc2 = clin.sample.event.individ.long$sc2_rpm_log10,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value) & !is.na(sc2))
  
  temp <- droplevels(temp)

  fit1 <- lmer(
      value ~ event_date + transplant + sc2 + (1|pid), 
      data = temp, REML = FALSE)
   
  fit2 <- lmer(
    value ~ event_date + transplant + event_date:transplant + sc2 + (1|pid),
    data = temp, REML = FALSE)

  aa <- anova(fit1, fit2, test = "LRT")
  res.long.2[i,"PVal"] <- aa[2,"Pr(>Chisq)"]
  res.long.2[i,"interaction.slope"] <- coef(summary(fit2))["event_date:transplant1","Estimate"]
}

# Correct P-values
res.long.2$adj.PVal <- p.adjust(res.long.2$PVal, method = "BH")
```

Which proteins are significant?

```{r}
subset(res.long.2, adj.PVal<0.05)
```

## Plotting longitudinal data

```{r}
p <- list()
for (i in c("CXCL11","CCL3","CCL4","IL22RA1")) {
  temp <- data.frame(
    value = olink.long[i,],
    event_date = clin.sample.event.individ.long$event_date,
    transplant = clin.sample.event.individ.long$transplant_final,
    sc2 = clin.sample.event.individ.long$sc2_rpm_log10,
    pid = clin.sample.event.individ.long$participant_id
  ) %>%
    subset(!is.na(value) & !is.na(sc2))

  # Rename transplant levels
  levels(temp$transplant) <- c("Control","Transplant")
  
  # Predict
  fit <- lmer(
        value ~ event_date + transplant + event_date:transplant + sc2 + (1|pid), 
        data = temp, REML = FALSE)
    
  my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","transplant"),
                                  type="fixed")
  
  colnames(my.pred)[colnames(my.pred)=="group"] <- "transplant"
  
  temp$yhat <- predict(fit)
  
  p[[i]] <- ggplot(data = temp, aes(x=event_date, y=value)) +
    geom_line(aes(group=pid), color="gray", alpha=0.4) +
    geom_point(color="black", shape=19, size=0.4, alpha=0.8) +
    geom_line(inherit.aes=FALSE, data=my.pred, 
              aes(x,predicted, color=transplant),
              linewidth=1.5) +
    geom_ribbon(inherit.aes=FALSE, data=my.pred,
                aes(x=x, y=predicted,
                    ymin=conf.low, ymax=conf.high,
                    fill=transplant),
                alpha=0.4) +
    labs(x="Days from admission", 
         y="Protein expression",
         title=i) +
    scale_color_manual(values=c("cornflowerblue", "orange"),
                       guide="none") +
    scale_fill_manual(values=c("cornflowerblue", "orange"),
                       guide="none") +
    facet_wrap(~ transplant, nrow=1) + theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5, size=12, face="plain"),
      axis.text.x = element_text(size=12, color="black"),
      axis.text.y = element_text(size=12, color="black"),
      text = element_text(size=12, family="Arial"),
      plot.margin = unit(c(0,1,0.7,0), "cm"),
      panel.spacing = unit(0.5, "cm"),
      panel.grid.minor=element_blank()
    )
}
p[[1]] + p[[2]] + 
  p[[3]] + p[[4]] + plot_layout(ncol=2) # olink_longitudinal_examples_viralload.pdf, 10w x 6.4h in
```
