---
title: "GWAS SNP Array QC"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

### Load libraries
```{r load-packages}

suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(plotrix))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(tinytex))
suppressPackageStartupMessages(library(latexpdf))
suppressPackageStartupMessages(library(plotrix))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(qqman))

```

### Load codebase.R, load data matrices and set path to a local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
config = read.table("../../Codebase/config.txt", header = T)
data_base_dir = config$value[config$key == "data_base_dir"]

# set directories in R
out_dir = "../output"
tmp_dir = "../tmp_files"
if(!dir.exists(tmp_dir)) {
  dir.create(tmp_dir)
}

# set directories in bash
Sys.setenv(inp_dir = file.path(data_base_dir, "bld-gwas/bld-gwas-snparray/"))
Sys.setenv(ped = file.path(data_base_dir, "../raw-data/bld-gwas-snparray/bld-snparray-Counts_G.ped"))
Sys.setenv(map = file.path(data_base_dir, "../raw-data/bld-gwas-snparray/bld-snparray-RowFeature.map"))

Sys.setenv(out_dir = "../output")
Sys.setenv(tmp_dir = "../tmp_files")

```

### Functions: data loading
```{bash, message=FALSE, warning=FALSE, results=FALSE}

source /programs/biogrids.shrc

# Convert .ped/map files to .bed/bim/fam files

plink --ped ${ped} \
      --file ${inp_dir} \
      --map ${map} \
      --make-bed \
      --out ${tmp_dir}/bld-snparray_Counts 

```

### Functions: data analysis

Objective: Heterozygosity and Missingness QC.

1. Check for elevated missing data & heterozygosity rate.

### Supplementary Panel A & B: analysis code
```{bash, message=FALSE, warning=FALSE, results=FALSE}

source /programs/biogrids.shrc

# Check for elevated missing data rates. 
plink    --bfile ${tmp_dir}/bld-snparray_Counts  \
         --missing \
         --out ${tmp_dir}/bld-snparray_Counts.mis
         
# Check for elevated heterozygosity rate.
plink   --bfile ${tmp_dir}/bld-snparray_Counts  \
        --het \
        --out ${tmp_dir}/bld-snparray_Counts

```

Objective: Take Heterozygosity results and visualize them in R.

- Calculate the observed heterozygosity rate per individual using the formula (N(NM) - O(Hom))/N(NM). The third column in the .het.het file gives the observed number of homozygous genotypes [O(Hom)] and the fifth column gives the number of non-missing genotypes [N(NM)], per individual. After doing this calculation, we suggest removing individuals who deviate ±3 SD from the samples' heterozygosity rate mean. For missingness, we suggest removing individuals with greater than 5% missingness (mis.rate = 0.05). Then, in order to visualize the results, we will plot the proportion of missing genotypes and the heterozygosity rate. 

### Supplementary Panel A: output generation code
```{r message=FALSE, warning=FALSE, results=TRUE}
# Generate figure for missingness and heterozygosity.

het=read.table(file.path(tmp_dir, "bld-snparray_Counts.het"), header = TRUE)
mis=read.table(file.path(tmp_dir, "bld-snparray_Counts.mis.imiss"), header = TRUE)

mishet=data.frame(FID=het$FID, IID=het$IID, het.rate=(het$N.NM- het$O.HOM)/het$N.NM, mis.rate=mis$F_MISS)

upper_het_3sd = mean(mishet$het.rate) + 3*sd(mishet$het.rate)
lower_het_3sd = mean(mishet$het.rate) - 3*sd(mishet$het.rate)

# Plot the proportion of missing genotypes and the heterozygosity rate

plot(y=mishet$het.rate, x=mishet$mis.rate, ylab="Heterozygosity rate", xlab="Proportion of missing genotypes", ylim = c(0,0.8)) 
abline(v=0.05, lty=2, col = "red") #vertical dashed line 
abline(h= upper_het_3sd, lty=2, col = "red") #horizontal dashed line
abline(h= lower_het_3sd, lty=2, col = "red") #horizontal dashed line

mishet_plot = recordPlot()       
plot.new()
mishet_plot

pdf(file.path(out_dir, "suppl_A.pdf"), width = 5, height = 5)
print(mishet_plot)
dev.off()

# identify individuals with mis.rate > 0.05 (Greater than 5% Missingness)
fail_mis_qc=mishet[mishet$mis.rate > 0.05,]

# identify individuals with het.rate < 0.103 or with het.rate > upper_het_3sd
fail_het_qc=mishet[mishet$het.rate < lower_het_3sd | mishet$het.rate > upper_het_3sd,]

# Create files fail_mis_qc.txt and fail_het_qc.txt with the individuals that do not pass the thresholds:
write.table(fail_mis_qc, file.path(tmp_dir, "fail_mis_qc.txt"), row.names=F, col.names=T, quote=F)
write.table(fail_het_qc, file.path(tmp_dir, "fail_het_qc.txt"), row.names=F, col.names=T, quote=F)

```

```{bash, message=FALSE, warning=FALSE, results=FALSE}

source /programs/biogrids.shrc

# Remove samples from "fail.het_qc.txt" (deviated ±3 SD from the samples' heterozygosity rate mean).
# Remove samples from "fail.mis_qc.txt" ( greater than 5% missingness).

plink --bfile ${tmp_dir}/bld-snparray_Counts \
      --remove ${tmp_dir}/fail_mis_qc.txt \
      --make-bed \
      --out ${tmp_dir}/bld-snparray_Counts_rmmishet.ind

```

Results from Heterozygosity and Missingness QC.

3. Result: Total genotyping rate is 0.989271.
4. 8 individuals were removed for having high missingness 

QC Progress: 483 samples remaining.

### Supplementary Panel B: output generation code
```{r message=FALSE, warning=FALSE, results=TRUE}

#Plot a histogram of the fraction of missing genotypes to determine the threshold to remove SNPs with missing data (typically about 5% missing data SNPs are removed).

lmis = read.table(file.path(tmp_dir, "bld-snparray_Counts.mis.lmiss"), header = TRUE)

# plot histogram of the fraction of missing genotypes:
hist(lmis$F_MISS, breaks=35, ylim = c(0,2000000), ylab="Number of SNPs",xlab="Fraction of missing genotypes", main="Missing Data Analysis")
# dashed vertical line corresponding to 5% missing data
abline(v=0.05,lty=2, col = "red")


snp_miss_hist = recordPlot()       
plot.new()
snp_miss_hist

pdf(file.path(out_dir, "suppl_B.pdf"), width = 5, height = 5)
print(snp_miss_hist)
dev.off()

```

```{bash, results=FALSE}
source /programs/biogrids.shrc

# Remove high-missingness SNPs & very rare SNPs with MAF < 0.01.
# SNPs with a low MAF are rare, therefore power is lacking for detecting SNP‐phenotype associations. 

plink --bfile ${tmp_dir}/bld-snparray_Counts_rmmishet.ind \
      --maf 0.01 \
      --geno 0.05 \
      --make-bed \
      --out ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno.ind
      
# Result: 

# Total genotyping rate is 0.989271.
# 24581 variants removed due to missing genotype data (--geno).
# 815644 variants removed due to minor allele threshold(s)
# 1074710 variants and 483 people pass filters and QC.

```

Objective: Sex Check QC.

Note: (Sex Check is done after removing samples and SNPs with high missingness and removing rare SNPs with MAF <5%, so as to not bias results).

1. Select individuals with mismatched SNPSEX and PEDSEX in the file bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_sexcheck_sexcheck.sexcheck
2. Remove samples that failed sex check. 

### Supplementary Panel C: analysis code
```{bash, results=FALSE}
source /programs/biogrids.shrc

# Objective: Sex Check QC
plink --bfile ${tmp_dir}/bld-snparray_Counts_rmmishet.ind \
      --maf 0.05 \
      --geno 0.05 \
      --make-bed \
      --out ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_sexcheck.ind
      
# Result (THIS IS AN INTERMEDIATE FILE FOR SEX CHECK ONLY):
plink --bfile ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_sexcheck.ind \
      --check-sex 0.5 0.8 \
      --out ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_sexcheck
        
# Try to identify the problem, and write the IDs of individuals with discordant sex info to a file called "fail_sex_check.txt"

grep "PROBLEM" ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_sexcheck.sexcheck > ${tmp_dir}/fail_sex_check.txt

# 5 samples had problematic sex, as determined by mismatches between SNPSEX and PEDSEX (clinical sex).

```

### Supplementary Panel C: output generation code
```{r message=FALSE, warning=FALSE, results=TRUE}
# make a file for the 5 samples in PLINK format in order to use --remove.

failed_sex_check = read.table(file.path(tmp_dir, "fail_sex_check.txt"), header = F, as.is = T)

failed_sex_check_to_remove = select(failed_sex_check, "V1", "V2") %>%
  rename("FID" = "V1") %>%
  rename("IID" = "V2")

write.table(failed_sex_check_to_remove, file.path(tmp_dir, "failed_sex_check_to_remove.txt"), row.names=F, col.names=T, quote=F)

# Make a plot of the sex_check into a histogram to show the distribution of the x-chromosome heterozygosity.

sex_check = read.table(file.path(tmp_dir, "bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_sexcheck.sexcheck"),  header = T, as.is = T)

hist(sex_check$F, ylab="Frequency", xlab="XHE", main="Phase 1 & 2 Sex Check", breaks = 65, xaxp=c(-0.1,1, 10), yaxp=c(0,500, 25), col= "blue")
abline(v=0.5, lty=2, col = "red") #vertical dashed line 
abline(v=0.8, lty=2, col = "red")
                       
sexcheck_plot = recordPlot()
plot.new()
sexcheck_plot
  
pdf(file.path(out_dir, "suppl_C.pdf"), width = 8, height = 8)
print(sexcheck_plot)
dev.off()
                       
```

Results from Sex Check QC:
Exclude 5 mismatches between PEDSEX and SNPSEX. 

QC Progress: 478 individuals remaining.

```{bash, results=FALSE}
source /programs/biogrids.shrc

plink --bfile ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno.ind \
      --remove ${tmp_dir}/failed_sex_check_to_remove.txt \
      --make-bed \
      --out ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_rmsexcheck.ind
        
```

Objective: Duplicated or related individuals check 

1. Duplicated or related individuals can be identified using the identity-by-descent (IBD). IBD report is constructed from a reduced subset of frequent SNPs. We first have to identify frequent SNPs by pruning to create a subset of frequent snps (MAF > 0.35) while filtering out SNPs that have > 5% missing rate and SNPs that don’t pass the HWE test at p-value = 0.00000001. Start this pruning using the files where we've already removed samples with high missingness and sex mismatches. People who on average share two alleles IBD are either monozygotic twins or duplicates. An IBD > 0.98 identifies duplicates.

### Supplementary Panel D: analysis code
```{bash, results=FALSE}
source /programs/biogrids.shrc

plink --bfile ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_rmsexcheck.ind \
      --maf 0.35 \
      --geno 0.05 \
      --hwe 0.00000001 \
      --make-bed \
      --out ${tmp_dir}/bld-snparray_Counts_frequent

plink --bfile ${tmp_dir}/bld-snparray_Counts_frequent \
      --indep-pairwise 50 5 0.2 \
      --out ${tmp_dir}/bld-snparray_Counts_prunedsnplist
      
# Reduce the subset of frequent SNPs by pruning so that no pair of SNPs (within 50 base-pairs) has an r2 greater that a 0.2.
# Pruning complete. 121247 of 172432 variants removed.

# Generate the identity-by-descent (IBD) report from the reduced subset of frequent SNPs:
plink --bfile ${tmp_dir}/bld-snparray_Counts_frequent \
      --extract ${tmp_dir}/bld-snparray_Counts_prunedsnplist.prune.in \
      --genome \
      --out ${tmp_dir}/bld-snparray_Counts_pruned

```	  

Results from duplicated or related individuals check.

 1. 12 twins / duplicates found
 2. Four pairs of related individuals found, with IBD > 0.1875.

Note: For Phase 1 & 2, we decided to keep all four pairs of related individuals, but to remove 1 of each pair of duplicates.

```{r message=FALSE, warning=FALSE, results=TRUE}
# Identify pairs of individuals with PI_HAT > 0.9, to remove duplicate samples.
# For current manuscript, we decided to keep pairs of related individuals and ONLY to remove duplicates.

genome = read.table(file.path(tmp_dir, "bld-snparray_Counts_pruned.genome"), header=T, as.is=T)

genome_new=genome[genome$PI_HAT > 0.9,] %>%
  select(FID1, IID1, FID2, IID2, PI_HAT) %>%
  arrange(desc(PI_HAT))

# Append the mis.rate information onto each one of the pairs to identify which one had the higher mis.rate & remove the individual with the higher mis.rate.

missingness_genome = left_join(x = genome_new, y = mishet, by = c("IID1" = "IID")) %>%
  select(-"het.rate", - "FID") %>% 
  rename ("mis.rate1" = "mis.rate")

missingness_genome = left_join(x = missingness_genome, y = mishet, by = c("IID2" = "IID")) %>%
  select(-"het.rate", - "FID") %>% 
  rename ("mis.rate2" = "mis.rate")

# Compare the miss.rates for each pair to determine which has a HIGHER miss rate. Print a list of all of the HIGHER missingness rates, i.e. the failures.

failed_ibd_qc = c() 
failed_ibd_qc = ifelse(missingness_genome$mis.rate1 > missingness_genome$mis.rate2, missingness_genome$IID1, missingness_genome$IID2)
failed_ibd_qc = data.frame(failed_ibd_qc) %>% rename("IID" = "failed_ibd_qc")

failed_ibd_qc = left_join(failed_ibd_qc, mishet, by = 'IID') %>% 
  select("FID", "IID")

# write a table of the FID and IID to print out for "failed_ibd_qc.txt"
write.table(failed_ibd_qc , file.path(tmp_dir,"fail_ibd_qc.txt"), row.names=F, col.names=T, quote=F)

```

```{bash, echo=FALSE, results=FALSE}
source /programs/biogrids.shrc

# Remove one of each pair of individuals that failed IBD QC. 12 individuals to remove.

plink --bfile ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_rmsexcheck.ind   \
	    --remove ${tmp_dir}/fail_ibd_qc.txt \
	    --make-bed \
	    --out ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_rmsexcheck_rmIBD.ind  
	    
```

### Supplementary Panel D: output generation code
```{r message=FALSE, warning=FALSE, results=TRUE}

# Plot the IBD by PI_HAT to see the distribution of IBD.
hist(genome$PI_HAT, ylab="Frequency", xlab="Pi-Hat",main="Sample Relatedness Check", breaks = 100, ylim = c (0,50)) # plot histogram of frequencies
abline(v=0.1875,lty=2, col = "red") 

ibd_plot = recordPlot()
plot.new()
ibd_plot
  
pdf(file.path(out_dir, "suppl_D.pdf"), width = 8, height = 8)
print(ibd_plot)
dev.off()

```

Results from duplicated or related individuals check.

1. 12 twins / duplicates found.

Objective: Identify SNPs that fail QC 

```{bash, results=FALSE}

source /programs/biogrids.shrc

# Removes SNPs with MAF<0.01 and SNPs with > 5% missing genotypes and SNPs deviating from HWE (exclude 1e-10) 

plink --bfile ${tmp_dir}/bld-snparray_Counts_rmmishet_rmSNPmis_rmSNPgeno_rmsexcheck_rmIBD.ind \
	    --maf 0.01 \
	    --hwe 1e-10 \
	    --geno 0.05 \
	    --make-bed \
	    --out ${tmp_dir}/bld-snparray_Counts.final.qc
```	  

Results: 

1. Total genotyping rate is 0.998694.
2. 101 variants removed due to missing genotype data (--geno).
3. 3320 variants removed due to Hardy-Weinberg exact test.
4. 8776 variants removed due to minor allele threshold(s)

```{r message=FALSE, warning=FALSE, results=TRUE, eval=FALSE}

# Concatenate all files listing individuals failing any QC steps into a single file "fail_qc.txt"

fail_mis_qc=read.table(file.path(tmp_dir, "fail_mis_qc.txt"),header=T,as.is=T) %>% select("FID", "IID")
fail_het_qc=read.table(file.path(tmp_dir, "fail_het_qc.txt"),header=T,as.is=T) %>% select("FID", "IID")
fail_ibd_qc=read.table(file.path(tmp_dir, "fail_ibd_qc.txt"),header=T,as.is=T)
fail_sex_qc=read.table(file.path(tmp_dir, "failed_sex_check_to_remove.txt"),header=T,as.is=T)


fail_qc = bind_rows(fail_mis_qc, fail_het_qc, fail_ibd_qc, fail_sex_qc)
fail_qc = unique(fail_qc)


write.table(fail_qc, file.path(tmp_dir, "fail_qc.txt"), row.names=F, col.names=T, quote=F)
            
```

### Supplementary Panel E & F: analysis code

```{bash, results=FALSE}

source /programs/biogrids.shrc

# Do the association between group 4+5 vs. 1+2+3.
# In .ped file,each patient is coded as 2 (affected) if in group 4+5 or 1 (unaffected) in group 1+2+3.

plink --bfile ${tmp_dir}/bld-snparray_Counts.final.qc \
      --chr 1-22 \
      --logistic beta \
      --ci 0.95 \
      --out ${tmp_dir}/group45.results
      
# We have genomic inflation. lambda = 2.038317. 

# Steps to doing a population stratification correction to account for genomic inflation. 

# 1. Prune data 
# 2. Select autosomal SNPs only.
# 3. MAF > 5%
# 4. Visually inspect PCA space, and select PCA that are associated to disease. 
# 5. Use those selected PCs as covariates in regression analysis. 

# Set a MAF for variants > 5%, missing genotype rate of 5% and autosomal chromosomes only. 

plink --bfile ${tmp_dir}/bld-snparray_Counts.final.qc \
      --maf 0.05 \
      --geno 0.05 \
      --chr 1-22 \
      --make-bed \
      --out ${tmp_dir}/bld-snparray_Counts.pop.strat
      
# Prune dataset for LD.      

plink --bfile ${tmp_dir}/bld-snparray_Counts.pop.strat \
      --indep-pairwise 50 5 0.2 \
      --out ${tmp_dir}/bld-snparray_Counts.pop.strat

# Calculate pairwise IBS between samples & measure the extent of their genetic similarity across the set of independent SNPs.  

plink --bfile ${tmp_dir}/bld-snparray_Counts.pop.strat \
      --extract ${tmp_dir}/bld-snparray_Counts.pop.strat.prune.in \
      --genome \
      --out ${tmp_dir}/bld-snparray_Counts.pop.strat
      
# We perform MDS of the pairwise IBS.

plink --bfile ${tmp_dir}/bld-snparray_Counts.pop.strat \
      --read-genome ${tmp_dir}/bld-snparray_Counts.pop.strat.genome \
      --cluster \
      --mds-plot 10 \
      --out ${tmp_dir}/bld-snparray_Counts.pop.strat

# Use "significant" PCs as covariates in our trend test of association. 
# Go back to the original Phase1_2.final.qc.updated file where you had the MAF cut-off at 0.01 in order to do the new association.
# Using MAF of 0.01 leads to a QQ plot with deflated p-values. Use the file with 0.05 MAF to see what the QQ Plot looks like. 

plink --bfile ${tmp_dir}/bld-snparray_Counts.pop.strat \
      --logistic beta \
      --ci 0.95 \
      --covar ${tmp_dir}/bld-snparray_Counts.pop.strat.mds \
      --covar-name C1, C3, C9 \
      --hide-covar \
      --out ${tmp_dir}/bld-snparray_Counts.pop.strat.covar
 
```	  

### Supplementary Panel E: output generation code
```{r message=FALSE, warning=FALSE, results=TRUE}

# Calculate the genomic control inflation factor:
group45_assoc = read.table(file.path(tmp_dir, "group45.results.assoc.logistic"), header = T)
  

median(qchisq(group45_assoc$P,df=1,lower.tail=F),na.rm=T)/0.456

# 2.038317 lambda = genomic control factor

# QQ plot of trajectory group 4 & 5 vs. 1,2,3 post-assoc test and corrected for population stratification.
group45_assoc_covar = read.table(file.path(tmp_dir, "bld-snparray_Counts.pop.strat.covar.assoc.logistic"), header=T)

#observed –log10 p-values 
p.obs_covar=-log10(sort(group45_assoc_covar$P,decreasing=F))

#expected –log10 p-values
p.exp_covar=-log10( 1:length(p.obs_covar)/length(p.obs_covar) )

#plot observed vs. expected –log10 p-values.
plot(p.exp_covar, p.obs_covar, pch=19, main= "QQ plot Corrected for Population Stratification", xlab="expected -log10 p-value", ylab="observed -log10 p-value")
abline(0, 1, lty=2, col="red") #equality line


qqplot = recordPlot()
plot.new()
qqplot
  
pdf(file.path(out_dir, "suppl_E.pdf"), width = 8, height = 8)
print(qqplot)
dev.off()
            
# calculate the genomic inflation coefficient
median(qchisq(group45_assoc_covar$P,df=1,lower.tail=F),na.rm=T)/0.456

```

### Supplementary Panel F: output generation code
```{r message=FALSE, warning=FALSE, results=TRUE}

# Manhattan Plot
manhattan(x = group45_assoc_covar, main = "TG 4&5 vs. TG 1-3", chr="CHR", bp="BP", snp="SNP", p="P", logp = TRUE, ylim = c(0, 9), suggestiveline = F)

manhattanplot = recordPlot()
plot.new()
manhattanplot
  
pdf(file.path(out_dir, "suppl_F.pdf"), width = 8, height = 8)
print(manhattanplot)
dev.off()
            

```

### Session info
```{r, print-session-info}
sessionInfo()
```



