---
title: "Nasal transcriptomics data analysis for core-assay manuscript"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ggnetwork"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "GSA"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))
suppressPackageStartupMessages(library(package = "enrichR"))


# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```


### Load codebase.R, load data matrices and set path to your local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# load codebase.R, viral load data and clinical information
# source("~/bitbucket/impacc/Analysis/data_analysis_template_codebase.R")
source("../../Codebase/codebase_v2.R")
data_env <- load_IMPACC_datasets(DATA_VERSION                = "2022-01-01", 
                                 KEEP_COVID19_POS            = TRUE, 
                                 FILTER_BY_CORE_ASSAY_COHORT = TRUE,
                                 ASSAY_NAME = "nasal_transcriptomics")

for (n in grep(pattern = "nasal_transcriptomics_counts|clinical_data|nasal_transcriptomics_rowfeature", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}


# Setup the assay specific output directory
out_dir = "../output"



```


#### Prepare nasal transcriptome QC data

```{r PrepareNasalTranscriptomeDataTemp, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

# This file is needed for median CV values to be used in PVCA analysis and 
# batch correction 
#nasal_transcriptome_metadata_file = file.path("legacy/2021-11-12", "nasal-transcriptomics-Metadata.csv")
nasal_transcriptome_metadata_data <- data_env$nasal_transcriptomics_metadata %>% rownames_to_column("sample_id") %>% data.frame(check.names = T) # read.csv(nasal_transcriptome_metadata_file)

```


&nbsp;

&nbsp;


#### Normalization of counts data
```{r ComputeCPMCountsData, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

# Select only protein-coding genes
genes_pc <- nasal_transcriptomics_rowfeature %>%
  # get rownames as an explicit column
  tibble::rownames_to_column(var = "gene_id") %>%
  # remove row names
  tibble::remove_rownames() %>%
  # keep only gene ids with valid hgnc symbols
  tidyr::drop_na(gene_name) %>%
  # keep only protein coding genes
  dplyr::filter(gene_biotype == "protein_coding") %>%
  # remove redundant genes if any and make unique
  dplyr::distinct(gene_name, .keep_all = TRUE) %>%
  # add gene ids as row names
  column_to_rownames(var = "gene_id")

# get counts data only for pc genes
counts1 <- nasal_transcriptomics_counts %>%
  t()
counts1 <- counts1[rownames(genes_pc), ]  # 19835  1088

nasal_transcriptome_metadata_data_2 <- nasal_transcriptome_metadata_data %>%
  select(sample_id,
  #       core_specific_ID,
         phase, 
         plate = plate_num,
         fastq_total_reads = `fastq_total_reads_.QCMetrics.`,
         percent_aligned = `percent_aligned_.QCMetrics.`,
         QCMetrics_aligned_counts = `aligned_counts_.QCMetrics.`,
         median_cv_coverage = `median_cv_coverage_.QCMetrics.`,
         passQC = `passQC_.QCMetrics.`)

# combine both clinical and QC data
clinical_meta_data <- clinical_data %>%
  select(-phase) %>%
  mutate_at(vars(participant_id, respiratory_status, sex, event_type, respiratory_status_day14, respiratory_status_day28, enrollment_site, race, ethnicity, trajectory_group), list(factor)) %>%
  inner_join(nasal_transcriptome_metadata_data_2, by = "sample_id") %>%
  mutate(phase_plate = paste0(phase, "_", plate)) %>%
  mutate_at(vars(phase_plate, phase, plate), list(factor)) %>%
  # keep only samples which has counts data
  filter(sample_id %in% colnames(counts1)) %>%
  arrange(match(sample_id, colnames(counts1)))
  
# sanity checks to make sure that counts and meta data are ordered consistent
if(!(all(rownames(counts1) == rownames(genes_pc)))){
  stop("make sure genes (features) are ordered consistent between counts data and gene annotation data")
} else {message("genes (features) are ordered consistent between counts data and gene annotation data")}

if(!(all(colnames(counts1) == clinical_meta_data$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}


# compute CPM values
dge_list <- edgeR::DGEList(counts = counts1, genes = genes_pc)
dge_list <- edgeR::calcNormFactors(dge_list)

# select genes which are expressed in at least 5% of samples
cut.filter <- 0.05 
keepRows <- rowSums(round(edgeR::cpm(dge_list$counts)) >= 1) >= cut.filter*ncol(counts1)
curDGE <- dge_list[keepRows,]
curDGE <- edgeR::calcNormFactors(curDGE) # 16363  1088
# dim(curDGE) # 16363  1088

# sanity check
if(!(all(rownames(curDGE$samples) == clinical_meta_data$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}

# compute voom counts
voomCounts_1 <- limma::voom(curDGE, design = NULL, plot = FALSE, save.plot = FALSE)

```

&nbsp;

&nbsp;

## PVCA analysis with technical+biological variables on batch uncorrected data
## Based on normalized limma voom counts

```{r pvcaBatchUncorrected, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

# step 1: load necessary package if not already
# library(package = "pvca") # devtools::install_github("dleelab/pvca")
# VarCorr <- lme4::VarCorr

# First make sure metadata and counts are ordered consistent
if(!(all(rownames(voomCounts_1$targets) == clinical_meta_data$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}

# step 2: specify which assay is going to be used as input
pvca_input_matrix <- as.matrix(voomCounts_1$E)
if (sum(is.na(pvca_input_matrix)) > 0) {
  stop("pvca input matrix should not have missing values")
}

pvca_input_metadata <- clinical_meta_data

# step3: select variables to be included in PVCA
#   event_data_week: day from admission to to the hospital coded as week (categorical variable)
#   sympt_date_week: day from onset of symptoms coded as week (categorical variable)
#   discretized_admit_age_quantile: age at admission coded as quintiles

pvca_input_phenodata <- clinical_data %>%
  mutate(event_date_week = findInterval(event_date, 
                                        vec        = c(0, 7, 14, 21, 28), 
                                        all.inside = TRUE),
         day_from_sympt  = event_date - symptom_date,
         sympt_date_week = findInterval(day_from_sympt, 
                                        vec        = c(0, 7, 14, 21, 28, Inf),
                                        all.inside = TRUE),
         age_decate      =  findInterval(admit_age, 
                                         vec = seq(from = 10, to = 100, by = 10), 
                                         all.inside = TRUE)) %>%
    select(sample_id, event_date_week, event_type, respiratory_status,
           sex, sympt_date_week, event_location,
           discretized_admit_age_quantile, ethnicity, participant_id, race, respiratory_status_day14,
           respiratory_status_day28, trajectory_group, enrollment_site)


# step 4 append meta data to pvca phenodata
#   core site: where are the sample processed (this chunk of code need to be modified for each core assay)
#   plate = interaction term of phase and plate

pvca_input_phenodata <- pvca_input_metadata %>%
  # rownames_to_column(var = "sample_id") %>%
  mutate(plate = interaction(phase, plate, sep = "_", drop = TRUE)) %>%
  mutate(medianCV_decate = findInterval(median_cv_coverage, 
                                        vec        = seq(from = 0, to = 1.5, by = 0.1),
                                        all.inside = TRUE)) %>%
  select(sample_id, plate, phase, medianCV_decate) %>%
  inner_join(pvca_input_phenodata,
        by    = "sample_id")

# check counts and meta data are ordered consistent
if(!(all(colnames(pvca_input_matrix) == pvca_input_phenodata$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}

# convert variables into factor
pvca_input_phenodata2 <- pvca_input_phenodata %>% 
  mutate_at(vars(participant_id, respiratory_status, sex, event_type, respiratory_status, respiratory_status_day14, respiratory_status_day28, enrollment_site, race, ethnicity, event_location, trajectory_group, medianCV_decate), list(factor)) %>%
  tibble::column_to_rownames(var = "sample_id") 

# step 5: run PVCA
fit <- pvca::PVCA(counts    = as.matrix(pvca_input_matrix),
            meta      = pvca_input_phenodata2,
            inter     = FALSE,
            threshold = 0.6)

# step 6: plot barplot with PVCA results
pvca_barplot_data <- data.frame(explained = as.vector(fit),
                                effect    = names(fit)) %>%
  arrange(explained) %>%
  mutate(effect = factor(effect, levels = effect))


pvca_barplot <- ggplot(data    = pvca_barplot_data,
       mapping = aes(x = effect, y = explained)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = signif(explained, digits = 3)),
            nudge_y   = 0.01,
            size      = 3) +
  labs(x = NULL, y = "Proportion of the variance explained") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1))

# pvca_barplot


```

#### Batch correction of expression data for phase_plate and median CV
```{r BatchCorrectionData, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

clinical_meta_data_batch_corr <- clinical_meta_data %>%
  tidyr::drop_na(trajectory_group, event_type)

curDGE_noNA <- curDGE[, clinical_meta_data_batch_corr$sample_id] # 16363  1088

if(!(all(rownames(curDGE_noNA$samples) == clinical_meta_data_batch_corr$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}


phase_plate <- clinical_meta_data_batch_corr$phase_plate
medCV <- clinical_meta_data_batch_corr$median_cv_coverage

designMat <- model.matrix(~trajectory_group*event_type, data = clinical_meta_data_batch_corr)

# compute voom counts
voomCounts_batch_corr_input <- limma::voom(curDGE_noNA, design = NULL, plot = FALSE, save.plot = FALSE) # 16363  1088

# batch correcting the data
exp_batch_corr_output_with_design <- limma::removeBatchEffect(voomCounts_batch_corr_input, batch = phase_plate, covariates = medCV, design = designMat)



```

#### Select highly variable genes based on median absolute deviation (MAD)

```{r Select highly variable genes MAD, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

mad_val = apply(exp_batch_corr_output_with_design,1,mad)
df_mad <- data.frame(gene_id = rownames(exp_batch_corr_output_with_design), mad_val = mad_val)

# You will see a bi-modal distribution and select cut off value at the center 
# ggplot(df_mad, aes(x=mad_val)) +
#   geom_histogram(bins = 1000) 

mad_genes <- df_mad %>%
  filter(mad_val > 1) %>%
  pull(gene_id)

exp_batch_corr_output_with_design_mad <- exp_batch_corr_output_with_design[mad_genes, ]
# dim(exp_batch_corr_output_with_design_mad) # 5569 1088
```


&nbsp;

&nbsp;

#### WGCNA module building on mad genes

#### Compute soft threshold power

```{r ComputeSoftThresholdPower, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

# compute soft-threshold
curPower <- c(c(1:10), seq(from = 11, to=40, by=1))
sft_tuned_mad = tune_soft_threshold_WGCNA(data_df = t(exp_batch_corr_output_with_design_mad),
           networkType = "signed", corFnc = "cor", powers = curPower)


cor <- WGCNA::cor

# module creation
  ret = generate_WGCNA_modules(
                   data_df = t(exp_batch_corr_output_with_design_mad), 
                   # Preprocessed data in data.frame class
                   networkType = "signed",     
                   # Indicate the type of network to construct
                   power = 6,              
                   # Provide a power value chosen from the first step, this parameter is mandatory.
                   minModuleSize = 20,      
                   # Provide a value for the minimum number of features a module can have; use a larger value (> 30) for datasets with thousands of features, and a smaller value otherwise.
                   corType = "bicor",         
                   # Correlation function
                   maxPOutliers = 0.1, 
                   # It is safer to use a smaller value for this parameter when corType = bicor is used.
                   assay_alias = "NasalRNAseq",         
                   # Append this assay_alias to the module names as prefix.
                   reassignThreshold = 1e-6,   
                   # This and rest of the parameters use default values, but users may want to tweak them based on the assay type and data.
                   mergeCutHeight = 0,
                   minKMEtoStay = 0.3,
                   minCoreKME = 0.5,
                   deepSplit = 4,
                   plot_heatmap = FALSE) 


```



&nbsp;

&nbsp;

### Supp panel A analysis code 
### PCA plot colored by enrollment site
```{r, Supp panel A analysis, fig.width=6, fig.height=4.5, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

#get principal components
pc <- prcomp(t(exp_batch_corr_output_with_design))

plotDF <- pc$x[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  # merge(y = clinical_data, by = "sample_id") %>%
  merge(y  = clinical_meta_data_batch_corr, by = "sample_id") 

enrollmentSite2color <- rainbow(n = length(unique(plotDF$enrollment_site))) %>%
  setNames(nm = unique(plotDF$enrollment_site))

```


### Supp panel A: output generation code 
### PCA plot colored by enrollment site
```{r, Supp panel A output, fig.width=6, fig.height=4.5, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

plotPCA <- ggplot(data = plotDF,
       mapping = aes(x = PC1, y = PC2, color = enrollment_site))+
  geom_point(size = 3, alpha = 0.7) +
  scale_color_manual(values = enrollmentSite2color) +
  scale_shape_manual(values = c(21, 23))+
  stat_ellipse()+
  labs(x     = paste0("1st dimension (",
                      round((pc$sdev^2/sum(pc$sdev^2))[1] * 100,1),
                      "%)"),
       y     = paste0("2nd dimension (",
                      round((pc$sdev^2/sum(pc$sdev^2))[2] * 100,1),
                      "%)")) +
  theme(legend.text     = element_text(size = 9),
        legend.key.size = unit(0.01, units = "npc")) +
  theme_bw() +
  theme(legend.position = "none")
  # labs(color="")

print(plotPCA)

ggsave(
 filename = file.path(out_dir, "suppl_A.pdf"),
 width = 6,
 height = 5
)

print(plotPCA)


```


&nbsp;

&nbsp;


### Supp panel B analysis code 
## PVCA analysis with technical+biological variables


```{r Supp panel B analysis code, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

# step 1: load required packages
# library(package = "pvca") # devtools::install_github("dleelab/pvca")
# VarCorr <- lme4::VarCorr

# First make sure metadata and counts are ordered consistent
if(!(all(colnames(as.matrix(exp_batch_corr_output_with_design)) == clinical_meta_data_batch_corr$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}


# step 2: specify which assay is going to be used as input
pvca_input_matrix <- as.matrix(exp_batch_corr_output_with_design)
if (sum(is.na(pvca_input_matrix)) > 0) {
  stop("pvca input matrix should not have missing values")
}
# pvca_input_raw <- as.matrix(normCounts2)
pvca_input_metadata <- clinical_meta_data_batch_corr

# step3: select variable to be included in PVCA
#   event_data_week: day from admission to to the hospital coded as week (categorical variable)
#   sympt_date_week: day from onset of symptoms coded as week (categorical variable)
#   discretized_admit_age_quantile: age at admission coded as quintiles

pvca_input_phenodata <- clinical_data %>%
  mutate(event_date_week = findInterval(event_date, 
                                        vec        = c(0, 7, 14, 21, 28), 
                                        all.inside = TRUE),
         day_from_sympt  = event_date - symptom_date,
         sympt_date_week = findInterval(day_from_sympt, 
                                        vec        = c(0, 7, 14, 21, 28, Inf),
                                        all.inside = TRUE),
         age_decate      =  findInterval(admit_age, 
                                         vec = seq(from = 10, to = 100, by = 10), 
                                         all.inside = TRUE)) %>%
    select(sample_id, event_date_week, event_type, respiratory_status,
           sex, sympt_date_week, event_location,
           discretized_admit_age_quantile, ethnicity, participant_id, race, trajectory_group, enrollment_site)


# step 4 append meta data to pvca phenodata
#   core site: where are the sample processed (this chunk of code need to be modified for each core assay)
#   plate = interaction term of phase and plate

pvca_input_phenodata <- pvca_input_metadata %>%
  # rownames_to_column(var = "sample_id") %>%
  mutate(plate = interaction(phase, plate, sep = "_", drop = TRUE)) %>%
  mutate(medianCV_decate = findInterval(median_cv_coverage, 
                                        vec        = seq(from = 0, to = 1.5, by = 0.1),
                                        all.inside = TRUE)) %>%
  select(sample_id, plate, phase, medianCV_decate) %>%
  inner_join(pvca_input_phenodata,
        by    = "sample_id")

# check counts and meta data are ordered consistent
if(!(all(colnames(pvca_input_matrix) == pvca_input_phenodata$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
}else{message("samples are ordered consistent between counts data and metadata")}

# convert variables into factor
pvca_input_phenodata2 <- pvca_input_phenodata %>% 
  mutate_at(vars(participant_id, respiratory_status, sex, event_type, respiratory_status, enrollment_site, race, ethnicity, event_location, trajectory_group, medianCV_decate), list(factor)) %>%
  tibble::column_to_rownames(var = "sample_id") 

# step 5: run PVCA
fit <- pvca::PVCA(counts    = as.matrix(pvca_input_matrix),
            meta      = pvca_input_phenodata2,
            inter     = FALSE,
            threshold = 0.6)

# step 6: plot barplot with PVCA results
pvca_barplot_data <- data.frame(explained = as.vector(fit),
                                effect    = names(fit)) %>%
  arrange(explained) %>%
  mutate(effect = factor(effect, levels = effect))


```


### Supp panel B output generation code 
## PVCA analysis with technical+biological variables


```{r Supp panel B output code, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE, fig.width=6, fig.height=4}

pvca_barplot <- ggplot(data    = pvca_barplot_data,
       mapping = aes(x = effect, y = explained)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = signif(explained, digits = 3)),
            nudge_y   = 0.01,
            size      = 3,
            hjust = 0.1) +
  labs(x = NULL, y = "Proportion of the variance explained") + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 0,
                                   vjust = 1,
                                   hjust = 1),
        text = element_text(size=12)) +
  coord_flip(ylim = c(0, 0.7)) 
  # coord_cartesian(ylim = c(0, 0.8))

# pvca_barplot

print(pvca_barplot)
ggsave(
 filename = file.path(out_dir, "suppl_B.pdf"),
 width = 6,
 height = 3
)




```

### Supp panel C output generation code 
### heatmap


```{r Supp panel C output code, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE, }

mat <- t(exp_batch_corr_output_with_design_mad) %>%
  scale()

columnAnnotDF <- 
  ret$module_membership[match(colnames(mat), 
                                   table = 
                                     ret$module_membership$feature), ] %>%
  remove_rownames() %>%
  column_to_rownames(var = "feature") 
columnAnnot <- 
  HeatmapAnnotation(df  = columnAnnotDF,
                    col = list(module = setNames(c("grey", 
                                                   standardColors()[1:7]),
                               nm     = 
                                 sort(unique(ret$module_membership$module)))))
rowAnnotDF <- clinical_data %>%
  select(sample_id, trajectory_group) %>%
  mutate(trajectory_group = paste0("TG", trajectory_group)) %>%
  .[match(rownames(mat), table = .$sample_id), ] %>%
  `rownames<-`(NULL) %>%
  column_to_rownames(var = "sample_id") 
rowAnnot <- rowAnnotation(df  = rowAnnotDF,
                          col = list(trajectory_group = c("TG1" = "#639A21",
                                                          "TG2" = "#39828C",
                                                          "TG3" = "#6371AD", 
                                                          "TG4" = "#BD7D31",
                                                          "TG5" = "#9C3418")),
                          show_annotation_name = FALSE)
# set.seed(seed = 1)
trajectory_group <- rowAnnotDF$trajectory_group
TrajGrpCol <- c("TG1" = "#639A21", "TG2" = "#39828C", "TG3" = "#6371AD", "TG4" = "#BD7D31", "TG5" = "#9C3418")
p = ComplexHeatmap::Heatmap(matrix            = mat,
            column_title_rot = 90, 
            column_title_gp = gpar(fontsize=10), 
            column_split = columnAnnotDF$module,
            row_split = rowAnnotDF$trajectory_group,
            heatmap_legend_param = list(title = "z-score"),
            show_column_names = FALSE, 
            show_row_names = FALSE, 
            left_annotation = rowAnnotation(traj_grp = rowAnnotDF$trajectory_group, col=list(traj_grp = TrajGrpCol)),
            top_annotation = columnAnnot)

# draw(p, padding = unit(c(2, 2, 15, 5), "mm"))


print(p)
pdf(file = file.path(out_dir, "suppl_C.pdf"), width = 8, height = 5)
print(p)
dev.off()

# pdf(file = file.path("/users/ndjayavelu/njayavelu/Scripts/impacc-public/nasal-transcriptomics/output", "suppl_C.pdf"), width = 8, height = 5)
# print(p)
# dev.off()


```


### Main Panel B: analysis code
```{r Main panel-B-analysis, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

clinical_subsets = clinical_data
rownames(clinical_subsets) = clinical_subsets$sample_id
row_ids = rownames(ret$MEs)
clinical_subsets = clinical_subsets[row_ids,]
visit1_row_ids = rownames(clinical_subsets[clinical_subsets$event_type == 'Visit 1',])

# Use the visit1_row_ids to only get samples from 'Visit 1':
modules_filtered = ret$MEs[visit1_row_ids ,]
clinical_subsets = clinical_subsets[visit1_row_ids ,]
data_use = modules_filtered 


if(!(all(rownames(data_use) == clinical_subsets$sample_id))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}


# Add the respiratory_status_day14 as $endpoints
data_use$trajectory_group = as.factor(clinical_subsets$trajectory_group)
data_use$enrollment_site = clinical_subsets$enrollment_site
data_use$discretized_admit_age_quantile = clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex

# Set module names:
module_names = colnames(modules_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$trajectory_group),]

# a model with only intercept and random effect across sites
res_table_ordinal = data.frame(matrix(0, ncol = 4, nrow = length(module_names)))
colnames(res_table_ordinal) = c("AIC", "pval", "qval", "coef")

rownames(res_table_ordinal) = module_names
for(j in 1:length(module_names)){
  print(j)
  my.formula0 <- paste0('trajectory_group~ 1+(1|enrollment_site)+discretized_admit_age_quantile+sex')
  my.formula1 <- paste0('trajectory_group~',
                        module_names[j],'+(1|enrollment_site)+','discretized_admit_age_quantile+sex')
  res_table_ordinal[j,1:2]=mixed_ordinal(my.formula0, my.formula1,data_use)
  res_table_ordinal[j,4] <- coef(ordinal::clmm(formula(my.formula1), data = data_use))[module_names[j]]
}

res_table_ordinal$qval =  qvalue::qvalue(res_table_ordinal$pval, fdr.level = 0.05, pi0 = 1)$qvalues


```


### Main Panel B: output generation code
```{r Main panel-B-output-gen, fig.height=5, fig.width=4, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

visit1_mod3 <- data_use %>%
  rownames_to_column(var = "sample_id") %>%
  select(sample_id, NasalRNAseq_mod3, trajectory_group) %>%
  pivot_longer(cols = -c(sample_id, trajectory_group)) %>%
  mutate(name             = "NasalRNAseq.mod3:\nECM formation",
         trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = trajectory_group, y = value)) +
  geom_quasirandom(mapping = aes(color = trajectory_group), cex = 1.5) +
  geom_boxplot(fill = "transparent", outlier.color= "transparent") +
  facet_wrap(facets= ~name, ncol = 1) + 
  labs(y = "Eigenvalue", x = "Clinical trajectory group") +
  scale_color_manual(values = c("TG1" = "#639A21",
                                "TG2" = "#39828C",
                                "TG3" = "#6371AD",
                                "TG4" = "#BD7D31",
                                "TG5" = "#9C3418")) +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=14)) 

print(visit1_mod3)
ggsave(
 filename = file.path(out_dir, "main_B.pdf"),
 width = 3,
 height = 3
)


```



### Main Panel D: output generation code
```{r Main panel-D-output-gen, fig.height=5, fig.width=4, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

visit1_mod4 <- data_use %>%
  rownames_to_column(var = "sample_id") %>%
  select(sample_id, NasalRNAseq_mod4, trajectory_group) %>%
  pivot_longer(cols = -c(sample_id, trajectory_group)) %>%
  mutate(name             = "NasalRNAseq.mod4:\nEpithelium Cornification\nPeptidases",
         trajectory_group = paste0("TG", trajectory_group)) %>%
  ggplot(mapping = aes(x = trajectory_group, y = value)) +
  geom_quasirandom(mapping = aes(color = trajectory_group), cex = 1.5) +
  geom_boxplot(fill = "transparent", outlier.color= "transparent") +
  facet_wrap(facets= ~name, ncol = 1) + 
  labs(y = "Eigenvalue", x = "Clinical trajectory group") +
  scale_color_manual(values = c("TG1" = "#639A21",
                                "TG2" = "#39828C",
                                "TG3" = "#6371AD",
                                "TG4" = "#BD7D31",
                                "TG5" = "#9C3418")) +
  theme_bw() +
  theme(legend.position = "none", 
        axis.text.x = element_text(angle = 0, hjust = 0.5, size = 12),
        axis.text.y = element_text(size = 14),
        text = element_text(size=14)) 

print(visit1_mod4)
ggsave(
 filename = file.path(out_dir, "main_E.pdf"),
 width = 3,
 height = 3
)


```



### Main Panel C: analysis code
```{r Main panel-C-analysis, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

clinical_subsets = clinical_data
rownames(clinical_subsets) = clinical_subsets$sample_id
row_ids = rownames(ret$MEs)
clinical_subsets = clinical_subsets[row_ids, ]
modules_scores = ret$MEs
modules_filtered = modules_scores

if(!(all(rownames(modules_filtered) == rownames(clinical_subsets)))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}


data_use_logitudinal = cbind(modules_filtered, clinical_subsets[c("trajectory_group", "event_date", "enrollment_site", "participant_id", "sex", "discretized_admit_age_quantile")])
data_use_logitudinal$trajectory_group = as.factor(data_use_logitudinal$trajectory_group)

inputDF <- data_use_logitudinal %>%
  rownames_to_column(var = "sample_id") %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "sex", "discretized_admit_age_quantile",
                         "enrollment_site", "trajectory_group"
                         )) %>%
  filter(!is.na(value) & !is.na(trajectory_group))

inputDF[["trajectory_group"]] = ordered(as.factor(as.character(inputDF[["trajectory_group"]])), levels =
                                       c("1", "2", "3", "4", "5"))
inputDF$name <- as.factor(inputDF$name)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)


smooth_spline_model_loop <- model_loop(inputDF, modelType = "smoothSpline", endpoint = "trajectory_group")

smooth_spline_model_loop_df <- smooth_spline_model_loop %>%
  rownames_to_column(var = "module_name") # %>%
  # left_join(df_module_size) %>%
  # relocate(module_size, .after = module_name)


```

### Main Panel C: output generation code
```{r Main panel-C-output-gen, fig.height=4, fig.width=5, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

plotExample <- inputDF %>%
  filter(name %in% "NasalRNAseq_mod3") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5")))

temp_plot <- plot_model(plotExample, model_loop = smooth_spline_model_loop, modelType = "smoothSpline",
           p_adjust = F, endpoint = "trajectory_group", signif_markers = F, remove_NS = T,
           title = "NasalRNAseq.mod3: ECM formation",
           individual_points_size = 0.5,
           individual_paths_size = 0.5,
           group_trendline_line_width = 1,
           title_size = 12,
           bar_height = 5,
           custom_theme_graph = 'text = element_text(size=10)',
           # custom_theme_bars = 'text = element_text(size=6)',
           knot_lines = F,
           ylabel = "Eigenvalue",
           CI = FALSE,
           individual_trendlines = T, individual_points = T, individual_paths = T,
           p_value_text_size = 1.5,
           return_multi_obj = T)
 
print(temp_plot[[1]]) 
ggsave(filename = file.path(out_dir, "main_C.pdf"),
        plot = temp_plot[[1]], 
        width = 5, 
        height = 2)

```


### Main Panel F: output generation code
```{r Main panel-F-output-gen, fig.height=4, fig.width=5, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

plotExample <- inputDF %>%
  filter(name %in% "NasalRNAseq_mod4") %>%
  mutate(trajectory_group = ordered(as.factor(as.character(trajectory_group)), 
                                    levels = c("1", "2", "3", "4", "5")))

temp_plot <- plot_model(plotExample, model_loop = smooth_spline_model_loop, modelType = "smoothSpline",
           p_adjust = F, endpoint = "trajectory_group", signif_markers = F, remove_NS = T,
           title = "NasalRNAseq.mod4: Epithelium Cornification Peptidases",
           individual_points_size = 0.5,
           individual_paths_size = 0.5,
           group_trendline_line_width = 1,
           title_size = 12,
           bar_height = 5,
           custom_theme_graph = 'text = element_text(size=10)',
           # custom_theme_bars = 'text = element_text(size=6)',
           knot_lines = F,
           ylabel = "Eigenvalue",
           CI = FALSE,
           individual_trendlines = T, individual_points = T, individual_paths = T,
           p_value_text_size = 1.5,
           return_multi_obj = T)
 
print(temp_plot[[1]])  
ggsave(filename = file.path(out_dir, "main_F.pdf"),
        plot = temp_plot[[1]], 
        width = 5, 
        height = 2)

```


#### Supp panel D analysis code  
#### Enrichment of modules for GO, pathways and MSIGDB hallmark genesets


```{r Supp panel D analysis, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}
# using enrichR package
# library(enrichR)
listEnrichrSites()
setEnrichrSite("Enrichr") # Human genes
websiteLive <- TRUE

dbs <- listEnrichrDbs()

# select databases of interest
dbs_select <- c("GO_Biological_Process_2021", "KEGG_2021_Human", "COVID-19_Related_Gene_Sets_2021", "MSigDB_Hallmark_2020", "Reactome_2016")


# define an empty list object to collect enriched terms for each module
# modules_enriched_terms_list <- vector("list", length(module_names))
modules_enriched_terms_list <- vector("list", length(module_names))
names(modules_enriched_terms_list) <- module_names

nasal_transcriptomics_rowfeature_v2 <- nasal_transcriptomics_rowfeature %>%
  tibble::rownames_to_column(var = "gene_id")

module_membership_df <- ret$module_membership %>%
  inner_join(nasal_transcriptomics_rowfeature_v2, by = c("feature" = "gene_id")) %>%
  select(gene_id = feature, module, gene_name)

for(k in 1:length(module_names)){
  # select a module
  module_select <- module_names[k]
  # get the gene lists for selected module
  genes_input <- module_membership_df %>% 
    filter(module == module_select) %>% 
    pull(gene_name)
  enriched_out <- try(enrichr(genes_input, dbs_select))
   
  # bind_rows function causes errors when there are no enriched terms (empty data frame)
  # so remove them before hand before merging
  enriched_out_temp <- enriched_out
  for(m in 1:length(enriched_out_temp)){
    if(nrow(enriched_out_temp[[m]]) == 0){
      enriched_out[[m]] = NULL}
  }
  
  module_enriched <- bind_rows(enriched_out, .id = "enrichment_category")
  modules_enriched_terms_list[[k]] <- module_enriched
  cat(sprintf("########################################################\n"))
  cat(sprintf("Enrichment analysis completed for module : %s, k = %d\n", module_select, k))
  cat(sprintf("########################################################\n"))
}

# bind_rows function causes errors when there are no enriched terms (empty data frame)
# so remove them before hand before merging
modules_enriched_terms_list_temp <- modules_enriched_terms_list
  for(m in 1:length(modules_enriched_terms_list_temp)){
    if(nrow(modules_enriched_terms_list_temp[[m]]) == 0){
      modules_enriched_terms_list[[m]] = NULL}
  }
modules_enriched_terms_bind_df <- bind_rows(modules_enriched_terms_list, .id = "module_name")



GO_mod1 <- modules_enriched_terms_bind_df %>%
  filter(module_name == "NasalRNAseq_mod1") %>%
  filter(Adjusted.P.value < 0.2) %>%
  filter(enrichment_category == "MSigDB_Hallmark_2020" | enrichment_category == "KEGG_2021_Human" | enrichment_category == "REACTOME_2016_Human" | enrichment_category == "GO_Biological_Process_2021")

GO_mod2 <- modules_enriched_terms_bind_df %>%
  filter(module_name == "NasalRNAseq_mod2") %>%
  filter(Adjusted.P.value < 0.2) %>%
  filter(enrichment_category == "MSigDB_Hallmark_2020" | enrichment_category == "KEGG_2021_Human" | enrichment_category == "REACTOME_2016_Human" | enrichment_category == "GO_Biological_Process_2021")

GO_mod3 <- modules_enriched_terms_bind_df %>%
  filter(module_name == "NasalRNAseq_mod3") %>%
  filter(Adjusted.P.value < 0.2) %>%
  filter(enrichment_category == "MSigDB_Hallmark_2020" | enrichment_category == "KEGG_2021_Human" | enrichment_category == "REACTOME_2016_Human" | enrichment_category == "GO_Biological_Process_2021")

GO_mod4 <- modules_enriched_terms_bind_df %>%
  filter(module_name == "NasalRNAseq_mod4") %>%
  filter(Adjusted.P.value < 0.2) %>%
  # filter(enrichment_category == "MSigDB_Hallmark_2020")
  filter(enrichment_category == "MSigDB_Hallmark_2020" | enrichment_category == "KEGG_2021_Human" | enrichment_category == "REACTOME_2016_Human" | enrichment_category == "GO_Biological_Process_2021")

GO_mod5 <- modules_enriched_terms_bind_df %>%
  filter(module_name == "NasalRNAseq_mod5") %>%
  filter(Adjusted.P.value < 0.2) %>%
  filter(enrichment_category == "MSigDB_Hallmark_2020" | enrichment_category == "KEGG_2021_Human" | enrichment_category == "REACTOME_2016_Human" | enrichment_category == "GO_Biological_Process_2021")

GO_mod6 <- modules_enriched_terms_bind_df %>%
  filter(module_name == "NasalRNAseq_mod6") %>%
  filter(Adjusted.P.value < 0.2) %>%
  filter(enrichment_category == "MSigDB_Hallmark_2020" | enrichment_category == "KEGG_2021_Human" | enrichment_category == "REACTOME_2016_Human" | enrichment_category == "GO_Biological_Process_2021")

GO_mod7 <- modules_enriched_terms_bind_df %>%
  filter(module_name == "NasalRNAseq_mod7") %>%
  filter(Adjusted.P.value < 0.2) %>%
  filter(enrichment_category == "MSigDB_Hallmark_2020" | enrichment_category == "KEGG_2021_Human" | enrichment_category == "REACTOME_2016_Human" | enrichment_category == "GO_Biological_Process_2021")


```

#### Supp panel D output generation code  
#### Heatmap of enriched terms


```{r Supp panel D output gen, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE, fig.width=14, fig.height=11}
# using enrichR package

go_terms_mod1 <- c("TNF-alpha Signaling via NF-kB",
              "Inflammatory Response",
              "Interferon Gamma Response",
              "Interferon Alpha Response",
              "IL-6/JAK/STAT3 Signaling",
              "IL-2/STAT5 Signaling",
              "JAK-STAT signaling pathway",
              "cytokine-mediated signaling pathway",
              "defense response to virus",
              "neutrophil activation involved in immune response",
              "neutrophil mediated immunity",
              "Neutrophil extracellular trap formation",
              # "neutrophil degranulation",
              # "neutrophil migration",
              # "neutrophil chemotaxis",
              "Epithelial Mesenchymal Transition",
              "Apical Junction",
              "Allograft Rejection",
              "Complement and coagulation cascades")

go_terms_mod2 <- c("cilium movement",
              "cilium assembly",
              "cilium-dependent cell motility",
              "Metabolism of xenobiotics by cytochrome P450",
              "axoneme assembly")

go_terms_mod3 <- c("Epithelial Mesenchymal Transition",
              "Apical Junction",
              "extracellular structure organization",
              "collagen fibril organization",
              "cell junction assembly")

go_terms_mod4 <- c("epidermis development",
              "skin development",
              "keratinocyte differentiation",
              "establishment of skin barrier")

go_terms_mod5 <- c("T cell receptor signaling pathway",
              "Th1 and Th2 cell differentiation",
              "Th17 cell differentiation",
              "Natural killer cell mediated cytotoxicity")

go_terms_mod6 <- c("Tyrosine metabolism",
              "Xenobiotic Metabolism")

go_terms_mod7 <- c("G2-M Checkpoint",
              "E2F Targets",
              "mitotic spindle organization",
              "Cell cycle")


go_terms_mod <- unique(c(go_terms_mod1, go_terms_mod2, go_terms_mod3, go_terms_mod4, go_terms_mod5, go_terms_mod6, go_terms_mod7)) 

data_plot_GO <- modules_enriched_terms_bind_df %>%
  tidyr::separate(col = Term, into = c("Term1", "temp"), sep = " \\(") %>%
  # filter(Term %in% go_terms_mod) # %>%
  # filter(grepl(go_terms_mod, Term))
  select(module_name, Term1, Adjusted.P.value, Overlap) %>%
  filter(Term1 %in% go_terms_mod) %>%
  mutate(Term = factor(Term1, levels = rev(go_terms_mod))) %>%
  tidyr::separate(col = Overlap, into = c("gene_count", "total_GOgenes"), sep = "/") %>%
  mutate(gene_count = as.numeric(gene_count),
         total_GOgenes = as.numeric(total_GOgenes)) %>%
  mutate(percent_genes = gene_count/total_GOgenes*100) 

data_plot_GO$`Adjusted.P.value`[data_plot_GO$`Adjusted.P.value` < 1e-10] <- 1e-10 

heatmap_go <- data_plot_GO %>%
  filter(module_name != "NasalRNAseq_mod0") %>%
  ggplot(aes(x = module_name, y = Term, fill = -log10(Adjusted.P.value), size = percent_genes)) + 
  geom_point(shape = 21) +
  theme_bw() + 
  ylab("") + xlab("") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        text = element_text(size=18)) + 
  labs(fill = "-log10(P)",
       size = "% Genes")

print(heatmap_go)
ggsave(
 filename = file.path(out_dir, "suppl_D.pdf"),
 width = 14,
 height = 11
)


```


# SuppTable 1
# Generate Modules membership file (ASSAY_Modules.csv)

```{r Supplementary tables 1, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

ASSAY <- "NasalRNAseq"

nasal_transcriptomics_rowfeature_v2 <- nasal_transcriptomics_rowfeature %>%
  tibble::rownames_to_column(var = "gene_id")
  
module_membership_df <- ret$module_membership %>%
  inner_join(nasal_transcriptomics_rowfeature_v2, by = c("feature" = "gene_id")) %>%
  select(gene_id = feature, module, gene_name)


module_membership_df_out <- module_membership_df %>%
  select(Module = module, gene_id, gene_name) %>%
  arrange(Module)
  
module_membership_df_out %>%
  head() %>%
  kable()

write.csv(module_membership_df_out, 
          file = file.path(out_dir, "NasalRNAseq_Modules.csv"), row.names = FALSE)



```


# SuppTable 2
# Generate Modules annotation file (ASSAY_Annotation.csv)

```{r Supplementary tables 2, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

supTab_annot <- modules_enriched_terms_bind_df %>%
  filter(Adjusted.P.value <= 0.05) %>%
  filter(enrichment_category == "MSigDB_Hallmark_2020" | enrichment_category == "KEGG_2021_Human" | enrichment_category == "REACTOME_2016_Human" | enrichment_category == "GO_Biological_Process_2021") %>%
  select(Module = module_name,
         enrichment_category,
         Term,
         Overlap,
         Adjusted.P.value,
         Genes
         )
 
supTab_annot %>%
  head() %>%
  kable()

write.csv(supTab_annot, 
          file = file.path(out_dir, "NasalRNAseq_Modules_Annotation.csv"), row.names = FALSE)


```

# SuppTable 3
# Generate modules asssociation with trajectory groups (ASSAY_Results.csv)



```{r Supplementary tables 3 Visit 1, warning=FALSE, echo=TRUE, eval=TRUE, message=FALSE}

### generate results visit 1 overall test (extracting coefficient for directionality)

# a model with only intercept and random effect across sites
res_table_ordinal <- data.frame(matrix(NA, ncol = 4, nrow = length(module_names)))
colnames(res_table_ordinal) <- c("AIC", "pval", "qval", "coef")
rownames(res_table_ordinal) <- module_names
# endpoints = "trajectory_group"
data_use$endpoints <- data_use$trajectory_group

for(j in 1:length(module_names)) {
  my.formula0 <- formula('endpoints~ 1+(1|enrollment_site)+discretized_admit_age_quantile+sex')
  my.formula1 <- paste0('endpoints~',module_names[j],'+(1|enrollment_site)+discretized_admit_age_quantile+sex')
  res_table_ordinal[j,1:2] <- mixed_ordinal(my.formula0, my.formula1,data_use)
  res_table_ordinal[j,4] <- ordinal::clmm(formula(my.formula1), 
                                          data = data_use) %>%
    coef() %>%
    .[module_names[j]]
}
res_table_ordinal$qval <- qvalue::qvalue(res_table_ordinal$pval, 
                                         fdr.level = 0.05, 
                                         pi0       = 1)$qvalues

# generate supplementary table for visit 1 overall
supTab_resultsVisit1 <- res_table_ordinal %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  filter(pval <= 0.05) %>%
  mutate(`Module (or Feature)`,
         Analysis              = "Visit 1 analysis - overall",
         Direction             = ifelse(test = sign(coef) %in% 1,
                                        yes  = "Severe",
                                        no   = "Mild")) %>%
  rename(`P value` = pval) %>%
  select(`Module (or Feature)`,
          Analysis,
         `P value`,
         Direction) %>%
  `rownames<-`(NULL)

supTab_resultsVisit1 %>%
  head() %>%
  kable()

# Step 4: do pairwise comparison at visit 1 and extract coefficient of regression

# subfunction use to fit ordinal regression and extract reg. coef
mixed_pairwise_coef <- function(my.formula0, my.formula1, data_use) {
  endpoints0 <- sort(unique(data_use$endpoints))
  pair_names <- c()
  res_table <- c()
  for(i in 1:(length(endpoints0)-1)){
    for(j in (i+1):(length(endpoints0))){
      pair_names <- c(pair_names, paste0(endpoints0[i],"|", endpoints0[j]))
      data_tmp <- data_use[data_use$endpoints %in% endpoints0[c(i, j)], ]
      fit <- lme4::lmer(my.formula1, data = data_tmp)
      res_table <- c(res_table, 
                     unique(coef(fit)$enrollment_site[,grep(pattern = "endpoint", 
                                               names(coef(fit)$enrollment_site))]))
    }
  }
  names(res_table) <- pair_names
  return(value = res_table)
}

res_table_pairwise <- NULL
res_table_pairwise_coef <- NULL
for(j in 1:length(module_names)){
  my.formula0 = paste0(module_names[j],"~ (1|enrollment_site)+discretized_admit_age_quantile+sex")
  my.formula1 = paste0(module_names[j],"~ endpoints + (1|enrollment_site)+discretized_admit_age_quantile+sex")
  res_table_pairwise <- rbind(res_table_pairwise,
                              mixed_pairwise(my.formula0, my.formula1, data_use))
  res_table_pairwise_coef <- rbind(res_table_pairwise_coef,
                                   mixed_pairwise_coef(my.formula0, 
                                                       my.formula1, 
                                                       data_use))
}
rownames(res_table_pairwise) <- module_names
rownames(res_table_pairwise_coef) <- module_names

# Step 5: append pairwise comparison to visit1 results
supTab_resultsVisit1tmp <- res_table_pairwise %>%
  as.data.frame() %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  pivot_longer(cols      = -`Module (or Feature)`, 
               names_to  = "Analysis", 
               values_to = "P value")
res_table_pairwise_coef %>%
  as.data.frame() %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  pivot_longer(cols      = -`Module (or Feature)`,
               names_to  = "Analysis",
               values_to = "Direction") %>%
  merge(x  = supTab_resultsVisit1tmp, 
        by = c("Module (or Feature)", "Analysis")) -> supTab_resultsVisit1tmp

supTab_resultsVisit1tmp <- supTab_resultsVisit1tmp %>%
  mutate(`Module (or Feature)` = `Module (or Feature)`,
         Analysis              = paste0("Visit 1 analysis - ", Analysis),
         Direction             = ifelse(test = sign(Direction) %in% 1,
                                        yes  = "Severe",
                                        no   = "Mild")) %>%
  filter(`P value` <= 0.05)

supTab_resultsVisit1 <- rbind(supTab_resultsVisit1,
                              supTab_resultsVisit1tmp)

supTab_resultsVisit1 %>%
  tail() %>%
  kable()

### longitudinal analysis

inputDF <- inputDF %>%
  select(sample_id, trajectory_group, event_date, 
         enrollment_site, participant_id, discretized_admit_age_quantile, 
         sex, name,	value)

# Step 7: regenerate smooth spline regression and impute directionality from linear regression
## Smooth spline takes a long time, demonstrating by cutting down to just ten factors
smooth_spline_model_loop <- model_loop(inputDF, 
                                       modelType = "smoothSpline", 
                                       endpoint  = "trajectory_group")
# convert ordered factor to factor for trajectory group
inputDF <- inputDF %>%
  mutate(trajectory_group = factor(as.character(trajectory_group), levels = c("1", "2", "3", "4", "5")))
# fetch directionality from lme (TG1 vs TG5)
smooth_spline_model_loop$trajectory_group5 <- NA
smooth_spline_model_loop$"event_date:trajectory_group5" <- NA
for (modName in rownames(smooth_spline_model_loop)) {
  fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
             random = ~1|enrollment_site/participant_id,
             data   = filter(inputDF, name %in% modName))
  smooth_spline_model_loop[modName, "trajectory_group5"] <-   unique(coef(fit)["trajectory_group5"])
  smooth_spline_model_loop[modName, "event_date:trajectory_group5"] <- unique(coef(fit)["event_date:trajectory_group5"])
}


# Step 8: generate supplementary table with longitudinal analysis overall
supTab_resultsLongitudinal <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  select(`Module (or Feature)`, p.slope, p.intercept, trajectory_group5, `event_date:trajectory_group5`) %>%
  rename(coef.intercept = trajectory_group5, 
         coef.slope = `event_date:trajectory_group5`) %>%
  pivot_longer(cols = -`Module (or Feature)`, names_to = c("cname", "Analysis"), names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = cname, values_from = value) %>%
  filter(p <= 0.05) %>%
  mutate(Analysis = recode(Analysis,
                           `slope` = "Longitudinal analysis - overall shape",
                           `intercept` = "Longitudinal analysis - overall average"),
         Direction = ifelse(test = sign(coef) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
    rename(`P value` = p) %>%
  select(`Module (or Feature)`,
         Analysis,
         `P value`,
         Direction) %>%
  `rownames<-`(NULL)

supTab_resultsLongitudinal  %>% 
  head() %>%
  kable()


# Step 9: do pairwise regression and extract coefficient from linear model
# use linear regression to infer direction
combMat <- combn(unique(inputDF$trajectory_group), m = 2)
coefCombMat <- apply(combMat, MARGIN = 2, FUN = function(comparison) {
  inputTemp <- filter(inputDF, trajectory_group %in% comparison)

  coefDF <- lapply(rownames(smooth_spline_model_loop), FUN = function(modName) {
    fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
               random = ~1|enrollment_site/participant_id,
               data   = filter(inputTemp, name %in% modName))
    average <- unique(coef(fit)[grep(pattern = "^trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    shape <- unique(coef(fit)[grep(pattern = "^event_date:trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    return(value = data.frame(`Module (or Feature)` = modName,
                              average, 
                              shape,
                              comparison = paste(sort(comparison), collapse = "|"),
                              check.names = FALSE))
  }) %>%
    do.call(what = rbind)
  return(value = coefDF)
}) %>%
  do.call(what = rbind)

# Step 10: append pairwise comparison to longitudinal results
supTab_resultsLongitudinalPairwise <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  select(`Module (or Feature)`, 
         matches(match = "^p.slope_.v.$"),
         matches(match = "^p.intercept_.v.$")) %>%
  pivot_longer(cols      = -`Module (or Feature)`,
               names_to  = "Analysis",
               values_to = "P value") %>%
  mutate(Analysis = gsub(pattern     = "slope",
                         replacement = "shape",
                         Analysis),
         Analysis = gsub(pattern     = "intercept",
                         replacement = "average",
                         Analysis),
         Analysis = gsub(pattern = "^p\\.(.+)_(.)v(.)$",
                         replacement = "Longitudinal analysis - \\2|\\3 \\1",
                         Analysis))
coefCombMat %>%
  pivot_longer(cols      = -c(`Module (or Feature)`, comparison),
               names_to  = "Analysis",
               values_to = "Direction") %>%
  mutate(Analysis = paste("Longitudinal analysis -",
                          comparison,
                          Analysis),
         Direction = ifelse(test = sign(Direction) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
  select(-comparison) %>%
  merge(x  = supTab_resultsLongitudinalPairwise,
        by = c("Module (or Feature)", "Analysis")) -> supTab_resultsLongitudinalPairwise

# append to overall results
supTab_resultsLongitudinalPairwise %>%
  filter(`P value` <= 0.05) %>%
  rbind(supTab_resultsLongitudinal, .) -> supTab_resultsLongitudinal

supTab_resultsLongitudinal %>%
  tail() %>%
  kable()



### Longitudinal data DFSO
### create inputDF for longitudinal DFSO

clinical_subsets = clinical_data
rownames(clinical_subsets) = clinical_subsets$sample_id

row_ids = rownames(ret$MEs)
clinical_subsets = clinical_subsets[row_ids, ]
modules_scores = ret$MEs

modules_filtered = modules_scores
# clinical_subsets = clinical_subsets

if(!(all(rownames(modules_filtered) == rownames(clinical_subsets)))){
  stop("make sure samples are ordered consistent between counts data and metadata")
} else {message("samples are ordered consistent between counts data and metadata")}


data_use_dfso = cbind(modules_filtered, clinical_subsets[c("trajectory_group",  "event_date", "symptom_date", "enrollment_site", "participant_id", "sex", "discretized_admit_age_quantile")])
data_use_dfso$trajectory_group = as.factor(data_use_dfso$trajectory_group)

# replace event_date = event_date-symptom_date
data_use_dfso <- data_use_dfso %>%
  mutate(event_date = event_date - symptom_date) %>%
  select(-symptom_date)

inputDF_dfso <- data_use_dfso %>%
  rownames_to_column(var = "sample_id") %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "sex", "discretized_admit_age_quantile",
                         "enrollment_site", "trajectory_group"
                         )) %>%
  filter(!is.na(value) & !is.na(trajectory_group))

# inputDF$trajectory_group <- as.factor(inputDF$trajectory_group)
inputDF_dfso[["trajectory_group"]] = ordered(as.factor(as.character(inputDF_dfso[["trajectory_group"]])), levels =
                                       c("1", "2", "3", "4", "5"))
inputDF_dfso$name <- as.factor(inputDF_dfso$name)
inputDF_dfso$sex <- factor(inputDF_dfso$sex, levels = c("Female", "Male"))
inputDF_dfso$discretized_admit_age_quantile <- as.factor(inputDF_dfso$discretized_admit_age_quantile)
inputDF_dfso$participant_id <- as.factor(inputDF_dfso$participant_id)
inputDF_dfso$enrollment_site <- as.factor(inputDF_dfso$enrollment_site)

inputDF_dfso <- inputDF_dfso %>%
  select(sample_id, trajectory_group, event_date, 
         enrollment_site, participant_id, discretized_admit_age_quantile, 
         sex, name,	value) %>%
  tidyr::drop_na(event_date)

# Step 12: generate smooth spline regression with symptom onset as reference and impute directionality from linear regression
smooth_spline_model_loop_dfso <- model_loop(inputDF_dfso, modelType = "smoothSpline", endpoint = "trajectory_group")

# convert ordered factor to factor for trajectory group
inputDF_dfso <- inputDF_dfso %>%
  mutate(trajectory_group = factor(as.character(trajectory_group), levels = c("1", "2", "3", "4", "5")))


# fetch directionality from lme
smooth_spline_model_loop_dfso$trajectory_group5 <- NA
smooth_spline_model_loop_dfso$"event_date:trajectory_group5" <- NA
for (modName in rownames(smooth_spline_model_loop_dfso)) {
  fit <- lme(fixed= value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
      random = ~1|enrollment_site/participant_id,
      data = filter(inputDF_dfso, name %in% modName))
smooth_spline_model_loop_dfso[modName, "trajectory_group5"] <- unique(coef(fit)["trajectory_group5"])
smooth_spline_model_loop_dfso[modName, "event_date:trajectory_group5"] <- unique(coef(fit)["event_date:trajectory_group5"])
}

# Step 13: generate supplementary table with longitudinal analaysis overall
supTab_resultsLongitudinalDFSO <- smooth_spline_model_loop_dfso %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  select(`Module (or Feature)`, p.slope, p.intercept, trajectory_group5, `event_date:trajectory_group5`) %>%
  rename(coef.intercept = trajectory_group5, 
         coef.slope = `event_date:trajectory_group5`) %>%
  pivot_longer(cols = -`Module (or Feature)`, names_to = c("cname", "Analysis"), names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = cname, values_from = value) %>%
  filter(p <= 0.05) %>%
  mutate(Analysis = recode(Analysis,
                           `slope` = "Longitudinal analysis DFSO - overall shape",
                           `intercept` = "Longitudinal analysis DFSO- overall average"),
         Direction = ifelse(test = sign(coef) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
    rename(`P value` = p) %>%
  select(`Module (or Feature)`,
         Analysis,
         `P value`,
         Direction) %>%
  `rownames<-`(NULL)

supTab_resultsLongitudinalDFSO %>%
  head() %>%
  kable()


# Step 14: do pairwise regression with DFSO as time and extract coefficient from linear model
# use linear regression to infer direction
combMat <- combn(unique(inputDF_dfso$trajectory_group), m = 2)
coefCombMat <- apply(combMat, MARGIN = 2, FUN = function(comparison) {
  inputTemp <- filter(inputDF_dfso, trajectory_group %in% comparison)

  coefDF <- lapply(rownames(smooth_spline_model_loop_dfso), FUN = function(modName) {
    fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
               random = ~1|enrollment_site/participant_id,
               data   = filter(inputTemp, name %in% modName))
    average <- unique(coef(fit)[grep(pattern = "^trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    shape <- unique(coef(fit)[grep(pattern = "^event_date:trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    return(value = data.frame(`Module (or Feature)` = modName,
                              average, 
                              shape,
                              comparison = paste(sort(comparison), collapse = "|"),
                              check.names = FALSE))
  }) %>%
    do.call(what = rbind)
  return(value = coefDF)
}) %>%
  do.call(what = rbind)

# Step 15: append pairwise comparison to longitudinal results
supTab_resultsLongitudinalDFSOPairwise <- smooth_spline_model_loop_dfso %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  select(`Module (or Feature)`, 
         matches(match = "^p.slope_.v.$"),
         matches(match = "^p.intercept_.v.$")) %>%
  pivot_longer(cols      = -`Module (or Feature)`,
               names_to  = "Analysis",
               values_to = "P value") %>%
  mutate(Analysis = gsub(pattern     = "slope",
                         replacement = "shape",
                         Analysis),
         Analysis = gsub(pattern     = "intercept",
                         replacement = "average",
                         Analysis),
         Analysis = gsub(pattern = "^p\\.(.+)_(.)v(.)$",
                         replacement = "Longitudinal analysis DFSO - \\2|\\3 \\1",
                         Analysis))
coefCombMat %>%
  pivot_longer(cols      = -c(`Module (or Feature)`, comparison),
               names_to  = "Analysis",
               values_to = "Direction") %>%
  mutate(Analysis = paste("Longitudinal analysis DFSO -",
                          comparison,
                          Analysis),
         Direction = ifelse(test = sign(Direction) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
  select(-comparison) %>%
  merge(x  = supTab_resultsLongitudinalDFSOPairwise,
        by = c("Module (or Feature)", "Analysis")) -> supTab_resultsLongitudinalDFSOPairwise

# append to overall results
supTab_resultsLongitudinalDFSOPairwise %>%
  filter(`P value` <= 0.05) %>%
  rbind(supTab_resultsLongitudinalDFSO, .) -> supTab_resultsLongitudinalDFSO


### merge all results table into one csv file

supTab_results <- rbind(supTab_resultsVisit1,
                        supTab_resultsLongitudinal,
                        supTab_resultsLongitudinalDFSO) %>%
  rename(Module = `Module (or Feature)`)

supTab_resultsLongitudinalDFSO %>%
  tail() %>%
  kable()



write.csv(supTab_results, 
          file =  file.path(out_dir, "NasalRNAseq_Modules_Results.csv"), row.names = FALSE)



```


### Session info
```{r, print-session-info}
sessionInfo()
```

<!--
R version 4.0.2 (2020-06-22)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Ubuntu 18.04.4 LTS

Matrix products: default
BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so

locale:
 [1] LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8       
 [4] LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8    LC_MESSAGES=C.UTF-8   
 [7] LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C          
[10] LC_TELEPHONE=C         LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

attached base packages:
[1] parallel  grid      stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] WGCNA_1.69-81         fastcluster_1.2.3     dynamicTreeCut_1.63-1 doParallel_1.0.17    
 [5] doRNG_1.8.2           rngtools_1.5.2        missForest_1.4        itertools_0.1-3      
 [9] iterators_1.0.14      foreach_1.5.2         randomForest_4.6-10   ComplexHeatmap_2.11.1
[13] corrr_0.4.3           gridExtra_2.3         rlang_1.0.1           RColorBrewer_1.1-2   
[17] cowplot_1.1.1         ggpubr_0.4.0          pals_1.7              enrichR_3.0          
[21] forcats_0.5.1         stringr_1.4.0         dplyr_1.0.8           purrr_0.3.4          
[25] readr_2.1.2           tidyr_1.2.0           tibble_3.1.6          tidyverse_1.3.1      
[29] nlme_3.1-147          lme4_1.1-28           Matrix_1.2-18         GSA_1.03.2           
[33] pvca_0.1.0            ggeffects_1.1.1       ggbeeswarm_0.6.0      qvalue_2.22.0        
[37] ordinal_2019.12-10    ggnetwork_0.5.10      ggplot2_3.3.5         igraph_1.3.0         
[41] impute_1.64.0         knitr_1.37           

loaded via a namespace (and not attached):
  [1] readxl_1.3.1          backports_1.4.1       circlize_0.4.14       Hmisc_4.6-0          
  [5] plyr_1.8.6            splines_4.0.2         digest_0.6.29         htmltools_0.5.2      
  [9] GO.db_3.12.1          fansi_1.0.2           checkmate_2.0.0       memoise_1.1.0        
 [13] magrittr_2.0.2        cluster_2.1.0         tzdb_0.2.0            limma_3.46.0         
 [17] modelr_0.1.8          matrixStats_0.61.0    jpeg_0.1-9            colorspace_2.0-3     
 [21] blob_1.2.2            rvest_1.0.2           haven_2.4.3           xfun_0.29            
 [25] crayon_1.5.0          jsonlite_1.7.3        survival_3.1-12       glue_1.6.1           
 [29] gtable_0.3.0          GetoptLong_1.0.5      car_3.0-12            shape_1.4.6          
 [33] BiocGenerics_0.36.1   maps_3.4.0            abind_1.4-5           scales_1.1.1         
 [37] DBI_1.1.2             edgeR_3.32.1          rstatix_0.7.0         Rcpp_1.0.8           
 [41] htmlTable_2.4.0       clue_0.3-60           foreign_0.8-79        bit_4.0.4            
 [45] mapproj_1.2.8         preprocessCore_1.52.1 Formula_1.2-4         stats4_4.0.2         
 [49] htmlwidgets_1.5.4     httr_1.4.2            ellipsis_0.3.2        pkgconfig_2.0.3      
 [53] farver_2.1.0          nnet_7.3-14           dbplyr_2.1.1          locfit_1.5-9.4       
 [57] utf8_1.2.2            AnnotationDbi_1.52.0  tidyselect_1.1.2      labeling_0.4.2       
 [61] reshape2_1.4.4        munsell_0.5.0         cellranger_1.1.0      tools_4.0.2          
 [65] cli_3.2.0             RSQLite_2.2.10        generics_0.1.2        sjlabelled_1.1.8     
 [69] broom_0.7.12          evaluate_0.14         fastmap_1.1.0         yaml_2.3.5           
 [73] bit64_4.0.5           fs_1.5.2              xml2_1.3.2            compiler_4.0.2       
 [77] rstudioapi_0.13       gamm4_0.2-6           beeswarm_0.4.0        curl_4.3             
 [81] png_0.1-7             ggsignif_0.6.3        reprex_2.0.1          stringi_1.7.6        
 [85] lattice_0.20-41       nloptr_2.0.0          vctrs_0.3.8           pillar_1.7.0         
 [89] lifecycle_1.0.1       GlobalOptions_0.1.2   ucminf_1.1-4          insight_0.16.0       
 [93] data.table_1.14.2     latticeExtra_0.6-29   R6_2.5.1              vipor_0.4.5          
 [97] IRanges_2.24.1        codetools_0.2-16      dichromat_2.0-0       boot_1.3-25          
[101] MASS_7.3-51.6         assertthat_0.2.1      rjson_0.2.21          withr_2.4.3          
[105] S4Vectors_0.28.1      mgcv_1.8-31           hms_1.1.1             rpart_4.1-15         
[109] minqa_1.2.4           snakecase_0.11.0      rmarkdown_2.11        carData_3.0-5        
[113] base64enc_0.1-3       Biobase_2.50.0        numDeriv_2016.8-1.1   lubridate_1.8.0
-->






