---
title: "nasal_viral_load"
output: html_document
---

Source: 20230411_viral_load.Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(dplyr)
library(ggplot2)
library(patchwork)
library(gamm4)

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )
```

# Data import

## Clinical metadata

```{r}
# Patient metadata
clin.individ <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-locked.csv",
  header = TRUE)

# Sample
clin.sample <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-sample-locked.csv",
  header = TRUE)

# Event
clin.event <- read.csv(
  "/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-event-locked.csv",
  header = TRUE)
```

Merge metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

Manually correct some typos.

```{r}
# Correct the symptom_date of patient 008-0103: change from -376 to -13 (confirmed with data management team)
clin.sample.event.individ[clin.sample.event.individ$participant_id=="008-0103","symptom_date"] <- -13

# Correct event_type of 004-0012-8: change from 2 to 32 (it's 32 in 2021-11-11-frozen data)
clin.sample.event.individ[clin.sample.event.individ$event_id=="004-0012-8","event_date"] <- 32
```

## Transplant metadata 

264 patients included in transplant study.

```{r}
transplant <- read.delim(
  "/scratch/immunosuppression/2023.03.09_txcasecontrol.txt",
  header = TRUE)

# Format factor
transplant$transplant_final <- as.factor(transplant$transplant_final)

# How many patients are there?
print(length(unique(transplant$studyid)))
```

Remove BMT patients and their controls.

```{r}
print(unique(transplant$organ_final))

transplant <- transplant %>%
  subset(!(studyid %in% c("002-0022","007-0008","001-0013","010-0009","004-0041","005-0047")))

# How many patients remain?
print(length(unique(transplant$studyid)))

print(unique(transplant$organ_final))
```

Only keep transplant patients in the clinical metadata

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event.individ[,c("participant_id","event_id","sample_id","event_date","event_type","event_location","respiratory_status")], 
  transplant,
  by.x="participant_id", by.y="studyid",
  all=FALSE
)

# How many patients are left?
print(length(unique(clin.sample.event.individ$participant_id)))
```

## Viral load data

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/current/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)

mngs.metadata <- mngs.metadata %>%
  subset(!is.na(sc2_rpm))
```

Only keep samples that pass QC, and merge with metadata 

```{r}
# mNGS QC
mngs.metadata <- mngs.metadata %>%
  filter(sample_status == "IMPACC sample assayed and passed QC")

# Add viral load to sample metadata
clin.sample.event.individ <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by="sample_id", all.x=FALSE, all.y=FALSE, sort=FALSE
)

# How many samples are left
print(nrow(clin.sample.event.individ))
```

Log-transform viral load

```{r}
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ$sc2_rpm_log10 <- log10(clin.sample.event.individ$sc2_rpm + 1)

# Any row with no sc2_rpm?
print(any(is.na(clin.sample.event.individ$sc2_rpm)))

# Any duplicated samples?
print(any(duplicated(clin.sample.event.individ$event_id)))
```

Filter for only samples at Visits 1-6

```{r}
clin.sample.event.individ <- clin.sample.event.individ %>%
        filter(!event_type %in% c("Escalation 1", "Escalation 2", NA))

print(table(clin.sample.event.individ$event_type))
```

## qPCR data

```{r}
# qPCR data
pcr <- read.csv(
  "/data/nasal-viralload/current/nasal-viralload-Counts.csv",
  header=TRUE, row.names=NULL,
  check.names=FALSE
)

pcr.metadata <- read.csv(
 "/data/nasal-viralload/current/nasal-viralload-Metadata.csv",
 header=TRUE, row.names=NULL,
 check.names=FALSE
)
```

```{r}
# qPCR QC
pcr.metadata <- subset(pcr.metadata,
                       sample_status=="IMPACC sample assayed and passed QC")
pcr <- subset(pcr, (sample_id %in% pcr.metadata$sample_id) &
                !is.na(N1_CT) & !is.na(N2_CT))

# Merge with clinical metadata
clin.sample.event.individ.pcr <- merge(
  clin.sample.event.individ, pcr[,c("sample_id","N1_CT","N2_CT")],
  by="sample_id", all=FALSE, sort=FALSE
)
```

# Visit 1 analysis

Filter for Visit 1 samples only

```{r}
# Keep visit 1 samples
clin.sample.event.individ.visit1 <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1")

# Check numbers
sum(clin.sample.event.individ.visit1$transplant_final == 0)
sum(clin.sample.event.individ.visit1$transplant_final == 1)
```

## Plot transplant vs viral load at visit 1 

```{r}
set.seed(0)
p <- ggplot(data=clin.sample.event.individ.visit1,
       aes(x=transplant_final, y=sc2_rpm_log10, color=transplant_final)) + 
  stat_boxplot(geom="errorbar", width=0.3) +
  geom_boxplot(fill="white", outlier.shape=NA, width=0.55) + 
  ggbeeswarm::geom_quasirandom(size=0.6, width=0.3) +
  scale_color_manual(values=c("0"="cornflowerblue", "1"="orange"), guide="none") +
  scale_x_discrete(labels=c("Control","Transplant")) +
  scale_y_continuous(limits=c(-0.2,6), expand=c(0,0)) +
  labs(x="", y="Viral load - log10(rpM)",
       title="SARS-CoV-2 viral load\nat Visit 1") +
  my.theme # viralload_visit1.pdf, 3 x 3.5 in
```

Calculate P-value with linear regression

```{r}
my.lm <- lm(sc2_rpm_log10 ~ transplant_final, data=clin.sample.event.individ.visit1)
summary(my.lm)
```

## Plot viral load vs tranplant types

```{r}
# Extract transplant patients
my.transplant <- clin.sample.event.individ.visit1 %>%
  subset(transplant_final == 1)

# Get transplant types
my.transplant <- my.transplant %>%
  mutate(organ = case_when(
    (organ_final == "kidney/heart") ~ "Heart",
    (organ_final == "liver/kidney") ~ "Liver", 
    (organ_final == "kidney") ~ "Kidney",
    (organ_final == "heart") ~ "Heart",
    (organ_final == "liver") ~ "Liver",
    (organ_final == "lung") ~ "Lung"
  )
  )

```

Plot.

```{r}
p1 <- ggplot(data=my.transplant,
             aes(x=organ, y=sc2_rpm_log10)) +
  stat_boxplot(geom="errorbar", width=0.3, color="orange") +
  geom_boxplot(fill="white", outlier.shape=NA, width=0.55, color="orange") + 
  ggbeeswarm::geom_quasirandom(
    size=0.6, width=0.3,
    color="orange") +
  # scale_color_manual(values=c("0"="cornflowerblue", "1"="orange"), guide="none") +
  # scale_x_discrete(labels=c("Control","Transplant")) +
  scale_y_continuous(limits=c(-0.2,6), expand=c(0,0)) +
  labs(x="Transplant organ", y="Viral load - log10(rpM)",
       title="SARS-CoV-2 viral load\nat Visit 1") +
  my.theme # viralload_organ_visit1.pdf, 3.5 x 3.51 in
```

Calculate P-value

```{r}
my.lm1 <- lm(sc2_rpm_log10 ~ organ, data=my.transplant)
my.lm2 <- lm(sc2_rpm_log10 ~ 1, data=my.transplant)
print(anova(my.lm2, my.lm1, test="LRT"))
```

## Compare rpM vs qPCR Ct

```{r}
# Add qPCR data
clin.sample.event.individ.visit1 <- merge(
  clin.sample.event.individ.visit1, clin.sample.event.individ.pcr[,c("event_id","N1_CT","N2_CT")],
  by="event_id", all=FALSE, sort=FALSE
)
```

```{r}
p1 <- ggplot(data=clin.sample.event.individ.visit1,
             aes(x=N1_CT, y=sc2_rpm_log10,
                 color=transplant_final)) +
  geom_point(size=1) +
  scale_color_manual(values=c("1"="orange","0"="cornflowerblue"),
                    guide="none") +
  labs(x="qPCR (N1 CT)", y="Metatranscriptomics\n(log-transformed rpM)") +
  my.theme

p2 <- ggplot(data=clin.sample.event.individ.visit1,
             aes(x=N2_CT, y=sc2_rpm_log10,
                 color=transplant_final)) +
  geom_point(size=1) +
  scale_color_manual(values=c("1"="orange","0"="cornflowerblue"),
                    guide="none") +
  labs(x="qPCR (N2 CT)", y="Metatranscriptomics\n(log-transformed rpM)") +
  my.theme

p1 + p2 + plot_layout(nrow=1) # viralload_mngs-vs-qpcr.pdf, 6 x 3 in
```

```{r}
print(cor.test(
  clin.sample.event.individ.visit1$N1_CT, clin.sample.event.individ.visit1$sc2_rpm_log10
))

print(cor.test(
  clin.sample.event.individ.visit1$N2_CT, clin.sample.event.individ.visit1$sc2_rpm_log10
))
```

# Longitudinal analysis

```{r}
# Cut off data at 30 days
clin.sample.event.individ.long <- clin.sample.event.individ %>%
  subset(event_date<=30)
```

## GAMM: Using days from admission

```{r}
temp <- clin.sample.event.individ.long[,c("sc2_rpm_log10","event_date","transplant_final","participant_id")]
temp$transplant_final <- ordered(temp$transplant_final) # convert to ordered factor

fit <- gamm4(
  sc2_rpm_log10 ~ s(event_date, bs="cr") + s(event_date, bs="cr", by=transplant_final) + transplant_final, 
  data = temp, random = ~(1|participant_id))

aa <- anova(fit$gam)
  
# Get P-values for the interaction term
aa1 <- sum(aa$s.table[-c(1), "F"]*aa$s.table[-c(1), "Ref.df"])
aa2 <- sum(aa$s.table[-c(1), "Ref.df"])
print(pchisq(aa1, aa2, lower.tail = FALSE))
```

Plot

```{r}
my.pred <- ggeffects::ggpredict(fit, terms=c("event_date","transplant_final"),
                                type="fixed")
colnames(my.pred)[colnames(my.pred)=="group"] <- "transplant_final"

temp$yhat <- predict(fit$mer)

p3 <- ggplot(data=temp, aes(x=event_date, y=sc2_rpm_log10, color=transplant_final)) +
  geom_line(aes(group=participant_id), color="gray", alpha=0.4) +
  geom_point(color="black", shape=16, size=0.7, alpha=0.8) +
  geom_line(inherit.aes=FALSE, data=my.pred, 
            aes(x,predicted, color=transplant_final),
            linewidth=1.5) +
  geom_ribbon(inherit.aes=FALSE, data=my.pred,
              aes(x=x, y=predicted,
                  ymin=conf.low, ymax=conf.high,
                  fill=transplant_final),
              alpha=0.4) +
  labs(x="Days from admission", y="Viral load") +
  scale_color_manual(values=c("0"="cornflowerblue", "1"="orange"),
                     guide="none") +
  scale_fill_manual(values=c("0"="cornflowerblue", "1"="orange"),
                    guide="none") +
  scale_y_continuous(limits=c(-0.5,6), expand=c(0,0)) +
  facet_wrap(~ transplant_final, nrow=1) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,0.7,0), "cm"),
    panel.grid.minor=element_blank(),
    panel.spacing = unit(0.5, "cm"),
  ) # viralload_longitudinal_gamm4.pdf, 5.4 x 3 in
```
