---
title: "Nasal Metagenomic Univariate Analysis"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "gamm4"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggeffects"))
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "tidyverse"))
suppressPackageStartupMessages(library(package = "vegan"))
# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
options(scipen=999)
set.seed(123)
```

### Load codebase.R, load data matrices and set path to a local output directory
```{r load-codebase, results="hide", fig.show="hide", message=FALSE, warning=FALSE}
# load codebase.R, nasal metagenomic data and clinical information
source("../../Codebase/codebase_v2.R")
data_env <- load_IMPACC_datasets(DATA_VERSION                = "2022-01-01", 
                                 KEEP_COVID19_POS            = TRUE, 
                                 FILTER_BY_CORE_ASSAY_COHORT = TRUE,
                                 ASSAY_NAME = "nasal_metagenomics")
#data_env <- readRDS("~/data_env.rds")
for (n in grep(pattern = "nasal_metagenomic|clinical_data", 
               names(data_env), 
               value   = TRUE)) {
  assign(n, value = data_env[[n]])
}

# Setup the assay specific output directory
out_dir <- "../output"


### Misc
run_NMDS_fresh = TRUE   # Set this value to TRUE if you haven't already calculated "Nasal_NMDS_k3_immportIds.RData". In that case, Nasal_NMDS_k3_immportIds.RData will be calculated.
precalculated_NMDS_path <- "Nasal_NMDS_k3_immportIds.RData"
#### Define the outlier samples, if any, that should be removed.
outlier_sample_ids <- c()
```

### Functions: data loading
```{r data-loading}


```

### Functions: data processing
```{r data-processing}

```

### Functions: data analysis
```{r data-analysis}
mixed_pairwise_coef <- function(my.formula0, my.formula1, data_use) {
  endpoints0 <- sort(unique(data_use$endpoints))
  pair_names <- c()
  res_table <- c()
  for(i in 1:(length(endpoints0)-1)){
    for(j in (i+1):(length(endpoints0))){
      pair_names <- c(pair_names, paste0(endpoints0[i],"|", endpoints0[j]))
      data_tmp <- data_use[data_use$endpoints %in% endpoints0[c(i, j)], ]
      fit <- lme4::lmer(my.formula1, data = data_tmp)
      res_table <- c(res_table,
                     unique(coef(fit)$sites[,grep(pattern = "endpoint",
                                               names(coef(fit)$sites))]))
    }
  }
  names(res_table) <- pair_names
  return(value = res_table)
}
```

### Panel A: analysis code
```{r panel-a-analysis, results="hide", fig.show="hide", message=FALSE}

meta_feature_data <- nasal_metagenomics_metadata %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = dplyr::select(clinical_data, sample_id, event_date, participant_id, respiratory_status_day14, respiratory_status_day28,enrollment_site, trajectory_group, sex, admit_age, enrollment_site, discretized_admit_age_quantile, event_type),
        by = c("sample_id", "participant_id")) %>%
  tidyr::drop_na(trajectory_group) %>%
  mutate(total_bacterial_rpm = log2(total_bacterial_rpm + 1)) %>% # Log Normalize only the total bacterial rpm
  pivot_longer(cols = c("total_bacterial_rpm", "alpha_diversity")) %>%
  mutate(
         enrollment_site = factor(enrollment_site),
         trajectory_group = ordered(factor(trajectory_group)),
         discretized_admit_age_quantile = factor(discretized_admit_age_quantile),
         sex = factor(sex),
         name = factor(name),
         participant_id = factor(participant_id))

metafeature_gam_loop <- model_loop(meta_feature_data, modelType = "smoothSpline")
```

### Panel A: output generation code
```{r panel-a-output-gen, fig.width=8, fig.height =2.5}

pa <- plot_model(meta_feature_data[meta_feature_data$name == "total_bacterial_rpm",],
           modelType = "smoothSpline", model_loop = metafeature_gam_loop, CI =F,
           p_adjust = NA, ylabel = "Total Bacterial\nAbundance (Log2 rpm)",
           endpoint = "trajectory_group", signif_markers = F,
           remove_NS = T, bar_height = 1, knot_lines = F) +
    labs(tag = "A")

print(pa)
pdf(file.path(out_dir, "main_A.pdf"), width=8, height = 2.5)
print(pa)
```
### Panel B: output generation code
```{r panel-b-output-gen, fig.width=8, fig.height =2.5}

pb <- plot_model(meta_feature_data[meta_feature_data$name == "alpha_diversity",],
           modelType = "smoothSpline", model_loop = metafeature_gam_loop, CI =F,
           p_adjust = NA, ylabel = "Alpha Diversity",
           endpoint = "trajectory_group", signif_markers = F,
           remove_NS = T, bar_height = 1, knot_lines = F) +
    labs(tag = "B")
print(pb)
pdf(file.path(out_dir, "main_B.pdf"), width=8, height = 2.5)
print(pb)

```
### Panel C: analysis code
```{r panel-c-analysis, results="hide", fig.show="hide", message=FALSE}

#### nasal_metagenomics_counts contains taxa read counts mapped to the genius level
print(paste0("Starting with ", ncol(nasal_metagenomics_counts), " taxa."))
nasal_metagenomics_counts <- nasal_metagenomics_counts[, colSums(nasal_metagenomics_counts) > 0]
nasal_metagenomics_counts <- nasal_metagenomics_counts[rowSums(nasal_metagenomics_counts) > 0, ]
print(paste0("After filtering Taxa with 0 counts, ", ncol(nasal_metagenomics_counts), " taxa remain."))
## Dropping 76 taxa that had no reads mapped

########### For some Analysis, only taxa that had greater than or equal to 30% presence in a trajectory group were used, here that data is prepped and log-transformed
taxa_selection <- nasal_metagenomics_counts %>%
  rownames_to_column("sample_id") %>%
  pivot_longer(-sample_id) %>%
  left_join(clinical_data, by = "sample_id") %>%
  group_by(trajectory_group, name) %>%
  dplyr::summarize(trj_group_freq_detected = sum(value > 0) / n()) %>%
  mutate(trajectory_group = paste0("trj", trajectory_group)) %>%
  pivot_wider(names_from = trajectory_group, values_from = trj_group_freq_detected) %>%
  rowwise() %>%
  mutate(trj_group_avg_freq = mean(c(trj1, trj2, trj3, trj4, trj5)),
         trj_group_max_freq = max(c(trj1, trj2, trj3, trj4, trj5)))

selected_nasal_taxa <- log2(nasal_metagenomics_counts[,colnames(nasal_metagenomics_counts) %in% taxa_selection$name[taxa_selection$trj_group_max_freq >= 0.3]] + 1)
## selected_nasal_taxa is already log transformed where as nasal_metagenomics_counts is not

taxa_inputDF <- selected_nasal_taxa %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = dplyr::select(clinical_data, sample_id, event_date, participant_id, respiratory_status_day14, respiratory_status_day28,enrollment_site, trajectory_group, sex, admit_age, enrollment_site, discretized_admit_age_quantile),
        by = "sample_id") %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "respiratory_status_day14",
                         "respiratory_status_day28",
                         "sex", "admit_age", "discretized_admit_age_quantile")) %>%
  dplyr::filter(!is.na(value) & !is.na(trajectory_group)) %>%
   mutate(
         enrollment_site = factor(enrollment_site),
         trajectory_group = ordered(factor(trajectory_group)),
         discretized_admit_age_quantile = factor(discretized_admit_age_quantile),
         sex = factor(sex),
         name = factor(name),
         participant_id = factor(participant_id))


taxa_gam_loop <- model_loop(taxa_inputDF, modelType = "smoothSpline")
```


### Panel C: output generation code
```{r panel-c-output-gen, fig.width=8, fig.height =2.5}

pc <- plot_model(taxa_inputDF[taxa_inputDF$name == "Bacteroides",],
           modelType = "smoothSpline", model_loop = taxa_gam_loop, CI =F,
           p_adjust = NA, ylabel = "Bacteroides (Log2 rpm)",
           endpoint = "trajectory_group", signif_markers = F,
           remove_NS = T, bar_height = 1, knot_lines = F) +
    labs(tag = "C")
print(pc)
pdf(file.path(out_dir, "main_C.pdf"), width=8, height = 2.5)
print(pc)
```

### Panel D: output generation code
```{r panel-d-output-gen, fig.width=8, fig.height =2.5}

pd <- plot_model(taxa_inputDF[taxa_inputDF$name == "Fusobacterium",],
           modelType = "smoothSpline", model_loop = taxa_gam_loop, CI =F,
           p_adjust = NA, ylabel = "Fusobacterium (Log2 rpm)",
           endpoint = "trajectory_group", signif_markers = F,
           remove_NS = T, bar_height = 1, knot_lines = F) +
    labs(tag = "D")

print(pd)
pdf(file.path(out_dir, "main_D.pdf"), width=8, height = 2.5)
print(pd)

```

### Panel E: output generation code
```{r panel-e-output-gen, fig.width=8, fig.height =2.5}

pe <- plot_model(taxa_inputDF[taxa_inputDF$name == "Prevotella",],
           modelType = "smoothSpline", model_loop = taxa_gam_loop, CI =F,
           p_adjust = NA, ylabel = "Prevotella (Log2 rpm)",
           endpoint = "trajectory_group", signif_markers = F,
           remove_NS = T, bar_height = 1, knot_lines = F) +
    labs(tag = "E")

print(pe)
pdf(file.path(out_dir, "main_E.pdf"), width=8, height = 2.5)
print(pe)
```

### Panel F: output generation code
```{r panel-f-output-gen, fig.width=8, fig.height =2.5}

pf <- plot_model(taxa_inputDF[taxa_inputDF$name == "Cutibacterium",],
           modelType = "smoothSpline", model_loop = taxa_gam_loop, CI =F,
           p_adjust = NA, ylabel = "Cutibacterium (Log2 rpm)",
           endpoint = "trajectory_group", signif_markers = F,
           remove_NS = T, bar_height = 1, knot_lines = F) +
    labs(tag = "F")

print(pf)
pdf(file.path(out_dir, "main_F.pdf"), width=8, height = 2.5)
print(pf)
```

### Supplementary Panel A: analysis code
```{r sup-panel-a-analysis, results="hide", fig.show="hide", message=FALSE}

set.seed(123) ## Seed critical for reproducing NMDS

if(run_NMDS_fresh){
  k_value = 3
  nasal_all_NMDS <- metaMDS(log2(nasal_metagenomics_counts + 1), k=k_value, trymax=5000, maxit = 999, sfgrmin = 1e-9)
  save(nasal_all_NMDS, file = precalculated_NMDS_path) #paste0("Nasal_NMDS_k", k_value, ".RData"))
} else {
  load(precalculated_NMDS_path, verbose = TRUE) ## See starting chunks for path
  }

nasal_prcomp <- prcomp(log2(nasal_metagenomics_counts + 1))
nasal_scores <- merge(scores(nasal_all_NMDS)$sites, nasal_prcomp$x[,1:5], by = 0)
rownames(nasal_scores) <- nasal_scores[,1]
nasal_scores <- nasal_scores[,-1]
plotDF <- nasal_scores %>%
  as.data.frame() %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = clinical_data, by = "sample_id") %>%
  merge(y  = rownames_to_column(nasal_metagenomics_counts, var = "sample_id"),
        by = "sample_id") %>%
  mutate(enrollment_site = ifelse(test = participant_type %in% "Healthy control (Emory)",
                                  yes  = "Emory-Ctrl",
                                  no   = enrollment_site))
enrollmentSite2color <- rainbow(n = length(unique(plotDF$enrollment_site))) %>%
  setNames(nm = unique(plotDF$enrollment_site))
enrollmentSite2color["Emory-Ctrl"] <- "black"
plotDF$trajectory_group <- as.factor(plotDF$trajectory_group)

################# PVCA #################################
pvca_input_matrix <- t(log2(nasal_metagenomics_counts + 1))
if (sum(is.na(pvca_input_matrix)) > 0) {
  stop("pvca input matrix should not have missing values")
}
pvca_input_raw <- t(nasal_metagenomics_counts)
pvca_input_metadata <- nasal_metagenomics_metadata

pvca_input_phenodata <- clinical_data %>%
  mutate(event_date_week = findInterval(event_date,
                                        vec        = c(0, 7, 14, 21, 28),
                                        all.inside = TRUE),
         day_from_sympt  = event_date - symptom_date,
         sympt_date_week = findInterval(day_from_sympt,
                                        vec        = c(0, 7, 14, 21, 28, Inf),
                                        all.inside = TRUE)) %>%
    dplyr::select(sample_id, event_date_week, event_type, respiratory_status,
           sex, sympt_date_week, event_location,
           discretized_admit_age_quantile, ethnicity, participant_id, race,
           trajectory_group, enrollment_site)

pvca_input_phenodata <- pvca_input_metadata %>%
  rownames_to_column(var = "sample_id") %>%
  mutate(
         plate     = interaction(phase, plate, drop = TRUE)) %>%
  dplyr::select(sample_id, plate, phase) %>%
  merge(x     = pvca_input_phenodata,
        by    = "sample_id",
        all.x = TRUE)

pvca_input_phenodata <- pvca_input_phenodata %>%
  column_to_rownames(var = "sample_id")
pvca_input_phenodata <- pvca_input_phenodata[colnames(pvca_input_matrix), , drop = FALSE]

pvca_input_phenodata$percent_missing <- factor(colMeans(is.na(pvca_input_raw[, rownames(pvca_input_phenodata)])))

pvca_fit <- PVCA(counts  = pvca_input_matrix,
            meta = pvca_input_phenodata[,c(1:7, 9:14)],
            inter = FALSE,
            threshold = 0.6)
```

### Supplementary Panel A: output generation code
```{r sup-panel-a-output-gen, fig.width=6, fig.height =4}
plotPCA <- ggplot(data = plotDF,
    mapping = aes(x = PC1, y = PC2, color = enrollment_site))+
  geom_point( size = 3, alpha = 0.7) +
  scale_color_manual(values = enrollmentSite2color) +
  scale_shape_manual(values = c(21, 23))+
  stat_ellipse()+
  labs(x = paste0("1st dimension (",
                  round((nasal_prcomp$sdev^2/sum(nasal_prcomp$sdev^2))[1] * 100),
                  "%)"),
       y = paste0("2nd dimension (",
                  round((nasal_prcomp$sdev^2/sum(nasal_prcomp$sdev^2))[2] * 100),
                  "%)"),
       tag = "A")+
  theme_classic() +
  theme(legend.position = "none")

print(plotPCA)
pdf(file.path(out_dir, "suppl_A.pdf"), width=6, height = 4)
print(plotPCA)
```

### Supplementary Panel B: output generation code
```{r sup-panel-b-output-gen, fig.width=6, fig.height =4}
plotNMDSBray <- ggplot(data = plotDF,
       mapping = aes(x = `NMDS1`, y = `NMDS2`, color = enrollment_site))+
  geom_point( size = 3, alpha = 0.7) +
  scale_color_manual(values = enrollmentSite2color) +
  scale_shape_manual(values = c(21, 23))+
  stat_ellipse()+
  theme_classic() +
  theme(legend.position = "none") +
  labs(tag = "B")
print(plotNMDSBray)
pdf(file.path(out_dir, "suppl_B.pdf"), width=6, height = 4)
print(plotNMDSBray)
```

### Supplementary Panel C: output generation code
```{r sup-panel-c-output-gen, fig.width=7, fig.height =4}
pvca_barplot_data <- data.frame(explained = as.vector(pvca_fit),
                                effect = names(pvca_fit)) %>%
  arrange(explained) %>%
  mutate(effect = factor(effect, levels = effect))

pvca_barplot <- ggplot(data = pvca_barplot_data,
       mapping = aes(x = effect, y = explained)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = signif(explained, digits = 3)),
            nudge_y   = 0.04,
            size      = 3) +
  labs(x = NULL, y = "Proportion of the variance explained",
       tag = "C") +
  theme_bw() + coord_flip()

print(pvca_barplot)
pdf(file.path(out_dir, "suppl_C.pdf"), width=7, height = 4)
print(pvca_barplot)
```

### Supplementary Table A: analysis code
```{r sup-table-a-analysis, message=FALSE, warning=FALSE, results="hide"}
ASSAY <- "Nasal_metagenomics"

################################################################
module_names <- colnames(selected_nasal_taxa)
data_use <- selected_nasal_taxa %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = dplyr::select(clinical_data, sample_id, event_date, participant_id, respiratory_status_day14, respiratory_status_day28,enrollment_site, trajectory_group, sex, admit_age, enrollment_site, discretized_admit_age_quantile, event_type),
        by = "sample_id") %>%
  tidyr::drop_na(trajectory_group) %>%
  dplyr::filter(event_type == "Visit 1") %>%
  mutate(
         sites = as.factor(enrollment_site),
         endpoints = as.factor(trajectory_group),
         age = as.factor(discretized_admit_age_quantile),
         sex = as.factor(sex))

################################################################
res_table_ordinal <- data.frame(matrix(NA, ncol = 4, nrow = length(module_names)))
colnames(res_table_ordinal) <- c("AIC", "pval", "qval", "coef")
rownames(res_table_ordinal) <- module_names
for(j in 1:length(module_names)) {
  my.formula0 <- paste0('endpoints~ 1+(1|sites)+age+sex')
  my.formula1 <- paste0('endpoints~',module_names[j],'+(1|sites)+age+sex')
  res_table_ordinal[j,1:2] <- mixed_ordinal(my.formula0, my.formula1,data_use)
  res_table_ordinal[j,4] <- ordinal::clmm(formula(my.formula1),
                                          data = data_use) %>%
    coef() %>%
    .[module_names[j]]
}
res_table_ordinal$qval <- qvalue::qvalue(res_table_ordinal$pval,
                                         fdr.level = 0.05,
                                         pi0       = 1)$qvalues


################################################################
supTab_resultsVisit1 <- res_table_ordinal %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  dplyr::filter(pval <= 0.05) %>%
  mutate(`Module (or Feature)` = paste0(ASSAY, ".", `Module (or Feature)`),
         Analysis              = "Visit 1 analysis - overall",
         Direction             = ifelse(test = sign(coef) %in% 1,
                                        yes  = "Severe",
                                        no   = "Mild")) %>%
  rename(`P value`      = pval,
         `Q value`      = qval) %>%
  dplyr::select(`Module (or Feature)`,
          Analysis,
         `P value`,
         `Q value`,
         Direction) %>%
  `rownames<-`(NULL)

################################################################

res_table_pairwise <- NULL
res_table_pairwise_coef <- NULL
for(j in 1:length(module_names)){
  my.formula0 = paste0(module_names[j],"~ (1|sites)+age+sex")
  my.formula1 = paste0(module_names[j],"~ endpoints + (1|sites)+age+sex")
  res_table_pairwise <- rbind(res_table_pairwise,
                              mixed_pairwise(my.formula0, my.formula1, data_use))
  res_table_pairwise_coef <- rbind(res_table_pairwise_coef,
                                   mixed_pairwise_coef(my.formula0,
                                                       my.formula1,
                                                       data_use))
}
rownames(res_table_pairwise) <- module_names
rownames(res_table_pairwise_coef) <- module_names

################################################################
supTab_resultsVisit1tmp <- res_table_pairwise %>%
  as.data.frame() %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  pivot_longer(cols      = -`Module (or Feature)`, 
               names_to  = "Analysis", 
               values_to = "P value") %>%
  mutate(`Q value` = qvalue::qvalue(`P value`, 
                                    fdr.level = 0.05, 
                                    pi0       = 1)$qvalues)
res_table_pairwise_coef %>%
  as.data.frame() %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  pivot_longer(cols      = -`Module (or Feature)`,
               names_to  = "Analysis",
               values_to = "Direction") %>%
  merge(x  = supTab_resultsVisit1tmp, 
        by = c("Module (or Feature)", "Analysis")) -> supTab_resultsVisit1tmp

supTab_resultsVisit1tmp <- supTab_resultsVisit1tmp %>%
  mutate(`Module (or Feature)` = paste0(ASSAY, ".", `Module (or Feature)`),
         Analysis              = paste0("Visit 1 analysis - ", Analysis),
         Direction             = ifelse(test = sign(Direction) %in% 1,
                                        yes  = "Severe",
                                        no   = "Mild")) %>%
  dplyr::filter(`P value` <= 0.05 & 
           `Module (or Feature)` %in% supTab_resultsVisit1$"Module (or Feature)")

supTab_resultsVisit1 <- rbind(supTab_resultsVisit1,
                              supTab_resultsVisit1tmp)


################################################################

inputDF <- selected_nasal_taxa %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y = dplyr::select(clinical_data, sample_id, event_date, participant_id, trajectory_group, sex, admit_age, enrollment_site, discretized_admit_age_quantile, symptom_date),
        by = "sample_id")  %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "symptom_date",
                         "sex", "admit_age", "discretized_admit_age_quantile")) %>%
  dplyr::filter(!is.na(value) & !is.na(trajectory_group))%>%
   mutate(
         enrollment_site = as.factor(enrollment_site),
         trajectory_group = as.factor(trajectory_group),
         discretized_admit_age_quantile = as.factor(discretized_admit_age_quantile),
         sex = as.factor(sex),
         name = as.factor(name),
         participant_id = as.factor(participant_id))

################################################################
smooth_spline_model_loop <- model_loop(inputDF,
                                       modelType = "smoothSpline",
                                       endpoint  = "trajectory_group")

# fetch directionality from lme (TG1 vs TG5)
smooth_spline_model_loop$trajectory_group5 <- NA
smooth_spline_model_loop$"event_date:trajectory_group5" <- NA
for (modName in rownames(smooth_spline_model_loop)) {
  fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
             random = ~1|enrollment_site/participant_id,
             data   = dplyr::filter(inputDF, name %in% modName))
  smooth_spline_model_loop[modName, "trajectory_group5"] <- unique(coef(fit)["trajectory_group5"])
  smooth_spline_model_loop[modName, "event_date:trajectory_group5"] <- unique(coef(fit)["event_date:trajectory_group5"])
}


################################################################
supTab_resultsLongitudinal <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  dplyr::select(`Module (or Feature)`, 
         p.slope, 
         p.intercept, 
         adjp.slope, 
         adjp.intercept, 
         trajectory_group5, 
         `event_date:trajectory_group5`) %>%
  rename(coef.intercept = trajectory_group5, 
         coef.slope = `event_date:trajectory_group5`) %>%
  pivot_longer(cols = -`Module (or Feature)`, names_to = c("cname", "Analysis"), names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = cname, values_from = value) %>%
  dplyr::filter(p <= 0.05) %>%
  mutate(Analysis = recode(Analysis,
                           `slope` = "Longitudinal analysis - overall shape",
                           `intercept` = "Longitudinal analysis - overall average"),
         Direction = ifelse(test = sign(coef) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
    rename(`P value` = p,
           `Q value` = adjp) %>%
  dplyr::select(`Module (or Feature)`,
         Analysis,
         `P value`,
         `Q value`,
         Direction) %>%
  `rownames<-`(NULL)

################################################################
# use linear regression to infer direction
combMat <- combn(unique(inputDF$trajectory_group), m = 2)
coefCombMat <- apply(combMat, MARGIN = 2, FUN = function(comparison) {
  inputTemp <- dplyr::filter(inputDF, trajectory_group %in% comparison)

  coefDF <- lapply(rownames(smooth_spline_model_loop), FUN = function(modName) {
    fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
               random = ~1|enrollment_site/participant_id,
               data   = dplyr::filter(inputTemp, name %in% modName))
    average <- unique(coef(fit)[grep(pattern = "^trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    shape <- unique(coef(fit)[grep(pattern = "^event_date:trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    return(value = data.frame(`Module (or Feature)` = modName,
                              average,
                              shape,
                              comparison = paste(sort(comparison), collapse = "|"),
                              check.names = FALSE))
  }) %>%
    do.call(what = rbind)
  return(value = coefDF)
}) %>%
  do.call(what = rbind)

################################################################
supTab_resultsLongitudinalPairwise <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  dplyr::select(`Module (or Feature)`, 
         matches(match = "^p.slope_.v.$"),
         matches(match = "^p.intercept_.v.$"),
         matches(match = "^p.slope.adj_.v.$"),
         matches(match = "^p.intercept.adj_.v.$")) %>%
  setNames(nm = gsub(pattern = "p(.+).adj_",
                     replacement = "adjp\\1_",
                     names(.))) %>%
  pivot_longer(cols          = -`Module (or Feature)`, 
               names_to      = c("cname", "Analysis"), 
               names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = cname, values_from = value) %>%
  mutate(Analysis = gsub(pattern     = "slope",
                         replacement = "shape",
                         Analysis),
         Analysis = gsub(pattern     = "intercept",
                         replacement = "average",
                         Analysis),
         Analysis = gsub(pattern = "^(.+)_(.)v(.)$",
                         replacement = "Longitudinal analysis - \\2|\\3 \\1",
                         Analysis)) %>%
  rename(`P value` = p,
         `Q value` = adjp)
coefCombMat %>%
  pivot_longer(cols      = -c(`Module (or Feature)`, comparison),
               names_to  = "Analysis",
               values_to = "Direction") %>%
  mutate(Analysis = paste("Longitudinal analysis -",
                          comparison,
                          Analysis),
         Direction = ifelse(test = sign(Direction) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
  dplyr::select(-comparison) %>%
  merge(x  = supTab_resultsLongitudinalPairwise,
        by = c("Module (or Feature)", "Analysis")) -> supTab_resultsLongitudinalPairwise

# append to overall results
supTab_resultsLongitudinalPairwise %>%
  dplyr::filter(`P value` <= 0.05 &
           `Module (or Feature)` %in% supTab_resultsLongitudinal$"Module (or Feature)") %>%
  rbind(supTab_resultsLongitudinal, .) -> supTab_resultsLongitudinal


################################################################
inputDF$symptom_date[is.na(inputDF$symptom_date)] <- 0
inputDF$event_date <- inputDF$event_date - inputDF$symptom_date

################################################################
smooth_spline_model_loop <- model_loop(inputDF, modelType = "smoothSpline", endpoint = "trajectory_group")

# fetch directionality from lme
smooth_spline_model_loop$trajectory_group5 <- NA
smooth_spline_model_loop$"event_date:trajectory_group5" <- NA
for (modName in rownames(smooth_spline_model_loop)) {
  fit <- lme(fixed= value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
      random = ~1|enrollment_site/participant_id,
      data = dplyr::filter(inputDF, name %in% modName))
smooth_spline_model_loop[modName, "trajectory_group5"] <- unique(coef(fit)["trajectory_group5"])
smooth_spline_model_loop[modName, "event_date:trajectory_group5"] <- unique(coef(fit)["event_date:trajectory_group5"])
}
################################################################
supTab_resultsLongitudinalDFSO <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  dplyr::select(`Module (or Feature)`, 
         p.slope, 
         p.intercept, 
         adjp.slope, 
         adjp.intercept, 
         trajectory_group5, 
         `event_date:trajectory_group5`) %>%
  rename(coef.intercept = trajectory_group5, 
         coef.slope = `event_date:trajectory_group5`) %>%
  pivot_longer(cols = -`Module (or Feature)`, names_to = c("cname", "Analysis"), names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = cname, values_from = value) %>%
  dplyr::filter(p <= 0.05) %>%
  mutate(Analysis = recode(Analysis,
                           `slope` = "Longitudinal analysis DFSO - overall shape",
                           `intercept` = "Longitudinal analysis DFSO- overall average"),
         Direction = ifelse(test = sign(coef) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
    rename(`P value` = p,
           `Q value` = adjp) %>%
  dplyr::select(`Module (or Feature)`,
         Analysis,
         `P value`,
         `Q value`,
         Direction) %>%
  `rownames<-`(NULL)

################################################################
# use linear regression to infer direction
combMat <- combn(unique(inputDF$trajectory_group), m = 2)
coefCombMat <- apply(combMat, MARGIN = 2, FUN = function(comparison) {
  inputTemp <- dplyr::filter(inputDF, trajectory_group %in% comparison)
  
  coefDF <- lapply(rownames(smooth_spline_model_loop), FUN = function(modName) {
    fit <- lme(fixed  = value ~ event_date * trajectory_group + sex + discretized_admit_age_quantile,
               random = ~1|enrollment_site/participant_id,
               data   = dplyr::filter(inputTemp, name %in% modName))
    average <- unique(coef(fit)[grep(pattern = "^trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    shape <- unique(coef(fit)[grep(pattern = "^event_date:trajectory_group",
                                     colnames(coef(fit)))]) %>%
      unlist() %>%
      unname()
    return(value = data.frame(`Module (or Feature)` = modName,
                              average, 
                              shape,
                              comparison = paste(sort(comparison), collapse = "|"),
                              check.names = FALSE))
  }) %>%
    do.call(what = rbind)
  return(value = coefDF)
}) %>%
  do.call(what = rbind)

################################################################
supTab_resultsLongitudinalDFSOPairwise <- smooth_spline_model_loop %>%
  rownames_to_column(var = "Module (or Feature)") %>%
  dplyr::select(`Module (or Feature)`, 
         matches(match = "^p.slope_.v.$"),
         matches(match = "^p.intercept_.v.$"),
         matches(match = "^p.slope.adj_.v.$"),
         matches(match = "^p.intercept.adj_.v.$")) %>%
  setNames(nm = gsub(pattern = "p(.+).adj_",
                     replacement = "adjp\\1_",
                     names(.))) %>%
  pivot_longer(cols          = -`Module (or Feature)`, 
               names_to      = c("cname", "Analysis"), 
               names_pattern = "(.*)\\.(.*)") %>%
  pivot_wider(names_from = cname, values_from = value) %>%
  mutate(Analysis = gsub(pattern     = "slope",
                         replacement = "shape",
                         Analysis),
         Analysis = gsub(pattern     = "intercept",
                         replacement = "average",
                         Analysis),
         Analysis = gsub(pattern = "^p\\.(.+)_(.)v(.)$",
                         replacement = "Longitudinal analysis DFSO - \\2|\\3 \\1",
                         Analysis)) %>%
  rename(`P value` = p,
         `Q value` = adjp)
coefCombMat %>%
  pivot_longer(cols      = -c(`Module (or Feature)`, comparison),
               names_to  = "Analysis",
               values_to = "Direction") %>%
  mutate(Analysis = paste("Longitudinal analysis DFSO -",
                          comparison,
                          Analysis),
         Direction = ifelse(test = sign(Direction) %in% 1,
                            yes  = "Severe",
                            no   = "Mild")) %>%
  dplyr::select(-comparison) %>%
  merge(x  = supTab_resultsLongitudinalDFSOPairwise,
        by = c("Module (or Feature)", "Analysis")) -> supTab_resultsLongitudinalDFSOPairwise

# append to overall results
supTab_resultsLongitudinalDFSOPairwise %>%
  dplyr::filter(`P value` <= 0.05 &
           `Module (or Feature)` %in% supTab_resultsLongitudinalDFSO$"Module (or Feature)") %>%
  rbind(supTab_resultsLongitudinalDFSO, .) -> supTab_resultsLongitudinalDFSO

```

### Supplementary Table A: output generation code
```{r sup-table-a-output-gen}
# combine all sup table results
supTab_results <- rbind(supTab_resultsVisit1,
                        supTab_resultsLongitudinal,
                        supTab_resultsLongitudinalDFSO)

write_csv(supTab_results, file = file.path(out_dir, "Nasal_Metagenomics_Results.csv"))
```

### Session info
```{r, print-session-info}
sessionInfo()
```
<!--
## R version 4.0.2 (2020-06-22)
## Platform: x86_64-pc-linux-gnu (64-bit)
## Running under: Ubuntu 18.04.4 LTS
## 
## Matrix products: default
## BLAS:   /usr/lib/x86_64-linux-gnu/openblas/libblas.so.3
## LAPACK: /usr/lib/x86_64-linux-gnu/libopenblasp-r0.2.20.so
## 
## locale:
##  [1] LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8       
##  [4] LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8    LC_MESSAGES=C.UTF-8   
##  [7] LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C          
## [10] LC_TELEPHONE=C         LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   
## 
## attached base packages:
## [1] parallel  grid      stats     graphics  grDevices utils     datasets 
## [8] methods   base     
## 
## other attached packages:
##  [1] doParallel_1.0.17    doRNG_1.8.2          rngtools_1.5.2      
##  [4] missForest_1.4       itertools_0.1-3      iterators_1.0.14    
##  [7] foreach_1.5.2        randomForest_4.6-14  ComplexHeatmap_2.6.2
## [10] corrr_0.4.3          gridExtra_2.3        rlang_1.0.2         
## [13] RColorBrewer_1.1-2   cowplot_1.1.1        ggpubr_0.4.0        
## [16] pals_1.7             vegan_2.6-2          lattice_0.20-41     
## [19] permute_0.9-7        forcats_0.5.1        stringr_1.4.0       
## [22] dplyr_1.0.8          purrr_0.3.4          readr_2.1.2         
## [25] tidyr_1.2.0          tibble_3.1.6         tidyverse_1.3.1     
## [28] ggbeeswarm_0.6.0     ggplot2_3.3.5        pvca_0.1.0          
## [31] ggeffects_1.1.1      qvalue_2.22.0        ordinal_2019.12-10  
## [34] gamm4_0.2-6          mgcv_1.8-31          nlme_3.1-148        
## [37] lme4_1.1-28          Matrix_1.2-18        knitr_1.37          
## 
## loaded via a namespace (and not attached):
##  [1] minqa_1.2.4         colorspace_2.0-3    ggsignif_0.6.3     
##  [4] rjson_0.2.21        ellipsis_0.3.2      sjlabelled_1.1.8   
##  [7] circlize_0.4.14     GlobalOptions_0.1.2 fs_1.5.2           
## [10] dichromat_2.0-0     clue_0.3-60         rstudioapi_0.13    
## [13] farver_2.1.0        bit64_4.0.5         fansi_0.4.1        
## [16] lubridate_1.8.0     xml2_1.3.2          codetools_0.2-16   
## [19] splines_4.0.2       jsonlite_1.8.0      nloptr_2.0.0       
## [22] Cairo_1.5-14        broom_0.7.12        cluster_2.1.0      
## [25] dbplyr_2.1.1        png_0.1-7           mapproj_1.2.8      
## [28] compiler_4.0.2      httr_1.4.2          backports_1.2.0    
## [31] assertthat_0.2.1    fastmap_1.1.0       cli_3.2.0          
## [34] htmltools_0.5.2     tools_4.0.2         gtable_0.3.0       
## [37] glue_1.6.2          reshape2_1.4.4      maps_3.4.0         
## [40] Rcpp_1.0.8          carData_3.0-5       cellranger_1.1.0   
## [43] vctrs_0.3.8         insight_0.16.0      xfun_0.30          
## [46] rvest_1.0.2         lifecycle_1.0.1     rstatix_0.7.0      
## [49] MASS_7.3-51.6       scales_1.1.1        vroom_1.5.7        
## [52] hms_1.1.1           yaml_2.2.1          stringi_1.5.3      
## [55] highr_0.8           ucminf_1.1-4        S4Vectors_0.28.1   
## [58] BiocGenerics_0.36.1 boot_1.3-25         shape_1.4.6        
## [61] pkgconfig_2.0.3     matrixStats_0.61.0  evaluate_0.14      
## [64] labeling_0.4.2      bit_4.0.4           tidyselect_1.1.2   
## [67] plyr_1.8.6          magrittr_2.0.2      R6_2.5.0           
## [70] IRanges_2.24.1      generics_0.1.2      DBI_1.1.2          
## [73] pillar_1.7.0        haven_2.4.3         withr_2.5.0        
## [76] abind_1.4-5         modelr_0.1.8        crayon_1.5.0       
## [79] car_3.0-12          utf8_1.1.4          tzdb_0.2.0         
## [82] rmarkdown_2.12      GetoptLong_1.0.5    readxl_1.3.1       
## [85] reprex_2.0.1        digest_0.6.27       numDeriv_2016.8-1.1
## [88] stats4_4.0.2        munsell_0.5.0       beeswarm_0.4.0     
## [91] vipor_0.4.5
-->

