---
title: "IMPACC_deep_dive_figure2"
output: html_document
date: "2023-12-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(DT)
library(e1071)
library(openxlsx)
library(stringr)
library(dplyr)
library(plyr)
source("~/Bitbucket/impacc-full-cohort/Analysis/IMPACC_data_analysis_codebase.R")
load("/scratch/metabolomics/plasma-metabolomics-global/scripts/deep_dive_met/deep_dive_data.Rdata")

PVCA <- function(counts, meta, threshold, inter){
  #counts = tmet
  #            meta = pdata
  #            inter = FALSE
  #            threshold = 0.6
  counts.center <- t(apply(counts, 1, scale, center=TRUE, scale=FALSE))
  cor.counts <- cor(counts.center)
  dim(cor.counts)
  eigen.counts <- eigen(cor.counts)
  eigen.mat <- eigen.counts$vectors
  eigen.val <- eigen.counts$values
  n.eigen <- length(eigen.val)
  eigen.val.sum <- sum(eigen.val)
  percents.pcs <- eigen.val/eigen.val.sum
  meta <- as.data.frame(meta)
  
  all <- 0
  npc.in <- 0
  for(i in 1:n.eigen){
    all <- all + percents.pcs[i]
    npc.in <- npc.in + 1
    if(all > threshold){break}
  }
  if (npc.in < 3) {npc <- 3}
  
  pred.list <- colnames(meta)
  meta <- droplevels(meta)
  
  n.preds <- ncol(meta) + 1
  if(inter) {n.preds <- n.preds + choose(ncol(meta),2)}
  
  ran.pred.list <- c()
  for(i in 1:ncol(meta)){
    ran.pred.list <- c(ran.pred.list, paste0("(1|", pred.list[i],")"))
  }
  ##interactions
  if(inter){
    for(i in 1:(ncol(meta)-1)){
      for(j in (i+1):ncol(meta)){
        ran.pred.list <- c(ran.pred.list, paste0("(1|", pred.list[i], ":", pred.list[j], ")"))
        pred.list <- c(pred.list, paste0(pred.list[i], ":", pred.list[j]))
      }
    }
  }
  formula <- paste(ran.pred.list, collapse = " + ")
  formula <- paste("pc", formula, sep=" ~ ")
  ran.var.mat <- NULL
  for(i in 1:npc.in){
    dat <- cbind(eigen.mat[,i],meta)
    colnames(dat) <- c("pc",colnames(meta))
    Rm1ML <- lme4::lmer(formula, dat, REML = TRUE, verbose = FALSE, na.action = na.omit)
    var.vec <- unlist(VarCorr(Rm1ML))
    ran.var.mat <- rbind(ran.var.mat, c(var.vec[pred.list], resid = sigma(Rm1ML)^2))
  }
  ran.var.mat.std <- ran.var.mat/rowSums(ran.var.mat)
  wgt.vec <- eigen.val/eigen.val.sum
  prop.var <- colSums(ran.var.mat.std*wgt.vec[1:npc.in])
  std.prop.var <- prop.var/sum(prop.var)
  std.prop.var
}
```

```{r}
Metabolomics_counts_conv = Metabolomics_counts_conv[clinical_data$sample_id,]
```

#PCA before and after
## QC script to generate supplimentary pannel A-C

```{r}
path <- "/scratch/metabolomics/plasma-metabolomics-global/rawfiles/phase4b/"
mets <- readxl::read_excel(paste0(path, "CHHB-0210-20DSML+_P1-P4B MERGED DATA TABLES_COMMON_2.XLSX"), sheet = "QC-norm Data Common")
mets_info <-  readxl::read_excel(paste0(path, "CHHB-0210-20DSML+_P1-P4B MERGED DATA TABLES_COMMON_1.XLSX"), sheet = "Chemical Annotation Common QC")
sam <- readxl::read_excel(paste0(path, "CHHB-0210-20DSML+_P1-P4B MERGED DATA TABLES_ALL_1.XLSX"), sheet = "Sample Meta Data")
```

```{r}
path2 <- "/scratch/metabolomics/plasma-metabolomics-global/rawfiles/phase3_merged/"
mets_info_2 <- read.csv(paste0(path2, "CHHB-0207-20DSML+_Chemical Annotations_updated 03Jan2022 with RT.csv"))
CHEM_ID <- mets_info_2$CHEM_ID
RT <- mets_info_2$RT
#mets_info_2 <- data.table(CHEM_ID, RT)

mets_info <- merge(mets_info, mets_info_2, by = "CHEM_ID")
```

```{r}
batch_layout <- read.csv(file =
"/scratch/metabolomics/plasma-metabolomics-global/rawfiles/phase1_2_merged/STEEN_Phase1and2_5_6_2021.csv")

batch_layout_2 <- read.csv(file =
"/scratch/metabolomics/plasma-metabolomics-global/rawfiles/phase3_merged/STEEN_LAB_Phase3_12_2_2021.csv")

batch_layout_2 <- read.csv(file =
"/impacc-study/qa/scripts/batchlayout_csv/Phase4a/STEEN_LAB_Samples_Lab_01_19_2022.csv")

batch_layout <- rbind.fill(batch_layout, batch_layout_2)
```


```{r}
batch_layout <- read.csv(file =
"/scratch/metabolomics/plasma-metabolomics-global/rawfiles/phase1_2_merged/STEEN_Phase1and2_5_6_2021.csv")

batch_layout_2 <- read.csv(file =
"/scratch/metabolomics/plasma-metabolomics-global/rawfiles/phase3_merged/STEEN_LAB_Phase3_12_2_2021.csv")
batch_layout_2$phase <- 3

batch_layout_3 <- read.csv(file =
"/impacc-study/qa/scripts/batchlayout_csv/Phase4a/STEEN_LAB_Samples_Lab_01_19_2022.csv")
batch_layout_3$phase <- 4

batch_layout <- rbind.fill(batch_layout, batch_layout_2)
batch_layout <- rbind.fill(batch_layout, batch_layout_3)
```

```{r}
batch_layout <- batch_layout[!duplicated(batch_layout$guspec),]
```

# 3. Convert to Data Table

```{r}
mets <- data.table(mets)
mets_info <- data.table(mets_info)
sam <- data.table(sam)
```

```{r}
check <- data.table(colnames(mets)[2:ncol(mets)], mets_info$CHEM_ID)
```

# 4. Drop Columns with no client identified

```{r}
check <- data.table(mets$PARENT_SAMPLE_NAME, sam$PARENT_SAMPLE_NAME)
```

# 5. Add to CORRECTED_SAMPLEID

```{r}
for (i in 1:nrow(sam)){
  if(is.na(sam$CLIENT_SAMPLE_ID[i])){
    sam$CLIENT_SAMPLE_ID[i] <- sam$CORRECTED_SAMPLEID[i]  
  }
}
```

```{r}
sam$CORRECTED_SAMPLEID <- sam$CLIENT_SAMPLE_ID
```

```{r}
#sam$CORRECTED_SAMPLEID <- sam$PARTICIPANT_ID
```

```{r}
sam_2 <-  read.csv("/scratch/metabolomics/plasma-metabolomics-global/scripts/kevin/Metabolon_SampleManifest_IMPACC-Phase2_04_06_2021.csv")
sam_2 <- sam_2[!is.na(sam_2$Unique.Tube.Label.ID.),]
sam_2 <- sam_2[!is.na(sam_2$Client.Sample.ID.),]
sam_2$CLIENT_IDENTIFIER <- str_sub(sam_2$Unique.Tube.Label.ID., 2, 20)
sam <- sam[!is.na(sam$CLIENT_IDENTIFIER),]

for(i in 1:nrow(sam)){
  for(j in 1:nrow(sam_2)){
    if(sam$CLIENT_IDENTIFIER[i] == sam_2$CLIENT_IDENTIFIER[j]){
      sam$CORRECTED_SAMPLEID[i] <- sam_2$Client.Sample.ID.[j]
      sam$CLIENT_MATRIX[i] <- sam_2$Client.Matrix.[j]
    }
  }
}

table(sam$CLIENT_MATRIX)
```


# 5. If URS include _ value

```{r}
for (i in 1:nrow(sam)){
  if (grepl("URS", sam$CORRECTED_SAMPLEID[i], fixed = TRUE) == TRUE){
    sam$CORRECTED_SAMPLEID[i] <- paste0(sam$CORRECTED_SAMPLEID[i], "_", sam$CLIENT_IDENTIFIER[i])
  }
}
```


# 4. Change sample_id to CORRECTED_SAMPLEID

```{r}
check <- data.table(mets$PARENT_SAMPLE_NAME, sam$PARENT_SAMPLE_NAME)
all(mets$PARENT_SAMPLE_NAME == sam$PARENT_SAMPLE_NAME)
```

```{r}
sample_id <- sam$CORRECTED_SAMPLEID
mets[, "PARENT_SAMPLE_NAME":=NULL] 

mets <- data.table(sample_id, mets)
colnames(mets) <- paste0("X", colnames(mets))
```

```{r}
colnames(mets)[1] <- "sample_id"
```

```{r}
mets <- mets[!is.na(mets$sample_id),]
sam <- sam[!is.na(sam$CORRECTED_SAMPLEID),]
```

# 5. Create Name Column

```{r}
# Name column matches metabolite IDs in mets
MET_ID <- paste0("X", mets_info$CHEM_ID)
mets_info <- data.table(MET_ID, mets_info)
```

# Remove duplicates

```{r}
mets <- mets[!duplicated(mets$sample_id),]
sam <- sam[!duplicated(sam$CORRECTED_SAMPLEID),]
#0865-005HTH00-005 -> row 906 and row 2183
```

# 6. Create Percentage Missing Column (by Metabolite)

```{r}
mets_list <- mets_info$MET_ID
x <- mets[, ..mets_list]
perc_missing <- colSums(is.na(x)) / nrow(x) * 100
mets_info$perc_missing <- perc_missing 

plot(hist(mets_info$perc_missing))
summary(mets_info$perc_missing)
```

# 7. Impute Missing Values (Half-Min Approach)

```{r}
impute_halfmin <- function(x) {replace(x, is.na(x), (min(x, na.rm = T)/2))}
mets_list <- mets_info$MET_ID

mets[, (mets_list) := lapply(.SD, impute_halfmin), .SDcols = mets_list] 
```
#F2_panel_A
```{r}
mets_list <- mets_info$MET_ID

# Create PCA
pca_pre <- prcomp(mets[, ..mets_list], scale. = F)
fit_pre <-summary(pca_pre)
pca_pre_scores <- as.data.frame(fit_pre$x)

# Plot PCA
F2_panel_A <- ggplot(data = pca_pre_scores,
       mapping = aes(x = PC1, y = PC2))+
  geom_point(size = 3, alpha = 0.7) +  ggtitle("PCA-PreQC")+
  labs(x     = paste0("1st dimension (",
                      round((pca_pre$sdev^2/sum(pca_pre$sdev^2))[1] * 100),
                      "%)"),
       y     = paste0("2nd dimension (",
                      round((pca_pre$sdev^2/sum(pca_pre$sdev^2))[2] * 100),
                      "%)")#,
       )+ #tag   = "A"
  theme_classic() +
  theme(legend.text     = element_text(size = 6),
        legend.key.size = unit(0.01, units = "npc"))+ theme(axis.title=element_text(size=15), legend.title= element_text (size=15))+theme(strip.text.x = element_text(size = 13)) + theme(axis.text = element_text(size = 15), legend.text = element_text(size=15)) 

```
# 8. Check IQR

```{r}
mets_list <- mets_info$MET_ID

iqr<-apply(mets[,..mets_list], 2, IQR, na.rm=T)
iqr<-as.data.frame(iqr)
summary(iqr)

mets_info$iqr_pre <- iqr

length(iqr[iqr == 0])

# 212 Metabolites with an IQR = 0, but 62 are not Xenobiotics
```

# 9. Drop Metabolites with IQR = 0 (for non-xenobiotics)

```{r}
mets_list_rm <- mets_info$MET_ID[mets_info$iqr == 0 & mets_info$SUPER_PATHWAY != "Xenobiotics"]
mets_list_rm <- mets_list_rm[complete.cases(mets_list_rm)]

mets[, c(mets_list_rm):=NULL] 
mets_info <- mets_info[!(mets_info$MET_ID %in% mets_list_rm), ]

# Check mets_info rows = 1358, and mets columns = 1358 + 1
dim(mets)
dim(mets_info)
```

# 10. Check Skewness

```{r}
mets_list <- mets_info$MET_ID

skew <-apply(mets[,..mets_list], 2, skewness, na.rm=T)
summary(skew)

skew_2<-subset(skew, skew>=2|skew<=-2)
length(skew_2)
length(mets_list)

mets_info$skew_pre <- skew

## 519 out of 658 Metabolites are skewed. Need to log10 transform.
```

# 11. Log10 Transformation & Pareto-Scale

```{r}
mets_list <- mets_info$MET_ID

# Log10 Transformation
mets[, (mets_list) := log10(.SD), .SDcols = mets_list] 

# Pareto-Scale
pareto_scaling <- function(x) {(x - mean(x, na.rm = T))/sqrt(sd(x, na.rm = T))}
mets[, (mets_list) := lapply(.SD, pareto_scaling) %>% as.data.table(), .SDcols = mets_list]
```

# 12. Re-Check IQR

```{r}
mets_list <- mets_info$MET_ID

iqr<-apply(mets[,..mets_list], 2, IQR, na.rm=T)
iqr<-as.data.frame(iqr)
summary(iqr)

mets_info$iqr_post <- iqr 

# No IQR = 0, except Xenobiotics
```

# 13. Re-Check Skewness

```{r}
mets_list <- mets_info$MET_ID

skew <-apply(mets[,..mets_list], 2, skewness, na.rm=T)
summary(skew)

skew_2<-subset(skew, skew>=2|skew<=-2)
length(skew_2)
length(mets_list)

mets_info$skew_post <- skew

## 34 out of 658 Metabolites are still skewed. 
```


#F2_panel_B
```{r}
mets_list <- mets_info$MET_ID

# Create PCA
pca_post <- prcomp(mets[, ..mets_list], scale. = F)
fit_post <-summary(pca_post)
pca_post_scores <- as.data.frame(fit_post$x)

# Plot PCA
F2_panel_B <- ggplot(data = pca_post_scores,
       mapping = aes(x = PC1, y = PC2))+ 
  geom_point(size = 3, alpha = 0.7) +  ggtitle("PCA-PostQC")+
  labs(x     = paste0("1st dimension (",
                      round((pca_post$sdev^2/sum(pca_post$sdev^2))[1] * 100),
                      "%)"),
       y     = paste0("2nd dimension (",
                      round((pca_post$sdev^2/sum(pca_post$sdev^2))[2] * 100),
                      "%)")#,
       )+ #tag   = "A"
  theme_classic() +
  theme(legend.text     = element_text(size = 6),
        legend.key.size = unit(0.01, units = "npc"))+ theme(axis.title=element_text(size=15), legend.title= element_text (size=15))+theme(strip.text.x = element_text(size = 13)) + theme(axis.text = element_text(size = 15), legend.text = element_text(size=15)) 
### No outliers identified?
```

#figure 2: panel C-D
```{r pvca, warning=FALSE}
##### Authors: Slim Fourati, Jingjing Qi, Brian Lee

# step 1: load required packages
library(package = "pvca") # devtools::install_github("dleelab/pvca")
VarCorr <- lme4::VarCorr


# step 2: specify which assay is going to be used as input
pvca_input_matrix <- t(Metabolomics_counts_conv)
if (sum(is.na(pvca_input_matrix)) > 0) {
  stop("pvca input matrix should not have missing values")
}
pvca_input_raw <- t(Metabolomics_counts_conv)
plasma_metabolomics_global_metadata = plasma_metabolomics_global_metadata[rownames(Metabolomics_counts_conv),]
pvca_input_metadata <- plasma_metabolomics_global_metadata
#clinical_data1 = subset(clinical_data, clinical_data$event_type == "Visit 1")

# step3: select variable to be included in PVCA
#   event_data_week: day from admission to to the hospital coded as week (categorical variable)
#   sympt_date_week: day from onset of symptoms coded as week (categorical variable)
#   age_decate: age at admission coded as decade (admit_age modulo 10)
pvca_input_phenodata <- clinical_data %>% #clinical_data1
  mutate(#event_date_week = findInterval(event_date, 
          #                              vec        = c(0, 7, 14, 21, 28), 
         #                               all.inside = TRUE),
         day_from_sympt  = event_date - symptom_date,
         sympt_date_week = findInterval(day_from_sympt, 
                                        vec        = c(0, 7, 14, 21, 28, Inf),
                                        all.inside = TRUE)) %>%
    dplyr::select(sample_id, adm_date_cat, event_type,
           sex, sympt_date_week, event_location, participant_id, 
           discretized_admit_age_quantile, ethnicity, race,  trajectory_group, enrollment_site, bmi, conv_cluster4,Type)
# step 4 append meta data to pvca phenodata
#   core site: where are the sample processed (this chunk of code need to be modified for each core assay)
#   plate = interaction term of phase and plate
pvca_input_phenodata <- pvca_input_metadata %>%
  #rownames_to_column(var = "sample_id") %>%
  mutate(         plate     = interaction(phase, plate, drop = TRUE)) %>%
  dplyr::select(sample_id, plate, phase) %>%
  merge(x     = pvca_input_phenodata,
        by    = "sample_id",
        all.x = TRUE)

# append rownames
pvca_input_phenodata <- pvca_input_phenodata %>%
  column_to_rownames(var = "sample_id")
pvca_input_phenodata <- pvca_input_phenodata[colnames(pvca_input_matrix), , drop = FALSE]

# add percent missing values (if processeed matrix required imputation otherwise comment line below)
#pvca_input_phenodata$percent_missing <- factor(colMeans(is.na(pvca_input_raw[, rownames(pvca_input_phenodata)])))

# step 5: run PVCA
fit <- PVCA(counts    = pvca_input_matrix,
            meta      = pvca_input_phenodata,
            inter     = FALSE,
            threshold = 0.6)

# step 6: plot barplot with PVCA results
pvca_barplot_data <- data.frame(explained = as.vector(fit),
                                effect    = names(fit)) %>%
  arrange(explained) %>%
  mutate(effect = factor(effect, levels = effect))

F2_panel_C<- ggplot(data    = pvca_barplot_data,
       mapping = aes(x = effect, y = explained)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = signif(explained, digits = 3)),
            nudge_y   = 0.02,
            size      = 3) +
  labs(x = NULL, y = "Proportion of the variance explained") + 
  theme_bw() + coord_flip () +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1)) + theme(axis.title=element_text(size=10), legend.title= element_text (size=10))+theme(strip.text.x = element_text(size = 10)) + theme(axis.text = element_text(size = 10), legend.text = element_text(size=10))

```
## PVCA analysis with technical+biological variables
# to use: modify step2 variable to your assay of interest
```{r pvca, warning=FALSE}
##### Authors: Slim Fourati, Jingjing Qi, Brian Lee

# step 1: load required packages
library(package = "pvca") # devtools::install_github("dleelab/pvca")
VarCorr <- lme4::VarCorr

# step 2: specify which assay is going to be used as input
pvca_input_matrix <- t(Metabolomics_counts_conv)
if (sum(is.na(pvca_input_matrix)) > 0) {
  stop("pvca input matrix should not have missing values")
}
pvca_input_raw <- t(Metabolomics_counts_conv)
pvca_input_metadata <- plasma_metabolomics_global_metadata

# step3: select variable to be included in PVCA
#   event_data_week: day from admission to to the hospital coded as week (categorical variable)
#   sympt_date_week: day from onset of symptoms coded as week (categorical variable)
#   discretized_admit_age_quantile: age at admission coded as quintiles 
clinical_data1 = subset(clinical_data, clinical_data$event_type == "Visit 1")
pvca_input_phenodata <- clinical_data1 %>%
  mutate(#event_date_week = findInterval(event_date, 
         #                               vec        = c(0, 7, 14, 21, 28), 
         #                               all.inside = TRUE),
         day_from_sympt  = event_date - symptom_date,
         sympt_date_week = findInterval(day_from_sympt, 
                                        vec        = c(0, 7, 14, 21, 28, Inf),
                                        all.inside = TRUE)) %>%
    dplyr::select(sample_id, adm_date_cat, respiratory_status,
           sex, sympt_date_week, 
           discretized_admit_age_quantile, ethnicity, race, admit_month,
           trajectory_group, enrollment_site, bmi, conv_cluster4,Type)

# step 4 append meta data to pvca phenodata
#   core site: where are the sample processed (this chunk of code need to be modified for each core assay)
#   plate = interaction term of phase and plate
pvca_input_phenodata <- pvca_input_metadata %>%
  #rownames_to_column(var = "sample_id") %>%
  mutate(         plate     = interaction(phase, plate, drop = TRUE)) %>%
  dplyr::select(sample_id, plate, phase) %>%
  merge(x     = pvca_input_phenodata,
        by    = "sample_id",
        all.x = TRUE)

# append rownames
pvca_input_phenodata <- pvca_input_phenodata %>%
  column_to_rownames(var = "sample_id")
pvca_input_matrix = pvca_input_matrix[,rownames(pvca_input_phenodata)]
pvca_input_phenodata <- pvca_input_phenodata[colnames(pvca_input_matrix), , drop = FALSE]

# add percent missing values (if processeed matrix required imputation otherwise comment line below)
#pvca_input_phenodata$percent_missing <- factor(colMeans(is.na(pvca_input_raw[, rownames(pvca_input_phenodata)])))

# step 5: run PVCA
fit <- PVCA(counts    = pvca_input_matrix,
            meta      = pvca_input_phenodata,
            inter     = FALSE,
            threshold = 0.6)

# step 6: plot barplot with PVCA results
pvca_barplot_data <- data.frame(explained = as.vector(fit),
                                effect    = names(fit)) %>%
  arrange(explained) %>%
  mutate(effect = factor(effect, levels = effect))

F2_panel_F <- ggplot(data    = pvca_barplot_data,
       mapping = aes(x = effect, y = explained)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = signif(explained, digits = 3)),
            nudge_y   = 0.02,
            size      = 3) +
  labs(x = NULL, y = "Proportion of the variance explained") + 
  theme_bw() + coord_flip () +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1)) + theme(axis.title=element_text(size=10), legend.title= element_text (size=10))+theme(strip.text.x = element_text(size = 10)) + theme(axis.text = element_text(size = 10), legend.text = element_text(size=10))

```
## PVCA analysis with technical+biological variables
# to use: modify step2 variable to your assay of interest
```{r pvca, warning=FALSE}
##### Authors: Slim Fourati, Jingjing Qi, Brian Lee

# step 1: load required packages
library(package = "pvca") # devtools::install_github("dleelab/pvca")
VarCorr <- lme4::VarCorr

# step 2: specify which assay is going to be used as input
pvca_input_matrix <- t(Metabolomics_counts_conv)
if (sum(is.na(pvca_input_matrix)) > 0) {
  stop("pvca input matrix should not have missing values")
}
pvca_input_raw <- t(Metabolomics_counts_conv)
pvca_input_metadata <- plasma_metabolomics_global_metadata

# step3: select variable to be included in PVCA
#   event_data_week: day from admission to to the hospital coded as week (categorical variable)
#   sympt_date_week: day from onset of symptoms coded as week (categorical variable)
#   discretized_admit_age_quantile: age at admission coded as quintiles 

pvca_input_phenodata <- clinical_data %>%
  mutate(#event_date_week = findInterval(event_date, 
         #                               vec        = c(0, 7, 14, 21, 28), 
         #                               all.inside = TRUE),
         day_from_sympt  = event_date - symptom_date,
         sympt_date_week = findInterval(day_from_sympt, 
                                        vec        = c(0, 7, 14, 21, 28, Inf),
                                        all.inside = TRUE)) %>%
    dplyr::select(sample_id, adm_date_cat, event_type, respiratory_status,
           sex, sympt_date_week, event_location,
           discretized_admit_age_quantile, ethnicity, race, admit_month,
           trajectory_group, enrollment_site, bmi, conv_cluster4,Type)

# step 4 append meta data to pvca phenodata
#   core site: where are the sample processed (this chunk of code need to be modified for each core assay)
#   plate = interaction term of phase and plate
pvca_input_phenodata <- pvca_input_metadata %>%
  #rownames_to_column(var = "sample_id") %>%
  mutate(         plate     = interaction(phase, plate, drop = TRUE)) %>%
  dplyr::select(sample_id, plate, phase) %>%
  merge(x     = pvca_input_phenodata,
        by    = "sample_id",
        all.x = TRUE)

# append rownames
pvca_input_phenodata <- pvca_input_phenodata %>%
  column_to_rownames(var = "sample_id")
pvca_input_phenodata <- pvca_input_phenodata[colnames(pvca_input_matrix), , drop = FALSE]

# add percent missing values (if processeed matrix required imputation otherwise comment line below)
#pvca_input_phenodata$percent_missing <- factor(colMeans(is.na(pvca_input_raw[, rownames(pvca_input_phenodata)])))

# step 5: run PVCA
fit <- PVCA(counts    = pvca_input_matrix,
            meta      = pvca_input_phenodata,
            inter     = FALSE,
            threshold = 0.6)

# step 6: plot barplot with PVCA results
pvca_barplot_data <- data.frame(explained = as.vector(fit),
                                effect    = names(fit)) %>%
  arrange(explained) %>%
  mutate(effect = factor(effect, levels = effect))

F2_panel_D <- ggplot(data    = pvca_barplot_data,
       mapping = aes(x = effect, y = explained)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = signif(explained, digits = 3)),
            nudge_y   = 0.02,
            size      = 3) +
  labs(x = NULL, y = "Proportion of the variance explained") + 
  theme_bw() + coord_flip () +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1)) + theme(axis.title=element_text(size=10), legend.title= element_text (size=10))+theme(strip.text.x = element_text(size = 10)) + theme(axis.text = element_text(size = 10), legend.text = element_text(size=10))

```
```{r}
ggarrange(F2_panel_A,F2_panel_B, F2_panel_C,F2_panel_F, ncol = 2 ,nrow = 2,labels = c("A","B","C","D"))
```

