---
title: "IMPACC_deep_dive_supplementary_TG"
output: html_document
date: "2023-12-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r define dataset version, echo = FALSE}
##### Authors: Ravi K. Patel, Jeremy Gygi;

# Indicate the version of the data to use ("current" or a date in "YYYY-MM-DD" format)
# If "current", the data files in "current" directories are used, otherwise, the data files in "legacy" directories that were current on the specified date are used.
DATA_VERSION <- "current"

# Indicate the phases that you want to include.
# NOTE: PHASES = 4 refers to 4a
#.      PHASES = 5 refers to 4b
#.      PHASES = 9 refers to phase 9 (leftover samples)
#.      PHASES = NA refers to phase NA
PHASES <- c(1,2,3,4,5,9, NA)

# Allowed QC? Defaults to pass (including questionable)
ALLOWED_SMPL_STATUS <- c("IMPACC sample assayed and passed QC", 
                         "IMPACC sample assayed with questionable QC")
#ALLOWED_SMPL_STATUS = "ALL"
# Keep only COVID-19 Positive individuals. Defaults to TRUE
# (TRUE will remove all COVID-19 Negative individuals)
KEEP_COVID19_POS <- TRUE
# Keep Escalation visits (Escalation 1, Escalation 2)? Defaults to FALSE
KEEP_ESCALATIONS = FALSE
# Keep Emory Controls? Defaults to FALSE
KEEP_EMORY_CONTROLS = FALSE

# Upper limit for event_date.
# Using higher values (e.g. 999) will include all visits (see KEEP_ESCALATIONS
#   for inclusion of Escalation visits).
EVENT_DATE_UPPER = 999

# Upper limit for event_type. For example, 6 would use up to 'Visit 6'.
# Using higher values (e.g. 999) will include all visits (see KEEP_ESCALATIONS
#   for inclusion of Escalation visits).
VISIT_LOWER = 1
VISIT_UPPER = 999

# Use 01/01/2023 locked data in "/data/clinical/current"?
USE_LOCKED_CLINICAL <- TRUE

# Filtering by manuscript cohorts:
# Currently supports core assay / integration manuscript cohort filtering.
FILTER_BY_CORE_ASSAY_COHORT = FALSE
FILTER_BY_INTEGRATION_COHORT = FALSE

# Provide vector of participant IDs to restrict loading the data to those participants only.
SELECT_PARTICIPANT_ID = NULL

# Provide vector of samples to exclude from the analysis
EXCLUDE_SAMPLES = NULL
```

<!-- ################################################################################################### -->
<!--           No need to edit any of the code below this point until end of Mandatory Code chunk        -->
<!-- ################################################################################################### -->

```{r Prepare environment, echo=FALSE}
##### Authors: Ravi K. Patel, Jeremy Gygi.
# Set options
options(encoding = "UTF-8") # instead of ANSI
options(scipen = 999) # non-scientific output
```

### ------------------------------------
#### **Study:** IMPACC
#### **Analysis title:** IMPACC Convalescent Data Loading
#### **Analysis description:** Loading combined data from all phases for analyses.
#### **License:** 
#### **Author(s):** Ravi K. Patel (ravi.patel2@ucsf.edu), Jeremy Gygi (jeremy.gygi@yale.edu), Steven Kleinstein (steven.kleinstein@yale.edu), Leying Guan (leying.guan@yale.edu)
#### **Date Created:** `r paste(format(Sys.time(), "%Y-%d-%m %X"), "UTC")`

#### **Source file:** `r knitr::current_input()`
#### **Dataset version:** `r ifelse(DATA_VERSION == "current", paste0("current (date: ", format(Sys.time(),"%Y-%m-%d"), ")"), paste0("legacy (date: ", DATA_VERSION, ")") )` `r paste0("(phases: ", paste0(PHASES, collapse=","), ")") `
#### **R version:** `r paste(R.Version()$version.string)`
#### **Platform:** `r paste(R.Version()$platform)`
#### **Running under**: `r paste(sessionInfo()$running)`

#### **Notes:** The observations for which the clinical data is unavailable or have failed QC are removed.
### ------------------------------------

```{r load header template codebase, echo=FALSE}
# Source the codebase file:
source("~/Bitbucket/impacc-full-cohort/Analysis/IMPACC_data_analysis_codebase.R")
```

```{r load datasets, message=FALSE, echo=FALSE}
# Load all available IMPACC datasets associated with DATA_VERSION.
data_env <- load_IMPACC_datasets(# Data version:
                                 DATA_VERSION = DATA_VERSION,
                                 # QC Sample Filtering:
                                 ALLOWED_SMPL_STATUS = ALLOWED_SMPL_STATUS,
                                 # COVID-19 Positive
                                 KEEP_COVID19_POS = KEEP_COVID19_POS,
                                 # Emory controls:
                                 KEEP_EMORY_CONTROLS = KEEP_EMORY_CONTROLS,
                                 # Escalations:
                                 KEEP_ESCALATIONS = KEEP_ESCALATIONS,
                                 # Use the locked clinical data (01/01/2023)
                                 USE_LOCKED_CLINICAL = USE_LOCKED_CLINICAL,
                                 
                                 # Don't restrict filtering by manuscript:
                                 FILTER_BY_CORE_ASSAY_COHORT = FILTER_BY_CORE_ASSAY_COHORT,
                                 FILTER_BY_INTEGRATION_COHORT = FILTER_BY_INTEGRATION_COHORT,
                                 
                                 # Select participant ID
                                 SELECT_PARTICIPANT_ID = SELECT_PARTICIPANT_ID,
                                 
                                 # Exclude samples
                                 EXCLUDE_SAMPLES = EXCLUDE_SAMPLES,
                                 
                                 # event_type / visit upper limits
                                 EVENT_DATE_UPPER = EVENT_DATE_UPPER,
                                 VISIT_LOWER = VISIT_LOWER,
                                 VISIT_UPPER = VISIT_UPPER,
                                 
                                 # Phases
                                 PHASES = PHASES)

# Load and assign the data objects into the environment
# User can call upon any assay via the name e.g. serum_olink_counts
for( n in names(data_env) ) {
  assign( n, data_env[[n]] )
}

```
## Metabolomics concentration

```{r}
# Concentration
Metabolomics_counts <- plasma_metabolomics_global_counts

# Row data
Metabolomics_rowfeature <- plasma_metabolomics_global_rowfeature
print(all(colnames(Metabolomics_counts) == rownames(Metabolomics_rowfeature)))

Metabolomics_rowfeature = subset(Metabolomics_rowfeature, Metabolomics_rowfeature$TYPE == "NAMED")
Metabolomics_counts = Metabolomics_counts[,rownames(Metabolomics_rowfeature)]

print(all(colnames(Metabolomics_counts) == rownames(Metabolomics_rowfeature)))
# metadata
Metabolomics_metadata <- plasma_metabolomics_global_metadata
Metabolomics_metadata$sample_id = rownames(Metabolomics_metadata)
print(all(rownames(Metabolomics_counts) == rownames(Metabolomics_metadata)))
```

```{r}
clinical_all = clinical_data
Metabolomics_metadata = subset(Metabolomics_metadata, Metabolomics_metadata$sample_id %in% clinical_all$sample_id)

clinical_df <- clinical_all
clinical_df = subset(clinical_df, ! is.na(clinical_df$trajectory_group))
metadata = subset(Metabolomics_metadata, Metabolomics_metadata$sample_id %in% clinical_df$sample_id)
clinical_df_met = subset(clinical_df, clinical_df$sample_id %in% metadata$sample_id)
participants = as.data.frame(unique(clinical_df_met$participant_id))
plasma_metabolomics_global_counts = Metabolomics_counts[metadata$sample_id,]
plasma_metabolomics_global_metadata = metadata
rownames(plasma_metabolomics_global_metadata) = plasma_metabolomics_global_metadata$sample_id
```


```{r}
clinical_data = clinical_df_met[!(clinical_df_met$sample_id %in% "0865-0017MC00-004"),]

clinical_data$adm_date_cat <- ifelse(clinical_data$event_date >= 0 & clinical_data$event_date <= 2, "Admission",
              ifelse(clinical_data$event_date >2 & clinical_data$event_date <= 8, "WK1",
              ifelse(clinical_data$event_date >8 & clinical_data$event_date <= 23, "WK2_3",
              ifelse(clinical_data$event_date >23 & clinical_data$event_date <= 35, "WK4",
              ifelse(clinical_data$event_date >35 & clinical_data$event_date <= 90, "3MO",
              ifelse(clinical_data$event_date >90 & clinical_data$event_date <= 200, "6MO",
              ifelse(clinical_data$event_date >200 & clinical_data$event_date <= 300, "9MO",
              ifelse(clinical_data$event_date >300 , "12MO", "Out"))))))))

table(clinical_data$adm_date_cat)
length(unique(clinical_data$participant_id))

```
```{r}
clinical_data = clinical_data[!(clinical_data$sample_id %in% c("0865-0017MC00-004","0865-00CW0A00-004","0865-00DZYB00-004","0865-00956A00-005")),]
visit1 = subset(clinical_data, clinical_data$event_type == "Visit 1")
range(visit1$event_date)
length(unique(visit1$sample_id))
visit1_in = subset(visit1, visit1$event_date >= 0 & visit1$event_date <= 3)
length(unique(visit1_in$sample_id))

visit2 = subset(clinical_data, clinical_data$event_type == "Visit 2")
range(visit2$event_date)
length(unique(visit2$sample_id))
visit2_in = subset(visit2, visit2$event_date > 3 & visit2$event_date <= 5)
length(unique(visit2_in$sample_id))

visit3 = subset(clinical_data, clinical_data$event_type == "Visit 3")
range(visit3$event_date)
length(unique(visit3$sample_id))
visit3_in = subset(visit3, visit3$event_date >= 6 & visit3$event_date <= 8)
length(unique(visit3_in$sample_id))


visit4 = subset(clinical_data, clinical_data$event_type == "Visit 4")
range(visit4$event_date)
length(unique(visit4$sample_id))
visit4_in = subset(visit4, visit4$event_date >= 12 & visit4$event_date <= 16)
length(unique(visit4_in$sample_id))

visit5 = subset(clinical_data, clinical_data$event_type == "Visit 5")
range(visit5$event_date)
length(unique(visit5$sample_id))
visit5_in = subset(visit5, visit5$event_date >= 19 & visit5$event_date <= 23)
length(unique(visit5_in$sample_id))


visit6 = subset(clinical_data, clinical_data$event_type == "Visit 6")
range(visit6$event_date)
length(unique(visit6$sample_id))
visit6_in = subset(visit6, visit6$event_date >= 26 & visit6$event_date <= 35)
length(unique(visit6_in$sample_id))

visit7 = subset(clinical_data, clinical_data$event_type == "Visit 7")
range(visit7$event_date)

length(unique(visit7$sample_id))
visit7_in = subset(visit7, visit7$event_date >= 76 & visit7$event_date <= 104)
length(unique(visit7_in$sample_id))


visit8 = subset(clinical_data, clinical_data$event_type == "Visit 8")
range(visit8$event_date)
length(unique(visit8$sample_id))
visit8_in = subset(visit8, visit8$event_date >= 166 & visit8$event_date <= 194)
length(unique(visit8_in$sample_id))

visit9 = subset(clinical_data, clinical_data$event_type == "Visit 9")
range(visit9$event_date)
length(unique(visit9$sample_id))
visit9_in = subset(visit9, visit9$event_date >= 256 & visit9$event_date <= 284)
length(unique(visit9_in$sample_id))

visit10 = subset(clinical_data, clinical_data$event_type == "Visit 10")
range(visit10$event_date)
length(unique(visit10$sample_id))
visit10_in = subset(visit10, visit10$event_date >= 346 & visit10$event_date <= 374)
length(unique(visit10_in$sample_id))

visit_all_in = rbind(visit1_in,visit2_in,visit3_in,visit4_in,visit5_in,visit6_in,visit7_in,visit8_in,visit9_in,visit10_in)

table(clinical_data$event_type)
```
```{r}
clinical_data$event_type = factor(clinical_data$event_type, levels = c("Visit 1","Visit 2","Visit 3","Visit 4","Visit 5","Visit 6","Visit 7","Visit 8","Visit 9","Visit 10"))
clinical_data = clinical_data %>% mutate(Type = case_when(
  sample_id %in% visit_all_in$sample_id ~ "Within",
  !sample_id %in% visit_all_in$sample_id ~ "Outlier",
))
bar_figure = ggplot(clinical_data, aes(event_type,fill = Type)) + geom_bar() + theme_bw() + ggtitle("Number of Samples") +scale_fill_discrete(drop = F) #+ geom_text(aes(Var1,Freq,label = Freq,fill = NULL), data = totals_H, vjust=-0.1)
print(bar_figure)
```
#PCA
```{r}
#all samples
pca_data<-plasma_metabolomics_global_counts
pca_data = as.data.frame(pca_data)
#pca_data$sample_id = str_replace_all(pca_data$sample_id, "-", "_")

pca_data[,] <- sapply(pca_data[, ], as.numeric)
pca_met<-prcomp (na.omit(pca_data, center = TRUE, scale = TRUE))
fit<-summary(pca_met)
pca<-as.data.frame(fit$x)
pca<-as.data.frame(pca[c(1:10)])
library(tidyverse)
pca <- pca %>% as.data.frame() %>% rownames_to_column("sample_id")

pca_table <- inner_join(clinical_data, pca,  by="sample_id")
 
  pca_p <- pca_table %>%
      ggplot(aes(PC1, PC2, colour = sex)) +

      geom_point(size = 3) +
      theme(legend.position="bottom") + theme_bw() +
      labs(x     = paste0("1st dimension (",
                          round((pca_met$sdev^2/sum(pca_met$sdev^2))[1] * 100),
                          "%)"),
           y     = paste0("2nd dimension (",
                          round((pca_met$sdev^2/sum(pca_met$sdev^2))[2] * 100),
                          "%)")) 
  pca_p
```
```{r}
#V1 samples
clinical_data_V1 = subset(clinical_data, clinical_data$event_type == "Visit 1")
pca_data<-plasma_metabolomics_global_counts[clinical_data_V1$sample_id,]
pca_data = as.data.frame(pca_data)
#pca_data$sample_id = str_replace_all(pca_data$sample_id, "-", "_")

pca_data[,] <- sapply(pca_data[, ], as.numeric)
pca_met<-prcomp (na.omit(pca_data, center = TRUE, scale = TRUE))
fit<-summary(pca_met)
pca<-as.data.frame(fit$x)
pca<-as.data.frame(pca[c(1:10)])
library(tidyverse)
pca <- pca %>% as.data.frame() %>% rownames_to_column("sample_id")

pca_table <- inner_join(clinical_data_V1, pca,  by="sample_id")
 
  pca_p_V1 <- pca_table %>%
      ggplot(aes(PC1, PC2, colour = sex)) +

      geom_point(size = 3) +
      theme(legend.position="bottom") + theme_bw() +
      labs(x     = paste0("1st dimension (",
                          round((pca_met$sdev^2/sum(pca_met$sdev^2))[1] * 100),
                          "%)"),
           y     = paste0("2nd dimension (",
                          round((pca_met$sdev^2/sum(pca_met$sdev^2))[2] * 100),
                          "%)")) 
  pca_p_V1
```
#SPLSDA
```{r}

library(mixOmics) # import the mixOmics library
run_SPLSDA = function(clinical_data,Metabolomics_counts_conv,plasma_metabolomics_global_rowfeature,visit_type){
  clinical_data = clinical_data
  Metabolomics_counts_conv = Metabolomics_counts_conv
  plasma_metabolomics_global_rowfeature = plasma_metabolomics_global_rowfeature
clinical_temp = subset(clinical_data, clinical_data$sample_id %in% rownames(Metabolomics_counts_conv))
clinical_temp_16_V = subset(clinical_temp, clinical_temp$event_type == visit_type)
plasma_metabolomics_global_counts_temp = Metabolomics_counts_conv[clinical_temp_16_V$sample_id,]



#X <- srbct$gene # use the gene expression data as the X matrix
X = plasma_metabolomics_global_counts_temp
plasma_metabolomics_global_rowfeature = plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_temp),]
colnames(X) = plasma_metabolomics_global_rowfeature$CHEMICAL_NAME
#Y <- srbct$class # use the class data as the Y matrix
rownames(clinical_temp) = clinical_temp$sample_id

Y = clinical_temp_16_V$sex

splsda.result <- splsda(X, Y) # run the method #, keepX = c(50,30)


background.mahal <-  background.predict(splsda.result,
                                        comp.predicted = 2,
                                        dist = 'mahalanobis.dist')

# plot the sample plot
figure = plotIndiv(splsda.result, group = Y,ellipse=TRUE, ind.names = FALSE,
          legend = TRUE, legend.title = 'Sex',
         
          background = background.mahal,title = visit_type)
a <- plotLoadings(splsda.result,
       comp = 1,
       ndisplay=50,
       contrib = "max",
       #legend.color = cols,
       legend.title = "Convalescent groups",
       legend = TRUE,
       size.title = rel(1),
       size.name = 0.7, title="Top 50 Metabolite Loadings"
       )
return(list(p1=figure,p2=a))
}


```

```{ r}
SPLSDA_V1 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 1")
SPLSDA_V2 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 2")
SPLSDA_V3 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 3")
SPLSDA_V4 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 4")
SPLSDA_V5 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 5")
SPLSDA_V6 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 6")
SPLSDA_V7 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 7")
SPLSDA_V8 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 8")
SPLSDA_V9 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 9")
SPLSDA_V10 = run_SPLSDA(clinical_data,plasma_metabolomics_global_counts,plasma_metabolomics_global_rowfeature,"Visit 10")

```

```{ r}
#all samples
  clinical_data = clinical_data
  Metabolomics_counts_conv = plasma_metabolomics_global_counts
  plasma_metabolomics_global_rowfeature = plasma_metabolomics_global_rowfeature
clinical_temp = subset(clinical_data, clinical_data$sample_id %in% rownames(Metabolomics_counts_conv))
clinical_temp_16_V = clinical_temp
plasma_metabolomics_global_counts_temp = Metabolomics_counts_conv[clinical_temp_16_V$sample_id,]



#X <- srbct$gene # use the gene expression data as the X matrix
X = plasma_metabolomics_global_counts_temp
plasma_metabolomics_global_rowfeature = plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_temp),]
colnames(X) = plasma_metabolomics_global_rowfeature$CHEMICAL_NAME
#Y <- srbct$class # use the class data as the Y matrix
rownames(clinical_temp) = clinical_temp$sample_id

Y = clinical_temp_16_V$sex

splsda.result <- splsda(X, Y) # run the method #, keepX = c(50,30)


background.mahal <-  background.predict(splsda.result,
                                        comp.predicted = 2,
                                        dist = 'mahalanobis.dist')

# plot the sample plot
figure = plotIndiv(splsda.result, group = Y,ellipse=TRUE, ind.names = FALSE,
          legend = TRUE, legend.title = 'Sex',
         
          background = background.mahal,title = "All")
a <- plotLoadings(splsda.result,
       comp = 1,
       ndisplay=50,
       contrib = "max",
       #legend.color = cols,
       legend.title = "Convalescent groups",
       legend = TRUE,
       size.title = rel(1),
       size.name = 0.7, title="Top 50 Metabolite Loadings"
       )
```

# Logistic regression baseline 1-4 vs. 5
```{r}
library("geepack")
library("data.table")
clinical_df = clinical_data
meta_data_raw <- plasma_metabolomics_global_metadata
meta_data_raw$sample_id = rownames(meta_data_raw)
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"))#,"phase"
met_data = subset(met_data, met_data$"trajectory_group" %in% c(1,2,3,4,5))
met_data_baseline = subset(met_data, met_data$event_type == "Visit 1")
met_data_baseline <-  met_data_baseline[order(met_data_baseline$participant_id),]
met_data_baseline <- met_data_baseline %>%  mutate(OUTCOME = case_when(
    trajectory_group %in% c(1,2,3,4) ~ 0,
    trajectory_group == 5  ~ 1
    ))

met_data_baseline <- met_data_baseline %>% drop_na(OUTCOME)
OUTCOME = met_data_baseline$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data_baseline$sex)
y <- as.numeric (OUTCOME)
age <- as.numeric(met_data_baseline$admit_age)
bmi <- as.numeric(met_data_baseline$bmi) #numerical
race <- as.factor(met_data_baseline$race)
ethnicity <- as.factor(met_data_baseline$ethnicity)
location <- as.factor(met_data_baseline$event_location)
participant <- as.factor(met_data_baseline$participant_id)
age_quantile <- as.factor(met_data_baseline$discretized_admit_age_quantile)
phase <- as.factor(met_data_baseline$phase.x)
plate <- as.factor(met_data_baseline$plate)
type = as.factor(met_data_baseline$Type)
#y <- as.factor (y)
plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data_baseline$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met,age, sex, participant, bmi, location,age_quantile,phase, plate,type)
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile","phase","plate","type") #"location",
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
     lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  
  
  lr_output$coef <- coef
  lr_output$std <- std
 
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AD_baseline <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AD_baseline$qval <= 0.05)
  
  output_all_AD = output_all
```
# Logistic regression baseline 4 vs. 5
```{ r}
library("geepack")
clinical_df = clinical_data
meta_data_raw <- plasma_metabolomics_global_metadata
meta_data_raw$sample_id = rownames(meta_data_raw)
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"))#,"phase"
met_data = subset(met_data, met_data$"trajectory_group" %in% c(4,5))
met_data_baseline = subset(met_data, met_data$event_type == "Visit 1")
met_data_baseline <-  met_data_baseline[order(met_data_baseline$participant_id),]
met_data_baseline <- met_data_baseline %>%  mutate(OUTCOME = case_when(
    trajectory_group == 4 ~ 0,
    trajectory_group == 5  ~ 1
    ))


met_data_baseline <- met_data_baseline %>% drop_na(OUTCOME)
OUTCOME = met_data_baseline$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data_baseline$sex)
y <- as.numeric (OUTCOME)
age <- as.numeric(met_data_baseline$admit_age)
bmi <- as.numeric(met_data_baseline$bmi) #numerical
race <- as.factor(met_data_baseline$race)
ethnicity <- as.factor(met_data_baseline$ethnicity)
location <- as.factor(met_data_baseline$event_location)
participant <- as.factor(met_data_baseline$participant_id)
age_quantile <- as.factor(met_data_baseline$discretized_admit_age_quantile)
phase <- as.factor(met_data_baseline$phase.x)
plate <- as.factor(met_data_baseline$plate)
type = as.factor(met_data_baseline$Type)
#y <- as.factor (y)
plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data_baseline$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met,age, sex, participant, bmi, location,age_quantile,phase, plate,type)
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile","phase","plate","type") #"location",
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  
  
  lr_output$coef <- coef
  lr_output$std <- std
 
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
 lr_output_45_baseline <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_45_baseline$qval <= 0.05)
  output_all_45 = output_all
```

# Logistic regression baseline 1-3 vs. 4
```{ r}
library("geepack")
clinical_df = clinical_data
meta_data_raw <- plasma_metabolomics_global_metadata
meta_data_raw$sample_id = rownames(meta_data_raw)
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"))#,"phase"
met_data = subset(met_data, met_data$"trajectory_group" %in% c(1,2,3,4))
met_data_baseline = subset(met_data, met_data$event_type == "Visit 1")
met_data_baseline <-  met_data_baseline[order(met_data_baseline$participant_id),]
met_data_baseline <- met_data_baseline %>%  mutate(OUTCOME = case_when(
    trajectory_group %in% c(1,2,3) ~ 0,
    trajectory_group == 4  ~ 1
    ))

met_data_baseline <- met_data_baseline %>% drop_na(OUTCOME)
OUTCOME = met_data_baseline$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data_baseline$sex)
y <- as.numeric (OUTCOME)
age <- as.numeric(met_data_baseline$admit_age)
bmi <- as.numeric(met_data_baseline$bmi) #numerical
race <- as.factor(met_data_baseline$race)
ethnicity <- as.factor(met_data_baseline$ethnicity)
location <- as.factor(met_data_baseline$event_location)
participant <- as.factor(met_data_baseline$participant_id)
age_quantile <- as.factor(met_data_baseline$discretized_admit_age_quantile)
phase <- as.factor(met_data_baseline$phase.x)
plate <- as.factor(met_data_baseline$plate)
type = as.factor(met_data_baseline$Type)
#y <- as.factor (y)
plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data_baseline$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met,age, sex, participant, bmi, location,age_quantile,phase, plate,type)
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile","phase","plate","type") #"location",
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  
  
  lr_output$coef <- coef
  lr_output$std <- std
 
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AS_baseline <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AS_baseline$qval <= 0.05)
  output_all_AS = output_all
```
# Logistic regression baseline 1-3 vs. 4,5
```{r}
library("geepack")
clinical_df = clinical_data
meta_data_raw <- plasma_metabolomics_global_metadata
meta_data_raw$sample_id = rownames(meta_data_raw)
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"))#,"phase"
met_data = subset(met_data, met_data$"trajectory_group" %in% c(1,2,3,4,5))
met_data_baseline = subset(met_data, met_data$event_type == "Visit 1")
met_data_baseline <-  met_data_baseline[order(met_data_baseline$participant_id),]
met_data_baseline <- met_data_baseline %>%  mutate(OUTCOME = case_when(
    trajectory_group %in% c(1,2,3) ~ 0,
    trajectory_group %in% c(4,5)  ~ 1
    ))

met_data_baseline <- met_data_baseline %>% drop_na(OUTCOME)
OUTCOME = met_data_baseline$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data_baseline$sex)
y <- as.numeric (OUTCOME)
age <- as.numeric(met_data_baseline$admit_age)
bmi <- as.numeric(met_data_baseline$bmi) #numerical
race <- as.factor(met_data_baseline$race)
ethnicity <- as.factor(met_data_baseline$ethnicity)
location <- as.factor(met_data_baseline$event_location)
participant <- as.factor(met_data_baseline$participant_id)
age_quantile <- as.factor(met_data_baseline$discretized_admit_age_quantile)
phase <- as.factor(met_data_baseline$phase.x)
plate <- as.factor(met_data_baseline$plate)
type = as.factor(met_data_baseline$Type)
#y <- as.factor (y)
plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data_baseline$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met,age, sex, participant, bmi, location,age_quantile,phase, plate,type)
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile","phase","plate","type") #"location",
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
     lr<- glm (fmla, data=df,family=binomial) #
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  
  
  lr_output$coef <- coef
  lr_output$std <- std
 
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_123_baseline <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_123_baseline$qval <= 0.05)
  output_all_123 = output_all
```
# Logistic regression baseline 1-3 vs. 4
```{r}
library("geepack")
MHTFR = read_csv(file = "/scratch/metabolomics/plasma-metabolomics-targeted/MTHFR/MTHFR.csv") %>% mutate_if(is.character, str_trim)
MHTFR$allele = paste(MHTFR$`allele 1`, MHTFR$`allele 2`)
AA = subset(MHTFR, MHTFR$allele == "A A")
clinical_df = clinical_data
meta_data_raw <- plasma_metabolomics_global_metadata
meta_data_raw$sample_id = rownames(meta_data_raw)
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"))#,"phase"
met_data = subset(met_data, met_data$"trajectory_group" %in% c(4,5))
met_data_baseline = subset(met_data, met_data$event_type == "Visit 1")
met_data_baseline <-  met_data_baseline[order(met_data_baseline$participant_id),]
met_data_baseline <- met_data_baseline %>%  mutate(OUTCOME = case_when(
    !met_data_baseline$participant_id %in% AA$participant_id ~ 0,
    met_data_baseline$participant_id %in% AA$participant_id ~ 1
    ))

met_data_baseline <- met_data_baseline %>% drop_na(OUTCOME)
OUTCOME = met_data_baseline$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data_baseline$sex)
y <- as.numeric (OUTCOME)
age <- as.numeric(met_data_baseline$admit_age)
bmi <- as.numeric(met_data_baseline$bmi) #numerical
race <- as.factor(met_data_baseline$race)
ethnicity <- as.factor(met_data_baseline$ethnicity)
location <- as.factor(met_data_baseline$event_location)
participant <- as.factor(met_data_baseline$participant_id)
age_quantile <- as.factor(met_data_baseline$discretized_admit_age_quantile)
phase <- as.factor(met_data_baseline$phase.x)
plate <- as.factor(met_data_baseline$plate)
type = as.factor(met_data_baseline$Type)
#y <- as.factor (y)
plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data_baseline$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met,age, sex, participant, bmi, location,age_quantile,phase, plate,type)
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile","phase","plate","type") #"location",
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  
  
  lr_output$coef <- coef
  lr_output$std <- std
 
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AA_baseline <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AA_baseline$qval <= 0.05)
  output_all_AA = output_all
```
# Logistic regression baseline 45 dead vs. no dead
```{r}
library("geepack")
clinical_df = clinical_data
meta_data_raw <- plasma_metabolomics_global_metadata
meta_data_raw$sample_id = rownames(meta_data_raw)
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"))#,"phase"
met_data = subset(met_data, met_data$"trajectory_group" %in% c(4,5))
met_data_baseline = subset(met_data, met_data$event_type == "Visit 1")
met_data_baseline <-  met_data_baseline[order(met_data_baseline$participant_id),]
met_data_baseline <- met_data_baseline %>%  mutate(OUTCOME = case_when(
    diedever == 0 ~ 0,
    diedever == 1 ~ 1
    ))

met_data_baseline <- met_data_baseline %>% drop_na(OUTCOME)
OUTCOME = met_data_baseline$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data_baseline$sex)
y <- as.numeric (OUTCOME)
age <- as.numeric(met_data_baseline$admit_age)
bmi <- as.numeric(met_data_baseline$bmi) #numerical
race <- as.factor(met_data_baseline$race)
ethnicity <- as.factor(met_data_baseline$ethnicity)
location <- as.factor(met_data_baseline$event_location)
participant <- as.factor(met_data_baseline$participant_id)
age_quantile <- as.factor(met_data_baseline$discretized_admit_age_quantile)
phase <- as.factor(met_data_baseline$phase.x)
plate <- as.factor(met_data_baseline$plate)
type = as.factor(met_data_baseline$Type)
#y <- as.factor (y)
plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data_baseline$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met,age, sex, participant, bmi, location,age_quantile,phase, plate,type)
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile","phase","plate","type") #"location",
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  
  
  lr_output$coef <- coef
  lr_output$std <- std
 
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_diedever_baseline <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_diedever_baseline$qval <= 0.05)
  output_all_diedever = output_all
```
```{r}
save(lr_output_diedever_baseline,lr_output_AD_baseline,lr_output_123_baseline, file = "/scratch/metabolomics/plasma-metabolomics-global/scripts/deep_dive_met/deep_dive_TG.Rdata")
```
```{ r}
load("/scratch/metabolomics/plasma-metabolomics-global/scripts/deep_dive_met/deep_dive_TG.Rdata")
```

```{r}
library(ggrepel)
volPlots2 <- function(df, titleArg = "Volcano Plot for omics features"){
  
  df <- df %>% as.data.frame() %>% dplyr::mutate(Direction = case_when(coef < 0 & qval < 0.05 ~ "Down", coef > 0 & qval < 0.05 ~ "Up", qval > 0.05 ~ "NotSig")) %>% dplyr::mutate(Direction = as.factor(Direction)) %>% dplyr::mutate(Direction = factor(Direction, levels = c("NotSig", "Down", "Up")))
   label_data <- df %>% dplyr::filter(qval < 0.05)
   my_subtitle <- glue::glue(
      #"{nrow(filter(df, Direction != 'NotSig'))} DE metabolites: ",
      "{nrow(filter(df, Direction == 'Up'))} Up, ",
      "{nrow(filter(df, Direction == 'Down'))} Down")
    ggplot(data = df)  +
    geom_point(aes(x = coef, y = -log10(pval), color = Direction), size = 3, pch = 16, alpha = 1) +
    #geom_label_repel(aes(x = coef, y = -log10(pval), label = Name), max.overlaps = 20, label.size = 0.05, label.padding = 0) +
    theme_classic() +
    scale_color_manual(values = c("NotSig" = "#C0C0C0", "Down"="darkblue","Up"="darkred" )) +
   #ylim(0, 16) +
    #xlim(-1.5, 1.5) + 
    geom_hline(yintercept = 1.30, linetype = "dashed", alpha = 0.8) +
    labs(title = titleArg,subtitle = my_subtitle, x = "log2FC", y = "-log10(Pvalue)") +
      theme(text = element_text(face = "bold"))
}
lr_output_123_baseline = lr_output_123_baseline[order(lr_output_123_baseline$qval),]
lr_output_123_baseline$name2 = ''
lr_output_123_baseline$name2[1:3] = lr_output_123_baseline$Name[1:3]
Volcano_metabolomics_123 <- volPlots2(lr_output_123_baseline, titleArg = "Severe(TG4_5) vs. Non-severe(TG1_3)") #+geom_label_repel(aes(x = coef, y = -log10(pval)), max.overlaps = 20, label.size = 0.05, label.padding = 0.1) + ylim

lr_output_diedever_baseline = lr_output_diedever_baseline[order(lr_output_diedever_baseline$qval),]
lr_output_diedever_baseline$name2 = ''
lr_output_diedever_baseline$name2[1:3] = lr_output_diedever_baseline$Name[1:3]
Volcano_metabolomics_diedever <- volPlots2(lr_output_diedever_baseline, titleArg = "Non-survivors(TG4 deaths + TG5) vs. Survivors(TG4 recovered)") #+geom_label_repel(aes(x = coef, y = -log10(pval)), max.overlaps = 20, label.size = 0.05, label.padding = 0.1)

lr_output_AD_baseline = lr_output_AD_baseline[order(lr_output_AD_baseline$qval),]
lr_output_AD_baseline$name2 = ''
lr_output_AD_baseline$name2[1:3] = lr_output_AD_baseline$Name[1:3]
Volcano_metabolomics_AD <- volPlots2(lr_output_AD_baseline, titleArg = "Fatal(TG5) vs. Non-fatal(TG1_4)") #+geom_label_repel(aes(x = coef, y = -log10(pval)), max.overlaps = 20, label.size = 0.05, label.padding = 0.1)

a = ggarrange(Volcano_metabolomics_AD,Volcano_metabolomics_123, Volcano_metabolomics_diedever, ncol = 1, common.legend = T, legend.position = "bottom")
a

```

```{ r}

x <- list(
  `Severe vs. Non-severe` = lr_output_AS_baseline[lr_output_AS_baseline$qval<= 0.05,]$Name, 
  `Fatal vs. Severe` =lr_output_45_baseline[lr_output_45_baseline$qval<= 0.05,]$Name, 
  `Fatal vs. Non-fatal` =lr_output_AD_baseline[lr_output_AD_baseline$qval<= 0.05,]$Name
  )
library(ggvenn)
ggvenn(
  x, 
  fill_color = c("#0000FF","#FF0000", "#a45ee9"),
  stroke_size = 0.5, set_name_size = 4
  )
```
```{ r}
library(data.table)
plasma_metabolomics_global_rowfeature$Compound_ID = rownames(plasma_metabolomics_global_rowfeature)
lr_output_cat1_all_glm_simple_sig = subset(lr_output_AD_baseline,lr_output_AD_baseline$qval <= 0.05)
lr_output_cat1_all_glm_simple_sig = lr_output_cat1_all_glm_simple_sig[,c("Compound_ID","coef","std","pval","qval")]
lr_output_cat1_all_glm_simple_sig_sub = plasma_metabolomics_global_rowfeature[lr_output_cat1_all_glm_simple_sig$Compound_ID,]
lr_output_cat1_all_glm_simple= lr_output_AD_baseline[,c("Compound_ID","coef","std","pval","qval")]
lr_output_cat2_all_glm_simple_sig = subset(lr_output_AS_baseline,lr_output_AS_baseline$qval <= 0.05)
lr_output_cat2_all_glm_simple_sig = lr_output_cat2_all_glm_simple_sig[,c("Compound_ID","coef","std","pval","qval")]
lr_output_cat2_all_glm_simple= lr_output_AS_baseline[,c("Compound_ID","coef","std","pval","qval")]
lr_output_cat3_all_glm_simple_sig = subset(lr_output_45_baseline,lr_output_45_baseline$qval <= 0.05)
lr_output_cat3_all_glm_simple_sig = lr_output_cat3_all_glm_simple_sig[,c("Compound_ID","coef","std","pval","qval")]
lr_output_cat3_all_glm_simple= lr_output_45_baseline[,c("Compound_ID","coef","std","pval","qval")]
lr_output_cat1_all_glm_simple_sig_sub = subset(plasma_metabolomics_global_rowfeature,plasma_metabolomics_global_rowfeature$Compound_ID %in% lr_output_cat1_all_glm_simple_sig$Compound_ID)
lr_output_cat2_all_glm_simple_sig_sub = subset(plasma_metabolomics_global_rowfeature,plasma_metabolomics_global_rowfeature$Compound_ID %in% lr_output_cat2_all_glm_simple_sig$Compound_ID)
lr_output_cat3_all_glm_simple_sig_sub = subset(plasma_metabolomics_global_rowfeature,plasma_metabolomics_global_rowfeature$Compound_ID %in% lr_output_cat3_all_glm_simple_sig$Compound_ID)

Sub_pathway_table2 = intersect(lr_output_cat1_all_glm_simple_sig_sub$SUB_PATHWAY,intersect(lr_output_cat2_all_glm_simple_sig_sub$SUB_PATHWAY, lr_output_cat3_all_glm_simple_sig_sub$SUB_PATHWAY))
Sub_pathway_table2 = unique(Sub_pathway_table2)

plasma_metabolomics_global_rowfeature_temp = subset(plasma_metabolomics_global_rowfeature, plasma_metabolomics_global_rowfeature$SUB_PATHWAY %in% Sub_pathway_table2)
figure_table = function(lr_table,plasma_metabolomics_global_rowfeature_temp){
  tmp <- lr_table
tmp = merge(tmp, plasma_metabolomics_global_rowfeature_temp, by = "Compound_ID", all.y = T)
tmp <- tmp[!is.na(tmp$SUPER_PATHWAY),]
tmp2 <- tmp[tmp$pval < 0.05, ]
tmp3 <- tmp[tmp$qval < 0.05, ]
dim(tmp3)
tmp3 <- tmp[tmp$pval < 0.05 / 578, ]
dim(tmp3)

pathways <- unique(tmp$SUB_PATHWAY)

# get a count of the total mets in each pathway 
df <- data.table(pathways) # list of pathways
df$total_pos <- 0 # create empty rows
df$total_neg <- 0 # create empty rows
for(i in 1:nrow(df)){
  for(j in 1:nrow(tmp)){
    if(df$pathways[i] == tmp$SUB_PATHWAY[j]){
      if(tmp$coef[j] > 0){
        df$total_pos[i] <- df$total_pos[i] + 1 
      }
      else{
         df$total_neg[i] <- df$total_neg[i] - 1
      }
     
    }
  }
}
# get a count of sig mets in each pathway at P<0.05
df$sig_neg <- 0
df$sig_pos <- 0
if(dim(tmp2)[1] > 0){
for(i in 1:nrow(df)){
  for(j in 1:nrow(tmp2)){
    if(df$pathways[i] == tmp2$SUB_PATHWAY[j]){
      ifelse(tmp2$coef[j] < 0, 
             df$sig_neg[i] <- df$sig_neg[i] - 1,
             df$sig_pos[i] <- df$sig_pos[i] + 1)
    }
  }
}
}
# get a count of sig mets in each pathway at FDR<0.05
df$sig_neg_fdr <- 0
df$sig_pos_fdr <- 0
if(dim(tmp3)[1] > 0){
  for(i in 1:nrow(df)){
  for(j in 1:nrow(tmp3)){
    if(df$pathways[i] == tmp3$SUB_PATHWAY[j]){
      ifelse(tmp3$coef[j] < 0, 
             df$sig_neg_fdr[i] <- df$sig_neg_fdr[i] - 1,
             df$sig_pos_fdr[i] <- df$sig_pos_fdr[i] + 1)
    }
  }
}
}


# add in super pathways for coloring
mets_info_all = plasma_metabolomics_global_rowfeature_temp
mets_info_tmp <- mets_info_all[, c("SUPER_PATHWAY", "SUB_PATHWAY")]
colnames(mets_info_tmp) <- c("SUPER_PATHWAY", "SUB_PATHWAY")
df <- merge(mets_info_tmp, df, by.x = "SUB_PATHWAY", by.y = "pathways", all.y = T)

# sort column and give idx
df <- df[order(df$SUPER_PATHWAY, df$SUB_PATHWAY),] 

df$idx <- as.factor(1:nrow(df))

# create long df for plotting
long <- melt(setDT(df), id.vars = c("SUB_PATHWAY","SUPER_PATHWAY", "idx"), variable.name = "key")
long$SUB_PATHWAY <- as.factor(long$SUB_PATHWAY)
long$color <- ifelse(long$SUPER_PATHWAY == "Amino Acid", "#FF689F",
              ifelse(long$SUPER_PATHWAY == "Carbohydrate", "#E76BF3",
              ifelse(long$SUPER_PATHWAY == "Cofactors and Vitamins", "#00A5FF",
              ifelse(long$SUPER_PATHWAY == "Lipid", "#00B81F",
              ifelse(long$SUPER_PATHWAY == "Nucleotide", "#BB9D00",
              ifelse(long$SUPER_PATHWAY == "Peptide", "#F8766D",
              ifelse(long$SUPER_PATHWAY == "Partially Characterized Molecules", "#E08B00",
              ifelse(long$SUPER_PATHWAY == "Xenobiotics", "firebrick2",
              ifelse(long$SUPER_PATHWAY == "Energy", "chocolate2", "pink")))))))))

long$name <- paste0(long$SUPER_PATHWAY, " | ", long$SUB_PATHWAY)

long2 <- long
long2 <- long[!duplicated(long$name), ]
rev_names =  rev(unique(long$name))
long$name = factor(long$name, levels = rev_names)
return(list(a = long, b= long2))
}
Callcat1 = figure_table(lr_output_cat1_all_glm_simple,plasma_metabolomics_global_rowfeature_temp)
Callcat2 = figure_table(lr_output_cat2_all_glm_simple,plasma_metabolomics_global_rowfeature_temp)
Callcat3 = figure_table(lr_output_cat3_all_glm_simple,plasma_metabolomics_global_rowfeature_temp)



p_cat1 = ggplot(Callcat1$a, aes(fill=key, y=name, x=value)) + 
  geom_bar(position="identity", stat="identity") +
  ylab('Metabolite Sub-Pathway') + xlab('# of metabolites') +
  labs(title = 'Fatal vs. Non-fatal') +
  scale_fill_manual("Legend", 
                    values = c("total_neg" = "gray83", 
                               "total_pos" = "gray83", 
                               "sig_neg" = "lightskyblue1", 
                               "sig_pos" = "lightcoral",
                               "sig_neg_fdr" = "deepskyblue4",
                               "sig_pos_fdr" = "firebrick4")) +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.y = element_text(angle = 0, size = 8, vjust = 0.5,
                                   color = rev(Callcat1$b$color)
                                   )) +geom_vline(xintercept = 0)
p_cat2 = ggplot(Callcat2$a, aes(fill=key, y=name, x=value)) + 
  geom_bar(position="identity", stat="identity") +
  ylab('Metabolite Sub-Pathway') + xlab('# of metabolites') +
  labs(title = 'Severe vs. Non-severe') +
  scale_fill_manual("Legend", 
                    values = c("total_neg" = "gray83", 
                               "total_pos" = "gray83", 
                               "sig_neg" = "lightskyblue1", 
                               "sig_pos" = "lightcoral",
                               "sig_neg_fdr" = "deepskyblue4",
                               "sig_pos_fdr" = "firebrick4")) +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.y = element_text(angle = 0, size = 8, vjust = 0.5,
                                   color = rev(Callcat2$b$color)
                                   )) +geom_vline(xintercept = 0)
p_cat3 = ggplot(Callcat3$a, aes(fill=key, y=name, x=value)) + 
  geom_bar(position="identity", stat="identity") +
  ylab('Metabolite Sub-Pathway') + xlab('# of metabolites') +
  labs(title = 'Fatal vs. Severe') +
  scale_fill_manual("Legend", 
                    values = c("total_neg" = "gray83", 
                               "total_pos" = "gray83", 
                               "sig_neg" = "lightskyblue1", 
                               "sig_pos" = "lightcoral",
                               "sig_neg_fdr" = "deepskyblue4",
                               "sig_pos_fdr" = "firebrick4")) +
  theme(panel.border = element_rect(colour = "black", fill=NA, size=1),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.text.y = element_text(angle = 0, size = 8, vjust = 0.5,
                                   color = rev(Callcat3$b$color)
                                   )) +geom_vline(xintercept = 0)


remove_y <- theme(
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  axis.title.y = element_blank()
  
)
p <- list(
  p_cat1 + 
  theme(legend.position="none")+ xlim(-30,30),# + remove_y
  p_cat2 + remove_y + xlim(-30,20)+
  theme(legend.position="none"),
  p_cat3+ remove_y + xlim(-30,30)+
      theme(legend.position="none")
 
 
)
library(patchwork)
p_TGcatall <- wrap_plots(p, nrow = 1) + plot_layout(guides = "collect")& theme(legend.position = 'bottom')
p_TGcatall

save(lr_output_cat1_all_glm_simple_sig, lr_output_cat2_all_glm_simple_sig, lr_output_cat3_all_glm_simple_sig, file = "Figure1_data.Rdata")
```
```{r}
bubble_plot <- function(significant_table, metabolomics_rf, alpha, begin, end){
  #
  #significant_table <- linear_geeglm_results_sig
  #metabolomics_rf <- plasma_metabolomics_global_rowfeature
  #alpha = 1
 # begin = 0
  #end = 1
  #
  Sub_size <- c()
  Sub_pathway <- c()
  if(is.factor(metabolomics_rf$SUB_PATHWAY)){
    metabolomics_rf$SUB_PATHWAY <- unfactor(metabolomics_rf$SUB_PATHWAY)
  }
  subpath_list <- unique(metabolomics_rf$SUB_PATHWAY)
  
  for(i in subpath_list){
    if (!is.na(i)){
      sub_table <- metabolomics_rf %>% filter(SUB_PATHWAY == i)
      
      sub <- unique(sub_table$SUB_PATHWAY)
      Sub_pathway <- c(Sub_pathway, sub)
      
      size_sub <- length(sub_table$SUB_PATHWAY)
      Sub_size <- c(Sub_size, size_sub)
    }
    
    
    
  }
  Sub_pathway_table <- data.table(Sub_pathway, Sub_size)
  results_sig <- significant_table
  results_sig_more <- merge(results_sig, Sub_pathway_table, by.x= "SUB_PATHWAY", by.y ="Sub_pathway")
  results_sig_more$SUPER_PATHWAY <- as.character(results_sig_more$SUPER_PATHWAY)
  pathwayfile <- results_sig_more
  pathwayfile$qval[pathwayfile$qval == 0] <- 0.00000000000000000001
  pathwayfile$`-log10FDR` <- -log10(pathwayfile$qval)
  subpath_list <- unique(pathwayfile$SUB_PATHWAY)
  
  Super_pathway <- c()
  Sub_pathway <- c()
  mean_neglog10FDR <- c()
  Size <- c()
  Impact_ratio <- c()
  Increase_ratio <- c()
  Size_all <- c()
  ###### impact ratio: # of significant / # of total in the subpathway 0-1
  for(i in subpath_list){
    sub_table <- pathwayfile %>% filter(SUB_PATHWAY == i)
    
    supp <- unique(sub_table$SUPER_PATHWAY)
    Super_pathway <- c(Super_pathway, supp)
    
    sub <- unique(sub_table$SUB_PATHWAY)
    Sub_pathway <- c(Sub_pathway, sub)
    
    mean_qval <- mean(sub_table$`-log10FDR`)
    mean_neglog10FDR <- c(mean_neglog10FDR, mean_qval)
    
    size_sub <- length(sub_table$SUB_PATHWAY)
    Size <- c(Size, size_sub)
    size_all <- unique(sub_table$Sub_size)
    Size_all <- c(Size_all, size_all)
    im_ratio <- size_sub/size_all
    Impact_ratio <- c(Impact_ratio, im_ratio)
    
    increased_table <- sub_table %>% filter(coef >= 0) 
    increased_num <- length(increased_table$Name) / size_sub
    Increase_ratio <- c(Increase_ratio, increased_num)
    
  }
  path_output <- data.table(Super_pathway, Sub_pathway, mean_neglog10FDR, Impact_ratio,Size , Increase_ratio,Size_all)#
  path_output$Super_pathway <- factor(path_output$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"))
  #path_output <- path_output[order(Super_pathway, desc(mean_p)),]
  path_output <- path_output[order(Super_pathway),]
  
  path_output$Super_pathway2 <- path_output$Super_pathway
  path_output$Super_pathway2 <- factor(path_output$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"), ordered = TRUE)
  #path_output$Super_pathway = unfactor(path_output$Super_pathway)
  p <- path_output %>%
    arrange(dplyr::desc(Super_pathway),mean_neglog10FDR) %>% mutate(Sub_pathway=factor(Sub_pathway, levels=Sub_pathway)) %>%
    ggplot(aes(x= Sub_pathway, y =mean_neglog10FDR)) +
    scale_colour_viridis_c( alpha = alpha,
                            begin = begin,
                            end = end,
                            direction = 1)+
    geom_tile(aes(fill = Super_pathway2), height = Inf, alpha =0.7) +
    geom_point(aes(size = Impact_ratio, colour = Increase_ratio)) + #Size
    geom_point(aes(size = Impact_ratio),shape = 1,colour = "black") +  #Size
    scale_fill_manual(values = c("Amino Acid" = "#fed0ed",
                                 "Carbohydrate" = "#f4d5fe",
                                 "Cofactors and Vitamins" = "#d0e1ff",
                                 "Energy" = "#b3eaf6",
                                   "Lipid" = "#b3ece2", 
                                 "Nucleotide" = "#b3eac3",
                                 "Partially Characterized Molecules" = "#dee5b3",
                                   "Peptide" = "#f2deb3", "Xenobiotics" = "#Fdd6d3"))+
    theme_light() +xlab("Sub_pathway")+
    coord_flip() #+ scale_x_reverse()
path_output = path_output %>%
    arrange(dplyr::desc(Super_pathway),mean_neglog10FDR) %>% mutate(Sub_pathway=factor(Sub_pathway, levels=Sub_pathway))
  path_output<- path_output[seq(dim(path_output)[1],1),]
  return(list(figure = p, table = path_output)) 
}
```

# bubble plot
```{r}
plasma_metabolomics_global_rowfeature_tmp = plasma_metabolomics_global_rowfeature
lr_output_123_baseline_sig = subset(lr_output_123_baseline, lr_output_123_baseline$qval <= 0.05)
bubble_plot_123_baseline_figure = bubble_plot(lr_output_123_baseline_sig, plasma_metabolomics_global_rowfeature_tmp,alpha= 1, begin = 0, end = 1)$figure
bubble_plot_123_baseline_table = bubble_plot(lr_output_123_baseline_sig, plasma_metabolomics_global_rowfeature_tmp,alpha= 1, begin = 0, end = 1)$table


lr_output_diedever_baseline_sig = subset(lr_output_diedever_baseline, lr_output_diedever_baseline$qval <= 0.05)
bubble_plot_diedever_baseline_figure = bubble_plot(lr_output_diedever_baseline_sig, plasma_metabolomics_global_rowfeature_tmp,alpha= 1, begin = 0, end = 1)$figure
bubble_plot_diedever_baseline_table = bubble_plot(lr_output_diedever_baseline_sig, plasma_metabolomics_global_rowfeature_tmp,alpha= 1, begin = 0, end = 1)$table

lr_output_AD_baseline_sig = subset(lr_output_AD_baseline, lr_output_AD_baseline$qval <= 0.05)
bubble_plot_AD_baseline_figure = bubble_plot(lr_output_AD_baseline_sig, plasma_metabolomics_global_rowfeature_tmp,alpha= 1, begin = 0, end = 1)$figure
bubble_plot_AD_baseline_table = bubble_plot(lr_output_AD_baseline_sig, plasma_metabolomics_global_rowfeature_tmp,alpha= 1, begin = 0, end = 1)$table

```

```{r}
all_pathways =  rbind(bubble_plot_diedever_baseline_table[,1:2], bubble_plot_AD_baseline_table[,1:2], bubble_plot_123_baseline_table[,1:2])
all_pathways <- unique(all_pathways)
bubble_plot_diedever_baseline_table_all = merge(all_pathways, bubble_plot_diedever_baseline_table, all.x = TRUE, by=c("Super_pathway","Sub_pathway"))
bubble_plot_AD_baseline_table_all = merge(all_pathways, bubble_plot_AD_baseline_table, all.x = TRUE, by=c("Super_pathway","Sub_pathway"))
bubble_plot_123_baseline_table_all = merge(all_pathways, bubble_plot_123_baseline_table, all.x = TRUE, by=c("Super_pathway","Sub_pathway"))
bubble_plot_diedever_baseline_table_less = bubble_plot_diedever_baseline_table_all[,c("Sub_pathway","Impact_ratio")]
bubble_plot_AD_baseline_table_less = bubble_plot_AD_baseline_table_all[,c("Sub_pathway","Impact_ratio")]
bubble_plot_123_baseline_table_less = bubble_plot_123_baseline_table_all[,c("Sub_pathway","Impact_ratio")]
names(bubble_plot_diedever_baseline_table_less)[2] = "Non-SurvivorsvsSurvivors"
names(bubble_plot_AD_baseline_table_less)[2] = "TG1_4vsTG5"
names(bubble_plot_123_baseline_table_less)[2] = "TG1_3vsTG4_5"
enrich_all <- cbind(bubble_plot_diedever_baseline_table_less, bubble_plot_AD_baseline_table_less, bubble_plot_123_baseline_table_less)
enrich_all <- as.data.frame(enrich_all[,c( "Sub_pathway","Non-SurvivorsvsSurvivors","TG1_4vsTG5","TG1_3vsTG4_5")])
enrich_all = enrich_all %>% mutate(Sub_pathway2 = ifelse(!is.na(`Non-SurvivorsvsSurvivors`) & !is.na(TG1_4vsTG5) & !is.na(TG1_3vsTG4_5), paste0(Sub_pathway,"*"), paste0(Sub_pathway)))
rownames(enrich_all) <- enrich_all$Sub_pathway2
enrich_all$Sub_pathway <- NULL
enrich_all$Sub_pathway2 <- NULL
enrich_all[is.na(enrich_all)] <- 0
enrich_all <- t(enrich_all)
enrich_all <- as.matrix(enrich_all) 
```

```{r}

path_output1 = bubble_plot_AD_baseline_table
path_output2 = bubble_plot_123_baseline_table
path_output3 = bubble_plot_diedever_baseline_table

all_pathway = merge(path_output1, path_output2, by = c("Super_pathway","Sub_pathway"))
all_pathway = merge(all_pathway, path_output3, by = c("Super_pathway","Sub_pathway"))

path_output1_all =subset(path_output1, path_output1$Sub_pathway %in% all_pathway$Sub_pathway)
path_output2_all =subset(path_output2, path_output2$Sub_pathway %in% all_pathway$Sub_pathway)
path_output3_all =subset(path_output3, path_output3$Sub_pathway %in% all_pathway$Sub_pathway)

path_output1_all$type = "Fatal vs. Non-fatal"
path_output2_all$type = "Severe vs. Non-severe"
path_output3_all$type = "Non-servivors vs. Servivors"

path_output = rbind(path_output1_all, path_output2_all, path_output3_all)
path_output = path_output %>%
    arrange(dplyr::desc(Super_pathway), Sub_pathway)
#p <- path_output %>%
 #   arrange(desc(Super_pathway)) %>% mutate(Sub_pathway=factor(Sub_pathway)) %>%
path_output$type = factor(path_output$type, levels = c("Fatal vs. Non-fatal", "Severe vs. Non-severe", "Non-servivors vs. Servivors"))
 p_bubble = ggplot(path_output, aes(x= factor(Sub_pathway), y =type)) +
    scale_colour_viridis_c( alpha = 1,
                            begin = 0,
                            end = 1,
                            direction = 1)+
    geom_tile(aes(fill = Super_pathway2), height = Inf, alpha =0.7) +
    geom_point(aes(size = Impact_ratio, colour = Increase_ratio)) + #Size
    geom_point(aes(size = Impact_ratio),shape = 1,colour = "black") +  #Size
    scale_fill_manual(values = c("Amino Acid" = "#fed0ed",
                                 "Carbohydrate" = "#f4d5fe",
                                 "Cofactors and Vitamins" = "#d0e1ff",
                                 "Energy" = "#b3eaf6",
                                   "Lipid" = "#b3ece2", 
                                 "Nucleotide" = "#b3eac3",
                                 "Partially Characterized Molecules" = "#dee5b3",
                                   "Peptide" = "#f2deb3", "Xenobiotics" = "#Fdd6d3"))+
    theme_light() +xlab("Sub_pathway")+ theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, size =10),axis.text.y = element_text(size =10)) +
    coord_flip() #+ scale_x_reverse()
 p_bubble
```
```{r}
summary_metab_data = path_output
summary_metab_data$Sub_pathway = as.character(summary_metab_data$Sub_pathway)
summary_metab_data$Super_pathway = as.character(summary_metab_data$Super_pathway)
summary_metab_data$type = factor(summary_metab_data$type, levels = c("Non-servivors vs. Servivors", "Severe vs. Non-severe", "Fatal vs. Non-fatal"))
gg1 <- summary_metab_data %>%
  arrange(Super_pathway, Sub_pathway) %>%
  filter(!is.na(Super_pathway)) %>%
  ggplot(aes(y=type, x=fct_reorder(Sub_pathway, Super_pathway))) +
  geom_point(aes(size = Impact_ratio, color = mean_neglog10FDR)) +
  ylab("Comparisons") +
  theme_light() +
  theme(axis.text.y = element_text(color = "black", size = 10),
        axis.text.x = element_blank(),
        panel.border = element_rect(color="black", size = 1),
        axis.ticks = element_line(color = "black"),
        axis.title.x = element_blank(),axis.title.y = element_blank(),legend.position = "bottom") +
   scale_colour_gradientn(colours = rev(c("darkred", "red", "orange")), name = "-log10 Transformed\nAdjusted p-values")

gg1
tmp.data <- summary_metab_data %>%
  distinct(Sub_pathway, Super_pathway) %>%
  arrange(Super_pathway, Sub_pathway) %>%
  mutate(Sub_pathway = as.factor(Sub_pathway)) 
Sub_pathway_levels = tmp.data$Sub_pathway
tmp.data$Sub_pathway <-factor(tmp.data$Sub_pathway, levels = Sub_pathway_levels)

gg2 <- tmp.data %>%
  ggplot(aes(y=1, x= Sub_pathway)) +#fct_reorder(SUB_PATHWAY, SUPER_PATHWAY)
  geom_tile(aes(fill = Super_pathway)) +
  theme_classic() +
  theme(axis.text.y = element_blank(),
       axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, color = "black", size = 10),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),axis.title.x = element_blank(),legend.position = "bottom") +
      scale_fill_manual(values = c("Amino Acid" = "#fed0ed",
                                 "Carbohydrate" = "#f4d5fe",
                                 "Cofactors and Vitamins" = "#d0e1ff",
                                "Energy" = "#b3eaf6",
                                   "Lipid" = "#b3ece2", 
                                 "Nucleotide" = "#b3eac3",
                                 "Partially Characterized Molecules" = "#dee5b3",
                                   "Peptide" = "#f2deb3", "Xenobiotics" = "#Fdd6d3"))
gg2 
#scale_fill_manual(values = c("Amino Acid" = "#972367",
#                                 "Carbohydrate" = "#ba1d27",
#                                 "Cofactors and Vitamins" = "orange",#58801d
#                                 "Energy" = "#ffd932",
#                                   "Lipid" = "#58801d", 
#                                 "Nucleotide" = "#b6e0f5",
#                                 "Partially Characterized Molecules" = "#76a4f8",
 #                                  "Peptide" = "purple", "Xenobiotics" = "#36332f"))
#
 

library(aplot)
Figure_B_1 = gg2 %>% insert_top(gg1, height = 4) 
Figure_B_1 = print(Figure_B_1) & theme(legend.position='bottom')
```
```{r}
Figure_B_1_1 = gg2 %>% insert_top(gg1, height = 4) 
Figure_B_1_1 = print(Figure_B_1_1) & theme(legend.position='bottom')

Figure_B_1_2 = gg2 %>% insert_top(gg1, height = 8) 
Figure_B_1_2 = print(Figure_B_1_2) & theme(legend.position='bottom')
```

```{r}
summary_metab_data = path_output
summary_metab_data$Sub_pathway = as.character(summary_metab_data$Sub_pathway)
summary_metab_data$Super_pathway = as.character(summary_metab_data$Super_pathway)
gg1 <- summary_metab_data %>%
  arrange(Super_pathway, Sub_pathway) %>%
  filter(!is.na(Super_pathway)) %>%
  ggplot(aes(x=type, y=fct_reorder(Sub_pathway, Super_pathway))) +
  geom_point(aes(size = Impact_ratio, color = mean_neglog10FDR)) +
  xlab("Comparisons") +
  theme_light() +
  theme(axis.text.x = element_text(vjust = 1, hjust=1,color = "black",angle = 45),
        axis.text.y = element_blank(),
        panel.border = element_rect(color="black", size = 1),
        axis.ticks = element_line(color = "black"),
        axis.title.y = element_blank()) +
   scale_colour_gradientn(colours = rev(c("darkred", "red", "orange")), name = "-log10 Transformed\nAdjusted p-values")


tmp.data <- summary_metab_data %>%
  distinct(Sub_pathway, Super_pathway) %>%
  arrange(Super_pathway, Sub_pathway) %>%
  mutate(Sub_pathway = as.factor(Sub_pathway)) 
Sub_pathway_levels = tmp.data$Sub_pathway
tmp.data$Sub_pathway <-factor(tmp.data$Sub_pathway, levels = Sub_pathway_levels)

gg2 <- tmp.data %>%
  ggplot(aes(x=1, y= Sub_pathway)) +#fct_reorder(SUB_PATHWAY, SUPER_PATHWAY)
  geom_tile(aes(fill = Super_pathway)) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
       axis.text.y = element_text(vjust = 1, hjust=1, color = "black"),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank()) +
      scale_fill_manual(values = c("Amino Acid" = "#fed0ed",
                                 "Carbohydrate" = "#f4d5fe",
                                 "Cofactors and Vitamins" = "#d0e1ff",
                                "Energy" = "#b3eaf6",
                                   "Lipid" = "#b3ece2", 
                                 "Nucleotide" = "#b3eac3",
                                 "Partially Characterized Molecules" = "#dee5b3",
                                   "Peptide" = "#f2deb3", "Xenobiotics" = "#Fdd6d3"))
gg2 
#scale_fill_manual(values = c("Amino Acid" = "#972367",
#                                 "Carbohydrate" = "#ba1d27",
#                                 "Cofactors and Vitamins" = "orange",#58801d
#                                 "Energy" = "#ffd932",
#                                   "Lipid" = "#58801d", 
#                                 "Nucleotide" = "#b6e0f5",
#                                 "Partially Characterized Molecules" = "#76a4f8",
 #                                  "Peptide" = "purple", "Xenobiotics" = "#36332f"))
#
 

library(aplot)
figure_B = gg2 %>% insert_right(gg1, width = 4)
```


```{r}

path_output1 = bubble_plot_AD_baseline_table
path_output2 = bubble_plot_123_baseline_table
path_output3 = bubble_plot_diedever_baseline_table

all_pathways =  rbind(bubble_plot_diedever_baseline_table[,1:2], bubble_plot_AD_baseline_table[,1:2], bubble_plot_123_baseline_table[,1:2])
all_pathways <- unique(all_pathways)

path_output1_all = merge(path_output1, all_pathways, by = c("Super_pathway" , "Sub_pathway"), all.y = TRUE)
path_output2_all = merge(path_output2, all_pathways, by = c("Super_pathway" , "Sub_pathway"), all.y = TRUE)
path_output3_all = merge(path_output3, all_pathways, by = c("Super_pathway" , "Sub_pathway"), all.y = TRUE)

path_output1_all$type = "Fatal vs. Non-fatal"
path_output2_all$type = "Severe vs. Non-severe"
path_output3_all$type = "Non-servivors vs. Servivors"

path_output = rbind(path_output1_all, path_output2_all, path_output3_all)
path_output = path_output %>%
    arrange(dplyr::desc(Super_pathway), Sub_pathway)
#p <- path_output %>%
 #   arrange(desc(Super_pathway)) %>% mutate(Sub_pathway=factor(Sub_pathway)) %>%
path_output$type = factor(path_output$type, levels = c("Fatal vs. Non-fatal", "Severe vs. Non-severe", "Non-servivors vs. Servivors"))
 
```
```{r}
summary_metab_data = path_output
summary_metab_data$Sub_pathway = as.character(summary_metab_data$Sub_pathway)
summary_metab_data$Super_pathway = as.character(summary_metab_data$Super_pathway)
summary_metab_data$type = factor(summary_metab_data$type, levels = c("Non-servivors vs. Servivors", "Severe vs. Non-severe", "Fatal vs. Non-fatal"))
gg1 <- summary_metab_data %>%
  arrange(Super_pathway, Sub_pathway) %>%
  filter(!is.na(Super_pathway)) %>%
  ggplot(aes(y=type, x=fct_reorder(Sub_pathway, Super_pathway))) +
  geom_point(aes(size = Impact_ratio, color = mean_neglog10FDR)) +
  ylab("Comparisons") +
  theme_light() +
  theme(axis.text.y = element_text(color = "black", size = 10),
        axis.text.x = element_blank(),
        panel.border = element_rect(color="black", size = 1),
        axis.ticks = element_line(color = "black"),
        axis.title.x = element_blank(),axis.title.y = element_blank(),legend.position = "bottom") +
   scale_colour_gradientn(colours = rev(c("darkred", "red", "orange")), name = "-log10 Transformed\nAdjusted p-values")


tmp.data <- summary_metab_data %>%
  distinct(Sub_pathway, Super_pathway) %>%
  arrange(Super_pathway, Sub_pathway) %>%
  mutate(Sub_pathway = as.factor(Sub_pathway)) 
Sub_pathway_levels = tmp.data$Sub_pathway
tmp.data$Sub_pathway <-factor(tmp.data$Sub_pathway, levels = Sub_pathway_levels)

gg2 <- tmp.data %>%
  ggplot(aes(y=1, x= Sub_pathway)) +#fct_reorder(SUB_PATHWAY, SUPER_PATHWAY)
  geom_tile(aes(fill = Super_pathway)) +
  theme_classic() +
  theme(axis.text.y = element_blank(),
       axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, color = "black", size = 10),
        axis.ticks.y = element_blank(),
        axis.title.y = element_blank(),axis.title.x = element_blank(),legend.position = "bottom") +
      scale_fill_manual(values = c("Amino Acid" = "#fed0ed",
                                 "Carbohydrate" = "#f4d5fe",
                                 "Cofactors and Vitamins" = "#d0e1ff",
                                "Energy" = "#b3eaf6",
                                   "Lipid" = "#b3ece2", 
                                 "Nucleotide" = "#b3eac3",
                                 "Partially Characterized Molecules" = "#dee5b3",
                                   "Peptide" = "#f2deb3", "Xenobiotics" = "#Fdd6d3"))
gg2 
#scale_fill_manual(values = c("Amino Acid" = "#972367",
#                                 "Carbohydrate" = "#ba1d27",
#                                 "Cofactors and Vitamins" = "orange",#58801d
#                                 "Energy" = "#ffd932",
#                                   "Lipid" = "#58801d", 
#                                 "Nucleotide" = "#b6e0f5",
#                                 "Partially Characterized Molecules" = "#76a4f8",
 #                                  "Peptide" = "purple", "Xenobiotics" = "#36332f"))
#
 

library(aplot)
Figure_B_1 = gg2 %>% insert_top(gg1, height = 4) 
Figure_B_1 = print(Figure_B_1) & theme(legend.position='bottom')
```


```{ r}
x_data_frame = lr_output_AD_baseline[,c("Compound_ID","coef","qval")]
y_data_frame = lr_output_123_baseline[,c("Compound_ID","coef","qval")]

names(x_data_frame)[2] = "Fatal"
names(x_data_frame)[3] = "qval_Fatal"
names(y_data_frame)[2] = "Severe"
names(y_data_frame)[3] = "qval_Severe"

data_all = merge(x_data_frame, y_data_frame, by = "Compound_ID")
data_all = subset(data_all, data_all$Compound_ID!= "X1442")

data_all <- data_all %>% as.data.frame() %>% dplyr::mutate(Direction = case_when(
  qval_Fatal <= 0.05 & qval_Severe <= 0.05 & Fatal > 0 &  Severe>0 ~ "Up sig both fatal and severe", 
  qval_Fatal <= 0.05 & qval_Severe <= 0.05 & Fatal < 0 &  Severe<0 ~ "Down sig both fatal and severe", 
  (qval_Fatal <= 0.05 | qval_Severe <= 0.05) & Fatal > 0 &  Severe>0 ~ "Up sig fatal or severe", 
  (qval_Fatal <= 0.05 | qval_Severe <= 0.05) & Fatal < 0 &  Severe<0 ~ "Down sig fatal or severe", 
  (qval_Fatal <= 0.05 & qval_Severe <= 0.05) & Fatal < 0 &  Severe>0 ~ "Sig opposite fatal and severe", 
  (qval_Fatal <= 0.05 & qval_Severe <= 0.05) & Fatal > 0 &  Severe<0 ~ "Sig opposite fatal and severe", 
  TRUE ~ "NotSig")) 


ggplot(data = data_all)  +
    geom_point(aes(x = Fatal, y =Severe,fill = Direction), size = 3, pch = 21, alpha = 0.5) +
    theme_classic()+ geom_hline(yintercept =0, linetype = "dashed", alpha = 0.8) + geom_vline(xintercept =0, linetype = "dashed", alpha = 0.8) +
    scale_fill_manual(values = c("NotSig" = "#C0C0C0", "Down sig both fatal and severe"="darkblue","Up sig both fatal and severe"="darkred", "Down sig fatal or severe"="lightblue", "Up sig fatal or severe"="pink", "Sig opposite fatal and severe"="yellow")) 
```


# Bubble by pathways_up and down 12
```{ r}
bubble_plot <- function(significant_table, metabolomics_rf, rowfeature_all){
  #lr_output_V1_12_glm_simple_up, plasma_metabolomics_global_rowfeature_temp,rowfeature_all
  significant_table <- significant_table
  metabolomics_rf <- metabolomics_rf
  rowfeature_all <- rowfeature_all
  alpha = 1
  begin = 0
  end = 1
  #
  Sub_size <- c()
  Sub_pathway <- c()
  if(is.factor(rowfeature_all$SUB_PATHWAY)){
    rowfeature_all$SUB_PATHWAY <- unfactor(rowfeature_all$SUB_PATHWAY)
  }
  subpath_list <- unique(rowfeature_all$SUB_PATHWAY)
  
  for(i in subpath_list){
    if (!is.na(i)){
      sub_table <- rowfeature_all %>% filter(SUB_PATHWAY == i)
      
      sub <- unique(sub_table$SUB_PATHWAY)
      Sub_pathway <- c(Sub_pathway, sub)
      
      size_sub <- length(sub_table$SUB_PATHWAY)
      Sub_size <- c(Sub_size, size_sub)
    }
    
    
    
  }
  Sub_pathway_table <- data.table(Sub_pathway, Sub_size)
  results_sig <- significant_table
  results_sig_more <- merge(results_sig, Sub_pathway_table, by.x ="SUB_PATHWAY.y", by.y = "Sub_pathway")
  results_sig_more$SUPER_PATHWAY <- as.character(results_sig_more$SUPER_PATHWAY.y)
  pathwayfile <- results_sig_more
 
  pathwayfile$`-log10FDR` <- -log10(pathwayfile$qval)
  subpath_list <- unique(pathwayfile$SUB_PATHWAY.y)
  
  Super_pathway <- c()
  Sub_pathway <- c()
  mean_neglog10FDR <- c()
  Size <- c()
  Impact_ratio <- c()
  Increase_ratio <- c()
  Size_all <- c()
  ###### impact ratio: # of significant / # of total in the subpathway 0-1
  for(i in subpath_list){
    sub_table <- pathwayfile %>% filter(SUB_PATHWAY.y == i)
    
    supp <- unique(sub_table$SUPER_PATHWAY)
    Super_pathway <- c(Super_pathway, supp)
    
    sub <- unique(sub_table$SUB_PATHWAY.y)
    Sub_pathway <- c(Sub_pathway, sub)
    
    mean_qval <- mean(sub_table$`-log10FDR`)
    mean_neglog10FDR <- c(mean_neglog10FDR, mean_qval)
    
    size_sub <- length(sub_table$SUB_PATHWAY.y)
    Size <- c(Size, size_sub)
    size_all <- unique(sub_table$Sub_size)
    Size_all <- c(Size_all, size_all)
    im_ratio <- size_sub/size_all
    Impact_ratio <- c(Impact_ratio, im_ratio)
    
    increased_table <- sub_table %>% filter(coef >= 0) 
    increased_num <- length(increased_table$mets_list) / size_sub
    Increase_ratio <- c(Increase_ratio, increased_num)
    
  }
  path_output <- data.table(Super_pathway, Sub_pathway, mean_neglog10FDR, Impact_ratio,Size , Increase_ratio,Size_all)#
  path_output$Super_pathway <- factor(path_output$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"))
  #path_output <- path_output[order(Super_pathway, desc(mean_p)),]
  path_output <- path_output[order(Super_pathway),]
  
  path_output$Super_pathway2 <- path_output$Super_pathway
  path_output$Super_pathway2 <- factor(path_output$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"), ordered = TRUE)
 
path_output = path_output %>%
    arrange(desc(Super_pathway),mean_neglog10FDR) %>% mutate(Sub_pathway=factor(Sub_pathway, levels=Sub_pathway))
  #path_output<- path_output[seq(dim(path_output)[1],1),]
return(path_output)
}
```
```{ r}

plasma_metabolomics_global_rowfeature_temp$Compound_ID = rownames(plasma_metabolomics_global_rowfeature_temp)
rowfeature_all = subset(plasma_metabolomics_global_rowfeature_temp, plasma_metabolomics_global_rowfeature_temp$SUB_PATHWAY %in% Sub_pathway_table2)[,c("Compound_ID","SUPER_PATHWAY","SUB_PATHWAY")]
lr_output_V1_12_glm_simple_sig = merge(lr_output_AD_baseline, plasma_metabolomics_global_rowfeature_temp, by = "Compound_ID")
lr_output_V1_12_glm_simple_sig = subset(lr_output_V1_12_glm_simple_sig,lr_output_V1_12_glm_simple_sig$qval <= 0.05)
lr_output_V1_12_glm_simple_up = subset(lr_output_V1_12_glm_simple_sig, lr_output_V1_12_glm_simple_sig$coef >= 0)
lr_output_V1_12_glm_simple_down = subset(lr_output_V1_12_glm_simple_sig, lr_output_V1_12_glm_simple_sig$coef < 0)
  
path_output_V1_21_up = bubble_plot(lr_output_V1_12_glm_simple_up, plasma_metabolomics_global_rowfeature_temp,rowfeature_all)
path_output_V1_21_down = bubble_plot(lr_output_V1_12_glm_simple_down, plasma_metabolomics_global_rowfeature_temp,rowfeature_all)
####V2
lr_output_V2_12_glm_simple_sig = merge(lr_output_AS_baseline, plasma_metabolomics_global_rowfeature_temp, by = "Compound_ID")
lr_output_V2_12_glm_simple_sig = subset(lr_output_V2_12_glm_simple_sig,lr_output_V2_12_glm_simple_sig$qval <= 0.05)
lr_output_V2_12_glm_simple_up = subset(lr_output_V2_12_glm_simple_sig, lr_output_V2_12_glm_simple_sig$coef >= 0)
lr_output_V2_12_glm_simple_down = subset(lr_output_V2_12_glm_simple_sig, lr_output_V2_12_glm_simple_sig$coef < 0)
  
path_output_V2_21_up = bubble_plot(lr_output_V2_12_glm_simple_up, plasma_metabolomics_global_rowfeature_temp,rowfeature_all)
path_output_V2_21_down = bubble_plot(lr_output_V2_12_glm_simple_down, plasma_metabolomics_global_rowfeature_temp,rowfeature_all)

####V3
lr_output_V3_12_glm_simple_sig = merge(lr_output_45_baseline, plasma_metabolomics_global_rowfeature_temp, by = "Compound_ID")
lr_output_V3_12_glm_simple_sig = subset(lr_output_V3_12_glm_simple_sig,lr_output_V3_12_glm_simple_sig$qval <= 0.05)
lr_output_V3_12_glm_simple_up = subset(lr_output_V3_12_glm_simple_sig, lr_output_V3_12_glm_simple_sig$coef >= 0)
lr_output_V3_12_glm_simple_down = subset(lr_output_V3_12_glm_simple_sig, lr_output_V3_12_glm_simple_sig$coef < 0)
  
path_output_V3_21_up = bubble_plot(lr_output_V3_12_glm_simple_up, plasma_metabolomics_global_rowfeature_temp,rowfeature_all)
path_output_V3_21_down = bubble_plot(lr_output_V3_12_glm_simple_down, plasma_metabolomics_global_rowfeature_temp,rowfeature_all)


```


```{ r}
Sub_pathway_table2 = unique(rowfeature_all[,c("SUPER_PATHWAY","SUB_PATHWAY")])
names(Sub_pathway_table2) = c("Super_pathway","Sub_pathway")
path_output_V1_21_more_up <- merge(Sub_pathway_table2, path_output_V1_21_up, by = c("Super_pathway","Sub_pathway"), all.x = TRUE)
path_output_V1_21_more_up$Super_pathway2 <- path_output_V1_21_more_up$Super_pathway
path_output_V1_21_more_down <- merge(Sub_pathway_table2, path_output_V1_21_down, by = c("Super_pathway","Sub_pathway"), all.x = TRUE)
path_output_V1_21_more_down$Super_pathway2 <- path_output_V1_21_more_down$Super_pathway

path_output_V2_21_more_up <- merge(Sub_pathway_table2, path_output_V2_21_up, by = c("Super_pathway","Sub_pathway"), all.x = TRUE)
path_output_V2_21_more_up$Super_pathway2 <- path_output_V2_21_more_up$Super_pathway
path_output_V2_21_more_down <- merge(Sub_pathway_table2, path_output_V2_21_down, by = c("Super_pathway","Sub_pathway"), all.x = TRUE)
path_output_V2_21_more_down$Super_pathway2 <- path_output_V2_21_more_down$Super_pathway

path_output_V3_21_more_up <- merge(Sub_pathway_table2, path_output_V3_21_up, by = c("Super_pathway","Sub_pathway"), all.x = TRUE)
path_output_V3_21_more_up$Super_pathway2 <- path_output_V3_21_more_up$Super_pathway
path_output_V3_21_more_down <- merge(Sub_pathway_table2, path_output_V3_21_down, by = c("Super_pathway","Sub_pathway"), all.x = TRUE)
path_output_V3_21_more_down$Super_pathway2 <- path_output_V3_21_more_down$Super_pathway


```

```{ r}
Super_pathway = path_output_V3_21_more_up$Super_pathway
########################V1
path_output_V1_21_more_up$type = "Up"
path_output_V1_21_more_down$type = "Down"

path_output_V1_21_more_all = rbind(path_output_V1_21_more_up, path_output_V1_21_more_down)
 path_output_V1_21_more_all$Super_pathway <- factor(path_output_V1_21_more_all$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"))
 
  path_output_V1_21_more_all <- path_output_V1_21_more_all[order(path_output_V1_21_more_all$Super_pathway),]
  
  path_output_V1_21_more_all$Super_pathway2 <- path_output_V1_21_more_all$Super_pathway
  path_output_V1_21_more_all$Super_pathway2 <- factor(path_output_V1_21_more_all$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"), ordered = TRUE)
   alpha = 1
  begin = 0
  end = 1
   Sub_pathway2 = rev(path_output_V1_21_more_up$Sub_pathway)
pV1 <-  path_output_V1_21_more_all %>%
    arrange(desc(Super_pathway)) %>% mutate(Sub_pathway=factor(Sub_pathway, levels=Sub_pathway2)) %>%
    ggplot(aes(x= Sub_pathway, y =mean_neglog10FDR)) +
    scale_colour_viridis_c( alpha = alpha,
                            begin = begin,
                            end = end,
                            direction = 1)+
    geom_tile(aes(fill = Super_pathway2), height = Inf, alpha =0.7) +
    geom_point(aes(size = Impact_ratio, shape = type)) + #Size
    #geom_point(aes(size = Impact_ratio),shape = 1,colour = "black") +  #Size
    scale_fill_manual(values = c("Amino Acid" = "#fed0ed",
                                 "Carbohydrate" = "#f4d5fe",
                                 "Cofactors and Vitamins" = "#d0e1ff",
                                 "Energy" = "#b3eaf6",
                                   "Lipid" = "#b3ece2", 
                                 "Nucleotide" = "#b3eac3",
                                 "Partially Characterized Molecules" = "#dee5b3",
                                   "Peptide" = "#f2deb3", "Xenobiotics" = "#Fdd6d3"))+scale_shape_manual(values=c("Up" = 2, 
                                                                                                                  "Down" = 6))+
    theme_light() +xlab("Sub_pathway")+labs(title = "Fatal vs. Non-fatal")+
    coord_flip() #+ scale_x_reverse()
pV1

########################V2
path_output_V2_21_more_up$type = "Up"
path_output_V2_21_more_down$type = "Down"

path_output_V2_21_more_all = rbind(path_output_V2_21_more_up, path_output_V2_21_more_down)
 path_output_V2_21_more_all$Super_pathway <- factor(path_output_V2_21_more_all$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"))
 
  path_output_V2_21_more_all <- path_output_V2_21_more_all[order(path_output_V2_21_more_all$Super_pathway),]
  
  path_output_V2_21_more_all$Super_pathway2 <- path_output_V2_21_more_all$Super_pathway
  path_output_V2_21_more_all$Super_pathway2 <- factor(path_output_V2_21_more_all$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"), ordered = TRUE)
   alpha = 1
  begin = 0
  end = 1
  
pV2 <-  path_output_V2_21_more_all %>%
    arrange(desc(Super_pathway)) %>% mutate(Sub_pathway=factor(Sub_pathway, levels=Sub_pathway2)) %>%
    ggplot(aes(x= Sub_pathway, y =mean_neglog10FDR)) +
    scale_colour_viridis_c( alpha = alpha,
                            begin = begin,
                            end = end,
                            direction = 1)+
    geom_tile(aes(fill = Super_pathway2), height = Inf, alpha =0.7) +
    geom_point(aes(size = Impact_ratio, shape = type)) + #Size
    #geom_point(aes(size = Impact_ratio),shape = 1,colour = "black") +  #Size
    scale_fill_manual(values = c("Amino Acid" = "#fed0ed",
                                 "Carbohydrate" = "#f4d5fe",
                                 "Cofactors and Vitamins" = "#d0e1ff",
                                 "Energy" = "#b3eaf6",
                                   "Lipid" = "#b3ece2", 
                                 "Nucleotide" = "#b3eac3",
                                 "Partially Characterized Molecules" = "#dee5b3",
                                   "Peptide" = "#f2deb3", "Xenobiotics" = "#Fdd6d3"))+scale_shape_manual(values=c("Up" = 2, 
                                                                                                                  "Down" = 6))+
    theme_light() +xlab("Sub_pathway")+labs(title = "Severe vs. Non-severe")+
    coord_flip() #+ scale_x_reverse()
pV2

########################V3
path_output_V3_21_more_up$type = "Up"
path_output_V3_21_more_down$type = "Down"

path_output_V3_21_more_all = rbind(path_output_V3_21_more_up)
 path_output_V3_21_more_all$Super_pathway <- factor(path_output_V3_21_more_all$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"))
 
  path_output_V3_21_more_all <- path_output_V3_21_more_all[order(path_output_V3_21_more_all$Super_pathway),]
  
  path_output_V3_21_more_all$Super_pathway2 <- path_output_V3_21_more_all$Super_pathway
  path_output_V3_21_more_all$Super_pathway2 <- factor(path_output_V3_21_more_all$Super_pathway, levels = c("Amino Acid" ,"Carbohydrate", "Cofactors and Vitamins", "Energy", "Lipid", "Nucleotide", "Partially Characterized Molecules", "Peptide","Xenobiotics"), ordered = TRUE)
   alpha = 1
  begin = 0
  end = 1
  
pV3 <-  path_output_V3_21_more_all %>%
    arrange(desc(Super_pathway)) %>% mutate(Sub_pathway=factor(Sub_pathway, levels=Sub_pathway2)) %>%
    ggplot(aes(x= Sub_pathway, y =mean_neglog10FDR)) +
    scale_colour_viridis_c( alpha = alpha,
                            begin = begin,
                            end = end,
                            direction = 1)+
    geom_tile(aes(fill = Super_pathway2), height = Inf, alpha =0.7) +
    geom_point(aes(size = Impact_ratio, shape = type)) + #Size
    #geom_point(aes(size = Impact_ratio),shape = 1,colour = "black") +  #Size
    scale_fill_manual(values = c("Amino Acid" = "#fed0ed",
                                 "Carbohydrate" = "#f4d5fe",
                                 "Cofactors and Vitamins" = "#d0e1ff",
                                 "Energy" = "#b3eaf6",
                                   "Lipid" = "#b3ece2", 
                                 "Nucleotide" = "#b3eac3",
                                 "Partially Characterized Molecules" = "#dee5b3",
                                   "Peptide" = "#f2deb3", "Xenobiotics" = "#Fdd6d3"))+scale_shape_manual(values=c("Up" = 2, 
                                                                                                                  "Down" = 6))+
    theme_light() +xlab("Sub_pathway")+labs(title = "Fatal vs. Severe")+
    coord_flip() #+ scale_x_reverse()
pV3




remove_y <- theme(
  axis.text.y = element_blank(),
  axis.ticks.y = element_blank(),
  axis.title.y = element_blank()
  
)
p <- list(
  pV1 + 
  theme(legend.position="none") +ylim(1, 13),# + remove_y
  pV2 + remove_y + ylim(1,13)+
  theme(legend.position="none"),
  pV3 + remove_y +ylim(1,13)+
      theme(legend.position="none")
  
)
library(patchwork)
p_co <- wrap_plots(p, nrow = 1) + plot_layout(guides = "collect")
p_co
```

```{r}
lr_output_diedever_baseline_all = merge(lr_output_diedever_baseline, output_all_diedever, by.x = "Compound_ID",by.y = "outc")
lr_output_diedever_baseline_all$type = "Non-survivors vs. Survivors"
lr_output_diedever_baseline_all = lr_output_diedever_baseline_all %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05  ~ "Not significant"
    ))

lr_output_AD_baseline_all = merge(lr_output_AD_baseline, output_all_AD, by.x = "Compound_ID",by.y = "outc")
lr_output_AD_baseline_all$type = "Fatal vs. Non-fatal"
lr_output_AD_baseline_all = lr_output_AD_baseline_all %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))
lr_output_123_baseline_all = merge(lr_output_123_baseline, output_all_123, by.x = "Compound_ID",by.y = "outc")
lr_output_123_baseline_all$type = "Severe vs. Non-severe"
lr_output_123_baseline_all = lr_output_123_baseline_all %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))

all_results_sig = rbind(lr_output_AD_baseline_all, lr_output_123_baseline_all, lr_output_diedever_baseline_all)
all_results_sig$type = factor(all_results_sig$type, levels = c("Non-survivors vs. Survivors", "Severe vs. Non-severe", "Fatal vs. Non-fatal"))

  b <-all_results_sig[SUB_PATHWAY %in% c("Tryptophan Metabolism")] 
  if(dim(b)[1] ==0){
    return("No metabolites found!")
  }else{
    b$data <- "Tryptophan Metabolism"
    if(dim(b)[1]>100){
      b <- b[order(qval)]
      b <- b[1:100,]
    }
  }
    p <- b%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Tryptophan Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
 
  names(b)[2] = "name"
library(ggforestplot)
p_1= forestplot(
  df = b,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Tryptophan Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19)) +
   theme(plot.title = element_text(size=12))
p_1
```

```{r}
lr_output_AD_baseline_all_heat = lr_output_AD_baseline_all[,c("Name","SUB_PATHWAY","coef","qval")]
names(lr_output_AD_baseline_all_heat)[3] = "Fatal vs. Non-fatal"
names(lr_output_AD_baseline_all_heat)[4] = "qval_AD"
lr_output_123_baseline_all_heat = lr_output_123_baseline_all[,c("Name","SUB_PATHWAY","coef","qval")]
names(lr_output_123_baseline_all_heat)[3] = "Severe vs. Non-severe"
names(lr_output_123_baseline_all_heat)[4] = "qval_123"
lr_output_diedever_baseline_all_heat = lr_output_diedever_baseline_all[,c("Name","SUB_PATHWAY","coef","qval")]

names(lr_output_diedever_baseline_all_heat)[3] = "Non-survivors vs. Survivors"
names(lr_output_diedever_baseline_all_heat)[4] = "qval_su"

lr_output_baseline_all_heat = merge(lr_output_AD_baseline_all_heat,lr_output_123_baseline_all_heat, by = c("Name","SUB_PATHWAY"))
lr_output_baseline_all_heat = merge(lr_output_baseline_all_heat, lr_output_diedever_baseline_all_heat, by = c("Name","SUB_PATHWAY"))
```
```{r}
lr_output_baseline_all_heat_Tryptophan = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Tryptophan Metabolism"))
rownames(lr_output_baseline_all_heat_Tryptophan) = lr_output_baseline_all_heat_Tryptophan$Name
p_data_Tryptophan = lr_output_baseline_all_heat_Tryptophan[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Tryptophan = lr_output_baseline_all_heat_Tryptophan[,c("qval_AD","qval_123","qval_su")]
Heatmap_Trypton = Heatmap(p_data_Tryptophan,column_title = "Tryptophan Metabolism"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Tryptophan[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_Trypton, heatmap_legend_side = "bottom")
Heatmap_Trypton
```
```{r}
lr_output_baseline_all_heat_Leucine = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Leucine, Isoleucine and Valine Metabolism"))
rownames(lr_output_baseline_all_heat_Leucine) = lr_output_baseline_all_heat_Leucine$Name
p_data_Leucine = lr_output_baseline_all_heat_Leucine[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_Leucine[,c("qval_AD","qval_123","qval_su")]

col_fun <- colorRamp2(
  c(min(p_data_Leucine), 0, max(p_data_Leucine)),
  c("blue", "white", "red")
)
Heatmap_Leucine = Heatmap(p_data_Leucine,column_title = "Leucine, Isoleucine and Valine Metabolism"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",col = col_fun,
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_Leucine, heatmap_legend_side = "bottom")
Heatmap_Leucine
```

```{r}
lr_output_baseline_all_heat_Dicarboxylate = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Fatty Acid, Dicarboxylate"))
rownames(lr_output_baseline_all_heat_Dicarboxylate) = lr_output_baseline_all_heat_Dicarboxylate$Name
p_data_Dicarboxylate = lr_output_baseline_all_heat_Dicarboxylate[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Dicarboxylate = lr_output_baseline_all_heat_Dicarboxylate[,c("qval_AD","qval_123","qval_su")]
Heatmap_Dicarboxylate = Heatmap(p_data_Dicarboxylate,column_title = "Fatty Acid, Dicarboxylate"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),col = col_fun,
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Dicarboxylate[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_Dicarboxylate, heatmap_legend_side = "bottom")
Heatmap_Dicarboxylate
```

```{ r}
lr_output_baseline_all_heat_POM = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Polyamine Metabolism"))
rownames(lr_output_baseline_all_heat_POM) = lr_output_baseline_all_heat_POM$Name
p_data_Leucine = lr_output_baseline_all_heat_POM[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_POM[,c("qval_AD","qval_123","qval_su")]
Heatmap_POM = Heatmap(p_data_Leucine,column_title = "Polyamine Metabolism"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_POM, heatmap_legend_side = "bottom")
Heatmap_POM
```

```{r}
lr_output_baseline_all_heat_POM = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Polyamine Metabolism"))
rownames(lr_output_baseline_all_heat_POM) = lr_output_baseline_all_heat_POM$Name
p_data_Leucine = lr_output_baseline_all_heat_POM[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_POM[,c("qval_AD","qval_123","qval_su")]
Heatmap_POM = Heatmap(p_data_Leucine,column_title = "Polyamine Metabolism"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),col = col_fun,
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_POM, heatmap_legend_side = "bottom")
Heatmap_POM
```
Sphingosines
```{r}
lr_output_baseline_all_heat_sph = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Sphingosines"))
rownames(lr_output_baseline_all_heat_sph) = lr_output_baseline_all_heat_sph$Name
p_data_Leucine = lr_output_baseline_all_heat_sph[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_sph[,c("qval_AD","qval_123","qval_su")]
Heatmap_sph = Heatmap(p_data_Leucine,column_title = "Sphingosines"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),col = col_fun,
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_sph, heatmap_legend_side = "bottom")
Heatmap_sph
```

Urea cycle; Arginine and Proline Metabolism
```{r}
lr_output_baseline_all_heat_uc = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Urea cycle; Arginine and Proline Metabolism"))
rownames(lr_output_baseline_all_heat_uc) = lr_output_baseline_all_heat_uc$Name
p_data_Leucine = lr_output_baseline_all_heat_uc[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_uc[,c("qval_AD","qval_123","qval_su")]
Heatmap_uc = Heatmap(p_data_Leucine,column_title = "Urea cycle; Arginine and Proline Metabolism"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),col = col_fun,
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_uc, heatmap_legend_side = "bottom")
Heatmap_uc
```
Aminosugar Metabolism"
```{r}
lr_output_baseline_all_heat_uc = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Aminosugar Metabolism"))
rownames(lr_output_baseline_all_heat_uc) = lr_output_baseline_all_heat_uc$Name
p_data_Leucine = lr_output_baseline_all_heat_uc[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_uc[,c("qval_AD","qval_123","qval_su")]
Heatmap_as = Heatmap(p_data_Leucine,column_title = "Aminosugar Metabolism"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),col = col_fun,
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_as, heatmap_legend_side = "bottom")
Heatmap_as
```

Aminosugar Metabolism"
```{r}
lr_output_baseline_all_heat_uc = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Pentose Metabolism"))
rownames(lr_output_baseline_all_heat_uc) = lr_output_baseline_all_heat_uc$Name
p_data_Leucine = lr_output_baseline_all_heat_uc[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_uc[,c("qval_AD","qval_123","qval_su")]
Heatmap_pm = Heatmap(p_data_Leucine,column_title = "Pentose Metabolism"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),col = col_fun,
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_pm, heatmap_legend_side = "bottom")
Heatmap_pm
```


```{r}
lr_output_baseline_all_heat_uc = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Pyrimidine Metabolism, Orotate containing"))
rownames(lr_output_baseline_all_heat_uc) = lr_output_baseline_all_heat_uc$Name
p_data_Leucine = lr_output_baseline_all_heat_uc[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_uc[,c("qval_AD","qval_123","qval_su")]
Heatmap_po = Heatmap(p_data_Leucine,column_title = "Pyrimidine Metabolism, Orotate containing"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),col = col_fun,
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_po, heatmap_legend_side = "bottom")
Heatmap_po
```

```{r}
lr_output_baseline_all_heat_uc = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Pentose Metabolism"))
rownames(lr_output_baseline_all_heat_uc) = lr_output_baseline_all_heat_uc$Name
p_data_Leucine = lr_output_baseline_all_heat_uc[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Leucine = lr_output_baseline_all_heat_uc[,c("qval_AD","qval_123","qval_su")]
Heatmap_pm = Heatmap(p_data_Leucine,column_title = "Pentose Metabolism"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),col = col_fun,
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Leucine[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_pm, heatmap_legend_side = "bottom")
Heatmap_pm
```
```{r}
ggarrange(grid.grabExpr(ComplexHeatmap::draw(Heatmap_Trypton, heatmap_legend_side = "bottom")),
          grid.grabExpr(ComplexHeatmap::draw(Heatmap_Leucine, heatmap_legend_side = "bottom")),
          grid.grabExpr(ComplexHeatmap::draw(Heatmap_Dicarboxylate, heatmap_legend_side = "bottom"))  , nrow = 1)
```

```{ r}
Fig5E = ComplexHeatmap::Heatmap(coef_table[,ii], #column_title = "Interaction with viral load",
             column_title_gp = gpar(fontsize = 25), column_title=gt_render("Interaction with viral load"),
             row_names_gp = gpar(fontsize = 15), column_names_gp = gpar(fontsize = 15), 
             heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "leftcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
             col = circlize::colorRamp2(c(-max(abs(coef_table[,ii])), 0, max(abs(coef_table[,ii]))), c("#284AA0", "white", "#BC4A4A")), na_col = "grey90", column_split = test_types, column_gap = unit(2, "mm"),
             cluster_columns = F, cluster_rows = F, cell_fun = function(j, i, x, y, w, h, fill) {
  if(pval_table_adj[i, j] < 0.0001) {
    grid.text("***", x, y, gp = gpar(cex=2), vjust=0.7)
  }else if(pval_table_adj[i, j] < 0.001) {
    grid.text("**", x, y, gp = gpar(cex=2), vjust=0.7)
  }else if(pval_table_adj[i, j] < 0.05) {
    grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
  }
})

ComplexHeatmap::draw(Fig5E, heatmap_legend_side = "bottom")
```

```{ r}
 b_Leucine <-all_results_sig[SUB_PATHWAY %in% c("Leucine, Isoleucine and Valine Metabolism")] 
  if(dim(b_Leucine)[1] ==0){
    return("No metabolites found!")
  }else{
    b_Leucine$data <- "Leucine, Isoleucine and Valine Metabolism"
    if(dim(b_Leucine)[1]>100){
      b_Leucine <- b_Leucine[order(qval)]
      b_Leucine <- b_Leucine[1:100,]
    }
  }
    p_Leucine <- b_Leucine%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Leucine, Isoleucine and Valine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
    
    names(b_Leucine)[2] = "name"
 p_2= forestplot(
  df = b_Leucine,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Leucine, Isoleucine and Valine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19)) +
   theme(plot.title = element_text(size=12))
p_2
```

#Fatty Acid, Dicarboxylate
```{ r}
b_Dicarboxylate <-all_results_sig[SUB_PATHWAY %in% c("Fatty Acid, Dicarboxylate")] 
  if(dim(b_Dicarboxylate)[1] ==0){
    return("No metabolites found!")
  }else{
    b_Dicarboxylate$data <- "Fatty Acid, Dicarboxylate"
    if(dim(b_Dicarboxylate)[1]>100){
      b_Dicarboxylate <- b_Dicarboxylate[order(qval)]
      b_Dicarboxylate <- b_Dicarboxylate[1:100,]
    }
  }
    p_Dicarboxylate <- b_Dicarboxylate%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Fatty Acid, Dicarboxylate", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
    names(b_Dicarboxylate)[2] = "name"
 p_3= forestplot(
  df = b_Dicarboxylate,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Fatty Acid, Dicarboxylate",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19)) +
   theme(plot.title = element_text(size=12))
p_3
```
```{r}
library(ggplotify)
part_2 =  ggarrange(grid.grabExpr(ComplexHeatmap::draw(Heatmap_pm, heatmap_legend_side = "bottom",padding = unit(c(20, 15, 15, 15), "mm"))),
          grid.grabExpr(ComplexHeatmap::draw(Heatmap_Leucine, heatmap_legend_side = "bottom",padding = unit(c(15, 15, 15, 15), "mm"))),
          grid.grabExpr(ComplexHeatmap::draw(Heatmap_Dicarboxylate, heatmap_legend_side = "bottom",padding = unit(c(15, 15, 15, 15), "mm")))  , nrow = 1,  labels = c("C"))
part_1 = ggarrange(a,as.ggplot(Figure_B_1)+ theme(plot.margin = unit(c(5,1,5,1), "cm")), ncol = 2, labels = c("A", "B"),widths = c(0.35, 1))
ggarrange(part_1, part_2, ncol = 1)
```
```{r}
part_2 =  ggarrange(grid.grabExpr(ComplexHeatmap::draw(Heatmap_pm, heatmap_legend_side = "bottom",padding = unit(c(20, 20, 15, 15), "mm"))),
          grid.grabExpr(ComplexHeatmap::draw(Heatmap_Leucine, heatmap_legend_side = "bottom",padding = unit(c(15, 15, 15, 15), "mm"))),
          grid.grabExpr(ComplexHeatmap::draw(Heatmap_Dicarboxylate, heatmap_legend_side = "bottom",padding = unit(c(15, 15, 15, 15), "mm")))  , nrow = 1,  labels = c("C"))
part_2
#part_1 = ggarrange(a,as.ggplot(Figure_B_1)+ theme(plot.margin = unit(c(5,1,5,1), "cm")), ncol = 2, labels = c("A", "B"),widths = c(0.35, 1))
b = ggarrange(Volcano_metabolomics_AD,Volcano_metabolomics_123, Volcano_metabolomics_diedever, ncol = 3, common.legend = T, legend="bottom")
b
as.ggplot(Figure_B_1_2)
ggarrange(b,as.ggplot(Figure_B_1_2), part_2, ncol = 1,labels = c("A", "B",""), heights = c(1,0.7,1)) #+ theme(plot.margin = unit(c(3,1,3,1), "cm")
```

```{r}
part_2 =  ggarrange(grid.grabExpr(ComplexHeatmap::draw(Heatmap_pm, heatmap_legend_side = "bottom",padding = unit(c(20, 20, 15, 15), "mm"))),
          grid.grabExpr(ComplexHeatmap::draw(Heatmap_Leucine, heatmap_legend_side = "bottom",padding = unit(c(15, 15, 15, 15), "mm"))),
          grid.grabExpr(ComplexHeatmap::draw(Heatmap_Dicarboxylate, heatmap_legend_side = "bottom",padding = unit(c(15, 15, 15, 15), "mm")))  , nrow = 1,  labels = c("C"))
part_1 = ggarrange(a,as.ggplot(figure_B) + theme(plot.margin = unit(c(3,1,1,3), "cm")), ncol = 2, labels = c("A", "B"),widths = c(0.35, 1))
ggarrange(part_1, part_2, ncol = 1)
```

```{r}
save(lr_output_baseline_all_heat,file ="/scratch/metabolomics/plasma-metabolomics-global/scripts/deep_dive_met/TG_data_use_heatmap.Rdata")
lr_output_baseline_all_heat_Sphingomyelins = as.data.frame(subset(lr_output_baseline_all_heat, lr_output_baseline_all_heat$SUB_PATHWAY == "Sphingomyelins"))
rownames(lr_output_baseline_all_heat_Sphingomyelins) = lr_output_baseline_all_heat_Sphingomyelins$Name
p_data_Sphingomyelins = lr_output_baseline_all_heat_Sphingomyelins[,c("Fatal vs. Non-fatal","Severe vs. Non-severe","Non-survivors vs. Survivors")]
sig_data_Sphingomyelins = lr_output_baseline_all_heat_Sphingomyelins[,c("qval_AD","qval_123","qval_su")]
Heatmap_Sphingomyelins = Heatmap(p_data_Sphingomyelins,column_title = "Sphingomyelins"
                          , cluster_rows = FALSE, cluster_columns = FALSE, column_names_rot = 45,
                          heatmap_legend_param = list(title = "Coefficient    ",
             direction = "horizontal", title_position = "topcenter",
             title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
                          cell_fun = function(j, i, x, y, w, h, fill) {
                            if(sig_data_Sphingomyelins[i, j] < 0.05) {
                              grid.text("*", x, y, gp = gpar(cex=2), vjust=0.7)
                                }
                          }
                          )
ComplexHeatmap::draw(Heatmap_Sphingomyelins, heatmap_legend_side = "bottom")
Heatmap_Sphingomyelins
```
```{r}
Metabolomics_rowfeature$Compound_ID = rownames(Metabolomics_rowfeature)
Metabolomics_rowfeature_sort =  Metabolomics_rowfeature %>%
  arrange(SUPER_PATHWAY, SUB_PATHWAY)
write_csv(Metabolomics_rowfeature_sort, file = "~/deep_dive_rowfeature.csv")

lr_output_123_baseline_sorted = lr_output_123_baseline[
  match(Metabolomics_rowfeature_sort$Compound_ID,
        lr_output_123_baseline$Compound_ID),
]
lr_output_123_baseline_sorted$FDR = p.adjust(lr_output_123_baseline_sorted$pval, method = 'BH')
lr_output_123_baseline_sorted = lr_output_123_baseline_sorted[,c("Name","coef","pval","qval","FDR")]
names(lr_output_123_baseline_sorted) = c("Name","coef","P","Adjusted Bonferroni","Adjusted FDR")
write_csv(lr_output_123_baseline_sorted, file = "~/deep_dive_rowfeature_TGSe.csv")


```
```{r}
lr_output_AD_baseline_sorted = lr_output_AD_baseline[
  match(Metabolomics_rowfeature_sort$Compound_ID,
        lr_output_AD_baseline$Compound_ID),
]
lr_output_AD_baseline_sorted$FDR = p.adjust(lr_output_AD_baseline_sorted$pval, method = 'BH')
lr_output_AD_baseline_sorted = lr_output_AD_baseline_sorted[,c("Name","coef","pval","qval","FDR")]
names(lr_output_AD_baseline_sorted) = c("Name","coef","P","Adjusted Bonferroni","Adjusted FDR")
write_csv(lr_output_AD_baseline_sorted, file = "~/deep_dive_rowfeature_TGF.csv")

```
```{r}
lr_output_diedever_baseline_sorted = lr_output_diedever_baseline[
  match(Metabolomics_rowfeature_sort$Compound_ID,
        lr_output_diedever_baseline$Compound_ID),
]
lr_output_diedever_baseline_sorted$FDR = p.adjust(lr_output_diedever_baseline_sorted$pval, method = 'BH')
lr_output_diedever_baseline_sorted = lr_output_diedever_baseline_sorted[,c("Name","coef","pval","qval","FDR")]
names(lr_output_diedever_baseline_sorted) = c("Name","coef","P","Adjusted Bonferroni","Adjusted FDR")
write_csv(lr_output_diedever_baseline_sorted, file = "~/deep_dive_rowfeature_TGSu.csv")

```

#Methionine, Cysteine, SAM and Taurine Metabolism
```{r}
b_Methionine <-all_results_sig[SUB_PATHWAY %in% c("Methionine, Cysteine, SAM and Taurine Metabolism")] 
  if(dim(b_Methionine)[1] ==0){
    return("No metabolites found!")
  }else{
    b_Methionine$data <- "Methionine, Cysteine, SAM and Taurine Metabolism"
    if(dim(b_Methionine)[1]>100){
      b_Methionine <- b_Methionine[order(qval)]
      b_Methionine <- b_Methionine[1:100,]
    }
  }
    p_Methionine <- b_Methionine%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Methionine, Cysteine, SAM and Taurine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
    names(b_Methionine)[2] = "name"
 p_3= forestplot(
  df = b_Methionine,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Methionine, Cysteine, SAM and Taurine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19))
p_3
```

#Glycine, Serine and Threonine Metabolism
```{r}
b_Glycine <-all_results_sig[SUB_PATHWAY %in% c("Glycine, Serine and Threonine Metabolism")] 
  if(dim(b_Glycine)[1] ==0){
    return("No metabolites found!")
  }else{
    b_Glycine$data <- "Glycine, Serine and Threonine Metabolism"
    if(dim(b_Glycine)[1]>100){
      b_Glycine <- b_Glycine[order(qval)]
      b_Glycine <- b_Glycine[1:100,]
    }
  }
    p_Glycine <- b_Glycine%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Methionine, Cysteine, SAM and Taurine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
    names(b_Glycine)[2] = "name"
 p_3= forestplot(
  df = b_Glycine,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Glycine, Serine and Threonine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19))
p_3
```

#Methionine, Cysteine, SAM and Taurine Metabolism for 34 vs. 123
```{r}
lr_output_123_baseline_all = merge(lr_output_123_baseline, output_all_123, by.x = "Compound_ID",by.y = "outc")
lr_output_123_baseline_all$type = "TG4_5 vs. TG1_3"
lr_output_123_baseline_all = lr_output_123_baseline_all %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))
b_Methionine <-lr_output_123_baseline_all[SUB_PATHWAY %in% c("Methionine, Cysteine, SAM and Taurine Metabolism")] 
  if(dim(b_Methionine)[1] ==0){
    return("No metabolites found!")
  }else{
    b_Methionine$data <- "Methionine, Cysteine, SAM and Taurine Metabolism"
    if(dim(b_Methionine)[1]>100){
      b_Methionine <- b_Methionine[order(qval)]
      b_Methionine <- b_Methionine[1:100,]
    }
  }
    p_Methionine <- b_Methionine%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Methionine, Cysteine, SAM and Taurine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
    names(b_Methionine)[2] = "name"
 p_3= forestplot(
  df = b_Methionine,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Methionine, Cysteine, SAM and Taurine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19))
p_3
```

#TG4_5, AA vs. Others(AG,GG)
```{r}
lr_output_AA_baseline_all = merge(lr_output_AA_baseline, output_all_AA, by.x = "Compound_ID",by.y = "outc")
lr_output_AA_baseline_all$type = "AA vs. Others"
lr_output_AA_baseline_all = lr_output_AA_baseline_all %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))
b_Methionine <-lr_output_AA_baseline_all[SUB_PATHWAY %in% c("Methionine, Cysteine, SAM and Taurine Metabolism")] 
  if(dim(b_Methionine)[1] ==0){
    return("No metabolites found!")
  }else{
    b_Methionine$data <- "Methionine, Cysteine, SAM and Taurine Metabolism"
    if(dim(b_Methionine)[1]>100){
      b_Methionine <- b_Methionine[order(qval)]
      b_Methionine <- b_Methionine[1:100,]
    }
  }
    p_Methionine <- b_Methionine%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Methionine, Cysteine, SAM and Taurine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
    names(b_Methionine)[2] = "name"
 p_3= forestplot(
  df = b_Methionine,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Methionine, Cysteine, SAM and Taurine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19))
p_3
```
