---
title: "IMPACC Targeted Metabolomics Analysis"
author: "Caitlin Syphurs"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
date: "`r format(Sys.time(), '%B %d, %Y')`"
editor_options:
  chunk_output_type: inline
---

Built with R version: `r getRversion()`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
setwd("/scratch/metabolomics/plasma-metabolomics-targeted/MTHFR")
```

### Load libraries
```{r load-packages}
suppressPackageStartupMessages(library(package = "knitr"))
suppressPackageStartupMessages(library(package = "impute"))
#suppressPackageStartupMessages(library(package = "igraph"))
suppressPackageStartupMessages(library(package = "ggnetwork"))
suppressPackageStartupMessages(library(package = "ordinal"))
suppressPackageStartupMessages(library(package = "qvalue"))
suppressPackageStartupMessages(library(package = "ggbeeswarm"))
suppressPackageStartupMessages(library(package = "ggeffects"))
#BiocManager::install("pvca")
suppressPackageStartupMessages(library(package = "pvca"))
suppressPackageStartupMessages(library(package = "GSA"))
suppressPackageStartupMessages(library(package = "lme4"))
suppressPackageStartupMessages(library(package = "nlme"))
suppressPackageStartupMessages(library(package = "tidyverse"))
suppressPackageStartupMessages(library(package = "ggfortify"))
suppressPackageStartupMessages(library(package = "epitools"))
suppressPackageStartupMessages(library(package = "circlize"))
suppressPackageStartupMessages(library(package = "rstatix"))
#BiocManager::install("sva")
suppressPackageStartupMessages(library(package = "sva"))
suppressPackageStartupMessages(library(package = "ComplexHeatmap"))
suppressPackageStartupMessages(library(package = "circlize"))
suppressPackageStartupMessages(library(package = "table1"))
suppressPackageStartupMessages(library(package = "dplyr"))
suppressPackageStartupMessages(library(package = "ggplot2"))
suppressPackageStartupMessages(library(package = "pROC"))
#install.packages(c("survival", "survminer"))
suppressPackageStartupMessages(library(package = "survival"))
suppressPackageStartupMessages(library(package = "survminer"))
suppressPackageStartupMessages(library(package = "geepack"))
suppressPackageStartupMessages(library(package = "table1"))
suppressPackageStartupMessages(library(package = "DT"))
suppressPackageStartupMessages(library(package = "data.table"))
suppressPackageStartupMessages(library(package = "readxl"))
suppressPackageStartupMessages(library(package = "lmtest"))

# set session options
options(show_col_types = FALSE)
VarCorr <- lme4::VarCorr
```

### Import convalescent annotations file
```{r read and process convalescent annotations}
convalescent_pro_groups <- read.csv("/scratch/clinical/current/impacc-convalescent.csv")
participant_status <- read.csv("/scratch/clinical/current/impacc_participant_status.csv")

# Convert PRO group cluster numbers to PRO groups as described in paper
# This mapping was communicated by Leying Guan via Teams on May 15th as the 
# original table is unclear about the exact PRO groups mappings
    
convalescent_pro_groups <- convalescent_pro_groups %>% 
  dplyr::mutate( pro_group_labels = case_match( 
    conv_cluster6, 
    c(1,3,4) ~ "MIN",
    c(2) ~ "PHY",
    c(5) ~ "COG",
    c(6) ~ "MLT"
  ))

# Check number of patients in each of the groups
conv_cohort_size <- length(unique(convalescent_pro_groups$participant_id))
cat("There are ", conv_cohort_size,
        " participants in the convalescent cohort.")
```

```{r define dataset version, echo = FALSE}
##### Authors: Ravi K. Patel, Jeremy Gygi;

# Indicate the version of the data to use ("current" or a date in "YYYY-MM-DD" format)
# If "current", the data files in "current" directories are used, otherwise, the data files in "legacy" directories that were current on the specified date are used.
DATA_VERSION <- "current"

# Indicate the phases that you want to include.
# NOTE: PHASES = 4 refers to 4a
#.      PHASES = 5 refers to 4b
#.      PHASES = 9 refers to phase 9 (leftover samples)
PHASES <- c(1,2,3,4,5,9,NA)

# Allowed QC? Defaults to pass (including questionable)
ALLOWED_SMPL_STATUS <- c("IMPACC sample assayed and passed QC", "IMPACC sample assayed with questionable QC")

# Keep only COVID-19 Positive individuals. Defaults to TRUE
# (TRUE will remove all COVID-19 Negative individuals)
KEEP_COVID19_POS <- TRUE
# Keep Escalation visits (Escalation 1, Escalation 2)? Defaults to FALSE
KEEP_ESCALATIONS = FALSE
# Keep Emory Controls? Defaults to FALSE
KEEP_EMORY_CONTROLS = FALSE

# Upper limit for event_date.
# Using higher values (e.g. 999) will include all visits (see KEEP_ESCALATIONS
#   for inclusion of Escalation visits).
EVENT_DATE_UPPER = 999

# Upper limit for event_type. For example, 6 would use up to 'Visit 6'.
# Using higher values (e.g. 999) will include all visits (see KEEP_ESCALATIONS
#   for inclusion of Escalation visits).
VISIT_LOWER = 1
VISIT_UPPER = 999

# Use 01/01/2023 locked data in "/data/clinical/current"?
USE_LOCKED_CLINICAL <- TRUE

# Filtering by manuscript cohorts:
# Currently supports core assay / integration manuscript cohort filtering.
FILTER_BY_CORE_ASSAY_COHORT = FALSE
FILTER_BY_INTEGRATION_COHORT = FALSE

# Filter by convalescent cohort participant ID
#SELECT_PARTICIPANT_ID = convalescent_pro_groups$participant_id
SELECT_PARTICIPANT_ID=NULL

# Exclude samples
EXCLUDE_SAMPLES = NULL
```

```{r Prepare environment, echo=FALSE}
##### Authors: Ravi K. Patel, Jeremy Gygi.
# Set options
options(encoding = "UTF-8") # instead of ANSI
options(scipen = 999) # non-scientific output
```

### ------------------------------------
#### **Study:** IMPACC
#### **Analysis title:** IMPACC Convalescent Data Loading
#### **Analysis description:** Loading combined data from all phases for analyses.
#### **License:** 
#### **Author(s):** Ravi K. Patel (ravi.patel2@ucsf.edu), Jeremy Gygi (jeremy.gygi@yale.edu), Steven Kleinstein (steven.kleinstein@yale.edu), Leying Guan (leying.guan@yale.edu)
#### **Date Created:** `r paste(format(Sys.time(), "%Y-%d-%m %X"), "UTC")`

#### **Source file:** `r knitr::current_input()`
#### **Dataset version:** `r ifelse(DATA_VERSION == "current", paste0("current (date: ", format(Sys.time(),"%Y-%m-%d"), ")"), paste0("legacy (date: ", DATA_VERSION, ")") )` `r paste0("(phases: ", paste0(PHASES, collapse=","), ")") `
#### **R version:** `r paste(R.Version()$version.string)`
#### **Platform:** `r paste(R.Version()$platform)`
#### **Running under**: `r paste(sessionInfo()$running)`

#### **Notes:** The observations for which the clinical data is unavailable are removed.
### ------------------------------------

```{r load header template codebase, echo=FALSE}
# Source the codebase file:
#source(path_to_codebase)
source("~/IMPACC_data_analysis_codebase.R")
```

```{r load datasets, message=FALSE, echo=FALSE}
# Load all available IMPACC datasets associated with DATA_VERSION.
data_env <- load_IMPACC_datasets(# Data version:
                                 DATA_VERSION = DATA_VERSION,
                                 # QC Sample Filtering:
                                 ALLOWED_SMPL_STATUS = ALLOWED_SMPL_STATUS,
                                 # COVID-19 Positive
                                 KEEP_COVID19_POS = KEEP_COVID19_POS,
                                 # Emory controls:
                                 KEEP_EMORY_CONTROLS = KEEP_EMORY_CONTROLS,
                                 # Escalations:
                                 KEEP_ESCALATIONS = KEEP_ESCALATIONS,
                                 # Use the locked clinical data (01/01/2023)
                                 USE_LOCKED_CLINICAL = USE_LOCKED_CLINICAL,
                                 
                                 # Don't restrict filtering by manuscript:
                                 FILTER_BY_CORE_ASSAY_COHORT = FILTER_BY_CORE_ASSAY_COHORT,
                                 FILTER_BY_INTEGRATION_COHORT = FILTER_BY_INTEGRATION_COHORT,
                                 
                                 # Select participant IDs
                                 SELECT_PARTICIPANT_ID = SELECT_PARTICIPANT_ID,
                                 
                                 # Exclude samples
                                 EXCLUDE_SAMPLES = EXCLUDE_SAMPLES,
                                 
                                 # event_type / visit upper limits
                                 EVENT_DATE_UPPER = 999,
                                 VISIT_LOWER = VISIT_LOWER,
                                 VISIT_UPPER = 6,
                                 
                                 # Phases
                                 PHASES = PHASES)

# Load and assign the data objects into the environment
# User can call upon any assay via the name e.g. serum_olink_counts
for( n in names(data_env) ) {
  assign( n, data_env[[n]] )
}


clinical_data <- data_env[[ "clinical_data" ]]

```

### Structure clinical data
* Add convalescent PRO clusters
* Subset for metabolomics samples (sample_type=="PROTEOMICS/MET")
```{r, results="hide", message=FALSE}
names(convalescent_pro_groups)[3] <- "conv_cluster4"
clinical_data <- merge(clinical_data, convalescent_pro_groups[,c(1,3)], by = "participant_id", all.x = TRUE)

# combine conv clusters 1, 3, and 4 -> 1 = 1,3,4
clinical_data$conv_cluster4 <- ifelse(clinical_data$conv_cluster4 == 2, 2, 
                                     ifelse(clinical_data$conv_cluster4 == 5, 5,
                                            ifelse(clinical_data$conv_cluster4 == 6, 6, 1)))
clinical_data$conv_cluster4 <- as.numeric(clinical_data$conv_cluster4)

# subset clinical data for metabolomics samples
clinical_data <- subset(clinical_data, clinical_data$sample_type=="PROTEOMICS/MET")
```

### Import MTHFR dataset
* Merge MTHFR data with clinical data
```{r}
alleles <- read.csv("/scratch/metabolomics/plasma-metabolomics-targeted/MTHFR/MTHFR.csv", header = TRUE)
# create MTHFR variable 
alleles$MTHFR <- paste(alleles$allele.1, alleles$allele.2, sep="")
alleles<-subset(alleles,alleles$MTHFR!="00") # remove 3
alleles$MTHFR<-as.factor(alleles$MTHFR)
table(alleles$MTHFR)
alleles <- alleles[,c(1,4)]

#merge MTHFR data with clinical data by participant_id
clin_alleles <- merge(clinical_data, alleles, by="participant_id")
length(unique(clin_alleles$participant_id)) # 1035 pts with MTHFR and met sample
clin_alleles$trajectory_group <- factor(clin_alleles$trajectory_group, levels = c(1,2,3,4,5))
```

### Import and structure targeted metabolomics data (data_use_targeted)
```{r}
### Import targeted phase 1-3 data
#targeted_raw <- read.csv("/impacc-study/Boston/plasma-metabolomics-targeted/2023-06-15/TargetedPlasmaMetabolomics-Counts.csv")
# DMAC QC'd targeted metabolomics
targeted_raw <- read.csv("/impacc-study/Boston/plasma-metabolomics-targeted/TargetedPlasmaMetabolomics-Counts-QCd.csv")
rownames(targeted_raw) <- targeted_raw[,1]
targeted <- pivot_longer(targeted_raw,cols = 2:69, names_to = "metabolite")

targeted_meta_raw <- read.csv("/impacc-study/Boston/plasma-metabolomics-targeted/2023-06-15/TargetedPlasmaMetabolomics-Metadata.csv")
#filter for passed QC
targeted_meta <- distinct(targeted_meta_raw) %>% filter(sample_status == "IMPACC sample assayed and passed QC")
paste0("Phases")
table(targeted_meta$phase)
paste0("Visits")
table(targeted_meta$visitnum)

#merge count and metadata
targeted_meta <- merge(targeted, targeted_meta, by = "guspec")

### Structure targeted data
clinical_df = clin_alleles
meta_data_raw <- targeted_meta
colnames(meta_data_raw)[1] <- "sample_id"
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"), all.x = TRUE)#,"phase"
met_data = subset(met_data, met_data$"trajectory_group" %in% c(1,2,3,4,5))
met_data <-  met_data[order(met_data$participant_id),]

targeted_metabolomics_counts_tmp <- targeted_raw[met_data$sample_id,]
targeted_metabolomics_counts_tmp <- drop_na(targeted_metabolomics_counts_tmp)
targeted_metabolomics_counts_tmp = targeted_metabolomics_counts_tmp[!duplicated(targeted_metabolomics_counts_tmp),]
rownames(targeted_metabolomics_counts_tmp) <- NULL
targeted_metabolomics_counts_tmp <- tibble::column_to_rownames(targeted_metabolomics_counts_tmp, var="guspec")

data_use_targeted <- targeted_metabolomics_counts_tmp %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = dplyr::select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                    sex,
                    bmi,
                    participant_id,event_location,conv_cluster4,event_type),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "discretized_admit_age_quantile", "sex", "bmi", "event_location","conv_cluster4","event_type")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))
data_use_targeted$conv_cluster4 = as.factor(data_use_targeted$conv_cluster4)
data_use_targeted <- merge(data_use_targeted,alleles,by="participant_id")
data_use_targeted$MTHFR = as.factor(data_use_targeted$MTHFR)
data_use_targeted$OUTCOME = data_use_targeted$MTHFR

data_use_targeted_acute = subset(data_use_targeted, data_use_targeted$event_type %in% c("Visit 1","Visit 2","Visit 4","Visit 6"))#,"Visit 3","Visit 5"

```

### Structure global metabolomics data (data_use_global)
```{r}
# identify metabolites in methionine pathway and glycine pathway
Metabolomics_rowfeature <- read.csv("/data/metabolomics/plasma-metabolomics-global/current/GlobalPlasmaMetabolomics-RowFeature.csv")
pathways <- Metabolomics_rowfeature %>% filter(SUB_PATHWAY %in% c('Methionine, Cysteine, SAM and Taurine Metabolism','Glycine, Serine and Threonine Metabolism','Creatine Metabolism') )

clinical_df = clin_alleles
meta_data_raw <- plasma_metabolomics_global_metadata
meta_data_raw <- rownames_to_column(meta_data_raw, "sample_id")
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"))#,"phase"
met_data <-  met_data[order(met_data$participant_id),]

  
plasma_metabolomics_global_counts_tmp <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_counts_tmp_long <- rownames_to_column(plasma_metabolomics_global_counts_tmp, var = "sample_id")
plasma_metabolomics_global_counts_tmp_long <- plasma_metabolomics_global_counts_tmp_long %>% pivot_longer(!sample_id, names_to = "MET_ID", values_to = "value")
plasma_metabolomics_global_counts_tmp_long <- subset(plasma_metabolomics_global_counts_tmp_long, MET_ID %in% pathways$MET_ID)
plasma_metabolomics_global_counts_tmp <- plasma_metabolomics_global_counts_tmp_long %>% pivot_wider(names_from = "MET_ID", values_from = "value")
plasma_metabolomics_global_counts_tmp <- column_to_rownames(plasma_metabolomics_global_counts_tmp, var = "sample_id")
colnames(plasma_metabolomics_global_counts_tmp) = pathways$CHEMICAL_NAME

data_use_global <- plasma_metabolomics_global_counts_tmp %>%
  rownames_to_column(var = "sample_id") %>%
  merge(y  = dplyr::select(clinical_data, 
                    sample_id, 
                    event_date,
                    trajectory_group,
                    enrollment_site,
                    discretized_admit_age_quantile,
                   admit_age,
                    sex,
                    bmi,
                    participant_id,event_location,conv_cluster4,event_type),
        by = "sample_id") %>%
  filter(!is.na(trajectory_group)) %>%
  filter(duplicated(participant_id) | duplicated(participant_id, fromLast = TRUE)) %>%
  mutate_if(is.character, as.factor) %>%
  mutate(trajectory_group = factor(trajectory_group)) %>%
  pivot_longer(cols = -c("sample_id", "event_date", "participant_id",
                         "enrollment_site", "trajectory_group",
                         "discretized_admit_age_quantile", 
                       "admit_age",
                         "sex", "bmi", "event_location","conv_cluster4","event_type")) %>%
  mutate(discretized_admit_age_quantile= factor(discretized_admit_age_quantile)) %>%
  filter(!is.na(value))
data_use_global$conv_cluster4 = as.factor(data_use_global$conv_cluster4)
data_use_global <- merge(data_use_global,alleles,by="participant_id")
data_use_global$MTHFR = as.factor(data_use_global$MTHFR)
data_use_global$OUTCOME = data_use_global$MTHFR

data_use_global_acute = subset(data_use_global, data_use_global$event_type %in% c("Visit 1","Visit 2","Visit 3","Visit 4","Visit 5","Visit 6"))
```

### Figure 1E
* Logistic regression  - TG123 vs TG45 based on baseline metabolites
```{r}
clinical_df = clinical_data
meta_data_raw <- plasma_metabolomics_global_metadata
meta_data_raw$sample_id = rownames(meta_data_raw)
met_data <- merge(meta_data_raw, clinical_df, by = c("sample_id"))#,"phase"
met_data = subset(met_data, met_data$"trajectory_group" %in% c(1,2,3,4,5))
met_data_baseline = subset(met_data, met_data$event_type == "Visit 1")
met_data_baseline <-  met_data_baseline[order(met_data_baseline$participant_id),]
met_data_baseline <- met_data_baseline %>%  mutate(OUTCOME = case_when(
    trajectory_group %in% c(1,2,3) ~ 0,
    trajectory_group %in% c(4,5)  ~ 1
    ))

met_data_baseline <- met_data_baseline %>% drop_na(OUTCOME)
OUTCOME = met_data_baseline$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data_baseline$sex)
y <- as.numeric (OUTCOME)
age <- as.numeric(met_data_baseline$admit_age)
bmi <- as.numeric(met_data_baseline$bmi) #numerical
race <- as.factor(met_data_baseline$race)
ethnicity <- as.factor(met_data_baseline$ethnicity)
location <- as.factor(met_data_baseline$event_location)
participant <- as.factor(met_data_baseline$participant_id)
age_quantile <- as.factor(met_data_baseline$discretized_admit_age_quantile)
phase <- as.factor(met_data_baseline$phase.x)
plate <- as.factor(met_data_baseline$plate)
type = as.factor(met_data_baseline$Type)
y <- as.factor (y)
plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data_baseline$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met,age, sex, participant, bmi,age_quantile,phase, plate)#,location,type
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile","phase","plate") #"location",,"type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
     lr<- glm (fmla, data=df,family=binomial) #
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  
  
  lr_output$coef <- coef
  lr_output$std <- std
 
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_123v45 <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_123v45$qval <= 0.05)
  output_all_123v45 = output_all


# order results based pvalue
select <- c("Methionine, Cysteine, SAM and Taurine Metabolism")
all_results_temp <- subset(lr_output_123v45, lr_output_123v45$SUB_PATHWAY %in% select)
all_results_temp = all_results_temp[order(all_results_temp$pval),]

all_results_temp = all_results_temp %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))
all_results_temp <- as.data.table(all_results_temp)

#  b <-all_results_sig[SUB_PATHWAY %in% select] 
 b <- all_results_temp
  if(dim(b)[1] ==0){
    return("No metabolites found!")
  }else{
    b$data <- ""
    if(dim(b)[1]>100){
      b <- b[order(qval)]
      b <- b[1:100,]
    }
  }
  
names(b)[2] = "name"
library(ggforestplot)

p_1= forestplot(
  df = b,
  estimate = coef,
  logodds = FALSE,
  #colour = type,
  shape = sig,
  title = "",
 se = std,
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19)) +
   theme(plot.title = element_text(size=12))
p_1

```


### Figure 4A
* PASC t-statistic deficit heatmap
```{r, fig.height=3, fig.width=3.5}
PRO_tstatistics <- data.frame(
  PRO = c("EQ-5D-5L","Health", "Cognitive", "Dyspnea", "Mental", "Impact", "Physical"),
  MIN = c(17.4, 12.75, 9.45, 7.9, 23.79, 18.82, 17.87),
  PHY   = c(-5.33, -2.69, -3.01, -2.84, -4.94, -1, -11.93),
  COG   = c(-6.33, -5.59, -8.8, -3.13, -18.5, -17.04, -4.31),
  MLT   = c(-18.01, -13.72, -3.36, -5.12, -21.88, -16.73, -15.64))
PRO_tstatistics <- PRO_tstatistics %>% column_to_rownames(var="PRO")
PRO_tstatistics <- as.matrix(PRO_tstatistics)

col_fun = circlize::colorRamp2(c(-20, 0, 20), c("blue", "white", "red"))
heat_pro <- Heatmap(PRO_tstatistics, name = "T-statistic", col = col_fun,column_names_rot = 45,
                    column_title = "PASC", column_title_side = "top",
                  #  row_title = "Survey", row_title_side = "right",
    show_column_names = TRUE, show_row_dend = FALSE, show_column_dend = FALSE,
    cluster_rows = FALSE, cluster_columns = FALSE, column_names_side = "top",
    column_names_gp = gpar(fontface = "bold"),
   column_title_gp = gpar(fontface = "bold"),
       #row_names_gp = gpar(fontface = "bold"),
   # row_title_gp = gpar(fontface = "bold")
    )
heat_pro
```

### Figure 4B
* Logistic regression AA vs. OTHERS (AG, GG) for MIN and OTHERS PASC groups based on baseline metabolites
```{r}
# subset for Visit 1
data_use_global_acute_v1 <- subset(data_use_global_acute, data_use_global_acute$event_type=="Visit 1")
# generate age quantile variable
data_use_global_acute_v1$discretized_admit_age_quantile <- ifelse(data_use_global_acute_v1$admit_age >= 18 & data_use_global_acute_v1$admit_age <= 35,"[18,35]",
                                                  ifelse(data_use_global_acute_v1$admit_age >= 36 & data_use_global_acute_v1$admit_age <= 51, "[36,51]", 
                                                         ifelse(data_use_global_acute_v1$admit_age >=52 & data_use_global_acute_v1$admit_age <= 66, "[52,66]",
                                                                ifelse(data_use_global_acute_v1$admit_age >=67 & data_use_global_acute_v1$admit_age <=81,"[67,81]",
                                                                       ifelse(data_use_global_acute_v1$admit_age >=82 & data_use_global_acute_v1$admit_age <=96,"[82,96]",NA)))))

# add PASC groups
data_use_global_acute_v1 <- merge(data_use_global_acute_v1,convalescent_pro_groups[,c(1,8)], by="participant_id", all.x=TRUE)
data_use_global_acute_v1$pro_group_labels <- factor(data_use_global_acute_v1$pro_group_labels, levels=c("MIN","PHY","COG","MLT"))

## PASC = MIN regression analysis 

# subset global V1 data for PASC=="MIN"
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$pro_group_labels=="MIN")

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$MTHFR=="AA","AA","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "OTHER" ~ 0,
    outcome == "AA" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)

plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AAvOTHER_min <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AAvOTHER_min$qval <= 0.05) # 258 sig mets
  output_all_AAvOTHER_min = output_all
```
```{r}
## PASC = OTHERS regression analysis

# subset global V1 data for PASC=="combined"
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$pro_group_labels %in% c("PHY","COG","MLT"))

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$MTHFR=="AA","AA","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "OTHER" ~ 0,
    outcome == "AA" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)

plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AAvOTHER_other <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AAvOTHER_other$qval <= 0.05) # 224 sig mets
  output_all_AAvOTHER_other = output_all
```
```{r}
## AA vs OTHERS forest plot - 'MIN' & 'OTHER' PASC groups

# Combine logistic regression results for PASC==MIN and PASC==OTHER (combined)
lr_output_AAvOTHER_min = merge(lr_output_AAvOTHER_min, output_all_AAvOTHER_min, by.x = "Compound_ID",by.y = "outc")
lr_output_AAvOTHER_min$type = "MIN"
lr_output_AAvOTHER_min = lr_output_AAvOTHER_min %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05  ~ "Not significant"
    ))

lr_output_AAvOTHER_other = merge(lr_output_AAvOTHER_other, output_all_AAvOTHER_other, by.x = "Compound_ID",by.y = "outc")
lr_output_AAvOTHER_other$type = "OTHER"
lr_output_AAvOTHER_other = lr_output_AAvOTHER_other %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))



lr_output_AAvOTHER_min <- lr_output_AAvOTHER_min[match(lr_output_AAvOTHER_other$Name, lr_output_AAvOTHER_min$Name), ]

all_results_sig = rbind(lr_output_AAvOTHER_min, lr_output_AAvOTHER_other)
all_results_sig$type = factor(all_results_sig$type, levels = c("MIN", "OTHER"))

# order results based on OTHER pvalue
select <- c("Methionine, Cysteine, SAM and Taurine Metabolism", "Glycine, Serine and Threonine Metabolism")
all_results_temp <- subset(all_results_sig, all_results_sig$SUB_PATHWAY %in% select)
min <- subset(all_results_temp, all_results_temp$type=="MIN")
other <- subset(all_results_temp, all_results_temp$type=="OTHER")
other = other[order(other$pval),]
min <- min[match(other$Name, min$Name), ]

all_results_sig <- rbind(other,min)

#  b <-all_results_sig[SUB_PATHWAY %in% select] 
 b <- all_results_sig
  if(dim(b)[1] ==0){
    return("No metabolites found!")
  }else{
    b$data <- ""
    if(dim(b)[1]>100){
      b <- b[order(qval)]
      b <- b[1:100,]
    }
  }
  
names(b)[2] = "name"
library(ggforestplot)

p_1= forestplot(
  df = b,
  estimate = coef,
  logodds = FALSE,
  colour = sig,
  shape = type,
  title = "",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19)) +
   theme(plot.title = element_text(size=12))
p_1

```

### Figure 4C
* Logistic regression MIN vs. OTHERS (combined) based on baseline metabolites for each MTHFR allele group
```{r}
## Allele = aa regression analysis 

# subset for MTHFR==AA
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$MTHFR=="AA")

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$pro_group_labels=="MIN","MIN","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "MIN" ~ 0,
    outcome == "OTHER" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)

plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_MINvOTHER_aa <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_MINvOTHER_aa$qval <= 0.05) # 322 sig mets
  output_all_MINvOTHER_aa = output_all
```
```{r}
## Allele = gg regression analysis

# subset global V1 data for MTHFR==GG
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$MTHFR=="GG")

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$pro_group_labels=="MIN","MIN","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "MIN" ~ 0,
    outcome == "OTHER" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)

plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_MINvOTHER_gg <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_MINvOTHER_gg$qval <= 0.05) # 256 sig mets
  output_all_MINvOTHER_gg = output_all
```
```{r}
## Allele = ag regression analysis

# subset global V1 data for MTHFR==AG
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$MTHFR=="AG")

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$pro_group_labels=="MIN","MIN","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "MIN" ~ 0,
    outcome == "OTHER" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)

plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_MINvOTHER_ag <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_MINvOTHER_ag$qval <= 0.05) # 286 sig mets
  output_all_MINvOTHER_ag = output_all
```
```{r}
### OTHER vs MIN forest plot

# combine logistic regression outputs for each MTHFR allele
lr_output_MINvOTHER_ag = merge(lr_output_MINvOTHER_ag, output_all_MINvOTHER_ag, by.x = "Compound_ID",by.y = "outc")
lr_output_MINvOTHER_ag$type = "AG"
lr_output_MINvOTHER_ag = lr_output_MINvOTHER_ag %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05  ~ "Not significant"
    ))

lr_output_MINvOTHER_gg = merge(lr_output_MINvOTHER_gg, output_all_MINvOTHER_gg, by.x = "Compound_ID",by.y = "outc")
lr_output_MINvOTHER_gg$type = "GG"
lr_output_MINvOTHER_gg = lr_output_MINvOTHER_gg %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))
lr_output_MINvOTHER_aa = merge(lr_output_MINvOTHER_aa, output_all_MINvOTHER_aa, by.x = "Compound_ID",by.y = "outc")
lr_output_MINvOTHER_aa$type = "AA"
lr_output_MINvOTHER_aa = lr_output_MINvOTHER_aa %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))

all_results_sig = rbind(lr_output_MINvOTHER_gg, lr_output_MINvOTHER_ag, lr_output_MINvOTHER_aa)
all_results_sig$type = factor(all_results_sig$type, levels = c("GG", "AG", "AA"))

  b <-all_results_sig[SUB_PATHWAY %in% c("Methionine, Cysteine, SAM and Taurine Metabolism")] 
  if(dim(b)[1] ==0){
    return("No metabolites found!")
  }else{
    b$data <- "Methionine, Cysteine, SAM and Taurine Metabolism"
    if(dim(b)[1]>100){
      b <- b[order(qval)]
      b <- b[1:100,]
    }
  }
  
    p <- b%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = sig), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = sig, shape = type), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Methionine, Cysteine, SAM and Taurine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
 
  names(b)[2] = "name"
library(ggforestplot)
p_1= forestplot(
  df = b,
  estimate = coef,
  logodds = FALSE,
  colour = sig,
  shape = type,
  title = "Methionine, Cysteine, SAM and Taurine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19,15)) +
   theme(plot.title = element_text(size=12))
p_1


b <-all_results_sig[SUB_PATHWAY %in% c("Glycine, Serine and Threonine Metabolism")] 
  if(dim(b)[1] ==0){
    return("No metabolites found!")
  }else{
    b$data <- "Glycine, Serine and Threonine Metabolism"
    if(dim(b)[1]>100){
      b <- b[order(qval)]
      b <- b[1:100,]
    }
  }
  
    p <- b%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = sig), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = sig, shape = type), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Glycine, Serine and Threonine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
 
  names(b)[2] = "name"
library(ggforestplot)
p_2= forestplot(
  df = b,
  estimate = coef,
  logodds = FALSE,
  colour = sig,
  shape = type,
  title = "Glycine, Serine and Threonine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19,15)) +
   theme(plot.title = element_text(size=12))
p_2


select <- c("Methionine, Cysteine, SAM and Taurine Metabolism", "Glycine, Serine and Threonine Metabolism")
b_combined <-all_results_sig[SUB_PATHWAY %in% select] 
  if(dim(b_combined)[1] ==0){
    return("No metabolites found!")
  }else{
    b_combined$data <- "Glycine, Serine and Threonine Metabolism"
    if(dim(b_combined)[1]>100){
      b_combined <- b_combined[order(qval)]
      b_combined <- b_combined[1:100,]
    }
  }

b_combined = b_combined[order(b_combined$pval),]
names(b_combined)[2] = "name"
library(ggforestplot)
p_3= forestplot(
  df = b_combined,
  estimate = coef,
  logodds = FALSE,
  colour = sig,
  shape = type,
  title = "",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19,15)) +
   theme(plot.title = element_text(size=12)) #+ facet_grid(. ~ type, scales = "free") + scale_color_manual("Legend", values = c("FDR significant" = "red", "Not significant" = "ivory4", "Raw significant" = "lightsalmon")) + theme(legend.position = "bottom")
p_3

```


### Targeted vs Global metabolite correlations

#### Targeted vs global methionine pearson correlation plot 
```{r, eval=F}
#subset for methionine
global_methionine <- subset(data_use_global_acute, data_use_global_acute$name == "methionine")
targeted_methionine <- subset(data_use_targeted_acute, data_use_targeted_acute$name == "methionine")

#match targeted and global methionine values by participants and visit
global_methionine$UID <- paste(global_methionine$participant_id, global_methionine$event_type, sep = "-")
targeted_methionine$UID <- paste(targeted_methionine$participant_id, targeted_methionine$event_type, sep = "-")
no_match <- anti_join(targeted_methionine, global_methionine, by = "UID") 
#print(no_match, n=Inf) # 1 UID in targeted that is not in global
global_methionine <- subset(global_methionine, UID %in% targeted_methionine$UID)
targeted_methionine <- subset(targeted_methionine, UID %in% global_methionine$UID)
length(unique(targeted_methionine$UID))
length(unique(global_methionine$UID))

global_methionine <- global_methionine[,c(17,14)]
colnames(global_methionine)[2] <- "global_value"
targeted_methionine <- targeted_methionine[,c(16,13)]
colnames(targeted_methionine)[2] <- "targeted_value"

methionine <- merge(global_methionine,targeted_methionine, by = "UID")

#log10 values 
methionine$log10_global_value <- log10(methionine$global_value)
methionine$log10_targeted_value <- log10(methionine$targeted_value)

corplot_methionine <- ggscatter(methionine, x = "global_value", y = "targeted_value",
   add = "reg.line",  # Add regressin line
   add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
   conf.int = TRUE, # Add confidence interval
   xlab = "Global Methionine Value", ylab = "Targeted Methionine Value"
   )
# Add correlation coefficient
corplot_methionine + stat_cor(method = "pearson",p.accuracy = 0.001, r.accuracy = 0.01)

methionine_unique <- methionine
methionine_unique$participant_id <- substr(methionine_unique$UID, 0, 8)
length(unique(methionine_unique$participant_id)) # 74 pts
length(unique(methionine_unique$UID)) # 293 matched samples
```

#### Targeted vs global creatinine pearson correlation plot 
```{r, eval=F}
#subset for creatinine
global_creatinine <- subset(data_use_global_acute, data_use_global_acute$name == "creatinine")
targeted_creatinine <- subset(data_use_targeted_acute, data_use_targeted_acute$name == "creatinine")

#match targeted and global methionine values by participants and visit
global_creatinine$UID <- paste(global_creatinine$participant_id, global_creatinine$event_type, sep = "-")
targeted_creatinine$UID <- paste(targeted_creatinine$participant_id, targeted_creatinine$event_type, sep = "-")
no_match <- anti_join(targeted_creatinine, global_creatinine, by = "UID") 
#print(no_match, n=Inf) # 1 UID in targeted that is not in global
global_creatinine <- subset(global_creatinine, UID %in% targeted_creatinine$UID)
targeted_creatinine <- subset(targeted_creatinine, UID %in% global_creatinine$UID)
length(unique(targeted_creatinine$UID))
length(unique(global_creatinine$UID))

global_creatinine <- global_creatinine[,c(17,14)]
colnames(global_creatinine)[2] <- "global_value"
targeted_creatinine <- targeted_creatinine[,c(16,13)]
colnames(targeted_creatinine)[2] <- "targeted_value"

creatinine <- merge(global_creatinine,targeted_creatinine, by = "UID")

#log10 values 
creatinine$log10_global_value <- log10(creatinine$global_value)
creatinine$log10_targeted_value <- log10(creatinine$targeted_value)

corplot_creatinine <- ggscatter(creatinine, x = "global_value", y = "targeted_value",
   add = "reg.line",  # Add regressin line
   add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
   conf.int = TRUE, # Add confidence interval
   xlab = "Global Creatinine Value", ylab = "Targeted Creatinine Value"
   )
# Add correlation coefficient
corplot_creatinine + stat_cor(method = "pearson",p.accuracy = 0.001, r.accuracy = 0.01)

creatinine_unique <- creatinine
creatinine_unique$participant_id <- substr(creatinine_unique$UID, 0, 8)
length(unique(creatinine_unique$participant_id)) # 74 pts
length(unique(creatinine_unique$UID)) # 293 matched samples
```

#### Targeted vs global creatine pearson correlation plot 
```{r, eval=F}
#subset for creatine
global_creatine <- subset(data_use_global_acute, data_use_global_acute$name == "creatine")
targeted_creatine <- subset(data_use_targeted_acute, data_use_targeted_acute$name == "creatine")

#match targeted and global methionine values by participants and visit
global_creatine$UID <- paste(global_creatine$participant_id, global_creatine$event_type, sep = "-")
targeted_creatine$UID <- paste(targeted_creatine$participant_id, targeted_creatine$event_type, sep = "-")
no_match <- anti_join(targeted_creatine, global_creatine, by = "UID") 
#print(no_match, n=Inf) # 1 UID in targeted that is not in global
global_creatine <- subset(global_creatine, UID %in% targeted_creatine$UID)
targeted_creatine <- subset(targeted_creatine, UID %in% global_creatinine$UID)
length(unique(targeted_creatine$UID))
length(unique(global_creatine$UID))

global_creatine <- global_creatine[,c(17,14)]
colnames(global_creatine)[2] <- "global_value"
targeted_creatine <- targeted_creatine[,c(16,13)]
colnames(targeted_creatine)[2] <- "targeted_value"

creatine <- merge(global_creatine,targeted_creatine, by = "UID")

#log10 values 
creatine$log10_global_value <- log10(creatine$global_value)
creatine$log10_targeted_value <- log10(creatine$targeted_value)

corplot_creatinine <- ggscatter(creatine, x = "global_value", y = "targeted_value",
   add = "reg.line",  # Add regressin line
   add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
   conf.int = TRUE, # Add confidence interval
   xlab = "Global Creatine Value", ylab = "Targeted Creatine Value"
   )
# Add correlation coefficient
corplot_creatinine + stat_cor(method = "pearson",p.accuracy = 0.001, r.accuracy = 0.01)

creatine_unique <- creatine
creatine_unique$participant_id <- substr(creatine_unique$UID, 0, 8)
length(unique(creatine_unique$participant_id)) # 74 pts
length(unique(creatine_unique$UID)) # 293 matched samples
```

### Supplementary Figure 1D
* Ordinal regression - TGs (1-5) and mortality
```{r}
# import baseline global ordinal regression results from JC
ordinal <- read.csv("/scratch/metabolomics/plasma-metabolomics-global/scripts/deep_dive_met/ordinal_v1.csv")

ordinal$type = "ordinal"
ordinal = ordinal %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))
ordinal <- as.data.table(ordinal)

b_Methionine <- ordinal[SUB_PATHWAY %in% c("Methionine, Cysteine, SAM and Taurine Metabolism") ]
  if(dim(b_Methionine)[1] ==0){
    return("No metabolites found!")
  }else{
    b_Methionine$data <- "Methionine, Cysteine, SAM and Taurine Metabolism"
    if(dim(b_Methionine)[1]>100){
      b_Methionine <- b_Methionine[order(qval)]
      b_Methionine <- b_Methionine[1:100,]
    }
  }
    p_Methionine <- b_Methionine%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Methionine, Cysteine, SAM and Taurine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
    names(b_Methionine)[2] = "name"
    
library(ggforestplot)
 p_3= forestplot(
  df = b_Methionine,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  se = std,
  title = "Methionine, Cysteine, SAM and Taurine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05, ci=0.95
) +scale_shape_manual(values=c(17,19))
p_3
```

### Distribution of trajectory groups and mortality across MTHFR alleles
```{r}
# subset for unique participants
clin_alleles_supp = clin_alleles[!duplicated(clin_alleles$participant_id),]
clin_alleles_supp$MTHFR <- factor(clin_alleles_supp$MTHFR, levels = c("GG","AG","AA"))

## TGs 
MTHFR_TG <- clin_alleles_supp %>%
  dplyr::count(MTHFR, trajectory_group) %>%       
  group_by(MTHFR) %>%
  mutate(pct= prop.table(n) * 100) %>%
  ggplot() + aes(MTHFR, pct, fill=trajectory_group) +
  scale_fill_manual(values=c('#639A21', '#39828C', '#0072B2', '#E69F00', '#9C3418')) +
  geom_bar(stat="identity") +
  ylab("Proportion of Participants") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct),"%")),
            position=position_stack(vjust=0.5)) +
  ggtitle("TG distribution across MTHFR alleles") +
  theme_bw() 
MTHFR_TG

table(clin_alleles_supp$MTHFR)
table(clin_alleles_supp$MTHFR, clin_alleles_supp$trajectory_group)

# chi square test for homogeneity
conttable <- table(clin_alleles_supp$MTHFR, clin_alleles_supp$trajectory_group)
chisq.test(conttable)


## MTHFR alleles
clin_alleles_supp$diedever <- ifelse(clin_alleles_supp$diedever == 0, "No","Yes")
clin_alleles_supp$diedever <- factor(clin_alleles_supp$diedever, levels = c("No","Yes"))

MTHFR_mortality <- clin_alleles_supp %>%
  dplyr::count(MTHFR, diedever) %>%       
  group_by(MTHFR) %>%
  mutate(pct= prop.table(n) * 100) %>%
  ggplot() + aes(MTHFR, pct, fill=diedever) +
  scale_fill_manual(values=c("royalblue3", "gray")) +
  geom_bar(stat="identity") +
  ylab("Proportion of Participants") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct),"%")),
            position=position_stack(vjust=0.5)) +
  ggtitle("Mortality distribution across MTHFR alleles") +
  theme_bw() +
  theme(panel.grid = element_blank())
MTHFR_mortality

table(clin_alleles_supp$MTHFR)
table(clin_alleles_supp$MTHFR, clin_alleles_supp$diedever)

conttable2 <- table(clin_alleles_supp$MTHFR, clin_alleles_supp$diedever)
chisq.test(conttable2)
```

### Hazard ratio analysis
* Univariate cox hazards model (status = diedever, predictor = MTHFR)
* Hazards ratio forest plot
```{r}
# distinct participant-level data (clin_alleles_supp)
clin_alleles_supp$death_date <- as.numeric(clin_alleles_supp$death_date)
clin_alleles_supp$death_date <- ifelse(is.na(clin_alleles_supp$death_date), 0, clin_alleles_supp$death_date)
clin_alleles_supp$death_date <- ifelse(clin_alleles_supp$death_date==-99, 0, clin_alleles_supp$death_date)
clin_alleles_supp$diedever <- ifelse(clin_alleles_supp$diedever == "Yes", 1,0)

clin_alleles_supp$MTHFR <- factor(clin_alleles_supp$MTHFR, levels = c("GG", "AG", "AA"))

univ_model <- coxph(Surv(death_date, diedever)~ MTHFR, data = clin_alleles_supp)
summary(univ_model)
# HR forest plot
MTHFR_HR <- ggforest(univ_model, data = clin_alleles_supp)

# survival curve 
#allele_fit <- survfit(Surv(death_date, diedever) ~ MTHFR, data=clin_alleles_supp)
#allele_fit
#ggsurvplot(allele_fit, data = clin_alleles_supp, pval = TRUE)

print(paste0("N=",length(unique(clin_alleles_supp$participant_id))))
clin_alleles_supp %>%
  group_by(MTHFR) %>%
  distinct(participant_id) %>%
  dplyr::summarise(n = n())

```

### eBV distribution across MTHFR alleles and TGs
```{ r}
# import eBV dataset
ebv <- read.csv("/scratch/metabolomics/plasma-metabolomics-targeted/MTHFR/EBV.csv", header = TRUE)
colnames(ebv)[1] <- "participant_id"

# subset clinical data for only participants with ebv data
clin_stanford <- subset(clinical_data, participant_id %in% ebv$participant_id)
# merge with ebv df
clin_stanford_ebv <- merge(clin_stanford, ebv, by = "participant_id")

clin_stanford_ebv$EBV_status <- factor(clin_stanford_ebv$EBV_status, levels = c("+","-"))
clin_stanford_ebv$trajectory_group <- factor(clin_stanford_ebv$trajectory_group, levels = c("1","2","3","4","5"))

clin_stanford_ebv = clin_stanford_ebv[!duplicated(clin_stanford_ebv$participant_id),]

# frequency barplot
clin_stanford_ebv %>%
  dplyr::count(trajectory_group, EBV_status) %>%       
  group_by(trajectory_group) %>%
  mutate(pct= prop.table(n) * 100) %>%
  ggplot() + aes(trajectory_group, pct, fill=EBV_status) +
  scale_fill_manual(values=c("royalblue3", "gray")) +
  geom_bar(stat="identity") +
  ylab("Proportion of Participants") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct),"%")),
            position=position_stack(vjust=0.5)) +
  ggtitle("eBV distribution across TGs") +
  theme_bw() +
  theme(panel.grid = element_blank())

cont.table1 <- table(clin_stanford_ebv$trajectory_group,clin_stanford_ebv$EBV_status)
cont.table1
# chi-square test for homogeneity
chisq.test(cont.table1)



# subset MTHFR data for participants in ebv (Stanford)
alleles_ebv <- subset(alleles, participant_id %in% ebv$participant_id)
# merge clinical ebv df with MTHFR alleles
clin_stanford_ebv_alleles <- merge(clin_stanford_ebv, alleles_ebv, by = "participant_id")
clin_stanford_ebv_alleles$MTHFR <- factor(clin_stanford_ebv_alleles$MTHFR, levels=c("GG","AG","AA"))

# frequency barplot
clin_stanford_ebv_alleles %>%
  dplyr::count(MTHFR, EBV_status) %>%       
  group_by(MTHFR) %>%
  mutate(pct= prop.table(n) * 100) %>%
  ggplot() + aes(MTHFR, pct, fill=EBV_status) +
  scale_fill_manual(values=c("royalblue3", "gray")) +
  geom_bar(stat="identity") +
  ylab("Proportion of Participants") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct),"%")),
            position=position_stack(vjust=0.5)) +
  ggtitle("eBV distribution across MTHFR alleles") +
  theme_bw() +
  theme(panel.grid = element_blank())

cont.table2 <- table(clin_stanford_ebv_alleles$MTHFR, clin_stanford_ebv_alleles$EBV_status)
cont.table2
# chi-square test for homogeneity
chisq.test(cont.table2)
```

### Distribution of 'anti-RBD IgG AUC' and 'anti-spike IgG' values across TGs
```{ r}
# import bcell data
bcell <- read.csv("/data/serum-rbd-abtiters/current/serum-rbd-abtiters-Counts.csv")
#log-transform 
bcell$log2_AUC.RBD.IgG <- log2(bcell$AUC.RBD.IgG)
bcell$log2_AUC.Spike.IgG <- log2(bcell$AUC.Spike.IgG)

#merge with complete clinical data (all sample types)
clinical <- data_env[[ "clinical_data" ]]
bcell_clin <- merge(bcell, clinical, by = "sample_id")
#match for pts in metabolomics clinical data (i.e. clinical_data)
bcell_clin <- subset(bcell_clin, participant_id %in% clinical_data$participant_id)
bcell_clin$trajectory_group <- factor(bcell_clin$trajectory_group, levels = c(1,2,3,4,5))

# anti-RBD IgG AUC plot
my_comparisons <- list( c("1", "2"), c("1", "3"), c("1", "4"), c("1", "5"), c("2", "3"), c("2", "4"), c("2", "5"), c("3", "4"), c("3", "5"), c("4", "5") )
ggplot(data = bcell_clin, aes(x=trajectory_group, y=log2_AUC.RBD.IgG)) + 
  geom_boxplot(aes(fill=trajectory_group)) +
  stat_compare_means(comparisons = my_comparisons, label="p.signif", method = "wilcox.test") + 
  scale_fill_manual(values=c("#639A21", "#39828C", "#0072B2", "#E69F00", "#9C3418")) +
  labs(title="anti-RBD IgG", y="log2(AUC)") + theme_classic()

length(unique(bcell_clin$participant_id)) #1104 pts
bcell_clin_sub_unique <- distinct(bcell_clin, participant_id, .keep_all = TRUE)
table(bcell_clin_sub_unique$trajectory_group)


# anti-spike IgG plot
sum(is.na(bcell_clin$log2_AUC.Spike.IgG)) #8 NA values - remove
bcell_clin <- bcell_clin %>% drop_na(log2_AUC.Spike.IgG)

ggplot(data = bcell_clin, aes(x=trajectory_group, y=log2_AUC.Spike.IgG)) + 
  geom_boxplot(aes(fill=trajectory_group)) +
  stat_compare_means(comparisons = my_comparisons, label="p.signif", method = "wilcox.test") + 
  scale_fill_manual(values=c("#639A21", "#39828C", "#0072B2", "#E69F00", "#9C3418")) +
  labs(title="anti-spike IgG", y="log2(AUC)") + theme_classic()

length(unique(bcell_clin$participant_id)) #1103 pts
bcell_clin_sub_unique <- distinct(bcell_clin, participant_id, .keep_all = TRUE)
table(bcell_clin_sub_unique$trajectory_group)
```

### Supplmentary Figure 4C (LRT results) & Fig3E (ROC curves)
* Likelihood ratio test (LRT)
* Null model: diedever ~ MTHFR
* Full model: diedever ~ MTHFR + SAH (baseline) + methionine-sulfoxide (baseline) + methionine (baseline)
* Compare goodness of fit of models using ROC curves
```{r}
# merge allele data (clin_alleles) with global V1 metabolite data for SAH (X197), methionine-sulfoxide (X100000039), and methionine (X415) (merge by sample_id)
counts <- plasma_metabolomics_global_counts %>% rownames_to_column(var = "sample_id")
clin_alleles_temp <- merge(clin_alleles, counts[,c(1,15,44,144)], by="sample_id", all.x=TRUE)
clin_alleles_temp <- clin_alleles_temp[,c(1,7,122,181,183:186)]
clin_alleles_temp <- subset(clin_alleles_temp, clin_alleles_temp$event_type=="Visit 1")
clin_alleles_temp$diedever <- factor(clin_alleles_temp$diedever, levels = c(0,1))

# subset for samples with data for metabolites at V1
clin_alleles_temp <- clin_alleles_temp %>%
  filter(!is.na(X197) & !is.na(X100000039) & !is.na(X415))

table(clin_alleles_temp$diedever)

clin_alleles_temp$MTHFR <- factor(clin_alleles_temp$MTHFR, levels = c("GG","AG","AA"))

# build model based on alleles only
model_alleles <- glm(diedever ~ MTHFR, data = clin_alleles_temp, family=binomial)
summary(model_alleles)
# build model based on alleles and metabolites
model_full <- glm(diedever ~ MTHFR + X197 + X100000039 + X415, data = clin_alleles_temp, family=binomial)
summary(model_full)
# perform likelihood ratio test
likelihood_ratio_test <- lrtest(model_alleles, model_full)
likelihood_ratio_test 
# Interpretation of likelihood ratio test results
if (likelihood_ratio_test$"Pr(>Chisq)"[2] < 0.05) 
{
  cat("Reject the null hypothesis. The full model is significantly better than 
        the null model.\n")
} else {
  cat("Fail to reject the null hypothesis. The null model is sufficient.\n")
}
# Calculate AIC values
AIC_null <- AIC(model_alleles)
AIC_full <- AIC(model_full)
# Calculate log-likelihood values
loglik_null <- logLik(model_alleles)
loglik_full <- logLik(model_full)
# Comparison of AIC and log-likelihood values
cat("AIC value (null model):", AIC_null, "\n")
cat("AIC value (full model):", AIC_full, "\n")
cat("Log-likelihood value (null model):", loglik_null, "\n")
cat("Log-likelihood value (full model):", loglik_full, "\n")

library(stargazer)
stargazer(model_alleles, model_full, title="Comparison of Nested Logistic Regression Models", type = "html", align=TRUE, out="/scratch/metabolomics/plasma-metabolomics-targeted/MTHFR/LRT_models.html",
           dep.var.labels=c("Death"),
          covariate.labels=c("MTHFR AG allele","MTHFR AA allele", 
                             "SAH","methionine-sulfoxide","methionine", "Reference (MTHFR GG allele)"),
          single.row = TRUE, ci=TRUE, ci.level=0.95, column.labels=c("MTHFR alleles only","MTHFR alleles + baseline metabolites"))
```
```{r}
### plot ROC curves

# predicted probabilities for both models
pred_alleles <- predict(model_alleles, type = "response")
pred_full <- predict(model_full, type = "response")

# roc
roc1 <- roc(clin_alleles_temp$diedever, pred_alleles)  #, smooth=TRUE
roc2 <- roc(clin_alleles_temp$diedever, pred_full)   #, smooth=TRUE
# print AUC values
auc(roc1)
auc(roc2)

# plot roc curves
plot(roc1, col = "blue", main = "ROC Curve Comparison", lwd = 2, xlab = "False Positive Rate (1 - Specificity)", ylab = "True Positive Rate (Sensitivity)")
lines(roc2, col = "red", lwd = 2)
legend("bottomright", legend = c("MTHFR alleles only", "MTHFR alleles + baseline metabolites"), col = c("blue", "red"), lwd = 2)

```

### Second LRT comparison
* Likelihood ratio test (LRT)
* Null model: diedever ~ SAH (baseline) + methionine-sulfoxide (baseline) + methionine (baseline)
* Full model: diedever ~ SAH (baseline) + methionine-sulfoxide (baseline) + methionine (baseline) + MTHFR 
* Compare goodness of fit of models using ROC curves
```{r}
# build model based on metabolites only
model_mets <- glm(diedever ~ X197 + X100000039 + X415, data = clin_alleles_temp, family=binomial)
summary(model_mets)
# build model based on alleles and metabolites
model_full <- glm(diedever ~ X197 + X100000039 + X415 + MTHFR, data = clin_alleles_temp, family=binomial)
summary(model_full)
# perform likelihood ratio test
likelihood_ratio_test <- lrtest(model_mets, model_full)
likelihood_ratio_test 
# Interpretation of likelihood ratio test results
if (likelihood_ratio_test$"Pr(>Chisq)"[2] < 0.05) 
{
  cat("Reject the null hypothesis. The full model is significantly better than 
        the null model.\n")
} else {
  cat("Fail to reject the null hypothesis. The null model is sufficient.\n")
}
# Calculate AIC values
AIC_null <- AIC(model_mets)
AIC_full <- AIC(model_full)
# Calculate log-likelihood values
loglik_null <- logLik(model_mets)
loglik_full <- logLik(model_full)
# Comparison of AIC and log-likelihood values
cat("AIC value (null model):", AIC_null, "\n")
cat("AIC value (full model):", AIC_full, "\n")
cat("Log-likelihood value (null model):", loglik_null, "\n")
cat("Log-likelihood value (full model):", loglik_full, "\n")


stargazer(model_mets, model_full, title="Comparison of Nested Logistic Regression Models", type = "text", align=TRUE,
           dep.var.labels=c("Death"),
          covariate.labels=c("SAH","methionine-sulfoxide","methionine",
                             "MTHFR AG allele","MTHFR AA allele", 
                             "Reference (MTHFR GG allele)"),
          single.row = TRUE, ci=TRUE, ci.level=0.95, 
         column.labels=c("Baseline mets only","Baseline mets + MTHFR alleles")
         )

```
```{r}
### plot ROC curves

# predicted probabilities for both models
pred_mets <- predict(model_mets, type = "response")
pred_full <- predict(model_full, type = "response")

# roc
roc3 <- roc(clin_alleles_temp$diedever, pred_mets)  #, smooth=TRUE
roc4 <- roc(clin_alleles_temp$diedever, pred_full)   #, smooth=TRUE
# print AUC values
auc(roc3)
auc(roc4)

# plot roc curves
plot(roc3, col = "blue", main = "ROC Curve Comparison", lwd = 2, xlab = "False Positive Rate (1 - Specificity)", ylab = "True Positive Rate (Sensitivity)")
lines(roc4, col = "red", lwd = 2)
legend("bottomright", legend = c("Baseline metabolites only", "Baseline metabolites + MTHFR alleles"), col = c("blue", "red"), lwd = 2)
```

### Supplementary Figure 5
* Baseline (V1) methionine pathway metabolite levels by PASC groups (MIN, COMBINED) and MTHFR allele groups
```{r, fig.keep='none'}
# add pasc groups
data_use_global_acute_conv <- merge(data_use_global_acute,convalescent_pro_groups[,c(1,8)], by="participant_id", all.x=TRUE)
data_use_global_acute_conv <- subset(data_use_global_acute_conv, !is.na(data_use_global_acute_conv$pro_group_labels))
data_use_global_acute_conv$pasc_binary <- ifelse(data_use_global_acute_conv$pro_group_labels=="MIN","MIN","OTHER")
data_use_global_acute_conv$pasc_binary <- factor(data_use_global_acute_conv$pasc_binary, levels=c("MIN","OTHER"))
data_use_global_acute_conv$pro_group_labels <- factor(data_use_global_acute_conv$pro_group_labels, levels=c("MIN","PHY","COG","MLT"))
data_use_global_acute_conv$MTHFR <- factor(data_use_global_acute_conv$MTHFR, levels=c("GG","AG","AA"))

# table - MTHFR allele and PASC - pt level
data_use_global_acute_conv_unique <-distinct(data_use_global_acute_conv, participant_id, .keep_all = TRUE)
print(paste0("N=",length(data_use_global_acute_conv_unique$participant_id))) # 415 pts
table(data_use_global_acute_conv_unique$MTHFR, data_use_global_acute_conv_unique$pasc_binary)

# subset for only V1
data_use_global_sub <- subset(data_use_global_acute_conv, data_use_global_acute_conv$event_type=="Visit 1")

# subset for metabolites in methionine pathway
rowfeature_methionine <- subset(Metabolomics_rowfeature, Metabolomics_rowfeature$SUB_PATHWAY=="Methionine, Cysteine, SAM and Taurine Metabolism") 
names <- rowfeature_methionine$CHEMICAL_NAME
data_use_global_methionine <- subset(data_use_global_sub, data_use_global_sub$name %in% names)


### methionine sulfoxide 
data_use_global_sulfoxide <- subset(data_use_global_sub, data_use_global_sub$name=="methionine sulfoxide")

stat.test <- data_use_global_sulfoxide %>%
  group_by(pasc_binary) %>%
  wilcox_test(value ~ MTHFR) %>%
  add_y_position() %>%
  add_x_position(x = "pasc_binary", dodge = 1)
stat.test <- 
  rstatix::add_significance(
  stat.test,
  p.col = "p",
  output.col = "p.signif",
  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
  symbols = c("****", "***", "**", "*", "ns")
)

plot_sulfoxide_pasc <- ggplot(data_use_global_sulfoxide, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
  labs(x="PASC Group", y="methionine sulfoxide") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
 # scale_fill_manual(values = c("#999999", "#E69F00"))
plot_sulfoxide_pasc


### S-adenosylhomocysteine (SAH) 
#data_use_global_SAH <- subset(data_use_global_sub, data_use_global_sub$name=="S-adenosylhomocysteine (SAH)")

#stat.test <- data_use_global_SAH %>%
 # group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
#  rstatix::add_significance(
#  stat.test,
#  p.col = "p",
#  output.col = "p.signif",
#  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
#  symbols = c("****", "***", "**", "*", "ns")
#)

#plot_SAH_pasc <- ggplot(data_use_global_SAH, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="S-adenosylhomocysteine (SAH)") + theme_bw() + stat_pvalue_manual(stat.test,  label = #"p.signif")
#plot_SAH_pasc


### 2-hydroxy-4-(methylthio)butanoic acid 
#data_use_global_butonic <- subset(data_use_global_sub, data_use_global_sub$name=="2-hydroxy-4-(methylthio)butanoic acid")

#stat.test <- data_use_global_butonic %>%
#  group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
#  add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
#  rstatix::add_significance(
#  stat.test,
#  p.col = "p",
#  output.col = "p.signif",
#  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
#  symbols = c("****", "***", "**", "*", "ns")
#)

#plot_butonic_pasc <- ggplot(data_use_global_butonic, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
#  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="2-hydroxy-4-(methylthio)butanoic acid") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_butonic_pasc


### betaine
#data_use_global_betaine <- subset(data_use_global_sub, data_use_global_sub$name=="betaine")

#stat.test <- data_use_global_betaine %>%
 # group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_betaine_pasc <- ggplot(data_use_global_betaine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="betaine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_betaine_pasc

### creatine
data_use_global_creatine <- subset(data_use_global_sub, data_use_global_sub$name=="creatine")

stat.test <- data_use_global_creatine %>%
  group_by(pasc_binary) %>%
  wilcox_test(value ~ MTHFR) %>%
  add_y_position() %>%
  add_x_position(x = "pasc_binary", dodge = 1)
stat.test <- 
  rstatix::add_significance(
  stat.test,
  p.col = "p",
  output.col = "p.signif",
  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
  symbols = c("****", "***", "**", "*", "ns")
)

plot_creatine_pasc <- ggplot(data_use_global_creatine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
  labs(x="PASC Group", y="creatine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
plot_creatine_pasc


### creatinine
#data_use_global_creatinine <- subset(data_use_global_sub, data_use_global_sub$name=="creatinine")

#stat.test <- data_use_global_creatinine %>%
#  group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
#  add_y_position() %>%
#  add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
#  rstatix::add_significance(
#  stat.test,
#  p.col = "p",
#  output.col = "p.signif",
#  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
#  symbols = c("****", "***", "**", "*", "ns")
#)

#plot_creatinine_pasc <- ggplot(data_use_global_creatinine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="creatinine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_creatinine_pasc

### cysteine
#data_use_global_cysteine <- subset(data_use_global_sub, data_use_global_sub$name=="cysteine")

#stat.test <- data_use_global_cysteine %>%
#  group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_cysteine_pasc <- ggplot(data_use_global_cysteine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="cysteine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_cysteine_pasc

### cysteine s-sulfate
#data_use_global_cysteinesulfate <- subset(data_use_global_sub, data_use_global_sub$name=="cysteine s-sulfate")

#stat.test <- data_use_global_cysteinesulfate %>%
#  group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
#  add_y_position() %>%
#  add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_cysteinesulfate_pasc <- ggplot(data_use_global_cysteinesulfate, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
#  labs(x="PASC Group", y="cysteine s-sulfate") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_cysteinesulfate_pasc

### cystine
#data_use_global_cystine <- subset(data_use_global_sub, data_use_global_sub$name=="cystine")

#stat.test <- data_use_global_cystine %>%
 # group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_cystine_pasc <- ggplot(data_use_global_cystine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="cystine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_cystine_pasc

### dimethylglycine
data_use_global_dimethylglycine <- subset(data_use_global_sub, data_use_global_sub$name=="dimethylglycine")

stat.test <- data_use_global_dimethylglycine %>%
  group_by(pasc_binary) %>%
  wilcox_test(value ~ MTHFR) %>%
  add_y_position() %>%
  add_x_position(x = "pasc_binary", dodge = 1)
stat.test <- 
  rstatix::add_significance(
  stat.test,
  p.col = "p",
  output.col = "p.signif",
  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
  symbols = c("****", "***", "**", "*", "ns")
)

plot_dimethylglycine_pasc <- ggplot(data_use_global_dimethylglycine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
  labs(x="PASC Group", y="dimethylglycine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
plot_dimethylglycine_pasc


### glycine
#data_use_global_glycine <- subset(data_use_global_sub, data_use_global_sub$name=="glycine")

#stat.test <- data_use_global_glycine %>%
#  group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
#  add_y_position() %>%
#  add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_glycine_pasc <- ggplot(data_use_global_glycine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
#  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
#  labs(x="PASC Group", y="glycine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_glycine_pasc

### guanidinoacetate
data_use_global_guanidinoacetate <- subset(data_use_global_sub, data_use_global_sub$name=="guanidinoacetate")

stat.test <- data_use_global_guanidinoacetate %>%
  group_by(pasc_binary) %>%
  wilcox_test(value ~ MTHFR) %>%
  add_y_position() %>%
  add_x_position(x = "pasc_binary", dodge = 1)
stat.test <- 
  rstatix::add_significance(
  stat.test,
  p.col = "p",
  output.col = "p.signif",
  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
  symbols = c("****", "***", "**", "*", "ns")
)

plot_guanidinoacetate_pasc <- ggplot(data_use_global_guanidinoacetate, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
  labs(x="PASC Group", y="guanidinoacetate") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
plot_guanidinoacetate_pasc


### methionine
#data_use_global_methionine <- subset(data_use_global_sub, data_use_global_sub$name=="methionine")

#stat.test <- data_use_global_methionine %>%
#  group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
#  add_y_position() %>%
#  add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
#  rstatix::add_significance(
#  stat.test,
#  p.col = "p",
#  output.col = "p.signif",
#  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_methionine_pasc <- ggplot(data_use_global_methionine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
#  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="methionine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_methionine_pasc

### methionine sulfone
data_use_global_methioninesulfone <- subset(data_use_global_sub, data_use_global_sub$name=="methionine sulfone")

stat.test <- data_use_global_methioninesulfone %>%
  group_by(pasc_binary) %>%
  wilcox_test(value ~ MTHFR) %>%
  add_y_position() %>%
  add_x_position(x = "pasc_binary", dodge = 1)
stat.test <- 
  rstatix::add_significance(
  stat.test,
  p.col = "p",
  output.col = "p.signif",
  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
  symbols = c("****", "***", "**", "*", "ns")
)

plot_methioninesulfone_pasc <- ggplot(data_use_global_methioninesulfone, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
  labs(x="PASC Group", y="methionine sulfone") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
plot_methioninesulfone_pasc

### N-acetylglycine
data_use_global_acetylglycine <- subset(data_use_global_sub, data_use_global_sub$name=="N-acetylglycine")

stat.test <- data_use_global_acetylglycine %>%
  group_by(pasc_binary) %>%
  wilcox_test(value ~ MTHFR) %>%
  add_y_position() %>%
  add_x_position(x = "pasc_binary", dodge = 1)
stat.test <- 
  rstatix::add_significance(
  stat.test,
  p.col = "p",
  output.col = "p.signif",
  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
  symbols = c("****", "***", "**", "*", "ns")
)

plot_acetylglycine_pasc <- ggplot(data_use_global_acetylglycine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
  labs(x="PASC Group", y="N-acetylglycine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
plot_acetylglycine_pasc


### N-acetylmethionine
#data_use_global_acetylmethionine <- subset(data_use_global_sub, data_use_global_sub$name=="N-acetylmethionine")

#stat.test <- data_use_global_acetylmethionine %>%
#  group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
#  p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_acetylmethionine_pasc <- ggplot(data_use_global_acetylmethionine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="N-acetylmethionine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_acetylmethionine_pasc

### N-acetylserine
#data_use_global_acetylserine <- subset(data_use_global_sub, data_use_global_sub$name=="N-acetylserine")

#stat.test <- data_use_global_acetylserine %>%
#  group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
#  output.col = "p.signif",
#  cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
#  symbols = c("****", "***", "**", "*", "ns")
#)

#plot_acetylserine_pasc <- ggplot(data_use_global_acetylserine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="N-acetylserine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_acetylserine_pasc

### N-acetylthreonine
#data_use_global_acetylthreonine <- subset(data_use_global_sub, data_use_global_sub$name=="N-acetylthreonine")

#stat.test <- data_use_global_acetylthreonine %>%
#  group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
#  rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_acetylthreonine_pasc <- ggplot(data_use_global_acetylthreonine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="N-acetylthreonine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_acetylthreonine_pasc

### N-formylmethionine
#data_use_global_formylmethionine <- subset(data_use_global_sub, data_use_global_sub$name=="N-formylmethionine")

#stat.test <- data_use_global_formylmethionine %>%
#  group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
#  add_y_position() %>%
#  add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
#  p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_formylmethionine_pasc <- ggplot(data_use_global_formylmethionine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
#  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="N-formylmethionine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_formylmethionine_pasc

### S-carboxyethylcysteine
#data_use_global_carboxyethylcysteine <- subset(data_use_global_sub, data_use_global_sub$name=="S-carboxyethylcysteine")

#stat.test <- data_use_global_carboxyethylcysteine %>%
 # group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_carboxyethylcysteine_pasc <- ggplot(data_use_global_carboxyethylcysteine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="S-carboxyethylcysteine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_carboxyethylcysteine_pasc

### S-methylcysteine
#data_use_global_methylcysteine <- subset(data_use_global_sub, data_use_global_sub$name=="S-methylcysteine")

#stat.test <- data_use_global_methylcysteine %>%
#  group_by(pasc_binary) %>%
#  wilcox_test(value ~ MTHFR) %>%
#  add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_methylcysteine_pasc <- ggplot(data_use_global_methylcysteine, aes(x = pasc_binary, y =  value, colour=MTHFR)) +
#  geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
  #labs(x="PASC Group", y="S-methylcysteine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_methylcysteine_pasc

### S-methylcysteine sulfoxide
#data_use_global_methylcysteinesulfoxide <- subset(data_use_global_sub, data_use_global_sub$name=="S-methylcysteine sulfoxide")

#stat.test <- data_use_global_methylcysteinesulfoxide %>%
 # group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_methylcysteinesulfoxide_pasc <- ggplot(data_use_global_methylcysteinesulfoxide, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="S-methylcysteine sulfoxide") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_methylcysteinesulfoxide_pasc

### sarcosine
#data_use_global_sarcosine <- subset(data_use_global_sub, data_use_global_sub$name=="sarcosine")

#stat.test <- data_use_global_sarcosine %>%
 # group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_sarcosine_pasc <- ggplot(data_use_global_sarcosine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="sarcosine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_sarcosine_pasc

### serine
#data_use_global_serine <- subset(data_use_global_sub, data_use_global_sub$name=="serine")

#stat.test <- data_use_global_serine %>%
 # group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_serine_pasc <- ggplot(data_use_global_serine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="serine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_serine_pasc

### taurine
#data_use_global_taurine <- subset(data_use_global_sub, data_use_global_sub$name=="taurine")

#stat.test.taurine <- data_use_global_taurine %>%
 # group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test.taurine <- 
 # rstatix::add_significance(
 # stat.test.taurine,
 # p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_taurine_pasc <- ggplot(data_use_global_taurine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="taurine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_taurine_pasc

### threonine
#data_use_global_threonine <- subset(data_use_global_sub, data_use_global_sub$name=="threonine")

#stat.test <- data_use_global_threonine %>%
 # group_by(pasc_binary) %>%
 # wilcox_test(value ~ MTHFR) %>%
 # add_y_position() %>%
 # add_x_position(x = "pasc_binary", dodge = 1)
#stat.test <- 
 # rstatix::add_significance(
 # stat.test,
#  p.col = "p",
 # output.col = "p.signif",
 # cutpoints = c(0, 1e-04, 0.001, 0.01, 0.05, 1),
 # symbols = c("****", "***", "**", "*", "ns")
#)

#plot_threonine_pasc <- ggplot(data_use_global_threonine, aes(x = pasc_binary, y =  value, colour=MTHFR)) + 
 # geom_point(size=1,position = position_jitterdodge(), alpha=0.5) +
 # labs(x="PASC Group", y="threonine") + theme_bw() + stat_pvalue_manual(stat.test,  label = "p.signif")
#plot_threonine_pasc 

```
### Figure S5 panel
```{r}
ggarrange(plot_sulfoxide_pasc, plot_creatine_pasc, plot_dimethylglycine_pasc, plot_guanidinoacetate_pasc, plot_methioninesulfone_pasc, plot_acetylglycine_pasc, common.legend = TRUE, legend = "bottom")
```

### Table 1 - patient demographics stratified by MTHFR alleles
* targeted and global metabolomics
* sex, age, TG, visit, MTHFR alleles, PASC
```{r}
temp <- clin_alleles
# distinct participant
temp <- temp %>% distinct(participant_id, .keep_all = TRUE)

# table 
temp$sex <- 
  factor(temp$sex, levels=c("Male","Female"),
         labels=c("Male", 
                  "Female"))
temp$event_type <-
  factor(temp$event_type, levels=c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6"),
         labels = c("Visit 1", "Visit 2", 
                    "Visit 3", "Visit 4", 
                    "Visit 5", "Visit 6"))

temp$trajectory_group <- 
  factor(temp$trajectory_group, levels=c("1","2","3","4","5"),
         labels=c("TG1","TG2","TG3","TG4","TG5"))

temp$conv_cluster4 <- 
  factor(temp$conv_cluster4, levels=c("1","2","5","6"),
         labels=c("MIN","PHY","COG","MLT"))

temp$MTHFR <-
  factor(temp$MTHFR, levels=c("GG","AG","AA"),
         labels=c("GG","AG","AA"))

temp$admit_age <- as.integer(temp$admit_age)
#temp$discretized_admit_age_quantile <- 
  #factor(temp$discretized_admit_age_quantile, levels=c("[18,35]","[36,51]","[52,66]","[67,81]","[82,96]"),
     #    labels=c("[18,35]","[36,51]","[52,66]","[67,81]","[82,96]"))

label(temp$sex) <- "Sex"
label(temp$admit_age) <- "Age at admission (years)"
#label(temp$discretized_admit_age_quantile) <- "Discretized Age Quantile (years)"
label(temp$event_type)       <- "Visit"
label(temp$trajectory_group)     <- "Trajectory Group"
label(temp$conv_cluster4) <- "Convalescent PASC Group"
label(temp$MTHFR) <- "MTHFR allele"


pvalue <- function(x, ...) {
  x <- x[-length(x)]  # Remove "overall" group
  # Construct vectors of data y, and groups (strata) g
  y <- unlist(x)
  g <- factor(rep(1:length(x), times=sapply(x, length)))
  if (is.numeric(y)) {
    # For numeric variables, perform an ANOVA
    p <- summary(aov(y ~ g))[[1]][["Pr(>F)"]][1]
  } else {
    # For categorical variables, perform a chi-squared test of independence
    p <- chisq.test(table(y, g))$p.value
  }
  # Format the p-value, using an HTML entity for the less-than sign.
  # The initial empty string places the output on the line below the variable label.
  c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
}

#pdf(file = "Table1_MTHFR.pdf",width=20,height=20)
table1_mthfr <- table1(~ sex + admit_age + event_type + trajectory_group + conv_cluster4 | MTHFR, data=temp, overall=c(left="Total"), topclass="Rtable1-zebra", extra.col = list(`P-value`=pvalue)) #, footnote=footnote, caption=caption
#dev.off()

write.table (table1_mthfr , "/users/csyphurs/Table1_MTHFR.csv", col.names = T, row.names=F, sep=',')

```

### Logistic regression AA vs. OTHERS (AG, GG) based on baseline metabolites for each PASC group (MIN, PHY, COG, MLT)
```{r, eval=F}
## PASC = MIN regression analysis

# subset global V1 data for PASC=="MIN"
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$pro_group_labels=="MIN")

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$MTHFR=="AA","AA","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "OTHER" ~ 0,
    outcome == "AA" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)
#age_quantile <- as.numeric(met_data$admit_age)


plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AAvOTHER_min <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AAvOTHER_min$qval <= 0.05) # 258 sig mets
  output_all_AAvOTHER_min = output_all
```
```{r, eval=F}
## PASC = PHY regression analysis

# subset global V1 data for PASC="PHY"
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$pro_group_labels=="PHY")

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$MTHFR=="AA","AA","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "OTHER" ~ 0,
    outcome == "AA" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)

plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AAvOTHER_phy <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AAvOTHER_phy$qval <= 0.05) # 270 sig mets
  output_all_AAvOTHER_phy = output_all
```
```{r, eval=F}
## PASC = COG regression analysis 

# subset global V1 data for PASC=="COG"
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$pro_group_labels=="COG")

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$MTHFR=="AA","AA","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "OTHER" ~ 0,
    outcome == "AA" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)

plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AAvOTHER_cog <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AAvOTHER_cog$qval <= 0.05) # 341 sig mets
  output_all_AAvOTHER_cog = output_all
```
```{r, eval=F}
## PASC = MLT regression analysis

# subset global V1 data for PASC=="MLT"
met_data <- subset(data_use_global_acute_v1, data_use_global_acute_v1$pro_group_labels=="MLT")

# structure outcome variable
met_data <-  met_data[order(met_data$participant_id),]
met_data$outcome <- ifelse(met_data$MTHFR=="AA","AA","OTHER")
met_data <- met_data %>%  mutate(OUTCOME = case_when(
    outcome == "OTHER" ~ 0,
    outcome == "AA" ~ 1
    ))

met_data <- met_data %>% drop_na(OUTCOME)
OUTCOME = met_data$OUTCOME
table(OUTCOME)
  
sex <- as.factor(met_data$sex)
y <- as.numeric (OUTCOME)
bmi <- as.numeric(met_data$bmi) #numerical
race <- as.factor(met_data$race)
ethnicity <- as.factor(met_data$ethnicity)
participant <- as.factor(met_data$participant_id)
age_quantile <- as.factor(met_data$discretized_admit_age_quantile)

plasma_metabolomics_global_counts_baseline <- plasma_metabolomics_global_counts[met_data$sample_id,]
plasma_metabolomics_global_rowfeature_tmp <- plasma_metabolomics_global_rowfeature[colnames(plasma_metabolomics_global_counts_baseline),]
  
x_met <- data.table(plasma_metabolomics_global_counts_baseline)
pval <- c()
coef <- c()
zval <- c()
std <- c()
output_all <- data.frame(matrix(NA, ncol = 8))
colnames(output_all) <- c("outc", "metabolite", "beta", "se", "zval", "pval_raw", "lowerCI", "upperCI")

for (i in 1:ncol(x_met)){
    # Get metab x
    #i = 1
    met_i <- x_met[, ..i]
    met_i_name <-colnames(x_met)[i]
    
    
    # Build data frame
    
    df <- cbind(y, x_met, sex, participant, bmi,age_quantile) #, location,phase, plate,type,age
    df <- df[complete.cases(df),]
    vars <- c("bmi","sex", "age_quantile") #"location","phase","plate","type"
    xvars <- c(met_i_name, vars)
   
    fmla <- as.formula(paste("y ~ ", paste(xvars, collapse= "+")))
    fmla
    
    # Build model and keep pvalue
    #lr<- geeglm (fmla, data=df, corstr="exchangeable",family=binomial, id=participant) #
    lr<- glm (fmla, data=df, family=binomial) 
    coef_2 <- summary(lr)$coef %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    coef_2 <- coef_2[rowname == met_i_name, ]
    conf <- confint.default(lr) %>% as.data.frame() %>% rownames_to_column() %>% as.data.table()
    conf <- conf[rowname == met_i_name, -c("rowname")]
    output <- cbind(met_i_name, coef_2,conf)
    setnames(output, c("outc", "metabolite", "beta", "se", "zval", "pval_raw","lowerCI","upperCI"))
    output[, (colnames(output)[-c(1:2)]) := lapply(.SD, as.numeric), .SDcols = colnames(output)[-c(1:2)]]
    output_all <- rbind(output_all, output)
  
    # Extract
    pval <- c(pval, coef(summary(lr))[2,4]) #sign
    #zval <- c(zval, coef(summary(lr))[2,3]) #zscore
    coef <- c(coef, coef(summary(lr))[2,1]) #beta value
    std <- c(std, coef(summary(lr))[2,2])  #stdev
  }
  feature_list = colnames(x_met)
  lr_output <- data.table(feature_list)
  lr_output$coef <- coef
  lr_output$std <- std
  lr_output$pval<- pval
  lr_output$qval  <- p.adjust(pval, method = 'bonferroni')
  
  Compound_ID <- rownames (plasma_metabolomics_global_rowfeature_tmp)
  Name <- plasma_metabolomics_global_rowfeature_tmp$CHEMICAL_NAME
  SUPER_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUPER_PATHWAY
  SUB_PATHWAY <- plasma_metabolomics_global_rowfeature_tmp$SUB_PATHWAY
  HMDB <- plasma_metabolomics_global_rowfeature_tmp$HMDB
  KEGG <- plasma_metabolomics_global_rowfeature_tmp$KEGG
  Platform <- plasma_metabolomics_global_rowfeature_tmp$PLATFORM
  
  lr_output_AAvOTHER_mlt <- data.table(Compound_ID, Name, Platform, SUPER_PATHWAY, SUB_PATHWAY, HMDB, lr_output)
  sum(lr_output_AAvOTHER_mlt$qval <= 0.05) # 301 sig mets
  output_all_AAvOTHER_mlt = output_all
```
```{r, eval=F}
## AA vs OTHERS forest plot

# Combine logistic regression results for each PASC group
lr_output_AAvOTHER_min = merge(lr_output_AAvOTHER_min, output_all_AAvOTHER_min, by.x = "Compound_ID",by.y = "outc")
lr_output_AAvOTHER_min$type = "MIN"
lr_output_AAvOTHER_min = lr_output_AAvOTHER_min %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05  ~ "Not significant"
    ))

lr_output_AAvOTHER_phy = merge(lr_output_AAvOTHER_phy, output_all_AAvOTHER_phy, by.x = "Compound_ID",by.y = "outc")
lr_output_AAvOTHER_phy$type = "PHY"
lr_output_AAvOTHER_phy = lr_output_AAvOTHER_phy %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))
lr_output_AAvOTHER_cog = merge(lr_output_AAvOTHER_cog, output_all_AAvOTHER_cog, by.x = "Compound_ID",by.y = "outc")
lr_output_AAvOTHER_cog$type = "COG"
lr_output_AAvOTHER_cog = lr_output_AAvOTHER_cog %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))
lr_output_AAvOTHER_mlt = merge(lr_output_AAvOTHER_mlt, output_all_AAvOTHER_mlt, by.x = "Compound_ID",by.y = "outc")
lr_output_AAvOTHER_mlt$type = "MLT"
lr_output_AAvOTHER_mlt = lr_output_AAvOTHER_mlt %>%  mutate(sig = case_when(
    qval <= 0.05 ~ "significant",
    qval > 0.05 ~ "Not significant"
    ))


all_results_sig = rbind(lr_output_AAvOTHER_min, lr_output_AAvOTHER_phy, lr_output_AAvOTHER_cog,lr_output_AAvOTHER_mlt)
all_results_sig$type = factor(all_results_sig$type, levels = c("MIN", "PHY", "COG", "MLT"))

  b <-all_results_sig[SUB_PATHWAY %in% c("Methionine, Cysteine, SAM and Taurine Metabolism")] 
  if(dim(b)[1] ==0){
    return("No metabolites found!")
  }else{
    b$data <- "Methionine, Cysteine, SAM and Taurine Metabolism"
    if(dim(b)[1]>100){
      b <- b[order(qval)]
      b <- b[1:100,]
    }
  }
  
    p <- b%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Methionine, Cysteine, SAM and Taurine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
 
  names(b)[2] = "name"
library(ggforestplot)
p_1= forestplot(
  df = b,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Methionine, Cysteine, SAM and Taurine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19)) +
   theme(plot.title = element_text(size=12))
p_1


b <-all_results_sig[SUB_PATHWAY %in% c("Glycine, Serine and Threonine Metabolism")] 
  if(dim(b)[1] ==0){
    return("No metabolites found!")
  }else{
    b$data <- "Glycine, Serine and Threonine Metabolism"
    if(dim(b)[1]>100){
      b <- b[order(qval)]
      b <- b[1:100,]
    }
  }
  
    p <- b%>%
     ggplot( aes(x =Name, y = coef)) +
      facet_grid(. ~ data) +
      geom_errorbar(aes(ymin = lowerCI, ymax = upperCI, color = type), width = 0.5, size = 1) +
      geom_pointrange(aes(ymin = lowerCI, ymax = upperCI, color = type, shape = sig), size = 1) +#shape = 15, 
      
      geom_hline(yintercept = 0, linetype = 2) +
      labs(title = "Glycine, Serine and Threonine Metabolism", x = "Metabolites", y = "coefficient (95% confidence interval)") +
      
      theme(title = element_text(size = 2),
            axis.title = element_text(size = 2),
            axis.text = element_text(size = 2),
            strip.text.x = element_text(size = 2),
            legend.position = "right",
            legend.title = element_text(size = 2),
            legend.text = element_text(size = 2)) +
      theme_bw() +
      coord_flip()
 
  names(b)[2] = "name"
library(ggforestplot)
p_2= forestplot(
  df = b,
  estimate = coef,
  logodds = FALSE,
  colour = type,
  shape = sig,
  title = "Glycine, Serine and Threonine Metabolism",
  xlab = "coefficient",pvalue = qval, psignif = 0.05
) +scale_shape_manual(values=c(17,19)) +
   theme(plot.title = element_text(size=12))
p_2

```

## Supplementary: check other MTHFR SNP distributions (rs1801131)
```{r}
# import allele data
alleles_other <- read.csv("/scratch/metabolomics/plasma-metabolomics-targeted/MTHFR/two_MTHFR_snps.csv")
alleles_other <- alleles_other[,c("participant_id","rs1801131")]

#merge MTHFR data with clinical data by participant_id
clin_alleles_other <- merge(clinical_data, alleles_other, by="participant_id")
length(unique(clin_alleles_other$participant_id)) # 1031 pts with MTHFR and met sample
clin_alleles_other$trajectory_group <- factor(clin_alleles_other$trajectory_group, levels = c(1,2,3,4,5))

# subset for unique participants
clin_alleles_other_supp = clin_alleles_other[!duplicated(clin_alleles_other$participant_id),]
clin_alleles_other_supp$rs1801131 <- factor(clin_alleles_other_supp$rs1801131, levels = c("TT","TG","GG"))

## TGs proportions
rs1801131_TG <- clin_alleles_other_supp %>%
  dplyr::count(rs1801131, trajectory_group) %>%       
  group_by(rs1801131) %>%
  mutate(pct= prop.table(n) * 100) %>%
  ggplot() + aes(rs1801131, pct, fill=trajectory_group) +
  scale_fill_manual(values=c('#639A21', '#39828C', '#0072B2', '#E69F00', '#9C3418')) +
  geom_bar(stat="identity") +
  ylab("Proportion of Participants") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct),"%")),
            position=position_stack(vjust=0.5)) +
  ggtitle("TG distribution across MTHFR (rs1801131) alleles") +
  theme_bw() 
rs1801131_TG

table(clin_alleles_other_supp$rs1801131)
table(clin_alleles_other_supp$rs1801131, clin_alleles_other_supp$trajectory_group)

# chi square test for homogeneity
conttable <- table(clin_alleles_other_supp$rs1801131, clin_alleles_other_supp$trajectory_group)
chisq.test(conttable)


## mortality proportions
clin_alleles_other_supp$diedever <- ifelse(clin_alleles_other_supp$diedever == 0, "No","Yes")
clin_alleles_other_supp$diedever <- factor(clin_alleles_other_supp$diedever, levels = c("No","Yes"))

rs1801131_mortality <- clin_alleles_other_supp %>%
  dplyr::count(rs1801131, diedever) %>%       
  group_by(rs1801131) %>%
  mutate(pct= prop.table(n) * 100) %>%
  ggplot() + aes(rs1801131, pct, fill=diedever) +
  scale_fill_manual(values=c("royalblue3", "gray")) +
  geom_bar(stat="identity") +
  ylab("Proportion of Participants") +
  geom_text(aes(label=paste0(sprintf("%1.1f", pct),"%")),
            position=position_stack(vjust=0.5)) +
  ggtitle("Mortality distribution across MTHFR (rs1801131) alleles") +
  theme_bw() +
  theme(panel.grid = element_blank())
rs1801131_mortality

table(clin_alleles_other_supp$rs1801131)
table(clin_alleles_other_supp$rs1801131, clin_alleles_other_supp$diedever)

conttable2 <- table(clin_alleles_other_supp$rs1801131, clin_alleles_other_supp$diedever)
chisq.test(conttable2)
```

```{r}
# distinct participant-level data (clin_alleles_other_supp)
clin_alleles_other_supp$death_date <- as.numeric(clin_alleles_other_supp$death_date)
clin_alleles_other_supp$death_date <- ifelse(is.na(clin_alleles_other_supp$death_date), 0, clin_alleles_other_supp$death_date)
clin_alleles_other_supp$death_date <- ifelse(clin_alleles_other_supp$death_date==-99, 0, clin_alleles_supp$death_date)
clin_alleles_other_supp$diedever <- ifelse(clin_alleles_other_supp$diedever=="Yes",1,0)

clin_alleles_other_supp$rs1801131 <- factor(clin_alleles_other_supp$rs1801131, levels = c("TT", "TG", "GG"))

univ_model <- coxph(Surv(death_date, diedever) ~ rs1801131, data = clin_alleles_other_supp)
summary(univ_model)
# HR forest plot
rs1801131_HR <- ggforest(univ_model, data = clin_alleles_other_supp)
rs1801131_HR

print(paste0("N=",length(unique(clin_alleles_other_supp$participant_id))))
clin_alleles_other_supp %>%
  group_by(rs1801131) %>%
  distinct(participant_id) %>%
  dplyr::summarise(n = n())

```


# Panel for Figure S4A
```{r}
ggarrange(MTHFR_TG, MTHFR_mortality,
  rs1801131_TG, rs1801131_mortality, 
  nrow = 2, ncol = 2)
```

# Panel for Figure S4B
```{r}
ggarrange(MTHFR_HR, rs1801131_HR,
  nrow = 2, ncol = 1)
```














