---
title: "comorbidity"
output: pdf_document
---

```{r load library}
require(openxlsx)
library(factoextra)
library(cluster)
require(nlme) # directly provides p-values, unlike lme4
require(splines)
require(dplyr)
library(fgsea)
library(knitr)
library(kableExtra)
library(tidyverse)
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source(paste0(codebase_path,"/codebase_subtype.R"))

organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")

intermediate_results_outdir_DR = "/scratch/data-integration/molecular_subtyping/DimensionReduction/"
mcia_results = readRDS(paste0(intermediate_results_outdir_DR,"MCIA_phase123_20230726.rds"))

data_path <- "/scratch/data-integration/pipelines/processedData/"
data_env0 <- readRDS(paste0(data_path, "data_env_ravi_20220513.RDS"))

base_obj = readRDS(file = "/scratch/data-integration/cleanedup/mod1_interpretation_base.rds")
```

# Load preliminary results
```{r load subtype assignment and clinical data}
# load subtype assignment and clinical data
nclust_tree_all = 6
subtype_hcl = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
clin_data_aligned = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/clin_data_aligned.rds")
extended_data = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/extended_data.rds")
```

# Data preprocessing
```{r Data preprocessing}
# viral load
ct_train = data.frame(
  "event_id" = rownames(base_obj$others$nasal_viralload$train$nasal_viralload_counts),
  "N1_CT" = base_obj$others$nasal_viralload$train$nasal_viralload_counts$N1_CT,
  "RP_CT" = base_obj$others$nasal_viralload$train$nasal_viralload_counts$RP_CT)
ct_test = data.frame(
  "event_id" = rownames(base_obj$others$nasal_viralload$test$nasal_viralload_counts),
  "N1_CT" = base_obj$others$nasal_viralload$test$nasal_viralload_counts$N1_CT,
  "RP_CT" = base_obj$others$nasal_viralload$test$nasal_viralload_counts$RP_CT)
ct_df = rbind(ct_train, ct_test)
ct_df = ct_df[order(ct_df$event_id),]
rownames(ct_df) = ct_df$event_id
# remove samples with undetected RP_CT values
ct_df = ct_df[!is.na(ct_df$RP_CT), ]
# replace NA (undetermined) in N1_CT to 40
ct_df$N1_CT[is.na(ct_df$N1_CT)] = 40
# replace values > 40 to 40
ct_df$N1_CT[ct_df$N1_CT > 40] = 40
# flip the sign of N1_CT
ct_df$N1_CT <- -ct_df$N1_CT
ct_df = subset(ct_df, select = -c(event_id, RP_CT))
ct_df = as.data.frame(quantileNormalizeNoTies(ct_df))
ct_df$event_id = rownames(ct_df)
colnames(ct_df) <- c("negative N1-CT", "event_id")
# antibody
auc_train = data.frame(
  "event_id" = rownames(base_obj$others$serum_rbd_abtiters$train$serum_rbd_abtiters_counts),
  "log2_AUC" = base_obj$others$serum_rbd_abtiters$train$serum_rbd_abtiters_counts$`AUC RBD IgG`)
auc_test = data.frame(
  "event_id" = rownames(base_obj$others$serum_rbd_abtiters$test$serum_rbd_abtiters_counts),
  "log2_AUC" = base_obj$others$serum_rbd_abtiters$test$serum_rbd_abtiters_counts$`AUC RBD IgG`)
auc_df = rbind(auc_train, auc_test)
auc_df = auc_df[order(auc_df$event_id),]
rownames(auc_df) = auc_df$event_id
auc_df = subset(auc_df, select = -event_id)
auc_df = as.data.frame(quantileNormalizeNoTies(auc_df))
auc_df$event_id = rownames(auc_df)
colnames(auc_df) <- c("AUC RBD IgG", "event_id")

# cytof parent preprocess
cytof_parent_mat_train = base_obj$others$bld_cytof$train$bld_cytof_counts
cytof_parent_mat_train = base_obj$others$bld_cytof$preprocess_bld_cytof_parent_granulo_separate(counts = cytof_parent_mat_train, rowfeature = base_obj$others$bld_cytof$train$bld_cytof_rowfeatures)
cytof_parent_mat_test = base_obj$others$bld_cytof$test$bld_cytof_counts
cytof_parent_mat_test = base_obj$others$bld_cytof$preprocess_bld_cytof_parent_granulo_separate(counts = cytof_parent_mat_test, rowfeature = base_obj$others$bld_cytof$test$bld_cytof_rowfeatures)

cytof_parent_mat = rbind(cytof_parent_mat_train, cytof_parent_mat_test)
cytof_parent_mat$event_id = rownames(cytof_parent_mat)
cytof_parent_mat <- cytof_parent_mat %>%
  left_join(ct_df %>% left_join(auc_df, by = "event_id"), by = "event_id")
cytof_parent_feature_names = setdiff(colnames(cytof_parent_mat), "event_id")

cytof_parent_mat = cytof_mat_preprocess(
  cytof_parent_mat, subtype_hcl, clin_data_aligned)
```

# Selected cytof for speparate analysis as in main figures
```{r selected cytof}
select_cytof = c("B Cell", "CD4+ T Cell", "CD8+ T Cell", "Monocytes",
                 "Plasmacytoid DC", "Basophil", "Neutrophil",
                 "negative N1-CT", "AUC RBD IgG")
```

# Cytof parent multi-group comparison (Extended Data Fig10B)
```{r Cytof parent multi-group comparison, fig.width=8.5, fig.height=4.2}
# scale data
cytof_parent_mat[,cytof_parent_feature_names] <- 
  scale(cytof_parent_mat[,cytof_parent_feature_names])
df_comorb <- clin_data_aligned[, c("event_id", "comorb_htn", 
                                   "comorb_anyresp_noasthma", 
                                   "comorb_isaric_cardiac",
                                   "comorb_isaric_ckd", "comorb_hxtrans")]
cytof_parent_mat <- left_join(cytof_parent_mat, df_comorb, by = "event_id")

max_cutoff = 0.5
cytof_parent_test = list()
cytof_parent_test[["early"]]= multi_group_mixed_test_comorb(
  feature_name = cytof_parent_feature_names, data_mat = cytof_parent_mat, 
  visit_numbers = c("V1", "V2", "V3"))
cytof_parent_test[["late"]]= multi_group_mixed_test_comorb(
  feature_name = cytof_parent_feature_names, data_mat = cytof_parent_mat, 
  visit_numbers = c("V4", "V5", "V6"))

est_mat = cbind(t(cytof_parent_test[["early"]]$est), 
                t(cytof_parent_test[["late"]]$est))
pval_mat = cbind(t(cytof_parent_test[["early"]]$pval), 
                 t(cytof_parent_test[["late"]]$pval))
rownames(est_mat)[rownames(est_mat) == "CD4 T Cell"] = "CD4+ T Cell"
rownames(est_mat)[rownames(est_mat) == "CD8 T Cell"] = "CD8+ T Cell"
rownames(est_mat)[rownames(est_mat) == "Hematopoetic progrenitor Cell"] = 
  "Hematopoetic progrenitor"
rownames(pval_mat) = rownames(est_mat)
pvec = as.vector(as.matrix(pval_mat))
pvec_bh = p.adjust(pvec, method = "BH")
adj_p = matrix(pvec_bh, nrow = nrow(pval_mat), ncol = ncol(pval_mat),
               dimnames = list(rownames(pval_mat), colnames(pval_mat)))
# adj_p = apply(pval_mat, 2, function(p) p.adjust(p, method = "BH"))

test_type = factor(colnames(est_mat), 
                   levels = c("EF vs ABC", "F vs E", 
                              "A vs BC", "B vs AC", "C vs AB"))
colnames(est_mat) = rep(c("V1-V3","V4-V6"), each = 5)
colnames(pval_mat) = rep(c("V1-V3","V4-V6"), each = 5)
est_mat[est_mat > max_cutoff] = max_cutoff
est_mat[est_mat < -max_cutoff] = -max_cutoff

est_mat = est_mat[select_cytof,]
adj_p = adj_p[select_cytof,]

cytof_multi_comparison = ComplexHeatmap::Heatmap(
  as.matrix(est_mat), 
  col = circlize::colorRamp2(c(-max_cutoff, 0, max_cutoff),
                             c("#284AA0", "white", "#BC4A4A")),
  show_heatmap_legend = TRUE, na_col = "grey90", border=T, 
  column_title_gp = gpar(fontsize = 15), 
  border_gp = gpar(col = "darkgrey", lwd = 0.5), 
  rect_gp = gpar(col = "darkgrey", lwd = 0.5),
  row_names_gp = gpar(fontsize = 15), 
  column_names_gp = gpar(fontsize = 15),
  heatmap_legend_param = list(
    title = "Coefficient    ",
    direction = "horizontal", title_position = "leftcenter",
    title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
  column_split = test_type, cluster_columns = F,
  cluster_rows = F, cell_fun = function(j, i, x, y, w, h, fill) {
  if(adj_p[i, j] < 0.001) {
    grid.text("***", x, y-h*0.12)
  }else if(adj_p[i, j] < 0.01) {
    grid.text("**", x, y-h*0.12)
  }else if(adj_p[i, j] < 0.05) {
    grid.text("*", x, y-h*0.12)
  }
})

ComplexHeatmap::draw(cytof_multi_comparison, heatmap_legend_side = "bottom")
```

# Test missingness adjusted for comorbidities (Extended Data Fig10A)
```{r test missingness adjusted for comorbidities, fig.height=6, fig.width=2.3}
sample_meta_participant = extended_data$sample_meta_participant_level
sample_meta_participant$death_date[sample_meta_participant$death_date<0] = 999
censoring_date_death = 200
death_surv = cbind(sample_meta_participant$death_date,
                   sample_meta_participant$death_date)
death_surv[death_surv[,2]>censoring_date_death,2] = censoring_date_death+1
death_surv[,1] = ifelse(death_surv[,2]<=censoring_date_death,1,0)
death_surv = data.frame(death_surv)
colnames(death_surv) = c("state", "time")

id_post_acute <- which(death_surv$time > 28 & subtype_hcl$label != "D")
post_acute_label = array("Non-respondent",
                         nrow(extended_data$sample_meta_participant_level))
idx_PASC = which(!is.na(subtype_hcl$PASC))
post_acute_label[idx_PASC] = "Any Deficit"
post_acute_label[subtype_hcl$PASC == "minimal"] = "Minimal Deficit"
post_acute_label = factor(post_acute_label, 
                          levels = c("Minimal Deficit", "Any Deficit",
                                     "Non-respondent"))

clin_data_patient <- clin_data_aligned[clin_data_aligned$event_type == "Visit 1",]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
ww <- ifelse(post_acute_label == "Non-respondent", 1, 0)
sig_comorb = c("comorb_htn", "comorb_anyresp_noasthma", "comorb_isaric_cardiac",
               "comorb_isaric_ckd", "comorb_hxtrans")
comorb_data_adjust = clin_data_patient[, sig_comorb]

idx_non_resp = which(post_acute_label == "Minimal Deficit" |
                       post_acute_label == "Non-respondent")
idx_non_resp = intersect(idx_non_resp, id_post_acute)

missing_demo_comorb_adjusted_test <- test_clinical_binomial_demo(
  subtype = subtype_hcl$label[idx_non_resp],
  response_binary = ww[idx_non_resp],
  demo_data = cbind(demo_data_adjust,
                    comorb_data_adjust)[idx_non_resp,])

heatmap_binomial(missing_demo_comorb_adjusted_test, col_name = "Non-\nrespondent")
```

# Test any deficit adjusted for cormorbidities (Extended Data Fig10A)
```{r test any deficit adjusted for cormorbidities, fig.height=6, fig.width=2.3}
id_pasc = which(!is.na(subtype_hcl$PASC) & subtype_hcl$label != "D")
ww = ifelse(subtype_hcl$PASC == "minimal", 0, 1)
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
sig_comorb = c("comorb_htn", "comorb_anyresp_noasthma", "comorb_isaric_cardiac",
               "comorb_isaric_ckd", "comorb_hxtrans")
comorb_data_adjust = clin_data_patient[, sig_comorb]

any_deficit_demo_comorb_adjusted_test <- test_clinical_binomial_demo(
  subtype = subtype_hcl$label[id_pasc],
  response_binary = ww[id_pasc],
  demo_data = cbind(demo_data_adjust,
                    comorb_data_adjust)[id_pasc,])

heatmap_binomial(any_deficit_demo_comorb_adjusted_test, col_name = "Any Deficit")
```

# Test PASC adjusted for comorbidities (Extended Data Fig10A)
```{r test PASC adjusted for comorbidities, fig.height=6, fig.width=4.2}
full_clust_test = list(); full_clust_test[["PASC"]] = list()
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
sig_comorb = c("comorb_htn", "comorb_anyresp_noasthma", "comorb_isaric_cardiac",
               "comorb_isaric_ckd", "comorb_hxtrans")
comorb_data_adjust = clin_data_patient[, sig_comorb]

full_clust_test[["PASC"]][["regression_comorb"]] = test_multi_pasc_demo(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = cbind(demo_data_adjust, comorb_data_adjust))

heatmap_pasc_multi(full_clust_test[["PASC"]][["regression_comorb"]],
                   "Demographics + Comorbidities")
```

# PASC comparison for A-C (Extended Data Fig10A)
```{r PASC comparison for A-C}
# any deficit (A vs C)
id_pasc = which(!is.na(subtype_hcl$PASC) & subtype_hcl$label %in% c("A", "C"))
ww = ifelse(subtype_hcl$PASC == "minimal", 0, 1)
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
sig_comorb = c("comorb_htn", "comorb_anyresp_noasthma", "comorb_isaric_cardiac",
               "comorb_isaric_ckd", "comorb_hxtrans")
comorb_data_adjust = clin_data_patient[, sig_comorb]

df_binom = data.frame(PASC = ww, label = subtype_hcl$label, 
                      demo_data_adjust, comorb_data_adjust)
df_binom = df_binom[id_pasc,]

model_pasc_AC = glm(PASC ~ ., family = binomial, data = df_binom)
coef(summary(model_pasc_AC))

# physical deficit (A vs C)
id_pasc = which(subtype_hcl$PASC %in% c("physical", "minimal") & 
                  subtype_hcl$label %in% c("A", "C"))
ww = ifelse(subtype_hcl$PASC == "minimal", 0, 1)
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
sig_comorb = c("comorb_htn", "comorb_anyresp_noasthma", "comorb_isaric_cardiac",
               "comorb_isaric_ckd", "comorb_hxtrans")
comorb_data_adjust = clin_data_patient[, sig_comorb]

df_binom = data.frame(PASC = ww, label = subtype_hcl$label, 
                      demo_data_adjust, comorb_data_adjust)
df_binom = df_binom[id_pasc,]

model_pasc_AC = glm(PASC ~ ., family = binomial, data = df_binom)
coef(summary(model_pasc_AC))
```

# Test PASC adjusted for comorbidities - global test (Extended Data Fig10A)
```{r test PASC adjusted for comorbidities - global test, fig.height=6, fig.width=2}
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
sig_comorb = c("comorb_htn", "comorb_anyresp_noasthma", "comorb_isaric_cardiac",
               "comorb_isaric_ckd", "comorb_hxtrans")
comorb_data_adjust = clin_data_patient[, sig_comorb]

global_test_comorb = test_global_multi_pasc(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = cbind(demo_data_adjust, comorb_data_adjust))

heatmap_multi_global(global_test_comorb)
```






