---
title: "mechanical ventilation"
output: pdf_document
---

```{r load library}
library(ade4) #used in mcia
library(mogsa)
library(factoextra)
library(cluster)
library(nnet)
library(survival)
library(ordinal)
library(ggeffects)
library(openxlsx)
library(network)
library(GGally)
library(ggalt)
library(metafor)
library(cowplot)
library(survminer)
require(nlme) # directly provides p-values, unlike lme4
require(splines)
require(dplyr)
library(fgsea)
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source(paste0(codebase_path,"/codebase_subtype.R"))

organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")

intermediate_results_outdir_DR = "/scratch/data-integration/molecular_subtyping/DimensionReduction/"
mcia_results = readRDS(paste0(intermediate_results_outdir_DR,"MCIA_phase123_20230726.rds"))

data_path <- "/scratch/data-integration/pipelines/processedData/"
data_env0 <- readRDS(paste0(data_path, "data_env_ravi_20220513.RDS"))
```

# Load subtype assignment and clinical data
```{r load subtype assignment and clinical data}
nclust_tree_all = 6
subtype_hcl = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
clin_data_aligned = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/clin_data_aligned.rds")
extended_data = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/extended_data.rds")
```

# Mechanical ventilation
```{r}
clin_data_patient <- clin_data_aligned[clin_data_aligned$event_type == "Visit 1",]
icu_table = table(clin_data_patient$ever_icu, subtype_hcl$label)
icu_table[rownames(icu_table) == 1,] / colSums(icu_table) * 100

resp_table = table(clin_data_patient$respiratory_status, subtype_hcl$label)
resp_table[rownames(resp_table) == 6, ] / colSums(resp_table) * 100

resp_table_LC = table(clin_data_patient$respiratory_status[!is.na(subtype_hcl$PASC)], 
                      subtype_hcl$label[!is.na(subtype_hcl$PASC)])
resp_table_LC[rownames(resp_table_LC) == 6, ] / colSums(resp_table_LC) * 100
```

# Clinical names correspondence
```{r clinical names correspondence}
clin_names <- c("sex" = "Sex (Female)",
                "ethnicity" = "Ethnicity (Hispanic or Latino)",
                "admit_age" = "Admission age",
                "discretized_admit_age_quantile" = "Admission age quantile",
                "raceAmerican Indian / Alaska Native" = "American Indian/Alaska Native",
                "raceAsian" = "Asian",
                "raceBlack / African American" = "Black/African American",
                "raceBlack...African.American" = "Black/African American",
                "raceMultiple" = "Multiple Races",
                "raceNative Hawaiian / Pacific Islander" = "Native Hawaiian/Pacific Islander",
                "raceWhite" = "White",
                "bmicat" = "BMI category",
                "comorb_htn" = "Hypertension",
                "comorb_isaric_dm" = "Diabetes",
                "comorb_anyresp_noasthma" = "Pulmonary (excludes asthma)",
                "comorb_isaric_asthma" = "Asthma",
                "comorb_isaric_cardiac" = "Chronic cardiac disease",
                "comorb_isaric_ckd" = "Chronic kidney disease",
                "comorb_isaric_neoplasm" = "Malignant neoplasm",
                "comorb_neuro" = "Chronic neurological disorder",
                "comorb_isaric_liver" = "Liver disease",
                "comorb_hxtrans" = "History of solid organ or bone marrow transplant",
                "comorb_hiv2" = "HIV",
                "comorb_eversmkvape" = "Current or former smoking and/or vaping",
                "comorb_anysubstance" = "Drug or alcohol abuse or cannabis use",
                "comp_seizure" = "Seizures",
                "comp_meningitis" = "Meningitis/encephalitis",
                "comp_stroke" = "Stroke/CVA",
                "comp_afib" = "Atrial fibrillation",
                "comp_ventricular_arr" = "Ventricular arrhythmia",
                "comp_endo" = "Endocarditis",
                "comp_myocard" = "Myocarditis/pericarditis",
                "comp_cardarrest" = "Cardiac arrest",
                "comp_stemi" = "ST-elevation MI (STEMI)",
                "comp_nstemi" = "Non-ST elevation MI (NSTEMI) or other cardiac ischemia",
                "comp_chf" = "CHF/cardiomyopathy",
                "comp_pneumonia" = "Pneumonia",
                "comp_bronch" = "Bronchiolitis",
                "comp_pneumo" = "Pneumothorax",
                "comp_pneumomed" = "Pneumomediastinum",
                "comp_aat" = "Acute arterial thromboembolism excluding stroke/CVA or myocardial infarction",
                "comp_avthromb" = "Acute venous thromboembolism",
                "comp_shock" = "Shock (use of vasopressors)",
                "comp_bac" = "Bacteremia",
                "comp_coag" = "Coagulation disorder/disseminated intravascular coagulation",
                "comp_hypogly" = "Symptomatic hypoglycemia",
                "comp_bleed" = "Bleeding",
                "comp_anemia" = "Anemia",
                "comp_pancreas" = "Pancreatitis",
                "comp_renal" = "Acute renal injury/failure",
                "comp_liver" = "Liver dysfunction/failure",
                "comp_othlab" = "Other lab abnormality",
                "comp_electrolyte" = "Electrolyte abnormality",
                "comp_othinf" = "Other infection (non-pulmonary)",
                "comp_othpulm" = "Other pulmonary complication",
                "comp_unddisease" = "Exacerbation of underlying disease",
                "comp_enceph" = "Encephalopathy",
                "comp_othneuro" = "Other neurologic complication",
                "comp_hypergly" = "Hyperglycemia",
                "comp_psych" = "Psychiatric complication",
                "comp_mis" = "Multisystem inflammatory syndrome (MIS)",
                "comp_gastroent" = "Gastroenteritis",
                "comp_pulmvas" = "Pulmonary vascular disease",
                "ever_azithro" = "Azithromycin",
                "ever_icu" = "Ever in ICU",
                "ever_otherabx" = "Other Antibiotics",
                "ever_remdes" = "Remdesivir",
                "ever_steroids" = "Steroids",
                "ever_ever_esc" = "Ever required escalation of care",
                "ever_immsupp" = "Immunosuppressives",
                "baseline_anyimg" = "Any chest imaging at baseline",
                "baseline_img_infil" = "Radiographic findings on chest imaging",
                "baseline_lymph_abn" = "Abnormal Lymphocyte count",
                "baseline_platelets_abn" = "Abnormal Platelets",
                "baseline_alt_abn" = "Abnormal ALT",
                "baseline_cr_abn" = "Abnormal Creatinine",
                "baseline_crp_abn" = "Abnormal CRP",
                "baseline_ddimer_abn" = "Abnormal D-dimer",
                "baseline_trop_abn" = "Abnormal Troponin")
```

```{r}
interest_clin_feature = c(
  "comorb_htn", "comorb_anyresp_noasthma", "comorb_isaric_cardiac", 
  "comorb_isaric_ckd", "comorb_hxtrans", "comp_stroke", "comp_afib",
  "comp_ventricular_arr", "comp_cardarrest", "comp_chf", "comp_pneumonia", 
  "comp_avthromb", "comp_shock", "comp_bac", "comp_bleed", "comp_anemia", 
  "comp_renal", "ever_icu", "ever_ever_esc", "baseline_lymph_abn", 
  "baseline_platelets_abn", "baseline_alt_abn", "baseline_cr_abn", 
  "baseline_crp_abn", "baseline_ddimer_abn", "baseline_trop_abn", 
  "baseline_img_infil")
```

# Clinical heatmap adjusted for mechanical ventilation (Extended Data Fig10C)
```{r clinical heatmap adjusted for mechanical ventilation, fig.height=8, fig.width=11, warning=FALSE}
clin_data_patient <- clin_data_aligned[clin_data_aligned$event_type == "Visit 1",]
demo_data <- clin_data_patient[, c("admit_age", "sex", "ethnicity",
                                   "discretized_admit_age_quantile",
                                   "bmicat", "enrollment_site", "race")]
demo_data$ethnicity[demo_data$ethnicity == "Not Specified"] <- NA
demo_data$bmicat[demo_data$bmicat == 999] <- NA
demo_data$bmicat_test <- demo_data$bmicat
# group obesity together
demo_data$bmicat_test[demo_data$bmicat_test == 5] <- 4
demo_data$race <- as.factor(demo_data$race)
race_one_hot <- model.matrix(~ race-1, data = demo_data)
race_one_hot <- race_one_hot[, !(colnames(race_one_hot) %in%
                                   c("raceOther / Declined", 
                                     "raceUnknown / Unavailable"))]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", 
                                          "sex", "respiratory_status")]
demo_data_adjust$respiratory_status <-
  ifelse(demo_data_adjust$respiratory_status == 6, 1, 0)
demo_data_adjust$respiratory_status <- factor(demo_data_adjust$respiratory_status)

clin_data_logit <- clin_data_patient %>%
  dplyr::select(matches("^(comorb|comp)"), ever_icu, ever_ever_esc) %>%
  dplyr::select(-comorb_count, -comorb_countcat, -comp_other2) %>%
  select_if(~sum(.x, na.rm = TRUE) > 0)
clin_data_demo_logit <- cbind(demo_data[, c("sex", "ethnicity")], race_one_hot)
clin_data_demo_logit$sex <- ifelse(clin_data_demo_logit$sex == "Female", 1, 0)
clin_data_demo_logit$ethnicity <- ifelse(clin_data_demo_logit$ethnicity == 
                                           "Hispanic or Latino", 1, 0)

clin_data_baseline <- clin_data_patient %>%
  dplyr::select(matches("^baseline")) %>%
  dplyr::select(-baseline_anyimg) %>%
  mutate(across(everything(), ~replace(., . == 999, NA)))
clin_data_baseline_logit <- clin_data_baseline %>%
  dplyr::select(ends_with("_abn"))

sub_idx <- which(subtype_hcl$label != "D")
clin_test_logit <- test_clinical_logit_demo(
  subtype_hcl$label[sub_idx],
  clin_data_logit[sub_idx,],
  demo_data_adjust[sub_idx,])
# demo_test_logit <- test_clinical_logit(
#   subtype_hcl$label[sub_idx],
#   clin_data_demo_logit[sub_idx,])
baseline_test_logit <- test_clinical_logit_demo(
  subtype_hcl$label[sub_idx],
  clin_data_baseline_logit[sub_idx,],
  demo_data_adjust[sub_idx,])
# bmi_test_ordinal <- test_ordinal(
#   subtype_hcl$label[sub_idx],
#   demo_data$bmicat_test[sub_idx])
# age_test_ordinal <- test_ordinal(
#   subtype_hcl$label[sub_idx],
#   demo_data$discretized_admit_age_quantile[sub_idx])
baseline_test_ordinal <- test_ordinal_demo(
  subtype_hcl$label[sub_idx],
  clin_data_baseline$baseline_img_infil[sub_idx],
  demo_data_adjust[sub_idx,])

clin_test_logit$p_lr_adj <- p.adjust(clin_test_logit$p_lr, method = "BH")
# demo_test_logit$p_lr_adj <- p.adjust(demo_test_logit$p_lr, method = "BH")

# clinical test featuring variables of interest
clin_test <- clin_test_logit
# interest_clin_idx <- which(clin_test_logit$p_lr_adj < 0.001)
# interest_demo_idx <- which(demo_test_logit$p_lr_adj < 0.001)
clin_test$est_logit <- cbind(clin_test_logit$est_logit, 
                             # demo_test_logit$est_logit[,interest_demo_idx],
                             baseline_test_logit$est_logit,
                             # bmi_test_ordinal$est_ordinal,
                             # age_test_ordinal$est_ordinal,
                             baseline_test_ordinal$est_ordinal)
# colnames(clin_test$est_logit)[ncol(clin_test$est_logit)-2] <- "bmicat"
# colnames(clin_test$est_logit)[ncol(clin_test$est_logit)-1] <- 
#   "discretized_admit_age_quantile"
colnames(clin_test$est_logit)[ncol(clin_test$est_logit)] <- "baseline_img_infil"
clin_test$est_logit = clin_test$est_logit[,interest_clin_feature]

clin_test$p_logit <- cbind(clin_test_logit$p_logit, 
                           # demo_test_logit$p_logit[,interest_demo_idx],
                           baseline_test_logit$p_logit,
                           # bmi_test_ordinal$p_ordinal,
                           # age_test_ordinal$p_ordinal,
                           baseline_test_ordinal$p_ordinal)
# colnames(clin_test$p_logit)[ncol(clin_test$p_logit)-2] <- "bmicat"
# colnames(clin_test$p_logit)[ncol(clin_test$p_logit)-1] <- 
#   "discretized_admit_age_quantile"
colnames(clin_test$p_logit)[ncol(clin_test$p_logit)] <- "baseline_img_infil"

pvec <- as.vector(as.matrix(clin_test$p_logit))
pvec_bh <- p.adjust(pvec, method = "BH")
clin_test$p_adj <- matrix(pvec_bh, nrow = nrow(clin_test$p_logit), 
                          ncol = ncol(clin_test$p_logit))
rownames(clin_test$p_adj) <- rownames(clin_test$p_logit)
colnames(clin_test$p_adj) <- colnames(clin_test$p_logit)

clin_test$p_logit = clin_test$p_logit[, interest_clin_feature]
clin_test$p_adj = clin_test$p_adj[, interest_clin_feature]

# clin_test$p_adj <- apply(clin_test$p_logit, 2,
#                          function(p) p.adjust(p, method = "BH"))

heatmap_clinical(clin_test, clin_names, adjust_p = TRUE)
```

# Test missingness adjusted for mv (Extended Data Fig10D)
```{r test missingness adjusted for mv, fig.height=6, fig.width=2.3}
sample_meta_participant = extended_data$sample_meta_participant_level
sample_meta_participant$death_date[sample_meta_participant$death_date<0] = 999
censoring_date_death = 200
death_surv = cbind(sample_meta_participant$death_date,
                   sample_meta_participant$death_date)
death_surv[death_surv[,2]>censoring_date_death,2] = censoring_date_death+1
death_surv[,1] = ifelse(death_surv[,2]<=censoring_date_death,1,0)
death_surv = data.frame(death_surv)
colnames(death_surv) = c("state", "time")

id_post_acute <- which(death_surv$time > 28 & subtype_hcl$label != "D")
post_acute_label = array("Non-respondent",
                         nrow(extended_data$sample_meta_participant_level))
idx_PASC = which(!is.na(subtype_hcl$PASC))
post_acute_label[idx_PASC] = "Any Deficit"
post_acute_label[subtype_hcl$PASC == "minimal"] = "Minimal Deficit"
post_acute_label = factor(post_acute_label, 
                          levels = c("Minimal Deficit", "Any Deficit",
                                     "Non-respondent"))

clin_data_patient <- clin_data_aligned[clin_data_aligned$event_type == "Visit 1",]
ww <- ifelse(post_acute_label == "Non-respondent", 1, 0)

demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", 
                                          "sex", "respiratory_status")]
demo_data_adjust$respiratory_status <-
  ifelse(demo_data_adjust$respiratory_status == 6, 1, 0)
demo_data_adjust$respiratory_status <- factor(demo_data_adjust$respiratory_status)

idx_non_resp = which(post_acute_label == "Minimal Deficit" |
                       post_acute_label == "Non-respondent")
idx_non_resp = intersect(idx_non_resp, id_post_acute)

missing_mv_adjusted_test <- test_clinical_binomial_demo(
  subtype_hcl$label[idx_non_resp],
  ww[idx_non_resp],
  demo_data_adjust[idx_non_resp,])

heatmap_binomial(missing_mv_adjusted_test, col_name = "Non-\nrespondent")
```

# Test any deficit adjusted for mechanical ventilation (Extended Data Fig10D)
```{r test any deficit adjusted for mechanical ventilation, fig.height=6, fig.width=2.3}
id_pasc = which(!is.na(subtype_hcl$PASC) & subtype_hcl$label != "D")
ww = ifelse(subtype_hcl$PASC == "minimal", 0, 1)
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", 
                                          "sex", "respiratory_status")]
demo_data_adjust$respiratory_status <-
  ifelse(demo_data_adjust$respiratory_status == 6, 1, 0)
demo_data_adjust$respiratory_status <- factor(demo_data_adjust$respiratory_status)

any_deficit_demo_me_adjusted_test <- test_clinical_binomial_demo(
  subtype = subtype_hcl$label[id_pasc],
  response_binary = ww[id_pasc],
  demo_data = demo_data_adjust[id_pasc,])

heatmap_binomial(any_deficit_demo_me_adjusted_test, col_name = "Any Deficit")
```

# Test PASC adjusted for mechanical ventilation (Extended Data Fig10D)
```{r test PASC adjusted for mechanical ventilation, fig.height=6, fig.width=4.2}
full_clust_test = list(); full_clust_test[["PASC"]] = list()
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", 
                                          "sex", "respiratory_status")]
demo_data_adjust$respiratory_status <-
  ifelse(demo_data_adjust$respiratory_status == 6, 1, 0)
demo_data_adjust$respiratory_status <- factor(demo_data_adjust$respiratory_status)

full_clust_test[["PASC"]][["regression_me"]] = test_multi_pasc_demo(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = demo_data_adjust)

heatmap_pasc_multi(full_clust_test[["PASC"]][["regression_me"]],
                   "Demographics + Mechanical Ventilation")
```

# PASC comparison for E-F
```{r PASC comparison for E-F}
# any deficit (F vs E)
id_pasc = which(!is.na(subtype_hcl$PASC) & subtype_hcl$label %in% c("E", "F"))
ww = ifelse(subtype_hcl$PASC == "minimal", 0, 1)
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", 
                                          "sex", "respiratory_status")]
demo_data_adjust$respiratory_status <-
  ifelse(demo_data_adjust$respiratory_status == 6, 1, 0)
demo_data_adjust$respiratory_status <- factor(demo_data_adjust$respiratory_status)

df_binom = data.frame(PASC = ww, label = subtype_hcl$label, demo_data_adjust)
df_binom = df_binom[id_pasc,]

model_pasc_EF = glm(PASC ~ ., family = binomial, data = df_binom)
coef(summary(model_pasc_EF))

# cognitive deficit (F vs E)
id_pasc = which(subtype_hcl$PASC %in% c("cognitive", "minimal") & 
                  subtype_hcl$label %in% c("E", "F"))
ww = ifelse(subtype_hcl$PASC == "minimal", 0, 1)
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", 
                                          "sex", "respiratory_status")]
demo_data_adjust$respiratory_status <-
  ifelse(demo_data_adjust$respiratory_status == 6, 1, 0)
demo_data_adjust$respiratory_status <- factor(demo_data_adjust$respiratory_status)

df_binom = data.frame(PASC = ww, label = subtype_hcl$label, demo_data_adjust)
df_binom = df_binom[id_pasc,]

model_pasc_EF = glm(PASC ~ ., family = binomial, data = df_binom)
coef(summary(model_pasc_EF))
```

# Test PASC adjusted for mechanical ventilation - global test (Extended Data Fig10D)
```{r test PASC adjusted for mechanical ventilation - global test, fig.height=6, fig.width=2}
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", 
                                          "sex", "respiratory_status")]
demo_data_adjust$respiratory_status <-
  ifelse(demo_data_adjust$respiratory_status == 6, 1, 0)
demo_data_adjust$respiratory_status <- factor(demo_data_adjust$respiratory_status)

global_test_demo = test_global_multi_pasc(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = demo_data_adjust)

heatmap_multi_global <- function(test_global) {
  df_global <- data.frame("Global" = -log10(test_global))
  df_global <- pivot_longer(df_global, cols = everything())
  df_global$subtype <- c("A", "B", "C", "E", "F")
  df_global$subtype <- factor(df_global$subtype)
  df_global$subtype <- factor(df_global$subtype,
                              levels = rev(levels(df_global$subtype)))
  
  df_global %>% ggplot(mapping = aes(x = name, y = subtype, size = value)) +
    geom_point(color = "grey30") +
    cowplot::theme_cowplot() +
    theme(axis.line  = element_blank()) +
    ylab("") +
    xlab("") +
    theme(axis.ticks = element_blank(),
          axis.text.y = element_text(size = 15, hjust = 0),
          axis.text.x = element_text(size = 15, angle = 90, vjust = 0.5),
          legend.position = "left",
          legend.title = element_text(size = 15, angle = 90),
          legend.text = element_text(size = 15, angle = 90),
          plot.margin = margin(b = 20)) +
    labs(size = expression(-log[10]*"("*italic("P")*"-value"*")")) +
    geom_text(aes(label = ifelse(abs(value) >= -log10(0.001), "***",
                                 ifelse(abs(value) >= -log10(0.01), "**",
                                        ifelse(abs(value) >= -log10(0.05), "*", "")))),
              size = 7, vjust = 0.78, color = "white") +
    scale_size(range = c(3, 12), limits = c(0, 2.0),
               breaks = c(0.4, 0.8, 1.2, 1.6, 2.0), 
               labels = c("0.4", "0.8", "1.2", "1.6", "2.0"))
}

heatmap_multi_global(global_test_demo)
```



