---
title: "medication"
output: pdf_document
---

```{r load library}
library(ade4) #used in mcia
library(mogsa)
library(factoextra)
library(cluster)
library(nnet)
library(survival)
library(ordinal)
library(ggeffects)
library(openxlsx)
library(network)
library(GGally)
library(ggalt)
library(metafor)
library(cowplot)
library(survminer)
require(nlme) # directly provides p-values, unlike lme4
require(splines)
require(dplyr)
library(fgsea)
codebase_path = "~/impacc/Analysis/molecular_subtyping/codebase"
source(paste0(codebase_path,"/data_analysis_template_codebase.R"))
source(paste0(codebase_path,"/codebase_basic.R"))
source(paste0(codebase_path,"/codebase_prediction.R"))
source(paste0(codebase_path,"/codebase_interpretation.R"))
source(paste0(codebase_path,"/class_structures.R"))
source(paste0(codebase_path,"/codebase_mod.R"))
source(paste0(codebase_path,"/codebase_subtype.R"))

organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")

intermediate_results_outdir_DR = "/scratch/data-integration/molecular_subtyping/DimensionReduction/"
mcia_results = readRDS(paste0(intermediate_results_outdir_DR,"MCIA_phase123_20230726.rds"))

data_path <- "/scratch/data-integration/pipelines/processedData/"
data_env0 <- readRDS(paste0(data_path, "data_env_ravi_20220513.RDS"))
```

# Load subtype assignment and clinical data
```{r load subtype assignment and clinical data}
nclust_tree_all = 6
subtype_hcl = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
clin_data_aligned = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/clin_data_aligned.rds")
extended_data = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/extended_data.rds")
```

# Medication (Extended Data Fig10G)
```{r, fig.height=6, fig.width=5.5, warning=FALSE}
clin_data_patient <- clin_data_aligned[clin_data_aligned$event_type == "Visit 1",]
demo_data <- clin_data_patient[, c("admit_age", "sex", "ethnicity",
                                   "discretized_admit_age_quantile",
                                   "bmicat", "enrollment_site", "race")]
demo_data_adjust <- demo_data[, c("discretized_admit_age_quantile", "sex")]

med_data_logit <- clin_data_patient %>%
  dplyr::select(matches("^ever")) %>%
  dplyr::select(-ever_ever_esc, -ever_icu) %>%
  select_if(~sum(.x, na.rm = TRUE) > 0)

sub_idx <- which(subtype_hcl$label != "D")
med_test_logit <- test_clinical_logit_demo(subtype_hcl$label[sub_idx],
                                           med_data_logit[sub_idx,],
                                           demo_data_adjust[sub_idx,])
med_test_logit$p_lr_adj <- p.adjust(med_test_logit$p_lr, method = "BH")

pvec <- as.vector(as.matrix(med_test_logit$p_logit))
pvec_bh <- p.adjust(pvec, method = "BH")
med_test_logit$p_adj <- matrix(pvec_bh, nrow = nrow(med_test_logit$p_logit), 
                          ncol = ncol(med_test_logit$p_logit))
rownames(med_test_logit$p_adj) <- rownames(med_test_logit$p_logit)
colnames(med_test_logit$p_adj) <- colnames(med_test_logit$p_logit)
# med_test_logit$p_adj <- apply(med_test_logit$p_logit, 2, 
#                               function(p) p.adjust(p, method = "BH"))

med_names <- c("ever_remdes" = "Remdesivir",
               "ever_steroids" = "Steroids",
               "ever_azithro" = "Azithromycin",
               "ever_otherabx" = "Other Antibiotics",
               "ever_conplas" = "Convalescent Plasma",
               "ever_monoclonal" = "Monoclonal Antibodies",
               "ever_immsupp" = "Immunosuppressives")
heatmap_med(med_test_logit, med_names)
```

# Medication - global test (Extended Data Fig10G)
```{r medication - global test, fig.height=1.5, fig.width=5.5, warning=FALSE}
heatmap_med_global(med_test_logit, med_names)
```

# Test missingness adjusted for medications (Extended Data Fig10H)
```{r test missingness adjusted for medications, fig.height=6, fig.width=2.3}
sample_meta_participant = extended_data$sample_meta_participant_level
sample_meta_participant$death_date[sample_meta_participant$death_date<0] = 999
censoring_date_death = 200
death_surv = cbind(sample_meta_participant$death_date,
                   sample_meta_participant$death_date)
death_surv[death_surv[,2]>censoring_date_death,2] = censoring_date_death+1
death_surv[,1] = ifelse(death_surv[,2]<=censoring_date_death,1,0)
death_surv = data.frame(death_surv)
colnames(death_surv) = c("state", "time")

id_post_acute <- which(death_surv$time > 28 & subtype_hcl$label != "D")
post_acute_label = array("Non-respondent",
                         nrow(extended_data$sample_meta_participant_level))
idx_PASC = which(!is.na(subtype_hcl$PASC))
post_acute_label[idx_PASC] = "Any Deficit"
post_acute_label[subtype_hcl$PASC == "minimal"] = "Minimal Deficit"
post_acute_label = factor(post_acute_label, 
                          levels = c("Minimal Deficit", "Any Deficit",
                                     "Non-respondent"))

clin_data_patient <- clin_data_aligned[clin_data_aligned$event_type == "Visit 1",]
demo_data_adjust <- clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
ww <- ifelse(post_acute_label == "Non-respondent", 1, 0)
sig_med = names(med_test_logit$p_lr_adj)[med_test_logit$p_lr_adj <= 0.05]
med_data_adjust = clin_data_patient[, sig_med]

idx_non_resp = which(post_acute_label == "Minimal Deficit" |
                       post_acute_label == "Non-respondent")
idx_non_resp = intersect(idx_non_resp, id_post_acute)

missing_demo_adjusted_test <- test_clinical_binomial_demo(
  subtype_hcl$label[idx_non_resp],
  ww[idx_non_resp],
  demo_data_adjust[idx_non_resp,])

missing_demo_rem_adjusted_test <- test_clinical_binomial_demo(
  subtype_hcl$label[idx_non_resp],
  ww[idx_non_resp],
  cbind(demo_data_adjust, 
        clin_data_patient[,c("ever_remdes")])[idx_non_resp,])

missing_demo_med_adjusted_test <- test_clinical_binomial_demo(
  subtype_hcl$label[idx_non_resp],
  ww[idx_non_resp],
  cbind(demo_data_adjust,
        med_data_adjust)[idx_non_resp,])

heatmap_binomial(missing_demo_adjusted_test, col_name = "Non-\nrespondent")
heatmap_binomial(missing_demo_rem_adjusted_test, col_name = "Non-\nrespondent")
heatmap_binomial(missing_demo_med_adjusted_test, col_name = "Non-\nrespondent")
```

# Test any deficit adjusted for medications (Extended Data Fig10H)
```{r test any deficit adjusted for medications, fig.height=6, fig.width=2.3}
id_pasc = which(!is.na(subtype_hcl$PASC) & subtype_hcl$label != "D")
ww = ifelse(subtype_hcl$PASC == "minimal", 0, 1)
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
sig_med = names(med_test_logit$p_lr_adj)[med_test_logit$p_lr_adj <= 0.05]
med_data_adjust = clin_data_patient[, sig_med]

any_deficit_demo_adjusted_test <- test_clinical_binomial_demo(
  subtype = subtype_hcl$label[id_pasc],
  response_binary = ww[id_pasc],
  demo_data = demo_data_adjust[id_pasc,])

any_deficit_demo_rem_adjusted_test <- test_clinical_binomial_demo(
  subtype = subtype_hcl$label[id_pasc],
  response_binary = ww[id_pasc],
  demo_data = cbind(demo_data_adjust,
                    clin_data_patient[, c("ever_remdes")])[id_pasc,])

any_deficit_demo_med_adjusted_test <- test_clinical_binomial_demo(
  subtype = subtype_hcl$label[id_pasc],
  response_binary = ww[id_pasc],
  demo_data = cbind(demo_data_adjust,
                    med_data_adjust)[id_pasc,])

heatmap_binomial(any_deficit_demo_adjusted_test, col_name = "Any Deficit")
heatmap_binomial(any_deficit_demo_rem_adjusted_test, col_name = "Any Deficit")
heatmap_binomial(any_deficit_demo_med_adjusted_test, col_name = "Any Deficit")
```

# Test PASC adjusted for medications (Extended Data Fig10H)
```{r test PASC adjusted for medications, fig.height=6, fig.width=4.2}
full_clust_test = list(); full_clust_test[["PASC"]] = list()
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
sig_med = names(med_test_logit$p_lr_adj)[med_test_logit$p_lr_adj <= 0.05]
med_data_adjust = clin_data_patient[, sig_med]

full_clust_test[["PASC"]][["regression_demo"]] = test_multi_pasc_demo(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = demo_data_adjust)

full_clust_test[["PASC"]][["regression_remdes"]] = test_multi_pasc_demo(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = cbind(demo_data_adjust, 
                    clin_data_patient[,c("ever_remdes")]))

full_clust_test[["PASC"]][["regression_med"]] = test_multi_pasc_demo(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = cbind(demo_data_adjust, med_data_adjust))

heatmap_pasc_multi(full_clust_test[["PASC"]][["regression_demo"]],
                   "Demographics Adjusted")
heatmap_pasc_multi(full_clust_test[["PASC"]][["regression_remdes"]],
                   "Demographics + Remdesivir")
heatmap_pasc_multi(full_clust_test[["PASC"]][["regression_med"]],
                   "Demographics + Medications")
```

# Test PASC adjusted for medications - global test (Extended Data Fig10H)
```{r test PASC adjusted for medications - global test, fig.height=6, fig.width=2}
clin_data_patient = clin_data_aligned[clin_data_aligned$event_type == "Visit 1", ]
demo_data_adjust = clin_data_patient[, c("discretized_admit_age_quantile", "sex")]
sig_med = names(med_test_logit$p_lr_adj)[med_test_logit$p_lr_adj <= 0.05]
med_data_adjust = clin_data_patient[, sig_med]

global_test_demo = test_global_multi_pasc(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = demo_data_adjust)

global_test_remdes = test_global_multi_pasc(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = cbind(demo_data_adjust, 
                    clin_data_patient[,c("ever_remdes")]))

global_test_med = test_global_multi_pasc(
  subtype = subtype_hcl$label,
  response = subtype_hcl$PASC,
  demo_data = cbind(demo_data_adjust, med_data_adjust))

heatmap_multi_global(global_test_demo)
heatmap_multi_global(global_test_remdes)
heatmap_multi_global(global_test_med)
```



