---
title: "Code_Yutong"
output: html_document
date: "2024-01-23"
---

Prediction for PASC outcomes
```{r}
library(dplyr)
library(pROC)
library(caret)
library(grid)
library(ComplexHeatmap)
```

```{r}
# factor scores
mcia_results <- readRDS("/scratch/data-integration/molecular_subtyping/DimensionReduction/MCIA_phase123_20230726.rds")
organized_data = readRDS("/scratch/data-integration/molecular_subtyping/Data Preprocessing/phases123_combined_jeremy_072023.rds")
clin_data_all = organized_data$clinical_data[,setdiff(colnames(organized_data$clinical_data), c("sample_id","sample_type", "phase1_2","samples_collected","phase","enrollment_site"))]
clin_data_all = unique(clin_data_all)
clin_data_aligned = data_frame(event_id=rownames(organized_data$datasets$plasma_proteomics_targeted))
clin_data_aligned <- clin_data_aligned %>% left_join(clin_data_all, by = c("event_id"))
ll = which(clin_data_aligned$event_type %in% paste0("Visit ",1:6))
clin_data_aligned <- clin_data_aligned[ll,]
mcia_factors = mcia_results$mcia_factors
rownames(mcia_factors) <- gsub('-([0-9])$', '-0\\1', rownames(organized_data$datasets_imputed[[1]]))
mcia_factors = mcia_factors[ll,]
n_factor = 40
factor_scores = mcia_factors[,1:n_factor]
factor_scores = apply(factor_scores, 2,function(z) qqnorm(z, plot.it = F)$x)
# update factors that explain more consistent variation across assays
aa = (apply(sqrt(mcia_results$mcia_factor_evaluation$ave_variance_explained[1:n_factor,]),1,sum))^2
factor_scores = t(apply(factor_scores,1,function(z) z*sqrt(aa)))
rownames(factor_scores) <- rownames(mcia_factors)
factor_scores <- as.data.frame(factor_scores)
factor_scores$participant_id <- substr(rownames(factor_scores), start = 1, stop = 8)

# ema of factor scores
mov_ave <- function(a){
  mu <- 1 - 2/(length(a)+2)
  res <- a[1]
  for (k in 1:length(a))  res <- mu*a[k] + (1-mu)*res
  return(res)
}
factor_scores_ema <- aggregate(. ~ participant_id, data = factor_scores, FUN = mov_ave)
rownames(factor_scores_ema) <- factor_scores_ema$participant_id
factor_scores_ema$participant_id <- NULL

# V1 of factor scores
factor_scores_v1 <- factor_scores[!duplicated(factor_scores$participant_id),]
# factor_scores_v1 <- factor_scores[substr(rownames(factor_scores), start = 10, stop = 11)=="01",]
rownames(factor_scores_v1) <- factor_scores_v1$participant_id
factor_scores_v1$participant_id <- NULL
```

```{r}
# subtype assignments
subtype_labels <- readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/subtype_hcl.rds")
subtype_labels$merged_labels <- subtype_labels$label
label_map <- c(A = 1, B = 2, C = 3, D = 4, E = 5, F = 6)
subtype_labels$merged_labels <- label_map[subtype_labels$merged_labels]

# probability vector
factor_scores$visit <- clin_data_aligned$event_type
factor_scores$subtype <- subtype_labels[factor_scores$participant_id,]$merged_labels
factor_scores <- na.omit(factor_scores)
subtype_centers <- list()
for (i in 1:6){
  tmp <- factor_scores[factor_scores$subtype ==i,]
  visits <- unique(tmp$visit)
  subtype_centers[[i]] <- 1:40
  for (v in visits){
    subtype_centers[[i]] <- rbind(subtype_centers[[i]], colMeans(tmp[tmp$visit == v, 1:40]))
  }
  subtype_centers[[i]] <- as.data.frame(subtype_centers[[i]][-1,])
  rownames(subtype_centers[[i]]) <- visits
}
copy <- factor_scores
copy$dist <- rowSums((copy[,1:40]-subtype_centers[[1]][copy$visit,])^2)
copy <- na.omit(copy)
dist_to_subtype <- aggregate(dist ~ participant_id, data = copy, FUN = mean)
rownames(dist_to_subtype) <- dist_to_subtype$participant_id
dist_to_subtype$participant_id <- NULL
colnames(dist_to_subtype) <- c("S1")
for (i in 2:6){
  copy <- factor_scores
  copy$dist <- rowSums((copy[,1:40]-subtype_centers[[i]][copy$visit,])^2)
  copy <- na.omit(copy)
  dist <- aggregate(dist ~ participant_id, data = copy, FUN = mean)
  dist_to_subtype[dist$participant_id, paste0("S",i)] <- dist$dist
}
softmax <- function(x) {
  exp_x <- exp(x - max(x))
  return(exp_x / sum(exp_x))
}
probs <- as.data.frame(t(apply(-dist_to_subtype, FUN = softmax, MARGIN = 1)))
colnames(probs) <- c("p1","p2","p3","p4","p5","p6")
log_probs <- log(probs)

# check assignment accuracy
print(sum(apply(probs,1,which.max)==subtype_labels$merged_labels)/dim(probs)[1])
```

```{r}
# clinical data
clinical_data <- clin_data_aligned[!duplicated(clin_data_aligned$participant_id),]
clinical_data <- as.data.frame(clinical_data)
rownames(clinical_data) <- clinical_data$participant_id
clinical_data <- clinical_data %>% dplyr::select(participant_id, admit_age, sex,  
                                                  bmicat, los, ncomp, 
                                                  baseline_sofa_score_new, 
                                                  matches("comorb"))
clinical_data$comorb_countcat <- NULL
clinical_data[clinical_data$sex=='Female',]$sex <- 1
clinical_data[clinical_data$sex=='Male',]$sex <- 0
clinical_data$sex <- as.numeric(clinical_data$sex)
base_obj <- readRDS('/scratch/data-integration/cleanedup/mod1_interpretation_base.rds')
spike_IgG <- rbind(base_obj$others$serum_rbd_abtiters$train$serum_rbd_abtiters_counts,
                   base_obj$others$serum_rbd_abtiters$test$serum_rbd_abtiters_counts)

spike_IgG$N1_CT <- rbind(base_obj$others$nasal_viralload$train$nasal_viralload_counts,
                         base_obj$others$nasal_viralload$test$nasal_viralload_counts)$N1_CT
spike_IgG <- spike_IgG[,c('AUC Spike IgG','N1_CT')]
spike_IgG <- na.omit(spike_IgG)
spike_IgG$participant_id <- substr(rownames(spike_IgG), start = 1, stop = 8)

# ema of viral loads and antibody
spike_IgG_ema <- aggregate(. ~ participant_id, data = spike_IgG, FUN=mov_ave)
rownames(spike_IgG_ema) <- spike_IgG_ema$participant_id

# V1 of viral loads and antibody
spike_IgG_v1 <- spike_IgG[!duplicated(spike_IgG$participant_id),]
# spike_IgG_v1 <- spike_IgG[substr(rownames(spike_IgG), start = 10, stop = 11)=="01",]
rownames(spike_IgG_v1) <- spike_IgG_v1$participant_id

clinical_data_ema <- merge(clinical_data, spike_IgG_ema, by='participant_id')
rownames(clinical_data_ema) <- clinical_data_ema$participant_id
clinical_data_ema$participant_id <- NULL
clinical_data_ema <- na.omit(clinical_data_ema)
clinical_data_ema <- clinical_data_ema[clinical_data_ema$bmicat!=999,]

clinical_data_v1 <- merge(clinical_data, spike_IgG_v1, by='participant_id')
rownames(clinical_data_v1) <- clinical_data_v1$participant_id
clinical_data_v1$participant_id <- NULL
clinical_data_v1 <- na.omit(clinical_data_v1)
clinical_data_v1 <- clinical_data_v1[clinical_data_v1$bmicat!=999,]
```

```{r}
# PASC
conval_raw <- organized_data$clinical_data$conv_cluster6
names(conval_raw) <- organized_data$clinical_data$event_id
conval_raw <- na.omit(conval_raw)
conval_raw[conval_raw == 3 | conval_raw == 4] = 1
conval_raw[conval_raw == 5] = 3
conval_raw[conval_raw == 6] = 4
conval_raw <- conval_raw[!duplicated(gsub('-[0-9]+$', '', names(conval_raw)))]
names(conval_raw) <- gsub('-[0-9]+$', '', names(conval_raw))
conval_raw <- conval_raw[sort(names(conval_raw))]
```

```{r}
# data for prediction
# ema data
id <- Reduce(intersect, list(rownames(clinical_data_ema), 
                             rownames(factor_scores_ema), 
                             rownames(log_probs),
                             names(conval_raw)))
data_pred_ema <- list()
data_pred_ema[['conval']] <- conval_raw[id]
data_pred_ema[['clinical']] <- clinical_data_ema[id,]
data_pred_ema[['subtype']] <- log_probs[id,]
data_pred_ema[['factor']] <- factor_scores_ema[id,]

# V1 data
id_v1 <- Reduce(intersect, list(rownames(clinical_data_v1), 
                             rownames(factor_scores_v1), 
                             rownames(log_probs),
                             names(conval_raw)))
data_pred_v1 <- list()
data_pred_v1[['conval']] <- conval_raw[id_v1]
data_pred_v1[['clinical']] <- clinical_data_v1[id_v1,]
data_pred_v1[['subtype']] <- log_probs[id_v1,]
data_pred_v1[['factor']] <- factor_scores_v1[id_v1,]
```

```{r}
prediction <- function(data, pred_model) { 
  models <- list()
  test_sets <- list()
  conval <- data$conval
  repeats <- 10
  rocs <- vector("list", repeats)
  prcs <- vector("list", repeats)
  aucs <- rep(0, repeats)
  auprcs <- rep(0, repeats)
  set.seed(3261) 
  for (s in 1:repeats) {
    # split data
    ind_train <- sample(1:length(conval), floor(length(conval)*0.7))
    conval_train <- conval[ind_train]
    conval_test <- conval[-ind_train]

    # binary response
    conval01_train <- conval_train
    conval01_train[conval01_train != 1] <- 0
    conval01_test <- conval_test
    conval01_test[conval01_test != 1] <- 0

    d_train <- data.frame(Column1 = 1:length(conval01_train))
    d_test <- data.frame(Column1 = 1:length(conval01_test))
    rownames(d_train) <- names(conval01_train)
    rownames(d_test) <- names(conval01_test)
    # clinical
    d_train <- cbind(d_train, data$clinical[rownames(d_train),])
    d_test <- cbind(d_test, data$clinical[rownames(d_test),])
    # subtype
    if (pred_model != "CLINICAL") {
      d_train <- cbind(d_train, data$subtype[rownames(d_train),])
      d_test <- cbind(d_test, data$subtype[rownames(d_test),])
    }
    # factor scores
    if (pred_model == "CLINICAL-SUBTYPE+") {
      d_train <- cbind(d_train, data$factor[rownames(d_train),])
      d_test <- cbind(d_test, data$factor[rownames(d_test),])
    }
    
    d_train$Column1 <- NULL
    d_test$Column1 <- NULL

    d_train$conval <- conval01_train
    d_train$conval[d_train$conval == 0] <- "level0"
    d_train$conval[d_train$conval == 1] <- "level1"
    d_train$conval <- factor(d_train$conval)
    d_test$conval <- conval01_test
    d_test$conval[d_test$conval == 0] <- "level0"
    d_test$conval[d_test$conval == 1] <- "level1"
    d_test$conval <- factor(d_test$conval)

    control <- trainControl(method = "cv", number = 5, summaryFunction = twoClassSummary, classProbs = TRUE)
    model = train(conval ~ ., data = d_train, trControl = control, method = "svmRadial", metric = "ROC")
    pred <- predict(model, d_test, type = 'prob')[,2]

    rocs[[s]] <- pROC::roc(d_test$conval, pred)
    aucs[s] <- rocs[[s]]$auc
    print(aucs[s])
    
    prcs[[s]] <- PRROC::pr.curve(scores.class0 = 1-pred, weights.class0 = 1-conval01_test, curve = TRUE)
    auprcs[s] <- prcs[[s]]$auc.integral
    print(auprcs[s])
    
    models[[s]] <- model
    test_sets[[s]] <- d_test
  }
  print(paste0("auc mean: ", mean(aucs)))
  print(paste0("auc std: ", sd(aucs)))
  print(paste0("auprc mean: ", mean(auprcs)))
  print(paste0("auprc std: ", sd(auprcs)))
  res <- list(aucs, rocs, auprcs, prcs, models, test_sets)
  names(res) <- c("aucs", "rocs", "auprcs", "prcs", "models", "test_sets")
  return(res)
}
```

```{r}
res <- prediction(data_pred_ema, "CLINICAL")
aucs_clinical <- res$aucs
rocs_clinical <- res$rocs
auprcs_clinical <- res$auprcs
prcs_clinical <- res$prcs
models_clinical <- res$models
test_sets_clinical <- res$test_sets

res <- prediction(data_pred_ema, "CLINICAL-SUBTYPE")
aucs_clinical_prob <- res$aucs
rocs_clinical_prob <- res$rocs
auprcs_clinical_prob <- res$auprcs
prcs_clinical_prob <- res$prcs
models_clinical_prob <- res$models
test_sets_clinical_prob <- res$test_sets

res <- prediction(data_pred_ema, "CLINICAL-SUBTYPE+")
aucs_clinical_factor_prob <- res$aucs
rocs_clinical_factor_prob <- res$rocs
auprcs_clinical_factor_prob <- res$auprcs
prcs_clinical_factor_prob <- res$prcs
models_clinical_factor_prob <- res$models
test_sets_clinical_factor_prob <- res$test_sets
```



# Visualization
# ROC plots
```{r}
library(cutpointr)
library(tidyverse)

# prepare data
fig_names <- c('CLINICAL', 'CLINICAL-SUBTYPE', 'CLINICAL-SUBTYPE+')

aucs <- list(aucs_clinical, aucs_clinical_prob, aucs_clinical_factor_prob)
rocs <- list(rocs_clinical, rocs_clinical_prob, rocs_clinical_factor_prob)
roc_data <- NULL
ci_data <- NULL
for (i in 1:length(fig_names)){
  fig_name <- fig_names[i]
  rocs_svm <- rocs[[i]]
  aucs_svm <- aucs[[i]]
  
  # compute average roc curves
  predictions <- rocs_svm[[1]]$predictor
  for (j in 2:10){
    predictions <- rbind(predictions, rocs_svm[[j]]$predictor)
  }
  
  responses <- rocs_svm[[1]]$response
  for (j in 2:10){
    responses <- rbind(responses, rocs_svm[[j]]$response)
  }
  
  predictions_10_samples <- data.frame(
      Sample = rep(c(1:10), times = length(rocs_svm[[1]]$predictor)),
      PredictionValues = as.numeric(predictions),
      RealClass = as.numeric(responses)
  )
  
  mean_roc <- function(data, cutoffs = seq(from = -1, to = 2, by = 0.01)) {
      map_df(cutoffs, function(cp) {
          out <- cutpointr(data = data, x = PredictionValues, class = RealClass,
                           subgroup = Sample, method = oc_manual, cutpoint = cp,
                           pos_class = 2, direction = ">=")
          data.frame(cutoff = cp,
                     sensitivity = mean(out$sensitivity),
                     specificity = mean(out$specificity))
      })
  }
  mr <- mean_roc(predictions_10_samples)
  
  df <- data.frame(
    CI_low = pmax(rev(mr$sensitivity) - sd(aucs_svm), 0),
    CI_high = pmin(rev(mr$sensitivity) + sd(aucs_svm), 1),
    FPR = rev(1 - mr$specificity),
    Model = fig_name
  )
  ci_data <- rbind(ci_data, df)
  
  df <- data.frame(
    TPR = rev(mr$sensitivity),
    FPR = rev(1 - mr$specificity),
    Label = "average",
    Model = fig_name
  )
  roc_data <- rbind(roc_data, df)
  
  df <- data.frame(
    TPR = seq(0, 1, 0.01),
    FPR = seq(0, 1, 0.01),
    Label = "chance",
    Model = fig_name
  )
  roc_data <- rbind(roc_data, df)
  
  for (j in 1:10){
    df <- data.frame(
      TPR = rev(rocs_svm[[j]]$sensitivities),
      FPR = rev(1 - rocs_svm[[j]]$specificities),
      Label = toString(j),
      Model = fig_name
    )
    roc_data <- rbind(roc_data, df)
  }
}
```

```{r fig.height=6, fig.width=18}
# ROC plots
mycol_purple <- rgb(171, 130, 255, max = 255, alpha = 70)
mycol_red <- rgb(233, 150, 122, max = 255, alpha = 70)
custom_colors <- c("1" = "mistyrose", "2" = "antiquewhite2", "3" = "paleturquoise",
                   "4" = "darkseagreen2", "5" = mycol_red, "6" = "peachpuff",
                   "7" = "honeydew3", "8" = "khaki1", "9" = "thistle1",
                   "10" = mycol_purple, "average" = "black", "chance" = "red")

custom_line_types <- c("1" = "solid", "2" = "solid", "3" = "solid", "4" = "solid", "5" = "solid",
                       "6" = "solid", "7" = "solid", "8" = "solid", "9" = "solid", "10" = "solid",
                       "average" = "solid", "chance" = "dashed")

custom_line_width <- c("1" = 0.8, "2" = 0.8, "3" = 0.8, "4" = 0.8, "5" = 0.8,
                       "6" = 0.8, "7" = 0.8, "8" = 0.8, "9" = 0.8, "10" = 0.8,
                       "average" = 1, "chance" = 0.8)

aucs <- list(aucs_clinical, aucs_clinical_prob, aucs_clinical_factor_prob)
texts <- c()
for (i in 1:length(aucs)){
  aucs_svm <- aucs[[i]]
  M <- mean(aucs_svm)
  SD <- sd(aucs_svm)
  texts[i] <- paste0("AUROC: ", sprintf("%.3f",round(M,3)), " (SD ", sprintf("%.3f",round(SD,3)), ")")
}

facet_labels <- data.frame(
  Model = c("CLINICAL", "CLINICAL-SUBTYPE", "CLINICAL-SUBTYPE+"),
  Text = texts,
  xPos = rep(0.7, 3), 
  yPos = rep(0.4, 3)  
)

ggplot() +
geom_line(data = roc_data, aes(x = FPR, y = TPR, color = Label, linewidth = Label, linetype = Label)) + 
scale_color_manual(values = custom_colors, breaks = c("chance", "average"),
                   labels = c("Chance (random classifier)        ", "Average testing AUROC        ")) +
scale_linewidth_manual(values = custom_line_width, breaks = c("chance", "average"), 
                       labels = c("Chance (random classifier)        ", "Average testing AUROC        ")) +  
scale_linetype_manual(values = custom_line_types, breaks = c("chance", "average"), 
                      labels = c("Chance (random classifier)        ", "Average testing AUROC        ")) + 
# Add the confidence intervals
geom_ribbon(data = ci_data, aes(x = FPR, ymin = CI_low, ymax = CI_high, fill = "CI"), alpha = 0.4) +
scale_fill_manual(values = c("CI" = "darkgray"), labels = "±1 SD") +
facet_wrap(~ Model, scales = "free_y") +
labs(x = "False positive rate", y = "True positive rate") +
scale_x_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
scale_y_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
# Customize the theme 
theme_minimal() +
theme(panel.spacing = unit(2, "lines"),
      panel.grid.major = element_blank(),  # Remove major grid lines
      panel.grid.minor = element_blank(),  # Remove minor grid lines
      panel.border = element_blank(),  # Remove panel border
      axis.line.x = element_line(linewidth = 1),  # Add x axis line
      axis.line.y = element_line(linewidth = 1),  # Add y axis line
      axis.ticks = element_line(linewidth = 1),  # axis ticks
      axis.text.x = element_text(color = "black", size = 15),  # axis text
      axis.text.y = element_text(color = "black", size = 15),  # axis text
      axis.title = element_text(color = "black", size = 15),  # axis titles 
      axis.ticks.length = unit(0.15, "cm"),  # Adjust the length of axis ticks
      strip.text = element_text(color = "black", size = 15, margin = margin(b = 20)),
      legend.position = "top",  # Position the legend at the top
      legend.justification = "left",  # Justify the legend to the left
      legend.title = element_blank(),
      legend.text = element_text(color = "black", size = 15),
      legend.margin = margin(t = 5, b = 10),
      legend.spacing.x = unit(1, "mm"),
      legend.key.width = unit(2, "cm"),
      plot.margin = margin(t = 10, r = 25, b = 10, l = 15),
      aspect.ratio = 0.85
    ) + 
guides(color = guide_legend(order = 1), linewidth = guide_legend(order = 1),
       linetype = guide_legend(order = 1), fill = guide_legend(order = 2)) + 
geom_text(color = "black", data = facet_labels, aes(x = xPos, y = yPos, label = Text), inherit.aes = FALSE, size = 5)
```


# PRC plots
```{r}
library(cutpointr)
library(tidyverse)

# prepare data
fig_names <- c('CLINICAL', 'CLINICAL-SUBTYPE', 'CLINICAL-SUBTYPE+')

auprcs <- list(auprcs_clinical, auprcs_clinical_prob, auprcs_clinical_factor_prob)
prcs <- list(prcs_clinical, prcs_clinical_prob, prcs_clinical_factor_prob)
rocs <- list(rocs_clinical, rocs_clinical_prob, rocs_clinical_factor_prob)
test_sets <- list(test_sets_clinical, test_sets_clinical_prob, test_sets_clinical_factor_prob)
prc_data <- NULL
ci_data <- NULL

for (i in 1:length(fig_names)){
  fig_name <- fig_names[i]
  rocs_svm <- rocs[[i]]
  prcs_svm <- prcs[[i]]
  auprcs_svm <- auprcs[[i]]
  test_sets_svm <- test_sets[[i]]
  
  # compute average prc
  predictions_10_samples <- list()
  for (j in 1:10){
    predictions_10_samples[[j]] <- data.frame(predictions = 1-rocs_svm[[j]]$predictor,
                                              responses = ifelse(rocs_svm[[j]]$response=="level0", 1, 0)
                                             )
  }
  
  cutoffs <- seq(from = 0, to = 1, by = 0.01)
  # Function to calculate precision and recall
  calculate_metrics <- function(data, cutoff) {
    predicted_labels <- ifelse(data$predictions >= cutoff, 1, 0)
    labels <- data$responses
    true_positives <- sum(predicted_labels == 1 & labels == 1)
    false_positives <- sum(predicted_labels == 1 & labels == 0)
    false_negatives <- sum(predicted_labels == 0 & labels == 1)
    
    precision <- ifelse((true_positives + false_positives) > 0, true_positives / (true_positives + false_positives), 1)
    recall <- true_positives / max((true_positives + false_negatives),1)
    
    return(data.frame(precision=precision, recall=recall))
  }

  # Initialize a data frame to store average precision and recall for each cutoff
  mr <- data.frame(cutoff = numeric(), precision = numeric(), recall = numeric())
  
  # Calculate average precision and recall for each cutoff
  for (cutoff in cutoffs) {
    metrics_list <- t(sapply(predictions_10_samples, calculate_metrics, cutoff = cutoff))
    mr <- rbind(mr, data.frame(cutoff = cutoff, precision = mean(unlist(metrics_list[,1])), 
                               recall = mean(unlist(metrics_list[,2]))))
  }
  
  df <- data.frame(
    CI_low = pmax(rev(mr$precision) - sd(auprcs_svm), 0),
    CI_high = pmin(rev(mr$precision) + sd(auprcs_svm), 1),
    recall = rev(mr$recall),
    Model = fig_name
  )
  ci_data <- rbind(ci_data, df)
  
  df <- data.frame(
    precision = rev(mr$precision),
    recall = rev(mr$recall),
    Label = "average",
    Model = fig_name
  )
  prc_data <- rbind(prc_data, df)
  
  baseline = 0
  for (j in 1:10) {
    baseline = baseline + table(test_sets_svm[[j]]$conval)[1]/length(test_sets_svm[[j]]$conval)
  }
  baseline = baseline/10
  df <- data.frame(
    precision = rep(baseline, 101),
    recall = seq(0, 1, 0.01),
    Label = "chance",
    Model = fig_name
  )
  prc_data <- rbind(prc_data, df)
  
  for (j in 1:10){
    df <- data.frame(
      precision = rev(prcs_svm[[j]]$curve[,2]),
      recall = rev(prcs_svm[[j]]$curve[,1]),
      Label = toString(j),
      Model = fig_name
    )
    prc_data <- rbind(prc_data, df)
  }
}
```

```{r fig.height=6, fig.width=18}
# PRC plots
mycol_purple <- rgb(171, 130, 255, max = 255, alpha = 70)
mycol_red <- rgb(233, 150, 122, max = 255, alpha = 70)
custom_colors <- c("1" = "mistyrose", "2" = "antiquewhite2", "3" = "paleturquoise",
                   "4" = "darkseagreen2", "5" = mycol_red, "6" = "peachpuff",
                   "7" = "honeydew3", "8" = "khaki1", "9" = "thistle1",
                   "10" = mycol_purple, "average" = "black", "chance" = "red")

custom_line_types <- c("1" = "solid", "2" = "solid", "3" = "solid", "4" = "solid", "5" = "solid",
                       "6" = "solid", "7" = "solid", "8" = "solid", "9" = "solid", "10" = "solid",
                       "average" = "solid", "chance" = "dashed")

custom_line_width <- c("1" = 0.8, "2" = 0.8, "3" = 0.8, "4" = 0.8, "5" = 0.8,
                       "6" = 0.8, "7" = 0.8, "8" = 0.8, "9" = 0.8, "10" = 0.8,
                       "average" = 1, "chance" = 0.8)

auprcs <- list(auprcs_clinical, auprcs_clinical_prob, auprcs_clinical_factor_prob)
texts <- c()
for (i in 1:length(auprcs)){
  auprcs_svm <- auprcs[[i]]
  M <- mean(auprcs_svm)
  SD <- sd(auprcs_svm)
  texts[i] <- paste0("AUPRC: ", sprintf("%.3f",round(M,3)), " (SD ", sprintf("%.3f",round(SD,3)), ")")
}

facet_labels <- data.frame(
  Model = c("CLINICAL", "CLINICAL-SUBTYPE", "CLINICAL-SUBTYPE+"),
  Text = texts,
  xPos = rep(0.5, 3), 
  yPos = rep(0.2, 3)  
)

ggplot() +
geom_line(data = prc_data, aes(x = recall, y = precision, color = Label, linewidth = Label, linetype = Label)) + 
scale_color_manual(values = custom_colors, breaks = c("chance", "average"),
                   labels = c("Chance (random classifier)        ", "Average testing AUPRC        ")) +
scale_linewidth_manual(values = custom_line_width, breaks = c("chance", "average"), 
                       labels = c("Chance (random classifier)        ", "Average testing AUPRC        ")) +  
scale_linetype_manual(values = custom_line_types, breaks = c("chance", "average"), 
                      labels = c("Chance (random classifier)        ", "Average testing AUPRC        ")) + 
# Add the confidence intervals
geom_ribbon(data = ci_data, aes(x = recall, ymin = CI_low, ymax = CI_high, fill = "CI"), alpha = 0.4) +
scale_fill_manual(values = c("CI" = "darkgray"), labels = "±1 SD") +
facet_wrap(~ Model, scales = "free_y") +
labs(x = "Recall", y = "Precision") +
scale_x_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
scale_y_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
# Customize the theme 
theme_minimal() +
theme(panel.spacing = unit(2, "lines"),
      panel.grid.major = element_blank(),  # Remove major grid lines
      panel.grid.minor = element_blank(),  # Remove minor grid lines
      panel.border = element_blank(),  # Remove panel border
      axis.line.x = element_line(linewidth = 1),  # Add x axis line
      axis.line.y = element_line(linewidth = 1),  # Add y axis line
      axis.ticks = element_line(linewidth = 1),  # axis ticks
      axis.text.x = element_text(color = "black", size = 15),  # axis text
      axis.text.y = element_text(color = "black", size = 15),  # axis text
      axis.title = element_text(color = "black", size = 15),  # axis titles 
      axis.ticks.length = unit(0.15, "cm"),  # Adjust the length of axis ticks
      strip.text = element_text(color = "black", size = 15, margin = margin(b = 20)),
      legend.position = "top",  # Position the legend at the top
      legend.justification = "left",  # Justify the legend to the left
      legend.title = element_blank(),
      legend.text = element_text(color = "black", size = 15),
      legend.margin = margin(t = 5, b = 10),
      legend.spacing.x = unit(1, "mm"),
      legend.key.width = unit(2, "cm"),
      plot.margin = margin(t = 10, r = 25, b = 10, l = 15),
      aspect.ratio = 0.85
    ) + 
guides(color = guide_legend(order = 1), linewidth = guide_legend(order = 1),
       linetype = guide_legend(order = 1), fill = guide_legend(order = 2)) + 
geom_text(color = "black", data = facet_labels, aes(x = xPos, y = yPos, label = Text), inherit.aes = FALSE, size = 5)
```


# barplot of AUROC, AUPRC
```{r fig.height=6, fig.width=6}
df <- data.frame(
  category = c("Average AUROC", "Average AUPRC"),
  value1 = c(mean(aucs_clinical), mean(auprcs_clinical)),
  value2 = c(mean(aucs_clinical_prob), mean(auprcs_clinical_prob)),
  value3 = c(mean(aucs_clinical_factor_prob), mean(auprcs_clinical_factor_prob)),
  value4 = c(0.5, 0.3715517)
)

df <- tidyr::pivot_longer(df, cols = c(value1, value2, value3, value4), 
                          names_to = "model", values_to = "value")
# df$category <- factor(df$category, levels = c("Average AUROC", "Average AUPRC"))

aucs <- list(aucs_clinical, aucs_clinical_prob, aucs_clinical_factor_prob,
             auprcs_clinical, auprcs_clinical_prob, auprcs_clinical_factor_prob)
texts <- c()
for (i in 1:length(aucs)){
  aucs_svm <- aucs[[i]]
  M <- mean(aucs_svm)
  SD <- sd(aucs_svm)
  texts[i] <- paste0(sprintf("%.3f",round(M,3)), "\n(SD ", sprintf("%.3f",round(SD,3)), ")")
}
texts <- c(texts[1:3], "0.500", texts[4:6], "0.372")
df$text <- texts

# Create the bar plot
ggplot(df, aes(x = category, y = value, fill = model)) +
  geom_bar(data = subset(df, model != "value4"), 
           stat = "identity", position = "dodge") +
  geom_bar(data = subset(df, model == "value4"),
           stat = "identity", position = "dodge", alpha = 0.8) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 0.8), breaks = seq(0, 0.8, by = 0.2)) +
  scale_fill_manual(values = c(value1="#F8766D",value2="#7CAE00",value3="#00BFC4",value4="darkgray"),
                    labels = c("CLINICAL","CLINICAL-SUBTYPE","CLINICAL-SUBTYPE+","Chance (random classifier)")) +
  guides(fill = guide_legend(ncol = 2)) + 
  labs(x = "", y = "", fill = "") +
  theme_minimal() +
  theme(
        panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank(),  
        panel.border = element_blank(),  
        axis.line.x = element_line(linewidth = 1),  
        axis.line.y = element_line(linewidth = 1),  
        axis.ticks = element_line(linewidth = 1),  
        axis.text.x = element_text(color = "black", size = 15),  
        axis.text.y = element_text(color = "black", size = 15),  
        axis.title = element_text(color = "black", size = 15),  
        axis.ticks.length = unit(0.15, "cm"),  
        legend.position = "top", 
        legend.justification = "left",
        legend.direction = "vertical",
        legend.text = element_text(size = 15),
        plot.margin = margin(r = 20, l = 10),
        aspect.ratio = 0.85
  ) +
  geom_text(data = subset(df, model != "value4"),
            aes(label = text), position = position_dodge(width = 0.9),  
            vjust = -0.5, size = 3) +
  geom_text(data = subset(df, model == "value4"),
            aes(label = text), position = position_dodge(width = 0.9),  
            vjust = -0.5, size = 3)
```


# boxplot of AUROC and AUPRC increment
```{r fig.height=6, fig.width=6}
# Categories
categories <- rep(c("AUROC", "AUPRC"), each = 10)

# Values for the first box
values1 <- c(aucs_clinical_prob-aucs_clinical, auprcs_clinical_prob-auprcs_clinical)

# Values for the second box
values2 <- c(aucs_clinical_factor_prob-aucs_clinical, auprcs_clinical_factor_prob-auprcs_clinical)

# Combine into a data frame
df <- data.frame(category = rep(categories, 2),
                 value = c(values1, values2),
                 group = rep(c("CLINICAL-SUBTYPE", "CLINICAL-SUBTYPE+"), each = 20))

# add p-values
p.val <- function(x){
  return(wilcox.test(rep(0,10), x, paired = TRUE, alternative = "less")$p.value)
}
p_values <- aggregate(value ~ category + group, data = df, FUN = p.val)
p_values$value <- c("0.003 **", "0.002 **", "0.007 **", "<0.001 ***")

ggplot(df, aes(x = category, y = value, color = group)) +
  geom_boxplot(position = position_dodge(width = 0.8), size = 1, width = 0.7, outlier.shape = NA) +
  geom_jitter(aes(color = group), position = position_dodge(width = 0.8), size = 2.5, alpha = 0.5) +
  scale_y_continuous(limits = c(-0.015, 0.1), breaks = seq(0, 0.1, by = 0.02)) + 
  labs(x = "", y = "Increment over CLINICAL", color = NULL) +
  theme_minimal() +
  theme(panel.grid.major = element_blank(),  
        panel.grid.minor = element_blank(),  
        panel.border = element_blank(),  
        axis.line.x = element_line(linewidth = 1),  
        axis.line.y = element_line(linewidth = 1),  
        axis.ticks = element_line(linewidth = 1),  
        axis.text.x = element_text(color = "black", size = 15),  
        axis.text.y = element_text(color = "black", size = 15),  
        axis.title = element_text(color = "black", size = 15),  
        axis.ticks.length = unit(0.15, "cm"),  
        legend.position = "top",  
        # legend.justification = "left",
        # legend.direction = "vertical",
        legend.text = element_text(size = 15),
        plot.margin = margin(r = 20, l = 10),
        aspect.ratio = 0.85
  ) + 
  geom_text(data = p_values, aes(label = value, y = 0.1, x=c(0.8,1.82,1.22,2.22)), size = 5, vjust = 0, 
            position = position_dodge(width = 0.8), color = "black") +
  coord_cartesian(clip = "off")
```


# boxplot of AUC increase
```{r fig.height=6, fig.width=6}
# one-sided paired t test
# p1 <- t.test(aucs_clinical, aucs_clinical_prob, paired = TRUE, alternative = "less")$p.value
# p2 <- t.test(aucs_clinical, aucs_clinical_factor_prob, paired = TRUE, alternative = "less")$p.value
# one-sided paired Wilcoxon test
p1 <- wilcox.test(aucs_clinical, aucs_clinical_prob, paired = TRUE, alternative = "less")$p.value
p2 <- wilcox.test(aucs_clinical, aucs_clinical_factor_prob, paired = TRUE, alternative = "less")$p.value

data <- data.frame(
  Group = factor(rep(c("CLINICAL-SUBTYPE", "CLINICAL-SUBTYPE+"), each = 10)),
  Value = c(aucs_clinical_prob-aucs_clinical, aucs_clinical_factor_prob-aucs_clinical)
)

# Create the box plot
ggplot(data, aes(x = Group, y = Value, color = Group)) +
geom_boxplot(size = 1, outlier.shape = NA, width = 0.7) +
geom_jitter(size = 2.5, alpha = 0.7, aes(color = Group)) +  
# scale_color_manual(values = c("coral3", "cadetblue")) +
scale_y_continuous(expand = c(0, 0), limits = c(-0.005, 0.08), breaks = seq(0, 0.08, by = 0.02)) +
scale_x_discrete(expand = expansion(add = 0.55)) +
theme_minimal() +
theme(panel.grid.major = element_blank(),  # Remove major grid lines
      panel.grid.minor = element_blank(),  # Remove minor grid lines
      panel.border = element_blank(),  # Remove panel border
      axis.line.x = element_line(linewidth = 1),  # Add x axis line
      axis.line.y = element_line(linewidth = 1),  # Add y axis line
      axis.ticks = element_line(linewidth = 1),  # axis ticks
      axis.text.x = element_text(color = "black", size = 15),  
      axis.text.y = element_text(color = "black", size = 15),  
      axis.title = element_text(color = "black", size = 15),  # axis titles 
      axis.ticks.length = unit(0.15, "cm"),  # Adjust the length of axis ticks
      legend.position = "none",  # Position the legend at the top
      plot.margin = margin(t = 75, r = 20, b = 0, l = 10),
      aspect.ratio = 0.85
) +
labs(y = "AUROC increment over CLINICAL", x = "Models") + 
annotate("text", x = 0.75, y = 0.08, label = paste0("p.val = ", ifelse(p1 < 0.001, format(p1, scientific = TRUE, digits = 2), format(p1, digits = 2))), size = 5, color = "black", hjust = 0, vjust = 0) +
annotate("text", x = 1.75, y = 0.08, label = paste0("p.val = ", ifelse(p2 < 0.001, format(p2, scientific = TRUE, digits = 2), format(p2, digits = 2))), size = 5, color = "black", hjust = 0, vjust = 0) +
coord_cartesian(clip = "off")
```


# importance evaluation
```{r}
test_sets <- test_sets_clinical_factor_prob
models <- models_clinical_factor_prob
repeats <- 10
importance <- matrix(0, nrow = repeats*nrow(test_sets[[1]]), ncol = (ncol(test_sets[[1]])-1))
colnames(importance) <- colnames(test_sets[[1]])[-ncol(test_sets[[1]])]

for (s in 1:repeats){
  print(paste0("split ", s))
  for (j in 1:ncol(test_sets[[s]])){
    f <- colnames(test_sets[[s]])[j]
    
    if (f %in% c("p1","p2","p3","p4","p5","p6")) {
      if (f == "p1") {
        tmp <- test_sets[[s]][,j:(j+5)]
        # SHAP
        pred_raw <- mean(predict(models[[s]], test_sets[[s]], type = 'prob')[,2])
        for (i in 1:dim(tmp)[1]) {
          cond <- tmp[i,]
          test_sets[[s]][,j:(j+5)] <- cond
          pred <- mean(predict(models[[s]], test_sets[[s]], type = 'prob')[,2])
          importance[(s-1)*dim(tmp)[1]+i,f] <- (pred - pred_raw)
        }
        test_sets[[s]][,j:(j+5)] <- tmp
      }
      next
    }
    
    if (f=="conval"){next}
    tmp <- test_sets[[s]][,j]
    # SHAP
    pred_raw <- mean(predict(models[[s]], test_sets[[s]], type = 'prob')[,2])
    for (i in 1:length(tmp)) {
      cond <- tmp[i]
      test_sets[[s]][,j] <- cond
      pred <- mean(predict(models[[s]], test_sets[[s]], type = 'prob')[,2])
      importance[(s-1)*length(tmp)+i,f] <- (pred - pred_raw)
    }
    test_sets[[s]][,j] <- tmp
  }
}
importance <- as.data.frame(importance)
colnames(importance) <- gsub('^V([0-9]+)$', 'Factor \\1', colnames(importance))
colnames(importance) <- gsub('^p([0-9]+)$', 'Subtype \\1', colnames(importance))
importance_rank <- rbind(colMeans(abs(importance)), apply(abs(importance), 2, sd))
colnames(importance_rank) <- colnames(importance)
rownames(importance_rank) <- c("mean", "std")
order <- order(-importance_rank[1,])
importance_rank <- round(t(importance_rank[,order]), 3)
print(importance_rank[1:10, 1])

saveRDS(importance, "~/importance.RDS")
saveRDS(importance_rank, "~/importance_rank.RDS")
```

# Violin plots
```{r  fig.height=6, fig.width=24}
# importance <- as.data.frame(readRDS('~/importance.RDS'))
# importance_rank <- readRDS('~/importance_rank.RDS')
n_features <- 10
test_sets <- test_sets_clinical_factor_prob

participant_id <- NULL
for (i in 1:10){
  participant_id <- c(participant_id, rownames(test_sets[[i]]))
}

importance$participant_id <- participant_id
importance <- aggregate(. ~ participant_id, data = importance, FUN = mean)
conval_tmp <- conval_raw[importance$participant_id]
importance$participant_id <- NULL


importance_minimal <- importance[conval_tmp == 1, rownames(importance_rank)[1:n_features]]
importance_physical <- importance[conval_tmp == 2, rownames(importance_rank)[1:n_features]]
importance_cognitive <- importance[conval_tmp == 3, rownames(importance_rank)[1:n_features]]
importance_multiple <- importance[conval_tmp == 4, rownames(importance_rank)[1:n_features]]
importance_others <- rbind(importance_physical,importance_cognitive,importance_multiple)


# construct p-value matrix
l <- dim(importance_minimal)[1]
p1 <- apply(rbind(importance_minimal, importance_physical), FUN = function(s){return(t.test(s[1:l], s[l+1:length(s)], alternative = "greater")$p.value)}, MARGIN = 2)
p2 <- apply(rbind(importance_minimal, importance_cognitive), FUN = function(s){return(t.test(s[1:l], s[l+1:length(s)], alternative = "greater")$p.value)}, MARGIN = 2)
p3 <- apply(rbind(importance_minimal, importance_multiple), FUN = function(s){return(t.test(s[1:l], s[l+1:length(s)], alternative = "greater")$p.value)}, MARGIN = 2)
p4 <- apply(rbind(importance_minimal, importance_others), FUN = function(s){return(t.test(s[1:l], s[l+1:length(s)], alternative = "greater")$p.value)}, MARGIN = 2)
p <- as.data.frame(t(rbind(p1,p2,p3,p4)))
print(p)


importance_minimal <- -importance[conval_tmp == 1,]
importance_physical <- -importance[conval_tmp == 2,]
importance_cognitive <- -importance[conval_tmp == 3,]
importance_multiple <- -importance[conval_tmp == 4,]

# violin plots
df_violin <- as.data.frame(t(importance_minimal[,rownames(importance_rank)[1:n_features]]))
df_violin <- pivot_longer(df_violin, cols = everything())
df_violin$feature = rep(rownames(importance_rank)[1:n_features], each = nrow(importance_minimal))
df_violin$group = "Minimal"

df <- as.data.frame(t(importance_physical[,rownames(importance_rank)[1:n_features]])) 
df <- pivot_longer(df, cols = everything())
df$feature = rep(rownames(importance_rank)[1:n_features], each = nrow(importance_physical))
df$group = "Physical"
df_violin <- rbind(df_violin, df)

df <- as.data.frame(t(importance_cognitive[,rownames(importance_rank)[1:n_features]])) 
df <- pivot_longer(df, cols = everything())
df$feature = rep(rownames(importance_rank)[1:n_features], each = nrow(importance_cognitive))
df$group = "Cognitive"
df_violin <- rbind(df_violin, df)

df <- as.data.frame(t(importance_multiple[,rownames(importance_rank)[1:n_features]]))
df <- pivot_longer(df, cols = everything())
df$feature = rep(rownames(importance_rank)[1:n_features], each = nrow(importance_multiple))
df$group = "Multiple"
df_violin <- rbind(df_violin, df)

df_violin$group <- factor(df_violin$group, levels = c("Minimal", "Physical", 
                                                      "Cognitive", "Multiple"))

maps <- c("Chronic cardiac disease", "Subtype", "Chronic neurological disorder", "Sex (Female)", "Factor 30", "Diabetes", "Number of comorbidities", "Factor 38", "Factor 21", "BMI")
names(maps) <- rownames(importance_rank)[1:n_features]
df_violin$feature <- maps[df_violin$feature]
df_violin$feature <- factor(df_violin$feature, levels = maps[1:n_features])


texts <- c("                Minimal vs Others: 0.002 **\n                0.02 *         0.22        0.004 **", 
           "               Minimal vs Others: <0.001 ***\n               0.004 **       0.11        0.014 *",
           "               Minimal vs Others: <0.001 ***\n               0.006 **      0.37      <0.001 ***", 
           "                Minimal vs Others: 0.005 **\n                0.03 *        0.03 *        0.11",
           "                Minimal vs Others: 0.01 *\n                0.02 *        0.04 *        0.31",
           "                Minimal vs Others: 0.007 **\n                0.02 *         0.20         0.03 *",
           "             Minimal vs Others: <0.001 ***\n             <0.001 ***     0.24      <0.001 ***",
           "                 Minimal vs Others: 0.17\n                 0.41          0.27          0.13",
           "                Minimal vs Others: 0.20\n                0.052         0.76          0.35",
           "             Minimal vs Others: <0.001 ***\n             <0.001 ***     0.17          0.03 *")
facet_labels <- data.frame(
  feature = factor(c("Chronic cardiac disease", "Subtype", "Chronic neurological disorder", "Sex (Female)", "Factor 30", "Diabetes", "Number of comorbidities", "Factor 38", "Factor 21", "BMI"), levels = maps[1:n_features]),
  Text = texts,
  xPos = rep(0.8, 10),  # x position of the text, adjust as needed
  yPos = c(0.084, 0.135, 0.152, 0.041, 0.056,
           0.031, 0.073, 0.074, 0.059, 0.032)
)

texts1 <- c("p.val:")
facet_labels1 <- data.frame(
  feature = factor(c("Chronic cardiac disease", "Subtype", "Chronic neurological disorder", "Sex (Female)", "Factor 30", "Diabetes", "Number of comorbidities", "Factor 38", "Factor 21", "BMI"), levels = maps[1:n_features]),
  Text = texts1,
  xPos = rep(0.8, 10),  # x position of the text, adjust as needed
  yPos = c(0.084, 0.135, 0.152, 0.041, 0.056,
           0.031, 0.073, 0.074, 0.059, 0.032)
)

ggplot(df_violin, aes(x = group, y = value, color = group)) +
geom_jitter(size = 2.5, alpha = 0.4, aes(color = group)) +
geom_violin(size = 1, alpha = 0) +
geom_boxplot(width = 0.1, outlier.shape = NA, color = "darkslategrey", alpha = 0) +
scale_color_manual(values = c(Minimal="#999999",Physical="#58B5EA",Cognitive="#EAAE2A",Multiple="#FF1820")) +
facet_wrap(~ feature, ncol = 5, scales = "free_y") +
theme_minimal() +
theme(panel.grid.major = element_blank(),  # Remove major grid lines
      panel.grid.minor = element_blank(),  # Remove minor grid lines
      panel.border = element_blank(),  # Remove panel border
      axis.line.x = element_line(linewidth = 1),  # Add x axis line
      axis.line.y = element_line(linewidth = 1),  # Add y axis line
      axis.ticks = element_line(linewidth = 1),  # axis ticks
      axis.text.x = element_text(color = "black", size = 15),  
      axis.text.y = element_text(color = "black", size = 15),  
      axis.title = element_text(color = "black", size = 15),  # axis titles 
      axis.ticks.length = unit(0.15, "cm"),  # Adjust the length of axis ticks
      strip.text.x = element_text(size = 15, color = "black", margin = margin(b = 20)),
      legend.position = "none"  # Position the legend at the top
) +
labs(y = "Sample-level SHAP values", x = "") +
geom_text(data = facet_labels, aes(x = xPos, y = yPos, label = Text), inherit.aes = FALSE, size = 5, hjust = 0, vjust = 0.2, lineheight = 0.9) +
geom_text(data = facet_labels1, aes(x = xPos, y = yPos, label = Text), inherit.aes = FALSE, size = 5, hjust = 0, vjust = -0.3) +
coord_cartesian(clip = "off")
```  


## predict for subtypes using different assays
```{r}
assays = organized_data$datasets_imputed
new_names <- gsub('-[0-9]+$', '', rownames(assays[[1]]))
subtype_labels_copy = subtype_labels[subtype_labels$merged_labels!=4,]
subtype_labels_copy$merged_labels[subtype_labels_copy$merged_labels==1] = "A"
subtype_labels_copy$merged_labels[subtype_labels_copy$merged_labels==2] = "B"
subtype_labels_copy$merged_labels[subtype_labels_copy$merged_labels==3] = "C"
subtype_labels_copy$merged_labels[subtype_labels_copy$merged_labels==5] = "E"
subtype_labels_copy$merged_labels[subtype_labels_copy$merged_labels==6] = "F"
id <- Reduce(intersect, list(new_names, rownames(subtype_labels_copy)))
for (i in 1:6){
  assay <- assays[[i]]
  # keep V1
  ind <- !duplicated(new_names)
  assay <- assay[ind,]
  rownames(assay) <- new_names[ind]

  assays[[i]] <- assay[id,]
}
```



```{r warning=FALSE, echo=FALSE}
library(glmnet)
set.seed(2024) 
y = factor(subtype_labels_copy[id,]$merged_labels)
train_id = sample(1:length(y), floor(length(y)*0.7))
y_train = y[train_id]
y_test = y[-train_id]
selected_features = list()
for (i in 1:6) {
  cat("\n")
  x <- as.matrix(assays[[i]])
  x_train = x[train_id,]
  x_test = x[-train_id,]
  
  glm_model <- cv.glmnet(x_train, y_train, family = "multinomial", alpha = 1, nfolds = 5)
  pred = predict(glm_model, x_test, s = "lambda.1se", type = 'class')
  probabilities = as.data.frame(predict(glm_model, x_test, s = "lambda.1se", type = 'response'))
  colnames(probabilities) = c("A","B","C","E","F")
  cat("Assay: ", names(assays)[i])
  cat("\n")
  cat("Test accuracy: ", mean(pred == y_test))
  cat("\n")
  auc = c()
  for (g in c("A","B","C","E","F")){
    f = names(which(coef(glm_model, s = "lambda.1se")[[g]][-1,] != 0))
    y_test1 = as.character(y_test)
    y_test1[y_test1 != g] <- 0
    y_test1[y_test1 == g] <- 1
    y_test1 = as.factor(y_test1)
    suppressMessages({
    auc = c(auc, pROC::roc(y_test1, probabilities[,g])$auc)
    })
  }
  cat("Test AUC: ", mean(auc), "(", auc, ")")
  cat("\n")
  # refit LASSO using all data
  glm_model = cv.glmnet(x, y, family = "multinomial", alpha = 1, nfolds = 5)
  selected = c()
  for (g in c("A","B","C","E","F")){
    f = names(which(coef(glm_model, s = "lambda.1se")[[g]][-1,] != 0))
    selected = unique(c(selected, f))
  }
  selected_features[[i]] = selected
  cat("Number of selected features: ", length(selected))
  cat("\n")
  # plot(glm_model)
}
```


```{r warning=FALSE, echo=FALSE}
set.seed(2024) 
rocs = list()
aucs = list()
for (i in 1:6) {
  cat("\n")
  df <- assays[[i]][,selected_features[[i]]]
  # df$subtype = y
  df_train = df[train_id,]
  df_test = df[-train_id,]
  
  svm_model <- e1071::svm(df_train, y_train, probability = TRUE, kernel = "radial")
  pred = predict(svm_model, df_test, probability = TRUE)
  probabilities <- attr(pred, "probabilities")
  pred = predict(svm_model, df_test, probability = F)
  # svm_model <- caret::train(subtype ~ ., data = df_train, method = "svmRadial", 
  #                           trControl = trainControl(method = "cv"), probability=T)
  # pred <- predict(svm_model, df_test)
  cat("Assay: ", names(assays)[i])
  cat("\n")
  cat("Test accuracy: ", mean(pred == y_test))
  cat("\n")
  auc = c()
  roc = list()
  for (g in c("A","B","C","E","F")){
    y_test1 = as.character(y_test)
    y_test1[y_test1 != g] <- 0
    y_test1[y_test1 == g] <- 1
    y_test1 = as.factor(y_test1)
    suppressMessages({
    roc[[g]] = pROC::roc(y_test1, probabilities[,g])
    auc = c(auc, roc[[g]]$auc)
    })
  }
  rocs[[i]] = roc
  aucs[[i]] = auc
  cat("Test AUC: ", mean(auc), "(", auc, ")")
  cat("\n")
  cat("Number of selected features: ", length(selected_features[[i]]))
  cat("\n")
}
```


```{r warning=FALSE, echo=FALSE}
set.seed(2024) 
df = data.frame(subtype = y)
for (i in 1:6){
  df <- cbind(df, assays[[i]][,selected_features[[i]]])
}
df <- df[,!duplicated(colnames(df))]
df$subtype = NULL
df_train = df[train_id,]
df_test = df[-train_id,]
svm_model <- e1071::svm(df_train, y_train, probability = TRUE, kernel = "radial")
pred = predict(svm_model, df_test, probability = TRUE)
probabilities <- attr(pred, "probabilities")
pred = predict(svm_model, df_test, probability = F)
# svm_model <- caret::train(subtype ~ ., data = df_train, method = "svmRadial", 
#                             trControl = trainControl(method = "cv"))
# pred <- predict(svm_model, df_test)
cat("Combined assays")
cat("\n")
cat("Test accuracy: ", mean(pred == y_test))
cat("\n")
auc = c()
roc = list()
for (g in c("A","B","C","E","F")){
  y_test1 = as.character(y_test)
  y_test1[y_test1 != g] <- 0
  y_test1[y_test1 == g] <- 1
  y_test1 = as.factor(y_test1)
  suppressMessages({
    roc[[g]] = pROC::roc(y_test1, probabilities[,g])
    auc = c(auc, roc[[g]]$auc)
  })
}
rocs[[7]] = roc
aucs[[7]] = auc
cat("Test AUC: ", mean(auc), "(", auc, ")")
```


## Use cell frequencies
```{r warning=FALSE, echo=FALSE}
# parent cytof
cytof_parent_mat_train = base_obj$others$bld_cytof$train$bld_cytof_counts
cytof_parent_mat_train = base_obj$others$bld_cytof$preprocess_bld_cytof_parent_granulo_separate(counts = cytof_parent_mat_train, rowfeature = base_obj$others$bld_cytof$train$bld_cytof_rowfeatures)
cytof_parent_mat_test = base_obj$others$bld_cytof$test$bld_cytof_counts
cytof_parent_mat_test = base_obj$others$bld_cytof$preprocess_bld_cytof_parent_granulo_separate(counts = cytof_parent_mat_test, rowfeature = base_obj$others$bld_cytof$test$bld_cytof_rowfeatures)

cytof = rbind(cytof_parent_mat_train, cytof_parent_mat_test)
new_names <- gsub('-[0-9]+$', '', rownames(cytof))
id <- Reduce(intersect, list(new_names, rownames(subtype_labels_copy)))

# keep V1
ind <- !duplicated(new_names)
cytof <- cytof[ind,]
rownames(cytof) <- new_names[ind]
cytof <- cytof[id,]

set.seed(2024)
cytof$subtype = factor(subtype_labels_copy[id,]$merged_labels)
cytof = na.omit(cytof)
y1 = cytof$subtype
 
train_id_1 = sample(1:nrow(cytof), floor(nrow(cytof)*0.7))
y_train_cytof = cytof$subtype[train_id_1]
y_test_cytof = cytof$subtype[-train_id_1]
cytof$subtype = NULL
df_train = cytof[train_id_1,]
df_test = cytof[-train_id_1,]

svm_model <- e1071::svm(df_train, y_train_cytof, probability = TRUE, kernel = "radial")
pred = predict(svm_model, df_test, probability = TRUE)
probabilities <- attr(pred, "probabilities")
pred = predict(svm_model, df_test, probability = F)
cat("Cell frequencies")
cat("\n")
cat("Test accuracy: ", mean(pred == y_test_cytof))
cat("\n")
auc = c()
roc = list()
for (g in c("A","B","C","E","F")){
  y_test1 = as.character(y_test_cytof)
  y_test1[y_test1 != g] <- 0
  y_test1[y_test1 == g] <- 1
  y_test1 = as.factor(y_test1)
  suppressMessages({
  roc[[g]] = pROC::roc(y_test1, probabilities[,g])
  auc = c(auc, roc[[g]]$auc)
  })
}
rocs[[8]] = roc
aucs[[8]] = auc
cat("Test AUC: ", mean(auc), "(", auc, ")")
cat("\n")
cat("Number of selected features: ", dim(cytof)[2])
```




```{r}
library(cutpointr)
library(tidyverse)
# prepare data
fig_names <- c('Targeted Proteomics (86)', 'Global Proteomics (120)', 'Serum Olink (74)', 
               'Plasma Metabolomics (168)', 'Nasal Transcriptomics (149)', 
               'PBMC Transcriptomics (157)', 'Combined (733)', 
               'Cell Frequencies / CyTOF (14)')

roc_data <- NULL
for (i in 1:length(fig_names)){
  fig_name <- fig_names[i]
  rocs_svm <- rocs[[i]]
  aucs_svm <- aucs[[i]]
  
  # compute average roc curves
  # predictions <- rocs_svm[[1]]$predictor
  # for (j in 2:length(rocs_svm)){
  #   predictions <- rbind(predictions, rocs_svm[[j]]$predictor)
  # }
  # 
  # responses <- rocs_svm[[1]]$response
  # for (j in 2:length(rocs_svm)){
  #   responses <- rbind(responses, rocs_svm[[j]]$response)
  # }
  # 
  # predictions_10_samples <- data.frame(
  #     Sample = rep(c(1:5), times = length(rocs_svm[[1]]$predictor)),
  #     PredictionValues = as.numeric(predictions),
  #     RealClass = as.numeric(responses)
  # )
  # 
  # mean_roc <- function(data, cutoffs = seq(from = -1, to = 2, by = 0.01)) {
  #     map_df(cutoffs, function(cp) {
  #         out <- cutpointr(data = data, x = PredictionValues, class = RealClass,
  #                          subgroup = Sample, method = oc_manual, cutpoint = cp,
  #                          pos_class = 2, direction = ">=")
  #         data.frame(cutoff = cp,
  #                    sensitivity = mean(out$sensitivity),
  #                    specificity = mean(out$specificity))
  #     })
  # }
  # mr <- mean_roc(predictions_10_samples)
  
  # df <- data.frame(
  #   TPR = rev(mr$sensitivity),
  #   FPR = rev(1 - mr$specificity),
  #   Label = "Average",
  #   Model = fig_name
  # )
  # roc_data <- rbind(roc_data, df)
  
  df <- data.frame(
    TPR = seq(0, 1, 0.01),
    FPR = seq(0, 1, 0.01),
    Label = "Chance",
    Model = fig_name
  )
  roc_data <- rbind(roc_data, df)
  
  for (j in c("A","B","C","E","F")){
    df <- data.frame(
      TPR = rev(rocs_svm[[j]]$sensitivities),
      FPR = rev(1 - rocs_svm[[j]]$specificities),
      Label = j,
      Model = fig_name
    )
    roc_data <- rbind(roc_data, df)
  }
}
```

```{r fig.height=12, fig.width=24}
# ROC plots
roc_data$Model = factor(roc_data$Model, levels = fig_names)
custom_colors <- c("A" = "#2ca02c","B" = "#1f77b4", "C"="#ff7f0e", "E"="#9C3418", "F"="#9467bd", 
                   "Chance" = "red")

custom_line_types <- c("A" = "solid", "B" = "solid", "C" = "solid", "E" = "solid", "F" = "solid",
                       "Chance" = "dashed")

custom_line_width <- c("A" = 0.8, "B" = 0.8, "C" = 0.8, "E" = 0.8, "F" = 0.8,
                       "Chance" = 0.8)

texts <- c()
for (i in 1:length(aucs)){
  aucs_svm <- aucs[[i]]
  M <- mean(aucs_svm)
  texts[i] <- paste0("Average AUROC: ", sprintf("%.3f",round(M,3)))
}

facet_labels <- data.frame(
  Model = factor(fig_names, levels = fig_names),
  Text = texts,
  xPos = rep(0.7, length(fig_names)), 
  yPos = rep(0.4, length(fig_names))  
)

ggplot() +
geom_line(data = roc_data, aes(x = FPR, y = TPR, color = Label, linewidth = Label, linetype = Label)) + 
scale_color_manual(values = custom_colors, breaks = c("A","B","C","E","F","Chance")) +
scale_linewidth_manual(values = custom_line_width, breaks = c("A","B","C","E","F","Chance")) +  
scale_linetype_manual(values = custom_line_types, breaks = c("A","B","C","E","F","Chance")) + 
facet_wrap(~ Model, scales = "free", ncol = 4) +
labs(x = "False positive rate", y = "True positive rate") +
scale_x_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
scale_y_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
# Customize the theme 
theme_minimal() +
theme(panel.spacing = unit(2, "lines"),
      panel.grid.major = element_blank(),  # Remove major grid lines
      panel.grid.minor = element_blank(),  # Remove minor grid lines
      panel.border = element_blank(),  # Remove panel border
      axis.line.x = element_line(linewidth = 1),  # Add x axis line
      axis.line.y = element_line(linewidth = 1),  # Add y axis line
      axis.ticks = element_line(linewidth = 1),  # axis ticks
      axis.text.x = element_text(color = "black", size = 15),  # axis text
      axis.text.y = element_text(color = "black", size = 15),  # axis text
      axis.title.x = element_text(color = "black", size = 15, margin = margin(t=10)),  # axis titles 
      axis.title.y = element_text(color = "black", size = 15, margin = margin(r=10)),
      axis.ticks.length = unit(0.15, "cm"),  # Adjust the length of axis ticks
      strip.text = element_text(color = "black", size = 15, margin = margin(b = 20)),
      legend.position = "top",
      legend.direction = "horizontal",
      legend.justification = "left",  # Justify the legend to the left
      legend.title = element_blank(),
      legend.text = element_text(color = "black", size = 15),
      #legend.margin = margin(l = -250, b = 100),
      legend.spacing.x = unit(0.5, "cm"),
      legend.key.width = unit(2, "cm"),
      # plot.margin = margin(r = 150),
      aspect.ratio = 0.85
    ) + 
guides(color = guide_legend(order = 1, keyheight = unit(1, "cm")), 
       linewidth = guide_legend(order = 1, keyheight = unit(1, "cm")),
       linetype = guide_legend(order = 1, keyheight = unit(1, "cm"))) + 
geom_text(color = "black", data = facet_labels, aes(x = xPos, y = yPos, label = Text), inherit.aes = FALSE, size = 5)
```



```{r}
library(cutpointr)
library(tidyverse)
# prepare data
fig_names <- c('Targeted Proteomics (86)', 'Global Proteomics (120)', 'Serum Olink (74)', 
               'Plasma Metabolomics (168)', 'Nasal Transcriptomics (149)', 
               'PBMC Transcriptomics (157)', 'Combined (733)', 
               'Cell Frequencies / CyTOF (14)')

roc_data <- NULL
for (i in 1:length(fig_names)){
  fig_name <- fig_names[i]
  rocs_svm <- rocs[[i]]
  aucs_svm <- aucs[[i]]
  
  # # compute average roc curves
  # predictions <- rocs_svm[[1]]$predictor
  # for (j in 2:length(rocs_svm)){
  #   predictions <- rbind(predictions, rocs_svm[[j]]$predictor)
  # }
  # 
  # responses <- rocs_svm[[1]]$response
  # for (j in 2:length(rocs_svm)){
  #   responses <- rbind(responses, rocs_svm[[j]]$response)
  # }
  # 
  # predictions_10_samples <- data.frame(
  #     Sample = rep(c(1:5), times = length(rocs_svm[[1]]$predictor)),
  #     PredictionValues = as.numeric(predictions),
  #     RealClass = as.numeric(responses)
  # )
  # 
  # mean_roc <- function(data, cutoffs = seq(from = -1, to = 2, by = 0.01)) {
  #     map_df(cutoffs, function(cp) {
  #         out <- cutpointr(data = data, x = PredictionValues, class = RealClass,
  #                          subgroup = Sample, method = oc_manual, cutpoint = cp,
  #                          pos_class = 2, direction = ">=")
  #         data.frame(cutoff = cp,
  #                    sensitivity = mean(out$sensitivity),
  #                    specificity = mean(out$specificity))
  #     })
  # }
  # mr <- mean_roc(predictions_10_samples)
  # 
  # df <- data.frame(
  #   TPR = rev(mr$sensitivity),
  #   FPR = rev(1 - mr$specificity),
  #   Label = "Average",
  #   Model = fig_name
  # )
  # roc_data <- rbind(roc_data, df)
  
  df <- data.frame(
    TPR = seq(0, 1, 0.01),
    FPR = seq(0, 1, 0.01),
    Label = "Chance",
    Model = fig_name
  )
  roc_data <- rbind(roc_data, df)
  
  for (j in c("A","B","C","E","F")){
    df <- data.frame(
      TPR = rev(rocs_svm[[j]]$sensitivities),
      FPR = rev(1 - rocs_svm[[j]]$specificities),
      Label = j,
      Model = fig_name
    )
    roc_data <- rbind(roc_data, df)
  }
}
```

```{r fig.height=12, fig.width=24}
# ROC plots
roc_data$Model = factor(roc_data$Model, levels = fig_names)
custom_colors <- c("A" = "#2ca02c","B" = "#1f77b4", "C"="#ff7f0e", "E"="#9C3418", 
                   "F"="#9467bd", "Average" = "black", "Chance" = "red")

custom_line_types <- c("A" = "solid", "B" = "solid", "C" = "solid", "E" = "solid", 
                       "F" = "solid", "Average" = "solid", "Chance" = "dashed")

custom_line_width <- c("A" = 0.8, "B" = 0.8, "C" = 0.8, "E" = 0.8, "F" = 0.8,
                      "Average" = 1.2,  "Chance" = 0.8)


texts <- c()
for (i in 1:length(aucs)){
  aucs_svm <- aucs[[i]]
  M <- mean(aucs_svm)
  texts[i] <- paste0("Average AUROC: ", sprintf("%.3f",round(M,3)),
                     "\nSubC AUROC: ", sprintf("%.3f",round(aucs_svm[3],3)),
                     "\nSubF AUROC: ", sprintf("%.3f",round(aucs_svm[5],3))
                     )
}

facet_labels <- data.frame(
  Model = factor(fig_names, levels = fig_names),
  Text = texts,
  xPos = rep(0.7, length(fig_names)), 
  yPos = rep(0.4, length(fig_names))  
)

ggplot() +
geom_line(data = roc_data, aes(x = FPR, y = TPR, color = Label, linewidth = Label, linetype = Label)) + 
scale_color_manual(values = custom_colors, breaks = c("A","B","C","E","F","Average","Chance")) +
scale_linewidth_manual(values = custom_line_width, breaks = c("A","B","C","E","F","Average","Chance")) +  
scale_linetype_manual(values = custom_line_types, breaks = c("A","B","C","E","F","Average","Chance")) + 
facet_wrap(~ Model, scales = "free", ncol = 4) +
labs(x = "False positive rate", y = "True positive rate") +
scale_x_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
scale_y_continuous(expand = c(0, 0), limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
# Customize the theme 
theme_minimal() +
theme(panel.spacing = unit(2, "lines"),
      panel.grid.major = element_blank(),  # Remove major grid lines
      panel.grid.minor = element_blank(),  # Remove minor grid lines
      panel.border = element_blank(),  # Remove panel border
      axis.line.x = element_line(linewidth = 1),  # Add x axis line
      axis.line.y = element_line(linewidth = 1),  # Add y axis line
      axis.ticks = element_line(linewidth = 1),  # axis ticks
      axis.text.x = element_text(color = "black", size = 15),  # axis text
      axis.text.y = element_text(color = "black", size = 15),  # axis text
      axis.title.x = element_text(color = "black", size = 15, margin = margin(t=10)),  # axis titles 
      axis.title.y = element_text(color = "black", size = 15, margin = margin(r=10)),
      axis.ticks.length = unit(0.15, "cm"),  # Adjust the length of axis ticks
      strip.text = element_text(color = "black", size = 15, margin = margin(b = 20)),
      legend.position = "top",  
      legend.justification = "left",  # Justify the legend to the left
      legend.title = element_blank(),
      legend.text = element_text(color = "black", size = 15),
      # legend.margin = margin(l = -250, b = 100),
      legend.spacing.x = unit(0.5, "cm"),
      legend.key.width = unit(2, "cm"),
      plot.margin = margin(r = 20),
      aspect.ratio = 0.85
    ) + 
guides(color = guide_legend(order = 1, keyheight = unit(1, "cm"), nrow = 1), 
       linewidth = guide_legend(order = 1, keyheight = unit(1, "cm")),
       linetype = guide_legend(order = 1, keyheight = unit(1, "cm"))) + 
geom_text(color = "black", data = facet_labels, aes(x = xPos, y = yPos, label = Text), inherit.aes = FALSE, size = 5)
```

```{r}
df = NULL

for (i in 1:6){
  tmp = data.frame(feature = selected_features[[i]], 
                 assay = rep(names(assays)[i],length(selected_features[[i]]))
                 )
  df = rbind(df, tmp)
}
# write.csv(df, "/scratch/yutong/selected_features.csv", row.names = FALSE)
```


## Association between predicted subtype and PASC
```{r}
set.seed(2024) 
prob_vec = list()
for (i in 1:6) {
  df <- assays[[i]][,selected_features[[i]]]
  svm_model <- e1071::svm(df, y, probability = TRUE, kernel = "radial")
  pred = predict(svm_model, df, probability = TRUE)
  probabilities <- attr(pred, "probabilities")
  prob_vec[[i]] = probabilities
}

df <- cytof
svm_model <- e1071::svm(df, y1, probability = TRUE, kernel = "radial")
pred = predict(svm_model, df, probability = TRUE)
probabilities <- attr(pred, "probabilities")
prob_vec[[8]] = probabilities
```


```{r fig.height=5.5, fig.width=3}
library(ComplexHeatmap)
fig_names <- c('Targeted Proteomics', 'Global Proteomics', 'Cytokines / Serum Olink', 
               'Plasma Metabolomics', 'Nasal Transcriptomics', 'PBMC Transcriptomics', 
               'Combined', 'Cell Frequencies / CyTOF')

for (i in c(3, 6, 8)){
  df = prob_vec[[i]]
  df = as.data.frame(df[,c("A","B","C","E","F")])
  df$PASC = subtype_labels_copy[rownames(df),]$PASC
  df1 = na.omit(df)
  df1[,c("A","B","C","E")] = df1[,c("A","B","C","E")] - df1[,"F"]
  df1$F = NULL
  model1 = nnet::multinom(PASC ~ ., data = df1)
  z_trees1 <- (summary(model1)$coefficients/summary(model1)$standard.errors)[,-1]
  p_trees1 <- (1 - pnorm(abs(z_trees1), 0, 1)) * 2
  
  df2 = na.omit(df)
  df2[,c("B","C","E","F")] = df2[,c("B","C","E","F")] - df2[,"A"]
  df2$A = NULL
  model2 = nnet::multinom(PASC ~ ., data = df2)
  z_trees2 <- (summary(model2)$coefficients/summary(model2)$standard.errors)[,-1]
  p_trees2 <- (1 - pnorm(abs(z_trees2), 0, 1)) * 2
  
  z_trees = cbind(z_trees1[,1], z_trees2)
  p_trees = cbind(p_trees1[,1], p_trees2)
  
  colnames(z_trees) = paste0("Sub", c("A","B","C","E","F"))
  colnames(p_trees) = paste0("Sub", c("A","B","C","E","F"))
  rownames(z_trees) = c("Physical","Cognitive","Multiple")
  rownames(p_trees) = c("Physical","Cognitive","Multiple")
  z_trees = t(z_trees)
  p_trees = t(p_trees)
  print(as.data.frame(p_trees))
  
  p = ComplexHeatmap::Heatmap(as.matrix(z_trees), row_names_side = "left",
               row_names_gp = gpar(fontsize = 15), column_names_gp = gpar(fontsize = 15), 
               column_names_rot = 90, #column_names_centered = T,
               column_title_gp = gpar(fontsize = 15),
               column_title=gt_render(fig_names[i]),
               heatmap_legend_param = list(title = "Coefficient",
               direction = "horizontal", title_position = "topcenter", labels_rot = 90,
               title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
               col = circlize::colorRamp2(c(-max(abs(z_trees)), 0, max(abs(z_trees))), c("#284AA0", "white", "#BC4A4A")), na_col = "grey90",
               cluster_columns = F, cluster_rows = F, cell_fun = function(j, i, x, y, w, h, fill) {
    if(p_trees[i, j] < 0.001) {
      grid.text("***", x, y, gp = gpar(cex=2))
    }else if(p_trees[i, j] < 0.01) {
      grid.text("**", x, y, gp = gpar(cex=2))
    }else if(p_trees[i, j] < 0.05) {
      grid.text("*", x, y, gp = gpar(cex=2))
    }
  })
  
  ComplexHeatmap::draw(p, heatmap_legend_side = "bottom")
}
```


```{r fig.height=5.5, fig.width=2.2}
for (i in c(3, 6, 8)){
  df = prob_vec[[i]]
  df = as.data.frame(df[,c("A","B","C","E","F")])
  df$PASC = as.character(subtype_labels_copy[rownames(df),]$PASC)
  df$PASC[is.na(df$PASC)] = "Missing"
  df[df$PASC %in% c("physical","cognitive","multiple"),]$PASC = "Any deficit"
  df[df$PASC == "minimal",]$PASC = "Minimal"
  
  # Any deficit vs minimal
  df1 = df[df$PASC %in% c("Minimal","Any deficit"),]
  df1[df1$PASC=="Minimal",]$PASC = 0
  df1[df1$PASC=="Any deficit",]$PASC = 1
  df1$PASC = as.numeric(df1$PASC)
  df1[,c("A","B","C","E")] = df1[,c("A","B","C","E")] - df1[,"F"]
  df1$F = NULL
  model = glm(PASC ~ ., data = df1)
  z_trees1 <- coef(summary(model))[-1,1]
  p_trees1 <- coef(summary(model))[-1,4]
  
  df1 = df[df$PASC %in% c("Minimal","Any deficit"),]
  df1[df1$PASC=="Minimal",]$PASC = 0
  df1[df1$PASC=="Any deficit",]$PASC = 1
  df1$PASC = as.numeric(df1$PASC)
  df1[,c("B","C","E","F")] = df1[,c("B","C","E","F")] - df1[,"A"]
  df1$A = NULL
  model = glm(PASC ~ ., data = df1)
  z_trees2 <- coef(summary(model))[-1,1]
  p_trees2 <- coef(summary(model))[-1,4]
  
  z_trees = as.data.frame(c(z_trees1[1], z_trees2))
  p_trees = as.data.frame(c(p_trees1[1], p_trees2))
  
  # Missing vs minimal
  df1 = df[df$PASC %in% c("Minimal","Missing"),]
  df1[df1$PASC=="Minimal",]$PASC = 0
  df1[df1$PASC=="Missing",]$PASC = 1
  df1$PASC = as.numeric(df1$PASC)
  df1[,c("A","B","C","E")] = df1[,c("A","B","C","E")] - df1[,"F"]
  df1$F = NULL
  model = glm(PASC ~ ., data = df1)
  z_trees1 <- coef(summary(model))[-1,1]
  p_trees1 <- coef(summary(model))[-1,4]
  
  df1 = df[df$PASC %in% c("Minimal","Missing"),]
  df1[df1$PASC=="Minimal",]$PASC = 0
  df1[df1$PASC=="Missing",]$PASC = 1
  df1$PASC = as.numeric(df1$PASC)
  df1[,c("B","C","E","F")] = df1[,c("B","C","E","F")] - df1[,"A"]
  df1$A = NULL
  model = glm(PASC ~ ., data = df1)
  z_trees2 <- coef(summary(model))[-1,1]
  p_trees2 <- coef(summary(model))[-1,4]
  
  z_trees = cbind(z_trees, c(z_trees1[1], z_trees2))
  p_trees = cbind(p_trees, c(p_trees1[1], p_trees2))
  rownames(z_trees) = paste0("Sub", c("A","B","C","E","F"))
  rownames(p_trees) = paste0("Sub", c("A","B","C","E","F"))
  colnames(z_trees) = c("Any deficit", "Missing")
  colnames(p_trees) = c("Any deficit", "Missing")
  print(p_trees)
  
  p = ComplexHeatmap::Heatmap(as.matrix(z_trees), row_names_side = "left",
               row_names_gp = gpar(fontsize = 15), column_names_gp = gpar(fontsize = 15), show_row_names = F,
               column_names_rot = 90, #column_names_centered = T,
               heatmap_legend_param = list(title = "Coefficient",
               direction = "horizontal", title_position = "topcenter",
               title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
               col = circlize::colorRamp2(c(-max(abs(z_trees)), 0, max(abs(z_trees))), c("#284AA0", "white", "#BC4A4A")), na_col = "grey90", column_title=gt_render(""),
               column_split = c("Any deficit","Missing"), column_gap = unit(1, "cm"),
               cluster_columns = F, cluster_rows = F, cell_fun = function(j, i, x, y, w, h, fill) {
    if(p_trees[i, j] < 0.001) {
      grid.text("***", x, y, gp = gpar(cex=2))
    }else if(p_trees[i, j] < 0.01) {
      grid.text("**", x, y, gp = gpar(cex=2))
    }else if(p_trees[i, j] < 0.05) {
      grid.text("*", x, y, gp = gpar(cex=2))
    }
  })
  
  ComplexHeatmap::draw(p, heatmap_legend_side = "bottom")
}
```

# Logistic regression for any deficit/non-respondent
```{r, fig.height=5.5, fig.width=2.2}
fig_names <- c('Targeted Proteomics', 'Global Proteomics', 'Cytokines / Serum Olink', 
               'Plasma Metabolomics', 'Nasal Transcriptomics', 'PBMC Transcriptomics', 
               'Combined', 'Cell Frequencies / CyTOF')

extended_data = readRDS("~/impacc/Analysis/molecular_subtyping/code/Fig1/src/extended_data.rds")
sample_meta_participant = extended_data$sample_meta_participant_level
sample_meta_participant$death_date[sample_meta_participant$death_date<0] = 999
censoring_date_death = 200
death_surv = cbind(sample_meta_participant$death_date,
                   sample_meta_participant$death_date)
death_surv[death_surv[,2]>censoring_date_death,2] = censoring_date_death+1
death_surv[,1] = ifelse(death_surv[,2]<=censoring_date_death,1,0)
death_surv = data.frame(death_surv)
colnames(death_surv) = c("state", "time")

death_time = death_surv$time[subtype_labels$label != "D"]
id_post_acute <- which(death_time > 28)

for (i in c(3, 6, 8)) {
  df = prob_vec[[i]]
  df = as.data.frame(df[,c("A","B","C","E","F")])
  df$PASC = as.character(subtype_labels_copy[rownames(df),]$PASC)
  df$PASC[is.na(df$PASC)] = "Non-respondent"
  df[df$PASC %in% c("physical","cognitive","multiple"),]$PASC = "Any deficit"
  df[df$PASC == "minimal",]$PASC = "Minimal"
  df = df[id_post_acute, ]
  
  # Any deficit vs minimal
  df1 = df[df$PASC %in% c("Minimal","Any deficit"),]
  df1[df1$PASC=="Minimal",]$PASC = 0
  df1[df1$PASC=="Any deficit",]$PASC = 1
  df1$PASC = as.numeric(df1$PASC)
  df1[,c("A","B","C","E")] = df1[,c("A","B","C","E")] - df1[,"F"]
  df1$F = NULL
  model = glm(PASC ~ ., family = binomial, data = df1)
  z_trees1 <- coef(summary(model))[-1,1]
  p_trees1 <- coef(summary(model))[-1,4]
  
  df1 = df[df$PASC %in% c("Minimal","Any deficit"),]
  df1[df1$PASC=="Minimal",]$PASC = 0
  df1[df1$PASC=="Any deficit",]$PASC = 1
  df1$PASC = as.numeric(df1$PASC)
  df1[,c("B","C","E","F")] = df1[,c("B","C","E","F")] - df1[,"A"]
  df1$A = NULL
  model = glm(PASC ~ ., family = binomial, data = df1)
  z_trees2 <- coef(summary(model))[-1,1]
  p_trees2 <- coef(summary(model))[-1,4]
  
  z_trees = as.data.frame(c(z_trees1[1], z_trees2))
  p_trees = as.data.frame(c(p_trees1[1], p_trees2))
  
  # Missing vs minimal
  df1 = df[df$PASC %in% c("Minimal", "Non-respondent"),]
  df1[df1$PASC=="Minimal",]$PASC = 0
  df1[df1$PASC=="Non-respondent",]$PASC = 1
  df1$PASC = as.numeric(df1$PASC)
  df1[,c("A","B","C","E")] = df1[,c("A","B","C","E")] - df1[,"F"]
  df1$F = NULL
  model = glm(PASC ~ ., family = binomial, data = df1)
  z_trees1 <- coef(summary(model))[-1,1]
  p_trees1 <- coef(summary(model))[-1,4]
  
  df1 = df[df$PASC %in% c("Minimal","Non-respondent"),]
  df1[df1$PASC=="Minimal",]$PASC = 0
  df1[df1$PASC=="Non-respondent",]$PASC = 1
  df1$PASC = as.numeric(df1$PASC)
  df1[,c("B","C","E","F")] = df1[,c("B","C","E","F")] - df1[,"A"]
  df1$A = NULL
  model = glm(PASC ~ ., family = binomial, data = df1)
  z_trees2 <- coef(summary(model))[-1,1]
  p_trees2 <- coef(summary(model))[-1,4]
  
  z_trees = cbind(z_trees, c(z_trees1[1], z_trees2))
  p_trees = cbind(p_trees, c(p_trees1[1], p_trees2))
  rownames(z_trees) = paste0("Sub", c("A","B","C","E","F"))
  rownames(p_trees) = paste0("Sub", c("A","B","C","E","F"))
  colnames(z_trees) = c("Any deficit", "Non-\nrespondent")
  colnames(p_trees) = c("Any deficit", "Non-\nrespondent")
  
  p = ComplexHeatmap::Heatmap(as.matrix(z_trees), row_names_side = "left",
               row_names_gp = gpar(fontsize = 15), column_names_gp = gpar(fontsize = 15), 
               show_row_names = F, column_names_centered = TRUE,
               column_names_rot = 90, #column_names_centered = T,
               heatmap_legend_param = list(title = "Coefficient",
               direction = "horizontal", title_position = "topcenter",
               title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
               col = circlize::colorRamp2(c(-2, 0, 2), c("#284AA0", "white", "#BC4A4A")), 
               na_col = "grey90", column_title=gt_render(""),
               column_split = c("Any deficit", "Non-\nrespondent"), 
               column_gap = unit(1, "cm"),
               cluster_columns = F, cluster_rows = F, cell_fun = function(j, i, x, y, w, h, fill) {
    if(p_trees[i, j] < 0.001) {
      grid.text("***", x, y-h*0.12, gp = gpar(cex=2))
    }else if(p_trees[i, j] < 0.01) {
      grid.text("**", x, y-h*0.12, gp = gpar(cex=2))
    }else if(p_trees[i, j] < 0.05) {
      grid.text("*", x, y-h*0.12, gp = gpar(cex=2))
    }
  })
  
  ComplexHeatmap::draw(p, heatmap_legend_side = "bottom")
}
```

# Logistic regression for PASC
```{r, fig.height=5.5, fig.width=2.5}
test_multi_pasc <- function(df_pasc) {
  response = df_pasc$PASC
  response = data.frame(model.matrix(~response-1))
  colnames(response) = gsub("response", "", colnames(response))
  pasc_names = setdiff(colnames(response), "minimal")
  subtype_names = setdiff(colnames(df_pasc), "PASC")
  nclass = length(subtype_names)
  
  pval_pasc = matrix(NA, nrow = nclass, ncol = length(pasc_names))
  colnames(pval_pasc) = pasc_names
  rownames(pval_pasc) = subtype_names
  coef_pasc = matrix(NA, nrow = nclass, ncol = length(pasc_names))
  colnames(coef_pasc) = pasc_names
  rownames(coef_pasc) = subtype_names
  
  for (fname in pasc_names) {
    idx_test = which(response[[fname]] == 1 |
                       response[["minimal"]] == 1)
    response_binary = response[idx_test, fname]
    
    x_pasc = df_pasc[, subtype_names]
    df_binom = cbind(data.frame(PASC = response_binary), x_pasc[idx_test, ])
    model1 <- glm(PASC ~ ., family = binomial, data = df_binom)
    coef_pasc[, fname] <- coef(summary(model1))[2:(nclass+1), "Estimate"]
    pval_pasc[, fname] <- coef(summary(model1))[2:(nclass+1), "Pr(>|z|)"]
  }
  return(list(pval_pasc = pval_pasc, coef_pasc = coef_pasc))
}

fig_names <- c('Targeted Proteomics', 'Global Proteomics', 'Cytokines / Serum Olink', 
               'Plasma Metabolomics', 'Nasal Transcriptomics', 'PBMC Transcriptomics', 
               'Combined', 'Cell Frequencies / CyTOF')

for (i in c(3, 6, 8)){
  df = prob_vec[[i]]
  df = as.data.frame(df[,c("A","B","C","E","F")])
  df$PASC = subtype_labels_copy[rownames(df),]$PASC
  df1 = na.omit(df)
  df1[,c("A","B","C","E")] = df1[,c("A","B","C","E")] - df1[,"F"]
  df1$F = NULL
  model1 = test_multi_pasc(df1)
  z_trees1 <- t(model1$coef_pasc)
  p_trees1 <- t(model1$pval_pasc)
  
  df2 = na.omit(df)
  df2[,c("B","C","E","F")] = df2[,c("B","C","E","F")] - df2[,"A"]
  df2$A = NULL
  model2 = test_multi_pasc(df2)
  z_trees2 <- t(model2$coef_pasc)
  p_trees2 <- t(model2$pval_pasc)
  
  z_trees = cbind(z_trees1[,1], z_trees2)
  p_trees = cbind(p_trees1[,1], p_trees2)
  
  colnames(z_trees) = c("A","B","C","E","F")
  colnames(p_trees) = c("A","B","C","E","F")
  rownames(z_trees) = c("Physical","Cognitive","Multiple")
  rownames(p_trees) = c("Physical","Cognitive","Multiple")
  z_trees = t(z_trees)
  p_trees = t(p_trees)
  print(as.data.frame(p_trees))
  
  p = ComplexHeatmap::Heatmap(as.matrix(z_trees), row_names_side = "left",
               row_names_gp = gpar(fontsize = 15), column_names_gp = gpar(fontsize = 15), 
               column_names_rot = 90, column_names_centered = T,
               column_title_gp = gpar(fontsize = 15),
               column_title=gt_render(fig_names[i]),
               heatmap_legend_param = list(title = "Coefficient",
               direction = "horizontal", title_position = "topcenter",
               title_gp = gpar(fontsize = 15), labels_gp = gpar(fontsize = 15)),
               col = circlize::colorRamp2(c(-2, 0, 2), c("#284AA0", "white", "#BC4A4A")),
               na_col = "grey90",
               cluster_columns = F, cluster_rows = F, cell_fun = function(j, i, x, y, w, h, fill) {
    if(p_trees[i, j] < 0.001) {
      grid.text("***", x, y-h*0.12, gp = gpar(cex=2))
    }else if(p_trees[i, j] < 0.01) {
      grid.text("**", x, y-h*0.12, gp = gpar(cex=2))
    }else if(p_trees[i, j] < 0.05) {
      grid.text("*", x, y-h*0.12, gp = gpar(cex=2))
    }
  })
  
  ComplexHeatmap::draw(p, heatmap_legend_side = "bottom")
}
```





