---
title: "pbmc-nasal_rnaseq_correlation"
output: html_document
---

Source: 20231004_data_integration_pbmc-ns.Rmd

```{r}
library(dplyr)
library(ggplot2)
library(patchwork)

library(limma)
library(edgeR)

library(fgsea)

my.theme <- theme_classic() + 
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm")
  )

```


# Data import

## Clinical metadata

```{r}
# Clinical characteristics
clin.individ <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-frozen.csv",
  header = TRUE, row.names = NULL
)
# Filter COVID positive samples
clin.individ <- subset(clin.individ, participant_type == "COVID-19 Positive")
print(head(clin.individ))

# Clinical characteristics supp
clin.individ.supp <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-individ-supp-frozen.csv",
  header = TRUE, row.names = NULL
)
print(clin.individ.supp[1:6,1:6])
```

Import event data.

```{r}
# For mapping sample_id to event_id
clin.sample <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-sample-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.sample))

# Explanation of event_id
clin.event <- read.csv(
  "/data/clinical/legacy/2021-11-11-frozen/2021-11-11-impacc-clin-event-frozen.csv",
  header = TRUE, row.names = NULL
)
print(head(clin.event))
```

I will annotate if an event_id has received steroids, by comparing clin.event's event_date with clin.individ.supp's steroids_start.

```{r}
clin.event <- merge(
  clin.event, clin.individ.supp[,c("participant_id","steroids_start")],
  by = "participant_id",
  all.x = TRUE, all.y = FALSE,
  sort = FALSE
)

clin.event$steroids <- clin.event$steroids_start <= clin.event$event_date
clin.event[is.na(clin.event$steroids),"steroids"] <- FALSE
clin.event$steroids <- as.factor(clin.event$steroids)

head(clin.event)
```

```{r}
# Convert ordinal score to factor
clin.event$respiratory_status <- as.factor(clin.event$respiratory_status)

# Convert to factor
clin.individ$sex <- as.factor(clin.individ$sex)
clin.individ$trajectory_group <- as.factor(clin.individ$trajectory_group)
clin.individ$enrollment_site <- as.factor(clin.individ$enrollment_site)
clin.individ$participant_id <- as.factor(clin.individ$participant_id)
```

### Merge clinical metadata

```{r}
# Combine sample and event metadata
clin.sample.event <- merge(
  clin.sample, clin.event,
  by = "event_id",
  all.x = TRUE, all.y = FALSE,
  suffixes = c("",".y"),
  sort = FALSE
)

# Add patient metadata
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ, by = "participant_id", 
  all = FALSE, sort = FALSE
)
```

Add patients' individual metadata.

```{r}
clin.sample.event.individ <- merge(
  clin.sample.event, clin.individ,
  by = "participant_id", all = FALSE, sort = FALSE
)
```

## RNA-seq data

PBMC RNA-seq

```{r}
pbmc.counts <- read.csv(
  "/data/pbmc-transcriptomics/legacy/2022-06-15/pbmc-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1,
  check.names = FALSE
)
pbmc.counts <- t(pbmc.counts)
print(pbmc.counts[1:6,1:6])

pbmc.metadata <- read.csv(
  "/data/pbmc-transcriptomics/legacy/2022-06-15/pbmc-transcriptomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
print(head(pbmc.metadata))

pbmc.rData <- read.csv(
  "/data/pbmc-transcriptomics/current/pbmc-transcriptomics-RowFeature.csv",
  header = FALSE, row.names = NULL,
  check.names = FALSE
)
stopifnot(rownames(pbmc.counts)==pbmc.rData$gene_id) # verify that gene ids match
colnames(pbmc.rData) <- pbmc.rData[2,]
pbmc.rData <- pbmc.rData[-(1:2),]
print(head(pbmc.rData))
```

NS RNA-seq

```{r}
nasal.counts <- read.csv(
  "/data/nasal-transcriptomics/legacy/2022-06-30/nasal-transcriptomics-Counts.csv",
  header = TRUE, row.names = 1,
  check.names = FALSE
)
nasal.counts <- t(nasal.counts)
print(nasal.counts[1:6,1:6])

nasal.metadata <- read.csv(
  "/data/nasal-transcriptomics/legacy/2022-06-30/nasal-transcriptomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
print(head(nasal.metadata))

nasal.rData <- read.csv(
  "/data/nasal-transcriptomics/current/nasal-transcriptomics-RowFeature.csv",
  header = FALSE, row.names = NULL,
  check.names = FALSE
)
stopifnot(rownames(nasal.counts)==nasal.rData$gene_id) # verify that gene ids match
colnames(nasal.rData) <- nasal.rData[2,]
nasal.rData <- nasal.rData[-(1:2),]
print(head(nasal.rData))
```

## Olink data

```{r}
# Concentration
olink <- read.csv(
  "/data/serum-olink/legacy/2023-01-06/Olink-Counts.csv",
  header = TRUE, row.names = 1
)
olink <- t(olink) # transpose

# Row data
olink.rData <- read.csv(
  "/data/serum-olink/legacy/2023-01-06/Olink-RowFeature.csv",
  header = TRUE, row.names = 3
)
stopifnot(rownames(olink) == rownames(olink.rData))

# Olink metadata
olink.metadata <- read.csv(
  "/data/serum-olink/legacy/2023-01-06/Olink-Metadata.csv",
  header = TRUE, row.names = NULL
)

# Verify that the sample_id of counts table and metadata match
stopifnot(colnames(olink) %in% olink.metadata$sample_id)
```

```{r}
# Set concentrations below LOD to 0
lod <- replicate(ncol(olink), olink.rData$LOD)
olink[olink < lod] <- 0

# Only keep cytokines detected in at least 20% of the samples
olink <- olink[rowSums(olink>0, na.rm = TRUE) >= 0.2*ncol(olink),]
```

## Viral load data

SARS-CoV-2 viral load is readily available in nasal-metagenomics-Metadata.csv, sc2_rpm column.

```{r}
# mNGS metadata
mngs.metadata <- read.csv(
  "/data/nasal-metagenomics/legacy/2022-09-03/nasal-metagenomics-Metadata.csv",
  header = TRUE, row.names = NULL,
  check.names = FALSE
)
```

## Only keep samples that pass QC

```{r}
# RNA-seq QC
pbmc.metadata <- subset(pbmc.metadata, sample_status == "IMPACC sample assayed and passed QC")
# Olink QC
nasal.metadata <- subset(nasal.metadata, sample_status=="IMPACC sample assayed and passed QC")

# mNGS QC
mngs.metadata <- subset(mngs.metadata,
                        sample_status == "IMPACC sample assayed and passed QC")
# Remove rows without sc2_rpm
mngs.metadata <- mngs.metadata[!is.na(mngs.metadata$sc2_rpm),]

# Add viral load to clin.sample.event.individ
clin.sample.event.individ.sc2 <- merge(
  clin.sample.event.individ, mngs.metadata[,c("sample_id","sc2_rpm")],
  by = "sample_id", all = FALSE, sort = FALSE
)
# Remove participant_id 014-2038 & 002-0040 visit 1, 010-0024 visit 6: they have 2 very different sc2_rpm for the duplicates
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "014-2038" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "002-0040" & event_type == "Visit 1"))
clin.sample.event.individ.sc2 <- subset(
  clin.sample.event.individ.sc2,
  !(participant_id == "010-0024" & event_type == "Visit 6"))

# Filter samples that pass QC
clin.sample.event.individ <- clin.sample.event.individ %>%
  subset((sample_id %in% pbmc.metadata$sample_id) |
           (sample_id %in% nasal.metadata$sample_id))
```

## Load pathways for GSEA

```{r}
# Reactome
reactome <- msigdbr::msigdbr(species = "Homo sapiens",
                             category = "C2", subcategory = "CP:REACTOME")
reactome.list <- split(x = reactome$ensembl_gene, f = reactome$gs_name)
reactome.list2 <- reactome %>% select(gs_name, ensembl_gene)
```

# Hospital visit 1 samples

```{r}
# Keep hospital visit 1 samples
clin.sample.event.individ.1 <- clin.sample.event.individ %>%
  subset(event_type=="Visit 1" & event_location=="Hospital")

# Add viral load
clin.sample.event.individ.1 <- merge(
  clin.sample.event.individ.1, clin.sample.event.individ.sc2[,c("event_id","sc2_rpm")],
  by = "event_id", all = FALSE, sort = FALSE
)
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.1$sc2_rpm <- log10(clin.sample.event.individ.1$sc2_rpm + 1)

# Drop unused levels
clin.sample.event.individ.1 <- droplevels(clin.sample.event.individ.1)

# Any row with no sc2_rpm?
print(any(is.na(clin.sample.event.individ.1$sc2_rpm)))
```

```{r}
# Subset PBMC RNA-seq visit 1 samples
pbmc.clin.sample.event.individ.1 <- clin.sample.event.individ.1 %>%
  subset(sample_id %in% colnames(pbmc.counts))

# Subset NS RNA-seq visit 1 samples
nasal.clin.sample.event.individ.1 <- clin.sample.event.individ.1 %>%
  subset(sample_id %in% colnames(nasal.counts))

# Any duplicated participant_id?
print(any(duplicated(pbmc.clin.sample.event.individ.1$participant_id)))
print(any(duplicated(nasal.clin.sample.event.individ.1$participant_id)))
```

Find overlapping event_id.

```{r}
my.event_id <- intersect(
  pbmc.clin.sample.event.individ.1$event_id,
  nasal.clin.sample.event.individ.1$event_id
)

# Subset
pbmc.clin.sample.event.individ.1 <- pbmc.clin.sample.event.individ.1 %>%
  subset(event_id %in% my.event_id)
nasal.clin.sample.event.individ.1 <- nasal.clin.sample.event.individ.1 %>%
  subset(event_id %in% my.event_id)

# Reorder based on event_id
pbmc.clin.sample.event.individ.1 <- pbmc.clin.sample.event.individ.1[order(pbmc.clin.sample.event.individ.1$event_id),]
nasal.clin.sample.event.individ.1 <- nasal.clin.sample.event.individ.1[order(nasal.clin.sample.event.individ.1$event_id),]

# Check if they are the same
print(all.equal(
  pbmc.clin.sample.event.individ.1,
  nasal.clin.sample.event.individ.1
  )
)
```

Extract PBMC and nasal data

```{r}
pbmc.counts.1 <- pbmc.counts[,pbmc.clin.sample.event.individ.1$sample_id]
nasal.counts.1 <- nasal.counts[,nasal.clin.sample.event.individ.1$sample_id]

# Rename sample_id to event_id
colnames(pbmc.counts.1) <- pbmc.clin.sample.event.individ.1$event_id
colnames(nasal.counts.1) <- nasal.clin.sample.event.individ.1$event_id
print(all(colnames(pbmc.counts.1) == colnames(nasal.counts.1)))

# QC on RNA-seq count data
pbmc.counts.1 <- pbmc.counts.1[pbmc.rData[pbmc.rData$gene_biotype=="protein_coding", "gene_id"],]
nasal.counts.1 <- nasal.counts.1[nasal.rData[nasal.rData$gene_biotype=="protein_coding", "gene_id"],]

# Keep genes with at least 10 counts in 20% of samples
keep <- rowSums(pbmc.counts.1>=10) >= 0.2*ncol(pbmc.counts.1)
pbmc.counts.1 <- pbmc.counts.1[keep,]
keep <- rowSums(nasal.counts.1>=10) >= 0.2*ncol(nasal.counts.1)
nasal.counts.1 <- nasal.counts.1[keep,]
```

## Explore data

```{r}
# Age distribution
hist(pbmc.clin.sample.event.individ.1$admit_age, breaks=20,
     xlab = "Admit age", main = "PBMC")

# Age by sex
boxplot(admit_age~sex, data=pbmc.clin.sample.event.individ.1, main = "PBMC")
```

Normalize data.

```{r}
# Normalize RNA-seq
pbmc.vwts <- voom(pbmc.counts.1,
                  normalize.method = "quantile",
                  plot = T)
nasal.vwts <- voom(nasal.counts.1,
                   normalize.method = "quantile",
                   plot = T)
```

## Gene-gene correlation

```{r}
res <- as.data.frame(matrix(NA, nrow=length(my.genes), ncol=4))
colnames(res) <- c("gene_id","all.corr","young.corr","old.corr")
rownames(res) <- my.genes
res$gene_id <- my.genes

for (i in rownames(res)) {
  res[i,"all.corr"] <- cor(pbmc.vwts$E[i,], nasal.vwts$E[i,],
                           method = "pearson")
  res[i,"young.corr"] <- cor(pbmc.vwts$E[i,pbmc.clin.sample.event.individ.1$admit_age<=46],
                             nasal.vwts$E[i,nasal.clin.sample.event.individ.1$admit_age<=46],
                             method = "pearson")
  res[i,"old.corr"] <- cor(pbmc.vwts$E[i,pbmc.clin.sample.event.individ.1$admit_age>=71],
                           nasal.vwts$E[i,nasal.clin.sample.event.individ.1$admit_age>=71],
                           method = "pearson")
}

# Add gene name
res <- merge(res, pbmc.rData[,c("gene_id","gene_name")],
             all.x = TRUE, all.y = FALSE,
             by = "gene_id", sort = FALSE)
```

Plot

```{r}
hist(res$all.corr,
     xlab="Pearson's correlation coefficient",
     main="Correlation of gene expression\nbetween PBMC and NS data")

plot(old.corr~young.corr, data=res,
     xlab="Pearson's correlation <= 46 yo",
     ylab="Pearson's correlation >= 71 yo",
     main="Correlation in gene expression between PBMC and NS")
```

Plot pretty

```{r}
# Figure 6D
p <- ggplot() +
  geom_point(data = res,
             aes(x=young.corr, y=old.corr,
                 color=as.factor(young.corr>0.5 & old.corr>0.5)),
             size=1) +
  geom_point(
    data=subset(res, gene_name %in% c("IFI6","IFI44","IFIT3","HLA-DQA2","HLA-DRB5")),
    aes(x=young.corr, y=old.corr), color="red", size=1
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_vline(xintercept = 0.5, linetype = "dashed") +
  scale_color_manual(values=c("TRUE"="black", "FALSE"="gray75"),
                     guide="none") +
  scale_x_continuous(breaks=c(-0.3,0,0.5,1), limits=c(-0.37,1), expand=c(0,0)) +
  scale_y_continuous(breaks=c(-0.3,0,0.5,1), limits=c(-0.37,1), expand=c(0,0)) +
  labs(
    x="Correlation coefficient in [18,46]",
    y="Correlation coefficient in [71,96]",
    title="PBMC-Nasal gene expression\ncorrelation at Visit 1"
  ) + my.theme + theme(aspect.ratio = 1) # pbmc_nasal_visit1_corr_age-group, 360 x 360
```

Enriched pathway and transcription factor.

```{r}
ora <- clusterProfiler::enricher(
  gene = res[res$young.corr>0.5 & res$old.corr>0.5,"gene_id"],
  TERM2GENE = reactome.list2,
  # TERM2GENE = hallmark.list2,
  universe = res$gene_id,
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  pAdjustMethod = "fdr")
ora.res <- ora@result

ora.c3 <- clusterProfiler::enricher(
  gene = res[res$young.corr>0.5 & res$old.corr>0.5,"gene_id"],
  TERM2GENE = tft.list2,
  # TERM2GENE = hallmark.list2,
  universe = res$gene_id,
  pvalueCutoff = 1,
  qvalueCutoff = 1,
  pAdjustMethod = "fdr")
ora.c3.res <- ora.c3@result
```

# Hospital Visit 1 analysis: PBMC and nasal RNA-seq, and olink

Here I will integrate these 3 data types and analyze oldest age group.

## Data preparation

```{r}
# Keep hospital visit 1 samples
clin.sample.event.individ.1 <- clin.sample.event.individ %>%
  subset((event_type=="Visit 1") & (event_location=="Hospital") &
           (admit_age>=71))

# Add viral load
clin.sample.event.individ.1 <- merge(
  clin.sample.event.individ.1, clin.sample.event.individ.sc2[,c("event_id","sc2_rpm")],
  by = "event_id", all = FALSE, sort = FALSE
)
# Convert viral load to log10(rpm + 1)
clin.sample.event.individ.1$sc2_rpm <- log10(clin.sample.event.individ.1$sc2_rpm + 1)

# Drop unused levels
clin.sample.event.individ.1 <- droplevels(clin.sample.event.individ.1)

# Any row with no sc2_rpm?
stopifnot(!is.na(clin.sample.event.individ.1$sc2_rpm))
```

```{r}
# Subset PBMC RNA-seq visit 1 samples
pbmc.clin.sample.event.individ.1 <- clin.sample.event.individ.1 %>%
  subset(sample_id %in% colnames(pbmc.counts))

# Subset NS RNA-seq visit 1 samples
nasal.clin.sample.event.individ.1 <- clin.sample.event.individ.1 %>%
  subset(sample_id %in% colnames(nasal.counts))

# Subset olink visit 1 samples
olink.clin.sample.event.individ.1 <- clin.sample.event.individ.1 %>%
  subset(sample_id %in% colnames(olink))

# Any duplicated participant_id?
stopifnot(!duplicated(pbmc.clin.sample.event.individ.1$participant_id))
stopifnot(!duplicated(nasal.clin.sample.event.individ.1$participant_id))
stopifnot(!duplicated(olink.clin.sample.event.individ.1$participant_id))
```

Merge the 3 omics.

```{r}
# Find overlapping event_id
my.event_id <- intersect(
  pbmc.clin.sample.event.individ.1$event_id,
  nasal.clin.sample.event.individ.1$event_id)
my.event_id <- intersect(
  my.event_id,
  olink.clin.sample.event.individ.1$event_id
)

# Subset
pbmc.clin.sample.event.individ.1 <- pbmc.clin.sample.event.individ.1 %>%
  subset(event_id %in% my.event_id)
nasal.clin.sample.event.individ.1 <- nasal.clin.sample.event.individ.1 %>%
  subset(event_id %in% my.event_id)
olink.clin.sample.event.individ.1 <- olink.clin.sample.event.individ.1 %>%
  subset(event_id %in% my.event_id)

# Reorder based on event_id
pbmc.clin.sample.event.individ.1 <- pbmc.clin.sample.event.individ.1[order(pbmc.clin.sample.event.individ.1$event_id),]
nasal.clin.sample.event.individ.1 <- nasal.clin.sample.event.individ.1[order(nasal.clin.sample.event.individ.1$event_id),]
olink.clin.sample.event.individ.1 <- olink.clin.sample.event.individ.1[order(olink.clin.sample.event.individ.1$event_id),]

# Check if they are the same
stopifnot(
  pbmc.clin.sample.event.individ.1$participant_id == nasal.clin.sample.event.individ.1$participant_id
)
stopifnot(
  pbmc.clin.sample.event.individ.1$participant_id == olink.clin.sample.event.individ.1$participant_id
)
stopifnot(
  pbmc.clin.sample.event.individ.1$event_id == nasal.clin.sample.event.individ.1$event_id
)
stopifnot(
  pbmc.clin.sample.event.individ.1$event_id == olink.clin.sample.event.individ.1$event_id
)
```

Extract Visit 1 counts PBMC and nasal data

```{r}
pbmc.counts.1 <- pbmc.counts[,pbmc.clin.sample.event.individ.1$sample_id]
nasal.counts.1 <- nasal.counts[,nasal.clin.sample.event.individ.1$sample_id]
olink.1 <- olink[,olink.clin.sample.event.individ.1$sample_id]

# Replace na with 0 in olink data
print(sprintf("Number of NA in olink data: %d", sum(is.na(olink.1))))
olink.1[is.na(olink.1)] <- 0

# Rename sample_id to event_id
colnames(pbmc.counts.1) <- pbmc.clin.sample.event.individ.1$event_id
colnames(nasal.counts.1) <- nasal.clin.sample.event.individ.1$event_id
colnames(olink.1) <- olink.clin.sample.event.individ.1$event_id

# Verify that the sample names of each count matrix match
stopifnot(colnames(pbmc.counts.1) == colnames(nasal.counts.1))
stopifnot(colnames(pbmc.counts.1) == colnames(olink.1))

# QC on RNA-seq count data
pbmc.counts.1 <- pbmc.counts.1[pbmc.rData[pbmc.rData$gene_biotype=="protein_coding", "gene_id"],]
nasal.counts.1 <- nasal.counts.1[nasal.rData[nasal.rData$gene_biotype=="protein_coding", "gene_id"],]

# Keep genes with at least 10 counts in 20% of samples
keep <- rowSums(pbmc.counts.1>=10) >= 0.2*ncol(pbmc.counts.1)
pbmc.counts.1 <- pbmc.counts.1[keep,]
keep <- rowSums(nasal.counts.1>=10) >= 0.2*ncol(nasal.counts.1)
nasal.counts.1 <- nasal.counts.1[keep,]

# Number of features per omics
print(dim(pbmc.counts.1))
print(dim(nasal.counts.1))
print(dim(olink.1))
```

Rename features to distinguish the 3 data types

```{r}
# Append ".pbmc"
rownames(pbmc.counts.1) <- paste0(
  rownames(pbmc.counts.1),
  ".pbmc")
# Append ".nasal"
rownames(nasal.counts.1) <- paste0(
  rownames(nasal.counts.1),
  ".nasal")
# Append ".olink"
rownames(olink.1) <- paste0(
  rownames(olink.1),
  ".olink")
```

Transform data

```{r}
# Calculate logCPM of gene counts
pbmc.cpm.1 <- edgeR::cpm(pbmc.counts.1, log=TRUE)
nasal.cpm.1 <- edgeR::cpm(nasal.counts.1, log=TRUE)

```

Concatenate the 3 data types together

```{r}
concat.1 <- rbind(
  pbmc.cpm.1, nasal.cpm.1, olink.1
)
print(dim(concat.1))
```

## PCA

```{r}
pca <- prcomp(
  t(concat.1), # prcomp expects samples-by-features shape
  center=TRUE, scale.=TRUE)
print(pca)
```

```{r}
library(ggfortify) # for plotting PCA
```

```{r}
autoplot(
  pca,
  scale=0) +
  geom_vline(xintercept=0, color="grey") +
  scale_x_continuous(
    limits=c(-320,100), expand=c(0,0)) +
  scale_y_continuous(
    limits=c(-230,500), expand=c(0,0), breaks=c(-200,0,250,500)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0.5,1,0.5,0), "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

```{r}
autoplot(
  pca, data=pbmc.clin.sample.event.individ.1,
  scale=0,
  color="sex")
autoplot(
  pca, data=pbmc.clin.sample.event.individ.1,
  scale=0,
  color="admit_age")
autoplot(
  pca, data=pbmc.clin.sample.event.individ.1,
  scale=0,
  color="respiratory_status")
autoplot(
  pca, data=pbmc.clin.sample.event.individ.1,
  scale=0,
  color="trajectory_group")
autoplot(
  pca, data=pbmc.clin.sample.event.individ.1 %>%
    mutate(severe=as.factor(respiratory_status %in% c("5","6"))),
  scale=0,
  color="severe")
```

```{r}
# Variance explained by each component
pca.sum <- summary(pca)

print(pca.sum)
```

```{r}
# Figure S13B
p <- ggplot(
  data=data.frame(var.exp=pca.sum$importance[2,1:50]),
  aes(x=seq(1,50),y=var.exp)) +
  geom_point(size=0.6) +
  geom_line() +
  scale_x_continuous(limits=c(0,51), expand=c(0,0)) +
  scale_y_continuous(limits=c(0,0.205), expand=c(0,0)) +
  labs(x="PC",
       y="Proportion of variance \nexplained by each PC") +
  my.theme # pbmc_nasal_olink_visit1_PCA_var_explained, 3.6w x 3.2h in
```

### Analysis of PC1+ and PC1- subgroups

```{r}
# Categorize patients into two groups based on positive or negative PC1
pc1 <- pca$x[,1,drop=FALSE] %>%
  as.data.frame() %>% 
  tibble::rownames_to_column(var="event_id")
pc1$PC1.group <- ifelse(
  pc1$PC1>0, "PC1+", "PC1-")

head(pc1)
```

```{r}
# Add PC1 group category to metadata
pbmc.clin.sample.event.individ.1 <- merge(
  pbmc.clin.sample.event.individ.1, pc1,
  by="event_id")
nasal.clin.sample.event.individ.1 <- merge(
  nasal.clin.sample.event.individ.1, pc1,
  by="event_id")
olink.clin.sample.event.individ.1 <- merge(
  olink.clin.sample.event.individ.1, pc1,
  by="event_id"
)
```

Plot pretty PCA plot

```{r}
# Figure S13A
p <- autoplot(
  pca,
  data=pbmc.clin.sample.event.individ.1,
  color="PC1.group",
  scale=0, size=0.6) +
  geom_vline(xintercept=0, color="black", linetype="dashed", linewidth=0.4) +
  scale_colour_brewer(palette="Dark2") +
  scale_x_continuous(
    limits=c(-320,100), expand=c(0,0)) +
  scale_y_continuous(
    limits=c(-230,500), expand=c(0,0), breaks=c(-200,0,250,500)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0.5,1,0.5,0), "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) # pbmc_nasal_olink_visit1_PCA, 4.5w x 3.2h in
```

Check potential confounders

```{r}
p1 <- autoplot(
  pca, data=pbmc.clin.sample.event.individ.1,
  scale=0,
  color="sex", size=0.6) +
  geom_vline(xintercept=0, color="grey", linewidth=0.8) +
  scale_color_discrete(name="Sex") +
  scale_x_continuous(
    limits=c(-320,100), expand=c(0,0)) +
  scale_y_continuous(
    limits=c(-230,500), expand=c(0,0), breaks=c(-200,0,250,500)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0.5,1,0.5,0), "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
p2 <- autoplot(
  pca, data=pbmc.clin.sample.event.individ.1,
  scale=0,
  color="admit_age", size=0.6) +
  geom_vline(xintercept=0, color="grey", linewidth=0.8) +
  viridis::scale_color_viridis(name="Age") +
  scale_x_continuous(
    limits=c(-320,100), expand=c(0,0)) +
  scale_y_continuous(
    limits=c(-230,500), expand=c(0,0), breaks=c(-200,0,250,500)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0.5,1,0.5,0), "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
p3 <- autoplot(
  pca, data=pbmc.clin.sample.event.individ.1,
  scale=0,
  color="respiratory_status", size=0.6) +
  geom_vline(xintercept=0, color="grey", linewidth=0.8) +
  scale_color_discrete(name="NIAID\nOS") +
  scale_x_continuous(
    limits=c(-320,100), expand=c(0,0)) +
  scale_y_continuous(
    limits=c(-230,500), expand=c(0,0), breaks=c(-200,0,250,500)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0.5,1,0.5,0), "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
p4 <- autoplot(
  pca, data=pbmc.clin.sample.event.individ.1,
  scale=0,
  color="trajectory_group", size=0.6) +
  geom_vline(xintercept=0, color="grey", linewidth=0.8) +
  scale_color_discrete(name="Trajectory\ngroup") +
  scale_x_continuous(
    limits=c(-320,100), expand=c(0,0)) +
  scale_y_continuous(
    limits=c(-230,500), expand=c(0,0), breaks=c(-200,0,250,500)) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0.5,1,0.5,0), "cm"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

p1 + p2 + p3 + p4 + plot_layout(nrow=2) # pbmc_nasal_olink_visit1_PCA_confounders, 9 x 6.5 in
```

Quantify the confounders between the two PC1 groups

```{r}
# Age
p1 <- ggplot(
  data=pbmc.clin.sample.event.individ.1,
  aes(x=admit_age, y=PC1.group, color=PC1.group)) +
  stat_boxplot(geom ="errorbar", width=0.3) + 
  geom_boxplot(fill="white", outlier.shape=NA, width=0.5) +
  ggbeeswarm::geom_quasirandom(width=0.35, size=0.5, orientation="y") +
  scale_colour_brewer(palette="Dark2", guide="none") +
  labs(x="Age", y="") +
  my.theme
print(wilcox.test(admit_age ~ PC1.group,
                  data=pbmc.clin.sample.event.individ.1))

# Sex
p2 <- ggplot(
  data=pbmc.clin.sample.event.individ.1,
  aes(fill=sex, y=PC1.group)) +
  geom_bar(stat="count", position=position_fill(reverse=TRUE), width=0.5) +
  scale_x_continuous(
    limits=c(0,1), expand=c(0,0)) +
  scale_fill_brewer(palette="Set1", name="Sex") +
  labs(x="Relative proportion\nof each sex",
       y="") +
  my.theme
print(chisq.test(pbmc.clin.sample.event.individ.1$PC1.group,
                 pbmc.clin.sample.event.individ.1$sex))

# NIAID OS
p3 <- ggplot(
  data=pbmc.clin.sample.event.individ.1,
  aes(fill=respiratory_status, y=PC1.group)) +
  geom_bar(stat="count", position=position_fill(reverse=TRUE), width=0.5) +
  scale_x_continuous(
    limits=c(0,1), expand=c(0,0)) +
  scale_fill_brewer(palette="Set1", name="NIAID\nOS") +
  labs(x="Relative proportion\nof each OS",
       y="") +
  my.theme
print(chisq.test(pbmc.clin.sample.event.individ.1$PC1.group,
                 pbmc.clin.sample.event.individ.1$respiratory_status))

# Trajectory group
p4 <- ggplot(
  data=pbmc.clin.sample.event.individ.1,
  aes(fill=trajectory_group, y=PC1.group)) +
  geom_bar(stat="count", position=position_fill(reverse=TRUE), width=0.5) +
  scale_x_continuous(
    limits=c(0,1), expand=c(0,0)) +
  scale_fill_brewer(palette="Set1", name="Trajectory\ngroup") +
  labs(x="Relative proportion\nof each trajectory group",
       y="") +
  my.theme
print(chisq.test(pbmc.clin.sample.event.individ.1$PC1.group,
                 pbmc.clin.sample.event.individ.1$trajectory_group))

# Figure S13C-F
p1 + p2 + p3 + p4 + plot_layout(nrow=2) # pbmc_nasal_olink_visit1_PC1_confounders, 8.9w x 5.5h in
```

#### PBMC RNA-seq data

```{r}
design <- model.matrix(~ PC1.group + sex + respiratory_status,
                       data=pbmc.clin.sample.event.individ.1)

vwts <- voom(pbmc.counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts, design=design)
vfit <- eBayes(vfit)

pbmc.res <- topTable(
  vfit, coef = "PC1.groupPC1+",
  sort.by = "none", number = Inf,
  adjust.method = "BH", p.value = 1)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

Volcano plot

```{r}
# Figure S14B
p <- ggplot(data=pbmc.res) +
  geom_point(aes(x=logFC, y=-log10(adj.P.Val)),
             color="gray75", size=0.5) +
  scale_x_continuous(limits=c(-2.1,2.1), expand=c(0,0)) +
  scale_y_continuous(limits=c(-0.02,1), expand=c(0,0)) +
  labs(x="log2(fold change)", y="-log10(adjusted P)",
       title="PBMC genes associated\nwith PC1 at Visit 1") +
  my.theme # pbmc_nasal_olink_visit1_PC1_pbmc_volcano, 3.2w x 3.5h in
```

Reactome GSEA

```{r}
# Sort gene list in descending order
gene.ranks <- pbmc.res$t
names(gene.ranks) <- rownames(pbmc.res) %>%
  sapply(FUN=function(x) strsplit(x,".",fixed=TRUE)[[1]][1]) # remove the ".pbmc" suffix
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
pbmc.reactome.gsea <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
```

#### Nasal RNA-seq data

```{r}
design <- model.matrix(~ PC1.group + sex + respiratory_status,
                       data=nasal.clin.sample.event.individ.1)

vwts <- voom(nasal.counts.1, 
             design = design,
             normalize.method = "quantile",
             plot = T)

vfit <- lmFit(vwts , design=design)
vfit <- eBayes(vfit)

nasal.res <- topTable(
  vfit, coef = "PC1.groupPC1+",
  sort.by = "none", number = Inf,
  adjust.method = "BH", p.value = 1)

summary(decideTests(vfit, adjust.method = "BH", p.value = 0.05))
```

Volcano plot

```{r}
# Figure S14A
p <- ggplot(
  data=nasal.res %>%
    mutate(tmp=ifelse(
      adj.P.Val>=0.05, "NS",
      ifelse(logFC>0, "PC1+", "PC1-")))) +
  geom_point(aes(x=logFC, y=-log10(adj.P.Val),
                 color=tmp), size=0.5) +
  scale_x_continuous(limits=c(-5.2,5.2), expand=c(0,0)) +
  scale_color_manual(values=c("PC1+"="#d95f02","PC1-"="#1b9e77","NS"="gray75"),
                     guide="none") +
  scale_y_continuous(limits=c(-0.8,40), expand=c(0,0)) +
  labs(x="log2(fold change)", y="-log10(adjusted P)",
       title="Nasal genes associated\nwith PC1 at Visit 1") +
  my.theme # pbmc_nasal_olink_visit1_PC1_nasal_volcano, 3.4 x 3.5 in
```

Reactome GSEA

```{r}
# Sort gene list in descending order
gene.ranks <- nasal.res$t
names(gene.ranks) <- rownames(nasal.res) %>%
  sapply(FUN=function(x) strsplit(x,".",fixed=TRUE)[[1]][1])
gene.ranks <- sort(gene.ranks, decreasing = TRUE)

set.seed(1)
nasal.reactome.gsea <- fgseaMultilevel(
  pathways = reactome.list,
  stats = gene.ranks,
  minSize = 15,
  maxSize = 500,
  nproc = 1
)
```

#### Olink data

```{r}
# Linear model
olink.res <- data.frame(matrix(NA, nrow=nrow(olink.1), ncol=4))
colnames(olink.res) <- c("cytokine","slope","PVal","adj.PVal")
rownames(olink.res) <- rownames(olink.1)
olink.res$cytokine <- rownames(olink.res)

for (i in rownames(olink.res)) {
  temp <- data.frame(conc=scale(olink.1[i,]), 
                     PC1=olink.clin.sample.event.individ.1$PC1.group,
                     sex=olink.clin.sample.event.individ.1$sex,
                     resp=olink.clin.sample.event.individ.1$respiratory_status) %>%
    subset(!is.na(conc))
  my.lm <- lm(conc ~ PC1 + sex + resp, data=temp)
  olink.res[i,"slope"] <- summary(my.lm)$coefficients["PC1PC1+",1]
  olink.res[i,"PVal"] <- summary(my.lm)$coefficients["PC1PC1+",4]
  
  remove(temp, my.lm)
}

# Correct P-values
olink.res$adj.PVal <- p.adjust(olink.res$PVal, method = "BH")
```

Volcano plot

```{r}
# Figure S14C
p <- ggplot(data=olink.res) +
  geom_point(aes(x=slope, y=-log10(adj.PVal)),
             color="gray75", size=0.5) +
  scale_x_continuous(limits=c(-1.05,1.05), expand=c(0,0)) +
  scale_y_continuous(limits=c(-0.02,1), expand=c(0,0)) +
  labs(x="log2(fold change)", y="-log10(adjusted P)",
       title="Proteins associated\nwith PC1 at Visit 1") +
  my.theme # pbmc_nasal_olink_visit1_PC1_olink_volcano, 3.2 x 3.5 in
```

#### Reactome of PBMC and nasal RNA-seq data

```{r}
# Manually choose pathways
my.pathways <- c("REACTOME_INTERFERON_ALPHA_BETA_SIGNALING",
                 "REACTOME_INTERFERON_GAMMA_SIGNALING",
                 "REACTOME_INTERLEUKIN_1_FAMILY_SIGNALING",
                 "REACTOME_INTERLEUKIN_2_FAMILY_SIGNALING",
                 "REACTOME_INTERLEUKIN_3_INTERLEUKIN_5_AND_GM_CSF_SIGNALING",
                 "REACTOME_INTERLEUKIN_6_FAMILY_SIGNALING","REACTOME_INTERLEUKIN_7_SIGNALING",
                 "REACTOME_INTERLEUKIN_10_SIGNALING","REACTOME_INTERLEUKIN_17_SIGNALING",
                 "REACTOME_INTERLEUKIN_4_AND_INTERLEUKIN_13_SIGNALING",
                 "REACTOME_TOLL_LIKE_RECEPTOR_CASCADES",
                 "REACTOME_TOLL_LIKE_RECEPTOR_TLR1_TLR2_CASCADE",
                 "REACTOME_MYD88_INDEPENDENT_TLR4_CASCADE",
                 "REACTOME_TOLL_LIKE_RECEPTOR_9_TLR9_CASCADE",
                 "REACTOME_COMPLEMENT_CASCADE",
                 "REACTOME_CASPASE_ACTIVATION_VIA_DEATH_RECEPTORS_IN_THE_PRESENCE_OF_LIGAND",
                 "REACTOME_TCR_SIGNALING",
                 "REACTOME_SIGNALING_BY_THE_B_CELL_RECEPTOR_BCR",
                 "REACTOME_DOWNSTREAM_SIGNALING_EVENTS_OF_B_CELL_RECEPTOR_BCR",
                 "REACTOME_PD_1_SIGNALING")
```

```{r}
# Combine PBMC and nasal GSEA results
combined.gsea <- rbind(
  pbmc.reactome.gsea %>%
    subset(pathway %in% my.pathways) %>%
    mutate(omics="PBMC"),
  nasal.reactome.gsea %>%
    subset(pathway %in% my.pathways) %>%
    mutate(omics="Nasal")
)

# Reformat pathway names
combined.gsea$pathway <- gsub("^REACTOME_","",combined.gsea$pathway)
combined.gsea$pathway <- gsub("_"," ",combined.gsea$pathway)
combined.gsea$pathway <- stringr::str_to_sentence(combined.gsea$pathway)

# Fix capitalization
my.dict <- list("tlr"="TLR", "Myd88"="MYD88",
                "Interleukin "="IL", "interleukin "="IL",
                "gm csf"="GM-CSF",
                "Toll like receptor 9"="",
                "Toll like receptor TLR1"="TLR1",
                "Toll like receptor"="TLR",
                "Pd 1"="PD1", "Tcr"="TCR",
                "b cell receptor "="", "bcr"="BCR",
                "Caspase activation via death receptors in the presence of ligand" = "Caspase activation via death receptors")
for (i in names(my.dict)) {
  combined.gsea$pathway <- gsub(i, my.dict[[i]], combined.gsea$pathway)
}
```

```{r}
# Figure S14D
p <- ggplot(data = combined.gsea, 
            aes(y=reorder(pathway,NES,FUN=function(x) x[2]), 
                x=NES)) +
  geom_point(aes(color=as.factor(NES>0)), fill="white",
             shape=21, size=2.5, stroke=1) +
  geom_point(data=. %>% subset(padj<0.05),
             aes(color=as.factor(NES>0), fill=as.factor(NES>0)),
             shape=21, size=2.5, stroke=1) +
  geom_text(
    data=. %>% subset(NES>0),
    aes(x=0,label=sprintf("P = %.1e", padj)), 
    hjust=1.08, size=3.5) +
  geom_text(
    data=. %>% subset(NES<0),
    aes(x=0, label=sprintf("P = %.1e", padj)),
    hjust=-0.08, size=3.5) +
  geom_vline(xintercept = 0, linewidth = 0.7, color = "black") +
  scale_color_manual(values=c("TRUE"="#d95f02", "FALSE"="#1b9e77"), guide="none") +
  scale_fill_manual(values=c("TRUE"="#d95f02", "FALSE"="#1b9e77"), guide="none") +
  scale_x_continuous(limits = c(-4,4), expand = c(0,0)) +
  labs(y="", x="GSEA normalized\nenrichment score",
       title="Pathways associated\nwith PC1 at Visit 1") +
  facet_wrap(~omics) +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size=12, face="plain"),
    axis.text = element_text(size=12, color="black", family="Arial"),
    text = element_text(size=12, family="Arial"),
    plot.margin = unit(c(0,1,1,0), "cm"),
    panel.spacing = unit(0.5, "cm"),
    panel.grid.minor=element_blank()
  ) # pbmc_nasal_olink_visit1_PC1_gsea-reactome, 6.5 x 8 in
```

