---
title: "Spear feature association analysis"
author: "Gisela Gabernet (gisela.gabernet@yale.edu)"
subtitle: PGX, SO, PMG, PPT, PPG, BCT
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
params:
  results_dir: "associate_top_spear_features_promis_physical_Factor1_test/"
  spear_object: "/scratch/data-fullcohort/spear_model/results/020624/032124_spearphysical_finalscores_allsamples.rds"
  features: "spear"
  partition: "test_train_nophy_nooverlap"
editor_options: 
  chunk_output_type: inline
---

```{r}

# list and install if needed CRAN packages
#install.packages("remotes")
#Specific version of Matrix needed
#remotes::install_version("Matrix", version = "1.6-1")
packages <- c("data.table", "RColorBrewer", "ggplot2", "tidyverse", "psych", "lme4", "ordinal", "nlme", "gamm4",
              "mgcv", "ggsignif", "kableExtra", "corrplot", "ggeffects", "ggfortify", "grid", "car", "rstatix", "ggpubr", "BiocManager", "stringr", "pheatmap", "ggbeeswarm")

new_pkg <- packages[!(packages %in% installed.packages())]
if (length(new_pkg)>0) {install.packages(new_pkg)}

# list (and install if needed) Bioconductor packages
packages_bioconductor <- c("qvalue", "MOFA2")
new_pkg_bioconductor <- packages_bioconductor[!(packages_bioconductor %in% installed.packages())]
if (length(new_pkg_bioconductor)>0) {BiocManager::install(new_pkg_bioconductor)}

# load packages
# load packages
packages <- sort(append(packages, packages_bioconductor)) # so the list of versions will be in order
for (n in seq_along(packages)) {
  suppressPackageStartupMessages(library(packages[n], character.only = TRUE))
  cat(paste0(packages[n], ": ", packageVersion(packages[n]), "\n")) # print simplified package versions
}

knitr::opts_chunk$set(warning = F, message=F)

print_kable <- function(table, kable_height = "500px", kable_width = "100%") {
  kable(table, digits=3) %>%
    kable_styling("striped") #%>%
    #scroll_box(height = kable_height, width = kable_width)
  
}

# save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
#    stopifnot(!missing(x))
#    stopifnot(!missing(filename))
#    pdf(filename, width=width, height=height)
#    grid::grid.newpage()
#    grid::grid.draw(x$gtable)
#    dev.off()
# }

```

# Running parameters

Running with parameters:
- features:
  - spear: get spear top features with joint probability >0.95
  - hallmark: get core set from hallmark heme metabolism pathway
- partition:
  - all: use all data for association
  - test: use the test dataset for association (with and without PROMIS PHY values)
  - PRO_PHY_NA_TEST: use all test dataset for association (with and without PROMIS PHY values, and the train dataset without PROMIS PHY values available)
- Spear object: spear object containing factor scores, and joint probabilities for extracting features.
- results_dir: name of results dir, will be created.

Example command to execute using an R script:

```{r eval=F}
# rmarkdown::render("associate_top_spear_features.Rmd",
#                   output_file="associate_top_spear_features_promis_physical_Factor1.html",
#                   params=list(
#                     results_dir = "associate_top_spear_features_promis_physical_Factor1/",
#                     spear_object = "/scratch/data-fullcohort/spear_model/results/factor_scores/031124_spearphysical_finalscores.rds",
#                     partition = "all",
#                     features = "spear"
#                   ))
```


# Explore Spear results

## Explore Feature correlation and variance explained

```{r fig.height=5, fig.width=10}

spear_object <- readRDS(params$spear_object)

data_env <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_convalescent_data_env_all_Feb_11_2024.RDS")
mofa_model_train <- readRDS("/scratch/data-fullcohort/MOFA_factors/MOFA_train_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240212.rds")
mofa_model_test <- readRDS("/scratch/data-fullcohort/MOFA_factors/MOFA_test_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240212.rds")

#results_dir <- "./associate_spear_factor_promis_physical_train/"
results_dir <- params$results_dir
dir.create(results_dir)

figures_dir <- "Figures_spear_features"
dir.create(figures_dir)


path_to_codebase <- "~/repos/impacc-full-cohort/Analysis/IMPACC_data_analysis_codebase.R"
source(path_to_codebase)

selected_factor_name <- "Factor1"
```


## Get clinical data

```{r}
clinical_data <- data_env$clinical_data

# Add additional symptom groups to metadata
symptom_groups <- read_csv("/scratch/clinical/current/impacc_conv_symptoms.csv", show_col_types = F)
PRO_scores <- read_csv("/scratch/clinical/current/PRO_scores_dates_updated_29Aug23.csv", show_col_types = F)

```


# Identify features to associate

Plotting features from `r params$features`.

```{r spear features}
if (params$features == "spear") {
  top_analyte_scores <- spear_object$analyte.scores %>% 
                              dplyr::filter(Factor == selected_factor_name) %>%
                              dplyr::mutate(joint.probability.rounded = round(joint.probability, 2)) %>%
                              dplyr::filter(joint.probability.rounded >= 0.95)
}


PRO_scores_PHY_avail <- PRO_scores %>% filter(!is.na(PROMIS_Physical_score))
event_with_PRO_PHY <- unique(PRO_scores_PHY_avail$event_id)

PRO_scores_PHY_na <- PRO_scores %>% filter(is.na(PROMIS_Physical_score))
event_without_PRO_PHY <- unique(PRO_scores_PHY_na$event_id)
```

# Create feature matrix

```{r}


train_data <- MOFA2::get_data(mofa_model_train)
test_data <- MOFA2::get_data(mofa_model_test)

detach("package:MOFA2", unload=TRUE)


feature_data_all <- list(
  SO = bind_rows(as.data.frame(t(train_data$SO$group1)),as.data.frame(t(test_data$SO$group1))),
  PMG = bind_rows(as.data.frame(t(train_data$PMG$group1)),as.data.frame(t(test_data$PMG$group1))),
  PPT = bind_rows(as.data.frame(t(train_data$PPT$group1)),as.data.frame(t(test_data$PPT$group1))),
  PPG = bind_rows(as.data.frame(t(train_data$PPG$group1)),as.data.frame(t(test_data$PPG$group1))),
  PGX = bind_rows(as.data.frame(t(train_data$PGX$group1)),as.data.frame(t(test_data$PGX$group1))),
  BCT = bind_rows(as.data.frame(t(train_data$BCT$group1)),as.data.frame(t(test_data$BCT$group1)))
)
  
if (params$partition == "all"){

  feature_data <- feature_data_all
  
} else if (params$partition == "test_train_nophy_nooverlap"){
  
  factor_scores <- rbind(spear_object$Test, spear_object$TestNoPhysical, spear_object$TrainNoPhysical_NoOverlap)
  event_ids <- rownames(factor_scores)
  # Rownames event_ids, colnames features
  feature_data <- lapply(feature_data_all, function(x) x[row.names(x) %in% event_ids,])

} else if (params$partition == "train_phy"){

  factor_scores <- spear_object$Train
  event_ids <- rownames(factor_scores)
  feature_data <- lapply(feature_data_all, function(x) x[row.names(x) %in% event_ids,])
  
} else {
  stop("Invalid partition parameter. Choose one of: all, test_train_nophy_nooverlap, train_phy")
}

print(lapply(feature_data,dim))

# Call top_feat_data_df <- as.data.frame(feature_data)

```

```{r}
# DFs for translating features to human readable
# PMG
metabolite_row_DF = data_env$plasma_metabolomics_global_rowfeature
metabolite_row_DF$CHEMICAL_NAME <- str_replace_all(metabolite_row_DF$CHEMICAL_NAME," ","_")
metabolite_row_DF$CHEMICAL_NAME <- str_replace_all(metabolite_row_DF$CHEMICAL_NAME,"[[:punct:]]","")

metabolomics_translate <- data.frame(metabolite_id = row.names(metabolite_row_DF),
                                     CHEMICAL_NAME = str_replace_all(metabolite_row_DF$CHEMICAL_NAME," ","_"))


gene_ensembl_PGX <- data_env$pbmc_transcriptomics_rowfeature %>%
                    rownames_to_column("ensembl_gene_id") %>%
                    dplyr::select(ensembl_gene_id, gene_name)

sample_event <- clinical_data %>% dplyr::select(sample_id, event_id) %>% distinct()

if (params$features == "spear") {

  top_feat_data <- list()
  for (assay in unique(top_analyte_scores$Assay)) {
    print(assay)
    
    selected_features_assay <- top_analyte_scores %>% filter(Assay == assay)
    
    top_feature_data <- list()
    
    if (nrow(selected_features_assay > 0)) {
      top_feat_data_assay <- feature_data[[assay]] %>%
                            dplyr::select(all_of(selected_features_assay$Analyte))
      
      if (assay == "PGX"){
        
        top_feat_data_assay_translate <- as.data.frame(t(top_feat_data_assay)) %>%
                                              rownames_to_column("ensembl_gene_id") %>%
                                              left_join(gene_ensembl_PGX, 
                                                        by="ensembl_gene_id") %>%
                                              dplyr::select(!ensembl_gene_id) %>%
                                              column_to_rownames("gene_name")
        top_feat_data_assay_translated <- as.data.frame(t(top_feat_data_assay_translate))
        
      } else if (assay == "PMG") {
        
        top_feat_data_assay_translate <- as.data.frame(t(top_feat_data_assay)) %>%
                                              rownames_to_column("metabolite_id") %>%
                                              left_join(metabolomics_translate,
                                                        by="metabolite_id") %>%
                                              dplyr::select(!metabolite_id) %>%
                                              column_to_rownames("CHEMICAL_NAME")
        top_feat_data_assay_translated <- as.data.frame(t(top_feat_data_assay_translate))
        
      } else {
        
        top_feat_data_assay_translated <- top_feat_data_assay
        top_feat_data_assay_translate <- NULL
        
      }
      
      colnames(top_feat_data_assay_translated) <- paste(assay,colnames(top_feat_data_assay_translated), sep = "_")
      top_feat_data[[assay]] <- top_feat_data_assay_translated
      rm(top_feat_data_assay_translated, 
         top_feat_data_assay_translate, top_feat_data_assay)
    }
  }
  
  top_feat_data_df <- bind_cols(top_feat_data)
  
  top_analyte_scores_omics <- split(top_analyte_scores, top_analyte_scores$Assay)
  top_analyte_scores_omics$PGX <- top_analyte_scores_omics$PGX %>%
                                        left_join(gene_ensembl_PGX, 
                                                  by=join_by(Analyte == ensembl_gene_id)) %>%
                                        dplyr::rename(translated_name = gene_name)
  top_analyte_scores_omics$PMG <- top_analyte_scores_omics$PMG %>%
                                        left_join(metabolomics_translate,
                                                  by=join_by(Analyte == metabolite_id)) %>%
                                        dplyr::rename(translated_name = CHEMICAL_NAME)
  top_analyte_scores_omics$PPG <- top_analyte_scores_omics$PPG %>%
                                        dplyr::mutate(translated_name = Analyte)
  top_analyte_scores_omics$SO <- top_analyte_scores_omics$SO %>%
                                        dplyr::mutate(translated_name = Analyte)
  top_analyte_scores_translated <- dplyr::bind_rows(top_analyte_scores_omics)

} else if (params$features == "hallmark") {
  
    GSEA_result_hallmark_Factor1_PGX <- read.table("enrichment_spear_factors_promis_physical_Factor1/GSEA_hallmark_Factor1_PGX_results.tsv", sep = "\t")
    top_analytes_core_list <- GSEA_result_hallmark_Factor1_PGX %>% 
                                filter(ID == "HALLMARK_HEME_METABOLISM") %>%
                                pull(core_enrichment)
    
    top_analytes_core <- str_split(top_analytes_core_list, "/")[[1]]
    
    feat_data_assay <- feature_data[["PGX"]]
    feat_data_assay_translate <- as.data.frame(t(feat_data_assay)) %>%
                                              rownames_to_column("ensembl_gene_id") %>%
                                              left_join(gene_ensembl_PGX, by="ensembl_gene_id") %>%
                                              dplyr::select(!ensembl_gene_id) %>%
                                              column_to_rownames("gene_name")
    feat_data_assay_translated <- as.data.frame(t(feat_data_assay_translate))
    
    top_feat_data_df  <- feat_data_assay_translated %>% 
                                        dplyr::select(all_of(top_analytes_core))
    colnames(top_feat_data_df) <- paste("PGX",colnames(top_feat_data_df), sep = "_")
    
}

```

# Associate features with outcomes

Here we will use a mixed linear model to associate features with outcomes

## Prepare metadata

```{r prepare the metadata for mixed linear model analysis, echo= FALSE, warning= FALSE}

correlation_metadata <- clinical_data %>%
  dplyr::select(event_id, resp_status_v1, sex, admit_age, event_after_first_vaccine_dose, 
                race, enrollment_site, event_date, 
                event_date_3weeks_after_dose, pro_group_labels, event_type, 
                discretized_admit_age_quantile, participant_id, partition)  %>%
  dplyr::mutate(pro_2_groups = 
      case_match( pro_group_labels,
      "MIN" ~ "MIN",
      c("PHY", "COG", "MLT")  ~ "OTHER"
      )
  ) %>%
  dplyr::distinct() %>%
  filter(event_id %in% rownames(top_feat_data_df))

clinical_subsets_all <- as.data.frame(correlation_metadata)
rownames(clinical_subsets_all) <- clinical_subsets_all$event_id


endpoint_coefficients_all <- data.frame(matrix(nrow=ncol(top_feat_data_df), ncol = 0))
rownames(endpoint_coefficients_all) <- colnames(top_feat_data_df)
```

## Check feature distribution over train and test set

```{r fig.height = 15, fig.width=8}
top_features_long <- top_feat_data_df %>%
                      rownames_to_column("event_id") %>%
                      tidyr::pivot_longer(!event_id, names_to = "feature", values_to="value")

plot_split_df <- top_features_long %>%
                    left_join(clinical_subsets_all)

p <- ggplot(plot_split_df, aes(x=partition, y=value, fill=pro_2_groups)) +
  #geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  geom_boxplot() +
  facet_wrap(~feature, drop = T, scales = "free_y", ncol = 4) + 
  theme_bw() #+ theme(legend.position = "none")
p

```

## Functions for association

```{r}

pro_min_colors <- c("#639A21", "#781B6C")
names(pro_min_colors) <- c("MIN", "OTHER")


pro_min_labels <- c("MIN", "LC")
names(pro_min_labels) <- c("MIN", "OTHER")

#original
#pro_colors <- c("#639A21", "#6371AD", "#FFC20A", "#9C3418")
pro_colors <- c("#639A21", "#6371AD", "#FFC20A", "#7D2D17")

col_annot = c("#8DD3C7", "#FFFFB3", "#BEBADA", "#FB8072")
names(col_annot) = c("SO", "PMG", "PPG", "PGX")

#pro_colors <- c("#3CE0B3", "#0072B2", "#F0E438", "#D55E00" )
names(pro_colors) <- c("MIN", "COG", "PHY", "MLT")

ordered_visits <- c("Visit 7", "Visit 8", "Visit 9", "Visit 10" )

scaleFUN <- function(x) sprintf("%.2e",x)
violin_plot_visits <- function(data, colors, limits=NULL, ncol=5) {
  p <- ggplot(data, aes(x=endpoints, y=value, fill=endpoints)) +
    geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
    facet_wrap(~Feature, drop = T, ncol = ncol, scales = "free_y") +
    theme_bw() + theme(legend.position = "none")
  if(!(is.null(limits))){
    p <- p + scale_y_continuous(limits = limits)
  } else {
    p <- p + scale_y_continuous(labels = scaleFUN)
  }
  if(!(missing(colors))){
    p <- p + scale_color_manual(values=colors)
    p <- p + scale_fill_manual(values=colors)
  }
  return(p)
}

pairwise_association_visit <- function(clinical_subsets_all, 
                                       feature_data, visit_id, 
                                       selected_features_names, 
                                       endpoint_var = "pro_2_groups",
                                       ordered_endpoint_var = c("MIN","OTHER"),
                                       colors = pro_min_colors
                                       ){
  cur_visit <- visit_id

  visit_row_ids = rownames(clinical_subsets_all[clinical_subsets_all$event_type == cur_visit,])
  
  # Use the visit7_row_ids to only get samples from 'Visit 7':
  features_filtered = feature_data[visit_row_ids , colnames(top_feat_data_df), drop=F]
  clinical_subsets = clinical_subsets_all[visit_row_ids ,]
  
  data_use = features_filtered
  
  # Add the respiratory_status_day14 as $endpoints
  data_use$endpoints = factor(clinical_subsets[[endpoint_var]], levels=ordered_endpoint_var)
  
  # Add the enrollment_site as $sites
  data_use$sites = clinical_subsets$enrollment_site
  data_use$age =  clinical_subsets$discretized_admit_age_quantile
  data_use$sex =  clinical_subsets$sex
  
  # Add the $control, needs to be different for each sample
  # required for clmm (always requires some random effect)
  data_use$control =  factor(1:nrow(data_use))
  
  # Set module names:
  feature_names = colnames(features_filtered)
  
  # Only use samples that have valid endpoints (not NA/missing)
  data_use = data_use[!is.na(data_use$endpoints),]

  for(j in 1:length(feature_names)){
    #print(j)
    my.formula0 = paste0(feature_names[j],"~ (1|sites)+age+sex")
    my.formula1 = paste0(feature_names[j],"~ endpoints + (1|sites)+age+sex")
    if(j == 1){
      tmp = mixed_pairwise(my.formula0, my.formula1, data_use)
      res_table_pairwise = data.frame(matrix(0, ncol = length(tmp$res_table), nrow = length(feature_names)))
      coef_endpoint_pairwise = data.frame(matrix(0, ncol= length(tmp$coef_endpoint), nrow=length(feature_names)))
      res_table_pairwise[j,] = tmp$res_table
      coef_endpoint_pairwise[j,] = tmp$coef_endpoint
      colnames(res_table_pairwise) = names(tmp$res_table)
      colnames(coef_endpoint_pairwise) = names(tmp$coef_endpoint)
    }else{
      tmp = mixed_pairwise(my.formula0, my.formula1, data_use)
      res_table_pairwise[j,] = tmp$res_table
      coef_endpoint_pairwise[j,] = tmp$coef_endpoint
    }
  }
  
  if (length(ordered_endpoint_var) > 2) {
      coef_endpoint_pairwise <- coef_endpoint_pairwise[,1:3]
      res_table_pairwise <- res_table_pairwise[,1:3]
  }
  coef_endpoint_pairwise$feature <- feature_names
  coef_endpoint_pairwise$visit <- cur_visit
  
  if (nrow(res_table_pairwise)>1) {
    res_table_pairwise_adj_df <- qvalue::qvalue(res_table_pairwise, fdr.level = 0.05, pi0=1)$qvalues
  } else {
    res_table_pairwise_adj_df <- res_table_pairwise
  }

  res_table_pairwise_adj_df$feature <- feature_names
  res_table_pairwise_adj_df$visit <- cur_visit
  
  plot_df <- data_use %>% dplyr::select(starts_with("SO"), 
                                        starts_with("PMG"),
                                        starts_with("PPT"),
                                        starts_with("PPG"),
                                        starts_with("PGX"),
                                        starts_with("BCT"),
                                        endpoints) %>%
            pivot_longer(!endpoints, names_to = "Feature", values_to = "value")
  plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))
  
  p <- violin_plot_visits(plot_df, colors = colors) + ggtitle(visit_id)
  
  res_list <- list(coefs_df = coef_endpoint_pairwise, 
                   res_df = res_table_pairwise_adj_df,
                   violin_plot = p)
  
  return(res_list)
}

```

## Functions for better pheatmap

```{r}
## Edit body of pheatmap:::draw_colnames, customizing it to your liking
draw_colnames_45 <- function (coln, ...) {
    m = length(coln)
    x = (1:m)/m - 1/2/m
    grid.text(coln, x = x, y = unit(0.96, "npc"), vjust = .5, 
        hjust = 1, rot = 45, gp = gpar(...)) ## Was 'hjust=0' and 'rot=270'
}

## For pheatmap_1.0.8 and later:
draw_colnames_45 <- function (coln, gaps, ...) {
    coord = pheatmap:::find_coordinates(length(coln), gaps)
    x = coord$coord - 0.5 * coord$size
    res = textGrob(coln, x = x, y = unit(1, "npc") - unit(3,"bigpts"), vjust = 0.5, hjust = 1, rot = 45, gp = gpar(...))
    return(res)}

## 'Overwrite' default draw_colnames with your own version 
assignInNamespace(x="draw_colnames", value="draw_colnames_45",
ns=asNamespace("pheatmap"))

p_val_only_palette <- colorRampPalette(brewer.pal(9,"Reds"))(10)

breaklist = seq(0,3,0.3)

pretty_metabolite <- function(analyte_name, metabolite_DF=metabolite_row_DF){
      if (analyte_name %in% metabolite_row_DF$CHEMICAL_NAME){
        pretty_name <- metabolite_row_DF[which(metabolite_row_DF$CHEMICAL_NAME==analyte_name),"PLOT_NAME"]
        return(pretty_name)
      } else {
        return(analyte_name)
      }
}
```

## Association MIN vs OTHER for each visit


We treat each factor expression as the covariate and consider its association with clinical endpoints 
(pro_groups) in the presence of potential heterogeneity across different clinical sites, 
which is modeled as a random effect.

We first examine if there is indeed a visible random effect from clinical sites, for both the intercepts and the slopes.
There may be some weak random effects for the intercept, but observe "nothing" regarding the slope. 
As a result, we consider only the random intercept effects.

We perform pairwise comparisons.


```{r fig.width=2, fig.height=2}
endpoint_coefficients_all <- data.frame()
res_table_pairwise_adj_all <- data.frame()

for (visit_id in ordered_visits){
  print(paste("Computing associations for visit:", visit_id))
  res_list <- pairwise_association_visit(clinical_subsets_all=clinical_subsets_all, 
                                          feature_data=top_feat_data_df, 
                                          visit_id=visit_id, 
                                          selected_features_names = colnames(top_feat_data_df),
                                          endpoint_var = "pro_2_groups")

  endpoint_coefficients_all <- bind_rows(endpoint_coefficients_all,
                                         res_list$coefs_df)
  res_table_pairwise_adj_all <- bind_rows(res_table_pairwise_adj_all, 
                                          res_list$res_df)
}


```

### Adj p-val all visits

```{r fig.height=2.5, fig.width=10}



res_table_pairwise_adj_all_plot <- res_table_pairwise_adj_all %>%
                                      pivot_wider(names_from = feature,
                                                  values_from = "MIN|OTHER") %>%
                                      column_to_rownames("visit")

print_kable(res_table_pairwise_adj_all_plot)
write.csv(res_table_pairwise_adj_all_plot, paste0(results_dir, "Adj_pval_all_visits_min_vs_other.csv"))
```


### Log p-value plot with coefficient sign

```{r fig.asp=0.5, fig.width=20}
endpoint_coefficient_plot <- endpoint_coefficients_all %>%
                              mutate(sign = if_else(`MIN|OTHER`>0, 1, -1)) %>%
                              dplyr::select(feature, visit, sign)

df_plot <- res_table_pairwise_adj_all %>%
              mutate(log10_adjpval = -log10(`MIN|OTHER`)) %>%
              inner_join(endpoint_coefficient_plot, by=c("feature", "visit")) %>%
              mutate(signed_adjp = log10_adjpval * sign)

df_plot$feature <- factor(df_plot$feature, levels = colnames(top_feat_data_df))
df_plot$visit <- factor(df_plot$visit, levels = unique(df_plot$visit))
df_plot <- df_plot %>% 
            mutate(annot =
                case_when(
                 abs(log10_adjpval) >= -log10(0.001) ~ "***",
                 abs(log10_adjpval) >= -log10(0.01) ~ "**",
                 abs(log10_adjpval) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 ))

thr <- 0.05
plot_title <- "Adj p-value feature association"


convert_df <- df_plot %>% 
              dplyr::select(feature, visit, signed_adjp) %>%
              pivot_wider(id_cols = c(visit), 
                          names_from = feature, 
                          values_from = signed_adjp) %>%
              column_to_rownames("visit")

convert_annot <- df_plot %>%
              dplyr::select(feature, visit, annot) %>%
              pivot_wider(id_cols = c(visit),
                          names_from = feature,
                          values_from = annot) %>%
              column_to_rownames("visit")

mat <- as.matrix(convert_df)
mat_annot <- as.matrix(convert_annot)

palette_length <- 50
abs_max <- max(abs(min(mat)), abs(max(mat)), 1.3)
breaks_list <- c(seq(-abs_max, 0, length.out=ceiling(palette_length/2) + 1), 
              seq(abs_max/palette_length, abs_max, length.out=floor(palette_length/2)))

pdf(file = paste0(results_dir, "Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 12)
p <- pheatmap::pheatmap(t(mat),
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = t(mat_annot),
         number_color = "black",
         fontsize_number = 10
         )
p
dev.off()

p
#save_pheatmap_pdf(p, paste0(results_dir,"Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 6)

```

```{r fig.asp=0.5, fig.width=20}
row_annot = data.frame(omics = str_split_i(rownames(t(mat)), "_", 1),
                       row.names = rownames(t(mat))
                       )
gp_row = which(diff(as.numeric(factor(row_annot$omics)))!=0)

pdf(file = paste0(results_dir, "Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 12)
p <- pheatmap::pheatmap(t(mat),
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = t(mat_annot),
         number_color = "black",
         annotation_row = row_annot, 
         annotation_names_row = F, 
         annotation_colors = list(omics= col_annot),
         gaps_row = gp_row,
         labels_row = str_split_i(rownames(t(mat)),"_",2),
         fontsize_number = 10
         )
p
dev.off()

p
#save_pheatmap_pdf(p, paste0(results_dir,"Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 6)

```

### Violin plots all visits

```{r }

features_filtered = top_feat_data_df[, colnames(top_feat_data_df), drop=F]
clinical_subsets = clinical_subsets_all[rownames(top_feat_data_df),]
  

data_use = features_filtered 

data_use$endpoints = as.factor(clinical_subsets$pro_2_groups)

# Add the enrollment_site as $sites
data_use$sites = clinical_subsets$enrollment_site
data_use$age =  clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex
data_use$visit = clinical_subsets$event_type

# Add the $control, needs to be different for each sample
# required for clmm (always requires some random effect)
data_use$control =  factor(1:nrow(data_use))

# Set module names:
feature_names = colnames(features_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$endpoints),]
```

Violin plots for all features, values from all visits

```{r fig.width=10, fig.height=15}

plot_df <- data_use %>% dplyr::select(starts_with("SO"), 
                                        starts_with("PMG"),
                                        starts_with("PPT"),
                                        starts_with("PPG"),
                                        starts_with("PGX"),
                                        starts_with("BCT"),
                                        endpoints) %>%
            pivot_longer(!endpoints, names_to = "Feature", values_to = "value")
plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))

p <- ggplot(plot_df, aes(x=endpoints, y=value, fill=endpoints)) +
  ggbeeswarm::geom_quasirandom(aes(fill=endpoints), shape=21, width=.3) +
  geom_boxplot(alpha=0.6, color="black", outlier.shape=NA) +
  #scale_y_continuous(labels = scaleFUN) +
  theme_bw() + theme(legend.position = "none") +
  scale_color_manual(values=pro_min_colors) +
  scale_fill_manual(values=pro_min_colors) +
  facet_wrap(~Feature, drop = T, scales = "free_y", ncol = 4)
p
```

Violin plots for all features, values per visit

```{r fig.height=2, fig.width=6}

plot_df <- data_use %>% dplyr::select(starts_with("SO"), 
                                        starts_with("PMG"),
                                        starts_with("PPT"),
                                        starts_with("PPG"),
                                        starts_with("PGX"),
                                        starts_with("BCT"),
                                        endpoints, visit) %>%
            pivot_longer(!c(endpoints,visit), names_to = "Feature", values_to = "value")
plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))
ordered_visits <- c("Visit 7", "Visit 8", "Visit 9", "Visit 10" )
plot_df$visit <- factor(plot_df$visit, levels=ordered_visits)

plot_list <- list()
for (feature in colnames(top_feat_data_df)) {
  plot_df_fact <- plot_df %>% filter(Feature == feature)
  
  p <- ggplot(plot_df_fact, aes(x=visit, y=value, fill=endpoints)) +
  ggbeeswarm::geom_quasirandom(aes(fill=endpoints), shape=21, width=.15, dodge.width = 0.75) +
  geom_boxplot(alpha=0.6, color="black", outlier.shape=NA) +
  #scale_y_continuous(labels = scaleFUN) +
  theme_bw() + 
  scale_color_manual(values=pro_min_colors, labels = pro_min_labels) +
  scale_fill_manual(values=pro_min_colors, labels = pro_min_labels) +
  #facet_wrap(~visit, drop = T, ncol = 4) +
  labs(x="", y=str_split_i(feature,"_",2), fill="")
  print(p)
  
  plot_list <- append(plot_list, list(p))
}
names(plot_list) <- colnames(top_feat_data_df)

```


```{r fig.height = 50, fig.width=6}
p_list <- ggarrange(plotlist=plot_list, ncol=1)
p_list

saveRDS(plot_list, file.path(figures_dir, "boxplots_top_spear_features.rds"))

```

```{r}
p_list <- ggarrange(plotlist=plot_list, ncol=1)
p_list

saveRDS(plot_list, file.path(figures_dir, "boxplots_top_spear_features.rds"))
```

## Association MIN vs all other groups

```{r fig.width=2, fig.height=2, warning=F}
endpoint_coefficients_all <- data.frame()
res_table_pairwise_adj_all <- data.frame()

for (visit_id in ordered_visits){
  tryCatch(
    {
      print(paste("Computing associations for visit:", visit_id))
      res_list <- pairwise_association_visit(clinical_subsets_all=clinical_subsets_all, 
                                              feature_data=top_feat_data_df, 
                                              visit_id=visit_id, 
                                              selected_features_names = colnames(top_feat_data_df),
                                              endpoint_var = "pro_group_labels", #TODO: update outcome variable here
                                              ordered_endpoint_var = names(pro_colors),
                                              colors = pro_colors
                                              )
    
      endpoint_coefficients_all <- bind_rows(endpoint_coefficients_all,
                                             res_list$coefs_df)
      res_table_pairwise_adj_all <- bind_rows(res_table_pairwise_adj_all, 
                                              res_list$res_df)
    },
    error = function(cond){
      message(paste("Not possible to calculate association for visit:",visit_id))
    }
  )
  
}


```

### Adj p-val all visits

```{r pairwise_heatmap, fig.height=2.5, fig.width=20}

res_table_pairwise_adj_all_plot <- res_table_pairwise_adj_all %>% 
                                    pivot_longer(cols=starts_with("MIN"),
                                                 values_to="padj",
                                                 names_to="comp") %>%
                                    mutate(comp_visit = paste0(comp, "_", visit)) %>%
                                    dplyr::select(-c(comp)) %>%
                                    pivot_wider(id_cols = c(comp_visit),
                                                names_from = feature,
                                                values_from = padj) %>%
                                    column_to_rownames("comp_visit")
print_kable(res_table_pairwise_adj_all_plot)
write.csv(res_table_pairwise_adj_all_plot, paste0(results_dir, "Adj_pval_all_visits_all_pro_groups.csv"))
```



### Log p-value plot with coefficient sign
```{r fig.asp=0.8, fig.width=20}
endpoint_coefficient_plot <- endpoint_coefficients_all %>%
                              pivot_longer(-c(feature,visit), 
                                           names_to = "comparison", 
                                           values_to = "coefficient") %>% 
                              mutate(sign = if_else(coefficient>0, 1, -1)) %>%
                              dplyr::select(feature, visit, comparison, sign)

df_plot <- res_table_pairwise_adj_all %>%
              pivot_longer(-c(feature,visit),
                           names_to = "comparison",
                           values_to = "adj_pval") %>%
              mutate(log10_adjpval = -log10(adj_pval)) %>%
              inner_join(endpoint_coefficient_plot, 
                         by=c("feature", "visit", "comparison")) %>%
              mutate(comp_visit = paste0(comparison, "_", visit)) %>%
              mutate(signed_adjp = log10_adjpval * sign)

df_plot$feature <- factor(df_plot$feature, levels = colnames(top_feat_data_df))
df_plot$comp_visit <- factor(df_plot$comp_visit, levels = unique(df_plot$comp_visit))
df_plot <- df_plot %>% 
            mutate(annot =
                case_when(
                 abs(log10_adjpval) >= -log10(0.001) ~ "***",
                 abs(log10_adjpval) >= -log10(0.01) ~ "**",
                 abs(log10_adjpval) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 ))

thr <- 0.05
plot_title <- "Adj p-value feature association"

convert_df <- df_plot %>% 
              dplyr::select(feature, comp_visit, signed_adjp) %>%
              pivot_wider(id_cols = c(comp_visit), 
                          names_from = feature, 
                          values_from = signed_adjp) %>%
              column_to_rownames("comp_visit")

convert_annot <- df_plot %>%
              dplyr::select(feature, comp_visit, annot) %>%
              pivot_wider(id_cols = c(comp_visit),
                          names_from = feature,
                          values_from = annot) %>%
              column_to_rownames("comp_visit")

mat <- as.matrix(convert_df)
mat_annot <- as.matrix(convert_annot)

palette_length <- 50
abs_max <- max(abs(min(mat)), abs(max(mat)), 1.3)
breaks_list <- c(seq(-abs_max, 0, length.out=ceiling(palette_length/2) + 1), 
              seq(abs_max/palette_length, abs_max, length.out=floor(palette_length/2)))

pdf(file = paste0(results_dir, "Adj_pval_all_visits_pro_groups_plot.pdf"), width = 12, height = 12)
p <- pheatmap::pheatmap(t(mat),
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = t(mat_annot),
         number_color = "black",
         fontsize_number = 10,
         )
p
dev.off()

p

#save_pheatmap_pdf(p, paste0(results_dir,"Adj_pval_all_visits_pro_groups_plot.pdf"), width = 12, height = 6)

```

```{r fig.asp=0.5, fig.width=20}
row_annot = data.frame(omics = str_split_i(rownames(t(mat)), "_", 1),
                       row.names = rownames(t(mat))
                       )
gp_row = which(diff(as.numeric(factor(row_annot$omics)))!=0)

pdf(file = paste0(results_dir, "Adj_pval_all_visits_pro_groups_plot.pdf"), width = 12, height = 12)
p <- pheatmap::pheatmap(t(mat),
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = t(mat_annot),
         number_color = "black",
         annotation_row = row_annot, 
         annotation_colors = list(omics= col_annot),
         annotation_names_row = F, 
         gaps_row = gp_row,
         labels_row = str_split_i(rownames(t(mat)),"_",2),
         fontsize_number = 10
         )
p
dev.off()

p
#save_pheatmap_pdf(p, paste0(results_dir,"Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 6)

```

### Violin plots all visits

```{r }

features_filtered = top_feat_data_df[, colnames(top_feat_data_df), drop=F]
clinical_subsets = clinical_subsets_all[rownames(top_feat_data_df),]
  

data_use = features_filtered 

data_use$endpoints = factor(clinical_subsets$pro_group_labels, levels = names(pro_colors))

# Add the enrollment_site as $sites
data_use$sites = clinical_subsets$enrollment_site
data_use$age =  clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex
data_use$visit = clinical_subsets$event_type

# Add the $control, needs to be different for each sample
# required for clmm (always requires some random effect)
data_use$control =  factor(1:nrow(data_use))

# Set module names:
feature_names = colnames(features_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$endpoints),]
```

Violin plots for all features, values from all visits

```{r fig.height=8, fig.width=15}

plot_df <- data_use %>% dplyr::select(starts_with("SO"), 
                                        starts_with("PMG"),
                                        starts_with("PPT"),
                                        starts_with("PPG"),
                                        starts_with("PGX"),
                                        starts_with("BCT"),
                                        endpoints) %>%
            pivot_longer(!endpoints, names_to = "Feature", values_to = "value")
plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))

p <- ggplot(plot_df, aes(x=endpoints, y=value, fill=endpoints)) +
  geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  scale_y_continuous(labels = scaleFUN) +
  theme_bw() + theme(legend.position = "none") +
  scale_color_manual(values=pro_colors) +
  scale_fill_manual(values=pro_colors) +
  facet_wrap(~Feature, drop = T, scales = "free_y", ncol = 4)
p
```

Violin plots for all features, values per visit

```{r fig.height=2, fig.width=6}

plot_df <- data_use %>% dplyr::select(starts_with("SO"), 
                                        starts_with("PMG"),
                                        starts_with("PPT"),
                                        starts_with("PPG"),
                                        starts_with("PGX"),
                                        starts_with("BCT"),
                                        endpoints, visit) %>%
            pivot_longer(!c(endpoints,visit), names_to = "Feature", values_to = "value")
plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))
ordered_visits <- c("Visit 7", "Visit 8", "Visit 9", "Visit 10" )
plot_df$visit <- factor(plot_df$visit, levels=ordered_visits)

for (feature in colnames(top_feat_data_df)) {
  plot_df_fact <- plot_df %>% filter(Feature == feature)
  
  p <- ggplot(plot_df_fact, aes(x=endpoints, y=value, fill=endpoints)) +
  geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  scale_y_continuous(labels = scaleFUN) +
  theme_bw() + theme(legend.position = "none") +
  scale_color_manual(values=pro_colors) +
  scale_fill_manual(values=pro_colors) +
  facet_wrap(~visit, drop = T, ncol = 4) +
  labs(title = feature, x="", y="value")
  print(p)
}

```

## Longitudinal analysis

With spear features

```{r data preparation, echo=TRUE}
clinical_subsets <- correlation_metadata
rownames(clinical_subsets) <- clinical_subsets$event_id

row_ids <-rownames(top_feat_data_df)
clinical_subsets <- clinical_subsets[row_ids,]


modules_filtered = top_feat_data_df %>% dplyr::select(all_of(colnames(top_feat_data_df)))

####%%%%%%%% spear Input %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

inputDF <- modules_filtered %>%
  rownames_to_column(var = "event_id") %>%
  merge(y = clinical_subsets,
        by = "event_id") %>%
  mutate(outcome = pro_2_groups) %>% #TODO: update outcome variable here
  pivot_longer(cols = -c("event_id", "event_date", "participant_id",
                        "enrollment_site", "pro_group_labels", "pro_2_groups", "event_type", "outcome",
                        "sex", "admit_age", "discretized_admit_age_quantile", "resp_status_v1",
                        "event_after_first_vaccine_dose", "partition", "event_date_3weeks_after_dose","race")) %>%
  dplyr::filter(!is.na(value) & !is.na(pro_2_groups))
inputDF$names <- factor(inputDF$name, levels=colnames(top_feat_data_df))


## Ensure proper variables are features
inputDF$pro_group_labels <- as.factor(inputDF$pro_group_labels)
inputDF$pro_2_groups <- as.factor(inputDF$pro_2_groups)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$race <- as.factor(inputDF$race)
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)
```


### PRO 2 groups

```{r, linear_model_loop, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T, 
                              model_type = "smoothSpline", 
                              spline_type = "ts", 
                              endpoint = "pro_2_groups") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])
write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_pro_2_groups.csv"))
```

```{r, longitudinal_model_loop_pheatmap, fig.width = 6, fig.height=10}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_min_vs_other.pdf"), width = 6, height = 10)

mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

row_annot = data.frame(omics = str_split_i(rownames(mat_log10pval), "_", 1),
                       row.names = rownames(mat_log10pval)
                       )
#row_annot_col = c()
#names(row_annot_col) <-  

gp_row = which(diff(as.numeric(factor(row_annot$omics)))!=0)

p1 <- pheatmap::pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "white",
         cluster_rows = F, 
         cluster_cols = F,
         annotation_row = row_annot, 
         annotation_colors = list(omics= col_annot),
         annotation_names_row = F, 
         gaps_row = gp_row,
         labels_row = str_split_i(rownames(mat_log10pval),"_",2),
         color = p_val_only_palette,
         breaks = breaklist)
p1
dev.off()
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_min_vs_other.pdf"), width = 3, height = 6)
p1


df_adjp_minother <- mat_log10pval %>% dplyr::select(adjp.intercept) %>% dplyr::rename(`MIN vs LC`=adjp.intercept)

df_adjp_minother_annot <- mat_annot %>% dplyr::select(adjp.intercept) %>% dplyr::rename(`PASC vs MIN`=adjp.intercept)
```

### Variability of features over time

```{r, fig.width=10, fig.height=50}
plots <- list()
plots_model <- list()
for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline",
                      colors = pro_min_colors,
                     spline_type = "ts",
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     CI = T,
                     endpoint = "pro_2_groups") #TODO: update outcome variable here
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_2_groups", #TODO: update outcome variable here
                   model_type = "smoothSpline",
                   colors = pro_min_colors,
                   spline_type = "ts",
                   knot_lines = F,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER.pdf"), width = 18, height = 45)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER.png"), width = 18, height = 50, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model.pdf"), width = 18, height = 45)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model.png"), width = 18, height = 50, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)

```

### All PRO groups

```{r, linear_model_loop_PRO_groups, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T,
                              model_type = "smoothSpline",
                              spline_type = "ts",
                              endpoint = "pro_group_labels") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, 
                               adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])

write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_all_pro_groups.csv"))
```

```{r linear_model_loop_pheatmap_allpro, fig.width = 6, fig.height=10}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_pro_groups.pdf"), width = 6, height = 10)
mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

row_annot = data.frame(omics = str_split_i(rownames(mat_log10pval), "_", 1),
                       row.names = rownames(mat_log10pval)
                       )
gp_row = which(diff(as.numeric(factor(row_annot$omics)))!=0)

p1 <- pheatmap::pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "white",
         cluster_rows = F, 
         cluster_cols = F,
         annotation_row = row_annot, 
         annotation_colors = list(omics= col_annot),
         annotation_names_row = F, 
         gaps_row = gp_row,
         labels_row = str_split_i(rownames(mat_log10pval),"_",2),
         color = p_val_only_palette,
         breaks = breaklist)
p1
dev.off()
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_min_vs_other.pdf"), width = 3, height = 6)
p1

df_adjp_pro <- mat_log10pval %>% dplyr::select(adjp.intercept) %>% dplyr::rename(`PRO group`=adjp.intercept)

df_adjp_pro_annot <- mat_annot %>% dplyr::select(adjp.intercept) %>% dplyr::rename(`PRO group`=adjp.intercept)
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_pro_groups.pdf"), width = 3, height = 6)

```

### Plotting features over the visits

```{r, fig.width=10, fig.height=50}
plots <- list()
plots_model <- list()

for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  plotDF$pro_group_labels <- factor(plotDF$pro_group_labels, levels = names(pro_colors))
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline", 
                     spline_type = "ts",
                     colors = pro_colors,
                     CI = T,
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     facet = T,
                     endpoint = "pro_group_labels")
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_group_labels",
                   model_type = "smoothSpline",
                   spline_type = "ts",
                   knot_lines = F,
                   colors = pro_colors,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right")
  plots_model <- c(plots_model, list(p2))
}

if(params$feature == "spear"){
  pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS.pdf"), width = 15, height = 50)
  gridExtra::grid.arrange(grobs = plots, ncol=1)
  dev.off()
  
  png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS.png"), width = 20, height = 60, units="cm", res = 300)
  gridExtra::grid.arrange(grobs = plots, ncol=1)
  dev.off()
  
  pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model.pdf"), width = 15, height = 50)
  gridExtra::grid.arrange(grobs = plots_model, ncol=1)
  dev.off()
  
  png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model.png"), width = 20, height = 60, units="cm", res = 300)
  gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

}

gridExtra::grid.arrange(grobs = plots, ncol=1)

```

## Analyte scores

```{r fig.width = 2, fig.height=7}
pdf(file = paste0(results_dir, "analyte_scores_longitudinal_min_pasc_and_pro_groups.pdf"), width = 6, height = 10)

top_analyte_scores_translated$Assay <- factor(top_analyte_scores_translated$Assay, levels= c("SO", "PMG", "PPG", "PGX"))
top_analyte_scores_mat <- top_analyte_scores_translated %>%
  dplyr::mutate(coefficient = projection.coefficient,
                analyte_name = paste0(Assay,"_",translated_name)) %>%
  dplyr::arrange(Assay, desc(abs(coefficient))) %>%
  dplyr::select(analyte_name, coefficient) %>%
  tibble::column_to_rownames("analyte_name")

top_analyte_scores_annot <- top_analyte_scores_translated %>%
  dplyr::mutate(annot = "",
                analyte_name = paste0(Assay,"_",translated_name)) %>%
  dplyr::select(analyte_name, annot) %>%
  tibble::column_to_rownames("analyte_name")

mat_log10pval <- bind_cols(df_adjp_minother, df_adjp_pro)
mat_annot <- bind_cols(df_adjp_minother_annot, df_adjp_pro_annot)
  
row_annot = data.frame(omics = str_split_i(rownames(top_analyte_scores_mat), "_", 1),
                       row.names = rownames(top_analyte_scores_mat)
                       )

row_annot$omics <- factor(row_annot$omics, levels = c("SO", "PMG", "PPG", "PGX"))


gp_row = which(diff(as.numeric(factor(row_annot$omics)))!=0)

p1 <- ComplexHeatmap::pheatmap(as.matrix(top_analyte_scores_mat),
         cellwidth = 10,
         cellheight = 10,
         fontsize = 8,
         cluster_rows = F, 
         cluster_cols = F,
         annotation_row = row_annot, 
         annotation_colors = list(omics= col_annot),
         annotation_names_row = F, 
         show_rownames = F,
         gaps_row = gp_row,
         annotation_legend = T,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
        heatmap_legend_param = list(
        title = "coefficient"
        ))
p1
dev.off()

spear_analyte_scores <- p1
spear_analyte_scores
saveRDS(spear_analyte_scores, file = file.path(figures_dir, "spear_analyte_scores.rds"))
```

```{r fig.width = 5, fig.height=7}

labels_row <- str_split_i(rownames(top_analyte_scores_mat),"_",2)
labels_row_pretty_metabolite <- sapply(labels_row, pretty_metabolite)

spear_analyte_scores_nolegend <- ComplexHeatmap::pheatmap(as.matrix(top_analyte_scores_mat),
         cellwidth = 10,
         cellheight = 10,
         fontsize = 8,
         cluster_rows = F, 
         cluster_cols = F,
         angle_col = "45",
         annotation_row = row_annot, 
         annotation_colors = list(omics= col_annot),
         annotation_names_row = F, 
         show_rownames = T,
         gaps_row = gp_row,
         labels_row = labels_row_pretty_metabolite,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         legend = T,
         annotation_legend = F,
        heatmap_legend_param = list(
        title = "coefficient"
        ))
saveRDS(spear_analyte_scores_nolegend, file = file.path(figures_dir, "spear_analyte_scores_nolegend_withrownames.rds"))
spear_analyte_scores_nolegend

```


## Analyte pval

```{r fig.width = 5, fig.height=7}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_min_pasc_and_pro_groups.pdf"), width = 6, height = 10)

#mat_log10pval <- bind_cols(df_adjp_minother, df_adjp_pro)
#mat_annot <- bind_cols(df_adjp_minother_annot, df_adjp_pro_annot)

mat_log10pval <- df_adjp_minother
mat_annot <- df_adjp_minother_annot
colnames(mat_log10pval) <- "adj. p-val"

# Order according to coefficient
mat_log10pval <- mat_log10pval[rownames(top_analyte_scores_mat),,drop=F]
mat_annot <- mat_annot[rownames(top_analyte_scores_mat),,drop=F]

row_annot = data.frame(omics = str_split_i(rownames(mat_log10pval), "_", 1),
                       row.names = rownames(mat_log10pval)
                       )
gp_row = which(diff(as.numeric(factor(row_annot$omics)))!=0)

labels_row <- str_split_i(rownames(mat_log10pval),"_",2)
labels_row_pretty_metabolite <- sapply(labels_row, pretty_metabolite)

p1 <- ComplexHeatmap::pheatmap(as.matrix(mat_log10pval),
         cellwidth = 10,
         cellheight = 10,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 8,
         fontsize = 8,
         number_color = "white",
         cluster_rows = F, 
         cluster_cols = F,
         angle_col = "45",
         #annotation_row = row_annot, 
         #annotation_colors = list(omics= col_annot),
         #annotation_names_row = F, 
         gaps_row = gp_row,
         labels_row = labels_row_pretty_metabolite,
         color = p_val_only_palette,
         breaks = breaklist,
        heatmap_legend_param = list(
        title = "-log10 adj. p"
        ))
p1
dev.off()

spear_analyte_pval <- p1
saveRDS(spear_analyte_pval, file.path(figures_dir, "spear_analyte_pval.rds"))


p1 
```

```{r fig.width = 5, fig.height=5}
combined_heatmap <- spear_analyte_scores_nolegend + spear_analyte_pval
saveRDS(combined_heatmap, file = file.path(figures_dir, "spear_analyte_scores_pvalue.rds"))

combined_heatmap
```

## Boxplots of significant analytes

```{r fig.width = 8, fig.height=8}
significant_analytes <- as.data.frame(mat_log10pval) %>% filter(`adj. p-val` >= -log10(0.05))
significant_analytes

boxplot_significant_list <- plot_list[rownames(significant_analytes)]

sig_boxplot <- ggarrange(plotlist=boxplot_significant_list, ncol =2, nrow=4)
sig_boxplot
ggsave(filename = file.path(figures_dir, "significant_analytes_boxplots.pdf"), plot=sig_boxplot, width = 8, height = 8)

saveRDS(sig_boxplot, file.path(figures_dir, "spear_analyte_significant_boxplots.rds"))
```

```{r fig.width = 8, fig.height=20}

boxplot_all_26_analytes <- plot_list

all_26_analytes_boxplot <- ggarrange(plotlist=boxplot_all_26_analytes, ncol =2, nrow=13)
all_26_analytes_boxplot
ggsave(filename = file.path(figures_dir, "all_26_analytes_boxplots.pdf"), plot=all_26_analytes_boxplot, width = 8, height = 20)
```

```{r}
sessionInfo()
```
