 ---
title: "Geometric mean plots"
author: "Jessica Maciuch"
date: "`r format(Sys.time(), '%b %d %Y')`"
output: html_document
---

Note: This script contains code to generate all geometric means plots, including versions that were not used (i.e. box plots with per-visit pvals, longitudinal plots, etc.) 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory
  workDir <- "~/impacc-full-cohort/Analysis/convalescent-project/Spear/clinical_associations"
  setwd(workDir) # For troubleshooting in the command line
  knitr::opts_knit$set(root.dir = workDir)

suppressPackageStartupMessages({
  library(tidyverse)
  library(magrittr)
  library(ggpubr)
  library(rstatix)
  library(patchwork)
  library(pheatmap)})

# Load codebase functions
  source("~/impacc-full-cohort/Analysis/IMPACC_data_analysis_codebase.R")

# Initialize list of variables to exclude from downstream environment cleaning
  keep_vars <- ls()
  keep_vars %<>% c("keep_vars")
```

# Calculate geometric means

If you have access to the geometric_means_clinical.csv file, skip to "Load geometric means" section.

## Dictionary for gene and metabolite names

```{r}

# Grab RowFeature tables for PMG and PGX
  PMG_dict <- read_csv("/data/metabolomics/plasma-metabolomics-global/current/GlobalPlasmaMetabolomics-RowFeature.csv")
  PMG_dict %<>% select(MET_ID, CHEMICAL_NAME)
  colnames(PMG_dict) <- c("Analyte", "Feature")
  
  PGX_dict <- read_csv("/data/pbmc-transcriptomics/current/pbmc-transcriptomics-RowFeature.csv", skip = 1)
  PGX_dict %<>% dplyr::select(gene_id, gene_name)
  colnames(PGX_dict) <- c("Analyte", "Feature")
  
  dict <- bind_rows(PMG_dict, PGX_dict)
  
# Clean environment
  keep_vars %<>% c("dict")
  rm(list = setdiff(ls(), keep_vars))
  gc()
```

## Load gene sets

```{r}

#### HEME METABOLISM AND ANDROGENIC STEROIDS ####
  gsea_files <- c("GSEA_subpath_Factor1_PMG_results.tsv",
                  "GSEA_hallmark_Factor1_PGX_results.tsv")
  gsea_files %<>% paste("~/impacc-full-cohort/Analysis/convalescent-project/Spear/enrichment_spear_factors_promis_physical_Factor1", ., sep = "/")
  
  results <- gsea_files |> map(\(f) read_delim(f, delim = "\t"))
  
  results %<>% bind_rows()
  results %<>% filter(ID %in% c("HALLMARK_HEME_METABOLISM",
                                "Androgenic Steroids"))
  
  # Separate individual features from their formatting in the table
  leadingEdge <- results$core_enrichment
  names(leadingEdge) <- results$ID
  
  leadingEdge %<>% map(\(l) str_split_i(l, pattern = "\t", i = -1))
  leadingEdge %<>% bind_rows()
  leadingEdge %<>% pivot_longer(cols = 1:last_col(),
                                names_to = "GeneSet",
                                values_to = "Feature")
  leadingEdge %<>% separate_longer_delim(cols = Feature,
                                         delim = "/")
  
  # Get assay names for features
  leadingEdge %<>% left_join(x = .,
                             y = dict,
                             by = "Feature")
  
  # Rename heme metabolism
  leadingEdge %<>% mutate(GeneSet = case_match(GeneSet,
                                               "HALLMARK_HEME_METABOLISM" ~ "Hallmark Heme Metabolism",
                                               .default = GeneSet))

#### SIGNIFICANT SPEAR ANALYTES ####
  spear <- readRDS("/scratch/data-fullcohort/spear_model/results/020624/032124_spearphysical_finalscores_allsamples.rds")
  
  top_analytes <- spear$analyte.scores
  top_analytes$joint.probability %<>% round(digits = 2)
  top_analytes %<>% filter(joint.probability >= 0.95)
  
  # Format same as leading edge data frames
  top_analytes %<>% mutate(GeneSet = "Significant SPEAR Analytes")
  top_analytes %<>% select(GeneSet, Analyte)
  
  # Get assay names for features
  top_analytes %<>% left_join(x = .,
                              y = dict,
                              by = "Analyte")
  
# Combine all gene sets into one data frame
  gene_sets <- rbind(leadingEdge, top_analytes)
  
  # If no readable feature name in dict, use original analyte name
  gene_sets$Feature %<>% if_else(is.na(.), gene_sets$Analyte, .)
  
  # Pull direction of each analyte in spear model
    gene_sets %<>% left_join(x = .,
                             y = spear$analyte.scores %>% select(Analyte, projection.coefficient),
                             by = "Analyte")
    
    gene_sets %<>% mutate(direction = sign(projection.coefficient))
  
#### COMBINED (i.e. Top Spear + Heme + steroids) ####
  Combined <- gene_sets %>% mutate(GeneSet = "Combined (Hallmark Heme Metabolism + Androgenic Steroids + Significant SPEAR Analytes)") %>% distinct()
  gene_sets %<>% rbind(Combined)
  
# Clean environment
  keep_vars %<>% c("spear", "gene_sets")
  rm(list = setdiff(ls(), keep_vars))
  gc()
```

## Load assay data for gene set features

```{r}

# Load pre-processed assay data
  acute_filepaths <- list("Train" = "/scratch/data-fullcohort/spear/acute_fullcohort_preprocessed_data_20240812/MOFA_acute_fullcohort_train_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240812.rds",
                          "Test" = "/scratch/data-fullcohort/spear/acute_fullcohort_preprocessed_data_20240812/MOFA_acute_fullcohort_test_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240812.rds")

  conv_filepaths <- list("Train" = "/scratch/data-fullcohort/MOFA_factors/MOFA_train_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240212.rds",
                                 "Test" = "/scratch/data-fullcohort/MOFA_factors/MOFA_test_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240212.rds")
  
  data <- list("Acute" = acute_filepaths,
               "Convalescent" = conv_filepaths) |> map_depth(.depth = 2, \(x) readRDS(x))
  
# Helper function to pull assay data from MOFA object
  pull_assay_data <- function(x, gene_sets) {
    data <- x@data |> lapply(\(x) x <- x[["group1"]])
    data %<>% purrr::reduce(rbind)
    data %<>% .[c(rownames(.) %in% gene_sets$Analyte),]
    data %<>% t() %>% cbind(x@samples_metadata %>% select(event_id))
    return(data)
  }
  
# Combine data from timepoints
  data %<>% map_depth(.depth = 2, \(x) pull_assay_data(x, gene_sets))
  data %<>% map(\(x) x %<>% bind_rows())
  data %<>% bind_rows(.id = "cohort")
  
# Convert to long form for geom mean function
  data %<>% pivot_longer(cols = which(colnames(data) %in% gene_sets$Analyte), names_to = "Analyte")
  
# Clean environment
  keep_vars %<>% c("data")
  rm(list = setdiff(ls(), keep_vars))
  gc()
```

## Calculate geometric mean

```{r}

# Function to calculate geometric mean
  geom_mean <- function(df, gene_set) {
    # Take only the features in the gene set
    df %<>% inner_join(x = .,
                       y = gene_set,
                       by = "Analyte")
    
    df %<>% 
      # Calculate separate mean for features in each direction for each event ID
      group_by(event_id, direction) %>% 
      summarise(mean = mean(value, na.rm = T)) %>%
      # Create separate columns for pos and neg mean
      pivot_wider(names_from = direction, values_from = mean) %>%
      # If all values are pos or neg, create all 0 column for other direction
      mutate("1" = if (!"1" %in% names(.)) {0} else {.data[["1"]]},
             "-1" = if (!"-1" %in% names(.)) {0} else {.data[["-1"]]}) 
    
    # Subtract neg from pos means
    df %<>% mutate(geom_mean = .data[["1"]] - .data[["-1"]]) 
    # If all directions were neg, flip sign again
    if (all(df[["1"]] == 0)) {df$geom_mean <- df$geom_mean * -1} 
    
    # Add pathway name to res
    res <- df %>% 
      select(event_id, geom_mean) %>%
      cbind(., pathway = unique(gene_set$GeneSet))
    return(res)
  }

# Separate data frame by gene set
  # Order gene sets for plotting
  order <- c("Hallmark Heme Metabolism",
             "Androgenic Steroids",
             "Significant SPEAR Analytes",
             "Combined (Hallmark Heme Metabolism + Androgenic Steroids + Significant SPEAR Analytes)")
  
  gene_sets$GeneSet %<>% factor(., levels = order)
  gene_sets %<>% split(., f = .[["GeneSet"]])
  
# Generate scores for each gene set
  scores <- gene_sets |> map(\(g) geom_mean(data, g))
  scores %<>% reduce(rbind)
  
# Save geometric means
  write_csv(scores, file = "./geometric_means.csv")
  
# Clean environment
  keep_vars %<>% c("scores")
  rm(list = setdiff(ls(), keep_vars))
  gc()
```

## Combine with clinical data

```{r}

# Load data environments for acute and convalescent
  data_env_conv <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_convalescent_data_env_all_Feb_11_2024.RDS")
  data_env_acute <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_allcohort_all_Aug_12_2024.RDS") 
  
# Pull clinical data
  clinical_data_conv <- data_env_conv$clinical_data
  clinical_data_acute <- data_env_acute$clinical_data
  clinical_data <- dplyr::bind_rows(list("convalescent" = clinical_data_conv, 
                                         "acute" = clinical_data_acute), 
                                    .id = "cohort")
  
# Select relevant columns
  clinical_data %<>% 
    select(all_of(c("participant_id", 
                    "event_id", 
                    "event_type",
                    "event_date",
                    "enrollment_site", 
                    "discretized_admit_age_quantile",
                    "sex",
                    "trajectory_group",
                    "pro_2_groups",
                    "cohort"))) %>% 
    distinct()
  
# Combine with scores
  scores %<>% left_join(x = .,
                      y = clinical_data,
                      by = "event_id")
  
# Update partition variable to reflect "extended test"
  test_eventIDs <- c(rownames(spear$Test),
                     rownames(spear$TestNoPhysical),
                     rownames(spear$TrainNoPhysical_NoOverlap))
  scores$partition <- if_else(scores$event_id %in% test_eventIDs, "test", "train")
  
  # If event in acute cohort, no partition
  scores$partition <- if_else(scores$cohort == "acute", "acute", scores$partition)
  
# Remove outlier/accidental acute data points after event date 50
  scores %<>% filter(!(cohort == "acute" & event_date > 50))
  
# Make sure categorical variables are plotted in consistent order
  scores$pro_2_groups <- if_else(scores$pro_2_groups == "MIN", "MIN", "LC")
  scores$pro_2_groups %<>% factor(., levels = c("MIN", "LC"))
  scores$sex %<>% factor(., levels = c("Male", "Female"))
  
# Format variables for stat functions
  scores %<>% dplyr::rename(value = geom_mean,
                            endpoints = pro_2_groups,
                            name = pathway)
  
# Save geometric means with clinical data
  write_csv(scores, file = "./geometric_means_clinical.csv")
```

# Load geometric means

```{r}

# Load csv
  scores <- read_csv("./geometric_means_clinical.csv")

# Format factors for proper plotting
  scores$endpoints %<>% factor(., levels = c("MIN", "LC"))
  scores$sex %<>% factor(., levels = c("Female", "Male"))
  scores$event_type %<>% factor(., levels = paste("Visit", 1:10))
  
  scores %<>% mutate(name = case_match(name, 
                                       "Combined (Hallmark Heme Metabolism + Androgenic Steroids + Significant SPEAR Analytes)" ~ "Combined (Hallmark Heme Metabolism +\nAndrogenic Steroids + Significant SPEAR Analytes)",
                                       .default = name))
  
  scores$name %<>% factor(., levels = c("Hallmark Heme Metabolism",
                                        "Androgenic Steroids",
                                        "Significant SPEAR Analytes",
                                        "Combined (Hallmark Heme Metabolism +\nAndrogenic Steroids + Significant SPEAR Analytes)"))
```

# Set color palette

```{r}

# Set color palette 
pro_2_groups_cols <- c("MIN" = "#649B20",
                       "LC" = "#7A186D")
```

# Per-visit box plots (global p-val)

This section generates plots of per-visit box plots with one global p-value correcting for age and visit # (plus random effects)

## Set filename vars

```{r}
# Set variables for filenames
  comparison <- "MIN_LC_geom_mean"
  plot_type <- "pervisit_globalpval"
```

## Plotting functions

```{r}

# Helper function to calculate pairwise p-val across all visits
  global_mixed_pairwise <- function(df,
                                    fixed_vars,
                                    predictor,
                                    random) {
    
    # Generate formulas for mixed_pairwise function
    var_list <- c(fixed_vars, random)
    
    terms <- paste(var_list, collapse = "+")
    formula.0 <- formula(paste("value~", terms, sep = ""))
    
    var_list %<>% c(predictor)
    terms <- paste(var_list, collapse = "+")
    formula.1 <- formula(paste("value~", terms, sep = ""))
    
    res_table <- mixed_pairwise(formula.0, formula.1, df)
    
    return(res_table$res_table)
  }

# Helper function to generate per-visit boxplots
  plot_geom_mean_pervisit <- function(df, pval_df, cols) {
    ggplot(df, aes(x = event_type, y = value)) +
      ggbeeswarm::geom_quasirandom(aes(fill = endpoints),
                                   shape = 21,
                                   width = .15, 
                                   dodge.width = 0.75) +
      geom_boxplot(aes(fill = endpoints),
                   alpha = 0.6, 
                   color = "black", 
                   outlier.shape = NA) +
      facet_wrap(~name, scales = "free_y") +
      theme_bw() + 
      scale_color_manual(values = cols) +
      scale_fill_manual(values = cols) +
      labs(x = NULL, 
           y = "Geometric Mean", 
           fill = "") +
      scale_y_continuous(expand = expansion(mult = c(0.1, 0.15))) + 
      geom_label(data = pval_df, 
                 inherit.aes = F, 
                 aes(label = label,
                     x = Inf,
                     y = Inf,
                     hjust = 1,
                     vjust = 1),
                 label.r = unit(0, "pt"),
                 size = 4) +
      theme(strip.background = element_blank(),
            text = element_text(size = 14)) -> plot

    return(plot)
  }
  
# Helper function to generate per visit box plots from starting df
  generate_geom_mean_pervisit_plots <- function(df) {
     df_list <- df %>% split(f = .[["name"]])
  
    # Calculate p-vals
    pval_list <- df_list |> 
      map(\(x) global_mixed_pairwise(df = x,
                                     fixed_vars = c("discretized_admit_age_quantile", "event_type", "sex"),
                                     predictor = "endpoints",
                                     random = "(1|enrollment_site/participant_id)")) |>
      as.numeric()
    
    # P-value adjustment
    adj_pval_list <- qvalue::qvalue(pval_list, fdr.level = 0.05, pi0 = 1)$qvalues
    
    # Create p-val df
    pval_df <- data.frame(name = names(df_list),
                          adjp = adj_pval_list)
    
    # Create p-val label
    pval_df %<>% mutate(label = paste("adj. p=", signif(adjp, digits = 2), sep = ""))
    
    # Order p-val df
    pval_df$name %<>% factor(., levels = levels(df$name))
    
    # Generate plot
    plot <- plot_geom_mean_pervisit(df, pval_df, pro_2_groups_cols)
    
    return(plot)
  }

```

## Convalescent plots

```{r}

# Pull only convalescent test set data
  df <- scores %>% filter(partition == "test")

# Generate plot
  plot <- generate_geom_mean_pervisit_plots(df)

# Save plots
  timepoint <- "convalescent"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)
```

## Acute plots

```{r}
# Pull only acute data
  df <- scores %>% filter(partition == "acute")

# Remove participants with no MIN/LC label
  df %<>% filter(!is.na(endpoints))

# Generate plot
  plot <- generate_geom_mean_pervisit_plots(df)

# Save plots
  timepoint <- "acute"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)
```

# Per-visit box plots by sex (global p-val)

This section generates plots of per-visit box plots with one global p-value correcting for age, sex, and visit # (plus random effects), separated by sex.

## Set filename vars

```{r}
# Set variables for filenames
  comparison <- "MIN_LC_geom_mean_bysex"
  plot_type <- "pervisit_globalpval"
```

## Plotting functions

```{r}

# Helper function to generate per-visit boxplots
  plot_geom_mean_pervisit_bysex <- function(df, cols) {
    ggplot(df, aes(x = event_type, y = value)) +
      ggbeeswarm::geom_quasirandom(aes(fill = endpoints),
                                   shape = 21,
                                   width = .15, 
                                   dodge.width = 0.75) +
      geom_boxplot(aes(fill = endpoints),
                   alpha = 0.6, 
                   color = "black", 
                   outlier.shape = NA) +
      facet_wrap(~sex) +
      theme_bw() + 
      scale_color_manual(values = cols) +
      scale_fill_manual(values = cols) +
      labs(x = NULL, 
           y = "Geometric Mean", 
           fill = "",
           title = unique(df$name)) +
      scale_y_continuous(expand = expansion(mult = c(0.1))) + 
      theme(strip.background = element_blank(),
            strip.text = element_text(size = 12),
            plot.title = element_text(hjust = 0.5, size = rel(1)),
            text = element_text(size = 14)) -> plot
    
    # Add p-values
    pval_df <- df %>% 
      select(name, sex, pval_label) %>% 
      distinct()
    
    plot <- plot + geom_label(data = pval_df,
                              inherit.aes = F,
                              aes(label = pval_label,
                                  x = Inf, 
                                  y = Inf, 
                                  hjust = 1,
                                  vjust = 1),
                              size = 4,
                              label.r = unit(0, "pt"))

    return(plot)
  }
  
# Helper function to generate per visit box plots by sex from starting df
  generate_geom_mean_pervisit_bysex_plots <- function(df) {
    # Split by pathway
    df_list <- df %>% split(f = .[["name"]])
     
    # Split by sex
    df_list %<>% map(\(df) df %<>% split(f = .[["sex"]]))
    
    # Calculate p-vals
    pval_list <- df_list |> map_depth(.depth = 2, \(df) global_mixed_pairwise(df = df,
                                                                              fixed_vars = c("discretized_admit_age_quantile", "event_type"),
                                                                              predictor = "endpoints",
                                                                              random = "(1|enrollment_site/participant_id)"))
  
    # Format pvals
    pvals <- pval_list %>% map(\(x) x %<>% bind_rows(.id = "sex"))
    pvals %<>% bind_rows(.id = "name")
      
    # P-value correction
    pvals$padj <- qvalue::qvalue(pvals$`MIN|LC`, fdr.level = 0.05, pi0 = 1)$qvalue
    
    # Create label for plots
    pvals$pval_label <- paste("adj. p=", 
                              signif(pvals$padj, digits = 2), 
                              sep = "")
      
    # Add p-vals to original df
    df %<>% left_join(x = .,
                      y = pvals,
                      by = c("name", "sex"))
    df$name %<>% factor(., levels = unique(pvals$name))
    
    # Split by pathway
    df_list <- df %>% split(f = .[["name"]])
      
    # Generate plots
    plot_list <- df_list |> 
        map(\(df) plot_geom_mean_pervisit_bysex(df, pro_2_groups_cols))
  
    # Combine plots
    plot <- wrap_plots(plot_list, ncol = 2, byrow = T)
  
    return(plot)
  }
```

## Convalescent plots by sex

```{r}

# Pull only convalescent test set data
  df <- scores %>% filter(partition == "test")

# Generate plots
  plot <- generate_geom_mean_pervisit_bysex_plots(df)

# Save plots
  timepoint <- "convalescent"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)
```

## Acute plots by sex

```{r}

# Pull only acute data
  df <- scores %>% filter(partition == "acute")

# Filter data points with MIN/LC label
  df %<>% filter(!is.na(endpoints))

# Generate plots
  plot <- generate_geom_mean_pervisit_bysex_plots(df)

# Save plots
  timepoint <- "acute"
  filename <- paste(workDir, "/plot_objects/additional_plots/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)
```

# MIN/LC longitudinal plots 

This section generates longitudinal plots of geometric means.

## Set filename vars

```{r}

# Set variables for filenames
  comparison <- "MIN_LC_geom_mean"
  plot_type <- "longitudinal"
```

## Plotting functions

```{r}

# Helper function to create longitudinal plot
  plot_geom_mean_MIN_LC_longitudinal <- function(df, cols) {
    # Generate model
    form <- as.formula("value ~ endpoints + discretized_admit_age_quantile + sex + s(event_date, bs = 'ts') + s(event_date, by = endpoints, bs = 'cr')")
    model <- gamm4::gamm4(form, random = ~(1|enrollment_site/participant_id), data = df)
    
    # Calculate confidence interval
    pred <- ggpredict(model, c("event_date", "endpoints"), type = "fixed",
                      ci.lvl = 0.95)
    
    # Generate plot
    plot <- ggplot(df, aes(x = event_date, y = value)) +
      geom_point(aes(color = endpoints),
                 size = 0.3) +
      geom_label(data = df %>% select(name, pval_label) %>% distinct(),
                 inherit.aes = F, 
                 aes(label = pval_label,
                     x = Inf, 
                     y = Inf, 
                     hjust = 1,
                     vjust = 1),
                 size = 4,
                 label.r = unit(0, "pt")) +
      geom_ribbon(inherit.aes = F, 
                  data = pred, 
                  aes(x = x, ymin = pred$conf.low, ymax = pred$conf.high, fill = group),
                  alpha = 0.3) +
      geom_line(inherit.aes = F, 
                data = pred, 
                aes(x = x, y = predicted, color = group),
                linewidth = 1.5) +
      scale_color_manual(values = cols) +
      scale_fill_manual(values = cols) +
      labs(title = unique(df$name),
           color = "",
           fill = "",
           x = "Days from Admission", 
           y = "Geometric Mean") +
      theme_bw() +
      theme(plot.title = element_text(hjust = 0.5))
    
    return(plot)
  }

# Helper function to generate longitudinal geom mean plots
 generate_geom_mean_MIN_LC_longitudinal_plots <- function(df) {
    # Split by pathway
    df_list <- df %>% split(f = .[["name"]])
    
    # Calcualte p-vals
    pval_list <- df_list |> map(\(df) model_loop(df, 
                                                 age_sex = T,
                                                 model_type = "smoothSpline",
                                                 spline_type = "ts",
                                                 endpoint ="trajectory_group"))
    pval_df <- bind_rows(pval_list, .id = "name")
    
    # P-value correction
    pval_df$padj <- qvalue::qvalue(pval_df$p.intercept, fdr.level = 0.05, pi0 = 1)$qvalues
    
    # Create p-value label
    pval_df %<>% select(name, padj)
    pval_df %<>% mutate(pval_label = paste("adj. p=", signif(padj, digits = 2), sep = ""))
     
    # Add p-vals to original df
    df %<>% left_join(x = .,
                      y = pval_df,
                      by = c("name"))
    df$name %<>% factor(., levels = unique(pval_df$name))
    
    # Split by pathway
    df_list <- df %>% split(f = .[["name"]])
    
    # Generate plots
    plot_list <- df_list |> 
        map(\(df) plot_geom_mean_MIN_LC_longitudinal(df, pro_2_groups_cols))
  
    # Combine plots
    plot <- wrap_plots(plot_list, ncol = 2, byrow = T)
  
    return(plot)
  }

```

## Acute plots

```{r}

# Pull only acute data
  df <- scores %>% filter(partition == "acute")

# Drop observations without PRO group label
  df %<>% filter(!is.na(endpoints))
    
# Generate plot
  plot <- generate_geom_mean_MIN_LC_longitudinal_plots(df)
  
# Save plots
  timepoint <- "acute"
  filename <- paste(workDir, "/plot_objects/additional_plots/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)
```

## Convalescent plots

```{r}

# Pull only convalescent test data
  df <- scores %>% filter(partition == "test")
    
# Generate plot
  plot <- generate_geom_mean_MIN_LC_longitudinal_plots(df)
  
# Save plots
  timepoint <- "convalescent"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)
```

