---
title: "Clinical features association analysis"
author: "Gisela Gabernet (gisela.gabernet@yale.edu)"
subtitle: PGX, SO, PMG, PPT, PPG, BCT
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
params:
  results_dir: "associate_clinical_features/"
  spear_object: "/scratch/data-fullcohort/spear_model/results/020624/032124_spearphysical_finalscores_allsamples.rds"
  partition: "all"
editor_options: 
  chunk_output_type: inline
---

```{r}
library(data.table)
#install.packages("RColorBrewer")
library(RColorBrewer)
library(ggplot2)
library(tidyverse)
#install.packages("psych")
library(psych)
library(dplyr)
# Additional libraries for mixed effect models
#BiocManager::install("qvalue")
library(qvalue)
library(lme4)
library(ordinal)
library(nlme)
library(gamm4)
library(mgcv)
library(ggsignif)
library(kableExtra)
library(corrplot)
library(ggeffects)
library(ggfortify)
library(grid)
library(pheatmap)

#install.packages("ggvenn")
library(ggvenn)

#install.packages("msigdbr")
library(msigdbr)

knitr::opts_chunk$set(warning = F, message=F)

print_kable <- function(table, kable_height = "500px", kable_width = "100%") {
  kable(table, digits=3) %>%
    kable_styling("striped") #%>%
    #scroll_box(height = kable_height, width = kable_width)
  
}

# save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
#    stopifnot(!missing(x))
#    stopifnot(!missing(filename))
#    pdf(filename, width=width, height=height)
#    grid::grid.newpage()
#    grid::grid.draw(x$gtable)
#    dev.off()
# }

```

# Running parameters

Running with parameters:
- features:
  - spear: get spear top features with joint probability >0.95
  - hallmark: get core set from hallmark heme metabolism pathway
- partition:
  - all: use all data for association
  - test: use the test dataset for association (with and without PROMIS PHY values)
  - PRO_PHY_NA_TEST: use all test dataset for association (with and without PROMIS PHY values, and the train dataset without PROMIS PHY values available)
- Spear object: spear object containing factor scores, and joint probabilities for extracting features.
- results_dir: name of results dir, will be created.

Example command to execute using an R script:

```{r eval=F}
# rmarkdown::render("associate_top_spear_features.Rmd",
#                   output_file="associate_top_spear_features_promis_physical_Factor1.html",
#                   params=list(
#                     results_dir = "associate_top_spear_features_promis_physical_Factor1/",
#                     spear_object = "/scratch/data-fullcohort/spear_model/results/factor_scores/031124_spearphysical_finalscores.rds",
#                     partition = "all",
#                     features = "spear"
#                   ))
```


# Explore Spear results

## Explore Feature correlation and variance explained

```{r fig.height=5, fig.width=10}

spear_object <- readRDS(params$spear_object)


#results_dir <- "./associate_spear_factor_promis_physical_train/"
results_dir <- params$results_dir
dir.create(results_dir)

figures_dir <- "./Figures_associate_clinical_features"
dir.create(figures_dir)


path_to_codebase <- "~/repos/impacc-full-cohort/Analysis/IMPACC_data_analysis_codebase.R"
source(path_to_codebase)

#selected_factor_name <- "Factor1"
```


## Get clinical data

```{r}
data_env_train <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_convalescent_data_env_train_Feb_11_2024.RDS")
clinical_data_train <- data_env_train$clinical_data
data_env_test <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_convalescent_data_env_test_Feb_11_2024.RDS")
clinical_data_test <- data_env_test$clinical_data

clinical_data_all <- bind_rows(clinical_data_train, clinical_data_test)

# Load clinical data (acute phase)
data_env_acute_train <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_train_Mar_23_2024.RDS")
clinical_data_acute_train <- data_env_acute_train$clinical_data
data_env_acute_test <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_test_Mar_23_2024.RDS")
clinical_data_acute_test <- data_env_acute_test$clinical_data

clinical_data_acute_all <- bind_rows(clinical_data_acute_train, clinical_data_acute_test)

# Add additional symptom groups to metadata
symptom_groups <- read_csv("/scratch/clinical/current/impacc_conv_symptoms.csv", show_col_types = F)
PRO_scores <- read_csv("/scratch/clinical/current/PRO_scores_dates_updated_29Aug23.csv", show_col_types = F)



```

# Associate all clinical features with the recovery factor

```{r}
factor_scores <- as.data.frame(rbind(spear_object$Train, 
                                     spear_object$TrainNoPhysical_Overlap, 
                                     spear_object$TrainNoPhysical_NoOverlap,
                                     spear_object$Test,
                                     spear_object$TestNoPhysical))

spear_factors <- as.data.frame(factor_scores)

selected_factor_names <- c("Factor1")

# Additional clinical variables
clinical_additional <- read.csv("/scratch/data-fullcohort/clinical_data_supplements/longitudinal_multiomics_lab_clinical_supplement.csv")
clinical_additional <- clinical_additional %>%
  dplyr::rename(participant_id = studyid)

clinical_data_filtered <- clinical_data_all %>%
  left_join(clinical_additional %>% select(participant_id, baseline_lab_hgb, baseline_lab_ferritin, 
                                           baseline_lab_hct, ever_nsaids, baseline_lab_bili
  ), by = join_by(participant_id)) %>%
  select(event_id, participant_id, sex, admit_age, discretized_admit_age_quantile, enrollment_site, 
         race, ethnicity, bmi, baseline_sofa_score_new, los,
         all_of(starts_with(c("comp_","comorb_","baseline_lab_"))),
                -symptom_date, -comorb_count, -comorb_countcat, -comp_stroke, 
         -comp_endo, -comp_othpulm,-comp_mis, ever_steroids, ever_nsaids, ever_immsupp, 
         -symptom_chestwall, -ever_ever_esc) %>% # comp and sympweren't observed or produce errors during association
  distinct()

df_association <- factor_scores %>%
  select(all_of(selected_factor_names)) %>%
  rownames_to_column("event_id") %>%
  left_join(clinical_data_filtered, by=join_by("event_id")) %>%
  mutate_if(is.character, as.factor) %>%
  mutate_at(vars(starts_with("comorb_")), as.factor) %>%
  mutate_at(vars(starts_with("comp_")), as.factor) %>%
  mutate_at(vars(starts_with("ever_")), as.factor) %>%
  mutate_at(vars(race, ethnicity, bmi, baseline_sofa_score_new), as.factor)


# Split columns by type
split_df_association <- df_association %>% 
  split.default(sapply(., class))

# Access the split dataframes
factor_df_association <- cbind(df_association[,c(selected_factor_names),drop=F],
                               split_df_association$factor)
numeric_df_association <- cbind(df_association[,c("event_id", "participant_id", "sex", "discretized_admit_age_quantile", "enrollment_site"),drop=F], 
                                split_df_association$numeric, 
                                split_df_association$integer)

factor_df_association_long <- factor_df_association %>%
  pivot_longer(-c("event_id", "participant_id", 
                  "discretized_admit_age_quantile", 
                  "enrollment_site", all_of(selected_factor_names))) %>%
  left_join(df_association %>% select(event_id, sex) %>% distinct(), by = join_by("event_id"))

numeric_df_association_long <- numeric_df_association %>%
   pivot_longer(-c("event_id", "participant_id", 
                   "sex", "discretized_admit_age_quantile", 
                   "enrollment_site", all_of(selected_factor_names)))
  
```

Here we will use a mixed linear model to associate features with outcomes

```{r}

out.p <- data.frame(matrix(nrow = 0, ncol=5))

for (data in list(factor_df_association_long, numeric_df_association_long)){
  for(active_factor in selected_factor_names){
    for(active_name in unique(data$name)){
      tmp <- data %>%
        filter(name == active_name) %>%
        filter(!is.na(value)) %>%
        filter(value != 999)
      
      if(active_name == "sex"){
        lme.formula.test <- formula(paste0(active_factor, " ~", "value + discretized_admit_age_quantile"))
        fit <- lme(fixed = lme.formula.test,
                   random = ~1|enrollment_site/participant_id, data = tmp)
        tmp.coef <- fit$coefficients$fixed[2]
        coef.note <- "Coefficient for Male Sex"
      } else if(active_name == "admit_age"){
        lme.formula.test <- formula(paste0(active_factor, " ~", "value + sex"))
        fit <- lme(fixed = lme.formula.test,
                   random = ~1|enrollment_site/participant_id, data = tmp)
        tmp.coef <- fit$coefficients$fixed[2]
        coef.note <- NA
       } else {
        lme.formula.test <- formula(paste0(active_factor, " ~", "value + sex+ discretized_admit_age_quantile"))
        fit <- lme(fixed = lme.formula.test,
                   random = ~1|enrollment_site/participant_id, data = tmp)
        tmp.coef <- fit$coefficients$fixed[2]
        coef.note <- NA
      }
      
      out.p <- rbind.data.frame(out.p, data.frame("clinical_feature" = active_name,
                                                  "Factor" = active_factor,
                                                  "p_val" = anova(fit, type = "marginal")$`p-value`[2],
                                                  "coefficient" =tmp.coef,
                                                  "coefficient_notes" = coef.note))
      rm(coef.note)
      
    }
    
  }
}

df_corr <- out.p %>%
  mutate(adj_pval = p.adjust(as.numeric(p_val), "fdr")) %>%
  mutate(sign = if_else(coefficient>=0, 1, -1))


```

```{r fig.height=3, fig.width=8}

clin.order <- c(
  "admit_age", "sex", "bmi", "los", "baseline_sofa_score_new", 
  "comorb_htn", "comorb_isaric_dm", "comorb_isaric_cardiac", "comorb_isaric_ckd", "comorb_isaric_neoplasm", "comorb_neuro", "comorb_isaric_liver", "comorb_hxtrans", "comorb_eversmkvape",
"comorb_isaric_asthma", "comorb_anyresp_noasthma", "comorb_anysubstance", "comorb_hiv2", "comorb_count",
 "comp_anemia", "ever_steroids", "ever_nsaids", "ever_immsupp",
"baseline_lab_hgb", "baseline_lab_hct", "baseline_lab_bili", "baseline_lab_ferritin", "baseline_crp_abn",
"baseline_lab_cr", "baseline_lab_platelets"
)

clin.labs <- c(
  "Age", "Sex", "BMI", "Length of hospital stay", "SOFA score",
  "Hypertension", "Diabetes", "Chronic Cardiac Disease", "Chronic Kidney Disease", "Malignant Neoplasm", 
  "Chronic Neurological Disorder", "Liver Disfunction/Failure", "Solid Organ/BM Transplant", "Smoking/Vaping", "Asthma", "Other Respiratory Disease", "Substance Use", "HIV",  "Comorbidity Count",
 "Anemia", "Steroids", "NSAIDs", "Immunosupressants",
"Hemoglobin", "Hematocrit", "Bilirubin", "Ferritin", "C-Reactive Protein", "Creatinine", "Platelets"
)


df_corr_filt <- df_corr %>%
  filter(adj_pval < 0.05) %>%
  mutate(adjpval_annot = if_else(adj_pval < 0.05, "*", ""),
         adjpval_annot = if_else(adj_pval < 0.01, "**", adjpval_annot),
         adjpval_annot = if_else(adj_pval < 0.001, "***", adjpval_annot),
         signed_adjpval = sign * -log10(adj_pval + 1e-13)
         ) %>%
  remove_rownames()

clin.order.filt <- clin.order[clin.order %in% df_corr_filt$clinical_feature]
clin.labs.filt <- clin.labs[clin.order %in% df_corr_filt$clinical_feature]

df_corr_ordered <- df_corr_filt
df_corr_ordered$clinical_feature <- factor(df_corr_ordered$clinical_feature, levels = clin.order.filt)

clinical_features <- ggplot(df_corr_ordered) +
  geom_point(aes(x = clinical_feature, y = Factor, fill = sign * -log10(adj_pval + 1e-13), size = -log10(adj_pval + 1e-13)), shape = 21) +
  scale_fill_gradient2(low = "darkblue", high = "darkred") +
  coord_equal() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 270, hjust = 0, vjust = .5)) +
  labs(fill = "Signed -log10(p-value)") +
  guides(size = "none") +
  scale_x_discrete(labels = clin.labs.filt, position = "top") +
  scale_y_discrete(labels = c("Recovery factor")) +
  xlab(NULL) +
  ylab(NULL) +
  theme(axis.text.x = element_text(angle = 45),
        panel.grid = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(r = 50, unit = "pt")) +
  geom_vline(xintercept = c(1:54) + .5, color = "grey80", linewidth = .5) +
  geom_hline(yintercept = c(1) + .5, color = "grey80", linewidth = .5) +
  geom_vline(xintercept = c(4, 12, 16) + .5, linewidth = .5)

saveRDS(clinical_features, file.path(figures_dir, "all_clinical_features_association.rds"))

clinical_features
```

# Associate the features used in the clinical model with the recovery factor

```{r}
# Load clinical data

# # Selecting clinical data from integration manuscript
# clinical_vars <- c("sex","race", "ethnicity", "discretized_admit_age_quantile", "bmi")
# include_vars_comorb <- c("comorb_htn", "comorb_isaric_dm", "comorb_anyresp_noasthma", "comorb_isaric_asthma", "comorb_isaric_cardiac", 
#                          "comorb_isaric_ckd", "comorb_isaric_neoplasm", "comorb_neuro", "comorb_isaric_liver", "comorb_hxtrans", 
#                          "comorb_hiv2", "comorb_eversmkvape", "comorb_anysubstance", "comorb_count", "comorb_countcat")
# include_vars_baseline <- c("baseline_img_infil", "baseline_lymph_abn", "baseline_platelets_abn", "baseline_alt_abn", "baseline_cr_abn",    
#                            "baseline_crp_abn", "baseline_ddimer_abn", "baseline_trop_abn" )
# all_clinical_vars <- c("event_id", clinical_vars, include_vars_comorb, include_vars_baseline)
# 
# # Clinical data train
# clinical_vars_train <- clinical_data_train %>%
#   dplyr::select(all_of(all_clinical_vars)) %>%
#   distinct() %>%
#   column_to_rownames("event_id")


# Read assays that are included in clinical model
antibody_titers_acute_env <- readRDS("/scratch/data-fullcohort/batch-effects/preprocessed_assay_antibody_titers_acute_20240701.rds")

nasal_viral_load_acute <- readRDS("/scratch/data-fullcohort/batch-effects/nasal_viral_load_acute_20240701.rds")

#Include vars average of acute time point of SPIKE IgG titers, N1-CT viral load
nasal_viral_load_acute_participant_mean <- nasal_viral_load_acute %>%
  select(participant_id, event_id, N1_CT) %>%
  pivot_longer(cols = -c(event_id, participant_id), names_to = "feature", values_to = "value") %>%
  group_by(participant_id, feature) %>%
  summarize(mean_value = mean(value, na.rm=T)) %>%
  pivot_wider(id_cols = participant_id, names_from = feature, values_from = mean_value)

antibody_titers_acute_participant_mean <- as.data.frame(antibody_titers_acute_env$preprocessed_data$serum_rbd_abtiters) %>%
  select(`AUC Spike IgG`) %>%
  rownames_to_column("sample_id") %>%
  left_join(clinical_data_acute_all %>%
              select(sample_id,event_id,participant_id), by = join_by(sample_id)) %>%
  select(-c(sample_id)) %>%
  filter(!is.na(event_id)) %>%
  pivot_longer(cols = -c(event_id, participant_id), names_to = "feature", values_to = "value") %>%
  group_by(participant_id, feature) %>%
  summarize(mean_value = mean(value, na.rm=T)) %>%
  pivot_wider(id_cols = participant_id, names_from = feature, values_from = mean_value)
  

# Selecting clinical data from subtyping manuscript
clinical_vars_subtyping <- c("sex","discretized_admit_age_quantile", "admit_age", "bmi", "los", "baseline_sofa_score_new", "enrollment_site")
include_vars_comorb_subtyping <- c("comorb_htn", "comorb_isaric_dm", "comorb_anyresp_noasthma", "comorb_isaric_asthma", "comorb_isaric_cardiac", 
                         "comorb_isaric_ckd", "comorb_isaric_neoplasm", "comorb_neuro", "comorb_isaric_liver", "comorb_hxtrans", 
                         "comorb_hiv2", "comorb_eversmkvape", "comorb_anysubstance", "comorb_count")

all_clinical_vars_subtyping <- c("event_id", clinical_vars_subtyping, include_vars_comorb_subtyping)

# Clinical data train
clinical_vars_subtyping_train <- clinical_data_all %>%
  dplyr::select(all_of(all_clinical_vars_subtyping), participant_id) %>%
  distinct() %>%
  left_join(nasal_viral_load_acute_participant_mean, by=join_by(participant_id)) %>%
  left_join(antibody_titers_acute_participant_mean, by=join_by(participant_id)) %>%
  select(-participant_id) %>%
  distinct() %>%
  column_to_rownames("event_id")

# Additional clinical variables
clinical_additional <- read.csv("/scratch/data-fullcohort/clinical_data_supplements/longitudinal_multiomics_lab_clinical_supplement.csv")
clinical_additional <- clinical_additional %>%
  dplyr::rename(participant_id = studyid)

clinical_vars_additional <- clinical_data_all %>%
  select(event_id, participant_id, baseline_crp_abn, comp_anemia, ever_steroids) %>%
  distinct() %>%
  left_join(clinical_additional %>% select(participant_id, baseline_lab_hgb, baseline_lab_ferritin, 
                                           baseline_lab_hct, ever_nsaids, baseline_lab_bili
  ), by = join_by(participant_id))

clinical_subtyping_with_additional <- as.data.frame(clinical_vars_subtyping_train) %>%
  rownames_to_column("event_id") %>%
  left_join(clinical_vars_additional, by=join_by("event_id"))

```

### Association clinical variables subtyping with additional
```{r}

factor_scores <- as.data.frame(rbind(spear_object$Train, 
                                     spear_object$TrainNoPhysical_Overlap, 
                                     spear_object$TrainNoPhysical_NoOverlap,
                                     spear_object$Test,
                                     spear_object$TestNoPhysical))

spear_factors <- as.data.frame(factor_scores)

selected_factor_names <- c("Factor1")

df_association <- factor_scores %>%
  select(all_of(selected_factor_names)) %>%
  rownames_to_column("event_id") %>%
  left_join(clinical_subtyping_with_additional, by=join_by("event_id")) %>%
  mutate_if(is.character, as.factor) %>%
  mutate_at(vars(starts_with("comorb_")), as.factor) %>%
  mutate_at(vars(starts_with("comp_")), as.factor) %>%
  mutate_at(vars(starts_with("ever_")), as.factor) 


# Split columns by type
split_df_association <- df_association %>% 
  split.default(sapply(., class))

# Access the split dataframes
factor_df_association <- cbind(df_association[,c(selected_factor_names),drop=F],
                               split_df_association$factor)
numeric_df_association <- cbind(df_association[,c("event_id", "participant_id", "sex", "discretized_admit_age_quantile", "enrollment_site"),drop=F], 
                                split_df_association$numeric, 
                                split_df_association$integer)

factor_df_association_long <- factor_df_association %>%
  pivot_longer(-c("event_id", "participant_id", 
                  "discretized_admit_age_quantile", 
                  "enrollment_site", all_of(selected_factor_names))) %>%
  left_join(df_association %>% select(event_id, sex) %>% distinct(), by = join_by("event_id"))

numeric_df_association_long <- numeric_df_association %>%
   pivot_longer(-c("event_id", "participant_id", 
                   "sex", "discretized_admit_age_quantile", 
                   "enrollment_site", all_of(selected_factor_names)))
  
```

Here we will use a mixed linear model to associate features with outcomes

```{r}

out.p <- data.frame(matrix(nrow = 0, ncol=5))

for (data in list(factor_df_association_long, numeric_df_association_long)){
  for(active_factor in selected_factor_names){
    for(active_name in unique(data$name)){
      tmp <- data %>%
        filter(name == active_name) %>%
        filter(!is.na(value))
      
      if(active_name == "sex"){
        lme.formula.test <- formula(paste0(active_factor, " ~", "value + discretized_admit_age_quantile"))
        fit <- lme(fixed = lme.formula.test,
                   random = ~1|enrollment_site/participant_id, data = tmp)
        tmp.coef <- fit$coefficients$fixed[2]
        coef.note <- "Coefficient for Male Sex"
      } else if(active_name == "admit_age"){
        lme.formula.test <- formula(paste0(active_factor, " ~", "value + sex"))
        fit <- lme(fixed = lme.formula.test,
                   random = ~1|enrollment_site/participant_id, data = tmp)
        tmp.coef <- fit$coefficients$fixed[2]
        coef.note <- NA
       } else {
        lme.formula.test <- formula(paste0(active_factor, " ~", "value + sex+ discretized_admit_age_quantile"))
        fit <- lme(fixed = lme.formula.test,
                   random = ~1|enrollment_site/participant_id, data = tmp)
        tmp.coef <- fit$coefficients$fixed[2]
        coef.note <- NA
      }
      
      out.p <- rbind.data.frame(out.p, data.frame("clinical_feature" = active_name,
                                                  "Factor" = active_factor,
                                                  "p_val" = anova(fit, type = "marginal")$`p-value`[2],
                                                  "coefficient" =tmp.coef,
                                                  "coefficient_notes" = coef.note))
      rm(coef.note)
      
    }
    
  }
}

df_corr <- out.p %>%
  mutate(adj_pval = p.adjust(as.numeric(p_val), "fdr")) %>%
  mutate(sign = if_else(coefficient>=0, 1, -1))


```

```{r fig.height=3, fig.width=8}

all_vars_subtyping <- c(clinical_vars_subtyping, include_vars_comorb_subtyping, "N1_CT", "AUC Spike IgG")

df_corr_subtyping <- df_corr %>% filter(clinical_feature %in% all_vars_subtyping)

clin.order <- c(
  "admit_age", "sex", "bmi", "los", "baseline_sofa_score_new", 
  "comorb_htn", "comorb_isaric_dm", "comorb_isaric_cardiac", "comorb_isaric_ckd", "comorb_isaric_neoplasm", "comorb_neuro", "comorb_isaric_liver", "comorb_hxtrans", "comorb_eversmkvape",
"comorb_isaric_asthma", "comorb_anyresp_noasthma", "comorb_anysubstance", "comorb_hiv2", "comorb_count",
"N1_CT",  "AUC Spike IgG"
)

clin.labs <- c(
  "Age", "Sex", "BMI", "Length of hospital stay", "SOFA score",
  "Hypertension", "Diabetes", "Chronic Cardiac Disease", "Chronic Kidney Disease", "Malignant Neoplasm", "Chronic Neurological Disorder", "Liver Disfunction/Failure", "Solid Organ/BM Transplant", "Smoking/Vaping", "Asthma", "Other Respiratory Disease", "Substance Use", "HIV",  "Comorbidity Count",
  "N1 CT", "Spike IgG titer"
)

df_corr_subtyping <- df_corr_subtyping %>%
  mutate(adjpval_annot = if_else(adj_pval < 0.05, "*", ""),
         adjpval_annot = if_else(adj_pval < 0.01, "**", adjpval_annot),
         adjpval_annot = if_else(adj_pval < 0.001, "***", adjpval_annot),
         signed_adjpval = sign * -log10(adj_pval + 1e-13)
         ) %>%
  remove_rownames() %>%
  column_to_rownames("clinical_feature")

df_corr_ordered <- df_corr_subtyping[clin.order, ]


heatmap1 <- ComplexHeatmap::pheatmap(t(as.matrix(df_corr_ordered[,"signed_adjpval"])),
        column_names_side = c("top"),
        row_names_side = c("left"),
        angle_col = "45",
         cellwidth = 15,
         cellheight = 15,
         display_numbers = t(as.matrix(df_corr_ordered[,"adjpval_annot"])),
         fontsize_number = 10,
         fontsize = 8,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F,
         labels_col = clin.labs,
         labels_row = c("Recovery factor"),
         show_rownames = T,
         color = colorRampPalette(c("darkblue", "white", "darkred"))(100),
         #breaks = breaklist,
        gaps_col = c(5, 19),
        heatmap_legend_param = list(
        title = "Signed -log10 adj. p",
        direction = "horizontal",
        position = "bottom"
        ))
p1 <- draw(heatmap1, heatmap_legend_side = "bottom")

# clinical_features <- ggplot(df_corr) +
#   geom_point(aes(x = clinical_feature, y = Factor, fill = sign * -log10(adj_pval + 1e-13), size = -log10(adj_pval + 1e-13)), shape = 21) +
#   scale_fill_gradient2(low = "darkblue", high = "darkred") +
#   coord_equal() +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 270, hjust = 0, vjust = .5)) +
#   labs(fill = "Signed -log10(p-value)") +
#   guides(size = "none") +
#   scale_x_discrete(labels = clin.labs, position = "top") +
#   scale_y_discrete(labels = c("Recovery factor")) +
#   xlab(NULL) +
#   ylab(NULL) +
#   theme(axis.text.x = element_text(angle = 45),
#         panel.grid = element_blank(),
#         legend.position = "bottom",
#         plot.margin = margin(r = 50, unit = "pt")) +
#   geom_vline(xintercept = c(1:54) + .5, color = "grey80", size = .5) +
#   geom_hline(yintercept = c(1) + .5, color = "grey80", size = .5) +
#   geom_vline(xintercept = c(2, 16, 22) + .5, size = .5)

saveRDS(p1, file.path(figures_dir, "clinical_subtyping_features_association.rds"))
```

```{r fig.height=3, fig.width=8}


df_corr_subtyping_additional <- df_corr

# clin.order.subtyping.additional <- c(
#   "admit_age", "sex", "comorb_htn", "comorb_isaric_dm", "comorb_isaric_cardiac", "comorb_isaric_ckd", "comorb_isaric_neoplasm", "comorb_neuro", "comorb_isaric_liver", "comorb_hxtrans", "comorb_eversmkvape",
# "comorb_isaric_asthma", "comorb_anyresp_noasthma", "comorb_anysubstance", "comorb_hiv2", "comorb_count",
# "bmi", "los", "baseline_sofa_score_new", "comp_anemia", "ever_steroids", "ever_nsaids",
# "baseline_lab_hgb", "baseline_lab_hct", "baseline_lab_bili", "baseline_lab_ferritin", "baseline_crp_abn"
# )
# 
# clin.labs.subtyping.additional <- c(
#   "Age", "Sex", 
#   "Hypertension", "Diabetes", "Chronic Cardiac Disease", "Chronic Kidney Disease", "Malignant Neoplasm", "Chronic Neurological Disorder", "Liver Disfunction/Failure", "Solid Organ/BM Transplant", "Smoking/Vaping", "Asthma", "Other Respiratory Disease", "Substance Use", "HIV",  "Comorbidity Count", 
# "BMI", "Length of hospital stay", "SOFA score", "Anemia", "Steroids", "NSAIDs",
# "Hemoglobin", "Hematocrit", "Bilirubin", "Ferritin", "C-Reactive Protein"
# )

clin.order <- c(
  "admit_age", "sex", "bmi", 
  "los", "baseline_sofa_score_new", 
  "comorb_htn", "comorb_isaric_dm", "comorb_isaric_cardiac", "comorb_isaric_ckd", "comorb_isaric_neoplasm", "comorb_neuro", "comorb_isaric_liver", "comorb_hxtrans", "comorb_eversmkvape",
"comorb_isaric_asthma", "comorb_anyresp_noasthma", "comorb_anysubstance", "comorb_hiv2", "comorb_count", 
"N1_CT", "AUC Spike IgG",
"comp_anemia", "ever_steroids", "ever_nsaids", "baseline_lab_hgb", "baseline_lab_hct", "baseline_lab_bili", "baseline_lab_ferritin", "baseline_crp_abn"
)

clin.labs <- c(
  "Age", "Sex", "BMI", "Length of hospital stay", "SOFA score",
  "Hypertension", "Diabetes", "Chronic Cardiac Disease", "Chronic Kidney Disease", "Malignant Neoplasm", "Chronic Neurological Disorder", "Liver Disfunction/Failure", "Solid Organ/BM Transplant", "Smoking/Vaping", "Asthma", "Other Respiratory Disease", "Substance Use", "HIV",  "Comorbidity Count",
"N1 CT", "Spike IgG titer",
"Anemia", "Steroids", "NSAIDs", "Hemoglobin", "Hematocrit", "Bilirubin", "Ferritin", "C-Reactive Protein"
)

df_corr_subtyping_additional <- df_corr_subtyping_additional %>%
  mutate(adjpval_annot = if_else(adj_pval < 0.05, "*", ""),
         adjpval_annot = if_else(adj_pval < 0.01, "**", adjpval_annot),
         adjpval_annot = if_else(adj_pval < 0.001, "***", adjpval_annot),
         signed_adjpval = sign * -log10(adj_pval + 1e-13)
         ) %>%
  remove_rownames() %>%
  column_to_rownames("clinical_feature")

df_corr_ordered <- df_corr_subtyping_additional[clin.order, ]


heatmap1 <- ComplexHeatmap::pheatmap(t(as.matrix(df_corr_ordered[,"signed_adjpval"])),
        column_names_side = c("top"),
        row_names_side = c("left"),
        angle_col = "45",
         cellwidth = 15,
         cellheight = 15,
         display_numbers = t(as.matrix(df_corr_ordered[,"adjpval_annot"])),
         fontsize_number = 10,
         fontsize = 8,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F,
         labels_col = clin.labs,
         labels_row = c("Recovery factor"),
         show_rownames = T,
         color = colorRampPalette(c("darkblue", "white", "darkred"))(100),
         #breaks = breaklist,
        gaps_col = c(5, 19, 21),
        heatmap_legend_param = list(
        title = "Signed -log10 adj. p",
        direction = "horizontal",
        position = "bottom"
        ))
p1 <- draw(heatmap1, heatmap_legend_side = "bottom")
# clinical_features <- ggplot(df_corr) +
#   geom_point(aes(x = clinical_feature, y = Factor, fill = sign * -log10(adj_pval + 1e-13), size = -log10(adj_pval + 1e-13)), shape = 21) +
#   scale_fill_gradient2(low = "darkblue", high = "darkred") +
#   coord_equal() +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 270, hjust = 0, vjust = .5)) +
#   labs(fill = "Signed -log10(p-value)") +
#   guides(size = "none") +
#   scale_x_discrete(labels = clin.labs, position = "top") +
#   scale_y_discrete(labels = c("Recovery factor")) +
#   xlab(NULL) +
#   ylab(NULL) +
#   theme(axis.text.x = element_text(angle = 45),
#         panel.grid = element_blank(),
#         legend.position = "bottom",
#         plot.margin = margin(r = 50, unit = "pt")) +
#   geom_vline(xintercept = c(1:54) + .5, color = "grey80", size = .5) +
#   geom_hline(yintercept = c(1) + .5, color = "grey80", size = .5) +
#   geom_vline(xintercept = c(2, 16, 22) + .5, size = .5)

saveRDS(p1, file.path(figures_dir, "clinical_subtyping_additional_features_association.rds"))
```