 ---
title: "Trajectory Group associations with Spear Factor1"
author: "Jessica Maciuch"
date: "`r format(Sys.time(), '%b %d %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory
  workDir <- "~/impacc-full-cohort/Analysis/convalescent-project/Spear/clinical_associations"
  setwd(workDir) # For troubleshooting in the command line
  knitr::opts_knit$set(root.dir = workDir)

suppressPackageStartupMessages({
  library(tidyverse)
  library(magrittr)
  library(ggpubr)
  library(rstatix)
  library(patchwork)
  library(pheatmap)
  library(ggeffects)
  library(RColorBrewer)})

# Load codebase functions
  source("~/impacc-full-cohort/Analysis/IMPACC_data_analysis_codebase.R")

# Initialize list of variables to exclude from downstream environment cleaning
  keep_vars <- ls()
  keep_vars %<>% c("keep_vars")
```

# Load data

## Clinical data

Pull clinical data for confounding factors (age, sex) and trajectory group designations

```{r}

# Load clinical data for acute and convalescent ppts
  data_env_conv <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_convalescent_data_env_all_Feb_11_2024.RDS")
  data_env_acute <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_allcohort_all_Aug_12_2024.RDS") 
  
  clinical_data_conv <- data_env_conv$clinical_data
  clinical_data_acute <- data_env_acute$clinical_data
  clinical_data <- dplyr::bind_rows(list(clinical_data_conv, 
                                         clinical_data_acute))
    
# Select variables
  clinical_data %<>% 
    select(all_of(c("participant_id", 
                    "event_id", 
                    "event_type",
                    "event_date",
                    "enrollment_site", 
                    "discretized_admit_age_quantile",
                    "sex",
                    "trajectory_group",
                    "pro_2_groups"))) %>% 
    distinct()

# Transform variables into factors
  factor_vars <- c("enrollment_site", 
                   "discretized_admit_age_quantile", 
                   "sex", 
                   "trajectory_group", 
                   "event_type", 
                   "participant_id",
                   "pro_2_groups")
  clinical_data %<>% mutate(across(all_of(factor_vars), as.factor))
  
# Specify order of variables for plotting
  clinical_data %<>% mutate(pro_2_groups = if_else(pro_2_groups == "MIN", "MIN", "LC"))
  clinical_data$pro_2_groups %<>% factor(., levels = c("MIN", "LC"))
  
  clinical_data$event_type %<>% factor(., levels = paste("Visit", 1:10))
  
  # Append "TG" to trajectory group variable
  clinical_data$trajectory_group %<>% 
    paste("TG", ., sep = "") %>%
    factor(., levels = paste("TG", 1:5, sep = "")) %>%
    droplevels()
  
# Clean environment
  keep_vars %<>% c("clinical_data")
  rm(list = setdiff(ls(), keep_vars))
  gc()
```

## Spear factors

```{r}

# Load spear factors
  spear_promis_phys_conv <- readRDS("/scratch/data-fullcohort/spear_model/results/020624/032124_spearphysical_finalscores_allsamples.rds")
  spear_promis_phys_acute <- readRDS("/scratch/data-fullcohort/spear/scores_data_acutefullcohort/082024_spearphysical_finalscores_acute_fullcohort_allsamples.rds")

# Separate test and train factors
  spear_scores <- list(test = rbind(spear_promis_phys_conv$Test,
                                    spear_promis_phys_conv$TestNoPhysical,
                                    spear_promis_phys_conv$TrainNoPhysical_NoOverlap),
                       acute = spear_promis_phys_acute$Acute_fullcohort)
    
# Add event_id as explicit column for downstream merging
  spear_scores %<>% map(\(df) df %<>% 
                          as.data.frame() %>% 
                          mutate(event_id = rownames(.)) %>% 
                          select(event_id, Factor1))
  
# Combine spear data with clinical data
  spear_scores %<>% bind_rows(.id = "partition") 
    
# Clean environment
  keep_vars %<>% c("spear_scores")
  rm(list = setdiff(ls(), keep_vars))
  gc()
```

## Prepare data frame

```{r}

# Combine spear and clinical data
  clinical_spear <- left_join(x = spear_scores,
                              y = clinical_data,
                              by = "event_id")

# Remove visit 6 outliers and accidentally included acute visits from conv phase
  clinical_spear %<>% filter(!(partition == "acute" & event_date > 50))
  
# Prepare data frame for codebase functions
  clinical_spear %<>% pivot_longer(cols = Factor1)

# Clean environment
  keep_vars %<>% c("clinical_spear")
  keep_vars %<>% .[which(!. %in% c("clinical_data", "spear_scores"))]
  rm(list = setdiff(ls(), keep_vars))
  gc()
```

# Color palettes

```{r}

# Set color palette for individual trajectory groups
  TG_cols <- c('#639A21',
               '#39828C', 
               '#6371AD',
               '#BD7D31',
               '#9C3418')
  names(TG_cols) <- levels(clinical_spear$trajectory_group)
  
# Set color palette for MIN/LC
  pro_2_groups_cols <- c("#649B20",
                         "#7A186D")
  names(pro_2_groups_cols) <- levels(clinical_spear$pro_2_groups)
  
# Clean environment
  keep_vars %<>% c("TG_cols", "pro_2_groups_cols")
  rm(list = setdiff(ls(), keep_vars))
  gc()
```

# MIN/LC per-visit box plots by trajectory group (global pvals)

## Set filename vars

```{r}
# Set variables for filenames
  comparison <- "MIN_LC_Factor1_byTG"
  plot_type <- "pervisit_global_pval"
```

## Plotting functions

```{r}

# Helper function to calculate p-value association
  global_mixed_pairwise <- function(df, 
                                    fixed_vars,
                                    predictor, 
                                    random) {
      
      # Generate formulas for mixed_pairwise function
      var_list <- c(fixed_vars, random)
      
      terms <- paste(var_list, collapse = "+")
      formula.0 <- formula(paste("value~", terms, sep = ""))
      
      var_list %<>% c(predictor)
      terms <- paste(var_list, collapse = "+")
      formula.1 <- formula(paste("value~", terms, sep = ""))
      
      res_table <- mixed_pairwise(formula.0, formula.1, df)
    
    return(res_table$res_table)
  }
  
# Helper function to generate per-visit boxplots
  plot_TG_MIN_LC <- function(df, cols) {
    ggplot(df, aes(x = trajectory_group, y = value)) +
      ggbeeswarm::geom_quasirandom(aes(fill = endpoints),
                                   shape = 21,
                                   width = .15, 
                                   dodge.width = 0.75) +
      geom_boxplot(aes(fill = endpoints),
                   alpha = 0.6, 
                   color = "black", 
                   outlier.shape = NA) +
      theme_bw() + 
      scale_color_manual(values = cols) +
      scale_fill_manual(values = cols) +
      scale_x_discrete(drop = F) +
      labs(x = "Trajectory Group", 
           y = "Recovery Factor", 
           fill = "") +
      geom_label(aes(x = Inf,
                     y = Inf, 
                     label = label,
                     hjust = 1,
                     vjust = 1),
                 label.r = unit(0, "pt"),
                 size = 4) +
      scale_y_continuous(expand = expansion(mult = c(0.1, 0.1))) +
      facet_wrap(vars(event_type)) +
      theme(text = element_text(size = 14),
            strip.background = element_blank(),
            strip.text = element_text(size = rel(1.25))) -> plot

    return(plot)
  }
  
# Helper function to generate completed plot from data frame
  generate_plot_TG_MIN_LC <- function(df) {
    # Change variable name for pval function
      df %<>% dplyr::rename("endpoints" = "pro_2_groups")
      
    # Drop observations without MIN/LC label
      df %<>% filter(!is.na(endpoints))
      
    # Drop TG5 from factor
      df$trajectory_group %<>% droplevels()
      
    # Split by visit
      df_list <- df %>% split(f = .[["event_type"]], drop = T)
      
    # Calculate p-values
      pval_list <- df_list |> 
        map(\(df) global_mixed_pairwise(df, 
                                        fixed_vars = c("discretized_admit_age_quantile",
                                                       "sex",
                                                       "trajectory_group"),
                                        predictor = "endpoints",
                                        random = "(1|enrollment_site)")) |> as.numeric()
      
    # P-value adjustment
      adjp_list <- qvalue::qvalue(pval_list, fdr.level = 0.05, pi0 = 1)$qvalues
      
    # Add p-vals as facet titles
      pval_df <- data.frame(pval = pval_list,
                            adj_pval = adjp_list,
                            event_type = names(df_list))
      
      pval_df %<>% mutate(across(contains("pval"), \(x) signif(x, digits = 2)))
      
      df %<>% left_join(x = .,
                        y = pval_df,
                        by = "event_type")
      
      df %<>% mutate(pval_label = paste("adj. p=", adj_pval, sep = ""))
      
    # Make sure visits are ordered correctly in facet
      df$event_type %<>% factor(., levels = paste("Visit", 1:10))
      
    # Add "n" for each visit
      n_df <- df %>% 
        select(event_type, participant_id) %>%
        group_by(event_type) %>%
        count() %>%
        mutate(n_label = paste("(n = ", n, ")", sep = "")) 
      
      df %<>% left_join(x = .,
                        y = n_df %>% select(event_type, n_label),
                        by = "event_type")
      
      df %<>% mutate(label = paste(pval_label, n_label, sep = "\n"))
      
    # Generate plot
      plot <- plot_TG_MIN_LC(df, pro_2_groups_cols)
      
      return(plot)
  }

```

## Acute plots

```{r}

# Pull only acute data
  df <- clinical_spear %>% filter(partition == "acute")
  
# Generate plot
  plot <- generate_plot_TG_MIN_LC(df)
  
# Save plots
  timepoint <- "acute"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)
```

## Convalescent plots

```{r}

# Pull only convalescent test data
  df <- clinical_spear %>% filter(partition == "test")
  
# Generate plot
  plot <- generate_plot_TG_MIN_LC(df)
  
# Save plots
  timepoint <- "convalesecent"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)
```

# MIN/LC longitudinal plots by trajectory group

## Set filename vars

```{r}

# Set variables for filenames
  comparison <- "MIN_LC_Factor1_byTG"
  plot_type <- "longitudinal"
```

## Plotting functions

```{r}

# Custom function to create longitudinal plots because of weird bug flipping order of pro_2_groups factor levels for geom_ribbon
  plot_TG_MIN_LC_longitudinal <- function(df, cols) {
    # Generate model
    form <- as.formula("value ~ pro_2_groups + discretized_admit_age_quantile + sex + s(event_date, bs = 'ts') + s(event_date, by = pro_2_groups, bs = 'cr')")
    model <- gamm4::gamm4(form, random = ~(1|enrollment_site/participant_id), data = df)
    
    # Calculate confidence interval
    pred <- ggpredict(model, c("event_date", "pro_2_groups"), type = "fixed",
                      ci.lvl = 0.95)
    
    # Generate plot
    plot <- ggplot(df, aes(x = event_date, y = value)) +
      geom_point(aes(color = pro_2_groups)) +
      geom_ribbon(inherit.aes = F, 
                  data = pred, 
                  aes(x = x, ymin = pred$conf.low, ymax = pred$conf.high, fill = group),
                  alpha = 0.3) +
      geom_line(inherit.aes = F, 
                data = pred, 
                aes(x = x, y = predicted, color = group),
                linewidth = 1.5) +
      scale_color_manual(values = cols) +
      scale_fill_manual(values = cols) +
      labs(title = unique(df$trajectory_group),
           color = "",
           fill = "",
           x = "Days from Admission", 
           y = "Recovery Factor") +
      theme_bw() +
      theme(text = element_text(size = 14),
            plot.title = element_text(hjust = 0.5, size = rel(1.25)))
    
    return(plot)
  }

# Helper function to generate longitudinal MIN/LC plots from starting df
generate_plot_TG_MIN_LC_longitudinal <- function(df) {
  # Drop observations without PRO group label
  df %<>% filter(!is.na(pro_2_groups))

  # Split data frame by trajectory group
  df_list <- df %>% split(f = .[["trajectory_group"]], drop = T)
  
  # Generate plots 
  plots <- df_list |> map(\(df) plot_TG_MIN_LC_longitudinal(df, pro_2_groups_cols))
  
  # Calculate longitudinal p-vals
  pval_list <- df_list |> 
    map(\(inputDF) model_loop(inputDF, 
                              model_type = "smoothSpline",
                              spline_type = "ts",
                              endpoint = "pro_2_groups"))
  pvals <- bind_rows(pval_list, .id = "trajectory_group")
  
  # P-value adjustment
  adj_pvals <- qvalue::qvalue(pvals$p.intercept, fdr.level = 0.05, pi0 = 1)$qvalues
  
  # Create p-val labels
  pval_labels <- paste("adj. p=", signif(adj_pvals, digits = 2), sep = "")
  
  # Add p-val labels to plots
  plots <- list(plots, pval_labels) |> 
    pmap(\(plot, pval) plot <- plot + annotate("label",  
                                               x = Inf, 
                                               y = Inf,
                                               label = pval, 
                                               vjust = 1, 
                                               hjust = 1, 
                                               size = 5, 
                                               label.r = unit(0, "pt")))
  
  # Combine plots
  plots %<>% wrap_plots(ncol = 2, byrow = T)
  
  # Enforce same y axis limits
  plots <- plots & scale_y_continuous(limits = c(min(df$value), max(df$value)),
                                      expand = expansion(mult = c(0.1, 0.1)))
  
  return(plots)
}

```

## Acute plots

```{r}

# Pull only acute data
  df <- clinical_spear %>% filter(partition == "acute")

# Generate plots
  plots <- generate_plot_TG_MIN_LC_longitudinal(df)
  
# Save plots
  timepoint <- "acute"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plots, filename)

```

## Convalescent plots

```{r}

# Pull only convalescent test data
  df <- clinical_spear %>% filter(partition == "test")

# Generate plots
  plots <- generate_plot_TG_MIN_LC_longitudinal(df)
  
# Save plots
  timepoint <- "convalescent"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plots, filename)

```

# Trajectory group vs. Factor1 longitudinal plots

## Set filename vars

```{r}

# Set variables for filenames
  comparison <- "trajectorygroups_Factor1"
  plot_type <- "longitudinal"
```

## Plotting functions

```{r}

# Helper fuction to generate longitudinal TG plot
generate_TG_Factor1_longitudinal_plot <- function(df) {
  # Drop missing TGs if any
  df$trajectory_group %<>% droplevels()
  
  # Calculate p-vals
  pvals <- model_loop(df, 
                      model_type = "smoothSpline",
                      spline_type = "ts",
                      endpoint ="trajectory_group")
  
  # Generate plot
  plot_model(df,
             pvals,
             model_type = "smoothSpline", 
             spline_type = "ts",
             colors = TG_cols,
             CI = T,
             knot_lines = F,
             individual_trendlines = F,
             individual_paths = F,
             group_trendline_color = "endpoint",
             facet = F,
             endpoint ="trajectory_group") -> plot
  
  # Add p-val to plot
  plot <- plot + annotate("label",  
                          x = Inf, 
                          y = Inf,
                          label = paste("p=", 
                                        signif(pvals$p.intercept, digits = 2), 
                                        sep = ""), 
                          vjust = 1, 
                          hjust = 1, 
                          size = 5, 
                          label.r = unit(0, "pt")) +
    scale_y_continuous(expand = expansion(mult = c(0.1))) +
    labs(fill = "Trajectory Group",
         color = "Trajectory Group",
         y = "Recovery Factor") + 
    theme(text = element_text(size = 14))
  
  return(plot)
}

```

## Acute plots

```{r}

# Pull only acute data
  df <- clinical_spear %>% filter(partition == "acute")

# Generate plot
  plot <- generate_TG_Factor1_longitudinal_plot(df)

# Save plot
  timepoint <- "acute"
  filename <- paste(workDir, "/plot_objects/plots_for_figures/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)

```

## Convalescent plots

```{r}

# Pull only convalescent test data
  df <- clinical_spear %>% filter(partition == "test")

# Generate plot
  plot <- generate_TG_Factor1_longitudinal_plot(df)

# Save plot
  timepoint <- "convalescent"
  filename <- paste(workDir, "/plot_objects/additional_plots/", comparison, "_", plot_type, "_", timepoint, ".Rds", sep = "")

  saveRDS(plot, filename)

```
