---
title: "Spear factor plotting acute and convalescent with different models"
author: "Gisela Gabernet (gisela.gabernet@yale.edu)"
subtitle: PGX, SO, PMG, PPT, PPG, BCT
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
params:
  fold: "acute_all_convalescent_test"
  results_dir: "longitudinal_spear_factor_1_scores_acute_all_conv_test_two_models/"
  spear_object: "/scratch/data-fullcohort/spear_model/results/020624/032124_spearphysical_finalscores_allsamples.rds"
editor_options: 
  chunk_output_type: inline
---

```{r}
library(data.table)
#install.packages("RColorBrewer")
library(RColorBrewer)
library(ggplot2)
library(tidyverse)
#install.packages("psych")
library(psych)
library(dplyr)
# Additional libraries for mixed effect models
#BiocManager::install("qvalue")
library(qvalue)
library(lme4)
library(ordinal)
library(nlme)
library(gamm4)
library(mgcv)
library(ggsignif)
library(kableExtra)
library(corrplot)
library(ggeffects)
library(ggfortify)
library(grid)
library(pheatmap)
library(ggbeeswarm)
library(ggpubr)
library(ggplotify)


#install.packages("ggvenn")
library(ggvenn)

#install.packages("msigdbr")
library(msigdbr)

knitr::opts_chunk$set(warning = F, message=F)

print_kable <- function(table, kable_height = "500px", kable_width = "100%") {
  kable(table, digits=3) %>%
    kable_styling("striped") #%>%
    #scroll_box(height = kable_height, width = kable_width)
  
}

# save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
#    stopifnot(!missing(x))
#    stopifnot(!missing(filename))
#    pdf(filename, width=width, height=height)
#    grid::grid.newpage()
#    grid::grid.draw(x$gtable)
#    dev.off()
# }

```

# Running parameters

Running with parameters


# Explore Spear results

## Explore Factor correlation and variance explained

```{r fig.height=5, fig.width=10}
# Read object from file

# binomial w1 object
#spear_object <- readRDS("/scratch/data-fullcohort/spear_model/results/factor_scores/022224_spearw1.rds")

# object with all spear models but old promis physical
#spear_object <- readRDS("/scratch/data-fullcohort/spear_model/results/factor_scores/022224_allspearfactors.rds")

#object with new promis physical (some participants were omitted in last model)
#spear_object <- readRDS("/scratch/data-fullcohort/spear_model/results/factor_scores/031124_spearphysical_finalscores.rds")

spear_object <- readRDS(params$spear_object)

data_env_conv <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_convalescent_data_env_all_Feb_11_2024.RDS")
data_env_acute <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_all_Mar_23_2024.RDS")

results_dir <- params$results_dir
dir.create(results_dir)

figures_dir <- "~/repos/impacc-full-cohort/Analysis/convalescent-project/Spear/Figures_spear_factors_association_acute"
dir.create(figures_dir)


path_to_codebase <- "~/repos/impacc-full-cohort/Analysis/IMPACC_data_analysis_codebase.R"
source(path_to_codebase)

selected_factors_names <- "Factor1"

```


## Get clinical data

```{r}
clinical_data_conv <- data_env_conv$clinical_data
clinical_data_acute <- data_env_acute$clinical_data

clinical_data <- dplyr::bind_rows(clinical_data_conv, clinical_data_acute)

# Add additional symptom groups to metadata
symptom_groups <- read_csv("/scratch/clinical/current/impacc_conv_symptoms.csv", show_col_types = F)
PRO_scores <- read_csv("/scratch/clinical/current/PRO_scores_dates_updated_29Aug23.csv", show_col_types = F)

## Description of spear object
# /scratch/data-fullcohort/spear_model/results/020624/032124_spearphysical_finalscores_allsamples.rds
# The following factor scores are stored (separating event_ids into the following categories):
# -Train (all event_ids used to train the SPEAR model, each with a PROMIS Physical score (not in the object))
# -CV (all event_ids used to train the SPEAR model, each with a PROMIS Physical score, using CV score)
# -Test (all event_ids with a PROMIS Physical score NOT included in the SPEAR model training)
# -TrainNoPhysical_Overlap - those who would pertain to the training set (80%) but do not have a PROMIS physical score, but DO overlap at least once with 'Train' (DON'T combine with Test!)
# -TrainNoPhysical_NoOverlap - those who would pertain to the training set (80%) but do not have a PROMIS physical score, but do not overlap at all with 'Train' above (can be used as additional Test data as we discussed)
# -TestNoPhysical (same as above, but pertaining to the 20%)
# -Acute (factor scores for all acute event_ids)

if (params$fold == "acuteconvalescent_test") {
  factor_scores_conv <- rbind(spear_object$Test, 
                              spear_object$TestNoPhysical, spear_object$TrainNoPhysical_NoOverlap)
  participants_conv <- clinical_data %>% filter(event_id %in% rownames(factor_scores_conv)) %>% pull(participant_id)
  diff_ids <- setdiff(rownames(spear_object$Acute), rownames(factor_scores_conv))
  factor_scores_acute <- spear_object$Acute[diff_ids,]
  events_acute <- unique(clinical_data_acute %>% filter(participant_id %in% participants_conv) %>% pull(event_id))
  factor_scores_acute_test_train_nophy_nooverlap <- factor_scores_acute[which(rownames(factor_scores_acute) %in% events_acute),]
  factor_scores <- rbind(factor_scores_conv, factor_scores_acute_test_train_nophy_nooverlap)
} else if (params$fold == "acute_all_convalescent_test") {
  # Acute phase all event_ids, convalescent phase only event_ids corresponding to the extended test set
  factor_scores_conv <- rbind(spear_object$Test, 
                              spear_object$TestNoPhysical, spear_object$TrainNoPhysical_NoOverlap)
  participants_conv <- clinical_data %>% filter(event_id %in% rownames(factor_scores_conv)) %>% pull(participant_id)
  event_id_conv <- rownames(factor_scores_conv)
  diff_ids <- setdiff(rownames(spear_object$Acute), rownames(factor_scores_conv))
  factor_scores_acute <- spear_object$Acute[diff_ids,]
  event_id_acute <- rownames(factor_scores_acute)
  factor_scores <- rbind(factor_scores_conv,factor_scores_acute)
  
} else if (params$fold == "acuteconvalescent_train") {
  factor_scores_conv <- spear_object$CV
  participants_conv <- clinical_data %>% filter(event_id %in% rownames(factor_scores_conv)) %>% pull(participant_id)
  diff_ids <- setdiff(rownames(spear_object$Acute), rownames(factor_scores_conv))
  factor_scores_acute <- spear_object$Acute[diff_ids,]
  events_acute <- unique(clinical_data_acute %>% filter(participant_id %in% participants_conv) %>% pull(event_id))
  factor_scores_acute_sel <- factor_scores_acute[which(rownames(factor_scores_acute) %in% events_acute),]
  factor_scores <- rbind(factor_scores_conv, factor_scores_acute_sel)
} else {
  stop("Params fold unrecognized, it should be one of: acuteconvalescent_test, acuteconvalescent_train, acute_all_convalescent_test")
}

dim(factor_scores)
```

# Associate factors with outcomes

Here we will use a mixed linear model to associate factors with outcomes

## Prepare metadata

```{r prepare the metadata for mixed linear model analysis, echo= FALSE, warning= FALSE}

correlation_metadata <- clinical_data %>%
  dplyr::select(event_id, resp_status_v1, sex, admit_age, event_after_first_vaccine_dose, 
                race, enrollment_site, event_date,
                event_date_3weeks_after_dose, pro_group_labels, event_type, 
                discretized_admit_age_quantile, participant_id)  %>%
  dplyr::mutate(pro_2_groups = 
      case_match( pro_group_labels,
      "MIN" ~ "MIN",
      c("PHY", "COG", "MLT")  ~ "OTHER"
      )
  ) %>%
  dplyr::distinct() %>%
  filter(event_id %in% rownames(factor_scores))

clinical_subsets_all <- as.data.frame(correlation_metadata)
rownames(clinical_subsets_all) <- clinical_subsets_all$event_id


endpoint_coefficients_all <- data.frame(matrix(nrow=ncol(factor_scores), ncol = 0))
rownames(endpoint_coefficients_all) <- colnames(factor_scores)
```

Get factor scores

```{r }
spear_factors <- as.data.frame(factor_scores)
```

## Functions for association

```{r}
factor_final_name = "Recovery Factor"

pro_min_colors <- c("#639A21", "#781B6C")
names(pro_min_colors) <- c("MIN", "OTHER")

pro_min_labels <- c("MIN", "LC")
names(pro_min_labels) <- c("MIN", "OTHER")

#original
#pro_colors <- c("#639A21", "#6371AD", "#FFC20A", "#9C3418")
pro_colors <- c("#639A21", "#6371AD", "#FFC20A", "#7D2D17")
names(pro_colors) <- c("MIN", "COG", "PHY", "MLT")

ordered_visits <- c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6", "Visit 7", "Visit 8", "Visit 9", "Visit 10")
acute_visits <- c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6")

scaleFUN <- function(x) sprintf("%.2e",x)
violin_plot_visits <- function(data, colors, limits=NULL, ncol=5) {
  p <- ggplot(data, aes(x=endpoints, y=value, fill=endpoints)) +
    geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
    facet_wrap(~Factor, drop = T, ncol = ncol, scales = "free_y") +
    theme_bw() + theme(legend.position = "none")
  if(!(is.null(limits))){
    p <- p + scale_y_continuous(limits = limits)
  } else {
    p <- p + scale_y_continuous(labels = scaleFUN)
  }
  if(!(missing(colors))){
    p <- p + scale_color_manual(values=colors)
    p <- p + scale_fill_manual(values=colors)
  }
  return(p)
}

pairwise_association_visit <- function(clinical_subsets_all, 
                                       spear_factors, visit_id, 
                                       selected_factors_names, 
                                       endpoint_var = "pro_2_groups",
                                       ordered_endpoint_var = c("MIN","OTHER"),
                                       colors = pro_min_colors
                                       ){
  cur_visit <- visit_id

  visit_row_ids = rownames(clinical_subsets_all[clinical_subsets_all$event_type == cur_visit,])
  
  # Use the visit7_row_ids to only get samples from 'Visit 7':
  factors_filtered = spear_factors[visit_row_ids , selected_factors_names, drop=F]
  clinical_subsets = clinical_subsets_all[visit_row_ids ,]
  
  data_use = factors_filtered
  
  # Add the respiratory_status_day14 as $endpoints
  data_use$endpoints = factor(clinical_subsets[[endpoint_var]], levels=ordered_endpoint_var)
  
  # Add the enrollment_site as $sites
  data_use$sites = clinical_subsets$enrollment_site
  data_use$age =  clinical_subsets$discretized_admit_age_quantile
  data_use$sex =  clinical_subsets$sex
  
  # Add the $control, needs to be different for each sample
  # required for clmm (always requires some random effect)
  data_use$control =  factor(1:nrow(data_use))
  
  # Set module names:
  factor_names = colnames(factors_filtered)
  
  # Only use samples that have valid endpoints (not NA/missing)
  data_use = data_use[!is.na(data_use$endpoints),]

  for(j in 1:length(factor_names)){
    #print(j)
    my.formula0 = paste0(factor_names[j],"~ (1|sites)+age+sex")
    my.formula1 = paste0(factor_names[j],"~ endpoints + (1|sites)+age+sex")
    if(j == 1){
      tmp = mixed_pairwise(my.formula0, my.formula1, data_use)
      res_table_pairwise = data.frame(matrix(0, ncol = length(tmp$res_table), nrow = length(factor_names)))
      coef_endpoint_pairwise = data.frame(matrix(0, ncol= length(tmp$coef_endpoint), nrow=length(factor_names)))
      res_table_pairwise[j,] = tmp$res_table
      coef_endpoint_pairwise[j,] = tmp$coef_endpoint
      colnames(res_table_pairwise) = names(tmp$res_table)
      colnames(coef_endpoint_pairwise) = names(tmp$coef_endpoint)
    }else{
      tmp = mixed_pairwise(my.formula0, my.formula1, data_use)
      res_table_pairwise[j,] = tmp$res_table
      coef_endpoint_pairwise[j,] = tmp$coef_endpoint
    }
  }
  
  if (length(ordered_endpoint_var) > 2) {
      coef_endpoint_pairwise <- coef_endpoint_pairwise[,1:3]
      res_table_pairwise <- res_table_pairwise[,1:3]
  }
  coef_endpoint_pairwise$factor <- factor_names
  coef_endpoint_pairwise$visit <- cur_visit
  
  if (nrow(res_table_pairwise)>1) {
    res_table_pairwise_adj_df <- qvalue::qvalue(res_table_pairwise, fdr.level = 0.05, pi0=1)$qvalues
  } else {
    res_table_pairwise_adj_df <- res_table_pairwise
  }

  res_table_pairwise_adj_df$factor <- factor_names
  res_table_pairwise_adj_df$visit <- cur_visit
  
  plot_df <- data_use %>% dplyr::select(starts_with("Factor"), endpoints) %>%
            pivot_longer(!endpoints, names_to = "Factor", values_to = "value")
  plot_df$Factor <- factor(plot_df$Factor, levels=selected_factors_names)
  
  p <- violin_plot_visits(plot_df, colors = colors) + ggtitle(visit_id)
  
  res_list <- list(coefs_df = coef_endpoint_pairwise, 
                   res_df = res_table_pairwise_adj_df,
                   violin_plot = p)
  
  return(res_list)
}

global_mixed_pairwise <- function(clinical_subsets_all, 
                                   spear_factors,
                                   selected_factor_name, 
                                   endpoint_var = "pro_2_groups",
                                   ordered_endpoint_var = c("MIN","OTHER")
                                       ){
  
  if(length(selected_factor_name) > 1) {
    stop("Global mixed pairwise can only be calculated one factor at a time!")
  }

  # Get selected factors
  factors_filtered = spear_factors[, selected_factor_name, drop=F]

  data_use = factors_filtered %>%
    rownames_to_column("event_id") %>%
    left_join(clinical_subsets_all, by="event_id")
  
  # Add the endpoint variable as endpoints
  data_use$endpoints = factor(data_use[[endpoint_var]], levels=ordered_endpoint_var)
  
  # Add the $control, needs to be different for each sample
  # required for clmm (always requires some random effect)
  data_use$control =  factor(1:nrow(data_use))
  
  # Only use samples that have valid endpoints (not NA/missing)
  data_use = data_use[!is.na(data_use$endpoints),]

  data_use$value <- data_use[[selected_factor_name]]
  
  # Generate formulas for mixed_pairwise function
  var_list <- c("discretized_admit_age_quantile",
                "sex",
                "event_type",
                "(1|enrollment_site/participant_id)")
  terms <- paste(var_list, collapse = "+")
  formula.0 <- formula(paste("value~", terms, sep = ""))
  var_list <- append(var_list, "endpoints")
  terms <- paste(var_list, collapse = "+")
  formula.1 <- formula(paste("value~", terms, sep = ""))
  res_table <- mixed_pairwise(formula.0, formula.1, data_use)
  
  return(res_table$res_table)
}


```


## Functions for better pheatmap

```{r}
## Edit body of pheatmap:::draw_colnames, customizing it to your liking
draw_colnames_45 <- function (coln, ...) {
    m = length(coln)
    x = (1:m)/m - 1/2/m
    grid.text(coln, x = x, y = unit(0.96, "npc"), vjust = .5, 
        hjust = 1, rot = 45, gp = gpar(...)) ## Was 'hjust=0' and 'rot=270'
}

## For pheatmap_1.0.8 and later:
draw_colnames_45 <- function (coln, gaps, ...) {
    coord = pheatmap:::find_coordinates(length(coln), gaps)
    x = coord$coord - 0.5 * coord$size
    res = textGrob(coln, x = x, y = unit(1, "npc") - unit(3,"bigpts"), vjust = 0.5, hjust = 1, rot = 45, gp = gpar(...))
    return(res)}

## 'Overwrite' default draw_colnames with your own version 
assignInNamespace(x="draw_colnames", value="draw_colnames_45",
ns=asNamespace("pheatmap"))

p_val_only_palette = colorRampPalette(brewer.pal(9,"Reds"))(10)
```

## Association MIN vs OTHER for each visit


We treat each factor expression as the covariate and consider its association with clinical endpoints 
(pro_groups) in the presence of potential heterogeneity across different clinical sites and phases, 
which is modeled as a random effect.


We perform pairwise comparisons.


```{r fig.width=2, fig.height=2}
endpoint_coefficients_all <- data.frame()
res_table_pairwise_adj_all <- data.frame()

for (visit_id in ordered_visits){
  print(paste("Computing associations for visit:", visit_id))
  res_list <- pairwise_association_visit(clinical_subsets_all=clinical_subsets_all, 
                                          spear_factors=spear_factors, 
                                          visit_id=visit_id, 
                                          selected_factors_names = selected_factors_names,
                                          endpoint_var = "pro_2_groups")

  endpoint_coefficients_all <- bind_rows(endpoint_coefficients_all,
                                         res_list$coefs_df)
  res_table_pairwise_adj_all <- bind_rows(res_table_pairwise_adj_all, 
                                          res_list$res_df)
}


```

### Adj p-val all visits

```{r fig.height=2.5, fig.width=10}

res_table_pairwise_adj_all_plot <- res_table_pairwise_adj_all %>%
                                      pivot_wider(names_from = factor,
                                                  values_from = "MIN|OTHER") %>%
                                      column_to_rownames("visit")

print_kable(res_table_pairwise_adj_all_plot)
write.csv(res_table_pairwise_adj_all_plot, paste0(results_dir, "Adj_pval_all_visits_min_vs_other.csv"))
```

### Log p-value plot with coefficient sign

```{r fig.asp=0.8, fig.width=4}
endpoint_coefficient_plot <- endpoint_coefficients_all %>%
                              mutate(sign = if_else(`MIN|OTHER`>0, 1, -1)) %>%
                              dplyr::select(factor, visit, sign)

df_plot <- res_table_pairwise_adj_all %>%
              mutate(log10_adjpval = -log10(`MIN|OTHER`)) %>%
              inner_join(endpoint_coefficient_plot, by=c("factor", "visit")) %>%
              mutate(signed_adjp = log10_adjpval * sign)

df_plot$factor <- factor(df_plot$factor, levels = selected_factors_names)
df_plot$visit <- factor(df_plot$visit, levels = unique(df_plot$visit))
df_plot <- df_plot %>% 
            mutate(annot =
                case_when(
                 abs(log10_adjpval) >= -log10(0.001) ~ "***",
                 abs(log10_adjpval) >= -log10(0.01) ~ "**",
                 abs(log10_adjpval) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 ))

thr <- 0.05
plot_title <- "Adj p-value factor association"


convert_df <- df_plot %>% 
              dplyr::select(factor, visit, signed_adjp) %>%
              pivot_wider(id_cols = c(visit), 
                          names_from = factor, 
                          values_from = signed_adjp) %>%
              column_to_rownames("visit")

convert_annot <- df_plot %>%
              dplyr::select(factor, visit, annot) %>%
              pivot_wider(id_cols = c(visit),
                          names_from = factor,
                          values_from = annot) %>%
              column_to_rownames("visit")

mat <- as.matrix(convert_df)
mat_annot <- as.matrix(convert_annot)

palette_length <- 50
abs_max <- max(abs(min(mat)), abs(max(mat)), 1.3)
breaks_list <- c(seq(-abs_max, 0, length.out=ceiling(palette_length/2) + 1), 
              seq(abs_max/palette_length, abs_max, length.out=floor(palette_length/2)))

pdf(file = paste0(results_dir, "Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 6)
p <- pheatmap(mat,
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = mat_annot,
         number_color = "black",
         fontsize_number = 10
         )
p
dev.off()

p
#save_pheatmap_pdf(p, paste0(results_dir,"Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 6)

```

### p-val only acute

```{r fig.asp=0.8, fig.width=4}
# Plotting only acute visits
convert_df <- df_plot %>% 
              dplyr::select(factor, visit, signed_adjp) %>%
              dplyr::filter(visit %in% acute_visits) %>%
              pivot_wider(id_cols = c(visit), 
                          names_from = factor, 
                          values_from = signed_adjp) %>%
              column_to_rownames("visit")

convert_annot <- df_plot %>%
              dplyr::select(factor, visit, annot) %>%
              dplyr::filter(visit %in% acute_visits) %>%
              pivot_wider(id_cols = c(visit),
                          names_from = factor,
                          values_from = annot) %>%
              column_to_rownames("visit")


mat <- as.matrix(convert_df)
mat_annot <- as.matrix(convert_annot)

palette_length <- 50
abs_max <- max(abs(min(mat)), abs(max(mat)), 1.3)
breaks_list <- c(seq(-abs_max, 0, length.out=ceiling(palette_length/2) + 1), 
              seq(abs_max/palette_length, abs_max, length.out=floor(palette_length/2)))

pdf(file = paste0(results_dir, "Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 6)
p <- pheatmap(mat,
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = mat_annot,
         number_color = "black",
         fontsize_number = 10
         )
p
dev.off()

p



```



### Violin plots all visits

```{r }

factors_filtered = spear_factors[, selected_factors_names, drop=F]
clinical_subsets = clinical_subsets_all[rownames(spear_factors),]
  

data_use = factors_filtered 

data_use$endpoints = as.factor(clinical_subsets$pro_2_groups)

# Add the enrollment_site as $sites
data_use$sites = clinical_subsets$enrollment_site
data_use$age =  clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex
data_use$visit = clinical_subsets$event_type
data_use$phase = clinical_subsets$phase

# Add the $control, needs to be different for each sample
# required for clmm (always requires some random effect)
data_use$control =  factor(1:nrow(data_use))

# Set module names:
factor_names = colnames(factors_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$endpoints),]
```

Violin plots for all factors, values from all visits

```{r fig.width=2, fig.height=2}

plot_df <- data_use %>% dplyr::select(starts_with("Factor"), endpoints) %>%
            pivot_longer(!endpoints, names_to = "Factor", values_to = "value")
plot_df$Factor <- factor(plot_df$Factor, levels=selected_factors_names)

p <- ggplot(plot_df, aes(x=endpoints, y=value, fill=endpoints)) +
  geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  scale_y_continuous(labels = scaleFUN) +
  theme_bw() + theme(legend.position = "none") +
  scale_color_manual(values=pro_min_colors) +
  scale_fill_manual(values=pro_min_colors) +
  facet_wrap(~Factor, drop = T, scales = "free_y", ncol = 4)
p
```

Violin plots for all factors, values per visit

```{r fig.height=2, fig.width=8}

plot_df <- data_use %>% dplyr::select(starts_with("Factor"), endpoints, visit) %>%
            pivot_longer(!c(endpoints,visit), names_to = "Factor", values_to = "value")
plot_df$Factor <- factor(plot_df$Factor, levels=selected_factors_names)
plot_df$visit <- factor(plot_df$visit, levels=ordered_visits)

for (factor in selected_factors_names) {
  plot_df_fact <- plot_df %>% filter(Factor == factor)
  
  p <- ggplot(plot_df_fact, aes(x=visit, y=value, fill=endpoints)) +
    ggbeeswarm::geom_quasirandom(aes(fill=endpoints),shape=21,width=.15, dodge.width = 0.75) +
    geom_boxplot(alpha=0.6, color="black") +
    theme_bw() + 
    scale_color_manual(values=pro_min_colors) +
    scale_fill_manual(values=pro_min_colors) +
    scale_y_continuous(limits = c(-1.5,1.5)) +
    #facet_wrap(~visit, drop = T, ncol = 4) +
    labs( x="", y=factor_final_name)
  print(p)
}

```


### Boxplot per visit
Violin plots for all factors, values per visit, only acute visits

```{r fig.height=3, fig.width=6}

plot_df <- data_use %>% dplyr::select(starts_with("Factor"), endpoints, visit) %>%
            pivot_longer(!c(endpoints,visit), names_to = "Factor", values_to = "value") %>%
            dplyr::filter(visit %in% acute_visits)
plot_df$Factor <- factor(plot_df$Factor, levels=selected_factors_names)
plot_df$visit <- factor(plot_df$visit, levels=acute_visits)

for (factor in selected_factors_names) {
  plot_df_fact <- plot_df %>% filter(Factor == factor)
  
  p <- ggplot(plot_df_fact, aes(x=visit, y=value, fill=endpoints)) +
    ggbeeswarm::geom_quasirandom(aes(fill=endpoints),shape=21,width=.15, dodge.width = 0.75) +
    geom_boxplot(alpha=0.6, color="black") +
    theme_bw() + 
    scale_color_manual(values=pro_min_colors, label=pro_min_labels) +
    scale_y_continuous(limits = c(-1.5,1.5)) +
    scale_fill_manual(values=pro_min_colors, label=pro_min_labels) +
    #facet_wrap(~visit, drop = T, ncol = 4) +
    labs( x="", y=factor_final_name, fill="", color="")
  print(p)
}


```

### Boxplot per visit with pval
Violin plots for all factors, values per visit, only acute visits

```{r}
pval_df <- global_mixed_pairwise(clinical_subsets_all=clinical_subsets_all %>% filter(event_type %in% acute_visits),
                                          spear_factors=spear_factors,
                                          selected_factor_name = "Factor1",
                                          endpoint_var = "pro_2_groups")
pval_title <- paste0("MIN vs LC p=", format(pval_df,scientific=F, digits=2))
```

```{r fig.height=3, fig.width=6}

for (factor in selected_factors_names) {
  plot_df_fact <- plot_df %>% filter(Factor == factor)
  
  format_corrected_pval <- res_table_pairwise_adj_all %>% 
    dplyr::filter(visit %in% acute_visits) %>%
    pivot_longer(cols=-c(factor,visit), names_to = "comparison", values_to = "adj_pval_sex_age_corrected") %>%
    dplyr::mutate(group1 = str_extract(comparison, "[^|]+"),
                  group2 = str_extract(comparison, "[^|]+$")) %>%
    dplyr::select(-c(comparison,factor))
  
  stat.test <- plot_df_fact %>%
    group_by(visit) %>%
    wilcox_test(value ~ endpoints) %>%
    adjust_pvalue(method = "hochberg") %>%
    left_join(format_corrected_pval, by=join_by(visit,group1,group2)) %>%
    add_significance("adj_pval_sex_age_corrected") %>%
    add_xy_position(x = "visit", dodge = 0.75) %>%
    dplyr::filter(adj_pval_sex_age_corrected < 0.05)
  
  p <- ggplot(plot_df_fact, aes(x=visit, y=value)) +
    ggbeeswarm::geom_quasirandom(aes(fill=endpoints),shape=21,width=.15, dodge.width = 0.75) +
    geom_boxplot(aes(fill=endpoints), alpha=0.6, color="black", outlier.shape = NA) +
    theme_bw() + 
    scale_color_manual(values=pro_min_colors, labels = pro_min_labels) +
    scale_fill_manual(values=pro_min_colors, labels = pro_min_labels) +
    #facet_wrap(~visit, drop = T, ncol = 4) +
    labs( x="", y=factor_final_name, fill="") +
    annotate("label",  x=Inf, y =Inf,label = pval_title, vjust=1, hjust=1, size = 3, label.r = unit(0, "pt")) +
    scale_y_continuous(limits = c(-1.1,1.8))
                      # expand = expansion(mult = c(0.1, 0.2)))
    #theme(plot.title = element_text(hjust = 0.5, size =8))
  
  print(p)
  
  if (nrow(stat.test) > 0){
    p_pval <- p + stat_pvalue_manual(stat.test, label = "adj_pval_sex_age_corrected.signif", y.position=1.3, tip.length = 0) +
      scale_y_continuous(limits = c(-1.1,1.8)) 
                         #expand = expansion(mult = c(0.1, 0.1)))
    print(p_pval)
  }
  
}

acute_boxplot_visits_minother_globalonly <- p
saveRDS(acute_boxplot_visits_minother_globalonly, file.path(figures_dir,"acute_boxplot_visits_minother_pval_globalonly.rds"))

acute_boxplot_visits_minother <- p_pval
saveRDS(acute_boxplot_visits_minother, file.path(figures_dir,"acute_boxplot_visits_minother_pval.rds"))

```

## Association MIN vs all other groups

```{r fig.width=2, fig.height=2}
endpoint_coefficients_all <- data.frame()
res_table_pairwise_adj_all <- data.frame()

for (visit_id in ordered_visits){
  print(paste("Computing associations for visit:", visit_id))
  res_list <- pairwise_association_visit(clinical_subsets_all=clinical_subsets_all, 
                                          spear_factors=spear_factors, 
                                          visit_id=visit_id, 
                                          selected_factors_names = selected_factors_names,
                                          endpoint_var = "pro_group_labels",
                                          ordered_endpoint_var = names(pro_colors),
                                          colors = pro_colors
                                          )

  endpoint_coefficients_all <- bind_rows(endpoint_coefficients_all,
                                         res_list$coefs_df)
  res_table_pairwise_adj_all <- bind_rows(res_table_pairwise_adj_all, 
                                          res_list$res_df)
}


```

### Adj p-val all visits

```{r pairwise_heatmap, fig.height=2.5, fig.width=20}

library(pheatmap)

res_table_pairwise_adj_all_plot <- res_table_pairwise_adj_all %>% 
                                    pivot_longer(cols=starts_with("MIN"),
                                                 values_to="padj",
                                                 names_to="comp") %>%
                                    mutate(comp_visit = paste0(comp, "_", visit)) %>%
                                    dplyr::select(-c(comp)) %>%
                                    pivot_wider(id_cols = c(comp_visit),
                                                names_from = factor,
                                                values_from = padj) %>%
                                    column_to_rownames("comp_visit")
print_kable(res_table_pairwise_adj_all_plot)
write.csv(res_table_pairwise_adj_all_plot, paste0(results_dir, "Adj_pval_all_visits_all_pro_groups.csv"))
```





### Log p-value plot with coefficient sign
```{r fig.height=4, fig.width=4}
endpoint_coefficient_plot <- endpoint_coefficients_all %>%
                              pivot_longer(-c(factor,visit), 
                                           names_to = "comparison", 
                                           values_to = "coefficient") %>% 
                              mutate(sign = if_else(coefficient>0, 1, -1)) %>%
                              dplyr::select(factor, visit, comparison, sign)

df_plot <- res_table_pairwise_adj_all %>%
              pivot_longer(-c(factor,visit),
                           names_to = "comparison",
                           values_to = "adj_pval") %>%
              mutate(log10_adjpval = -log10(adj_pval)) %>%
              inner_join(endpoint_coefficient_plot, 
                         by=c("factor", "visit", "comparison")) %>%
              mutate(comp_visit = paste0(comparison, "_", visit)) %>%
              mutate(signed_adjp = log10_adjpval * sign)

df_plot$factor <- factor(df_plot$factor, levels = selected_factors_names)
df_plot$comp_visit <- factor(df_plot$comp_visit, levels = unique(df_plot$comp_visit))
df_plot <- df_plot %>% 
            mutate(annot =
                case_when(
                 abs(log10_adjpval) >= -log10(0.001) ~ "***",
                 abs(log10_adjpval) >= -log10(0.01) ~ "**",
                 abs(log10_adjpval) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 ))

thr <- 0.05
plot_title <- "Adj p-value factor association"

convert_df <- df_plot %>% 
              dplyr::select(comparison, visit, signed_adjp) %>%
              pivot_wider(id_cols = c(visit), 
                          names_from = comparison, 
                          values_from = signed_adjp) %>%
              column_to_rownames("visit")

convert_annot <- df_plot %>%
              dplyr::select(comparison, visit, annot) %>%
              pivot_wider(id_cols = c(visit),
                          names_from = comparison,
                          values_from = annot) %>%
              column_to_rownames("visit")

mat <- as.matrix(convert_df)
mat_annot <- as.matrix(convert_annot)

palette_length <- 50
abs_max <- max(abs(min(mat)), abs(max(mat)), 1.3)
breaks_list <- c(seq(-abs_max, 0, length.out=ceiling(palette_length/2) + 1), 
              seq(abs_max/palette_length, abs_max, length.out=floor(palette_length/2)))

pdf(file = paste0(results_dir, "Adj_pval_all_visits_pro_groups_plot.pdf"), width = 12, height = 10)
p <- pheatmap(mat,
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = mat_annot,
         number_color = "black",
         fontsize_number = 10,
         )
p
dev.off()

p

#save_pheatmap_pdf(p, paste0(results_dir,"Adj_pval_all_visits_pro_groups_plot.pdf"), width = 12, height = 6)

```

### Log p-val acute only

```{r fig.height=4, fig.width=4}
endpoint_coefficient_plot <- endpoint_coefficients_all %>%
                              pivot_longer(-c(factor,visit), 
                                           names_to = "comparison", 
                                           values_to = "coefficient") %>% 
                              mutate(sign = if_else(coefficient>0, 1, -1)) %>%
                              dplyr::select(factor, visit, comparison, sign)

df_plot <- res_table_pairwise_adj_all %>%
              pivot_longer(-c(factor,visit),
                           names_to = "comparison",
                           values_to = "adj_pval") %>%
              mutate(log10_adjpval = -log10(adj_pval)) %>%
              inner_join(endpoint_coefficient_plot, 
                         by=c("factor", "visit", "comparison")) %>%
              dplyr::filter(visit %in% acute_visits) %>%
              mutate(comp_visit = paste0(comparison, "_", visit)) %>%
              mutate(signed_adjp = log10_adjpval * sign)

df_plot$factor <- factor(df_plot$factor, levels = selected_factors_names)
df_plot$comp_visit <- factor(df_plot$comp_visit, levels = unique(df_plot$comp_visit))
df_plot <- df_plot %>% 
            mutate(annot =
                case_when(
                 abs(log10_adjpval) >= -log10(0.001) ~ "***",
                 abs(log10_adjpval) >= -log10(0.01) ~ "**",
                 abs(log10_adjpval) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 ))

thr <- 0.05
plot_title <- "Adj p-value factor association"

convert_df <- df_plot %>% 
              dplyr::select(comparison, visit, signed_adjp) %>%
              pivot_wider(id_cols = c(visit), 
                          names_from = comparison, 
                          values_from = signed_adjp) %>%
              column_to_rownames("visit")

convert_annot <- df_plot %>%
              dplyr::select(comparison, visit, annot) %>%
              pivot_wider(id_cols = c(visit),
                          names_from = comparison,
                          values_from = annot) %>%
              column_to_rownames("visit")

mat <- as.matrix(convert_df)
mat_annot <- as.matrix(convert_annot)

palette_length <- 50
abs_max <- max(abs(min(mat)), abs(max(mat)), 1.3)
breaks_list <- c(seq(-abs_max, 0, length.out=ceiling(palette_length/2) + 1), 
              seq(abs_max/palette_length, abs_max, length.out=floor(palette_length/2)))

pdf(file = paste0(results_dir, "Adj_pval_all_visits_pro_groups_plot.pdf"), width = 12, height = 10)
p <- pheatmap(mat,
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = mat_annot,
         number_color = "black",
         fontsize_number = 10,
         )
p
dev.off()

p

```

### Violin plots all visits

```{r }

factors_filtered = spear_factors[, selected_factors_names, drop=F]
clinical_subsets = clinical_subsets_all[rownames(spear_factors),]
  

data_use = factors_filtered 

data_use$endpoints = factor(clinical_subsets$pro_group_labels, levels = names(pro_colors))

# Add the enrollment_site as $sites
data_use$sites = clinical_subsets$enrollment_site
data_use$age =  clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex
data_use$visit = clinical_subsets$event_type

# Add the $control, needs to be different for each sample
# required for clmm (always requires some random effect)
data_use$control =  factor(1:nrow(data_use))

# Set module names:
factor_names = colnames(factors_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$endpoints),]
```

Violin plots for all factors, values from all visits

```{r fig.height=2, fig.width=3}

plot_df <- data_use %>% dplyr::select(starts_with("Factor"), endpoints) %>%
            pivot_longer(!endpoints, names_to = "Factor", values_to = "value")
plot_df$Factor <- factor(plot_df$Factor, levels=selected_factors_names)

p <- ggplot(plot_df, aes(x=endpoints, y=value, fill=endpoints)) +
  geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  theme_bw() + theme(legend.position = "none") +
  scale_color_manual(values=pro_colors) +
  scale_fill_manual(values=pro_colors) +
  facet_wrap(~Factor, drop = T, scales = "free_y", ncol = 4)
p
```

Violin plots for all factors, values per visit

```{r fig.height=2, fig.width=10}

plot_df <- data_use %>% dplyr::select(starts_with("Factor"), endpoints, visit) %>%
            pivot_longer(!c(endpoints,visit), names_to = "Factor", values_to = "value")
plot_df$Factor <- factor(plot_df$Factor, levels=selected_factors_names)
plot_df$visit <- factor(plot_df$visit, levels=ordered_visits)

for (factor in selected_factors_names) {
  plot_df_fact <- plot_df %>% filter(Factor == factor)
  p <- ggplot(plot_df_fact, aes(x=visit, y=value, fill=endpoints)) +
    ggbeeswarm::geom_quasirandom(aes(fill=endpoints),shape=21,width=.1, dodge.width = 0.75) +
    geom_boxplot(alpha=0.6, color="black") +
    theme_bw() + #theme(legend.position = "none") +
    scale_y_continuous(limits = c(-1.5,1.5)) +
    scale_color_manual(values=pro_colors) +
    scale_fill_manual(values=pro_colors) +
    labs(x="", y=factor_final_name, fill="", color="")#+
    #facet_wrap(~Factor, drop = T, scales = "free_y", ncol = 4)
  print(p)
}

```
### Boxplots per visit
Violin plots for all factors, values per visit, only acute visits

```{r fig.height=2, fig.width=6}

plot_df <- data_use %>% dplyr::select(starts_with("Factor"), endpoints, visit) %>%
            pivot_longer(!c(endpoints,visit), names_to = "Factor", values_to = "value") %>%
            dplyr::filter(visit %in% acute_visits)
plot_df$Factor <- factor(plot_df$Factor, levels=selected_factors_names)
plot_df$visit <- factor(plot_df$visit, levels=acute_visits)

for (factor in selected_factors_names) {
  plot_df_fact <- plot_df %>% filter(Factor == factor)
  p <- ggplot(plot_df_fact, aes(x=visit, y=value, fill=endpoints)) +
    ggbeeswarm::geom_quasirandom(aes(fill=endpoints),shape=21,width=.1, dodge.width = 0.75) +
    geom_boxplot(alpha=0.6, color="black") +
    theme_bw() + #theme(legend.position = "none") +
    scale_y_continuous(limits = c(-1.5,1.5)) +
    scale_color_manual(values=pro_colors) +
    scale_fill_manual(values=pro_colors) +
    labs(x="", y=factor_final_name, fill="", color="")#+
    #facet_wrap(~Factor, drop = T, scales = "free_y", ncol = 4)
  print(p)
}

```

### Boxplots per visit with pval

```{r}
pval_df <- global_mixed_pairwise(clinical_subsets_all=clinical_subsets_all %>% filter(event_type %in% acute_visits),
                                          spear_factors=spear_factors,
                                          selected_factor_name = "Factor1",
                                          endpoint_var = "pro_group_labels",
                                          ordered_endpoint_var = names(pro_colors))

pval_title = paste0("MIN vs COG p=", format(pval_df[1],scientific=F, digits=1),
               "\nMIN vs PHY p=", format(pval_df[2],scientific=F, digits=1),
               "\nMIN vs MLT p=", format(pval_df[3],scientific=F, digits=1))
pval_title
```

```{r fig.height=3, fig.width=6}

for (factor in selected_factors_names) {
  plot_df_fact <- plot_df %>% filter(Factor == factor)
  
  format_corrected_pval <- res_table_pairwise_adj_all %>% 
    pivot_longer(cols=-c(factor,visit), names_to = "comparison", values_to = "adj_pval_sex_age_corrected") %>%
    dplyr::mutate(group1 = str_extract(comparison, "[^|]+"),
                  group2 = str_extract(comparison, "[^|]+$")) %>%
    dplyr::select(-c(comparison,factor))
  
  stat.test <- plot_df_fact %>%
    group_by(visit) %>%
    wilcox_test(value ~ endpoints) %>%
    adjust_pvalue(method = "hochberg") %>%
    left_join(format_corrected_pval, by=join_by(visit,group1,group2)) %>%
    add_significance("adj_pval_sex_age_corrected") %>%
    add_xy_position(x = "visit", dodge = 0.75) %>%
    dplyr::filter(group1 == "MIN") %>%
    dplyr::filter(adj_pval_sex_age_corrected < 0.05) #%>%
    #dplyr::mutate(y.position = if_else(y.position > 2, 1, y.position))

  
  p <- ggplot(plot_df_fact, aes(x=visit, y=value)) +
    ggbeeswarm::geom_quasirandom(aes(fill=endpoints),shape=21,width=.10, dodge.width = 0.75) +
    geom_boxplot(aes(fill=endpoints),alpha=0.6, color="black",outlier.shape=NA) +
    theme_bw() + 
    scale_color_manual(values=pro_colors) +
    scale_fill_manual(values=pro_colors) +
    #scale_y_continuous(expand = expansion(mult = c(0.1, 0.2))) +
    labs( x="", y=factor_final_name, fill="") +
    annotate("label",  x=Inf, y =Inf, label = pval_title, vjust=1, hjust=1, size = 3, label.r = unit(0, "pt")) +
    scale_y_continuous(limits = c(-1.1,1.8))#,expand = expansion(mult = c(0.1, 0.3)),
                                     
  print(p)

  if (nrow(stat.test) > 0){
    p_pval <- p + stat_pvalue_manual(stat.test, label = "adj_pval_sex_age_corrected.signif", tip.length = 0, label.size = 3,
                                     step.increase=0.00001, step.group.by = "visit") + 
      scale_y_continuous(limits = c(-1.1,1.8), 
                         expand = expansion(mult = c(0.1, 0.3)))
  }
  print(p_pval)
}

acute_boxplot_visits_progroups_globalonly <- p
saveRDS(acute_boxplot_visits_progroups_globalonly, file.path(figures_dir,"acute_boxplot_visits_progroups_pval_globalonly.rds"))

acute_boxplot_visits_progroups <- p_pval
saveRDS(acute_boxplot_visits_progroups, file.path(figures_dir,"acute_boxplot_visits_progroups_pval.rds"))

```

## Longitudinal analysis

With spear factors

```{r data preparation, echo=TRUE}
clinical_subsets <- correlation_metadata
rownames(clinical_subsets) <- clinical_subsets$event_id

row_ids <-rownames(spear_factors)
clinical_subsets <- clinical_subsets[row_ids,]


modules_filtered = spear_factors %>% dplyr::select(all_of(selected_factors_names))

####%%%%%%%% spear Input %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

inputDF <- modules_filtered %>%
  rownames_to_column(var = "event_id") %>%
  merge(y = clinical_subsets,
        by = "event_id") %>%
  mutate(outcome = pro_2_groups) %>%
  pivot_longer(cols = -c("event_id", "event_date", "participant_id",
                        "enrollment_site", "pro_group_labels", "pro_2_groups", "event_type", "outcome",
                        "sex", "admit_age", "discretized_admit_age_quantile", "resp_status_v1",
                        "event_after_first_vaccine_dose", "event_date_3weeks_after_dose","race")) %>%
  dplyr::filter(!is.na(value) & !is.na(pro_2_groups))
inputDF$names <- factor(inputDF$name, levels=selected_factors_names)


## Ensure proper variables are factors
inputDF$pro_group_labels <- as.factor(inputDF$pro_group_labels)
inputDF$pro_2_groups <- as.factor(inputDF$pro_2_groups)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$race <- as.factor(inputDF$race)
inputDF$outcome <- as.factor(inputDF$outcome)
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)
```


### PRO 2 groups acute model

```{r, linear_model_loop, fig.width = 3, fig.height=5}
## Can use inputDF from either the variable of WGCNA module forms in the previous chunk
inputDF_acute <- inputDF %>% filter(event_type %in% paste("Visit",c(1:6)),
                            event_date <= 50)

gamm_model_loop_acute <- model_loop(inputDF_acute, age_sex = T, 
                              model_type = "smoothSpline", 
                              spline_type = "ts", 
                              endpoint = "pro_2_groups")

gamm_model_loop_plot_acute <- gamm_model_loop_acute %>% 
                        dplyr::select(adjp.slope, adjp.intercept)

print_kable(gamm_model_loop_plot_acute[selected_factors_names,])
write.csv(gamm_model_loop_plot_acute[selected_factors_names,], paste0(results_dir, "global_p_value_longitudinal_pro_2_groups_acute.csv"))
```

```{r, longitudinal_model_loop_pheatmap, fig.width = 3, fig.height=3}
breaklist = seq(0,3,0.3)
pdf(file = paste0(results_dir, "global_p_value_longitudinal_min_vs_other_acute.pdf"), width = 3, height = 6)

mat_log10pval <- -log10(gamm_model_loop_plot_acute[selected_factors_names,])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "white",
         cluster_rows = F, 
         cluster_cols = F, 
         color = p_val_only_palette,
         breaks = breaklist)
p1
dev.off()

p1

pval_min_other_acute_value <- 10^(-mat_log10pval$adjp.intercept[[1]])
pval_annot <- paste0("MIN vs LC p=",format(pval_min_other_acute_value,scientific=T, digits=2))
pval_annot
```

### PRO 2 groups acute model plot

```{r, fig.width=6, fig.height=3}
plots <- list()
plots_model <- list()
for (feature_name in selected_factors_names) {
  plotDF <- inputDF_acute %>% filter(name == feature_name)
  p1_acute <- plot_model(plotDF, 
                     gamm_model_loop_acute,
                     model_type = "smoothSpline",
                     colors = pro_min_colors,
                     spline_type = "ts",
                     knot_lines = F,
                     group_trendline_color = "black",
                     CI = T,
                     endpoint = "pro_2_groups") +
              labs(y=factor_final_name) +
              scale_y_continuous(limits = c(-1.1,1.1)) +
              theme(strip.background = element_rect(fill="white"),
                    strip.text = element_text(color = "black", face = "bold"),
                    plot.title = element_text(hjust = 0.5),
                    legend.position = "none")
  plots <- c(plots, list(p1_acute))
  p2_acute <- plot_model(plotDF, 
                   gamm_model_loop_acute,
                   endpoint = "pro_2_groups",
                   model_type = "smoothSpline",
                   colors = pro_min_colors,
                   spline_type = "ts",
                   knot_lines = F,
                   individual_trendlines = F,
                   individual_points = T,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + 
                  theme(legend.position = "right") +
                  scale_fill_manual(values=pro_min_colors, labels = pro_min_labels) +
                  scale_color_manual(values=pro_min_colors, labels = pro_min_labels) +
                  annotate("label",  x=Inf, y =Inf, label = pval_annot, vjust=1, hjust=1, size = 3, label.r = unit(0, "pt")) +
                  scale_y_continuous(expand = expansion(mult = c(0.1, 0.2)),
                                     limits = c(-1.1,1.1)) +
                  labs(title = , fill = "", color = "", y=factor_final_name) +
                  theme(plot.title = element_text(hjust = 0.5))
  plots_model <- c(plots_model, list(p2_acute))
}

pdf(paste0(results_dir,"Factors_longitudinal_MIN_vs_OTHER_acute.pdf"), width = 18, height = 45)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

png(paste0(results_dir,"Factors_longitudinal_MIN_vs_OTHER_acute.png"), width = 18, height = 50, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

pdf(paste0(results_dir,"Factors_longitudinal_MIN_vs_OTHER_model_acute.pdf"), width = 18, height = 45)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

png(paste0(results_dir,"Factors_longitudinal_MIN_vs_OTHER_model_acute.png"), width = 18, height = 50, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=1)

gridExtra::grid.arrange(grobs = plots_model, ncol=1)

acute_longitudinal_min_other <- plots[[1]]
saveRDS(acute_longitudinal_min_other, file.path(figures_dir,"acute_longitudinal_min_other.rds"))

acute_longitudinal_min_other_model <- plots_model[[1]]
saveRDS(acute_longitudinal_min_other_model, file.path(figures_dir,"acute_longitudinal_min_other_model.rds"))


```


### All PRO groups acute

```{r, linear_model_loop_PRO_groups, fig.width = 3, fig.height=5}

gamm_model_loop_acute <- model_loop(inputDF_acute, age_sex = T,
                              model_type = "smoothSpline",
                              spline_type = "ts",
                              endpoint = "pro_group_labels")

gamm_model_loop_plot_acute <- gamm_model_loop_acute %>% 
                        dplyr::select(adjp.slope, 
                               adjp.intercept)

print_kable(gamm_model_loop_plot_acute[selected_factors_names,])

write.csv(gamm_model_loop_plot_acute[selected_factors_names,], paste0(results_dir, "global_p_value_longitudinal_all_pro_groups_acute.csv"))
```


```{r, longitudinal_model_loop_pheatmap2, fig.width = 3, fig.height=3}
breaklist = seq(0,3,0.3)
pdf(file = paste0(results_dir, "global_p_value_longitudinal_pro_groups_acute.pdf"), width = 3, height = 6)

mat_log10pval <- -log10(gamm_model_loop_plot_acute[selected_factors_names,])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "white",
         cluster_rows = F, 
         cluster_cols = F, 
         color = p_val_only_palette,
         breaks = breaklist)
p1
dev.off()
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_min_vs_other.pdf"), width = 3, height = 6)
p1

pval_min_pro_acute_value <- 10^(-mat_log10pval$adjp.intercept[[1]])
```

### All PRO groups acute model plot

```{r, fig.width=6, fig.height=3}
plots <- list()
plots_model <- list()

for (feature_name in selected_factors_names) {
  plotDF <- inputDF_acute %>% filter(name == feature_name)
  plotDF$pro_group_labels <- factor(plotDF$pro_group_labels, levels = names(pro_colors))
  p1_acute <- plot_model(plotDF, 
                     gamm_model_loop_acute,
                     model_type = "smoothSpline", 
                     spline_type = "ts",
                     colors = pro_colors,
                     CI = T,
                     knot_lines = F,
                     group_trendline_color = "black",
                     facet = T,
                     endpoint = "pro_group_labels") +
              scale_y_continuous(limits = c(-1.1,1.1)) +
              labs(y=factor_final_name) +
              theme(strip.background = element_rect(fill="white"),
                    strip.text = element_text(color = "black", face = "bold"),
                    plot.title = element_text(hjust = 0.5),
                    legend.position = "none")
  plots <- c(plots, list(p1_acute))
  p2_acute <- plot_model(plotDF, 
                   gamm_model_loop_acute,
                   endpoint = "pro_group_labels",
                   model_type = "smoothSpline",
                   spline_type = "ts",
                   knot_lines = F,
                   colors = pro_colors,
                   individual_trendlines = F,
                   individual_points = T,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + 
                   scale_color_manual(values = pro_colors) +
                   scale_fill_manual(values = pro_colors) +
                   scale_y_continuous(limits = c(-1.1,1.1)) +
                   theme(legend.position = "right") +
                   labs(y=factor_final_name, color="", fill="") +
                   guides(fill=FALSE) +
                   theme(plot.title = element_text(hjust = 0.5))
  plots_model <- c(plots_model, list(p2_acute))
}

pdf(paste0(results_dir,"Factors_longitudinal_PRO_GROUPS_acute.pdf"), width = 15, height = 50)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

png(paste0(results_dir,"Factors_longitudinal_PRO_GROUPS_acute.png"), width = 20, height = 60, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

pdf(paste0(results_dir,"Factors_longitudinal_PRO_GROUPS_model_acute.pdf"), width = 15, height = 50)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

png(paste0(results_dir,"Factors_longitudinal_PRO_GROUPS_model_acute.png"), width = 20, height = 60, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=1)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)

acute_longitudinal_progroups <- plots[[1]]
saveRDS(acute_longitudinal_progroups, file.path(figures_dir,"acute_longitudinal_progroups.rds"))

acute_longitudinal_progroups_model <- plots_model[[1]]
saveRDS(acute_longitudinal_progroups_model, file.path(figures_dir,"acute_longitudinal_progroups_model.rds"))



```

### Longitudinal pairwise comparison acute

```{r}
comps <- c("MIN","COG", "PHY", "MLT")
combs <- combn(comps,2)

df_pairwise_combs <- as.data.frame(gamm_model_loop_plot_acute[selected_factors_names,]) %>%
                        rownames_to_column("Factor")
df_pairwise_combs$comp <- "ALL_PRO"

for (i in c(1:dim(combs)[2])){
  combs_name <- paste(combs[,i], collapse="_")
  inputDF_comb <- inputDF_acute %>% filter(pro_group_labels %in% combs[,i])
  
  gamm_model_loop_acute <- model_loop(inputDF_comb, age_sex = T,
                              model_type = "smoothSpline",
                              spline_type = "ts",
                              endpoint = "pro_group_labels")

  gamm_model_loop_plot_acute <- gamm_model_loop_acute %>% 
                          dplyr::select(adjp.slope, 
                                 adjp.intercept)
  
  gamm_model_loop_plot_sel_acute <- as.data.frame(gamm_model_loop_plot_acute[selected_factors_names,]) %>%
                                mutate(comp = combs_name) %>%
                                rownames_to_column("Factor")
  df_pairwise_combs <- bind_rows(df_pairwise_combs, gamm_model_loop_plot_sel_acute)
}

```

### pairwise pval plot

```{r linear_model_loop_pheatmap_pairwise, fig.width = 3, fig.height=3}
mat_longitudinal_pairwise <- df_pairwise_combs %>%
                                         dplyr::select(-Factor, -adjp.slope) %>%
                                         dplyr::rename(`adj p-val` = adjp.intercept) %>%
                                         column_to_rownames("comp")


pdf(file = paste0(results_dir, "global_p_value_longitudinal_pro_groups_pairwise_acute.pdf"), width = 3, height = 6)
mat_log10pval <- -log10(mat_longitudinal_pairwise)
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "white",
         cluster_rows = F, 
         cluster_cols = F, 
         color = p_val_only_palette,
         breaks = breaklist)
p1
dev.off()

p1
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_pro_groups.pdf"), width = 3, height = 6)

acute_longitudinal_pairwise_progroup_pval_plot <- p1
saveRDS(acute_longitudinal_pairwise_progroup_pval_plot, file.path(figures_dir,"acute_longitudinal_pairwise_progroup_pval_plot.rds"))

```


```{r}
sessionInfo()
```
