---
title: "Viral load association analysis"
author: "Gisela Gabernet (gisela.gabernet@yale.edu)"
subtitle: PGX, SO, PMG, PPT, PPG, BCT
date: "`r format(Sys.time(), '%b %d %Y')`"
output:
  html_document:
    code_download: yes
    code_folding: hide
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: yes
      smooth_scroll: no
    theme: yeti
params:
  results_dir: "associate_antibody_titers/"
  spear_object: "/scratch/data-fullcohort/spear_model/results/020624/032124_spearphysical_finalscores_allsamples.rds"
  partition: "all"
editor_options: 
  chunk_output_type: inline
---

```{r}
library(data.table)
#install.packages("RColorBrewer")
library(RColorBrewer)
library(ggplot2)
library(tidyverse)
#install.packages("psych")
library(psych)
library(dplyr)
# Additional libraries for mixed effect models
#BiocManager::install("qvalue")
library(qvalue)
library(lme4)
library(ordinal)
library(nlme)
library(gamm4)
library(mgcv)
library(ggsignif)
library(kableExtra)
library(corrplot)
library(ggeffects)
library(ggfortify)
library(grid)
library(pheatmap)

#install.packages("ggvenn")
library(ggvenn)

#install.packages("msigdbr")
library(msigdbr)

knitr::opts_chunk$set(warning = F, message=F)

print_kable <- function(table, kable_height = "500px", kable_width = "100%") {
  kable(table, digits=3) %>%
    kable_styling("striped") #%>%
    #scroll_box(height = kable_height, width = kable_width)
  
}

# save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
#    stopifnot(!missing(x))
#    stopifnot(!missing(filename))
#    pdf(filename, width=width, height=height)
#    grid::grid.newpage()
#    grid::grid.draw(x$gtable)
#    dev.off()
# }

```

# Running parameters

Running with parameters:
- features:
  - spear: get spear top features with joint probability >0.95
  - hallmark: get core set from hallmark heme metabolism pathway
- partition:
  - all: use all data for association
  - test: use the test dataset for association (with and without PROMIS PHY values)
  - PRO_PHY_NA_TEST: use all test dataset for association (with and without PROMIS PHY values, and the train dataset without PROMIS PHY values available)
- Spear object: spear object containing factor scores, and joint probabilities for extracting features.
- results_dir: name of results dir, will be created.

Example command to execute using an R script:

```{r eval=F}
# rmarkdown::render("associate_top_spear_features.Rmd",
#                   output_file="associate_top_spear_features_promis_physical_Factor1.html",
#                   params=list(
#                     results_dir = "associate_top_spear_features_promis_physical_Factor1/",
#                     spear_object = "/scratch/data-fullcohort/spear_model/results/factor_scores/031124_spearphysical_finalscores.rds",
#                     partition = "all",
#                     features = "spear"
#                   ))
```


# Explore Spear results

## Explore Feature correlation and variance explained

```{r fig.height=5, fig.width=10}

#spear_object <- readRDS(params$spear_object)

data_env_conv <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_convalescent_data_env_all_Feb_11_2024.RDS")
data_env_acute <- readRDS("/scratch/data-fullcohort/batch-effects/impacc_acute_data_env_all_Mar_23_2024.RDS")
mofa_model_conv_train <- readRDS("/scratch/data-fullcohort/MOFA_factors/MOFA_train_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240212.rds")
mofa_model_conv_test <- readRDS("/scratch/data-fullcohort/MOFA_factors/MOFA_test_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240212.rds")
mofa_model_acute_train <- readRDS("/scratch/data-fullcohort/spear/acute_preprocessed_data_20240323/MOFA_acute_train_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240331.rds")
mofa_model_acute_test <- readRDS("/scratch/data-fullcohort/spear/acute_preprocessed_data_20240323/MOFA_acute_test_set_250_factors_SO_PPT_PPG_PMG_PGX_BCT_20240331.rds")

#results_dir <- "./associate_spear_factor_promis_physical_train/"
results_dir <- params$results_dir
dir.create(results_dir)

figures_dir <- "./Figures_viral_load_association"
dir.create(figures_dir)


path_to_codebase <- "~/repos/impacc-full-cohort/Analysis/IMPACC_data_analysis_codebase.R"
source(path_to_codebase)

#selected_factor_name <- "Factor1"
```


## Get clinical data

```{r}
clinical_data_conv <- data_env_conv$clinical_data
clinical_data_acute <- data_env_acute$clinical_data

# Add additional symptom groups to metadata
symptom_groups <- read_csv("/scratch/clinical/current/impacc_conv_symptoms.csv", show_col_types = F)
PRO_scores <- read_csv("/scratch/clinical/current/PRO_scores_dates_updated_29Aug23.csv", show_col_types = F)

```



# Create feature matrix

```{r}
library(MOFA2)

train_data_conv <- MOFA2::get_data(mofa_model_conv_train)
test_data_conv <- MOFA2::get_data(mofa_model_conv_test)
train_data_acute <- MOFA2::get_data(mofa_model_acute_train)
test_data_acute <- MOFA2::get_data(mofa_model_acute_test)

detach("package:MOFA2", unload=TRUE)


```

## Antibody titers subset to considered events

```{r}
# Need to reprocess clinical data because the data for the RBD titers was somehow not kept in the data_env

clinical_sample_data <- read.table("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-sample-locked.csv", sep = ",", header=T)
clinical_event_data <- read.table("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-event-locked.csv", sep = ",", header=T)
clinical_individ_data <- read.table("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-locked.csv", sep = ",", header=T)
clinical_individ_supp_data <- read.table("/data/clinical/current/2023-01-01-locked/2023-01-01-impacc-clin-individ-supp-locked.csv", sep = ",", header=T)

clinical_data <-
    dplyr::inner_join(clinical_sample_data,
               clinical_event_data,
               by = c("event_id" = "event_id")) %>%    # Join the clinical_sample_data and clinical_event_data tables using "event_id" column
    dplyr::mutate(
      participant_id = participant_id.x,
      participant_id.x = NULL,
      participant_id.y = NULL
    ) %>%   # Remove redundant "participant_id" columns
    dplyr::inner_join(clinical_individ_data,
               by = c("participant_id" = "participant_id"))   # Append the clinical_individ_data table
  
# Add new clinical data by participant_id:
clinical_data <- dplyr::left_join(clinical_data, clinical_individ_supp_data, by="participant_id")

viral_load_metadata <- bind_rows(data_env_acute$nasal_viralload_metadata, data_env_conv$nasal_viralload_metadata)

viral_load_metadata_fixed_phase <- viral_load_metadata %>%
  rownames_to_column("sample_id") %>%
  dplyr::select(-phase) %>%
  dplyr::left_join(clinical_data %>% select("sample_id","phase"), by="sample_id")
```


```{r}
data_env_acute$nasal_viralload_counts <- data_env_acute$nasal_viralload_counts %>% 
  dplyr::select(N1_CT, N2_CT)

data_env_acute$nasal_viralload_metadata <- viral_load_metadata_fixed_phase[rownames(data_env_acute$nasal_viralload_metadata),]

convalescent_pro_groups <- read.csv("/scratch/clinical/current/impacc-convalescent.csv")

# Convert PRO group cluster numbers to PRO groups as described in paper
# This mapping was communicated by Leying Guan via Teams on May 15th as the 
# original table is unclear about the exact PRO groups mappings
    
convalescent_pro_groups <- convalescent_pro_groups %>% 
  dplyr::mutate( pro_group_labels = case_match( 
    conv_cluster6,
    c(1,3,4) ~ "MIN",
    c(2) ~ "PHY",
    c(5) ~ "COG",
    c(6) ~ "MLT"
  ))

# Check number of patients in each of the groups
conv_cohort_size <- length(unique(convalescent_pro_groups$participant_id))
cat("There are ", conv_cohort_size,
        " participants in the convalescent cohort.")

clinical_data <- left_join(clinical_data, convalescent_pro_groups, by="participant_id") %>%
                dplyr::mutate(pro_2_groups =
                    case_match( pro_group_labels,
                    "MIN" ~ "MIN",
                    c("PHY", "COG", "MLT")  ~ "OTHER"
                    )
                )

vaccine_data <- read.csv("~/repos/impacc-full-cohort/Analysis/convalescent-project/vaccination/Convalescent_vaccination_data_with_event_dates.csv")


clinical_data <- left_join(clinical_data, vaccine_data, by=join_by("participant_id"==id))

clinical_data <- clinical_data %>% 
  rowwise() %>%
  mutate(event_after_first_vaccine_dose_NA = 
        dplyr::if_else(event_date >= date_anydose,
                       "yes",
                       "no"),
        event_after_first_vaccine_dose = dplyr::if_else(vac_reported_corrected=="No", 
                                                        "no", 
                                                        event_after_first_vaccine_dose_NA),
        event_date_relative_to_first_dose = event_date - date_anyfirstdose,
        event_date_relative_to_second_dose = event_date - Second,
        event_date_relative_to_booster_dose = event_date - Booster,
        event_date_3weeks_after_first_dose = ifelse((event_date_relative_to_first_dose >= 0 && 
                                                       event_date_relative_to_first_dose <= 21),
                                        1, 0),
        event_date_3weeks_after_second_dose = ifelse((event_date_relative_to_second_dose >= 0 &&
                                                       event_date_relative_to_second_dose <= 21),
                                        1, 0),
        event_date_3weeks_after_booster = ifelse((event_date_relative_to_booster_dose >= 0 &&
                                                       event_date_relative_to_booster_dose <= 21),
                                        1, 0),
        event_date_3weeks_after_dose_NA = ifelse(( event_date_3weeks_after_first_dose |  event_date_3weeks_after_second_dose | event_date_3weeks_after_booster ), "yes", "no"),
        event_date_3weeks_after_dose = ifelse(is.na(event_date_3weeks_after_dose_NA),
                                                    ifelse(event_after_first_vaccine_dose == "no", "no", event_date_3weeks_after_dose_NA),
                                                    event_date_3weeks_after_dose_NA),
        event_date_3weeks_after_dose = ifelse(is.na(event_date_3weeks_after_dose),
                                                    ifelse(vaccinated == "No", "no", event_date_3weeks_after_dose),
                                                    event_date_3weeks_after_dose)
        )

### Generate discretized age based in 5 quantiles.
admit_age_levels = sort(unique(clinical_data$admit_age))
admit_age_levels_tiles = ntile(admit_age_levels, 5)
label_df_tmp = data.frame( admit_age_levels_tiles, admit_age_levels) %>% 
  group_by(admit_age_levels_tiles) %>% 
  summarize(min=min(admit_age_levels), max=max(admit_age_levels)) %>% 
  mutate(labels = paste0("[",min, ",", max, "]") ) 
admit_age_levels_tiles_labels = label_df_tmp$labels %>% setNames(label_df_tmp$admit_age_levels_tiles)
admit_age_levels_labels = admit_age_levels_tiles_labels[admit_age_levels_tiles] %>% setNames(admit_age_levels)
discretized_admit_age_quantile = admit_age_levels_labels[ as.character(clinical_data$admit_age) ]

### Generate discretized age based on equal width ages.
discretized_admit_age_equalWidth = cut(clinical_data$admit_age, breaks = 5)
clinical_data$discretized_admit_age_quantile <- discretized_admit_age_quantile
clinical_data$discretized_admit_age_equalWidth <- discretized_admit_age_equalWidth

data_env_conv$clinical_data <- clinical_data
data_env_acute$clinical_data <- clinical_data


nasal_viral_load <- data_env_acute$nasal_viralload_counts  %>% 
  rownames_to_column("sample_id") %>%
  left_join(clinical_data %>% dplyr::select(-phase), by="sample_id") %>%
  left_join(viral_load_metadata_fixed_phase, by="sample_id") %>%
  mutate(phase_plate = paste(phase,"_",plate_num))

saveRDS(nasal_viral_load, file="/scratch/data-fullcohort/batch-effects/nasal_viral_load_acute_20240701.rds")

dim(nasal_viral_load)
```


```{r}
top_feat_data_df <- nasal_viral_load %>% 
  dplyr::filter(participant_id %in% convalescent_pro_groups$participant_id) %>%
  dplyr::select(event_id, N1_CT, N2_CT) %>% # Selecting only the interesting Nasal viral load measurements
  tibble::column_to_rownames("event_id")

colnames(top_feat_data_df) <- str_replace_all(colnames(top_feat_data_df), " ","_")
```

# Associate features with outcomes

Here we will use a mixed linear model to associate features with outcomes

## Prepare metadata

```{r prepare the metadata for mixed linear model analysis, echo= FALSE, warning= FALSE}
correlation_metadata <- clinical_data %>%
  dplyr::select(event_id, resp_status_v1, sex, admit_age, event_after_first_vaccine_dose, 
                race, enrollment_site, event_date, 
                event_date_3weeks_after_dose, pro_group_labels, event_type, 
                discretized_admit_age_quantile, participant_id)  %>%
  dplyr::mutate(pro_2_groups = 
      case_match( pro_group_labels,
      "MIN" ~ "MIN",
      c("PHY", "COG", "MLT")  ~ "OTHER"
      )
  ) %>%
  dplyr::distinct() %>%
  filter(event_id %in% rownames(top_feat_data_df))

clinical_subsets_all <- as.data.frame(correlation_metadata)
rownames(clinical_subsets_all) <- clinical_subsets_all$event_id


endpoint_coefficients_all <- data.frame(matrix(nrow=ncol(top_feat_data_df), ncol = 0))
rownames(endpoint_coefficients_all) <- colnames(top_feat_data_df)
```


## Functions for association

```{r}

pro_min_colors <- c("#639A21", "#781B6C")
names(pro_min_colors) <- c("MIN", "OTHER")

pro_min_labels <- c("MIN", "LC")
names(pro_min_labels) <- c("MIN", "OTHER")

#original
#pro_colors <- c("#639A21", "#6371AD", "#FFC20A", "#9C3418")
pro_colors <- c("#639A21", "#6371AD", "#FFC20A", "#7D2D17")
names(pro_colors) <- c("MIN", "COG", "PHY", "MLT")

ordered_visits <- c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6")
acute_visits <- c("Visit 1", "Visit 2", "Visit 3", "Visit 4", "Visit 5", "Visit 6")

scaleFUN <- function(x) sprintf("%.2e",x)
violin_plot_visits <- function(data, colors, limits=NULL, ncol=5) {
  p <- ggplot(data, aes(x=endpoints, y=value, fill=endpoints)) +
    geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
    facet_wrap(~Feature, drop = T, ncol = ncol, scales = "free_y") +
    theme_bw() + theme(legend.position = "none")
  if(!(is.null(limits))){
    p <- p + scale_y_continuous(limits = limits)
  } else {
    p <- p + scale_y_continuous(labels = scaleFUN)
  }
  if(!(missing(colors))){
    p <- p + scale_color_manual(values=colors)
    p <- p + scale_fill_manual(values=colors)
  }
  return(p)
}

pairwise_association_visit <- function(clinical_subsets_all, 
                                       feature_data, visit_id, 
                                       selected_features_names, 
                                       endpoint_var = "pro_2_groups",
                                       ordered_endpoint_var = c("MIN","OTHER"),
                                       colors = pro_min_colors
                                       ){
  cur_visit <- visit_id

  visit_row_ids = rownames(clinical_subsets_all[clinical_subsets_all$event_type == cur_visit,])
  
  # Use the visit7_row_ids to only get samples from 'Visit 7':
  features_filtered = feature_data[visit_row_ids , colnames(feature_data), drop=F]
  clinical_subsets = clinical_subsets_all[visit_row_ids ,]
  
  data_use = features_filtered
  
  # Add the respiratory_status_day14 as $endpoints
  data_use$endpoints = factor(clinical_subsets[[endpoint_var]], levels=ordered_endpoint_var)
  
  # Add the enrollment_site as $sites
  data_use$sites = clinical_subsets$enrollment_site
  data_use$age =  clinical_subsets$discretized_admit_age_quantile
  data_use$sex =  clinical_subsets$sex
  
  # Add the $control, needs to be different for each sample
  # required for clmm (always requires some random effect)
  data_use$control =  factor(1:nrow(data_use))
  
  # Set module names:
  feature_names = colnames(features_filtered)
  
  # Only use samples that have valid endpoints (not NA/missing)
  data_use = data_use[!is.na(data_use$endpoints),]

  for(j in 1:length(feature_names)){
    #print(j)
    my.formula0 = paste0(feature_names[j],"~ (1|sites)+age+sex")
    my.formula1 = paste0(feature_names[j],"~ endpoints + (1|sites)+age+sex")
    if(j == 1){
      tmp = mixed_pairwise(my.formula0, my.formula1, data_use)
      res_table_pairwise = data.frame(matrix(0, ncol = length(tmp$res_table), nrow = length(feature_names)))
      coef_endpoint_pairwise = data.frame(matrix(0, ncol= length(tmp$coef_endpoint), nrow=length(feature_names)))
      res_table_pairwise[j,] = tmp$res_table
      coef_endpoint_pairwise[j,] = tmp$coef_endpoint
      colnames(res_table_pairwise) = names(tmp$res_table)
      colnames(coef_endpoint_pairwise) = names(tmp$coef_endpoint)
    }else{
      tmp = mixed_pairwise(my.formula0, my.formula1, data_use)
      res_table_pairwise[j,] = tmp$res_table
      coef_endpoint_pairwise[j,] = tmp$coef_endpoint
    }
  }
  
  if (length(ordered_endpoint_var) > 2) {
      coef_endpoint_pairwise <- coef_endpoint_pairwise[,1:3]
      res_table_pairwise <- res_table_pairwise[,1:3]
  }
  coef_endpoint_pairwise$feature <- feature_names
  coef_endpoint_pairwise$visit <- cur_visit
  
  if (nrow(res_table_pairwise)>1) {
    res_table_pairwise_adj_df <- qvalue::qvalue(res_table_pairwise, fdr.level = 0.05, pi0=1)$qvalues
  } else {
    res_table_pairwise_adj_df <- res_table_pairwise
  }

  res_table_pairwise_adj_df$feature <- feature_names
  res_table_pairwise_adj_df$visit <- cur_visit
  
  plot_df <- data_use %>% dplyr::select(selected_features_names,
                                        endpoints) %>%
            pivot_longer(!endpoints, names_to = "Feature", values_to = "value")
  plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))
  
  p <- violin_plot_visits(plot_df, colors = colors) + ggtitle(visit_id)
  
  res_list <- list(coefs_df = coef_endpoint_pairwise, 
                   res_df = res_table_pairwise_adj_df,
                   violin_plot = p)
  
  return(res_list)
}

```

## Functions for better pheatmap

```{r}
## Edit body of pheatmap:::draw_colnames, customizing it to your liking
draw_colnames_45 <- function (coln, ...) {
    m = length(coln)
    x = (1:m)/m - 1/2/m
    grid.text(coln, x = x, y = unit(0.96, "npc"), vjust = .5, 
        hjust = 1, rot = 45, gp = gpar(...)) ## Was 'hjust=0' and 'rot=270'
}

## For pheatmap_1.0.8 and later:
draw_colnames_45 <- function (coln, gaps, ...) {
    coord = pheatmap:::find_coordinates(length(coln), gaps)
    x = coord$coord - 0.5 * coord$size
    res = textGrob(coln, x = x, y = unit(1, "npc") - unit(3,"bigpts"), vjust = 0.5, hjust = 1, rot = 45, gp = gpar(...))
    return(res)}

## 'Overwrite' default draw_colnames with your own version 
assignInNamespace(x="draw_colnames", value="draw_colnames_45",
ns=asNamespace("pheatmap"))

p_val_only_palette = colorRampPalette(brewer.pal(9,"Reds"))(10)
```

## Reproducing publication plot

### N1 Ct
```{r}
nasal_viral_load$pro_group_labels <- factor(nasal_viral_load$pro_group_labels, levels=names(pro_colors))
ggplot(nasal_viral_load, aes(x=pro_group_labels, y=N1_CT)) +
    ggbeeswarm::geom_quasirandom(aes(fill=pro_group_labels), shape=21, width=.1, dodge.width = 0.75) +
    geom_boxplot(aes(fill=pro_group_labels),alpha=0.6, color="black", outlier.shape=NA) +
    theme_bw() + 
    scale_y_reverse() +
    scale_color_manual(values=pro_colors) +
    scale_fill_manual(values=pro_colors) +
    labs( x="", y="N1 Ct", fill="")

```

### N2 Ct
```{r}
ggplot(nasal_viral_load, aes(x=pro_group_labels, y=N1_CT)) +
    ggbeeswarm::geom_quasirandom(aes(fill=pro_group_labels), shape=21, width=.1, dodge.width = 0.75) +
    geom_boxplot(aes(fill=pro_group_labels),alpha=0.6, color="black", outlier.shape=NA) +
    theme_bw() + 
    scale_y_reverse() +
    scale_color_manual(values=pro_colors) +
    scale_fill_manual(values=pro_colors) +
    labs( x="", y="N2 Ct", fill="")

```

## Association MIN vs OTHER for each visit


We treat each factor expression as the covariate and consider its association with clinical endpoints 
(pro_groups) in the presence of potential heterogeneity across different clinical sites, 
which is modeled as a random effect.

We first examine if there is indeed a visible random effect from clinical sites, for both the intercepts and the slopes.
There may be some weak random effects for the intercept, but observe "nothing" regarding the slope. 
As a result, we consider only the random intercept effects.

We perform pairwise comparisons.


```{r fig.width=2, fig.height=2}
endpoint_coefficients_all <- data.frame()
res_table_pairwise_adj_all <- data.frame()

# Need to subset to IgG as IgM has only acute measurements
for (visit_id in ordered_visits){
  print(paste("Computing associations for visit:", visit_id))
  res_list <- pairwise_association_visit(clinical_subsets_all=clinical_subsets_all,
                                          feature_data=top_feat_data_df,
                                          visit_id=visit_id,
                                          selected_features_names = colnames(top_feat_data_df),
                                          endpoint_var = "pro_2_groups") #TODO: update outcome variable here

  endpoint_coefficients_all <- bind_rows(endpoint_coefficients_all,
                                         res_list$coefs_df)
  res_table_pairwise_adj_all <- bind_rows(res_table_pairwise_adj_all, 
                                          res_list$res_df)
}


```

### Adj p-val all visits

```{r fig.height=2.5, fig.width=10}

res_table_pairwise_adj_all_plot <- res_table_pairwise_adj_all %>%
                                      pivot_wider(names_from = feature,
                                                  values_from = "MIN|OTHER") %>%
                                      column_to_rownames("visit")

print_kable(res_table_pairwise_adj_all_plot)
write.csv(res_table_pairwise_adj_all_plot, paste0(results_dir, "Adj_pval_all_visits_min_vs_other.csv"))
```


### Log p-value plot with coefficient sign

```{r fig.asp=0.5, fig.width=5}
endpoint_coefficient_plot <- endpoint_coefficients_all %>%
                              mutate(sign = if_else(`MIN|OTHER`>0, 1, -1)) %>%
                              dplyr::select(feature, visit, sign)

df_plot <- res_table_pairwise_adj_all %>%
              mutate(log10_adjpval = -log10(`MIN|OTHER`)) %>%
              inner_join(endpoint_coefficient_plot, by=c("feature", "visit")) %>%
              mutate(signed_adjp = log10_adjpval * sign)

df_plot$feature <- factor(df_plot$feature, levels = colnames(top_feat_data_df))
df_plot$visit <- factor(df_plot$visit, levels = unique(df_plot$visit))
df_plot <- df_plot %>% 
            mutate(annot =
                case_when(
                 abs(log10_adjpval) >= -log10(0.001) ~ "***",
                 abs(log10_adjpval) >= -log10(0.01) ~ "**",
                 abs(log10_adjpval) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 ))

thr <- 0.05
plot_title <- "Adj p-value feature association"


convert_df <- df_plot %>% 
              dplyr::select(feature, visit, signed_adjp) %>%
              pivot_wider(id_cols = c(visit), 
                          names_from = feature, 
                          values_from = signed_adjp) %>%
              column_to_rownames("visit")

convert_annot <- df_plot %>%
              dplyr::select(feature, visit, annot) %>%
              pivot_wider(id_cols = c(visit),
                          names_from = feature,
                          values_from = annot) %>%
              column_to_rownames("visit")

mat <- as.matrix(convert_df)
mat_annot <- as.matrix(convert_annot)

palette_length <- 50
abs_max <- max(abs(min(mat)), abs(max(mat)), 1.3)
breaks_list <- c(seq(-abs_max, 0, length.out=ceiling(palette_length/2) + 1), 
              seq(abs_max/palette_length, abs_max, length.out=floor(palette_length/2)))

pdf(file = paste0(results_dir, "Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 12)
p <- pheatmap(t(mat),
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = t(mat_annot),
         number_color = "black",
         fontsize_number = 10
         )
p
dev.off()

p
#save_pheatmap_pdf(p, paste0(results_dir,"Adj_pval_all_visits_min_vs_other_plot.pdf"), width = 12, height = 6)

```

### Violin plots all visits

```{r }

features_filtered = top_feat_data_df[, colnames(top_feat_data_df), drop=F]
clinical_subsets = clinical_subsets_all[rownames(top_feat_data_df),]
  

data_use = features_filtered 

data_use$endpoints = as.factor(clinical_subsets$pro_2_groups)

# Add the enrollment_site as $sites
data_use$sites = clinical_subsets$enrollment_site
data_use$age =  clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex
data_use$visit = clinical_subsets$event_type

# Add the $control, needs to be different for each sample
# required for clmm (always requires some random effect)
data_use$control =  factor(1:nrow(data_use))

# Set module names:
feature_names = colnames(features_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$endpoints),]
```

Violin plots for all features, values from all visits

```{r fig.width=4, fig.height=2}

plot_df <- data_use %>% dplyr::select(colnames(top_feat_data_df),
                                        endpoints) %>%
            pivot_longer(!endpoints, names_to = "Feature", values_to = "value")
plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))

p <- ggplot(plot_df, aes(x=endpoints, y=value, color=endpoints)) +
  geom_boxplot(fill=NA, outlier.shape = NA)+
  geom_jitter(width = 0.2, alpha=0.2) +
  #geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  theme_bw() + theme(legend.position = "none") +
  scale_color_manual(values=pro_min_colors) +
  scale_fill_manual(values=pro_min_colors) +
  facet_wrap(~Feature, drop = T, scales = "free_y", ncol = 4)
p
```

Violin plots for all features, values per visit

```{r fig.height=2, fig.width=8}

plot_df <- data_use %>% dplyr::select(colnames(top_feat_data_df),
                                        endpoints, visit) %>%
            pivot_longer(!c(endpoints,visit), names_to = "Feature", values_to = "value")
plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))
plot_df$visit <- factor(plot_df$visit, levels=ordered_visits)

for (feature in colnames(top_feat_data_df)) {
  plot_df_fact <- plot_df %>% filter(Feature == feature)
  
  p <- ggplot(plot_df_fact, aes(x=visit, y=value, fill=endpoints)) +
  ggbeeswarm::geom_quasirandom(aes(fill=endpoints), shape=21, width=.15, dodge.width = 0.75) +
  geom_boxplot(alpha=0.6, color="black", outlier.shape = NA) +
  #scale_y_continuous(labels = scaleFUN) +
  theme_bw() + 
  scale_y_reverse() + # Scale is reversed as higher Ct values indicate lower viral load
  scale_color_manual(values=pro_min_colors, labels = pro_min_labels) +
  scale_fill_manual(values=pro_min_colors, labels = pro_min_labels) +
  #facet_wrap(~visit, drop = T, ncol = 4) +
  labs( x="", y=feature, fill="")
  print(p)
}

```

### Violin plots per visit p-val

```{r fig.height=2, fig.width=6}

feature_violin_plots_list <- list()

for (feat in colnames(top_feat_data_df)) {
  plot_df_feat <- plot_df %>% filter(Feature == feat)
  
  format_corrected_pval <- res_table_pairwise_adj_all %>% 
    dplyr::filter(feature == feat) %>%
    pivot_longer(cols=-c(feature,visit), names_to = "comparison", values_to = "adj_pval_sex_age_corrected") %>%
    dplyr::mutate(group1 = str_extract(comparison, "[^|]+"),
                  group2 = str_extract(comparison, "[^|]+$")) %>%
    dplyr::select(-c(comparison,feature))
  
  stat.test <- plot_df_feat %>%
    group_by(visit) %>%
    wilcox_test(value ~ endpoints) %>%
    adjust_pvalue(method = "hochberg") %>%
    left_join(format_corrected_pval, by=join_by(visit,group1,group2)) %>%
    add_significance("adj_pval_sex_age_corrected") %>%
    add_xy_position(x = "visit", dodge = 0.75) %>%
    dplyr::filter(adj_pval_sex_age_corrected < 0.05)

  
  p <- ggplot(plot_df_feat, aes(x=visit, y=value)) +
    ggbeeswarm::geom_quasirandom(aes(fill=endpoints),shape=21,width=.15, dodge.width = 0.75) +
    geom_boxplot(aes(fill=endpoints),alpha=0.6, color="black", outlier.shape=NA) +
    theme_bw() + 
    scale_y_reverse(expand = expansion(mult = c(0.1, 0.1))) + # reverse y scale as low Ct values correspond to high viral load
    scale_color_manual(values=pro_min_colors, labels = pro_min_labels) +
    scale_fill_manual(values=pro_min_colors, labels = pro_min_labels) +
    labs( x="", y=str_replace_all(feat,"_"," "), fill="") +
    stat_pvalue_manual(stat.test, label = "adj_pval_sex_age_corrected.signif", y.position = 0)
  print(p)
  feature_violin_plots_list <- append(feature_violin_plots_list, list(p))
}

feature_visits_min_other_pvals <- feature_violin_plots_list
```
## Association with SPEAR factor


```{r}
# Helper function to plot association with viral load and Factor1
  plot_titer_association_Factor1 <- function(df, predictor) {
    # Filter to df where predictor values are not NA
    df <- df %>% filter(!is.na(!!as.symbol(predictor)))
    
    # Generate model without predictor
    form <- "Factor1~age + sex + (1|sites/participant_id) + visit"
    model.0 <- lmer(as.formula(form), data = df)
    
    # Generate model with predictor
    form <- paste(form, predictor, sep = "+")
    model.1 <- lmer(as.formula(form), data = df)
    
    # Run anova
    test <- anova(model.1, model.0)
    
    # Pull p-val
    pval <- test$`Pr(>Chisq)`[2]
    
    # Generate marginal means
    pred <- ggpredict(model.1, terms = predictor, type = "fixed", ci.lvl = 0.95)
    
    ggplot(df, aes(x = .data[[predictor]], y = Factor1)) +
      geom_point() +
      geom_ribbon(data = pred, 
                  aes(x = x, 
                      ymin = pred$conf.low, 
                      ymax = pred$conf.high),
                  alpha = 0.3,
                  inherit.aes = F,
                  fill = "grey") +
      geom_line(data = pred, 
                aes(x = x, y = predicted),
                linewidth = 1.5,
                inherit.aes = F,
                color = "grey30") +
      theme_bw() +
      labs(title = paste("p-val =", signif(pval, digits = 2)),
           x = str_replace_all(predictor, pattern = "_", replacement = " ")) +
      theme(plot.title = element_text(hjust = 0.5)) -> plot
    
    return(plot)
  }
```

```{r}
spear_object <- readRDS(params$spear_object)

# Load spear factors
factor_scores_conv <- rbind(spear_object$Test, 
                            spear_object$TestNoPhysical, spear_object$TrainNoPhysical_NoOverlap)
participants_conv <- convalescent_pro_groups$participant_id
event_id_conv <- rownames(factor_scores_conv)
diff_ids <- setdiff(rownames(spear_object$Acute), rownames(factor_scores_conv))
factor_scores_acute <- as.data.frame(spear_object$Acute[diff_ids,])
event_id_acute <- rownames(factor_scores_acute)
```

### Association acute factor with N1 CT

```{r}

# Add factor scores to df
data_use_factor <- data_use %>%
  rownames_to_column("event_id") %>%
  left_join(factor_scores_acute %>% rownames_to_column("event_id"), by="event_id") %>%
  left_join(clinical_data %>% select(event_id, participant_id) %>% distinct(), by="event_id") %>%
  column_to_rownames("event_id")
  
```

```{r}
corplot <- plot_titer_association_Factor1(data_use_factor, "N1_CT")
corplot + 
  scale_y_continuous(limits = c(-1.2,1.2)) +
  labs(x="N1 Ct", y = "Acute Factor 1")
```

### Association acute factor with N1 CT (only Visit 1)

```{r}
# Helper function to plot association with viral load and Factor1 at one visit
  plot_titer_association_Factor1_pervisit <- function(df, predictor) {
    # Filter to df where predictor values are not NA
    df <- df %>% filter(!is.na(!!as.symbol(predictor)))
    
    # Generate model without predictor
    form <- "Factor1~age + sex + (1|sites)"
    model.0 <- lmer(as.formula(form), data = df)
    
    # Generate model with predictor
    form <- paste(form, predictor, sep = "+")
    model.1 <- lmer(as.formula(form), data = df)
    
    # Run anova
    test <- anova(model.1, model.0)
    
    # Pull p-val
    pval <- test$`Pr(>Chisq)`[2]
    
    # Generate marginal means
    pred <- ggpredict(model.1, terms = predictor, type = "fixed", ci.lvl = 0.95)
    
    ggplot(df, aes(x = .data[[predictor]], y = Factor1)) +
      geom_point() +
      geom_ribbon(data = pred, 
                  aes(x = x, 
                      ymin = pred$conf.low, 
                      ymax = pred$conf.high),
                  alpha = 0.3,
                  inherit.aes = F,
                  fill = "grey") +
      geom_line(data = pred, 
                aes(x = x, y = predicted),
                linewidth = 1.5,
                inherit.aes = F,
                color = "grey30") +
      theme_bw() +
      labs(title = paste("p-val =", signif(pval, digits = 2)),
           x = str_replace_all(predictor, pattern = "_", replacement = " ")) +
      theme(plot.title = element_text(hjust = 0.5)) -> plot
    
    return(plot)
  }
```

```{r}
data_use_factor_visit1 <- data_use_factor %>%
  filter(visit == "Visit 1")

corplot <- plot_titer_association_Factor1_pervisit(data_use_factor_visit1, "N1_CT")
corplot + labs(x="N1 Ct (Visit 1)", y = "Acute Factor 1 (Visit 1)")
```


## Association MIN vs all other groups

```{r fig.width=2, fig.height=2}
endpoint_coefficients_all <- data.frame()
res_table_pairwise_adj_all <- data.frame()

for (visit_id in ordered_visits){
  tryCatch(
    {
      print(paste("Computing associations for visit:", visit_id))
      res_list <- pairwise_association_visit(clinical_subsets_all=clinical_subsets_all, 
                                              feature_data=top_feat_data_df, 
                                              visit_id=visit_id, 
                                              selected_features_names = colnames(top_feat_data_df),
                                              endpoint_var = "pro_group_labels", #TODO: update outcome variable here
                                              ordered_endpoint_var = names(pro_colors),
                                              colors = pro_colors
                                              )
    
      endpoint_coefficients_all <- bind_rows(endpoint_coefficients_all,
                                             res_list$coefs_df)
      res_table_pairwise_adj_all <- bind_rows(res_table_pairwise_adj_all, 
                                              res_list$res_df)
    },
    error = function(cond){
      message(paste("Not possible to calculate association for visit:",visit_id))
    }
  )
  
}


```

### Adj p-val all visits

```{r pairwise_heatmap, fig.height=2.5, fig.width=20}

library(pheatmap)

res_table_pairwise_adj_all_plot <- res_table_pairwise_adj_all %>% 
                                    pivot_longer(cols=starts_with("MIN"),
                                                 values_to="padj",
                                                 names_to="comp") %>%
                                    mutate(comp_visit = paste0(comp, "_", visit)) %>%
                                    dplyr::select(-c(comp)) %>%
                                    pivot_wider(id_cols = c(comp_visit),
                                                names_from = feature,
                                                values_from = padj) %>%
                                    column_to_rownames("comp_visit")
print_kable(res_table_pairwise_adj_all_plot)
write.csv(res_table_pairwise_adj_all_plot, paste0(results_dir, "Adj_pval_all_visits_all_pro_groups.csv"))
```

### Log p-value plot with coefficient sign
```{r fig.asp=0.4, fig.width=20}
endpoint_coefficient_plot <- endpoint_coefficients_all %>%
                              pivot_longer(-c(feature,visit), 
                                           names_to = "comparison", 
                                           values_to = "coefficient") %>% 
                              mutate(sign = if_else(coefficient>0, 1, -1)) %>%
                              dplyr::select(feature, visit, comparison, sign)

df_plot <- res_table_pairwise_adj_all %>%
              pivot_longer(-c(feature,visit),
                           names_to = "comparison",
                           values_to = "adj_pval") %>%
              mutate(log10_adjpval = -log10(adj_pval)) %>%
              inner_join(endpoint_coefficient_plot, 
                         by=c("feature", "visit", "comparison")) %>%
              mutate(comp_visit = paste0(comparison, "_", visit)) %>%
              mutate(signed_adjp = log10_adjpval * sign)

df_plot$feature <- factor(df_plot$feature, levels = colnames(top_feat_data_df))
df_plot$comp_visit <- factor(df_plot$comp_visit, levels = unique(df_plot$comp_visit))
df_plot <- df_plot %>% 
            mutate(annot =
                case_when(
                 abs(log10_adjpval) >= -log10(0.001) ~ "***",
                 abs(log10_adjpval) >= -log10(0.01) ~ "**",
                 abs(log10_adjpval) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 ))

thr <- 0.05
plot_title <- "Adj p-value feature association"

convert_df <- df_plot %>% 
              dplyr::select(feature, comp_visit, signed_adjp) %>%
              pivot_wider(id_cols = c(comp_visit), 
                          names_from = feature, 
                          values_from = signed_adjp) %>%
              column_to_rownames("comp_visit")

convert_annot <- df_plot %>%
              dplyr::select(feature, comp_visit, annot) %>%
              pivot_wider(id_cols = c(comp_visit),
                          names_from = feature,
                          values_from = annot) %>%
              column_to_rownames("comp_visit")

mat <- as.matrix(convert_df)
mat_annot <- as.matrix(convert_annot)

palette_length <- 50
abs_max <- max(abs(min(mat, na.rm = T)), abs(max(mat, na.rm = T)), 1.3)
breaks_list <- c(seq(-abs_max, 0, length.out=ceiling(palette_length/2) + 1), 
              seq(abs_max/palette_length, abs_max, length.out=floor(palette_length/2)))

pdf(file = paste0(results_dir, "Adj_pval_all_visits_pro_groups_plot.pdf"), width = 12, height = 12)
p <- pheatmap(t(mat),
         cluster_cols = F, cluster_rows = F,
         cellwidth = 15, cellheight = 15,
         color = colorRampPalette(rev(brewer.pal(9,"RdBu")))(palette_length),
         breaks = breaks_list,
         display_numbers = t(mat_annot),
         number_color = "black",
         fontsize_number = 10,
         )
p
dev.off()

p

#save_pheatmap_pdf(p, paste0(results_dir,"Adj_pval_all_visits_pro_groups_plot.pdf"), width = 12, height = 6)

```

### Violin plots all visits

```{r }

features_filtered = top_feat_data_df[, colnames(top_feat_data_df), drop=F]
clinical_subsets = clinical_subsets_all[rownames(top_feat_data_df),]
  

data_use = features_filtered 

data_use$endpoints = factor(clinical_subsets$pro_group_labels, levels = names(pro_colors))

# Add the enrollment_site as $sites
data_use$sites = clinical_subsets$enrollment_site
data_use$age =  clinical_subsets$discretized_admit_age_quantile
data_use$sex =  clinical_subsets$sex
data_use$visit = clinical_subsets$event_type

# Add the $control, needs to be different for each sample
# required for clmm (always requires some random effect)
data_use$control =  factor(1:nrow(data_use))

# Set module names:
feature_names = colnames(features_filtered)

# Only use samples that have valid endpoints (not NA/missing)
data_use = data_use[!is.na(data_use$endpoints),]
```

Violin plots for all features, values from all visits

```{r fig.height=3, fig.width=6}

plot_df <- data_use %>% dplyr::select(colnames(top_feat_data_df), 
                                        endpoints) %>%
            pivot_longer(!endpoints, names_to = "Feature", values_to = "value")
plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))

p <- ggplot(plot_df, aes(x=endpoints, y=value, color=endpoints)) +
  #geom_violin(draw_quantiles = c(0.25,0.5,0.75)) +
  geom_jitter(width=0.3, alpha=0.2) +
  geom_boxplot(fill=NA, outlier.shape=NA) +
  theme_bw() + theme(legend.position = "none") +
  scale_color_manual(values=pro_colors) +
  scale_fill_manual(values=pro_colors) +
  facet_wrap(~Feature, drop = T, scales = "free_y", ncol = 4)
p

```

Violin plots for all features, values per visit

```{r fig.height=3, fig.width=8}

plot_df <- data_use %>% dplyr::select(colnames(top_feat_data_df),
                                        endpoints, visit) %>%
            pivot_longer(!c(endpoints,visit), names_to = "Feature", values_to = "value")
plot_df$Feature <- factor(plot_df$Feature, levels=colnames(top_feat_data_df))
plot_df$visit <- factor(plot_df$visit, levels=ordered_visits)

for (feat in colnames(top_feat_data_df)) {
  plot_df_fact <- plot_df %>% filter(Feature == feat)
  
  p <- ggplot(plot_df_fact, aes(x=visit, y=value, fill=endpoints)) +
  ggbeeswarm::geom_quasirandom(aes(fill=endpoints), shape=21, width=.1, dodge.width = 0.75) +
  geom_boxplot(alpha=0.6, color="black", outlier.shape = NA) +
  scale_y_reverse() +
  theme_bw() + 
  scale_color_manual(values=pro_colors) +
  scale_fill_manual(values=pro_colors) +
  #facet_wrap(~visit, drop = T, ncol = 4) +
  labs( x="", y=feat, fill="")
  print(p)
}

```

### Violin plots per visit p-val

```{r fig.height=3, fig.width=8}

feature_violin_plots_list <- list()

for (feat in colnames(top_feat_data_df)) {
  plot_df_feat <- plot_df %>% filter(Feature == feat)
  
  format_corrected_pval <- res_table_pairwise_adj_all %>% 
    dplyr::filter(feature == feat) %>%
    pivot_longer(cols=-c(feature,visit), names_to = "comparison", values_to = "adj_pval_sex_age_corrected") %>%
    dplyr::mutate(group1 = str_extract(comparison, "[^|]+"),
                  group2 = str_extract(comparison, "[^|]+$")) %>%
    dplyr::select(-c(comparison,feature))
  
  stat.test <- plot_df_feat %>%
    group_by(visit) %>%
    wilcox_test(value ~ endpoints) %>%
    adjust_pvalue(method = "hochberg") %>%
    left_join(format_corrected_pval, by=join_by(visit,group1,group2)) %>%
    add_significance("adj_pval_sex_age_corrected") %>%
    add_xy_position(x = "visit", dodge = 0.75) %>%
    dplyr::filter(group1 == "MIN") %>%
    dplyr::filter(adj_pval_sex_age_corrected < 0.05) %>%
    dplyr::mutate(y.position = y.position -40)

  
  p <- ggplot(plot_df_feat, aes(x=visit, y=value)) +
    ggbeeswarm::geom_quasirandom(aes(fill=endpoints),shape=21,width=.1, dodge.width = 0.75) +
    geom_boxplot(aes(fill=endpoints),alpha=0.6, color="black", outlier.shape=NA) +
    theme_bw() + 
    scale_color_manual(values=pro_colors) +
    scale_fill_manual(values=pro_colors) +
    labs( x="", y=str_replace_all(feat,"_"," "), fill="") +
    stat_pvalue_manual(stat.test, label = "adj_pval_sex_age_corrected.signif", tip.length = 0) +
    scale_y_reverse(expand = expansion(mult = c(0.1, 0.1)))
  print(p)
  feature_violin_plots_list <- append(feature_violin_plots_list, list(p))
}

feature_visits_min_pro_pvals <- feature_violin_plots_list
```

### Fig viral load

```{r fig.width=8, fig.height=8}

figS10AB <- ggarrange(feature_visits_min_other_pvals[[1]], 
          feature_visits_min_other_pvals[[2]],
          feature_visits_min_pro_pvals[[1]],
          feature_visits_min_pro_pvals[[2]],
          ncol = 1, nrow = 4, labels = c("A", "", "B", ""))
figS10AB


ggsave(file.path(figures_dir, "viral_load_association.pdf"), figS10AB, height=8, width=8)
ggsave(file.path(figures_dir, "viral_load_association.png"), figS10AB, height=8, width=8)
saveRDS(feature_visits_min_other_pvals, file.path(figures_dir, "feature_visits_min_other_pvals.rds"))
saveRDS(feature_visits_min_pro_pvals, file.path(figures_dir, "feature_visits_min_pro_pvals.rds"))
```


## Longitudinal analysis over acute data


```{r data preparation, echo=TRUE}
clinical_subsets <- correlation_metadata
rownames(clinical_subsets) <- clinical_subsets$event_id

row_ids <-rownames(top_feat_data_df)
clinical_subsets <- clinical_subsets[row_ids,]

event_types <- c("Visit 1", "Visit 2","Visit 3","Visit 4", "Visit 5", "Visit 6")


modules_filtered = top_feat_data_df %>% dplyr::select(all_of(colnames(top_feat_data_df)))

####%%%%%%%%  Input %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

inputDF <- modules_filtered %>%
  rownames_to_column(var = "event_id") %>%
  merge(y = clinical_subsets,
        by = "event_id") %>%
  mutate(outcome = pro_2_groups) %>% #TODO: update outcome variable here
  pivot_longer(cols = -c("event_id", "event_date", "participant_id",
                        "enrollment_site", "pro_group_labels", "pro_2_groups", "event_type", "outcome",
                        "sex", "admit_age", "discretized_admit_age_quantile", "resp_status_v1",
                        "event_after_first_vaccine_dose", "event_date_3weeks_after_dose","race")) %>%
  dplyr::filter(!is.na(value) & !is.na(pro_2_groups)) %>%
  dplyr::filter(event_type %in% acute_visits) %>%
  dplyr::filter(event_date <= 50)
inputDF$names <- factor(inputDF$name, levels=colnames(top_feat_data_df))


## Ensure proper variables are features
inputDF$pro_group_labels <- as.factor(inputDF$pro_group_labels)
inputDF$pro_2_groups <- as.factor(inputDF$pro_2_groups)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$race <- as.factor(inputDF$race)
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)
```


### PRO 2 groups

```{r, linear_model_loop, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T, 
                              model_type = "smoothSpline", 
                              spline_type = "ts", 
                              endpoint = "pro_2_groups") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])
write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_pro_2_groups_acute.csv"))
```

```{r, longitudinal_model_loop_pheatmap, fig.width = 6, fig.height=6}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_min_vs_other_acute.pdf"), width = 6, height = 20)

mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F, 
         color = c("#000004", "#1B0A40", "#4A0C69", "#781B6C", "#A42C5F", 
                   "#CD4345", "#EC6824", "#FA990B", "#F7CF3D", "#FCFFA4", "#FCFFA4"),
         breaks = breaklist)
p1
dev.off()
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_min_vs_other.pdf"), width = 3, height = 6)
p1
```

### Variability of features over time

```{r, fig.width=10, fig.height=2}
plots <- list()
plots_model <- list()
for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline",
                      colors = pro_min_colors,
                     spline_type = "ts",
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     CI = T,
                     endpoint = "pro_2_groups") #TODO: update outcome variable here
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_2_groups", #TODO: update outcome variable here
                   model_type = "smoothSpline",
                   colors = pro_min_colors,
                   spline_type = "ts",
                   knot_lines = F,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_acute.pdf"), width = 18, height = 5)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_acute.png"), width = 18, height = 5, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model_acute.pdf"), width = 18, height = 5)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model_acute.png"), width = 18, height = 5, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)

```

### All PRO groups

```{r, linear_model_loop_PRO_groups, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T,
                              model_type = "smoothSpline",
                              spline_type = "ts",
                              endpoint = "pro_group_labels") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, 
                               adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])

write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_all_pro_groups_acute.csv"))
```

```{r linear_model_loop_pheatmap_allpro, fig.width = 6, fig.height=6}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_pro_groups_acute.pdf"), width = 6, height = 20)
mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F, 
         color = c("#000004", "#1B0A40", "#4A0C69", "#781B6C", "#A42C5F", 
                   "#CD4345", "#EC6824", "#FA990B", "#F7CF3D", "#FCFFA4", "#FCFFA4"),
         breaks = breaklist)
p1
dev.off()

p1
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_pro_groups.pdf"), width = 3, height = 6)

```

### Plotting features over the visits

```{r, fig.width=10, fig.height=2}
plots <- list()
plots_model <- list()

for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  plotDF$pro_group_labels <- factor(plotDF$pro_group_labels, levels = names(pro_colors))
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline", 
                     spline_type = "ts",
                     colors = pro_colors,
                     CI = T,
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     facet = T,
                     endpoint = "pro_group_labels")
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_group_labels",
                   model_type = "smoothSpline",
                   spline_type = "ts",
                   knot_lines = F,
                   colors = pro_colors,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_acute.pdf"), width = 15, height = 5)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_acute.png"), width = 20, height = 6, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model_acute.pdf"), width = 15, height = 5)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model_acute.png"), width = 20, height = 6, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)

```

## Longitudinal analysis over convalescent data

With spear features

```{r data preparation, echo=TRUE}
clinical_subsets <- correlation_metadata
rownames(clinical_subsets) <- clinical_subsets$event_id

row_ids <-rownames(top_feat_data_df)
clinical_subsets <- clinical_subsets[row_ids,]

event_types <- c("Visit 7", "Visit 8","Visit 9","Visit 10")


modules_filtered = top_feat_data_df %>% dplyr::select(all_of(colnames(top_feat_data_df)))

####%%%%%%%%  Input %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

inputDF <- modules_filtered %>%
  rownames_to_column(var = "event_id") %>%
  merge(y = clinical_subsets,
        by = "event_id") %>%
  mutate(outcome = pro_2_groups) %>% #TODO: update outcome variable here
  pivot_longer(cols = -c("event_id", "event_date", "participant_id",
                        "enrollment_site", "pro_group_labels", "pro_2_groups", "event_type", "outcome",
                        "sex", "admit_age", "discretized_admit_age_quantile", "resp_status_v1",
                        "event_after_first_vaccine_dose", "event_date_3weeks_after_dose","race")) %>%
  dplyr::filter(!is.na(value) & !is.na(pro_2_groups)) %>%
  dplyr::filter(event_type %in% event_types)
inputDF$names <- factor(inputDF$name, levels=colnames(top_feat_data_df))


## Ensure proper variables are features
inputDF$pro_group_labels <- as.factor(inputDF$pro_group_labels)
inputDF$pro_2_groups <- as.factor(inputDF$pro_2_groups)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$race <- as.factor(inputDF$race)
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)
```


### PRO 2 groups

```{r, linear_model_loop, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T, 
                              model_type = "smoothSpline", 
                              spline_type = "ts", 
                              endpoint = "pro_2_groups") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])
write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_pro_2_groups_convalescent.csv"))
```

```{r, longitudinal_model_loop_pheatmap, fig.width = 6, fig.height=6}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_min_vs_other_convalescent.pdf"), width = 6, height = 20)

mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F, 
         color = c("#000004", "#1B0A40", "#4A0C69", "#781B6C", "#A42C5F", 
                   "#CD4345", "#EC6824", "#FA990B", "#F7CF3D", "#FCFFA4", "#FCFFA4"),
         breaks = breaklist)
p1
dev.off()
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_min_vs_other.pdf"), width = 3, height = 6)
p1
```

### Variability of features over time

```{r, fig.width=10, fig.height=2}
plots <- list()
plots_model <- list()
for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline",
                      colors = pro_min_colors,
                     spline_type = "ts",
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     CI = T,
                     endpoint = "pro_2_groups") #TODO: update outcome variable here
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_2_groups", #TODO: update outcome variable here
                   model_type = "smoothSpline",
                   colors = pro_min_colors,
                   spline_type = "ts",
                   knot_lines = F,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_convalescent.pdf"), width = 18, height = 5)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_convalescent.png"), width = 18, height = 5, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model_convalescent.pdf"), width = 18, height = 5)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model_convalescent.png"), width = 18, height = 5, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)

```

### All PRO groups

```{r, linear_model_loop_PRO_groups, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T,
                              model_type = "smoothSpline",
                              spline_type = "ts",
                              endpoint = "pro_group_labels") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, 
                               adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])

write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_all_pro_groups_convalescent.csv"))
```

```{r linear_model_loop_pheatmap_allpro, fig.width = 6, fig.height=6}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_pro_groups_convalescent.pdf"), width = 6, height = 20)
mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F, 
         color = c("#000004", "#1B0A40", "#4A0C69", "#781B6C", "#A42C5F", 
                   "#CD4345", "#EC6824", "#FA990B", "#F7CF3D", "#FCFFA4", "#FCFFA4"),
         breaks = breaklist)
p1
dev.off()

p1
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_pro_groups.pdf"), width = 3, height = 6)

```

### Plotting features over the visits

```{r, fig.width=10, fig.height=2}
plots <- list()
plots_model <- list()

for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  plotDF$pro_group_labels <- factor(plotDF$pro_group_labels, levels = names(pro_colors))
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline", 
                     spline_type = "ts",
                     colors = pro_colors,
                     CI = T,
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     facet = T,
                     endpoint = "pro_group_labels")
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_group_labels",
                   model_type = "smoothSpline",
                   spline_type = "ts",
                   knot_lines = F,
                   colors = pro_colors,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_convalescent.pdf"), width = 15, height = 5)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_convalescent.png"), width = 20, height = 6, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model_convalescent.pdf"), width = 15, height = 5)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model_convalescent.png"), width = 20, height = 6, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)

```

## Longitudinal analysis over all data

With spear features

```{r data preparation, echo=TRUE}
clinical_subsets <- correlation_metadata
rownames(clinical_subsets) <- clinical_subsets$event_id

row_ids <-rownames(top_feat_data_df)
clinical_subsets <- clinical_subsets[row_ids,]

event_types <- c("Visit 1", "Visit 2","Visit 3","Visit 4", "Visit 5", "Visit 6", "Visit 7", "Visit 8","Visit 9","Visit 10")


modules_filtered = top_feat_data_df %>% dplyr::select(all_of(colnames(top_feat_data_df)))

####%%%%%%%%  Input %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

inputDF <- modules_filtered %>%
  rownames_to_column(var = "event_id") %>%
  merge(y = clinical_subsets,
        by = "event_id") %>%
  mutate(outcome = pro_2_groups) %>% #TODO: update outcome variable here
  pivot_longer(cols = -c("event_id", "event_date", "participant_id",
                        "enrollment_site", "pro_group_labels", "pro_2_groups", "event_type", "outcome",
                        "sex", "admit_age", "discretized_admit_age_quantile", "resp_status_v1",
                        "event_after_first_vaccine_dose", "event_date_3weeks_after_dose","race")) %>%
  dplyr::filter(!is.na(value) & !is.na(pro_2_groups)) %>%
  dplyr::filter(event_type %in% event_types)
inputDF$names <- factor(inputDF$name, levels=colnames(top_feat_data_df))


## Ensure proper variables are features
inputDF$pro_group_labels <- as.factor(inputDF$pro_group_labels)
inputDF$pro_2_groups <- as.factor(inputDF$pro_2_groups)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$race <- as.factor(inputDF$race)
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)
```


### PRO 2 groups

```{r, linear_model_loop, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T, 
                              model_type = "smoothSpline", 
                              spline_type = "ts", 
                              endpoint = "pro_2_groups") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])
write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_pro_2_groups_all.csv"))
```

```{r, longitudinal_model_loop_pheatmap, fig.width = 6, fig.height=6}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_min_vs_other_all.pdf"), width = 6, height = 20)

mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F, 
         color = c("#000004", "#1B0A40", "#4A0C69", "#781B6C", "#A42C5F", 
                   "#CD4345", "#EC6824", "#FA990B", "#F7CF3D", "#FCFFA4", "#FCFFA4"),
         breaks = breaklist)
p1
dev.off()
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_min_vs_other.pdf"), width = 3, height = 6)
p1
```

### Variability of features over time

```{r, fig.width=10, fig.height=2}
plots <- list()
plots_model <- list()
for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline",
                      colors = pro_min_colors,
                     spline_type = "ts",
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     CI = T,
                     endpoint = "pro_2_groups") #TODO: update outcome variable here
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_2_groups", #TODO: update outcome variable here
                   model_type = "smoothSpline",
                   colors = pro_min_colors,
                   spline_type = "ts",
                   knot_lines = F,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_all.pdf"), width = 18, height = 5)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_all.png"), width = 18, height = 5, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model_all.pdf"), width = 18, height = 5)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model_all.png"), width = 18, height = 5, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)

```

### All PRO groups

```{r, linear_model_loop_PRO_groups, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T,
                              model_type = "smoothSpline",
                              spline_type = "ts",
                              endpoint = "pro_group_labels") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, 
                               adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])

write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_all_pro_groups_all.csv"))
```

```{r linear_model_loop_pheatmap_allpro, fig.width = 6, fig.height=6}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_pro_groups_all.pdf"), width = 6, height = 20)
mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F, 
         color = c("#000004", "#1B0A40", "#4A0C69", "#781B6C", "#A42C5F", 
                   "#CD4345", "#EC6824", "#FA990B", "#F7CF3D", "#FCFFA4", "#FCFFA4"),
         breaks = breaklist)
p1
dev.off()

p1
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_pro_groups.pdf"), width = 3, height = 6)

```

### Plotting features over the visits

```{r, fig.width=10, fig.height=2}
plots <- list()
plots_model <- list()

for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  plotDF$pro_group_labels <- factor(plotDF$pro_group_labels, levels = names(pro_colors))
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline", 
                     spline_type = "ts",
                     colors = pro_colors,
                     CI = T,
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     facet = T,
                     endpoint = "pro_group_labels")
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_group_labels",
                   model_type = "smoothSpline",
                   spline_type = "ts",
                   knot_lines = F,
                   colors = pro_colors,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_all.pdf"), width = 15, height = 5)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_all.png"), width = 20, height = 6, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model_all.pdf"), width = 15, height = 5)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model_all.png"), width = 20, height = 6, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)

```

## Longitudinal analysis symptom onset date over acute data

```{r prepare the metadata for mixed linear model analysis, echo= FALSE, warning= FALSE}
correlation_metadata <- clinical_data %>%
  dplyr::select(event_id, resp_status_v1, sex, admit_age, event_after_first_vaccine_dose, 
                race, enrollment_site, event_date, symptom_date, pcr_date,
                event_date_3weeks_after_dose, pro_group_labels, event_type, 
                discretized_admit_age_quantile, participant_id)  %>%
  dplyr::mutate(date_after_symptom = event_date + symptom_date,
                date_after_pcr = event_date + pcr_date) %>%
  dplyr::mutate(pro_2_groups = 
      case_match( pro_group_labels,
      "MIN" ~ "MIN",
      c("PHY", "COG", "MLT")  ~ "OTHER"
      )
  ) %>%
  dplyr::distinct() %>%
  filter(event_id %in% rownames(top_feat_data_df))

clinical_subsets_all <- as.data.frame(correlation_metadata)
rownames(clinical_subsets_all) <- clinical_subsets_all$event_id


endpoint_coefficients_all <- data.frame(matrix(nrow=ncol(top_feat_data_df), ncol = 0))
rownames(endpoint_coefficients_all) <- colnames(top_feat_data_df)
```


Check how symptom onset correlates with event date

```{r}
symptom_date_avail <- clinical_subsets_all %>% 
  dplyr::select(participant_id, symptom_date) %>%
  dplyr::distinct() %>%
  dplyr::group_by(participant_id)

nrow(symptom_date_avail[!is.na(symptom_date_avail$symptom_date),])


ggplot(clinical_subsets_all, aes(x=event_date, y = date_after_symptom)) +
  geom_point()

# Participant 008-0103 excluded as very far away symptom onset date (-375)
```

Check days after symptom onset for Visit 1:
```{r}
correlation_metadata <- correlation_metadata %>%
  dplyr::filter(participant_id != "008-0103")

correlation_metadata$event_type <- factor(correlation_metadata$event_type, levels = c("Visit 1", "Visit 2","Visit 3","Visit 4", "Visit 5", "Visit 6", "Visit 7", "Visit 8","Visit 9","Visit 10"))
ggplot(correlation_metadata, aes(x=event_type, y=date_after_symptom)) +
  geom_boxplot() +
  geom_point()

```

```{r data preparation, echo=TRUE}
clinical_subsets <- correlation_metadata
rownames(clinical_subsets) <- clinical_subsets$event_id

row_ids <-rownames(top_feat_data_df)
clinical_subsets <- clinical_subsets[row_ids,]

event_types <- c("Visit 1", "Visit 2","Visit 3","Visit 4", "Visit 5", "Visit 6")


modules_filtered = top_feat_data_df %>% dplyr::select(all_of(colnames(top_feat_data_df)))

####%%%%%%%%  Input %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

inputDF <- modules_filtered %>%
  rownames_to_column(var = "event_id") %>%
  merge(y = clinical_subsets,
        by = "event_id") %>%
  mutate(outcome = pro_2_groups) %>% #TODO: update outcome variable here
  pivot_longer(cols = -c("event_id", "symptom_date", "pcr_date", 
                         "date_after_symptom", "date_after_pcr",
                         "participant_id",
                        "enrollment_site", "pro_group_labels", "pro_2_groups", "event_type", "outcome",
                        "sex", "admit_age", "discretized_admit_age_quantile", "resp_status_v1",
                        "event_after_first_vaccine_dose", "event_date_3weeks_after_dose","race")) %>%
  dplyr::filter(!is.na(value) & !is.na(pro_2_groups) & !is.na(date_after_symptom)) %>%
  dplyr::filter(event_type %in% event_types) %>%
  dplyr::filter(date_after_symptom <= 100)
inputDF$names <- factor(inputDF$name, levels=colnames(top_feat_data_df))


## Ensure proper variables are features
inputDF$pro_group_labels <- as.factor(inputDF$pro_group_labels)
inputDF$pro_2_groups <- as.factor(inputDF$pro_2_groups)
inputDF$sex <- factor(inputDF$sex, levels = c("Female", "Male"))
inputDF$race <- as.factor(inputDF$race)
inputDF$discretized_admit_age_quantile <- as.factor(inputDF$discretized_admit_age_quantile)
inputDF$participant_id <- as.factor(inputDF$participant_id)
inputDF$enrollment_site <- as.factor(inputDF$enrollment_site)
```


### PRO 2 groups

```{r, linear_model_loop, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T, 
                              model_type = "smoothSpline", 
                              spline_type = "ts",
                              time_var = "date_after_symptom",
                              endpoint = "pro_2_groups") 

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])
write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_pro_2_groups_acute_dateaftersymptom.csv"))
```

```{r, longitudinal_model_loop_pheatmap, fig.width = 6, fig.height=6}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_min_vs_other_acute_dateaftersymptom.pdf"), width = 6, height = 20)

mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F, 
         color = c("#000004", "#1B0A40", "#4A0C69", "#781B6C", "#A42C5F", 
                   "#CD4345", "#EC6824", "#FA990B", "#F7CF3D", "#FCFFA4", "#FCFFA4"),
         breaks = breaklist)
p1
dev.off()
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_min_vs_other.pdf"), width = 3, height = 6)
p1
```

### Variability of features over time

```{r, fig.width=10, fig.height=2}
plots <- list()
plots_model <- list()
for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline",
                     time_var = "date_after_symptom",
                      colors = pro_min_colors,
                     spline_type = "ts",
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     CI = T,
                     endpoint = "pro_2_groups") +
        labs(x="Days from symptom onset")
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_2_groups", #TODO: update outcome variable here
                   model_type = "smoothSpline",
                   time_var = "date_after_symptom",
                   colors = pro_min_colors,
                   spline_type = "ts",
                   knot_lines = F,
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right") +
    labs(x="Days from symptom onset")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_acute_dateaftersymptom.pdf"), width = 18, height = 5)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_acute_dateaftersymptom.png"), width = 18, height = 5, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=2)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model_acute_dateaftersymptom.pdf"), width = 18, height = 5)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

png(paste0(results_dir,"Features_longitudinal_MIN_vs_OTHER_model_acute_dateaftersymptom.png"), width = 18, height = 5, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)

```

### All PRO groups

```{r, linear_model_loop_PRO_groups, fig.width = 3, fig.height=5}

gamm_model_loop <- model_loop(inputDF, age_sex = T,
                              model_type = "smoothSpline",
                              spline_type = "ts",
                              time_var = "date_after_symptom",
                              endpoint = "pro_group_labels") #TODO: update outcome variable here

gamm_model_loop_plot <- gamm_model_loop %>% 
                        dplyr::select(adjp.slope, 
                               adjp.intercept)

print_kable(gamm_model_loop_plot[colnames(top_feat_data_df),])

write.csv(gamm_model_loop_plot[colnames(top_feat_data_df),], paste0(results_dir, "global_p_value_longitudinal_all_pro_groups_acute_daysfromsymptomonset.csv"))
```

```{r linear_model_loop_pheatmap_allpro, fig.width = 6, fig.height=6}
pdf(file = paste0(results_dir, "global_p_value_longitudinal_pro_groups_acute_dateaftersymptom.pdf"), width = 6, height = 20)
mat_log10pval <- -log10(gamm_model_loop_plot[colnames(top_feat_data_df),])
mat_annot <- mat_log10pval %>% 
  mutate(across(everything(), ~ case_when(
                 abs(.x) >= -log10(0.001) ~ "***",
                 abs(.x) >= -log10(0.01) ~ "**",
                 abs(.x) >= -log10(0.05) ~ "*",
                 .default = as.character("")
                 )))

p1 <- pheatmap(as.matrix(mat_log10pval),
         cellwidth = 15,
         cellheight = 15,
         display_numbers = as.matrix(mat_annot),
         fontsize_number = 10,
         number_color = "black",
         cluster_rows = F, 
         cluster_cols = F, 
         color = c("#000004", "#1B0A40", "#4A0C69", "#781B6C", "#A42C5F", 
                   "#CD4345", "#EC6824", "#FA990B", "#F7CF3D", "#FCFFA4", "#FCFFA4"),
         breaks = breaklist)
p1
dev.off()

p1
#save_pheatmap_pdf(p1, paste0(results_dir,"global_p_value_longitudinal_pro_groups.pdf"), width = 3, height = 6)

```

### Plotting features over the visits

```{r, fig.width=10, fig.height=2}
plots <- list()
plots_model <- list()

for (feature_name in colnames(top_feat_data_df)) {
  plotDF <- inputDF %>% filter(name == feature_name)
  plotDF$pro_group_labels <- factor(plotDF$pro_group_labels, levels = names(pro_colors))
  p1 <- plot_model(plotDF, 
                     gamm_model_loop,
                     model_type = "smoothSpline", 
                     spline_type = "ts",
                     colors = pro_colors,
                     time_var = "date_after_symptom",
                     CI = T,
                     knot_lines = F,
                     group_trendline_color = "endpoint",
                     facet = T,
                     endpoint = "pro_group_labels") +
    labs(x="Days after symptom onset")
  plots <- c(plots, list(p1))
  p2 <- plot_model(plotDF, 
                   gamm_model_loop,
                   endpoint = "pro_group_labels",
                   model_type = "smoothSpline",
                   spline_type = "ts",
                   knot_lines = F,
                   colors = pro_colors,
                   time_var = "date_after_symptom",
                   individual_trendlines = F,
                   individual_points = F,
                   individual_paths = F,
                   CI = T,
                   group_trendline_color = "endpoint",
                   facet = F) + theme(legend.position = "right") +
        labs(x="Days after symptom onset")
  plots_model <- c(plots_model, list(p2))
}

pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_acute_dateaftersymptom.pdf"), width = 15, height = 5)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_acute_dateaftersymptom.png"), width = 20, height = 6, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots, ncol=1)
dev.off()

pdf(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model_acute_dateaftersymptom.pdf"), width = 15, height = 5)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

png(paste0(results_dir,"Features_longitudinal_PRO_GROUPS_model_acute_dateaftersymptom.png"), width = 20, height = 6, units="cm", res = 300)
gridExtra::grid.arrange(grobs = plots_model, ncol=1)
dev.off()

gridExtra::grid.arrange(grobs = plots, ncol=2)
gridExtra::grid.arrange(grobs = plots_model, ncol=2)

```


```{r}
sessionInfo()
```
